### 2016-08-23 paper/00specification.tex
initial commit with up-to-date bounds-RMD files. However, large simulation results were purged from the repository

\documentclass[11pt]{article}

\usepackage[backgroundcolor=white,textsize=small,linecolor=red]{todonotes}

\usepackage[margin=30mm]{geometry}
\usepackage{natbib}

%--------------------
% Math
\usepackage[fleqn]{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{multicol}
\usepackage{graphicx}

\newcommand{\mvalueof}[1]{\llbracket#1\rrbracket}
\newcommand{\citeposs}[2][]{\citeauthor{#2}'s (\citeyear[#1]{#2})}
\newcommand{\tuple}[1]{\ensuremath{\left\langle #1 \right\rangle}}

\newcommand{\hloranj}[1]{\textcolor[rgb]{.8,.33,.0}{#1}}% prints in orange
\newcommand{\argmax}[1]{\underset{#1}{\operatorname{arg}\,\operatorname{max}}\;}
\newcommand{\argmin}[1]{\underset{#1}{\operatorname{arg}\,\operatorname{min}}\;}
\newcommand{\sbna}{\exists\lnot\forall}

%--------------------
%\usepackage{gb4e}
%
%\usepackage{setspace}
%\onehalfspacing
%
%\usepackage{soul} %underlining
%\setuldepth{a}

%\usepackage{inconsolata}

%-------------------

\title{Bounds, Semantics/Pragmatics-Divide, and Evolution}

\author{%\bf NAME1 and NAME2\\
    \today
}

\date{}

\begin{document}
\maketitle

\section{Short exposition}
The plan is to expand our work on the semantics/pragmatics divide by means of a more involved \& principled study on the lack of upper-bounds in the literal meaning of scalar expressions. The main idea of this extension, other than providing a more detailed treatment as well as a comparison with iterated learning, is to do away with the assumption of communal learning. Potential extensions include more scalar pairs, the addition of an inductive bias towards uniform lexica across scalar pairs, and a variation of state frequencies.

\section{Differences to CogSci setup.}
\subparagraph{(I) No communal learning.} Only parental learning is considered

\subparagraph{(II) Sequences and atomic observations.} Before, the set of all observations was $O =$\linebreak  $\{\tuple{\tuple{s_1,m_i},\tuple{s_2,m_j}} | m_i, m_j \in M\}$. A member of $O$ encodes that a teacher produced $m_i$ in state $s_1$ and $m_j$ in $s_2$, i.e., it encodes one witnessed message for each state. A datum $d$ was a sequence of length $k$ of members of $O$. Learners witnessed such data sequences. Now, more in line with \citet{griffiths+kalish:2007}, $O = \{\tuple{s_i,m_j} | s_i \in S, m_j \in M\}$ and $d$ is a sequence of length $k$ of members of $O$. The main difference is that now some $d$ do not provide any production information for some states.

\subparagraph{(III) Observations as production.} Instead of taking the space of all possible sequences of length $k$ into consideration, we take sample from $O$ $k$-times according to the production probabilities of each type; $P(o = \tuple{s,m} | t_i) = P(s) P(m|s,t_i)$. $n$ such $k$-length sequences are sampled for each type. As a consequence, the data used for computing $Q_i$ is not the same as that used for $j$ $(i \neq j)$.

\subparagraph{(IV) Parametrized learning} $Q_{ij} \propto \sum_d P(d|t_i) F(t_j,d)$, where $F(t_j,d) \propto P(t_j|d)^l$ and $l =1$ corresponds to probability matching and, as $l$ increases towards infinity, to MAP.

\section{Further modifications we may want to keep in mind (but currently don't make use of)}
\subparagraph{(I) Cost for pragmatic reasoning.} At least in the CogSci setup the effect of adding cost to pragmatic reasoning is unsurprising: High cost for pragmatic signaling lowers the prevalence of pragmatic types. Lexica that semantically encode an upper-bound benefit the most from this. However, the cost needed to be substantial to make the pragmatic English-like lexicon stop being the incumbent type (particularly when learning is communal).

\subparagraph{(II) Negative learning bias.} Instead of penalizing complex semantics (semantic upper-bounds) one may consider penalizing simple semantics (no upper-bounds). This is useful as a sanity check but also yields unsurprising results in the CogSci setup: The more learners are biased against simple semantics, the more prevalent are lexica that semantically encode upper-bounds.

\subparagraph{(III) Inductive bias.} A second learning bias that codifies the idea that lexica should be uniform, i.e. be biased towards either lexicalizing an upper-bound for all weaker alternatives in a scalar pair or for none.

\subparagraph{(IV)  Uncertainty.} The other advantage of non-upper bounded semantics lies in being non-committal to the negation of stronger alternatives when the speaker is uncertain. Adding this to the model requires the most changes to our present setup and some additional assumptions about the cues available to players to discern the speaker's knowledge about the state she is in.

\subparagraph{(V) More scalar pairs.} Taking into consideration more than one scalar pair. Preliminary results suggest that this does not influence the results in any meaningful way without further additions, e.g. by (III).

\subparagraph{(VI) More lexica.} Not necessary. Preliminary results suggest that considering more lexica has no noteworthy effect on the dynamics (tested with all possible 2x2 lexica).

\subparagraph{(VII) State frequencies.} Variations on state frequencies. This may have an interesting interaction with (III).

\subparagraph{(VIII) Reintroduction of communal learning.} One possibility: The probably $N_{ij}$ with which a child of $t_i$ adopts $t_j$ could be the weighted sum of $Q_{ij}$ (as before) and a vector we get from learning from all of the population: $L_j = \sum_d P(d | \vec{p})  P(t_j | d)$, where $P(d | \vec{p}) = \sum_{i} P(d | t_i)  \vec{p}_i$ is the probability of observing $d$ when learning from a random member of the present population distribution.

\section{General setup}
\subsection{Part 1: Signaling.}

We consider a population of players with two signaling behaviors, literal and Gricean (level $0$ and $1$ below), each equipped with one of $6$ lexicons. This yields a total of $12$ distinct player types $t \in T$. $|M| = |S| = 2$, i.e., a lexicon is a $(2,2)$-matrix. These are listed in Table \ref{tab:lexica}.

\begin{table}[h]
\centering 
\begin{tabular}{l c l}
$L_1$ = $\begin{pmatrix} 0 & 0 \\ 1 & 1 \end{pmatrix}$ & 
$L_2$ = $\begin{pmatrix} 1 & 1 \\ 0 & 0 \end{pmatrix}$ & 
$L_3$ = $\begin{pmatrix} 1 & 1 \\ 1 & 1 \end{pmatrix}$\\[0.5cm]

$L_4$ = $\begin{pmatrix} 0 & 1 \\ 1 & 0 \end{pmatrix}$ &
$L_5$ = $\begin{pmatrix} 0 & 1 \\ 1 & 1 \end{pmatrix}$ &
$L_6$ = $\begin{pmatrix} 1 & 1 \\ 1 & 0 \end{pmatrix}$
\end{tabular}
\caption{{\footnotesize Set of possible $(2,2)$-matrices, i.e., a lexicon.}}
\label{tab:lexica}
\end{table}

As in the CogSci paper, $L_4$ (semantic upper-bound for $m_2$) and $L_5$ (no semantic upper-bound for $m_2$) are the target lexica. Gricean $L_5$ users can convey/infer the bound pragmatically, while literal/Gricean $L_4$ users do so semantically.

\paragraph{Signaling behavior.} With $\lambda \geq 1$ (rationality parameter), $\alpha \in [0,1]$ (pragmatic violations) and $pr \in \Delta(S)$ a common prior over $S$ (uniform so far):

\vspace{-0.15cm}
\begin{flalign}
&R_{0}(s|m;L) \propto pr(s) L_{sm}\label{litl}\\
&S_{0}(m|s;L) \propto \exp(\lambda \; L_{sm}) \label{lits}\\
&R_{1}(s|m;L) \propto pr(s) S_{0}(m|s;L) \label{pragl}\\
&S_{1}(m|s;L) \propto  \exp(\lambda \; R_{0}(s|m;L)^\alpha) \label{prags}
\end{flalign}

\paragraph{Symmetrized expected utility.} With $P \in \Delta(S)$ (uniform so far; therefore $P = pr$):
\begin{itemize}

  \item $U(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)] / 2$
  \item $U_S(t_i,t_j) = \sum_s P(s) \; \sum_m P_S(m|s;t_i) \; \sum_{s'} P_R(s'|m,t_j) \; \delta(s,s')$, where $\delta(s,s')$ returns $1$ iff $s = s'$ and otherwise $0$
  \item $U_R(t_i,t_j) = U_S(t_j,t_i)$
\end{itemize}

\subsection{Part 2: Mutation \& replication.}
The proportion of players of type $i$, $x_i$, is initialized as an arbitrary distribution over $T$. $p^\star \in \Delta(T)$ is learning a prior over (player) types dependent only on the lexicon of the type. 
\begin{itemize}
    \item $f_i = \sum_j x_j U(x_i,x_j)$
    \item $\Phi = \sum_i x_i f_i$
    \item $Q_{ij} \propto [\sum_d P(d|t_i) \; P(t_j|d)]^l$, where $P(t_j|d) \propto p^\star(t_j) P(d|t_j)$, $d$ is a sequence of observations of length $k$ of the form \tuple{\tuple{s_i,m_j}, ... \tuple{s_k, m_l}}, and $l \geq 1$ is a learning parameter.
	\item For parental learning (standard RMD): $\dot x_i = \sum_j Q_{ji} \frac{x_j f_j}{\Phi}$
\end{itemize}

\subsection{Model parameters \& procedure} 
\begin{enumerate}
  \item Sequence length $k$
  \item Pragmatic production parameter $\alpha$
  \item Rationality parameter $\lambda$
  \item Learning prior over types (lexica); cost parameter $c$. $p^\star(t_i) \propto n - c \cdot r$ where $n$ is the total number of states and $r$ that of upper-bounded messages only true of $s_1$ in $t_i$'s lexicon (if only $s_1$ is true of a message, then this message encodes an upper-bound). Then the score for $L_1$, $L_3$, $L_5$ is $2$, that of $L_4$ and $L_6$ is $2-c$, and that of $L_2$ is $2-2c$; Normalization over lexica scores yields the prior over lexica (which is equal to the prior over types).   
  \item Prior over meanings ($pr$). We assume that $pr(s) = \frac{1}{|S|}$ for all $s$.
  \item True state distribution ($P$). We currently assume that $P = \frac{1}{|S|}$ but it may be interesting to vary this
  \item Learning parameter $l \geq 1$ with $1$ corresponding to probability matching, and MAP as $l$ approaches infinity
  \item $n$ is the sample of sequences of observations of length $k$ sampled from the production probabilities of each type
  \item Number of generations $g$
\end{enumerate}

\paragraph{Procedural description.} The game is initialized with some arbitrary distribution over player types. At the game's onset we compute $Q$ once based on the sets  of sequences $D$ (one for each parent type). Replicator dynamics are computed based on the fitness of each type in the current population as usual. $Q$ is computed anew for each independent run (of $g$ generations) given that it depends on $D$, which is sampled from production probabilities.

%\bibliographystyle{apacite}
\bibliographystyle{unsrtnat}

%\setlength{\bibleftmargin}{.125in}
%\setlength{\bibindent}{-\bibleftmargin}
\bibliography{./bounds-rmd}

\end{document}

### 2016-08-23 paper/paper.tex
initial commit with up-to-date bounds-RMD files. However, large simulation results were purged from the repository

\documentclass[a4paper]{article}

\usepackage{geometry}
\usepackage{natbib}
\bibpunct[,]{(}{)}{,}{a}{}{;}

%--------------------
\usepackage{gb4e}
\noautomath

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{amssymb}
%\usepackage{stmaryrd}
%\usepackage{multicol}
%\usepackage{graphicx}
\usepackage{color}
%\newcommand{\mvalueof}[1]{\llbracket#1\rrbracket}
\newcommand{\citeposs}[2][]{\citeauthor{#2}'s (\citeyear[#1]{#2})}
\newcommand{\tuple}[1]{\ensuremath{\left\langle #1 \right\rangle}}

\newcommand{\hl}[1]{\textcolor[rgb]{.8,.33,.0}{#1}}% prints in orange
%\newcommand{\argmax}[1]{\underset{#1}{\operatorname{arg}\,\operatorname{max}}\;}
%\newcommand{\argmin}[1]{\underset{#1}{\operatorname{arg}\,\operatorname{min}}\;}
%\newcommand{\sbna}{\exists\lnot\forall}

%--------------------
%
%\usepackage{setspace}
%\onehalfspacing
%
%-------------------

\title{Communicative pressures at the semantics-pragmatics interface}

\author{%\bf NAME1 and NAME2\\
    ( -- draft \today --- )
}

\date{}

\begin{document}
\maketitle
\section{The semantics-pragmatics divide}\label{sec:introduction}

\begin{abstract} Natural languages allow for the conventional pragmatic enrichment of semantic structure in a notably productive fashion. As a result, the information conveyed often goes beyond the literal meaning of an expression. This raises a challenge to explain the regular selection of particular semantics over others in light of pragmatics. To address this challenge, we propose a general model for the analysis of linguistic pressures that integrates (iterated) Bayesian learning in the replicator-mutator dynamics commonly used in evolutionary game theory. This model allows for population-level analyses of rational probabilistic language users with varied degrees of pragmatic sophistication and distinct languages. In an application to scalar implicatures, we show that semantics that allow for pragmatic enrichment are selected for when languages are pressured towards learnability. Crucially, this result hinges on pragmatics to compensate for the disadvantage in expressivity that users of such languages incur otherwise.
\end{abstract}

In linguistic theorizing, it is common to draw a distinction between semantics and pragmatics. Broadly speaking, the former concerns the truth-conditional content of expressions, whereas the latter concerns information beyond such literal meanings. For instance, the integration of context and the derivation of defeasible inferences. Thus, the information conveyed by an utterance is seldom, if ever, solely determined by semantics, but rather in tandem with pragmatics.

Much research in semantics and pragmatics is aimed at characterizing classes of expressions in terms of either domain, or their interplay. Thus, while there is disagreement on where their boundary lies, the distinction between semantics and pragmatics plays an important role in linguistics. However, an issue that has received little attention is the justification of semantic structure in light of pragmatics. That is, an account for the selection and pervasiveness of particular semantic structures over others under consideration of the informational enrichment provided by pragmatics.

%For instance, truth-conditionally, {\em Bill read five books} does not set an upper-bound to the amount of books that Bill read; he may have read six, seven, or more books. However, the (defeasible) inference that {\em Bill read no more than five books} may be drawn on pragmatic grounds. After all, if Bill had read, for example, six books, and had the speaker known this, she would have said so. In this manner, what is conveyed (pragmatically) can go beyond what is (literally) said.

%Prima facie, it is puzzling that {\em Bill read five books} does not semantically rule out its more informative or ``stronger'' alternatives. More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} or {\em big} were truth-conditionally incompatible with stronger alternatives such as, respectively, {\em hot}, {\em and}, {\em all} and {\em huge}? After all, analogously to the example sketched above, this is what a pragmatically enriched interpretation yields; an upper-bound that rules out these alternatives through a so-called scalar inference \citep{horn:1972,gazdar:1979}. The present investigation focuses on the lack of lexicalization of such upper-bounds in natural languages. That is, we seek to provide an explanation for the selection of the particular semantics of scalar expressions, which in turn hinges on the semantics-pragmatics distinction for what is ultimately conveyed to interlocutors. However, once a distinction between semantics and pragmatics is drawn, similar questions can be posed for other types of pragmatic inferences, such as ignorance and manner inferences.

A number of recent investigations have began to address issues pertaining to the development and selection of linguistic properties (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews). Our starting point is given by the overarching account of competing pressures that has crystallized across these approaches: Natural languages are pressured for successful communication, one the one hand, and for acquisition, on the other. That is, they need to be well-adapted to the communicative needs within a linguistic community, as well as be learnable to survive their faithful transmission across generations.

We proceed by instantiating these pressures in the replicator-mutator dynamics, commonly used in evolutionary game theory. This model allows for general and precise means to model the dynamics of linguistic pressures by combining functional pressure on successful communication, effects of learning biases on (iterated) Bayesian learning \citep{griffiths+kalish:2007}, and  probabilistic models of language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}. Drawing from the latter research strand allows us to make the distinction between semantics and pragmatics in communicative behavior precise. In this way, this model bridges the recent surge of synchronic probabilistic rational language use analyses with diachronic models of cultural evolution. We then analyze the prevalence of a lack of semantic upper-bounds in the literal meaning of weak scalar expressions. We show that a lack of upper-bounds is selected for when learners are biased towards simpler semantic representations, provided they have means to convey upper-bounds, i.e., provided that they can be derived via pragmatic reasoning.

\section{Simplicity and expressivity at the semantics-pragmatics interface, and beyond}

\begin{itemize}
  \item \hl{General introduction to the idea of competing pressures in shaping languages (\citealt{zipf:1949}, Piantadosi,...}
  %Zipf's principle of least effort as an account of human behavior. In language he interpreted it as minimal articulation with maximal expliciteness
  %Martinet 1962:139 describes language change as an interaction of `` ‘first,   the   requirements   of  communication,  the  need  for  the  speaker  to  convey  his  message,  and  second,  the principle of least effort, which makes him restrict his output of energy, both mental and physical, to the minimum compatible with achieving his ends
  \item \hl{Idea in pragmatics: \citealt{clark+wilkes-gibbs:1986}, \citealt{horn:1984}}
  %Principle of least collaborative effort The principle  of least  collaborative  effort:  In conversation, the  participants  try  to  minimize  their  collaborative  effort—the  work that both do from the initiation of each contribution to its mutual acceptance
  \item \hl{Discuss Q- and R-predictions for language change (cf. \citealt{traugott:2004})}
  \item \hl{Idea in LOT (\citealt{piantadosi+etal:underreview})}
\end{itemize}

\subsection{Competing pressures in cultural evolution}
\begin{itemize}
  \item \hl{discuss past research on interplay of pressures from a population-level perspective (see discussion in \citealt{kirby+etal:2015} for IL}
  \item \hl{Learnability in iterated learning}
  \item \hl{Expressivity in GT and some IL}
  \item \hl{Sketch of replicator-mutator dynamics}
\end{itemize}

\hl{More detailed discussion of models of cultural evolution. Short overview of past research with a focus on the difference between pure IL and functional pressure together with IL. {\bf Possibly add a direct comparison of IL and RMD in the appendix using the setup of Griffiths \& Kalish 2007. This may not be necessary.}}

The emergence and change of linguistic structure is driven by many factors, from biological and socio-ecological to cultural \citep{steels:2011,tamariz+kirby:2016}. Broadly put, social and ecological pressures determine communicative needs, while biology determines the architecture available for its use. Our focus is on the latter, cultural, factor, wherein linguistic structure is analyzed in terms of its use, as well as its transmission across generations.

As already mentioned in \S\ref{sec:introduction}, research on selectional forces that apply in the cultural evolution of language has focused on two main pressures: expressivity and learnability. However, while it is generally acknowledged that both play a pivotal role, past approaches have focused exclusively, or at least emphasized, the role of one over the other (a recent exception is \citealt{kirby+etal:2015}).

Expressiveness, or communicative efficiency, has been at the center of applications of evolutionary game theory to linguistics \citep{nowak+krakauer:1999,huttegger+zollman:2013},  \hl{explain RMD}

In contrast, the iterated learning paradigm has focused on the effects of language transmission from generations of speakers to the next. \hl{explain IL}

\section{Model}
\subsection{Replicator-mutator dynamics}
\subparagraph{(II) Sequences and atomic observations.} Before, the set of all observations was $O =$\linebreak  $\{\tuple{\tuple{s_1,m_i},\tuple{s_2,m_j}} | m_i, m_j \in M\}$. A member of $O$ encodes that a teacher produced $m_i$ in state $s_1$ and $m_j$ in $s_2$, i.e., it encodes one witnessed message for each state. A datum $d$ was a sequence of length $k$ of members of $O$. Learners witnessed such data sequences. Now, more in line with \citet{griffiths+kalish:2007}, $O = \{\tuple{s_i,m_j} | s_i \in S, m_j \in M\}$ and $d$ is a sequence of length $k$ of members of $O$. The main difference is that now some $d$ do not provide any production information for some states.

\subparagraph{(III) Observations as production.} Instead of taking the space of all possible sequences of length $k$ into consideration, we take sample from $O$ $k$-times according to the production probabilities of each type; $P(o = \tuple{s,m} | t_i) = P(s) P(m|s,t_i)$. $n$ such $k$-length sequences are sampled for each type. As a consequence, the data used for computing $Q_i$ is not the same as that used for $j$ $(i \neq j)$.

\subparagraph{(IV) Parametrized learning} $Q_{ij} \propto \sum_d P(d|t_i) F(t_j,d)$, where $F(t_j,d) \propto P(t_j|d)^l$ and $l =1$ corresponds to probability matching and, as $l$ increases towards infinity, to MAP.

The proportion of players of type $i$, $x_i$, is initialized as an arbitrary distribution over $T$. $p^\star \in \Delta(T)$ is learning a prior over (player) types dependent only on the lexicon of the type. 
\begin{itemize}
    \item $f_i = \sum_j x_j U(x_i,x_j)$
    \item $\Phi = \sum_i x_i f_i$
    \item $Q_{ij} \propto \sum_d P(d|t_i) \; P(t_j|d)$, where $P(t_j|d) \propto [P(t_j) P(d|t_j)]^l$, $d$ is a sequence of observations of length $k$ of the form \tuple{\tuple{s_i,m_j}, ... \tuple{s_k, m_l}}, and $l \geq 1$ is a learning parameter.
	\item For parental learning (standard RMD): $\dot x_i = \sum_j Q_{ji} \frac{x_j f_j}{\Phi}$
\end{itemize}

\subsection{Expressiveness as fitness-relative replication}

\paragraph{Symmetrized expected utility.} With $P \in \Delta(S)$ (uniform so far; $P = pr$):
\begin{itemize}

  \item $U(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)] / 2$
  \item $U_S(t_i,t_j) = \sum_s P(s) \; \sum_m P_S(m|s;t_i) \; \sum_{s'} P_R(s'|m,t_j) \; \delta(s,s')$, where $\delta(s,s')$ returns $1$ iff $s = s'$ and otherwise $0$
  \item $U_R(t_i,t_j) = U_S(t_j,t_i)$
\end{itemize}

\subsection{Iterated learning as acquisition-based mutation}

\subsection{Signaling behavior}
\hl{Leave languages undefined until application}

\paragraph{Signaling behavior.}
\hl{Exposition of signaling behavior as reasoning hierarchy, which we use to make a distinction between semantic and pragmatic language users}

With $\lambda \geq 1$ (rationality parameter), $\alpha \in [0,1]$ (pragmatic violations) and $pr \in \Delta(S)$ a common prior over $S$ (uniform so far):

\begin{flalign}
&R_{0}(s|m;L) \propto pr(s) L_{sm}\label{litl}\\
&S_{0}(m|s;L) \propto \exp(\lambda \; L_{sm}) \label{lits}\\
&R_{1}(s|m;L) \propto pr(s) S_{0}(m|s;L) \label{pragl}\\
&S_{1}(m|s;L) \propto  \exp(\lambda \; R_{0}(s|m;L)^\alpha) \label{prags}
\end{flalign}

\section{Lack of semantic upper-bounds in lexical meaning}

\paragraph{Procedural description.} The game is initialized with some arbitrary distribution over player types. At the game's onset we compute $Q$ once based on the sets  of sequences $D$ (one for each parent type). Replicator dynamics are computed based on the fitness of each type in the current population as usual. $Q$ is computed anew for each independent run (of $g$ generations) given that it depends on $D$, which is sampled from production probabilities.

\paragraph{Languages.} We consider a population of players with two signaling behaviors, literal and Gricean (level $0$ and $1$ below), each equipped with one of $6$ lexicons. This yields a total of $12$ distinct player types $t \in T$. $|M| = |S| = 2$, i.e., a lexicon is a $(2,2)$-matrix. These are listed in Table \ref{tab:lexica}.

\begin{table}[h]
\centering 
\begin{tabular}{l c l}
$L_1$ = $\begin{pmatrix} 0 & 0 \\ 1 & 1 \end{pmatrix}$ & 
$L_2$ = $\begin{pmatrix} 1 & 1 \\ 0 & 0 \end{pmatrix}$ & 
$L_3$ = $\begin{pmatrix} 1 & 1 \\ 1 & 1 \end{pmatrix}$\\[0.5cm]

$L_4$ = $\begin{pmatrix} 0 & 1 \\ 1 & 0 \end{pmatrix}$ &
$L_5$ = $\begin{pmatrix} 0 & 1 \\ 1 & 1 \end{pmatrix}$ &
$L_6$ = $\begin{pmatrix} 1 & 1 \\ 1 & 0 \end{pmatrix}$
\end{tabular}
\caption{{\footnotesize Set of considered lexica.}}
\label{tab:lexica}
\end{table}

As in the CogSci paper, $L_4$ (semantic upper-bound for $m_2$) and $L_5$ (no semantic upper-bound for $m_2$) are the target lexica. Gricean $L_5$ users can convey/infer the bound pragmatically, while literal/Gricean $L_4$ users do so semantically.

\subsection{Model parameters \& procedure} 
\begin{enumerate}
  \item Sequence length $k$
  \item Pragmatic production parameter $\alpha$
  \item Rationality parameter $\lambda$
  \item Learning prior over types (lexica); cost parameter $c$. $p^\star(t_i) \propto n - c \cdot r$ where $n$ is the total number of states and $r$ that of upper-bounded messages only true of $s_1$ in $t_i$'s lexicon (if only $s_1$ is true of a message, then this message encodes an upper-bound). Then the score for $L_1$, $L_3$, $L_5$ is $2$, that of $L_4$ and $L_6$ is $2-c$, and that of $L_2$ is $2-2c$; Normalization over lexica scores yields the prior over lexica (which is equal to the prior over types).   
  \item Prior over meanings ($pr$). We assume that $pr(s) = \frac{1}{|S|}$ for all $s$.
  \item True state distribution ($P$). We currently assume that $P = \frac{1}{|S|}$ but it may be interesting to vary this
  \item Learning parameter $l \geq 1$ with $1$ corresponding to probability matching, and MAP as $l$ approaches infinity
  \item $n$ is the sample of sequences of observations of length $k$ sampled from the production probabilities of each type
  \item Number of generations $g$
\end{enumerate}

\section{Discussion}

\section{Extensions}
\subparagraph{(I) Cost for pragmatic reasoning.} At least in the CogSci setup the effect of adding cost to pragmatic reasoning is unsurprising: High cost for pragmatic signaling lowers the prevalence of pragmatic types. Lexica that semantically encode an upper-bound benefit the most from this. However, the cost needed to be substantial to make the pragmatic English-like lexicon stop being the incumbent type (particularly when learning is communal).

\subparagraph{(II) Negative learning bias.} Instead of penalizing complex semantics (semantic upper-bounds) one may consider penalizing simple semantics (no upper-bounds). This is useful as a sanity check but also yields unsurprising results in the CogSci setup: The more learners are biased against simple semantics, the more prevalent are lexica that semantically encode upper-bounds.

\subparagraph{(III) Inductive bias.} A second learning bias that codifies the idea that lexica should be uniform, i.e. be biased towards either lexicalizing an upper-bound for all weaker alternatives in a scalar pair or for none.

\subparagraph{(IV)  Uncertainty.} The other advantage of non-upper bounded semantics lies in being non-committal to the negation of stronger alternatives when the speaker is uncertain. Adding this to the model requires the most changes to our present setup and some additional assumptions about the cues available to players to discern the speaker's knowledge about the state she is in.

\subparagraph{(V) More scalar pairs.} Taking into consideration more than one scalar pair. Preliminary results suggest that this does not influence the results in any meaningful way without further additions, e.g. by (III).

\subparagraph{(VI) More lexica.} Not necessary. Preliminary results suggest that considering more lexica has no noteworthy effect on the dynamics (tested with all possible 2x2 lexica).

\subparagraph{(VII) State frequencies.} Variations on state frequencies. This may have an interesting interaction with (III).

\subparagraph{(VIII) Reintroduction of communal learning.} One possibility: The probably $N_{ij}$ with which a child of $t_i$ adopts $t_j$ could be the weighted sum of $Q_{ij}$ (as before) and a vector we get from learning from all of the population: $L_j = \sum_d P(d | \vec{p})  P(t_j | d)$, where $P(d | \vec{p}) = \sum_{i} P(d | t_i)  \vec{p}_i$ is the probability of observing $d$ when learning from a random member of the present population distribution.

\section{Conclusion}

%%% Old snippets %%%
%\section{Conveying upper-bounds}
%Scalar inferences refer to the pragmatic derivation of an upper-bound for weak scalar expressions to the effect that stronger alternatives are inferred to not hold, e.g. {\em some students came} may be taken to convey that {\em not all students came}. The order of an expression such as {\em some} with respect to an alternative, e.g., {\em all}, is induced by entailment. For instance, {\em all students came} entails {\em most students came}, which in turn entails {\em some students came}. In this sense, {\em some} is weaker than {\em all}. A considerable class of natural language expressions do not lexicalize an upper-bound and can be ordered in this fashion, allowing for their pragmatic strengthening. As alluded at above, examples in English include numerals, scalar adjectives, quantifiers, modals, and connectives. \hl{possibly add some typological data on universality, frequency, monomorphemic status}
%
%The pragmatic enrichment of the semantic content of such expressions is enabled by mutual reasoning \citep{grice:1975}. More specifically, it is driven by interlocutors' mutual expectations of rational language use. The hearer reasons about the speaker's choice of a weak alternative over a stronger one. Had the speaker known that a stronger alternatives holds, she would have said so as this would have been more informative. Since she did not, the hearer can infer that the stronger alternative does not hold. Analogously, a speaker who reasons about her addressee may rely on her to derive this inference. In this way, a strengthened, upper-bounded, state of affairs can be conveyed without codifying the bound explicitly in the semantics.
%
%However, while pragmatics offers means to convey upper-bounds, the question why they are not part of the lexical meaning of these expressions remains. There are two main explanatory venues for this pattern. The first targets the functional advantages a lack of upper-bounds may confer to language users, whereas the second focuses on a learnability advantage of simpler over more complex semantic representations. 
%
%\paragraph{Function-based explanations.} Two assumptions are key to the pragmatic strengthening of weak alternatives: (the assumption of) cooperation and knowledge about the issue at hand. That is, the hearer needs to assume the speaker to be as informative as possible, i.e., not to withhold information, and that the speaker is knowlegeable, e.g., she knows whether {\em all students came}. Conversely, the speaker needs to assume the hearer to regard these conditions as satisfied. It is not difficult to imagine scenarios where either or both of these conditions are not given. For instance, the speaker may (be assumed to) not want to disclose all information about the students' attendance, or may have left early without being able to verify the attendance to a satisfactory degree. 
%
%\hl{Discussion of functional pressures for a lack of upper-bounds}

%\paragraph{Learning-based explanations.}
%
%\hl{Discussion of our main assumption that a lack of upper-bounds provides a learnability advantage framed in terms of relative representational simplicity over the codification of an upper-bound. {\bf Should this be made precise? If so, in which way?}}

%\bibliographystyle{apacite}
\bibliographystyle{unsrtnat}

%\setlength{\bibleftmargin}{.125in}
%\setlength{\bibindent}{-\bibleftmargin}
\bibliography{./bounds-rmd}

\end{document}

### 2016-08-23 1rmd-singlescalar.py

@@ -1,0 +1,163 @@
+#####
+#RMD with parametrized iterated parental learning 
+#1 pairs of scalar items, six lexica per scalar pair. 
+#2 possible signaling behaviors: literal or gricean
+#12 types (6 lexica * 2 signaling behaviors) per signaling pair
+#6 literal types, 6 gricean types, 12 types in total
+#####
+
+import numpy as np
+np.set_printoptions(threshold=np.nan)
+from random import sample
+from itertools import product
+from player import LiteralPlayer,GriceanPlayer
+import sys 
+import datetime
+import csv
+
+#####
+l1,l2,l3,l4,l5,l6 = np.array( [[0.,0.],[1.,1.]] ), np.array( [[1.,1.],[0.,0.]] ), np.array( [[1.,1.],[1.,1.]] ), np.array( [[0.,1.],[1.,0.]] ), np.array( [[0.,1.],[1.,1.]] ), np.array( [[1.,1.],[1.,0.]] )
+
+alpha = 1 # rate to control difference between semantic and pragmatic violations
+cost = 0.1 # cost for LOT-concept with upper bound
+lam = 30 # soft-max parameter
+k = 3  # length of observation sequences
+sample_amount = 15 #amount of k-length samples for each production type
+
+gens = 20 #number of generations per simulation run
+runs = 50 #number of independent simulation runs
+
+states = 2 #number of states
+messages = 2 #number of messages
+
+learning_parameter = 1 #prob-matching = 1, increments approach MAP
+state_freq = np.ones(states) / float(states) #frequency of states s_1,...,s_n 
+
+
+f = csv.writer(open('./results/singlescalar-a%d-c%.2f-l%d-k%d-samples%d-learn%.2f-g%d-r%d.csv' %(alpha,cost,lam,k,sample_amount, learning_parameter, gens,runs),'wb')) #file to store mean results
+
+f.writerow(["run_ID", "t1_initial", "t2_initial","t3_initial","t4_initial","t5_initial","t6_initial","t7_initial","t8_initial","t9_initial","t10_initial","t11_initial","t12_initial","alpha", "prior_cost_c", "lambda", "k", "sample_amount", "learning_parameter", "generations", "t1_final", "t2_final","t3_final","t4_final","t5_final","t6_final","t7_final","t8_final","t9_final","t10_final","t11_final","t12_final"])
+
[…]

### 2016-08-23 2rmd-alllexica.py

@@ -1,0 +1,177 @@
+##### Variant with all possible 2x2 lexica for a single scalar pair
+#RMD with parametrized iterated parental learning 
+#1 pairs of scalar items, six lexica per scalar pair. 
+#2 possible signaling behaviors: literal or gricean
+#12 types (6 lexica * 2 signaling behaviors) per signaling pair
+#6 literal types, 6 gricean types, 12 types in total
+#####
+
+import numpy as np
+np.set_printoptions(threshold=np.nan)
+from random import sample
+from itertools import product
+from player import LiteralPlayer,GriceanPlayer
+import sys 
+import datetime
+import csv
+
+#####
+
+alpha = 1# rate to control difference between semantic and pragmatic violations
+cost = 0.9 # cost for LOT-concept with upper bound
+lam = 10 # soft-max parameter
+k = 3  # length of observation sequences
+sample_amount = 10 #amount of k-length samples for each production type
+
+gens = 100 #number of generations per simulation run
+runs = 50 #number of independent simulation runs
+
+states = 2 #number of states
+messages = 2 #number of messages
+
+learning_parameter = 1 #prob-matching = 1, increments approach MAP
+state_freq = np.ones(states) / float(states) #frequency of states s_1,...,s_n 
+
+
+print '#Starting, ', datetime.datetime.now()
+
+def lexica(s,m):
+    values = list(product((0.,1.), repeat = s * m))
+    return [np.array([[i[0],i[1]],[i[2],i[3]]]) for i in values]
+lexica = lexica(states,messages)
+
+
+def types(l):
+    typeList = []
+    for i in l:
+        typeList.append(LiteralPlayer(alpha,i))
+    for i in l:
+        typeList.append(GriceanPlayer(alpha,lam,i))
+    return typeList 
+
+typeList = types(lexica)
+
+def lex_prior(l):
+    cost_v = np.zeros(len(lexica))
+    for j in range(len(l)):
+        lexicon_score = 0.
+        for i in range(l[j].shape[1]):
+            if (i == 0 or i == 1) and l[j][0,i] > 0 and l[j][1,i] == 0:
+                lexicon_score += cost
+        cost_v[j] = states - lexicon_score
+    return np.append(cost_v,cost_v)
+
+lex_prior = lex_prior(lexica)
+lexica_prior = lex_prior / sum(lex_prior)
+
+
[…]

### 2016-08-23 IL-comparison/il-comparison-rmd.py

@@ -1,0 +1,292 @@
+import numpy as np
+#np.set_printoptions(threshold=np.nan)
+from itertools import product,permutations
+from random import choice,sample
+from player import Player
+import sys 
+import datetime
+import csv
+
+def normalize(m):
+    return m / m.sum(axis=1)[:, np.newaxis]
+
+def norm(m):
+    m = m / m.sum(axis=1)[:, np.newaxis]
+    m[np.isnan(m)] = 1. / np.shape(m)[1]
+    return m
+
+#lexicon = np.array([[1,0,1],[0,1,1]])
+s = 4 #number of states
+n = 10 #observation sequence length
+generations = 10000 #generations per game
+games = 1000 #amount of independent games
+alpha = 0.5 #bias parameter
+epsilon = 0.05
+mc = 1000 #amount of samples
+
+
+f_il_samp_mean = csv.writer(open('./results/comparison-il-samp-mean-a%d-e%f-k%d-g%d-r%d.csv' %(alpha,epsilon,n,generations,games),'wb')) 
+f_unwgh_samp_mean = csv.writer(open('./results/comparison-unwgh-samp-mean-a%d-e%f-k%d-g%d-r%d.csv' %(alpha,epsilon,n,generations,games),'wb')) 
+f_wgh_samp_mean = csv.writer(open('./results/comparison-wgh-samp-mean-a%d-e%f-k%d-g%d-r%d.csv' %(alpha,epsilon,n,generations,games),'wb')) 
+
+f_il_map_mean= csv.writer(open('./results/comparison-il-map-mean-a%d-e%f-k%d-g%d-r%d.csv' %(alpha,epsilon,n,generations,games),'wb')) 
+f_unwgh_map_mean= csv.writer(open('./results/comparison-unwgh-map-mean-a%d-e%f-k%d-g%d-r%d.csv' %(alpha,epsilon,n,generations,games),'wb')) 
+f_wgh_map_mean = csv.writer(open('./results/comparison-wgh-map-mean-a%d-e%f-k%d-g%d-r%d.csv' %(alpha,epsilon,n,generations,games),'wb')) 
+
+f_prior = csv.writer(open('./results/prior-a%d-e%f-k%d-g%d-r%d.csv' %(alpha,epsilon,n,generations,games),'wb')) 
+
+print '#Starting, ', datetime.datetime.now()
+
+def lexica(s):
+    values = list(set(permutations((1.,0.,0.,0.), s )))
[…]

### 2016-08-23 IL-comparison/player.py

@@ -1,0 +1,27 @@
+import numpy as np
+
+def normalize(m):
+    m = m / m.sum(axis=1)[:, np.newaxis]
+    m[np.isnan(m)] = 0.
+    return m
+
+
+class Player:
+    def __init__(self,lexicon,epsilon):
+        self.states = [00,01,10,11]
+        self.messages = [00,01,10,11]
+        self.lexicon = lexicon
+        self.epsilon = epsilon
+
+    def senderMatrix(self):
+        out = np.zeros((len(self.states),len(self.messages)))
+        for i in range(np.shape(out)[0]):
+            for j in range(np.shape(out)[1]):
+                if self.lexicon[i,j] == 1:
+                    out[i,j] = 1 - self.epsilon
+                else:
+                    out[i,j] = self.epsilon / (len(self.messages) - 1)
+        return out
+
+    def receiverMatrix(self):
+        return normalize(np.transpose(self.senderMatrix()))

### 2016-08-23 IL-comparison/plot.py

@@ -1,0 +1,180 @@
+from math import sqrt
+import matplotlib.pyplot as plt
+import seaborn as sns
+sns.set_context(rc={'lines.markeredgewidth': 0.5})
+import csv
+import numpy as np
+import sys
+
+
+f1 = csv.reader(open('./results/comparison-il-samp-mean-a0-e0.050000-k3-g100-r10.csv'))
+f2 = csv.reader(open('./results/comparison-il-map-mean-a0-e0.050000-k3-g100-r10.csv'))
+f3 = csv.reader(open('./results/comparison-unwgh-samp-mean-a0-e0.050000-k3-g100-r10.csv'))
+f4 = csv.reader(open('./results/comparison-unwgh-map-mean-a0-e0.050000-k3-g100-r10.csv'))
+f5 = csv.reader(open('./results/comparison-wgh-samp-mean-a0-e0.050000-k3-g100-r10.csv'))
+f6 = csv.reader(open('./results/comparison-wgh-map-mean-a0-e0.050000-k3-g100-r10.csv'))
+f7 = csv.reader(open('./results/prior-a0-e0.050000-k3-g100-r10.csv'))
+
+compIndices = [45, 57, 108, 120]
+
+def results_in_list(f):
+    results = []
+    firstline = True
+    for i in f:
+        if firstline:
+            firstline = False
+            continue
+        results.append(float(i[1]))
+    return results
+
+il_samp = results_in_list(f1)
+il_map = results_in_list(f2)
+unwgh_samp = results_in_list(f3)
+unwgh_map = results_in_list(f4)
+wgh_samp = results_in_list(f5)
+wgh_map = results_in_list(f6)
+prior = results_in_list(f7)
+
+def reorder_by_compositionality(lst):
+    compo = [lst[compIndices[i]] for i in xrange(len(compIndices))]
+    print compo
+    for j in xrange(len(compIndices)):
+        lst.pop(compIndices[j])
+        lst.insert(j,compo[j])
+    return lst
+
+il_samp = reorder_by_compositionality(il_samp)
+il_map = reorder_by_compositionality(il_map)
+unwgh_samp = reorder_by_compositionality(unwgh_samp)
+unwgh_map = reorder_by_compositionality(unwgh_map)
+wgh_samp = reorder_by_compositionality(wgh_samp)
+wgh_map = reorder_by_compositionality(wgh_map)
[…]

### 2016-08-23 mf_rstuff/analysis.r

@@ -1,0 +1,21 @@
+require('ggplot2')
+require('dplyr')
+require('reshape2')
+
+d = read.csv('merged_select.csv')
+ds = select(d, prior_cost_c, k, learning_parameter, t11_initial, lambda, t11_final) %>% 
+  rename(c = prior_cost_c, l = learning_parameter, target = t11_final)
+
+mFull = lm(formula = target ~ c * k * l * lambda * t11_initial, data = ds)
+m = lm(formula = target ~ c * k * l, data = ds)
+mCKL = lm(formula = target ~ c + k + l + lambda, data = ds)
+mCK = lm(formula = target ~ c + k, data = ds)
+mCL = lm(formula = target ~ c + l, data = ds)
+mKL = lm(formula = target ~ k + l, data = ds)
+mC = lm(formula = target ~ c, data = ds)
+mK = lm(formula = target ~ k, data = ds)
+mL = lm(formula = target ~ l, data = ds)
+
+AIC(mFull,m,mCKL,mCL,mCK,mC,mKL,mL,mK)
+summary(mFull)
+

### 2016-08-23 modsinglescalar.py

@@ -1,0 +1,156 @@
+#####
+#RMD with parametrized iterated parental learning 
+#1 pairs of scalar items, six lexica per scalar pair. 
+#2 possible signaling behaviors: literal or gricean
+#12 types (6 lexica * 2 signaling behaviors) per signaling pair
+#6 literal types, 6 gricean types, 12 types in total
+#####
+
+import numpy as np
+np.set_printoptions(threshold=np.nan)
+from random import sample
+from itertools import product
+from player import LiteralPlayer,GriceanPlayer
+import sys 
+import datetime
+import csv
+
+def normalize(m):
+    return m / m.sum(axis=1)[:, np.newaxis]
+
+def summarize_counts(lst,states,messages):
+    """summarize counts for tuples of k-states and k-messages""" 
+    counter = [0 for _ in xrange(states**messages)]
+    for i in xrange(len(lst)):
+        s,m = lst[i][0] *2, lst[i][1]
+        counter[s+m] += 1
+    return counter
+
+def get_obs(k,states,messages,likelihoods,state_freq,sample_amount):
+    """Returns summarized counts of k-length <s_i,m_j> production observations as [#(<s_0,m_0>), #(<s_0,m_1), #(<s_1,m_0>, #(s_1,m_1)], ...]] = k"""
+    s = list(xrange(states))
+    m = list(xrange(messages))
+    atomic_observations = list(product(s,m))
+   
+    obs = [] #store all produced k-length (s,m) sequences 
+    for t in xrange(12):
+        produced_obs = [] #store k-length (s,m) sequences of a type
+        production_vector = likelihoods[t].flatten()
+        doubled_state_freq = np.column_stack((state_freq,state_freq)).flatten() # P(s)
+        sample_vector = production_vector * doubled_state_freq #P(s) * P(m|s,t_i)
+        for i in xrange(sample_amount):
+            sample_t = [np.random.choice(len(atomic_observations),p=sample_vector) for _ in xrange(k)]
+            sampled_obs = [atomic_observations[i] for i in sample_t]
+            produced_obs.append(summarize_counts(sampled_obs,states,messages))
+        obs.append(produced_obs)
+    return obs
+
+
[…]

### 2016-08-23 old-variants/1plot.py

@@ -1,0 +1,68 @@
+#####
+#Read in CSV, plot results.
+#####
+import numpy as np
+import matplotlib.pyplot as plt
+import seaborn as sns
+sns.set_context(rc={'lines.markeredgewidth': 0.5})
+import csv
+import sys
+
+f1 = csv.reader(open('./results/singlescalar-wgh-mean-a1-c0.100000-l30-k9-g30-r100.csv','rt'))
+f2 = csv.reader(open('./results/1multiscalar-wgh-mean-a1-c0.100000-l30-k9-g30-r100.csv','rt'))
+
+def results_by_lexica_multi(f):
+    p = np.zeros(6)
+    firstline = True
+    for i in f:
+        if firstline:
+            firstline = False
+            continue
+        p[int(i[1])] += float(i[-1]) /3.
+        p[int(i[2])] += float(i[-1]) /3.
+        p[int(i[3])] += float(i[-1]) /3.
+    return p
+
+def results_by_lexica_single(f):
+    p = np.zeros(6)
+    firstline = True
+    for i in f:
+        if firstline:
+            firstline = False
+            continue
+        p[int(i[1])] += float(i[-1])
+    return p
+
+Y_single = results_by_lexica_single(f1)
+Y_multi = results_by_lexica_multi(f2)
+
+print Y_single
+print Y_multi
+###Plots##
+
+X = np.arange(6) #number of lexica
+
+
+
+
+width = 0.35
+#ymin,ymax = 0,np.max(np.array([Ya[np.argmax(results3a)],Yb[np.argmax(results3b)],Yc[np.argmax(results3c)]]))+0.01
+#ax.set_xticks(np.array([len(compIndices)/2,(len(hypotheses)-len(compIndices))/2]))
+
+fig, ax = plt.subplots()
+bar1 = ax.bar(X, Y_single, width,color='r')
+bar2 = ax.bar(X+width,Y_multi,width,color='y')
+
+plt.ylabel('Mean in population')
+ax.set_xticks(X+width)
+ax.set_xticklabels(('L1', 'L2', 'L3', 'L4','L5','L6'))
+ax.legend((bar1[0], bar2[0]), ('single scalar', '3 scalars'))
+ax.margins(0.025, 0.025)
+#ax.text(.5,.9,'$\alpha$ = 1, c = %r, $\lambda$ = %r, pairs = %r, k = %r',
+#        horizontalalignment='center',
[…]

### 2016-08-23 old-variants/1rmd-multiscalar.py

@@ -1,0 +1,264 @@
+#####
+#RMD for unweighted and weighted (community) learning with Bayesian (iterated) learning as M. 
+#3 pairs of scalar items, six lexica per scalar pair. 
+#2 possible signaling behaviors: literal or gricean
+#12 types (6 lexica * 2 signaling behaviors) per signaling pair
+#6*6*6 = 216  literal types, 216 gricean types, 432 types in total
+#Set of observations is computed for one scalar pair first, taking the cartesian product of the likelihood matrix three times yields the full likelihood matrix
+#####
+
+
+import numpy as np
+#np.set_printoptions(threshold=np.nan)
+from random import sample
+from itertools import product, permutations, combinations_with_replacement
+from player import LiteralPlayer,GriceanPlayer
+import sys 
+import datetime
+import csv
+
+
+alpha = 1 # rate to control difference between semantic and pragmatic violations
+cost = 0.9 # cost for LOT-concept with upper bound
+lam = 30 # soft-max parameter
+k = 4  # number of learning observations
+sample_amount = 145 #amount of samples from OBS when k > 3
+
+gens = 30 #number of generations per simulation run
+runs = 100 #number of independent simulation runs
+
+
+lexical_pairs = 3 #number of 2x2 matrices
+states = 2 #number of states
+messages = 2 #number of messages
+
+
+f_unwgh_mean = csv.writer(open('./results/1multiscalar-unwgh-mean-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store each unweighted simulation run after n generations
+f_wgh_mean = csv.writer(open('./results/1multiscalar-wgh-mean-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store each weighted simulation run after n generations
+
+f_q = csv.writer(open('./results/1multiscalar-q-matrix-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store Q-matrix
+
+
+print '#Starting, ', datetime.datetime.now()
[…]

### 2016-08-23 old-variants/2plot.py

@@ -1,0 +1,69 @@
+#####
+#Read in CSV, plot results.
+#####
+import numpy as np
+import matplotlib.pyplot as plt
+import seaborn as sns
+sns.set_context(rc={'lines.markeredgewidth': 0.5})
+import csv
+import sys
+
+f1 = csv.reader(open('./results/2multiscalar-unwgh-mean-a1-c0.900000-l30-k9-g30-r100-s1.csv','rt'))
+f2 = csv.reader(open('./results/2multiscalar-wgh-mean-a1-c0.900000-l30-k9-g30-r100-s1.csv','rt'))
+
+
+def results_by_lexica_multi(f):
+    p = np.zeros(6)
+    firstline = True
+    for i in f:
+        if firstline:
+            firstline = False
+            continue
+        p[int(i[1])] += float(i[-1]) /3.
+        p[int(i[2])] += float(i[-1]) /3.
+        p[int(i[3])] += float(i[-1]) /3.
+    return p
+
+def results_by_lexica_single(f):
+    p = np.zeros(6)
+    firstline = True
+    for i in f:
+        if firstline:
+            firstline = False
+            continue
+        p[int(i[1])] += float(i[-1])
+    return p
+
+Y_single = results_by_lexica_single(f1)
+Y_multi = results_by_lexica_multi(f2)
+
+print Y_single
+print Y_multi
+###Plots##
+
+X = np.arange(6) #number of lexica
+
+
+
+
+width = 0.35
+#ymin,ymax = 0,np.max(np.array([Ya[np.argmax(results3a)],Yb[np.argmax(results3b)],Yc[np.argmax(results3c)]]))+0.01
+#ax.set_xticks(np.array([len(compIndices)/2,(len(hypotheses)-len(compIndices))/2]))
+
+fig, ax = plt.subplots()
+bar1 = ax.bar(X, Y_single, width,color='r')
+bar2 = ax.bar(X+width,Y_multi,width,color='y')
+
+plt.ylabel('Mean in population')
+ax.set_xticks(X+width)
+ax.set_xticklabels(('L1', 'L2', 'L3', 'L4','L5','L6'))
+ax.legend((bar1[0], bar2[0]), ('unweighted', 'weighted'))
+ax.margins(0.025, 0.025)
+#ax.text(.5,.9,'$\alpha$ = 1, c = %r, $\lambda$ = %r, pairs = %r, k = %r',
[…]

### 2016-08-23 old-variants/2rmd-multiscalar.py

@@ -1,0 +1,260 @@
+#####
+#RMD for unweighted and weighted (community) learning with Bayesian (iterated) learning as M. 
+#3 pairs of scalar items, six lexica per scalar pair. 
+#2 possible signaling behaviors: literal or gricean
+#12 types (6 lexica * 2 signaling behaviors) per signaling pair
+#6*6*6 = 216  literal types, 216 gricean types, 432 types in total
+#Set of observations of length k:
+##If k < 5: 6**k * 2**k observations
+##If k > 4: 20000 observations sampled from the set of all observations
+#####
+
+
+import numpy as np
+#np.set_printoptions(threshold=np.nan)
+from random import sample,choice
+from itertools import product, permutations, combinations_with_replacement
+from player import LiteralPlayer,GriceanPlayer
+import sys 
+import datetime
+import csv
+
+
+alpha = 1 # rate to control difference between semantic and pragmatic violations
+cost = 0.9 # cost for LOT-concept with upper bound
+lam = 30 # soft-max parameter
+k = 7  # number of learning observations
+sample_amount = 10000 #amount of samples from OBS when k > 4
+sample_with_replacement = False #To fix sampling problems. Set false to sample w/o replacement
+
+
+
+gens = 30 #number of generations per simulation run
+runs = 1000 #number of independent simulation runs
+
+lexical_pairs = 3 #number of 2x2 matrices
+states = 2 #number of states
+messages = 2 #number of messages
+
+
+f_unwgh_mean = csv.writer(open('./results/2multiscalar-unwgh-mean-a%d-c%f-l%d-k%d-g%d-r%d-s%d.csv' %(alpha,cost,lam,k,gens,runs,sample_with_replacement),'wb')) #file to store each unweighted simulation run after n generations
+f_wgh_mean = csv.writer(open('./results/2multiscalar-wgh-mean-a%d-c%f-l%d-k%d-g%d-r%d-s%d.csv' %(alpha,cost,lam,k,gens,runs,sample_with_replacement),'wb')) #file to store each weighted simulation run after n generations
+
+f_q = csv.writer(open('./results/2multiscalar-q-matrix-a%d-c%f-l%d-k%d-g%d-r%d-s%d.csv' %(alpha,cost,lam,k,gens,runs,sample_with_replacement),'wb')) #file to store Q-matrix
+
+
[…]

### 2016-08-23 old-variants/3plot.py

@@ -1,0 +1,69 @@
+#####
+#Read in CSV, plot results.
+#####
+import numpy as np
+import matplotlib.pyplot as plt
+import seaborn as sns
+sns.set_context(rc={'lines.markeredgewidth': 0.5})
+import csv
+import sys
+
+f1 = csv.reader(open('./results/3multiscalar-unwgh-mean-a1-c0.100000-l30-k50-g30-r100.csv','rt'))
+f2 = csv.reader(open('./results/3multiscalar-wgh-mean-a1-c0.100000-l30-k50-g30-r100.csv','rt'))
+
+
+def results_by_lexica_multi(f):
+    p = np.zeros(6)
+    firstline = True
+    for i in f:
+        if firstline:
+            firstline = False
+            continue
+        p[int(i[1])] += float(i[-1]) /3.
+        p[int(i[2])] += float(i[-1]) /3.
+        p[int(i[3])] += float(i[-1]) /3.
+    return p
+
+def results_by_lexica_single(f):
+    p = np.zeros(6)
+    firstline = True
+    for i in f:
+        if firstline:
+            firstline = False
+            continue
+        p[int(i[1])] += float(i[-1])
+    return p
+
+Y_single = results_by_lexica_single(f1)
+Y_multi = results_by_lexica_multi(f2)
+
+print Y_single
+print Y_multi
+###Plots##
+
+X = np.arange(6) #number of lexica
+
+
+
+
+width = 0.35
+#ymin,ymax = 0,np.max(np.array([Ya[np.argmax(results3a)],Yb[np.argmax(results3b)],Yc[np.argmax(results3c)]]))+0.01
+#ax.set_xticks(np.array([len(compIndices)/2,(len(hypotheses)-len(compIndices))/2]))
+
+fig, ax = plt.subplots()
+bar1 = ax.bar(X, Y_single, width,color='r')
+bar2 = ax.bar(X+width,Y_multi,width,color='y')
+
+plt.ylabel('Mean in population')
+ax.set_xticks(X+width)
+ax.set_xticklabels(('L1', 'L2', 'L3', 'L4','L5','L6'))
+ax.legend((bar1[0], bar2[0]), ('unweighted', 'weighted'))
+ax.margins(0.025, 0.025)
+#ax.text(.5,.9,'$\alpha$ = 1, c = %r, $\lambda$ = %r, pairs = %r, k = %r',
+#        horizontalalignment='center',
[…]

### 2016-08-23 old-variants/3rmd-multiscalar.py

@@ -1,0 +1,249 @@
+#####
+#RMD for unweighted and weighted (community) learning with Bayesian (iterated) learning as M. 
+#3 pairs of scalar items, six lexica per scalar pair. 
+#2 possible signaling behaviors: literal or gricean
+#12 types (6 lexica * 2 signaling behaviors) per signaling pair
+#6*6*6 = 216  literal types, 216 gricean types, 432 types in total
+#Set of observations of length k:
+##Sample n states from the set of all states, pair with n messages
+#####
+
+
+import numpy as np
+#np.set_printoptions(threshold=np.nan)
+from random import sample,choice
+from itertools import product, permutations, combinations_with_replacement
+from player import LiteralPlayer,GriceanPlayer
+import sys 
+import datetime
+import csv
+
+
+alpha = 1 # rate to control difference between semantic and pragmatic violations
+cost = 0.1 # cost for LOT-concept with upper bound
+lam = 30 # soft-max parameter
+k = 3  # number of learning observations
+sample_amount = 10000 #amount of samples
+
+
+gens = 30 #number of generations per simulation run
+runs = 100 #number of independent simulation runs
+
+lexical_pairs = 3 #number of 2x2 matrices
+states = 2 #number of states
+messages = 2 #number of messages
+
+
+
+f_unwgh_mean = csv.writer(open('./results/3multiscalar-unwgh-mean-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store each unweighted simulation run after n generations
+f_wgh_mean = csv.writer(open('./results/3multiscalar-wgh-mean-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store each weighted simulation run after n generations
+
+f_q = csv.writer(open('./results/3multiscalar-q-matrix-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store Q-matrix
+
+
+print '#Starting, ', datetime.datetime.now()
[…]

### 2016-08-23 old-variants/4-onlyparental-rmd-multiscalar.py

@@ -1,0 +1,279 @@
+#####
+#RMD for unweighted and weighted (community) learning with Bayesian (iterated) learning as M. 
+#3 pairs of scalar items, six lexica per scalar pair. 
+#2 possible signaling behaviors: literal or gricean
+#12 types (6 lexica * 2 signaling behaviors) per signaling pair
+#6*6*6 = 216  literal types, 216 gricean types, 432 types in total
+#Set of observations of length k:
+##Sample <s,m> tuples from a type's production probabilities, P(a = <s,m> | t_i) = P(s) * P(m|s,t_i)
+#
+#####
+
+
+import numpy as np
+#np.set_printoptions(threshold=np.nan)
+from random import sample,choice
+from itertools import product, permutations, combinations_with_replacement
+from player import LiteralPlayer,GriceanPlayer
+import sys 
+import datetime
+import csv
+
+
+alpha = 1 # rate to control difference between semantic and pragmatic violations
+cost = 0.9 # cost for LOT-concept with upper bound
+lam = 30 # soft-max parameter
+k = 50  # number of learning observations
+sample_amount = 400 #amount of samples per type production
+
+gens = 50 #number of generations per simulation run
+runs = 100 #number of independent simulation runs
+
+lexical_pairs = 3 #number of 2x2 matrices
+states = 2 #number of states
+messages = 2 #number of messages
+
+learning_parameter = 5 #prob-matching = 1, increments approach MAP
+
+state_freq = np.ones(states*lexical_pairs) / float(states*lexical_pairs) #frequency of states. 
+w_communal,w_parental = 0.5,0.5 #w_communal is the weight for communal learning, w_parental that of parental learning, w_communcal + w_parental = 1
+
+
+
+
+f_unwgh_mean = csv.writer(open('./results/4multiscalar-unwgh-mean-a%d-c%f-l%d-k%d-g%d-r%d-samp%d.csv' %(alpha,cost,lam,k,gens,runs,learning_parameter),'wb')) #file to store each unweighted simulation run after n generations
[…]

### 2016-08-23 old-variants/4-parental-separateobs-rmd-multiscalar.py

@@ -1,0 +1,284 @@
+#####
+#RMD for unweighted and weighted (community) learning with Bayesian (iterated) learning as M. 
+#3 pairs of scalar items, six lexica per scalar pair. 
+#2 possible signaling behaviors: literal or gricean
+#12 types (6 lexica * 2 signaling behaviors) per signaling pair
+#6*6*6 = 216  literal types, 216 gricean types, 432 types in total
+#Set of observations of length k:
+##Sample <s,m> tuples from a type's production probabilities, P(a = <s,m> | t_i) = P(s) * P(m|s,t_i)
+#
+#####
+
+
+import numpy as np
+#np.set_printoptions(threshold=np.nan)
+from random import sample,choice
+from itertools import product, permutations, combinations_with_replacement
+from player import LiteralPlayer,GriceanPlayer
+import sys 
+import datetime
+import csv
+
+
+alpha = 1 # rate to control difference between semantic and pragmatic violations
+cost = 0.4 # cost for LOT-concept with upper bound
+lam = 30 # soft-max parameter
+k = 10  # observation sequence length 
+sample_amount = 100 #amount of samples per type production
+
+gens = 50 #number of generations per simulation run
+runs = 100 #number of independent simulation runs
+
+lexical_pairs = 3 #number of 2x2 matrices
+states = 2 #number of states
+messages = 2 #number of messages
+
+learning_parameter = 5 #prob-matching = 1, increments approach MAP
+
+state_freq = np.ones(states*lexical_pairs) / float(states*lexical_pairs) #frequency of states. 
+w_communal,w_parental = 0.5,0.5 #w_communal is the weight for communal learning, w_parental that of parental learning, w_communcal + w_parental = 1
+
+
+
+
+f_unwgh_mean = csv.writer(open('./results/4multiscalar-unwgh-mean-a%d-c%f-l%d-k%d-g%d-r%d-samp%d.csv' %(alpha,cost,lam,k,gens,runs,learning_parameter),'wb')) #file to store each unweighted simulation run after n generations
[…]

### 2016-08-23 old-variants/4rmd-multiscalar.py

@@ -1,0 +1,281 @@
+#####
+#RMD for unweighted and weighted (community) learning with Bayesian (iterated) learning as M. 
+#3 pairs of scalar items, six lexica per scalar pair. 
+#2 possible signaling behaviors: literal or gricean
+#12 types (6 lexica * 2 signaling behaviors) per signaling pair
+#6*6*6 = 216  literal types, 216 gricean types, 432 types in total
+#Set of observations of length k:
+##Sample <s,m> tuples from a type's production probabilities, P(a = <s,m> | t_i) = P(s) * P(m|s,t_i)
+#
+#####
+
+
+import numpy as np
+#np.set_printoptions(threshold=np.nan)
+from random import sample,choice
+from itertools import product, permutations, combinations_with_replacement
+from player import LiteralPlayer,GriceanPlayer
+import sys 
+import datetime
+import csv
+
+
+alpha = 1 # rate to control difference between semantic and pragmatic violations
+cost = 0.7 # cost for LOT-concept with upper bound
+lam = 30 # soft-max parameter
+k = 5  # number of learning observations
+sample_amount = 50 #amount of samples per type production
+
+gens = 30 #number of generations per simulation run
+runs = 10 #number of independent simulation runs
+
+lexical_pairs = 3 #number of 2x2 matrices
+states = 2 #number of states
+messages = 2 #number of messages
+
+learning_parameter = 1 #prob-matching = 1, increments approach MAP
+
+state_freq = np.ones(states*lexical_pairs) / float(states*lexical_pairs) #frequency of states. 
+w_communal,w_parental = 0.5,0.5 #w_communal is the weight for communal learning, w_parental that of parental learning, w_communcal + w_parental = 1
+
+
+
+
+f_unwgh_mean = csv.writer(open('./results/4multiscalar-unwgh-mean-a%d-c%f-l%d-k%d-g%d-r%d-samp%d.csv' %(alpha,cost,lam,k,gens,runs,learning_parameter),'wb')) #file to store each unweighted simulation run after n generations
[…]

### 2016-08-23 old-variants/rmd-singlescalar.py

@@ -1,0 +1,196 @@
+#####
+#RMD for unweighted and weighted (community) learning with Bayesian (iterated) learning as M. 
+#1 pairs of scalar items, six lexica per scalar pair. 
+#2 possible signaling behaviors: literal or gricean
+#12 types (6 lexica * 2 signaling behaviors) per signaling pair
+#6 literal types, 6 gricean types, 12 types in total
+#####
+
+
+import numpy as np
+#np.set_printoptions(threshold=np.nan)
+from random import sample
+from itertools import product, permutations, combinations_with_replacement
+from player import LiteralPlayer,GriceanPlayer
+import sys 
+import datetime
+import csv
+
+
+l1,l2,l3,l4,l5,l6 = np.array( [[0.,0.],[1.,1.]] ), np.array( [[1.,1.],[0.,0.]] ), np.array( [[1.,1.],[1.,1.]] ), np.array( [[0.,1.],[1.,0.]] ), np.array( [[0.,1.],[1.,1.]] ), np.array( [[1.,1.],[1.,0.]] )
+
+alpha = 1# rate to control difference between semantic and pragmatic violations
+cost = 0.9 # cost for LOT-concept with upper bound
+lam = 30 # soft-max parameter
+k = 20  # number of learning observations
+sample_amount = 145 #amount of samples from OBS when k > 4
+
+gens = 30 #number of generations per simulation run
+runs = 100 #number of independent simulation runs
+
+states = 2 #number of states
+messages = 2 #number of messages
+
+
+f_unwgh_mean = csv.writer(open('./results/singlescalar-unwgh-mean-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store each unweighted simulation run after n generations
+f_wgh_mean = csv.writer(open('./results/singlescalar-wgh-mean-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store each weighted simulation run after n generations
+
+f_q = csv.writer(open('./results/singlescalar-q-matrix-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store Q-matrix
+
+
+
+print '#Starting, ', datetime.datetime.now()
+
[…]

### 2016-08-23 old-variants/test_ignore/2rmd-multiscalar.py

@@ -1,0 +1,254 @@
+#####
+#RMD for unweighted and weighted (community) learning with Bayesian (iterated) learning as M. 
+#3 pairs of scalar items, six lexica per scalar pair. 
+#2 possible signaling behaviors: literal or gricean
+#12 types (6 lexica * 2 signaling behaviors) per signaling pair
+#6*6*6 = 216  literal types, 216 gricean types, 432 types in total
+#Set of observations of length k:
+##If k < 5: 6**k * 2**k observations
+##If k > 4: 20000 observations sampled from the set of all observations
+#####
+
+
+import numpy as np
+#np.set_printoptions(threshold=np.nan)
+from random import sample
+from itertools import product, permutations, combinations_with_replacement
+from player import LiteralPlayer,GriceanPlayer
+import sys 
+import datetime
+import csv
+
+
+alpha = 1 # rate to control difference between semantic and pragmatic violations
+cost = 0.9 # cost for LOT-concept with upper bound
+lam = 30 # soft-max parameter
+k = 7  # number of learning observations
+sample_amount = 20000 #amount of samples from OBS when k > 4
+
+gens = 30 #number of generations per simulation run
+runs = 1000 #number of independent simulation runs
+
+lexical_pairs = 3 #number of 2x2 matrices
+states = 2 #number of states
+messages = 2 #number of messages
+
+
+f_unwgh_mean = csv.writer(open('./results/multiscalar-unwgh-mean-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store each unweighted simulation run after n generations
+f_wgh_mean = csv.writer(open('./results/multiscalar-wgh-mean-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store each weighted simulation run after n generations
+
+f_q = csv.writer(open('./results/multiscalar-q-matrix-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store Q-matrix
+
+
+print '#Starting, ', datetime.datetime.now()
[…]

### 2016-08-23 old-variants/test_ignore/rmd-multi-MAP.py

@@ -1,0 +1,261 @@
+#####
+#RMD for unweighted and weighted (community) learning with Bayesian (iterated) learning as M. 
+#3 pairs of scalar items, six lexica per scalar pair. 
+#2 possible signaling behaviors: literal or gricean
+#12 types (6 lexica * 2 signaling behaviors) per signaling pair
+#6*6*6 = 216  literal types, 216 gricean types, 432 types in total
+#Set of observations is computed for one scalar pair first, taking the cartesian product of the likelihood matrix three times yields the full likelihood matrix
+#####
+
+
+import numpy as np
+#np.set_printoptions(threshold=np.nan)
+from random import sample
+from itertools import product, permutations, combinations_with_replacement
+from player import LiteralPlayer,GriceanPlayer
+import sys 
+import datetime
+import csv
+
+
+alpha = 1 # rate to control difference between semantic and pragmatic violations
+cost = 0.4 # cost for LOT-concept with upper bound
+lam = 30 # soft-max parameter
+k = 3 # number of learning observations
+sample_amount = 145 #amount of samples from OBS when k > 3
+
+gens = 30 #number of generations per simulation run
+runs = 100 #number of independent simulation runs
+
+
+lexical_pairs = 3 #number of 2x2 matrices
+states = 2 #number of states
+messages = 2 #number of messages
+
+
+f_unwgh_mean = csv.writer(open('./results/1multiscalar-unwgh-mean-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store each unweighted simulation run after n generations
+f_wgh_mean = csv.writer(open('./results/1multiscalar-wgh-mean-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store each weighted simulation run after n generations
+
+f_q = csv.writer(open('./results/1multiscalar-q-matrix-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store Q-matrix
+
+
+print '#Starting, ', datetime.datetime.now()
[…]

### 2016-08-23 old-variants/test_ignore/rmd-multi-replication-last.py

@@ -1,0 +1,281 @@
+#####
+#RMD for unweighted and weighted (community) learning with Bayesian (iterated) learning as M. 
+#3 pairs of scalar items, six lexica per scalar pair. 
+#2 possible signaling behaviors: literal or gricean
+#12 types (6 lexica * 2 signaling behaviors) per signaling pair
+#6*6*6 = 216  literal types, 216 gricean types, 432 types in total
+#Set of observations is computed for one scalar pair first, taking the cartesian product of the likelihood matrix three times yields the full likelihood matrix
+#####
+
+
+import numpy as np
+#np.set_printoptions(threshold=np.nan)
+from random import sample
+from itertools import product, permutations, combinations_with_replacement
+from player import LiteralPlayer,GriceanPlayer
+import sys 
+import datetime
+import csv
+
+
+alpha = 1 # rate to control difference between semantic and pragmatic violations
+cost = 0.4 # cost for LOT-concept with upper bound
+lam = 30 # soft-max parameter
+k = 3  # number of learning observations
+sample_amount = 145 #amount of samples from OBS when k > 3
+
+gens = 30 #number of generations per simulation run
+runs = 100 #number of independent simulation runs
+
+
+lexical_pairs = 3 #number of 2x2 matrices
+states = 2 #number of states
+messages = 2 #number of messages
+
+
+f_unwgh_mean = csv.writer(open('./results/1multiscalar-unwgh-mean-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store each unweighted simulation run after n generations
+f_wgh_mean = csv.writer(open('./results/1multiscalar-wgh-mean-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store each weighted simulation run after n generations
+
+f_q = csv.writer(open('./results/1multiscalar-q-matrix-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store Q-matrix
+
+
+print '#Starting, ', datetime.datetime.now()
[…]

### 2016-08-23 old-variants/test_ignore/rmd-singlescalar-MAP.py

@@ -1,0 +1,205 @@
+#####
+#RMD for unweighted and weighted (community) learning with Bayesian (iterated) learning as M. 
+#1 pairs of scalar items, six lexica per scalar pair. 
+#2 possible signaling behaviors: literal or gricean
+#12 types (6 lexica * 2 signaling behaviors) per signaling pair
+#6 literal types, 6 gricean types, 12 types in total
+#####
+
+
+import numpy as np
+#np.set_printoptions(threshold=np.nan)
+from random import sample
+from itertools import product, permutations, combinations_with_replacement
+from player import LiteralPlayer,GriceanPlayer
+import sys 
+import datetime
+import csv
+
+
+l1,l2,l3,l4,l5,l6 = np.array( [[0.,0.],[1.,1.]] ), np.array( [[1.,1.],[0.,0.]] ), np.array( [[1.,1.],[1.,1.]] ), np.array( [[0.,1.],[1.,0.]] ), np.array( [[0.,1.],[1.,1.]] ), np.array( [[1.,1.],[1.,0.]] )
+
+alpha = 1# rate to control difference between semantic and pragmatic violations
+cost = 0.5 # cost for LOT-concept with upper bound
+lam = 30 # soft-max parameter
+k = 3  # number of learning observations
+sample_amount = 145 #amount of samples from OBS when k > 4
+
+gens = 30 #number of generations per simulation run
+runs = 100 #number of independent simulation runs
+
+states = 2 #number of states
+messages = 2 #number of messages
+
+
+f_unwgh_mean = csv.writer(open('./results/singlescalar-unwgh-mean-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store each unweighted simulation run after n generations
+f_wgh_mean = csv.writer(open('./results/singlescalar-wgh-mean-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store each weighted simulation run after n generations
+
+f_q = csv.writer(open('./results/singlescalar-q-matrix-a%d-c%f-l%d-k%d-g%d-r%d.csv' %(alpha,cost,lam,k,gens,runs),'wb')) #file to store Q-matrix
+
+
+
+print '#Starting, ', datetime.datetime.now()
+
[…]

### 2016-08-23 pd-csv.py

@@ -1,0 +1,220 @@
+###
+import pandas as pd
+import numpy as np 
+import matplotlib.pyplot as plt
+import seaborn as sns
+#sns.set_context(rc={'lines.markeredgewidth': 0.5})
+import os
+import glob
+import sys
+###
+
+###Column headers:###
+#run_ID,t1_initial,t2_initial,t3_initial,t4_initial,t5_initial,t6_initial,t7_initial,t8_initial,t9_initial,t10_initial,t11_initial,t12_initial,
+#alpha,prior_cost_c,lambda,k,sample_amount,learning_parameter,generations,
+#t1_final,t2_final,t3_final,t4_final,t5_final,t6_final,t7_final,t8_final,t9_final,t10_final,t11_final,t12_final
+#####################
+
+
+print 'Loading data'
+path =r'./resultssubset' 
+all_files = glob.glob(os.path.join(path, "*.csv"))     # advisable to use os.path.join as this makes concatenation OS independent
+df = pd.concat(pd.read_csv(f) for f in all_files)
+
+
+sys.exit()
+###Plot 1###
+a = 1
+c = 0.2
+k = 5
+lam = 10
+sample = 15
+learn = 1
+
+final_group = df.loc[:,'alpha':]
+final_group = final_group.loc[final_group['alpha'] == a]
+final_group = final_group.loc[final_group['prior_cost_c'] == c]
+final_group = final_group.loc[final_group['k'] == k]
+final_group = final_group.loc[final_group['lambda'] == lam]
+final_group = final_group.loc[final_group['sample_amount'] == sample]
+#final_group = final_group.loc[final_group['learning_parameter'] == learn]
+
+
+all_finals = final_group.groupby(['learning_parameter'])
+sub_finals = all_finals[['t10_final','t11_final','t12_final']].agg(np.average)
+
+print sub_finals
+
+ax = sub_finals.plot(title='alpha = %d, $\lambda = %d$, c = %.2f,k = %d, samples = %d' %(a,lam,c,k,sample))
+ax.set(ylabel="proportion",xlabel='learning parameter')
+plt.legend(["$pragmatic_{10}$","$pragmatic_{11}$","$pragmatic_{12}$"], loc='best')
+
+plt.show()
+
+
+
+###Plot 2###
+a = 1
+c = 0.5
+k = 5
+lam = 10
+sample = 15
+learn = 1
+
+final_group = df.loc[:,'alpha':]
[…]

### 2016-08-23 presentations/edindata/modsinglescalar.py

@@ -1,0 +1,185 @@
+#####
+#RMD with parametrized iterated parental learning 
+#1 pairs of scalar items, six lexica per scalar pair. 
+#2 possible signaling behaviors: literal or gricean
+#12 types (6 lexica * 2 signaling behaviors) per signaling pair
+#6 literal types, 6 gricean types, 12 types in total
+#####
+
+import numpy as np
+np.set_printoptions(threshold=np.nan)
+from random import sample
+from itertools import product
+from player import LiteralPlayer,GriceanPlayer
+import sys 
+import datetime
+import csv
+
+def normalize(m):
+    return m / m.sum(axis=1)[:, np.newaxis]
+
+def summarize_counts(lst,states,messages):
+    """summarize counts for tuples of k-states and k-messages""" 
+    counter = [0 for _ in xrange(states**messages)]
+    for i in xrange(len(lst)):
+        s,m = lst[i][0] *2, lst[i][1]
+        counter[s+m] += 1
+    return counter
+
+def get_obs(k,states,messages,likelihoods,state_freq,sample_amount):
+    """Returns summarized counts of k-length <s_i,m_j> production observations as [#(<s_0,m_0>), #(<s_0,m_1), #(<s_1,m_0>, #(s_1,m_1)], ...]] = k"""
+    s = list(xrange(states))
+    m = list(xrange(messages))
+    atomic_observations = list(product(s,m))
+   
+    obs = [] #store all produced k-length (s,m) sequences 
+    for t in xrange(12):
+        produced_obs = [] #store k-length (s,m) sequences of a type
+        production_vector = likelihoods[t].flatten()
+        doubled_state_freq = np.column_stack((state_freq,state_freq)).flatten() # P(s)
+        sample_vector = production_vector * doubled_state_freq #P(s) * P(m|s,t_i)
+        for i in xrange(sample_amount):
+            sample_t = [np.random.choice(len(atomic_observations),p=sample_vector) for _ in xrange(k)]
+            sampled_obs = [atomic_observations[i] for i in sample_t]
+            produced_obs.append(summarize_counts(sampled_obs,states,messages))
+        obs.append(produced_obs)
+    return obs
+
+
+def get_likelihood(obs,likelihoods):
[…]

### 2016-08-23 presentations/edindata/modsinglescalaronlyfitness.py

@@ -1,0 +1,183 @@
+#####
+#RMD with parametrized iterated parental learning 
+#1 pairs of scalar items, six lexica per scalar pair. 
+#2 possible signaling behaviors: literal or gricean
+#12 types (6 lexica * 2 signaling behaviors) per signaling pair
+#6 literal types, 6 gricean types, 12 types in total
+#####
+
+import numpy as np
+np.set_printoptions(threshold=np.nan)
+from random import sample
+from itertools import product
+from player import LiteralPlayer,GriceanPlayer
+import sys 
+import datetime
+import csv
+
+def normalize(m):
+    return m / m.sum(axis=1)[:, np.newaxis]
+
+def summarize_counts(lst,states,messages):
+    """summarize counts for tuples of k-states and k-messages""" 
+    counter = [0 for _ in xrange(states**messages)]
+    for i in xrange(len(lst)):
+        s,m = lst[i][0] *2, lst[i][1]
+        counter[s+m] += 1
+    return counter
+
+def get_obs(k,states,messages,likelihoods,state_freq,sample_amount):
+    """Returns summarized counts of k-length <s_i,m_j> production observations as [#(<s_0,m_0>), #(<s_0,m_1), #(<s_1,m_0>, #(s_1,m_1)], ...]] = k"""
+    s = list(xrange(states))
+    m = list(xrange(messages))
+    atomic_observations = list(product(s,m))
+   
+    obs = [] #store all produced k-length (s,m) sequences 
+    for t in xrange(12):
+        produced_obs = [] #store k-length (s,m) sequences of a type
+        production_vector = likelihoods[t].flatten()
+        doubled_state_freq = np.column_stack((state_freq,state_freq)).flatten() # P(s)
+        sample_vector = production_vector * doubled_state_freq #P(s) * P(m|s,t_i)
+        for i in xrange(sample_amount):
+            sample_t = [np.random.choice(len(atomic_observations),p=sample_vector) for _ in xrange(k)]
+            sampled_obs = [atomic_observations[i] for i in sample_t]
+            produced_obs.append(summarize_counts(sampled_obs,states,messages))
+        obs.append(produced_obs)
+    return obs
+
+
+def get_likelihood(obs,likelihoods):
[…]

### 2016-08-23 presentations/edindata/modsinglescalaronlylearning.py

@@ -1,0 +1,183 @@
+#####
+#RMD with parametrized iterated parental learning 
+#1 pairs of scalar items, six lexica per scalar pair. 
+#2 possible signaling behaviors: literal or gricean
+#12 types (6 lexica * 2 signaling behaviors) per signaling pair
+#6 literal types, 6 gricean types, 12 types in total
+#####
+
+import numpy as np
+np.set_printoptions(threshold=np.nan)
+from random import sample
+from itertools import product
+from player import LiteralPlayer,GriceanPlayer
+import sys 
+import datetime
+import csv
+
+def normalize(m):
+    return m / m.sum(axis=1)[:, np.newaxis]
+
+def summarize_counts(lst,states,messages):
+    """summarize counts for tuples of k-states and k-messages""" 
+    counter = [0 for _ in xrange(states**messages)]
+    for i in xrange(len(lst)):
+        s,m = lst[i][0] *2, lst[i][1]
+        counter[s+m] += 1
+    return counter
+
+def get_obs(k,states,messages,likelihoods,state_freq,sample_amount):
+    """Returns summarized counts of k-length <s_i,m_j> production observations as [#(<s_0,m_0>), #(<s_0,m_1), #(<s_1,m_0>, #(s_1,m_1)], ...]] = k"""
+    s = list(xrange(states))
+    m = list(xrange(messages))
+    atomic_observations = list(product(s,m))
+   
+    obs = [] #store all produced k-length (s,m) sequences 
+    for t in xrange(12):
+        produced_obs = [] #store k-length (s,m) sequences of a type
+        production_vector = likelihoods[t].flatten()
+        doubled_state_freq = np.column_stack((state_freq,state_freq)).flatten() # P(s)
+        sample_vector = production_vector * doubled_state_freq #P(s) * P(m|s,t_i)
+        for i in xrange(sample_amount):
+            sample_t = [np.random.choice(len(atomic_observations),p=sample_vector) for _ in xrange(k)]
+            sampled_obs = [atomic_observations[i] for i in sample_t]
+            produced_obs.append(summarize_counts(sampled_obs,states,messages))
+        obs.append(produced_obs)
+    return obs
+
+
+def get_likelihood(obs,likelihoods):
[…]

### 2016-08-23 presentations/edindata/player.py

@@ -1,0 +1,52 @@
+import numpy as np
+from math import exp
+
+def normalize(m):
+    m = m / m.sum(axis=1)[:, np.newaxis]
+    m[np.isnan(m)] = 0.
+    return m
+
+class LiteralPlayer:
+    def __init__(self,alpha,lexicon):
+        self.alpha = alpha
+        self.lexicon = lexicon
+        self.sender_matrix = self.sender_selection_matrix(lexicon)
+        self.receiver_matrix =  self.receiver_selection_matrix()
+   
+    def sender_selection_matrix(self,l):
+        #Compute e^([utility(column-cell given row-cell]**alpha)
+        m = np.zeros(np.shape(l))
+        for i in range(np.shape(l)[0]):
+            for j in range(np.shape(l)[1]):
+                m[i,j] = exp(l[i,j]**self.alpha)
+        return normalize(m)
+
+    def receiver_selection_matrix(self):
+        """Take transposed lexicon and normalize row-wise (prior over states plays no role as it's currently uniform)"""
+        m = normalize(np.transpose(self.lexicon))
+        for r in range(np.shape(m)[0]):
+            if sum(m[r]) == 0:
+                for c in range(np.shape(m)[1]):
+                    m[r,c] = 1. / np.shape(m)[0]
+
+        return m
+
+
+
+#######################################
+class GriceanPlayer:
+    def __init__(self,alpha, lam, lexicon):
+        self.alpha = alpha
+        self.lam = lam
+        self.lexicon = lexicon
+        self.sender_matrix = self.sender_selection_matrix(lexicon)
+        self.receiver_matrix = self.receiver_selection_matrix()
+
+    def sender_selection_matrix(self,l):
+        literalListener = normalize(np.transpose(l))
+        utils = np.transpose(literalListener)
+        return normalize(np.exp(self.lam*np.power(utils, self.alpha)))
+
+    def receiver_selection_matrix(self):
+        """Take transposed sender matrix and normalize row-wise (prior over states plays no role as it's currently uniform)"""
+        return normalize(np.transpose(self.sender_matrix))

### 2016-08-23 presentations/edindata/plot-fitness.py

@@ -1,0 +1,103 @@
+###
+import pandas as pd
+import numpy as np 
+import matplotlib.pyplot as plt
+import seaborn as sns
+#sns.set_context(rc={'lines.markeredgewidth': 0.5})
+import os
+import glob
+import sys
+###
+###Column headers:###
+#run_ID,t1_initial,t2_initial,t3_initial,t4_initial,t5_initial,t6_initial,t7_initial,t8_initial,t9_initial,t10_initial,t11_initial,t12_initial,
+#alpha,prior_cost_c,lambda,k,sample_amount,learning_parameter,generations,
+#t1_final,t2_final,t3_final,t4_final,t5_final,t6_final,t7_final,t8_final,t9_final,t10_final,t11_final,t12_final
+#####################
+
+
+print 'Loading data fitness'
+path =r'./results-fit' 
+all_files = glob.glob(os.path.join(path, "*.csv"))     # advisable to use os.path.join as this makes concatenation OS independent
+df = pd.concat(pd.read_csv(f) for f in all_files)
+
+print 'Loading data learn'
+path =r'./results-learn' 
+all_files = glob.glob(os.path.join(path, "*.csv"))     # advisable to use os.path.join as this makes concatenation OS independent
+dl = pd.concat(pd.read_csv(f) for f in all_files)
+
+def cust_mean(grp):
+    grp['mean'] = grp['t11_final'].mean()
+    return grp
+
+
+
+
+### development over cost ###
+
+a = 1
+k = 5
+lam = 30
+sample = 10
+learn = 1
+
+
+final_groupA = df.loc[:,'alpha':]
+final_groupB = dl.loc[:,'alpha':]
+
+groupA = final_groupA.loc[final_groupA['alpha'] == a]
+groupA = groupA.loc[groupA['k'] == k]
+groupA = groupA.loc[groupA['sample_amount'] == sample]
+groupA = groupA.loc[groupA['learning_parameter'] == learn]
+groupA = groupA.loc[groupA['prior_cost_c'] == 0.1]
+
+t_finalA = groupA.groupby(['lambda'])
[…]

### 2016-08-23 presentations/edindata/plots.py

@@ -1,0 +1,108 @@
+###
+import pandas as pd
+import numpy as np 
+import matplotlib.pyplot as plt
+import seaborn as sns
+#sns.set_context(rc={'lines.markeredgewidth': 0.5})
+import os
+import glob
+import sys
+###
+###Column headers:###
+#run_ID,t1_initial,t2_initial,t3_initial,t4_initial,t5_initial,t6_initial,t7_initial,t8_initial,t9_initial,t10_initial,t11_initial,t12_initial,
+#alpha,prior_cost_c,lambda,k,sample_amount,learning_parameter,generations,
+#t1_final,t2_final,t3_final,t4_final,t5_final,t6_final,t7_final,t8_final,t9_final,t10_final,t11_final,t12_final
+#####################
+
+
+print 'Loading data'
+path =r'./results1' 
+all_files = glob.glob(os.path.join(path, "*.csv"))     # advisable to use os.path.join as this makes concatenation OS independent
+df = pd.concat(pd.read_csv(f) for f in all_files)
+
+def cust_mean(grp):
+    grp['mean'] = grp['t11_final'].mean()
+    return grp
+
+
+### Heatplot prior/posterior ####
+a = 1
+k = 5
+lam = 30
+sample = 10
+
+final_group = df.loc[:,('alpha','prior_cost_c','lambda','k','sample_amount','learning_parameter','t11_final')]
+
+group = final_group.loc[final_group['alpha'] == a]
+group = group.loc[group['k'] == k]
+group = group.loc[group['lambda'] == lam]
+group = group.loc[group['sample_amount'] == sample]
+
+
+t11_final = group.groupby(['prior_cost_c', 'learning_parameter']).apply(cust_mean) #add new column with mean
+t11_rel = t11_final.loc[:,('prior_cost_c', 'learning_parameter','mean')] #ignore other columns, given that they are fixed
+t11_rel_uniq = t11_rel.drop_duplicates() #drop duplictes that arise from adding the mean column
+t11_rec = t11_rel_uniq.pivot('prior_cost_c','learning_parameter','mean') #reshape to have a prior_cost_c by learning_parameter table
+
+sns.set(font_scale=1.2)
+
+#yticks = np.arange(min(t11_rec.index),max(t11_rec.index),0.15)
+vacio = ["" for _ in xrange(9)]
[…]

### 2016-08-23 presentations/edindata/plots2.py

@@ -1,0 +1,173 @@
+###
+import pandas as pd
+import numpy as np 
+import matplotlib.pyplot as plt
+import seaborn as sns
+#sns.set_context(rc={'lines.markeredgewidth': 0.5})
+import os
+import glob
+import sys
+###
+###Column headers:###
+#run_ID,t1_initial,t2_initial,t3_initial,t4_initial,t5_initial,t6_initial,t7_initial,t8_initial,t9_initial,t10_initial,t11_initial,t12_initial,
+#alpha,prior_cost_c,lambda,k,sample_amount,learning_parameter,generations,
+#t1_final,t2_final,t3_final,t4_final,t5_final,t6_final,t7_final,t8_final,t9_final,t10_final,t11_final,t12_final
+#####################
+
+
+print 'Loading data'
+path =r'./results2' 
+all_files = glob.glob(os.path.join(path, "*.csv"))     # advisable to use os.path.join as this makes concatenation OS independent
+df = pd.concat(pd.read_csv(f) for f in all_files)
+
+def cust_mean(grp):
+    grp['mean'] = grp['t11_final'].mean()
+    return grp
+
+### Heatplot seq-length/lam ####
+a = 1
+sample = 10
+c = .4
+sample_amount = 20
+learn = 3
+gens = 20
+runs = 1000
+
+
+final_group = df.loc[:,('alpha','prior_cost_c','lambda','k','sample_amount','learning_parameter','t11_final')]
+
+group = final_group.loc[final_group['alpha'] == a]
+group = group.loc[group['prior_cost_c'] == c]
+group = group.loc[group['sample_amount'] == sample]
+group = group.loc[group['learning_parameter'] == learn]
+
+t11_final = group.groupby(['lambda','k']).apply(cust_mean) #add new column with mean
+t11_rel = t11_final.loc[:,('lambda','k','mean')] #ignore other columns, given that they are fixed
+t11_rel_uniq = t11_rel.drop_duplicates() #drop duplictes that arise from adding the mean column
+t11_rec = t11_rel_uniq.pivot('lambda','k','mean') #reshape to have a k by lambda table
+
+
+sns.set(font_scale=1.2)
+vacio = ["" for _ in xrange(4)]
[…]

### 2016-08-23 presentations/edindata/plots3.py

@@ -1,0 +1,105 @@
+###
+import pandas as pd
+import numpy as np 
+import matplotlib.pyplot as plt
+import seaborn as sns
+#sns.set_context(rc={'lines.markeredgewidth': 0.5})
+import os
+import glob
+import sys
+###
+###Column headers:###
+#run_ID,t1_initial,t2_initial,t3_initial,t4_initial,t5_initial,t6_initial,t7_initial,t8_initial,t9_initial,t10_initial,t11_initial,t12_initial,
+#alpha,prior_cost_c,lambda,k,sample_amount,learning_parameter,generations,
+#t1_final,t2_final,t3_final,t4_final,t5_final,t6_final,t7_final,t8_final,t9_final,t10_final,t11_final,t12_final
+#####################
+
+
+print 'Loading data fitness'
+path =r'./results-fit' 
+all_files = glob.glob(os.path.join(path, "*.csv"))     # advisable to use os.path.join as this makes concatenation OS independent
+df = pd.concat(pd.read_csv(f) for f in all_files)
+
+print 'Loading data learn'
+path =r'./results-learn' 
+all_files = glob.glob(os.path.join(path, "*.csv"))     # advisable to use os.path.join as this makes concatenation OS independent
+dl = pd.concat(pd.read_csv(f) for f in all_files)
+
+def cust_mean(grp):
+    grp['mean'] = grp['t11_final'].mean()
+    return grp
+
+
+
+
+### development over cost ###
+
+a = 1
+k = 5
+lam = 30
+sample = 10
+learn = 1
+
+
+final_groupA = df.loc[:,'alpha':]
+final_groupB = dl.loc[:,'alpha':]
+
+groupA = final_groupA.loc[final_groupA['alpha'] == a]
+groupA = groupA.loc[groupA['k'] == k]
+groupA = groupA.loc[groupA['sample_amount'] == sample]
+groupA = groupA.loc[groupA['learning_parameter'] == learn]
+groupA = groupA.loc[groupA['prior_cost_c'] == 0.1]
+
+t_finalA = groupA.groupby(['lambda'])
[…]

### 2016-08-23 presentations/edindata/run-small.py

@@ -1,0 +1,67 @@
+import modsinglescalar as rmd
+import modsinglescalaronlyfitness as fit
+import modsinglescalaronlylearning as learner
+#parameter order for run(*) is run(alpha,cost,lam,k,sample_amount, learning_parameter,gens,runs)
+
+
+def cost_to_learning():
+    a = 1
+    lam = 30
+    cost = [x/100. for x in xrange(100)]
+    seq_length = 5
+    sample_amount = 10
+    learning_parameter = [x for x in xrange(1,11)]
+    gens = 20
+    runs = 1000
+    
+    for c in cost:
+        for learn in learning_parameter:
+            rmd.run(a,c,lam,seq_length,sample_amount,learn,gens,runs)
+
+def ration_to_seq():
+    a = 1
+    lam = [x for x in xrange(1,50)]
+    c = .4
+    seq_length = [x for x in xrange(1,20)]
+    sample_amount = 20
+    learn = 3
+    gens = 20
+    runs = 1000
+    
+    for l in lam:
+        for k in seq_length:
+            rmd.run(a,c,l,k,sample_amount,learn,gens,runs)
+
+def alpha_to_seq():
+    a = [x/100. for x in xrange(1,101)]
+    lam = 20
+    c = .4
+    seq_length = [x for x in xrange(1,20)]
+    sample_amount = 20
+    learn = 3
+    gens = 20
+    runs = 1000
+    
+    for al in a:
+        for k in seq_length:
+            rmd.run(al,c,l,k,sample_amount,learn,gens,runs)
+
+def only_r_and_m():
+    a = 1
+    lam = 30
+    cost = [x/100. for x in xrange(100)]
+    learn = 1
+    seq_length = 5
+    sample_amount = 10
+    gens = 20
+    runs = 1000
+    
+    for c in cost:
+        print 'learn-only'
+        learner.run(a,c,lam,seq_length,sample_amount,learn,gens,runs)
+        print 'fit-only'
+        fit.run(a,c,lam,seq_length,sample_amount,learn,gens,runs)
+
+
+
+only_r_and_m()

### 2016-08-23 run-analysis.py

@@ -1,0 +1,18 @@
+import modsinglescalar as rmd
+#parameter order for run(*) is run(alpha,cost,lam,k,sample_amount, learning_parameter,gens,runs)
+
+alpha = [0.25,0.5,0.7,1]
+cost = [0.01,0.05,0.1,0.2,0.3,0.4,0.5,0.7,0.8,0.9]
+lam = [1,10,30,50]
+seq_length = [1,3,5,9,13,15,17,20]
+sample_amount = [1,3,5,7,9,12,15] + [20,40,60,80,100] #+ [200,300,400,500]
+learning_parameter = [1,2,4,6,8,10]
+gens = 20
+runs = 1000
+
+for a in alpha:
+    for c in cost:
+        for l in lam:
+            for k in seq_length:
+                for learn in learning_parameter:
+                    rmd.run(a,c,l,k,learn,gens,runs)

### 2016-08-23 run-small.py

@@ -1,0 +1,15 @@
+import modsinglescalar as rmd
+#parameter order for run(*) is run(alpha,cost,lam,k, learning_parameter,gens,runs)
+
+alpha = 1
+lam = 30
+cost = [x/100. for x in xrange(1,101)]
+seq_length = [1,3,5,9,13,15,17,20]
+learning_parameter = [1,2,4,6,8,10]
+gens = 20
+runs = 10000
+
+for c in cost:
+    for k in seq_length:
+        for learn in learning_parameter:
+                    rmd.run(alpha,c,lam,k,learn,gens,runs)

### 2016-08-24 paper/paper.tex
more text on communicative pressures in working paper

@@
 \usepackage{natbib}
-\bibpunct[,]{(}{)}{,}{a}{}{;}
+\bibpunct[:]{(}{)}{,}{a}{}{;}
 
 %--------------------
-\usepackage{gb4e}
-\noautomath
+%\usepackage{gb4e}
+%\noautomath
 
@@
 
-\title{Communicative pressures at the semantics-pragmatics interface}
+\title{Communicative pressures at the semantics-pragmatics interface:\\ Learning biases may prevent the lexicalization of pragmatic inferences}
 
@@
 
-\begin{abstract} Natural languages allow for the conventional pragmatic enrichment of semantic structure in a notably productive fashion.
-As a result, the information conveyed often goes beyond the literal meaning of an expression.
-This raises a challenge to explain the regular selection of particular semantics over others in light of pragmatics.
+\begin{abstract} Natural languages allow for conventional pragmatic enrichments in a notably productive fashion.
+As a consequence, the information conveyed often goes beyond the literal meaning of expressions.
+In light of pragmatics, this raises the challenge to explain the regular selection of particular literal meanings over other putative alternatives.
 To address this challenge, we propose a general model for the analysis of linguistic pressures that integrates (iterated) Bayesian learning in the replicator-mutator dynamics commonly used in evolutionary game theory.
 This model allows for population-level analyses of rational probabilistic language users with varied degrees of pragmatic sophistication and distinct languages.
-In an application to scalar implicatures, we show that semantics that allow for pragmatic enrichment are selected for when languages are pressured towards learnability.
-Crucially, this result hinges on pragmatics to compensate for the disadvantage in expressivity that users of such languages incur otherwise.
+In a case-study on the (failure of the) lexicalization of scalar implicatures, we show how simpler semantic representations are selected for when languages are pressured towards learnability provided that pragmatic reasoning can compensate for the disadvantage in expressivity that users of such languages would otherwise incur.
 \end{abstract}
@@
 Broadly speaking, the former concerns the truth-conditional content of expressions, whereas the latter concerns information beyond such literal meanings.
-For instance, the integration of context and the derivation of defeasible inferences.
-Thus, the information conveyed by an utterance is seldom, if ever, solely determined by semantics, but rather in tandem with pragmatics.
+For instance, defeasible inferences licensed by interlocutors' mutual reasoning about rational language use.
+Thus, under this view, the information conveyed by an utterance is seldom, if ever, solely determined by semantics, but rather in tandem with pragmatics.
 
-Much research in semantics and pragmatics is aimed at characterizing classes of expressions in terms of either domain, or their interplay.
-Thus, while there is disagreement on where their boundary lies, the distinction between semantics and pragmatics plays an important role in linguistics.
-However, an issue that has received little attention is the justification of semantic structure in light of pragmatics.
+Much research at the semantics-pragmatics interface has been aimed at characterizing expressions in terms of either domain, or their interplay.
+Therefore, while there is disagreement on where exactly their boundary lies, their distinction has played an important role in the field's development.
+Notwithstanding, an issue that has received little attention is the justification of semantic structure in light of pragmatics.
 That is, an account for the selection and pervasiveness of particular semantic structures over others under consideration of the informational enrichment provided by pragmatics.
+
+A number of recent investigations have began to address the development and selection of linguistic properties (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
+While their efforts have largely concentrated on compositional and combinatorial systems, our starting point is given by the overarching account of competing pressures that has crystallized across these approaches: Natural languages need to be well-adapted to the communicative needs within its linguistic community, but also need to be learnable to survive their faithful transmission across generations.
+More succinctly; natural languages are pressured for successful communication as well as acquisition.
+
+We proceed by modeling these pressures in the replicator-mutator dynamics commonly used in evolutionary game theory.
+This yields general and precise means to model the dynamics of linguistic pressures by combining functional pressure on successful communication, effects of learning biases on (iterated) Bayesian learning \citep{griffiths+kalish:2007}, and  probabilistic models of language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
+In this way, the model connects the recent surge of synchronic probabilistic rational language use analyses with diachronic models of cultural evolution.
+%We then analyze the prevalence of a lack of semantic upper-bounds in the literal meaning of weak scalar expressions.
+We show that a lack of upper-bounds is selected for when learners are biased towards simpler semantic representations, provided they have means to convey upper-bounds, i.e., provided that they can be derived via pragmatic reasoning.
 
@@
 
-A number of recent investigations have began to address issues pertaining to the development and selection of linguistic properties (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
-Our starting point is given by the overarching account of competing pressures that has crystallized across these approaches: Natural languages are pressured for successful communication, one the one hand, and for acquisition, on the other.
-That is, they need to be well-adapted to the communicative needs within a linguistic community, as well as be learnable to survive their faithful transmission across generations.
+\section{Simplicity and expressivity in semantics, pragmatics, and beyond}
+At least since \citeposs{zipf:1949} rationalization of the observation that word frequency rankings can be approximated by a power law distribution as competing hearer and speaker preferences, the idea that languages are shaped by trade-offs of competing pressures has played a pivotal role in synchronic and diachronic analyses (e.g.
+\citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007, piantadosi:2014,kirby+etal:2015}).
+A cornerstone principle, both in linguistics and cognitive science more generally, is that of simplicity \citep{chater+vitanyi:2003}.
 
-We proceed by instantiating these pressures in the replicator-mutator dynamics, commonly used in evolutionary game theory.
-This model allows for general and precise means to model the dynamics of linguistic pressures by combining functional pressure on successful communication, effects of learning biases on (iterated) Bayesian learning \citep{griffiths+kalish:2007}, and  probabilistic models of language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
-Drawing from the latter research strand allows us to make the distinction between semantics and pragmatics in communicative behavior precise.
-In this way, this model bridges the recent surge of synchronic probabilistic rational language use analyses with diachronic models of cultural evolution.
-We then analyze the prevalence of a lack of semantic upper-bounds in the literal meaning of weak scalar expressions.
-We show that a lack of upper-bounds is selected for when learners are biased towards simpler semantic representations, provided they have means to convey upper-bounds, i.e., provided that they can be derived via pragmatic reasoning.
-
-\section{Simplicity and expressivity at the semantics-pragmatics interface, and beyond}
-
-\begin{itemize}
-  \item \hl{General introduction to the idea of competing pressures in shaping languages (\citealt{zipf:1949}, Piantadosi,...}
 %Zipf's principle of least effort as an account of human behavior. In language he interpreted it as minimal articulation with maximal expliciteness
   %Martinet 1962:139 describes language change as an interaction of `` ‘first,   the   requirements   of  communication,  the  need  for  the  speaker  to  convey  his  message,  and  second,  the principle of least effort, which makes him restrict his output of energy, both mental and physical, to the minimum compatible with achieving his ends
-  \item \hl{Idea in pragmatics: \citealt{clark+wilkes-gibbs:1986}, \citealt{horn:1984}}
+
+More particular to natural communication, languages are pressured towards expressivity and learnability.
+Their opposition becomes clear when considering their consequences in the extreme (cf.
+\citealt{kemp+regier:2012,kirby+etal:2015}).
+On the one side, a language with a single form is extremely easy to learn but lacking in expressivity.
+On the other, a language that associates a distinct form with all possible meanings speakers may want to convey is expressive but challenging to acquire.
+However, acquiring a system to express a potentially infinite set of meanings through finite means, the so-called transmission bottleneck \citep{kirby:2002}, is not the only problem learners confront.
+Instead, the relevant challenge for our purpose is that of selecting between functionally similar, if not identical, lexical meanings.
+
+\hl{explain scalar implicatures and the challenge that arises, then go for simplicity in lexical entries being codified as a preference in the learners}
+
+%Synchronically, \citeposs{clark+wilkes-gibbs:1986} {\em principle of least collaborative effort} and \citeposs{horn:1984} Q- and R-principles, amongst others, seek to explain linguistic behavior from speaker and hearer economy.
+For instance, speakers are taken to prefer to minimize utterance length whereas hearers prefer them to be longer for better understanding.
+The balance of these forces is thereby understood to drive phenomena such as the iterative decrease in utterance length in dialog \citep{kim+etal:2011, pickering+ferreira:2008, brennan+clark:1998} and the association of frequent meanings with unmarked forms \citep{grice:1975}.
   %Principle of least collaborative effort The principle  of least  collaborative  effort:  In conversation, the  participants  try  to  minimize  their  collaborative  effort—the  work that both do from the initiation of each contribution to its mutual acceptance
-  \item \hl{Discuss Q- and R-predictions for language change (cf. \citealt{traugott:2004})}
+
+%While his rationalization is not in want of detractors (see \citealt{piantadosi:2014} for a recent overview), 
+\begin{itemize}
   \item \hl{Idea in LOT (\citealt{piantadosi+etal:underreview})}
@@
 \section{Lack of semantic upper-bounds in lexical meaning}
+
+The observation that monomorphemic expression that lexically rule out stronger alternatives (e.g.
+{not all} as hypothetical `nall') are unattested across languages has received substantial support (most prominently in \citealt[252-267]{horn:1984} but also e.g.
+in \citealt{horn:1972,traugott:2004,vdAuwera:2010}).
+To the best of our knowledge, this claim stands unchallenged.
 

### 2016-08-26 paper/paper.tex
small additions to section 2 in paper draft

@@
 A cornerstone principle, both in linguistics and cognitive science more generally, is that of simplicity \citep{chater+vitanyi:2003}.
+\hl{expand a bit here to prepare for linkage further down}
 
@@
 More particular to natural communication, languages are pressured towards expressivity and learnability.
-Their opposition becomes clear when considering their consequences in the extreme (cf.
+Their opposition becomes particularly clear when considering their consequences in the extreme (cf.
 \citealt{kemp+regier:2012,kirby+etal:2015}).
@@
 However, acquiring a system to express a potentially infinite set of meanings through finite means, the so-called transmission bottleneck \citep{kirby:2002}, is not the only problem learners confront.
-Instead, the relevant challenge for our purpose is that of selecting between functionally similar, if not identical, lexical meanings.
+Instead, for our purpose the relevant challenge is that of selecting between functionally similar, if not identical, pragmatically enriched lexical meanings.
 
-\hl{explain scalar implicatures and the challenge that arises, then go for simplicity in lexical entries being codified as a preference in the learners}
+A particularly well-studied type of conventional pragmatic enrichment are so-called {\em scalar implicatures} (cf.
+\citealt{horn:1972,gazdar:1979}).
+These inferences are licensed for groups of expressions ordered in terms of informativity, here understood an entailment induced order.\footnote{\hl{maybe discuss that there are other possible orders considering relevant alternatives}}.
+For instance, {\em some} is entailed by {\em all} -- if it were true that `All students came to class', it would also be true that `Some students came to class'.
+However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not necessarily what their use is taken to convey.
+Instead, the use of a less informative expression when a more informative one could have been used can license the inference that the more stronger alternative does not hold.
+That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that, since the speaker did not use a stronger alternative ({\em all}), this alternative must not hold.
+In this way, `Some students came to class' is strengthened as conveying `Some but not all students came to class' through a pragmatically supplied upper-bound that rules out stronger alternatives.
+Analogously, a speaker can rely on her interlocutor to draw this inference without having to express the bound overtly, e.g.
+by stating {\em some but not all}.
+
+Rephrasing our initial question in terms of scalar implicatures: Why is the lack of lexical upper-bounds in weak scalar alternatives regularly selected for over the alternative of incorporating the bound in the semantics?
+This question is particularly striking when considering the diversity of the classes of expressions that license such inferences across languages (\citealt{horn:1972},\citealt[252-267]{horn:1984},\citealt{traugott:2004,vdAuwera:2010}).
+In the following, we assume learners to be biased, a priori, towards simpler (more compressed) representations, corresponding to the argument that rational learners should prefer simpler over more complex explanations of data \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
+
+\hl{summarize assumptions and then move into the dynamics right away}
+
+%The observation that monomorphemic expression that lexically rule out stronger alternatives (e.g.
+{not all} as hypothetical `nall') are unattested across languages has received substantial support (most prominently in \citealt[252-267]{horn:1984} but also e.g.
+in \citealt{horn:1972,traugott:2004,vdAuwera:2010}).
+To the best of our knowledge, this claim stands unchallenged.
 
@@
 %While his rationalization is not in want of detractors (see \citealt{piantadosi:2014} for a recent overview),
-\begin{itemize}
-  \item \hl{Idea in LOT (\citealt{piantadosi+etal:underreview})}
-\end{itemize}
 
@@
 \section{Lack of semantic upper-bounds in lexical meaning}
-
-The observation that monomorphemic expression that lexically rule out stronger alternatives (e.g.
-{not all} as hypothetical `nall') are unattested across languages has received substantial support (most prominently in \citealt[252-267]{horn:1984} but also e.g.
-in \citealt{horn:1972,traugott:2004,vdAuwera:2010}).
-To the best of our knowledge, this claim stands unchallenged.
 

### 2016-09-14 paper/paper.tex
draft of expressiveness and mutation parts of the paper + partial adaptation of IL-comparison code to newest dynamics with MAP

--- 2016-08-23 paper/paper.tex
@@
 \usepackage{amssymb}
+\usepackage{mathrsfs}
+\usepackage{nicefrac}
 %\usepackage{stmaryrd}
@@
 \maketitle
+
+\begin{abstract} Certain semantic structures enable for pragmatic enrichments in a notably productive fashion.
+This raises the challenge to justify their regular selection over alternatives that codify semantically what is conveyed pragmatically.
+This issue is particularly puzzling under a purely functional perspective.
+To address this challenge, we propose a general model for the analysis of linguistic pressures that integrates iterated Bayesian learning in the replicator-mutator dynamics.
+This model allows for population-level analyses of the effects of such pressures on probabilistic language users with varied degrees of pragmatic sophistication and distinct languages.
+We showcase the model's predictions in a case study on the (lack of) lexicalization of scalar implicatures.
+The results suggest simpler semantic representations to be selected for when languages are pressured towards learnability, provided that pragmatic reasoning can compensate for the disadvantage in expressivity that users of such languages otherwise incur.
+
 \section{The semantics-pragmatics divide}\label{sec:introduction}
 
-\begin{abstract} Natural languages allow for the conventional pragmatic enrichment of semantic structure in a notably productive fashion.
-As a result, the information conveyed often goes beyond the literal meaning of an expression.
-This raises a challenge to explain the regular selection of particular semantics over others in light of pragmatics.
-To address this challenge, we propose a general model for the analysis of linguistic pressures that integrates (iterated) Bayesian learning in the replicator-mutator dynamics commonly used in evolutionary game theory.
-This model allows for population-level analyses of rational probabilistic language users with varied degrees of pragmatic sophistication and distinct languages.
-In an application to scalar implicatures, we show that semantics that allow for pragmatic enrichment are selected for when languages are pressured towards learnability.
-Crucially, this result hinges on pragmatics to compensate for the disadvantage in expressivity that users of such languages incur otherwise.
 \end{abstract}
@@
 In linguistic theorizing, it is common to draw a distinction between semantics and pragmatics.
-Broadly speaking, the former concerns the truth-conditional content of expressions, whereas the latter concerns information beyond such literal meanings.
-For instance, the integration of context and the derivation of defeasible inferences.
-Thus, the information conveyed by an utterance is seldom, if ever, solely determined by semantics, but rather in tandem with pragmatics.
+Broadly speaking, the former concerns the truth-conditional content of expressions, whereas the latter concerns information beyond literal meanings and their composition.
+Under this view, the information conveyed by an utterance is seldom, if ever, solely determined by semantics, but rather in tandem with pragmatics.
 
-Much research in semantics and pragmatics is aimed at characterizing classes of expressions in terms of either domain, or their interplay.
-Thus, while there is disagreement on where their boundary lies, the distinction between semantics and pragmatics plays an important role in linguistics.
+Much research at the semantics-pragmatics interface has been aimed at characterizing expressions in terms of either domain, or their interplay.
+As a consequence, their distinction has played an important role in the field's theoretical and experimental development.
 However, an issue that has received little attention is the justification of semantic structure in light of pragmatics.
-That is, an account for the selection and pervasiveness of particular semantic structures over others under consideration of the informational enrichment provided by pragmatics.
+That is, the selection and pervasiveness of particular semantics under consideration of the regular informational enrichment provided by pragmatics.
 
-%For instance, truth-conditionally, {\em Bill read five books} does not set an upper-bound to the amount of books that Bill read; he may have read six, seven, or more books.
-However, the (defeasible) inference that {\em Bill read no more than five books} may be drawn on pragmatic grounds.
-After all, if Bill had read, for example, six books, and had the speaker known this, she would have said so.
-In this manner, what is conveyed (pragmatically) can go beyond what is (literally) said.
+Similar questions have lead to a recent surge of models that aim to analyze the development and selection of linguistic features (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
+Our starting point is given by the overarching argument that has crystallized across this literature: Natural languages need to be well-adapted to communicative needs within a linguistic community, but also need to be learnable to survive their faithful transmission across generations.
+More succinctly; natural languages are pressured for expressiveness as well as learnability.
 
-%Prima facie, it is puzzling that {\em Bill read five books} does not semantically rule out its more informative or ``stronger'' alternatives.
-More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} or {\em big} were truth-conditionally incompatible with stronger alternatives such as, respectively, {\em hot}, {\em and}, {\em all} and {\em huge}?
-After all, analogously to the example sketched above, this is what a pragmatically enriched interpretation yields; an upper-bound that rules out these alternatives through a so-called scalar inference \citep{horn:1972,gazdar:1979}.
-The present investigation focuses on the lack of lexicalization of such upper-bounds in natural languages.
-That is, we seek to provide an explanation for the selection of the particular semantics of scalar expressions, which in turn hinges on the semantics-pragmatics distinction for what is ultimately conveyed to interlocutors.
-However, once a distinction between semantics and pragmatics is drawn, similar questions can be posed for other types of pragmatic inferences, such as ignorance and manner inferences.
+%While their efforts have largely concentrated on compositional and combinatorial systems, 
+We proceed by modeling these pressures using the replicator-mutator dynamics (see \citealt{hofbauer+sigmund:2003} for an overview).
+This allows us to inspect their dynamics by combining functional pressure on successful communication, effects of learning biases on (iterated) Bayesian learning \citep{griffiths+kalish:2007}, and  probabilistic models of language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
 
-A number of recent investigations have began to address issues pertaining to the development and selection of linguistic properties (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
-Our starting point is given by the overarching account of competing pressures that has crystallized across these approaches: Natural languages are pressured for successful communication, one the one hand, and for acquisition, on the other.
-That is, they need to be well-adapted to the communicative needs within a linguistic community, as well as be learnable to survive their faithful transmission across generations.
+%In this way, the model establishes a link between models of synchronic probabilistic rational language use and diachronic models of cultural evolution.
 
-We proceed by instantiating these pressures in the replicator-mutator dynamics, commonly used in evolutionary game theory.
-This model allows for general and precise means to model the dynamics of linguistic pressures by combining functional pressure on successful communication, effects of learning biases on (iterated) Bayesian learning \citep{griffiths+kalish:2007}, and  probabilistic models of language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
-Drawing from the latter research strand allows us to make the distinction between semantics and pragmatics in communicative behavior precise.
-In this way, this model bridges the recent surge of synchronic probabilistic rational language use analyses with diachronic models of cultural evolution.
-We then analyze the prevalence of a lack of semantic upper-bounds in the literal meaning of weak scalar expressions.
+%We then analyze the prevalence of a lack of semantic upper-bounds in the literal meaning of weak scalar expressions.
 We show that a lack of upper-bounds is selected for when learners are biased towards simpler semantic representations, provided they have means to convey upper-bounds, i.e., provided that they can be derived via pragmatic reasoning.
 
-\section{Simplicity and expressivity at the semantics-pragmatics interface, and beyond}
+\section{Simplicity, expressivity, and learnability}
 
-\begin{itemize}
-  \item \hl{General introduction to the idea of competing pressures in shaping languages (\citealt{zipf:1949}, Piantadosi,...}
-  %Zipf's principle of least effort as an account of human behavior. In language he interpreted it as minimal articulation with maximal expliciteness
-  %Martinet 1962:139 describes language change as an interaction of `` ‘first,   the   requirements   of  communication,  the  need  for  the  speaker  to  convey  his  message,  and  second,  the principle of least effort, which makes him restrict his output of energy, both mental and physical, to the minimum compatible with achieving his ends
-  \item \hl{Idea in pragmatics: \citealt{clark+wilkes-gibbs:1986}, \citealt{horn:1984}}
-  %Principle of least collaborative effort The principle  of least  collaborative  effort:  In conversation, the  participants  try  to  minimize  their  collaborative  effort—the  work that both do from the initiation of each contribution to its mutual acceptance
-  \item \hl{Discuss Q- and R-predictions for language change (cf. \citealt{traugott:2004})}
-  \item \hl{Idea in LOT (\citealt{piantadosi+etal:underreview})}
-\end{itemize}
+The emergence and change of linguistic structure is influenced by many factors, ranging from biological and socio-ecological to cultural \citep{steels:2011,tamariz+kirby:2016}.
+Social and ecological pressures determine communicative needs, while biology determines the architecture that enables and constrains their means of fulfillment.
+In the following, our focus lies on the latter, cultural factor, wherein certain processes of linguistic change are understood as shaped by use and transmission.
+That is, as a result of cultural evolution.
 
-\subsection{Competing pressures in cultural evolution}
-\begin{itemize}
-  \item \hl{discuss past research on interplay of pressures from a population-level perspective (see discussion in \citealt{kirby+etal:2015} for IL}
-  \item \hl{Learnability in iterated learning}
-  \item \hl{Expressivity in GT and some IL}
-  \item \hl{Sketch of replicator-mutator dynamics}
-\end{itemize}
+At latest since \citeposs{zipf:1949} rationalization of the observation that word frequency rankings can be approximated by a power law distribution as competing hearer and speaker preferences, the idea that linguistic selection and change are driven by communicative pressures has played a pivotal role in synchronic and diachronic analyses (e.g.
+\citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007, piantadosi:2014,kirby+etal:2015}).
+As noted above, expressivity and learnability are two key instances of such competing pressures.
+Their opposition becomes particularly clear when considering their consequences in the extreme (cf.
+\citealt{kemp+regier:2012,kirby+etal:2015}).
+On the one side, a language with a single form is easy to learn but lacking in expressivity for most applications.
+On the other, a language that associates a distinct form with all possible meanings its users may want to convey is maximally expressive but challenging to acquire.
 
-\hl{More detailed discussion of models of cultural evolution.
-Short overview of past research with a focus on the difference between pure IL and functional pressure together with IL.
-{\bf Possibly add a direct comparison of IL and RMD in the appendix using the setup of Griffiths \& Kalish 2007.
-This may not be necessary.}}
+The most prominent problem that arises from the tension between learnability and expressivity is that of acquiring a language to express a potentially infinite set of meanings through finite means \citep{kirby:2002}.
+However, this so-called transmission bottleneck is not the only challenge learners confront.
+More important for our purposes is the problem of selecting particular hypotheses out of a potentially infinite space of alternatives compatible with the data learners are exposed to.
+At the semantics-pragmatics interface this concerns the selection between functionally similar, if not identical, lexical meanings.
+In the following, we assume an integeral part of the answer to be that learners are a priori biased towards simpler, more compressed, representations.
+This corresponds to the argument that rational learners should prefer simpler over more complex explanations of data \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
+In linguistics, a drive for simplicity has been argued to underpin speaker preferences for brevity and ease of articulation, as well as to pressure languages towards lexical ambiguity and grammatical compression \citep{zipf:1949,grice:1975,piantadosi+etal:2012, kirby+etal:2015}.
+\hl{As a broader cognitive principle, the use of simplicity as means to select betweeen hypotheses that fit the data has a long standing tradition.
+Crucially, \citet{chater+vitanyi:2003} provide strong arguments for a prior towards simpler explanations.
+{\bf this needs to be expanded}}
 
-The emergence and change of linguistic structure is driven by many factors, from biological and socio-ecological to cultural \citep{steels:2011,tamariz+kirby:2016}.
-Broadly put, social and ecological pressures determine communicative needs, while biology determines the architecture available for its use.
-Our focus is on the latter, cultural, factor, wherein linguistic structure is analyzed in terms of its use, as well as its transmission across generations.
+The remainder of this section introduces the individual components of the model in more technical detail.
+That is: (i) languages and their use, (ii) pressures towards expressivity and learnability, regulated by the replicator and mutator dynamics, respectively, as well as (iii) a bias towards simpler semantic representations, codified as a language learner's prior.
+After laying out the model, we discuss its application to the lack of lexicalization of scalar implicatures.
 
-As already mentioned in \S\ref{sec:introduction}, research on selectional forces that apply in the cultural evolution of language has focused on two main pressures: expressivity and learnability.
-However, while it is generally acknowledged that both play a pivotal role, past approaches have focused exclusively, or at least emphasized, the role of one over the other (a recent exception is \citealt{kirby+etal:2015}).
+%The observation that monomorphemic expression that lexically rule out stronger alternatives (e.g.
+{not all} as hypothetical `nall') are unattested across languages has received substantial argumentative support (most prominently in \citealt[252-267]{horn:1984} but also e.g.
+in \citealt{horn:1972,traugott:2004,vdAuwera:2010}).
+To the best of our knowledge, this claim stands unchallenged.
 
-Expressiveness, or communicative efficiency, has been at the center of applications of evolutionary game theory to linguistics \citep{nowak+krakauer:1999,huttegger+zollman:2013},  \hl{explain RMD}
+\subsection{Languages and linguistic behavior}
+Lexica are taken to codify the truth-conditions of a language's expressions, i.e., its semantics.
+Given a state of affairs and a lexicon, language users can judge whether an expression is true or false.
+A convenient way to represent such lexica is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs and $M$ the set of messages in the lexicon \citep{franke+jaeger:2014}.
+For instance, the following two lexica fragments determine the truth-conditions of two messages, $m_1$ and $m_2$, for two states, $s_1$ and $s_2$:
 
-In contrast, the iterated learning paradigm has focused on the effects of language transmission from generations of speakers to the next. \hl{explain IL}
+\begin{centering}
+$L_a$ = \bordermatrix{~ & m_1 & m_2 \cr 
+                  s_1 & 1 & 0 \cr
+                  s_2 & 1 & 1 \cr} \hspace{2cm} $L_b$ = \bordermatrix{~ & m_1 & m_2 \cr 
+                  s_1 & 1 & 0 \cr
+                  s_2 & 0 & 1 \cr}\\[0.5cm]
+\end{centering}
 
-\section{Model}
-\subsection{Replicator-mutator dynamics}
+In words, according to lexicon $L_a$, $m_1$ is true of $s_1$ as well as $s_2$.
+In contrast, message $m_1$ is only true of $s_1$ in $L_b$.
+Otherwise, the two languages are truth-conditionally equivalent.
+
+To make the distinction betweeen semantics and pragmatics precise, we distinguish between two kinds of linguistic behavior.
+{\em Literal interlocutors} produce and interpret messages literally.
+That is, their linguistic choices are solely guided by their lexica.
+In contrast, {\em pragmatic interlocutors} engage in mutual reasoning to inform their choices.
+For instance, given that $m_2$ is unambiguously associated with state $s_1$, a rational speaker of $L_a$ who reasons about her addressee should use $m_1$ to signal state $s_2$.
+Analogously, should rational hearers expect their interlocutors to reason along these lines, they will interpret the messages accordingly.
+Note in particular that according to this sketch of the pragmatic strenghtening of $m_1$, $L_a$ is indistinguishable from $L_b$ in terms of expressiveness.
+
+Following models of rational language use such as Rational Speech Act models \citep{frank+goodman:2012} and their game-theoretic counterparts \citep{benz+etal:2005a,franke:2009,franke+jaeger:2014}, this kind of signaling behavior is captured by a hierarchy over reasoning types.
+The hierarchy's bottom, level $0$, corresponds to literal language use.
+Language users of level $n + 1$ behave rationally according to (expected) level $n$ behavior of their interlocutors.
+The behavior of literal level $0$ and pragmatic level $n+1$ hearers of a language $L$ is is captured by their respective selection functions in (\ref{h:level0}) and (\ref{h:leveln}).
+Mutatis mutandis for the speaker functions in (\ref{s:level0}) and (\ref{s:leveln}).
+
+\begin{flalign}
+&H_{0}(s|m;L) \propto pr(s) L_{sm} \label{h:level0}\\
+&S_{0}(m|s;L) \propto \exp(\lambda \; L_{sm}) \label{s:level0}\\
+&H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
+&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)^\alpha) \label{s:leveln}
+\end{flalign}
+
+According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state $s$ depends on her lexicon, weighted by her prior over states, $pr \in \Delta(S)$.
+The latter plays an important role when hearers face ambiguous messages about which the prior is informative.
+
+The literal speaker's choice in (\ref{s:level0}) is regulated by a soft-max parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}.
+As $\alpha$ increases, choices made in production are more rational; higher values lead to more deterministic expected utility maximizing behavior.
+
+Pragmatic behavior resembles its literal counterpart.
+As described above, the crucial difference is that level $n+1$ speakers/hearers reason about level $n$ hearer/speaker behavior.
+That is, they reason about how a rational level $n$ interlocutor would use or interpret a message and behave according to these expectations.
+Additionally, pragmatic production is further regulated by a parameter $\alpha$ which controls the tension between semantics and pragmatics, $\alpha \in (0,1]$.
+Lower values lead to more literal production, whereas higher values lead to more pragmatic behavior.
+
+The combination of a lexicon with its use, i.e., a degree of pragmatic sophistication, yields a type $t \in T$.
+Accordingly, $T$ contains all all lexica and reasoning level pairings under consideration.
+Types are the basic units on which our population dynamics operate.
+
+\subsection{Replication \& expressivity}
+On a population level, expressiveness, or communicative efficiency, has received particular attention from investigations using evolutionary game theory \citep{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}.
+Under this view, successful communication confers a higher fitness to some types relative to less successful ones.
+As a consequence, types with a higher fitness replicate and spread through the population, whereas the proportion of communicatively less efficient types decreases.
+This association of a type's communicative success within a population with changes in the types present in it creates a feedback loop that pressures the population towards greater expressivity.
+The replicator equation gives us the means to make these dynamics precise.
+
+The proportion of types in a given population is captured by a vector $x$, where $x_i$ is type $i$'s proportion in the population.
+The fitness of a type $i$, $f_i$, is given by its expected utility when interacting in this population.
+That is, its fitness is the sum of its weighted expected communicative success, $f_i = \sum_j x_j \text{EU}(t_i,t_j)$.
+The expected utility of $i$ and $j$ is obtained by considering the expected utility of speaker $i$ interacting with hearer $j$, and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$.
+$U_S(x,y)$ and $U_R(x,y)$ are respectively $\sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$ and $U_S(y,x)$ for $n$ and $o$ being the reasoning level of $x$ and $y$, and $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.\footnote{Note that the definition of $U_R(\cdot,\cdot)$ implies equal sender and receiver payoff in an interaction.
+This need not be so in the general case but suffices for our application.} Lastly, the average fitness of the population is captured by $\Phi$, $\Phi = \sum_i x_i f_i$.
+This term serves as a normalizing constant for the (discrete) replicator equation; $\dot{x}_i = \frac{x_i f_i}{\Phi}$
+
+Under its biological interpretation, the replicator equation captures the idea of fitness-relative selection whereby fitter types produce more offspring, leading to their propagation in subsequent generations.
+In analogy to biological replication, many aspects of natural language are subject to processees of change across varied time-spans.
+For example, the replicator equation can be understood as a learning across generations as e.g.
+in \citealt{nowak+etal:2002}, but also as a process of horizontal adaptation (see \citealt[\S3.3]{benz+etal:2005b} for discussion).
+In the following, we take the latter view and assume that interlocutors adapt their language's semantics and how it is used to that which works best within their population.
+In contrast to the effect of language acquisition from a generation to the next, this means that expressivity exherts its pressure on the current members of a linguistic community.
+For pressures of language learning from one generation to the next, we turn to iterated Bayesian learning as mutator dynamics.
+
+\subsection{Mutation \& learning}
+While past research using evolutionary game theory has mostly focused on communicative fitness driven change, the contribution of learning has similarly been argued to be one if not the central force guiding language change.
+In a nutshell, the iterated learning paradigm models language transmission as chains of parents and children \hl{citation IL.
+Look up which ones they use}.
+The parent produces linguistic data from which the child infers a language to adopt.
+In the next generation, previous children adopt the role of a teaching parent that will themselves produce data to be learnt by a new generation of na\"ive learners.
+The central idea is that this iterated learning process pressures languages towards greater learnability.
+In this way, linguistic complexity hindering may be reduced, and languages regularized, to overcome to transmission bottleneck.
+
+\hl{Some more technical exposition of IL. Mainly based on G+K07}
+
+In our present setup, the learners' task is to perform a joint inference over lexica and signaling behavior.
+As with the sole replicator dynamics, the mutator dynamics are deterministic.
+They are codified in a transition matrix $Q$, where $Q_{ij}$ indicates the probability of the children of a parent of type $i$ adopting type $j$.
+This, in turn, is proportional to the probability of $i$ producing the learning data and that of $j$ given the data.
+
+The elements of the set of learning data $D$ are sequences of length $k$ of state-message pairings.
+That is, a sequence of observations of language use.
+Put differently, a datum $d \in D$ contains $k$ members of the set $\{\tuple{s_i,m_j} | s_i \in S, m_j \in M\}$ and $D$ is the set of all such sequences.
+Having fixed $D$,
+\[
+ Q_{ij} \propto \sum_{d \in D} P(d|t_i) F(t_j,d),
+\]
+where $F(t_j,d) \propto P(t_j|d)^l$ and $P(t_j|d) \propto P(t_j) P(d|t_j)$.
+Given a type $i$, $P(d|t_i)$ can be straightforwardly computed based on $t_i$'s production behavior.
+However, the above involves two crucial aspects that require a more detailed discussion: the posterior parameter $l$ and the learning prior $P \in \Delta(\mathcal{T})$.
+
+\paragraph{Posterior parametrization.}
+
+Crucially, the posterior is parametrized by $l$, $l \geq 1$.
+When $l = 1$ learners sample from the posterior.
+As $l$ increases towards infinity, the learners tendency to posterior maximization increases.
+\hl{Some more discussion on posterior maximization}
+
+\paragraph{Learning prior.}
+
+\subsection{Summary}
+
+\section{Discussion}
+
 \subparagraph{(II) Sequences and atomic observations.} Before, the set of all observations was $O =$\linebreak  $\{\tuple{\tuple{s_1,m_i},\tuple{s_2,m_j}} | m_i, m_j \in M\}$.
@@
 Learners witnessed such data sequences.
-Now, more in line with \citet{griffiths+kalish:2007}, $O = \{\tuple{s_i,m_j} | s_i \in S, m_j \in M\}$ and $d$ is a sequence of length $k$ of members of $O$.
-The main difference is that now some $d$ do not provide any production information for some states.
 
@@
 
-\subparagraph{(IV) Parametrized learning} $Q_{ij} \propto \sum_d P(d|t_i) F(t_j,d)$, where $F(t_j,d) \propto P(t_j|d)^l$ and $l =1$ corresponds to probability matching and, as $l$ increases towards infinity, to MAP.
-
+\subparagraph{(IV) Parametrized learning} 
 The proportion of players of type $i$, $x_i$, is initialized as an arbitrary distribution over $T$.
@@
 \begin{itemize}
-    \item $f_i = \sum_j x_j U(x_i,x_j)$
-    \item $\Phi = \sum_i x_i f_i$
     \item $Q_{ij} \propto \sum_d P(d|t_i) \; P(t_j|d)$, where $P(t_j|d) \propto [P(t_j) P(d|t_j)]^l$, $d$ is a sequence of observations of length $k$ of the form \tuple{\tuple{s_i,m_j}, ...
@@
 \end{itemize}
-
-\subsection{Expressiveness as fitness-relative replication}
 
@@
 
-\subsection{Iterated learning as acquisition-based mutation}
+\section{Lack of semantic upper-bounds in lexical meaning}
 
-\subsection{Signaling behavior}
-\hl{Leave languages undefined until application}
+\hl{Note that this corresponds to the description of scalar implicatures given above: Reasoning about the alternatives at the speaker's disposition can pragmatically strengthen an expression's literal meaning by ruling out stronger alternatives not mentioned.}
 
-\paragraph{Signaling behavior.}
-\hl{Exposition of signaling behavior as reasoning hierarchy, which we use to make a distinction between semantic and pragmatic language users}
+\hl{A particularly well-studied type of conventional pragmatic enrichment are so-called {\em scalar implicatures} (cf.
+\citealt{horn:1972,gazdar:1979}).
+These inferences are licensed for groups of expressions ordered in terms of informativity, here understood as an entailment induced order.
+For instance, {\em some} is entailed by {\em all}; if it were true that `All students came to class', it would also be true that `Some students came to class'.
+However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not necessarily what their use is taken to convey.
+Instead, the use of a less informative expression when a more informative one could have been used can license the inference that the stronger alternative does not hold.
+That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that, since the speaker did not use a stronger alternative ({\em all}), this alternative must not hold.
+In this way, `Some students came to class' is strengthened as conveying `Some but not all students came to class'.
+Analogously, a speaker can rely on her interlocutor to draw this inference without having to express the bound overtly, e.g.
+by stating {\em some but not all}.
+In summary, mutual reasoning about rational language use supplies an upper-bound that rules out stronger alternatives pragmatically.}
 
-With $\lambda \geq 1$ (rationality parameter), $\alpha \in [0,1]$ (pragmatic violations) and $pr \in \Delta(S)$ a common prior over $S$ (uniform so far):
+\hl{We can now rephrase our initial question in terms of scalar implicatures: Why is the lack of lexical upper-bounds in weak scalar alternatives regularly selected for over the alternative of lexicalizing it?
+This question is particularly striking considering the number of classes of expressions that license such inferences across languages (\citealt{horn:1972}, \citealt[252-267]{horn:1984}, \citealt{traugott:2004,vdAuwera:2010}).}
 
-\begin{flalign}
-&R_{0}(s|m;L) \propto pr(s) L_{sm}\label{litl}\\
-&S_{0}(m|s;L) \propto \exp(\lambda \; L_{sm}) \label{lits}\\
-&R_{1}(s|m;L) \propto pr(s) S_{0}(m|s;L) \label{pragl}\\
-&S_{1}(m|s;L) \propto  \exp(\lambda \; R_{0}(s|m;L)^\alpha) \label{prags}
-\end{flalign}
+\hl{For instance, truth-conditionally, {\em Bill read five books} does not set an upper-bound to the amount of books that Bill read; he may have read six, seven, or more books.
+However, the (defeasible) inference that {\em Bill read no more than five books} may be drawn on pragmatic grounds.
+After all, if Bill had read, for example, six books, and had the speaker known this, she would have said so.
+In this manner, what is conveyed (pragmatically) can go beyond what is (literally) said.
+}
 
-\section{Lack of semantic upper-bounds in lexical meaning}
+\hl{Prima facie, it is puzzling that {\em Bill read five books} does not semantically rule out its more informative or ``stronger'' alternatives.
+More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} or {\em big} were truth-conditionally incompatible with stronger alternatives such as, respectively, {\em hot}, {\em and}, {\em all} and {\em huge}?
+After all, analogously to the example sketched above, this is what a pragmatically enriched interpretation yields; an upper-bound that rules out these alternatives through a so-called scalar inference \citep{horn:1972,gazdar:1979}.
+The present investigation focuses on the lack of lexicalization of such upper-bounds in natural languages.
+That is, we seek to provide an explanation for the selection of the particular semantics of scalar expressions, which in turn hinges on the semantics-pragmatics distinction for what is ultimately conveyed to interlocutors.
+However, once a distinction between semantics and pragmatics is drawn, similar questions can be posed for other types of pragmatic inferences, such as ignorance and manner inferences.}
 

### 2016-09-14 IL-comparison/il-comparison-rmd.py

@@ -21,17 +21,19 @@ n = 10 #observation sequence length
 generations = 10000 #generations per game
 games = 1000 #amount of independent games
 alpha = 0.5 #bias parameter
-epsilon = 0.05
+epsilon = 0#0.05
 mc = 1000 #amount of samples
 
 
 f_il_samp_mean = csv.writer(open('./results/comparison-il-samp-mean-a%d-e%f-k%d-g%d-r%d.csv' %(alpha,epsilon,n,generations,games),'wb')) 
-f_unwgh_samp_mean = csv.writer(open('./results/comparison-unwgh-samp-mean-a%d-e%f-k%d-g%d-r%d.csv' %(alpha,epsilon,n,generations,games),'wb')) 
-f_wgh_samp_mean = csv.writer(open('./results/comparison-wgh-samp-mean-a%d-e%f-k%d-g%d-r%d.csv' %(alpha,epsilon,n,generations,games),'wb')) 
+f_rmd_samp_mean = csv.writer(open('./results/comparison-rmd-samp-mean-a%d-e%f-k%d-g%d-r%d.csv' %(alpha,epsilon,n,generations,games),'wb')) 
+f_il_q = csv.writer(open('./results/comparison-il-samp-q-a%d-e%f-k%d.csv' %(alpha,epsilon,n),'wb')) 
+f_rmd_q = csv.writer(open('./results/comparison-rmd-samp-q-a%d-e%f-k%d.csv' %(alpha,epsilon,n),'wb')) 
 
 f_il_map_mean= csv.writer(open('./results/comparison-il-map-mean-a%d-e%f-k%d-g%d-r%d.csv' %(alpha,epsilon,n,generations,games),'wb')) 
-f_unwgh_map_mean= csv.writer(open('./results/comparison-unwgh-map-mean-a%d-e%f-k%d-g%d-r%d.csv' %(alpha,epsilon,n,generations,games),'wb')) 
-f_wgh_map_mean = csv.writer(open('./results/comparison-wgh-map-mean-a%d-e%f-k%d-g%d-r%d.csv' %(alpha,epsilon,n,generations,games),'wb')) 
+f_rmd_map_mean = csv.writer(open('./results/comparison-rmd-map-mean-a%d-e%f-k%d-g%d-r%d.csv' %(alpha,epsilon,n,generations,games),'wb'))
[…]

### 2016-09-15 paper/paper.tex
paper part of mutation draft and beginning of SI exposition

@@
 Similar questions have lead to a recent surge of models that aim to analyze the development and selection of linguistic features (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
-Our starting point is given by the overarching argument that has crystallized across this literature: Natural languages need to be well-adapted to communicative needs within a linguistic community, but also need to be learnable to survive their faithful transmission across generations.
+Our starting point is given by the overarching argument that has crystalized from accumulated mathematical, experimental and cross-linguistic evidence in this literature: Natural languages need to be well-adapted to communicative needs within a linguistic community, but also need to be learnable to survive their faithful transmission across generations.
 More succinctly; natural languages are pressured for expressiveness as well as learnability.
@@
 In contrast to the effect of language acquisition from a generation to the next, this means that expressivity exherts its pressure on the current members of a linguistic community.
-For pressures of language learning from one generation to the next, we turn to iterated Bayesian learning as mutator dynamics.
+
+Nowak \& et al.
+did not only consider replication, but also recognized the relevance of variation introduced by a language's transmition across generations, construed as mutation.
+As a consequence, the offspring of a type may end up adopting a different type than their parents.
+However, in this work mutation rates were construed as independent from a type.
+That is, the variation introduced by generational turnovers did not depend on a the relative learnability of a type based on learning data or learning biases.
+Therefore, to address the issue of selecting a particular type over (near) functional equivalents, we turn to a different strand of research in cultural evolution: {\em iterated learning}.
 
 \subsection{Mutation \& learning}
-While past research using evolutionary game theory has mostly focused on communicative fitness driven change, the contribution of learning has similarly been argued to be one if not the central force guiding language change.
-In a nutshell, the iterated learning paradigm models language transmission as chains of parents and children \hl{citation IL.
-Look up which ones they use}.
-The parent produces linguistic data from which the child infers a language to adopt.
-In the next generation, previous children adopt the role of a teaching parent that will themselves produce data to be learnt by a new generation of na\"ive learners.
-The central idea is that this iterated learning process pressures languages towards greater learnability.
-In this way, linguistic complexity hindering may be reduced, and languages regularized, to overcome to transmission bottleneck.
+Iterated learning is a process in which the behavior of one individual serves as learning input for another, who's behavior subsequently serves as input for a new learner, and so on.
+In linguistic terms, this process can be construed as chains of parents and children, where the parent produces linguistic data from which the child infers a language.
+The latter then goes on to produce linguistic data for a new generation of na\"ive learners.
+A central result of this paradigm is that this process leads to simpler and more regular languages in virtue of the pressure towards greater learnability exherted (surveys of experimental data and models are given in \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016}).
 
@@
 \subsection{Summary}
+Following past research, we have argued for expressivity, learnability and simplicity as central pressures that apply on language change.
+The dynamics we propose model them as fitness-relative replication, mutation based on iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings.
+Taken together the dynamics are described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:
 
-\section{Discussion}
+\[ 
+\hat{x_i} = \sum_j Q_{ji} \frac{x_jf_j}{\Phi}
+\]
 
-\subparagraph{(II) Sequences and atomic observations.} Before, the set of all observations was $O =$\linebreak  $\{\tuple{\tuple{s_1,m_i},\tuple{s_2,m_j}} | m_i, m_j \in M\}$.
+The basic units that these pressures apply to are a combination of semantics, i.e., a lexicon, and its use.
+The expressivity of these types depends on their communicative efficiency within a population.
+Their learnability depends on the fidelity by which they can be infered by a new generation of learners based on production data.
+In particular, learners perform a joint inference on types of linguistic behavior and lexical meaning.
+This process is influenced by the learners' preference for simpler lexical representations, codified as a learning bias.
+
+\hl{Summarize novelties \& advantages
+1. Advantage of connection between two strands of research
+2. Simple and general mean field dynamics}
+
+%\hl{So far, we have focused on highlighting the strands of research and assumptions present in the model.
+Less has been said about its departure from past approaches, beyond the virtues we argued their synthesis gives rise to.
+Given the different foci of these investigations, most comparisons on a general level are misplaced.
+However, \citet{kirby+etal:2015} ...}
+
+%%%Differences to our earlier CogSci proceedings paper:
+%Sequences and atomic observations.
+Before, the set of all observations was $O =$\linebreak  $\{\tuple{\tuple{s_1,m_i},\tuple{s_2,m_j}} | m_i, m_j \in M\}$.
 A member of $O$ encodes that a teacher produced $m_i$ in state $s_1$ and $m_j$ in $s_2$, i.e., it encodes one witnessed message for each state.
-A datum $d$ was a sequence of length $k$ of members of $O$.
-Learners witnessed such data sequences.
 
-\subparagraph{(III) Observations as production.} Instead of taking the space of all possible sequences of length $k$ into consideration, we take sample from $O$ $k$-times according to the production probabilities of each type; $P(o = \tuple{s,m} | t_i) = P(s) P(m|s,t_i)$.
+%Observations as production.
+Instead of taking the space of all possible sequences of length $k$ into consideration, we take sample from $O$ $k$-times according to the production probabilities of each type; $P(o = \tuple{s,m} | t_i) = P(s) P(m|s,t_i)$.
 $n$ such $k$-length sequences are sampled for each type.
@@
 
-\subparagraph{(IV) Parametrized learning} 
-The proportion of players of type $i$, $x_i$, is initialized as an arbitrary distribution over $T$.
-$p^\star \in \Delta(T)$ is learning a prior over (player) types dependent only on the lexicon of the type.
-\begin{itemize}
-    \item $Q_{ij} \propto \sum_d P(d|t_i) \; P(t_j|d)$, where $P(t_j|d) \propto [P(t_j) P(d|t_j)]^l$, $d$ is a sequence of observations of length $k$ of the form \tuple{\tuple{s_i,m_j}, ...
-\tuple{s_k, m_l}}, and $l \geq 1$ is a learning parameter.
-	\item For parental learning (standard RMD): $\dot x_i = \sum_j Q_{ji} \frac{x_j f_j}{\Phi}$
-\end{itemize}
-
-\paragraph{Symmetrized expected utility.} With $P \in \Delta(S)$ (uniform so far; $P = pr$):
-\begin{itemize}
-
-  \item $U(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)] / 2$
-  \item $U_S(t_i,t_j) = \sum_s P(s) \; \sum_m P_S(m|s;t_i) \; \sum_{s'} P_R(s'|m,t_j) \; \delta(s,s')$, where $\delta(s,s')$ returns $1$ iff $s = s'$ and otherwise $0$
-  \item $U_R(t_i,t_j) = U_S(t_j,t_i)$
-\end{itemize}
+%No communal learniner; only parental
+%%%
 
 \section{Lack of semantic upper-bounds in lexical meaning}
-
-\hl{Note that this corresponds to the description of scalar implicatures given above: Reasoning about the alternatives at the speaker's disposition can pragmatically strengthen an expression's literal meaning by ruling out stronger alternatives not mentioned.}
-
-\hl{A particularly well-studied type of conventional pragmatic enrichment are so-called {\em scalar implicatures} (cf.
+We set out to investigate the prevalence of lexical meanings that allow for regular pragmatic enrichments over other potential alternatives.
+A particularly well-studied type of conventional pragmatic enrichment are so-called {\em scalar implicatures} (cf.
 \citealt{horn:1972,gazdar:1979}).
 These inferences are licensed for groups of expressions ordered in terms of informativity, here understood as an entailment induced order.
+
 For instance, {\em some} is entailed by {\em all}; if it were true that `All students came to class', it would also be true that `Some students came to class'.
 However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not necessarily what their use is taken to convey.
-Instead, the use of a less informative expression when a more informative one could have been used can license the inference that the stronger alternative does not hold.
+Instead, the use of a less informative expression when a more informative one could have been used can license a defeasible inference that stronger alternatives do not hold.
 That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that, since the speaker did not use a stronger alternative ({\em all}), this alternative must not hold.
 In this way, `Some students came to class' is strengthened as conveying `Some but not all students came to class'.
-Analogously, a speaker can rely on her interlocutor to draw this inference without having to express the bound overtly, e.g.
+Analogously, a speaker can rely on her interlocutor to draw this inference without having to express this upper-bound overtly, e.g.
 by stating {\em some but not all}.
-In summary, mutual reasoning about rational language use supplies an upper-bound that rules out stronger alternatives pragmatically.}
+Put differently, mutual reasoning about rational language use supplies a bound that rules out stronger alternatives pragmatically.
 
-\hl{We can now rephrase our initial question in terms of scalar implicatures: Why is the lack of lexical upper-bounds in weak scalar alternatives regularly selected for over the alternative of lexicalizing it?
-This question is particularly striking considering the number of classes of expressions that license such inferences across languages (\citealt{horn:1972}, \citealt[252-267]{horn:1984}, \citealt{traugott:2004,vdAuwera:2010}).}
+Note that this corresponds to our previous description of the pragmatic use of lexicon $L_a$, repeated below for convenience.
+A pragmatic hearer who reasons about a speaker's use of message $m_1$ will associate it more strongly with $s_1$ than with $s_2$ given that the latter is unambiguously associated with the latter state.
+The strength of this association being dependent on their degree of rationality $\lambda$ and their prior over states.
+Conversely, a pragmatic speaker will reason about her interlocutors interpretation and use the messages accordingly.
 
-\hl{For instance, truth-conditionally, {\em Bill read five books} does not set an upper-bound to the amount of books that Bill read; he may have read six, seven, or more books.
-However, the (defeasible) inference that {\em Bill read no more than five books} may be drawn on pragmatic grounds.
-After all, if Bill had read, for example, six books, and had the speaker known this, she would have said so.
-In this manner, what is conveyed (pragmatically) can go beyond what is (literally) said.
-}
+\begin{centering}
+$L_a$ = \bordermatrix{~ & m_1 & m_2 \cr 
+                  s_1 & 1 & 0 \cr
+                  s_2 & 1 & 1 \cr} \hspace{2cm} $L_b$ = \bordermatrix{~ & m_1 & m_2 \cr 
+                  s_1 & 1 & 0 \cr
+                  s_2 & 0 & 1 \cr}\\[0.5cm]
+\end{centering}
 
-\hl{Prima facie, it is puzzling that {\em Bill read five books} does not semantically rule out its more informative or ``stronger'' alternatives.
+We can now rephrase our initial question in terms of scalar implicatures: Why is the lack of lexical upper-bounds in weak scalar alternatives, as $m_1$ in $L_a$, regularly selected for over the alternative of lexicalizing it as in $L_b$?
 More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} or {\em big} were truth-conditionally incompatible with stronger alternatives such as, respectively, {\em hot}, {\em and}, {\em all} and {\em huge}?
-After all, analogously to the example sketched above, this is what a pragmatically enriched interpretation yields; an upper-bound that rules out these alternatives through a so-called scalar inference \citep{horn:1972,gazdar:1979}.
-The present investigation focuses on the lack of lexicalization of such upper-bounds in natural languages.
-That is, we seek to provide an explanation for the selection of the particular semantics of scalar expressions, which in turn hinges on the semantics-pragmatics distinction for what is ultimately conveyed to interlocutors.
-However, once a distinction between semantics and pragmatics is drawn, similar questions can be posed for other types of pragmatic inferences, such as ignorance and manner inferences.}
+This question is particularly striking considering the number of classes of expressions that license such inferences across languages (\citealt{horn:1972}, \citealt[252-267]{horn:1984}, \citealt{traugott:2004,vdAuwera:2010}).
+In what follows we investigate the hypothesis that this preference is driven by the advantage in compression that lexical meanings lacking an upper-bound have over those that explicitly codify it.
+
+\hl{Other alternative explanations}
+
+\subsection{Application}
 

### 2016-09-16 paper/paper.tex
draft addition of discussion of prior and posterior + sketch of shared plot for fitness and learning only

@@
 %\usepackage{multicol}
-%\usepackage{graphicx}
+\usepackage{graphicx}
 \usepackage{color}
@@
 
-\begin{abstract} Certain semantic structures enable for pragmatic enrichments in a notably productive fashion.
+\begin{abstract} Many semantic structures enable for pragmatic enrichments in a notably productive fashion.
 This raises the challenge to justify their regular selection over alternatives that codify semantically what is conveyed pragmatically.
@@
 Iterated learning is a process in which the behavior of one individual serves as learning input for another, who's behavior subsequently serves as input for a new learner, and so on.
-In linguistic terms, this process can be construed as chains of parents and children, where the parent produces linguistic data from which the child infers a language.
+In linguistic terms, this process can be construed by chains of parents and children, where the parent produces linguistic data from which the child infers a language.
 The latter then goes on to produce linguistic data for a new generation of na\"ive learners.
-A central result of this paradigm is that this process leads to simpler and more regular languages in virtue of the pressure towards greater learnability exherted (surveys of experimental data and models are given in \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016}).
+Following \citet{griffiths+kalish:2007} we model learning as a process of Bayesian inference in which learners combine the likelihood of a type producing the learning data with prior inductive biases.
+They then select a type from the resulting posterior distribution according to their learning strategy.
 
-\hl{Some more technical exposition of IL. Mainly based on G+K07}
+A central result of iterated learning is that this it leads to simpler and more regular languages due to the pressure towards greater learnability it exherts (surveys of experimental data and models are given in \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016}).
+Importantly, experimental and mathematical results suggest the results of this process to reflect the learners' learning biases, codified as a prior $P \in \Delta(\mathcal{T})$.
+This bias can be thought of as the amount of data a learner would require in order to adopt a language or, in our case, a combination of a lexicon and a signaling behavior (cf.
+\citealt[450]{griffiths+kalish:2007}).
 
-In our present setup, the learners' task is to perform a joint inference over lexica and signaling behavior.
-As with the sole replicator dynamics, the mutator dynamics are deterministic.
-They are codified in a transition matrix $Q$, where $Q_{ij}$ indicates the probability of the children of a parent of type $i$ adopting type $j$.
+However, the extent of the influence of the prior on the dynamics was shown to depend on assumptions about the strategy employed by learners.
+While simulation results suggested that weak biases could be magnified by exposing learners to only small data samples \citep{brighton:2002}, the mathematical characterization provided by \citet{griffiths+kalish:2007} showed that, instead, iterated learning converged to the prior.
+That is, the distribution over languages in a population or, from an individual's perspective, the likelihood of learning a language corresponded to the learners' prior distribution -- irrespective of the amount of input given to learners.
+The divergence between these predictions crucially depends on the assumed learning strategy.
+On the one extreme, Griffith \& Kalish's prediction holds for learners that sample from the posterior, on the other, more deterministic strategies such as the selection of the type with the highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase the prior's influence.
+
+In the following, we parametrize the posterior, $P(t_i|d)^l$, to obtain a range of learning strategies that live in the range between posterior sampling and MAP, $l \geq 1$.
+When $l = 1$ learners sample from the posterior.
+As $l$ increases towards infinity, the learners tendency to posterior maximization increases.
+
+More generally, we combine the replicator dynamics with iterated learning by codifying the latter as a transition matrix $Q$, where $Q_{ij}$ indicates the probability of the children of a parent of type $i$ adopting type $j$.
 This, in turn, is proportional to the probability of $i$ producing the learning data and that of $j$ given the data.
+That is, the learner's task is to perform a joint inference over lexica and signaling behavior.
 
@@
 \]
-where $F(t_j,d) \propto P(t_j|d)^l$ and $P(t_j|d) \propto P(t_j) P(d|t_j)$.
-Given a type $i$, $P(d|t_i)$ can be straightforwardly computed based on $t_i$'s production behavior.
-However, the above involves two crucial aspects that require a more detailed discussion: the posterior parameter $l$ and the learning prior $P \in \Delta(\mathcal{T})$.
-
-\paragraph{Posterior parametrization.}
-
-Crucially, the posterior is parametrized by $l$, $l \geq 1$.
-When $l = 1$ learners sample from the posterior.
-As $l$ increases towards infinity, the learners tendency to posterior maximization increases.
-\hl{Some more discussion on posterior maximization}
-
-\paragraph{Learning prior.}
+where $F(t_j,d) \propto P(t_j|d)^l$ and $P(t_j|d) \propto P(t_j) P(d|t_j)$. Given a type $i$, $P(d|t_i)$ can be straightforwardly computed based on $t_i$'s production behavior.
 
 \subsection{Summary}
-Following past research, we have argued for expressivity, learnability and simplicity as central pressures that apply on language change.
-The dynamics we propose model them as fitness-relative replication, mutation based on iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings.
+So far, we have argued for expressivity, learnability and simplicity as central pressures that apply on the cultural evolution of language.
+The dynamics we propose model them as fitness-relative replication, mutation based on iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings, respectively.
 Taken together the dynamics are described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:
@@
 2. Simple and general mean field dynamics}
+
+\hl{Possibly put reference to Kalish in EVOLANG7, Spain, on selection + iterated learning, and our CogSci}
 
@@
 This question is particularly striking considering the number of classes of expressions that license such inferences across languages (\citealt{horn:1972}, \citealt[252-267]{horn:1984}, \citealt{traugott:2004,vdAuwera:2010}).
+
+\paragraph{Other hyp}
+\hl{The first is that pragmatic reasoning offers a general mechanism to strengthen the meaning of a wide range of expressions when the conditions outlined above hold.
+Consequently,  cases where cooperativity or  knowledge  are  not  likely  to  be  given  are  non-committal to whether stronger alternatives hold.
+If for all the speaker knows ‘some students came’ but she does not know whether ‘all came’, then the compatibility of some with (possibly) all
+succinctly conveys the speaker’s uncertainty about the latter.}
+
+\hl{Given that scalar expressions occur in contexts in which their
+upper-bounded reading is absent, one could argue for a func-
+tional advantage of a lack of semantic upper-bounds:  If ex-
+pressing such a state of affairs is relevant and contextual cues
+provide enough information for a hearer to discern when a
+bound is conveyed pragmatically, then doing so is preferred
+over enforcing the bound overtly through a longer (more com-
+plex) expression, e.g. by stating ‘some but not all’ explicitly.
+That is, all else being equal, speakers prefer to communicate
+as economically as possible, and pragmatic reasoning enables
+them to do so.  Additionally, this can be contrasted with the
+hypothetical alternative of lexicalizing two expressions – one
+with and one lacking an upper-bound.  Four conditions may
+pressure language to English-like semantics over this alter-
+native: (i) contextual cues are very reliable, morphosyntactic
+disambiguation is either (ii) not frequently necessary or (iii)
+not  very  costly,  or  (iv)  having  larger  lexica  is  more  costly
+than morphosyntactic disambiguation.   In a nutshell (i) and
+(ii) place a heavy burden on the ability to retrieve contextual
+cues to a degree that is unlikely to undercut the benefit of safe
+communication with more expressions.  As for (iii) and (iv),
+these seem mostly like technical solutions without a proper
+empirical basis.}
+
+\paragraph{Our hyp}
 In what follows we investigate the hypothesis that this preference is driven by the advantage in compression that lexical meanings lacking an upper-bound have over those that explicitly codify it.
 
-\hl{Other alternative explanations}
+At present, the contrast in compressability between lexical meanings, however, is not represented as lexica codify the truth-conditions of expressions at a lower level of granularity.
+In principle this difference could be made precise with an adequate representational language, e.g., through measures over representational complexity such as minimal description length.
+
+There is a growing effort to develop such
+empirically  testable  representational  languages.   For  instance,  the
+so-called
+language of thought
+has been put to test in various rational
+probabilistic models that show encouraging results (see e.g. Katz et
+al. 2008; Piantadosi et al. under review, 2012 and references therein).
+We think that our assumption is well-warranted as a working hypoth-
+esis and decide against such an enrichment given that the introduc-
+tion of a larger framework would also require further assumptions
+and justifications.
 
@@
 \end{enumerate}
+
+\subsection{Analysis}
+\hl{Possibly put only fitness and only learning in a single plot with shared y axis}
+
+%\begin{figure}
+%\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{../presentations/07fitness}
+%\end{figure}
+%
+%\begin{figure}
+%\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{../presentations/07learning}
+%\end{figure}
+
+\begin{figure}
+\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{../presentations/03cost-with-l1}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{../presentations/01heatmap}
+\end{figure}
 

### 2016-09-16 presentations/edindata/plot-only-fitness+only-learning.py

@@ -1,0 +1,117 @@
+###
+import pandas as pd
+import numpy as np 
+import matplotlib.pyplot as plt
+import seaborn as sns
+#sns.set_context(rc={'lines.markeredgewidth': 0.5})
+import os
+import glob
+import sys
+###
+###Column headers:###
+#run_ID,t1_initial,t2_initial,t3_initial,t4_initial,t5_initial,t6_initial,t7_initial,t8_initial,t9_initial,t10_initial,t11_initial,t12_initial,
+#alpha,prior_cost_c,lambda,k,sample_amount,learning_parameter,generations,
+#t1_final,t2_final,t3_final,t4_final,t5_final,t6_final,t7_final,t8_final,t9_final,t10_final,t11_final,t12_final
+#####################
+
+
+print 'Loading data fitness'
+path =r'./results-fit' 
+all_files = glob.glob(os.path.join(path, "*.csv"))     # advisable to use os.path.join as this makes concatenation OS independent
+df = pd.concat(pd.read_csv(f) for f in all_files)
+
+print 'Loading data learn'
+path =r'./results-learn' 
+all_files = glob.glob(os.path.join(path, "*.csv"))     # advisable to use os.path.join as this makes concatenation OS independent
+dl = pd.concat(pd.read_csv(f) for f in all_files)
+
+def cust_mean(grp):
+    grp['mean'] = grp['t11_final'].mean()
+    return grp
+
+
+
+
+### development over cost ###
+
+a = 1
+k = 5
+lam = 30
+sample = 10
+learn = 1
+
+
+final_groupA = df.loc[:,'alpha':]
+final_groupB = dl.loc[:,'alpha':]
+
+groupA = final_groupA.loc[final_groupA['alpha'] == a]
+groupA = groupA.loc[groupA['k'] == k]
+groupA = groupA.loc[groupA['sample_amount'] == sample]
+groupA = groupA.loc[groupA['learning_parameter'] == learn]
+groupA = groupA.loc[groupA['prior_cost_c'] == 0.1]
+
+t_finalA = groupA.groupby(['lambda'])
[…]

### 2016-09-18 paper/paper.tex
small draft fixes in introduction

@@
 
-\begin{abstract} Many semantic structures enable for pragmatic enrichments in a notably productive fashion.
-This raises the challenge to justify their regular selection over alternatives that codify semantically what is conveyed pragmatically.
-This issue is particularly puzzling under a purely functional perspective.
-To address this challenge, we propose a general model for the analysis of linguistic pressures that integrates iterated Bayesian learning in the replicator-mutator dynamics.
-This model allows for population-level analyses of the effects of such pressures on probabilistic language users with varied degrees of pragmatic sophistication and distinct languages.
-We showcase the model's predictions in a case study on the (lack of) lexicalization of scalar implicatures.
-The results suggest simpler semantic representations to be selected for when languages are pressured towards learnability, provided that pragmatic reasoning can compensate for the disadvantage in expressivity that users of such languages otherwise incur.
+\begin{abstract} Certain lexical meanings enable for pragmatic enrichments in a notably productive fashion.
+This raises the challenge to justify their regular selection, in particular, over alternatives that codify semantically what is conveyed pragmatically.
+To address this challenge, we propose a general model that integrates iterated Bayesian learning in the replicator-mutator dynamics.
+This model allows for population-level analyses of the effects of linguistic pressures on probabilistic language users with varied degrees of pragmatic sophistication and distinct languages.
+We showcase the model's use and predictions in a case study on the (lack of) lexicalization of scalar implicatures.
+The results suggest simpler semantic representations to be selected for when languages are pressured towards learnability and compression, provided that pragmatic reasoning can compensate for the disadvantage in expressivity that users of such languages otherwise incur.
 
@@
 Broadly speaking, the former concerns the truth-conditional content of expressions, whereas the latter concerns information beyond literal meanings and their composition.
-Under this view, the information conveyed by an utterance is seldom, if ever, solely determined by semantics, but rather in tandem with pragmatics.
+An important consequence of this distinction is that the information conveyed by an utterance is seldom, if ever, solely determined by semantics, but rather in tandem with pragmatics.
 
 Much research at the semantics-pragmatics interface has been aimed at characterizing expressions in terms of either domain, or their interplay.
-As a consequence, their distinction has played an important role in the field's theoretical and experimental development.
 However, an issue that has received little attention is the justification of semantic structure in light of pragmatics.
-That is, the selection and pervasiveness of particular semantics under consideration of the regular informational enrichment provided by pragmatics.
+The present investigation seeks to fill this gap by analyzing the effects linguistic pressures have on the selection and pervasiveness of particular lexical meanings under consideration of pragmatic enrichments.
 
-Similar questions have lead to a recent surge of models that aim to analyze the development and selection of linguistic features (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
+In recent years, similar questions about the emergence and change of linguistic features have lead to a surge in models to address them (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
 Our starting point is given by the overarching argument that has crystalized from accumulated mathematical, experimental and cross-linguistic evidence in this literature: Natural languages need to be well-adapted to communicative needs within a linguistic community, but also need to be learnable to survive their faithful transmission across generations.
@@
 
-%While their efforts have largely concentrated on compositional and combinatorial systems, 
-We proceed by modeling these pressures using the replicator-mutator dynamics (see \citealt{hofbauer+sigmund:2003} for an overview).
-This allows us to inspect their dynamics by combining functional pressure on successful communication, effects of learning biases on (iterated) Bayesian learning \citep{griffiths+kalish:2007}, and  probabilistic models of language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
+We build on these insights by modeling these pressures using the replicator-mutator dynamics (see \citealt{hofbauer+sigmund:2003} for an overview).
+This allows for the inspection of their interaction by combining functional pressure on successful communication with effects of learning biases on (iterated) Bayesian learning \citep{griffiths+kalish:2007}.
+The semantics-pragmatics distinction and its effect on production and comprehension is made precise by considering probabilistic models of rational language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
 
@@
 
-The emergence and change of linguistic structure is influenced by many factors, ranging from biological and socio-ecological to cultural \citep{steels:2011,tamariz+kirby:2016}.
+The emergence and change of linguistic structure is influenced by many intertwined factors, ranging from biological and socio-ecological to cultural \citep{steels:2011,tamariz+kirby:2016}.
 Social and ecological pressures determine communicative needs, while biology determines the architecture that enables and constrains their means of fulfillment.
-In the following, our focus lies on the latter, cultural factor, wherein certain processes of linguistic change are understood as shaped by use and transmission.
+In the following, our focus lies on the latter, cultural factor, wherein processes of linguistic change are understood as shaped by its use and transmission.
 That is, as a result of cultural evolution.
 
-At latest since \citeposs{zipf:1949} rationalization of the observation that word frequency rankings can be approximated by a power law distribution as competing hearer and speaker preferences, the idea that linguistic selection and change are driven by communicative pressures has played a pivotal role in synchronic and diachronic analyses (e.g.
+At latest since \citeposs{zipf:1949} rationalization of the observation that word frequency rankings can be approximated by a power law distribution as competing hearer and speaker preferences, the idea that linguistic change is influenced by communicative pressures has played a pivotal role in synchronic and diachronic analyses (e.g.
 \citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007, piantadosi:2014,kirby+etal:2015}).
-As noted above, expressivity and learnability are two key instances of such competing pressures.
+
+As noted above, expressivity and learnability are two major competing pressures.
 Their opposition becomes particularly clear when considering their consequences in the extreme (cf.
 \citealt{kemp+regier:2012,kirby+etal:2015}).
-On the one side, a language with a single form is easy to learn but lacking in expressivity for most applications.
+On the one side, a language with a single form is easy to learn but lacking in expressivity for most purposes.
 On the other, a language that associates a distinct form with all possible meanings its users may want to convey is maximally expressive but challenging to acquire.
+The most prominent problem that arises from this tension is that of acquiring a language to express a potentially infinite set of meanings through finite means \citep{kirby:2002}.
+However, this so-called transmission bottleneck is not the only challenge learners confront.
 
-The most prominent problem that arises from the tension between learnability and expressivity is that of acquiring a language to express a potentially infinite set of meanings through finite means \citep{kirby:2002}.
-However, this so-called transmission bottleneck is not the only challenge learners confront.
 More important for our purposes is the problem of selecting particular hypotheses out of a potentially infinite space of alternatives compatible with the data learners are exposed to.
 At the semantics-pragmatics interface this concerns the selection between functionally similar, if not identical, lexical meanings.
-In the following, we assume an integeral part of the answer to be that learners are a priori biased towards simpler, more compressed, representations.
+In the following, we argue an integral part of the answer to be that learners are a priori biased towards simpler, more compressed, lexical representations.
 This corresponds to the argument that rational learners should prefer simpler over more complex explanations of data \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
 In linguistics, a drive for simplicity has been argued to underpin speaker preferences for brevity and ease of articulation, as well as to pressure languages towards lexical ambiguity and grammatical compression \citep{zipf:1949,grice:1975,piantadosi+etal:2012, kirby+etal:2015}.
-\hl{As a broader cognitive principle, the use of simplicity as means to select betweeen hypotheses that fit the data has a long standing tradition.
-Crucially, \citet{chater+vitanyi:2003} provide strong arguments for a prior towards simpler explanations.
-{\bf this needs to be expanded}}
+As a broader cognitive principle, the use of simplicity as means to select between hypotheses has a long standing tradition.
+Crucially, \citet{chater+vitanyi:2003} give a number of compelling arguments for simplicity on both mathematical and empirical grounds.
 
-The remainder of this section introduces the individual components of the model in more technical detail.
-That is: (i) languages and their use, (ii) pressures towards expressivity and learnability, regulated by the replicator and mutator dynamics, respectively, as well as (iii) a bias towards simpler semantic representations, codified as a language learner's prior.
+The remainder of this section introduces the individual components of the model in more detail, as well as the assumptions underlying them.
+These are: (i) languages and their use, (ii) pressures towards expressivity and learnability, regulated by the replicator and mutator dynamics, respectively, as well as (iii) a bias towards simpler semantic representations, codified as a language learner's prior.
 After laying out the model, we discuss its application to the lack of lexicalization of scalar implicatures.

### 2016-09-19 paper/paper.tex
shared y-axis for R and M only results, small overhaul for paper up to section 3

@@
 \subsection{Languages and linguistic behavior}
-Lexica are taken to codify the truth-conditions of a language's expressions, i.e., its semantics.
-Given a state of affairs and a lexicon, language users can judge whether an expression is true or false.
-A convenient way to represent such lexica is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs and $M$ the set of messages in the lexicon \citep{franke+jaeger:2014}.
+Lexica codify the truth-conditions of a language's expressions, i.e., its semantics.
+A convenient way to represent such lexica is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs, or meanings, to convey and $M$ a set of messages of the language \citep{franke+jaeger:2014}.
 For instance, the following two lexica fragments determine the truth-conditions of two messages, $m_1$ and $m_2$, for two states, $s_1$ and $s_2$:
@@
 
-In words, according to lexicon $L_a$, $m_1$ is true of $s_1$ as well as $s_2$.
+In words, according to lexicon $L_a$, $m_1$ is true of state $s_1$ as well as of $s_2$.
 In contrast, message $m_1$ is only true of $s_1$ in $L_b$.
@@
 {\em Literal interlocutors} produce and interpret messages literally.
-That is, their linguistic choices are solely guided by their lexica.
+That is, their linguistic choices are guided by their lexica only.
 In contrast, {\em pragmatic interlocutors} engage in mutual reasoning to inform their choices.
-For instance, given that $m_2$ is unambiguously associated with state $s_1$, a rational speaker of $L_a$ who reasons about her addressee should use $m_1$ to signal state $s_2$.
-Analogously, should rational hearers expect their interlocutors to reason along these lines, they will interpret the messages accordingly.
-Note in particular that according to this sketch of the pragmatic strenghtening of $m_1$, $L_a$ is indistinguishable from $L_b$ in terms of expressiveness.
+For instance, a rational speaker of $L_a$ who reasons about her addressee should use $m_1$ to signal state $s_1$ given that $m_2$ is already unambiguously associated with state $s_2$.
+Analogously, should rational hearers expect their interlocutors to reason along these lines, they will interpret ambiguous $m_1$ accordingly.
+Note in particular that according to this strenghtening of $m_1$, $L_a$ is indistinguishable from $L_b$ in terms of expressiveness if its users are pragmatic reasoners.
 
@@
 The hierarchy's bottom, level $0$, corresponds to literal language use.
-Language users of level $n + 1$ behave rationally according to (expected) level $n$ behavior of their interlocutors.
-The behavior of literal level $0$ and pragmatic level $n+1$ hearers of a language $L$ is is captured by their respective selection functions in (\ref{h:level0}) and (\ref{h:leveln}).
+Pragmatic language users of level $n + 1$ behave rationally according to (expected) level $n$ behavior of their interlocutors.
+The behavior of literal and pragmatic hearers of a language $L$ is given by their respective selection functions in (\ref{h:level0}) and (\ref{h:leveln}).
 Mutatis mutandis for the speaker functions in (\ref{s:level0}) and (\ref{s:leveln}).
@@
 
-According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state $s$ depends on her lexicon, weighted by her prior over states, $pr \in \Delta(S)$.
-The latter plays an important role when hearers face ambiguous messages about which the prior is informative.
+According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state $s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$.
+The literal speaker's choice in (\ref{s:level0}) is regulated by a soft-max parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}.
+As $\lambda$ increases, choices made in production are more rational in that higher values lead to more deterministic in line with expected utility maximization.
 
-The literal speaker's choice in (\ref{s:level0}) is regulated by a soft-max parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}.
-As $\alpha$ increases, choices made in production are more rational; higher values lead to more deterministic expected utility maximizing behavior.
-
-Pragmatic behavior resembles its literal counterpart.
-As described above, the crucial difference is that level $n+1$ speakers/hearers reason about level $n$ hearer/speaker behavior.
+For the most part, pragmatic behavior mirrors its literal counterpart.
+As described above, their difference lies in that level $n+1$ speakers/hearers reason about level $n$ hearer/speaker behavior instead of solely relaying on their lexicon.
 That is, they reason about how a rational level $n$ interlocutor would use or interpret a message and behave according to these expectations.
 Additionally, pragmatic production is further regulated by a parameter $\alpha$ which controls the tension between semantics and pragmatics, $\alpha \in (0,1]$.
-Lower values lead to more literal production, whereas higher values lead to more pragmatic behavior.
+Lower values lead to more literal production, whereas higher values lead to stronger pragmatic behavior.
 
-The combination of a lexicon with its use, i.e., a degree of pragmatic sophistication, yields a type $t \in T$.
-Accordingly, $T$ contains all all lexica and reasoning level pairings under consideration.
-Types are the basic units on which our population dynamics operate.
+The combination of a lexicon with its use, i.e., a particular level of linguistic sophistication, yields a type $t$. Types are the basic units on which our population dynamics operate.
 
 \subsection{Replication \& expressivity}
-On a population level, expressiveness, or communicative efficiency, has received particular attention from investigations using evolutionary game theory \citep{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}.
-Under this view, successful communication confers a higher fitness to some types relative to less successful ones.
-As a consequence, types with a higher fitness replicate and spread through the population, whereas the proportion of communicatively less efficient types decreases.
+Communicative efficiency, or expressiveness, has received particular attention from investigations using evolutionary game theory \citep{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}.
+Under this view, a type's success in communication confers it a higher fitness relative to less successful ones.
+As a consequence they replicate more than other types, increasing their proportion in the population.
 This association of a type's communicative success within a population with changes in the types present in it creates a feedback loop that pressures the population towards greater expressivity.
@@
 The proportion of types in a given population is captured by a vector $x$, where $x_i$ is type $i$'s proportion in the population.
-The fitness of a type $i$, $f_i$, is given by its expected utility when interacting in this population.
-That is, its fitness is the sum of its weighted expected communicative success, $f_i = \sum_j x_j \text{EU}(t_i,t_j)$.
+The fitness of a type $i$, $f_i$, is given by its expected utility in this population, $f_i = \sum_j x_j \text{EU}(t_i,t_j)$.
+That is, its fitness is the sum of its expected communicative success with other types weighted by the latter type's population share.
 The expected utility of $i$ and $j$ is obtained by considering the expected utility of speaker $i$ interacting with hearer $j$, and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$.
 $U_S(x,y)$ and $U_R(x,y)$ are respectively $\sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$ and $U_S(y,x)$ for $n$ and $o$ being the reasoning level of $x$ and $y$, and $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.\footnote{Note that the definition of $U_R(\cdot,\cdot)$ implies equal sender and receiver payoff in an interaction.
-This need not be so in the general case but suffices for our application.} Lastly, the average fitness of the population is captured by $\Phi$, $\Phi = \sum_i x_i f_i$.
+This need not be so in the general case but suffices for our application.} This quantity is symmetric, reflecting the probability of two types' mutual understanding.
+Lastly, the average fitness of the population is captured by $\Phi$, $\Phi = \sum_i x_i f_i$.
 This term serves as a normalizing constant for the (discrete) replicator equation; $\dot{x}_i = \frac{x_i f_i}{\Phi}$
@@
 Under its biological interpretation, the replicator equation captures the idea of fitness-relative selection whereby fitter types produce more offspring, leading to their propagation in subsequent generations.
-In analogy to biological replication, many aspects of natural language are subject to processees of change across varied time-spans.
+In analogy to this kind of replication, many aspects of natural language are subject to processees of transmission and change across varied time-spans.
 For example, the replicator equation can be understood as a learning across generations as e.g.
 in \citealt{nowak+etal:2002}, but also as a process of horizontal adaptation (see \citealt[\S3.3]{benz+etal:2005b} for discussion).
-In the following, we take the latter view and assume that interlocutors adapt their language's semantics and how it is used to that which works best within their population.
-In contrast to the effect of language acquisition from a generation to the next, this means that expressivity exherts its pressure on the current members of a linguistic community.
+In the following, we take the latter view in assuming that interlocutors adapt their lexica and their use to that which works best within their population.
+It should be stressed, however, that the model itself is compatible with either view.
 
-Nowak \& et al.
-did not only consider replication, but also recognized the relevance of variation introduced by a language's transmition across generations, construed as mutation.
-As a consequence, the offspring of a type may end up adopting a different type than their parents.
-However, in this work mutation rates were construed as independent from a type.
-That is, the variation introduced by generational turnovers did not depend on a the relative learnability of a type based on learning data or learning biases.
-Therefore, to address the issue of selecting a particular type over (near) functional equivalents, we turn to a different strand of research in cultural evolution: {\em iterated learning}.
+Nowak et al.
+did not only consider replication, but also recognized the important role of the variation that is introduced by a language's transmition across generations, construed as mutation.
+Due to this process, the offspring of a type may end up adopting a different type than that of its parent.
+Crucially, in this work mutation rates were modelled as begin independent from a type.
+This means that the variation introduced by generational turnovers did not depend on factors such as the relative learnability of a type.
+To address the issue of selecting a particular type over (near) functional equivalents, we turn to a different strand of research in cultural evolution: {\em iterated learning}.
 
@@
 Iterated learning is a process in which the behavior of one individual serves as learning input for another, who's behavior subsequently serves as input for a new learner, and so on.
-In linguistic terms, this process can be construed by chains of parents and children, where the parent produces linguistic data from which the child infers a language.
-The latter then goes on to produce linguistic data for a new generation of na\"ive learners.
+For linguistic purposes this process can be thought of as chains of parents and children, where the parent produces linguistic data from which the child infers a language.
+The latter, now a parent, goes on to produce linguistic data for a new generation of na\"ive learners.
 Following \citet{griffiths+kalish:2007} we model learning as a process of Bayesian inference in which learners combine the likelihood of a type producing the learning data with prior inductive biases.
-They then select a type from the resulting posterior distribution according to their learning strategy.
+They then select a type to adopt from the resulting posterior distribution.
 
-A central result of iterated learning is that this it leads to simpler and more regular languages due to the pressure towards greater learnability it exherts (surveys of experimental data and models are given in \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016}).
-Importantly, experimental and mathematical results suggest the results of this process to reflect the learners' learning biases, codified as a prior $P \in \Delta(\mathcal{T})$.
-This bias can be thought of as the amount of data a learner would require in order to adopt a language or, in our case, a combination of a lexicon and a signaling behavior (cf.
+Due to the pressure towards learnability it exherts, iterated learning generally leads to simpler and more regular languages (surveys of empirical data and models are given in \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016}).
+Importantly, experimental and mathematical results suggest the results of this process to reflect learners' learning biases, codified in the following as a prior $P \in \Delta(\mathcal{T})$.
+A way to think about this bias is as the amount of data a learner would require in order to adopt a language -- or, in our case, a combination of a lexicon and a signaling behavior (cf.
 \citealt[450]{griffiths+kalish:2007}).
-
-However, the extent of the influence of the prior on the dynamics was shown to depend on assumptions about the strategy employed by learners.
+Crucially, the extent of the prior's influence has been shown to strongly depend on the learning strategy assumed to underly the inference process.
 While simulation results suggested that weak biases could be magnified by exposing learners to only small data samples \citep{brighton:2002}, the mathematical characterization provided by \citet{griffiths+kalish:2007} showed that, instead, iterated learning converged to the prior.
-That is, the distribution over languages in a population or, from an individual's perspective, the likelihood of learning a language corresponded to the learners' prior distribution -- irrespective of the amount of input given to learners.
-The divergence between these predictions crucially depends on the assumed learning strategy.
-On the one extreme, Griffith \& Kalish's prediction holds for learners that sample from the posterior, on the other, more deterministic strategies such as the selection of the type with the highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase the prior's influence.
-
+That is, the distribution over languages in a population or, from an individual's perspective, the likelihood of learning a language corresponds to the learners' prior distribution, irrespective of the amount of input given to learners.
+This divergence in predictions can be traced back to differences in the selection of hypotheses from the posterior.
+On the one extreme, Griffith \& Kalish's convergence to the prior holds for learners that sample from the posterior.
+On the other, more deterministic strategies such as the selection of the type with the highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase the prior's influence \citep{griffiths+kalish:2007,kirby+etal:2007}.
 In the following, we parametrize the posterior, $P(t_i|d)^l$, to obtain a range of learning strategies that live in the range between posterior sampling and MAP, $l \geq 1$.
 When $l = 1$ learners sample from the posterior.
-As $l$ increases towards infinity, the learners tendency to posterior maximization increases.
+As $l$ increases towards infinity, the learners' tendency maximize the posterior increases.
 
-More generally, we combine the replicator dynamics with iterated learning by codifying the latter as a transition matrix $Q$, where $Q_{ij}$ indicates the probability of the children of a parent of type $i$ adopting type $j$.
-This, in turn, is proportional to the probability of $i$ producing the learning data and that of $j$ given the data.
-That is, the learner's task is to perform a joint inference over lexica and signaling behavior.
+More generally, we combine the replicator dynamics with iterated learning by codifying the latter as a transition matrix $Q$.
+Just as in standard mutator dynamics, $Q_{ij}$ indicates the probability of the children of a parent of type $i$ adopting type $j$.
+However, to make this process depend on a type's learnbility, this quantity is proportional to the probability of $i$ producing the learning data and that of $j$ given the data.
 
@@
 So far, we have argued for expressivity, learnability and simplicity as central pressures that apply on the cultural evolution of language.
-The dynamics we propose model them as fitness-relative replication, mutation based on iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings, respectively.
+These are modelled as fitness-relative replication, mutation based on iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings, respectively.
 Taken together the dynamics are described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:
@@
 The basic units that these pressures apply to are a combination of semantics, i.e., a lexicon, and its use.
-The expressivity of these types depends on their communicative efficiency within a population.
-Their learnability depends on the fidelity by which they can be infered by a new generation of learners based on production data.
-In particular, learners perform a joint inference on types of linguistic behavior and lexical meaning.
-This process is influenced by the learners' preference for simpler lexical representations, codified as a learning bias.
+Their expressivity depends on their communicative efficiency within a population, whereas their learnability depends on the fidelity by which they can be infered by a new generation of learners.
+Importantly, the latter is influenced by the learners' preference for simpler lexical representations, codified as a learning bias.
 
@@
 \section{Lack of semantic upper-bounds in lexical meaning}
-We set out to investigate the prevalence of lexical meanings that allow for regular pragmatic enrichments over other potential alternatives.
-A particularly well-studied type of conventional pragmatic enrichment are so-called {\em scalar implicatures} (cf.
-\citealt{horn:1972,gazdar:1979}).
+With this model, we set out to investigate the prevalence of lexical meanings that allow for regular pragmatic enrichments over other alternatives.
+A particularly well-studied type of conventional pragmatic enrichment are so-called {\em scalar implicatures}.
 These inferences are licensed for groups of expressions ordered in terms of informativity, here understood as an entailment induced order.
-
 For instance, {\em some} is entailed by {\em all}; if it were true that `All students came to class', it would also be true that `Some students came to class'.
 However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not necessarily what their use is taken to convey.
-Instead, the use of a less informative expression when a more informative one could have been used can license a defeasible inference that stronger alternatives do not hold.
-That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that, since the speaker did not use a stronger alternative ({\em all}), this alternative must not hold.
-In this way, `Some students came to class' is strengthened as conveying `Some but not all students came to class'.
+Instead, the use of a less informative expression when a more informative one could have been used can license a defeasible inference that stronger alternatives do not hold (cf.
+\citealt{horn:1972,gazdar:1979}).
+That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that, since the speaker did not use a stronger alternative, e.g.
+{\em all}, this alternative must not hold.
+In this way, `Some students came to class' is strengthened to convey `Some but not all students came to class'.
 Analogously, a speaker can rely on her interlocutor to draw this inference without having to express this upper-bound overtly, e.g.
 by stating {\em some but not all}.
-Put differently, mutual reasoning about rational language use supplies a bound that rules out stronger alternatives pragmatically.
+In other words, mutual reasoning about rational language use supplies a bound that rules out stronger alternatives pragmatically.
 
-Note that this corresponds to our previous description of the pragmatic use of lexicon $L_a$, repeated below for convenience.
-A pragmatic hearer who reasons about a speaker's use of message $m_1$ will associate it more strongly with $s_1$ than with $s_2$ given that the latter is unambiguously associated with the latter state.
-The strength of this association being dependent on their degree of rationality $\lambda$ and their prior over states.
-Conversely, a pragmatic speaker will reason about her interlocutors interpretation and use the messages accordingly.
+This corresponds to our previous description of the pragmatic use of lexicon $L_a$, repeated below for convenience.
+A pragmatic hearer who reasons about a speaker's use of message $m_1$ will associate it more strongly with $s_1$ than with $s_2$ given that the latter is unambiguously associated with $s_2$.
+The strength of this association dependens on the individuals' degree of rationality $\lambda$ and their prior over states.
+Conversely, a pragmatic speaker will reason about her interlocutor's interpretation and use the messages accordingly.
 
@@
 
-We can now rephrase our initial question in terms of scalar implicatures: Why is the lack of lexical upper-bounds in weak scalar alternatives, as $m_1$ in $L_a$, regularly selected for over the alternative of lexicalizing it as in $L_b$?
+Our initial question can now be rephrased in terms of scalar implicatures by asking for justifiations for the lack of lexical upper-bounds in weak scalar alternatives.
+That is, why semantics such as those of message $m_1$ in $L_a$ are regularly selected for over the alternative of lexicalizing it as in $L_b$.
 More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} or {\em big} were truth-conditionally incompatible with stronger alternatives such as, respectively, {\em hot}, {\em and}, {\em all} and {\em huge}?
-This question is particularly striking considering the number of classes of expressions that license such inferences across languages (\citealt{horn:1972}, \citealt[252-267]{horn:1984}, \citealt{traugott:2004,vdAuwera:2010}).
+This question is particularly striking considering the number of expressions that license such inferences across languages (\citealt{horn:1972}, \citealt[252-267]{horn:1984}, \citealt{traugott:2004,vdAuwera:2010}).
 
@@
 
-%\begin{figure}
-%\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{../presentations/07fitness}
-%\end{figure}
+\begin{figure}
+\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{./only-R-or-M}
+\end{figure}
 %

### 2016-09-19 presentations/edindata/plot-only-fitness+only-learning.py

@@ -75,12 +75,14 @@ d4 = t_finalB.loc[:,('t9_final','t10_final', 't11_final')]
 
 fig = plt.figure()
 
-ax1 = fig.add_subplot(1,2,1, aspect = "equal")
-ax2 = fig.add_subplot(1,2,2, aspect = "equal", sharey = ax1)  #Share y-axes with subplot 1
+#ax1 = fig.add_subplot(1,2,1, aspect = "equal")
+ax1 = fig.add_subplot(1,2,1, xlabel=r'Rationality parameter $\lambda$', ylabel='Proportion in population')
+
+ax2 = fig.add_subplot(1,2,2, sharey = ax1, xlabel='Prior parameter c')  #Share y-axes with subplot 1
 
 #Set y-ticks of subplot 2 invisible
 plt.setp(ax2.get_yticklabels(), visible=False)
-
+plt.tight_layout(pad=0.4, w_pad=0.5, h_pad=1.0)
 #Plot data
 im1 = ax1.plot(d2)
 im2 = ax2.plot(d4)

### 2016-09-20 paper/paper.tex
small modifications on paper-draft section 3

@@
 Our starting point is given by the overarching argument that has crystalized from accumulated mathematical, experimental and cross-linguistic evidence in this literature: Natural languages need to be well-adapted to communicative needs within a linguistic community, but also need to be learnable to survive their faithful transmission across generations.
-More succinctly; natural languages are pressured for expressiveness as well as learnability.
+More succinctly; natural languages are pressured for expressivity as well as learnability.
 
@@
 
-%The observation that monomorphemic expression that lexically rule out stronger alternatives (e.g.
-{not all} as hypothetical `nall') are unattested across languages has received substantial argumentative support (most prominently in \citealt[252-267]{horn:1984} but also e.g.
-in \citealt{horn:1972,traugott:2004,vdAuwera:2010}).
-To the best of our knowledge, this claim stands unchallenged.
-
 \subsection{Languages and linguistic behavior}
@@
 Analogously, should rational hearers expect their interlocutors to reason along these lines, they will interpret ambiguous $m_1$ accordingly.
-Note in particular that according to this strenghtening of $m_1$, $L_a$ is indistinguishable from $L_b$ in terms of expressiveness if its users are pragmatic reasoners.
+Note in particular that according to this strenghtening of $m_1$, $L_a$ is indistinguishable from $L_b$ in terms of expressivity if its users are pragmatic reasoners.
 
@@
 \subsection{Replication \& expressivity}
-Communicative efficiency, or expressiveness, has received particular attention from investigations using evolutionary game theory \citep{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}.
+Communicative efficiency, or expressivity, has received particular attention from investigations using evolutionary game theory \citep{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}.
 Under this view, a type's success in communication confers it a higher fitness relative to less successful ones.
@@
 \subsection{Summary}
-So far, we have argued for expressivity, learnability and simplicity as central pressures that apply on the cultural evolution of language.
-These are modelled as fitness-relative replication, mutation based on iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings, respectively.
-Taken together the dynamics are described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:
+We argued for expressivity, learnability and simplicity as important pressures that apply on the cultural evolution of language.
+They are respectively modelled as communicative efficiency-relative replication, iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings.
+Taken together these evolutionary dynamics are described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:
 
@@
 
-The basic units that these pressures apply to are a combination of semantics, i.e., a lexicon, and its use.
-Their expressivity depends on their communicative efficiency within a population, whereas their learnability depends on the fidelity by which they can be infered by a new generation of learners.
-Importantly, the latter is influenced by the learners' preference for simpler lexical representations, codified as a learning bias.
+The basic units that the dynamics operate on are a combination of lexica lexicon and their use; a type.
+A type's expressivity depends on its communicative efficiency within a population while its learnability depends on the fidelity by which it is infered by new generations of na\"ive learners.
 
-\hl{Summarize novelties \& advantages
-1. Advantage of connection between two strands of research
-2. Simple and general mean field dynamics}
-
-\hl{Possibly put reference to Kalish in EVOLANG7, Spain, on selection + iterated learning, and our CogSci}
+In sum, the innovation of this model lies in its combination of functional pressure on successful communication, effects of learning biases on (iterated) Bayesian language learning \citealt{griffiths+kalish:2007}, and probabilistic models of language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014,bergen+etal:2016}.
+In particular, this synthesis enables for the investigation of the effects of communicative pressures on the semantics-pragmatics interface.
+In doing so, it links the previously disconnected areas of rational probabilistic language use and cultural evoltion.
 
@@
 
-\section{Lack of semantic upper-bounds in lexical meaning}
+\section{Lack of lexicalization of pragmatic inferences}
+
+%
 With this model, we set out to investigate the prevalence of lexical meanings that allow for regular pragmatic enrichments over other alternatives.
@@
 
-\paragraph{Other hyp}
-\hl{The first is that pragmatic reasoning offers a general mechanism to strengthen the meaning of a wide range of expressions when the conditions outlined above hold.
-Consequently,  cases where cooperativity or  knowledge  are  not  likely  to  be  given  are  non-committal to whether stronger alternatives hold.
-If for all the speaker knows ‘some students came’ but she does not know whether ‘all came’, then the compatibility of some with (possibly) all
-succinctly conveys the speaker’s uncertainty about the latter.}
+We see two main explanations for the lack of upper-bounds in the lexical meaning of weak scalar expressions.
+The first is that their truth-conditional compatibility with stronger expressions lends them a broader range of application.
+When the speaker is unwilling or unable to assert a stronger alternative, e.g.
+because she lacks knowledge, weaker alternatives can still be used.
+That is, cases where cooperativity or knowlege are not likely to be given are non-commital to whether stronger alternatives hold.
+For instance, if for all the speaker knows `Some students came' but she doesn't know whether `All came', then the use of {\em some} succintly conveys her uncertainty about the latter.
+Thus, given that scalar expressions occur in contexts in which their upper-bounded reading is absent, one could argue for a functional advantage of a lack of semantic upper-bounds: should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then these means are preferred over expressing it overtly through a longer expression, e.g.
+by saying `some but not all' explicitly.
+Furthermore, not lexicalizing an upper-bound would also be prefered over the hypothetical alternative of lexicalizing two expressions -- one with and one lacking an upper-bound.\footnote{The observation that such monomorphemic expression that lexically rule out stronger alternatives are unattested across languages has received substantial argumentative support (most prominently in \citealt[252-267]{horn:1984} but also e.g.
+in \citealt{horn:1972,traugott:2004,vdAuwera:2010}).
+To the best of our knowledge, this claim stands unchallenged.}
 
-\hl{Given that scalar expressions occur in contexts in which their
-upper-bounded reading is absent, one could argue for a func-
-tional advantage of a lack of semantic upper-bounds:  If ex-
-pressing such a state of affairs is relevant and contextual cues
-provide enough information for a hearer to discern when a
-bound is conveyed pragmatically, then doing so is preferred
-over enforcing the bound overtly through a longer (more com-
-plex) expression, e.g. by stating ‘some but not all’ explicitly.
-That is, all else being equal, speakers prefer to communicate
-as economically as possible, and pragmatic reasoning enables
-them to do so.  Additionally, this can be contrasted with the
-hypothetical alternative of lexicalizing two expressions – one
-with and one lacking an upper-bound.  Four conditions may
-pressure language to English-like semantics over this alter-
-native: (i) contextual cues are very reliable, morphosyntactic
-disambiguation is either (ii) not frequently necessary or (iii)
-not  very  costly,  or  (iv)  having  larger  lexica  is  more  costly
-than morphosyntactic disambiguation.   In a nutshell (i) and
-(ii) place a heavy burden on the ability to retrieve contextual
-cues to a degree that is unlikely to undercut the benefit of safe
-communication with more expressions.  As for (iii) and (iv),
-these seem mostly like technical solutions without a proper
-empirical basis.}
+In a nutshell, this argument posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so.
+However, this explanation relies on cues being strongly reliable or morphosyntactic disambiguition being seldom necessary.
+These assumptions put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of safe communication.
+Two related conditions that might pressure language to English-like semantics are that either morphosyntactic disambiguation is not very costly on its own, or at least less costly than having larger lexica (in this case two for each scalar expression).
+However, these strike us as technical solutions without a proper empirical basis.
+
+\hl{See notes on Desktop and finish up saying something about probably being a factor, but we see }
 
@@
 At present, the contrast in compressability between lexical meanings, however, is not represented as lexica codify the truth-conditions of expressions at a lower level of granularity.
-In principle this difference could be made precise with an adequate representational language, e.g., through measures over representational complexity such as minimal description length.
-
-There is a growing effort to develop such
-empirically  testable  representational  languages.   For  instance,  the
-so-called
-language of thought
-has been put to test in various rational
-probabilistic models that show encouraging results (see e.g. Katz et
-al. 2008; Piantadosi et al. under review, 2012 and references therein).
-We think that our assumption is well-warranted as a working hypoth-
-esis and decide against such an enrichment given that the introduc-
-tion of a larger framework would also require further assumptions
-and justifications.
+\hl{In principle this difference could be made precise with an adequate representational language, e.g., through measures over representational complexity such as minimal description length.
+There is a growing effort to develop such empirically  testable  representational  languages.
+For  instance, the so-called language of thought has been put to test in various rational probabilistic models that show encouraging results (see e.g.
+Katz et al.
+2008; Piantadosi et al.
+under review, 2012 and references therein).
+We think that our assumption is well-warranted as a working hypothesis and decide against such an enrichment given that the introduction of a larger framework would also require further assumptions and justifications.}
 

### 2016-09-21 cogsci16-paper/cogsci-final-cameraready.tex
added cogsci files, extended draft section 3

%% Change "letterpaper" in the following line to "a4paper" if you must.

\documentclass[10pt,a4paper]{article}

\usepackage{cogsci}
\usepackage{pslatex}
\usepackage{apacite}
\usepackage{natbib}

\usepackage{color}
\usepackage[fleqn]{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{multicol}
\usepackage{graphicx}

\newcommand{\mvalueof}[1]{\llbracket#1\rrbracket}
\newcommand{\citeposs}[2][]{\citeauthor{#2}'s (\citeyear[#1]{#2})}
\newcommand{\tuple}[1]{\ensuremath{\left\langle #1 \right\rangle}}

\newcommand{\hl}[1]{\textcolor[rgb]{.8,.33,.0}{#1}}% prints in orange
\newcommand{\hlblue}[1]{\textcolor[rgb]{0.13,0.67,0.8}{#1}}
\newcommand{\argmax}[1]{\underset{#1}{\operatorname{arg}\,\operatorname{max}}\;}
\newcommand{\argmin}[1]{\underset{#1}{\operatorname{arg}\,\operatorname{min}}\;}
\newcommand{\sbna}{\exists\lnot\forall}

\title{Learning biases may prevent lexicalization of pragmatic inferences:\\ a case study combining iterated (Bayesian) learning and functional selection}

\author{{\large \bf Thomas Brochhagen (t.s.brochhagen@uva.nl)} \\
  Institute for Logic, Language and Computation, University of Amsterdam\\
  Science Park, Kruislaan 107, 1098 XG Amsterdam, The Netherlands
  \AND {\large \bf Michael Franke  (mchfranke@gmail.com)} \\
  Department of Linguistics, University of T\"ubingen\\
 Wilhelmstrasse 19, 72074 T\"ubingen, Germany 
    \AND {\large \bf Robert van Rooij (r.a.m.vanRooij@uva.nl)} \\
  Institute for Logic, Language and Computation, University of Amsterdam\\
Science Park, Kruislaan 107, 1098 XG Amsterdam, The Netherlands}

\begin{document}

\maketitle

\begin{abstract}
Natural languages exhibit properties that are difficult to explain from a purely functional perspective. One of these properties is the systematic lack of upper-bounds in the literal meaning of scalar expressions. This investigation addresses the development and selection of such semantics from a space of possible alternatives. To do so we put forward a model that integrates Bayesian learning into the replicator-mutator dynamics commonly used in evolutionary game theory. We argue this synthesis to provide a suitable and general model to analyze the dynamics involved in the use and transmission of language. Our results shed light on the semantics-pragmatics divide and show how a learning bias in tandem with functional pressure may prevent the lexicalization of pragmatic inferences.

\textbf{Keywords:} 
semantics; pragmatics; iterated learning; evolutionary game theory; scalar expressions
\end{abstract}

\section{Introduction}
Why are natural languages structured the way they are and not differently? In particular, what factors are involved in the selection of semantic structure from a space of possible alternatives? A number of recent studies have began to investigate such issues, pertaining to the development and emergence of linguistic features (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews). While different methodologies have been put forward to this end, the overarching argument is one of competing pressures: natural languages need to enable successful communication, as well as be suited for acquisition. That is, they need to be well-adapted to the communicative needs of their users but they also need to be learnable in order to survive their faithful transmission across generations.

The present investigation focuses on the interplay of such selective forces by means of a case study on the lexicalization of semantic upper-bounds. Using evolutionary game theory, framed in terms of language use and learning, we investigate the prevalence of a lack of upper-bounds in the literal meaning of scalar expressions. The innovation of the model lies in its combination of functional pressure on successful communication, effects of learning biases on (iterated) Bayesian language learning \citep{griffiths+kalish:2007}, and probabilistic models of language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014,bergen+etal:2015}. Our results show that a learning bias for simple semantic representations coupled with communal language exposure can lead to a prevalence of a lack of semantic upper-bounds provided bounds can be inferred via pragmatic reasoning. This, in turn, offers an explanation to why certain pragmatic inferences fail to lexicalize.

\section{Conveying and lexicalizing upper-bounds}
A considerably large class of natural language expressions do not lexicalize an upper-bound. That is, they are truth-conditionally compatible with more informative or ``stronger'' alternatives. Here, informativity is understood as an order over entailment. Examples in English include, among many others, numerals such as {\em five} and {\em six}, scalar adjectives like {\em cold} and {\em big}, as well as quantifiers like {\em some} and {\em many}. The commonality of these expressions lies in their literal meaning being compatible with the truth of stronger (relevant) alternatives. For example, if it is true that `Bill read five books' it may well be true that `Bill read six/seven/... books'. Crucially, the latter entails the former; if it is true that `some students came to class' then this would also hold if it were true that `all students came to class'. Prima facie this pattern may seem surprising. If it were true that `some (but not all) students came', then it would putatively serve interlocutors better if {\em some} semantically ruled out the stronger {\em all} case.

In practice, however, what is communicated often goes beyond the literal meaning of expressions. In particular, mutual reasoning between interlocutors can lead to the enrichment of semantic content \citep{grice:1975}. Such {\em pragmatic reasoning} is driven by interlocutors' mutual expectations of rational language use. In this case the use of a less informative expression when a more informative one could have been used can license a defeasible inference that the stronger alternative does not hold (cf. \citealt{horn:1972,gazdar:1979a}). In this way, a hearer who assumes the speaker to be knowledgeable and cooperative, i.e., able and willing to supply all relevant information at her disposition, can infer {\em some} to be pragmatically strengthened to convey an upper-bounded interpretation.

This kind of reasoning was already hinted at above: a speaker would have used the more informative alternative if it were true. Since she did not, her addressee can infer that it does not hold. Conversely, a speaker who reasons about her addressee can rely on her drawing this inference. Notwithstanding, while pragmatic reasoning provides an answer to how interlocutors may derive and convey upper-bounds, the question why they are not part of the literal meaning still stands. More generally, while a divide between semantics and pragmatics is commonly agreed upon by linguists, under consideration of purely functional pressures this raises the challenge of providing justifications for the former's structure.

We see two main explanations for a lack of semantic upper-bounds in scalar expressions. The first is that pragmatic reasoning offers a general mechanism to strengthen the meaning of a wide range of expressions when the conditions outlined above hold. Consequently, cases where cooperativity or knowledge are not likely to be given are non-committal to whether stronger alternatives hold. If for all the speaker knows `some students came' but she does not know whether `all came', then the compatibility of {\em some} with (possibly) {\em all} succinctly conveys the speaker's uncertainty about the latter. Given that scalar expressions occur in contexts in which their upper-bounded reading is absent, one could argue for a functional advantage of a lack of semantic upper-bounds: If expressing such a state of affairs is relevant and contextual cues provide enough information for a hearer to discern when a bound is conveyed pragmatically, then doing so is preferred over enforcing the bound overtly through a longer (more complex) expression, e.g. by stating `some but not all' explicitly. That is, all else being equal, speakers prefer to communicate as economically as possible, and pragmatic reasoning enables them to do so. Additionally, this can be contrasted with the hypothetical alternative of lexicalizing two expressions -- one with and one lacking an upper-bound. Four conditions may pressure language to English-like semantics over this alternative: (i) contextual cues are very reliable, morphosyntactic disambiguation is either (ii) not frequently necessary or (iii) not very costly, or (iv) having larger lexica is more costly than morphosyntactic disambiguation. In a nutshell (i) and (ii) place a heavy burden on the ability to retrieve contextual cues to a degree that is unlikely to undercut the benefit of safe communication with more expressions. As for (iii) and (iv), these seem mostly like technical solutions without a proper empirical basis.

A second explanation targets the contrast of the underlying semantic representation of upper-bounds and a lack thereof, positing a learning advantage for the latter. That is, by virtue of their simpler representations, expressions lacking upper-bounds are more readily learnable than their bounded counterparts. If a bound can be supplied pragmatically, this difference may explain the prevalence of such semantics. In the following we explore this hypothesis. While we do not want to argue that functional pressures may not play a role, we see a clear benefit in exploring whether learnability would not give us additional leverage. Furthermore, it should also be stressed that these explanations may well be complementary in that a full-fledged account can reasonably be expected to involve an interplay between them.

\section{Communication and selective dynamics}
%Broadly speaking, two intertwined factors will drive this process: communicative success and the learnability of a language. On the one hand, behaviors and languages that are communicatively successful prosper and, as a consequence, have a higher chance of propagation in following generations. On the other hand, the transmission of a language is affected by the players' learning bias and its aptitude to explain exposure to linguistic data. Neither component alone inevitably leads to the selection of a particular language. Instead, this framework provides a more nuanced view in which only their combination tells the full story.

We employ evolutionary game theory to investigate the selection dynamics of language fragments that capture the contrast between upper-bounded meanings and their unbounded counterparts. Evolutionary game theory offers general and precise means to model the dynamics of linguistic pressures \citep{nowak+krakauer:1999, huttegger+zollman:2013}. More concretely, our model uses the replicator-mutator dynamics (see \citealt{hofbauer+sigmund:2003} for an overview). In contrast to previous models we integrate (iterated) Bayesian learning in the dynamics \citep{griffiths+kalish:2007}, as well as probabilistic language users of varied degrees of pragmatic sophistication \citep{frank+goodman:2012,franke+jaeger:2014} together with multiple lexica \citep{bergen+etal:2015}. In this way the model synthesizes core insights of previous proposals of language use and learning.

Linguistic interactions are represented by signaling games \citep{lewis:1969}. An interaction involves two players; a speaker and a hearer. The speaker has some private information about the state of the world and tries to convey this information to the hearer by sending a message. The hearer receives the message and interprets it. Communication is successful if the message's interpretation matches the speaker's information, i.e., when the message is interpreted correctly. We restrict our attention to games with two states, $S = \{s_1,s_2\}$, and two messages, $M = \{m_1,m_2\}$. This allows for a minimal contrast between weaker (non-)upper-bounded expressions and stronger alternatives in states were a bound is to be conveyed. Players base their choices on lexica that specify the semantics of expressions. Formally, a lexicon $L$ is a Boolean $(|S|,|M|)$-matrix such as $L_{a}$ and $L_{b}$ below.\\

%\vspace{0.01cm}
\begin{centering}
$L_a$ = \bordermatrix{~ & m_1 & m_2 \cr 
                  s_1 & 0 & 1 \cr
                  s_2 & 1 & 0 \cr} \hspace{2cm} $L_b$ = \bordermatrix{~ & m_1 & m_2 \cr 
                  s_1 & 1 & 1 \cr
                  s_2 & 1 & 0 \cr}\\[0.5cm]
\end{centering}

According to $L_a$, message $m_1$ is true in $s_2$ and false in $s_1$. The converse holds for $m_2$. In $L_b$, however, $m_1$ is true in both $s_1$ and $s_2$. In this way $L_a$ stands for a language that encodes an upper-bound for $m_1$, e.g. {\em some but not all}, whereas $L_b$ stands for one where $m_1$ lacks such a bound, e.g. {\em some}.

To contrast literal language use to its pragmatic counterpart we consider two kinds of players. {\em Literal players} communicate according to the semantics of their lexica. {\em Pragmatic players} reason about their interlocutors and base their choices on this reasoning. In other words, pragmatic speakers take into account how listeners would interpret a message and pragmatic listeners take the speaker into account when interpreting. This kind of signaling behavior shares the core features common to models of rational language use such as Rational Speech Act models \citep{frank+goodman:2012} and Quantal- and Best-Response models \citep{franke:2009,franke+jaeger:2014}. Following this line of research, player behavior can be captured by a hierarchy of reasoning types. Literal players constitute the bottom of the hierarchy, level $0$, and are solely guided by the semantics of their language. Players of level $n+1$ behave rationally according to level $n$ behavior of their interlocutors. Presently, it suffices to consider players of no order higher than $1$ as the cases considered here offer little room for further pragmatic refinement. Literal hearer and speaker behavior are summarized in (\ref{litl}) and (\ref{lits}), and their pragmatic counterparts in (\ref{pragl}) and (\ref{prags}).
\vspace{-0.15cm}
\begin{flalign}
&R_{0}(s|m;L) \propto P^*(s) L_{sm}\label{litl}\\
&S_{0}(m|s;L) \propto \exp(\lambda \; L_{sm}) \label{lits}\\
&R_{1}(s|m;L) \propto P^*(s) S_{0}(m|s;L) \label{pragl}\\
&S_{1}(m|s;L) \propto  \exp(\lambda \; R_{0}(s|m;L)^\alpha) \label{prags}
\end{flalign}

Speakers strive to maximize their communicative success but may occasionally make mistakes in computing their expected utility. This is regulated by the soft-max parameter $\lambda$, $\lambda > 0$ \citep{luce:1959, sutton+barto:1998}. Intuitively, the larger $\lambda$ the more faithful a speaker's choices are to the maximization of expected success. A second parameter $\alpha$, $\alpha \in [0,1]$, controls the tension between semantics and pragmatics. Lower $\alpha$-values lead to more literal signaling whereas larger values lead to stronger pragmatic behavior. $P^* \in \Delta(S)$ is a prior over states. For simplicity in the following we assume this prior to be common and uniform. Lastly, a {\em player type} is a combination of signaling behavior, i.e., either literal or pragmatic, and a lexicon.

With these player types at hand population-level dynamics can be analyzed. The dynamics involve two key components: communicative fitness and lexicon learnability. The fitness of a type $i$ is given by its expected utility. That is, the fitness of type $i$, $f_i$, in a population $x$ is the sum of its weighted expected utility, $f_i = \sum_j x_j U(x_i,x_j)$, where $x_j$ is the proportion of players of type $j$ in $x$ and $U(x_i,x_j)$ is the symmetrized expected utility of $x_i$ and $x_j$.\footnote{$U(x_i,x_j) = [U_S(x_i,x_j) + U_R(x_i,x_j)]/2$. $U_S(x_i,x_j)$ and $U_R(x_i,x_j)$, the expected utility of speaker type $x_i$ and hearer type $x_j$ and vice versa, are respectively $\sum_s P^*(s) \sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$ and $U_S(x_j,x_i)$ for $n$ and $o$ being the reasoning level of $i$ and $j$, and $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.}  A type's fitness indicates how well it communicates within a population. The average fitness of the population is  $\Phi$, $\Phi = \sum_i x_i f_i$.

The second component is given by a learning matrix $Q$, which adds (iterated) Bayesian learning to the dynamics \citep{griffiths+kalish:2007}. It operationalizes the assumption that new generations of learners combine prior learning biases with observations over language use. This determines the probability of adopting a type. As a consequence, lexica that better explain observations are more likely to be passed onto the next generation faithfully. Here we assume that the adoption of a lexicon is not solely dependent on how well a parent's lexicon explains the data but also on its proportional representation in the population. The intuition is that it may be possible that a lexicon that is -- in principle -- well-suited for acquisition may nevertheless not be adopted when only few types of the preceding population used it. More concretely, $Q_{ji}$ specifies the probability that a child of type $j$ will be of type $i$. This is weighted by the probability of being exposed to type $i$, $N_{ji} = x'_i Q_{ji}$ where $x'_i$ is proportion of $i$ in the previous generation. $Q$ itself depends on observations of linguistic utterances and a learning prior. The set of possible observations is given by all combinations between state-message pairs, $O =  \{\tuple{\tuple{s_1,m_i},\tuple{s_2,m_j}} | m_i, m_j \in M\}$. A member of $O$ encodes that a teacher produced $m_i$ in state $s_1$ and $m_j$ in $s_2$. It encodes one witnessed message for each state. A datum $d$ is a sequence of length $k$ of members of $O$. Learners witness such data sequences. Accordingly, $Q_{ji} \propto \sum_d P(d|t_j) P(t_i|d)$ with $P(t_i|d) \propto P(t_i)P(d|t_i)$. The prior  $P(t)$ gives the probability of a type without witnessing data, i.e., it encodes the learning bias of players prior to linguistic exposure. The likelihood $P(d|t)$ gives the probability of data being produced by a type. The posterior thusly yields a combination of prior expectations and data that captures a type's fit to the data. This, in turn, regulates the transmission of its lexicon. Putting these components together, the discrete selection dynamics are captured by $\hat{x}_i = \sum_j N_{ji} \frac{x_j f_j}{\Phi}$ (cf. \citealt{hofbauer+sigmund:2003}).

As mentioned above, the prior captures the learning bias of players. For simplicity we assume it to be solely dependent on lexica and not on a type's signaling behavior. In particular it biases learners for simpler semantic representations over more complex ones. In the following the prior encodes the intuition that the semantic representation of an upper-bound is more complex than a lack thereof. Since semantics are only implicitly represented through a lexicon's Boolean matrix, the bias is regulated by a parameter $c$ that operates over cells of a lexicon.\footnote{In principle this contrast could be made more precise with an adequate representational language, e.g., through measures over representational complexity. There is a growing effort to develop such empirically testable representational languages. For instance, the so-called {\em language of thought} has been put  to test in various rational probabilistic models that show encouraging results (see e.g. \citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012a} and references therein). We think that our assumption is well-warranted as a working hypothesis and decide against such an enrichment given that the introduction of a larger framework would also require further assumptions and justifications.} As illustrated above by lexica $L_a$ and $L_b$, we let $s_1$ stand for a ``some but not all''-state and $s_2$ for an ``all''-state. Accordingly, the prior biases players against lexica in which a message $m$ holds true only of the former and not the latter, as in $L_a$. The literal meaning of English {\em some} corresponds to a message true of $s_1$ and $s_2$ in such a fragment. All other semantics are a priori equally probable. Thus, $P(t_i) \propto n - c \cdot r$ where $n$ is the total number of states and $r$ that of upper-bounded messages only true of $s_1$ in $t_i$'s lexicon.

\section{Analysis}
We consider a total of $12$ player types, obtained by pairing six lexica with literal and pragmatic signaling behavior. The lexica are specified in Table \ref{tab:lexica}. Lexica $L_1$ to $L_3$ are not optimal for communication because they assign the same meaning to all their messages. They were included to showcase the selection process for a larger hypothesis space. $L_4$ and $L_5$ are the crucial lexica that represent upper-bounded semantics and a lack thereof, respectively. Lastly, $L_6$ is similar to $L_5$ in that two messages are true of the same state but differs from it in assigning upper-bounded semantics to $m_2$.

In the following we focus our analysis on the contrast between literal and pragmatic types using lexica $L_4$ and $L_5$. Note in particular that the linguistic behavior of pragmatic $L_4$ and $L_5$ can be close to indistinguishable as both are able to convey upper-bounds. The former semantically, the latter pragmatically. Depending on the parameters, however, there might be slight differences between the probability with which speakers would (erroneously) use a semantically false description and that they would (erroneously) use a pragmatically suboptimal description. This contrasts with literal $L_5$, which lacks the means to convey an upper-bound by $m_2$. Differences between pragmatic $L_4$ and $L_5$ are expected to mainly depend on the learning bias. Things are less clear for literal $L_5$ contrasted with literal/pragmatic $L_4$. The former has a learning advantage but is expected to fare worse in terms of communicative fitness in virtue of ambiguous $m_2$.

\begin{table}
\centering 
\begin{tabular}{l c l}
$L_1$ = $\begin{pmatrix} 0 & 0 \\ 1 & 1 \end{pmatrix}$ & 
$L_2$ = $\begin{pmatrix} 1 & 1 \\ 0 & 0 \end{pmatrix}$ & 
$L_3$ = $\begin{pmatrix} 1 & 1 \\ 1 & 1 \end{pmatrix}$\\[0.5cm]

$L_4$ = $\begin{pmatrix} 0 & 1 \\ 1 & 0 \end{pmatrix}$ &
$L_5$ = $\begin{pmatrix} 0 & 1 \\ 1 & 1 \end{pmatrix}$ &
$L_6$ = $\begin{pmatrix} 1 & 1 \\ 1 & 0 \end{pmatrix}$

\end{tabular}

\caption{Space of possible lexicon fragments considered.}
\label{tab:lexica}
\end{table}

The learning matrix $Q$ was computed with data sequences of length $20$. Pilot simulations showed that changes in sequence length influence the population in a predictable way: smaller values lead to more heterogeneous populations whereas larger ones lead to more pronounced differences. This is expected insofar as the likelihood that a sequence of length $1$ was produced by any type is relatively uniform (modulo prior) whereas the likelihood of types with lexica $L_1$ - $L_3$ to produce, for instance, a sequence of $10$ observations consistently with the same state-message combination is less likely than for pragmatic players using $L_4$ - $L_6$ or literal $L_4$. Thus, while noteworthy, sequence length has no direct bearing on the main contrast of interest. Similar considerations hold for $\alpha$ and $\lambda$ -- set to $1$ and $50$ in the following. Overall, lower rationality in $\lambda$ or more pragmatic violations in $\alpha$ lead to a higher selection of lexica with semantic upper-bounds. The fitness of pragmatic behavior increases with higher $\lambda$-/$\alpha$-values. In other words, these parameters level the functional contrast between $L_4$ and $L_5$. Due to space restrictions we leave a more detailed analysis of their interaction to future research and focus on the learning bias instead.

All simulations were initialized with an arbitrary distribution over types, constituting the population's first generation, and were run for $20$ generations. This corresponds to a developmental plateau after which no noteworthy change was registered.

\vspace{-0.2cm}

\paragraph{Results.} We report the mean results obtained from $1000$ independent simulations for each set of parameters. Figure \ref{fig:c-85} illustrates the development of the target types over $20$ generations with $c = 0.85$, i.e., a strong bias against semantic complexity. Figure \ref{fig:cost} shows the effect of the learning bias after $20$ generations for $c \in [0,1]$. More detailed results for all types across a sample of $c$-values are presented in Table \ref{tab:numeric-results}.
\begin{figure}[t] %htb for exact position
	\centering
	\resizebox{9.5cm}{!}{\includegraphics{figure1}}
	\caption{Mean proportions of target types across $20$ generations of $1000$ populations ($\alpha = 1, \lambda = 50, k = 20, c = 0.85$).}
	\label{fig:c-85}
\end{figure}

\begin{figure}[t] %htb for exact position
	\centering
	\resizebox{9.5cm}{!}{\includegraphics{figure2-cost}}
	\caption{Mean proportion of pragmatic $L_4$ and $L_5$ types after $20$ generations with $c \in [0,1]$ in $1000$ populations ($\alpha = 1, \lambda = 50, k = 20$).}
	\label{fig:cost}
\end{figure}

\begin{table}
\centering 
\begin{tabular}{l c c c c c c c}
~ & 0 & .01 & 0.03 & 0.05 & 0.07 & 0.09 & 0.85\\ \hline \hline
lit. $L_1$& .001 & .002 & .002& .004&.004& .002 & .004\\
lit. $L_2$& .001 & .002 & .001& .001&.001&0& 0\\
lit. $L_3$& .001 & .003 & .001& .002&.002& .001& .004\\
lit. $L_4$& .193 & .193 & .165& .177&.187&.142& .022\\
lit. $L_5$& .022 & .022 & .039& .04&.033&.049& .042\\
lit. $L_6$& .023 & .023 & .028& .027&.028&.025& .002\\ \hline
prag. $L_1$ & .001 & .001 &.001& .002&.004&.003& .005 \\
prag. $L_2$ & .001 & .001 &.001& .001&.001&0 & 0 \\
prag. $L_3$ & .001 & .002 &.002& .002&.004&.002& .006 \\ 
prag. $L_4$ & .257 & .234 &.23& .209&.205&.208& 0 \\
prag. $L_5$ & .249 & .26 & .29& .328& .329& .379& .914 \\
prag. $L_6$ & .25 & .237 & .24& .207& .202&.188& 0

\end{tabular}
\caption{Mean proportion of types in $1000$ populations after $20$ generations across different bias values $c$ ($\alpha = 1, \lambda = 50, k = 20$).}

\label{tab:numeric-results}
\end{table}

In sum, the results show that in the present setup a weak bias is sufficient to lead to a selection of $L_5$ over $L_4$. This effect increases with the bias' strength provided $L_5$ users are pragmatic. That is, such a learning bias may prevent the lexicalization of pragmatic inferences and explain the prevalence of $L_5$-like semantics. However, note that literal $L_5$ is underrepresented across all values of $c$. This highlights that, while the bias plays a major role for the contrast between $L_4$ and $L_5$, in itself it does not enable types that fail to convey an upper-bound to establish themselves in the population. This suggests pragmatic reasoning to be paramount to the selection of $L_5$-like semantics. That is, neither the learning bias nor functional pressure alone but their combination lead to the systematic lack of semantic upper-bounds in scalar expressions.

\section{Discussion}
Broadly speaking these results confirm our expectations that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome selective pressures and stabilize in a population provided there is a bias for simpler representations. This outcome is particularly encouraging in light of the potential additional advantages of a lack of semantic upper-bounds discussed above that were not considered here.

The model gives a justification for lexicalization patterns found in natural language, as well as for the failure to lexicalize certain pragmatic inferences. That is, while the puzzle raised by semantics is hard to explain by purely functional means, we suggest part of the answer to lie in learnability. Simpler semantic representations are more likely to be learned, and pragmatic reasoning can counteract functional disadvantages  otherwise incurred. This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics and offers an account of why (certain) pragmatic inferences are not part of the literal meaning of expressions. It furthermore leaves open the possibility of such inferences to fossilize when they do not compete against a lexical simplicity bias.

A virtue of this model is that it allows for analysis specific modifications and extensions. Straightforward extensions include larger hypothesis spaces, as well as larger or different lexicon fragments. Another worthwhile modification relates to possible disadvantages of pragmatic reasoning. We tacitly assumed pragmatic reasoning to come at no cost. However, there is experimental evidence for the assumption that the pragmatic derivation of upper-bounds costs effort and takes additional processing time (cf. \citealt{deNeys+schaeken:2007, huang+snedeker:2009}). This raises the question at which point such usage-based cost undercuts the learnability advantage of simpler semantic representations.\footnote{In the present setup this modification has a straightforward effect: A penalty for pragmatic signaling lowers the fitness of pragmatic types, to the advantage of literal types. However, the penalty needs to be substantial to counteract the functional advantage pragmatic $L_5$ has over all but $L_4$ together with its learning advantage.}

\begin{figure}[t] %htb for exact position
	\centering
	\resizebox{9.5cm}{!}{\includegraphics{figure3-parental}}
	\caption{Mean proportion of pragmatic $L_4$ and $L_5$ types after $20$ generations of parental teaching with $c \in [0,1]$ in $1000$ populations ($\alpha = 1, \lambda = 50, k = 20$).}
	\label{fig:figure3-parental}
\end{figure}

Lastly, a crucial component of this model is the learning matrix $Q$ together with its weighted variant $N$. The latter describes ``communal teaching'' where learning a lexicon is influenced by its proportional representation in the preceding generation. This diverges from the mean field dynamics of classical (dynamic) evolutionary game theory and iterated learning. Replacing the weighted matrix $N$ by $Q$, i.e. not weighting the learning matrix, allows for an inspection of the model's predicted outcome for a more classical case of ``parental teaching''. Figure \ref{fig:figure3-parental} shows the effect of the learning bias for this case (cf. Figure \ref{fig:cost}).

Note first that the main result of the prior driving selection of $L_5$ over $L_4$ holds here as well. However, the resulting populations are more heterogeneous than in the communal setting.  The intuition behind communal learning was that learning is influenced by the proportion in which languages are used in a population. However, a stage of exposure to communal language use is currently not represented in the model but only implicit in the weights applied to the Bayesian learning matrix. This step may be regarded as conceptually problematic given that it assumes ``implicit exposure'' to the true distribution of types in the population. Notwithstanding, we see it as a first step in the right direction to capture what constitutes an important factor in learning pressures. Further, it retains but strengthens the results obtained from a more standard learning setup. It is nevertheless important to highlight the weaker results obtained from parental learning as they suggest a missing piece to our overall account. Ideally one would expect the dynamics to favor a well-adapted lexicon over time even when only learning from parents, offering a stronger justification for the advantage of a lexicon without a population's involvement in the learning process. It is possible that a more complex setup where the speaker wants to convey not only bounded but also upper-unbounded states may shed light on this issue, as discussed in \S 2. Alternatively, a similarity bias over lexica may enable for a gradual adoption of particular languages.  A more involved analysis and comparison between static learning matrices and weighted variants, as well as an explicit representation of exposure to linguistic data from the population raise many interesting technical and conceptual challenges we hope to see addressed in the future.

\section{Conclusion}
The development of natural languages is driven by complex intertwined pressures. Drawing from past insights we put forward a model that allows for a general and malleable integration of core aspects involved in this process. Chiefly, the model combines functional pressure, iterated Bayesian learning, and probabilistic speaker and hearer models.  In particular, this analysis addresses longstanding issues concerning the semantics-pragmatics divide. It shows that when pressured for learnability and expressivity, the former force drives for simpler semantic representations inasmuch as pragmatics can compensate for them in language use. Consequently, semantic patterns can be explained in virtue of the linguistic behavior of their users and their complexity. Furthermore, the ease of acquisition of simpler semantics offers an answer to why natural languages do not lexicalize certain pragmatic inferences.

\section{Acknowledgments}
TB acknowledges funding from the European Community's Seventh Framework Programme (FP7/2007-2013) under grant agreement no. 567652 /{\em ESSENCE}/. MF gratefully acknowledges financial support by the Institutional Strategy of the University of T\"ubingen (Deutsche Forschungsgemeinschaft, ZUK 63). RvR acknowledges support by the `Language in Interaction' Gravitation project, funded by NWO.

%\bibliographystyle{unsrtnat}

\bibliographystyle{apacite}

\setlength{\bibleftmargin}{.125in}
\setlength{\bibindent}{-\bibleftmargin}
\bibliography{./cogsci-bib}

\end{document}

### 2016-09-21 paper/paper.tex
added cogsci files, extended draft section 3

@@
 We see two main explanations for the lack of upper-bounds in the lexical meaning of weak scalar expressions.
-The first is that their truth-conditional compatibility with stronger expressions lends them a broader range of application.
-When the speaker is unwilling or unable to assert a stronger alternative, e.g.
-because she lacks knowledge, weaker alternatives can still be used.
-That is, cases where cooperativity or knowlege are not likely to be given are non-commital to whether stronger alternatives hold.
-For instance, if for all the speaker knows `Some students came' but she doesn't know whether `All came', then the use of {\em some} succintly conveys her uncertainty about the latter.
-Thus, given that scalar expressions occur in contexts in which their upper-bounded reading is absent, one could argue for a functional advantage of a lack of semantic upper-bounds: should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then these means are preferred over expressing it overtly through a longer expression, e.g.
-by saying `some but not all' explicitly.
-Furthermore, not lexicalizing an upper-bound would also be prefered over the hypothetical alternative of lexicalizing two expressions -- one with and one lacking an upper-bound.\footnote{The observation that such monomorphemic expression that lexically rule out stronger alternatives are unattested across languages has received substantial argumentative support (most prominently in \citealt[252-267]{horn:1984} but also e.g.
+The first is that their truth-conditional compatibility with stronger expressions endows them a broader range of application.
+Scalar expressions occur in contexts in which their upper-bounded reading is absent.
+This can happen when embedded in downward-entailing contexts, when the speaker is likely uncertain about whether the upper bounded reading is true, or when the distinction between an upper-bounded reading and the simple, only lower-bounded reading, is not relevant.
+For instance, if for all the speaker knows `Some students came' but she doesn't know whether `All came', then the use of {\em some} succinctly conveys her uncertainty about the latter.
+This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then these means are preferred over expressing it overtly through a longer expression, e.g., by stating `some but not all' explicitly.
+Importantly, although dispreferred due to its relative length and complexity, morphosyntactic disambiguation still allows speakers to enforce an upper-bound to override contextual cues that might otherwise mislead the hearer.
+
+In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so.
+Compare this with a hypothetical language lexicalizes two expressions instead of a single scalar one; one with and one lacking an upper-bound.\footnote{The observation that monomorphemic expression that lexically rule out stronger alternatives are unattested across languages has received substantial argumentative support (most prominently in \citealt[252-267]{horn:1984} but also e.g.
 in \citealt{horn:1972,traugott:2004,vdAuwera:2010}).
-To the best of our knowledge, this claim stands unchallenged.}
+To the best of our knowledge, this claim stands unchallenged.} Should the explanation rest on purely functional grounds, we see four conditions that may pressure languages for English-like semantics over this alternative.
+First, contextual cues are strongly reliable.
+Second, morphosyntactic disambiguation is seldom necessary.
+Third, morphosyntactic disambiguation is only marginally dispreferred.
+Fourth, larger lexica are costly.
 
-In a nutshell, this argument posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so.
-However, this explanation relies on cues being strongly reliable or morphosyntactic disambiguition being seldom necessary.
-These assumptions put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of safe communication.
-Two related conditions that might pressure language to English-like semantics are that either morphosyntactic disambiguation is not very costly on its own, or at least less costly than having larger lexica (in this case two for each scalar expression).
-However, these strike us as technical solutions without a proper empirical basis.
+However, these conditions are not convincing in their role as central explanatory devices for such a wide-spread phenomenon.
+The first two put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of .
+It is likely that human language users are very good at retrieving cues from contexts, but to stipulate that they are so good as to undercut the benefit of safe communication provided by the hypothetical alternative strikes us as too strong of an assumption.
+As for conditions (3) and (4), these seem mostly like technical solutions, without a proper empirical basis.
 
-\hl{See notes on Desktop and finish up saying something about probably being a factor, but we see }
+Instead, in what follows we investigate the hypothesis that the lack of lexicalization of scalar inferences is driven by the advantage in compression that lexical meanings lacking an upper-bound have over those that explicitly codify it.
+Note however that we do not represent this contrast in compression between lexical meanings explicitly in lexica.
+Instead, the bias towards a lack of upper-bounds in weak scalar alternatives is directly encoded in the learners' prior over types.
 
-\paragraph{Our hyp}
-In what follows we investigate the hypothesis that this preference is driven by the advantage in compression that lexical meanings lacking an upper-bound have over those that explicitly codify it.
-
-At present, the contrast in compressability between lexical meanings, however, is not represented as lexica codify the truth-conditions of expressions at a lower level of granularity.
-\hl{In principle this difference could be made precise with an adequate representational language, e.g., through measures over representational complexity such as minimal description length.
+ In principle this difference could be made precise with an adequate representational language, e.g., through measures over representational complexity such as minimal description length.
 There is a growing effort to develop such empirically  testable  representational  languages.
 For  instance, the so-called language of thought has been put to test in various rational probabilistic models that show encouraging results (see e.g.
-Katz et al.
-2008; Piantadosi et al.
-under review, 2012 and references therein).
-We think that our assumption is well-warranted as a working hypothesis and decide against such an enrichment given that the introduction of a larger framework would also require further assumptions and justifications.}
+\citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012} and references therein).
+We think that our assumption is well-warranted as a working hypothesis and decide against such an enrichment given that the introduction of a larger framework would also require further assumptions and justifications.\footnote{\hl{T: We could possibly show the length difference of the lexical meanings with and without an upper-bound using a LOT grammar in the appendix.
+I don't know if it is worth it though.}}
 
-\subsection{Application}
+In sum, while we do not want to argue that functionalist pressure may not play a role, we do see a clear benefit in exploring whether matters of learnability would not give us additional leverage.
 
-\paragraph{Procedural description.} The game is initialized with some arbitrary distribution over player types.
-At the game's onset we compute $Q$ once based on the sets  of sequences $D$ (one for each parent type).
-Replicator dynamics are computed based on the fitness of each type in the current population as usual.
-$Q$ is computed anew for each independent run (of $g$ generations) given that it depends on $D$, which is sampled from production probabilities.
-
-\paragraph{Languages.} We consider a population of players with two signaling behaviors, literal and Gricean (level $0$ and $1$ below), each equipped with one of $6$ lexicons.
-This yields a total of $12$ distinct player types $t \in T$.
-$|M| = |S| = 2$, i.e., a lexicon is a $(2,2)$-matrix.
+\subsection{Preliminaries}
+The dynamics are initialized with an arbitrary distribution over types.
+We consider populations with two signaling behaviors, literal or Gricean (level $0$ and $1$ reasoners), each equipped with one of $6$ lexicons.
+This yields a total of $12$ distinct types $t \in T$.
+As in our examples above, $|M| = |S| = 2$, i.e., lexica are $(2,2)$-matrices.
 These are listed in Table \ref{tab:lexica}.
 
-\begin{table}[h]
+\begin{table}[t]
 \centering 
@@
 
-As in the CogSci paper, $L_4$ (semantic upper-bound for $m_2$) and $L_5$ (no semantic upper-bound for $m_2$) are the target lexica.
-Gricean $L_5$ users can convey/infer the bound pragmatically, while literal/Gricean $L_4$ users do so semantically.
+Target lexica....
+
+A language that has conventionalized upper and lower bounds to realize a (quasi-)partition of the relevant semantic space will produce speaker behavior that is /almost/ indistinguishable from that of a language with only the respective upper bounds, but with Gricean speakers.
+Almost, because there may be slight differences between the probability with which speakers would (erroneously) use a semantically false description and the probability with which speakers would (erroneously) use a pragmatically suboptimal description.
+\hl{explain why only these and not others}
+
+\begin{table}
+\centering
+\begin{tabular}{l|l|l}
+    \multicolumn{1}{c}{parameter} & \multicolumn{1}{c}{explanation} & \multicolumn{1}{c}{locus}\\ \hline
+    $\lambda \geq 1$ & rationality parameter & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\
+    $\alpha \in (0,1)$ & semantics-pragmatics tension & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\ 
+    $|D|$ & learning data produced per parent type & $P(d|t_j)P(t_i|d)$\\
+    $k = |d|$ & number of observations per datum& $P(d|t_j)P(t_i|d)$\\
+    $l \geq 1$ & posterior parameter from sampling to MAP & $P(t_i|d) \propto [P(t_i)P(d|t_i)]^l$\\
+    $c \in (0,1]$ & learning bias for upper-bound lack &  $P(t_i)$
+\end{tabular}
+\end{table}
+
+\paragraph{Procedural description.} At the game's onset we compute $Q$ once based on the sets  of sequences $D$ (one for each parent type).
+Replicator dynamics are computed based on the fitness of each type in the current population as usual.
+$Q$ is computed anew for each independent run (of $g$ generations) given that it depends on $D$, which is sampled from production probabilities.
 
@@
 \subsection{Analysis}
-\hl{Possibly put only fitness and only learning in a single plot with shared y axis}
-
 \begin{figure}
-\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{./only-R-or-M}
+\centering
+\includegraphics[scale=.5]{./only-R-or-M}
 \end{figure}
-%
-%\begin{figure}
-%\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{../presentations/07learning}
-%\end{figure}
-
 \begin{figure}
-\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{../presentations/03cost-with-l1}
+\centering
+\includegraphics[scale=.5]{../presentations/03cost-with-l1}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{../presentations/01heatmap}
+\centering
+\includegraphics[scale=.5]{../presentations/01heatmap}
 \end{figure}
 
-\section{Discussion}
+\subsection{Discussion}
 
-\section{Extensions}
+- There is experimental evidence that scalar implicature calculation costs effort and takes additional processing time (Schaeken, Snedeker, ...).
+
+\section{General discussion}
+
+The model combines:
+\begin{itemize}
+  \item game-theoretical models of functional pressure towards efficient communication
+  \item effects of learning biases on (iterated) language learning
+  \item assumptions about particular learning biases (Piantadosi et al. 20XX)
+  \item probabilistic speaker and listener types of various pragmatic sophistication (Frank \& Goodman, 2012; Franke \& Jäger, 2014)
+  \item speaker and listener types with different lexica (Bergen et al. 2012, to appear)
+\end{itemize}
+
+Discussion about expressivity as external to learning (cf. Stadler, replicator-papers by Kenny Smith, Kirby et al 2015). Possibly add appendix with direct comparison between IL and RMD.
+
+{\bf Extensions:}
+
 \subparagraph{(I) Cost for pragmatic reasoning.} At least in the CogSci setup the effect of adding cost to pragmatic reasoning is unsurprising: High cost for pragmatic signaling lowers the prevalence of pragmatic types.

### 2016-09-23 paper/paper.tex
paper section 4 draft on only pressure for expressivity/learnability

@@
 In contrast, {\em pragmatic interlocutors} engage in mutual reasoning to inform their choices.
-For instance, a rational speaker of $L_a$ who reasons about her addressee should use $m_1$ to signal state $s_1$ given that $m_2$ is already unambiguously associated with state $s_2$.
+For instance, a rational speaker of $L_a$ who reasons about her addressee should use $m_1$ to signal state $s_1$ given that $s_2$ can unambiguously be conveyed with $m_2$.
 Analogously, should rational hearers expect their interlocutors to reason along these lines, they will interpret ambiguous $m_1$ accordingly.
@@
 
-\subsection{Preliminaries}
-The dynamics are initialized with an arbitrary distribution over types.
-We consider populations with two signaling behaviors, literal or Gricean (level $0$ and $1$ reasoners), each equipped with one of $6$ lexicons.
-This yields a total of $12$ distinct types $t \in T$.
-As in our examples above, $|M| = |S| = 2$, i.e., lexica are $(2,2)$-matrices.
-These are listed in Table \ref{tab:lexica}.
+\subsection{Setup}
+ We consider populations with two signaling behaviors, either literal or pragmatic.
+As mentioned earlier, the former correspond to level $0$ reasoners who only take their lexica into consideration.
+In the following, pragmatic types correspond to level $1$ reasoners since higher level reasoning is not required to derive scalar implicatures from the lexica fragments we consider here.
+The prior over states is assumed to be uniform for all behaviors.
+
+The lexica are listed in Table \ref{tab:lexica}.
+As in our previous examples, they are $(2,2)$-Boolean matrices, i.e., $|M| = |S| = 2$.
+Lexica $L_1$ to $L_3$ are not optimal for communication because they assign the same meaning to all their messages.
+They were included to showcase the selection process for a larger hypothesis space.
+$L_4$ and $L_5$ are our target lexica, codifying upper-bounded semantics for message $m_1$ (the first column in a lexicon) and a lack thereof, respectively.
+Lastly, $L_6$ is similar to $L_5$ in that two messages are true of the same state but differs from it in assigning upper-bounded semantics to $m_1$.
+Combining a signaling behavior with one of these $6$ lexica yields a total of $12$ distinct types.\footnote{While there is a total of $16$ possible $(2,2)$-matrices, a number of them are identical both in terms of expressivity and the learning bias against lexical upper-bounds.
+The competition between such types is determined by their  proportions in the initial population but this fact can be obscured when averaging across simulations.
+We therefore focus only on the subset that exhibits the properties we set out to explain.}
 
@@
 
-$L_4$ = $\begin{pmatrix} 0 & 1 \\ 1 & 0 \end{pmatrix}$ &
-$L_5$ = $\begin{pmatrix} 0 & 1 \\ 1 & 1 \end{pmatrix}$ &
-$L_6$ = $\begin{pmatrix} 1 & 1 \\ 1 & 0 \end{pmatrix}$
+$L_4$ = $\begin{pmatrix} 1 & 0 \\ 0 & 1 \end{pmatrix}$ &
+$L_5$ = $\begin{pmatrix} 1 & 0 \\ 1 & 1 \end{pmatrix}$ &
+$L_6$ = $\begin{pmatrix} 1 & 1 \\ 0 & 1 \end{pmatrix}$
 \end{tabular}
-\caption{{\footnotesize Set of considered lexica.}}
+\caption{Space of possible lexicon fragments considered.}
 \label{tab:lexica}
@@
 
-Target lexica....
+We focus our analysis on the contrast between literal and pragmatic types using lexica $L_4$ and $L_5$.
+Note in particular  that a type that has conventionalized upper and lower bounds to realize a (quasi-)partition of the relevant semantic space, such as $L_4$, will produce speaker behavior that is {\em almost} indistinguishable from that of a language with only the respective upper bounds, but with Gricean speakers, such as $L_5$.
+Almost, because there may be slight differences between the probability with which speakers would (erroneously) use a semantically false description and the probability with which speakers would (erroneously) use a pragmatically suboptimal description.
+This further contrasts with literal $L_5$, a type lacks means to convey an upper-bound with $m_1$.
+Due to the possible marginal difference in signaling behavior between pragmatic $L_4$ and $L_5$, the selection of one type over the other is expected to mainly depend on the learning bias.
+Things are less clear for literal $L_5$ contrasted with literal/pragmatic $L_4$.
+The former has a learning advantage but is expected to fare worse in terms of communicative fitness in virtue of ambiguous $m_1$.
 
-A language that has conventionalized upper and lower bounds to realize a (quasi-)partition of the relevant semantic space will produce speaker behavior that is /almost/ indistinguishable from that of a language with only the respective upper bounds, but with Gricean speakers.
-Almost, because there may be slight differences between the probability with which speakers would (erroneously) use a semantically false description and the probability with which speakers would (erroneously) use a pragmatically suboptimal description.
-\hl{explain why only these and not others}
+As implicit in the discussion above, one may think of $s_1$ as a ``some but not all''-state and $s_2$ for an ``all''-state.
+The literal meaning of weak scalar expressions such as {\em some} then corresponds to a message true of both $s_1$ and $s_2$ in these fragments.
+Following our assumption for a preference for simple lexical representations, the prior biases learners against lexica in which a message holds true only of the former and not the latter.
+All other semantics are a priori equally probable.
+This prior is given by $P(t_i) \propto n - c \cdot r$, where $n$ is the total number of states and $r$ is the number of messages only true of $s_1$ in $t_i$'s lexicon, $c \in [0,1]$.
+In sum, an increase in $c$ brings about a stronger learning bias against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.
+
+\subsection{Analysis}
+The dynamics are initialized with an arbitrary distribution over types, constituting the population's first generation.
+If not stated otherwise, the results for a given parameter setting were obtained from $1000$ independent runs.
+Each run consisted of $20$ generations.
+This corresponds to a developmental plateau after which no noteworthy change was registered.
+The model's parameters are summarized in Table \ref{tab:summary}.
 
@@
     $\lambda \geq 1$ & rationality parameter & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\
-    $\alpha \in (0,1)$ & semantics-pragmatics tension & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\ 
+    $\alpha \in [0,1)$ & semantics-pragmatics tension & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\ 
     $|D|$ & learning data produced per parent type & $P(d|t_j)P(t_i|d)$\\
@@
     $l \geq 1$ & posterior parameter from sampling to MAP & $P(t_i|d) \propto [P(t_i)P(d|t_i)]^l$\\
-    $c \in (0,1]$ & learning bias for upper-bound lack &  $P(t_i)$
+    $c \in [0,1]$ & learning bias for lack of upper-bounds &  $P(t_i)$
 \end{tabular}
+\caption{Summary of model parameters.} 
+\label{tab:summary}
 \end{table}
 
-\paragraph{Procedural description.} At the game's onset we compute $Q$ once based on the sets  of sequences $D$ (one for each parent type).
-Replicator dynamics are computed based on the fitness of each type in the current population as usual.
-$Q$ is computed anew for each independent run (of $g$ generations) given that it depends on $D$, which is sampled from production probabilities.
+\hl{MENTION SAMPLING FROM PRODUCTION}
 
-\subsection{Model parameters \& procedure} 
-\begin{enumerate}
-  \item Sequence length $k$
-  \item Pragmatic production parameter $\alpha$
-  \item Rationality parameter $\lambda$
-  \item Learning prior over types (lexica); cost parameter $c$.
-$p^\star(t_i) \propto n - c \cdot r$ where $n$ is the total number of states and $r$ that of upper-bounded messages only true of $s_1$ in $t_i$'s lexicon (if only $s_1$ is true of a message, then this message encodes an upper-bound).
-Then the score for $L_1$, $L_3$, $L_5$ is $2$, that of $L_4$ and $L_6$ is $2-c$, and that of $L_2$ is $2-2c$; Normalization over lexica scores yields the prior over lexica (which is equal to the prior over types).
-  \item Prior over meanings ($pr$). We assume that $pr(s) = \frac{1}{|S|}$ for all $s$.
-  \item True state distribution ($P$). We currently assume that $P = \frac{1}{|S|}$ but it may be interesting to vary this
-  \item Learning parameter $l \geq 1$ with $1$ corresponding to probability matching, and MAP as $l$ approaches infinity
-  \item $n$ is the sample of sequences of observations of length $k$ sampled from the production probabilities of each type
-  \item Number of generations $g$
-\end{enumerate}
+According to our hypothesis, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $L_5$-like semantics.
+However, it is instructive to first inspect the effect of these pressures in isolation.
+We will focus three pragmatic types.\footnote{Pragmatic reasoning allows hearers and speakers to refine their (possibly erroneous) interpretations, rendering it advantageous even for those types that codify an upper-bound lexically.} Users of $L_3$, representing a type that is lacking in expressivity but is a prior preferred for its lack of upper-bounds.
+Users of $L_4$, a type that is functionally advantageous but biased against.
+And users of $L_5$, combining the virtues of the latter two.
 
-\subsection{Analysis}
+\paragraph{Expressivity only.} Recall that the outcome of the replicator dynamics are influenced by $\lambda$ and $\alpha$ as they have a bearing on a type's fitness.
+Low $\alpha$ disadvantages types that rely on pragmatic reasoning for more deterministic signaling behavior to the gain of those that codify this information semantically.
+The rationality parameter $\lambda$ has a similar effect for different reasons.
+Less utility maximizing behavior decreases the association of an ambiguous message with a single state, even when other states are uniquely associated with a different one.
+That is, $\lambda$ regulates the strength by which users of $L_5$ associate non-upper-bounded $m_1$ exclusively with the ``some''-state $s_1$ over the ``all''-state $s_2$.
+
+The effect of the rationality parameter using only the replicator dynamics is shown in the left-hand side of Figure \ref{fig:either-R-or-M}.
+As expected, the less expressive type using $L_3$ fares worse and shows little variation across $\lambda$-values.
+Crucially, values of $\lambda \leq 10$ lead to an increase in $L_4$ and a decrease in $L_5$.
+As the rationality parameter increases, the functional difference between $L_4$ and $L_5$ is levelled.
+Overall, the final populations that result only from a pressure towards expressively approximate an even share of pragmatic $L_4, L_5$ and $L_6$ types.
+The latter follows the same trajectory as $L_4$ in Figure \ref{fig:either-R-or-M}.
+
 \begin{figure}
@@
 \includegraphics[scale=.5]{./only-R-or-M}
+\caption{Mean proportions of target types after $20$ generations in $1000$ populations with only replication (left; $\alpha = 1$) and only mutation (right; $\lambda = 20, k = 5$).}
+\label{fig:either-R-or-M}
 \end{figure}
+
+\paragraph{Learnability only.} To get an impression of the effect of iterated learning without a pressure towards expressivity, we first consider (relatively) deterministic parental data production ($\lambda = 20, \alpha = 1$).
+In this way small data sequences suffice for learners to differentiate types that produce strongly diverging expressions.
+In line with \citeposs{griffiths+kalish:2007} analysis, under these conditions the resulting populations converge to the learner's prior distribution when sampling from the posterior ($l = 1$).
+This is shown in the right-hand side of Figure \ref{fig:either-R-or-M}, which directly reflects the prior distribution for each value of $c$.
+
+Inspecting the effects of these dynamics separately not only gives some intuitions about the parameters' influence, but also highlights some further-reaching issues in our setup.
+First and foremost, neither dynamic comes close to converging to a monomorphic population under most parameter configurations.
+For instance, while types using $L_4$ can come to populate a large proportion of the final population, a very low degree of rationality needs to be assumed.
+
+\hl{How about IL with higher L? Take a look at the prior please!}
+
+\begin{itemize}
+  \item Discuss RMD;
+  \item Discuss prior and posterior;
+\end{itemize}
+
 \begin{figure}
@@
 
+\paragraph{Results.} We report the mean results obtained from $1000$ independent simulations for each set of parameters.
+Figure \ref{fig:c-85} illustrates the development of the target types over $20$ generations with $c = 0.85$, i.e., a strong bias against semantic complexity.
+Figure \ref{fig:cost} shows the effect of the learning bias after $20$ generations for $c \in [0,1]$.
+More detailed results for all types across a sample of $c$-values are presented in Table \ref{tab:numeric-results}.
+
+\begin{table}
+\centering 
+\begin{tabular}{l c c c c c c c c c c c c c}
+~ & \multicolumn{5}{c}{$l = 1$} & ~ & \multicolumn{5}{c}{$l = 10$}\\ \hline \hline
+c  & 0 & .01 & .05 & .1 & .8 & empty & 0 & .01 & .05 & .1 & .8\\ \hline \hline
+lit. $L_1$& ~ & ~ & ~ & ~  & ~& ~ & ~& ~ & ~ & ~ & ~\\ 
+lit. $L_2$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~\\
+lit. $L_3$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~\\
+lit. $L_4$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~\\
+lit. $L_5$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~\\
+lit. $L_6$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~\\ \hline
+prg. $L_1$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~ \\
+prg. $L_2$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~ \\
+prg. $L_3$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~ \\ 
+prg. $L_4$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~ \\
+prg. $L_5$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~ \\
+prg. $L_6$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~
+
+\end{tabular}
+\caption{Mean proportion of types in $1000$ populations after $20$ generations across different bias values $c$ ($\alpha = 1, \lambda = 50, k = 20$).}
+
+\label{tab:numeric-results}
+\end{table}
+
+In sum, the results show that in the present setup a weak bias is sufficient to lead to a selection of $L_5$ over $L_4$.
+This effect increases with the bias' strength provided $L_5$ users are pragmatic.
+That is, such a learning bias may prevent the lexicalization of pragmatic inferences and explain the prevalence of $L_5$-like semantics.
+However, note that literal $L_5$ is underrepresented across all values of $c$.
+This highlights that, while the bias plays a major role for the contrast between $L_4$ and $L_5$, in itself it does not enable types that fail to convey an upper-bound to establish themselves in the population.
+This suggests pragmatic reasoning to be paramount to the selection of $L_5$-like semantics.
+That is, neither the learning bias nor functional pressure alone but their combination lead to the systematic lack of semantic upper-bounds in scalar expressions.
+
 \subsection{Discussion}
+changes in sequence length influence the population in a predictable way: smaller values lead to more heterogeneous populations whereas larger ones lead to more pronounced differences.
+This is expected insofar as the likelihood that a sequence of length $1$ was produced by any type is relatively uniform (modulo prior) whereas the likelihood of types with lexica $L_1$ - $L_3$ to produce, for instance, a sequence of $10$ observations consistently with the same state-message combination is less likely than for pragmatic players using $L_4$ - $L_6$ or literal $L_4$.
+Thus, while noteworthy, sequence length has no direct bearing on the main contrast of interest.
+Similar considerations hold for $\alpha$ and $\lambda$ -- set to $1$ and $50$ in the following.
+Overall, lower rationality in $\lambda$ or more pragmatic violations in $\alpha$ lead to a higher selection of lexica with semantic upper-bounds.
+The fitness of pragmatic behavior increases with higher $\lambda$-/$\alpha$-values.
+In other words, these parameters level the functional contrast between $L_4$ and $L_5$.
 

### 2016-09-26 paper/paper.tex
new labels and style for paper plots + Added more to the analysis section

@@
 
-\subsection{Mutation \& learning}
+\subsection{Mutation \& learning}\label{sec:mutation}
 Iterated learning is a process in which the behavior of one individual serves as learning input for another, who's behavior subsequently serves as input for a new learner, and so on.
@@
 This corresponds to a developmental plateau after which no noteworthy change was registered.
+As specified in \ref{sec:mutation}, the learning transition matrix $Q$ can be obtained by considering all possible state-message sequences of length $k$.
+Given that this is intractable for large $k$, matrices with $k > 5$ were approximated by using Monte Carlo with $10$ sequences sampled from each type's production probabilities and a type's children being exposed only to this subset.
 The model's parameters are summarized in Table \ref{tab:summary}.
@@
 
-\hl{MENTION SAMPLING FROM PRODUCTION}
-
 According to our hypothesis, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $L_5$-like semantics.
 However, it is instructive to first inspect the effect of these pressures in isolation.
-We will focus three pragmatic types.\footnote{Pragmatic reasoning allows hearers and speakers to refine their (possibly erroneous) interpretations, rendering it advantageous even for those types that codify an upper-bound lexically.} Users of $L_3$, representing a type that is lacking in expressivity but is a prior preferred for its lack of upper-bounds.
+For this purpose, we focus on three pragmatic types.\footnote{Pragmatic reasoning allows language users to refine their (possibly erroneous) choices.
+Therefore, it is advantageous even for those types that codify an upper-bound lexically.} Users of $L_3$, representing a type that is lacking in expressivity but is a prior preferred for its lack of upper-bounds.
 Users of $L_4$, a type that is functionally advantageous but biased against.
@@
 \includegraphics[scale=.5]{./only-R-or-M}
-\caption{Mean proportions of target types after $20$ generations in $1000$ populations with only replication (left; $\alpha = 1$) and only mutation (right; $\lambda = 20, k = 5$).}
+\caption{Mean proportions of target types after $20$ generations in $1000$ populations with only replication (A; $\alpha = 1$) and only mutation (B; $\alpha =1, \lambda = 30, k = 5$).}
 \label{fig:either-R-or-M}
@@
 In this way small data sequences suffice for learners to differentiate types that produce strongly diverging expressions.
-In line with \citeposs{griffiths+kalish:2007} analysis, under these conditions the resulting populations converge to the learner's prior distribution when sampling from the posterior ($l = 1$).
+In line with \citeposs{griffiths+kalish:2007} analysis, under these conditions the resulting populations converge to the learner's prior distribution when sampling from the posterior.
 This is shown in the right-hand side of Figure \ref{fig:either-R-or-M}, which directly reflects the prior distribution for each value of $c$.
 
-Inspecting the effects of these dynamics separately not only gives some intuitions about the parameters' influence, but also highlights some further-reaching issues in our setup.
+Inspecting the effects of these dynamics separately not only gives some intuitions about the parameters' influence, but also highlights some of their broader implications.
 First and foremost, neither dynamic comes close to converging to a monomorphic population under most parameter configurations.
-For instance, while types using $L_4$ can come to populate a large proportion of the final population, a very low degree of rationality needs to be assumed.
+For instance, types using $L_4$ can come to occupy a large proportion of the final population.
+However, this holds only for a restricted range of low degrees of rationality.
+Apart from polymorphy, both dynamics make some undesirable predictions.
+A pressure only towards expressivity leads to the selection of types using $L_4$ to $L_6$ and to the ejection of $L_1$ to $L_3$.
+However, it can not explain the regular selection of $L_5$-like semantics over either of these functionally similar alternatives.
+In contrast, a pressure only towards learnability has a modest but clear effect in differentiating $L_5$ from these alternatives but fails to rule out functionally suboptimal types such as tautological $L_3$.
+This showcases that, while the bias plays a major role for the contrast between $L_4$ and $L_5$, on its own it does not enable types that fail to convey an upper-bound to establish themselves in the population.
+In sum, neither dynamic on its own is a suitable candidate to provide a justification for the predicted prevalence of $L_5$-like semantics.
 
-\hl{How about IL with higher L? Take a look at the prior please!}
-
-\begin{itemize}
-  \item Discuss RMD;
-  \item Discuss prior and posterior;
-\end{itemize}
+\paragraph{Expressivity and learnability.} Figure \ref{fig:cost} illustrates the effect of the learning bias after $20$ generations across values of $c$ with $l = 1$ (A) and $l =3$ (B).
+More detailed results for all types across a sample of $c$-values are presented in Table \ref{tab:numeric-results}.
+Overall, these results suggest that in the present setup a weak bias is sufficient to lead to a selection of $L_5$ over $L_4$.
+As in the simulations that only considered learnability, this effect increases with the bias' strength provided $L_5$ users are pragmatic.
+Importantly, the addition of a pressure towards expressivity magnifies this effect by dampening the proliferation of functionally suboptimal types advantaged by the learning bias.
+As stressed above, this suggests that neither the learning bias nor functional pressure alone but their combination lead to the systematic lack of semantic upper-bounds in scalar expressions.
 
@@
 \centering
-\includegraphics[scale=.5]{../presentations/03cost-with-l1}
+\includegraphics[scale=.5]{./fig2-rmd}
+\label{fig:cost}
+\caption{Mean proportions of target types after $20$ generations in $1000$ populations across bias values $c \in [0,1]$ with $l =1$ in A and $l = 3$ in B ($\alpha =1, \lambda = 20, k = 5$).}
 \end{figure}
+
+\begin{table}
+\centering 
+\begin{tabular}{l | c c c c c| c c c c c c c c}
+\multicolumn{1}{c}{~} & \multicolumn{5}{c}{$l = 1$} & ~ & \multicolumn{5}{c}{$l = 10$}\\ \hline \hline
+c  & 0 & .01 & .05 & .1 & .8 & ~ & 0 & .01 & .05 & .1 & .8\\ \hline \hline
+lit. $L_1$& .03 & .03 & .03 & .03  & .04 & ~ & $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\ 
+lit. $L_2$ & .03 & .03 & .03 & .03 & .01 & ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
+lit. $L_3$ & .03 & .03 & .03 & .03 & .04 & ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
+lit. $L_4$ & .07 & .07 & .07 & .07 & .06 & ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
+lit. $L_5$ & .04 & .04 & .05 & .05 & .06& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
+lit. $L_6$ & .04 & .04 & .04 & .04 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\ \hline
+prg. $L_1$ & .03 & .03 & .03 & .03 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ \\
+prg. $L_2$ & .03 & .03 & .03 & .03 & .01& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ \\
+prg. $L_3$ & .03 & .03 & .03 & .03 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ \\ 
+prg. $L_4$ & .22 & .22 & .22 & .22 & .18& ~ & .33& .33 & .32 & .31 & .15 \\
+prg. $L_5$ & .22 & .22 & .22 & .23 & .3& ~ & .33& .33 & .35 & .37 & .7 \\
+prg. $L_6$ & .22 & .22 & .22 & .22 & .18& ~ & .33& .33 & .32 & .31 & .15   
+\end{tabular}
+\caption{Mean proportions of types in $1000$ populations after $20$ generations across bias values $c \in [0,1]$ with $l =1$ and $l = 3$ ($\alpha = 1, \lambda = 30, k = 5$), $\epsilon < 0.005$}
+\label{tab:numeric-results}
+\end{table}
+
+This data suggests that the proportion of scalar implicature users that is predicted primarily hinges on three aspects of the model.
+First, the degree to which linguistic behavior is deterministic, controlled by $\lambda$ and $\alpha$, plays a role both for expressivity as well as for producing data that allows learners to discriminate this type from others.
+Second, the learning bias $c$, leads learners to discriminate and prefer $L_5$ over $L_4$ and $L_5$.
+Lastly, the posterior parameter $l$ magnifies the effects of the learning bias in tandem with replication.
+This interaction is shown in \ref{fig:prior-posterior}.
+In the present setup posterior sampling can lead to the incumbency of pragmatic $L_5$, but not even a strong favorable learning bias manages to completely drive out competing types (cf.
+\ref{fig:cost}.A).
+However, as posterior maximization increases, the range of bias values within which this type takes over the population increases drastically.
+
+\hl{More  discussion about parameter interaction}
 
@@
 \includegraphics[scale=.5]{../presentations/01heatmap}
+\caption{Mean proportion of pragmatic $L_5$ in $1000$ populations after $20$ generations ($\alpha = 1, \lambda = 30, k = 5$)}
+\label{fig:prior-posterior}
 \end{figure}
-
-\paragraph{Results.} We report the mean results obtained from $1000$ independent simulations for each set of parameters.
-Figure \ref{fig:c-85} illustrates the development of the target types over $20$ generations with $c = 0.85$, i.e., a strong bias against semantic complexity.
-Figure \ref{fig:cost} shows the effect of the learning bias after $20$ generations for $c \in [0,1]$.
-More detailed results for all types across a sample of $c$-values are presented in Table \ref{tab:numeric-results}.
-
-\begin{table}
-\centering 
-\begin{tabular}{l c c c c c c c c c c c c c}
-~ & \multicolumn{5}{c}{$l = 1$} & ~ & \multicolumn{5}{c}{$l = 10$}\\ \hline \hline
-c  & 0 & .01 & .05 & .1 & .8 & empty & 0 & .01 & .05 & .1 & .8\\ \hline \hline
-lit. $L_1$& ~ & ~ & ~ & ~  & ~& ~ & ~& ~ & ~ & ~ & ~\\ 
-lit. $L_2$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~\\
-lit. $L_3$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~\\
-lit. $L_4$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~\\
-lit. $L_5$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~\\
-lit. $L_6$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~\\ \hline
-prg. $L_1$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~ \\
-prg. $L_2$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~ \\
-prg. $L_3$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~ \\ 
-prg. $L_4$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~ \\
-prg. $L_5$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~ \\
-prg. $L_6$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~
-
-\end{tabular}
-\caption{Mean proportion of types in $1000$ populations after $20$ generations across different bias values $c$ ($\alpha = 1, \lambda = 50, k = 20$).}
-
-\label{tab:numeric-results}
-\end{table}
-
-In sum, the results show that in the present setup a weak bias is sufficient to lead to a selection of $L_5$ over $L_4$.
-This effect increases with the bias' strength provided $L_5$ users are pragmatic.
-That is, such a learning bias may prevent the lexicalization of pragmatic inferences and explain the prevalence of $L_5$-like semantics.
-However, note that literal $L_5$ is underrepresented across all values of $c$.
-This highlights that, while the bias plays a major role for the contrast between $L_4$ and $L_5$, in itself it does not enable types that fail to convey an upper-bound to establish themselves in the population.
-This suggests pragmatic reasoning to be paramount to the selection of $L_5$-like semantics.
-That is, neither the learning bias nor functional pressure alone but their combination lead to the systematic lack of semantic upper-bounds in scalar expressions.
 

### 2016-09-26 presentations/edindata/rmd-plots-sidebyside.py

@@ -15,21 +15,14 @@ import sys
 #####################
 
 
-print 'Loading data fitness'
-path =r'./results-fit' 
+print 'Loading data'
+path =r'./results1' 
 all_files = glob.glob(os.path.join(path, "*.csv"))     # advisable to use os.path.join as this makes concatenation OS independent
 df = pd.concat(pd.read_csv(f) for f in all_files)
-
-print 'Loading data learn'
-path =r'./results-learn' 
-all_files = glob.glob(os.path.join(path, "*.csv"))     # advisable to use os.path.join as this makes concatenation OS independent
-dl = pd.concat(pd.read_csv(f) for f in all_files)
 
 def cust_mean(grp):
     grp['mean'] = grp['t11_final'].mean()
     return grp
-
-
 
 
 ### development over cost ###
@@ -38,47 +31,36 @@ a = 1
 k = 5
 lam = 30
 sample = 10
+learn = 3
+
+
+final_group = df.loc[:,('alpha','prior_cost_c','lambda','k','sample_amount','learning_parameter','t9_final','t10_final','t11_final')]
+
+group = final_group.loc[final_group['alpha'] == a]
+group = group.loc[group['k'] == k]
+group = group.loc[group['lambda'] == lam]
+group = group.loc[group['sample_amount'] == sample]
+group = group.loc[group['learning_parameter'] == learn]
+
+t_final = group.groupby(['prior_cost_c'])
+t_final = t_final[['t9_final','t10_final','t11_final']].agg(np.average) 
+
+##
 learn = 1
 
-
-final_groupA = df.loc[:,'alpha':]
-final_groupB = dl.loc[:,'alpha':]
-
-groupA = final_groupA.loc[final_groupA['alpha'] == a]
-groupA = groupA.loc[groupA['k'] == k]
-groupA = groupA.loc[groupA['sample_amount'] == sample]
-groupA = groupA.loc[groupA['learning_parameter'] == learn]
-groupA = groupA.loc[groupA['prior_cost_c'] == 0.1]
-
-t_finalA = groupA.groupby(['lambda'])
[…]

### 2016-09-26 presentations/edindata/tabular-info.py

@@ -1,0 +1,90 @@
+###
+import pandas as pd
+import numpy as np 
+import os
+import glob
+import sys
+###
+###Column headers:###
+#run_ID,t1_initial,t2_initial,t3_initial,t4_initial,t5_initial,t6_initial,t7_initial,t8_initial,t9_initial,t10_initial,t11_initial,t12_initial,
+#alpha,prior_cost_c,lambda,k,sample_amount,learning_parameter,generations,
+#t1_final,t2_final,t3_final,t4_final,t5_final,t6_final,t7_final,t8_final,t9_final,t10_final,t11_final,t12_final
+#####################
+
+
+print 'Loading data'
+path =r'./results1' 
+all_files = glob.glob(os.path.join(path, "*.csv"))     # advisable to use os.path.join as this makes concatenation OS independent
+df = pd.concat(pd.read_csv(f) for f in all_files)
+
+def cust_mean(grp):
+    grp['mean'] = grp['t11_final'].mean()
+    return grp
+
+### development over cost ###
+
+a = 1
+k = 5
+lam = 30
+sample = 10
+
+
+final_group = df.loc[:,('alpha','prior_cost_c','lambda','k','sample_amount','learning_parameter','t1_final','t2_final', 't3_final', 't4_final', 't5_final', 't6_final', 't7_final', 't8_final', 't9_final','t10_final','t11_final', 't12_final')]
+
+group = final_group.loc[final_group['alpha'] == a]
+group = group.loc[group['k'] == k]
+group = group.loc[group['lambda'] == lam]
+group = group.loc[group['sample_amount'] == sample]
+group1 = group.loc[group['learning_parameter'] == 1]
+
+group1 = group1.groupby(['prior_cost_c']).mean()
+
+group2 = group.loc[group['learning_parameter'] == 3]
+group2 = group2.groupby(['prior_cost_c']).mean()
+
+sys.exit()
+##
+learn = 1
+
+groupB = final_group.loc[final_group['alpha'] == a]
+groupB = groupB.loc[groupB['k'] == k]
+groupB = groupB.loc[groupB['lambda'] == lam]
+groupB = groupB.loc[groupB['sample_amount'] == sample]
+groupB = groupB.loc[groupB['learning_parameter'] == learn]
+
[…]

### 2016-09-27 paper/paper.tex
draft analysis section for SIs

@@
 
-\begin{abstract} Certain lexical meanings enable for pragmatic enrichments in a notably productive fashion.
-This raises the challenge to justify their regular selection, in particular, over alternatives that codify semantically what is conveyed pragmatically.
-To address this challenge, we propose a general model that integrates iterated Bayesian learning in the replicator-mutator dynamics.
-This model allows for population-level analyses of the effects of linguistic pressures on probabilistic language users with varied degrees of pragmatic sophistication and distinct languages.
-We showcase the model's use and predictions in a case study on the (lack of) lexicalization of scalar implicatures.
-The results suggest simpler semantic representations to be selected for when languages are pressured towards learnability and compression, provided that pragmatic reasoning can compensate for the disadvantage in expressivity that users of such languages otherwise incur.
-
-\section{The semantics-pragmatics divide}\label{sec:introduction}
-
+\begin{abstract} Certain classes of lexical meanings enable for pragmatic enrichments in a notably productive fashion.
+This raises the challenge to justify their regular selection over alternatives that codify semantically what is conveyed pragmatically.
+To address this issue, we propose a general model that integrates iterated Bayesian learning in the replicator-mutator dynamics.
+This model generates predictions about the effects of linguistic pressures on selection and transmission in populations of probabilistic language users with varied degrees of pragmatic sophistication.
+We apply this model in a case study on the (lack of) lexicalization of scalar implicatures.
+The results show that simpler semantic representations are selected for when languages are pressured towards learnability and compression, provided that pragmatic reasoning can compensate for the disadvantage in expressivity that users of such languages otherwise incur.
+We argue this result to shed light on the lack of lexicalization of scalar inferences, as well as the semantics-pragmatics distinction more generally.
 \end{abstract}
 
-In linguistic theorizing, it is common to draw a distinction between semantics and pragmatics.
-Broadly speaking, the former concerns the truth-conditional content of expressions, whereas the latter concerns information beyond literal meanings and their composition.
+\section{Simplicity, expressivity, and learnability at the semantics-pragmatics interface}\label{sec:introduction}
+
+In linguistic theorizing it is common to draw a distinction between semantics and pragmatics.
+Broadly speaking, the former concerns the truth-conditional content of expressions whereas the latter concerns information beyond literal meanings and their composition.
 An important consequence of this distinction is that the information conveyed by an utterance is seldom, if ever, solely determined by semantics, but rather in tandem with pragmatics.
+Much research at the semantics-pragmatics interface has been aimed at characterizing expressions in terms of either domain or their interplay.
+However, an issue that has received little attention is the justification of semantic structure in light of pragmatics.
+The present investigation seeks to fill this gap by analyzing the effects linguistic pressures have on the selection and pervasiveness of particular lexical meanings under consideration of their possible pragmatic enrichments.
 
-Much research at the semantics-pragmatics interface has been aimed at characterizing expressions in terms of either domain, or their interplay.
-However, an issue that has received little attention is the justification of semantic structure in light of pragmatics.
-The present investigation seeks to fill this gap by analyzing the effects linguistic pressures have on the selection and pervasiveness of particular lexical meanings under consideration of pragmatic enrichments.
-
-In recent years, similar questions about the emergence and change of linguistic features have lead to a surge in models to address them (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
-Our starting point is given by the overarching argument that has crystalized from accumulated mathematical, experimental and cross-linguistic evidence in this literature: Natural languages need to be well-adapted to communicative needs within a linguistic community, but also need to be learnable to survive their faithful transmission across generations.
-More succinctly; natural languages are pressured for expressivity as well as learnability.
+In recent years questions about the emergence and stability of linguistic features have lead to a surge in models that analyze their transmission within and across generations (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
+Our starting point is given by the overarching argument that has crystalized from the accumulated mathematical, experimental and cross-linguistic evidence in this literature: Natural languages need to be well-adapted to communicative needs within a linguistic community, but also need to be learnable to survive their faithful transmission across generations.
+More succinctly; natural languages are pressured for expressivity and learnability.
 
@@
 This allows for the inspection of their interaction by combining functional pressure on successful communication with effects of learning biases on (iterated) Bayesian learning \citep{griffiths+kalish:2007}.
-The semantics-pragmatics distinction and its effect on production and comprehension is made precise by considering probabilistic models of rational language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
-
-%In this way, the model establishes a link between models of synchronic probabilistic rational language use and diachronic models of cultural evolution.
+The semantics-pragmatics distinction and its effect on production and comprehension are made precise by considering probabilistic models of rational language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
+\hl{SOMETHING MISSING HERE} \hl{In this way, the model establishes a link between models of synchronic probabilistic rational language use and diachronic models of cultural evolution.
+}
 
@@
 
-\section{Simplicity, expressivity, and learnability}
-
-The emergence and change of linguistic structure is influenced by many intertwined factors, ranging from biological and socio-ecological to cultural \citep{steels:2011,tamariz+kirby:2016}.
-Social and ecological pressures determine communicative needs, while biology determines the architecture that enables and constrains their means of fulfillment.
-In the following, our focus lies on the latter, cultural factor, wherein processes of linguistic change are understood as shaped by its use and transmission.
-That is, as a result of cultural evolution.
+The emergence and change of linguistic structure is influenced by many intertwined factors.
+These range from biological and socio-ecological to cultural \citep{steels:2011,tamariz+kirby:2016}.
+Social and ecological pressures determine communicative needs, while biology determines the architecture that enables and constrains the means by which they can be fulfilled.
+Our focus lies on the latter, cultural, factor wherein processes of linguistic change are understood as shaped by language use and its transmission.
+That is, as a result of a process of cultural evolution.
 
@@
 
-As noted above, expressivity and learnability are two major competing pressures.
+As noted above, expressivity and learnability have been identified as two central competing pressures.
 Their opposition becomes particularly clear when considering their consequences in the extreme (cf.
 \citealt{kemp+regier:2012,kirby+etal:2015}).
-On the one side, a language with a single form is easy to learn but lacking in expressivity for most purposes.
+On one extreme, a language with a single form is easy to learn but lacking in expressivity.
 On the other, a language that associates a distinct form with all possible meanings its users may want to convey is maximally expressive but challenging to acquire.
@@
 However, this so-called transmission bottleneck is not the only challenge learners confront.
-
-More important for our purposes is the problem of selecting particular hypotheses out of a potentially infinite space of alternatives compatible with the data learners are exposed to.
+Instead, a problem that is more important for our present purpose is that of selecting particular hypotheses out of a (potentially infinite) space of alternatives compatible with the data learners are exposed to.
 At the semantics-pragmatics interface this concerns the selection between functionally similar, if not identical, lexical meanings.
 In the following, we argue an integral part of the answer to be that learners are a priori biased towards simpler, more compressed, lexical representations.
-This corresponds to the argument that rational learners should prefer simpler over more complex explanations of data \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
+That is, rational learners should prefer simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
 In linguistics, a drive for simplicity has been argued to underpin speaker preferences for brevity and ease of articulation, as well as to pressure languages towards lexical ambiguity and grammatical compression \citep{zipf:1949,grice:1975,piantadosi+etal:2012, kirby+etal:2015}.
-As a broader cognitive principle, the use of simplicity as means to select between hypotheses has a long standing tradition.
-Crucially, \citet{chater+vitanyi:2003} give a number of compelling arguments for simplicity on both mathematical and empirical grounds.
+\hl{As a broader cognitive principle, the use of simplicity as means to select between hypotheses has a long standing tradition.
+Crucially, \citet{chater+vitanyi:2003} give a number of compelling arguments for simplicity on both mathematical and empirical grounds.}
 
 The remainder of this section introduces the individual components of the model in more detail, as well as the assumptions underlying them.
-These are: (i) languages and their use, (ii) pressures towards expressivity and learnability, regulated by the replicator and mutator dynamics, respectively, as well as (iii) a bias towards simpler semantic representations, codified as a language learner's prior.
+These are: (i) the representation of languages and their use, (ii) pressures towards expressivity and learnability, regulated by the replicator and mutator dynamics, respectively, as well as (iii) a bias towards simpler semantic representations, codified as a language learner's prior.
 After laying out the model, we discuss its application to the lack of lexicalization of scalar implicatures.
 
-\subsection{Languages and linguistic behavior}
+\subsection{Lexica and linguistic behavior}
 Lexica codify the truth-conditions of a language's expressions, i.e., its semantics.
-A convenient way to represent such lexica is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs, or meanings, to convey and $M$ a set of messages of the language \citep{franke+jaeger:2014}.
-For instance, the following two lexica fragments determine the truth-conditions of two messages, $m_1$ and $m_2$, for two states, $s_1$ and $s_2$:
+A convenient way to represent them is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs -- or meanings -- to convey and $M$ a set of messages of the language \citep{franke+jaeger:2014}.
+For example, the following two fragments determine the truth-conditions of two messages, $m_1$ and $m_2$, for two states, $s_1$ and $s_2$.
 
@@
 
-In words, according to lexicon $L_a$, $m_1$ is true of state $s_1$ as well as of $s_2$.
-In contrast, message $m_1$ is only true of $s_1$ in $L_b$.
+According to lexicon $L_a$ the message $m_1$ is true of state $s_1$ as well as of $s_2$.
+In contrast, $m_1$ is only true of $s_1$ in $L_b$.
 Otherwise, the two languages are truth-conditionally equivalent.
@@
 {\em Literal interlocutors} produce and interpret messages literally.
-That is, their linguistic choices are guided by their lexica only.
+That is, their linguistic choices are guided only by their lexica.
 In contrast, {\em pragmatic interlocutors} engage in mutual reasoning to inform their choices.
-For instance, a rational speaker of $L_a$ who reasons about her addressee should use $m_1$ to signal state $s_1$ given that $s_2$ can unambiguously be conveyed with $m_2$.
-Analogously, should rational hearers expect their interlocutors to reason along these lines, they will interpret ambiguous $m_1$ accordingly.
-Note in particular that according to this strenghtening of $m_1$, $L_a$ is indistinguishable from $L_b$ in terms of expressivity if its users are pragmatic reasoners.
+For instance, a rational speaker of $L_a$ who reasons about her addressee may use $m_1$ to exclusively signal state $s_1$ given that $s_2$ can unambiguously be conveyed through $m_2$.
+Accordingly, rational hearers that expect their interlocutor to reason along these lines will interpret ambiguous $m_1$ as $s_1$.
+An important consequence of this pragmatic process is that it renders $L_a$ indistinguishable from $L_b$ in terms of expressivity.
 
-Following models of rational language use such as Rational Speech Act models \citep{frank+goodman:2012} and their game-theoretic counterparts \citep{benz+etal:2005a,franke:2009,franke+jaeger:2014}, this kind of signaling behavior is captured by a hierarchy over reasoning types.
+Following models of rational language use such as Rational Speech Act models \citep{frank+goodman:2012} and their game-theoretic counterparts \citep{benz+etal:2005a,franke:2009,franke+jaeger:2014}, this kind of signaling behavior is captured by a resaoning hierarchy.
 The hierarchy's bottom, level $0$, corresponds to literal language use.
 Pragmatic language users of level $n + 1$ behave rationally according to (expected) level $n$ behavior of their interlocutors.
-The behavior of literal and pragmatic hearers of a language $L$ is given by their respective selection functions in (\ref{h:level0}) and (\ref{h:leveln}).
-Mutatis mutandis for the speaker functions in (\ref{s:level0}) and (\ref{s:leveln}).
+(\ref{h:level0}) and (\ref{h:leveln}) specify the behavior of literal and pragmatic hearers of a language $L$.
+Mutatis mutandis for the literal and pragmatic speakers in (\ref{s:level0}) and (\ref{s:leveln}).
 
@@
 
-According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state $s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$.
-The literal speaker's choice in (\ref{s:level0}) is regulated by a soft-max parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}.
-As $\lambda$ increases, choices made in production are more rational in that higher values lead to more deterministic in line with expected utility maximization.
+According to (\ref{h:level0}) a literal hearer's interpretation of a message $m$ as a state $s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$.
+The literal speaker's behavior as given in (\ref{s:level0}) is regulated by a soft-max parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}.
+As $\lambda$ increases, choices made in production are more rational.
+That is, higher values lead to behavior that is increasingly in line with expected utility maximization.
 
-For the most part, pragmatic behavior mirrors its literal counterpart.
-As described above, their difference lies in that level $n+1$ speakers/hearers reason about level $n$ hearer/speaker behavior instead of solely relaying on their lexicon.
-That is, they reason about how a rational level $n$ interlocutor would use or interpret a message and behave according to these expectations.
-Additionally, pragmatic production is further regulated by a parameter $\alpha$ which controls the tension between semantics and pragmatics, $\alpha \in (0,1]$.
-Lower values lead to more literal production, whereas higher values lead to stronger pragmatic behavior.
+Pragmatic behavior is similar to its literal counterpart.
+As in our informal sketch, their difference lies in that level $n+1$ speakers/hearers reason about level $n$ hearer/speaker behavior instead of solely relaying on their lexicon.
+That is, they reason about how a rational level $n$ interlocutor would use or interpret a message, and behave according to these expectations.
+Pragmatic production is further regulated by a parameter $\alpha$ which controls the tension between semantics and pragmatics, $\alpha \in (0,1]$.
+Lower values lead to more literal production whereas higher values lead to stronger pragmatic behavior.
 
-The combination of a lexicon with its use, i.e., a particular level of linguistic sophistication, yields a type $t$. Types are the basic units on which our population dynamics operate.
+The combination of a lexicon with its use, i.e., a level in the reasoning hierarchy, is called a type. These are the basic units on which our population dynamics operate.
 
@@
 In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so.
-Compare this with a hypothetical language lexicalizes two expressions instead of a single scalar one; one with and one lacking an upper-bound.\footnote{The observation that monomorphemic expression that lexically rule out stronger alternatives are unattested across languages has received substantial argumentative support (most prominently in \citealt[252-267]{horn:1984} but also e.g.
-in \citealt{horn:1972,traugott:2004,vdAuwera:2010}).
-To the best of our knowledge, this claim stands unchallenged.} Should the explanation rest on purely functional grounds, we see four conditions that may pressure languages for English-like semantics over this alternative.
+Compare this with a hypothetical language lexicalizes two expressions instead of a single scalar one; one with and one lacking an upper-bound.
+Should the explanation rest on purely functional grounds, we see four conditions that may pressure languages for English-like semantics over this alternative.
 First, contextual cues are strongly reliable.
@@
 
-\hl{More  discussion about parameter interaction}
+\hl{More  discussion about parameter interaction: changes in sequence length influence the population in a predictable way: smaller values lead to more heterogeneous populations whereas larger ones lead to more pronounced differences.
+This is expected insofar as the likelihood that a sequence of length $1$ was produced by any type is relatively uniform (modulo prior) whereas the likelihood of types with lexica $L_1$ - $L_3$ to produce, for instance, a sequence of $10$ observations consistently with the same state-message combination is less likely than for pragmatic players using $L_4$ - $L_6$ or literal $L_4$.
+Thus, while noteworthy, sequence length has no direct bearing on the main contrast of interest.
+Similar considerations hold for $\alpha$ and $\lambda$ -- set to $1$ and $50$ in the following.
+Overall, lower rationality in $\lambda$ or more pragmatic violations in $\alpha$ lead to a higher selection of lexica with semantic upper-bounds.
+The fitness of pragmatic behavior increases with higher $\lambda$-/$\alpha$-values.
+In other words, these parameters level the functional contrast between $L_4$ and $L_5$.
+}
 
@@
 \subsection{Discussion}
-changes in sequence length influence the population in a predictable way: smaller values lead to more heterogeneous populations whereas larger ones lead to more pronounced differences.
-This is expected insofar as the likelihood that a sequence of length $1$ was produced by any type is relatively uniform (modulo prior) whereas the likelihood of types with lexica $L_1$ - $L_3$ to produce, for instance, a sequence of $10$ observations consistently with the same state-message combination is less likely than for pragmatic players using $L_4$ - $L_6$ or literal $L_4$.
-Thus, while noteworthy, sequence length has no direct bearing on the main contrast of interest.
-Similar considerations hold for $\alpha$ and $\lambda$ -- set to $1$ and $50$ in the following.
-Overall, lower rationality in $\lambda$ or more pragmatic violations in $\alpha$ lead to a higher selection of lexica with semantic upper-bounds.
-The fitness of pragmatic behavior increases with higher $\lambda$-/$\alpha$-values.
-In other words, these parameters level the functional contrast between $L_4$ and $L_5$.
+Broadly speaking these results suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome selective pressures and stabilize in a population provided there is a bias for simpler representations.
+This outcome is particularly encouraging in light of the other advantages a lack of semantic upper-bounds may confer that were discussed above.
+\hl{The model gives a justification for lexicalization patterns found in natural language, as well as for the failure to lexicalize certain pragmatic inferences.
+That is, while the puzzle raised by semantics is hard to explain by purely functionalmeans,  we  suggest  part  of  the  answer  to  lie  in  learnability.
+Simpler semantic representations are more likely to be learned,  and pragmatic reasoning can counteract functional disadvantages otherwise incurred.
+This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics and offers an account of why (certain) pragmatic inferences are not part of the literal meaning of expressions.
+It furthermore leaves open the possibility of such inferences to fossilize when they do not compete against a lexical simplicity bias}
 
-- There is experimental evidence that scalar implicature calculation costs effort and takes additional processing time (Schaeken, Snedeker, ...).
+The model predicts this result to hinge on four conditions.
+First, types need to be pressured toward both expressivity and learnability.
+Second, language use can not be too unpredictable; low $\alpha$ or $\lambda$ values render non-upper-bounded lexical meanings more prone to communicative failure and more difficult to learn.
+Third, learners should prefer simpler over more complex lexical representations.
+Fourth, the strength by which they need to be preferred depends on the inferential mechanism of learners.
+That is, to which degree learners maximize the posterior.
+Put differently, high rationality in learning and choice requires a weaker bias towards simpler representations.
+The selection of lexical meanings lacking upper-bounds is robust against parameter perturbations under these conditions.
+
+A large range of parameter values lead to the stable incumbency of scalar implicatures.
+However, while this is predicted by the literature, it is less clear to what proportion other types should be represented.
+On the one hand, there are  reasons to expect functionally suboptimal types $L_1$-$L_3$ to be largely ruled out.
+They fail to enable their users to communicate without substantial error both amongst themselves and with other types.
+In short, these lexica stand at odds with past theoretical and empirical observations.
+On the other, this is not so for $L_4$.\footnote{$L_6$ presents a special case.
+In our current setup, it mirrors $L_5$ in allowing for pragmatic strenghtening of message $m_2$ rather than $m_1$.
+However, its association of $s_1$ with $m_1$ and $s_2$ with $m_2$ under favourable parameter conditions is achieved by ruling out the ``some but not all''-state $s_1$ and not, as with scalar implicatures, the ``all''-state $s_2$.
+$L_6$ speakers thusly have a ``some''-message strengthened to convey ``all but not some but not all''.
+The observation that monomorphemic expression that lexically rule out stronger alternatives are unattested across languages has received substantial argumentative support (most prominently in \citealt[252-267]{horn:1984} but also e.g.
+in \citealt{horn:1972,traugott:2004,vdAuwera:2010}).
+However, our present setup is blind to these differences.} Notwithstanding, it is possible that natural language communities are composed of such mixed populations or that a single speaker may entertain $L_4$-like semantics for one scalar expression and $L_5$-like semantics for another \hl{CITATIONS FOR THIS PURPOSE}.
+
+Lastly, there are a number of extensions one may want to consider.
+For instance, the consideration of larger lexica that consider epistemic uncertainty.
+Another interesting additon relates to possible disadvantages of pragmatic reasoning and the effect of state frequencies on fossilization.
+We tacitly assumed pragmatic reasoning to come at no cost.
+However, there is experimental evidence that suggests  that the pragmatic derivation of upper-bounds costs effort and takes additional processing time (cf.
+\citealt{deNeys+schaeken:2007, huang+snedeker:2009}.
+This raises the question at which point such usage-based cost undercuts the learnability advantage of simpler semantic representations.
+As noted above, this might also depend on a given scalar expression and its frequency.
+Frequently drawn scalar implicatures might fossilize to avoid cost, while infrequent ones could still be computed on-line.
 
 \section{General discussion}
+The model gives a justification for lexicalization patterns found in natural language, as well as for the failure to lexicalize certain pragmatic inferences.
+That is, while the puzzle raised by semantics is hard to explain by purely functional means, we suggest part of the answer to lie in learnability.
+Simpler semantic representations are more likely to be learned, and pragmatic reasoning can counteract functional disadvantages  otherwise incurred.
+This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics and offers an account of why (certain) pragmatic inferences are not part of the literal meaning of expressions.
+It furthermore leaves open the possibility of such inferences to fossilize when they do not compete against a lexical simplicity bias.
+
+A virtue of this model is that it allows for analysis specific modifications and extensions.
+Straightforward extensions include larger hypothesis spaces, as well as larger or different lexicon fragments.
+Another worthwhile modification relates to possible disadvantages of pragmatic reasoning.
+We tacitly assumed pragmatic reasoning to come at no cost.
+However, there is experimental evidence for the assumption that the pragmatic derivation of upper-bounds costs effort and takes additional processing time (cf.
+\citealt{deNeys+schaeken:2007, huang+snedeker:2009}).
+This raises the question at which point such usage-based cost undercuts the learnability advantage of simpler semantic representations.\footnote{In the present setup this modification has a straightforward effect: A penalty for pragmatic signaling lowers the fitness of pragmatic types, to the advantage of literal types.
+However, the penalty needs to be substantial to counteract the functional advantage pragmatic $L_5$ has over all but $L_4$ together with its learning advantage.}
 
@@
 
-{\bf Extensions:}
+\section{Conclusion}
 
-\subparagraph{(I) Cost for pragmatic reasoning.} At least in the CogSci setup the effect of adding cost to pragmatic reasoning is unsurprising: High cost for pragmatic signaling lowers the prevalence of pragmatic types.
-Lexica that semantically encode an upper-bound benefit the most from this.
-However, the cost needed to be substantial to make the pragmatic English-like lexicon stop being the incumbent type (particularly when learning is communal).
-
-\subparagraph{(II) Negative learning bias.} Instead of penalizing complex semantics (semantic upper-bounds) one may consider penalizing simple semantics (no upper-bounds).
-This is useful as a sanity check but also yields unsurprising results in the CogSci setup: The more learners are biased against simple semantics, the more prevalent are lexica that semantically encode upper-bounds.
-
-\subparagraph{(III) Inductive bias.} A second learning bias that codifies the idea that lexica should be uniform, i.e.
-be biased towards either lexicalizing an upper-bound for all weaker alternatives in a scalar pair or for none.
-
-\subparagraph{(IV)  Uncertainty.} The other advantage of non-upper bounded semantics lies in being non-committal to the negation of stronger alternatives when the speaker is uncertain.
-Adding this to the model requires the most changes to our present setup and some additional assumptions about the cues available to players to discern the speaker's knowledge about the state she is in.
-
-\subparagraph{(V) More scalar pairs.} Taking into consideration more than one scalar pair.
-Preliminary results suggest that this does not influence the results in any meaningful way without further additions, e.g.
-by (III).
-
-\subparagraph{(VI) More lexica.} Not necessary. Preliminary results suggest that considering more lexica has no noteworthy effect on the dynamics (tested with all possible 2x2 lexica).
-
-\subparagraph{(VII) State frequencies.} Variations on state frequencies. This may have an interesting interaction with (III).
-
-\subparagraph{(VIII) Reintroduction of communal learning.} One possibility: The probably $N_{ij}$ with which a child of $t_i$ adopts $t_j$ could be the weighted sum of $Q_{ij}$ (as before) and a vector we get from learning from all of the population: $L_j = \sum_d P(d | \vec{p})  P(t_j | d)$, where $P(d | \vec{p}) = \sum_{i} P(d | t_i)  \vec{p}_i$ is the probability of observing $d$ when learning from a random member of the present population distribution.
-
-\section{Conclusion}
+\hl{The development of natural languages is driven by complex intertwined pressures.
+Drawing from past insights we put forward a model that allows for a general and malleable integration of core aspects involved in this process.
+Chiefly, the model combines functional pressure, iterated Bayesian learning, and probabilistic speaker and hearer models.
+In particular,  this analysis addresses longstanding issues concerning
+the  semantics-pragmatics  divide.
+It  shows  that  when  pressured for learnability and expressivity, the former force drives for simpler semantic representations inasmuch as pragmatics can compensate for them in language use.
+Consequently, semantic  patterns  can  be  explained  in  virtue  of  the  linguistic behavior  of  their  users  and  their  complexity.
+Furthermore, the ease of acquisition of simpler semantics offers an answer to why natural languages do not lexicalize certain pragmatic inferences.}
 

### 2016-09-28 paper/paper.tex
minor changes in paper draft

@@
 \subsection{Replication \& expressivity}
-Communicative efficiency, or expressivity, has received particular attention from investigations using evolutionary game theory \citep{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}.
-Under this view, a type's success in communication confers it a higher fitness relative to less successful ones.
-As a consequence they replicate more than other types, increasing their proportion in the population.
-This association of a type's communicative success within a population with changes in the types present in it creates a feedback loop that pressures the population towards greater expressivity.
+Expressivity has received particular attention from investigations using evolutionary game theory \citep{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}.
+Under this view, a type's communicative success confers it a higher fitness, a measure that is relative to the success of other types in the population.
+In the simplest models fitness directly translates into the proportion of types present in the subsequent population.
+As a consequence, fitter types flourish while unfit ones are driven out.
+This association of communicative success within a population with changes in the proportion of types present in it creates a feedback loop that pressures the population towards greater expressivity.
 The replicator equation gives us the means to make these dynamics precise.
 
-The proportion of types in a given population is captured by a vector $x$, where $x_i$ is type $i$'s proportion in the population.
+The proportion of types in a given population is codified in a vector $x$, where $x_i$ is type $i$'s proportion.
 The fitness of a type $i$, $f_i$, is given by its expected utility in this population, $f_i = \sum_j x_j \text{EU}(t_i,t_j)$.
-That is, its fitness is the sum of its expected communicative success with other types weighted by the latter type's population share.
-The expected utility of $i$ and $j$ is obtained by considering the expected utility of speaker $i$ interacting with hearer $j$, and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$.
-$U_S(x,y)$ and $U_R(x,y)$ are respectively $\sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$ and $U_S(y,x)$ for $n$ and $o$ being the reasoning level of $x$ and $y$, and $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.\footnote{Note that the definition of $U_R(\cdot,\cdot)$ implies equal sender and receiver payoff in an interaction.
-This need not be so in the general case but suffices for our application.} This quantity is symmetric, reflecting the probability of two types' mutual understanding.
-Lastly, the average fitness of the population is captured by $\Phi$, $\Phi = \sum_i x_i f_i$.
-This term serves as a normalizing constant for the (discrete) replicator equation; $\dot{x}_i = \frac{x_i f_i}{\Phi}$
+That is, its fitness is the sum of its expected communicative success in interacting with other types weighted by the latter's population share.
+The expected utility of $i$ and $j$ is obtained by considering the expected utility of speaker $i$ interacting with hearer $j$ and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$.
+$U_S(x,y)$ and $U_R(x,y)$ are $\sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$ and $U_S(y,x)$, respectively, for $n$ and $o$ being the reasoning level of $x$ and $y$, and $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.
+This quantity is symmetric, reflecting the probability of two types' mutual understanding.
+Lastly, the average fitness of the population is given by $\Phi$, $\Phi = \sum_i x_i f_i$.
+This term serves as a normalizing constant for the (discrete) replicator equation: $\dot{x}_i = \frac{x_i f_i}{\Phi}$
 
-Under its biological interpretation, the replicator equation captures the idea of fitness-relative selection whereby fitter types produce more offspring, leading to their propagation in subsequent generations.
-In analogy to this kind of replication, many aspects of natural language are subject to processees of transmission and change across varied time-spans.
-For example, the replicator equation can be understood as a learning across generations as e.g.
+Under its biological interpretation the replicator equation captures the idea of fitness-relative selection whereby fitter types produce more offspring, leading to their propagation in subsequent generations.
+Similarly, many aspects of natural language are subject to processees of transmission and change across varied time-spans.
+Amongst others, the replicator equation can be understood as a model of language acquisition across generations, as e.g.
 in \citealt{nowak+etal:2002}, but also as a process of horizontal adaptation (see \citealt[\S3.3]{benz+etal:2005b} for discussion).
-In the following, we take the latter view in assuming that interlocutors adapt their lexica and their use to that which works best within their population.
-It should be stressed, however, that the model itself is compatible with either view.
+In the following, we take the latter view in assuming that interlocutors may adapt their types to that which works best within their population.
 
-Nowak et al.
-did not only consider replication, but also recognized the important role of the variation that is introduced by a language's transmition across generations, construed as mutation.
-Due to this process, the offspring of a type may end up adopting a different type than that of its parent.
-Crucially, in this work mutation rates were modelled as begin independent from a type.
-This means that the variation introduced by generational turnovers did not depend on factors such as the relative learnability of a type.
-To address the issue of selecting a particular type over (near) functional equivalents, we turn to a different strand of research in cultural evolution: {\em iterated learning}.
+In their series of papers on language evolution, Nowak and colleagues did not only consider replication, but also recognized the central role that variation plays in its transmission.
+In keeping the analogy to evolutionary processes, this variation can be likened to mutation to the effect that a type's offspring may adopt a different type than that of its parent.
+However, mutation rates were modeled as independent from a type \hl{CITATION} meaning that the variation introduced by generational turnovers did not depend on factors such as the relative learnability of a type.
+In the following, we interpret mutation as learnability to address the issue of selecting a particular type over (near) functional equivalents.
+For this purpose, we turn to a different strand of research in cultural evolution: {\em iterated learning}.
 
@@
 Iterated learning is a process in which the behavior of one individual serves as learning input for another, who's behavior subsequently serves as input for a new learner, and so on.
-For linguistic purposes this process can be thought of as chains of parents and children, where the parent produces linguistic data from which the child infers a language.
+In language this process can be thought of as a progression through chains of parents and children where the parent produces linguistic data from which the child infers a language.
 The latter, now a parent, goes on to produce linguistic data for a new generation of na\"ive learners.
-Following \citet{griffiths+kalish:2007} we model learning as a process of Bayesian inference in which learners combine the likelihood of a type producing the learning data with prior inductive biases.
+Following \citet{griffiths+kalish:2007} we model iterated learning as a process of Bayesian inference in which learners combine the likelihood of a type producing the learning data with prior inductive biases.
 They then select a type to adopt from the resulting posterior distribution.
 
-Due to the pressure towards learnability it exherts, iterated learning generally leads to simpler and more regular languages (surveys of empirical data and models are given in \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016}).
-Importantly, experimental and mathematical results suggest the results of this process to reflect learners' learning biases, codified in the following as a prior $P \in \Delta(\mathcal{T})$.
-A way to think about this bias is as the amount of data a learner would require in order to adopt a language -- or, in our case, a combination of a lexicon and a signaling behavior (cf.
+Due to the pressure towards learnability it exherts, iterated learning generally leads to simpler and more regular languages (see \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016} for recent surveys).
+Importantly, experimental and mathematical results suggest the outcome of this process to reflect the learners' a priori biases.
+In a Bayesian setting these biases are codified in the prior $P \in \Delta(\mathcal{T})$.
+A way to think about this prior is as the amount of data a learner would require in order to adopt a language (cf.
 \citealt[450]{griffiths+kalish:2007}).
-Crucially, the extent of the prior's influence has been shown to strongly depend on the learning strategy assumed to underly the inference process.
+Or, in our case, a combination of a lexicon and a signaling behavior.
+\hl{INTRODUCE SIMPLICITY BIAS HERE}
+
+The extent of the prior's influence has been shown to strongly depend on the learning strategy assumed to underly the inference process.
 While simulation results suggested that weak biases could be magnified by exposing learners to only small data samples \citep{brighton:2002}, the mathematical characterization provided by \citet{griffiths+kalish:2007} showed that, instead, iterated learning converged to the prior.
-That is, the distribution over languages in a population or, from an individual's perspective, the likelihood of learning a language corresponds to the learners' prior distribution, irrespective of the amount of input given to learners.
+That is, the distribution over languages in a population, or the likelihood of learning a language when taking an individual's perspective, corresponds to the learners' prior distribution irrespective of the amount of input given to learners.
 This divergence in predictions can be traced back to differences in the selection of hypotheses from the posterior.
-On the one extreme, Griffith \& Kalish's convergence to the prior holds for learners that sample from the posterior.
+On the one extreme, Griffith \& Kalish's prediction of convergence to the prior holds for learners that sample from the posterior.
 On the other, more deterministic strategies such as the selection of the type with the highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase the prior's influence \citep{griffiths+kalish:2007,kirby+etal:2007}.
-In the following, we parametrize the posterior, $P(t_i|d)^l$, to obtain a range of learning strategies that live in the range between posterior sampling and MAP, $l \geq 1$.
+In the following, we parametrize the posterior, $P(t_i|d)^l$, to obtain a range of learning strategies that live between posterior sampling and MAP, $l \geq 1$.
 When $l = 1$ learners sample from the posterior.
-As $l$ increases towards infinity, the learners' tendency maximize the posterior increases.
+As $l$ increases towards infinity, the learners' propensity to maximize the posterior increases.
 
-More generally, we combine the replicator dynamics with iterated learning by codifying the latter as a transition matrix $Q$.
+We combine the replicator dynamics with iterated learning by codifying the latter as a transition matrix $Q$.
 Just as in standard mutator dynamics, $Q_{ij}$ indicates the probability of the children of a parent of type $i$ adopting type $j$.
-However, to make this process depend on a type's learnbility, this quantity is proportional to the probability of $i$ producing the learning data and that of $j$ given the data.
+However, to make this process depend on a type's learnbility, this quantity is proportional to the probability of $i$ producing the learning data and that of inferring $j$ given the data.
 
-The elements of the set of learning data $D$ are sequences of length $k$ of state-message pairings.
+The elements of the set of learning data $D$ are $k$-length sequences of state-message pairings.
 That is, a sequence of observations of language use.
-Put differently, a datum $d \in D$ contains $k$ members of the set $\{\tuple{s_i,m_j} | s_i \in S, m_j \in M\}$ and $D$ is the set of all such sequences.
-Having fixed $D$,
+Put differently, a datum $d \in D$ contains $k$ members of the set $\{\tuple{s_i,m_j} | s_i \in S, m_j \in M\}$.
+$D$ is the set of all such sequences.
+With the at hand $D$, the learning matrix is constructed as follows:
 \[
@@
 \subsection{Summary}
-We argued for expressivity, learnability and simplicity as important pressures that apply on the cultural evolution of language.
+We argued that expressivity, learnability and simplicity are central pressures that apply on the cultural evolution of language.
 They are respectively modelled as communicative efficiency-relative replication, iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings.
-Taken together these evolutionary dynamics are described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:
+Taken together their dynamics can be described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:
 
 \[ 
-\hat{x_i} = \sum_j Q_{ji} \frac{x_jf_j}{\Phi}
+\hat{x}_i = \sum_j Q_{ji} \frac{x_jf_j}{\Phi}
 \]
 
-The basic units that the dynamics operate on are a combination of lexica lexicon and their use; a type.
+The units that the dynamics operate on are a combination of a lexicon and linguistic behavior which determines its use, i.e., a type.
 A type's expressivity depends on its communicative efficiency within a population while its learnability depends on the fidelity by which it is infered by new generations of na\"ive learners.
 
-In sum, the innovation of this model lies in its combination of functional pressure on successful communication, effects of learning biases on (iterated) Bayesian language learning \citealt{griffiths+kalish:2007}, and probabilistic models of language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014,bergen+etal:2016}.
+%In sum, the innovation of this model lies in its combination of functional pressure on successful communication, effects of learning biases on (iterated) Bayesian language learning \citealt{griffiths+kalish:2007}, and probabilistic models of language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014,bergen+etal:2016}.
 In particular, this synthesis enables for the investigation of the effects of communicative pressures on the semantics-pragmatics interface.

### 2016-09-29 paper/paper.tex
minimal draft changes

@@
 
-\section{Lack of lexicalization of pragmatic inferences}
-
+\section{Scalar implicatures}
 %
-With this model, we set out to investigate the prevalence of lexical meanings that allow for regular pragmatic enrichments over other alternatives.
-A particularly well-studied type of conventional pragmatic enrichment are so-called {\em scalar implicatures}.
+A particularly well-studied type of conventional pragmatic enrichments are so-called {\em scalar implicatures}.
 These inferences are licensed for groups of expressions ordered in terms of informativity, here understood as an entailment induced order.
-For instance, {\em some} is entailed by {\em all}; if it were true that `All students came to class', it would also be true that `Some students came to class'.
-However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not necessarily what their use is taken to convey.
+For instance, {\em some} is entailed by {\em all}.
+If it were true that `All students came to class', it would also be true that `Some students came to class'.
+However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what their use is normally taken to convey.
 Instead, the use of a less informative expression when a more informative one could have been used can license a defeasible inference that stronger alternatives do not hold (cf.
@@
 In this way, `Some students came to class' is strengthened to convey `Some but not all students came to class'.
-Analogously, a speaker can rely on her interlocutor to draw this inference without having to express this upper-bound overtly, e.g.
-by stating {\em some but not all}.
-In other words, mutual reasoning about rational language use supplies a bound that rules out stronger alternatives pragmatically.
+Conversely, speakers can rely on their interlocutors to draw this inference.
+In a nutshell, mutual reasoning about rational language use supplies a bound that rules out stronger alternatives pragmatically which is absent from the semantics of weak scalar expressions.
+\hl{This entire par needs to be more reader friendly}
 
-This corresponds to our previous description of the pragmatic use of lexicon $L_a$, repeated below for convenience.
-A pragmatic hearer who reasons about a speaker's use of message $m_1$ will associate it more strongly with $s_1$ than with $s_2$ given that the latter is unambiguously associated with $s_2$.
-The strength of this association dependens on the individuals' degree of rationality $\lambda$ and their prior over states.
+This kind of strengthening corresponds to our previous description of the pragmatic use of lexicon $L_a$.
+A pragmatic hearer who reasons about a speaker's use of message $m_1$ will associate it more strongly with $s_1$ than with $s_2$ since the latter is already unambiguously associated with $s_2$.
 Conversely, a pragmatic speaker will reason about her interlocutor's interpretation and use the messages accordingly.
@@
 
-Our initial question can now be rephrased in terms of scalar implicatures by asking for justifiations for the lack of lexical upper-bounds in weak scalar alternatives.
-That is, why semantics such as those of message $m_1$ in $L_a$ are regularly selected for over the alternative of lexicalizing it as in $L_b$.
-More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} or {\em big} were truth-conditionally incompatible with stronger alternatives such as, respectively, {\em hot}, {\em and}, {\em all} and {\em huge}?
-This question is particularly striking considering the number of expressions that license such inferences across languages (\citealt{horn:1972}, \citealt[252-267]{horn:1984}, \citealt{traugott:2004,vdAuwera:2010}).
+Our initial question can now be narrowed in terms of scalar implicatures by asking for a justification for the lack of lexical upper-bounds in weak scalar alternatives.
+That is, why semantics such as those of message $m_1$ in $L_a$ are regularly selected for over the alternative of codifying the bound semantically as in $L_b$.
+More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} and {\em big} were truth-conditionally incompatible with stronger alternatives such as {\em hot}, {\em and}, {\em all} and {\em huge}?
+This question is particularly striking considering the number of expressions that license such inferences across languages and thusly lack a lexicalized upper-bound (\citealt{horn:1972}, \citealt[252-267]{horn:1984}, \citealt{traugott:2004,vdAuwera:2010}).
 
 We see two main explanations for the lack of upper-bounds in the lexical meaning of weak scalar expressions.
-The first is that their truth-conditional compatibility with stronger expressions endows them a broader range of application.
-Scalar expressions occur in contexts in which their upper-bounded reading is absent.
+The first is that their truth-conditional compatibility with stronger expressions endows them with broader applicability, allowing them to occur in contexts in which their upper-bounded reading is absent.
 This can happen when embedded in downward-entailing contexts, when the speaker is likely uncertain about whether the upper bounded reading is true, or when the distinction between an upper-bounded reading and the simple, only lower-bounded reading, is not relevant.
-For instance, if for all the speaker knows `Some students came' but she doesn't know whether `All came', then the use of {\em some} succinctly conveys her uncertainty about the latter.
-This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then these means are preferred over expressing it overtly through a longer expression, e.g., by stating `some but not all' explicitly.
-Importantly, although dispreferred due to its relative length and complexity, morphosyntactic disambiguation still allows speakers to enforce an upper-bound to override contextual cues that might otherwise mislead the hearer.
+For instance, if for all the speaker knows `Some students came' but she doesn't know whether `All came' then the use of unbounded {\em some} succinctly conveys her uncertainty about the latter.
+This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then this is preferred over expressing it overtly through longer expressions.
+For example by saying {\em some but not all} explicitly.
+Importantly, although morphosyntactic disambiguation is dispreferred due to its relative length and complexity, it allows speakers to enforce an upper-bound and override contextual cues that might otherwise mislead the hearer.
 
 In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so.
-Compare this with a hypothetical language lexicalizes two expressions instead of a single scalar one; one with and one lacking an upper-bound.
-Should the explanation rest on purely functional grounds, we see four conditions that may pressure languages for English-like semantics over this alternative.
+Compare this with a hypothetical language that lexicalizes two expressions for each English scalar expression; one with and one lacking an upper-bound.
+Along this functionalist explanation, we see four conditions that may pressure languages for English-like semantics over this alternative.
 First, contextual cues are strongly reliable.
@@
 Fourth, larger lexica are costly.
+Overall, neither condition is convincing in its role as a central explanatory devices for such a wide-spread phenomenon.
+The first two put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of unambiguous communication.
+It is likely that human language users are very good at retrieving cues from contexts, but to stipulate that they are so good as to undercut the benefit of safe communication provided by our hypothetical alternative strikes us as too strong of an assumption.
+As for the third and fourth condition, these seem mostly like technical solutions without a proper empirical basis.
 
-However, these conditions are not convincing in their role as central explanatory devices for such a wide-spread phenomenon.
-The first two put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of .
-It is likely that human language users are very good at retrieving cues from contexts, but to stipulate that they are so good as to undercut the benefit of safe communication provided by the hypothetical alternative strikes us as too strong of an assumption.
-As for conditions (3) and (4), these seem mostly like technical solutions, without a proper empirical basis.
-
-Instead, in what follows we investigate the hypothesis that the lack of lexicalization of scalar inferences is driven by the advantage in compression that lexical meanings lacking an upper-bound have over those that explicitly codify it.
+\hl{All the rest of this section needs to be tidied up} Instead, in what follows we investigate the hypothesis that the lack of lexicalization of scalar inferences is driven by the advantage in compression that lexical meanings lacking an upper-bound have over those that explicitly codify it.
 Note however that we do not represent this contrast in compression between lexical meanings explicitly in lexica.

### 2016-10-01 paper/paper.tex
changes in draft discussion+conclusion

@@
 
-In recent years questions about the emergence and stability of linguistic features have lead to a surge in models that analyze their transmission within and across generations (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
-Our starting point is given by the overarching argument that has crystalized from the accumulated mathematical, experimental and cross-linguistic evidence in this literature: Natural languages need to be well-adapted to communicative needs within a linguistic community, but also need to be learnable to survive their faithful transmission across generations.
-More succinctly; natural languages are pressured for expressivity and learnability.
-
-We build on these insights by modeling these pressures using the replicator-mutator dynamics (see \citealt{hofbauer+sigmund:2003} for an overview).
-This allows for the inspection of their interaction by combining functional pressure on successful communication with effects of learning biases on (iterated) Bayesian learning \citep{griffiths+kalish:2007}.
-The semantics-pragmatics distinction and its effect on production and comprehension are made precise by considering probabilistic models of rational language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
-\hl{SOMETHING MISSING HERE} \hl{In this way, the model establishes a link between models of synchronic probabilistic rational language use and diachronic models of cultural evolution.
-}
-
-%We then analyze the prevalence of a lack of semantic upper-bounds in the literal meaning of weak scalar expressions.
-We show that a lack of upper-bounds is selected for when learners are biased towards simpler semantic representations, provided they have means to convey upper-bounds, i.e., provided that they can be derived via pragmatic reasoning.
-
 The emergence and change of linguistic structure is influenced by many intertwined factors.
@@
 
-At latest since \citeposs{zipf:1949} rationalization of the observation that word frequency rankings can be approximated by a power law distribution as competing hearer and speaker preferences, the idea that linguistic change is influenced by communicative pressures has played a pivotal role in synchronic and diachronic analyses (e.g.
+The idea that linguistic change is influenced by communicative pressures has played a pivotal role in synchronic and diachronic analyses at latest since \citeposs{zipf:1949} rationalization of the approximability of word frequency rankings by a power law distribution as competing hearer and speaker preferences (e.g.
 \citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007, piantadosi:2014,kirby+etal:2015}).
+In recent years this line of research has led to a surge in models that analyze communicative pressures within and across generations (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
+Our starting point is given by the overarching argument that has crystalized from the accumulated mathematical, experimental and cross-linguistic evidence in this literature: Natural languages need to be well-adapted to communicative needs within a linguistic community, but also need to be learnable to survive their faithful transmission across generations.
+More succinctly; natural languages are pressured for expressivity and learnability.
 
-As noted above, expressivity and learnability have been identified as two central competing pressures.
-Their opposition becomes particularly clear when considering their consequences in the extreme (cf.
+The opposition of expressivity and learnability becomes particularly clear when considering their consequences in the extreme (cf.
 \citealt{kemp+regier:2012,kirby+etal:2015}).
-On one extreme, a language with a single form is easy to learn but lacking in expressivity.
+On one end, a language with a single form is easy to learn but lacking in expressivity.
 On the other, a language that associates a distinct form with all possible meanings its users may want to convey is maximally expressive but challenging to acquire.
@@
 However, this so-called transmission bottleneck is not the only challenge learners confront.
-Instead, a problem that is more important for our present purpose is that of selecting particular hypotheses out of a (potentially infinite) space of alternatives compatible with the data learners are exposed to.
+A more pressing problem for our purpose is that of selecting particular hypotheses out of a (potentially infinite) space of alternatives compatible with the data learners are exposed to.
 At the semantics-pragmatics interface this concerns the selection between functionally similar, if not identical, lexical meanings.
-In the following, we argue an integral part of the answer to be that learners are a priori biased towards simpler, more compressed, lexical representations.
+We assume an integral part of the answer to be that learners are a priori biased towards simpler, more compressed, lexical representations.
 That is, rational learners should prefer simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
-In linguistics, a drive for simplicity has been argued to underpin speaker preferences for brevity and ease of articulation, as well as to pressure languages towards lexical ambiguity and grammatical compression \citep{zipf:1949,grice:1975,piantadosi+etal:2012, kirby+etal:2015}.
-\hl{As a broader cognitive principle, the use of simplicity as means to select between hypotheses has a long standing tradition.
-Crucially, \citet{chater+vitanyi:2003} give a number of compelling arguments for simplicity on both mathematical and empirical grounds.}
 
-The remainder of this section introduces the individual components of the model in more detail, as well as the assumptions underlying them.
-These are: (i) the representation of languages and their use, (ii) pressures towards expressivity and learnability, regulated by the replicator and mutator dynamics, respectively, as well as (iii) a bias towards simpler semantic representations, codified as a language learner's prior.
-After laying out the model, we discuss its application to the lack of lexicalization of scalar implicatures.
+In the following, we model these pressures using the replicator-mutator dynamics by combining functional pressure on successful communication with effects of learning biases on (iterated) Bayesian learning \citep{griffiths+kalish:2007}.
+The semantics-pragmatics distinction and its effect on production and comprehension are made precise through probabilistic models of rational language use in populations and multiple lexica \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
+The remainder of this section introduces the individual components of the model as well as the assumptions underlying them.
+These are: the representation of languages and their use (\S\ref{sec:languages+use}), pressures towards expressivity (\S\ref{sec:expressivity}) and learnability (\S\ref{sec:learnability}), regulated by the replicator and mutator dynamics, respectively, as well as a bias towards simpler semantic representations, codified as the learners' prior.
+After laying out the model, we analyze its predictions in a case-study on the lack of lexicalization of scalar implicatures in \S\ref{sec:si-case-study}.
 
-\subsection{Lexica and linguistic behavior}
+\subsection{Lexica and linguistic behavior}\label{sec:languages+use}
 Lexica codify the truth-conditions of a language's expressions, i.e., its semantics.
@@
 
-\subsection{Replication \& expressivity}
+\subsection{Replication \& expressivity}\label{sec:expressivity}
 Expressivity has received particular attention from investigations using evolutionary game theory \citep{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}.
-Under this view, a type's communicative success confers it a higher fitness, a measure that is relative to the success of other types in the population.
+Under this view, a type's ability to convey and interpret information successfully confers it a higher fitness, this measure being relative to the success of other types in the population.
 In the simplest models fitness directly translates into the proportion of types present in the subsequent population.
@@
 That is, its fitness is the sum of its expected communicative success in interacting with other types weighted by the latter's population share.
-The expected utility of $i$ and $j$ is obtained by considering the expected utility of speaker $i$ interacting with hearer $j$ and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$.
+The expected utility of $i$ and $j$ is obtained by considering the average communcative success of $i$ conveying information to $j$ and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$.
 $U_S(x,y)$ and $U_R(x,y)$ are $\sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$ and $U_S(y,x)$, respectively, for $n$ and $o$ being the reasoning level of $x$ and $y$, and $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.
 This quantity is symmetric, reflecting the probability of two types' mutual understanding.
-Lastly, the average fitness of the population is given by $\Phi$, $\Phi = \sum_i x_i f_i$.
+The average fitness of the population is given by $\Phi$, $\Phi = \sum_i x_i f_i$.
 This term serves as a normalizing constant for the (discrete) replicator equation: $\dot{x}_i = \frac{x_i f_i}{\Phi}$
@@
 
-In their series of papers on language evolution, Nowak and colleagues did not only consider replication, but also recognized the central role that variation plays in its transmission.
-In keeping the analogy to evolutionary processes, this variation can be likened to mutation to the effect that a type's offspring may adopt a different type than that of its parent.
-However, mutation rates were modeled as independent from a type \hl{CITATION} meaning that the variation introduced by generational turnovers did not depend on factors such as the relative learnability of a type.
-In the following, we interpret mutation as learnability to address the issue of selecting a particular type over (near) functional equivalents.
+In their series of papers on language evolution, Nowak and colleagues did not only consider expressivity but also recognized the central role that learnability plays in its transmission.
+On the one hand, linguistic production can be prone to errors \citep{nowak+krakauer:1999}.
+On the other, multiple languages may be compatible with the data learners are exposed to \citep{nowak+etal:2002}.
+Both sources introduce variation in language acquistion.
+In keeping the analogy to evolutionary processes this variation can be likened to mutation to the effect that a type's offspring may adopt a different type than that of its parent.
+Importantly, the resulting generational turnovers should depend on the relative learnability of a type.
 For this purpose, we turn to a different strand of research in cultural evolution: {\em iterated learning}.
 
-\subsection{Mutation \& learning}\label{sec:mutation}
+\subsection{Mutation \& learning}\label{sec:learnability}
 Iterated learning is a process in which the behavior of one individual serves as learning input for another, who's behavior subsequently serves as input for a new learner, and so on.
@@
 \hl{INTRODUCE SIMPLICITY BIAS HERE}
+
+\hl{ In linguistics, a drive for simplicity has been argued to underpin speaker preferences for brevity and ease of articulation, as well as to pressure languages towards lexical ambiguity and grammatical compression \citep{zipf:1949,grice:1975,piantadosi+etal:2012, kirby+etal:2015}.
+As a broader cognitive principle, the use of simplicity as means to select between hypotheses has a long standing tradition.
+Crucially, \citet{chater+vitanyi:2003} give a number of compelling arguments for simplicity on both mathematical and empirical grounds.}
 
@@
 
-\section{Scalar implicatures}
+\section{Scalar implicatures}\label{sec:si-case-study}
 %
@@
 This outcome is particularly encouraging in light of the other advantages a lack of semantic upper-bounds may confer that were discussed above.
-\hl{The model gives a justification for lexicalization patterns found in natural language, as well as for the failure to lexicalize certain pragmatic inferences.
-That is, while the puzzle raised by semantics is hard to explain by purely functionalmeans,  we  suggest  part  of  the  answer  to  lie  in  learnability.
-Simpler semantic representations are more likely to be learned,  and pragmatic reasoning can counteract functional disadvantages otherwise incurred.
-This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics and offers an account of why (certain) pragmatic inferences are not part of the literal meaning of expressions.
-It furthermore leaves open the possibility of such inferences to fossilize when they do not compete against a lexical simplicity bias}
 
@@
 
-\section{General discussion}
-The model gives a justification for lexicalization patterns found in natural language, as well as for the failure to lexicalize certain pragmatic inferences.
-That is, while the puzzle raised by semantics is hard to explain by purely functional means, we suggest part of the answer to lie in learnability.
-Simpler semantic representations are more likely to be learned, and pragmatic reasoning can counteract functional disadvantages  otherwise incurred.
-This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics and offers an account of why (certain) pragmatic inferences are not part of the literal meaning of expressions.
-It furthermore leaves open the possibility of such inferences to fossilize when they do not compete against a lexical simplicity bias.
-
 A virtue of this model is that it allows for analysis specific modifications and extensions.
@@
 
-The model combines:
-\begin{itemize}
-  \item game-theoretical models of functional pressure towards efficient communication
-  \item effects of learning biases on (iterated) language learning
-  \item assumptions about particular learning biases (Piantadosi et al. 20XX)
-  \item probabilistic speaker and listener types of various pragmatic sophistication (Frank \& Goodman, 2012; Franke \& Jäger, 2014)
-  \item speaker and listener types with different lexica (Bergen et al. 2012, to appear)
-\end{itemize}
+\section{General discussion}
+We laid out a model that combines game-theoretical models of functional pressure towards efficient communication \citep{nowak+krakauer:1999}, effects of learning biases on (iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012, franke+jaeger:2014} as well as different lexica \citep{bergen+etal:2012,bergen+etal:2016}.This model generates predictions about  lexicalization patterns found in natural language and the failure to lexicalize of certain pragmatic inferences.
+While the puzzle raised by semantics is hard to explain by purely functional means, it suggests part of the answer to lie in learnability.
+Simpler semantic representations are more likely to be learned, and pragmatic reasoning can counteract functional disadvantages otherwise incurred.
+This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics by offering an account of why (certain) pragmatic inferences are not part of the literal meaning of expressions.
+It furthermore leaves the possibility of such inferences to fossilize open when they do not compete against a lexical simplicity bias.
 
@@
 
+In contrast to past research using the replicator-mutator dynamics, the Bayesian setting afforded by iterated learning allows for a straightforward integration of a learning prior.
+Furthermore, less idiosyncratic assumptions about the variation introduced by learning without having to resort to similarity matrices between vocabularies \citep{nowak+krakauer:1999} nor languages \citep{nowak+etal:2002}.
+
 \section{Conclusion}
+Language change is affected by complex intertwined pressures.
+Drawing from past insights we put forward a model that allows for a general integration of core aspects involved in this process.
+In particular, the model combines functional pressure, iterated Bayesian learning, and probabilistic speaker and hearer models.
+We argued this combination of cultural evolution dynamics with models of probabilistic language use to be suitable for the analysis of longstanding issues concerning the semantics-pragmatics divide.
 
-\hl{The development of natural languages is driven by complex intertwined pressures.
-Drawing from past insights we put forward a model that allows for a general and malleable integration of core aspects involved in this process.
-Chiefly, the model combines functional pressure, iterated Bayesian learning, and probabilistic speaker and hearer models.
-In particular,  this analysis addresses longstanding issues concerning
-the  semantics-pragmatics  divide.
-It  shows  that  when  pressured for learnability and expressivity, the former force drives for simpler semantic representations inasmuch as pragmatics can compensate for them in language use.
-Consequently, semantic  patterns  can  be  explained  in  virtue  of  the  linguistic behavior  of  their  users  and  their  complexity.
-Furthermore, the ease of acquisition of simpler semantics offers an answer to why natural languages do not lexicalize certain pragmatic inferences.}
+In a case-study on the lack of lexicalization of scalar implicatures, the model predicts that,  when  pressured for learnability and expressivity, the former force drives for simpler semantic representations inasmuch as pragmatics can compensate for them in language use.
+Consequently, semantic patterns observed in natural languages can be explained in  virtue of the linguistic behavior exhibited by their users and their representational complexity.
+In particular, the relative ease of acquisition of simpler semantics may offer an answer to why natural languages do not lexicalize certain pragmatic inferences.
 

### 2016-10-02 paper/paper.tex
changes in paper on Kirby+etal draft

@@
 
-\begin{abstract} Certain classes of lexical meanings enable for pragmatic enrichments in a notably productive fashion.
+\begin{abstract} 
+Mutual reasoning endows speakers with the ability to convey information that goes beyond the literal meaning of expressions.
+A particularity of certain classes of lexical meanings is that they enable for such pragmatic enrichments in a notably productive fashion.
 This raises the challenge to justify their regular selection over alternatives that codify semantically what is conveyed pragmatically.
-To address this issue, we propose a general model that integrates iterated Bayesian learning in the replicator-mutator dynamics.
-This model generates predictions about the effects of linguistic pressures on selection and transmission in populations of probabilistic language users with varied degrees of pragmatic sophistication.
+That is, to justify cross-linguistically pervasive semantic structure in light of pragmatics.
+We provide an explanatory account for the roles of both domains by means of an analysis of the linguistic pressures that apply at the semantics-pragmatics interface.
+To this end, we propose a general model that combines functional pressure on language use and a learnability pressure in acquisition in populations of probabilistic language users with varied degrees of pragmatic sophistication.
 We apply this model in a case study on the (lack of) lexicalization of scalar implicatures.
@@
 
-\section{Simplicity, expressivity, and learnability at the semantics-pragmatics interface}\label{sec:introduction}
-
+\section{Introduction}\label{sec:introduction}
+\hl{This should start more generally, stressing mutual reasoning and what it brings to the table.}
 In linguistic theorizing it is common to draw a distinction between semantics and pragmatics.
@@
 
-The idea that linguistic change is influenced by communicative pressures has played a pivotal role in synchronic and diachronic analyses at latest since \citeposs{zipf:1949} rationalization of the approximability of word frequency rankings by a power law distribution as competing hearer and speaker preferences (e.g.
+The idea that linguistic change is influenced by communicative pressures has played a pivotal role in synchronic and diachronic analyses at latest since \citeposs{zipf:1949} rationalization of the approximation of word frequency rankings by a power law distribution as competing hearer and speaker preferences (e.g.
 \citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007, piantadosi:2014,kirby+etal:2015}).
@@
 \citealt{kemp+regier:2012,kirby+etal:2015}).
-On one end, a language with a single form is easy to learn but lacking in expressivity.
-On the other, a language that associates a distinct form with all possible meanings its users may want to convey is maximally expressive but challenging to acquire.
+A language with a single form is easy to learn but lacking in expressivity.
+Conversely, a language that associates a distinct form with all possible meanings its users may want to convey is maximally expressive but challenging to acquire.
 The most prominent problem that arises from this tension is that of acquiring a language to express a potentially infinite set of meanings through finite means \citep{kirby:2002}.
@@
 
-To make the distinction betweeen semantics and pragmatics precise, we distinguish between two kinds of linguistic behavior.
+While these language representations are very simple, they suffice to make precise the distinction between semantics and pragmatics.
+For this purpose we distinguish between two kinds of linguistic behavior.
 {\em Literal interlocutors} produce and interpret messages literally.
@@
 \hl{INTRODUCE SIMPLICITY BIAS HERE}
+
+\hl{Mention similarity to Piantadosi+etal's prior}
 
@@
 As in our previous examples, they are $(2,2)$-Boolean matrices, i.e., $|M| = |S| = 2$.
+These are the simplest matrices that allow us to make the presence or lack of an upper-bound and the use of scalar implicatures precise.
 Lexica $L_1$ to $L_3$ are not optimal for communication because they assign the same meaning to all their messages.
@@
 
+\hl{Summarize main predictions: Only expressivity, only learnability, both together; expressivity nudging learning in the right direction and learning leading to a uniform preference for thing in the population}
+
 Discussion about expressivity as external to learning (cf. Stadler, replicator-papers by Kenny Smith, Kirby et al 2015). Possibly add appendix with direct comparison between IL and RMD.
+
+\hl{\citet{kirby+etal:2015} propose dynamics that involve expressivity and learnability.
+In contrast to our proposal, the former pressure is not fully decoupled from the latter in that is restricted to the production of learning data as a form of ambiguity avoidance (cf.
+brighton+etal:2005b cited on p.89l, who make render such ambiguity avoidance as a ``expressivity bias'' in learning).
+That is, a measure of the degree of understanding achieved by interlocutors is not a part of the theoretical model.
+}\footnote{\hl{We could say more about this}}
 
@@
 \section{Conclusion}
-Language change is affected by complex intertwined pressures.
-Drawing from past insights we put forward a model that allows for a general integration of core aspects involved in this process.
-In particular, the model combines functional pressure, iterated Bayesian learning, and probabilistic speaker and hearer models.
-We argued this combination of cultural evolution dynamics with models of probabilistic language use to be suitable for the analysis of longstanding issues concerning the semantics-pragmatics divide.
+Language change is affected by intertwined pressures.
+Drawing from past research, we put forward a general model that integrates central aspects involved in this process by combining cultural evolution dynamics with models of probabilistic language use.
+In the case of the lack of lexicalization of scalar implicatures, the model predicts that, when  pressured for learnability and expressivity, the former force drives for simpler semantic representations inasmuch as pragmatics can compensate for their lack of expressivity in use.
+That is, the relative learnability advantage of simpler semantics may offer an answer to why natural languages do not lexicalize certain pragmatic inferences.
 
-In a case-study on the lack of lexicalization of scalar implicatures, the model predicts that,  when  pressured for learnability and expressivity, the former force drives for simpler semantic representations inasmuch as pragmatics can compensate for them in language use.
-Consequently, semantic patterns observed in natural languages can be explained in  virtue of the linguistic behavior exhibited by their users and their representational complexity.
-In particular, the relative ease of acquisition of simpler semantics may offer an answer to why natural languages do not lexicalize certain pragmatic inferences.
+More broadly, we argued that semantic patterns can be explained by taking into consideration the way in which they are used in interactions as well as their representational complexity.
+As use more often than not concerns pragmatics, this kind of analysis not only offers an explanatory account for the phenomenon at hand, but sheds light on the semantics-pragmatics distinction by elucidating their respective roles in language change.
 

### 2016-10-03 paper/paper.tex
draft additions and corrections for section 1

@@
 A particularity of certain classes of lexical meanings is that they enable for such pragmatic enrichments in a notably productive fashion.
-This raises the challenge to justify their regular selection over alternatives that codify semantically what is conveyed pragmatically.
-That is, to justify cross-linguistically pervasive semantic structure in light of pragmatics.
-We provide an explanatory account for the roles of both domains by means of an analysis of the linguistic pressures that apply at the semantics-pragmatics interface.
-To this end, we propose a general model that combines functional pressure on language use and a learnability pressure in acquisition in populations of probabilistic language users with varied degrees of pragmatic sophistication.
-We apply this model in a case study on the (lack of) lexicalization of scalar implicatures.
-The results show that simpler semantic representations are selected for when languages are pressured towards learnability and compression, provided that pragmatic reasoning can compensate for the disadvantage in expressivity that users of such languages otherwise incur.
+This raises the challenge to justify their regular selection over other alternatives that, e.g., do not require such enrichments.
+We suggest that such an account can be provided by an analysis of the linguistic pressures that apply at the semantics-pragmatics interface.
+To this end, we lay out a general model that combines functional pressure on language use and a pressure for learnability in acquisition in populations of probabilistic language users with varied degrees of pragmatic sophistication.
+We then use this model in a case study on the (lack of) lexicalization of scalar implicatures.
+The results suggest that such inferences fail to lexicalize when languages are pressured towards simpler semantic representations in learning, provided that pragmatic reasoning can compensate for the disadvantage in expressivity that users of such languages otherwise incur.
 We argue this result to shed light on the lack of lexicalization of scalar inferences, as well as the semantics-pragmatics distinction more generally.
@@
 \section{Introduction}\label{sec:introduction}
-\hl{This should start more generally, stressing mutual reasoning and what it brings to the table.}
-In linguistic theorizing it is common to draw a distinction between semantics and pragmatics.
-Broadly speaking, the former concerns the truth-conditional content of expressions whereas the latter concerns information beyond literal meanings and their composition.
-An important consequence of this distinction is that the information conveyed by an utterance is seldom, if ever, solely determined by semantics, but rather in tandem with pragmatics.
-Much research at the semantics-pragmatics interface has been aimed at characterizing expressions in terms of either domain or their interplay.
-However, an issue that has received little attention is the justification of semantic structure in light of pragmatics.
+What is conveyed usually goes beyond what is said.
+A request for a blanket can be politely veiled by uttering ``I'm cold'', a temporal succession of events conveyed by the order in which conjunts appear, as in``I traveled to Paris and got married'', or an invitation be declined by saying ``I have to work''.
+An influential explanation of the relation between the literal meaning (semantics) of such expressions and what they are intended and interpreted to convey (pragmatics) sees the latter as a product of a process of mutual reasoning \citep{grice:1975}.
+For instance, under the assumption that the speaker is cooperative and strives to be relevant, ``I have to work'' may be interpreted as providing a reason why the speaker will not be able to accept the invitation -- why would she have said so otherwise?
+
+Nowadays, it is common to draw a distinction between semantics and pragmatics in linguistic theorizing.
+Importantly, many classes of lexical meanings allow for pragmatic enrichments in a notably productive fashion.
+For instance, while John can eat all the cookies and still truthfully claim ``I ate some cookies'', the use of {\em some} over {\em all} is usually taken to convey that the latter does not hold.
+That is, ``I ate some cookies'' is pragmatically strengthened to convey that {\em some but not all} were eaten (\hl{citations for experimental evidence}).
+This inference is not restricted to quantifiers such as {\em some} and {\em all} but can be licensed for a large class of natural language expressions ranging from numerals such as {\em four} and {\em five} to adjectives such as {\em cold} and {\em big}.
+
+These pragmatic inferences, so-called {\em scalar implicatures}, have been the focus of many theoretical and experimental studies at the semantics-pragmatics interface (\hl{citations}).
+However, an issue that has received little attention is the justification of semantic structure in light of the systematic enrichment offered by pragmatics.
 The present investigation seeks to fill this gap by analyzing the effects linguistic pressures have on the selection and pervasiveness of particular lexical meanings under consideration of their possible pragmatic enrichments.
 
+\subsection{Linguistic pressures at the semantics-pragmatics interface}
 The emergence and change of linguistic structure is influenced by many intertwined factors.
@@
 Social and ecological pressures determine communicative needs, while biology determines the architecture that enables and constrains the means by which they can be fulfilled.
-Our focus lies on the latter, cultural, factor wherein processes of linguistic change are understood as shaped by language use and its transmission.
+In the following, our focus lies on the remaining cultural aspect, wherein processes of linguistic change are viewed as shaped by language use and its transmission.
 That is, as a result of a process of cultural evolution.
 
-The idea that linguistic change is influenced by communicative pressures has played a pivotal role in synchronic and diachronic analyses at latest since \citeposs{zipf:1949} rationalization of the approximation of word frequency rankings by a power law distribution as competing hearer and speaker preferences (e.g.
-\citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007, piantadosi:2014,kirby+etal:2015}).
-In recent years this line of research has led to a surge in models that analyze communicative pressures within and across generations (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
+The idea that language is influenced by communicative pressures has played a pivotal role in synchronic and diachronic analyses at latest since \citeposs{zipf:1949} rationalization of the approximation of word frequency rankings by a power law distribution as competing hearer and speaker preferences (e.g.
+in \citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007, piantadosi:2014,kirby+etal:2015}).
+In recent years this line of research has led to a surge of approaches that seek to analyze the effects of such pressures on language from multiple angles ranging from simulations to experiments with human and robotic agents (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
 Our starting point is given by the overarching argument that has crystalized from the accumulated mathematical, experimental and cross-linguistic evidence in this literature: Natural languages need to be well-adapted to communicative needs within a linguistic community, but also need to be learnable to survive their faithful transmission across generations.
@@
 The most prominent problem that arises from this tension is that of acquiring a language to express a potentially infinite set of meanings through finite means \citep{kirby:2002}.
-However, this so-called transmission bottleneck is not the only challenge learners confront.
-A more pressing problem for our purpose is that of selecting particular hypotheses out of a (potentially infinite) space of alternatives compatible with the data learners are exposed to.
-At the semantics-pragmatics interface this concerns the selection between functionally similar, if not identical, lexical meanings.
+However, this is not the only challenge learners confront.
+A more central issue to our explanandum is that of selecting particular hypotheses out of a potentially infinite space of alternatives compatible with the data learners are exposed to.
+At the semantics-pragmatics interface this concerns the selection between functionally (near-)equivalent lexical meanings.
+Put differently, what is systematically conveyed through pragmatics could alternatively be codified lexically.
+However, lexical meanings that allow interlocutors to draw these inferences seem to resist their lexicalization.
+The question is why.
 We assume an integral part of the answer to be that learners are a priori biased towards simpler, more compressed, lexical representations.
 That is, rational learners should prefer simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
+In a nutshell, our hypothesis is that codifying more semantically is a priori dispreferred by learners.
+Crucially, in cases where pragmatics offers conventional means of enrichment, the communicative disadvantage that speakers would otherwise incurr is leveled.
+In this way, semantics and pragmatics play a synergic role in overcoming linguistic pressures for expressivity and learnability which may prevent the lexicalization of pragmatic inferences.
 
-In the following, we model these pressures using the replicator-mutator dynamics by combining functional pressure on successful communication with effects of learning biases on (iterated) Bayesian learning \citep{griffiths+kalish:2007}.
-The semantics-pragmatics distinction and its effect on production and comprehension are made precise through probabilistic models of rational language use in populations and multiple lexica \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
-The remainder of this section introduces the individual components of the model as well as the assumptions underlying them.
+In the following, we model these components using the replicator-mutator dynamics, combining functional pressure on successful communication with effects of learning biases on (iterated) Bayesian learning \citep{griffiths+kalish:2007}.
+The semantics-pragmatics distinction and its effect on production and comprehension are captured by a probabilistic model of rational language use with different degrees of pragmatic sophisitication and languages \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
+The remainder of this section introduces these components individually together with the assumptions underlying them.
 These are: the representation of languages and their use (\S\ref{sec:languages+use}), pressures towards expressivity (\S\ref{sec:expressivity}) and learnability (\S\ref{sec:learnability}), regulated by the replicator and mutator dynamics, respectively, as well as a bias towards simpler semantic representations, codified as the learners' prior.
@@
 Lexica codify the truth-conditions of a language's expressions, i.e., its semantics.
-A convenient way to represent them is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs -- or meanings -- to convey and $M$ a set of messages of the language \citep{franke+jaeger:2014}.
+Following e.g.
+\citet{franke+jaeger:2014}, a convenient way to represent lexica is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs (meanings) to convey and $M$ a set of messages (forms available in the language).
 For example, the following two fragments determine the truth-conditions of two messages, $m_1$ and $m_2$, for two states, $s_1$ and $s_2$.
@@
 
-While these language representations are very simple, they suffice to make precise the distinction between semantics and pragmatics.
-For this purpose we distinguish between two kinds of linguistic behavior.
+While these representations are very coarse and simple, they suffice to make the distinction between semantics and pragmatics precise.
+We distinguish between two kinds of linguistic behavior.
 {\em Literal interlocutors} produce and interpret messages literally.
@@
 In contrast, {\em pragmatic interlocutors} engage in mutual reasoning to inform their choices.
-For instance, a rational speaker of $L_a$ who reasons about her addressee may use $m_1$ to exclusively signal state $s_1$ given that $s_2$ can unambiguously be conveyed through $m_2$.
+For instance, a rational speaker of $L_a$ who reasons about her addressee may use $m_1$ to exclusively signal state $s_1$ since $s_2$ can unambiguously be conveyed through $m_2$.
 Accordingly, rational hearers that expect their interlocutor to reason along these lines will interpret ambiguous $m_1$ as $s_1$.
-An important consequence of this pragmatic process is that it renders $L_a$ indistinguishable from $L_b$ in terms of expressivity.
+An important consequence of this process of mutual reasoning for our purposes is that it can render $L_a$ (nearly) indistinguishable from $L_b$ in terms of language use.
 
-Following models of rational language use such as Rational Speech Act models \citep{frank+goodman:2012} and their game-theoretic counterparts \citep{benz+etal:2005a,franke:2009,franke+jaeger:2014}, this kind of signaling behavior is captured by a resaoning hierarchy.
+Following models of rational language use such as Rational Speech Act models \citep{frank+goodman:2012} and their game-theoretic predecessors, the Iterated Best/Quantal Response models \citep{franke:2009,franke+jaeger:2014}, this kind of signaling behavior is captured by a reasoning hierarchy.
 The hierarchy's bottom, level $0$, corresponds to literal language use.
-Pragmatic language users of level $n + 1$ behave rationally according to (expected) level $n$ behavior of their interlocutors.
-(\ref{h:level0}) and (\ref{h:leveln}) specify the behavior of literal and pragmatic hearers of a language $L$.
-Mutatis mutandis for the literal and pragmatic speakers in (\ref{s:level0}) and (\ref{s:leveln}).
+Pragmatic language users of level $n + 1$ behave rationally according to expected level $n$ behavior of their interlocutors.
+Below, (\ref{h:level0}) and (\ref{h:leveln}) specify the behavior of literal and pragmatic hearers of a language $L$.
+Literal and pragmatic speakers are specified in (\ref{s:level0}) and (\ref{s:leveln}).
 
@@
 
-According to (\ref{h:level0}) a literal hearer's interpretation of a message $m$ as a state $s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$.
-The literal speaker's behavior as given in (\ref{s:level0}) is regulated by a soft-max parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}.
-As $\lambda$ increases, choices made in production are more rational.
-That is, higher values lead to behavior that is increasingly in line with expected utility maximization.
+According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state $s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$.
+In the following this prior is assumed to be uniform across hearers.
+The behavior of literal speakers, given in (\ref{s:level0}), is regulated by a soft-max parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}.
+As $\lambda$ increases, choices made in production are more rational in that higher values lead to behavior that is increasingly in line with expected utility maximization.
 
-Pragmatic behavior is similar to its literal counterpart.
-As in our informal sketch, their difference lies in that level $n+1$ speakers/hearers reason about level $n$ hearer/speaker behavior instead of solely relaying on their lexicon.
-That is, they reason about how a rational level $n$ interlocutor would use or interpret a message, and behave according to these expectations.
+Pragmatic behavior is similar to its literal counterparts.
+As hinted at in the informal sketch above, their difference lies in that level $n+1$ speakers/hearers reason about level $n$ hearer/speaker behavior instead of solely relaying on their lexicon.
+That is, they reason about the way a rational level $n$ interlocutor would use or interpret a message, and behave according to these expectations.
 Pragmatic production is further regulated by a parameter $\alpha$ which controls the tension between semantics and pragmatics, $\alpha \in (0,1]$.
@@
 
-The combination of a lexicon with its use, i.e., a level in the reasoning hierarchy, is called a type. These are the basic units on which our population dynamics operate.
+We call the combination of a lexicon with its use, i.e., a level in the reasoning hierarchy, a type. These are the basic units on which our population dynamics operate.
 
 \subsection{Replication \& expressivity}\label{sec:expressivity}
-Expressivity has received particular attention from investigations using evolutionary game theory \citep{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}.
-Under this view, a type's ability to convey and interpret information successfully confers it a higher fitness, this measure being relative to the success of other types in the population.
-In the simplest models fitness directly translates into the proportion of types present in the subsequent population.
+Expressivity has received particular attention from investigations using evolutionary game theory (e.g.
+\citealt{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}).
+Under this view, a type's ability to convey and interpret information successfully confers it a higher fitness, a measure that is relative to the success of other types in the population.
+In the simplest models fitness directly translates into the proportion of types present in the population after a generational turnover.
 As a consequence, fitter types flourish while unfit ones are driven out.
 This association of communicative success within a population with changes in the proportion of types present in it creates a feedback loop that pressures the population towards greater expressivity.
+
 The replicator equation gives us the means to make these dynamics precise.
-
 The proportion of types in a given population is codified in a vector $x$, where $x_i$ is type $i$'s proportion.
-The fitness of a type $i$, $f_i$, is given by its expected utility in this population, $f_i = \sum_j x_j \text{EU}(t_i,t_j)$.
+As noted above, the fitness of type $i$ is equal to its relative communicative success within this population, $f_i = \sum_j x_j \text{EU}(t_i,t_j)$.
 That is, its fitness is the sum of its expected communicative success in interacting with other types weighted by the latter's population share.
-The expected utility of $i$ and $j$ is obtained by considering the average communcative success of $i$ conveying information to $j$ and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$.
-$U_S(x,y)$ and $U_R(x,y)$ are $\sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$ and $U_S(y,x)$, respectively, for $n$ and $o$ being the reasoning level of $x$ and $y$, and $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.
+Communicating well with types that appear often thusly confers a higher fitness than doing so with underrepresented ones.
+The expected communicative success of $i$ and $j$ is obtained by considering the average success of $i$ conveying information to $j$ and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$.
+$U_S(x,y)$ and $U_R(x,y)$ are, respectively, $\sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$ and $U_S(y,x)$, for $n$ and $o$ being the reasoning level of $x$ and $y$, and $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.
 This quantity is symmetric, reflecting the probability of two types' mutual understanding.
 The average fitness of the population is given by $\Phi$, $\Phi = \sum_i x_i f_i$.
-This term serves as a normalizing constant for the (discrete) replicator equation: $\dot{x}_i = \frac{x_i f_i}{\Phi}$
+This term serves as a normalizing constant for the discrete replicator equation: $\dot{x}_i = \frac{x_i f_i}{\Phi}$
 
-Under its biological interpretation the replicator equation captures the idea of fitness-relative selection whereby fitter types produce more offspring, leading to their propagation in subsequent generations.
-Similarly, many aspects of natural language are subject to processees of transmission and change across varied time-spans.
-Amongst others, the replicator equation can be understood as a model of language acquisition across generations, as e.g.
-in \citealt{nowak+etal:2002}, but also as a process of horizontal adaptation (see \citealt[\S3.3]{benz+etal:2005b} for discussion).
-In the following, we take the latter view in assuming that interlocutors may adapt their types to that which works best within their population.
+Under its biological interpretation, this equation captures the idea of fitness-relative selection whereby fitter types produce more offspring, leading to their propagation in subsequent generations.
+As noted in \S\ref{sec:introduction}, many aspects of language are subject to processees of transmission and change that can be likened to these biological dynamics.
+Amongst others, replication can be construed as modelling language acquisition, as e.g.
+in \citealt{nowak+etal:2002}, but also as a process of horizontal adaptation in a single generation (see \citealt[\S3.3]{benz+etal:2005b} for discussion).
 
-In their series of papers on language evolution, Nowak and colleagues did not only consider expressivity but also recognized the central role that learnability plays in its transmission.
-On the one hand, linguistic production can be prone to errors \citep{nowak+krakauer:1999}.
-On the other, multiple languages may be compatible with the data learners are exposed to \citep{nowak+etal:2002}.
-Both sources introduce variation in language acquistion.
+In their series of papers on language evolution, Nowak and colleagues did not only consider expressivity but also recognized the central role played by the fidelity by which languages are transmitted.
+Linguistic production can be prone to errors and multiple languages may be compatible with the data learners are exposed to.
+These sources of uncertainty introduce variation in their faithful acquistion.
 In keeping the analogy to evolutionary processes this variation can be likened to mutation to the effect that a type's offspring may adopt a different type than that of its parent.
@@
 
+%errors were considered in N+K:1999, mutation in N+etal:2002
+
 \subsection{Mutation \& learning}\label{sec:learnability}
-Iterated learning is a process in which the behavior of one individual serves as learning input for another, who's behavior subsequently serves as input for a new learner, and so on.
-In language this process can be thought of as a progression through chains of parents and children where the parent produces linguistic data from which the child infers a language.
+Iterated learning is a process in which the behavior of one individual serves as learning input for another.
+This learner, upon acquisition of this behavior, then goes on to produce behavior that serves as input for a new learner.
+This process can be thought of as a progression through chains of parents and children; the parent produces linguistic data from which the child infers a language.
 The latter, now a parent, goes on to produce linguistic data for a new generation of na\"ive learners.
-Following \citet{griffiths+kalish:2007} we model iterated learning as a process of Bayesian inference in which learners combine the likelihood of a type producing the learning data with prior inductive biases.
-They then select a type to adopt from the resulting posterior distribution.
+Following \citet{griffiths+kalish:2007} we model iterated learning as a repeated process of Bayesian inference in which learners combine the likelihood of a type producing the received learning input with prior inductive biases.
 
 Due to the pressure towards learnability it exherts, iterated learning generally leads to simpler and more regular languages (see \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016} for recent surveys).
-Importantly, experimental and mathematical results suggest the outcome of this process to reflect the learners' a priori biases.
-In a Bayesian setting these biases are codified in the prior $P \in \Delta(\mathcal{T})$.
+Importantly, experimental and mathematical results suggest that the outcome of this process reflects learners' a priori biases.
+In a Bayesian setting these biases can be codified in a prior $P \in \Delta(T)$.
 A way to think about this prior is as the amount of data a learner would require in order to adopt a language (cf.
@@
 Or, in our case, a combination of a lexicon and a signaling behavior.
-\hl{INTRODUCE SIMPLICITY BIAS HERE}
 
-\hl{Mention similarity to Piantadosi+etal's prior}
+As touched upon in \S\ref{sec:introduction} we assume learners to be biased towards simpler semantics.
+That is, learners have a preference for types that can explain the data by codifying less in their lexical representations.
+More generally, a drive for simplicity has been argued to underpin speaker preferences for brevity and ease of articulation, as well as to pressure languages towards lexical ambiguity and grammatical compression \citep{zipf:1949,grice:1975,piantadosi+etal:2012, kirby+etal:2015}.
+As shown by \citet{chater+vitanyi:2003} simplicity not only provides a solution to problems of induction by allowing for the distinction between equally explanatory hypotheses, but it has also proven its empirical worth as a predictive principle in cognitive modelling.
 
-\hl{ In linguistics, a drive for simplicity has been argued to underpin speaker preferences for brevity and ease of articulation, as well as to pressure languages towards lexical ambiguity and grammatical compression \citep{zipf:1949,grice:1975,piantadosi+etal:2012, kirby+etal:2015}.
-As a broader cognitive principle, the use of simplicity as means to select between hypotheses has a long standing tradition.
-Crucially, \citet{chater+vitanyi:2003} give a number of compelling arguments for simplicity on both mathematical and empirical grounds.}
+The extent of the prior's influence has been shown to heavily depend on the learning strategy assumed to underly the inference process.
+On the one hand, simulation results suggested that weak biases could be magnified by exposing learners to only small data samples \citep{brighton:2002}.
+On the other, \citeposs{griffiths+kalish:2007} mathematical characterization showed that iterated learning converged to the prior.
+That is, the distribution over languages obtained from this process corresponded to the learners' prior distribution and was independent of the amount of input given to them.
+This difference in predictions can be traced back to differences in the selection of hypotheses from the posterior.
+Griffith \& Kalish's convergence to the prior holds for learners that sample from the posterior.
+More deterministic strategies such as the choice for the type with the highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase the influence of both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}.
+In the following, we parametrize the posterior by $l \geq 1$.
+In doing so, we obtain a range of learning strategies that live between posterior sampling and MAP.
+When $l = 1$ learners sample from the posterior.
+As $l$ increases towards infinity their propensity to maximize the posterior grows as well.
 
-The extent of the prior's influence has been shown to strongly depend on the learning strategy assumed to underly the inference process.
-While simulation results suggested that weak biases could be magnified by exposing learners to only small data samples \citep{brighton:2002}, the mathematical characterization provided by \citet{griffiths+kalish:2007} showed that, instead, iterated learning converged to the prior.
-That is, the distribution over languages in a population, or the likelihood of learning a language when taking an individual's perspective, corresponds to the learners' prior distribution irrespective of the amount of input given to learners.
-This divergence in predictions can be traced back to differences in the selection of hypotheses from the posterior.
-On the one extreme, Griffith \& Kalish's prediction of convergence to the prior holds for learners that sample from the posterior.
-On the other, more deterministic strategies such as the selection of the type with the highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase the prior's influence \citep{griffiths+kalish:2007,kirby+etal:2007}.
-In the following, we parametrize the posterior, $P(t_i|d)^l$, to obtain a range of learning strategies that live between posterior sampling and MAP, $l \geq 1$.
-When $l = 1$ learners sample from the posterior.
-As $l$ increases towards infinity, the learners' propensity to maximize the posterior increases.
+The data learners are exposed to is described by a set $D$ containing sequences of state-message pairs, e.g., $\tuple{\tuple{s_i,m_v},...\tuple{s_j,m_w}}$.
+These correspond to sequences of of language use witnessed by the learners.
+In the following, we denote the length of these sequences by $k$.
+Put differently, each datum $d \in D$ contains $k$ members of the set $\{\tuple{s_i,m_v} | s_i \in S, m_v \in M\}$.
 
-We combine the replicator dynamics with iterated learning by codifying the latter as a transition matrix $Q$.
-Just as in standard mutator dynamics, $Q_{ij}$ indicates the probability of the children of a parent of type $i$ adopting type $j$.
-However, to make this process depend on a type's learnbility, this quantity is proportional to the probability of $i$ producing the learning data and that of inferring $j$ given the data.
+To summarize, the parametrized posterior, $P(d|t)^l$, is obtained from combining the prior $P(t)$ and the likelihood $P(d|t)$.
+Since types are combinations of a lexicon and its use, the latter can be computed straightforwardly from the type's production.
+We combine the replicator dynamics with iterated learning by codifying these components in a transition matrix $Q$.
+$Q_{ij}$ indicates the probability that a child of a parent of type $i$ adopts type $j$.
+This quantity is proportional to the probability of $i$ producing the learning data and that of inferring $j$ given the data:
 
-The elements of the set of learning data $D$ are $k$-length sequences of state-message pairings.
-That is, a sequence of observations of language use.
-Put differently, a datum $d \in D$ contains $k$ members of the set $\{\tuple{s_i,m_j} | s_i \in S, m_j \in M\}$.
-$D$ is the set of all such sequences.
-With the at hand $D$, the learning matrix is constructed as follows:
 \[
@@
 \]
-where $F(t_j,d) \propto P(t_j|d)^l$ and $P(t_j|d) \propto P(t_j) P(d|t_j)$. Given a type $i$, $P(d|t_i)$ can be straightforwardly computed based on $t_i$'s production behavior.
+where $F(t_j,d) \propto P(t_j|d)^l$ and $P(t_j|d) \propto P(t_j) P(d|t_j)$.
 
 \subsection{Summary}
-We argued that expressivity, learnability and simplicity are central pressures that apply on the cultural evolution of language.
-They are respectively modelled as communicative efficiency-relative replication, iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings.
-Taken together their dynamics can be described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:
+We argued that expressivity, learnability and simplicity are central to the cultural evolution of language.
+In the present setup they are modelled as communicative efficiency-relative replication, iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings.
+Taken together their dynamics are described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:
 
@@
 
-The units that the dynamics operate on are a combination of a lexicon and linguistic behavior which determines its use, i.e., a type.
+The units that the dynamics operate on are a combination of a lexicon and a degree of pragmatic sophistication which determines its use.
+We call this combination a type.
 A type's expressivity depends on its communicative efficiency within a population while its learnability depends on the fidelity by which it is infered by new generations of na\"ive learners.
-
-%In sum, the innovation of this model lies in its combination of functional pressure on successful communication, effects of learning biases on (iterated) Bayesian language learning \citealt{griffiths+kalish:2007}, and probabilistic models of language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014,bergen+etal:2016}.
-In particular, this synthesis enables for the investigation of the effects of communicative pressures on the semantics-pragmatics interface.
-In doing so, it links the previously disconnected areas of rational probabilistic language use and cultural evoltion.
-
-%\hl{So far, we have focused on highlighting the strands of research and assumptions present in the model.
-Less has been said about its departure from past approaches, beyond the virtues we argued their synthesis gives rise to.
-Given the different foci of these investigations, most comparisons on a general level are misplaced.
-However, \citet{kirby+etal:2015} ...}
-
-%%%Differences to our earlier CogSci proceedings paper:
-%Sequences and atomic observations.
-Before, the set of all observations was $O =$\linebreak  $\{\tuple{\tuple{s_1,m_i},\tuple{s_2,m_j}} | m_i, m_j \in M\}$.
-A member of $O$ encodes that a teacher produced $m_i$ in state $s_1$ and $m_j$ in $s_2$, i.e., it encodes one witnessed message for each state.
-
-%Observations as production.
-Instead of taking the space of all possible sequences of length $k$ into consideration, we take sample from $O$ $k$-times according to the production probabilities of each type; $P(o = \tuple{s,m} | t_i) = P(s) P(m|s,t_i)$.
-$n$ such $k$-length sequences are sampled for each type.
-As a consequence, the data used for computing $Q_i$ is not the same as that used for $j$ $(i \neq j)$.
-
-%No communal learniner; only parental
-%%%
+With this model at hand we turn to the analysis of the lack of lexicalization of productive pragmatic inferences.
 
@@
 %
-A particularly well-studied type of conventional pragmatic enrichments are so-called {\em scalar implicatures}.
-These inferences are licensed for groups of expressions ordered in terms of informativity, here understood as an entailment induced order.
+Scalar implicatures are a particularly well-studied type of conventional pragmatic inference.
+They are licensed for groups of expressions ordered in terms of informativity, here understood as an entailment induced order.
 For instance, {\em some} is entailed by {\em all}.
 If it were true that `All students came to class', it would also be true that `Some students came to class'.
-However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what their use is normally taken to convey.
+However, as already noted in \S\ref{sec:introduction}, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what their use is normally taken to convey.
 Instead, the use of a less informative expression when a more informative one could have been used can license a defeasible inference that stronger alternatives do not hold (cf.
 \citealt{horn:1972,gazdar:1979}).
-That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that, since the speaker did not use a stronger alternative, e.g.
-{\em all}, this alternative must not hold.
-In this way, `Some students came to class' is strengthened to convey `Some but not all students came to class'.
+That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that stronger alternatives do not hold since the speaker did not use them.
+In this way, `Some students came to class' is strengthened to convey that some but not all students came to class.
 Conversely, speakers can rely on their interlocutors to draw this inference.
-In a nutshell, mutual reasoning about rational language use supplies a bound that rules out stronger alternatives pragmatically which is absent from the semantics of weak scalar expressions.
-\hl{This entire par needs to be more reader friendly}
+The bound that rules out stronger alternatives is thusly not codified in the lexical meaning of weak alternatives but instead pragmatically supplied.
 
@@
 More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} and {\em big} were truth-conditionally incompatible with stronger alternatives such as {\em hot}, {\em and}, {\em all} and {\em huge}?
-This question is particularly striking considering the number of expressions that license such inferences across languages and thusly lack a lexicalized upper-bound (\citealt{horn:1972}, \citealt[252-267]{horn:1984}, \citealt{traugott:2004,vdAuwera:2010}).
+This question is particularly striking considering the number of expressions suggested to license such inferences across languages (\citealt{horn:1972}, \citealt[252-267]{horn:1984}, \citealt{traugott:2004,vdAuwera:2010}).
 
 We see two main explanations for the lack of upper-bounds in the lexical meaning of weak scalar expressions.
-The first is that their truth-conditional compatibility with stronger expressions endows them with broader applicability, allowing them to occur in contexts in which their upper-bounded reading is absent.
+The first is that their truth-conditional compatibility with stronger expressions endows them with a broader range of applicability by allowing them to occur in contexts in which their upper-bounded reading is absent.
 This can happen when embedded in downward-entailing contexts, when the speaker is likely uncertain about whether the upper bounded reading is true, or when the distinction between an upper-bounded reading and the simple, only lower-bounded reading, is not relevant.
-For instance, if for all the speaker knows `Some students came' but she doesn't know whether `All came' then the use of unbounded {\em some} succinctly conveys her uncertainty about the latter.
+For instance, if for all the speaker knows `Some students came' but she doesn't know whether all came, then the use of unbounded {\em some} succinctly conveys her uncertainty.
 This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then this is preferred over expressing it overtly through longer expressions.
-For example by saying {\em some but not all} explicitly.
+For example, by saying {\em some but not all} explicitly.
 Importantly, although morphosyntactic disambiguation is dispreferred due to its relative length and complexity, it allows speakers to enforce an upper-bound and override contextual cues that might otherwise mislead the hearer.
-
 In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so.
-Compare this with a hypothetical language that lexicalizes two expressions for each English scalar expression; one with and one lacking an upper-bound.
+Compare this with a hypothetical language that lexicalizes two expressions for each weak scalar expression -- one with and one lacking an upper-bound.
 Along this functionalist explanation, we see four conditions that may pressure languages for English-like semantics over this alternative.
-First, contextual cues are strongly reliable.
+First, contextual cues are very reliable.
 Second, morphosyntactic disambiguation is seldom necessary.
@@
 Fourth, larger lexica are costly.
-Overall, neither condition is convincing in its role as a central explanatory devices for such a wide-spread phenomenon.
-The first two put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of unambiguous communication.
-It is likely that human language users are very good at retrieving cues from contexts, but to stipulate that they are so good as to undercut the benefit of safe communication provided by our hypothetical alternative strikes us as too strong of an assumption.
+Overall, neither condition is convincing as a pivotal explanatory device for such a wide-spread phenomenon.
+The first two conditions put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of unambiguous communication.
+It is likely that human language users are very good at retrieving cues from context, but to stipulate that they are so good as to undercut the benefit of safe communication provided by our hypothetical alternative strikes us as too strong of an assumption.
 As for the third and fourth condition, these seem mostly like technical solutions without a proper empirical basis.
 
-\hl{All the rest of this section needs to be tidied up} Instead, in what follows we investigate the hypothesis that the lack of lexicalization of scalar inferences is driven by the advantage in compression that lexical meanings lacking an upper-bound have over those that explicitly codify it.
-Note however that we do not represent this contrast in compression between lexical meanings explicitly in lexica.
-Instead, the bias towards a lack of upper-bounds in weak scalar alternatives is directly encoded in the learners' prior over types.
+Instead, in what follows we investigate the hypothesis that the lack of lexicalization of scalar inferences can be accounted for through the relative simplicity of lexical meanings lacking an upper-bound over those that explicitly codify it.
+Note however that we do not represent the contrast between lexical representations explicitly.
+Instead, the learning bias towards a lack of upper-bounds in weak scalar alternatives is directly encoded in the learners' prior over types.
 
@@
 \citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012} and references therein).
-We think that our assumption is well-warranted as a working hypothesis and decide against such an enrichment given that the introduction of a larger framework would also require further assumptions and justifications.\footnote{\hl{T: We could possibly show the length difference of the lexical meanings with and without an upper-bound using a LOT grammar in the appendix.
-I don't know if it is worth it though.}}
+We think that our assumption is well-warranted as a working hypothesis and decide against such an enrichment given that the introduction of a larger framework would also require further assumptions and justifications.
 
-In sum, while we do not want to argue that functionalist pressure may not play a role, we do see a clear benefit in exploring whether matters of learnability would not give us additional leverage.
+While we do not want to argue that functionalist pressure may not play a role, we do see a clear benefit in exploring whether matters of learnability would not give us additional leverage.
 
@@
 
+%In sum, the innovation of this model lies in its combination of functional pressure on successful communication, effects of learning biases on (iterated) Bayesian language learning \citealt{griffiths+kalish:2007}, and probabilistic models of language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014,bergen+etal:2016}.
+In particular, this synthesis enables for the investigation of the effects of communicative pressures on the semantics-pragmatics interface.
+In doing so, it links the previously disconnected areas of rational probabilistic language use and cultural evoltion.
+
 \hl{Summarize main predictions: Only expressivity, only learnability, both together; expressivity nudging learning in the right direction and learning leading to a uniform preference for thing in the population}
@@
 
+Following \citet{piantadosi+etal:quantifiers} we assumed that, at some level, lexical meanings are expressed in a representational language and that the learners' task is to infer the lexicon that best explains the linguistic data they witness.
+We go beyond this by inferring type of use
+
 \section{Conclusion}
 Language change is affected by intertwined pressures.
-Drawing from past research, we put forward a general model that integrates central aspects involved in this process by combining cultural evolution dynamics with models of probabilistic language use.
-In the case of the lack of lexicalization of scalar implicatures, the model predicts that, when  pressured for learnability and expressivity, the former force drives for simpler semantic representations inasmuch as pragmatics can compensate for their lack of expressivity in use.
+We investigated the selection of non-upper-bounded lexical meanings in light of pragmatic inference as a case study on role of semantics and pragmatics in overcoming such pressures.
+We argued a central difficulty for their selection to lie in the distinction between functional (near-)equivalents.
+The model we put forward predicts that, when  pressured for learnability and expressivity, the former force drives for simpler semantic representations inasmuch as pragmatics can compensate for their lack of expressivity in use.
 That is, the relative learnability advantage of simpler semantics may offer an answer to why natural languages do not lexicalize certain pragmatic inferences.
-
 More broadly, we argued that semantic patterns can be explained by taking into consideration the way in which they are used in interactions as well as their representational complexity.
-As use more often than not concerns pragmatics, this kind of analysis not only offers an explanatory account for the phenomenon at hand, but sheds light on the semantics-pragmatics distinction by elucidating their respective roles in language change.
+As use more often than not involves pragmatics, this kind of analysis also sheds light on the semantics-pragmatics distinction by elucidating their respective roles in language change.
 

### 2016-10-04 paper/paper.tex
minor changes paper draft

@@
 \subsection{Summary}
-We argued that expressivity, learnability and simplicity are central to the cultural evolution of language.
+Drawing from past research, we argued that expressivity, learnability and simplicity are central to the cultural evolution of language.
 In the present setup they are modelled as communicative efficiency-relative replication, iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings.
@@
 
-\subsection{Setup}
- We consider populations with two signaling behaviors, either literal or pragmatic.
-As mentioned earlier, the former correspond to level $0$ reasoners who only take their lexica into consideration.
-In the following, pragmatic types correspond to level $1$ reasoners since higher level reasoning is not required to derive scalar implicatures from the lexica fragments we consider here.
-The prior over states is assumed to be uniform for all behaviors.
+\subsection{Analysis}
+We analyze the model's predictions in populations of types with two signaling behaviors; literal or pragmatic.
+The former correspond to level $0$ reasoners who only take their lexica into consideration and the latter to level $1$ reasoners.
+Higher level reasoning is not required to derive scalar implicatures from the lexica we consider here nor would it provide substantial pragmatic refinement.
 
-The lexica are listed in Table \ref{tab:lexica}.
-As in our previous examples, they are $(2,2)$-Boolean matrices, i.e., $|M| = |S| = 2$.
-These are the simplest matrices that allow us to make the presence or lack of an upper-bound and the use of scalar implicatures precise.
-Lexica $L_1$ to $L_3$ are not optimal for communication because they assign the same meaning to all their messages.
-They were included to showcase the selection process for a larger hypothesis space.
-$L_4$ and $L_5$ are our target lexica, codifying upper-bounded semantics for message $m_1$ (the first column in a lexicon) and a lack thereof, respectively.
-Lastly, $L_6$ is similar to $L_5$ in that two messages are true of the same state but differs from it in assigning upper-bounded semantics to $m_1$.
-Combining a signaling behavior with one of these $6$ lexica yields a total of $12$ distinct types.\footnote{While there is a total of $16$ possible $(2,2)$-matrices, a number of them are identical both in terms of expressivity and the learning bias against lexical upper-bounds.
-The competition between such types is determined by their  proportions in the initial population but this fact can be obscured when averaging across simulations.
-We therefore focus only on the subset that exhibits the properties we set out to explain.}
+The space of possible lexica is given in Table \ref{tab:lexica}.
+As in the preceding examples, lexica are $(2,2)$-Boolean matrices.
+These are the simplest matrices that allow us to make the contrast between the presence or lack of an upper-bound and the use of scalar implicatures precise.
+As implicit in the discussion so far, one may think of $s_1$ as a ``some but not all''-state and of $s_2$ as an ``all''-state.
+The literal meaning of weak scalar expressions such as English {\em some} then corresponds to a message true of both $s_1$ and $s_2$ in these fragments.
+Following our assumption for a preference for simple lexical representations, the prior biases learners against lexica in which a message holds true only of state $s_1$.
+That is, those messages that lexicalize an upper-bound that rules out the ``all''-state $s_2$.
+All other semantics are assumed to be a priori equally probable.
+Accordingly, this prior is captured by $P(t_i) \propto n - c \cdot r$, where $n$ is the total number of states and $r$ is the number of messages only true of $s_1$ in $t_i$'s lexicon, $c \in [0,1]$.
+In short, increments in parameter $c$ bring about a stronger preference against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.
 
@@
 \end{tabular}
-\caption{Space of possible lexicon fragments considered.}
+\caption{Space of possible lexica.}
 \label{tab:lexica}
@@
 
-We focus our analysis on the contrast between literal and pragmatic types using lexica $L_4$ and $L_5$.
-Note in particular  that a type that has conventionalized upper and lower bounds to realize a (quasi-)partition of the relevant semantic space, such as $L_4$, will produce speaker behavior that is {\em almost} indistinguishable from that of a language with only the respective upper bounds, but with Gricean speakers, such as $L_5$.
+While there is a total of $16$ possible $(2,2)$-lexica, a number of them are identical both in terms of expressivity and the learning bias.
+The competition between such types is determined by the initial configuration of a population, i.e., the proportion with which each type starts out.
+However, this fact can be obscured when averaging across simulations.
+We therefore focus on a smaller representative subset.\footnote{Simulations conducted with the full space of possible lexica confirm that the results reported here do not hinge on their exclusion.} Lexica $L_1$ to $L_3$ are not optimal for communication because they assign the same state to all their messages.
+This failure to associate a single form to a meaning both semantically and pragmatically inevitably leads to their disadvantage in language use.
+$L_4$ and $L_5$ are our target lexica, previously labeled as lexica $L_b$ and $L_a$.
+They codify upper-bounded semantics for message $m_1$ and a lack thereof, respectively.
+Lastly, $L_6$ is similar to $L_5$ in that two messages are true of the same state but differs from it in assigning upper-bounded semantics to $m_1$.
+
+Combining a signaling behavior with each of these $6$ lexica yields a total of $12$ distinct types.
+Note in particular  that a type that has conventionalized upper bounds to realize a (quasi-)partition of the relevant semantic space, such as $L_4$, will produce speaker behavior that is {\em almost} indistinguishable from that of a language that lacks upper bounds, but with pragmatic speakers, such as $L_5$.
 Almost, because there may be slight differences between the probability with which speakers would (erroneously) use a semantically false description and the probability with which speakers would (erroneously) use a pragmatically suboptimal description.
-This further contrasts with literal $L_5$, a type lacks means to convey an upper-bound with $m_1$.
-Due to the possible marginal difference in signaling behavior between pragmatic $L_4$ and $L_5$, the selection of one type over the other is expected to mainly depend on the learning bias.
+Due to this possibly marginal difference between pragmatic $L_4$ and $L_5$, the selection of one type over the other is expected to mainly depend on the learning bias.
 Things are less clear for literal $L_5$ contrasted with literal/pragmatic $L_4$.
-The former has a learning advantage but is expected to fare worse in terms of communicative fitness in virtue of ambiguous $m_1$.
+The former has a learning advantage but is expected to fare worse in communicative  terms in virtue of ambiguous $m_1$.
 
-As implicit in the discussion above, one may think of $s_1$ as a ``some but not all''-state and $s_2$ for an ``all''-state.
-The literal meaning of weak scalar expressions such as {\em some} then corresponds to a message true of both $s_1$ and $s_2$ in these fragments.
-Following our assumption for a preference for simple lexical representations, the prior biases learners against lexica in which a message holds true only of the former and not the latter.
-All other semantics are a priori equally probable.
-This prior is given by $P(t_i) \propto n - c \cdot r$, where $n$ is the total number of states and $r$ is the number of messages only true of $s_1$ in $t_i$'s lexicon, $c \in [0,1]$.
-In sum, an increase in $c$ brings about a stronger learning bias against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.
-
-\subsection{Analysis}
 The dynamics are initialized with an arbitrary distribution over types, constituting the population's first generation.
-If not stated otherwise, the results for a given parameter setting were obtained from $1000$ independent runs.
-Each run consisted of $20$ generations.
+The results for each parameter setting were obtained from $1000$ independent runs, each consisting of $20$ generations.
 This corresponds to a developmental plateau after which no noteworthy change was registered.
-As specified in \ref{sec:mutation}, the learning transition matrix $Q$ can be obtained by considering all possible state-message sequences of length $k$.
-Given that this is intractable for large $k$, matrices with $k > 5$ were approximated by using Monte Carlo with $10$ sequences sampled from each type's production probabilities and a type's children being exposed only to this subset.
-The model's parameters are summarized in Table \ref{tab:summary}.
+As specified in \S\ref{sec:learnability}, the learning matrix $Q$ can be obtained by considering all possible state-message sequences of length $k$.
+Given that this is intractable for large $k$, matrices were approximated by sampling $10$ sequences from each type's production probabilities and a type's children being exposed only to this subset.
+For convenience, the model's parameters are summarized in Table \ref{tab:summary}.
 
@@
 According to our hypothesis, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $L_5$-like semantics.
-However, it is instructive to first inspect the effect of these pressures in isolation.
-For this purpose, we focus on three pragmatic types.\footnote{Pragmatic reasoning allows language users to refine their (possibly erroneous) choices.
-Therefore, it is advantageous even for those types that codify an upper-bound lexically.} Users of $L_3$, representing a type that is lacking in expressivity but is a prior preferred for its lack of upper-bounds.
-Users of $L_4$, a type that is functionally advantageous but biased against.
-And users of $L_5$, combining the virtues of the latter two.
+It is instructive to first inspect the effect of these pressures in isolation.
+For this purpose we focus our attention on three pragmatic types.\footnote{Pragmatic reasoning allows language users to refine their (possibly erroneous) choices.
+Therefore, it is advantageous even for those types that codify an upper-bound lexically.} Pragmatic $L_3$, a type that is lacking in expressivity but is a priori preferred for its lack of upper-bounds.
+Pragmatic $L_4$, a type that is functionally advantageous but biased against.
+And pragmatic $L_5$, combining virtues of the latter two.
 
-\paragraph{Expressivity only.} Recall that the outcome of the replicator dynamics are influenced by $\lambda$ and $\alpha$ as they have a bearing on a type's fitness.
-Low $\alpha$ disadvantages types that rely on pragmatic reasoning for more deterministic signaling behavior to the gain of those that codify this information semantically.
-The rationality parameter $\lambda$ has a similar effect for different reasons.
-Less utility maximizing behavior decreases the association of an ambiguous message with a single state, even when other states are uniquely associated with a different one.
+\paragraph{Expressivity only.} Recall that the replicator dynamics are sensitive to $\lambda$ and $\alpha$ as both have a bearing on a type's fitness.
+In particular, low $\alpha$ disadvantages types that rely on pragmatic reasoning to the gain of those that codify more semantically.
+The rationality parameter $\lambda$ has a similar effect for different reasons: Low rationality leads to a less pronounced preference for the choice(s) expected to succeed best in communication.
 That is, $\lambda$ regulates the strength by which users of $L_5$ associate non-upper-bounded $m_1$ exclusively with the ``some''-state $s_1$ over the ``all''-state $s_2$.
+Speakers of $L_4$ need not rely on $\lambda$ for this, as as the association of $m_1$ with $s_1$ is already part of their language's semantics.
 

### 2016-10-04 paper/paper.tex
changes in draft + tabular presentation of detailed results now shows different values of c

@@
 More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} and {\em big} were truth-conditionally incompatible with stronger alternatives such as {\em hot}, {\em and}, {\em all} and {\em huge}?
-This question is particularly striking considering the number of expressions suggested to license such inferences across languages (\citealt{horn:1972}, \citealt[252-267]{horn:1984}, \citealt{traugott:2004,vdAuwera:2010}).
+This question is particularly striking considering the number of expressions suggested to license such inferences across languages.
 
@@
 
-Instead, in what follows we investigate the hypothesis that the lack of lexicalization of scalar inferences can be accounted for through the relative simplicity of lexical meanings lacking an upper-bound over those that explicitly codify it.
+Taking the regularity and spread of scalar implicatures together with the observation that monomorphemic expressions that lexically rule out stronger alternatives are unattested across languages (\citealt[252-267]{horn:1984},\citealt{horn:1972,traugott:2004,vdAuwera:2010}) suggests that other forces may be play.
+In what follows we investigate the hypothesis that the lack of lexicalization of scalar inferences can be accounted for through the relative simplicity of lexical meanings lacking an upper-bound over those that explicitly codify it.
+While we do not want to argue that functionalist pressure may not play a role, we do see a clear benefit in exploring whether matters of learnability would not give us additional leverage.
+
 Note however that we do not represent the contrast between lexical representations explicitly.
 Instead, the learning bias towards a lack of upper-bounds in weak scalar alternatives is directly encoded in the learners' prior over types.
-
 In principle this difference could be made precise with an adequate representational language, e.g., through measures over representational complexity such as minimal description length.
@@
 We think that our assumption is well-warranted as a working hypothesis and decide against such an enrichment given that the introduction of a larger framework would also require further assumptions and justifications.
-
-While we do not want to argue that functionalist pressure may not play a role, we do see a clear benefit in exploring whether matters of learnability would not give us additional leverage.
 
@@
     $\alpha \in [0,1)$ & semantics-pragmatics tension & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\ 
-    $|D|$ & learning data produced per parent type & $P(d|t_j)P(t_i|d)$\\
+    $|D|$ & number of data produced per parent type & $P(d|t_j)P(t_i|d)$\\
     $k = |d|$ & number of observations per datum& $P(d|t_j)P(t_i|d)$\\
@@
 
-The effect of the rationality parameter using only the replicator dynamics is shown in the left-hand side of Figure \ref{fig:either-R-or-M}.
-As expected, the less expressive type using $L_3$ fares worse and shows little variation across $\lambda$-values.
-Crucially, values of $\lambda \leq 10$ lead to an increase in $L_4$ and a decrease in $L_5$.
-As the rationality parameter increases, the functional difference between $L_4$ and $L_5$ is levelled.
-Overall, the final populations that result only from a pressure towards expressively approximate an even share of pragmatic $L_4, L_5$ and $L_6$ types.
-The latter follows the same trajectory as $L_4$ in Figure \ref{fig:either-R-or-M}.
+The effect of the rationality parameter for $\alpha = 1$ is depicted in Figure \ref{fig:either-R-or-M}.A.
+As expected, the less expressive $L_3$ speakers fare the worse and are influenced the least by variations in $\lambda$.
+In contrast, low values of $\lambda$ result in a higher proportion of $L_4$ speakers relative to $L_5$.
+This is expected given role of rationality in producing more deterministic linguistic behavior in users of $L_5$-like languages.
+As the rationality parameter increases the functional difference between $L_4$ and $L_5$ is levelled.
+Overall, the final populations that result only from a pressure towards expressively come close to an even share of pragmatic $L_4, L_5$ and $L_6$ types (the latter follows the same trajectory as $L_5$ in Figure \ref{fig:either-R-or-M}).
+That is, expressivity alone can not differentiate between these lexica in populations of rational pragmatic language users.
 
@@
 
-\paragraph{Learnability only.} To get an impression of the effect of iterated learning without a pressure towards expressivity, we first consider (relatively) deterministic parental data production ($\lambda = 20, \alpha = 1$).
-In this way small data sequences suffice for learners to differentiate types that produce strongly diverging expressions.
-In line with \citeposs{griffiths+kalish:2007} analysis, under these conditions the resulting populations converge to the learner's prior distribution when sampling from the posterior.
-This is shown in the right-hand side of Figure \ref{fig:either-R-or-M}, which directly reflects the prior distribution for each value of $c$.
+\paragraph{Learnability only.} The effect of iterated learning without a pressure for expressivity is shown in Figure \ref{fig:either-R-or-M}.B with posterior sampling.
+In line with our expectations, the share of $L_4$ speakers decreases as the bias against this lexicon increases.
+In turn, this benefits $L_3$ and, in particular, $L_4$.
+However, note that even a strong bias against lexical upper-bounds leads only to a moderate advanage for $L_5$ over $L_4$.
+Furthermore, a pressure only towards learnability can promote functionally defective languages $L_3$.
 
-Inspecting the effects of these dynamics separately not only gives some intuitions about the parameters' influence, but also highlights some of their broader implications.
+Inspecting these pressures separately not only gives some intuitions about the parameters' influence, but also highlights some of their broader implications.
 First and foremost, neither dynamic comes close to converging to a monomorphic population under most parameter configurations.
-For instance, types using $L_4$ can come to occupy a large proportion of the final population.
-However, this holds only for a restricted range of low degrees of rationality.
-Apart from polymorphy, both dynamics make some undesirable predictions.
+For instance, while $L_4$ speakers can come to take over a substantial proportion of the population, this only happens in a restricted range of low degrees of rationality.
+Apart from polymorphy, both pressures make undesirable predictions.
 A pressure only towards expressivity leads to the selection of types using $L_4$ to $L_6$ and to the ejection of $L_1$ to $L_3$.
@@
 In contrast, a pressure only towards learnability has a modest but clear effect in differentiating $L_5$ from these alternatives but fails to rule out functionally suboptimal types such as tautological $L_3$.
-This showcases that, while the bias plays a major role for the contrast between $L_4$ and $L_5$, on its own it does not enable types that fail to convey an upper-bound to establish themselves in the population.
 In sum, neither dynamic on its own is a suitable candidate to provide a justification for the predicted prevalence of $L_5$-like semantics.
-
-\paragraph{Expressivity and learnability.} Figure \ref{fig:cost} illustrates the effect of the learning bias after $20$ generations across values of $c$ with $l = 1$ (A) and $l =3$ (B).
-More detailed results for all types across a sample of $c$-values are presented in Table \ref{tab:numeric-results}.
-Overall, these results suggest that in the present setup a weak bias is sufficient to lead to a selection of $L_5$ over $L_4$.
-As in the simulations that only considered learnability, this effect increases with the bias' strength provided $L_5$ users are pragmatic.
-Importantly, the addition of a pressure towards expressivity magnifies this effect by dampening the proliferation of functionally suboptimal types advantaged by the learning bias.
-As stressed above, this suggests that neither the learning bias nor functional pressure alone but their combination lead to the systematic lack of semantic upper-bounds in scalar expressions.
 
@@
 \includegraphics[scale=.5]{./fig2-rmd}
+\caption{Mean proportions of target types after $20$ generations in $1000$ populations across bias values $c \in [0,1]$ with $l =1$ in A and $l = 3$ in B ($\alpha =1, \lambda = 20, k = 5$).}
 \label{fig:cost}
-\caption{Mean proportions of target types after $20$ generations in $1000$ populations across bias values $c \in [0,1]$ with $l =1$ in A and $l = 3$ in B ($\alpha =1, \lambda = 20, k = 5$).}
 \end{figure}
+
+\paragraph{Expressivity and learnability.} Figure \ref{fig:cost} illustrates the effect of the learning bias for posterior sampling (\ref{fig:cost}.A) and slightly more MAP-like learning (\ref{fig:cost}.B).
+More detailed results for all types across a sample of $c$-values are presented in Table \ref{tab:numeric-results}.
+Overall, these results suggest that in the present setup a weak bias is sufficient to lead to a selection of $L_5$ over $L_4$.
+As in the simulations that only considered learnability, this effect increases with the bias' strength provided $L_5$ users are pragmatic.
+Importantly, the addition of a pressure towards expressivity magnifies this effect and dampens the proliferation of functionally suboptimal types advantaged by the learning bias.
+As stressed above, this suggests that neither the learning bias nor functional pressure alone but their combination may lead to the lack of upper-bounds in the lexical meaning of scalar expressions.
 
@@
 \begin{tabular}{l | c c c c c| c c c c c c c c}
-\multicolumn{1}{c}{~} & \multicolumn{5}{c}{$l = 1$} & ~ & \multicolumn{5}{c}{$l = 10$}\\ \hline \hline
-c  & 0 & .01 & .05 & .1 & .8 & ~ & 0 & .01 & .05 & .1 & .8\\ \hline \hline
-lit. $L_1$& .03 & .03 & .03 & .03  & .04 & ~ & $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\ 
-lit. $L_2$ & .03 & .03 & .03 & .03 & .01 & ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
-lit. $L_3$ & .03 & .03 & .03 & .03 & .04 & ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
-lit. $L_4$ & .07 & .07 & .07 & .07 & .06 & ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
-lit. $L_5$ & .04 & .04 & .05 & .05 & .06& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
-lit. $L_6$ & .04 & .04 & .04 & .04 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\ \hline
-prg. $L_1$ & .03 & .03 & .03 & .03 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ \\
-prg. $L_2$ & .03 & .03 & .03 & .03 & .01& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ \\
-prg. $L_3$ & .03 & .03 & .03 & .03 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ \\ 
-prg. $L_4$ & .22 & .22 & .22 & .22 & .18& ~ & .33& .33 & .32 & .31 & .15 \\
-prg. $L_5$ & .22 & .22 & .22 & .23 & .3& ~ & .33& .33 & .35 & .37 & .7 \\
-prg. $L_6$ & .22 & .22 & .22 & .22 & .18& ~ & .33& .33 & .32 & .31 & .15   
+\multicolumn{1}{c}{~} & \multicolumn{5}{c}{$l = 1$} & ~ & \multicolumn{5}{c}{$l = 3$}\\ \hline \hline
+  c        &  0  & .1  & .5  & .8  & .9 & ~ & 0 & .1 & .5 & .8 & .9\\ \hline \hline
+lit. $L_1$ & .03 & .03 & .04 & .04 & .04& ~ & $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\ 
+lit. $L_2$ & .03 & .03 & .02 & .01 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
+lit. $L_3$ & .03 & .03 & .04 & .04 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
+lit. $L_4$ & .07 & .07 & .06 & .06 & .05& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
+lit. $L_5$ & .04 & .05 & .05 & .06 & .06& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
+lit. $L_6$ & .04 & .04 & .04 & .04 & .03& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\ \hline
+prg. $L_1$ & .03 & .03 & .04 & .04 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ \\
+prg. $L_2$ & .03 & .03 & .02 & .01 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ \\
+prg. $L_3$ & .03 & .03 & .04 & .04 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ \\ 
+prg. $L_4$ & .22 & .22 & .2 & .18 & .17& ~ & .33& .31 & .23 & .15 & .12 \\
+prg. $L_5$ & .22 & .23 & .27 & .3  & .32& ~ & .33& .37 & .54 & .7 & .75 \\
+prg. $L_6$ & .22 & .22 & .2 & .18 & .17& ~ & .33& .31 & .23 & .15 & .12
 \end{tabular}
-\caption{Mean proportions of types in $1000$ populations after $20$ generations across bias values $c \in [0,1]$ with $l =1$ and $l = 3$ ($\alpha = 1, \lambda = 30, k = 5$), $\epsilon < 0.005$}
+\caption{Mean proportions of types in $1000$ populations after $20$ generations across bias values $c \in [0,1]$ with $l =1$ and $l = 3$ ($\alpha = 1, \lambda = 30, k = 5$), $\epsilon < 0.005$.}
 \label{tab:numeric-results}
@@
 
-This data suggests that the proportion of scalar implicature users that is predicted primarily hinges on three aspects of the model.
-First, the degree to which linguistic behavior is deterministic, controlled by $\lambda$ and $\alpha$, plays a role both for expressivity as well as for producing data that allows learners to discriminate this type from others.
-Second, the learning bias $c$, leads learners to discriminate and prefer $L_5$ over $L_4$ and $L_5$.
-Lastly, the posterior parameter $l$ magnifies the effects of the learning bias in tandem with replication.
-This interaction is shown in \ref{fig:prior-posterior}.
-In the present setup posterior sampling can lead to the incumbency of pragmatic $L_5$, but not even a strong favorable learning bias manages to completely drive out competing types (cf.
-\ref{fig:cost}.A).
-However, as posterior maximization increases, the range of bias values within which this type takes over the population increases drastically.
+The resulting proportion of pragmatic $L_5$ speakers primarily hinges on three parameters.
+First, the degree to which linguistic behavior is deterministic, controlled by $\lambda$ and $\alpha$.
+As stressed before, this plays a role both for expressivity as well as for producing data that allows learners to discriminate this type from others.
+Second,  the learning bias $c$ which controlls the learners preference for simpler lexical representations -- leading to the selection of $L_5$ over $L_4$.
+Lastly, the posterior parameter $l$ which magnifies the effects of the learning bias in tandem with replication.
 
-\hl{More  discussion about parameter interaction: changes in sequence length influence the population in a predictable way: smaller values lead to more heterogeneous populations whereas larger ones lead to more pronounced differences.
-This is expected insofar as the likelihood that a sequence of length $1$ was produced by any type is relatively uniform (modulo prior) whereas the likelihood of types with lexica $L_1$ - $L_3$ to produce, for instance, a sequence of $10$ observations consistently with the same state-message combination is less likely than for pragmatic players using $L_4$ - $L_6$ or literal $L_4$.
-Thus, while noteworthy, sequence length has no direct bearing on the main contrast of interest.
-Similar considerations hold for $\alpha$ and $\lambda$ -- set to $1$ and $50$ in the following.
-Overall, lower rationality in $\lambda$ or more pragmatic violations in $\alpha$ lead to a higher selection of lexica with semantic upper-bounds.
-The fitness of pragmatic behavior increases with higher $\lambda$-/$\alpha$-values.
-In other words, these parameters level the functional contrast between $L_4$ and $L_5$.
-}
+As discussed in relation to Figure \ref{fig:cost}.A, posterior sampling can lead to the incumbency of pragmatic $L_5$.
+However, not even a strong favorable learning bias combined with a pressure for expressivity completely drives out competing types.
+This is not so for more posterior maximizing behavior.
+Crucially, as shown in Figure \ref{fig:prior-posterior} the range of bias values within which $L_5$ takes over the population increases with MAP-like learning.
+In other words, the strength of the learning bias required for a given final proportion of $L_5$ speakers strongly depends on the learning strategy adopted evinced by it.
+As for the effect of the other parameters not mentioned so far, changes in sequence length influence the population in a predictable way: smaller values lead to more heterogeneous populations that reflect the learner's prior more faithfully whereas larger ones lead to more pronounced differences amongst equally preferred types.
+This is expected insofar as the likelihood that a sequence of length $1$ was produced by any type is relatively uniform (modulo prior) whereas the likelihood of types with lexica $L_1$ - $L_3$ to produce, for instance, a sequence of $10$ observations consistently with the same state-message combination is less likely than for pragmatic types using $L_4$ - $L_6$, or literal $L_4$.
+Thus, while noteworthy, sequence length has no direct bearing on the  contrast of interest.
 
@@
 Broadly speaking these results suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome selective pressures and stabilize in a population provided there is a bias for simpler representations.
-This outcome is particularly encouraging in light of the other advantages a lack of semantic upper-bounds may confer that were discussed above.
+This outcome is particularly encouraging in light of other advantages a lack of semantic upper-bounds may confer.
 
-The model predicts this result to hinge on four conditions.
-First, types need to be pressured toward both expressivity and learnability.
-Second, language use can not be too unpredictable; low $\alpha$ or $\lambda$ values render non-upper-bounded lexical meanings more prone to communicative failure and more difficult to learn.
-Third, learners should prefer simpler over more complex lexical representations.
-Fourth, the strength by which they need to be preferred depends on the inferential mechanism of learners.
-That is, to which degree learners maximize the posterior.
-Put differently, high rationality in learning and choice requires a weaker bias towards simpler representations.
-The selection of lexical meanings lacking upper-bounds is robust against parameter perturbations under these conditions.
+This predictions hinges on three assumptions.
+First, that language is pressured toward both expressivity and learnability.
+Second, that language use is not too unpredictable: low $\alpha$ or $\lambda$ values renders languages that rely on pragmatics too prone to miscommunication and more difficult to learn.
+Third, that learners prefer simpler over more complex lexical representations.
+Importantly, a combination of rationality in learning and choice requires a weaker bias towards simpler representations.
+Under these conditions the selection of lexical meanings lacking upper-bounds in populations of pragmatic speakers is robust against parameter perturbations.
 
-A large range of parameter values lead to the stable incumbency of scalar implicatures.
-However, while this is predicted by the literature, it is less clear to what proportion other types should be represented.
-On the one hand, there are  reasons to expect functionally suboptimal types $L_1$-$L_3$ to be largely ruled out.
-They fail to enable their users to communicate without substantial error both amongst themselves and with other types.
-In short, these lexica stand at odds with past theoretical and empirical observations.
-On the other, this is not so for $L_4$.\footnote{$L_6$ presents a special case.
-In our current setup, it mirrors $L_5$ in allowing for pragmatic strenghtening of message $m_2$ rather than $m_1$.
-However, its association of $s_1$ with $m_1$ and $s_2$ with $m_2$ under favourable parameter conditions is achieved by ruling out the ``some but not all''-state $s_1$ and not, as with scalar implicatures, the ``all''-state $s_2$.
-$L_6$ speakers thusly have a ``some''-message strengthened to convey ``all but not some but not all''.
-The observation that monomorphemic expression that lexically rule out stronger alternatives are unattested across languages has received substantial argumentative support (most prominently in \citealt[252-267]{horn:1984} but also e.g.
-in \citealt{horn:1972,traugott:2004,vdAuwera:2010}).
-However, our present setup is blind to these differences.} Notwithstanding, it is possible that natural language communities are composed of such mixed populations or that a single speaker may entertain $L_4$-like semantics for one scalar expression and $L_5$-like semantics for another \hl{CITATIONS FOR THIS PURPOSE}.
+While a great representation of non upper-bounded lexical meanings is predicted by the literature, it is less clear to what extent other types should be represented -- if at all.
+There are reasons to expect functionally suboptimal types $L_1$-$L_3$ to be ruled out since they fail to enable their users to communicate effectively.
+However, this is not true of $L_4$.\footnote{$L_6$ presents a special case.
+In our current setup, it mirrors $L_5$ in enabling for a pragmatic strenghtening of message $m_2$ rather than $m_1$.
+However, this association of $s_2$ with $m_2$ under favourable parameter conditions is achieved by ruling out the ``some but not all''-state $s_1$ and not, as with scalar implicatures, the ``all''-state $s_2$.
+$L_6$ speakers therefore strenghten a ``some''-message to convey something paraphrasable as `some but not [some but not all]'.
+However, our present setup is blind to this redundancy.} Notwithstanding, it is possible that natural language communities are composed of homogeneous populations or that a single speaker may entertain $L_4$-like semantics for one scalar expression and $L_5$-like semantics for another.
 
-Lastly, there are a number of extensions one may want to consider.
-For instance, the consideration of larger lexica that consider epistemic uncertainty.
-Another interesting additon relates to possible disadvantages of pragmatic reasoning and the effect of state frequencies on fossilization.
+This empirical issue relates to two aspects left undiscussed in our present analysis: disadvantages of pragmatic reasoning and the effect of state frequencies on the fossilization of pragmatic inferences.
 We tacitly assumed pragmatic reasoning to come at no cost.
@@
 This raises the question at which point such usage-based cost undercuts the learnability advantage of simpler semantic representations.
-As noted above, this might also depend on a given scalar expression and its frequency.
-Frequently drawn scalar implicatures might fossilize to avoid cost, while infrequent ones could still be computed on-line.
-
-A virtue of this model is that it allows for analysis specific modifications and extensions.
-Straightforward extensions include larger hypothesis spaces, as well as larger or different lexicon fragments.
-Another worthwhile modification relates to possible disadvantages of pragmatic reasoning.
-We tacitly assumed pragmatic reasoning to come at no cost.
-However, there is experimental evidence for the assumption that the pragmatic derivation of upper-bounds costs effort and takes additional processing time (cf.
-\citealt{deNeys+schaeken:2007, huang+snedeker:2009}).
-This raises the question at which point such usage-based cost undercuts the learnability advantage of simpler semantic representations.\footnote{In the present setup this modification has a straightforward effect: A penalty for pragmatic signaling lowers the fitness of pragmatic types, to the advantage of literal types.
-However, the penalty needs to be substantial to counteract the functional advantage pragmatic $L_5$ has over all but $L_4$ together with its learning advantage.}
+Should cost play a role, it's effect is bound to depend on the frequency with which a given scalar expression is used.
+We speculate that frequently drawn scalar implicatures might fossilize to avoid cost, while infrequent ones could still be computed on-line.
 
 \section{General discussion}
-We laid out a model that combines game-theoretical models of functional pressure towards efficient communication \citep{nowak+krakauer:1999}, effects of learning biases on (iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012, franke+jaeger:2014} as well as different lexica \citep{bergen+etal:2012,bergen+etal:2016}.This model generates predictions about  lexicalization patterns found in natural language and the failure to lexicalize of certain pragmatic inferences.
-While the puzzle raised by semantics is hard to explain by purely functional means, it suggests part of the answer to lie in learnability.
-Simpler semantic representations are more likely to be learned, and pragmatic reasoning can counteract functional disadvantages otherwise incurred.
+We laid out a model that combines game-theoretical models of functional pressure towards efficient communication \citep{nowak+krakauer:1999}, effects of learning biases on (iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012, franke+jaeger:2014} as well as different lexica \citep{bergen+etal:2012,bergen+etal:2016}.
+This model generates predictions about lexicalization patterns found in natural language.
+We argued that, while the puzzle raised by semantics in light of pragmatics is hard to explain on purely functional grounds, part of the answer lies in learnability: Simpler semantic representations are more likely to be learned while pragmatic reasoning can counteract functional disadvantages otherwise incurred.
 This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics by offering an account of why (certain) pragmatic inferences are not part of the literal meaning of expressions.
-It furthermore leaves the possibility of such inferences to fossilize open when they do not compete against a lexical simplicity bias.
 
@@
 Language change is affected by intertwined pressures.
-We investigated the selection of non-upper-bounded lexical meanings in light of pragmatic inference as a case study on role of semantics and pragmatics in overcoming such pressures.
-We argued a central difficulty for their selection to lie in the distinction between functional (near-)equivalents.
-The model we put forward predicts that, when  pressured for learnability and expressivity, the former force drives for simpler semantic representations inasmuch as pragmatics can compensate for their lack of expressivity in use.
+We investigated the selection of non-upper-bounded lexical meanings as a case study on role of semantics and pragmatics in satisfying such pressures.
+We showed that, when pressured for learnability and expressivity, the former drives for simpler semantic representations inasmuch as pragmatics can compensate for their lack of expressivity in use.
 That is, the relative learnability advantage of simpler semantics may offer an answer to why natural languages do not lexicalize certain pragmatic inferences.
 More broadly, we argued that semantic patterns can be explained by taking into consideration the way in which they are used in interactions as well as their representational complexity.
-As use more often than not involves pragmatics, this kind of analysis also sheds light on the semantics-pragmatics distinction by elucidating their respective roles in language change.
 

### 2016-10-05 paper/paper.tex
draft modifications, additions to general discussion and general clean-up

@@
 To this end, we lay out a general model that combines functional pressure on language use and a pressure for learnability in acquisition in populations of probabilistic language users with varied degrees of pragmatic sophistication.
-We then use this model in a case study on the (lack of) lexicalization of scalar implicatures.
+We then apply this model in a case study on the (lack of) lexicalization of scalar implicatures.
 The results suggest that such inferences fail to lexicalize when languages are pressured towards simpler semantic representations in learning, provided that pragmatic reasoning can compensate for the disadvantage in expressivity that users of such languages otherwise incur.
-We argue this result to shed light on the lack of lexicalization of scalar inferences, as well as the semantics-pragmatics distinction more generally.
+This result sheds light on the lack of lexicalization of scalar inferences, as well as the semantics-pragmatics distinction more generally.
 \end{abstract}
@@
 What is conveyed usually goes beyond what is said.
-A request for a blanket can be politely veiled by uttering ``I'm cold'', a temporal succession of events conveyed by the order in which conjunts appear, as in``I traveled to Paris and got married'', or an invitation be declined by saying ``I have to work''.
-An influential explanation of the relation between the literal meaning (semantics) of such expressions and what they are intended and interpreted to convey (pragmatics) sees the latter as a product of a process of mutual reasoning \citep{grice:1975}.
-For instance, under the assumption that the speaker is cooperative and strives to be relevant, ``I have to work'' may be interpreted as providing a reason why the speaker will not be able to accept the invitation -- why would she have said so otherwise?
+A request for a blanket can be politely veiled by uttering ``I'm cold'', a temporal succession of events conveyed by the order in which conjuncts appear as in``I traveled to Paris and got married'', or an invitation be declined by saying ``I have to work''.
+An influential explanation of the relation between the literal meaning of such expressions (semantics) and what they are intended and interpreted to convey (pragmatics) due to \citet{grice:1975} sees the latter as a product of a process of mutual reasoning.
+For instance, under the assumption that the speaker is cooperative and relevant, ``I have to work'' may be interpreted as providing a reason why the speaker will not be able to accept the invitation -- why would she have said so otherwise?
 
@@
 Importantly, many classes of lexical meanings allow for pragmatic enrichments in a notably productive fashion.
-For instance, while John can eat all the cookies and still truthfully claim ``I ate some cookies'', the use of {\em some} over {\em all} is usually taken to convey that the latter does not hold.
-That is, ``I ate some cookies'' is pragmatically strengthened to convey that {\em some but not all} were eaten (\hl{citations for experimental evidence}).
-This inference is not restricted to quantifiers such as {\em some} and {\em all} but can be licensed for a large class of natural language expressions ranging from numerals such as {\em four} and {\em five} to adjectives such as {\em cold} and {\em big}.
+For instance, while John can eat all the cookies and still truthfully say ``I ate some cookies'', the use of {\em some} over {\em all} is usually taken to convey that the latter does not hold.
+That is, ``I ate some cookies'' is pragmatically strengthened to convey that {\em some but not all} were eaten (\hl{references for experimental evidence that people in fact draw these inferences}).
+This inference is not restricted to quantifiers such as {\em some} and {\em all} but can be licensed for a large class of natural language expressions,  ranging from numerals such as {\em four} and {\em five} to adjectives such as {\em cold} and {\em big}.
 
-These pragmatic inferences, so-called {\em scalar implicatures}, have been the focus of many theoretical and experimental studies at the semantics-pragmatics interface (\hl{citations}).
+These pragmatic inferences, so-called {\em scalar implicatures}, have been the focus of many studies at the semantics-pragmatics interface.
 However, an issue that has received little attention is the justification of semantic structure in light of the systematic enrichment offered by pragmatics.
@@
 The emergence and change of linguistic structure is influenced by many intertwined factors.
-These range from biological and socio-ecological to cultural \citep{steels:2011,tamariz+kirby:2016}.
-Social and ecological pressures determine communicative needs, while biology determines the architecture that enables and constrains the means by which they can be fulfilled.
+These range from biological and socio-ecological to cultural \citep{steels:2011}.
+Social and ecological pressures determine communicative needs while biology determines the architecture that enables and constrains the means by which they can be fulfilled.
 In the following, our focus lies on the remaining cultural aspect, wherein processes of linguistic change are viewed as shaped by language use and its transmission.
@@
 The idea that language is influenced by communicative pressures has played a pivotal role in synchronic and diachronic analyses at latest since \citeposs{zipf:1949} rationalization of the approximation of word frequency rankings by a power law distribution as competing hearer and speaker preferences (e.g.
-in \citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007, piantadosi:2014,kirby+etal:2015}).
-In recent years this line of research has led to a surge of approaches that seek to analyze the effects of such pressures on language from multiple angles ranging from simulations to experiments with human and robotic agents (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
+in \citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007, piantadosi:2014,kirby+etal:2015} among many).
+In recent years this line of research has led to a surge of approaches that seek to analyze the effects of such pressures from multiple angles ranging from simulations to experiments with human and robotic agents (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
 Our starting point is given by the overarching argument that has crystalized from the accumulated mathematical, experimental and cross-linguistic evidence in this literature: Natural languages need to be well-adapted to communicative needs within a linguistic community, but also need to be learnable to survive their faithful transmission across generations.
@@
 \citealt{kemp+regier:2012,kirby+etal:2015}).
-A language with a single form is easy to learn but lacking in expressivity.
+A language with a single form-meaning association is easy to learn but lacking in expressivity.
 Conversely, a language that associates a distinct form with all possible meanings its users may want to convey is maximally expressive but challenging to acquire.
@@
 A more central issue to our explanandum is that of selecting particular hypotheses out of a potentially infinite space of alternatives compatible with the data learners are exposed to.
-At the semantics-pragmatics interface this concerns the selection between functionally (near-)equivalent lexical meanings.
-Put differently, what is systematically conveyed through pragmatics could alternatively be codified lexically.
-However, lexical meanings that allow interlocutors to draw these inferences seem to resist their lexicalization.
+At the semantics-pragmatics interface this concerns the selection between functionally (near-)equivalent lexical meanings, noting in particular that what is systematically conveyed through pragmatics could alternatively be codified lexically.
+Notwithstanding, often lexical meanings that allow interlocutors to draw these inferences seem to resist their lexicalization.
 The question is why.
+
 We assume an integral part of the answer to be that learners are a priori biased towards simpler, more compressed, lexical representations.
 That is, rational learners should prefer simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
-In a nutshell, our hypothesis is that codifying more semantically is a priori dispreferred by learners.
-Crucially, in cases where pragmatics offers conventional means of enrichment, the communicative disadvantage that speakers would otherwise incurr is leveled.
+In a nutshell, our hypothesis is that codifying more semantically is a priori dispreferred by learners and in cases where pragmatics offers conventional means of enrichment, the communicative disadvantage that speakers would otherwise incurr is leveled.
 In this way, semantics and pragmatics play a synergic role in overcoming linguistic pressures for expressivity and learnability which may prevent the lexicalization of pragmatic inferences.
@@
 In the following, we model these components using the replicator-mutator dynamics, combining functional pressure on successful communication with effects of learning biases on (iterated) Bayesian learning \citep{griffiths+kalish:2007}.
-The semantics-pragmatics distinction and its effect on production and comprehension are captured by a probabilistic model of rational language use with different degrees of pragmatic sophisitication and languages \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
+The semantics-pragmatics distinction and its bearing on production and comprehension are captured by a probabilistic model of rational language use with different degrees of pragmatic sophisitication and languages \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
 The remainder of this section introduces these components individually together with the assumptions underlying them.
-These are: the representation of languages and their use (\S\ref{sec:languages+use}), pressures towards expressivity (\S\ref{sec:expressivity}) and learnability (\S\ref{sec:learnability}), regulated by the replicator and mutator dynamics, respectively, as well as a bias towards simpler semantic representations, codified as the learners' prior.
+These are: the representation of languages and their use (\S\ref{sec:languages+use}), pressures towards expressivity (\S\ref{sec:expressivity}) and learnability (\S\ref{sec:learnability}) regulated by the replicator and mutator dynamics, respectively, as well as a bias towards simpler semantic representations, codified in the learners' prior.
 After laying out the model, we analyze its predictions in a case-study on the lack of lexicalization of scalar implicatures in \S\ref{sec:si-case-study}.
@@
 Lexica codify the truth-conditions of a language's expressions, i.e., its semantics.
-Following e.g.
-\citet{franke+jaeger:2014}, a convenient way to represent lexica is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs (meanings) to convey and $M$ a set of messages (forms available in the language).
+Following \citet{franke+jaeger:2014}, a convenient way to represent lexica is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs (meanings) to convey and $M$ a set of messages (forms available in the language).
 For example, the following two fragments determine the truth-conditions of two messages, $m_1$ and $m_2$, for two states, $s_1$ and $s_2$.
@@
 
-According to lexicon $L_a$ the message $m_1$ is true of state $s_1$ as well as of $s_2$.
+According to lexicon $L_a$ message $m_1$ is true of state $s_1$ as well as of $s_2$.
 In contrast, $m_1$ is only true of $s_1$ in $L_b$.
 Otherwise, the two languages are truth-conditionally equivalent.
+While these representations are very simple, they suffice to make the distinction between semantics and pragmatics precise.
 
-While these representations are very coarse and simple, they suffice to make the distinction between semantics and pragmatics precise.
 We distinguish between two kinds of linguistic behavior.
@@
 Accordingly, rational hearers that expect their interlocutor to reason along these lines will interpret ambiguous $m_1$ as $s_1$.
-An important consequence of this process of mutual reasoning for our purposes is that it can render $L_a$ (nearly) indistinguishable from $L_b$ in terms of language use.
+For our purposes, an important consequence of this process of mutual reasoning is that it can render $L_a$ nearly indistinguishable from $L_b$ in terms of language use.
 
@@
 That is, its fitness is the sum of its expected communicative success in interacting with other types weighted by the latter's population share.
-Communicating well with types that appear often thusly confers a higher fitness than doing so with underrepresented ones.
+Communicating well with types that are well-represented in the population therefore confers a higher fitness than doing so with underrepresented ones.
 The expected communicative success of $i$ and $j$ is obtained by considering the average success of $i$ conveying information to $j$ and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$.
@@
 Under its biological interpretation, this equation captures the idea of fitness-relative selection whereby fitter types produce more offspring, leading to their propagation in subsequent generations.
-As noted in \S\ref{sec:introduction}, many aspects of language are subject to processees of transmission and change that can be likened to these biological dynamics.
+As noted above, many aspects of language are subject to processees of transmission and change that can be likened to these biological dynamics.
 Amongst others, replication can be construed as modelling language acquisition, as e.g.
@@
 
-In their series of papers on language evolution, Nowak and colleagues did not only consider expressivity but also recognized the central role played by the fidelity by which languages are transmitted.
-Linguistic production can be prone to errors and multiple languages may be compatible with the data learners are exposed to.
-These sources of uncertainty introduce variation in their faithful acquistion.
+In their series of papers on language evolution, Nowak and colleagues did not only consider expressivity but also recognized the central role of the fidelity by which languages are transmitted.
+Linguistic production can be prone to errors, and multiple languages may be compatible with the data learners are exposed to.
+These sources of uncertainty introduce variation in their transmission from one generation to the next.
 In keeping the analogy to evolutionary processes this variation can be likened to mutation to the effect that a type's offspring may adopt a different type than that of its parent.
@@
 This learner, upon acquisition of this behavior, then goes on to produce behavior that serves as input for a new learner.
-This process can be thought of as a progression through chains of parents and children; the parent produces linguistic data from which the child infers a language.
+This process can be thought of as a progression through chains of parents and children; the parent produces linguistic data from which the child infers a language (\hl{reference for main representative work of IL}).
 The latter, now a parent, goes on to produce linguistic data for a new generation of na\"ive learners.
@@
 Importantly, experimental and mathematical results suggest that the outcome of this process reflects learners' a priori biases.
-In a Bayesian setting these biases can be codified in a prior $P \in \Delta(T)$.
-A way to think about this prior is as the amount of data a learner would require in order to adopt a language (cf.
+In a Bayesian setting these biases can be codified in a prior $P \in \Delta(T)$ and a way to think about it is as the amount of data a learner would require in order to adopt a language (cf.
 \citealt[450]{griffiths+kalish:2007}).
@@
 That is, learners have a preference for types that can explain the data by codifying less in their lexical representations.
-More generally, a drive for simplicity has been argued to underpin speaker preferences for brevity and ease of articulation, as well as to pressure languages towards lexical ambiguity and grammatical compression \citep{zipf:1949,grice:1975,piantadosi+etal:2012, kirby+etal:2015}.
-As shown by \citet{chater+vitanyi:2003} simplicity not only provides a solution to problems of induction by allowing for the distinction between equally explanatory hypotheses, but it has also proven its empirical worth as a predictive principle in cognitive modelling.
+More generally, a drive for simplicity has been argued to underpin speaker preferences for brevity and ease of articulation, as well as to pressure languages towards lexical ambiguity and grammatical compression \citep{zipf:1949,grice:1975,piantadosi+etal:2012b, kirby+etal:2015}.
+As shown by \citet{chater+vitanyi:2003} simplicity not only provides a solution to induction problems by providing a distinction between equally explanatory hypotheses, but has also proven its empirical worth as a longstanding predictive principle in cognitive modelling.
 
-The extent of the prior's influence has been shown to heavily depend on the learning strategy assumed to underly the inference process.
-On the one hand, simulation results suggested that weak biases could be magnified by exposing learners to only small data samples \citep{brighton:2002}.
-On the other, \citeposs{griffiths+kalish:2007} mathematical characterization showed that iterated learning converged to the prior.
-That is, the distribution over languages obtained from this process corresponded to the learners' prior distribution and was independent of the amount of input given to them.
+In iterated learning, the extent of the prior's influence has been shown to heavily depend on the learning strategy assumed to underly the inference process.
+On the one hand, simulation results suggested that weak biases could be magnified by exposing learners to only small data samples (e.g.
+in \citealt{brighton:2002}).
+On the other, \citeposs{griffiths+kalish:2007} mathematical characterization showed that iterated learning converged to the prior, i.e., the resulting distribution over languages corresponds to the learners' prior distribution and is not influenced by the amount of input given to them.
 This difference in predictions can be traced back to differences in the selection of hypotheses from the posterior.
@@
 More deterministic strategies such as the choice for the type with the highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase the influence of both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}.
-In the following, we parametrize the posterior by $l \geq 1$.
+In the following, we parametrize the posterior with a parameter $l$, $l \geq 1$.
 In doing so, we obtain a range of learning strategies that live between posterior sampling and MAP.
@@
 The data learners are exposed to is described by a set $D$ containing sequences of state-message pairs, e.g., $\tuple{\tuple{s_i,m_v},...\tuple{s_j,m_w}}$.
-These correspond to sequences of of language use witnessed by the learners.
-In the following, we denote the length of these sequences by $k$.
+These are sequences of language use witnessed by learners, the length of which we denote by $k$.
 Put differently, each datum $d \in D$ contains $k$ members of the set $\{\tuple{s_i,m_v} | s_i \in S, m_v \in M\}$.
@@
 To summarize, the parametrized posterior, $P(d|t)^l$, is obtained from combining the prior $P(t)$ and the likelihood $P(d|t)$.
-Since types are combinations of a lexicon and its use, the latter can be computed straightforwardly from the type's production.
-We combine the replicator dynamics with iterated learning by codifying these components in a transition matrix $Q$.
-$Q_{ij}$ indicates the probability that a child of a parent of type $i$ adopts type $j$.
+Since types are combinations of a lexicon and its use, the latter can be computed straightforwardly from a type's production behavior.
+We combine the replicator dynamics with iterated learning by codifying the latter in a transition matrix $Q$, where $Q_{ij}$ indicates the probability that a child of a parent of type $i$ adopts type $j$.
 This quantity is proportional to the probability of $i$ producing the learning data and that of inferring $j$ given the data:
@@
 Drawing from past research, we argued that expressivity, learnability and simplicity are central to the cultural evolution of language.
-In the present setup they are modelled as communicative efficiency-relative replication, iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings.
-Taken together their dynamics are described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:
+We propose these components to be modelled respectively as communicative efficiency-relative replication, iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings.
+Taken together their interaction is described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:
 
@@
 
-The units that the dynamics operate on are a combination of a lexicon and a degree of pragmatic sophistication which determines its use.
+The units that the dynamics operate on are a combination of a lexicon and a degree of pragmatic sophistication determining its use.
 We call this combination a type.
-A type's expressivity depends on its communicative efficiency within a population while its learnability depends on the fidelity by which it is infered by new generations of na\"ive learners.
-With this model at hand we turn to the analysis of the lack of lexicalization of productive pragmatic inferences.
+A type's expressivity depends on its communicative efficiency within a population while its learnability depends on the fidelity by which it is inferred by new generations of na\"ive learners.
+The learners' task is consequently to perform a joint inference over types of linguistic behavior and lexical meaning.
+With this model at hand we turn to the analysis of the lack of lexicalization of productive pragmatic inferences in a case study on scalar implicatures.
 
@@
 \citealt{horn:1972,gazdar:1979}).
-That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that stronger alternatives do not hold since the speaker did not use them.
+That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that stronger alternatives do not hold since the speaker did not choose them.
 In this way, `Some students came to class' is strengthened to convey that some but not all students came to class.
@@
 A pragmatic hearer who reasons about a speaker's use of message $m_1$ will associate it more strongly with $s_1$ than with $s_2$ since the latter is already unambiguously associated with $s_2$.
-Conversely, a pragmatic speaker will reason about her interlocutor's interpretation and use the messages accordingly.
+Conversely, a pragmatic speaker will reason about her interlocutor's expected interpretation and use the messages at her disposition accordingly.
 
@@
 
-Our initial question can now be narrowed in terms of scalar implicatures by asking for a justification for the lack of lexical upper-bounds in weak scalar alternatives.
+Our initial question can now be narrowed to the case of scalar implicatures by asking for a justification for the lack of lexical upper-bounds in weak scalar alternatives.
 That is, why semantics such as those of message $m_1$ in $L_a$ are regularly selected for over the alternative of codifying the bound semantically as in $L_b$.
 More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} and {\em big} were truth-conditionally incompatible with stronger alternatives such as {\em hot}, {\em and}, {\em all} and {\em huge}?
-This question is particularly striking considering the number of expressions suggested to license such inferences across languages.
+This question is particularly striking considering the number of expressions that license such inferences across languages.
 
@@
 This can happen when embedded in downward-entailing contexts, when the speaker is likely uncertain about whether the upper bounded reading is true, or when the distinction between an upper-bounded reading and the simple, only lower-bounded reading, is not relevant.
-For instance, if for all the speaker knows `Some students came' but she doesn't know whether all came, then the use of unbounded {\em some} succinctly conveys her uncertainty.
+For instance, if for all the speaker knows `Some students came' but she doesn't know whether all came, then the use of not upper-bounded {\em some} succinctly conveys her uncertainty.
 This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then this is preferred over expressing it overtly through longer expressions.
 For example, by saying {\em some but not all} explicitly.
-Importantly, although morphosyntactic disambiguation is dispreferred due to its relative length and complexity, it allows speakers to enforce an upper-bound and override contextual cues that might otherwise mislead the hearer.
+Importantly, although morphosyntactic disambiguation is dispreferred due to its relative length and complexity \citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and override contextual cues that might otherwise mislead the hearer.
 In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so.
 Compare this with a hypothetical language that lexicalizes two expressions for each weak scalar expression -- one with and one lacking an upper-bound.
-Along this functionalist explanation, we see four conditions that may pressure languages for English-like semantics over this alternative.
+We see four conditions along this functionalist explanation that may pressure languages for English-like semantics over this alternative.
 First, contextual cues are very reliable.
@@
 The first two conditions put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of unambiguous communication.
-It is likely that human language users are very good at retrieving cues from context, but to stipulate that they are so good as to undercut the benefit of safe communication provided by our hypothetical alternative strikes us as too strong of an assumption.
+It is likely that human language users are very good at retrieving cues from context, but to stipulate that they are so good as to undercut the benefit of safe communication provided by this hypothetical alternative strikes us as too strong of an assumption.
 As for the third and fourth condition, these seem mostly like technical solutions without a proper empirical basis.
 
-Taking the regularity and spread of scalar implicatures together with the observation that monomorphemic expressions that lexically rule out stronger alternatives are unattested across languages (\citealt[252-267]{horn:1984},\citealt{horn:1972,traugott:2004,vdAuwera:2010}) suggests that other forces may be play.
-In what follows we investigate the hypothesis that the lack of lexicalization of scalar inferences can be accounted for through the relative simplicity of lexical meanings lacking an upper-bound over those that explicitly codify it.
-While we do not want to argue that functionalist pressure may not play a role, we do see a clear benefit in exploring whether matters of learnability would not give us additional leverage.
+Instead, considering the regularity and spread of scalar implicatures together with the observation that monomorphemic expressions that lexically rule out stronger alternatives are unattested across languages (\citealt[252-267]{horn:1984},\citealt{horn:1972,traugott:2004,vdAuwera:2010}) suggests that other forces may be at play.
+In what follows we investigate the hypothesis that the lack of lexicalization of scalar inferences can be accounted for by the relative representational simplicity of lexical meanings lacking an upper-bound over those that explicitly codify it.
+While we do not want to argue that functional aspects as the ones discussed above do not play a role, we do see a clear benefit in exploring whether matters of learnability would not give us additional explanatory leverage.
 
@@
 There is a growing effort to develop such empirically  testable  representational  languages.
-For  instance, the so-called language of thought has been put to test in various rational probabilistic models that show encouraging results (see e.g.
-\citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012} and references therein).
-We think that our assumption is well-warranted as a working hypothesis and decide against such an enrichment given that the introduction of a larger framework would also require further assumptions and justifications.
+For  instance, the so-called {\em language of thought} has been put to test in various rational probabilistic models that show encouraging results (see e.g.
+\citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012a} and references therein).
+We think that our assumption is well-warranted as a working hypothesis and decide against such an enrichment at present in order to focus on the effects of linguistic pressures predicted by our model instead.
 
@@
 The former correspond to level $0$ reasoners who only take their lexica into consideration and the latter to level $1$ reasoners.
-Higher level reasoning is not required to derive scalar implicatures from the lexica we consider here nor would it provide substantial pragmatic refinement.
+Higher level reasoning is not required to derive scalar implicatures from the lexica we consider here, nor do they leave room for substantial pragmatic refinement.
 
-The space of possible lexica is given in Table \ref{tab:lexica}.
+The space of lexica we consider is given in Table \ref{tab:lexica}.
 As in the preceding examples, lexica are $(2,2)$-Boolean matrices.
-These are the simplest matrices that allow us to make the contrast between the presence or lack of an upper-bound and the use of scalar implicatures precise.
+These are the simplest matrices that allow us to make the contrast between the presence or abscence of an upper-bound and the use of scalar implicatures precise.
 As implicit in the discussion so far, one may think of $s_1$ as a ``some but not all''-state and of $s_2$ as an ``all''-state.
@@
 Accordingly, this prior is captured by $P(t_i) \propto n - c \cdot r$, where $n$ is the total number of states and $r$ is the number of messages only true of $s_1$ in $t_i$'s lexicon, $c \in [0,1]$.
-In short, increments in parameter $c$ bring about a stronger preference against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.
+Increments in the prior parameter $c$ therefore bring about a stronger preference against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.
 
@@
 However, this fact can be obscured when averaging across simulations.
-We therefore focus on a smaller representative subset.\footnote{Simulations conducted with the full space of possible lexica confirm that the results reported here do not hinge on their exclusion.} Lexica $L_1$ to $L_3$ are not optimal for communication because they assign the same state to all their messages.
+We therefore focus on this smaller representative subset.\footnote{Simulations conducted with the full space of possible lexica confirm that the results reported here do not hinge on their exclusion.} Lexica $L_1$ to $L_3$ are not optimal for communication because they assign the same state to all their messages.
 This failure to associate a single form to a meaning both semantically and pragmatically inevitably leads to their disadvantage in language use.
-$L_4$ and $L_5$ are our target lexica, previously labeled as lexica $L_b$ and $L_a$.
-They codify upper-bounded semantics for message $m_1$ and a lack thereof, respectively.
+$L_4$ and $L_5$ are our target lexica, previously respectively labeled as lexica $L_b$ and $L_a$.
+They codify upper-bounded semantics for message $m_1$ and a lack thereof.
 Lastly, $L_6$ is similar to $L_5$ in that two messages are true of the same state but differs from it in assigning upper-bounded semantics to $m_1$.
@@
 Given that this is intractable for large $k$, matrices were approximated by sampling $10$ sequences from each type's production probabilities and a type's children being exposed only to this subset.
-For convenience, the model's parameters are summarized in Table \ref{tab:summary}.
 
-\begin{table}
-\centering
-\begin{tabular}{l|l|l}
-    \multicolumn{1}{c}{parameter} & \multicolumn{1}{c}{explanation} & \multicolumn{1}{c}{locus}\\ \hline
-    $\lambda \geq 1$ & rationality parameter & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\
-    $\alpha \in [0,1)$ & semantics-pragmatics tension & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\ 
-    $|D|$ & number of data produced per parent type & $P(d|t_j)P(t_i|d)$\\
-    $k = |d|$ & number of observations per datum& $P(d|t_j)P(t_i|d)$\\
-    $l \geq 1$ & posterior parameter from sampling to MAP & $P(t_i|d) \propto [P(t_i)P(d|t_i)]^l$\\
-    $c \in [0,1]$ & learning bias for lack of upper-bounds &  $P(t_i)$
-\end{tabular}
-\caption{Summary of model parameters.} 
-\label{tab:summary}
-\end{table}
+%For convenience, the model's parameters are summarized in Table \ref{tab:summary}.
+%
+%\begin{table}
+%\centering
+%\begin{tabular}{l|l|l}
+%    \multicolumn{1}{c}{parameter} & \multicolumn{1}{c}{explanation} & \multicolumn{1}{c}{locus}\\ \hline
+%    $\lambda \geq 1$ & rationality parameter & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\
+%    $\alpha \in [0,1)$ & semantics-pragmatics tension & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\ 
+%    $|D|$ & number of data produced per parent type & $P(d|t_j)P(t_i|d)$\\
+%    $k = |d|$ & number of observations per datum& $P(d|t_j)P(t_i|d)$\\
+%    $l \geq 1$ & posterior parameter from sampling to MAP & $P(t_i|d) \propto [P(t_i)P(d|t_i)]^l$\\
+%    $c \in [0,1]$ & learning bias for lack of upper-bounds &  $P(t_i)$
+%\end{tabular}
+%\caption{Summary of model parameters.} 
+%\label{tab:summary}
+%\end{table}
 
@@
 That is, $\lambda$ regulates the strength by which users of $L_5$ associate non-upper-bounded $m_1$ exclusively with the ``some''-state $s_1$ over the ``all''-state $s_2$.
-Speakers of $L_4$ need not rely on $\lambda$ for this, as as the association of $m_1$ with $s_1$ is already part of their language's semantics.
+Speakers of $L_4$ need not rely on $\lambda$ for this as the association of $m_1$ with $s_1$ is already part of their language's semantics.
 
-The effect of the rationality parameter for $\alpha = 1$ is depicted in Figure \ref{fig:either-R-or-M}.A.
+The influence of the rationality parameter for $\alpha = 1$ is depicted in Figure \ref{fig:either-R-or-M}.A.
 As expected, the less expressive $L_3$ speakers fare the worse and are influenced the least by variations in $\lambda$.
 In contrast, low values of $\lambda$ result in a higher proportion of $L_4$ speakers relative to $L_5$.
-This is expected given role of rationality in producing more deterministic linguistic behavior in users of $L_5$-like languages.
-As the rationality parameter increases the functional difference between $L_4$ and $L_5$ is levelled.
-Overall, the final populations that result only from a pressure towards expressively come close to an even share of pragmatic $L_4, L_5$ and $L_6$ types (the latter follows the same trajectory as $L_5$ in Figure \ref{fig:either-R-or-M}).
+This is expected given role of rationality in producing more deterministic behavior in users of $L_5$-like languages.
+As the rationality parameter increases, the functional difference between $L_4$ and $L_5$ is leveled.
+Overall, the final populations that result only from a pressure towards expressively come close to an even share of pragmatic $L_4, L_5$ and $L_6$ types (the latter follows the same trajectory as $L_5$ in Figure \ref{fig:either-R-or-M}.A).
 That is, expressivity alone can not differentiate between these lexica in populations of rational pragmatic language users.
@@
 \includegraphics[scale=.5]{./only-R-or-M}
-\caption{Mean proportions of target types after $20$ generations in $1000$ populations with only replication (A; $\alpha = 1$) and only mutation (B; $\alpha =1, \lambda = 30, k = 5$).}
+\caption{Mean proportions of target types after $20$ generations in $1000$ populations with only a pressure for expressivity in A ($\alpha = 1$) and only for learnability in B ($\alpha =1, \lambda = 30, k = 5$, $l=1$).}
 \label{fig:either-R-or-M}
@@
 
-\paragraph{Learnability only.} The effect of iterated learning without a pressure for expressivity is shown in Figure \ref{fig:either-R-or-M}.B with posterior sampling.
-In line with our expectations, the share of $L_4$ speakers decreases as the bias against this lexicon increases.
+\paragraph{Learnability only.} The effect of iterated learning without a pressure for expressivity is shown in Figure \ref{fig:either-R-or-M}.B for posterior sampling.
+In line with our expectations, the share of $L_4$ speakers decreases as the bias against upper-bounds increases.
 In turn, this benefits $L_3$ and, in particular, $L_4$.
-However, note that even a strong bias against lexical upper-bounds leads only to a moderate advanage for $L_5$ over $L_4$.
-Furthermore, a pressure only towards learnability can promote functionally defective languages $L_3$.
+However, note that even a strong bias against lexical upper-bounds leads only to a moderate advantage for $L_5$ over $L_4$.
+Furthermore, a pressure only towards learnability promotes functionally defective languages $L_3$.
 
@@
 Apart from polymorphy, both pressures make undesirable predictions.
-A pressure only towards expressivity leads to the selection of types using $L_4$ to $L_6$ and to the ejection of $L_1$ to $L_3$.
-However, it can not explain the regular selection of $L_5$-like semantics over either of these functionally similar alternatives.
+A pressure only towards expressivity leads to the to the ejection of $L_1$, $L_2$ and $L_3$ from the population.
+However, it can not explain the regular selection of $L_5$-like semantics over either of its functionally similar alternatives $L_4$ and $L_6$.
 In contrast, a pressure only towards learnability has a modest but clear effect in differentiating $L_5$ from these alternatives but fails to rule out functionally suboptimal types such as tautological $L_3$.
-In sum, neither dynamic on its own is a suitable candidate to provide a justification for the predicted prevalence of $L_5$-like semantics.
 
@@
 \paragraph{Expressivity and learnability.} Figure \ref{fig:cost} illustrates the effect of the learning bias for posterior sampling (\ref{fig:cost}.A) and slightly more MAP-like learning (\ref{fig:cost}.B).
-More detailed results for all types across a sample of $c$-values are presented in Table \ref{tab:numeric-results}.
+More detailed results for all types across a sample of $c$-values for $l = 1$ and $l = 3$ are presented in Table \ref{tab:numeric-results}.
 Overall, these results suggest that in the present setup a weak bias is sufficient to lead to a selection of $L_5$ over $L_4$.
@@
 
-The resulting proportion of pragmatic $L_5$ speakers primarily hinges on three parameters.
-First, the degree to which linguistic behavior is deterministic, controlled by $\lambda$ and $\alpha$.
-As stressed before, this plays a role both for expressivity as well as for producing data that allows learners to discriminate this type from others.
-Second,  the learning bias $c$ which controlls the learners preference for simpler lexical representations -- leading to the selection of $L_5$ over $L_4$.
-Lastly, the posterior parameter $l$ which magnifies the effects of the learning bias in tandem with replication.
+The resulting proportion of pragmatic $L_5$ speakers primarily hinges on four parameters.
+First, the degree to which linguistic behavior is deterministic, controlled by $\lambda$ and $\alpha$, as it plays a role both for expressivity as well as in producing data that allows learners to discriminate this type from others.
+Second,  the learning bias $c$ which controls the learners preference for simpler lexical representations -- leading to the selection of $L_5$ over $L_4$.
+Lastly, the posterior parameter $l$, which magnifies the effects of the learning bias in tandem with replication.
 
@@
 Crucially, as shown in Figure \ref{fig:prior-posterior} the range of bias values within which $L_5$ takes over the population increases with MAP-like learning.
-In other words, the strength of the learning bias required for a given final proportion of $L_5$ speakers strongly depends on the learning strategy adopted evinced by it.
+In other words, the strength of the learning bias required for a given final proportion of $L_5$ speakers strongly depends on learners' inference mechanism.
 As for the effect of the other parameters not mentioned so far, changes in sequence length influence the population in a predictable way: smaller values lead to more heterogeneous populations that reflect the learner's prior more faithfully whereas larger ones lead to more pronounced differences amongst equally preferred types.
@@
 Broadly speaking these results suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome selective pressures and stabilize in a population provided there is a bias for simpler representations.
+This prediction hinges on three assumptions.
+First, that language is pressured toward both expressivity and learnability.
+Second, that language use is not too unpredictable: low $\alpha$ or $\lambda$ render languages that rely on pragmatics too prone to communicative failure and more difficult to learn.
+Third, that learners prefer simpler over more complex lexical representations.
+An important remark on this third condition is that a combination of rationality in choice and maximization in learning requires a weaker bias towards simpler representations.
+Under these conditions the selection of lexical meanings lacking upper-bounds in populations of pragmatic speakers is robust against parameter perturbations.
 This outcome is particularly encouraging in light of other advantages a lack of semantic upper-bounds may confer.
 
-This predictions hinges on three assumptions.
-First, that language is pressured toward both expressivity and learnability.
-Second, that language use is not too unpredictable: low $\alpha$ or $\lambda$ values renders languages that rely on pragmatics too prone to miscommunication and more difficult to learn.
-Third, that learners prefer simpler over more complex lexical representations.
-Importantly, a combination of rationality in learning and choice requires a weaker bias towards simpler representations.
-Under these conditions the selection of lexical meanings lacking upper-bounds in populations of pragmatic speakers is robust against parameter perturbations.
-
-While a great representation of non upper-bounded lexical meanings is predicted by the literature, it is less clear to what extent other types should be represented -- if at all.
-There are reasons to expect functionally suboptimal types $L_1$-$L_3$ to be ruled out since they fail to enable their users to communicate effectively.
+While a major share of non upper-bounded lexical meanings is predicted by the literature, it is less clear to what extent other types should be present in the final population -- if at all.
+There are strong reasons to expect functionally suboptimal types $L_1$-$L_3$ to be ruled out since they fail to enable their users to communicate effectively.
 However, this is not true of $L_4$.\footnote{$L_6$ presents a special case.
-In our current setup, it mirrors $L_5$ in enabling for a pragmatic strenghtening of message $m_2$ rather than $m_1$.
-However, this association of $s_2$ with $m_2$ under favourable parameter conditions is achieved by ruling out the ``some but not all''-state $s_1$ and not, as with scalar implicatures, the ``all''-state $s_2$.
-$L_6$ speakers therefore strenghten a ``some''-message to convey something paraphrasable as `some but not [some but not all]'.
-However, our present setup is blind to this redundancy.} Notwithstanding, it is possible that natural language communities are composed of homogeneous populations or that a single speaker may entertain $L_4$-like semantics for one scalar expression and $L_5$-like semantics for another.
-
-This empirical issue relates to two aspects left undiscussed in our present analysis: disadvantages of pragmatic reasoning and the effect of state frequencies on the fossilization of pragmatic inferences.
+In our current setup, it mirrors $L_5$ in enabling for a pragmatic strengthening of message $m_2$ rather than $m_1$.
+However, this association of $s_2$ with $m_2$ under favorable parameter conditions is achieved by ruling out the ``some but not all''-state $s_1$ and not, as with scalar implicatures, the ``all''-state $s_2$.
+$L_6$ speakers therefore strengthen a ``some''-message to convey something paraphrasable as `some but not [some but not all]'.
+However, our representation of lexica as Boolean matrices is blind to this.} Notwithstanding, it is not unthinkable that natural language communities are homogeneous or that a single speaker may entertain $L_4$-like semantics for one scalar expression and $L_5$-like semantics for another (\hl{references on individual level differences in the derivation of scalar implicatures}).
+Alternatively, a stronger tendency for posterior maximization has to be assumed (cf.
+Figure \ref{fig:prior-posterior}).
+This empirical issue relates to other two aspects left undiscussed in our present analysis: disadvantages of pragmatic reasoning and the effect of state frequencies on the fossilization of pragmatic inferences.
 We tacitly assumed pragmatic reasoning to come at no cost.
@@
 This raises the question at which point such usage-based cost undercuts the learnability advantage of simpler semantic representations.
-Should cost play a role, it's effect is bound to depend on the frequency with which a given scalar expression is used.
-We speculate that frequently drawn scalar implicatures might fossilize to avoid cost, while infrequent ones could still be computed on-line.
+Should cost play a role then its effect is bound to depend on the frequency with which a given scalar expression is used.
+We conject that frequently drawn scalar implicatures might fossilize to avoid cost, while infrequent ones could still be computed on-line.
+In turn, this opens a possible venue to address the previous question about expected use of $L_4$-like semantics.
 
@@
 We laid out a model that combines game-theoretical models of functional pressure towards efficient communication \citep{nowak+krakauer:1999}, effects of learning biases on (iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012, franke+jaeger:2014} as well as different lexica \citep{bergen+etal:2012,bergen+etal:2016}.
-This model generates predictions about lexicalization patterns found in natural language.
-We argued that, while the puzzle raised by semantics in light of pragmatics is hard to explain on purely functional grounds, part of the answer lies in learnability: Simpler semantic representations are more likely to be learned while pragmatic reasoning can counteract functional disadvantages otherwise incurred.
-This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics by offering an account of why (certain) pragmatic inferences are not part of the literal meaning of expressions.
+This model generates predictions about lexicalization patterns and, more generally, effects of communicative pressures on language.
+We argued that the puzzle raised by semantics in light of pragmatics is hard to explain on purely functional grounds and that part of the answer instead may lie in learnability: Simpler semantic representations are more likely to be learned while pragmatic reasoning can counteract functional disadvantages otherwise incurred.
+This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics by offering an account of why (certain) pragmatic inferences fail to lexicalize.
 
-%In sum, the innovation of this model lies in its combination of functional pressure on successful communication, effects of learning biases on (iterated) Bayesian language learning \citealt{griffiths+kalish:2007}, and probabilistic models of language use in populations with distinct lexica \citep{frank+goodman:2012,franke+jaeger:2014,bergen+etal:2016}.
-In particular, this synthesis enables for the investigation of the effects of communicative pressures on the semantics-pragmatics interface.
-In doing so, it links the previously disconnected areas of rational probabilistic language use and cultural evoltion.
+The main innovations of the model are its modular separation of expressivity and learnability, which allows for their isolated and combined analysis, as well as the learning process involving a joint inference over types of pragmatic behavior and lexical meaning.
+The goal to decouple but model both expressivity and learnability has also recently been addressed by \citet{kirby+etal:2015}.
+In contrast to the present proposal, however, expressivity is modelled as exerting its force in the production of learning data only.
+This model's expressivity parameter thereby fulfills a similar role  to high values of $\lambda$ and $\alpha$ in making speaker behavior more deterministic.
+Consequently, it ``favors'' unambiguous languages.
+Crucially, the degree of mutual understanding of interlocutors central to replication and to our notion of expressivity is not taken into consideration.
+Our reasoning behind its inclusion lies in the empirical and theoretical observation that learnability alone can have the undesirable outcome of favoring functionally defective languages.
+This outcome was showcased for a tautological language in our analysis but similar effects have been reported in a number laboratory experiments with human subjects where the participants' task was to learn and subsequently reproduce the language produced by a previous participant, leading to a proliferation of languages were a large number of meanings are associated with a single form (\hl{references}).
+In contrast, adding an interactive component to the experimental setup has been found to foster languages that enable interlocutors to distinguish meanings accurately  (\hl{references}; for a review of these studies and further references see \citealt{kirby+etal:2015, tamariz+kirby:2016}).
+More generally, there is an important difference between a language generating learnable linguistic data and its actual performance as a means of information transfer.
+The former solely depends on the mechanism by which speakers associate form and meaning.
+The latter additionally hinges on the hearers linguistic experience and her capability to interpret linguistic input based on this experience.
+We contend that successful information transfer in a linguistic community is central the adoption of a communication system and that this measure is not faithfully reflected by taking only production into consideration.
 
-\hl{Summarize main predictions: Only expressivity, only learnability, both together; expressivity nudging learning in the right direction and learning leading to a uniform preference for thing in the population}
-
-Discussion about expressivity as external to learning (cf. Stadler, replicator-papers by Kenny Smith, Kirby et al 2015). Possibly add appendix with direct comparison between IL and RMD.
-
-\hl{\citet{kirby+etal:2015} propose dynamics that involve expressivity and learnability.
-In contrast to our proposal, the former pressure is not fully decoupled from the latter in that is restricted to the production of learning data as a form of ambiguity avoidance (cf.
-brighton+etal:2005b cited on p.89l, who make render such ambiguity avoidance as a ``expressivity bias'' in learning).
-That is, a measure of the degree of understanding achieved by interlocutors is not a part of the theoretical model.
-}\footnote{\hl{We could say more about this}}
-
-In contrast to past research using the replicator-mutator dynamics, the Bayesian setting afforded by iterated learning allows for a straightforward integration of a learning prior.
-Furthermore, less idiosyncratic assumptions about the variation introduced by learning without having to resort to similarity matrices between vocabularies \citep{nowak+krakauer:1999} nor languages \citep{nowak+etal:2002}.
-
-Following \citet{piantadosi+etal:quantifiers} we assumed that, at some level, lexical meanings are expressed in a representational language and that the learners' task is to infer the lexicon that best explains the linguistic data they witness.
-We go beyond this by inferring type of use
+A central assumption to the model's predictions is the learning bias for simpler lexical representations.
+Albeit implicit in our formalization, the conceptual foundation for a compression preference at the lexical level is grounded in work in the language of thought tradition put forward by \citet{piantadosi+etal:2012a,piantadosi+etal:underreview}.
+Following Piantadosi and colleagues we assumed that lexical meanings are expressed in a representational language and that the learners' task is to infer lexica that best explain the linguistic witnessed.
+Furthermore, we assumed learners to not only infer lexical meaning but types of linguistic use.
+Whether language use is culturally transmitted is an empirical matter that can not be settled here.
+However, this idealized model suggests that semantic structure evinced cross-linguistically that lends itself for pragmatic exploitation reflects lexical information that is both easy to learn and, more interestingly, also easy to learn how to exploit via mutual reasoning.
 
@@
 We showed that, when pressured for learnability and expressivity, the former drives for simpler semantic representations inasmuch as pragmatics can compensate for their lack of expressivity in use.
-That is, the relative learnability advantage of simpler semantics may offer an answer to why natural languages do not lexicalize certain pragmatic inferences.
-More broadly, we argued that semantic patterns can be explained by taking into consideration the way in which they are used in interactions as well as their representational complexity.
+That is, the relative learnability advantage of simpler semantics may offer an answer to why natural languages fail to lexicalize certain pragmatic inferences.
+More broadly, we argued that semantic patterns can be explained by taking into consideration the way in which they are used in interaction as well as their representational complexity.
 
-%%% Old snippets %%%
-%\section{Conveying upper-bounds}
-%Scalar inferences refer to the pragmatic derivation of an upper-bound for weak scalar expressions to the effect that stronger alternatives are inferred to not hold, e.g.
-{\em some students came} may be taken to convey that {\em not all students came}.
-The order of an expression such as {\em some} with respect to an alternative, e.g., {\em all}, is induced by entailment.
-For instance, {\em all students came} entails {\em most students came}, which in turn entails {\em some students came}.
-In this sense, {\em some} is weaker than {\em all}.
-A considerable class of natural language expressions do not lexicalize an upper-bound and can be ordered in this fashion, allowing for their pragmatic strengthening.
-As alluded at above, examples in English include numerals, scalar adjectives, quantifiers, modals, and connectives.
-\hl{possibly add some typological data on universality, frequency, monomorphemic status}
-%
-%The pragmatic enrichment of the semantic content of such expressions is enabled by mutual reasoning \citep{grice:1975}.
-More specifically, it is driven by interlocutors' mutual expectations of rational language use.
-The hearer reasons about the speaker's choice of a weak alternative over a stronger one.
-Had the speaker known that a stronger alternatives holds, she would have said so as this would have been more informative.
-Since she did not, the hearer can infer that the stronger alternative does not hold.
-Analogously, a speaker who reasons about her addressee may rely on her to derive this inference.
-In this way, a strengthened, upper-bounded, state of affairs can be conveyed without codifying the bound explicitly in the semantics.
-%
-%However, while pragmatics offers means to convey upper-bounds, the question why they are not part of the lexical meaning of these expressions remains.
-There are two main explanatory venues for this pattern.
-The first targets the functional advantages a lack of upper-bounds may confer to language users, whereas the second focuses on a learnability advantage of simpler over more complex semantic representations.
-%
-%\paragraph{Function-based explanations.} Two assumptions are key to the pragmatic strengthening of weak alternatives: (the assumption of) cooperation and knowledge about the issue at hand.
-That is, the hearer needs to assume the speaker to be as informative as possible, i.e., not to withhold information, and that the speaker is knowlegeable, e.g., she knows whether {\em all students came}.
-Conversely, the speaker needs to assume the hearer to regard these conditions as satisfied.
-It is not difficult to imagine scenarios where either or both of these conditions are not given.
-For instance, the speaker may (be assumed to) not want to disclose all information about the students' attendance, or may have left early without being able to verify the attendance to a satisfactory degree.
-%
-%\hl{Discussion of functional pressures for a lack of upper-bounds}
-
-%\paragraph{Learning-based explanations.}
-%
-%\hl{Discussion of our main assumption that a lack of upper-bounds provides a learnability advantage framed in terms of relative representational simplicity over the codification of an upper-bound.
-{\bf Should this be made precise?
-If so, in which way?}}
-
-%\bibliographystyle{apacite}
-\bibliographystyle{unsrtnat}
-
-%\setlength{\bibleftmargin}{.125in}
-%\setlength{\bibindent}{-\bibleftmargin}
+\bibliographystyle{plainnat}
 \bibliography{./bounds-rmd}
 
+\newpage
+
+\section{Some thoughts + what may be missing}
+\begin{itemize}
+  \item I still need to review, read and find some references I'm withholding in the current draft (marked as \hl{references} in color). 
+  \item The general discussion should be expanded and polished. In particular concerning the discussion of lexical representations and the language of thought approach;
+  \item I still want to add references and clarify the discussion on homogeneous populations and frequency effects in the discussion of the model's predictions;
+  \item The analysis may benefit from further data and less of a qualitative overview of the parameters' effect (e.g. the regression analysis Michael conducted);
+  \item As you will have noticed when reading footnote $3$ I think the performance of $L_6$ being on par with $L_4$ is somewhat problematic;
+  \item There's a bit of repetition due to the early stages of the draft and the fact that scalar implicatures and some matrices are discussed in non-contiguous parts of the paper.
+Whether they stay depends on whether we roughly keep the current structure and length;
+  \item In principle, one could make a direct comparison between the case-study of \citet{kalish+griffiths:2007} on compositionality using iterated learning and iterated learning + functional pressure.
+This could be relegated to the appendix but I think this is somewhat tangential and wouldn't add much.
+The same goes for an explicit comparison of {\em some} vs.
+{\em some but not all} using a language of thought-style representation.
+I currently decided against it, but let me know if you see worth in either option.
+\end{itemize}
+
 \end{document}

### 2016-10-06 paper/paper.tex
first draft long CogSci version

@@
 \subsection{Discussion}
-Broadly speaking these results suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome selective pressures and stabilize in a population provided there is a bias for simpler representations.
+Broadly speaking these results suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome communicative pressures and stabilize in a population provided there is a bias for simpler representations.
 This prediction hinges on three assumptions.
 First, that language is pressured toward both expressivity and learnability.
-Second, that language use is not too unpredictable: low $\alpha$ or $\lambda$ render languages that rely on pragmatics too prone to communicative failure and more difficult to learn.
+Second, that language use is not too unpredictable -- low $\alpha$ or $\lambda$ render languages that rely on pragmatics too prone to communicative failure and more difficult to learn.
 Third, that learners prefer simpler over more complex lexical representations.
-An important remark on this third condition is that a combination of rationality in choice and maximization in learning requires a weaker bias towards simpler representations.
+An important addendum to this third condition being that a combination of rationality in choice and maximization in learning requires a weaker bias towards simplicity.
 Under these conditions the selection of lexical meanings lacking upper-bounds in populations of pragmatic speakers is robust against parameter perturbations.
@@
 
-While a major share of non upper-bounded lexical meanings is predicted by the literature, it is less clear to what extent other types should be present in the final population -- if at all.
-There are strong reasons to expect functionally suboptimal types $L_1$-$L_3$ to be ruled out since they fail to enable their users to communicate effectively.
+While a major share of non upper-bounded lexical meanings is predicted by the literature, it is less clear to what extent other types should be present in the final population, if at all.
+It seems reasonable to expect functionally suboptimal types $L_1$, $L_2$ and $L_3$ to be ruled out because they fail to enable their users to communicate effectively.
 However, this is not true of $L_4$.\footnote{$L_6$ presents a special case.
-In our current setup, it mirrors $L_5$ in enabling for a pragmatic strengthening of message $m_2$ rather than $m_1$.
+In our current setup, it mirrors $L_5$ in enabling for the pragmatic narrowing of message $m_2$ rather than $m_1$.
 However, this association of $s_2$ with $m_2$ under favorable parameter conditions is achieved by ruling out the ``some but not all''-state $s_1$ and not, as with scalar implicatures, the ``all''-state $s_2$.
 $L_6$ speakers therefore strengthen a ``some''-message to convey something paraphrasable as `some but not [some but not all]'.
-However, our representation of lexica as Boolean matrices is blind to this.} Notwithstanding, it is not unthinkable that natural language communities are homogeneous or that a single speaker may entertain $L_4$-like semantics for one scalar expression and $L_5$-like semantics for another (\hl{references on individual level differences in the derivation of scalar implicatures}).
+Our representation of lexica as Boolean matrices is blind to this anomaly.} Notwithstanding, the prediction that natural language communities are homogeneous or that a single speaker may entertain $L_4$-like semantics for one scalar expression and $L_5$-like semantics for another is is not implausible (\hl{references on individual level differences in the derivation of scalar implicatures}).
 Alternatively, a stronger tendency for posterior maximization has to be assumed (cf.
@@
 However, there is experimental evidence that suggests  that the pragmatic derivation of upper-bounds costs effort and takes additional processing time (cf.
-\citealt{deNeys+schaeken:2007, huang+snedeker:2009}.
+\citealt{deNeys+schaeken:2007, huang+snedeker:2009}).
 This raises the question at which point such usage-based cost undercuts the learnability advantage of simpler semantic representations.
-Should cost play a role then its effect is bound to depend on the frequency with which a given scalar expression is used.
+Should cost play a role, then its effect is bound to depend on the frequency with which a given scalar expression is used.
 We conject that frequently drawn scalar implicatures might fossilize to avoid cost, while infrequent ones could still be computed on-line.
-In turn, this opens a possible venue to address the previous question about expected use of $L_4$-like semantics.
+This opens a possible venue to address our previous question about the expected presence of $L_4$-like semantics.
 
@@
 This model generates predictions about lexicalization patterns and, more generally, effects of communicative pressures on language.
-We argued that the puzzle raised by semantics in light of pragmatics is hard to explain on purely functional grounds and that part of the answer instead may lie in learnability: Simpler semantic representations are more likely to be learned while pragmatic reasoning can counteract functional disadvantages otherwise incurred.
+We argued that the puzzle raised by semantics in light of pragmatics is hard to explain on purely functional grounds and that part of the answer may instead lie in learnability: Simpler semantic representations are more likely to be learned while pragmatic reasoning can counteract functional disadvantages otherwise incurred.
 This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics by offering an account of why (certain) pragmatic inferences fail to lexicalize.
 
-The main innovations of the model are its modular separation of expressivity and learnability, which allows for their isolated and combined analysis, as well as the learning process involving a joint inference over types of pragmatic behavior and lexical meaning.
+The main innovations of the model are its modular separation of expressivity and learnability, allowing for their isolated and combined analysis, as well as the learning process involving a joint inference over types of pragmatic behavior and lexical meaning.
 The goal to decouple but model both expressivity and learnability has also recently been addressed by \citet{kirby+etal:2015}.
-In contrast to the present proposal, however, expressivity is modelled as exerting its force in the production of learning data only.
+In contrast to our proposal, however, Kirby et al.
+model expressivity as exerting its force only in the production of learning data.
 This model's expressivity parameter thereby fulfills a similar role  to high values of $\lambda$ and $\alpha$ in making speaker behavior more deterministic.
-Consequently, it ``favors'' unambiguous languages.
+In this way, it ``favors'' unambiguous languages.
 Crucially, the degree of mutual understanding of interlocutors central to replication and to our notion of expressivity is not taken into consideration.
 Our reasoning behind its inclusion lies in the empirical and theoretical observation that learnability alone can have the undesirable outcome of favoring functionally defective languages.
-This outcome was showcased for a tautological language in our analysis but similar effects have been reported in a number laboratory experiments with human subjects where the participants' task was to learn and subsequently reproduce the language produced by a previous participant, leading to a proliferation of languages were a large number of meanings are associated with a single form (\hl{references}).
-In contrast, adding an interactive component to the experimental setup has been found to foster languages that enable interlocutors to distinguish meanings accurately  (\hl{references}; for a review of these studies and further references see \citealt{kirby+etal:2015, tamariz+kirby:2016}).
-More generally, there is an important difference between a language generating learnable linguistic data and its actual performance as a means of information transfer.
+This outcome is predicted by our analysis and similar effects have also been reported in a number laboratory experiments where the participants' task was to learn and subsequently reproduce the language produced by a previous participant.
+This process was shown to generally lead to a proliferation of languages that associated a large number of meanings with a single form (\hl{references}).
+In contrast, adding an experiments involving an interactive component have been found to foster languages that enable interlocutors to distinguish meanings accurately  (\hl{references}; for a review of these studies and further references see \citealt{kirby+etal:2015, tamariz+kirby:2016}).
+More generally, we think there is an important difference between a language generating learnable linguistic data and its actual performance as a means of information transfer.
 The former solely depends on the mechanism by which speakers associate form and meaning.
 The latter additionally hinges on the hearers linguistic experience and her capability to interpret linguistic input based on this experience.
-We contend that successful information transfer in a linguistic community is central the adoption of a communication system and that this measure is not faithfully reflected by taking only production into consideration.
+In sum, we contend that successful information transfer in a linguistic community is central the adoption of a communication system and that this measure is not faithfully reflected by taking only production into consideration.
 
-A central assumption to the model's predictions is the learning bias for simpler lexical representations.
+One of our central  assumptions is the learning bias for simpler lexical representations.
 Albeit implicit in our formalization, the conceptual foundation for a compression preference at the lexical level is grounded in work in the language of thought tradition put forward by \citet{piantadosi+etal:2012a,piantadosi+etal:underreview}.
-Following Piantadosi and colleagues we assumed that lexical meanings are expressed in a representational language and that the learners' task is to infer lexica that best explain the linguistic witnessed.
+Following Piantadosi and colleagues we assumed that lexical meanings are expressed in a representational language and that the learners' task is to infer lexica that best explain the data witnessed.
 Furthermore, we assumed learners to not only infer lexical meaning but types of linguistic use.
 Whether language use is culturally transmitted is an empirical matter that can not be settled here.
-However, this idealized model suggests that semantic structure evinced cross-linguistically that lends itself for pragmatic exploitation reflects lexical information that is both easy to learn and, more interestingly, also easy to learn how to exploit via mutual reasoning.
+However, this idealized model suggests that semantic structure evinced cross-linguistically that lends itself for pragmatic exploitation reflects lexical information that is both easy to learn and, perhaps more interestingly, also easy to learn how to exploit via mutual reasoning.
 
@@
 Language change is affected by intertwined pressures.
-We investigated the selection of non-upper-bounded lexical meanings as a case study on role of semantics and pragmatics in satisfying such pressures.
+We investigated the selection of lexical meanings lacking upper-bounds in a case study on role of semantics and pragmatics in satisfying such pressures.
 We showed that, when pressured for learnability and expressivity, the former drives for simpler semantic representations inasmuch as pragmatics can compensate for their lack of expressivity in use.
 That is, the relative learnability advantage of simpler semantics may offer an answer to why natural languages fail to lexicalize certain pragmatic inferences.
-More broadly, we argued that semantic patterns can be explained by taking into consideration the way in which they are used in interaction as well as their representational complexity.
+More broadly, we argued that semantic patterns can be explained by taking into consideration the way in which they are used in interaction as well as their representational complexity and learnability.
 
@@
 \begin{itemize}
-  \item I still need to review, read and find some references I'm withholding in the current draft (marked as \hl{references} in color). 
-  \item The general discussion should be expanded and polished. In particular concerning the discussion of lexical representations and the language of thought approach;
-  \item I still want to add references and clarify the discussion on homogeneous populations and frequency effects in the discussion of the model's predictions;
+  \item I still need to review, read and find some references I'm withholding in the current draft (marked as ``\hl{references}'' in color). 
+  \item The discussion (sub) sections should be expanded and polished.
+In particular the discussion concerning lexical representations and the language of thought.
+I'm aware that this part is not as readable as the rest;
   \item The analysis may benefit from further data and less of a qualitative overview of the parameters' effect (e.g. the regression analysis Michael conducted);
-  \item As you will have noticed when reading footnote $3$ I think the performance of $L_6$ being on par with $L_4$ is somewhat problematic;
-  \item There's a bit of repetition due to the early stages of the draft and the fact that scalar implicatures and some matrices are discussed in non-contiguous parts of the paper.
-Whether they stay depends on whether we roughly keep the current structure and length;
-  \item In principle, one could make a direct comparison between the case-study of \citet{kalish+griffiths:2007} on compositionality using iterated learning and iterated learning + functional pressure.
-This could be relegated to the appendix but I think this is somewhat tangential and wouldn't add much.
+  \item As you will have noticed when reading footnote $3$, I think the performance of $L_6$ being on par with $L_4$ is somewhat problematic;
+  \item There's a fair share of repetition, this is partly due to the fact that scalar implicatures and some matrices are discussed in non-contiguous parts of the paper.
+Whether the repetition is necessary depends on whether we roughly keep the current structure and length or change things;
+  \item In principle, we could make a direct comparison between the case-study of \citet{griffiths+kalish:2007} on compositionality using iterated learning and iterated learning + functional pressure.
+This could be relegated to the appendix.
+However, I'm inclined to think this is somewhat tangential and wouldn't add much.
 The same goes for an explicit comparison of {\em some} vs.
 {\em some but not all} using a language of thought-style representation.
-I currently decided against it, but let me know if you see worth in either option.
+Let me know if you see worth in either option.
 \end{itemize}

### 2016-10-06 paper/paper.tex
references on laboratory experiments in IL and their somewhat choppy inclusion in the main text

@@
 Crucially, the degree of mutual understanding of interlocutors central to replication and to our notion of expressivity is not taken into consideration.
-Our reasoning behind its inclusion lies in the empirical and theoretical observation that learnability alone can have the undesirable outcome of favoring functionally defective languages.
-This outcome is predicted by our analysis and similar effects have also been reported in a number laboratory experiments where the participants' task was to learn and subsequently reproduce the language produced by a previous participant.
-This process was shown to generally lead to a proliferation of languages that associated a large number of meanings with a single form (\hl{references}).
-In contrast, adding an experiments involving an interactive component have been found to foster languages that enable interlocutors to distinguish meanings accurately  (\hl{references}; for a review of these studies and further references see \citealt{kirby+etal:2015, tamariz+kirby:2016}).
-More generally, we think there is an important difference between a language generating learnable linguistic data and its actual performance as a means of information transfer.
+That is, while our proposal combines bidirectional horizontal transmission with its vertical and unidirectional counterpart, Kirby et al.'s model only considers the latter's influence.
+Our reasoning behind the inclusion of the former lies in the empirical and theoretical observation that learnability alone can lead the selection of functionally defective languages, as showcased by the tautological language $L_3$ in our analysis.
+This outcome has been reported in a number laboratory experiments where the participants' task was to learn and subsequently reproduce the language produced by a previous participant, leading to a proliferation of languages that associated a large number of meanings with a single form (see e.g.
+\citealt{silvey+etal:2014} and experiment 1 in \citealt{kirby+etal:2008}).
+In contrast, experiments involving an interactive component have been found to foster languages that enable interlocutors to distinguish meanings accurately  (e.g.
+\citealt{fay+etal:2013}; for a review of laboratory results under the iterated learning paradigm and further discussion see \citealt{kirby+etal:2015, tamariz+kirby:2016}).
+It is not evident how to compare these empirical predictions given their variation in meaning spaces, modes of transmission, iterations and feedback given to participants.
+However, we take these results to suggest that there is an important difference between a language generating learnable linguistic data and its actual performance as a means of information transfer.
 The former solely depends on the mechanism by which speakers associate form and meaning.

### 2016-10-15 3rmd-singlescalar_withNoise.py
fixed minor issues with 1rmd-singlescalar: setting cost to 0 and initial export of noise-perturbed alternative  dynamics

@@ -18,11 +18,13 @@ import csv
 #####
 l1,l2,l3,l4,l5,l6 = np.array( [[0.,0.],[1.,1.]] ), np.array( [[1.,1.],[0.,0.]] ), np.array( [[1.,1.],[1.,1.]] ), np.array( [[0.,1.],[1.,0.]] ), np.array( [[0.,1.],[1.,1.]] ), np.array( [[1.,1.],[1.,0.]] )
 
-alpha = 1 # rate to control difference between semantic and pragmatic violations
-cost = 0.1 # cost for LOT-concept with upper bound
-lam = 30 # soft-max parameter
+alpha = 10 # rate to control difference between semantic and pragmatic violations
+cost = 0 # cost for LOT-concept with upper bound
+lam = 20 # soft-max parameter
 k = 3  # length of observation sequences
-sample_amount = 15 #amount of k-length samples for each production type
+sample_amount = 50 #amount of k-length samples for each production type
+deltaE = 0.3 # probability of perceiving S-all, when true state is S-sbna
+deltaA = 0.1 # probability of perceiving S-sbna, when true state is S-all
 
 gens = 20 #number of generations per simulation run
 runs = 50 #number of independent simulation runs
@@ -30,19 +32,8 @@ runs = 50 #number of independent simulation runs
 states = 2 #number of states
 messages = 2 #number of messages
 
-learning_parameter = 1 #prob-matching = 1, increments approach MAP
+learning_parameter = 10 #prob-matching = 1, increments approach MAP
 state_freq = np.ones(states) / float(states) #frequency of states s_1,...,s_n 
-
-
-f = csv.writer(open('./results/singlescalar-a%d-c%.2f-l%d-k%d-samples%d-learn%.2f-g%d-r%d.csv' %(alpha,cost,lam,k,sample_amount, learning_parameter, gens,runs),'wb')) #file to store mean results
-
-f.writerow(["run_ID", "t1_initial", "t2_initial","t3_initial","t4_initial","t5_initial","t6_initial","t7_initial","t8_initial","t9_initial","t10_initial","t11_initial","t12_initial","alpha", "prior_cost_c", "lambda", "k", "sample_amount", "learning_parameter", "generations", "t1_final", "t2_final","t3_final","t4_final","t5_final","t6_final","t7_final","t8_final","t9_final","t10_final","t11_final","t12_final"])
-
[…]

### 2016-10-16 mf_stuff/note-on-noisy-RMD/note-on-noisy-RMD.tex
model extension note

\documentclass[fleqn,reqno,10pt]{article}

\usepackage{myarticlestyledefault}

\title{Note on a noisy variant of the RMD}
\author{Michael Franke}
\date{}

\begin{document}
\maketitle

\subsection*{Motivation}

Our present model computes the mutation probability that a teacher type $t_i$ has
offspring of type $t_j$ as:

\begin{align*}
  Q_{ij}  \propto \sum_{d \in D} P(d|t_i) F(t_j,d) \,, \ \  \text{where} \ \
F(t_j,d)  \propto P(t_j|d)^l \ \  \text{and} \ \ P(t_j|d) \propto P(t_j) P(d|t_j)
\end{align*}

\noindent What makes for the evolution of a natural lower-bound-only semantics for \emph{some}
is the \textbf{cognitive bias} encoded in the prior $P({t_j})$. While it is not unnatural or
indefensible to assume such a cognitive bias, we also do not have a strong argument in support
of it. We also do not want to argue that \emph{this} is the right explanation, do we? Our main
achievement is rather a perspicuous alignment of conceptual bits and pieces into a working
formal model. The main ingredients of this model are (i) a combination of learning biases and
selective pressure towards efficient communication, and (ii) the distinction between semantic
meaning and pragmatic use, both of which are shaped and honed by evolution at once. This is
great, but maybe we might want to have more.

Let's focus on (i) for a moment. In a recent paper, Pedro Correia and I have explored an
evolutionary dynamic that ensues from ``noise-perturbed imitation''
\citep{FrankeCorreia2016:Vagueness-and-I}. The idea is that agents update their strategies by
occasionally imitating choices with a probability proportional to how good these observed
choices are. Crucially, however, agents may not always perceive a state correctly. We applied
this to vagueness, where this is most natural: some actual state $s_a$ has a chance
$P_N(s_p \mid s_a)$ of being observed as $s_p$, e.g., maybe
$P_N(s_p \mid s_a) = \text{Gaussian}(s_p, \mu = s_a, \sigma = \text{something})$ is a normal
distribution around the actual state. What we showed in this paper is that noisy perception of
states not only leads to vagueness in signal meaning, but also speeds up the evolutionary
process by unifying and regularizing agents' strategies. We interpreted this as a sort of
emergent ``\emph{as-if} generalization'': at the population level it seems as if agents would
extrapolate and generalize what they have learned about one state to nearby similar states. But
this is not, of course, what actually happens. The mere presence of systematic noise in the
transmission of strategies can introduce regularization that looks as if the agents have a
learning bias. But, most importantly, this disturbance of transmission fidelity is due to
perceptual noise and properties of the environment, not learning biases. In other words,
learning biases are clearly not the only transmission biases that can shape evolution alongside
functional pressure. Environmental and perceptual noise can play a role too. This also
  ties in with a bunch of work on work that tries to explain some features of language as being
  optimal adaptations to a ``noisy-channel'' model (work by Ted Gibson, Steve Piantadosi
  etc.). However, in this work the noise is rather in the signal $m$ not the state perception
  $t$. In general, then, I believe that there is a certain lacuna here: paying attention to
  the effects of systematic distortions of state perceptions on the cultural evolution of
  language.

So, I asked myself whether noisy perception of states could also be included into our model and
whether this could have a similar effect to a cognitive bias in favor of lower-bound-only
semantics, without actually making use of such a cognitive bias. The answers in short are:
``yes, yes''. However, perhaps unsurprisingly, not all conceivable kinds of perception error
lead to this outcome. So, if we wanted to defend that such-and-such a model is the right
explanation we'd be hard pressed to defend a particular noise structure. I don't think that we
can do this. The main message, in my view, should rather be: look we have two possible
lines of explanation for lower-bound semantics, one with cognitive biases, the other with
perceptual errors. Our main contribution is showing how two things can interact in cultural
evolution, namely (i) transmission biases and (ii) functional pressure, and that, crucially,
transmission biases need not only be cognitive/learning biases but can also ensue from
perception and the environment. We then have a case study that puts everything into place and
demonstrates the latter claim. That's the story line that we could adopt if we like the
following model extension and whatever comes with it.

\printbibliography[heading=bibintoc]

\end{document}

### 2016-10-16 mf_stuff/note-on-noisy-RMD/note-on-noisy-RMD.tex
extension model sketch final; minor tweaks to extension model code

@@
 offspring of type $t_j$ as:
-
 \begin{align*}
@@
 
+\subsection*{RMD with noisy perception}
+
+This model is implemented in file \texttt{3rmd-singlescalar\_withNoise.py} in the new repository
+\texttt{rmd-bounds2}.
+
+We have two states: $s_\exists$ (some but not all) and $s_\forall$ (all). Let's assume that the
+probabilities for observing the state in the column when the actual state is the state in the
+row are given by this table:
+\begin{center}
+  \begin{tabular}{ccc}
+    & $s_\exists$ & $s_\forall$ \\ \midrule 
+    $s_\exists$ & $1-\epsilon$ & $\epsilon$ \\
+    $s_\forall$ & $\delta$ & $1-\delta$
+  \end{tabular}
+\end{center}
+So, $\epsilon$ is the error probability when the true state is $s_\exists$ and $\delta$ is the
+error probability when the true state is $s_\forall$. How does this affect our model of
+Bayesian learning?
+
+Let's denote the probability that the teacher (learner) observes state $s_t$ ($s_l$) when the
+actual state is $s_a$ as $P_N(s_t \mid s_a)$ ($P_N(s_l \mid s_a)$ ). This is given by the table
+above. We can then derive the following probabilities. Firstly, the probability that $s_a$ is
+the actual state when the learner observes $s_l$ is just Bayes rule, combining prior of $s_a$
+with the noise from above:
+\begin{align*}
+  P_N(s_a \mid s_l) \propto P(s_a) \ P_N(s_l \mid s_a)\,.
+\end{align*}
+Secondly, the probability that the teacher observes $s_t$ when the learner observes $s_l$ is:
+\begin{align*}
+  P_N(s_t \mid s_l) = \sum_{s_a} P(s_a \mid s_l) \ P_N(s_t \mid s_a)\,.
+\end{align*}
+Finally, this gives us the probability that a teacher of type $t$ produces a datum that is
+perceived by the listener as $d = \tuple{s_l, m}$:
+\begin{align*}
+  P_N(\tuple{s_l, m} \mid t) = \sum_{s_t} P_N(s_t \mid s_l) \ P(m \mid s_t, t)\,.
+\end{align*}
+Generalize this to a sequence of perceived data $d_l$ and write $P_N(d_l \mid t)$. We then
+define the noise-perturbed mutation matrix as:
+\begin{align*}
+  Q_{ij}  \propto \sum_{d_l \in D} P(d_l \mid t_i) F(t_j,d_l) \,, \ \  \text{where $F(t_j,d)$
+    as before.}
+\end{align*}
+In words, it may be the case that learner and/or teacher do not perceive the actual state as
+what it is. They are not aware of this, and produce/learn as if what they observed was the
+actual state. In particular, the learner does not reason about noise when he tries to infer the
+speaker's type. He takes what he observes a state to be as the actual state that the teacher
+has seen as well and computes which type would have most likely generated the message to this
+state. This can lead to biases of inferring the ``wrong'' teacher type, if the noise makes some
+types err in a way that resembles the noiseless behavior of other types. I.e., this could, in
+principle, induce transmission biases that look as is there was a cognitive bias in favor of a
+particular type, simply because that type better explains the noise.
+
+On top of changing the mutation matrix in this way, we also need to adapt the calculation of
+expected utilities, taking into consideration that states are perceived noisily. So, where
+before we had:
+\begin{align*}
+  U_S(t_i,t_j) = \sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')\,,
+\end{align*}
+we now have:
+\begin{align*}
+  U_S(t_i,t_j) = \sum_{s_a}  P(s_a) \sum_{s_t} P_N(s_t \mid s_a) \sum_m S_n(m|s_t;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')\,.
+\end{align*}
+
+\subsection*{Results}
+
+In a nutshel: this works for some parameter settings but not for others. I have not fully
+explored the whole space. I would be content with a paper that says: there are parameter values
+that produce behavior as if there are cognitive biases without cognitive biases. I believe that
+this would be an additional and very interesting contribution, even if there are noise
+structures that have other effects.
+
+One parameter setting that gives us the desired result (most prevalent type is $t_{11}$) is:
+\begin{verbatim}
+alpha = 5 # rate to control difference between semantic and pragmatic violations
+cost = 0 # cost for LOT-concept with upper bound
+lam = 20 # soft-max parameter
+k = 3  # length of observation sequences
+sample_amount = 50 #amount of k-length samples for each production type
+epsilon = 0.3 # probability of perceiving S-all, when true state is S-sbna
+delta = 0.1 # probability of perceiving S-sbna, when true state is S-all
+gens = 20 #number of generations per simulation run
+runs = 50 #number of independent simulation runs
+states = 2 #number of states
+messages = 2 #number of messages
+learning_parameter = 10 #prob-matching = 1, increments approach MAP
+\end{verbatim}
+One sample result is these parameter values is:
+\begin{verbatim}
+[ 0.01 ,  0.01 ,  0.01 ,  0.012,  0.028,  0.024,  0.01 ,  0.01 ,
+        0.01 ,  0.301,  0.471,  0.102]
+\end{verbatim}
+Importantly, we don't need costs ($c = 0$)! Unfortunately, we do
+need to tinker with $\alpha$ (for defensible, techical reasons) and we need to assume that
+$\epsilon > \delta$, otherwise either $t_{10}$ or even $t_{12}$ may dominate.
+
+That means that if we wanted to also argue that this is a plausible explanation for
+lower-bounds semantics, we'd need to argue why it is natural that $\epsilon > \delta$. Maybe
+such a story can be given, but I presently do not see why/how any such story is obviously
+better than any other story I could come up with for any other relation of $\epsilon$ and
+$\delta$.
+
+One reasoning for $\epsilon > \delta$ could be this. When the true state is $s_a$, we would
+actually need negative evidence that some item/person/whatever does not have a
+property. Negative evidence of this kind might be relatively infrequent, especially if
+untrue. On the other hand, perceiving $s_a$ when the true state is $s_e$ could be more
+frequent, because it would, for example, require wrong information about the domain size. It
+could also be a cause of overemphasizing: speakers tend to exaggerate and want to claim
+\emph{all}, when in fact that is not true \citep[e.g.][for a model that has the speakers'
+tendency to overemphasize as a motor of language change]{Schaden2012:Modelling-the-A}.
+
 \printbibliography[heading=bibintoc]

### 2016-10-16 3rmd-singlescalar_withNoise.py

@@ -18,13 +18,13 @@ import csv
 #####
 l1,l2,l3,l4,l5,l6 = np.array( [[0.,0.],[1.,1.]] ), np.array( [[1.,1.],[0.,0.]] ), np.array( [[1.,1.],[1.,1.]] ), np.array( [[0.,1.],[1.,0.]] ), np.array( [[0.,1.],[1.,1.]] ), np.array( [[1.,1.],[1.,0.]] )
 
-alpha = 10 # rate to control difference between semantic and pragmatic violations
+alpha = 5 # rate to control difference between semantic and pragmatic violations
 cost = 0 # cost for LOT-concept with upper bound
 lam = 20 # soft-max parameter
 k = 3  # length of observation sequences
 sample_amount = 50 #amount of k-length samples for each production type
-deltaE = 0.3 # probability of perceiving S-all, when true state is S-sbna
-deltaA = 0.1 # probability of perceiving S-sbna, when true state is S-all
+epsilon  = 0.3 # probability of perceiving S-all, when true state is S-sbna
+delta = 0.1 # probability of perceiving S-sbna, when true state is S-all
 
 gens = 20 #number of generations per simulation run
 runs = 50 #number of independent simulation runs
@@ -50,17 +50,13 @@ def normalize(m):
     return m / m.sum(axis=1)[:, np.newaxis]
 
 ## state confusability
-state_confusion_matrix = np.array([[1-deltaE, deltaE],
-                                   [deltaA, 1-deltaA]])
+state_confusion_matrix = np.array([[1-epsilon , epsilon ],
+                                   [delta, 1-delta]])
 lh_perturbed = likelihoods
 PosteriorState = normalize(np.array([[state_freq[sActual] * state_confusion_matrix[sActual, sPerceived] for sActual in xrange(states)] \
  for sPerceived in xrange(states)])) # probability of actual state given a perceived state
 DoublePerception = np.array([[np.sum([ state_confusion_matrix[sActual, sTeacher] * PosteriorState[sLearner,sActual] \
  for sActual in xrange(states)]) for sTeacher in xrange(states) ] for sLearner in xrange(states)])# probability of teacher observing column, given that learner observes row
-
-print PosteriorState
-print DoublePerception
-
 for t in xrange(len(likelihoods)):
    for sLearner in xrange(len(likelihoods[t])):
        for m in xrange(len(likelihoods[t][sLearner])):
[…]

### 2016-11-01 paper/paper-v2.tex
second draft version

--- 2016-10-06 paper/paper.tex
@@
 \usepackage{color}
+\usepackage{booktabs}
 %\newcommand{\mvalueof}[1]{\llbracket#1\rrbracket}
@@
 
-\title{Communicative pressures at the semantics-pragmatics interface:\\ Learning biases may prevent the lexicalization of pragmatic inferences}
+\title{Communicative pressures at the semantics-pragmatics interface}
 
@@
 \begin{abstract}
-Mutual reasoning endows speakers with the ability to convey information that goes beyond the literal meaning of expressions.
+According to standard linguistic theory, the conventional meaning of expressions is systematically modified by general pragmatic rules governing their use.
+This implies that models of the cultural evolution of meaning should likewise take into consideration that observable language use is a complex interaction of semantic representations and pragmatic rules.
+To this end, we present a game theoretic model of the cultural evolution of language where communicative pressures work on abstract semantic representations and pragmatic usage rules.
+Our model traces two evolutionary forces and their interaction: (i) fitness-based pressure towards communicative efficiency and (ii) systematic transmission perturbations when linguistic knowledge is transferred from one agent to another.
+The latter can contain general cognitive or learning biases, but can also contain other sources of systematic noise, e.g., resulting from errors in perception.
+We illustrate the model based on a case study showing that cognitive biases that favor simple semantic representations can prevent the lexicalization of pragmatic inferences.
+We also show that, technically speaking, it is possible that environmental factors, such as perceptual errors during acquisition, can produce evolutionary outcomes that look as if such cognitive biases are present even if they are not.
+\end{abstract}
+
+%Mutual reasoning endows speakers with the ability to convey information that goes beyond the literal meaning of expressions.
 A particularity of certain classes of lexical meanings is that they enable for such pragmatic enrichments in a notably productive fashion.
@@
 This result sheds light on the lack of lexicalization of scalar inferences, as well as the semantics-pragmatics distinction more generally.
-\end{abstract}
 
@@
 What is conveyed usually goes beyond what is said.
-A request for a blanket can be politely veiled by uttering ``I'm cold'', a temporal succession of events conveyed by the order in which conjuncts appear as in``I traveled to Paris and got married'', or an invitation be declined by saying ``I have to work''.
-An influential explanation of the relation between the literal meaning of such expressions (semantics) and what they are intended and interpreted to convey (pragmatics) due to \citet{grice:1975} sees the latter as a product of a process of mutual reasoning.
+A request for a blanket can be politely veiled by uttering ``I'm cold'', a temporal succession of events can be conveyed by the order in which conjuncts appear as in ``I traveled to Paris and got married'', or an invitation can be declined by saying ``I have to work''.
+An influential explanation of the relation between the literal meaning of expressions (semantics) and what they are intended and interpreted to convey (pragmatics) due to \citet{grice:1975} views the latter as a product of a process of mutual reasoning about rational language use.
 For instance, under the assumption that the speaker is cooperative and relevant, ``I have to work'' may be interpreted as providing a reason why the speaker will not be able to accept the invitation -- why would she have said so otherwise?
+Crucially, many classes of lexical meanings allow for pragmatic enrichments in a notably systematic fashion.
 
-Nowadays, it is common to draw a distinction between semantics and pragmatics in linguistic theorizing.
-Importantly, many classes of lexical meanings allow for pragmatic enrichments in a notably productive fashion.
-For instance, while John can eat all the cookies and still truthfully say ``I ate some cookies'', the use of {\em some} over {\em all} is usually taken to convey that the latter does not hold.
-That is, ``I ate some cookies'' is pragmatically strengthened to convey that {\em some but not all} were eaten (\hl{references for experimental evidence that people in fact draw these inferences}).
-This inference is not restricted to quantifiers such as {\em some} and {\em all} but can be licensed for a large class of natural language expressions,  ranging from numerals such as {\em four} and {\em five} to adjectives such as {\em cold} and {\em big}.
-
-These pragmatic inferences, so-called {\em scalar implicatures}, have been the focus of many studies at the semantics-pragmatics interface.
-However, an issue that has received little attention is the justification of semantic structure in light of the systematic enrichment offered by pragmatics.
-The present investigation seeks to fill this gap by analyzing the effects linguistic pressures have on the selection and pervasiveness of particular lexical meanings under consideration of their possible pragmatic enrichments.
+Productive pragmatic inferences such as scalar and manner implicatures have been at the focus of many studies on the semantics-pragmatics interface.
+However, an issue that has received little attention is the justification of semantic structure in light of the systematic enrichments offered by pragmatics.
+Similarly, multiple models of the cultural evolution of language have been put forward to analyze properties such as compositionality and combinatoriality, but the selection of particular semantics in light of pragmatics remains uncharted territory.
+The present investigation seeks to fill these gaps by analyzing the effects linguistic pressures have on the selection and pervasiveness of particular lexical meanings under consideration of their systematic pragmatic modification.
+Accordingly, our main goal is to uncover how transmission biases and functional pressure interact in the cultural evolution of meaning when pragmatics is taken into account.
+Crucially, we argue that transmission biases need not be cognitive in nature but can also ensue from perception and the environment.
 
@@
 The emergence and change of linguistic structure is influenced by many intertwined factors.
-These range from biological and socio-ecological to cultural \citep{steels:2011}.
+These range from biological and socio-ecological to cultural ones \citep{steels:2011}.
 Social and ecological pressures determine communicative needs while biology determines the architecture that enables and constrains the means by which they can be fulfilled.
-In the following, our focus lies on the remaining cultural aspect, wherein processes of linguistic change are viewed as shaped by language use and its transmission.
+In the following, our focus lies on cultural aspects, wherein processes of linguistic change are viewed as shaped by language use and its transmission.
 That is, as a result of a process of cultural evolution.
@@
 The idea that language is influenced by communicative pressures has played a pivotal role in synchronic and diachronic analyses at latest since \citeposs{zipf:1949} rationalization of the approximation of word frequency rankings by a power law distribution as competing hearer and speaker preferences (e.g.
-in \citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007, piantadosi:2014,kirby+etal:2015} among many).
+in \citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007, piantadosi:2014,kirby+etal:2015}).
 In recent years this line of research has led to a surge of approaches that seek to analyze the effects of such pressures from multiple angles ranging from simulations to experiments with human and robotic agents (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
-Our starting point is given by the overarching argument that has crystalized from the accumulated mathematical, experimental and cross-linguistic evidence in this literature: Natural languages need to be well-adapted to communicative needs within a linguistic community, but also need to be learnable to survive their faithful transmission across generations.
+Our starting point is given by the overarching argument that has crystalized from this literature: Natural languages need to be well-adapted to communicative needs within a linguistic community, but also need to be learnable to survive their faithful transmission across generations.
 More succinctly; natural languages are pressured for expressivity and learnability.
@@
 A language with a single form-meaning association is easy to learn but lacking in expressivity.
-Conversely, a language that associates a distinct form with all possible meanings its users may want to convey is maximally expressive but challenging to acquire.
+Conversely, a language that associates a distinct form with all possible meanings one may want to convey is maximally expressive but challenging to acquire.
 The most prominent problem that arises from this tension is that of acquiring a language to express a potentially infinite set of meanings through finite means \citep{kirby:2002}.
 However, this is not the only challenge learners confront.
-A more central issue to our explanandum is that of selecting particular hypotheses out of a potentially infinite space of alternatives compatible with the data learners are exposed to.
+More central to our explanandum is the issue of selecting particular hypotheses out of a potentially infinite space of alternatives compatible with the data learners are exposed to.
 At the semantics-pragmatics interface this concerns the selection between functionally (near-)equivalent lexical meanings, noting in particular that what is systematically conveyed through pragmatics could alternatively be codified lexically.
-Notwithstanding, often lexical meanings that allow interlocutors to draw these inferences seem to resist their lexicalization.
+Notwithstanding, lexical meanings that allow for pragmatic modifications often seem to resist the lexicalization of their pragmatic component.
 The question is why.
 
-We assume an integral part of the answer to be that learners are a priori biased towards simpler, more compressed, lexical representations.
-That is, rational learners should prefer simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
-In a nutshell, our hypothesis is that codifying more semantically is a priori dispreferred by learners and in cases where pragmatics offers conventional means of enrichment, the communicative disadvantage that speakers would otherwise incurr is leveled.
-In this way, semantics and pragmatics play a synergic role in overcoming linguistic pressures for expressivity and learnability which may prevent the lexicalization of pragmatic inferences.
+We assume an integral part of the answer to lie in the effects of transmission perturbations on language when passed from one generation to the next.
+Such perturbations may take the form of learning biases, but may also stem from extraneous factors such as a noisy perception of the environment.
+In particular, we show how the presence of systematic noise can introduce regularizations that resemble those obtained under the assumption of a learning bias.
 
-In the following, we model these components using the replicator-mutator dynamics, combining functional pressure on successful communication with effects of learning biases on (iterated) Bayesian learning \citep{griffiths+kalish:2007}.
+In the following, we model these components using the replicator-mutator dynamics, combining functional pressure on successful communication with effects of transmission perturbations on (iterated) Bayesian learning \citep{griffiths+kalish:2007}.
 The semantics-pragmatics distinction and its bearing on production and comprehension are captured by a probabilistic model of rational language use with different degrees of pragmatic sophisitication and languages \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
 The remainder of this section introduces these components individually together with the assumptions underlying them.
-These are: the representation of languages and their use (\S\ref{sec:languages+use}), pressures towards expressivity (\S\ref{sec:expressivity}) and learnability (\S\ref{sec:learnability}) regulated by the replicator and mutator dynamics, respectively, as well as a bias towards simpler semantic representations, codified in the learners' prior.
-After laying out the model, we analyze its predictions in a case-study on the lack of lexicalization of scalar implicatures in \S\ref{sec:si-case-study}.
+These are: the representation of languages and their use (\S\ref{sec:languages+use}), pressures towards expressivity (\S\ref{sec:expressivity}) and learnability (\S\ref{sec:learnability}) regulated by the replicator and mutator dynamics, respectively.
+After laying out the model, we analyze its predictions in a case-study on the lack of lexicalization of scalar implicatures in \S\ref{sec:si-case-study} and compare the outcome predicted by different transmission biases in \S\ref{subsec:bias} and \S\ref{subsec:noise}.
 
 \subsection{Lexica and linguistic behavior}\label{sec:languages+use}
-Lexica codify the truth-conditions of a language's expressions, i.e., its semantics.
+Lexica codify the truth-conditions of a language's expressions.
 Following \citet{franke+jaeger:2014}, a convenient way to represent lexica is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs (meanings) to convey and $M$ a set of messages (forms available in the language).
-For example, the following two fragments determine the truth-conditions of two messages, $m_1$ and $m_2$, for two states, $s_1$ and $s_2$.
-
-\begin{centering}
-$L_a$ = \bordermatrix{~ & m_1 & m_2 \cr 
-                  s_1 & 1 & 0 \cr
-                  s_2 & 1 & 1 \cr} \hspace{2cm} $L_b$ = \bordermatrix{~ & m_1 & m_2 \cr 
-                  s_1 & 1 & 0 \cr
-                  s_2 & 0 & 1 \cr}\\[0.5cm]
-\end{centering}
-
-According to lexicon $L_a$ message $m_1$ is true of state $s_1$ as well as of $s_2$.
-In contrast, $m_1$ is only true of $s_1$ in $L_b$.
-Otherwise, the two languages are truth-conditionally equivalent.
-While these representations are very simple, they suffice to make the distinction between semantics and pragmatics precise.
 
 We distinguish between two kinds of linguistic behavior.
-{\em Literal interlocutors} produce and interpret messages literally.
-That is, their linguistic choices are guided only by their lexica.
+{\em Literal interlocutors} produce and interpret messages literally, being guided only by their lexica.
 In contrast, {\em pragmatic interlocutors} engage in mutual reasoning to inform their choices.
-For instance, a rational speaker of $L_a$ who reasons about her addressee may use $m_1$ to exclusively signal state $s_1$ since $s_2$ can unambiguously be conveyed through $m_2$.
-Accordingly, rational hearers that expect their interlocutor to reason along these lines will interpret ambiguous $m_1$ as $s_1$.
-For our purposes, an important consequence of this process of mutual reasoning is that it can render $L_a$ nearly indistinguishable from $L_b$ in terms of language use.
-
-Following models of rational language use such as Rational Speech Act models \citep{frank+goodman:2012} and their game-theoretic predecessors, the Iterated Best/Quantal Response models \citep{franke:2009,franke+jaeger:2014}, this kind of signaling behavior is captured by a reasoning hierarchy.
+Following models of rational language use such as Rational Speech Act models \citep{frank+goodman:2012} and their game-theoretic predecessors, the Iterated Best/Quantal Response models \citep{franke:2009,franke+jaeger:2014}, these behaviors are captured by a reasoning hierarchy.
 The hierarchy's bottom, level $0$, corresponds to literal language use.
@@
 Below, (\ref{h:level0}) and (\ref{h:leveln}) specify the behavior of literal and pragmatic hearers of a language $L$.
-Literal and pragmatic speakers are specified in (\ref{s:level0}) and (\ref{s:leveln}).
+Their speaker counterparts are given in (\ref{s:level0}) and (\ref{s:leveln}).
 
@@
 &H_{0}(s|m;L) \propto pr(s) L_{sm} \label{h:level0}\\
-&S_{0}(m|s;L) \propto \exp(\lambda \; L_{sm}) \label{s:level0}\\
+&S_{0}(m|s;L) \propto \exp(L_{sm}^\alpha) \label{s:level0}\\
 &H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
@@
 In the following this prior is assumed to be uniform across hearers.
-The behavior of literal speakers, given in (\ref{s:level0}), is regulated by a soft-max parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}.
+The behavior of literal speakers, given in (\ref{s:level0}), is regulated by a parameter $\alpha$ which controls the sensitivity to which speakers prefer one signal over another based on its expected communicative success.
+
+Pragmatic behavior is similar to its literal counterparts.
+Their difference lies in that level $n+1$ speakers/hearers reason about level $n$ hearer/speaker behavior instead of solely relaying on their lexicon.
+That is, they reason about the way a rational level $n$ interlocutor would use or interpret a message, and behave according to these expectations.
+Speaker behavior is further influenced by a soft-max parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}.
 As $\lambda$ increases, choices made in production are more rational in that higher values lead to behavior that is increasingly in line with expected utility maximization.
 
-Pragmatic behavior is similar to its literal counterparts.
-As hinted at in the informal sketch above, their difference lies in that level $n+1$ speakers/hearers reason about level $n$ hearer/speaker behavior instead of solely relaying on their lexicon.
-That is, they reason about the way a rational level $n$ interlocutor would use or interpret a message, and behave according to these expectations.
-Pragmatic production is further regulated by a parameter $\alpha$ which controls the tension between semantics and pragmatics, $\alpha \in (0,1]$.
-Lower values lead to more literal production whereas higher values lead to stronger pragmatic behavior.
-
-We call the combination of a lexicon with its use, i.e., a level in the reasoning hierarchy, a type. These are the basic units on which our population dynamics operate.
+We call the combination of a lexicon with its use, i.e., a level in the reasoning hierarchy, a {\em type}. These are the basic units on which our model's dynamics operate.
 
@@
 Under this view, a type's ability to convey and interpret information successfully confers it a higher fitness, a measure that is relative to the success of other types in the population.
-In the simplest models fitness directly translates into the proportion of types present in the population after a generational turnover.
-As a consequence, fitter types flourish while unfit ones are driven out.
+In the simplest models, fitness directly translates into the proportion of types present in the population after a generational turnover.
 This association of communicative success within a population with changes in the proportion of types present in it creates a feedback loop that pressures the population towards greater expressivity.
@@
 As noted above, the fitness of type $i$ is equal to its relative communicative success within this population, $f_i = \sum_j x_j \text{EU}(t_i,t_j)$.
-That is, its fitness is the sum of its expected communicative success in interacting with other types weighted by the latter's population share.
-Communicating well with types that are well-represented in the population therefore confers a higher fitness than doing so with underrepresented ones.
 The expected communicative success of $i$ and $j$ is obtained by considering the average success of $i$ conveying information to $j$ and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$.
@@
 In their series of papers on language evolution, Nowak and colleagues did not only consider expressivity but also recognized the central role of the fidelity by which languages are transmitted.
-Linguistic production can be prone to errors, and multiple languages may be compatible with the data learners are exposed to.
+Crucially, linguistic production can be prone to errors, states or messages may be percieved incorrectly, and multiple languages may be compatible with the data learners are exposed to.
 These sources of uncertainty introduce variation in their transmission from one generation to the next.
 In keeping the analogy to evolutionary processes this variation can be likened to mutation to the effect that a type's offspring may adopt a different type than that of its parent.
-Importantly, the resulting generational turnovers should depend on the relative learnability of a type.
+Importantly, the resulting generational turnovers should depend on the relative learnability of a type instead of being a constant that is equal to all types as in \citet{nowak+etal:2002}.
 For this purpose, we turn to a different strand of research in cultural evolution: {\em iterated learning}.
@@
 This learner, upon acquisition of this behavior, then goes on to produce behavior that serves as input for a new learner.
-This process can be thought of as a progression through chains of parents and children; the parent produces linguistic data from which the child infers a language (\hl{reference for main representative work of IL}).
-The latter, now a parent, goes on to produce linguistic data for a new generation of na\"ive learners.
+This process can be thought of as a progression through chains of parents and children; the parent produces linguistic data from which the child infers a language.
+The latter, now a parent, goes on to produce linguistic data for a new generation of learners.
 Following \citet{griffiths+kalish:2007} we model iterated learning as a repeated process of Bayesian inference in which learners combine the likelihood of a type producing the received learning input with prior inductive biases.
@@
 Or, in our case, a combination of a lexicon and a signaling behavior.
-
-As touched upon in \S\ref{sec:introduction} we assume learners to be biased towards simpler semantics.
-That is, learners have a preference for types that can explain the data by codifying less in their lexical representations.
-More generally, a drive for simplicity has been argued to underpin speaker preferences for brevity and ease of articulation, as well as to pressure languages towards lexical ambiguity and grammatical compression \citep{zipf:1949,grice:1975,piantadosi+etal:2012b, kirby+etal:2015}.
-As shown by \citet{chater+vitanyi:2003} simplicity not only provides a solution to induction problems by providing a distinction between equally explanatory hypotheses, but has also proven its empirical worth as a longstanding predictive principle in cognitive modelling.
-
-In iterated learning, the extent of the prior's influence has been shown to heavily depend on the learning strategy assumed to underly the inference process.
+The extent of the prior's influence has been shown to heavily depend on the learning strategy assumed to underly the inference process.
 On the one hand, simulation results suggested that weak biases could be magnified by exposing learners to only small data samples (e.g.
@@
 Griffith \& Kalish's convergence to the prior holds for learners that sample from the posterior.
-More deterministic strategies such as the choice for the type with the highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase the influence of both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}.
-In the following, we parametrize the posterior with a parameter $l$, $l \geq 1$.
-In doing so, we obtain a range of learning strategies that live between posterior sampling and MAP.
+More deterministic strategies such as the adoption of the type with the highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase the influence of both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}.
+In the following, we parametrize the posterior with a parameter $l \geq 1$.
+In doing so, we obtain a range of learning strategies between posterior sampling and MAP.
 When $l = 1$ learners sample from the posterior.
@@
 These are sequences of language use witnessed by learners, the length of which we denote by $k$.
-Put differently, each datum $d \in D$ contains $k$ members of the set $\{\tuple{s_i,m_v} | s_i \in S, m_v \in M\}$.
+%Put differently, each datum $d \in D$ contains $k$ members of the set $\{\tuple{s_i,m_v} | s_i \in S, m_v \in M\}$.
 
-To summarize, the parametrized posterior, $P(d|t)^l$, is obtained from combining the prior $P(t)$ and the likelihood $P(d|t)$.
+%To summarize, the parametrized posterior, $P(d|t)^l$, is obtained from combining the prior $P(t)$ and the likelihood $P(d|t)$.
 Since types are combinations of a lexicon and its use, the latter can be computed straightforwardly from a type's production behavior.
-We combine the replicator dynamics with iterated learning by codifying the latter in a transition matrix $Q$, where $Q_{ij}$ indicates the probability that a child of a parent of type $i$ adopts type $j$.
+We integrated iterated learning in the replicator dynamics by codifying the former in a transition matrix $Q$, where $Q_{ij}$ indicates the probability that a child of a parent of type $i$ adopts type $j$.
 This quantity is proportional to the probability of $i$ producing the learning data and that of inferring $j$ given the data:
@@
 \[
- Q_{ij} \propto \sum_{d \in D} P(d|t_i) F(t_j,d),
+ Q_{ij} \propto \sum_{d \in D} P(d|t_i) F(t_j,d), \; \text{where } F(t_j,d) \propto P(t_j|d)^l \text{ and } P(t_j|d) \propto P(t_j) P(d|t_j).
 \]
-where $F(t_j,d) \propto P(t_j|d)^l$ and $P(t_j|d) \propto P(t_j) P(d|t_j)$.
 
 \subsection{Summary}
-Drawing from past research, we argued that expressivity, learnability and simplicity are central to the cultural evolution of language.
-We propose these components to be modelled respectively as communicative efficiency-relative replication, iterated Bayesian learning, and a prior that biases learners for compressed lexical meanings.
+Drawing from past research, we argued that expressivity and learnability are central to the cultural evolution of language.
+We propose these components to be modelled respectively as communicative efficiency-relative replication and (iterated) Bayesian learning.
 Taken together their interaction is described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:
@@
 We call this combination a type.
-A type's expressivity depends on its communicative efficiency within a population while its learnability depends on the fidelity by which it is inferred by new generations of na\"ive learners.
+A type's expressivity depends on its communicative efficiency within a population while its learnability depends on the fidelity by which it is inferred by new generations of learners.
 The learners' task is consequently to perform a joint inference over types of linguistic behavior and lexical meaning.
@@
 %
+\hl{We assume an integral part of the answer to be that learners are a priori biased towards simpler, more compressed, lexical representations.
+That is, rational learners should prefer simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
+More narrowly, codifying more semantically is a priori dispreferred and in cases where pragmatics offers conventional means of enrichment, the communicative disadvantage that speakers would otherwise incurr by the use of simpler semantics is leveled.
+In this way, these domains play a synergic role in overcoming linguistic pressures which may prevent the lexicalization of pragmatic inferences.
+Crucially, we also show how a different kind of transmission bias in the form of a noisy perception of the environment can lead to the same outcome as that predicted by a learning bias.}
+
+\hl{The mere presence of systematic noise in the transmission of strategies can introduce regularization that looks as if the agents have a learning bias.
+But, most importantly, this disturbance of transmission fidelity is due to perceptual noise and properties of the environment, not learning biases.
+In other words, learning biases are clearly not the only transmission biases that can shape evolution alongside functional pressure.
+Environmental and perceptual noise can play a role too.}
+
+\hl{As touched upon in \S\ref{sec:introduction} we assume learners to be biased towards simpler semantics.
+That is, learners have a preference for types that can explain the data by codifying less in their lexical representations.
+More generally, a drive for simplicity has been argued to underpin speaker preferences for brevity and ease of articulation, as well as to pressure languages towards lexical ambiguity and grammatical compression \citep{zipf:1949,grice:1975,piantadosi+etal:2012b, kirby+etal:2015}.
+As shown by \citet{chater+vitanyi:2003} simplicity not only provides a solution to induction problems by providing a distinction between equally explanatory hypotheses, but has also proven its empirical worth as a longstanding predictive principle in cognitive modelling.}
+
 Scalar implicatures are a particularly well-studied type of conventional pragmatic inference.
@@
 %\end{table}
-
+\subsection{Learning bias}\label{subsec:bias}
 According to our hypothesis, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $L_5$-like semantics.
@@
 
+\subsection{Noise}\label{subsec:noise}
+We have two states: $s_\exists$ (some but not all) and $s_\forall$ (all). Let's assume that the
+probabilities for observing the state in the column when the actual state is the state in the
+row are given by this table:
+\begin{center}
+  \begin{tabular}{ccc}
+    & $s_\exists$ & $s_\forall$ \\ \midrule 
+    $s_\exists$ & $1-\epsilon$ & $\epsilon$ \\
+    $s_\forall$ & $\delta$ & $1-\delta$
+  \end{tabular}
+\end{center}
+So, $\epsilon$ is the error probability when the true state is $s_\exists$ and $\delta$ is the
+error probability when the true state is $s_\forall$. How does this affect our model of
+Bayesian learning?
+
+Let's denote the probability that the teacher (learner) observes state $s_t$ ($s_l$) when the
+actual state is $s_a$ as $P_N(s_t \mid s_a)$ ($P_N(s_l \mid s_a)$ ). This is given by the table
+above. We can then derive the following probabilities. Firstly, the probability that $s_a$ is
+the actual state when the learner observes $s_l$ is just Bayes rule, combining prior of $s_a$
+with the noise from above:
+\begin{align*}
+  P_N(s_a \mid s_l) \propto P(s_a) \ P_N(s_l \mid s_a)\,.
+\end{align*}
+Secondly, the probability that the teacher observes $s_t$ when the learner observes $s_l$ is:
+\begin{align*}
+  P_N(s_t \mid s_l) = \sum_{s_a} P(s_a \mid s_l) \ P_N(s_t \mid s_a)\,.
+\end{align*}
+Finally, this gives us the probability that a teacher of type $t$ produces a datum that is
+perceived by the listener as $d = \tuple{s_l, m}$:
+\begin{align*}
+  P_N(\tuple{s_l, m} \mid t) = \sum_{s_t} P_N(s_t \mid s_l) \ P(m \mid s_t, t)\,.
+\end{align*}
+Generalize this to a sequence of perceived data $d_l$ and write $P_N(d_l \mid t)$. We then
+define the noise-perturbed mutation matrix as:
+\begin{align*}
+  Q_{ij}  \propto \sum_{d_l \in D} P(d_l \mid t_i) F(t_j,d_l) \,, \ \  \text{where $F(t_j,d)$
+    as before.}
+\end{align*}
+In words, it may be the case that learner and/or teacher do not perceive the actual state as
+what it is. They are not aware of this, and produce/learn as if what they observed was the
+actual state. In particular, the learner does not reason about noise when he tries to infer the
+speaker's type. He takes what he observes a state to be as the actual state that the teacher
+has seen as well and computes which type would have most likely generated the message to this
+state. This can lead to biases of inferring the ``wrong'' teacher type, if the noise makes some
+types err in a way that resembles the noiseless behavior of other types. I.e., this could, in
+principle, induce transmission biases that look as is there was a cognitive bias in favor of a
+particular type, simply because that type better explains the noise.
+
+On top of changing the mutation matrix in this way, we also need to adapt the calculation of
+expected utilities, taking into consideration that states are perceived noisily. So, where
+before we had:
+\begin{align*}
+  U_S(t_i,t_j) = \sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')\,,
+\end{align*}
+we now have:
+\begin{align*}
+  U_S(t_i,t_j) = \sum_{s_a}  P(s_a) \sum_{s_t} P_N(s_t \mid s_a) \sum_m S_n(m|s_t;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')\,.
+\end{align*}
+
+\subsection*{Results}
+
+In a nutshel: this works for some parameter settings but not for others. I have not fully
+explored the whole space. I would be content with a paper that says: there are parameter values
+that produce behavior as if there are cognitive biases without cognitive biases. I believe that
+this would be an additional and very interesting contribution, even if there are noise
+structures that have other effects.
+
+Importantly, we don't need costs ($c = 0$)! Unfortunately, we do
+need to tinker with $\alpha$ (for defensible, techical reasons) and we need to assume that
+$\epsilon > \delta$, otherwise either $t_{10}$ or even $t_{12}$ may dominate.
+
+That means that if we wanted to also argue that this is a plausible explanation for
+lower-bounds semantics, we'd need to argue why it is natural that $\epsilon > \delta$. Maybe
+such a story can be given, but I presently do not see why/how any such story is obviously
+better than any other story I could come up with for any other relation of $\epsilon$ and
+$\delta$.
+
+One reasoning for $\epsilon > \delta$ could be this. When the true state is $s_a$, we would
+actually need negative evidence that some item/person/whatever does not have a
+property. Negative evidence of this kind might be relatively infrequent, especially if
+untrue. On the other hand, perceiving $s_a$ when the true state is $s_e$ could be more
+frequent, because it would, for example, require wrong information about the domain size. It
+could also be a cause of overemphasizing: speakers tend to exaggerate and want to claim
+\emph{all}, when in fact that is not true \citep[e.g.][for a model that has the speakers'
+tendency to overemphasize as a motor of language change]{Schaden2012:Modelling-the-A}.
+
 \subsection{Discussion}

### 2016-11-01 3rmd-singlescalar_withNoise.py

@@ -23,8 +23,8 @@ cost = 0 # cost for LOT-concept with upper bound
 lam = 20 # soft-max parameter
 k = 3  # length of observation sequences
 sample_amount = 50 #amount of k-length samples for each production type
-epsilon  = 0.1 # probability of perceiving S-all, when true state is S-sbna
-delta = 0.5 # probability of perceiving S-sbna, when true state is S-all
+epsilon  = 0.6 # probability of perceiving S-all, when true state is S-sbna
+delta = 0.1 # probability of perceiving S-sbna, when true state is S-all
 learning_parameter = 10 #prob-matching = 1, increments approach MAP
 
 gens = 50 #number of generations per simulation run

### 2016-11-01 player.py

@@ -17,7 +17,7 @@ class LiteralPlayer:
         m = np.zeros(np.shape(l))
         for i in range(np.shape(l)[0]):
             for j in range(np.shape(l)[1]):
-                m[i,j] = np.exp(l[i,j]**self.alpha)
+                m[i,j] = np.exp(np.power(l[i,j],self.alpha))
         return normalize(m)
 
     def receiver_selection_matrix(self):

### 2016-11-02 paper/paper-v2.tex
draft v2 minor changes

@@
 
-\subsection{Replication \& expressivity}\label{sec:expressivity}
+\subsection{Expressivity}\label{sec:expressivity}
 Expressivity has received particular attention from investigations using evolutionary game theory (e.g.
@@
 
-\subsection{Mutation \& learning}\label{sec:learnability}
+\subsection{Learnability}\label{sec:learnability}
 Iterated learning is a process in which the behavior of one individual serves as learning input for another.
@@
 %
-\hl{We assume an integral part of the answer to be that learners are a priori biased towards simpler, more compressed, lexical representations.
-That is, rational learners should prefer simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
-More narrowly, codifying more semantically is a priori dispreferred and in cases where pragmatics offers conventional means of enrichment, the communicative disadvantage that speakers would otherwise incurr by the use of simpler semantics is leveled.
-In this way, these domains play a synergic role in overcoming linguistic pressures which may prevent the lexicalization of pragmatic inferences.
-Crucially, we also show how a different kind of transmission bias in the form of a noisy perception of the environment can lead to the same outcome as that predicted by a learning bias.}
-
-\hl{The mere presence of systematic noise in the transmission of strategies can introduce regularization that looks as if the agents have a learning bias.
-But, most importantly, this disturbance of transmission fidelity is due to perceptual noise and properties of the environment, not learning biases.
-In other words, learning biases are clearly not the only transmission biases that can shape evolution alongside functional pressure.
-Environmental and perceptual noise can play a role too.}
-
-\hl{As touched upon in \S\ref{sec:introduction} we assume learners to be biased towards simpler semantics.
-That is, learners have a preference for types that can explain the data by codifying less in their lexical representations.
-More generally, a drive for simplicity has been argued to underpin speaker preferences for brevity and ease of articulation, as well as to pressure languages towards lexical ambiguity and grammatical compression \citep{zipf:1949,grice:1975,piantadosi+etal:2012b, kirby+etal:2015}.
-As shown by \citet{chater+vitanyi:2003} simplicity not only provides a solution to induction problems by providing a distinction between equally explanatory hypotheses, but has also proven its empirical worth as a longstanding predictive principle in cognitive modelling.}
-
-Scalar implicatures are a particularly well-studied type of conventional pragmatic inference.
+Scalar implicatures are a particularly well-studied type of conventional pragmatic inferences.
 They are licensed for groups of expressions ordered in terms of informativity, here understood as an entailment induced order.
@@
 If it were true that `All students came to class', it would also be true that `Some students came to class'.
-However, as already noted in \S\ref{sec:introduction}, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what their use is normally taken to convey.
+However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what their use is normally taken to convey.
 Instead, the use of a less informative expression when a more informative one could have been used can license a defeasible inference that stronger alternatives do not hold (cf.
 \citealt{horn:1972,gazdar:1979}).
-That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that stronger alternatives do not hold since the speaker did not choose them.
+That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that stronger alternatives do not hold because the speaker used a weaker alternative instead.
 In this way, `Some students came to class' is strengthened to convey that some but not all students came to class.
 Conversely, speakers can rely on their interlocutors to draw this inference.
-The bound that rules out stronger alternatives is thusly not codified in the lexical meaning of weak alternatives but instead pragmatically supplied.
+A bound that rules out stronger alternatives is thusly not codified in the lexical meaning of weak alternatives but instead pragmatically supplied.
 
-This kind of strengthening corresponds to our previous description of the pragmatic use of lexicon $L_a$.
-A pragmatic hearer who reasons about a speaker's use of message $m_1$ will associate it more strongly with $s_1$ than with $s_2$ since the latter is already unambiguously associated with $s_2$.
+This kind of strengthening is captured by the signaling behavior of pragmatic types introduced in \S\ref{sec:languages+use}: A pragmatic hearer who reasons about a speaker's use of message involving {\em some} will associate it more strongly with a state in which {\em some but not all} holds than with one in which the stronger alternative {\em all} holds, as the {\em all} already unambiguously conveys the latter state.
 Conversely, a pragmatic speaker will reason about her interlocutor's expected interpretation and use the messages at her disposition accordingly.
 
-\begin{centering}
-$L_a$ = \bordermatrix{~ & m_1 & m_2 \cr 
-                  s_1 & 1 & 0 \cr
-                  s_2 & 1 & 1 \cr} \hspace{2cm} $L_b$ = \bordermatrix{~ & m_1 & m_2 \cr 
-                  s_1 & 1 & 0 \cr
-                  s_2 & 0 & 1 \cr}\\[0.5cm]
-\end{centering}
-
 Our initial question can now be narrowed to the case of scalar implicatures by asking for a justification for the lack of lexical upper-bounds in weak scalar alternatives.
-That is, why semantics such as those of message $m_1$ in $L_a$ are regularly selected for over the alternative of codifying the bound semantically as in $L_b$.
+That is, why they are regularly selected for over other alternatives such as that of codifying the bound semantically.
 More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} and {\em big} were truth-conditionally incompatible with stronger alternatives such as {\em hot}, {\em and}, {\em all} and {\em huge}?
@@
 This can happen when embedded in downward-entailing contexts, when the speaker is likely uncertain about whether the upper bounded reading is true, or when the distinction between an upper-bounded reading and the simple, only lower-bounded reading, is not relevant.
-For instance, if for all the speaker knows `Some students came' but she doesn't know whether all came, then the use of not upper-bounded {\em some} succinctly conveys her uncertainty.
+For instance, if for all the speaker knows `Some students came' but she doesn't know whether all came, then the use of {\em some} lacking an upper-bound succinctly conveys her uncertainty.
 This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then this is preferred over expressing it overtly through longer expressions.
 For example, by saying {\em some but not all} explicitly.
-Importantly, although morphosyntactic disambiguation is dispreferred due to its relative length and complexity \citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and override contextual cues that might otherwise mislead the hearer.
+Importantly, although morphosyntactic disambiguation may be dispreferred due to its relative length and complexity \citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and override contextual cues that might otherwise mislead the hearer.
 In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so.
@@
 
-Instead, considering the regularity and spread of scalar implicatures together with the observation that monomorphemic expressions that lexically rule out stronger alternatives are unattested across languages (\citealt[252-267]{horn:1984},\citealt{horn:1972,traugott:2004,vdAuwera:2010}) suggests that other forces may be at play.
-In what follows we investigate the hypothesis that the lack of lexicalization of scalar inferences can be accounted for by the relative representational simplicity of lexical meanings lacking an upper-bound over those that explicitly codify it.
-While we do not want to argue that functional aspects as the ones discussed above do not play a role, we do see a clear benefit in exploring whether matters of learnability would not give us additional explanatory leverage.
+Instead, the systematicity and typological spread of scalar implicatures together with the observation that monomorphemic expressions that lexically rule out stronger alternatives are unattested across languages (\citealt[252-267]{horn:1984}, \citealt{horn:1972,traugott:2004,vdAuwera:2010}) suggests that other forces may be at play.
+In what follows we investigate the hypothesis that the lack of lexicalization of scalar inferences may be accounted for by the relative representational simplicity of lexical meanings lacking an upper-bound over those that explicitly codify it.
+This difference is reflected in a learning bias towards more compressed lexical representations.
+That is, rational learners should prefer simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
+More narrowly, codifying more semantically is a priori dispreferred and in cases where pragmatics offers conventional means of enrichment, the communicative disadvantage that speakers would otherwise incurr by the use of simpler semantics is leveled.
+In this way, these domains play a synergic role in overcoming linguistic pressures which may prevent the lexicalization of pragmatic inferences.
 
+While we do not want to argue that functional aspects as the ones discussed above do not play a role, we do see a clear benefit in exploring whether matters of transmission biases would not give us additional explanatory leverage.
 Note however that we do not represent the contrast between lexical representations explicitly.
@@
 We think that our assumption is well-warranted as a working hypothesis and decide against such an enrichment at present in order to focus on the effects of linguistic pressures predicted by our model instead.
+In line with our aim to analyze the effects of transmission perturbations more broadly, we also show how noisy perception of the environment can lead to the same outcome as that predicted by a learning bias.
 
@@
 The space of lexica we consider is given in Table \ref{tab:lexica}.
-As in the preceding examples, lexica are $(2,2)$-Boolean matrices.
-These are the simplest matrices that allow us to make the contrast between the presence or abscence of an upper-bound and the use of scalar implicatures precise.
-As implicit in the discussion so far, one may think of $s_1$ as a ``some but not all''-state and of $s_2$ as an ``all''-state.
-The literal meaning of weak scalar expressions such as English {\em some} then corresponds to a message true of both $s_1$ and $s_2$ in these fragments.
-Following our assumption for a preference for simple lexical representations, the prior biases learners against lexica in which a message holds true only of state $s_1$.
-That is, those messages that lexicalize an upper-bound that rules out the ``all''-state $s_2$.
-All other semantics are assumed to be a priori equally probable.
-Accordingly, this prior is captured by $P(t_i) \propto n - c \cdot r$, where $n$ is the total number of states and $r$ is the number of messages only true of $s_1$ in $t_i$'s lexicon, $c \in [0,1]$.
-Increments in the prior parameter $c$ therefore bring about a stronger preference against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.
+These $(2,2)$-Boolean matrices are the simplest ones that allow us to make the contrast between the presence or abscence of an upper-bound and the use of scalar implicatures precise.
+One may think of the state corresponding to the first row of a matrix as a ``some but not all''-state and the second as an ``all''-state.
+The literal meaning of weak scalar expressions such as English {\em some} then corresponds to a message true of both rows and rows in these fragments.
+While there are $16$ possible $(2,2)$-lexica, a number of them are identical both in terms of expressivity and the learning bias.
+The competition between such types is determined by the initial configuration of a population -- the proportion with which each type starts out.
+However, this fact can be obscured when averaging across simulations.
+Simulations conducted with the full space of possible lexica confirm that the results reported here do not hinge on their exclusion.
+We therefore focus on this smaller representative subset.
+Lexica $L_1$ to $L_3$ are not optimal for communication because they assign the same state to all their messages.
+This failure to associate a single form to a meaning both semantically and pragmatically inevitably leads to their disadvantage in language use.
+$L_4$ and $L_5$ are our target lexica.
+They codify upper-bounded semantics for the message corresponding to the first matrix's column and a lack thereof.
+Lastly, $L_6$ is similar to $L_5$ in that two messages are true of the same state but differs from it in assigning upper-bounded semantics to the first column's message.
 
@@
 
-While there is a total of $16$ possible $(2,2)$-lexica, a number of them are identical both in terms of expressivity and the learning bias.
-The competition between such types is determined by the initial configuration of a population, i.e., the proportion with which each type starts out.
-However, this fact can be obscured when averaging across simulations.
-We therefore focus on this smaller representative subset.\footnote{Simulations conducted with the full space of possible lexica confirm that the results reported here do not hinge on their exclusion.} Lexica $L_1$ to $L_3$ are not optimal for communication because they assign the same state to all their messages.
-This failure to associate a single form to a meaning both semantically and pragmatically inevitably leads to their disadvantage in language use.
-$L_4$ and $L_5$ are our target lexica, previously respectively labeled as lexica $L_b$ and $L_a$.
-They codify upper-bounded semantics for message $m_1$ and a lack thereof.
-Lastly, $L_6$ is similar to $L_5$ in that two messages are true of the same state but differs from it in assigning upper-bounded semantics to $m_1$.
-
 Combining a signaling behavior with each of these $6$ lexica yields a total of $12$ distinct types.
@@
 Almost, because there may be slight differences between the probability with which speakers would (erroneously) use a semantically false description and the probability with which speakers would (erroneously) use a pragmatically suboptimal description.
-Due to this possibly marginal difference between pragmatic $L_4$ and $L_5$, the selection of one type over the other is expected to mainly depend on the learning bias.
-Things are less clear for literal $L_5$ contrasted with literal/pragmatic $L_4$.
-The former has a learning advantage but is expected to fare worse in communicative  terms in virtue of ambiguous $m_1$.
+Due to this possibly marginal difference between pragmatic $L_4$ and $L_5$, the selection of one type over the other is expected to mainly depend on the transmission bias.
+Things are less clear for literal $L_5$ contrasted with literal/pragmatic $L_4$ as the former has a learning advantage but is expected to fare worse in communicative terms.
 
@@
 \subsection{Learning bias}\label{subsec:bias}
+Following our assumption of a preference for simple lexical representations, the prior biases learners against lexica in which a message holds true only of the first row.
+That is, those messages that lexicalize an upper-bound that rules out the ``all''-state.
+All other semantics are assumed to be a priori equally probable.
+Accordingly, this prior is captured by $P(t_i) \propto n - c \cdot r$, where $n$ is the total number of states and $r$ is the number of messages only true of row $1$ in $t_i$'s lexicon, $c \in [0,1]$.
+Increments in the prior parameter $c$ therefore bring about a stronger bias against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.
+
 According to our hypothesis, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $L_5$-like semantics.
@@
 
-\paragraph{Expressivity only.} Recall that the replicator dynamics are sensitive to $\lambda$ and $\alpha$ as both have a bearing on a type's fitness.
-In particular, low $\alpha$ disadvantages types that rely on pragmatic reasoning to the gain of those that codify more semantically.
-The rationality parameter $\lambda$ has a similar effect for different reasons: Low rationality leads to a less pronounced preference for the choice(s) expected to succeed best in communication.
-That is, $\lambda$ regulates the strength by which users of $L_5$ associate non-upper-bounded $m_1$ exclusively with the ``some''-state $s_1$ over the ``all''-state $s_2$.
-Speakers of $L_4$ need not rely on $\lambda$ for this as the association of $m_1$ with $s_1$ is already part of their language's semantics.
-
+\paragraph{Expressivity only.} The replicator dynamics are sensitive to $\lambda$ and $\alpha$ as both have a bearing on a type's fitness.
 The influence of the rationality parameter for $\alpha = 1$ is depicted in Figure \ref{fig:either-R-or-M}.A.
@@
 As the rationality parameter increases, the functional difference between $L_4$ and $L_5$ is leveled.
-Overall, the final populations that result only from a pressure towards expressively come close to an even share of pragmatic $L_4, L_5$ and $L_6$ types (the latter follows the same trajectory as $L_5$ in Figure \ref{fig:either-R-or-M}.A).
-That is, expressivity alone can not differentiate between these lexica in populations of rational pragmatic language users.
+Overall, the outcome from only a pressure towards expressivity come close to an even share of pragmatic $L_4, L_5$ and $L_6$ types (the latter follows the same trajectory as $L_5$ in Figure \ref{fig:either-R-or-M}.A).
+This illustrates the problem we started out with: expressivity alone can not differentiate between (near-)functional equivalents to a degree that justifies the systematic prevalence of $L_5$-like semantics.
+
+%In particular, low $\alpha$ disadvantages types that rely on pragmatic reasoning to the gain of those that codify more semantically.
+The rationality parameter $\lambda$ has a similar effect for different reasons: Low rationality leads to a less pronounced preference for the choice(s) expected to succeed best in communication.
+That is, $\lambda$ regulates the strength by which users of $L_5$ associate non-upper-bounded $m_1$ exclusively with the ``some''-state $s_1$ over the ``all''-state $s_2$.
+Speakers of $L_4$ need not rely on $\lambda$ for this as the association of $m_1$ with $s_1$ is already part of their language's semantics.
 
@@
 
-\paragraph{Learnability only.} The effect of iterated learning without a pressure for expressivity is shown in Figure \ref{fig:either-R-or-M}.B for posterior sampling.
+\paragraph{Learnability only.} The effect of iterated learning with posterior sampling but without a pressure for expressivity is shown in Figure \ref{fig:either-R-or-M}.B.
 In line with our expectations, the share of $L_4$ speakers decreases as the bias against upper-bounds increases.
@@
 However, note that even a strong bias against lexical upper-bounds leads only to a moderate advantage for $L_5$ over $L_4$.
-Furthermore, a pressure only towards learnability promotes functionally defective languages $L_3$.
+Crucially, a pressure only towards learnability can promote functionally defective languages such as $L_3$.
 
@@
 Apart from polymorphy, both pressures make undesirable predictions.
-A pressure only towards expressivity leads to the to the ejection of $L_1$, $L_2$ and $L_3$ from the population.
-However, it can not explain the regular selection of $L_5$-like semantics over either of its functionally similar alternatives $L_4$ and $L_6$.
+A pressure only towards expressivity leads to the to the ejection of $L_1$, $L_2$ and $L_3$ from the population but it can not explain the regular selection of $L_5$-like semantics over either of its functionally similar alternatives.
 In contrast, a pressure only towards learnability has a modest but clear effect in differentiating $L_5$ from these alternatives but fails to rule out functionally suboptimal types such as tautological $L_3$.
@@
 
-\paragraph{Expressivity and learnability.} Figure \ref{fig:cost} illustrates the effect of the learning bias for posterior sampling (\ref{fig:cost}.A) and slightly more MAP-like learning (\ref{fig:cost}.B).
-More detailed results for all types across a sample of $c$-values for $l = 1$ and $l = 3$ are presented in Table \ref{tab:numeric-results}.
-Overall, these results suggest that in the present setup a weak bias is sufficient to lead to a selection of $L_5$ over $L_4$.
-As in the simulations that only considered learnability, this effect increases with the bias' strength provided $L_5$ users are pragmatic.
+\paragraph{Expressivity and learnability.} Figure \ref{fig:cost} illustrates the effect of the learning bias for posterior sampling (\ref{fig:cost}.A) and slightly more MAP-like learning (\ref{fig:cost}.B) when pressured for both expressivity and learnability.\footnote{\hl{Robert suggests to illustrate the effect with a higher l-value.
+E.g., l = 1 and l = 10}} More detailed results for all types across a sample of $c$-values for $l = 1$ and $l = 3$ are presented in Table \ref{tab:numeric-results}.
+These results show that a weak bias is sufficient to lead to a selection of $L_5$ over $L_4$ in the present setup.
+As in the outcomes that only considered learnability, this effect increases with the bias' strength provided $L_5$ users are pragmatic.
 Importantly, the addition of a pressure towards expressivity magnifies this effect and dampens the proliferation of functionally suboptimal types advantaged by the learning bias.
-As stressed above, this suggests that neither the learning bias nor functional pressure alone but their combination may lead to the lack of upper-bounds in the lexical meaning of scalar expressions.
+As stressed above, this indicates that neither a learning bias nor functional pressure alone but their combination may lead to the lack of upper-bounds in the lexical meaning of scalar expressions.
 
@@
   c        &  0  & .1  & .5  & .8  & .9 & ~ & 0 & .1 & .5 & .8 & .9\\ \hline \hline
-lit. $L_1$ & .03 & .03 & .04 & .04 & .04& ~ & $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\ 
-lit. $L_2$ & .03 & .03 & .02 & .01 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
-lit. $L_3$ & .03 & .03 & .04 & .04 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
-lit. $L_4$ & .07 & .07 & .06 & .06 & .05& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
-lit. $L_5$ & .04 & .05 & .05 & .06 & .06& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
-lit. $L_6$ & .04 & .04 & .04 & .04 & .03& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\ \hline
-prg. $L_1$ & .03 & .03 & .04 & .04 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ \\
-prg. $L_2$ & .03 & .03 & .02 & .01 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ \\
-prg. $L_3$ & .03 & .03 & .04 & .04 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ \\ 
+lit. $L_1$ & .03 & .03 & .04 & .04 & .04& ~ & 0 &0 & 0 & 0 & 0\\ 
+lit. $L_2$ & .03 & .03 & .02 & .01 & .04& ~ & 0& 0 & 0 & 0 & 0\\
+lit. $L_3$ & .03 & .03 & .04 & .04 & .04& ~ & 0& 0 & 0 & 0 & 0\\
+lit. $L_4$ & .07 & .07 & .06 & .06 & .05& ~ & 0& 0 & 0 & 0 & 0\\
+lit. $L_5$ & .04 & .05 & .05 & .06 & .06& ~ & 0& 0 & 0 & 0 & 0\\
+lit. $L_6$ & .04 & .04 & .04 & .04 & .03& ~ & 0& 0 & 0 & 0 & 0\\ \hline
+prg. $L_1$ & .03 & .03 & .04 & .04 & .04& ~ & 0& 0 & 0 & 0 & 0 \\
+prg. $L_2$ & .03 & .03 & .02 & .01 & .04& ~ & 0& 0 & 0 & 0 & 0 \\
+prg. $L_3$ & .03 & .03 & .04 & .04 & .04& ~ & 0& 0 & 0 & 0 & 0 \\ 
 prg. $L_4$ & .22 & .22 & .2 & .18 & .17& ~ & .33& .31 & .23 & .15 & .12 \\
@@
 \end{tabular}
-\caption{Mean proportions of types in $1000$ populations after $20$ generations across bias values $c \in [0,1]$ with $l =1$ and $l = 3$ ($\alpha = 1, \lambda = 30, k = 5$), $\epsilon < 0.005$.}
+\caption{Mean proportions of types in $1000$ populations after $20$ generations across bias values $c \in [0,1]$ with $l =1$ and $l = 3$ ($\alpha = 1, \lambda = 30, k = 5$)}%, $\epsilon < 0.005$.}
 \label{tab:numeric-results}
@@
 First, the degree to which linguistic behavior is deterministic, controlled by $\lambda$ and $\alpha$, as it plays a role both for expressivity as well as in producing data that allows learners to discriminate this type from others.
-Second,  the learning bias $c$ which controls the learners preference for simpler lexical representations -- leading to the selection of $L_5$ over $L_4$.
+Second,  the learning bias $c$ which controls the learners preference for simpler lexical representations.
 Lastly, the posterior parameter $l$, which magnifies the effects of the learning bias in tandem with replication.
@@
 This is not so for more posterior maximizing behavior.
-Crucially, as shown in Figure \ref{fig:prior-posterior} the range of bias values within which $L_5$ takes over the population increases with MAP-like learning.
+As shown in Figure \ref{fig:prior-posterior}, the range of bias values within which $L_5$ takes over the population increases with MAP-like learning.
 In other words, the strength of the learning bias required for a given final proportion of $L_5$ speakers strongly depends on learners' inference mechanism.
-As for the effect of the other parameters not mentioned so far, changes in sequence length influence the population in a predictable way: smaller values lead to more heterogeneous populations that reflect the learner's prior more faithfully whereas larger ones lead to more pronounced differences amongst equally preferred types.
-This is expected insofar as the likelihood that a sequence of length $1$ was produced by any type is relatively uniform (modulo prior) whereas the likelihood of types with lexica $L_1$ - $L_3$ to produce, for instance, a sequence of $10$ observations consistently with the same state-message combination is less likely than for pragmatic types using $L_4$ - $L_6$, or literal $L_4$.
+As for the effect of the other parameters not mentioned so far, changes in sequence length influence the population in a predictable way: smaller values lead to more heterogeneous populations that reflect the learner's prior more faithfully.
+Larger ones lead to more pronounced differences amongst equally preferred types.
+This is expected insofar as the likelihood that a small sequence was produced by any type is relatively uniform (modulo prior) whereas the likelihood of types with lexica $L_1$ - $L_3$ to produce larger sequences with the same state-message combination is less likely than for pragmatic types using $L_4$ - $L_6$, or literal $L_4$.
 Thus, while noteworthy, sequence length has no direct bearing on the contrast of interest.
@@
 
-\subsection{Noise}\label{subsec:noise}
-We have two states: $s_\exists$ (some but not all) and $s_\forall$ (all). Let's assume that the
-probabilities for observing the state in the column when the actual state is the state in the
-row are given by this table:
-\begin{center}
-  \begin{tabular}{ccc}
-    & $s_\exists$ & $s_\forall$ \\ \midrule 
-    $s_\exists$ & $1-\epsilon$ & $\epsilon$ \\
-    $s_\forall$ & $\delta$ & $1-\delta$
-  \end{tabular}
-\end{center}
-So, $\epsilon$ is the error probability when the true state is $s_\exists$ and $\delta$ is the
-error probability when the true state is $s_\forall$. How does this affect our model of
-Bayesian learning?
+\subsection{Noisy transmission}\label{subsec:noise}
+The assumption of a learning bias was pivotal in introducing transmission perturbations that differentiate similar types and homogenize the population.
+However, such perturbations need not necessarily arise from cognitive factors.
+Instead, as shown in the following, similar outcomes can arise from environmental actors, such as perceptual errors during production and comprehension.
+We should stress that our goal is not to suggest noisy perception to underlie the selection of a lack of semantic upper-bounds in scalar implicatures, but to elucidate the role of transmission perturbations in this process and the cultural evolution of meaning more broadly.
 
-Let's denote the probability that the teacher (learner) observes state $s_t$ ($s_l$) when the
-actual state is $s_a$ as $P_N(s_t \mid s_a)$ ($P_N(s_l \mid s_a)$ ). This is given by the table
-above. We can then derive the following probabilities. Firstly, the probability that $s_a$ is
-the actual state when the learner observes $s_l$ is just Bayes rule, combining prior of $s_a$
-with the noise from above:
+The core components of the model remain as before.
+All that is required is an adjustment to the possibility of confusing one state of affairs with another.
+Letting $\epsilon$ stand for the probability of confusing the first row state with the second, and vice-versa for $\delta$, we denote the probability that the teacher (learner) observes state $s_t$ ($s_l$) when the actual state is $s_a$ as $P_N(s_t \mid s_a)$ ($P_N(s_l \mid s_a)$).
+The probability that $s_a$ is the actual state when the learner observes $s_l$ is just Bayes rule, combining prior of $s_a$ with the noise from above:
 \begin{align*}
@@
 \end{align*}
-In words, it may be the case that learner and/or teacher do not perceive the actual state as
-what it is. They are not aware of this, and produce/learn as if what they observed was the
-actual state. In particular, the learner does not reason about noise when he tries to infer the
-speaker's type. He takes what he observes a state to be as the actual state that the teacher
-has seen as well and computes which type would have most likely generated the message to this
-state. This can lead to biases of inferring the ``wrong'' teacher type, if the noise makes some
-types err in a way that resembles the noiseless behavior of other types. I.e., this could, in
-principle, induce transmission biases that look as is there was a cognitive bias in favor of a
-particular type, simply because that type better explains the noise.
+In words, it may be the case that learner and/or teacher do not perceive the actual state as what it is.
+They are not aware of this, and produce/learn as if what they observed was the actual state.
+In particular, the learner does not reason about noise when he tries to infer the speaker's type.
+He takes what he observes a state to be as the actual state that the teacher has seen as well and infers which type would have most likely generated the message to this state.
+This can lead to biases of inferring the ``wrong'' teacher type, if the noise makes some types err in a way that resembles the noiseless behavior of other types.
+That is, such environmental factors can, in principle, induce transmission biases that look as is there was a cognitive bias in favor of a particular type, simply because that type better explains the noise.
 
-On top of changing the mutation matrix in this way, we also need to adapt the calculation of
-expected utilities, taking into consideration that states are perceived noisily. So, where
-before we had:
-\begin{align*}
-  U_S(t_i,t_j) = \sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')\,,
-\end{align*}
-we now have:
+Apart from changing the mutation matrix in this way, we also need to adapt the calculation of expected utilities, taking into consideration that states are perceived noisily:
 \begin{align*}
@@
 
-\subsection*{Results}
+\paragraph{Results.} As shown in Table \ref{tab:num-noise}, a prevalence of pragmatic $L_5$ can also arise from noisy transmission without a prior  learning bias for particular types.
+In contrast to the outcome of the non-noisy model, favourable outcomes for this type concentrate in noise patterns with $\epsilon > \delta$, $\alpha \geq 5$, $1 < k < 20$ and $l > 3$.
+Crucially, these results show that the mere presence of systematic noise in the transmission of strategies can introduce regularization that looks as if the agents have a learning bias.
+But, most importantly, this disturbance of transmission fidelity is due to perceptual noise and properties of the environment, not learning biases.
+In other words, learning biases are clearly not the only transmission biases that can shape evolution alongside functional pressure.
+Environmental and perceptual noise can play a role too.
 
-In a nutshel: this works for some parameter settings but not for others. I have not fully
-explored the whole space. I would be content with a paper that says: there are parameter values
-that produce behavior as if there are cognitive biases without cognitive biases. I believe that
-this would be an additional and very interesting contribution, even if there are noise
-structures that have other effects.
-
-Importantly, we don't need costs ($c = 0$)! Unfortunately, we do
-need to tinker with $\alpha$ (for defensible, techical reasons) and we need to assume that
-$\epsilon > \delta$, otherwise either $t_{10}$ or even $t_{12}$ may dominate.
-
-That means that if we wanted to also argue that this is a plausible explanation for
-lower-bounds semantics, we'd need to argue why it is natural that $\epsilon > \delta$. Maybe
-such a story can be given, but I presently do not see why/how any such story is obviously
-better than any other story I could come up with for any other relation of $\epsilon$ and
-$\delta$.
-
-One reasoning for $\epsilon > \delta$ could be this. When the true state is $s_a$, we would
-actually need negative evidence that some item/person/whatever does not have a
-property. Negative evidence of this kind might be relatively infrequent, especially if
-untrue. On the other hand, perceiving $s_a$ when the true state is $s_e$ could be more
-frequent, because it would, for example, require wrong information about the domain size. It
-could also be a cause of overemphasizing: speakers tend to exaggerate and want to claim
-\emph{all}, when in fact that is not true \citep[e.g.][for a model that has the speakers'
-tendency to overemphasize as a motor of language change]{Schaden2012:Modelling-the-A}.
+\begin{table}
+\centering 
+\begin{tabular}{l | c c c c c| c c c c c c c c}
+\multicolumn{1}{c}{~} & \multicolumn{5}{c}{$l = $} & ~ & \multicolumn{5}{c}{$l = 15$}\\ \hline \hline
+ ($\epsilon,\delta$)        &  (.1,.1)  & (.1,.3)  & (.3,.1)  & (.8,.1)  & (.1,.8) & ~ & (.1,.1) & (.1,.3) & (.3,.1) & (.8,.1) & (.1,.8)\\ \hline \hline
+lit. $L_1$ & .01 & .02 & .03 & .03 & .03& ~ & 0& .02&  .02 & .01 & .01\\ 
+lit. $L_2$ & .01 & .02 & .03 & .03 & .03& ~ & 0& .02&  .02 & .01 & .01\\
+lit. $L_3$ & .01 & .02 & .03 & .01 & .03& ~ & 0& .02 & .02 & .01 & .01\\
+lit. $L_4$ & .03 & .03 & .03 & .05 & .01& ~ & .04& .03 & .03 & .01 & .01\\
+lit. $L_5$ & .02 & .04 & .03 & .02 & .02& ~ & .01& .03 & .02 & .02 & .01\\
+lit. $L_6$ & .02 & .03 & .04 & .03 & .05& ~ & .01& .03 & .04 & .01 & .02\\ \hline
+prg. $L_1$ & .01 & .02 & .03 & .03 & .03& ~ & 0&  .02& .02 & .01 &  .01\\
+prg. $L_2$ & .01 & .02 & .03 & .03 & .03& ~ & 0&  .02& .02& .01&  .01\\
+prg. $L_3$ & .01 & .02 & .03 & .03 & .03& ~ & 0&  .02& .02 & .01 & .01 \\ 
+prg. $L_4$ & .45 & .11 & .11 & .02 & .02& ~ & .44& .11 & .1 & .22 & .02 \\
+prg. $L_5$ & .22 & .17 & .47 & .04 & .68& ~ & .24& .18 & .56 & .02 & .84\\
+prg. $L_6$ & .22 & .48 & .17 & .69 & .04& ~ & .24& .52 & .16 & .84 & .02
+\end{tabular}
+\caption{\hl{Currently: Mean proportions of types from $50$ runs per $100$ independently generated Q-matrices per parameter setting after $20$ generations across noisy values for $\epsilon$ and $\delta$ with $l = 5$ and $l = 10$ ($\alpha = 10, \lambda = 20, k = 5$).
+In principle, we could skip every second row because pragmatic $L5$ and $L6$ mirror each other.}}
+\label{tab:num-noise}
+\end{table}
 
 \subsection{Discussion}
-Broadly speaking these results suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome communicative pressures and stabilize in a population provided there is a bias for simpler representations.
+Under the assumption of a learning bias for simpler representations, our results suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome communicative pressures and stabilize in a population.
 This prediction hinges on three assumptions.
@@
 Second, that language use is not too unpredictable -- low $\alpha$ or $\lambda$ render languages that rely on pragmatics too prone to communicative failure and more difficult to learn.
-Third, that learners prefer simpler over more complex lexical representations.
+Lastly, that learners prefer simpler over more complex lexical representations.
 An important addendum to this third condition being that a combination of rationality in choice and maximization in learning requires a weaker bias towards simplicity.
@@
 We conject that frequently drawn scalar implicatures might fossilize to avoid cost, while infrequent ones could still be computed on-line.
-This opens a possible venue to address our previous question about the expected presence of $L_4$-like semantics.
+This opens a possible venue to address the question about the expected presence of $L_4$-like semantics.
+
+More broadly, we have shown that environmental factors such as noisy perception by language learners and users can lead to outcomes indistinguishible from those generated by cognitive biases.
+We hope this finding to invite further investigation on causes and interactions of such transmission perturbations.
+\hl{elaborate}
 
 \section{General discussion}
-We laid out a model that combines game-theoretical models of functional pressure towards efficient communication \citep{nowak+krakauer:1999}, effects of learning biases on (iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012, franke+jaeger:2014} as well as different lexica \citep{bergen+etal:2012,bergen+etal:2016}.
+We laid out a model that combines game-theoretical models of functional pressure towards efficient communication \citep{nowak+krakauer:1999}, effects of transmission biases on (iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012, franke+jaeger:2014} as well as different lexica \citep{bergen+etal:2012,bergen+etal:2016}.
 This model generates predictions about lexicalization patterns and, more generally, effects of communicative pressures on language.
-We argued that the puzzle raised by semantics in light of pragmatics is hard to explain on purely functional grounds and that part of the answer may instead lie in learnability: Simpler semantic representations are more likely to be learned while pragmatic reasoning can counteract functional disadvantages otherwise incurred.
+We argued that the puzzle raised by semantics in light of pragmatics is hard to explain on purely functional grounds and that part of the answer may instead lie in learnability.
+In particular, in perturbations in the transmission of language.
+In the realm of learning biases, simpler semantic representations are more likely to be learned while pragmatic reasoning can counteract functional disadvantages otherwise incurred.
 This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics by offering an account of why (certain) pragmatic inferences fail to lexicalize.
+More broadly, we showed that systematic noise in perception can produce outcomes that are indistinguishible from those generated by cognitive biases.
 
-The main innovations of the model are its modular separation of expressivity and learnability, allowing for their isolated and combined analysis, as well as the learning process involving a joint inference over types of pragmatic behavior and lexical meaning.
+The main innovations of the model are its modular separation of expressivity and learnability, allowing for their isolated and combined analysis, the learning process involving a joint inference over types of pragmatic behavior and lexical meaning, as well as in its accommodation of a range of transmission perturbations that go beyond learning biases.
 The goal to decouple but model both expressivity and learnability has also recently been addressed by \citet{kirby+etal:2015}.
-In contrast to our proposal, however, Kirby et al.
+In contrast to our proposal, Kirby et al.
 model expressivity as exerting its force only in the production of learning data.
-This model's expressivity parameter thereby fulfills a similar role  to high values of $\lambda$ and $\alpha$ in making speaker behavior more deterministic.
+This model's expressivity parameter thereby fulfills a similar role  to high values of $\lambda$ in making speaker behavior more deterministic.
 In this way, it ``favors'' unambiguous languages.
@@
 Our reasoning behind the inclusion of the former lies in the empirical and theoretical observation that learnability alone can lead the selection of functionally defective languages, as showcased by the tautological language $L_3$ in our analysis.
-This outcome has been reported in a number laboratory experiments where the participants' task was to learn and subsequently reproduce the language produced by a previous participant, leading to a proliferation of languages that associated a large number of meanings with a single form (see e.g.
+This outcome has been reported in a number of laboratory experiments where the participants' task was to learn and subsequently reproduce the language produced by a previous participant, leading to a proliferation of languages that associated a large number of meanings with a single form (see e.g.
 \citealt{silvey+etal:2014} and experiment 1 in \citealt{kirby+etal:2008}).
@@
 \citealt{fay+etal:2013}; for a review of laboratory results under the iterated learning paradigm and further discussion see \citealt{kirby+etal:2015, tamariz+kirby:2016}).
-It is not evident how to compare these empirical predictions given their variation in meaning spaces, modes of transmission, iterations and feedback given to participants.
+It is not evident how to compare these findings given that they consider distinct meaning spaces, modes of transmission, iterations and feedback given to participants.
 However, we take these results to suggest that there is an important difference between a language generating learnable linguistic data and its actual performance as a means of information transfer.
 The former solely depends on the mechanism by which speakers associate form and meaning.
-The latter additionally hinges on the hearers linguistic experience and her capability to interpret linguistic input based on this experience.
-In sum, we contend that successful information transfer in a linguistic community is central the adoption of a communication system and that this measure is not faithfully reflected by taking only production into consideration.
+The latter additionally hinges on the addressee's linguistic experience and her ability to interpret linguistic input based on this experience.
+In sum, we contend that successful information transfer in a linguistic community is central to the adoption of a communication system and that this measure is not adequately reflected by production alone.
 
-One of our central  assumptions is the learning bias for simpler lexical representations.
+\hl{One of our central  assumptions is the learning bias for simpler lexical representations.
 Albeit implicit in our formalization, the conceptual foundation for a compression preference at the lexical level is grounded in work in the language of thought tradition put forward by \citet{piantadosi+etal:2012a,piantadosi+etal:underreview}.
@@
 Whether language use is culturally transmitted is an empirical matter that can not be settled here.
-However, this idealized model suggests that semantic structure evinced cross-linguistically that lends itself for pragmatic exploitation reflects lexical information that is both easy to learn and, perhaps more interestingly, also easy to learn how to exploit via mutual reasoning.
+However, this idealized model suggests that semantic structure evinced cross-linguistically that lends itself for pragmatic exploitation reflects lexical information that is both easy to learn and, perhaps more interestingly, also easy to learn how to exploit via mutual reasoning.}
+\hl{Shift this a bit, and involve the noisy transmission}
 
 \section{Conclusion}
+\hl{add noise}
 Language change is affected by intertwined pressures.
@@
   \item I still need to review, read and find some references I'm withholding in the current draft (marked as ``\hl{references}'' in color). 
-  \item The discussion (sub) sections should be expanded and polished.
-In particular the discussion concerning lexical representations and the language of thought.
-I'm aware that this part is not as readable as the rest;
+  \item The data used in plots and slides should be generated as uniformly as possible.
+I will do this once we settle for everything else.
+Any input on this is welcome, but I think it boils down to adjusting the non-noisy simulations a bit to the noisy ones (the number of samples is what mostly diverges).
+It's also up to debate whether we should run the non-noisy simulations with multiple Qs, as their noisy counterparts.
   \item The analysis may benefit from further data and less of a qualitative overview of the parameters' effect (e.g. the regression analysis Michael conducted);
-  \item As you will have noticed when reading footnote $3$, I think the performance of $L_6$ being on par with $L_4$ is somewhat problematic;
-  \item There's a fair share of repetition, this is partly due to the fact that scalar implicatures and some matrices are discussed in non-contiguous parts of the paper.
-Whether the repetition is necessary depends on whether we roughly keep the current structure and length or change things;
   \item In principle, we could make a direct comparison between the case-study of \citet{griffiths+kalish:2007} on compositionality using iterated learning and iterated learning + functional pressure.

### 2016-11-04 paper/paper-v2.tex
minor changes in draft

@@
 
-\title{Communicative pressures at the semantics-pragmatics interface}
+\title{Tracing the cultural evolution of meaning at the semantics-pragmatics interface}
 
@@
 Our model traces two evolutionary forces and their interaction: (i) fitness-based pressure towards communicative efficiency and (ii) systematic transmission perturbations when linguistic knowledge is transferred from one agent to another.
-The latter can contain general cognitive or learning biases, but can also contain other sources of systematic noise, e.g., resulting from errors in perception.
+The latter can arise from general cognitive or learning biases, but also from other sources of systematic noise, e.g., as a result of errors in perception.
 We illustrate the model based on a case study showing that cognitive biases that favor simple semantic representations can prevent the lexicalization of pragmatic inferences.
@@
 
-%Mutual reasoning endows speakers with the ability to convey information that goes beyond the literal meaning of expressions.
-A particularity of certain classes of lexical meanings is that they enable for such pragmatic enrichments in a notably productive fashion.
-This raises the challenge to justify their regular selection over other alternatives that, e.g., do not require such enrichments.
-We suggest that such an account can be provided by an analysis of the linguistic pressures that apply at the semantics-pragmatics interface.
-To this end, we lay out a general model that combines functional pressure on language use and a pressure for learnability in acquisition in populations of probabilistic language users with varied degrees of pragmatic sophistication.
-We then apply this model in a case study on the (lack of) lexicalization of scalar implicatures.
-The results suggest that such inferences fail to lexicalize when languages are pressured towards simpler semantic representations in learning, provided that pragmatic reasoning can compensate for the disadvantage in expressivity that users of such languages otherwise incur.
-This result sheds light on the lack of lexicalization of scalar inferences, as well as the semantics-pragmatics distinction more generally.
-
 \section{Introduction}\label{sec:introduction}
 What is conveyed usually goes beyond what is said.
-A request for a blanket can be politely veiled by uttering ``I'm cold'', a temporal succession of events can be conveyed by the order in which conjuncts appear as in ``I traveled to Paris and got married'', or an invitation can be declined by saying ``I have to work''.
-An influential explanation of the relation between the literal meaning of expressions (semantics) and what they are intended and interpreted to convey (pragmatics) due to \citet{grice:1975} views the latter as a product of a process of mutual reasoning about rational language use.
-For instance, under the assumption that the speaker is cooperative and relevant, ``I have to work'' may be interpreted as providing a reason why the speaker will not be able to accept the invitation -- why would she have said so otherwise?
+A request for a blanket can be politely veiled by uttering ``I'm cold'', the temporal succession of events can be conveyed by the order in which conjuncts appear as in ``I traveled to Paris and got married'', or an invitation be declined by saying ``I have to work''.
+An influential explanation of the relation between the literal meaning of expressions and what they are intended and interpreted to convey due to \citet{grice:1975} views the latter as a product of a process of mutual reasoning about rational language use.
+For instance, under the assumption that the speaker is cooperative and relevant, ``I have to work'' may be interpreted as providing a reason why the speaker will not be able to accept the invitation, going beyond its literal meaning.
 Crucially, many classes of lexical meanings allow for pragmatic enrichments in a notably systematic fashion.
 
-Productive pragmatic inferences such as scalar and manner implicatures have been at the focus of many studies on the semantics-pragmatics interface.
+Productive pragmatic inferences, such as scalar and manner implicatures, have been at the focus of many studies on the semantics-pragmatics interface.
 However, an issue that has received little attention is the justification of semantic structure in light of the systematic enrichments offered by pragmatics.
-Similarly, multiple models of the cultural evolution of language have been put forward to analyze properties such as compositionality and combinatoriality, but the selection of particular semantics in light of pragmatics remains uncharted territory.
-The present investigation seeks to fill these gaps by analyzing the effects linguistic pressures have on the selection and pervasiveness of particular lexical meanings under consideration of their systematic pragmatic modification.
-Accordingly, our main goal is to uncover how transmission biases and functional pressure interact in the cultural evolution of meaning when pragmatics is taken into account.
-Crucially, we argue that transmission biases need not be cognitive in nature but can also ensue from perception and the environment.
+Similarly, differnt models of the cultural evolution of language have been put forward in order to analyze properties such as compositionality and combinatoriality (e.g.
+\citealt{kirby+etal:2015,verhoef+etal:2014}), but the selection of meaning in light of pragmatics remains uncharted territory.
+The present investigation seeks to fill these gaps by analyzing the effects linguistic pressures have on the selection and pervasiveness of particular lexical meanings under consideration of their pragmatic modification.
+Accordingly, our primary goal is to uncover how transmission perturbations and functional pressure interact in the cultural evolution of meaning.
+Additionally, we show how systematic disturbances in the fidelity by which linguistic knowledge is transmitted that stem from environmental factors can lead to outcomes akin to those predicted under the assumption of inductive learning biases.
+That is, that the regularization standardly predicted to follow from learning biases can also emerge as a byproduct of noise.
 
-\subsection{Linguistic pressures at the semantics-pragmatics interface}
+\subsection{Communicative pressures at the semantics-pragmatics interface}
 The emergence and change of linguistic structure is influenced by many intertwined factors.
 These range from biological and socio-ecological to cultural ones \citep{steels:2011}.
-Social and ecological pressures determine communicative needs while biology determines the architecture that enables and constrains the means by which they can be fulfilled.
-In the following, our focus lies on cultural aspects, wherein processes of linguistic change are viewed as shaped by language use and its transmission.
-That is, as a result of a process of cultural evolution.
+Social and ecological pressures determine communicative needs, while biology determines the architecture that enables and constrains the means by which they can be fulfilled.
+In the following, our focus lies on the cultural aspects, wherein processes of linguistic change are viewed as shaped by language use and its transmission, i.e., as a result of a process of cultural evolution.
 
@@
 in \citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007, piantadosi:2014,kirby+etal:2015}).
-In recent years this line of research has led to a surge of approaches that seek to analyze the effects of such pressures from multiple angles ranging from simulations to experiments with human and robotic agents (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
+In recent years, this line of research has led to a surge of approaches that seek to analyze the effects of such pressures from a multidisciplinary perspective, ranging from simulations to experiments with human and robotic agents (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
 Our starting point is given by the overarching argument that has crystalized from this literature: Natural languages need to be well-adapted to communicative needs within a linguistic community, but also need to be learnable to survive their faithful transmission across generations.
@@
 A language with a single form-meaning association is easy to learn but lacking in expressivity.
-Conversely, a language that associates a distinct form with all possible meanings one may want to convey is maximally expressive but challenging to acquire.
+Conversely, a language that associates a distinct form with all possible meanings a speaker may want to convey is maximally expressive but challenging to acquire.
 The most prominent problem that arises from this tension is that of acquiring a language to express a potentially infinite set of meanings through finite means \citep{kirby:2002}.
@@
 More central to our explanandum is the issue of selecting particular hypotheses out of a potentially infinite space of alternatives compatible with the data learners are exposed to.
-At the semantics-pragmatics interface this concerns the selection between functionally (near-)equivalent lexical meanings, noting in particular that what is systematically conveyed through pragmatics could alternatively be codified lexically.
+At the semantics-pragmatics interface this concerns the selection between functional (near-)equivalents, noting in particular that what is systematically conveyed through pragmatics could alternatively be codified lexically.
 Notwithstanding, lexical meanings that allow for pragmatic modifications often seem to resist the lexicalization of their pragmatic component.
@@
 
-We assume an integral part of the answer to lie in the effects of transmission perturbations on language when passed from one generation to the next.
-Such perturbations may take the form of learning biases, but may also stem from extraneous factors such as a noisy perception of the environment.
-In particular, we show how the presence of systematic noise can introduce regularizations that resemble those obtained under the assumption of a learning bias.
+We assume an integral part of the answer to lie in the effects of transmission perturbations that arise when linguistic knowledge is acquired by new learners.
+As shown in the following, such perturbations may take the form of learning biases, but may also stem from extraneous factors such as a noisy perception of the environment.
 
-In the following, we model these components using the replicator-mutator dynamics, combining functional pressure on successful communication with effects of transmission perturbations on (iterated) Bayesian learning \citep{griffiths+kalish:2007}.
-The semantics-pragmatics distinction and its bearing on production and comprehension are captured by a probabilistic model of rational language use with different degrees of pragmatic sophisitication and languages \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
-The remainder of this section introduces these components individually together with the assumptions underlying them.
-These are: the representation of languages and their use (\S\ref{sec:languages+use}), pressures towards expressivity (\S\ref{sec:expressivity}) and learnability (\S\ref{sec:learnability}) regulated by the replicator and mutator dynamics, respectively.
-After laying out the model, we analyze its predictions in a case-study on the lack of lexicalization of scalar implicatures in \S\ref{sec:si-case-study} and compare the outcome predicted by different transmission biases in \S\ref{subsec:bias} and \S\ref{subsec:noise}.
+We model these components using the replicator-mutator dynamics, combining functional pressure on successful communication with effects of transmission perturbations on (iterated) Bayesian learning \citep{griffiths+kalish:2007}.
+The semantics-pragmatics distinction and its bearing on production and comprehension are captured by a probabilistic model of rational language use with different degrees of pragmatic sophisitication and different lexica \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
+The remainder of this section introduces these components together with the assumptions underlying them.
+These are: the representation of languages and their use (\S\ref{sec:languages+use}), as well as pressures towards expressivity (\S\ref{sec:expressivity}) and learnability (\S\ref{sec:learnability}).
+After laying out the model, we analyze its predictions in a case-study on the lack of lexicalization of scalar implicatures in \S\ref{sec:si-case-study}.
+Additionally, \S\ref{subsec:noise} shows how the outcome predicted under the assumption of an inductive bias for simplicity in \S\ref{subsec:bias} can also arise from as a byproduct of the noisy perception of language users and learners.
+Section \S\ref{sec:discussion} discusses the general predictions of the model, as well as the implications drawn from its noisy counterpart.
 
@@
 Lexica codify the truth-conditions of a language's expressions.
-Following \citet{franke+jaeger:2014}, a convenient way to represent lexica is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs (meanings) to convey and $M$ a set of messages (forms available in the language).
+Following \citet{franke+jaeger:2014}, a convenient way to represent lexica is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs (meanings) and $M$ a set of messages (forms available in the language).
 
@@
 {\em Literal interlocutors} produce and interpret messages literally, being guided only by their lexica.
-In contrast, {\em pragmatic interlocutors} engage in mutual reasoning to inform their choices.
+{\em Pragmatic interlocutors} instead engage in mutual reasoning to inform their choices.
 Following models of rational language use such as Rational Speech Act models \citep{frank+goodman:2012} and their game-theoretic predecessors, the Iterated Best/Quantal Response models \citep{franke:2009,franke+jaeger:2014}, these behaviors are captured by a reasoning hierarchy.
@@
 Pragmatic language users of level $n + 1$ behave rationally according to expected level $n$ behavior of their interlocutors.
-Below, (\ref{h:level0}) and (\ref{h:leveln}) specify the behavior of literal and pragmatic hearers of a language $L$.
+(\ref{h:level0}) and (\ref{h:leveln}) specify the behavior of literal and pragmatic hearers of a language $L$.
 Their speaker counterparts are given in (\ref{s:level0}) and (\ref{s:leveln}).
@@
 &H_{0}(s|m;L) \propto pr(s) L_{sm} \label{h:level0}\\
-&S_{0}(m|s;L) \propto \exp(L_{sm}^\alpha) \label{s:level0}\\
+&S_{0}(m|s;L) \propto \exp((L_{sm})^\alpha) \label{s:level0}\\
 &H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
@@
 According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state $s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$.
-In the following this prior is assumed to be uniform across hearers.
+For simplicity, in the following this prior is assumed to be uniform across hearers.
 The behavior of literal speakers, given in (\ref{s:level0}), is regulated by a parameter $\alpha$ which controls the sensitivity to which speakers prefer one signal over another based on its expected communicative success.
@@
 
-We call the combination of a lexicon with its use, i.e., a level in the reasoning hierarchy, a {\em type}. These are the basic units on which our model's dynamics operate.
+We call the combination of a lexicon with its use, i.e., a level in the reasoning hierarchy, a {\em type}. These are the basic units on which the model's dynamics operate.
 
@@
 As noted above, the fitness of type $i$ is equal to its relative communicative success within this population, $f_i = \sum_j x_j \text{EU}(t_i,t_j)$.
-The expected communicative success of $i$ and $j$ is obtained by considering the average success of $i$ conveying information to $j$ and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$.
+The expected communicative success of $i$ and $j$ is simply the average success of $i$ conveying information to $j$ and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$.
 $U_S(x,y)$ and $U_R(x,y)$ are, respectively, $\sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$ and $U_S(y,x)$, for $n$ and $o$ being the reasoning level of $x$ and $y$, and $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.
 This quantity is symmetric, reflecting the probability of two types' mutual understanding.
-The average fitness of the population is given by $\Phi$, $\Phi = \sum_i x_i f_i$.
-This term serves as a normalizing constant for the discrete replicator equation: $\dot{x}_i = \frac{x_i f_i}{\Phi}$
+The average fitness of the population is given by $\Phi$, $\Phi = \sum_i x_i f_i$, serving as a normalizing constant for the discrete replicator equation: $\dot{x}_i = \frac{x_i f_i}{\Phi}$
 
-Under its biological interpretation, this equation captures the idea of fitness-relative selection whereby fitter types produce more offspring, leading to their propagation in subsequent generations.
-As noted above, many aspects of language are subject to processees of transmission and change that can be likened to these biological dynamics.
+Under its biological interpretation, the replicator dynamics capture the idea of fitness-relative selection whereby fitter types produce more offspring, leading to their propagation in subsequent generations.
+Many aspects of language are subject to transmission and change that can be likened to such biological processes.
 Amongst others, replication can be construed as modelling language acquisition, as e.g.
@@
 
-In their series of papers on language evolution, Nowak and colleagues did not only consider expressivity but also recognized the central role of the fidelity by which languages are transmitted.
-Crucially, linguistic production can be prone to errors, states or messages may be percieved incorrectly, and multiple languages may be compatible with the data learners are exposed to.
-These sources of uncertainty introduce variation in their transmission from one generation to the next.
-In keeping the analogy to evolutionary processes this variation can be likened to mutation to the effect that a type's offspring may adopt a different type than that of its parent.
-Importantly, the resulting generational turnovers should depend on the relative learnability of a type instead of being a constant that is equal to all types as in \citet{nowak+etal:2002}.
-For this purpose, we turn to a different strand of research in cultural evolution: {\em iterated learning}.
+In their series of papers on language evolution, Nowak and colleagues did not only consider expressivity but also recognized the central role of the fidelity by which language is transmitted.
+Among others, linguistic production can be prone to errors, states or messages may be percieved incorrectly, and multiple languages may be compatible with the data learners are exposed to.
+These sources of uncertainty introduce variation in their transmission from one generation to the next which can be likened to mutation to the effect that a type's offspring may adopt a different type than that of its parent as a result of transmission perturbations.
+Importantly, this variation should be relative to the learnability of a type under these perturbations (instead of, e.g., being a constant that is equal across all types as in \citealt{nowak+etal:2002}).
+For this purpose, we turn to a different strand of research in cultural evolution: iterated learning.
 
@@
 This process can be thought of as a progression through chains of parents and children; the parent produces linguistic data from which the child infers a language.
-The latter, now a parent, goes on to produce linguistic data for a new generation of learners.
+The latter, now a parent, goes on to produce data for a new generation of learners.
 Following \citet{griffiths+kalish:2007} we model iterated learning as a repeated process of Bayesian inference in which learners combine the likelihood of a type producing the received learning input with prior inductive biases.
@@
 Due to the pressure towards learnability it exherts, iterated learning generally leads to simpler and more regular languages (see \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016} for recent surveys).
-Importantly, experimental and mathematical results suggest that the outcome of this process reflects learners' a priori biases.
-In a Bayesian setting these biases can be codified in a prior $P \in \Delta(T)$ and a way to think about it is as the amount of data a learner would require in order to adopt a language (cf.
+Importantly, experimental and mathematical results have been taken to suggest that the outcome of this process reflects learners' inductive biases.
+In a Bayesian setting these biases can be codified in a prior $P \in \Delta(T)$, which can be thought of as the amount of data a learner would require in order to adopt a language (cf.
 \citealt[450]{griffiths+kalish:2007}).
-Or, in our case, a combination of a lexicon and a signaling behavior.
+Or, in our case, a combination of a lexicon and its use.
 The extent of the prior's influence has been shown to heavily depend on the learning strategy assumed to underly the inference process.
-On the one hand, simulation results suggested that weak biases could be magnified by exposing learners to only small data samples (e.g.
+On the one hand, early simulation results suggested that weak biases could be magnified by exposing learners to only small data samples (e.g.
 in \citealt{brighton:2002}).
@@
 More deterministic strategies such as the adoption of the type with the highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase the influence of both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}.
-In the following, we parametrize the posterior with a parameter $l \geq 1$.
+In the following, we parametrize the posterior by $l \geq 1$.
 In doing so, we obtain a range of learning strategies between posterior sampling and MAP.
-When $l = 1$ learners sample from the posterior.
-As $l$ increases towards infinity their propensity to maximize the posterior grows as well.
+When $l = 1$ learners sample from the posterior and the learners propensity to maximize the posterior grows as $l$ increases.
 
@@
 Since types are combinations of a lexicon and its use, the latter can be computed straightforwardly from a type's production behavior.
-We integrated iterated learning in the replicator dynamics by codifying the former in a transition matrix $Q$, where $Q_{ij}$ indicates the probability that a child of a parent of type $i$ adopts type $j$.
+We combine the above with the replicator dynamics by codifying iterated learning in a transition matrix $Q$, where $Q_{ij}$ indicates the probability that a child of a parent of type $i$ adopts type $j$.
 This quantity is proportional to the probability of $i$ producing the learning data and that of inferring $j$ given the data:
@@
 
-\subsection{Summary}
+\subsection{Model summary}
 Drawing from past research, we argued that expressivity and learnability are central to the cultural evolution of language.
@@
 In this way, `Some students came to class' is strengthened to convey that some but not all students came to class.
-Conversely, speakers can rely on their interlocutors to draw this inference.
 A bound that rules out stronger alternatives is thusly not codified in the lexical meaning of weak alternatives but instead pragmatically supplied.
 
-This kind of strengthening is captured by the signaling behavior of pragmatic types introduced in \S\ref{sec:languages+use}: A pragmatic hearer who reasons about a speaker's use of message involving {\em some} will associate it more strongly with a state in which {\em some but not all} holds than with one in which the stronger alternative {\em all} holds, as the {\em all} already unambiguously conveys the latter state.
+This kind of strengthening is captured by the linguistic behavior of pragmatic types introduced in \S\ref{sec:languages+use}: A pragmatic hearer who reasons about a speaker's use of a message involving a weak scalar alternative will associate it more strongly with upper-bounded states than with ones in which stronger alternatives hold because these alternatives already unambiguously convey the latter states.
 Conversely, a pragmatic speaker will reason about her interlocutor's expected interpretation and use the messages at her disposition accordingly.
@@
 For instance, if for all the speaker knows `Some students came' but she doesn't know whether all came, then the use of {\em some} lacking an upper-bound succinctly conveys her uncertainty.
-This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then this is preferred over expressing it overtly through longer expressions.
-For example, by saying {\em some but not all} explicitly.
+This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: Should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then this is preferred over expressing it overtly through longer expressions, e.g., by saying {\em some but not all} explicitly.
 Importantly, although morphosyntactic disambiguation may be dispreferred due to its relative length and complexity \citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and override contextual cues that might otherwise mislead the hearer.
@@
 Fourth, larger lexica are costly.
-Overall, neither condition is convincing as a pivotal explanatory device for such a wide-spread phenomenon.
+Overall, neither condition seems convincing as a pivotal explanatory device for such a widespread phenomenon.
 The first two conditions put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of unambiguous communication.
@@
 In what follows we investigate the hypothesis that the lack of lexicalization of scalar inferences may be accounted for by the relative representational simplicity of lexical meanings lacking an upper-bound over those that explicitly codify it.
-This difference is reflected in a learning bias towards more compressed lexical representations.
-That is, rational learners should prefer simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
-More narrowly, codifying more semantically is a priori dispreferred and in cases where pragmatics offers conventional means of enrichment, the communicative disadvantage that speakers would otherwise incurr by the use of simpler semantics is leveled.
-In this way, these domains play a synergic role in overcoming linguistic pressures which may prevent the lexicalization of pragmatic inferences.
+This difference is reflected in a learning bias towards more compressed lexical representation, i.e., in a preference of learners for simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
 
@@
 Note however that we do not represent the contrast between lexical representations explicitly.
-Instead, the learning bias towards a lack of upper-bounds in weak scalar alternatives is directly encoded in the learners' prior over types.
+Instead, the bias is directly encoded in the learners' prior over types.
 In principle this difference could be made precise with an adequate representational language, e.g., through measures over representational complexity such as minimal description length.
@@
 \citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012a} and references therein).
-We think that our assumption is well-warranted as a working hypothesis and decide against such an enrichment at present in order to focus on the effects of linguistic pressures predicted by our model instead.
-In line with our aim to analyze the effects of transmission perturbations more broadly, we also show how noisy perception of the environment can lead to the same outcome as that predicted by a learning bias.
+We think that our assumption is well-warranted as a working hypothesis and decide against such an enrichment at present in order to focus on the effects of linguistic pressures predicted the model instead.
 
 \subsection{Analysis}
-We analyze the model's predictions in populations of types with two signaling behaviors; literal or pragmatic.
+We analyze the model's predictions in populations of types with two signaling behaviors; literal and pragmatic.
 The former correspond to level $0$ reasoners who only take their lexica into consideration and the latter to level $1$ reasoners.
@@
 
-The space of lexica we consider is given in Table \ref{tab:lexica}.
+The space of possible lexica is given in Table \ref{tab:lexica}.
 These $(2,2)$-Boolean matrices are the simplest ones that allow us to make the contrast between the presence or abscence of an upper-bound and the use of scalar implicatures precise.
-One may think of the state corresponding to the first row of a matrix as a ``some but not all''-state and the second as an ``all''-state.
-The literal meaning of weak scalar expressions such as English {\em some} then corresponds to a message true of both rows and rows in these fragments.
+One may think of the state corresponding to the first row of any such lexicon as a ``some but not all''-state and the second as an ``all''-state.
+The literal meaning of weak scalar expressions such as English {\em some} then corresponds to a message true of both rows in these fragments.
 While there are $16$ possible $(2,2)$-lexica, a number of them are identical both in terms of expressivity and the learning bias.
-The competition between such types is determined by the initial configuration of a population -- the proportion with which each type starts out.
-However, this fact can be obscured when averaging across simulations.
-Simulations conducted with the full space of possible lexica confirm that the results reported here do not hinge on their exclusion.
-We therefore focus on this smaller representative subset.
+The competition between such types is determined by the initial configuration of a population.
+However, this can be obscured when averaging across simulations.
+We focus on this smaller representative subset as simulations conducted with the full space confirm that the general results reported here do not hinge on this choice.
+
 Lexica $L_1$ to $L_3$ are not optimal for communication because they assign the same state to all their messages.
-This failure to associate a single form to a meaning both semantically and pragmatically inevitably leads to their disadvantage in language use.
+This failure to be able to associate a single form to a state inevitably leads to a communicative disadvantage in their use.
 $L_4$ and $L_5$ are our target lexica.
-They codify upper-bounded semantics for the message corresponding to the first matrix's column and a lack thereof.
+They codify upper-bounded semantics for the message corresponding to the first matrix's column and a lack thereof, respectively.
 Lastly, $L_6$ is similar to $L_5$ in that two messages are true of the same state but differs from it in assigning upper-bounded semantics to the first column's message.
@@
 
-Combining a signaling behavior with each of these $6$ lexica yields a total of $12$ distinct types.
-Note in particular  that a type that has conventionalized upper bounds to realize a (quasi-)partition of the relevant semantic space, such as $L_4$, will produce speaker behavior that is {\em almost} indistinguishable from that of a language that lacks upper bounds, but with pragmatic speakers, such as $L_5$.
+Combining a linguistic behavior with each of these $6$ lexica yields a total of $12$ distinct types.
+Note in particular  that a type that has conventionalized upper-bounds to realize a (quasi-)partition of the relevant semantic space, such as $L_4$, will produce speaker behavior that is {\em almost} indistinguishable from that of a language that lacks upper-bounds, but with pragmatic users, such as $L_5$.
 Almost, because there may be slight differences between the probability with which speakers would (erroneously) use a semantically false description and the probability with which speakers would (erroneously) use a pragmatically suboptimal description.
-Due to this possibly marginal difference between pragmatic $L_4$ and $L_5$, the selection of one type over the other is expected to mainly depend on the transmission bias.
-Things are less clear for literal $L_5$ contrasted with literal/pragmatic $L_4$ as the former has a learning advantage but is expected to fare worse in communicative terms.
+Due to this possibly marginal difference between pragmatic $L_4$ and $L_5$, the selection of one type over the other is expected to mainly depend on their transmission to new learners.
+Things are less clear for literal $L_5$ contrasted with literal/pragmatic $L_4$ as the former has a learning advantage under the inductive bias but is expected to fare worse in communicative terms.
 
@@
 %\end{table}
-\subsection{Learning bias}\label{subsec:bias}
-Following our assumption of a preference for simple lexical representations, the prior biases learners against lexica in which a message holds true only of the first row.
-That is, those messages that lexicalize an upper-bound that rules out the ``all''-state.
+\subsection{Transmission bias}\label{subsec:bias}
+Following our assumption of a preference for simple lexical representations, the prior biases learners against lexica in which a message holds true only of the first row, i.e., against messages that lexicalize an upper-bound that rules out the ``all''-state.
 All other semantics are assumed to be a priori equally probable.
-Accordingly, this prior is captured by $P(t_i) \propto n - c \cdot r$, where $n$ is the total number of states and $r$ is the number of messages only true of row $1$ in $t_i$'s lexicon, $c \in [0,1]$.
-Increments in the prior parameter $c$ therefore bring about a stronger bias against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.
+This is captured by $P(t_i) \propto n - c \cdot r$, where $n$ is the total number of states and $r$ is the number of messages only true of row $1$ in $t_i$'s lexicon, $c \in [0,1]$.
+Increments in the value of $c$ therefore bring about a stronger bias against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.
 
@@
 \paragraph{Expressivity only.} The replicator dynamics are sensitive to $\lambda$ and $\alpha$ as both have a bearing on a type's fitness.
-The influence of the rationality parameter for $\alpha = 1$ is depicted in Figure \ref{fig:either-R-or-M}.A.
-As expected, the less expressive $L_3$ speakers fare the worse and are influenced the least by variations in $\lambda$.
+The influence of the rationality parameter for is depicted in Figure \ref{fig:either-R-or-M}.A.
+The less expressive $L_3$ speakers fare the worse and are influenced the least by change in $\lambda$.
 In contrast, low values of $\lambda$ result in a higher proportion of $L_4$ speakers relative to $L_5$.
@@
 As the rationality parameter increases, the functional difference between $L_4$ and $L_5$ is leveled.
-Overall, the outcome from only a pressure towards expressivity come close to an even share of pragmatic $L_4, L_5$ and $L_6$ types (the latter follows the same trajectory as $L_5$ in Figure \ref{fig:either-R-or-M}.A).
+Overall, the outcome from only a pressure towards expressivity approximates an even share of pragmatic $L_4, L_5$ and $L_6$ types.
+The latter follows the same trajectory as $L_5$ in Figure \ref{fig:either-R-or-M}.A.
 This illustrates the problem we started out with: expressivity alone can not differentiate between (near-)functional equivalents to a degree that justifies the systematic prevalence of $L_5$-like semantics.
@@
 In line with our expectations, the share of $L_4$ speakers decreases as the bias against upper-bounds increases.
-In turn, this benefits $L_3$ and, in particular, $L_4$.
-However, note that even a strong bias against lexical upper-bounds leads only to a moderate advantage for $L_5$ over $L_4$.
-Crucially, a pressure only towards learnability can promote functionally defective languages such as $L_3$.
+In turn, this benefits $L_3$ and, in particular, $L_5$.
+However, even a strong bias against lexical upper-bounds leads only to a moderate advantage of $L_5$ over $L_4$.
+More importantly, a pressure only towards learnability can promote functionally defective languages such as $L_3$.
 
-Inspecting these pressures separately not only gives some intuitions about the parameters' influence, but also highlights some of their broader implications.
+Inspecting these pressures separately not only showcases the influence of some model components but also highlights some of their broader implications.
 First and foremost, neither dynamic comes close to converging to a monomorphic population under most parameter configurations.
 For instance, while $L_4$ speakers can come to take over a substantial proportion of the population, this only happens in a restricted range of low degrees of rationality.
-Apart from polymorphy, both pressures make undesirable predictions.
-A pressure only towards expressivity leads to the to the ejection of $L_1$, $L_2$ and $L_3$ from the population but it can not explain the regular selection of $L_5$-like semantics over either of its functionally similar alternatives.
-In contrast, a pressure only towards learnability has a modest but clear effect in differentiating $L_5$ from these alternatives but fails to rule out functionally suboptimal types such as tautological $L_3$.
+Apart from polymorphy, both pressures make undesirable predictions when considered in isolation.
+A pressure only towards expressivity leads to the expulsion of communicatively suboptimal $L_1$, $L_2$ and $L_3$ from the population but can not explain the regular selection of $L_5$-like semantics over either of its functionally similar alternatives.
+A pressure only towards learnability has a modest but clear effect in differentiating $L_5$ from these alternatives but fails to rule out functionally suboptimal types such as tautological $L_3$.
 
@@
 E.g., l = 1 and l = 10}} More detailed results for all types across a sample of $c$-values for $l = 1$ and $l = 3$ are presented in Table \ref{tab:numeric-results}.
-These results show that a weak bias is sufficient to lead to a selection of $L_5$ over $L_4$ in the present setup.
+These results show that a weak bias is sufficient to lead to a selection of $L_5$ over $L_4$.
 As in the outcomes that only considered learnability, this effect increases with the bias' strength provided $L_5$ users are pragmatic.
@@
 
-The resulting proportion of pragmatic $L_5$ speakers primarily hinges on four parameters.
-First, the degree to which linguistic behavior is deterministic, controlled by $\lambda$ and $\alpha$, as it plays a role both for expressivity as well as in producing data that allows learners to discriminate this type from others.
-Second,  the learning bias $c$ which controls the learners preference for simpler lexical representations.
-Lastly, the posterior parameter $l$, which magnifies the effects of the learning bias in tandem with replication.
+Other than the consideration of both pressures, the resulting proportion of pragmatic $L_5$ speakers primarily hinges on three factors.
+First, the degree to which linguistic behavior is deterministic, as it plays a role both for expressivity as well as in producing data that allows learners to discriminate this type from others.
+Second, the inductive bias, which controls the learners preference for simpler lexical representations.
+Lastly, the posterior parameter, which magnifies the effects of the learning bias in tandem with replication.
 
@@
 As shown in Figure \ref{fig:prior-posterior}, the range of bias values within which $L_5$ takes over the population increases with MAP-like learning.
-In other words, the strength of the learning bias required for a given final proportion of $L_5$ speakers strongly depends on learners' inference mechanism.
+In other words, the strength of the learning bias required for a given final proportion of $L_5$ speakers strongly depends on learners' inferential strategy.
 As for the effect of the other parameters not mentioned so far, changes in sequence length influence the population in a predictable way: smaller values lead to more heterogeneous populations that reflect the learner's prior more faithfully.
 Larger ones lead to more pronounced differences amongst equally preferred types.
-This is expected insofar as the likelihood that a small sequence was produced by any type is relatively uniform (modulo prior) whereas the likelihood of types with lexica $L_1$ - $L_3$ to produce larger sequences with the same state-message combination is less likely than for pragmatic types using $L_4$ - $L_6$, or literal $L_4$.
-Thus, while noteworthy, sequence length has no direct bearing on the contrast of interest.
+This is due to the fact that the likelihood that a small sequence was produced by any type is relatively uniform (modulo prior) compared to that of types with lexica $L_1$ - $L_3$ to produce larger sequences with the same state-message combination in contrast to pragmatic speakers of $L_4$ - $L_6$, or literal $L_4$.
 
@@
 
+\subsection{Discussion}
+Under the assumption of a learning bias for simpler representations, our results suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome communicative pressures and stabilize in a population.
+This prediction hinges on three assumptions.
+First, that language is pressured toward both expressivity and learnability.
+Second, that language use is not too unpredictable -- low $\alpha$ or $\lambda$ render languages that rely on pragmatics too prone to communicative failure and more difficult to learn.
+Lastly, that learners prefer simpler over more complex lexical representations.
+An important addendum to this third condition being that a combination of rationality in choice and maximization in learning requires a weaker bias towards simplicity.
+Under these conditions the selection of lexical meanings lacking upper-bounds in populations of pragmatic speakers is robust against parameter perturbations.
+This outcome is particularly encouraging in light of other advantages a lack of semantic upper-bounds may confer.
+
+While non upper-bounded lexical meanings in weak scalar alternatives are predicted by the literature, it not clear to what extent other types should be present in the final population, if at all.
+It seems reasonable to expect functionally suboptimal types $L_1$, $L_2$ and $L_3$ to be ruled out because they fail to enable their users to communicate effectively.
+However, this is not true of $L_4$.\footnote{$L_6$ presents a special case.
+In our current setup, it mirrors $L_5$ in enabling for the pragmatic narrowing of a message that does not codify an upper-bound lexically.
+However, this is achieved by ruling out the ``some but not all''-state and not, as with scalar implicatures, the ``all''-state.
+$L_6$ speakers therefore strengthen a ``some''-message to convey something paraphrasable as `some but not [some but not all]'.
+Our representation of lexica as Boolean matrices is blind to this anomaly.} Notwithstanding, the prediction that natural language communities are homogeneous or that a single speaker may entertain $L_4$-like semantics for one scalar expression and $L_5$-like semantics for another is not implausible (cf.
+\citealt{franke+degen:2016}).
+Alternatively, a stronger tendency for posterior maximization has to be assumed (cf.
+Figure \ref{fig:prior-posterior}).
+This empirical issue relates to other two aspects left undiscussed: disadvantages of pragmatic reasoning and the effect of state frequencies on the fossilization of pragmatic inferences.
+We tacitly assumed pragmatic reasoning to come at no cost.
+However, there is experimental evidence that suggests  that the pragmatic derivation of upper-bounds costs effort and takes additional processing time (cf.
+\citealt{deNeys+schaeken:2007, huang+snedeker:2009}).
+This raises the question at which point such usage-based cost undercuts the learnability advantage of simpler semantic representations.
+Should cost play a role, then its effect is bound to depend on the frequency with which a given scalar expression is used.
+We conject that frequently drawn scalar implicatures might fossilize to avoid cost, while infrequent ones could still be computed on-line.
+This opens a possible venue to address the first question about the expected presence of $L_4$-like semantics.
+
+\hl{More narrowly, codifying more semantically is a priori dispreferred and in cases where pragmatics offers conventional means of enrichment, the communicative disadvantage that speakers would otherwise incurr by the use of simpler semantics is leveled.
+In this way, these domains play a synergic role in overcoming linguistic pressures which may prevent the lexicalization of pragmatic inferences.
+}
+
 \subsection{Noisy transmission}\label{subsec:noise}
-The assumption of a learning bias was pivotal in introducing transmission perturbations that differentiate similar types and homogenize the population.
-However, such perturbations need not necessarily arise from cognitive factors.
-Instead, as shown in the following, similar outcomes can arise from environmental actors, such as perceptual errors during production and comprehension.
-We should stress that our goal is not to suggest noisy perception to underlie the selection of a lack of semantic upper-bounds in scalar implicatures, but to elucidate the role of transmission perturbations in this process and the cultural evolution of meaning more broadly.
+The assumption of a learning bias was pivotal in introducing transmission perturbations that differentiate functionally similar types.
+However, such perturbations need not necessarily be a consequence of such biases.
+As shown in the following, indistinguishable outcomes can come about through environmental factors such as perceptual errors during production and comprehension.
+We should stress that we do not want to suggest noisy perception to underlie the selection of a lack of semantic upper-bounds in scalar implicatures.
+Instead, our goal is to elucidate the role of transmission perturbations in this process and the cultural evolution of meaning more broadly, as well as to highlight the explanatory potential of less considered forces.
 
 The core components of the model remain as before.
-All that is required is an adjustment to the possibility of confusing one state of affairs with another.
+All that is required is an adjustment to the possibility of confusing one state with another.
 Letting $\epsilon$ stand for the probability of confusing the first row state with the second, and vice-versa for $\delta$, we denote the probability that the teacher (learner) observes state $s_t$ ($s_l$) when the actual state is $s_a$ as $P_N(s_t \mid s_a)$ ($P_N(s_l \mid s_a)$).
-The probability that $s_a$ is the actual state when the learner observes $s_l$ is just Bayes rule, combining prior of $s_a$ with the noise from above:
+The probability that $s_a$ is the actual state when the learner observes $s_l$ is therefore:
+
 \begin{align*}
@@
 \end{align*}
-Secondly, the probability that the teacher observes $s_t$ when the learner observes $s_l$ is:
+
+Accordingly, the probability that the teacher observes $s_t$ when the learner observes $s_l$ is:
 \begin{align*}
@@
 \begin{align*}
-  P_N(\tuple{s_l, m} \mid t) = \sum_{s_t} P_N(s_t \mid s_l) \ P(m \mid s_t, t)\,.
+  P_N(\tuple{s_l, m} \mid t) = \sum_{s_t} P_N(s_t \mid s_l) \ P(m \mid s_t; t)\,.
 \end{align*}
@@
   Q_{ij}  \propto \sum_{d_l \in D} P(d_l \mid t_i) F(t_j,d_l) \,, \ \  \text{where $F(t_j,d)$
-    as before.}
+    is as before.}
 \end{align*}
@@
 They are not aware of this, and produce/learn as if what they observed was the actual state.
-In particular, the learner does not reason about noise when he tries to infer the speaker's type.
-He takes what he observes a state to be as the actual state that the teacher has seen as well and infers which type would have most likely generated the message to this state.
+In particular, the learner does not reason about noise when she tries to infer the speaker's type.
+She takes what she observes a state to be as the actual state that the teacher has seen as well and infers which type would have most likely generated the message to this state.
 This can lead to biases of inferring the ``wrong'' teacher type, if the noise makes some types err in a way that resembles the noiseless behavior of other types.
-That is, such environmental factors can, in principle, induce transmission biases that look as is there was a cognitive bias in favor of a particular type, simply because that type better explains the noise.
+That is, such environmental factors can, in principle, induce transmission biases that look as if there was a cognitive bias in favor of a particular type, simply because that type better explains the noise.
 
@@
 
-\paragraph{Results.} As shown in Table \ref{tab:num-noise}, a prevalence of pragmatic $L_5$ can also arise from noisy transmission without a prior  learning bias for particular types.
-In contrast to the outcome of the non-noisy model, favourable outcomes for this type concentrate in noise patterns with $\epsilon > \delta$, $\alpha \geq 5$, $1 < k < 20$ and $l > 3$.
+\paragraph{Results.} As shown in Table \ref{tab:num-noise}, a prevalence of pragmatic $L_5$ can also arise from noisy transmission without a learning bias for particular types.
+In contrast to the outcome of the non-noisy model, favourable outcomes for this type concentrate in parameter configurations  with $\epsilon > \delta$, $\alpha \geq 5$, $1 < k < 20$ and $l > 3$.
 Crucially, these results show that the mere presence of systematic noise in the transmission of strategies can introduce regularization that looks as if the agents have a learning bias.
@@
 \begin{tabular}{l | c c c c c| c c c c c c c c}
-\multicolumn{1}{c}{~} & \multicolumn{5}{c}{$l = $} & ~ & \multicolumn{5}{c}{$l = 15$}\\ \hline \hline
+\multicolumn{1}{c}{~} & \multicolumn{5}{c}{$l = 5$} & ~ & \multicolumn{5}{c}{$l = 15$}\\ \hline \hline
  ($\epsilon,\delta$)        &  (.1,.1)  & (.1,.3)  & (.3,.1)  & (.8,.1)  & (.1,.8) & ~ & (.1,.1) & (.1,.3) & (.3,.1) & (.8,.1) & (.1,.8)\\ \hline \hline
@@
 
-\subsection{Discussion}
-Under the assumption of a learning bias for simpler representations, our results suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome communicative pressures and stabilize in a population.
-This prediction hinges on three assumptions.
-First, that language is pressured toward both expressivity and learnability.
-Second, that language use is not too unpredictable -- low $\alpha$ or $\lambda$ render languages that rely on pragmatics too prone to communicative failure and more difficult to learn.
-Lastly, that learners prefer simpler over more complex lexical representations.
-An important addendum to this third condition being that a combination of rationality in choice and maximization in learning requires a weaker bias towards simplicity.
-Under these conditions the selection of lexical meanings lacking upper-bounds in populations of pragmatic speakers is robust against parameter perturbations.
-This outcome is particularly encouraging in light of other advantages a lack of semantic upper-bounds may confer.
-
-While a major share of non upper-bounded lexical meanings is predicted by the literature, it is less clear to what extent other types should be present in the final population, if at all.
-It seems reasonable to expect functionally suboptimal types $L_1$, $L_2$ and $L_3$ to be ruled out because they fail to enable their users to communicate effectively.
-However, this is not true of $L_4$.\footnote{$L_6$ presents a special case.
-In our current setup, it mirrors $L_5$ in enabling for the pragmatic narrowing of message $m_2$ rather than $m_1$.
-However, this association of $s_2$ with $m_2$ under favorable parameter conditions is achieved by ruling out the ``some but not all''-state $s_1$ and not, as with scalar implicatures, the ``all''-state $s_2$.
-$L_6$ speakers therefore strengthen a ``some''-message to convey something paraphrasable as `some but not [some but not all]'.
-Our representation of lexica as Boolean matrices is blind to this anomaly.} Notwithstanding, the prediction that natural language communities are homogeneous or that a single speaker may entertain $L_4$-like semantics for one scalar expression and $L_5$-like semantics for another is is not implausible (\hl{references on individual level differences in the derivation of scalar implicatures}).
-Alternatively, a stronger tendency for posterior maximization has to be assumed (cf.
-Figure \ref{fig:prior-posterior}).
-This empirical issue relates to other two aspects left undiscussed in our present analysis: disadvantages of pragmatic reasoning and the effect of state frequencies on the fossilization of pragmatic inferences.
-We tacitly assumed pragmatic reasoning to come at no cost.
-However, there is experimental evidence that suggests  that the pragmatic derivation of upper-bounds costs effort and takes additional processing time (cf.
-\citealt{deNeys+schaeken:2007, huang+snedeker:2009}).
-This raises the question at which point such usage-based cost undercuts the learnability advantage of simpler semantic representations.
-Should cost play a role, then its effect is bound to depend on the frequency with which a given scalar expression is used.
-We conject that frequently drawn scalar implicatures might fossilize to avoid cost, while infrequent ones could still be computed on-line.
-This opens a possible venue to address the question about the expected presence of $L_4$-like semantics.
-
-More broadly, we have shown that environmental factors such as noisy perception by language learners and users can lead to outcomes indistinguishible from those generated by cognitive biases.
-We hope this finding to invite further investigation on causes and interactions of such transmission perturbations.
-\hl{elaborate}
-
-\section{General discussion}
-We laid out a model that combines game-theoretical models of functional pressure towards efficient communication \citep{nowak+krakauer:1999}, effects of transmission biases on (iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012, franke+jaeger:2014} as well as different lexica \citep{bergen+etal:2012,bergen+etal:2016}.
+\section{General discussion}\label{sec:discussion}
+We laid out a model that combines game theoretical models of functional pressure towards efficient communication \citep{nowak+krakauer:1999}, effects of transmission perturbations on (iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012, franke+jaeger:2014} as well as different lexica \citep{bergen+etal:2012,bergen+etal:2016}.
 This model generates predictions about lexicalization patterns and, more generally, effects of communicative pressures on language.
-We argued that the puzzle raised by semantics in light of pragmatics is hard to explain on purely functional grounds and that part of the answer may instead lie in learnability.
-In particular, in perturbations in the transmission of language.
+We argued that the puzzle raised by semantics in light of pragmatics is hard to explain on purely functional grounds and that part of the answer may instead lie the way transmission shapes the outcome of cultural evolution in tandem with a pressure for successful information transfer.
 In the realm of learning biases, simpler semantic representations are more likely to be learned while pragmatic reasoning can counteract functional disadvantages otherwise incurred.
 This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics by offering an account of why (certain) pragmatic inferences fail to lexicalize.
-More broadly, we showed that systematic noise in perception can produce outcomes that are indistinguishible from those generated by cognitive biases.
+More generally, we showed that systematic noise in perception can produce outcomes that are indistinguishable from those generated by learning biases.
+This adds to \citeposs{franke+correia:toappear} argument that linguistic regularities may arise as a byproduct of noise, rather than through inductive biases.
 
-The main innovations of the model are its modular separation of expressivity and learnability, allowing for their isolated and combined analysis, the learning process involving a joint inference over types of pragmatic behavior and lexical meaning, as well as in its accommodation of a range of transmission perturbations that go beyond learning biases.
+The main innovations of the model are its modular separation of expressivity and learnability, allowing for their isolated and combined analysis, the learning process involving a joint inference over types of pragmatic behavior and lexical meaning, as well as in its accommodation of different transmission perturbations that go beyond learning biases.
 The goal to decouple but model both expressivity and learnability has also recently been addressed by \citet{kirby+etal:2015}.
@@
 In this way, it ``favors'' unambiguous languages.
-Crucially, the degree of mutual understanding of interlocutors central to replication and to our notion of expressivity is not taken into consideration.
+However, the degree of mutual understanding of interlocutors central to replication and to our notion of expressivity is not taken into consideration.
 That is, while our proposal combines bidirectional horizontal transmission with its vertical and unidirectional counterpart, Kirby et al.'s model only considers the latter's influence.
@@
 
-\hl{One of our central  assumptions is the learning bias for simpler lexical representations.
-Albeit implicit in our formalization, the conceptual foundation for a compression preference at the lexical level is grounded in work in the language of thought tradition put forward by \citet{piantadosi+etal:2012a,piantadosi+etal:underreview}.
-Following Piantadosi and colleagues we assumed that lexical meanings are expressed in a representational language and that the learners' task is to infer lexica that best explain the data witnessed.
-Furthermore, we assumed learners to not only infer lexical meaning but types of linguistic use.
-Whether language use is culturally transmitted is an empirical matter that can not be settled here.
-However, this idealized model suggests that semantic structure evinced cross-linguistically that lends itself for pragmatic exploitation reflects lexical information that is both easy to learn and, perhaps more interestingly, also easy to learn how to exploit via mutual reasoning.}
-\hl{Shift this a bit, and involve the noisy transmission}
+The demonstration that noise can lead to regularized evolutionary outcomes that are indistinguishable from those generated by prior learning biases is relevant not for the case study at hand, but more so for the broader project of analyzes of the cultural evolution of language.
+On the one hand, the plurality of sources of transmission perturbations admitted by these models paints a cautionary tale for the design of studies that purport to provide explanatory accounts of linguistic phenomena.
+In particular, when the outcome is interpreted as informative about the perturbation assumed to generates it (cf.
+\citealt{tamariz+kirby:2016}).
+On the other, and most importantly, it showcases how regularities can arise as a byproduct of systematic noise rather than from standardly assumed inductive biases.
 
 \section{Conclusion}
-\hl{add noise}
 Language change is affected by intertwined pressures.
@@
 That is, the relative learnability advantage of simpler semantics may offer an answer to why natural languages fail to lexicalize certain pragmatic inferences.
+We also considered an alternate instantiating of the model which shows that systematic noise in state perception can give rise to evolutionary outcomes that are indistinguishable from those predicted by inductive biases.
 More broadly, we argued that semantic patterns can be explained by taking into consideration the way in which they are used in interaction as well as their representational complexity and learnability.
@@
 \begin{itemize}
-  \item I still need to review, read and find some references I'm withholding in the current draft (marked as ``\hl{references}'' in color). 
   \item The data used in plots and slides should be generated as uniformly as possible.
@@
   \item The analysis may benefit from further data and less of a qualitative overview of the parameters' effect (e.g. the regression analysis Michael conducted);
-  \item In principle, we could make a direct comparison between the case-study of \citet{griffiths+kalish:2007} on compositionality using iterated learning and iterated learning + functional pressure.
-This could be relegated to the appendix.
-However, I'm inclined to think this is somewhat tangential and wouldn't add much.
-The same goes for an explicit comparison of {\em some} vs.
-{\em some but not all} using a language of thought-style representation.
-Let me know if you see worth in either option.
 \end{itemize}

### 2016-11-06 paper/paper-v2.tex
minor changes draft discussion

@@
 First, that language is pressured toward both expressivity and learnability.
-Second, that language use is not too unpredictable -- low $\alpha$ or $\lambda$ render languages that rely on pragmatics too prone to communicative failure and more difficult to learn.
+Second, that language use is relatively deterministic -- low $\lambda$ renders languages that rely on pragmatics to associate one state to one message too prone to communicative failure and more difficult to learn.
 Lastly, that learners prefer simpler over more complex lexical representations.
@@
 However, this is not true of $L_4$.\footnote{$L_6$ presents a special case.
-In our current setup, it mirrors $L_5$ in enabling for the pragmatic narrowing of a message that does not codify an upper-bound lexically.
+In our current setup, it mirrors $L_5$ in enabling for the pragmatic strengthening of a message that does not codify an upper-bound lexically.
 However, this is achieved by ruling out the ``some but not all''-state and not, as with scalar implicatures, the ``all''-state.
 $L_6$ speakers therefore strengthen a ``some''-message to convey something paraphrasable as `some but not [some but not all]'.
-Our representation of lexica as Boolean matrices is blind to this anomaly.} Notwithstanding, the prediction that natural language communities are homogeneous or that a single speaker may entertain $L_4$-like semantics for one scalar expression and $L_5$-like semantics for another is not implausible (cf.
+The current representation of lexica as Boolean matrices is blind to this anomaly without further restrictions.} Notwithstanding, the prediction that natural language communities are homogeneous or that a single speaker may entertain $L_4$-like semantics for one scalar expression and $L_5$-like semantics for another is not implausible (cf.
 \citealt{franke+degen:2016}).
-Alternatively, a stronger tendency for posterior maximization has to be assumed (cf.
-Figure \ref{fig:prior-posterior}).
+Alternatively, a stronger tendency for posterior maximization has to be assumed (see Figure \ref{fig:prior-posterior}).
 This empirical issue relates to other two aspects left undiscussed: disadvantages of pragmatic reasoning and the effect of state frequencies on the fossilization of pragmatic inferences.
@@
 Should cost play a role, then its effect is bound to depend on the frequency with which a given scalar expression is used.
-We conject that frequently drawn scalar implicatures might fossilize to avoid cost, while infrequent ones could still be computed on-line.
-This opens a possible venue to address the first question about the expected presence of $L_4$-like semantics.
-
-\hl{More narrowly, codifying more semantically is a priori dispreferred and in cases where pragmatics offers conventional means of enrichment, the communicative disadvantage that speakers would otherwise incurr by the use of simpler semantics is leveled.
-In this way, these domains play a synergic role in overcoming linguistic pressures which may prevent the lexicalization of pragmatic inferences.
-}
+It is therefore plausible that frequently drawn scalar implicatures might fossilize to avoid cost, while infrequent ones could still be computed on-line.
+This opens a possible venue to address the first question about the expected presence of $L_4$-like semantics, but further empirical evidence is needed to assess these matters beyond speculation.
 
@@
 The assumption of a learning bias was pivotal in introducing transmission perturbations that differentiate functionally similar types.
-However, such perturbations need not necessarily be a consequence of such biases.
-As shown in the following, indistinguishable outcomes can come about through environmental factors such as perceptual errors during production and comprehension.
-We should stress that we do not want to suggest noisy perception to underlie the selection of a lack of semantic upper-bounds in scalar implicatures.
-Instead, our goal is to elucidate the role of transmission perturbations in this process and the cultural evolution of meaning more broadly, as well as to highlight the explanatory potential of less considered forces.
+However, these perturbations need not necessarily arise from inductive biases.
+As shown in the following, outcomes that are indistinguishable from the ones predicted above can come about through environmental factors such as perceptual errors during production and comprehension as well.
+We should stress that we do not want to suggest noisy perception to underlie the selection of a lack of semantic upper-bounds in scalar items.
+Instead, our goal is to stress the role that transmission perturbations play in this process and the cultural evolution of meaning more broadly, as well as to highlight the explanatory potential environmental forces, which have received less attention than their inductive counterparts.
 
 The core components of the model remain as before.
-All that is required is an adjustment to the possibility of confusing one state with another.
+All that is required is their adjustment to the possibility of confusing one state with another by language users and learners.
 Letting $\epsilon$ stand for the probability of confusing the first row state with the second, and vice-versa for $\delta$, we denote the probability that the teacher (learner) observes state $s_t$ ($s_l$) when the actual state is $s_a$ as $P_N(s_t \mid s_a)$ ($P_N(s_l \mid s_a)$).
@@
 \end{align*}
-Generalize this to a sequence of perceived data $d_l$ and write $P_N(d_l \mid t)$. We then
-define the noise-perturbed mutation matrix as:
+Generalize this to a sequence of perceived data $d_l$ and write $P_N(d_l \mid t)$. Then, the noise-perturbed mutation matrix is defined as:
 \begin{align*}
@@
 
-\paragraph{Results.} As shown in Table \ref{tab:num-noise}, a prevalence of pragmatic $L_5$ can also arise from noisy transmission without a learning bias for particular types.
-In contrast to the outcome of the non-noisy model, favourable outcomes for this type concentrate in parameter configurations  with $\epsilon > \delta$, $\alpha \geq 5$, $1 < k < 20$ and $l > 3$.
-Crucially, these results show that the mere presence of systematic noise in the transmission of strategies can introduce regularization that looks as if the agents have a learning bias.
-But, most importantly, this disturbance of transmission fidelity is due to perceptual noise and properties of the environment, not learning biases.
-In other words, learning biases are clearly not the only transmission biases that can shape evolution alongside functional pressure.
+\paragraph{Results.} As shown in Table \ref{tab:num-noise}, a prevalence of pragmatic $L_5$ can also arise from noisy transmission without a learning bias for particular types $(c = 0)$.
+In contrast to the outcome predicted by its noiseless counterpart, favourable outcomes for this type concentrate in parameter configurations  with $\epsilon > \delta$, $\alpha \geq 5$, $1 < k < 20$ and $l > 3$.
+As stressed above, we are not concerned with the interpretation of these particular values for the case study at hand, but rather with the technical result that shows that the mere presence of systematic noise in the transmission of strategies can introduce regularization that looks as if the agents have a learning bias.
+In other words, learning biases are clearly not the only transmission perturbations that shape cultural evolution alongside functional pressure.
 Environmental and perceptual noise can play a role too.
@@
 We laid out a model that combines game theoretical models of functional pressure towards efficient communication \citep{nowak+krakauer:1999}, effects of transmission perturbations on (iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012, franke+jaeger:2014} as well as different lexica \citep{bergen+etal:2012,bergen+etal:2016}.
-This model generates predictions about lexicalization patterns and, more generally, effects of communicative pressures on language.
-We argued that the puzzle raised by semantics in light of pragmatics is hard to explain on purely functional grounds and that part of the answer may instead lie the way transmission shapes the outcome of cultural evolution in tandem with a pressure for successful information transfer.
-In the realm of learning biases, simpler semantic representations are more likely to be learned while pragmatic reasoning can counteract functional disadvantages otherwise incurred.
+This model generates predictions about lexicalization patterns and, more generally, effects of communicative pressures on the cultural evolution of language.
+We argued that the puzzle raised by semantics in light of pragmatics is hard to explain on purely functional grounds and that part of the answer may instead lie in the way transmission shapes the outcome of cultural evolution in tandem with a pressure for successful information transfer.
+In the realm of inductive biases, we adopted the assumption that simpler semantic representations are more likely to be learned (cf.
+\citealt{chater+vitanyi:2003}).
+Under this view, semantics and pragmatics play a synergic role in that representational simplicity is supplemented by pragmatic reasoning to counteract functional disadvantages otherwise incurred.
+As a consequence, iterated transmission and use of language lead to a regularization that may explain the lack of lexicalization of systematic pragmatic enrichments.
 This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics by offering an account of why (certain) pragmatic inferences fail to lexicalize.
-More generally, we showed that systematic noise in perception can produce outcomes that are indistinguishable from those generated by learning biases.
+More generally, we showed that systematic noise in perception can produce outcomes that are indistinguishable from those generated by inductive biases.
 This adds to \citeposs{franke+correia:toappear} argument that linguistic regularities may arise as a byproduct of noise, rather than through inductive biases.

### 2016-11-07 paper/paper-v2.tex
minor changes draft conclusion

@@
 \section{Conclusion}
-Language change is affected by intertwined pressures.
-We investigated the selection of lexical meanings lacking upper-bounds in a case study on role of semantics and pragmatics in satisfying such pressures.
-We showed that, when pressured for learnability and expressivity, the former drives for simpler semantic representations inasmuch as pragmatics can compensate for their lack of expressivity in use.
-That is, the relative learnability advantage of simpler semantics may offer an answer to why natural languages fail to lexicalize certain pragmatic inferences.
-We also considered an alternate instantiating of the model which shows that systematic noise in state perception can give rise to evolutionary outcomes that are indistinguishable from those predicted by inductive biases.
-More broadly, we argued that semantic patterns can be explained by taking into consideration the way in which they are used in interaction as well as their representational complexity and learnability.
+The cultural evolution of language is influenced by intertwined pressures.
+We set out to investigate this process by putting forward a model that combines a pressure toward efficient and successful information transfer with perturbations that may arise from the transmission of linguistic knowledge in acquisition.
+Additionally, we argued for the necessity of considering the role of pragmatics in investigations on the cultural evolution of meaning.
+These components and their mutual influence were highlighted in a case study on the lack of lexical upper-bounds in weak scalar expressions that showed that, when pressured for learnability and expressivity, the former drives for simpler semantic representations inasmuch as pragmatics can compensate for their lack of expressivity in use.
+That is, the relative advantage in learning of simpler semantics in tandem with a functional pressure in use may offer an answer to why natural languages fail to lexicalize systematic pragmatic inferences.
+
+We also considered an alternate instantiation of the model, which shows that systematic noise in state perception can give rise to evolutionary outcomes that are indistinguishable from those predicted by inductive biases.
+This stresses the fact that that learning and typology are not necessarily close reflections of each other \citep{bowerman:2010}.
+In particular, language use and environmental factors can play an important role in language change, making them central variables in explanatory accounts of natural language properties.
 

### 2016-11-11 paper/paper-v2.tex
paper intro extended and rewritten

@@
 
+\definecolor{Red}{RGB}{178,34,34}
+\newcommand{\mf}[1]{\textcolor{Red}{[mf: #1]}}
+
 %--------------------
@@
 \begin{abstract}
-According to standard linguistic theory, the conventional meaning of expressions is systematically modified by general pragmatic rules governing their use.
-This implies that models of the cultural evolution of meaning should likewise take into consideration that observable language use is a complex interaction of semantic representations and pragmatic rules.
-To this end, we present a game theoretic model of the cultural evolution of language where communicative pressures work on abstract semantic representations and pragmatic usage rules.
-Our model traces two evolutionary forces and their interaction: (i) fitness-based pressure towards communicative efficiency and (ii) systematic transmission perturbations when linguistic knowledge is transferred from one agent to another.
-The latter can arise from general cognitive or learning biases, but also from other sources of systematic noise, e.g., as a result of errors in perception.
-We illustrate the model based on a case study showing that cognitive biases that favor simple semantic representations can prevent the lexicalization of pragmatic inferences.
-We also show that, technically speaking, it is possible that environmental factors, such as perceptual errors during acquisition, can produce evolutionary outcomes that look as if such cognitive biases are present even if they are not.
+  According to standard linguistic theory, the meaning of an utterance is the product of a
+  conventional semantic meaning of the used expression and general pragmatic reasoning applied
+  to the context of utterance. This implies that models of the cultural evolution of meaning
+  should likewise take into consideration that observable language use is a complex interaction
+  of semantic representations and pragmatic use. To this end, we present a game theoretic model
+  of the cultural evolution of language where communicative pressures work on abstract semantic
+  representations and pragmatic patterns of use. Our model traces two evolutionary forces and
+  their interaction: (i) fitness-based pressure towards communicative efficiency and (ii)
+  systematic transmission perturbations when linguistic knowledge is transferred from one agent
+  to another. The latter can arise from general cognitive or learning biases, but also from
+  other sources of systematic noise, e.g., as a result of errors in perception. We illustrate
+  the model based on a case study showing that cognitive biases that favor simple semantic
+  representations can prevent the lexicalization of pragmatic inferences. We also show that,
+  technically speaking, it is possible that environmental factors, such as perceptual errors
+  during acquisition, can produce evolutionary outcomes that look as if such cognitive biases
+  are present even if they are not.
 \end{abstract}
@@
 \section{Introduction}\label{sec:introduction}
-What is conveyed usually goes beyond what is said.
-A request for a blanket can be politely veiled by uttering ``I'm cold'', the temporal succession of events can be conveyed by the order in which conjuncts appear as in ``I traveled to Paris and got married'', or an invitation be declined by saying ``I have to work''.
-An influential explanation of the relation between the literal meaning of expressions and what they are intended and interpreted to convey due to \citet{grice:1975} views the latter as a product of a process of mutual reasoning about rational language use.
-For instance, under the assumption that the speaker is cooperative and relevant, ``I have to work'' may be interpreted as providing a reason why the speaker will not be able to accept the invitation, going beyond its literal meaning.
-Crucially, many classes of lexical meanings allow for pragmatic enrichments in a notably systematic fashion.
+What is conveyed usually goes beyond what is said. A request for a blanket can be politely
+veiled by uttering ``I'm cold.'' The temporal succession of events can be communicated by the
+order in which conjuncts appear as in ``I traveled to Paris and got married.'' An invitation
+can be declined by saying ``I have to work.'' An influential explanation of the relation
+between the literal meaning of expressions and what they are intended and interpreted to convey
+is due to \citet{grice:1975}. It characterizes pragmatic use and interpretation as a process of
+mutual reasoning about rational language use. For instance, under the assumption that the
+speaker is cooperative and relevant, ``I have to work'' may be interpreted as providing a
+reason why the speaker will not be able to accept the invitation, going beyond its literal
+meaning. Some of these enrichments are rather \emph{ad hoc}. Others show a striking
+regularities, such as the use of questions for polite requests, or certain enrichments of
+lexical meanings such as \emph{and} to mean \emph{and then}.
 
-Productive pragmatic inferences, such as scalar and manner implicatures, have been at the focus of many studies on the semantics-pragmatics interface.
-However, an issue that has received little attention is the justification of semantic structure in light of the systematic enrichments offered by pragmatics.
-Similarly, differnt models of the cultural evolution of language have been put forward in order to analyze properties such as compositionality and combinatoriality (e.g.
-\citealt{kirby+etal:2015,verhoef+etal:2014}), but the selection of meaning in light of pragmatics remains uncharted territory.
-The present investigation seeks to fill these gaps by analyzing the effects linguistic pressures have on the selection and pervasiveness of particular lexical meanings under consideration of their pragmatic modification.
-Accordingly, our primary goal is to uncover how transmission perturbations and functional pressure interact in the cultural evolution of meaning.
-Additionally, we show how systematic disturbances in the fidelity by which linguistic knowledge is transmitted that stem from environmental factors can lead to outcomes akin to those predicted under the assumption of inductive learning biases.
-That is, that the regularization standardly predicted to follow from learning biases can also emerge as a byproduct of noise.
+A particularly productive and well studied class of pragmatic enrichments are scalar
+implicatures
+\citep{horn:1984,Hirschberg1985:A-Theory-of-Sca,LevinsonPragmatics1983,Geurts2010:Quantity-Implic}. Usually,
+an utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
+that the speaker does not own all of them. This is because, if the speaker had them all, he
+could have used the stronger word \emph{all} instead of \emph{some} in his utterance and
+thereby would have made a more informative statement. Scalar implicatures, especially the
+inference from \emph{some} to \emph{some but not all}, have been studied extensively, both
+theoretically
+\citep[e.g.][]{Sauerland2004:Scalar-Implicat,ChierchiaFox2008:The-Grammatical,Rooyvan-RooijJagerde-Jager2012:Explaining-Quan}
+as well as experimentally
+\citep[e.g.][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. While
+there is much dispute in this domain and many interesting details, a position endorsed by a
+clear majority is that a word like \emph{some} is underspecified to mean \emph{some and maybe
+  all} and that the enrichment to \emph{some but not all} is part of some regular enrichment
+process with roots in pragmatics.
+
+If this majority view of underspecified lexical meaning and pragmatic enrichment routines is
+correct, the question arises how such a system could have emerged in language evolution. Models
+of meaning evolution abound. There are simulation-based models studying the evolution of
+meaning in populations of communicating agents
+\citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
+and there are mathematical models meaning evolution, mostly rooted in game theory
+\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
+work has focused on explaining basic properties such as compositionality and combinatoriality
+\citep[e.g.][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax}.
+But
+little attention has been paid to the interaction between conventional meaning and pragmatic
+use.
+
+To fill this gap, we here spell out a model of the co-evolution of conventional meaning and
+pragmatic reasoning types. In other words, the object of evolutionary selection are pairs of
+lexical meanings and general types of pragmatic behavior. Our model combines state-of-the-art
+probabilistic cognitive models of pragmatic language use
+\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
+with a general and established model of language evolution
+\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,Nowak2006:Evolutionary-Dy}. The
+latter allows us to study the interaction between (i) evolutionary pressure towards
+communicative efficiency and (ii) possible infidelity in the transmission of linguistic
+knowledge, such as from inductive learning biases or systematic perceptual errors. This model
+contains the most versatile dynamic from evolutionary game theory, the replicator dynamic
+\citep{TaylorJonker1978:Evolutionary-St}, and a well-understood model of iterated Bayesian
+learning \citep{griffiths+kalish:2007} as two special cases. The latter is particularly
+important because in language learning agents need to infer unobservable conventional meanings
+and pragmatic reasoning types from observable language use of their parent
+generation. Section~\ref{sec:model} introduces this model.
+
+Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. We
+demonstrate a setting in which the majority view of underspecified lexical meanings and
+pragmatic enrichments emerges, but only if selection and transmission infidelity are
+combined. In particular, we show that inductive learning biases of Bayesian learners that favor
+simpler lexical meanings can lead to the desired outcome. Additionally, we show how systematic
+disturbances from environmental factors can lead to outcomes akin to those predicted under the
+assumption of inductive learning biases. That is, that the regularization standardly predicted
+to follow from learning biases can also emerge as a byproduct of noise.
+
+% We see the main contribution of this work as conceptual and technical, not as a definite answer
+% to the question why scalar implicatures emerged. The work here rather demonstrates how current
+% probabilistic cognitive modeling of language use and evolutionary modeling can be fruitfully
+% combined to study the co-evolution of semantics and pragmatics side-by-side.
+
+\mf{on cultural evolution: \citet{Pagel2009:Human-Language-}}
+
+\section{Model}
+\label{sec:model}
 

### 2016-11-12 paper/paper-v2.tex
paper V2 rewrite: pulling RMD to the front in section 2

@@
 reason why the speaker will not be able to accept the invitation, going beyond its literal
-meaning. Some of these enrichments are rather \emph{ad hoc}. Others show a striking
-regularities, such as the use of questions for polite requests, or certain enrichments of
-lexical meanings such as \emph{and} to mean \emph{and then}.
+meaning. Some of these enrichments are rather \emph{ad hoc}. Others show striking regularities,
+such as the use of ability questions for polite requests (``Could you please \dots?''), or
+certain enrichments of lexical meanings such as \emph{and} to mean \emph{and then}.
 
@@
 \citep[e.g.][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. While
-there is much dispute in this domain and many interesting details, a position endorsed by a
+there is much dispute in this domain about many interesting details, a position endorsed by a
 clear majority is that a word like \emph{some} is underspecified to mean \emph{some and maybe
@@
 
-If this majority view of underspecified lexical meaning and pragmatic enrichment routines is
-correct, the question arises how such a system could have emerged in language evolution. Models
-of meaning evolution abound. There are simulation-based models studying the evolution of
-meaning in populations of communicating agents
+If this majority view is correct, the question arises how such a division of labor between
+semantics and pragmatics could have evolved. Models of meaning evolution abound. There are
+simulation-based models studying the evolution of meaning in populations of communicating
+agents
 \citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
-and there are mathematical models meaning evolution, mostly rooted in game theory
+and there are mathematical models of meaning evolution, mostly rooted in game theory
 \citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
@@
 little attention has been paid to the interaction between conventional meaning and pragmatic
-use.
+use. What is more, many mathematical models explain evolved meaning as a regularity in the
+behavior of agents which maps objective states of the world to observable signals. There is no
+room in such a purely extensional approach to address the semantics-pragmatics division
+directly. We would need to look at richer representations of cognizing agents and their
+communicative interaction.
 
 To fill this gap, we here spell out a model of the co-evolution of conventional meaning and
-pragmatic reasoning types. In other words, the object of evolutionary selection are pairs of
-lexical meanings and general types of pragmatic behavior. Our model combines state-of-the-art
+pragmatic reasoning types. The objects of replication and selection are pairs of lexical
+meanings and general types of pragmatic behavior, which we represent using state-of-the-art
 probabilistic cognitive models of pragmatic language use
-\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
-with a general and established model of language evolution
-\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,Nowak2006:Evolutionary-Dy}. The
-latter allows us to study the interaction between (i) evolutionary pressure towards
+\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
+and selection is described by the \emph{replicator mutator dynamic}, a general and established
+model of language evolution
+\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. The
+approach allows us to study the interaction between (i) evolutionary pressure towards
 communicative efficiency and (ii) possible infidelity in the transmission of linguistic
-knowledge, such as from inductive learning biases or systematic perceptual errors. This model
-contains the most versatile dynamic from evolutionary game theory, the replicator dynamic
-\citep{TaylorJonker1978:Evolutionary-St}, and a well-understood model of iterated Bayesian
-learning \citep{griffiths+kalish:2007} as two special cases. The latter is particularly
-important because in language learning agents need to infer unobservable conventional meanings
-and pragmatic reasoning types from observable language use of their parent
-generation. Section~\ref{sec:model} introduces this model.
+knowledge, such as from inductive learning biases or systematic perceptual errors. The latter
+is particularly important because neither semantic meanings nor pragmatic usage patterns are
+directly observable. Language learners have to infer these from observable behavior which is
+the product of the former. We formalize this acquisition process as a form of Bayesian
+inference. Our approach thereby contains a well-understood model of iterated Bayesian learning
+\citep{griffiths+kalish:2007} as a special cases, but combines it with functional selection,
+here formalized as the most versatile dynamic from evolutionary game theory, the replicator
+dynamic \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this
+model.
 
 Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. We
-demonstrate a setting in which the majority view of underspecified lexical meanings and
-pragmatic enrichments emerges, but only if selection and transmission infidelity are
-combined. In particular, we show that inductive learning biases of Bayesian learners that favor
-simpler lexical meanings can lead to the desired outcome. Additionally, we show how systematic
-disturbances from environmental factors can lead to outcomes akin to those predicted under the
-assumption of inductive learning biases. That is, that the regularization standardly predicted
-to follow from learning biases can also emerge as a byproduct of noise.
+discuss a setting in which the majority view of underspecified lexical meanings and pragmatic
+enrichments emerges if selection and transmission infidelity are combined. In particular, we
+show that inductive learning biases of Bayesian learners that favor simpler lexical meanings
+can lead to the desired outcome. Additionally, we show that a similar outcome can be achieved
+without assuming any cognitive biases, simply as an epiphenomenon of systematic disturbances from
+environmental factors. This formal results highlights the frequently overlooked possibility
+that channel noise in evolutionary replication can mimic effects of inductive biases.
 
-% We see the main contribution of this work as conceptual and technical, not as a definite answer
-% to the question why scalar implicatures emerged. The work here rather demonstrates how current
-% probabilistic cognitive modeling of language use and evolutionary modeling can be fruitfully
-% combined to study the co-evolution of semantics and pragmatics side-by-side.
-
-\mf{on cultural evolution: \citet{Pagel2009:Human-Language-}}
+We see the main contribution of this work as conceptual and technical, not as a definite answer
+to the question why scalar implicatures emerged. The work here rather demonstrates how current
+probabilistic cognitive modeling of language use and evolutionary modeling can be fruitfully
+combined to study the co-evolution of semantics and pragmatics side-by-side. Reversely, the
+approach taken here may be seen as a first step towards giving an evolutionary rationale for
+empirically successful probabilistic models of language use that embrace the majority view of
+the division of labor between semantics and pragmatics. Section~\ref{sec:discussion} elaborates
+on these points.
 
@@
 
-\subsection{Communicative pressures at the semantics-pragmatics interface}
-The emergence and change of linguistic structure is influenced by many intertwined factors.
-These range from biological and socio-ecological to cultural ones \citep{steels:2011}.
-Social and ecological pressures determine communicative needs, while biology determines the architecture that enables and constrains the means by which they can be fulfilled.
-In the following, our focus lies on the cultural aspects, wherein processes of linguistic change are viewed as shaped by language use and its transmission, i.e., as a result of a process of cultural evolution.
+\subsection{Expressivity and learnability at the semantics-pragmatics interface}
 
-The idea that language is influenced by communicative pressures has played a pivotal role in synchronic and diachronic analyses at latest since \citeposs{zipf:1949} rationalization of the approximation of word frequency rankings by a power law distribution as competing hearer and speaker preferences (e.g.
-in \citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007, piantadosi:2014,kirby+etal:2015}).
-In recent years, this line of research has led to a surge of approaches that seek to analyze the effects of such pressures from a multidisciplinary perspective, ranging from simulations to experiments with human and robotic agents (see \citealt{steels:2015} and \citealt{tamariz+kirby:2016} for recent overviews).
-Our starting point is given by the overarching argument that has crystalized from this literature: Natural languages need to be well-adapted to communicative needs within a linguistic community, but also need to be learnable to survive their faithful transmission across generations.
-More succinctly; natural languages are pressured for expressivity and learnability.
+The emergence and change of linguistic structure is influenced by many intertwined
+factors. These range from biological and socio-ecological to cultural ones \citep{steels:2011}
+\mf{more references?}. Social and ecological pressures determine communicative needs, while
+biology determines the architecture that enables and constrains the means by which they can be
+fulfilled. In the following, our focus lies on the cultural aspects, wherein processes of
+linguistic change are viewed as shaped by language use and its transmission, i.e., as a result
+of a process of cultural evolution
+\citep{Pagel2009:Human-Language-,ThompsonKirby2016:Culture-Shapes-}.
 
-The opposition of expressivity and learnability becomes particularly clear when considering their consequences in the extreme (cf.
-\citealt{kemp+regier:2012,kirby+etal:2015}).
-A language with a single form-meaning association is easy to learn but lacking in expressivity.
-Conversely, a language that associates a distinct form with all possible meanings a speaker may want to convey is maximally expressive but challenging to acquire.
-The most prominent problem that arises from this tension is that of acquiring a language to express a potentially infinite set of meanings through finite means \citep{kirby:2002}.
-However, this is not the only challenge learners confront.
-More central to our explanandum is the issue of selecting particular hypotheses out of a potentially infinite space of alternatives compatible with the data learners are exposed to.
-At the semantics-pragmatics interface this concerns the selection between functional (near-)equivalents, noting in particular that what is systematically conveyed through pragmatics could alternatively be codified lexically.
-Notwithstanding, lexical meanings that allow for pragmatic modifications often seem to resist the lexicalization of their pragmatic component.
-The question is why.
+The idea that language is an adaptation to serve a communicative function has played a pivotal
+role in synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of
+word frequency rankings as a result of competing hearer and speaker preferences (e.g. in
+\citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
+  piantadosi:2014,kirby+etal:2015}). If processes of selection, such as conditional imitation
+or reinforcement, favor more communicatively efficient types of behavior, languages are driven
+towards semantic expressivity \citep[e.g.][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
+pressure towards communicative efficiency is not the only force that shapes
+language. Learnability is another. Natural languages need to be learnable to survive their
+faithful transmission across generations. Clearly, an unlearnable code will not make it past
+the one happy fellow who invented it. Moreover, even small biases implicit in the acquisition
+of a language can built up and have quite striking effects on an evolving language in a process
+of iterated learning
+\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}. In
+sum, natural languages are pressured for expressivity and learnability. But expressivity and
+learnability may pull in opposite directions. The opposition becomes particularly clear when
+considering the extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). A language with a
+single form-meaning association is easy to learn but lacking in expressivity. Conversely, a
+language that associates a distinct form with all possible meanings a speaker may want to
+convey is maximally expressive but challenging to acquire.
+
+An elegant formal approach to modeling the competition between expressivity and learnability is
+the \emph{replicator mutator dynamic}
+\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. In
+its simplest, discrete-time formulation, the RMD defines the frequencies $x'_i$ of each type
+$i$ in a population at the next time step as a function of: (i) the frequencies $x_i$ of each
+type $i$ before the update step, (ii) each the fitness $f_i$ of type $i$, and the probability
+$Q_{ji}$ that an agent who wants to imitate, adopt or learn the type of an agent with type $j$
+ends up acquiring type $i$:
+\begin{align}
+  \label{eq:RMD_discrete}
+  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}
+\end{align}
+The replicator mutator dynamic consists of two components: fitness-based selection and
+transmission biases. This becomes most transparent when we consider an equivalent formulation
+in terms of a step-wise application of the discrete-time replicator dynamic
+\citep{TaylorJonker1978:Evolutionary-St} and subsequent multiplication with a mutation matrix
+$Q$ of the initial population vector $vec{x}$:
+\begin{align}
+  \label{eq:RMD_discrete_recast}
+  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i \ \ \text{, where} \\
+  (\text{M}(\vec{x}))_i & = (\vec{x} \cdot Q)_i = \left ( \sum_j
+          x_j Q_{ji} \right)_i \ \  \text{and}  \\
+        \left ( \text{RD}(\vec{x}) \right )_i 
+        & = \frac{x_i f_i}{\sum_k x_k f_k}
+\end{align}
+
+thus conceived: replication then mutation; if $f_i$ is the same for all $i$ we get only
+multiplication by $Q$, if $Q_{ji} = 1$ exactly when $i=j$, we get the RD, la di da
+
+main question what are types, what is fitness, what is $Q$ answered next
+
+\bigskip
+
+The replicator equation gives us the means to make these dynamics precise. The proportion of
+types in a given population is codified in a vector $x$, where $x_i$ is type $i$'s
+proportion. As noted above, the fitness of type $i$ is equal to its relative communicative
+success within this population, $f_i = \sum_j x_j \text{EU}(t_i,t_j)$. The expected
+communicative success of $i$ and $j$ is simply the average success of $i$ conveying information
+to $j$ and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$. $U_S(x,y)$ and
+$U_R(x,y)$ are, respectively, $\sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$
+and $U_S(y,x)$, for $n$ and $o$ being the reasoning level of $x$ and $y$, and
+$\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise. This quantity is symmetric, reflecting the
+probability of two types' mutual understanding. The average fitness of the population is given
+by $\Phi$, $\Phi = \sum_i x_i f_i$, serving as a normalizing constant for the discrete
+replicator equation: $\dot{x}_i = \frac{x_i f_i}{\Phi}$
+
+\bigskip
+
+However, this is
+not the only challenge learners confront. More central to our explanandum is the issue of
+selecting particular hypotheses out of a potentially infinite space of alternatives compatible
+with the data learners are exposed to. At the semantics-pragmatics interface this concerns the
+selection between functional (near-)equivalents, noting in particular that what is
+systematically conveyed through pragmatics could alternatively be codified
+lexically. Notwithstanding, lexical meanings that allow for pragmatic modifications often seem
+to resist the lexicalization of their pragmatic component. The question is why.
 

### 2016-11-13 paper/paper-v2.tex
paper rewrite

@@
 \citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. In
-its simplest, discrete-time formulation, the RMD defines the frequencies $x'_i$ of each type
-$i$ in a population at the next time step as a function of: (i) the frequencies $x_i$ of each
-type $i$ before the update step, (ii) each the fitness $f_i$ of type $i$, and the probability
+its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$
+in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
+$i$ before the update step, (ii) the fitness $f_i$ of each type $i$, and (iii) the probability
 $Q_{ji}$ that an agent who wants to imitate, adopt or learn the type of an agent with type $j$
@@
   \label{eq:RMD_discrete}
-  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}
+  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}\,.
 \end{align}
@@
 \citep{TaylorJonker1978:Evolutionary-St} and subsequent multiplication with a mutation matrix
-$Q$ of the initial population vector $vec{x}$:
+$Q$ of the initial population vector $\vec{x}$:
 \begin{align}
   \label{eq:RMD_discrete_recast}
-  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i \ \ \text{, where} \\
-  (\text{M}(\vec{x}))_i & = (\vec{x} \cdot Q)_i = \left ( \sum_j
-          x_j Q_{ji} \right)_i \ \  \text{and}  \\
+  x'_i & = (\text{M}(\text{RD}(\vec{x})))_i\,,
+\end{align}
+where
+\begin{align*}
+  (\text{M}(\vec{x}))_i = (\vec{x} \cdot Q)_i = \left ( \sum_j
+          x_j Q_{ji} \right)_i \ \ \ \ \text{and} \ \ \ \ 
       \left ( \text{RD}(\vec{x}) \right )_i 
-        & = \frac{x_i f_i}{\sum_k x_k f_k}
-\end{align}
+         = \frac{x_i f_i}{\sum_k x_k f_k}\,.
+ \end{align*}
+ If the transmission matrix $Q$ is trivial in the sense that $Q_{ij}=1$ whenever $i=j$, the
+ dynamic reduces to the replicator dynamic. The replicator dynamic is a model of fitness-based
+ selection in which the relative frequency of type $i$ will increase with a gradient
+ proportional to its average fitness in the population. The replicator dynamic is popular and
+ versatile because it can be derived from many abstract processes of biological and cultural
+ transmission and selection \citep[for overview and any derivations
+ see][]{Sandholm2010:Population-Game}, including conditional imitation
+ \citep[e.g.][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
+ learning \citep[e.g.][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. If $f_i$
+ is the same for all types $i$, the replicator step is the identity map
+ $ \left ( \text{RD}(\vec{x}) \right )_i = x_i$ and the dynamic reduces to a process of
+ iteration of the transmission bias encoded in $Q$. In this way, the process in
+ (\ref{eq:RMD_discrete}), equivalently (\ref{eq:RMD_discrete_recast}), contains a model of
+ iterated learning \citep{griffiths+kalish:2007}.
 
-thus conceived: replication then mutation; if $f_i$ is the same for all $i$ we get only
-multiplication by $Q$, if $Q_{ji} = 1$ exactly when $i=j$, we get the RD, la di da
-
-main question what are types, what is fitness, what is $Q$ answered next
+Where our goal is an application of this dynamic to the case of co-evolution of semantic
+meaning and pragmatic use, we need to fix what the relevant types are, how fitness is measured
+and how the mutation matrix is computed. These issues will be addressed, one by one, in the
+following.
 
@@
 
-\subsection{Lexica and linguistic behavior}\label{sec:languages+use}
+\subsection{Types: Lexica and linguistic behavior}
+\label{sec:languages+use}
+
 Lexica codify the truth-conditions of a language's expressions.

### 2016-11-18 paper/paper-v2.tex
paper rewrite model section

@@
 \definecolor{Red}{RGB}{178,34,34}
-\newcommand{\mf}[1]{\textcolor{Red}{[mf: #1]}}
+\newcommand{\mf}[1]{\textcolor{Red}{[MF: #1]}} 
+\usepackage{blkarray}
+\usepackage{xspace}
+
+%%% MF's commands
+\newcommand{\set}[1]{\left\{#1\right\}}
+\newcommand{\card}[1]{\left \lvert \, #1 \, \right\rvert}
+\newcommand{\abs}[1]{\lvert #1 \rvert}
+\newcommand{\States}{\ensuremath{S}\xspace}		% Set of States
+\newcommand{\state}{\ensuremath{s}\xspace}		% single states
+\newcommand{\mystate}[1]{\ensuremath{\state_{\text{#1}}}\xspace} %meaningful states
+\newcommand{\Messgs}{\ensuremath{M}\xspace}		% Set of Messages
+\newcommand{\messg}{\ensuremath{m}\xspace}		% single messages
+\newcommand{\mymessg}[1]{\ensuremath{\messg_{\text{#1}}}\xspace} %meaningful messages
+\newcommand{\ssome}{\mystate{\ensuremath{\exists\neg\forall}}}
+\newcommand{\sall}{\mystate{\ensuremath{\forall}}}
+\newcommand{\msome}{\mymessg{some}}
+\newcommand{\mall}{\mymessg{all}}
+\newcommand{\asome}{\myact{\ensuremath{\exists\neg\forall}}}
+\newcommand{\aall}{\myact{\ensuremath{\forall}}}
+\definecolor{mygray}{cmyk}{0.35,0.35,0.35,0.35}
+\newcommand{\mygray}[1]{{\textcolor{mygray}{#1}}}
+%%%
 
@@
 \citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
-and there are mathematical models of meaning evolution, mostly rooted in game theory
+and there are mathematical models of meaning evolution, mostly coming from in game theory
 \citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
@@
 and selection is described by the \emph{replicator mutator dynamic}, a general and established
-model of language evolution
+model of evolutionary change in large and homogeneous populations
 \citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. The
@@
 communicative efficiency and (ii) possible infidelity in the transmission of linguistic
-knowledge, such as from inductive learning biases or systematic perceptual errors. The latter
-is particularly important because neither semantic meanings nor pragmatic usage patterns are
-directly observable. Language learners have to infer these from observable behavior which is
-the product of the former. We formalize this acquisition process as a form of Bayesian
-inference. Our approach thereby contains a well-understood model of iterated Bayesian learning
-\citep{griffiths+kalish:2007} as a special cases, but combines it with functional selection,
-here formalized as the most versatile dynamic from evolutionary game theory, the replicator
-dynamic \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this
-model.
+knowledge, such as from inductive learning biases or systematic perceptual errors. Considering
+transmission of linguistic knowledge is important because neither semantic meanings nor
+pragmatic usage patterns are directly observable. Language learners have to infer these
+unobservables from the observable behavior in which they result. We formalize this acquisition
+process as a form of Bayesian inference. Our approach thereby contains a well-understood model
+of iterated Bayesian learning \citep{griffiths+kalish:2007} as a special case, but combines it
+with functional selection, here formalized as the most versatile dynamic from evolutionary game
+theory, the replicator dynamic
+\citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this model.
 
@@
 
-An elegant formal approach to modeling the competition between expressivity and learnability is
-the \emph{replicator mutator dynamic}
+An elegant formal approach to capturing the interaction between expressivity and learnability
+is the \emph{replicator mutator dynamic}
 \citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. In
@@
 in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
-$i$ before the update step, (ii) the fitness $f_i$ of each type $i$, and (iii) the probability
-$Q_{ji}$ that an agent who wants to imitate, adopt or learn the type of an agent with type $j$
-ends up acquiring type $i$:
+$i$ before the update, (ii) the fitness $f_i$ of each type $i$ before the update, and (iii) the
+probability $Q_{ji}$ that an agent who wants to imitate, adopt or learn the type of an agent
+with type $j$ ends up acquiring type $i$:
 \begin{align}
@@
 \end{align}
-The replicator mutator dynamic consists of two components: fitness-based selection and
-transmission biases. This becomes most transparent when we consider an equivalent formulation
-in terms of a step-wise application of the discrete-time replicator dynamic
-\citep{TaylorJonker1978:Evolutionary-St} and subsequent multiplication with a mutation matrix
-$Q$ of the initial population vector $\vec{x}$:
+The RMD consists of two components: fitness-based selection and transmission biases. This
+becomes most transparent when we consider an equivalent formulation in terms of a step-wise
+application of the discrete-time replicator dynamic \citep{TaylorJonker1978:Evolutionary-St}
+and subsequent multiplication with a mutation matrix $Q$ of the initial population vector
+$\vec{x}$:
 \begin{align}
@@
 \begin{align*}
+      \left ( \text{RD}(\vec{x}) \right )_i 
+         = \frac{x_i f_i}{\sum_k x_k f_k}
+ \ \ \ \ \text{and} \ \ \ \ 
   (\text{M}(\vec{x}))_i = (\vec{x} \cdot Q)_i = \left ( \sum_j
-          x_j Q_{ji} \right)_i \ \ \ \ \text{and} \ \ \ \ 
-      \left ( \text{RD}(\vec{x}) \right )_i 
-         = \frac{x_i f_i}{\sum_k x_k f_k}\,.
+          x_j Q_{ji} \right)_i\,.
 \end{align*}
- If the transmission matrix $Q$ is trivial in the sense that $Q_{ij}=1$ whenever $i=j$, the
+If the transmission matrix $Q$ is trivial in the sense that $Q_{ji}=1$ whenever $j=i$, the
 dynamic reduces to the replicator dynamic. The replicator dynamic is a model of fitness-based
@@
 versatile because it can be derived from many abstract processes of biological and cultural
- transmission and selection \citep[for overview and any derivations
+transmission and selection \citep[for overview and several derivations
 see][]{Sandholm2010:Population-Game}, including conditional imitation
 \citep[e.g.][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
- learning \citep[e.g.][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. If $f_i$
- is the same for all types $i$, the replicator step is the identity map
+learning \citep[e.g.][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. If fitness
+$f_i$ is the same for all types $i$, the replicator step is the identity map
 $ \left ( \text{RD}(\vec{x}) \right )_i = x_i$ and the dynamic reduces to a process of
@@
 (\ref{eq:RMD_discrete}), equivalently (\ref{eq:RMD_discrete_recast}), contains a model of
- iterated learning \citep{griffiths+kalish:2007}.
+iterated learning \citep{griffiths+kalish:2007}. \mf{should we include a simple example here? I
+have an example from a lecture ready at hand; it's a simple coordination game in a
+one-population setting.}
 
@@
 
-\bigskip
-
-The replicator equation gives us the means to make these dynamics precise. The proportion of
-types in a given population is codified in a vector $x$, where $x_i$ is type $i$'s
-proportion. As noted above, the fitness of type $i$ is equal to its relative communicative
-success within this population, $f_i = \sum_j x_j \text{EU}(t_i,t_j)$. The expected
-communicative success of $i$ and $j$ is simply the average success of $i$ conveying information
-to $j$ and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$. $U_S(x,y)$ and
-$U_R(x,y)$ are, respectively, $\sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$
-and $U_S(y,x)$, for $n$ and $o$ being the reasoning level of $x$ and $y$, and
-$\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise. This quantity is symmetric, reflecting the
-probability of two types' mutual understanding. The average fitness of the population is given
-by $\Phi$, $\Phi = \sum_i x_i f_i$, serving as a normalizing constant for the discrete
-replicator equation: $\dot{x}_i = \frac{x_i f_i}{\Phi}$
-
-\bigskip
-
-However, this is
-not the only challenge learners confront. More central to our explanandum is the issue of
-selecting particular hypotheses out of a potentially infinite space of alternatives compatible
-with the data learners are exposed to. At the semantics-pragmatics interface this concerns the
-selection between functional (near-)equivalents, noting in particular that what is
-systematically conveyed through pragmatics could alternatively be codified
-lexically. Notwithstanding, lexical meanings that allow for pragmatic modifications often seem
-to resist the lexicalization of their pragmatic component. The question is why.
-
-We assume an integral part of the answer to lie in the effects of transmission perturbations that arise when linguistic knowledge is acquired by new learners.
-As shown in the following, such perturbations may take the form of learning biases, but may also stem from extraneous factors such as a noisy perception of the environment.
-
-We model these components using the replicator-mutator dynamics, combining functional pressure on successful communication with effects of transmission perturbations on (iterated) Bayesian learning \citep{griffiths+kalish:2007}.
-The semantics-pragmatics distinction and its bearing on production and comprehension are captured by a probabilistic model of rational language use with different degrees of pragmatic sophisitication and different lexica \citep{frank+goodman:2012,franke+jaeger:2014, bergen+etal:2016}.
-The remainder of this section introduces these components together with the assumptions underlying them.
-These are: the representation of languages and their use (\S\ref{sec:languages+use}), as well as pressures towards expressivity (\S\ref{sec:expressivity}) and learnability (\S\ref{sec:learnability}).
-After laying out the model, we analyze its predictions in a case-study on the lack of lexicalization of scalar implicatures in \S\ref{sec:si-case-study}.
-Additionally, \S\ref{subsec:noise} shows how the outcome predicted under the assumption of an inductive bias for simplicity in \S\ref{subsec:bias} can also arise from as a byproduct of the noisy perception of language users and learners.
-Section \S\ref{sec:discussion} discusses the general predictions of the model, as well as the implications drawn from its noisy counterpart.
-
 \subsection{Types: Lexica and linguistic behavior}
@@
 
-Lexica codify the truth-conditions of a language's expressions.
-Following \citet{franke+jaeger:2014}, a convenient way to represent lexica is by $(|S|,|M|)$-Boolean matrices, where $S$ is a set of states of affairs (meanings) and $M$ a set of messages (forms available in the language).
+Types are what cultural evolution operates on. In standard applications of evolutionary game
+theory, types correspond to ways of acting in a game, e.g., either cooperating or defecting in
+a prisoner's dilemma. \mf{maybe good to refer back to the example from before if there was
+  one?} For our purposes, we would rather like to identify a type by its cognitive make-up. A
+type is a pair consisting of a lexicon and a pragmatic strategy. This is because we are
+interested in the question under which conditions processes of cultural evolution will favor
+specific divisions of labor between lexical meaning and pragmatic use.
 
-We distinguish between two kinds of linguistic behavior.
-{\em Literal interlocutors} produce and interpret messages literally, being guided only by their lexica.
-{\em Pragmatic interlocutors} instead engage in mutual reasoning to inform their choices.
-Following models of rational language use such as Rational Speech Act models \citep{frank+goodman:2012} and their game-theoretic predecessors, the Iterated Best/Quantal Response models \citep{franke:2009,franke+jaeger:2014}, these behaviors are captured by a reasoning hierarchy.
-The hierarchy's bottom, level $0$, corresponds to literal language use.
-Pragmatic language users of level $n + 1$ behave rationally according to expected level $n$ behavior of their interlocutors.
-(\ref{h:level0}) and (\ref{h:leveln}) specify the behavior of literal and pragmatic hearers of a language $L$.
-Their speaker counterparts are given in (\ref{s:level0}) and (\ref{s:leveln}).
+Lexica codify the truth-conditions of a language's expressions. A convenient way to represent
+lexica is by $(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of
+states of affairs (meanings) and $M$ a set of messages (forms available in the language). For
+example, suppose that there are two relevant world states $\States = \set{\msome, \mall}$:
+$\msome$ where Chris owns some but not all of Johnny Cash's albums and $\mall$ where Chris owns
+them all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
+short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
+same sentence with \emph{some} replaced by \emph{all}. Lexica for this case would assign a
+truth value for each state-message pair. The following two lexica exemplify the distinction
+between a lexicalized upper-bound for \emph{some} in $L_4$ and the widely assumed logical
+semantics with only a lower-bound in $L_5$. \mf{can we not give these guys more mnemonic names?}
+\begin{align*}
+  L_{4} & = \begin{blockarray}{lcc}
+    & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[cc]}
+      \mygray{\ssome} & 1 & 0 \\
+      \mygray{\sall}  & 0 & 1 \\
+    \end{block}
+  \end{blockarray} &
+  L_{5} & = \begin{blockarray}{lcc}
+    & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[cc]}
+      \mygray{\ssome} & 1 & 0 \\
+      \mygray{\sall}  & 1 & 1 \\
+    \end{block}
+  \end{blockarray}
+\end{align*}
+
+We distinguish between two kinds of linguistic behavior. {\em Literal interlocutors} produce
+and interpret messages literally, being guided only by their lexica. {\em Pragmatic
+  interlocutors} instead engage in mutual reasoning to inform their choices. Following models
+of rational language use such as Rational Speech Act models \citep{frank+goodman:2012} and
+their game-theoretic predecessors, the Iterated Best/Quantal Response models
+\citep{franke:2009,franke+jaeger:2014}, these behaviors are captured by a reasoning
+hierarchy. The hierarchy's bottom, level $0$, corresponds to literal language use. Pragmatic
+language users of level $n + 1$ behave rationally according to expected level $n$ behavior of
+their interlocutors. (\ref{h:level0}) and (\ref{h:leveln}) specify the behavior of literal and
+pragmatic hearers of a language $L$. Their speaker counterparts are given in (\ref{s:level0})
+and (\ref{s:leveln}).
 
@@
 
-According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state $s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$.
-For simplicity, in the following this prior is assumed to be uniform across hearers.
-The behavior of literal speakers, given in (\ref{s:level0}), is regulated by a parameter $\alpha$ which controls the sensitivity to which speakers prefer one signal over another based on its expected communicative success.
+According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state
+$s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in
+the following this prior is assumed to be uniform across hearers. The behavior of literal
+speakers, given in (\ref{s:level0}), is regulated by a parameter $\alpha$ which controls the
+sensitivity to which speakers prefer one signal over another based on its expected
+communicative success.
 
-Pragmatic behavior is similar to its literal counterparts.
-Their difference lies in that level $n+1$ speakers/hearers reason about level $n$ hearer/speaker behavior instead of solely relaying on their lexicon.
-That is, they reason about the way a rational level $n$ interlocutor would use or interpret a message, and behave according to these expectations.
-Speaker behavior is further influenced by a soft-max parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}.
-As $\lambda$ increases, choices made in production are more rational in that higher values lead to behavior that is increasingly in line with expected utility maximization.
+Pragmatic behavior is similar to its literal counterparts. Their difference lies in that level
+$n+1$ speakers/hearers reason about level $n$ hearer/speaker behavior instead of solely
+relaying on their lexicon. That is, they reason about the way a rational level $n$ interlocutor
+would use or interpret a message, and behave according to these expectations.  Speaker behavior
+is further influenced by a soft-max parameter $\lambda$, $\lambda \geq 1$
+\citep{luce:1959,sutton+barto:1998}. As $\lambda$ increases, choices made in production are
+more rational in that higher values lead to behavior that is increasingly in line with expected
+utility maximization.
 
-We call the combination of a lexicon with its use, i.e., a level in the reasoning hierarchy, a {\em type}. These are the basic units on which the model's dynamics operate.
+We call the combination of a lexicon with its use, i.e., a level in the reasoning hierarchy, a
+{\em type}. These are the basic units on which the model's dynamics operate.
 

### 2016-11-18 paper/paper-v2.tex
paper revision

@@
 
-We distinguish between two kinds of linguistic behavior. {\em Literal interlocutors} produce
-and interpret messages literally, being guided only by their lexica. {\em Pragmatic
-  interlocutors} instead engage in mutual reasoning to inform their choices. Following models
-of rational language use such as Rational Speech Act models \citep{frank+goodman:2012} and
-their game-theoretic predecessors, the Iterated Best/Quantal Response models
-\citep{franke:2009,franke+jaeger:2014}, these behaviors are captured by a reasoning
-hierarchy. The hierarchy's bottom, level $0$, corresponds to literal language use. Pragmatic
-language users of level $n + 1$ behave rationally according to expected level $n$ behavior of
-their interlocutors. (\ref{h:level0}) and (\ref{h:leveln}) specify the behavior of literal and
-pragmatic hearers of a language $L$. Their speaker counterparts are given in (\ref{s:level0})
-and (\ref{s:leveln}).
+We distinguish between two kinds of pragmatic behavior. {\em Literal interlocutors} produce and
+interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
+instead engage in mutual reasoning to inform their choices. Recent probabilistic models of
+rational language use
+\citep{franke:2009,frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
+capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
+level $0$, corresponds to literal language use. Pragmatic language users of level $n + 1$
+act (approximately) rational with respect to level $n$ behavior of their
+interlocutors. (\ref{h:level0}) and (\ref{s:level0}) define probabilistic behavior of literal
+hearers and speakers respectively: \mf{is it a problem that $S$ denotes speakers and the set of
+states?}
+\begin{flalign}
+&H_{0}(s|m;L) \propto pr(s) L_{sm} \label{h:level0}\\
+&S_{0}(m|s;L) \propto \exp((L_{sm})^\alpha) \label{s:level0}
+\end{flalign}
+According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state
+$s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in
+the following this prior is assumed to be uniform. A literal interpreter with lexicon $L_4$
+from above would then assign $\ssome$ a probability of one after hearing $\msome$, while a
+literal interpreter with lexicon $L_5$ would assign $\ssome$ probability $.5$. The simplest
+definition of a literal speaker would just be $S_{0}(m|s;L) \propto \exp((L_{sm})^\alpha)$, but
+the more complex formulation in (\ref{s:level0})
+
+\begin{itemize}
+\item no prob zero
+\item $\alpha$ adjusts the level of avoidance of untrue signaling
+\end{itemize}
+
+ The behavior of literal speakers, given in
+(\ref{s:level0}), is regulated by a parameter $\alpha$ which controls the sensitivity to which
+speakers prefer one signal over another based on its expected communicative success.
 
 \begin{flalign}
-&H_{0}(s|m;L) \propto pr(s) L_{sm} \label{h:level0}\\
-&S_{0}(m|s;L) \propto \exp((L_{sm})^\alpha) \label{s:level0}\\
 &H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
@@
 \end{flalign}
-
-According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state
-$s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in
-the following this prior is assumed to be uniform across hearers. The behavior of literal
-speakers, given in (\ref{s:level0}), is regulated by a parameter $\alpha$ which controls the
-sensitivity to which speakers prefer one signal over another based on its expected
-communicative success.
 

### 2016-11-19 paper/paper-v2.tex
paper rewrite section on types

@@
 a prisoner's dilemma. \mf{maybe good to refer back to the example from before if there was
-  one?} For our purposes, we would rather like to identify a type by its cognitive make-up. A
-type is a pair consisting of a lexicon and a pragmatic strategy. This is because we are
+  one?} For our purposes here, types are identified by their cognitive make-up. Since we are
 interested in the question under which conditions processes of cultural evolution will favor
-specific divisions of labor between lexical meaning and pragmatic use.
+specific divisions of labor between lexical meaning and pragmatic use, a type is a pair
+consisting of a lexicon and a pragmatic strategy.
 
-Lexica codify the truth-conditions of a language's expressions. A convenient way to represent
-lexica is by $(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of
-states of affairs (meanings) and $M$ a set of messages (forms available in the language). For
-example, suppose that there are two relevant world states $\States = \set{\msome, \mall}$:
-$\msome$ where Chris owns some but not all of Johnny Cash's albums and $\mall$ where Chris owns
-them all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
+Lexica codify the truth-conditions of expressions. A convenient way to represent lexica is by
+$(\card{\States}, \card{\Messgs})$-Boolean matrices, where $\States$ is a set of states
+(meanings) and $M$ a set of messages (forms available in the language). For example, suppose
+that there are two relevant world states $\States = \set{\ssome, \sall}$. In state $\ssome$
+Chris owns some but not all of Johnny Cash's albums while in $\sall$ Chris owns them
+all. Suppose that there are two messages $\Messgs = \set{\msome, \mall}$ where $\msome$ is
 short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
@@
 between a lexicalized upper-bound for \emph{some} in $L_4$ and the widely assumed logical
-semantics with only a lower-bound in $L_5$. \mf{can we not give these guys more mnemonic names?}
+semantics with only a lower-bound in $L_5$. \mf{can we not give these guys more mnemonic
+  names?}
 \begin{align*}
@@
 level $0$, corresponds to literal language use. Pragmatic language users of level $n + 1$
-act (approximately) rational with respect to level $n$ behavior of their
+act (approximately) rational with respect to level-$n$ behavior of their
 interlocutors. (\ref{h:level0}) and (\ref{s:level0}) define probabilistic behavior of literal
 hearers and speakers respectively: \mf{is it a problem that $S$ denotes speakers and the set of
-states?}
+states?}\mf{Literal speaker definition had $\alpha$ but that has no effect if we have Boolean
+semantics; I left this out.}
 \begin{flalign}
-&H_{0}(s|m;L) \propto pr(s) L_{sm} \label{h:level0}\\
-&S_{0}(m|s;L) \propto \exp((L_{sm})^\alpha) \label{s:level0}
+&H_{0}(s \mid m;L) \propto pr(s) L_{sm} \label{h:level0}\\
+&S_{0}(m \mid s;L) \propto \exp(L_{sm}) \label{s:level0}
 \end{flalign}
@@
 the following this prior is assumed to be uniform. A literal interpreter with lexicon $L_4$
-from above would then assign $\ssome$ a probability of one after hearing $\msome$, while a
-literal interpreter with lexicon $L_5$ would assign $\ssome$ probability $.5$. The simplest
-definition of a literal speaker would just be $S_{0}(m|s;L) \propto \exp((L_{sm})^\alpha)$, but
-the more complex formulation in (\ref{s:level0})
+from above would assign $\ssome$ a probability of $H_0(\ssome \mid \msome; L_4) = 1$ after
+hearing $\msome$, while a literal interpreter with lexicon $L_5$ would assign $\ssome$
+probability $H_0(\ssome \mid \msome; L_5) = 0.5$. A literal speaker chooses any true message
+for a state with equal probability and prefers a true message over a false one. The formulation
+in (\ref{s:level0}) allows also false messages to be sent with a small positive probability in
+analogy to the definition of pragmatic speakers in probabilistic pragmatic models. This is
+necessary to guarantee a mutation matrix with only positive entries (see below). A literal
+speaker with lexicon $L_4$ or $L_5$ produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
+\ssome;L_{4,5}) \approx .73$. The probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
+\sall;L_4) \approx .27$ for $L_4$ and $S_{0}(\msome \mid
+\sall;L_4) \approx .5$ for $L_5$. \mf{I realize that the probability of producing a false
+  message for literal speakers is actually rather high. That's iffy because our pragmatic types
+are able (given favorable parameter values) to perform much better. That means that we
+implicitly have engineered in a possible fitness advantage of pragmatic types! Should we do
+something about this? Use definition $S_{0}(m \mid s;L) \propto L_{sm} + 0.001$? Make sure we
+only look at parameters for pragmatic types that make them not much better than the literal guys?}
 
-\begin{itemize}
-\item no prob zero
-\item $\alpha$ adjusts the level of avoidance of untrue signaling
-\end{itemize}
-
- The behavior of literal speakers, given in
-(\ref{s:level0}), is regulated by a parameter $\alpha$ which controls the sensitivity to which
-speakers prefer one signal over another based on its expected communicative success.
-
+Pragmatic behavior of level-$n+1$ is similar to its literal counterparts but uses the
+interpretation or production behavior of a level-$n$ player instead of the lexical meaning:
 \begin{flalign}
@@
 \end{flalign}
-
-Pragmatic behavior is similar to its literal counterparts. Their difference lies in that level
-$n+1$ speakers/hearers reason about level $n$ hearer/speaker behavior instead of solely
-relaying on their lexicon. That is, they reason about the way a rational level $n$ interlocutor
-would use or interpret a message, and behave according to these expectations.  Speaker behavior
-is further influenced by a soft-max parameter $\lambda$, $\lambda \geq 1$
-\citep{luce:1959,sutton+barto:1998}. As $\lambda$ increases, choices made in production are
-more rational in that higher values lead to behavior that is increasingly in line with expected
-utility maximization.
-
-We call the combination of a lexicon with its use, i.e., a level in the reasoning hierarchy, a
-{\em type}. These are the basic units on which the model's dynamics operate.
+As usual in probabilistic pragmatics models, speaker behavior is regulated by a soft-max
+parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}. As $\lambda$
+increases, choices made in production are more rational in that higher values lead to behavior
+that is increasingly in line with expected utility maximization. Expected utility of a message
+$\messg$ in state $\state$ is here defined as $H_{n}(s|m;L)^\alpha$. The probability
+$H_{n}(s|m;L)$ is the probability that the hearer will assign to or choose the correct
+meaning. The transformation with $\alpha$ allows to differentiate theoretically interesting
+speaker production behavior. To see this, notice that for two states and two messages (the
+minimal scenario we will investigate), the possible values for $H_{0}(s|m;L)$ are restricted to
+zero, $0.5$ and one. Consider the case of $H_0$ with $L_5$, which assigns the following
+probabilities to states after hearing messages:
+\begin{align*}
+  H_{0}(\messg \mid \state; L_5) & = \begin{blockarray}{lcc}
+    & \mygray{\ssome} & \mygray{\sall} \\
+    \begin{block}{l[cc]}
+      \mygray{\msome} & 0.5 & 0.5 \\
+      \mygray{\mall}  & 0 & 1 \\
+    \end{block}
+  \end{blockarray}
+\end{align*}
+A level-1 speaker's utilities are then:
+\begin{align*}
+  \begin{blockarray}{lcc}
+    & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[cc]}
+      \mygray{\ssome} & 0.5^\alpha & 0^\alpha \\
+      \mygray{\sall}  & 0.5^\alpha & 1^\alpha \\
+    \end{block}
+  \end{blockarray}
+\end{align*}
+As the soft-max choice rule is sensitive to differences in utilities, no matter what $\lambda$
+is we will get $S_1(\msome \mid \ssome; L_5) = S_1(\mall \mid \sall; L_5)$ if $\alpha =
+1$. This means that we cannot distinguish pragmatically deviant choices (selecting a less
+informative message instead of a more informative one), from semantically deviant choices
+(selecting a false message instead of a true one). If $\alpha > 1$ pragmatically deviant
+choices will get less likely than semantically deviant choices; if $\alpha < 0$ it is the other
+way around. In sum, parameter $\lambda$ regulates the overall tendency to avoid pragmatically
+or semantically deviant choices, $\alpha$ further distinguishes between pragmatically and
+semantically deviant behavior. Variation in $\alpha$ is only necessary for the technical
+extension to environmental transmission noise in Section~XYZ. \mf{I would suggest to just give
+  the intuitive summary here and have a little appendix paragraph with the example. What do you
+say?}
 

### 2016-11-19 paper/paper-v2.tex
revised section on expressivity

@@
 
-\subsection{Expressivity}\label{sec:expressivity}
-Expressivity has received particular attention from investigations using evolutionary game theory (e.g.
-\citealt{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}).
-Under this view, a type's ability to convey and interpret information successfully confers it a higher fitness, a measure that is relative to the success of other types in the population.
-In the simplest models, fitness directly translates into the proportion of types present in the population after a generational turnover.
-This association of communicative success within a population with changes in the proportion of types present in it creates a feedback loop that pressures the population towards greater expressivity.
+\subsection{Fitness \& fitness-based selection based on expressivity}\label{sec:expressivity}
 
-The replicator equation gives us the means to make these dynamics precise.
-The proportion of types in a given population is codified in a vector $x$, where $x_i$ is type $i$'s proportion.
-As noted above, the fitness of type $i$ is equal to its relative communicative success within this population, $f_i = \sum_j x_j \text{EU}(t_i,t_j)$.
-The expected communicative success of $i$ and $j$ is simply the average success of $i$ conveying information to $j$ and vice versa: $\text{EU}(t_i,t_j) = [U_S(t_i,t_j) + U_R(t_i,t_j)]/2$.
-$U_S(x,y)$ and $U_R(x,y)$ are, respectively, $\sum_s P(s)\sum_m S_n(m|s;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')$ and $U_S(y,x)$, for $n$ and $o$ being the reasoning level of $x$ and $y$, and $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.
-This quantity is symmetric, reflecting the probability of two types' mutual understanding.
-The average fitness of the population is given by $\Phi$, $\Phi = \sum_i x_i f_i$, serving as a normalizing constant for the discrete replicator equation: $\dot{x}_i = \frac{x_i f_i}{\Phi}$
+Most evolutionary dynamics assume that the proportion of type $i$ in a population will increase
+or decrease as a function of its relative fitness $f_i$. In the context of language evolution,
+fitness is frequently associated with expressivity, i.e., the ability to successfully
+communicate with other language users from the same population
+\citep[e.g.][]{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Under a biological
+interpretation, the assumption is that organisms have a higher chance of survival and
+reproduction if they are able to share and receive useful information via communication with
+peers. Under a cultural interpretation, the picture is that agents themselves strive towards
+communicative success and therefore adapt or revise their behavior occasionally to achieve a
+higher communicative success (see \citealt[\S3.3]{benz+etal:2005b} for discussion).
 
-Under its biological interpretation, the replicator dynamics capture the idea of fitness-relative selection whereby fitter types produce more offspring, leading to their propagation in subsequent generations.
-Many aspects of language are subject to transmission and change that can be likened to such biological processes.
-Amongst others, replication can be construed as modelling language acquisition, as e.g.
-in \citealt{nowak+etal:2002}, but also as a process of horizontal adaptation in a single generation (see \citealt[\S3.3]{benz+etal:2005b} for discussion).
+The replicator equation gives us the means to make the ensuing dynamics precise, without
+necessarily committing to a biological or cultural interpretation. The proportion of types in a
+given population is codified in a vector $\vec{x}$, where $x_i$ is type $i$'s proportion. The
+fitness of type $i$ is type $i$'s average expected communicative success, or \emph{expected
+  utility} (EU), given the current frequencies of types in the current population:
+\begin{align*}
+  f_i = \sum_j x_j \text{EU}(t_i,t_j)\,.
+\end{align*}
+The expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
+average success of $i$ when talking or listening to $j$. Assuming that agents are speakers half
+of the time this yields:
+\begin{align*}
+  \text{EU}(t_i,t_j) = \nicefrac{1}{2} \, U_S(t_i,t_j) + \nicefrac{1}{2} \, U_H(t_i,t_j)\,,
+\end{align*}
+where $\text{EU}_S(t_i,t_j)$ and $\text{EU}_H(t_i,t_j)$ are the expected utilities for $i$ as a
+speaker and as a hearer when communicating with $j$, defined as follows where $n_i$ and $n_j$
+are type $i$'s and $j$'s pragmatic reasoning types and $L_i$ and $L_j$ are their lexica:
+\begin{flalign*}
+  & \text{EU}_S(t_i,t_j)  = \sum_s P(s)\sum_m S_{n_i}(m \mid s;L_i) \sum_{s'} R_{n_j}(s' \mid m;L_j)
+  \delta(s,s') \\
+ & \text{EU}_H(t_i,t_j)  = \text{EU}_S(t_j,t_i)
+\end{flalign*}
+As usual, $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise. % The average fitness of the
+% population is given by $\Phi$, $\Phi = \sum_i x_i f_i$, serving as a normalizing constant for
+% the discrete replicator equation: $x'_i = \frac{x_i f_i}{\Phi}$.
 
-In their series of papers on language evolution, Nowak and colleagues did not only consider expressivity but also recognized the central role of the fidelity by which language is transmitted.
-Among others, linguistic production can be prone to errors, states or messages may be percieved incorrectly, and multiple languages may be compatible with the data learners are exposed to.
-These sources of uncertainty introduce variation in their transmission from one generation to the next which can be likened to mutation to the effect that a type's offspring may adopt a different type than that of its parent as a result of transmission perturbations.
-Importantly, this variation should be relative to the learnability of a type under these perturbations (instead of, e.g., being a constant that is equal across all types as in \citealt{nowak+etal:2002}).
-For this purpose, we turn to a different strand of research in cultural evolution: iterated learning.
+\bigskip
+
+In their series of papers on language evolution, Nowak and colleagues did not only consider
+expressivity but also recognized the central role of the fidelity by which language is
+transmitted. Among others, linguistic production can be prone to errors, states or messages may
+be percieved incorrectly, and multiple languages may be compatible with the data learners are
+exposed to. These sources of uncertainty introduce variation in their transmission from one
+generation to the next which can be likened to mutation to the effect that a type's offspring
+may adopt a different type than that of its parent as a result of transmission
+perturbations. Importantly, this variation should be relative to the learnability of a type
+under these perturbations (instead of, e.g., being a constant that is equal across all types as
+in \citealt{nowak+etal:2002}). For this purpose, we turn to a different strand of research in
+cultural evolution: iterated learning.
 
@@
 
-\subsection{Learnability}\label{sec:learnability}
-Iterated learning is a process in which the behavior of one individual serves as learning input for another.
-This learner, upon acquisition of this behavior, then goes on to produce behavior that serves as input for a new learner.
-This process can be thought of as a progression through chains of parents and children; the parent produces linguistic data from which the child infers a language.
-The latter, now a parent, goes on to produce data for a new generation of learners.
-Following \citet{griffiths+kalish:2007} we model iterated learning as a repeated process of Bayesian inference in which learners combine the likelihood of a type producing the received learning input with prior inductive biases.
+\subsection{Learnability}
+\label{sec:learnability}
 
-Due to the pressure towards learnability it exherts, iterated learning generally leads to simpler and more regular languages (see \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016} for recent surveys).
-Importantly, experimental and mathematical results have been taken to suggest that the outcome of this process reflects learners' inductive biases.
-In a Bayesian setting these biases can be codified in a prior $P \in \Delta(T)$, which can be thought of as the amount of data a learner would require in order to adopt a language (cf.
-\citealt[450]{griffiths+kalish:2007}).
-Or, in our case, a combination of a lexicon and its use.
-The extent of the prior's influence has been shown to heavily depend on the learning strategy assumed to underly the inference process.
-On the one hand, early simulation results suggested that weak biases could be magnified by exposing learners to only small data samples (e.g.
-in \citealt{brighton:2002}).
-On the other, \citeposs{griffiths+kalish:2007} mathematical characterization showed that iterated learning converged to the prior, i.e., the resulting distribution over languages corresponds to the learners' prior distribution and is not influenced by the amount of input given to them.
-This difference in predictions can be traced back to differences in the selection of hypotheses from the posterior.
-Griffith \& Kalish's convergence to the prior holds for learners that sample from the posterior.
-More deterministic strategies such as the adoption of the type with the highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase the influence of both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}.
-In the following, we parametrize the posterior by $l \geq 1$.
-In doing so, we obtain a range of learning strategies between posterior sampling and MAP.
-When $l = 1$ learners sample from the posterior and the learners propensity to maximize the posterior grows as $l$ increases.
+Iterated learning is a process in which the behavior of one individual serves as learning input
+for another. This learner, upon acquisition of this behavior, then goes on to produce behavior
+that serves as input for a new learner. This process can be thought of as a progression through
+chains of parents and children; the parent produces linguistic data from which the child infers
+a language. The latter, now a parent, goes on to produce data for a new generation of
+learners. Following \citet{griffiths+kalish:2007} we model iterated learning as a repeated
+process of Bayesian inference in which learners combine the likelihood of a type producing the
+received learning input with prior inductive biases.
+
+Due to the pressure towards learnability it exherts, iterated learning generally leads to
+simpler and more regular languages (see \citealt{kirby+etal:2014} and
+\citealt{tamariz+kirby:2016} for recent surveys). Importantly, experimental and mathematical
+results have been taken to suggest that the outcome of this process reflects learners'
+inductive biases. In a Bayesian setting these biases can be codified in a prior
+$P \in \Delta(T)$, which can be thought of as the amount of data a learner would require in
+order to adopt a language (cf. \citealt[450]{griffiths+kalish:2007}). Or, in our case, a
+combination of a lexicon and its use. The extent of the prior's influence has been shown to
+heavily depend on the learning strategy assumed to underly the inference process. On the one
+hand, early simulation results suggested that weak biases could be magnified by exposing
+learners to only small data samples (e.g. in \citealt{brighton:2002}). On the other,
+\citeposs{griffiths+kalish:2007} mathematical characterization showed that iterated learning
+converged to the prior, i.e., the resulting distribution over languages corresponds to the
+learners' prior distribution and is not influenced by the amount of input given to them. This
+difference in predictions can be traced back to differences in the selection of hypotheses from
+the posterior. Griffith \& Kalish's convergence to the prior holds for learners that sample
+from the posterior. More deterministic strategies such as the adoption of the type with the
+highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase
+the influence of both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}. In
+the following, we parametrize the posterior by $l \geq 1$. In doing so, we obtain a range of
+learning strategies between posterior sampling and MAP. When $l = 1$ learners sample from the
+posterior and the learners propensity to maximize the posterior grows as $l$ increases.
 

### 2016-11-20 paper/paper-v2.tex
paper rewrite section 2 first pass finished

@@
 \end{flalign*}
-As usual, $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise. % The average fitness of the
-% population is given by $\Phi$, $\Phi = \sum_i x_i f_i$, serving as a normalizing constant for
-% the discrete replicator equation: $x'_i = \frac{x_i f_i}{\Phi}$.
-
-\bigskip
-
-In their series of papers on language evolution, Nowak and colleagues did not only consider
-expressivity but also recognized the central role of the fidelity by which language is
-transmitted. Among others, linguistic production can be prone to errors, states or messages may
-be percieved incorrectly, and multiple languages may be compatible with the data learners are
-exposed to. These sources of uncertainty introduce variation in their transmission from one
-generation to the next which can be likened to mutation to the effect that a type's offspring
-may adopt a different type than that of its parent as a result of transmission
-perturbations. Importantly, this variation should be relative to the learnability of a type
-under these perturbations (instead of, e.g., being a constant that is equal across all types as
-in \citealt{nowak+etal:2002}). For this purpose, we turn to a different strand of research in
-cultural evolution: iterated learning.
-
-%errors were considered in N+K:1999, mutation in N+etal:2002
+As usual, $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.
 
@@
 
-Iterated learning is a process in which the behavior of one individual serves as learning input
-for another. This learner, upon acquisition of this behavior, then goes on to produce behavior
-that serves as input for a new learner. This process can be thought of as a progression through
-chains of parents and children; the parent produces linguistic data from which the child infers
-a language. The latter, now a parent, goes on to produce data for a new generation of
-learners. Following \citet{griffiths+kalish:2007} we model iterated learning as a repeated
-process of Bayesian inference in which learners combine the likelihood of a type producing the
-received learning input with prior inductive biases.
+Languages are shaped not only by functionalist forces towards greater expressivity. Another
+important factor is the fidelity by which language is transmitted. Among others, linguistic
+production can be prone to errors, states or messages may be perceived incorrectly, and
+multiple languages may be compatible with the data learners are exposed to. These sources of
+uncertainty introduce variation in their transmission from one generation to the next. Biases
+or systematic noise in the iterated transmission process can influence language evolution
+substantially.
 
-Due to the pressure towards learnability it exherts, iterated learning generally leads to
-simpler and more regular languages (see \citealt{kirby+etal:2014} and
-\citealt{tamariz+kirby:2016} for recent surveys). Importantly, experimental and mathematical
-results have been taken to suggest that the outcome of this process reflects learners'
-inductive biases. In a Bayesian setting these biases can be codified in a prior
-$P \in \Delta(T)$, which can be thought of as the amount of data a learner would require in
-order to adopt a language (cf. \citealt[450]{griffiths+kalish:2007}). Or, in our case, a
-combination of a lexicon and its use. The extent of the prior's influence has been shown to
-heavily depend on the learning strategy assumed to underly the inference process. On the one
-hand, early simulation results suggested that weak biases could be magnified by exposing
-learners to only small data samples (e.g. in \citealt{brighton:2002}). On the other,
-\citeposs{griffiths+kalish:2007} mathematical characterization showed that iterated learning
-converged to the prior, i.e., the resulting distribution over languages corresponds to the
-learners' prior distribution and is not influenced by the amount of input given to them. This
-difference in predictions can be traced back to differences in the selection of hypotheses from
-the posterior. Griffith \& Kalish's convergence to the prior holds for learners that sample
-from the posterior. More deterministic strategies such as the adoption of the type with the
-highest posterior probability, so-called {\it maximum a posterior estimation} (MAP), increase
-the influence of both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}. In
-the following, we parametrize the posterior by $l \geq 1$. In doing so, we obtain a range of
-learning strategies between posterior sampling and MAP. When $l = 1$ learners sample from the
-posterior and the learners propensity to maximize the posterior grows as $l$ increases.
+In biological evolution, where types are expressed genetically, transmission infidelity comes
+into the picture through infrequent and mostly random genetic mutations. But an agent's lexicon
+and pragmatic reasoning behavior is not inherited genetically. They need to be learned from
+observation. Concretely, when agents of type $j$ want to adopt or imitate the linguistic
+behavior of type $i$, they observe the linguistic behavior of type $i$ and need to infer what
+their type is from that. Iterated learning is a process in which languages are learned repeatedly from the observation
+of linguistic behavior of agents who have acquired the language from observation and inference
+in the same way before. In the simplest case, there is a single teacher and a single
+learner. After sufficient training, each learner becomes a teacher and produces behavior that
+serves as input for a new learner. Due to the pressure towards learnability it exerts, iterated
+learning generally leads to simpler and more regular languages (see \citealt{kirby+etal:2014}
+and \citealt{tamariz+kirby:2016} for recent surveys).
 
-The data learners are exposed to is described by a set $D$ containing sequences of state-message pairs, e.g., $\tuple{\tuple{s_i,m_v},...\tuple{s_j,m_w}}$.
-These are sequences of language use witnessed by learners, the length of which we denote by $k$.
-%Put differently, each datum $d \in D$ contains $k$ members of the set $\{\tuple{s_i,m_v} | s_i \in S, m_v \in M\}$.
+Following \citet{griffiths+kalish:2007} we model language acquisition as a process of Bayesian
+inference in which learners combine the likelihood of a type producing the received learning
+input with prior inductive biases. Experimental and mathematical results on iterated learning
+suggest that the outcome of this process reflects learners' inductive biases
+\citep[e.g.][]{kirby+etal:2014}. In a Bayesian setting these biases can be codified in a prior
+$P \in \Delta(T)$, which reflect the amount of data a learner requires to faithfully acquire
+the language of the teacher \citep[c.f.][450]{griffiths+kalish:2007}. The extent of the prior's
+influence has been shown to heavily depend on the learning strategy assumed to underly the
+inference process. On the one hand, early simulation results suggested that weak biases could
+be magnified by exposing learners to only small data samples (e.g. in
+\citealt{brighton:2002}). On the other, \citeposs{griffiths+kalish:2007} mathematical
+characterization showed that iterated learning converged to the prior, i.e., the resulting
+distribution over languages corresponds to the learners' prior distribution and is not
+influenced by the amount of input given to them. This difference in predictions can be traced
+back to differences in the selection of hypotheses from the posterior. Griffith \& Kalish's
+convergence to the prior holds for learners that sample from the posterior. More deterministic
+strategies such as the adoption of the type with the highest posterior probability, so-called
+{\it maximum a posterior estimation} (MAP), increase the influence of both the prior and the
+data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we use a paremeter
+$l\ge1$ to modulate between posterior sampling and the MAP strategy. When $l = 1$ learners
+sample from the posterior and the learners propensity to maximize the posterior grows as $l$
+increases.
 
-%To summarize, the parametrized posterior, $P(d|t)^l$, is obtained from combining the prior $P(t)$ and the likelihood $P(d|t)$.
-Since types are combinations of a lexicon and its use, the latter can be computed straightforwardly from a type's production behavior.
-We combine the above with the replicator dynamics by codifying iterated learning in a transition matrix $Q$, where $Q_{ij}$ indicates the probability that a child of a parent of type $i$ adopts type $j$.
-This quantity is proportional to the probability of $i$ producing the learning data and that of inferring $j$ given the data:
-
-\[
- Q_{ij} \propto \sum_{d \in D} P(d|t_i) F(t_j,d), \; \text{where } F(t_j,d) \propto P(t_j|d)^l \text{ and } P(t_j|d) \propto P(t_j) P(d|t_j).
-\]
+Let $D$ be the set of possible data that learners may be exposed to. This set $D$ contains all
+sequences of length $k$ of state-message pairs, e.g.,
+$\tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$. The number $k$ of observations is another model
+parameters. As $k$ increases, learners have more data to base their inference on and so tend to
+recover the true types that generated a given sequence with higher probability. The mutation
+matrix $Q$ of the replicator mutator dynamics in (\ref{eq:RMD_discrete}) can then be defined as
+follows: $Q_{ji}$ is the probability that a learner acquires type $i$ when learning from an
+agent of type $j$. The learner observes one length-$k$ sequence $d$ of state-message pairs, but
+the probability $P(d \mid t_j)$ with which sequence $d = \tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$ is observed depends on type $j$'s
+behavior:
+\begin{align*}
+  P(d = \tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}} \mid t_j) = \prod_{i = 1}^k S_{n_j}(m_i
+  \mid t_i; L_{j})\,,
+\end{align*}
+where, as before, $n_j$ is $j$'s pragmatic reasoning type and $L_j$ is $j$'s lexicon. For a
+given observation $d$, the probability of acquiring type $i$ is $F(t_i \mid d)$, so that:
+\mf{check if we need ``proportional to'' here or whether ``equals to'' is okay}
+\begin{flalign*}
+  Q_{ji} \propto \sum_{d \in D} P(d \mid t_j) F(t_i \mid d)\,.
+\end{flalign*}
+The acquisition probability $F(t_i \mid
+d)$ given datum $d$ is obtained by probability matching $l = 1$ or a tendency towards choosing
+the most likely type $l > 1$ from the posterior distribution $P(\cdot \mid d)$ over types given
+the data, which is calculated by Bayes' rule:
+\begin{flalign*}
+  & F(t_i \mid d) \propto P(t_i \mid d)^l \; \text{ and }\\
+  & P(t_i \mid d) \propto P(t_i) P(d \mid t_i)\,.
+\end{flalign*}
 
 \subsection{Model summary}
-Drawing from past research, we argued that expressivity and learnability are central to the cultural evolution of language.
-We propose these components to be modelled respectively as communicative efficiency-relative replication and (iterated) Bayesian learning.
-Taken together their interaction is described by the replicator-mutator dynamics \citep{hofbauer+sigmund:2003}:
 
-\[ 
-\hat{x}_i = \sum_j Q_{ji} \frac{x_jf_j}{\Phi}
-\]
-
-The units that the dynamics operate on are a combination of a lexicon and a degree of pragmatic sophistication determining its use.
-We call this combination a type.
-A type's expressivity depends on its communicative efficiency within a population while its learnability depends on the fidelity by which it is inferred by new generations of learners.
-The learners' task is consequently to perform a joint inference over types of linguistic behavior and lexical meaning.
-With this model at hand we turn to the analysis of the lack of lexicalization of productive pragmatic inferences in a case study on scalar implicatures.
+Expressivity and learnability are central to the cultural evolution of language. These
+components can be modelled, respectively, as replication based on a measure of fitness in terms
+of communicative efficiency and iterated Bayesian learning. Their interaction is described by
+the discrete time replicator mutator dynamics in (\ref{eq:RMD_discrete}), repeated here:
+\begin{align*}
+  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}\,.
+\end{align*}
+This equation defines the frequency $x'_i$ of type $i$ at the next time step, based on its
+frequency $x_i$ before the update, its fitness $f_i$ and the probability that a learner infers
+$i$ when observing the behavior of a type-$j$ agent. Fitness-based selection can be thought of
+as biological (fitness as expected relative number of offspring) or cultural (fitness of
+likelihood of being imitated or repeated). The types that the dynamic operates on are pairs
+consisting of a lexicon and a pragmatic use pattern. A type's expressivity depends on its
+communicative efficiency within a population while its learnability depends on the fidelity by
+which it is inferred by new generations of learners. The learners' task is consequently to
+perform a joint inference over types of linguistic behavior and lexical meaning.
 

### 2016-11-24 paper/paper-v2.tex
draft v2 changes, cleaner (temporary) plot and data files in paper-folder

@@
 \newcommand{\mf}[1]{\textcolor{Red}{[MF: #1]}} 
+\newcommand{\tb}[1]{\textcolor[rgb]{.8,.33,.0}{[TB: #1]}}% prints in orange
+
 \usepackage{blkarray}
@@
 that the speaker does not own all of them. This is because, if the speaker had them all, he
-could have used the stronger word \emph{all} instead of \emph{some} in his utterance and
+could have used the stronger word \emph{all} instead of \emph{some} in his utterance, and
 thereby would have made a more informative statement. Scalar implicatures, especially the
@@
 clear majority is that a word like \emph{some} is underspecified to mean \emph{some and maybe
-  all} and that the enrichment to \emph{some but not all} is part of some regular enrichment
+  all} and that the enrichment to \emph{some but not all} is part of some regular 
 process with roots in pragmatics.
@@
 If this majority view is correct, the question arises how such a division of labor between
-semantics and pragmatics could have evolved. Models of meaning evolution abound. There are
-simulation-based models studying the evolution of meaning in populations of communicating
+semantics and pragmatics could have evolved. Models of language evolution abound. There are
+simulation-based models studying the evolution of language in populations of communicating
 agents
 \citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
-and there are mathematical models of meaning evolution, mostly coming from in game theory
+and there are mathematical models of language evolution, mostly coming from game theory
 \citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
 work has focused on explaining basic properties such as compositionality and combinatoriality
-\citep[e.g.][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax}.
-But
+\citep[e.g.][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax}, but
 little attention has been paid to the interaction between conventional meaning and pragmatic
@@
 room in such a purely extensional approach to address the semantics-pragmatics division
-directly. We would need to look at richer representations of cognizing agents and their
+directly. Instead, we would need to look at richer representations of cognizing agents and their
 communicative interaction.
 
-To fill this gap, we here spell out a model of the co-evolution of conventional meaning and
+To fill this gap, we spell out a model of the co-evolution of conventional meaning and
 pragmatic reasoning types. The objects of replication and selection are pairs of lexical
@@
 \citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
-and selection is described by the \emph{replicator mutator dynamic}, a general and established
+and selection are described by the \emph{replicator mutator dynamic}, a general and established
 model of evolutionary change in large and homogeneous populations
@@
 communicative efficiency and (ii) possible infidelity in the transmission of linguistic
-knowledge, such as from inductive learning biases or systematic perceptual errors. Considering
+knowledge, caused by factors such as inductive learning biases or systematic perceptual errors. Considering
 transmission of linguistic knowledge is important because neither semantic meanings nor
-pragmatic usage patterns are directly observable. Language learners have to infer these
-unobservables from the observable behavior in which they result. We formalize this acquisition
+pragmatic usage patterns are directly observable. Instead, language learners have to infer these
+unobservables from the observable behavior in which they result. We formalize this 
 process as a form of Bayesian inference. Our approach thereby contains a well-understood model
@@
 with functional selection, here formalized as the most versatile dynamic from evolutionary game
-theory, the replicator dynamic
+theory; the replicator dynamic
 \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this model.
@@
 show that inductive learning biases of Bayesian learners that favor simpler lexical meanings
-can lead to the desired outcome. Additionally, we show that a similar outcome can be achieved
-without assuming any cognitive biases, simply as an epiphenomenon of systematic disturbances from
+can lead to this outcome. Additionally, we show that a similar outcome can be reached 
+without assuming any cognitive biases, but simply as an epiphenomenon of systematic disturbances from
 environmental factors. This formal results highlights the frequently overlooked possibility
@@
 We see the main contribution of this work as conceptual and technical, not as a definite answer
-to the question why scalar implicatures emerged. The work here rather demonstrates how current
+to the question why scalar implicatures emerge. The work here rather demonstrates how current
 probabilistic cognitive modeling of language use and evolutionary modeling can be fruitfully
@@
 The emergence and change of linguistic structure is influenced by many intertwined
-factors. These range from biological and socio-ecological to cultural ones \citep{steels:2011}
-\mf{more references?}. Social and ecological pressures determine communicative needs, while
+factors. These range from biological and socio-ecological to cultural ones \citep{benz+etal:2005b,steels:2011,tamariz+kirby:2016}. Social and ecological pressures determine communicative needs, while
 biology determines the architecture that enables and constrains the means by which they can be
@@
 pressure towards communicative efficiency is not the only force that shapes
-language. Learnability is another. Natural languages need to be learnable to survive their
+language. Learnability is another as natural languages need to be learnable to survive their
 faithful transmission across generations. Clearly, an unlearnable code will not make it past
-the one happy fellow who invented it. Moreover, even small biases implicit in the acquisition
-of a language can built up and have quite striking effects on an evolving language in a process
+the one happy fellow who invented it. Importantly, even small biases implicit in acquisition
+can build up and have quite striking effects on an evolving language in a process
 of iterated learning
-\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}. In
-sum, natural languages are pressured for expressivity and learnability. But expressivity and
-learnability may pull in opposite directions. The opposition becomes particularly clear when
+\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.
+
+While natural languages are pressured for both expressivity and learnability these forces may pull in opposite directions. The opposition becomes particularly clear when
 considering the extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). A language with a
@@
 $i$ before the update, (ii) the fitness $f_i$ of each type $i$ before the update, and (iii) the
-probability $Q_{ji}$ that an agent who wants to imitate, adopt or learn the type of an agent
+probability $Q_{ji}$ that an agent who wants to imitate, adopt, or learn the type of an agent
 with type $j$ ends up acquiring type $i$:
@@
 becomes most transparent when we consider an equivalent formulation in terms of a step-wise
-application of the discrete-time replicator dynamic \citep{TaylorJonker1978:Evolutionary-St}
-and subsequent multiplication with a mutation matrix $Q$ of the initial population vector
-$\vec{x}$:
+application of the discrete-time replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$
+and its subsequent multiplication with a mutation matrix $Q$:
+
 \begin{align}
@@
 selection in which the relative frequency of type $i$ will increase with a gradient
-proportional to its average fitness in the population. The replicator dynamic is popular and
+proportional to its average fitness in the population. This dynamic is popular and
 versatile because it can be derived from many abstract processes of biological and cultural
@@
 have an example from a lecture ready at hand; it's a simple coordination game in a
-one-population setting.}
+one-population setting.} \tb{Yes, I think most readers would appreciate it. We can always cut it out again if the reviewers prefer more compression.}
 
@@
 a prisoner's dilemma. \mf{maybe good to refer back to the example from before if there was
-  one?} For our purposes here, types are identified by their cognitive make-up. Since we are
+  one?} \tb{Yes!} For our purposes here, types are identified by their cognitive make-up. Since we are
 interested in the question under which conditions processes of cultural evolution will favor
@@
 semantics with only a lower-bound in $L_5$. \mf{can we not give these guys more mnemonic
-  names?}
+  names?} \tb{``bound'' and ``lack''? They would need to be pretty short to not be too disruptive as an inline indices.}
 \begin{align*}
@@
 hearers and speakers respectively: \mf{is it a problem that $S$ denotes speakers and the set of
-states?}\mf{Literal speaker definition had $\alpha$ but that has no effect if we have Boolean
-semantics; I left this out.}
+states?} \tb{I think that's fine.
+$S_n(\cdot \mid \cdot)$ is not used beyond this section and the set $S$ only by mention of its members.
+Otherwise we can, e.g., relabel $M$ as $F$(orms) and use $M$(eanings) instead of $S$(tates).
+I'm fine with either alternative.}
 \begin{flalign}
@@
 for a state with equal probability and prefers a true message over a false one. The formulation
-in (\ref{s:level0}) allows also false messages to be sent with a small positive probability in
+in (\ref{s:level0}) also allows false messages to be sent with a small positive probability in
 analogy to the definition of pragmatic speakers in probabilistic pragmatic models. This is
 necessary to guarantee a mutation matrix with only positive entries (see below). A literal
-speaker with lexicon $L_4$ or $L_5$ produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
+speaker with either lexicon $L_4$ or $L_5$ produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
 \ssome;L_{4,5}) \approx .73$. The probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
 \sall;L_4) \approx .27$ for $L_4$ and $S_{0}(\msome \mid
-\sall;L_4) \approx .5$ for $L_5$. \mf{I realize that the probability of producing a false
+\sall;L_5) \approx .5$ for $L_5$. \mf{I realize that the probability of producing a false
   message for literal speakers is actually rather high. That's iffy because our pragmatic types
@@
 something about this? Use definition $S_{0}(m \mid s;L) \propto L_{sm} + 0.001$? Make sure we
-only look at parameters for pragmatic types that make them not much better than the literal guys?}
+only look at parameters for pragmatic types that make them not much better than the literal guys?} \tb{That's true.
+We could also just let them soft-maximize.
+Let's decide on this via email.
+I'll adapt the simulation results from the next section according to what we decide.}
 
@@
 meaning. The transformation with $\alpha$ allows to differentiate theoretically interesting
-speaker production behavior. To see this, notice that for two states and two messages (the
-minimal scenario we will investigate), the possible values for $H_{0}(s|m;L)$ are restricted to
-zero, $0.5$ and one. Consider the case of $H_0$ with $L_5$, which assigns the following
-probabilities to states after hearing messages:
-\begin{align*}
-  H_{0}(\messg \mid \state; L_5) & = \begin{blockarray}{lcc}
-    & \mygray{\ssome} & \mygray{\sall} \\
-    \begin{block}{l[cc]}
-      \mygray{\msome} & 0.5 & 0.5 \\
-      \mygray{\mall}  & 0 & 1 \\
-    \end{block}
-  \end{blockarray}
-\end{align*}
-A level-1 speaker's utilities are then:
-\begin{align*}
-  \begin{blockarray}{lcc}
-    & \mygray{\msome} & \mygray{\mall} \\
-    \begin{block}{l[cc]}
-      \mygray{\ssome} & 0.5^\alpha & 0^\alpha \\
-      \mygray{\sall}  & 0.5^\alpha & 1^\alpha \\
-    \end{block}
-  \end{blockarray}
-\end{align*}
-As the soft-max choice rule is sensitive to differences in utilities, no matter what $\lambda$
-is we will get $S_1(\msome \mid \ssome; L_5) = S_1(\mall \mid \sall; L_5)$ if $\alpha =
-1$. This means that we cannot distinguish pragmatically deviant choices (selecting a less
-informative message instead of a more informative one), from semantically deviant choices
-(selecting a false message instead of a true one). If $\alpha > 1$ pragmatically deviant
+speaker production behavior: If $\alpha > 1$ pragmatically deviant
 choices will get less likely than semantically deviant choices; if $\alpha < 0$ it is the other
-way around. In sum, parameter $\lambda$ regulates the overall tendency to avoid pragmatically
-or semantically deviant choices, $\alpha$ further distinguishes between pragmatically and
+way around. See Appendix \ref{app:sig-example} for an example that elucidates the role of $\alpha$ \tb{Did you mean it like this?}.
+
+In sum, parameter $\lambda$ regulates the overall tendency to avoid pragmatically or semantically deviant choices, $\alpha$ further distinguishes between pragmatically and
 semantically deviant behavior. Variation in $\alpha$ is only necessary for the technical
-extension to environmental transmission noise in Section~XYZ. \mf{I would suggest to just give
-  the intuitive summary here and have a little appendix paragraph with the example. What do you
-say?}
+extension to environmental transmission noise in Section \ref{subsec:noise}.
 
@@
 peers. Under a cultural interpretation, the picture is that agents themselves strive towards
-communicative success and therefore adapt or revise their behavior occasionally to achieve a
+communicative success and therefore occasionally adapt or revise their behavior to achieve a
 higher communicative success (see \citealt[\S3.3]{benz+etal:2005b} for discussion).
@@
 The replicator equation gives us the means to make the ensuing dynamics precise, without
-necessarily committing to a biological or cultural interpretation. The proportion of types in a
+necessarily committing to a biological or cultural interpretation. As above, the proportion of types in a
 given population is codified in a vector $\vec{x}$, where $x_i$ is type $i$'s proportion. The
-fitness of type $i$ is type $i$'s average expected communicative success, or \emph{expected
-  utility} (EU), given the current frequencies of types in the current population:
+fitness of type $i$ is its average expected communicative success, or \emph{expected
+  utility} (EU), given the frequencies of types in the current population:
 \begin{align*}
@@
 \begin{align*}
-  \text{EU}(t_i,t_j) = \nicefrac{1}{2} \, U_S(t_i,t_j) + \nicefrac{1}{2} \, U_H(t_i,t_j)\,,
+  \text{EU}(t_i,t_j) = \nicefrac{1}{2} \, \text{EU}_S(t_i,t_j) + \nicefrac{1}{2} \, \text{EU}_H(t_i,t_j)\,,
 \end{align*}
 where $\text{EU}_S(t_i,t_j)$ and $\text{EU}_H(t_i,t_j)$ are the expected utilities for $i$ as a
-speaker and as a hearer when communicating with $j$, defined as follows where $n_i$ and $n_j$
+speaker and as a hearer when communicating with $j$, defined as follows, where $n_i$ and $n_j$
 are type $i$'s and $j$'s pragmatic reasoning types and $L_i$ and $L_j$ are their lexica:
@@
 In biological evolution, where types are expressed genetically, transmission infidelity comes
-into the picture through infrequent and mostly random genetic mutations. But an agent's lexicon
+into the picture through infrequent and mostly random genetic mutations. However, an agent's lexicon
 and pragmatic reasoning behavior is not inherited genetically. They need to be learned from
@@
 of linguistic behavior of agents who have acquired the language from observation and inference
-in the same way before. In the simplest case, there is a single teacher and a single
-learner. After sufficient training, each learner becomes a teacher and produces behavior that
+before as well. In the simplest case there is a single teacher and a single
+learner. After sufficient training the learner becomes a teacher and produces behavior that
 serves as input for a new learner. Due to the pressure towards learnability it exerts, iterated
@@
 {\it maximum a posterior estimation} (MAP), increase the influence of both the prior and the
-data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we use a paremeter
+data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we use a parameter
 $l\ge1$ to modulate between posterior sampling and the MAP strategy. When $l = 1$ learners
-sample from the posterior and the learners propensity to maximize the posterior grows as $l$
+sample from the posterior. The learners propensity to maximize the posterior grows as $l$
 increases.
@@
 Let $D$ be the set of possible data that learners may be exposed to. This set $D$ contains all
-sequences of length $k$ of state-message pairs, e.g.,
+sequences of state-message pairs of length $k$, e.g.,
 $\tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$. The number $k$ of observations is another model
@@
 follows: $Q_{ji}$ is the probability that a learner acquires type $i$ when learning from an
-agent of type $j$. The learner observes one length-$k$ sequence $d$ of state-message pairs, but
+agent of type $j$. The learner observes a length-$k$ sequence $d$ of state-message pairs, but
 the probability $P(d \mid t_j)$ with which sequence $d = \tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$ is observed depends on type $j$'s
@@
   P(d = \tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}} \mid t_j) = \prod_{i = 1}^k S_{n_j}(m_i
-  \mid t_i; L_{j})\,,
+  \mid s_i; L_{j})\,,
 \end{align*}
@@
 given observation $d$, the probability of acquiring type $i$ is $F(t_i \mid d)$, so that:
-\mf{check if we need ``proportional to'' here or whether ``equals to'' is okay}
+\mf{check if we need ``proportional to'' here or whether ``equals to'' is okay} \tb{``proportional to'' is needed. ``equal'' would work if for all $d$; $\sum_j P(d \mid t_j) = 1$}
 \begin{flalign*}
@@
 
-This kind of strengthening is captured by the linguistic behavior of pragmatic types introduced in \S\ref{sec:languages+use}: A pragmatic hearer who reasons about a speaker's use of a message involving a weak scalar alternative will associate it more strongly with upper-bounded states than with ones in which stronger alternatives hold because these alternatives already unambiguously convey the latter states.
+This kind of strengthening is captured by the linguistic behavior of pragmatic types introduced in \S\ref{sec:languages+use}: A pragmatic hearer who reasons about the use of a message involving a weak scalar alternative will associate it more with a state in which stronger alternatives do not hold.
+This is so because a rational speaker would use a more informative message when in such a state.
 Conversely, a pragmatic speaker will reason about her interlocutor's expected interpretation and use the messages at her disposition accordingly.
 
-Our initial question can now be narrowed to the case of scalar implicatures by asking for a justification for the lack of lexical upper-bounds in weak scalar alternatives.
-That is, why they are regularly selected for over other alternatives such as that of codifying the bound semantically.
+Our initial question about the division of labor between semantics and pragmatics can be narrowed to the case of scalar implicatures by asking for a justification for the lack of lexical upper-bounds in weak scalar alternatives.
+That is, we ask why lexical meanings that lack upper-bounds and convey it pragmatically are regularly selected for over alternatives such as that of codifying the bound semantically.
 More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} and {\em big} were truth-conditionally incompatible with stronger alternatives such as {\em hot}, {\em and}, {\em all} and {\em huge}?
-This question is particularly striking considering the number of expressions that license such inferences across languages.
+This question is particularly striking considering the number of expressions that license such inferences across natural languages.
 
@@
 This can happen when embedded in downward-entailing contexts, when the speaker is likely uncertain about whether the upper bounded reading is true, or when the distinction between an upper-bounded reading and the simple, only lower-bounded reading, is not relevant.
-For instance, if for all the speaker knows `Some students came' but she doesn't know whether all came, then the use of {\em some} lacking an upper-bound succinctly conveys her uncertainty.
+For instance, if for all the speaker knows `Some students came' but she does not know whether all came, then the use of {\em some} lacking an upper-bound succinctly conveys her uncertainty.
 This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: Should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then this is preferred over expressing it overtly through longer expressions, e.g., by saying {\em some but not all} explicitly.
@@
 Instead, the systematicity and typological spread of scalar implicatures together with the observation that monomorphemic expressions that lexically rule out stronger alternatives are unattested across languages (\citealt[252-267]{horn:1984}, \citealt{horn:1972,traugott:2004,vdAuwera:2010}) suggests that other forces may be at play.
-In what follows we investigate the hypothesis that the lack of lexicalization of scalar inferences may be accounted for by the relative representational simplicity of lexical meanings lacking an upper-bound over those that explicitly codify it.
-This difference is reflected in a learning bias towards more compressed lexical representation, i.e., in a preference of learners for simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
+In what follows we investigate the predictions of our model under the assumption that the lack of lexicalization of scalar inferences may be accounted for by the relative representational simplicity of lexical meanings lacking an upper-bound over those that explicitly codify it.
+This difference is reflected in a learning bias towards more compressed lexical representations.
+That is, in a preference of learners for simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
 
@@
 Note however that we do not represent the contrast between lexical representations explicitly.
-Instead, the bias is directly encoded in the learners' prior over types.
-In principle this difference could be made precise with an adequate representational language, e.g., through measures over representational complexity such as minimal description length.
+Instead, the bias is directly encoded in the learners' prior over types.\footnote{In principle this difference could be made precise with an adequate representational language, e.g., through measures over representational complexity such as minimal description length.
 There is a growing effort to develop such empirically  testable  representational  languages.
@@
 \citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012a} and references therein).
-We think that our assumption is well-warranted as a working hypothesis and decide against such an enrichment at present in order to focus on the effects of linguistic pressures predicted the model instead.
+At present we decide against such an enrichment in favor of a stronger focus on the general predictions of the model and the interaction of its components.}
 
 \subsection{Analysis}
-We analyze the model's predictions in populations of types with two signaling behaviors; literal and pragmatic.
-The former correspond to level $0$ reasoners who only take their lexica into consideration and the latter to level $1$ reasoners.
+We analyze the model's predictions in populations of types with one of the two signaling behaviors introduced earlier; literal or pragmatic.
+The former correspond to level $0$ reasoners and the latter to ones of level $1$.
 Higher level reasoning is not required to derive scalar implicatures from the lexica we consider here, nor do they leave room for substantial pragmatic refinement.
@@
 The space of possible lexica is given in Table \ref{tab:lexica}.
-These $(2,2)$-Boolean matrices are the simplest ones that allow us to make the contrast between the presence or abscence of an upper-bound and the use of scalar implicatures precise.
-One may think of the state corresponding to the first row of any such lexicon as a ``some but not all''-state and the second as an ``all''-state.
+These $(2,2)$-Boolean matrices are the simplest ones that allow us to make the contrast between the presence or absence of an upper-bound and the use of scalar implicatures precise.
+As illustrated in Section \ref{sec:languages+use}, one may think of the state corresponding to the first row of any such lexicon as a ``some but not all''-state, $\ssome$, and the second as an ``all''-state, $\sall$.
 The literal meaning of weak scalar expressions such as English {\em some} then corresponds to a message true of both rows in these fragments.
-While there are $16$ possible $(2,2)$-lexica, a number of them are identical both in terms of expressivity and the learning bias.
-The competition between such types is determined by the initial configuration of a population.
-However, this can be obscured when averaging across simulations.
-We focus on this smaller representative subset as simulations conducted with the full space confirm that the general results reported here do not hinge on this choice.
+While there are $16$ possible $(2,2)$-lexica, a number of them are identical both in terms of expressivity and on whether they codify upper-bounds lexically.
+The competition between such types is therefore determined by the initial configuration of a population.
+This fact can be obscured when averaging across outcomes.
+We therefore focus on this smaller representative subset.
+Simulations conducted with the full space confirm that the general results reported here do not hinge on this choice.
 
-Lexica $L_1$ to $L_3$ are not optimal for communication because they assign the same state to all their messages.
-This failure to be able to associate a single form to a state inevitably leads to a communicative disadvantage in their use.
+Lexica $L_1$ to $L_3$ are not optimal for communication because they assign all their messages to the same state(s).
+This failure to be able to associate a state to single a form inevitably leads to a communicative disadvantage in their use.
 $L_4$ and $L_5$ are our target lexica.
-They codify upper-bounded semantics for the message corresponding to the first matrix's column and a lack thereof, respectively.
-Lastly, $L_6$ is similar to $L_5$ in that two messages are true of the same state but differs from it in assigning upper-bounded semantics to the first column's message.
+The former assings upper-bounded semantics to \msome (the first matrix's column) whereas the latter does not.
+Lastly, $L_6$ is similar to $L_5$ in that one message is true of the same state but differs from it in assigning upper-bounded semantics to \msome.
 
@@
 Almost, because there may be slight differences between the probability with which speakers would (erroneously) use a semantically false description and the probability with which speakers would (erroneously) use a pragmatically suboptimal description.
-Due to this possibly marginal difference between pragmatic $L_4$ and $L_5$, the selection of one type over the other is expected to mainly depend on their transmission to new learners.
-Things are less clear for literal $L_5$ contrasted with literal/pragmatic $L_4$ as the former has a learning advantage under the inductive bias but is expected to fare worse in communicative terms.
+Due to this possibly marginal difference between pragmatic $L_4$ and $L_5$, the selection of one type over the other is expected to mainly depend on how well each can be transmitted to new learners.
+Things are less clear for literal $L_5$ contrasted with literal/pragmatic $L_4$.
+The former has a learning advantage when learners are biased against upper-bounds but is expected to fare worse in communicative terms.
 
-The dynamics are initialized with an arbitrary distribution over types, constituting the population's first generation.
+The dynamic is initialized with an arbitrary distribution over types, constituting the population's first generation.
 The results for each parameter setting were obtained from $1000$ independent runs, each consisting of $20$ generations.
 This corresponds to a developmental plateau after which no noteworthy change was registered.
-As specified in \S\ref{sec:learnability}, the learning matrix $Q$ can be obtained by considering all possible state-message sequences of length $k$.
+As specified in \S\ref{sec:learnability}, the mutation matrix $Q$ can be obtained by considering all possible state-message sequences of length $k$.
 Given that this is intractable for large $k$, matrices were approximated by sampling $10$ sequences from each type's production probabilities and a type's children being exposed only to this subset.
@@
 \subsection{Transmission bias}\label{subsec:bias}
-Following our assumption of a preference for simple lexical representations, the prior biases learners against lexica in which a message holds true only of the first row, i.e., against messages that lexicalize an upper-bound that rules out the ``all''-state.
+Following our assumption of an inductive bias that favors simpler lexical representations, the prior biases learners against lexica in which a message holds true only of the first row, i.e., against messages that lexicalize an upper-bound that rules out state \sall.
 All other semantics are assumed to be a priori equally probable.
-This is captured by $P(t_i) \propto n - c \cdot r$, where $n$ is the total number of states and $r$ is the number of messages only true of row $1$ in $t_i$'s lexicon, $c \in [0,1]$.
-Increments in the value of $c$ therefore bring about a stronger bias against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.
+This is captured by $P(t_i) \propto n - c \cdot r$, where $n$ is the total number of states and $r$ is the number of messages only true of \ssome in $t_i$'s lexicon, $c \in [0,1]$.
+Increments in the value of $c$ thereby bring about a stronger bias against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.
 
-According to our hypothesis, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $L_5$-like semantics.
-It is instructive to first inspect the effect of these pressures in isolation.
+Drawing from our preceding discussion, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $L_5$-like semantics.
+However, it is instructive to first inspect the effect of these pressures in isolation.
 For this purpose we focus our attention on three pragmatic types.\footnote{Pragmatic reasoning allows language users to refine their (possibly erroneous) choices.
-Therefore, it is advantageous even for those types that codify an upper-bound lexically.} Pragmatic $L_3$, a type that is lacking in expressivity but is a priori preferred for its lack of upper-bounds.
+Therefore, it is advantageous even for types that codify more lexically.} Pragmatic $L_3$, a type that is lacking in expressivity but is a priori preferred for its lack of upper-bounds.
 Pragmatic $L_4$, a type that is functionally advantageous but biased against.
@@
 
-\paragraph{Expressivity only.} The replicator dynamics are sensitive to $\lambda$ and $\alpha$ as both have a bearing on a type's fitness.
-The influence of the rationality parameter for is depicted in Figure \ref{fig:either-R-or-M}.A.
-The less expressive $L_3$ speakers fare the worse and are influenced the least by change in $\lambda$.
+\paragraph{Expressivity only.} The replicator dynamic is sensitive to $\lambda$ and $\alpha$ as both have a bearing on a type's fitness.
+The influence of $\lambda$ is depicted in Figure \ref{fig:either-R-or-M}.A.
+The less expressive $L_3$ speakers fare the worst and are influenced the least by change in $\lambda$.
 In contrast, low values of $\lambda$ result in a higher proportion of $L_4$ speakers relative to $L_5$.
-This is expected given role of rationality in producing more deterministic behavior in users of $L_5$-like languages.
-As the rationality parameter increases, the functional difference between $L_4$ and $L_5$ is leveled.
+This is expected given the central role of rationality in producing more deterministic behavior in users of $L_5$-like languages.
+Consequently, as the rationality parameter increases the functional difference between $L_4$ and $L_5$ is leveled.
 Overall, the outcome from only a pressure towards expressivity approximates an even share of pragmatic $L_4, L_5$ and $L_6$ types.
 The latter follows the same trajectory as $L_5$ in Figure \ref{fig:either-R-or-M}.A.
-This illustrates the problem we started out with: expressivity alone can not differentiate between (near-)functional equivalents to a degree that justifies the systematic prevalence of $L_5$-like semantics.
+This illustrates an important issue in the evolution of meaning at the semantics-pragmatics interface: expressivity alone can not differentiate between (near-)functional equivalents to a degree that justifies the systematic prevalence of $L_5$-like semantics.
 
@@
 \centering
-\includegraphics[scale=.5]{./only-R-or-M}
+\includegraphics[scale=.5]{./fig1-only-R-or-M}
 \caption{Mean proportions of target types after $20$ generations in $1000$ populations with only a pressure for expressivity in A ($\alpha = 1$) and only for learnability in B ($\alpha =1, \lambda = 30, k = 5$, $l=1$).}
@@
 
-Inspecting these pressures separately not only showcases the influence of some model components but also highlights some of their broader implications.
-First and foremost, neither dynamic comes close to converging to a monomorphic population under most parameter configurations.
+Inspecting these pressures separately not only showcases the contribution of the model's components but also highlights some of their broader implications.
+First and foremost, neither dynamic on its own comes close to converging to a monomorphic population under most parameter configurations.
 For instance, while $L_4$ speakers can come to take over a substantial proportion of the population, this only happens in a restricted range of low degrees of rationality.
-Apart from polymorphy, both pressures make undesirable predictions when considered in isolation.
+Apart from polymorphy, both pressures make undesirable predictions in isolation.
 A pressure only towards expressivity leads to the expulsion of communicatively suboptimal $L_1$, $L_2$ and $L_3$ from the population but can not explain the regular selection of $L_5$-like semantics over either of its functionally similar alternatives.
@@
 \includegraphics[scale=.5]{./fig2-rmd}
-\caption{Mean proportions of target types after $20$ generations in $1000$ populations across bias values $c \in [0,1]$ with $l =1$ in A and $l = 3$ in B ($\alpha =1, \lambda = 20, k = 5$).}
+\caption{Mean proportions of target types after $20$ generations in $1000$ populations across bias values $c \in [0,1]$ with $l =1$ in A and $l = 5$ in B ($\alpha =1, \lambda = 20, k = 5$).}
 \label{fig:cost}
@@
 
-\paragraph{Expressivity and learnability.} Figure \ref{fig:cost} illustrates the effect of the learning bias for posterior sampling (\ref{fig:cost}.A) and slightly more MAP-like learning (\ref{fig:cost}.B) when pressured for both expressivity and learnability.\footnote{\hl{Robert suggests to illustrate the effect with a higher l-value.
-E.g., l = 1 and l = 10}} More detailed results for all types across a sample of $c$-values for $l = 1$ and $l = 3$ are presented in Table \ref{tab:numeric-results}.
+\paragraph{Expressivity and learnability.} Figure \ref{fig:cost} illustrates the effect of the learning bias for posterior sampling (\ref{fig:cost}.A) and slightly more MAP-like learning (\ref{fig:cost}.B) when pressured for both expressivity and learnability.
+More detailed results for all types across a sample of $c$-values for $l = 1$ and $l = 5$ are presented in Table \ref{tab:numeric-results}.
 These results show that a weak bias is sufficient to lead to a selection of $L_5$ over $L_4$.
-As in the outcomes that only considered learnability, this effect increases with the bias' strength provided $L_5$ users are pragmatic.
+As when only learnability is considered, this effect increases with the bias' strength, provided $L_5$ users are pragmatic.
 Importantly, the addition of a pressure towards expressivity magnifies this effect and dampens the proliferation of functionally suboptimal types advantaged by the learning bias.
@@
 \begin{tabular}{l | c c c c c| c c c c c c c c}
-\multicolumn{1}{c}{~} & \multicolumn{5}{c}{$l = 1$} & ~ & \multicolumn{5}{c}{$l = 3$}\\ \hline \hline
+\multicolumn{1}{c}{~} & \multicolumn{5}{c}{$l = 1$} & ~ & \multicolumn{5}{c}{$l = 5$}\\ \hline \hline
   c        &  0  & .1  & .5  & .8  & .9 & ~ & 0 & .1 & .5 & .8 & .9\\ \hline \hline
@@
 prg. $L_3$ & .03 & .03 & .04 & .04 & .04& ~ & 0& 0 & 0 & 0 & 0 \\ 
-prg. $L_4$ & .22 & .22 & .2 & .18 & .17& ~ & .33& .31 & .23 & .15 & .12 \\
-prg. $L_5$ & .22 & .23 & .27 & .3  & .32& ~ & .33& .37 & .54 & .7 & .75 \\
-prg. $L_6$ & .22 & .22 & .2 & .18 & .17& ~ & .33& .31 & .23 & .15 & .12
+prg. $L_4$ & .22 & .22 & .2 & .18 & .17& ~ & .33& .30 & .16 & .07 & .05 \\
+prg. $L_5$ & .22 & .23 & .27 & .3  & .32& ~ & .33& .39 & .68 & .86 & .90 \\
+prg. $L_6$ & .22 & .22 & .2 & .18 & .17& ~ & .33& .30 & .16 & .07 & .05
 \end{tabular}
-\caption{Mean proportions of types in $1000$ populations after $20$ generations across bias values $c \in [0,1]$ with $l =1$ and $l = 3$ ($\alpha = 1, \lambda = 30, k = 5$)}%, $\epsilon < 0.005$.}
+\caption{Mean proportions of types in $1000$ populations after $20$ generations across bias values $c \in [0,1]$ with $l =1$ and $l = 5$ ($\alpha = 1, \lambda = 30, k = 5$)}%, $\epsilon < 0.005$.}
 \label{tab:numeric-results}
@@
 
-Other than the consideration of both pressures, the resulting proportion of pragmatic $L_5$ speakers primarily hinges on three factors.
-First, the degree to which linguistic behavior is deterministic, as it plays a role both for expressivity as well as in producing data that allows learners to discriminate this type from others.
+Other than the involvement of both pressures, the resulting proportion of pragmatic $L_5$ speakers primarily hinges on three factors.
+First, the degree to which linguistic behavior is deterministic.
+This plays a role both for expressivity as well as in producing data that allows learners to discriminate this type from others.
 Second, the inductive bias, which controls the learners preference for simpler lexical representations.
-Lastly, the posterior parameter, which magnifies the effects of the learning bias in tandem with replication.
+Lastly, the posterior parameter, as it magnifies the effects of the learning bias in tandem with replication.
 
-As discussed in relation to Figure \ref{fig:cost}.A, posterior sampling can lead to the incumbency of pragmatic $L_5$.
-However, not even a strong favorable learning bias combined with a pressure for expressivity completely drives out competing types.
-This is not so for more posterior maximizing behavior.
+As showcased by Figure \ref{fig:cost}.A, posterior sampling can lead to the incumbency of pragmatic $L_5$.
+However, not even a strong favorable learning bias combined with a pressure for expressivity completely drives out competing types under this inference strategy.
+This is not so for stronger posterior maximizing learning.
 As shown in Figure \ref{fig:prior-posterior}, the range of bias values within which $L_5$ takes over the population increases with MAP-like learning.
-In other words, the strength of the learning bias required for a given final proportion of $L_5$ speakers strongly depends on learners' inferential strategy.
+Put differently, the strength of the learning bias required for a given final proportion of $L_5$ speakers strongly depends on learners' inferential strategy.
 As for the effect of the other parameters not mentioned so far, changes in sequence length influence the population in a predictable way: smaller values lead to more heterogeneous populations that reflect the learner's prior more faithfully.
-Larger ones lead to more pronounced differences amongst equally preferred types.
+Larger ones lead to more pronounced differences among equally preferred types.
 This is due to the fact that the likelihood that a small sequence was produced by any type is relatively uniform (modulo prior) compared to that of types with lexica $L_1$ - $L_3$ to produce larger sequences with the same state-message combination in contrast to pragmatic speakers of $L_4$ - $L_6$, or literal $L_4$.
+\tb{We need to discuss $\alpha$ here as well if noisy perception is taken out}
 
@@
 \subsection{Discussion}
-Under the assumption of a learning bias for simpler representations, our results suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome communicative pressures and stabilize in a population.
+Under the assumption of a learning bias for simpler lexical representations, our results suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome communicative pressures and stabilize in a population.
 This prediction hinges on three assumptions.
 First, that language is pressured toward both expressivity and learnability.
-Second, that language use is relatively deterministic -- low $\lambda$ renders languages that rely on pragmatics to associate one state to one message too prone to communicative failure and more difficult to learn.
+Second, that language use is relatively deterministic.
 Lastly, that learners prefer simpler over more complex lexical representations.
@@
 Under these conditions the selection of lexical meanings lacking upper-bounds in populations of pragmatic speakers is robust against parameter perturbations.
-This outcome is particularly encouraging in light of other advantages a lack of semantic upper-bounds may confer.
+This outcome is particularly encouraging in light of other advantages a lack of semantic upper-bounds may confer, as discussed at the beginning of Section \ref{sec:si-case-study}.
 
-While non upper-bounded lexical meanings in weak scalar alternatives are predicted by the literature, it not clear to what extent other types should be present in the final population, if at all.
+A lack of upper-bounded in the lexical meaning of weak scalar expressions constitutes the majority view in the literature.
+However, it not clear to what extent other types should be present in the final population.
 It seems reasonable to expect functionally suboptimal types $L_1$, $L_2$ and $L_3$ to be ruled out because they fail to enable their users to communicate effectively.
@@
 In our current setup, it mirrors $L_5$ in enabling for the pragmatic strengthening of a message that does not codify an upper-bound lexically.
-However, this is achieved by ruling out the ``some but not all''-state and not, as with scalar implicatures, the ``all''-state.
+However, this is achieved by ruling out $\ssome$ and not, as with scalar implicatures, $\sall$.
 $L_6$ speakers therefore strengthen a ``some''-message to convey something paraphrasable as `some but not [some but not all]'.
-The current representation of lexica as Boolean matrices is blind to this anomaly without further restrictions.} Notwithstanding, the prediction that natural language communities are homogeneous or that a single speaker may entertain $L_4$-like semantics for one scalar expression and $L_5$-like semantics for another is not implausible (cf.
+The current representation of lexica as Boolean matrices is blind to this anomaly.} The prediction that natural language communities are homogeneous or that a single speaker may entertain $L_4$-like semantics for one scalar expression and $L_5$-like semantics for another is not implausible (cf.
 \citealt{franke+degen:2016}).
@@
 Should cost play a role, then its effect is bound to depend on the frequency with which a given scalar expression is used.
-It is therefore plausible that frequently drawn scalar implicatures might fossilize to avoid cost, while infrequent ones could still be computed on-line.
-This opens a possible venue to address the first question about the expected presence of $L_4$-like semantics, but further empirical evidence is needed to assess these matters beyond speculation.
+It is therefore plausible that frequently drawn scalar implicatures might fossilize to avoid cost, while infrequent ones are still derived on-line.
+This also opens a possible venue to address the preceding question about the expected presence of $L_4$-like semantics, but further empirical evidence is needed to assess these matters beyond speculation.
 
 \subsection{Noisy transmission}\label{subsec:noise}
-The assumption of a learning bias was pivotal in introducing transmission perturbations that differentiate functionally similar types.
+Above, the assumption of a learning bias was pivotal in introducing transmission perturbations that differentiate functionally similar types.
 However, these perturbations need not necessarily arise from inductive biases.
 As shown in the following, outcomes that are indistinguishable from the ones predicted above can come about through environmental factors such as perceptual errors during production and comprehension as well.
-We should stress that we do not want to suggest noisy perception to underlie the selection of a lack of semantic upper-bounds in scalar items.
-Instead, our goal is to stress the role that transmission perturbations play in this process and the cultural evolution of meaning more broadly, as well as to highlight the explanatory potential environmental forces, which have received less attention than their inductive counterparts.
+With this we do not want to suggest noisy perception to underlie the selection of a lack of semantic upper-bounds in scalar items.
+Instead, we want to stress the role that transmission perturbations play in this process, and the cultural evolution of meaning more broadly, as well as to highlight the explanatory potential  of environmental factors, which have received less attention than their inductive counterparts (but see \citealt{oconnor:2015, franke+correia:toappear}).
 
@@
 All that is required is their adjustment to the possibility of confusing one state with another by language users and learners.
-Letting $\epsilon$ stand for the probability of confusing the first row state with the second, and vice-versa for $\delta$, we denote the probability that the teacher (learner) observes state $s_t$ ($s_l$) when the actual state is $s_a$ as $P_N(s_t \mid s_a)$ ($P_N(s_l \mid s_a)$).
+Letting $\epsilon$ stand for the probability of confusing \ssome with \sall, and vice-versa for $\delta$, we denote the probability that the teacher (learner) observes state $s_t$ ($s_l$) when the actual state is $s_a$ as $P_N(s_t \mid s_a)$ ($P_N(s_l \mid s_a)$).
 The probability that $s_a$ is the actual state when the learner observes $s_l$ is therefore:
@@
 She takes what she observes a state to be as the actual state that the teacher has seen as well and infers which type would have most likely generated the message to this state.
-This can lead to biases of inferring the ``wrong'' teacher type, if the noise makes some types err in a way that resembles the noiseless behavior of other types.
+This can lead to biases of inferring the ``wrong'' teacher type if the noise makes some types err in a way that resembles the noiseless behavior of other types.
 That is, such environmental factors can, in principle, induce transmission biases that look as if there was a cognitive bias in favor of a particular type, simply because that type better explains the noise.
@@
 In contrast to the outcome predicted by its noiseless counterpart, favourable outcomes for this type concentrate in parameter configurations  with $\epsilon > \delta$, $\alpha \geq 5$, $1 < k < 20$ and $l > 3$.
-As stressed above, we are not concerned with the interpretation of these particular values for the case study at hand, but rather with the technical result that shows that the mere presence of systematic noise in the transmission of strategies can introduce regularization that looks as if the agents have a learning bias.
+As stressed above, we are not concerned with the interpretation of these particular values for the case study at hand, but rather with the technical result that shows that the mere presence of systematic noise in transmission can introduce regularization that looks as if the agents have a learning bias.
 In other words, learning biases are clearly not the only transmission perturbations that shape cultural evolution alongside functional pressure.
@@
 This model generates predictions about lexicalization patterns and, more generally, effects of communicative pressures on the cultural evolution of language.
+
 We argued that the puzzle raised by semantics in light of pragmatics is hard to explain on purely functional grounds and that part of the answer may instead lie in the way transmission shapes the outcome of cultural evolution in tandem with a pressure for successful information transfer.
@@
 As a consequence, iterated transmission and use of language lead to a regularization that may explain the lack of lexicalization of systematic pragmatic enrichments.
-This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics by offering an account of why (certain) pragmatic inferences fail to lexicalize.
-More generally, we showed that systematic noise in perception can produce outcomes that are indistinguishable from those generated by inductive biases.
-This adds to \citeposs{franke+correia:toappear} argument that linguistic regularities may arise as a byproduct of noise, rather than through inductive biases.
+This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics.
+It offers an account of why (certain) pragmatic inferences fail to lexicalize.
+More generally, we showed that systematic noise in perception can produce outcomes that are similar from those generated by inductive biases.
 
@@
 That is, while our proposal combines bidirectional horizontal transmission with its vertical and unidirectional counterpart, Kirby et al.'s model only considers the latter's influence.
-Our reasoning behind the inclusion of the former lies in the empirical and theoretical observation that learnability alone can lead the selection of functionally defective languages, as showcased by the tautological language $L_3$ in our analysis.
+Our reasoning behind the inclusion of the former lies in the empirical and theoretical observation that learnability alone can lead the selection of functionally defective languages, as illustrated by the tautological language $L_3$ in our analysis.
 This outcome has been reported in a number of laboratory experiments where the participants' task was to learn and subsequently reproduce the language produced by a previous participant, leading to a proliferation of languages that associated a large number of meanings with a single form (see e.g.
 \citealt{silvey+etal:2014} and experiment 1 in \citealt{kirby+etal:2008}).
-In contrast, experiments involving an interactive component have been found to foster languages that enable interlocutors to distinguish meanings accurately  (e.g.
+In contrast, experiments involving an interactive component have been found to foster languages that enable interlocutors to distinguish meanings more accurately  (e.g.
 \citealt{fay+etal:2013}; for a review of laboratory results under the iterated learning paradigm and further discussion see \citealt{kirby+etal:2015, tamariz+kirby:2016}).
-It is not evident how to compare these findings given that they consider distinct meaning spaces, modes of transmission, iterations and feedback given to participants.
+It is not evident how to compare these empirical findings given that they consider distinct meaning spaces, modes of transmission, iterations and feedback given to participants.
 However, we take these results to suggest that there is an important difference between a language generating learnable linguistic data and its actual performance as a means of information transfer.
@@
 
-The demonstration that noise can lead to regularized evolutionary outcomes that are indistinguishable from those generated by prior learning biases is relevant not for the case study at hand, but more so for the broader project of analyzes of the cultural evolution of language.
+The demonstration that noise can lead to regularized evolutionary outcomes that are similar to those generated by prior learning biases is relevant not for the case study at hand, but more so for the broader project of investigating the cultural evolution of language.
 On the one hand, the plurality of sources of transmission perturbations admitted by these models paints a cautionary tale for the design of studies that purport to provide explanatory accounts of linguistic phenomena.
-In particular, when the outcome is interpreted as informative about the perturbation assumed to generates it (cf.
+In particular when the outcome is interpreted as being informative about the perturbation assumed to generate it (cf.
 \citealt{tamariz+kirby:2016}).
@@
 The cultural evolution of language is influenced by intertwined pressures.
-We set out to investigate this process by putting forward a model that combines a pressure toward efficient and successful information transfer with perturbations that may arise from the transmission of linguistic knowledge in acquisition.
+We set out to investigate this process by putting forward a model that combines a pressure toward efficient and successful information transfer with perturbations that may arise in the transmission of linguistic knowledge in acquisition.
 Additionally, we argued for the necessity of considering the role of pragmatics in investigations on the cultural evolution of meaning.
-These components and their mutual influence were highlighted in a case study on the lack of lexical upper-bounds in weak scalar expressions that showed that, when pressured for learnability and expressivity, the former drives for simpler semantic representations inasmuch as pragmatics can compensate for their lack of expressivity in use.
-That is, the relative advantage in learning of simpler semantics in tandem with a functional pressure in use may offer an answer to why natural languages fail to lexicalize systematic pragmatic inferences.
+These components and their mutual influence were highlighted in a case study on the lack of lexical upper-bounds in weak scalar expressions that showed that, when pressured for learnability and expressivity, the former drives for simpler semantic representations inasmuch as pragmatics can compensate for lack of expressivity in use.
+That is, the relative learning advantage of simpler semantics in tandem with a functional pressure in use may offer an answer to why natural languages fail to lexicalize systematic pragmatic inferences.
 
-We also considered an alternate instantiation of the model, which shows that systematic noise in state perception can give rise to evolutionary outcomes that are indistinguishable from those predicted by inductive biases.
+We also considered an alternate instantiation of the model, which shows that systematic noise in state perception can give rise to evolutionary outcomes that are similar to those predicted by inductive biases.
 This stresses the fact that that learning and typology are not necessarily close reflections of each other \citep{bowerman:2010}.
@@
 
+\appendix
+\section{Illustration of the role of $\alpha$ in linguistic choice}\label{app:sig-example}
+To clarify the role of $\alpha$, notice that for two states and two messages the possible values for $H_{0}(s|m;L)$ are restricted to
+zero, $0.5$ and one. Consider the case of $H_0$ with $L_5$, which assigns the following
+probabilities to states after hearing messages:
+\begin{align*}
+  H_{0}(\messg \mid \state; L_5) & = \begin{blockarray}{lcc}
+    & \mygray{\ssome} & \mygray{\sall} \\
+    \begin{block}{l[cc]}
+      \mygray{\msome} & 0.5 & 0.5 \\
+      \mygray{\mall}  & 0 & 1 \\
+    \end{block}
+  \end{blockarray}
+\end{align*}
+A level-1 speaker's utilities are then:
+\begin{align*}
+  \begin{blockarray}{lcc}
+    & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[cc]}
+      \mygray{\ssome} & 0.5^\alpha & 0^\alpha \\
+      \mygray{\sall}  & 0.5^\alpha & 1^\alpha \\
+    \end{block}
+  \end{blockarray}
+\end{align*}
+As the soft-max choice rule is sensitive to differences in utilities, no matter what $\lambda$
+is we will get $S_1(\msome \mid \ssome; L_5) = S_1(\mall \mid \sall; L_5)$ if $\alpha =
+1$. This means that we cannot distinguish pragmatically deviant choices (selecting a less
+informative message instead of a more informative one), from semantically deviant choices
+(selecting a false message instead of a true one). In contrast, if $\alpha > 1$ the former will be less likely than the latter -- and the other way around for $\alpha < 0$.
+
 \bibliographystyle{plainnat}
 \bibliography{./bounds-rmd}
-
-\newpage
-
-\section{Some thoughts + what may be missing}
-\begin{itemize}
-  \item The data used in plots and slides should be generated as uniformly as possible.
-I will do this once we settle for everything else.
-Any input on this is welcome, but I think it boils down to adjusting the non-noisy simulations a bit to the noisy ones (the number of samples is what mostly diverges).
-It's also up to debate whether we should run the non-noisy simulations with multiple Qs, as their noisy counterparts.
-  \item The analysis may benefit from further data and less of a qualitative overview of the parameters' effect (e.g. the regression analysis Michael conducted);
-\end{itemize}
-
 \end{document}

### 2016-11-24 paper/data/plots.py

@@ -1,0 +1,275 @@
+###
+import pandas as pd
+import numpy as np 
+import matplotlib.pyplot as plt
+import seaborn as sns
+#sns.set_context(rc={'lines.markeredgewidth': 0.5})
+import os
+import glob
+import sys
+###
+###Column headers:###
+#run_ID,t1_initial,t2_initial,t3_initial,t4_initial,t5_initial,t6_initial,t7_initial,t8_initial,t9_initial,t10_initial,t11_initial,t12_initial,
+#alpha,prior_cost_c,lambda,k,sample_amount,learning_parameter,generations,
+#t1_final,t2_final,t3_final,t4_final,t5_final,t6_final,t7_final,t8_final,t9_final,t10_final,t11_final,t12_final
+#####################
+
+
+def fitness_only_plot():
+    print 'Loading data fitness'
+    df = pd.read_csv('./mean-1000games-replication-only.csv')
+
+    
+    a = 1
+    lam = 30
+    
+    final_groupA = df.loc[:,'alpha':]
+    
+    groupA = final_groupA.loc[final_groupA['alpha'] == a]
+    
+    t_finalA = groupA.groupby(['lambda'])
+    t_finalA = t_finalA[['t1_final','t2_final','t3_final','t4_final','t5_final','t6_final','t7_final','t8_final','t9_final','t10_final','t11_final','t12_final']].agg(np.average) 
+    
+    d1 = t_finalA.loc[:,('t1_final','t2_final','t3_final', 't4_final', 't5_final', 't6_final')]
+    d2 = t_finalA.loc[:,('t9_final','t10_final', 't11_final')]
+    
+    ax2 = plt.plot(d2)
+    #ax1 = plt.plot(d1,marker='d',markevery=5)
+    plt.title(r'($\alpha$ = %d)' %(a))
+    plt.ylabel("Proportion in population")
+    plt.xlabel(r'Rationality parameter $\lambda$')
+    
+    plt.legend(["prag. L-taut", "prag. L-bound","prag. L-lack"], loc='best')
+    plt.xlim(0,48)
+    plt.show() 
+
+
+def learning_only_plot():
+    print 'Loading data learn'
+    dl = pd.read_csv('./mean-1000games-mutation-only.csv')
+    
+    a = 1
+    lam = 30
+    k = 5
+    sample = 10
+    learn = 1
+    
+    final_groupB = dl.loc[:,'alpha':]
[…]

### 2016-11-24 presentations/edindata/run-small.py

@@ -4,7 +4,7 @@ import modsinglescalaronlylearning as learner
 #parameter order for run(*) is run(alpha,cost,lam,k,sample_amount, learning_parameter,gens,runs)
 
 
-def cost_to_learning():
+def cost_to_learning(): #results 1
     a = 1
     lam = 30
     cost = [x/100. for x in xrange(100)]
@@ -64,4 +64,4 @@ def only_r_and_m():
 
 
 
-only_r_and_m()
+#only_r_and_m()

### 2016-12-29 paper/new_simulation_setup.tex
new simulation specs initial

\documentclass[a4paper]{article}

\usepackage{geometry}
\usepackage{natbib}
\bibpunct[:]{(}{)}{,}{a}{}{;}

%--------------------
%\usepackage{gb4e}
%\noautomath

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{mathrsfs}
\usepackage{nicefrac}
%\usepackage{stmaryrd}
%\usepackage{multicol}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}
%\newcommand{\mvalueof}[1]{\llbracket#1\rrbracket}
\newcommand{\citeposs}[2][]{\citeauthor{#2}'s (\citeyear[#1]{#2})}
\newcommand{\tuple}[1]{\ensuremath{\left\langle #1 \right\rangle}}

\newcommand{\hl}[1]{\textcolor[rgb]{.8,.33,.0}{#1}}% prints in orange
%\newcommand{\argmax}[1]{\underset{#1}{\operatorname{arg}\,\operatorname{max}}\;}
%\newcommand{\argmin}[1]{\underset{#1}{\operatorname{arg}\,\operatorname{min}}\;}
%\newcommand{\sbna}{\exists\lnot\forall}

\definecolor{Red}{RGB}{178,34,34}
\newcommand{\mf}[1]{\textcolor{Red}{[MF: #1]}} 
\newcommand{\tb}[1]{\textcolor[rgb]{.8,.33,.0}{[TB: #1]}}% prints in orange

\usepackage{blkarray}
\usepackage{xspace}

%%% MF's commands
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\card}[1]{\left \lvert \, #1 \, \right\rvert}
\newcommand{\abs}[1]{\lvert #1 \rvert}
\newcommand{\States}{\ensuremath{S}\xspace}		% Set of States
\newcommand{\state}{\ensuremath{s}\xspace}		% single states
\newcommand{\mystate}[1]{\ensuremath{\state_{\text{#1}}}\xspace} %meaningful states
\newcommand{\Messgs}{\ensuremath{M}\xspace}		% Set of Messages
\newcommand{\messg}{\ensuremath{m}\xspace}		% single messages
\newcommand{\mymessg}[1]{\ensuremath{\messg_{\text{#1}}}\xspace} %meaningful messages
\newcommand{\ssome}{\mystate{\ensuremath{\exists\neg\forall}}}
\newcommand{\sall}{\mystate{\ensuremath{\forall}}}
\newcommand{\snone}{\mystate{\ensuremath{\emptyset}}}
\newcommand{\msome}{\mymessg{some}}
\newcommand{\mall}{\mymessg{all}}
\newcommand{\asome}{\myact{\ensuremath{\exists\neg\forall}}}
\newcommand{\aall}{\myact{\ensuremath{\forall}}}
\definecolor{mygray}{cmyk}{0.35,0.35,0.35,0.35}
\newcommand{\mygray}[1]{{\textcolor{mygray}{#1}}}
%%%

%--------------------
%
%\usepackage{setspace}
%\onehalfspacing
%
%-------------------

\title{Tracing the cultural evolution of meaning at the semantics-pragmatics interface}

\author{%\bf NAME1 and NAME2\\
    ( -- draft \today --- )
}

\date{}

\begin{document}

We consider three states $\States = \set{\snone, \ssome, \sall}$. This state space can be
thought of as a partition of possible worlds into cells where none, some or all of the $A$s are
$B$s, for some arbitrary fixed predicates $A$ and $B$. Eight concepts can be distinguished
based on their truth or falsity in three world states, six of which are not contradictory
(always false) or tautologous (always true). These are listed with mnemonic names in
Table~\ref{tab:concepts}.

\begin{table}
  \centering
\begin{center}
  \begin{tabular}{lccccl}
    \toprule
    intuitive name
    & \snone
    & \ssome
    & \sall
    & min.~derivation length
    & formula
    \\ \midrule
    ``all''
    & 0
    & 0
    & 1
    & $3$
    & $A \subseteq B$
    \\
    ``some but not all''
    & 0
    & 1
    & 0
    & $7$
    & $A \cap B \neq \emptyset \wedge A \neq \emptyset$
    \\    
    ``some''
    & 0
    & 1
    & 1
    & $4$
    & $A \cap B \neq \emptyset$
    \\
    ``none''
    & 1
    & 0
    & 0
    & $4$
    & $A \cap B = \emptyset$
    \\
    ``none or all''
    & 1
    & 0
    & 1
    & $8$
    & $\neg(A \cap B \neq \emptyset \wedge A \neq \emptyset)$
    \\
    ``not all''
    & 1
    & 1
    & 0
    & $4$
    & $\neg (A \subseteq B)$
    \\
    \bottomrule
  \end{tabular}
\end{center}
\caption{Available concepts and their minimal derivation length}
\label{tab:concepts}
\end{table}

Identifying three world states requires three messages in $\Messgs$. As before, we call a
lexicon $L$ a mapping $\Messgs \rightarrow C$ from messages to concepts. There are $6^3 = 343$
possible lexica, but many of these assign the same concept to more than one message. For
simplicity, attention could be restricted to only those lexica which avoid expressive
redundancy. Hardcoding such an \emph{mutual exclusivity bias}
\citep[e.g.][]{Clark2009:Lexical-Meaning}, there are 20 non-redundant
lexica.\footnote{Functional pressure towards efficient communication will evidently favor
  non-redundant lexica, so this restriction is fairly innocuous but practical.}

As before, the type of a player is a combination of a lexicon and a manner of pragmatic
language use: literal or pragmatic. With this, there are 40 types in this model.

\begin{table}
  \centering
  \begin{align*}
    C & \rightarrow C \wedge C 
    & 
    X & \rightarrow \set{A,B} \\
    C & \rightarrow\neg C 
    & 
    X & \rightarrow X \cap X \\
    C & \rightarrow X \subseteq X
    & 
    X & \rightarrow X \cup X \\
    C & \rightarrow X \neq \emptyset
    & 
    X & \rightarrow X \setminus X \\
    C & \rightarrow X = \emptyset     
    & 
    X & \rightarrow \overline{M}
  \end{align*}
  \caption{Toy grammar in a set-theoretic "language of thought"}
  \label{tab:grammar}
\end{table}

The prior probability of a type is just the prior probability of its lexicon. The prior of a
lexicon is a function of the complexity of the concepts in its image set. Lexica that use
simpler concepts are \emph{a priori} more likely. This can be motivated by assuming that
learners beam search for suitable concepts to map onto overt signals by (probabilistically)
considering simpler concepts first. There are many ways for defining complexity of a
concept. Following \citet{piantadosi+etal:underreview} and related work, we define complexity
of a concept as a function of its derivation length in a generative ``language of
thought''. For concreteness of example, consider the toy grammar of concepts in
Table~\ref{tab:grammar}. This grammar uses basic set-theoretic operations to form expressions
which can be evaluated as true or false in our three world states. Table~\ref{tab:concepts}
lists the minimal length of a derivation of an expression that represents each concept. A
simple way of forming priors over a lexicon is:
\begin{align*}
  P(L)  & = \prod_{c \in Im(L)} P(c)  \ \ \ \text{, with} & 
  P(c) & \propto \max_{c'}Compl(c') - Compl(c')\,,
\end{align*}
where $Compl(c)$ is the minimum length of a derivation of concept $c$.

\bibliographystyle{plainnat}
\bibliography{./bounds-rmd}
\end{document}

### 2016-12-29 paper/new_simulation_setup.tex
new simulation specs update

@@
 \begin{center}
-  \begin{tabular}{lccccl}
+  \begin{tabular}{lccclc}
     \toprule
@@
     & \sall
-    & min.~derivation length
-    & formula
+    & least complex formula
+    & complexity
     \\ \midrule
@@
     & 1
+    & $A \subseteq B$
     & $3$
-    & $A \subseteq B$
     \\
@@
     & 0
-    & $7$
     & $A \cap B \neq \emptyset \wedge A \neq \emptyset$
+    & $8$
     \\    
@@
     & 1
+    & $A \cap B \neq \emptyset$
     & $4$
-    & $A \cap B \neq \emptyset$
     \\
@@
     & 0
+    & $A \cap B = \emptyset$
     & $4$
-    & $A \cap B = \emptyset$
     \\
@@
     & 1
-    & $8$
     & $\neg(A \cap B \neq \emptyset \wedge A \neq \emptyset)$
+    & $9$
     \\
@@
     & 0
-    & $4$
     & $\neg (A \subseteq B)$
+    & $5$
     \\
@@
 
-Identifying three world states requires three messages in $\Messgs$. As before, we call a
-lexicon $L$ a mapping $\Messgs \rightarrow C$ from messages to concepts. There are $6^3 = 343$
-possible lexica, but many of these assign the same concept to more than one message. For
-simplicity, attention could be restricted to only those lexica which avoid expressive
-redundancy. Hardcoding such an \emph{mutual exclusivity bias}
-\citep[e.g.][]{Clark2009:Lexical-Meaning}, there are 20 non-redundant
-lexica.\footnote{Functional pressure towards efficient communication will evidently favor
-  non-redundant lexica, so this restriction is fairly innocuous but practical.}
+A lexicon $L$ is a mapping $\Messgs \rightarrow C$ from messages to concepts. With three
+messages there are $6^3 = 343$ possible lexica, but many of these assign the same concept to
+more than one message. For simplicity, we restrict attention to only those lexica which avoid
+expressive redundancy. Hardcoding such an \emph{mutual exclusivity bias}
+\citep[e.g.][]{Clark2009:Lexical-Meaning}, there are 20 non-redundant lexica. Functional
+pressure towards efficient communication will evidently favor non-redundant lexica, so this
+restriction is fairly innocuous but practical.
 
@@
   \begin{align*}
-    C & \rightarrow C \wedge C 
+    C & \rightarrow_2 C \wedge C 
     & 
-    X & \rightarrow \set{A,B} \\
-    C & \rightarrow\neg C 
+    X & \rightarrow_1 \set{A,B} \\
+    C & \rightarrow_2 \neg C 
     & 
-    X & \rightarrow X \cap X \\
-    C & \rightarrow X \subseteq X
+    X & \rightarrow_1 X \cap X \\
+    C & \rightarrow_1 X \subseteq X
     & 
-    X & \rightarrow X \cup X \\
-    C & \rightarrow X \neq \emptyset
-    & 
-    X & \rightarrow X \setminus X \\
-    C & \rightarrow X = \emptyset     
-    & 
-    X & \rightarrow \overline{M}
+    X & \rightarrow_1 X \cup X \\
+    C & \rightarrow_1 X \neq \emptyset \\
+    C & \rightarrow_1 X = \emptyset     
   \end{align*}
@@
 learners beam search for suitable concepts to map onto overt signals by (probabilistically)
-considering simpler concepts first. There are many ways for defining complexity of a
-concept. Following \citet{piantadosi+etal:underreview} and related work, we define complexity
-of a concept as a function of its derivation length in a generative ``language of
+considering simpler concepts first. Many ways for defining complexity of a concept are
+conceivable. If strong empirical claims were at stake here, empirically motivated measures of
+complexity should be used. For the sake of a non-trivial example, we follow
+\citet{piantadosi+etal:underreview} and related work to define complexity of a concept as a
+function of its derivation cost in a (weighed or probabilistic) generative ``language of
 thought''. For concreteness of example, consider the toy grammar of concepts in
 Table~\ref{tab:grammar}. This grammar uses basic set-theoretic operations to form expressions
-which can be evaluated as true or false in our three world states. Table~\ref{tab:concepts}
-lists the minimal length of a derivation of an expression that represents each concept. A
-simple way of forming priors over a lexicon is:
+which can be evaluated as true or false in our three world states. Applications of generative
+rules have a cost attached to them. (Alternatively, a probability.) Here we simply assume that
+Boolean combinations of concepts are more complex than ``atomic'' concepts and that otherwise
+each rule application adds the same cost unit. Table~\ref{tab:concepts} lists, for each
+concept, the least complex formula derivable in this grammar that has the appropriate truth
+conditions. A simple way of defining priors over a lexicon is:
 \begin{align*}
@@
 \end{align*}
-where $Compl(c)$ is the minimum length of a derivation of concept $c$.
+where $Compl(c)$ is the complexity of each concepts.
 

### 2017-02-01 code-LOT-extension/dynamics.py
code for new rmd + some plots

@@ -5,152 +5,100 @@
 #Target type: index 24 in type/lexica/prior list
 #####
 
-import numpy as np
-#np.set_printoptions(threshold=np.nan)
-from random import sample
-from lexica import get_lexica,get_prior
-from player import LiteralPlayer,GriceanPlayer
-from rmd import get_utils,get_mutation_matrix
-import sys 
-import datetime
-import csv
-
+from rmd import run_dynamics
 
 ##### Parameters & setup #####
-alpha = 1 # rate to control difference between semantic and pragmatic violations
-lam = 20 # soft-max parameter
-k = 5  # length of observation sequences
-sample_amount = 200 #amount of k-length samples for each production type
+a = [1] # rate to control difference between semantic and pragmatic violations
+lamb = [10,30] # soft-max parameter
+seq_length = [5,10,15,20]  # length of observation sequences
+samples = [200] #amount of k-length samples for each production type
+l = [1,5,10,15] #prob-matching = 1, increments approach MAP
 
-gens = 50 #number of generations per simulation run
-runs = 50 #number of independent simulation runs
+g = [50] #number of generations per simulation run
+r = [1000] #number of independent simulation runs
+
+dynamics = ['r','m','rmd'] #kind is the type of dynamics, either 'rmd', 'm' or 'r'
 
 states = 3 #number of states
 messages = 3 #number of messages
 
-learning_parameter = 1 #prob-matching = 1, increments approach MAP
-state_freq = np.ones(states) / float(states) #frequency of states s_1,...,s_n 
+
+for alpha in a:
+    for lam in lamb:
+        for k in seq_length:
+            for sample_amount in samples:
+                for learning_parameter in l:
+                    for gens in g:
+                        for runs in r:
+                            for kind in dynamics:
+#                                run_dynamics(alpha,lam,k,sample_amount,gens,runs,states,messages,learning_parameter,kind) 
+
+
+
+
[…]

### 2017-02-01 code-LOT-extension/lexica.py

@@ -39,5 +39,3 @@ def get_prior(lexica_list):
         out.append(lex_val)
     out = out + out #double for two types of linguistic behavior
     return np.array(out) / np.sum(out)
-
-

### 2017-02-01 code-LOT-extension/plots.py

@@ -1,0 +1,142 @@
+###
+import pandas as pd
+import numpy as np 
+import matplotlib.pyplot as plt
+import seaborn as sns
+#sns.set_context(rc={'lines.markeredgewidth': 0.5})
+import os
+import glob
+import sys
+###
+###Column headers:###
+#####################
+
+
+def get_prior_plot():
+    from lexica import get_prior, get_lexica
+    prior = get_prior(get_lexica(3,3,mutual_exclusivity=True))
+    
+    X = np.arange(40)
+    Y_target = [0 for _ in xrange(24)] + [prior[24]] + [0 for _ in xrange(15)]
+    print prior[24]
+    prior[24] = 0
+    Y_rest = prior
+    print prior[24]
+    
+    fig, ax = plt.subplots()
+    ax.bar(X,Y_target,width=1, color='green')
+    ax.bar(X,Y_rest,width=1)
+    plt.show()
+
+#get_prior_plot()
+
+def get_utility_heatmap(s_amount,m_amount,lam,alpha):
+    print 'Loading U-matrix'
+    df = pd.read_csv('./matrices/umatrix-s%d-m%d-lam%d-a%d.csv' %(s_amount,m_amount,lam,alpha))
+    yticks = [0] + ["" for _ in xrange(9)] + [10] + ["" for _ in xrange(9)] +\
+             [20] + ["" for _ in xrange(9)] + [30] + ["" for _ in xrange(9)]
+    
+    ax = sns.heatmap(df, yticklabels=yticks, xticklabels=yticks)#, annot=True) 
+#    plt.yticks(rotation=0)
+    from matplotlib.patches import Rectangle
+    
+    ax.add_patch(Rectangle((24,15), 1, 1, fill=False, edgecolor='blue', lw=3))
+    ax.invert_yaxis()
+    plt.show()
+
+#get_utility_heatmap(3,3,10,1)
+
+
+def get_mutation_heatmap(s_amount,m_amount,lam,alpha,k,samples,l):
+    print 'Loading Q-matrix'
+    df = pd.read_csv('./matrices/qmatrix-s%d-m%d-lam%d-a%d-k%d-samples%d-l%d.csv' %(s_amount,m_amount,lam,alpha,k,samples,l))
[…]

### 2017-02-01 code-LOT-extension/rmd.py

@@ -5,9 +5,12 @@ np.set_printoptions(threshold=np.nan)
 from random import sample
 from itertools import product
 from player import LiteralPlayer,GriceanPlayer
+from lexica import get_lexica,get_prior
+
 import sys 
 import datetime
 import csv
+import os.path
 
 def normalize(m):
     return m / m.sum(axis=1)[:, np.newaxis]
@@ -47,62 +50,114 @@ def get_likelihood(obs,likelihoods):
             out[lhi,o] = np.prod([flat_lhi[x]**obs[o][x] for x in xrange(len(obs[o]))]) 
     return out
 
-def get_mutation_matrix(s_amount,m_amount,state_freq, likelihoods,lexica_prior,learning_parameter,sample_amount,k):
-    obs = get_obs(s_amount,m_amount,k,likelihoods,state_freq,sample_amount) #get production data from all types
-    out = np.zeros([len(likelihoods),len(likelihoods)]) #matrix to store Q
+def get_mutation_matrix(s_amount,m_amount,state_freq, likelihoods,lexica_prior,learning_parameter,sample_amount,k,lam,alpha):
 
-    for parent_type in xrange(len(likelihoods)):
-        type_obs = obs[parent_type] #Parent production data
-        lhs = get_likelihood(type_obs,likelihoods) #P(parent data|t_i) for all types
-        post = normalize(lexica_prior * np.transpose(lhs)) #P(t_j|parent data) for all types; P(d|t_j)P(t_j)
-        parametrized_post = normalize(post**learning_parameter)
+    if os.path.isfile('./matrices/qmatrix-s%d-m%d-lam%d-a%d-k%d-samples%d-l%d.csv' %(s_amount,m_amount,lam,alpha,k,sample_amount,learning_parameter)):
+        print '#Loading mutation matrix, ', datetime.datetime.now()
+        return np.genfromtxt('./matrices/qmatrix-s%d-m%d-lam%d-a%d-k%d-samples%d-l%d.csv' %(s_amount,m_amount,lam,alpha,k,sample_amount,learning_parameter), delimiter=',')
+    else:
+        print '#Computing mutation matrix, ', datetime.datetime.now()
+    
 
-        out[parent_type] = np.dot(np.transpose(lhs[parent_type]),parametrized_post)
+        obs = get_obs(s_amount,m_amount,k,likelihoods,state_freq,sample_amount) #get production data from all types
[…]

### 2017-02-01 code-LOT-extension/plots.py
r/m/rm comparison plots

@@ -67,76 +67,121 @@ def get_mutation_heatmap(s_amount,m_amount,lam,alpha,k,samples,l):
 #get_mutation_heatmap(3,3,30,1,15,200,10)
 #
 
-group1 = [30,1,5,200,1,50,1000]
-#def get_some_analysis(group1,group2,group3):
-print 'Loading data'
-df_r = pd.read_csv('./results/00mean-r-s3-m3-g50-r1000.csv')
-df_m = pd.read_csv('./results/00mean-m-s3-m3-g50-r1000.csv')
-df_rm = pd.read_csv('./results/00mean-rmd-s3-m3-g50-r1000.csv')
-
-df_r = df_r.loc[df_r['lam'] == group1[0]]
-df_r = df_r.loc[df_r['alpha'] == group1[1]]
-df_r = df_r.loc[df_r['k'] == group1[2]]
-df_r = df_r.loc[df_r['samples'] == group1[3]]
-df_r = df_r.loc[df_r['l'] == group1[4]]
-df_r = df_r.loc[df_r['gens'] == group1[5]]
-df_r = df_r.loc[df_r['runs'] == group1[6]]
-final_r = df_r.loc[:,'t_mean0':]
-
-#target_lex = final_r['t_mean24']
-
-r_array = map(list,final_r.values)
-r_array = r_array[0]
-
-target = r_array[24]
-r_sorted = sorted(r_array,reverse=True)
-r_top = r_sorted[:3]
-idx = []
-for i in r_top:
-    index = r_array.index(i)
-    idx.append(index)
-
-if 24 not in idx:
-    r_top.pop()
-    idx.pop()
-    r_top.append(target)
-    idx.append(24)
 
 
-X = np.arange(27)
-empty = [0 for _ in xrange(3)]
-Y_r = empty + r_top + empty * 7
+def get_some_analysis(group1,group2,group3):
+    print 'Loading data'
+    df_r = pd.read_csv('./results/00mean-r-s3-m3-g50-r1000.csv')
+    df_m = pd.read_csv('./results/00mean-m-s3-m3-g50-r1000.csv')
+    df_rm = pd.read_csv('./results/00mean-rmd-s3-m3-g50-r1000.csv')
+    
+    df_r = df_r.loc[df_r['lam'] == group1[0]]
+    df_r = df_r.loc[df_r['alpha'] == group1[1]]
+    df_r = df_r.loc[df_r['k'] == group1[2]]
[…]

### 2017-02-02 paper/new_simulation_setup.tex
mistake corrected in new simulation set up
comments new simulation setup

@@
 \usepackage{booktabs}
+
+\usepackage{caption}
+\usepackage{subcaption}
+
 %\newcommand{\mvalueof}[1]{\llbracket#1\rrbracket}
@@
 \newcommand{\mall}{\mymessg{all}}
+\newcommand{\mnone}{\mymessg{none}}
 \newcommand{\asome}{\myact{\ensuremath{\exists\neg\forall}}}
@@
     & $\neg(A \cap B \neq \emptyset \wedge A \neq \emptyset)$
-    & $9$
+    & \tb{$10$}
     \\
@@
   P(L)  & = \prod_{c \in Im(L)} P(c)  \ \ \ \text{, with} & 
-  P(c) & \propto \max_{c'}Compl(c') - Compl(c')\,,
+  P(c) & \propto \max_{c'}Compl(c') - Compl(c') + 1\,,
 \end{align*}
 where $Compl(c)$ is the complexity of each concepts.
+\tb{I changed `$P(L) = $' to `$P(L) \propto$', as well as   `$P(c) \propto \max_{c'}Compl(c') - Compl(c) + 1$'.
+I hope both were intended like this}.
+
+\section*{Some results} To get an idea of how this changes our previous results, I inspected some of the new setup's properties. Overall, the results are quite similar. A couple remarks: 
+\begin{enumerate}
+\item So far there is no parameter to regulate the strength of the inductive bias (before controlled by $c$). This means that we could get stronger/weaker results than the ones presented here
+\item Our target type is $t_{24}$, made explicit below, which currently has (i) the most a priori probable lexicon (see Figure \ref{fig:priors}), but is (ii) not the most learnable (see Figure \ref{fig:q-matrices}), nor (iii) the fittest type (see Figure \ref{fig:u-matrix}).
+\item In terms of the dynamics, $t_{24}$ fares worse than in our target type in the previous setup under only replication, OK but far from perfectly under mutation, but fares surprisingly well under when both dynamics are involved (see Figure \ref{fig:outcomes})
+\item Functional deficiency of $t_{24}$ is due to its (Gricean) receiver behavior.
+In principle, this would not be so if either (i) receivers were one level higher in the reasoning type hierarchy or (ii) receivers soft-maximized.
+\item As noted earlier by Michael, having our target be the most a priori probable is not so nice.
+On the other hand, we could stress the fact that its not the most communicatively efficient nor the most learnable language (when learning from a parent of the same type).
+\item Overall, this seems to work.
+We have a lot more types and more complex lexica but the results are roughly the same even without involving parameter $c$.
+The most important issue is to settle on a good way to transform a LOT-like representation into a prior over lexica.
+Personally, I'm torn between a probabilistic conception (skipping complexity and considering the probability of each rule application directly) and something akin to what Michael proposed here.
+I'll read up more on how other people have done it in the past in the meantime.
+Let me know if you have suggestions about this.
+Afterwards, I can run a more proper analysis.
+\end{enumerate}
+
+\section*{Details}
+
+Our target, Gricean $t_{24}$ lexicalizes messages corresponding to the \sall, \snone and \{\sall,\ssome\} (cf. Table \ref{tab:concepts}). Its lexicon is accordingly (same as literal player $t_4$):
+
+\begin{centering}
+\bordermatrix{~ &	  \mall & \msome & \mnone \cr 
+                  \snone &  0 & 0 & 1 \cr
+                  \ssome & 0 & 1 & 0 \cr
+		  \sall & 1 & 1 & 0 \cr}
+\end{centering}
+
+Following the setup above and the complexity values in Table \ref{tab:concepts}, the priors of our $40$ types are shown here:
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[scale=.35]{../code-LOT-extension/priors.png} %width = 0.75\linewidth
+  \caption{Prior over types (lexica). The green prior corresponds to $t_{24}$.}
+  \label{fig:priors}
+\end{figure}
+
+I thought it would be instructive to inspect different fitness and Q-matrices.
+As before, the matrix for expected utility only depends on parameters $\alpha$ and $\lambda$.
+There do not seem to be any particular surprises; $t_{24}$ is not the best but also far from the worse (see remark 4 above).
+With $\alpha = 1$ and $\lambda = 30$ we get $\text{EU}(t_{24},t_{24}) \approx .88$.
+Here's a visualization of the entire u-matrix:
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[scale=.25]{../code-LOT-extension/u-l30-a1.png} %width = 0.75\linewidth
+  \caption{Utility-Matrix for $\alpha = 1$ and $\lambda = 30$. The boxed value corresponds to $\text{EU}(t_{24},t_{24})$.}
+  \label{fig:u-matrix}
+\end{figure}
+
+For $Q$ there are more values to test.
+For sequence length $k$ there's not that much change as long as it is not too small and possible noise due to the sampling of learning data is accounted for (it's still important, but I'll background this for now).
+As before, the posterior parameter $l$ is very important.
+If $l \geq 10$ we get very faithful transmission of our target type.
+Lower values lead to more mutation.
+Here's a visual impression of two $Q$ matrices for different $l$ values:
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[scale=0.4]{../code-LOT-extension/q-l30-a1-k5-l1.png}
+  \end{subfigure}
+  ~
+   \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[scale=0.4]{../code-LOT-extension/q-l30-a1-k5-l10.png}
+  \end{subfigure}
+  \caption{$Q$-matrices for $\alpha = 1, \lambda = 30, k = 5$, $200$ samples per parent type with either $l = 1$ (left) and $l = 10$ (right). The boxed value corresponds to $P(t_{24} | t_{24})$.}
+  \label{fig:q-matrices}
+\end{figure}
+
+\newpage
+
+Finally, we can see how the above affects actual applications of only replication, only mutation, and replication \emph{and} mutation using the matrices shown above.
+To construct these plots I took the mean top $3$ types from $1000$ simulations using either of the three dynamics.
+If $t_{24}$ was not in the top $3$, I took out the third most prolific type and added $t_{24}$ instead.
+
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[scale=0.4]{../code-LOT-extension/rmd-lam30-a1-k5-s200-l1.png}
+  \end{subfigure}
+  ~
+   \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[scale=0.4]{../code-LOT-extension/rmd-lam30-a1-k5-s200-l10.png}
+  \end{subfigure}
+  \caption{Mean proportion of top $2/3$ types and target $t_{24}$ under replication only (green plot), mutation only (blue plot), and replication {\em and} mutation (red plot) after $50$ generations with $\alpha = 1, \lambda = 30, k = 5$, $200$ samples per parent type and $l = 1$ (left figure) and $l = 10$ (right figure).}
+  \label{fig:outcomes}
+\end{figure}
 

### 2017-02-02 code-LOT-extension/dynamics.py

@@ -23,16 +23,15 @@ states = 3 #number of states
 messages = 3 #number of messages
 
 
-for alpha in a:
-    for lam in lamb:
-        for k in seq_length:
-            for sample_amount in samples:
-                for learning_parameter in l:
-                    for gens in g:
-                        for runs in r:
-                            for kind in dynamics:
+#for alpha in a:
+#    for lam in lamb:
+#        for k in seq_length:
+#            for sample_amount in samples:
+#                for learning_parameter in l:
+#                    for gens in g:
+#                        for runs in r:
+#                            for kind in dynamics:
 #                                run_dynamics(alpha,lam,k,sample_amount,gens,runs,states,messages,learning_parameter,kind) 
-
 
 
 
@@ -45,23 +44,24 @@ for alpha in a:
 #import datetime
 #import csv
 #import os.path
-
-
+#
+#lam = 30
+#alpha = 1
+#
 #
 #state_freq = np.ones(states) / float(states) #frequency of states s_1,...,s_n 
-
+#
 ##### Auxiliary functions #####
 #def m_max(m): #aux function for convenience
 #    return np.unravel_index(m.argmax(), m.shape)
-
-#print '#Starting, ', datetime.datetime.now()
 #
+#print '#Starting, ', datetime.datetime.now()
 #lexica = get_lexica(states,messages,mutual_exclusivity=True)
 #l_prior = get_prior(lexica)
 #typeList = [LiteralPlayer(lam,lex) for lex in lexica] + [GriceanPlayer(alpha,lam,lex) for lex in lexica]
 #
 #likelihoods = np.array([t.sender_matrix for t in typeList])
-#
+
 #u = get_utils(typeList,states,messages,lam,alpha)
 #q = get_mutation_matrix(states, messages, state_freq, likelihoods,l_prior,learning_parameter,sample_amount,k,lam,alpha)
 #

### 2017-02-02 code-LOT-extension/dynamics.py
comments on dynamics script

@@ -1,10 +1,4 @@
 ##### Main file to run dynamics
-#RMD with parametrized iterated parental learning 
-#three states (none,sbna,all), three messages 
-#2 possible signaling behaviors: literal or gricean
-#Target type: index 24 in type/lexica/prior list
-#####
-
 from rmd import run_dynamics
 
 ##### Parameters & setup #####
@@ -23,17 +17,17 @@ states = 3 #number of states
 messages = 3 #number of messages
 
 
-#for alpha in a:
-#    for lam in lamb:
-#        for k in seq_length:
-#            for sample_amount in samples:
-#                for learning_parameter in l:
-#                    for gens in g:
-#                        for runs in r:
-#                            for kind in dynamics:
-#                                run_dynamics(alpha,lam,k,sample_amount,gens,runs,states,messages,learning_parameter,kind) 
+for alpha in a:
+    for lam in lamb:
+        for k in seq_length:
+            for sample_amount in samples:
+                for learning_parameter in l:
+                    for gens in g:
+                        for runs in r:
+                            for kind in dynamics:
+                                run_dynamics(alpha,lam,k,sample_amount,gens,runs,states,messages,learning_parameter,kind) 
 
-
+#### This is all that is needed to run either replication only, mutation only or both together #####
 
 ############################### Decomment for testing individual components ########################################
 #import numpy as np

### 2017-02-08 paper/new_simulation_setup.tex
results for larger state space + some code adjustments for plots and simulations

@@
 A lexicon $L$ is a mapping $\Messgs \rightarrow C$ from messages to concepts. With three
-messages there are $6^3 = 343$ possible lexica, but many of these assign the same concept to
+messages there are $6^3 = 216$ possible lexica, but many of these assign the same concept to
 more than one message. For simplicity, we restrict attention to only those lexica which avoid
@@
 
-\section*{Details}
+\section*{Details for mutual exclusivity type space}
 
@@
   \centering
-  \includegraphics[scale=.35]{../code-LOT-extension/priors.png} %width = 0.75\linewidth
+  \includegraphics[scale=.35]{../code-LOT-extension/plots/priors-meTrue.png} %width = 0.75\linewidth
   \caption{Prior over types (lexica). The green prior corresponds to $t_{24}$.}
@@
   \centering
-  \includegraphics[scale=.25]{../code-LOT-extension/u-l30-a1.png} %width = 0.75\linewidth
+  \includegraphics[scale=.25]{../code-LOT-extension/plots/u-l30-a1-meTrue.png} %width = 0.75\linewidth
   \caption{Utility-Matrix for $\alpha = 1$ and $\lambda = 30$. The boxed value corresponds to $\text{EU}(t_{24},t_{24})$.}
@@
   \begin{subfigure}[b]{0.45\textwidth}
-    \includegraphics[scale=0.4]{../code-LOT-extension/q-l30-a1-k5-l1.png}
+    \includegraphics[scale=0.4]{../code-LOT-extension/plots/q-l30-a1-k5-l1-meTrue.png}
   \end{subfigure}
@@
    \begin{subfigure}[b]{0.45\textwidth}
-    \includegraphics[scale=0.4]{../code-LOT-extension/q-l30-a1-k5-l10.png}
+    \includegraphics[scale=0.4]{../code-LOT-extension/plots/q-l30-a1-k5-l10-meTrue.png}
   \end{subfigure}
@@
   \begin{subfigure}[b]{0.45\textwidth}
-    \includegraphics[scale=0.4]{../code-LOT-extension/rmd-lam30-a1-k5-s200-l1.png}
+    \includegraphics[scale=0.4]{../code-LOT-extension/plots/rmd-lam30-a1-k5-s200-l1-meTrue.png}
   \end{subfigure}
@@
    \begin{subfigure}[b]{0.45\textwidth}
-    \includegraphics[scale=0.4]{../code-LOT-extension/rmd-lam30-a1-k5-s200-l10.png}
+    \includegraphics[scale=0.4]{../code-LOT-extension/plots/rmd-lam30-a1-k5-s200-l10-meTrue.png}
   \end{subfigure}
   \caption{Mean proportion of top $2/3$ types and target $t_{24}$ under replication only (green plot), mutation only (blue plot), and replication {\em and} mutation (red plot) after $50$ generations with $\alpha = 1, \lambda = 30, k = 5$, $200$ samples per parent type and $l = 1$ (left figure) and $l = 10$ (right figure).}
+  \label{fig:outcomes}
+\end{figure}
+
+\newpage
+
+\section*{Details for larger type space}
+
+Our target is now Gricean $t_{68}$ with the same lexicon as $t_{24}$ in the reduced type space.
+Following the setup above and the complexity values in Table \ref{tab:concepts}, the priors of our now $112$ types are shown here:
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[scale=.35]{../code-LOT-extension/plots/priors-meFalse.png} %width = 0.75\linewidth
+  \caption{Prior over types (lexica). The green prior corresponds to $t_{68}$.}
+  \label{fig:priors}
+\end{figure}
+
+As before,  $t_{68}$ is not the best but also far from the worse. Here's a visualization of the entire u-matrix:
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[scale=.25]{../code-LOT-extension/plots/u-l30-a1-meFalse.png} %width = 0.75\linewidth
+  \caption{Utility-Matrix for $\alpha = 1$ and $\lambda = 30$. The boxed value corresponds to $\text{EU}(t_{68},t_{68})$.}
+  \label{fig:u-matrix}
+\end{figure}
+
+Here's a visual impression of two $Q$ matrices for different $l$ values with the larger type space:
+
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[scale=0.4]{../code-LOT-extension/plots/q-l30-a1-k5-l1-meFalse.png}
+  \end{subfigure}
+  ~
+   \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[scale=0.4]{../code-LOT-extension/plots/q-l30-a1-k5-l10-meFalse.png}
+  \end{subfigure}
+  \caption{$Q$-matrices for $\alpha = 1, \lambda = 30, k = 5$, $200$ samples per parent type with either $l = 1$ (left) and $l = 10$ (right).
+The boxed value corresponds to $P(t_{68} | t_{68})$.
+You have to zoom in all the way to see something.
+Otherwise; it's similar to the mutation value of our target in the reduced space}
+  \label{fig:q-matrices}
+\end{figure}
+
+\newpage
+
+Finally, we can see how the above affects actual applications of only replication, only mutation, and replication \emph{and} mutation using the matrices shown above.
+To construct these plots I took the mean top $3$ types from $1000$ simulations using either of the three dynamics.
+If $t_{68}$ was not in the top $3$, I took out the third most prolific type and added $t_{68}$ instead.
+
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[scale=0.4]{../code-LOT-extension/plots/rmd-lam30-a1-k5-s200-l1-meFalse.png}
+  \end{subfigure}
+  ~
+   \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[scale=0.4]{../code-LOT-extension/plots/rmd-lam30-a1-k5-s200-l10-meFalse.png}
+  \end{subfigure}
+  \caption{Mean proportion of top $2/3$ types and target $t_{68}$ under replication only (green plot), mutation only (blue plot), and replication {\em and} mutation (red plot) after $50$ generations with $\alpha = 1, \lambda = 30, k = 5$, $200$ samples per parent type and $l = 1$ (left figure) and $l = 10$ (right figure).}
   \label{fig:outcomes}

### 2017-02-08 code-LOT-extension/dynamics.py

@@ -15,48 +15,55 @@ dynamics = ['r','m','rmd'] #kind is the type of dynamics, either 'rmd', 'm' or '
 
 states = 3 #number of states
 messages = 3 #number of messages
+me = [False] #mutual exclusivity
 
-
-for alpha in a:
-    for lam in lamb:
-        for k in seq_length:
-            for sample_amount in samples:
-                for learning_parameter in l:
-                    for gens in g:
-                        for runs in r:
-                            for kind in dynamics:
-                                run_dynamics(alpha,lam,k,sample_amount,gens,runs,states,messages,learning_parameter,kind) 
-
+#for alpha in a:
+#    for lam in lamb:
+#        for k in seq_length:
+#            for sample_amount in samples:
+#                for learning_parameter in l:
+#                    for gens in g:
+#                        for runs in r:
+#                            for kind in dynamics:
+#                                for m_excl in me:
+#                                    run_dynamics(alpha,lam,k,sample_amount,gens,runs,states,messages,learning_parameter,kind,m_excl) 
+#
 #### This is all that is needed to run either replication only, mutation only or both together #####
 
-############################### Decomment for testing individual components ########################################
-#import numpy as np
-#from lexica import get_lexica,get_prior
-#from player import LiteralPlayer,GriceanPlayer
-#from rmd import get_utils,get_mutation_matrix,run_dynamics
-#import sys 
-#import datetime
-#import csv
-#import os.path
-#
-#lam = 30
-#alpha = 1
-#
-#
-#state_freq = np.ones(states) / float(states) #frequency of states s_1,...,s_n 
-#
-##### Auxiliary functions #####
-#def m_max(m): #aux function for convenience
-#    return np.unravel_index(m.argmax(), m.shape)
-#
-#print '#Starting, ', datetime.datetime.now()
-#lexica = get_lexica(states,messages,mutual_exclusivity=True)
-#l_prior = get_prior(lexica)
-#typeList = [LiteralPlayer(lam,lex) for lex in lexica] + [GriceanPlayer(alpha,lam,lex) for lex in lexica]
-#
-#likelihoods = np.array([t.sender_matrix for t in typeList])
 
-#u = get_utils(typeList,states,messages,lam,alpha)
+
+
+############################### Decomment for testing individual components ########################################
[…]

### 2017-02-08 code-LOT-extension/lexica.py

@@ -1,12 +1,12 @@
 ##generation of lexica and prior
 import numpy as np
-from itertools import product,combinations
+from itertools import product,combinations,combinations_with_replacement
 
 def get_lexica(s_amount,m_amount,mutual_exclusivity=True):
     columns = list(product([0.,1.],repeat=s_amount))
+    columns.remove((0,0,0)) #remove message false of all states
+    columns.remove((1,1,1)) #remove message true of all states
     if mutual_exclusivity:
-        columns.remove((0,0,0)) #remove message false of all states
-        columns.remove((1,1,1)) #remove message true of all states
         matrix = list(combinations(columns,r=m_amount)) #no concept assigned to more than one message
         out = []
         for mrx in matrix:
@@ -14,8 +14,8 @@ def get_lexica(s_amount,m_amount,mutual_exclusivity=True):
             lex = np.transpose(np.array([mrx[i] for i in xrange(s_amount)]))
             out.append(lex)
     else:
-        columns.remove((0,0,0)) #remove message false of all states
-        matrix = list(product(columns,repeat=m_amount)) 
+#        matrix = list(product(columns,repeat=m_amount)) #If we allow for symmetric lexica
+        matrix = list(combinations_with_replacement(columns,m_amount)) 
         out = []
         for mrx in matrix:
             lex = np.array([mrx[i] for i in xrange(s_amount)])

### 2017-02-08 code-LOT-extension/plots.py

@@ -12,68 +12,85 @@ import sys
 #####################
 
 
-def get_prior_plot():
+def get_prior_plot(m_excl):
     from lexica import get_prior, get_lexica
-    prior = get_prior(get_lexica(3,3,mutual_exclusivity=True))
+    prior = get_prior(get_lexica(3,3,m_excl))
     
-    X = np.arange(40)
-    Y_target = [0 for _ in xrange(24)] + [prior[24]] + [0 for _ in xrange(15)]
-    print prior[24]
-    prior[24] = 0
+    X = np.arange(len(prior))
+    if m_excl:
+        target = 24
+    else:
+        target = 68
+    Y_target = [0 for _ in xrange(target)] + [prior[target]] + [0 for _ in xrange(len(prior)-target-1)]
+    
+    prior[target] = 0
     Y_rest = prior
-    print prior[24]
     
     fig, ax = plt.subplots()
     ax.bar(X,Y_target,width=1, color='green')
     ax.bar(X,Y_rest,width=1)
     plt.show()
 
-#get_prior_plot()
+#get_prior_plot(False)
 
-def get_utility_heatmap(s_amount,m_amount,lam,alpha):
+def get_utility_heatmap(s_amount,m_amount,lam,alpha,m_excl):
     print 'Loading U-matrix'
-    df = pd.read_csv('./matrices/umatrix-s%d-m%d-lam%d-a%d.csv' %(s_amount,m_amount,lam,alpha))
-    yticks = [0] + ["" for _ in xrange(9)] + [10] + ["" for _ in xrange(9)] +\
-             [20] + ["" for _ in xrange(9)] + [30] + ["" for _ in xrange(9)]
+    df = pd.read_csv('./matrices/umatrix-s%d-m%d-lam%d-a%d-me%s.csv' %(s_amount,m_amount,lam,alpha,str(m_excl)))
     
-    ax = sns.heatmap(df, yticklabels=yticks, xticklabels=yticks)#, annot=True) 
+    axlabels=["" for _ in xrange(df.shape[1])]
+    show = np.arange(0,df.shape[1]+1,10)
+    for i in xrange(len(show)):
+        axlabels[show[i]] = show[i]
+
+    ax = sns.heatmap(df, xticklabels=axlabels, yticklabels=axlabels)#, annot=True) 
+    ax.invert_yaxis()
+
+
 #    plt.yticks(rotation=0)
[…]

### 2017-02-08 code-LOT-extension/rmd.py

@@ -50,11 +50,11 @@ def get_likelihood(obs,likelihoods):
             out[lhi,o] = np.prod([flat_lhi[x]**obs[o][x] for x in xrange(len(obs[o]))]) 
     return out
 
-def get_mutation_matrix(s_amount,m_amount,state_freq, likelihoods,lexica_prior,learning_parameter,sample_amount,k,lam,alpha):
+def get_mutation_matrix(s_amount,m_amount,state_freq, likelihoods,lexica_prior,learning_parameter,sample_amount,k,lam,alpha,mutual_exclusivity):
 
-    if os.path.isfile('./matrices/qmatrix-s%d-m%d-lam%d-a%d-k%d-samples%d-l%d.csv' %(s_amount,m_amount,lam,alpha,k,sample_amount,learning_parameter)):
+    if os.path.isfile('./matrices/qmatrix-s%d-m%d-lam%d-a%d-k%d-samples%d-l%d-me%s.csv' %(s_amount,m_amount,lam,alpha,k,sample_amount,learning_parameter,str(mutual_exclusivity))):
         print '#Loading mutation matrix, ', datetime.datetime.now()
-        return np.genfromtxt('./matrices/qmatrix-s%d-m%d-lam%d-a%d-k%d-samples%d-l%d.csv' %(s_amount,m_amount,lam,alpha,k,sample_amount,learning_parameter), delimiter=',')
+        return np.genfromtxt('./matrices/qmatrix-s%d-m%d-lam%d-a%d-k%d-samples%d-l%d-me%s.csv' %(s_amount,m_amount,lam,alpha,k,sample_amount,learning_parameter,str(mutual_exclusivity)), delimiter=',')
     else:
         print '#Computing mutation matrix, ', datetime.datetime.now()
     
@@ -71,17 +71,17 @@ def get_mutation_matrix(s_amount,m_amount,state_freq, likelihoods,lexica_prior,l
             out[parent_type] = np.dot(np.transpose(lhs[parent_type]),parametrized_post)
     
         q = normalize(out)
-        f_q = csv.writer(open('./matrices/qmatrix-s%d-m%d-lam%d-a%d-k%d-samples%d-l%d.csv' %(s_amount,m_amount,lam,alpha,k,sample_amount,learning_parameter),'wb'))
[…]

### 2017-02-13 code-LOT-extension/dynamics.py
lexica as concept permutation + some results

@@ -17,89 +17,89 @@ states = 3 #number of states
 messages = 3 #number of messages
 me = [False] #mutual exclusivity
 
-#for alpha in a:
-#    for lam in lamb:
-#        for k in seq_length:
-#            for sample_amount in samples:
-#                for learning_parameter in l:
-#                    for gens in g:
-#                        for runs in r:
-#                            for kind in dynamics:
-#                                for m_excl in me:
-#                                    run_dynamics(alpha,lam,k,sample_amount,gens,runs,states,messages,learning_parameter,kind,m_excl) 
-#
+for alpha in a:
+    for lam in lamb:
+        for k in seq_length:
+            for sample_amount in samples:
+                for learning_parameter in l:
+                    for gens in g:
+                        for runs in r:
+                            for kind in dynamics:
+                                for m_excl in me:
+                                    run_dynamics(alpha,lam,k,sample_amount,gens,runs,states,messages,learning_parameter,kind,m_excl) 
+
 #### This is all that is needed to run either replication only, mutation only or both together #####
 
 
 
 
 ############################### Decomment for testing individual components ########################################
-import numpy as np
-from lexica import get_lexica,get_prior
-from player import LiteralPlayer,GriceanPlayer
-from rmd import get_utils,get_mutation_matrix,run_dynamics
-import sys 
-import datetime
-import csv
-import os.path
-
-lam = 30
-alpha = 1
-mutual_exclusivity=False
-
-
-state_freq = np.ones(states) / float(states) #frequency of states s_1,...,s_n 
-
-#### Auxiliary functions #####
-def m_max(m): #aux function for convenience
-    return np.unravel_index(m.argmax(), m.shape)
-
-print '#Starting, ', datetime.datetime.now()
-lexica = get_lexica(states,messages,mutual_exclusivity=False)
-l_prior = get_prior(lexica)
-typeList = [LiteralPlayer(lam,lex) for lex in lexica] + [GriceanPlayer(alpha,lam,lex) for lex in lexica]
-
[…]

### 2017-02-13 code-LOT-extension/lexica.py

@@ -14,14 +14,43 @@ def get_lexica(s_amount,m_amount,mutual_exclusivity=True):
             lex = np.transpose(np.array([mrx[i] for i in xrange(s_amount)]))
             out.append(lex)
     else:
-#        matrix = list(product(columns,repeat=m_amount)) #If we allow for symmetric lexica
-        matrix = list(combinations_with_replacement(columns,m_amount)) 
+        matrix = list(product(columns,repeat=m_amount)) #If we allow for symmetric lexica
+#        matrix = list(combinations_with_replacement(columns,m_amount)) #only 112 types
         out = []
         for mrx in matrix:
             lex = np.array([mrx[i] for i in xrange(s_amount)])
             lex = np.transpose(np.array([mrx[i] for i in xrange(s_amount)]))
             out.append(lex)
     return out 
+
+def get_lexica_bins(lexica_list):
+    concepts = [[0,0,1],[0,1,0],[0,1,1],\
+                    [1,0,0],[1,0,1],[1,1,0]]
+    lexica_concepts = []
+    for lex_idx in xrange(len(lexica_list)):
+        concept_indices = []
+        current_lex = np.transpose(lexica_list[lex_idx])
+        for concept_idx in xrange(len(current_lex)):
+            concept_indices.append(concepts.index(list(current_lex[concept_idx])))
+        lexica_concepts.append(concept_indices)
+    
+    bin_counter = []
+    bins = []
+    for lex_idx in xrange(len(lexica_list)):
+        sorted_lexica_concepts = lexica_concepts[lex_idx]
+        sorted_lexica_concepts.sort()
+        if not(sorted_lexica_concepts in bin_counter):
+            bin_counter.append(sorted_lexica_concepts)
+            bins.append([lex_idx])
+        else:
+            bins[bin_counter.index(sorted_lexica_concepts)].append(lex_idx)
+    ### up to here we get bins for a single linguistic behavior. Now we double that for Literal/Gricean split 
+    gricean_bins = []
+    for b in bins:
+        g_bin = [x+len(lexica_list) for x in b]
+        gricean_bins.append(g_bin)
+    bins = bins+gricean_bins
+    return bins
 
 def get_prior(lexica_list):
     concepts = [[0,0,1],[0,1,0],[0,1,1],\

### 2017-02-13 code-LOT-extension/rmd.py

@@ -5,7 +5,7 @@ np.set_printoptions(threshold=np.nan)
 from random import sample
 from itertools import product
 from player import LiteralPlayer,GriceanPlayer
-from lexica import get_lexica,get_prior
+from lexica import get_lexica,get_prior,get_lexica_bins
 
 import sys 
 import datetime
@@ -97,6 +97,12 @@ def get_utils(typeList,states,messages,lam,alpha,mutual_exclusivity):
 
         return out
 
+def get_type_bin(type_idx,bins):
+    for b in xrange(len(bins)):
+        if type_idx in bins[b]:
+            return b
+    
+
 def run_dynamics(alpha,lam,k,sample_amount,gens,runs,states,messages,learning_parameter,kind,mutual_exclusivity):
 
     state_freq = np.ones(states) / float(states) #frequency of states s_1,...,s_n 
@@ -104,6 +110,7 @@ def run_dynamics(alpha,lam,k,sample_amount,gens,runs,states,messages,learning_pa
     print '#Starting, ', datetime.datetime.now()
     
     lexica = get_lexica(states,messages,mutual_exclusivity)
+    bins = get_lexica_bins(lexica) #To bin types with identical lexica
     l_prior = get_prior(lexica)
     typeList = [LiteralPlayer(lam,lex) for lex in lexica] + [GriceanPlayer(alpha,lam,lex) for lex in lexica]
     
@@ -150,6 +157,9 @@ def run_dynamics(alpha,lam,k,sample_amount,gens,runs,states,messages,learning_pa
     f_mean.writerow([kind,str(lam),str(alpha),str(k),str(sample_amount),str(learning_parameter),str(gens),str(runs),str(mutual_exclusivity)] +\
                         [str(p_mean[x]) for x in xrange(len(typeList))])
     
+    
+    inc = np.argmax(p_mean)
+    inc_bin = get_type_bin(inc,bins)
 
     print 
     print '##### Mean results#####'
@@ -157,8 +167,11 @@ def run_dynamics(alpha,lam,k,sample_amount,gens,runs,states,messages,learning_pa
     print 'dynamics= %s, alpha = %d, lambda = %d, k = %d, samples per type = %d, learning parameter = %.2f, generations = %d, runs = %d' % (kind, alpha, lam, k, sample_amount, learning_parameter, gens, runs)
     print '#######################'
     print 
-    print 'Incumbent type:', np.argmax(p_mean), ' with proportion ', p_mean[np.argmax(p_mean)]
[…]

### 2017-02-13 paper/new_simulation_setup.tex
new report draft + plots

@@
   \centering
-  \includegraphics[scale=.35]{../code-LOT-extension/plots/priors-meFalse.png} %width = 0.75\linewidth
+  \includegraphics[scale=.35]{../code-LOT-extension/plots/priors-meFalse112.png} %width = 0.75\linewidth
   \caption{Prior over types (lexica). The green prior corresponds to $t_{68}$.}
@@
   \centering
-  \includegraphics[scale=.25]{../code-LOT-extension/plots/u-l30-a1-meFalse.png} %width = 0.75\linewidth
+  \includegraphics[scale=.25]{../code-LOT-extension/plots/u-l30-a1-meFalse112.png} %width = 0.75\linewidth
   \caption{Utility-Matrix for $\alpha = 1$ and $\lambda = 30$. The boxed value corresponds to $\text{EU}(t_{68},t_{68})$.}
@@
   \begin{subfigure}[b]{0.45\textwidth}
-    \includegraphics[scale=0.4]{../code-LOT-extension/plots/q-l30-a1-k5-l1-meFalse.png}
+    \includegraphics[scale=0.4]{../code-LOT-extension/plots/q-l30-a1-k5-l1-meFalse112.png}
   \end{subfigure}
@@
    \begin{subfigure}[b]{0.45\textwidth}
-    \includegraphics[scale=0.4]{../code-LOT-extension/plots/q-l30-a1-k5-l10-meFalse.png}
+    \includegraphics[scale=0.4]{../code-LOT-extension/plots/q-l30-a1-k5-l10-meFalse112.png}
   \end{subfigure}
@@
   \begin{subfigure}[b]{0.45\textwidth}
-    \includegraphics[scale=0.4]{../code-LOT-extension/plots/rmd-lam30-a1-k5-s200-l1-meFalse.png}
+    \includegraphics[scale=0.4]{../code-LOT-extension/plots/rmd-lam30-a1-k5-s200-l1-meFalse112.png}
   \end{subfigure}
@@
    \begin{subfigure}[b]{0.45\textwidth}
-    \includegraphics[scale=0.4]{../code-LOT-extension/plots/rmd-lam30-a1-k5-s200-l10-meFalse.png}
+    \includegraphics[scale=0.4]{../code-LOT-extension/plots/rmd-lam30-a1-k5-s200-l10-meFalse112.png}
   \end{subfigure}
@@
 
+\newpage
+
+\section*{Details for an even larger type space}
+
+Allowing for all possible concept permutations gives us $216$ lexica and accordingly $432$ types.
+Binning equivalent lexica of identical signaling behavior gives us $112$ type-bins.
+There are $6$ concept orderings that are equivalent to our Gricean target type; $t_{231}$, $t_{236}$, $t_{291}$, $t_{306}$, $t_{326}$ and $t_{336}$.
+When binned, I will refer to this class of types as $b_{68}$.
+
+Following the setup above and the complexity values in Table \ref{tab:concepts}, the priors of our now $432$ types are shown here:
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[scale=.35]{../code-LOT-extension/plots/priors-meFalse432.png} %width = 0.75\linewidth
+  \caption{Prior over types (lexica). The red priors corresponds to our targets.}
+  \label{fig:priors}
+\end{figure}
+
+Note however that, when binned (summing values of types corresponding to each bin), our target bin is again the most probable.
+The main reason being that there are comparably many types in this bin (some have only one type).
+\begin{figure}[h!]
+  \centering
+  \includegraphics[scale=.35]{../code-LOT-extension/plots/priors-binned-meFalse432.png} %width = 0.75\linewidth
+  \caption{Prior over bins. The red prior corresponds to our target bin.}
+  \label{fig:priors}
+\end{figure}
+
+Inspecting the entire utility or mutation matrix visually makes little sense.
+Let us instead first compare the {\bf mean} results of replication only, mutation only, and replication and mutation together.
+Both with the full space and with bins.
+As before, the graphs show the mean top $3$ types.
+This time, I refrained from adding our target if it wasn't in the top three.
+Full space first:
+
+%
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[scale=0.25]{../code-LOT-extension/plots/rmd-l30-a1-k5-l1-meFalse432.png}
+  \end{subfigure}
+  ~
+   \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[scale=0.25]{../code-LOT-extension/plots/rmd-l30-a1-k5-l10-meFalse432.png}
+  \end{subfigure}
+  \caption{Mean proportion of top $3$ under replication only (green plot), mutation only (blue plot), and replication {\em and} mutation (red plot) after $50$ generations with $\alpha = 1, \lambda = 30, k = 5$, $200$ samples per parent type and $l = 1$ (left figure) and $l = 10$ (right figure).}
+  \label{fig:outcomes}
+\end{figure}
+%
+
+What these results suggest (if you squint) is that: (i) as before, replication alone does not help our targets much, (ii) mutation helps, particularly when $l$ is high, and (iii) replication and mutation only seems to do slightly better than mutation alone.
+Note also that the proportion of the incumbent types is pretty low for these parameter values.
+My first thought was that we may not see the effects very clearly because we are looking only at the top $3$ but actually have $6$ targets.
+Therefore, I looked at the binned state space (recall $b_{68}$ is our target):
+
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[scale=0.25]{../code-LOT-extension/plots/rmd-l30-a1-k5-l1-meFalse432_binned.png}
+  \end{subfigure}
+  ~
+   \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[scale=0.25]{../code-LOT-extension/plots/rmd-l30-a1-k5-l10-meFalse432_binned.png}
+  \end{subfigure}
+  \caption{Mean proportion of top $3$ under replication only (green plot), mutation only (blue plot), and replication {\em and} mutation (red plot) after $50$ generations with $\alpha = 1, \lambda = 30, k = 5$, $200$ samples per parent type and $l = 1$ (left figure) and $l = 10$ (right figure).}
+\end{figure}
+
+This pattern is very similar to the one before.
+We get higher incumbeny proportions but replication seems to add little to what mutation alone already achieves.
+However, we are looking at the mean across $1000$ simulations -- binned or not -- instead of individual population trajectories.
+This seems to make all the difference.
+I haven't had a good idea how to visually show this for all simulations and all types, but it's already instructive to look at a subset of just the first $30$ independent simulations and only our targets.
+
+Here's the trajectory of $30$ populations under only replication:
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[scale=.45]{../code-LOT-extension/plots/r-lam30-a1-k15-s200-l10-meFalse432-nomean.png} %width = 0.75\linewidth
+  \caption{Proportions of target types in $30$ independent populations under replication after $50$ generations ($\lambda = 20, \alpha = 1)$.}
+\end{figure}
+
+The only thing to note here seems to be that replication does not favor any particular target and, in each population, rules out all but one of them. Contrast this with mutation:
+
+\begin{figure}[h!]
+  \centering
+    \includegraphics[scale=.8]{../code-LOT-extension/plots/m-lam30-a1-k15-s200-l10-meFalse432-nomean.png}
+  \caption{Mean proportion of target types in $30$ independent populations under mutation after $50$ generations with $\alpha = 1, \lambda = 30, k = 15$, $200$ samples per parent type and $l = 10$.}
+\end{figure}
+
+This is pretty messy, but the take-home message is that mutation alone leads to cohabitation of our targets.
+A similar pattern seems to obtain for other values of $l$ and $k$.
+Finally, we can look at replication and mutation:
+
+\begin{figure}[h!]
+  \centering
+    \includegraphics[scale=0.8]{../code-LOT-extension/plots/rm-lam30-a1-k15-s200-l10-meFalse432-nomean.png}
+  \caption{Mean proportion of target types in $30$ independent populations under mutation after $50$ generations with $\alpha = 1, \lambda = 30, k = 15$, $200$ samples per parent type and $l = 10$.}
+\end{figure}
+
+And what we see is that, indeed, in each population only one of our types is an incumbent under replication {\em and} mutation under good learning conditions.
+We still see a similar trend for lower $l$ and $k$, but a bit more noisy, as expected.
+For example, with $l = 1$ and $k = 5$:
+
+\begin{figure}[h!]
+  \centering
+    \includegraphics[scale=0.4]{../code-LOT-extension/plots/rm-lam30-a1-k5-s200-l1-meFalse432-nomean.png}
+  \caption{Mean proportion of target types in $30$ independent populations under mutation after $50$ generations with $\alpha = 1, \lambda = 30, k = 5$, $200$ samples per parent type and $l = 1$.}
+\end{figure}
+
 \bibliographystyle{plainnat}

### 2017-02-13 code-LOT-extension/lexica.py

@@ -52,6 +52,9 @@ def get_lexica_bins(lexica_list):
     bins = bins+gricean_bins
     return bins
 
+lex = get_lexica(3,3,False)
+bins = get_lexica_bins(lex)
+
 def get_prior(lexica_list):
     concepts = [[0,0,1],[0,1,0],[0,1,1],\
                     [1,0,0],[1,0,1],[1,1,0]]

### 2017-02-13 code-LOT-extension/plots.py

@@ -12,178 +12,166 @@ import sys
 #####################
 
 
+
+
 def get_prior_plot(m_excl):
     from lexica import get_prior, get_lexica
+
     prior = get_prior(get_lexica(3,3,m_excl))
     
     X = np.arange(len(prior))
     if m_excl:
         target = 24
     else:
-        target = 68
-    Y_target = [0 for _ in xrange(target)] + [prior[target]] + [0 for _ in xrange(len(prior)-target-1)]
-    
-    prior[target] = 0
+        targets = [231,236,291,306,326,336]
+       
+    Y_target = [0 for _ in xrange(len(prior))]
+    for target in targets:
+        Y_target[target] = prior[target]
+        prior[target] = 0
+   
     Y_rest = prior
     
     fig, ax = plt.subplots()
-    ax.bar(X,Y_target,width=1, color='green')
+    ax.bar(X,Y_target,width=1, color='red')
     ax.bar(X,Y_rest,width=1)
     plt.show()
 
 #get_prior_plot(False)
 
-def get_utility_heatmap(s_amount,m_amount,lam,alpha,m_excl):
-    print 'Loading U-matrix'
-    df = pd.read_csv('./matrices/umatrix-s%d-m%d-lam%d-a%d-me%s.csv' %(s_amount,m_amount,lam,alpha,str(m_excl)))
+def get_binned_prior_plot(m_excl):
+    from lexica import get_prior, get_lexica, get_lexica_bins
+    lexica = get_lexica(3,3,m_excl)
+    prior = get_prior(lexica)
+    bins = get_lexica_bins(lexica)
+
+    binned_prior = []
+    for b in xrange(len(bins)):
+        binned_prior.append(np.sum([prior[x] for x in bins[b]]))
+
+    X = np.arange(len(binned_prior))
     
-    axlabels=["" for _ in xrange(df.shape[1])]
-    show = np.arange(0,df.shape[1]+1,10)
-    for i in xrange(len(show)):
-        axlabels[show[i]] = show[i]
-
-    ax = sns.heatmap(df, xticklabels=axlabels, yticklabels=axlabels)#, annot=True) 
-    ax.invert_yaxis()
-
-
-#    plt.yticks(rotation=0)
[…]

### 2017-02-14 paper/new_simulation_setup.tex
new summary of results + plots for type space with 432 types

--- 2017-02-08 paper/new_simulation_setup.tex
@@
 
+\newpage
+
+\section*{Details for an even larger type space}
+
+Allowing for all possible concept permutations gives us $216$ lexica ($432$ types).
+There are $6$ concept orderings that are equivalent to our previous Gricean target type; $t_{231}$, $t_{236}$, $t_{291}$, $t_{306}$, $t_{326}$ and $t_{336}$.
+Binning equivalent lexica of identical signaling behavior gives us $112$ type-bins.
+When binned, I will refer to the bin of our targets as $b_{68}$.
+
+Following the setup above and the complexity values in Table \ref{tab:concepts}, the priors of our now $432$ types are shown here:
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[scale=.45]{../code-LOT-extension/plots/priors-meFalse432.png} %width = 0.75\linewidth
+  \caption{Prior over types (lexica). The red priors corresponds to our targets.}
+\end{figure}
+
+Note that when binned our target bin is the most probable. The main reason being that there are comparably many types in this bin and each of them has a comparably high prior probability.
+\begin{figure}[h!]
+  \centering
+  \includegraphics[scale=.45]{../code-LOT-extension/plots/priors-binned-meFalse432.png} %width = 0.75\linewidth
+  \caption{Prior over bins. The red prior corresponds to our target bin.}
+\end{figure}
+
+\newpage
+Inspecting the full utility or mutation matrix visually makes little sense.
+Let us instead first compare the {\bf mean} results of (i) replication only, (ii) mutation only, and (iii) replication and mutation together (over $1000$ independent populations).
+As before, the graphs show the mean top $3$ types.
+This time, I refrained from adding our target if it wasn't in the top three (as we have many to chose from).
+Here are some results without binning:
+
+%
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[scale=0.4]{../code-LOT-extension/plots/rmd-l30-a1-k5-l1-meFalse432.png}
+  \end{subfigure}
+  ~
+   \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[scale=0.4]{../code-LOT-extension/plots/rmd-l30-a1-k5-l10-meFalse432.png}
+  \end{subfigure}
+  \caption{Mean proportion of top $3$ under replication only (green plot), mutation only (blue plot), and replication {\em and} mutation (red plot) after $50$ generations with $\alpha = 1, \lambda = 30, k = 5$, $200$ samples per parent type and $l = 1$ (left figure) and $l = 10$ (right figure).}
+  \label{fig:outcomes}
+\end{figure}
+%
+
+This seems to suggest that: (i) as before, replication alone does not help our targets much, (ii) mutation helps, particularly when $l$ is high, and (iii) replication and mutation only seems to do slightly better than mutation alone.
+Note also that the proportion of the incumbent types is low for these parameters compared to what we got in smaller spaces.
+My first thought was that we may not see the effects very clearly because we are looking only at the top $3$, but actually have $6$ targets.
+Therefore, I looked at the binned state space (recall $b_{68}$ is our target):
+
+\begin{figure}[h!]
+  \centering
+  \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[scale=0.4]{../code-LOT-extension/plots/rmd-l30-a1-k5-l1-meFalse432_binned.png}
+  \end{subfigure}
+  ~
+   \begin{subfigure}[b]{0.45\textwidth}
+    \includegraphics[scale=0.4]{../code-LOT-extension/plots/rmd-l30-a1-k5-l10-meFalse432_binned.png}
+  \end{subfigure}
+  \caption{Mean proportion of top $3$ under replication only (green plot), mutation only (blue plot), and replication {\em and} mutation (red plot) after $50$ generations with $\alpha = 1, \lambda = 30, k = 5$, $200$ samples per parent type and $l = 1$ (left figure) and $l = 10$ (right figure).}
+\end{figure}
+
+This pattern is very similar to the one before.
+We get more monomorphic populations but replication seems to add little to what mutation alone already achieves.
+Now for the punch line.
+So far, we have looked at the mean across $1000$ simulations -- binned or not -- instead of individual population trajectories.
+This seems to make all the difference.
+I don't know yet how to best visualize this, so let us focus at a subset of just the first $30$ independent populations and only our targets.
+
+Here's the development of $30$ populations under only replication (y-axis gives the population \#):
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[scale=.4]{../code-LOT-extension/plots/r-lam30-a1-k15-s200-l10-meFalse432-nomean.png} %width = 0.75\linewidth
+  \caption{Proportions of target types in $30$ independent populations under replication after $50$ generations ($\lambda = 20, \alpha = 1)$.}
+\end{figure}
+
+The main thing to note here is that replication does not favor any particular target type, and, more importantly, replication rules out all targets but one in each individual population.
+Contrast this with mutation:
+
+\newpage
+
+\begin{figure}[h!]
+  \centering
+    \includegraphics[scale=.65]{../code-LOT-extension/plots/m-lam30-a1-k15-s200-l10-meFalse432-nomean.png}
+  \caption{Mean proportion of target types in $30$ independent populations under mutation after $50$ generations with $\alpha = 1, \lambda = 30, k = 15$, $200$ samples per parent type and $l = 10$.}
+\end{figure}
+
+So it seems that mutation alone leads to cohabitation of our targets.
+This also explains why the mean results for mutation only were so high compared to replication only.
+Similar patterns seems to obtain for other values of $l$ and $k$.
+Finally, we can look at replication and mutation:
+\newpage
+\begin{figure}[h!]
+  \centering
+    \includegraphics[scale=0.65]{../code-LOT-extension/plots/rm-lam30-a1-k15-s200-l10-meFalse432-nomean.png}
+  \caption{Mean proportion of target types in $30$ independent populations under mutation after $50$ generations with $\alpha = 1, \lambda = 30, k = 15$, $200$ samples per parent type and $l = 10$.}
+\end{figure}
+
+And what we see is that in each population only one of our types is an incumbent under replication {\em and} mutation under good learning conditions.
+We still see a similar trend for lower $l$ and $k$ with more ``target cohabitation'' because transmission is noisier.
+For example, with $l = 1$ and $k = 5$:
+
+\begin{figure}[h!]
+  \centering
+    \includegraphics[scale=0.35]{../code-LOT-extension/plots/rm-lam30-a1-k5-s200-l1-meFalse432-nomean.png}
+  \caption{Mean proportion of target types in $30$ independent populations under mutation after $50$ generations with $\alpha = 1, \lambda = 30, k = 5$, $200$ samples per parent type and $l = 1$.}
+\end{figure}
+
+\newpage
 \bibliographystyle{plainnat}

### 2017-02-14 code-LOT-extension/plots.py

@@ -134,12 +134,12 @@ def get_bins_and_mean(group1,group2,group3,m_excl):
         idx_bin_rm.append(index)
     
     
-    X = np.arange(21)
-    empty = [0 for _ in xrange(3)]
+    X = np.arange(11)
+    empty = [0 for _ in xrange(1)]
     
-    Y_r = empty + r_mean_top + empty * 5
-    Y_m = empty * 3 + m_mean_top + empty * 3
-    Y_rm = empty * 5 + rm_mean_top + empty
+    Y_r = r_mean_top + empty * 8
+    Y_m = empty * 4 + m_mean_top + empty * 4
+    Y_rm = empty * 8 + rm_mean_top
     
     
     fig, ax = plt.subplots()
@@ -151,16 +151,18 @@ def get_bins_and_mean(group1,group2,group3,m_excl):
     m_labels = ['t'+str(x) for x in idx_mean_m]
     rm_labels = ['t'+str(x) for x in idx_mean_rm]
     
-    ax.set_xticks(xrange(21))
-    ax.set_xticklabels(['', '', '', r_labels[0], r_labels[1],r_labels[2], '', '', '', m_labels[0], m_labels[1],m_labels[2], '', '', '',\
-                        rm_labels[0], rm_labels[1],rm_labels[2], '', '', ''])
+    ax.set_xticks(xrange(11))
+    ax.set_xticklabels([r_labels[0], r_labels[1],r_labels[2], '', m_labels[0], m_labels[1],m_labels[2], '', \
+                        rm_labels[0], rm_labels[1],rm_labels[2]])
     
+
+
     plt.show()
 
     ### Bins ###
-    Y_r = empty + r_bin_top+ empty * 5
-    Y_m = empty * 3 + m_bin_top + empty * 3
-    Y_rm = empty * 5 + rm_bin_top + empty
+    Y_r = r_bin_top + empty * 8
+    Y_m = empty * 4 + m_bin_top + empty * 4
+    Y_rm = empty * 8 + rm_bin_top
     
     
     fig, ax = plt.subplots()
@@ -172,16 +174,16 @@ def get_bins_and_mean(group1,group2,group3,m_excl):
     m_labels = ['b'+str(x) for x in idx_bin_m]
     rm_labels = ['b'+str(x) for x in idx_bin_rm]
     
[…]

### 2017-03-08 paper/paper-v3.tex
revision up to Analysis

--- 2016-11-24 paper/paper-v2.tex
@@
 \newcommand{\mystate}[1]{\ensuremath{\state_{\text{#1}}}\xspace} %meaningful states
+\newcommand{\mylang}[1]{\ensuremath{L_{\text{#1}}}\xspace} %meaningful states
 \newcommand{\Messgs}{\ensuremath{M}\xspace}		% Set of Messages
@@
 \newcommand{\sall}{\mystate{\ensuremath{\forall}}}
+\newcommand{\snone}{\mystate{\ensuremath{\emptyset}}}
 \newcommand{\msome}{\mymessg{some}}
 \newcommand{\mall}{\mymessg{all}}
+\newcommand{\mnone}{\mymessg{none}}
+\newcommand{\Lbound}{\mylang{bound}}
+\newcommand{\Llack}{\mylang{lack}}
 \newcommand{\asome}{\myact{\ensuremath{\exists\neg\forall}}}
 \newcommand{\aall}{\myact{\ensuremath{\forall}}}
+
 \definecolor{mygray}{cmyk}{0.35,0.35,0.35,0.35}
@@
 \begin{abstract}
-  According to standard linguistic theory, the meaning of an utterance is the product of a
-  conventional semantic meaning of the used expression and general pragmatic reasoning applied
-  to the context of utterance. This implies that models of the cultural evolution of meaning
+  According to standard linguistic theory, the meaning of an utterance is the product of 
+  conventional semantic meaning and general pragmatic reasoning applied
+  to the context of utterance. This implies that models of cultural evolution of meaning
   should likewise take into consideration that observable language use is a complex interaction
   of semantic representations and pragmatic use. To this end, we present a game theoretic model
-  of the cultural evolution of language where communicative pressures work on abstract semantic
+  of cultural language evolution where communicative pressures work on abstract semantic
   representations and pragmatic patterns of use. Our model traces two evolutionary forces and
@@
   systematic transmission perturbations when linguistic knowledge is transferred from one agent
-  to another. The latter can arise from general cognitive or learning biases, but also from
-  other sources of systematic noise, e.g., as a result of errors in perception. We illustrate
-  the model based on a case study showing that cognitive biases that favor simple semantic
-  representations can prevent the lexicalization of pragmatic inferences. We also show that,
-  technically speaking, it is possible that environmental factors, such as perceptual errors
-  during acquisition, can produce evolutionary outcomes that look as if such cognitive biases
-  are present even if they are not.
+  to another.  We illustrate the model based on a case study showing that learning biases that favor simple semantic
+  representations can prevent the lexicalization of pragmatic inferences as long as mutual reasoning counteracts a potential loss in expressivity from such semantics. 
 \end{abstract}
@@
 between the literal meaning of expressions and what they are intended and interpreted to convey
-is due to \citet{grice:1975}. It characterizes pragmatic use and interpretation as a process of
+is due to \citet{grice:1975}, characterizing pragmatic use and interpretation as a process of
 mutual reasoning about rational language use. For instance, under the assumption that the
@@
 \citep{horn:1984,Hirschberg1985:A-Theory-of-Sca,LevinsonPragmatics1983,Geurts2010:Quantity-Implic}. Usually,
-an utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
+the utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
 that the speaker does not own all of them. This is because, if the speaker had them all, he
-could have used the stronger word \emph{all} instead of \emph{some} in his utterance, and
-thereby would have made a more informative statement. Scalar implicatures, especially the
+could have used the stronger word \emph{all} instead of \emph{some} in his utterance, 
+thereby making a more informative statement. Scalar implicatures, especially the
 inference from \emph{some} to \emph{some but not all}, have been studied extensively, both
 theoretically
-\citep[e.g.][]{Sauerland2004:Scalar-Implicat,ChierchiaFox2008:The-Grammatical,Rooyvan-RooijJagerde-Jager2012:Explaining-Quan}
+\citep[e.g.,][]{Sauerland2004:Scalar-Implicat,ChierchiaFox2008:The-Grammatical,Rooyvan-RooijJagerde-Jager2012:Explaining-Quan}
 as well as experimentally
-\citep[e.g.][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. While
+\citep[e.g.,][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. While
 there is much dispute in this domain about many interesting details, a position endorsed by a
@@
 and there are mathematical models of language evolution, mostly coming from game theory
-\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
-work has focused on explaining basic properties such as compositionality and combinatoriality
-\citep[e.g.][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax}, but
+\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}.
+Much of this work has focused on explaining basic properties such as compositionality and combinatoriality
+\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax}, but
 little attention has been paid to the interaction between conventional meaning and pragmatic
@@
 communicative efficiency and (ii) possible infidelity in the transmission of linguistic
-knowledge, caused by factors such as inductive learning biases or systematic perceptual errors. Considering
+knowledge, caused by factors such as inductive learning biases and general aspects of learnability. Considering
 transmission of linguistic knowledge is important because neither semantic meanings nor
@@
 show that inductive learning biases of Bayesian learners that favor simpler lexical meanings
-can lead to this outcome. Additionally, we show that a similar outcome can be reached 
-without assuming any cognitive biases, but simply as an epiphenomenon of systematic disturbances from
-environmental factors. This formal results highlights the frequently overlooked possibility
-that channel noise in evolutionary replication can mimic effects of inductive biases.
+can lead to this outcome.
 
 We see the main contribution of this work as conceptual and technical, not as a definite answer
-to the question why scalar implicatures emerge. The work here rather demonstrates how current
+to the question why scalar implicatures emerge. It rather demonstrates how current
 probabilistic cognitive modeling of language use and evolutionary modeling can be fruitfully
@@
 biology determines the architecture that enables and constrains the means by which they can be
-fulfilled. In the following, our focus lies on the cultural aspects, wherein processes of
+fulfilled. In the following, our focus lies on cultural aspects, wherein processes of
 linguistic change are viewed as shaped by language use and its transmission, i.e., as a result
@@
 role in synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of
-word frequency rankings as a result of competing hearer and speaker preferences (e.g. in
-\citealt{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
-  piantadosi:2014,kirby+etal:2015}). If processes of selection, such as conditional imitation
+word frequency rankings as a result of competing hearer and speaker preferences \citep[e.g., in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
+  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation
 or reinforcement, favor more communicatively efficient types of behavior, languages are driven
-towards semantic expressivity \citep[e.g.][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
+towards semantic expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
 pressure towards communicative efficiency is not the only force that shapes
-language. Learnability is another as natural languages need to be learnable to survive their
+language. Learnability is another, as natural languages need to be learnable to survive their
 faithful transmission across generations. Clearly, an unlearnable code will not make it past
@@
 
-While natural languages are pressured for both expressivity and learnability these forces may pull in opposite directions. The opposition becomes particularly clear when
+While natural languages are pressured for both expressivity and learnability these forces may pull in opposite directions. Their opposition becomes particularly clear when
 considering the extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). A language with a
@@
 in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
-$i$ before the update, (ii) the fitness $f_i$ of each type $i$ before the update, and (iii) the
+$i$ before the step, (ii) the fitness $f_i$ of each type $i$ before the step, and (iii) the
 probability $Q_{ji}$ that an agent who wants to imitate, adopt, or learn the type of an agent
@@
 see][]{Sandholm2010:Population-Game}, including conditional imitation
-\citep[e.g.][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
-learning \citep[e.g.][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. If fitness
+\citep[e.g.,][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
+learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. If fitness
 $f_i$ is the same for all types $i$, the replicator step is the identity map
@@
 a prisoner's dilemma. \mf{maybe good to refer back to the example from before if there was
-  one?} \tb{Yes!} For our purposes here, types are identified by their cognitive make-up. Since we are
+  one?} \tb{Yes!} For our present purpose, types are identified by their cognitive make-up. Since we are
 interested in the question under which conditions processes of cultural evolution will favor
@@
 short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
-same sentence with \emph{some} replaced by \emph{all}. Lexica for this case would assign a
+same sentence with \emph{some} replaced by \emph{all}.
+\tb{I left $\snone$/$\mnone$ out because they are not relevant to the strengthening of $\msome$, but that means that it needs to be introduced a bit later.
+Feel free to add it if you think that's better, exposition-wise.}
+Lexica for this case would assign a
 truth value for each state-message pair. The following two lexica exemplify the distinction
-between a lexicalized upper-bound for \emph{some} in $L_4$ and the widely assumed logical
-semantics with only a lower-bound in $L_5$. \mf{can we not give these guys more mnemonic
-  names?} \tb{``bound'' and ``lack''? They would need to be pretty short to not be too disruptive as an inline indices.}
+between a lexicalized upper-bound for \emph{some} in $\Lbound$ and the widely assumed logical
+semantics with only a lower-bound in $\Llack$.
+
 \begin{align*}
-  L_{4} & = \begin{blockarray}{lcc}
+  \Lbound & = \begin{blockarray}{lcc}
     & \mygray{\msome} & \mygray{\mall} \\
@@
   \end{blockarray} &
-  L_{5} & = \begin{blockarray}{lcc}
+  \Llack & = \begin{blockarray}{lcc}
     & \mygray{\msome} & \mygray{\mall} \\
@@
 &H_{0}(s \mid m;L) \propto pr(s) L_{sm} \label{h:level0}\\
-&S_{0}(m \mid s;L) \propto \exp(L_{sm}) \label{s:level0}
+&S_{0}(m \mid s;L) \propto \exp(\lambda \; L_{sm}) \label{s:level0}
 \end{flalign}
+
 According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state
 $s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in
-the following this prior is assumed to be uniform. A literal interpreter with lexicon $L_4$
-from above would assign $\ssome$ a probability of $H_0(\ssome \mid \msome; L_4) = 1$ after
-hearing $\msome$, while a literal interpreter with lexicon $L_5$ would assign $\ssome$
-probability $H_0(\ssome \mid \msome; L_5) = 0.5$. A literal speaker chooses any true message
-for a state with equal probability and prefers a true message over a false one. The formulation
+the following this prior is assumed to be uniform. A literal interpreter with lexicon $\Lbound$
+from above would assign $\ssome$ a probability of $H_0(\ssome \mid \msome; \Lbound) = 1$ after
+hearing $\msome$, while a literal interpreter with lexicon $\Llack$ would assign $\ssome$
+probability $H_0(\ssome \mid \msome; \Llack) = 0.5$. As usual in probabilistic pragmatics models, speaker behavior is regulated by a soft-max
+parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}. As $\lambda$
+increases, choices made in production are more rational in that higher values lead to behavior
+that is increasingly in line with expected utility maximization. Expected utility of a message
+$\messg$ in state $\state$ for a level $n+1$ speaker is here defined as $H_{n}(s|m;L)$, the probability that the hearer will assign to or choose the correct meaning.
+Put differently, how likely the hearer is thought to infer $\state$ when sent $\messg$.
+The formulation
 in (\ref{s:level0}) also allows false messages to be sent with a small positive probability in
 analogy to the definition of pragmatic speakers in probabilistic pragmatic models. This is
-necessary to guarantee a mutation matrix with only positive entries (see below). A literal
-speaker with either lexicon $L_4$ or $L_5$ produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
-\ssome;L_{4,5}) \approx .73$. The probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
-\sall;L_4) \approx .27$ for $L_4$ and $S_{0}(\msome \mid
-\sall;L_5) \approx .5$ for $L_5$. \mf{I realize that the probability of producing a false
-  message for literal speakers is actually rather high. That's iffy because our pragmatic types
-are able (given favorable parameter values) to perform much better. That means that we
-implicitly have engineered in a possible fitness advantage of pragmatic types! Should we do
-something about this? Use definition $S_{0}(m \mid s;L) \propto L_{sm} + 0.001$? Make sure we
-only look at parameters for pragmatic types that make them not much better than the literal guys?} \tb{That's true.
-We could also just let them soft-maximize.
-Let's decide on this via email.
-I'll adapt the simulation results from the next section according to what we decide.}
+necessary to guarantee a mutation matrix with only positive entries (see below). If $\lambda = 1$ a literal
+speaker with either lexicon $\Lbound$ or $\Llack$ produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
+\ssome;L_{\text{bound},\text{lack}}) \approx .73$. The probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
+\sall;\Lbound) \approx .27$ for $\Lbound$ and $S_{0}(\msome \mid
+\sall;\Llack) \approx .5$ for $\Llack$. By contrast, if $\lambda = 20$, a literal speaker with either lexicon produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
+\ssome;L_{\text{bound},\text{lack}}) \approx 1$. But the probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
+\sall;\Lbound) \approx 1$ for $\Lbound$ whereas it is still $S_{0}(\msome \mid \sall;\Llack) \approx .5$ for $\Llack$.
 
@@
 &H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
-&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)^\alpha) \label{s:leveln}
+&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)) \label{s:leveln}
 \end{flalign}
-As usual in probabilistic pragmatics models, speaker behavior is regulated by a soft-max
-parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}. As $\lambda$
-increases, choices made in production are more rational in that higher values lead to behavior
-that is increasingly in line with expected utility maximization. Expected utility of a message
-$\messg$ in state $\state$ is here defined as $H_{n}(s|m;L)^\alpha$. The probability
-$H_{n}(s|m;L)$ is the probability that the hearer will assign to or choose the correct
-meaning. The transformation with $\alpha$ allows to differentiate theoretically interesting
-speaker production behavior: If $\alpha > 1$ pragmatically deviant
-choices will get less likely than semantically deviant choices; if $\alpha < 0$ it is the other
-way around. See Appendix \ref{app:sig-example} for an example that elucidates the role of $\alpha$ \tb{Did you mean it like this?}.
 
-In sum, parameter $\lambda$ regulates the overall tendency to avoid pragmatically or semantically deviant choices, $\alpha$ further distinguishes between pragmatically and
-semantically deviant behavior. Variation in $\alpha$ is only necessary for the technical
-extension to environmental transmission noise in Section \ref{subsec:noise}.
+Particularly important to our purpose is the fact that players of a level greater than $0$ using $\Llack$ can {\em pragmatically} associate $\msome$ with $\ssome$ over $\sall$, in contrast to their literal counterparts.
+Intuitively, this is so because hearers reason a speaker to convey $\sall$ with $\mall$, as this has a higher chance of leading to mutual understanding than using underspecified $\msome$.
+Therefore they will expect $\msome$ to be more strongly associated with $\ssome$ than with $\sall$ as it is already possible to clearly convey the latter state with a different message.
+The converse reasoning pattern holds for the speaker.
 
@@
 communicate with other language users from the same population
-\citep[e.g.][]{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Under a biological
+\citep[e.g.,][]{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Under a biological
 interpretation, the assumption is that organisms have a higher chance of survival and
@@
 \end{flalign*}
-As usual, $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.
+As usual for cooperative communication, $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.
 
@@
 multiple languages may be compatible with the data learners are exposed to. These sources of
-uncertainty introduce variation in their transmission from one generation to the next. Biases
-or systematic noise in the iterated transmission process can influence language evolution
+uncertainty introduce variation in their transmission from one generation to the next. In particular, learning biases in the iterated transmission process can influence language evolution
 substantially.
@@
 Following \citet{griffiths+kalish:2007} we model language acquisition as a process of Bayesian
-inference in which learners combine the likelihood of a type producing the received learning
+inference in which learners combine the likelihood of a type producing the witnessed learning
 input with prior inductive biases. Experimental and mathematical results on iterated learning
 suggest that the outcome of this process reflects learners' inductive biases
-\citep[e.g.][]{kirby+etal:2014}. In a Bayesian setting these biases can be codified in a prior
-$P \in \Delta(T)$, which reflect the amount of data a learner requires to faithfully acquire
-the language of the teacher \citep[c.f.][450]{griffiths+kalish:2007}. The extent of the prior's
+\citep[e.g.,][]{kirby+etal:2014}. In a Bayesian setting these biases can be codified in a prior
+$P \in \Delta(T)$, which reflects the amount of data a learner requires to faithfully acquire
+the language of the teacher \citep[cf.][450]{griffiths+kalish:2007}. The extent of the prior's
 influence has been shown to heavily depend on the learning strategy assumed to underly the
 inference process. On the one hand, early simulation results suggested that weak biases could
-be magnified by exposing learners to only small data samples (e.g. in
-\citealt{brighton:2002}). On the other, \citeposs{griffiths+kalish:2007} mathematical
+be magnified by exposing learners to only small data samples \citep[e.g., in][]{brighton:2002}. On the other, \citeposs{griffiths+kalish:2007} mathematical
 characterization showed that iterated learning converged to the prior, i.e., the resulting
@@
 sequences of state-message pairs of length $k$, e.g.,
-$\tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$. The number $k$ of observations is another model
-parameters. As $k$ increases, learners have more data to base their inference on and so tend to
+$\tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$. As $k$ increases, learners have more data to base their inference on and so tend to
 recover the true types that generated a given sequence with higher probability. The mutation
@@
 given observation $d$, the probability of acquiring type $i$ is $F(t_i \mid d)$, so that:
-\mf{check if we need ``proportional to'' here or whether ``equals to'' is okay} \tb{``proportional to'' is needed. ``equal'' would work if for all $d$; $\sum_j P(d \mid t_j) = 1$}
 \begin{flalign*}
@@
 This equation defines the frequency $x'_i$ of type $i$ at the next time step, based on its
-frequency $x_i$ before the update, its fitness $f_i$ and the probability that a learner infers
+frequency $x_i$ before the step, its fitness $f_i$ and the probability that a learner infers
 $i$ when observing the behavior of a type-$j$ agent. Fitness-based selection can be thought of
@@
 For instance, {\em some} is entailed by {\em all}.
-If it were true that `All students came to class', it would also be true that `Some students came to class'.
+If it were true that `Chris owns all of Johnny Cash's albums', it would also be true that `Chris owns some of Johnny Cash's albums'.
 However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what their use is normally taken to convey.
@@
 That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that stronger alternatives do not hold because the speaker used a weaker alternative instead.
-In this way, `Some students came to class' is strengthened to convey that some but not all students came to class.
+In this way, `Chris owns some of Johnny Cash's albums' is strengthened to convey that he owns some but not all albums.
 A bound that rules out stronger alternatives is thusly not codified in the lexical meaning of weak alternatives but instead pragmatically supplied.
 
-This kind of strengthening is captured by the linguistic behavior of pragmatic types introduced in \S\ref{sec:languages+use}: A pragmatic hearer who reasons about the use of a message involving a weak scalar alternative will associate it more with a state in which stronger alternatives do not hold.
+As noted earlier, this kind of strengthening is captured by the linguistic behavior of pragmatic types introduced in \S\ref{sec:languages+use}: A pragmatic hearer who reasons about the use of a message involving a weak scalar alternative will associate it more with a state in which stronger alternatives do not hold.
 This is so because a rational speaker would use a more informative message when in such a state.
@@
 This can happen when embedded in downward-entailing contexts, when the speaker is likely uncertain about whether the upper bounded reading is true, or when the distinction between an upper-bounded reading and the simple, only lower-bounded reading, is not relevant.
-For instance, if for all the speaker knows `Some students came' but she does not know whether all came, then the use of {\em some} lacking an upper-bound succinctly conveys her uncertainty.
+For instance, if for all the speaker knows `Chris owns some of Johnny Cash's albums' but she does not know whether he owns all, then the use of {\em some} lacking an upper-bound succinctly conveys her uncertainty.
 This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: Should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then this is preferred over expressing it overtly through longer expressions, e.g., by saying {\em some but not all} explicitly.
@@
 That is, in a preference of learners for simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
-
 While we do not want to argue that functional aspects as the ones discussed above do not play a role, we do see a clear benefit in exploring whether matters of transmission biases would not give us additional explanatory leverage.
-Note however that we do not represent the contrast between lexical representations explicitly.
-Instead, the bias is directly encoded in the learners' prior over types.\footnote{In principle this difference could be made precise with an adequate representational language, e.g., through measures over representational complexity such as minimal description length.
-There is a growing effort to develop such empirically  testable  representational  languages.
-For instance, the so-called {\em language of thought} has been put to test in various rational probabilistic models that show encouraging results (see e.g.
-\citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012a} and references therein).
-At present we decide against such an enrichment in favor of a stronger focus on the general predictions of the model and the interaction of its components.}
 
@@
 Higher level reasoning is not required to derive scalar implicatures from the lexica we consider here, nor do they leave room for substantial pragmatic refinement.
+
+\paragraph{Lexica and transmission bias.} We consider a state space with three states $\States = \set{\snone, \ssome, \sall}$. This space can be
+thought of as a partition of possible worlds into cells where none, some or all of the $A$s are
+$B$s, for some arbitrary fixed predicates $A$ and $B$. Eight concepts can be distinguished
+based on their truth or falsity in three world states, six of which are not contradictory
+(always false) or tautologous (always true). These are listed with mnemonic names in
+Table~\ref{tab:concepts}.
+
+\begin{table}
+  \centering
+\begin{center}
+  \begin{tabular}{lccclc}
+    \toprule
+    intuitive name
+    & \snone
+    & \ssome
+    & \sall
+    & least complex formula
+    & complexity
+    \\ \midrule
+    ``all''
+    & 0
+    & 0
+    & 1
+    & $A \subseteq B$
+    & $3$
+    \\
+    ``some but not all''
+    & 0
+    & 1
+    & 0
+    & $A \cap B \neq \emptyset \wedge A \neq \emptyset$
+    & $8$
+    \\    
+    ``some''
+    & 0
+    & 1
+    & 1
+    & $A \cap B \neq \emptyset$
+    & $4$
+    \\
+    ``none''
+    & 1
+    & 0
+    & 0
+    & $A \cap B = \emptyset$
+    & $4$
+    \\
+    ``none or all''
+    & 1
+    & 0
+    & 1
+    & $\neg(A \cap B \neq \emptyset \wedge A \neq \emptyset)$
+    & \tb{$10$}
+    \\
+    ``not all''
+    & 1
+    & 1
+    & 0
+    & $\neg (A \subseteq B)$
+    & $5$
+    \\
+    \bottomrule
+  \end{tabular}
+\end{center}
+\caption{Available concepts and their minimal derivation length}
+\label{tab:concepts}
+\end{table}
+
+A lexicon $L$ is a mapping $\Messgs \rightarrow C$ from messages to concepts. With three
+messages there are $6^3 = 216$ possible lexica. \tb{Many of these assign the same concept to
+more than one message. For simplicity, we restrict attention to only those lexica which avoid
+expressive redundancy. Hardcoding such an \emph{mutual exclusivity bias}
+\citep[e.g.][]{Clark2009:Lexical-Meaning}, there are 20 non-redundant lexica. Functional
+pressure towards efficient communication will evidently favor non-redundant lexica, so this
+restriction is fairly innocuous but practical.}
+
+As before, the type of a player is a combination of a lexicon and a manner of pragmatic
+language use: literal or pragmatic. With this, there are $432$ types in this model.
+
+\begin{table}
+  \centering
+  \begin{align*}
+    C & \rightarrow_2 C \wedge C 
+    & 
+    X & \rightarrow_1 \set{A,B} \\
+    C & \rightarrow_2 \neg C 
+    & 
+    X & \rightarrow_1 X \cap X \\
+    C & \rightarrow_1 X \subseteq X
+    & 
+    X & \rightarrow_1 X \cup X \\
+    C & \rightarrow_1 X \neq \emptyset \\
+    C & \rightarrow_1 X = \emptyset     
+  \end{align*}
+  \caption{Toy grammar in a set-theoretic "language of thought"}
+  \label{tab:grammar}
+\end{table}
+
+\tb{ Note however that we do not represent the contrast between lexical representations explicitly.
+Instead, the bias is directly encoded in the learners' prior over types.\footnote{In principle this difference could be made precise with an adequate representational language, e.g., through measures over representational complexity such as minimal description length.
+There is a growing effort to develop such empirically  testable  representational  languages.
+For instance, the so-called {\em language of thought} has been put to test in various rational probabilistic models that show encouraging results (see e.g.
+\citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012a} and references therein).
+At present we decide against such an enrichment in favor of a stronger focus on the general predictions of the model and the interaction of its components.}}
+
+Following our assumption of an inductive bias that favors simpler lexical representations, the prior biases learners against lexica in which a message holds true only of the first row, i.e., against messages that lexicalize an upper-bound that rules out state \sall.
+All other semantics are assumed to be a priori equally probable.
+This is captured by $P(t_i) \propto n - c \cdot r$, where $n$ is the total number of states and $r$ is the number of messages only true of \ssome in $t_i$'s lexicon, $c \in [0,1]$.
+Increments in the value of $c$ thereby bring about a stronger bias against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.
+
+\tb{This entire paragraphs needs to add: mention to LOT, other possible priors in LOT and why we use ours} The prior probability of a type is just the prior probability of its lexicon. The prior of a
+lexicon is a function of the complexity of the concepts in its image set. Lexica that use
+simpler concepts are \emph{a priori} more likely. This can be motivated by assuming that
+learners beam search for suitable concepts to map onto overt signals by (probabilistically)
+considering simpler concepts first. Many ways for defining complexity of a concept are
+conceivable. If strong empirical claims were at stake here, empirically motivated measures of
+complexity should be used. For the sake of a non-trivial example, we follow
+\citet{piantadosi+etal:underreview} and related work to define complexity of a concept as a
+function of its derivation cost in a (weighed or probabilistic) generative ``language of
+thought''. For concreteness of example, consider the toy grammar of concepts in
+Table~\ref{tab:grammar}. This grammar uses basic set-theoretic operations to form expressions
+which can be evaluated as true or false in our three world states. Applications of generative
+rules have a cost attached to them. (Alternatively, a probability.) Here we simply assume that
+Boolean combinations of concepts are more complex than ``atomic'' concepts and that otherwise
+each rule application adds the same cost unit. Table~\ref{tab:concepts} lists, for each
+concept, the least complex formula derivable in this grammar that has the appropriate truth
+conditions. A simple way of defining priors over a lexicon is:
+\begin{align*}
+  P(L)  & = \prod_{c \in Im(L)} P(c)  \ \ \ \text{, with} & 
+  P(c) & \propto \max_{c'}Compl(c') - Compl(c') + 1\,,
+\end{align*}
+where $Compl(c)$ is the complexity of each concepts.
 
@@
 
-Combining a linguistic behavior with each of these $6$ lexica yields a total of $12$ distinct types.
+\hl{Instead of this: Give an example of one of our target lexica, one of our competing (functional) lexica and one that is functionally deficient} Combining a linguistic behavior with each of these $6$ lexica yields a total of $12$ distinct types.
 Note in particular  that a type that has conventionalized upper-bounds to realize a (quasi-)partition of the relevant semantic space, such as $L_4$, will produce speaker behavior that is {\em almost} indistinguishable from that of a language that lacks upper-bounds, but with pragmatic users, such as $L_5$.
@@
 Given that this is intractable for large $k$, matrices were approximated by sampling $10$ sequences from each type's production probabilities and a type's children being exposed only to this subset.
-
-%For convenience, the model's parameters are summarized in Table \ref{tab:summary}.
-%
-%\begin{table}
-%\centering
-%\begin{tabular}{l|l|l}
-%    \multicolumn{1}{c}{parameter} & \multicolumn{1}{c}{explanation} & \multicolumn{1}{c}{locus}\\ \hline
-%    $\lambda \geq 1$ & rationality parameter & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\
-%    $\alpha \in [0,1)$ & semantics-pragmatics tension & $S_{n+1}(m|s;L) \propto \text{exp}(\lambda \; H_{n}(s|m;L)^\alpha)$\\ 
-%    $|D|$ & number of data produced per parent type & $P(d|t_j)P(t_i|d)$\\
-%    $k = |d|$ & number of observations per datum& $P(d|t_j)P(t_i|d)$\\
-%    $l \geq 1$ & posterior parameter from sampling to MAP & $P(t_i|d) \propto [P(t_i)P(d|t_i)]^l$\\
-%    $c \in [0,1]$ & learning bias for lack of upper-bounds &  $P(t_i)$
-%\end{tabular}
-%\caption{Summary of model parameters.} 
-%\label{tab:summary}
-%\end{table}
-\subsection{Transmission bias}\label{subsec:bias}
-Following our assumption of an inductive bias that favors simpler lexical representations, the prior biases learners against lexica in which a message holds true only of the first row, i.e., against messages that lexicalize an upper-bound that rules out state \sall.
-All other semantics are assumed to be a priori equally probable.
-This is captured by $P(t_i) \propto n - c \cdot r$, where $n$ is the total number of states and $r$ is the number of messages only true of \ssome in $t_i$'s lexicon, $c \in [0,1]$.
-Increments in the value of $c$ thereby bring about a stronger bias against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.
 
@@
 
-\subsection{Noisy transmission}\label{subsec:noise}
-Above, the assumption of a learning bias was pivotal in introducing transmission perturbations that differentiate functionally similar types.
-However, these perturbations need not necessarily arise from inductive biases.
-As shown in the following, outcomes that are indistinguishable from the ones predicted above can come about through environmental factors such as perceptual errors during production and comprehension as well.
-With this we do not want to suggest noisy perception to underlie the selection of a lack of semantic upper-bounds in scalar items.
-Instead, we want to stress the role that transmission perturbations play in this process, and the cultural evolution of meaning more broadly, as well as to highlight the explanatory potential  of environmental factors, which have received less attention than their inductive counterparts (but see \citealt{oconnor:2015, franke+correia:toappear}).
-
-The core components of the model remain as before.
-All that is required is their adjustment to the possibility of confusing one state with another by language users and learners.
-Letting $\epsilon$ stand for the probability of confusing \ssome with \sall, and vice-versa for $\delta$, we denote the probability that the teacher (learner) observes state $s_t$ ($s_l$) when the actual state is $s_a$ as $P_N(s_t \mid s_a)$ ($P_N(s_l \mid s_a)$).
-The probability that $s_a$ is the actual state when the learner observes $s_l$ is therefore:
-
-\begin{align*}
-  P_N(s_a \mid s_l) \propto P(s_a) \ P_N(s_l \mid s_a)\,.
-\end{align*}
-
-Accordingly, the probability that the teacher observes $s_t$ when the learner observes $s_l$ is:
-\begin{align*}
-  P_N(s_t \mid s_l) = \sum_{s_a} P(s_a \mid s_l) \ P_N(s_t \mid s_a)\,.
-\end{align*}
-Finally, this gives us the probability that a teacher of type $t$ produces a datum that is
-perceived by the listener as $d = \tuple{s_l, m}$:
-\begin{align*}
-  P_N(\tuple{s_l, m} \mid t) = \sum_{s_t} P_N(s_t \mid s_l) \ P(m \mid s_t; t)\,.
-\end{align*}
-Generalize this to a sequence of perceived data $d_l$ and write $P_N(d_l \mid t)$. Then, the noise-perturbed mutation matrix is defined as:
-\begin{align*}
-  Q_{ij}  \propto \sum_{d_l \in D} P(d_l \mid t_i) F(t_j,d_l) \,, \ \  \text{where $F(t_j,d)$
-    is as before.}
-\end{align*}
-In words, it may be the case that learner and/or teacher do not perceive the actual state as what it is.
-They are not aware of this, and produce/learn as if what they observed was the actual state.
-In particular, the learner does not reason about noise when she tries to infer the speaker's type.
-She takes what she observes a state to be as the actual state that the teacher has seen as well and infers which type would have most likely generated the message to this state.
-This can lead to biases of inferring the ``wrong'' teacher type if the noise makes some types err in a way that resembles the noiseless behavior of other types.
-That is, such environmental factors can, in principle, induce transmission biases that look as if there was a cognitive bias in favor of a particular type, simply because that type better explains the noise.
-
-Apart from changing the mutation matrix in this way, we also need to adapt the calculation of expected utilities, taking into consideration that states are perceived noisily:
-\begin{align*}
-  U_S(t_i,t_j) = \sum_{s_a}  P(s_a) \sum_{s_t} P_N(s_t \mid s_a) \sum_m S_n(m|s_t;L) \sum_{s'} R_o(s'|m;L) \delta(s,s')\,.
-\end{align*}
-
-\paragraph{Results.} As shown in Table \ref{tab:num-noise}, a prevalence of pragmatic $L_5$ can also arise from noisy transmission without a learning bias for particular types $(c = 0)$.
-In contrast to the outcome predicted by its noiseless counterpart, favourable outcomes for this type concentrate in parameter configurations  with $\epsilon > \delta$, $\alpha \geq 5$, $1 < k < 20$ and $l > 3$.
-As stressed above, we are not concerned with the interpretation of these particular values for the case study at hand, but rather with the technical result that shows that the mere presence of systematic noise in transmission can introduce regularization that looks as if the agents have a learning bias.
-In other words, learning biases are clearly not the only transmission perturbations that shape cultural evolution alongside functional pressure.
-Environmental and perceptual noise can play a role too.
-
-\begin{table}
-\centering 
-\begin{tabular}{l | c c c c c| c c c c c c c c}
-\multicolumn{1}{c}{~} & \multicolumn{5}{c}{$l = 5$} & ~ & \multicolumn{5}{c}{$l = 15$}\\ \hline \hline
- ($\epsilon,\delta$)        &  (.1,.1)  & (.1,.3)  & (.3,.1)  & (.8,.1)  & (.1,.8) & ~ & (.1,.1) & (.1,.3) & (.3,.1) & (.8,.1) & (.1,.8)\\ \hline \hline
-lit. $L_1$ & .01 & .02 & .03 & .03 & .03& ~ & 0& .02&  .02 & .01 & .01\\ 
-lit. $L_2$ & .01 & .02 & .03 & .03 & .03& ~ & 0& .02&  .02 & .01 & .01\\
-lit. $L_3$ & .01 & .02 & .03 & .01 & .03& ~ & 0& .02 & .02 & .01 & .01\\
-lit. $L_4$ & .03 & .03 & .03 & .05 & .01& ~ & .04& .03 & .03 & .01 & .01\\
-lit. $L_5$ & .02 & .04 & .03 & .02 & .02& ~ & .01& .03 & .02 & .02 & .01\\
-lit. $L_6$ & .02 & .03 & .04 & .03 & .05& ~ & .01& .03 & .04 & .01 & .02\\ \hline
-prg. $L_1$ & .01 & .02 & .03 & .03 & .03& ~ & 0&  .02& .02 & .01 &  .01\\
-prg. $L_2$ & .01 & .02 & .03 & .03 & .03& ~ & 0&  .02& .02& .01&  .01\\
-prg. $L_3$ & .01 & .02 & .03 & .03 & .03& ~ & 0&  .02& .02 & .01 & .01 \\ 
-prg. $L_4$ & .45 & .11 & .11 & .02 & .02& ~ & .44& .11 & .1 & .22 & .02 \\
-prg. $L_5$ & .22 & .17 & .47 & .04 & .68& ~ & .24& .18 & .56 & .02 & .84\\
-prg. $L_6$ & .22 & .48 & .17 & .69 & .04& ~ & .24& .52 & .16 & .84 & .02
-\end{tabular}
-\caption{\hl{Currently: Mean proportions of types from $50$ runs per $100$ independently generated Q-matrices per parameter setting after $20$ generations across noisy values for $\epsilon$ and $\delta$ with $l = 5$ and $l = 10$ ($\alpha = 10, \lambda = 20, k = 5$).
-In principle, we could skip every second row because pragmatic $L5$ and $L6$ mirror each other.}}
-\label{tab:num-noise}
-\end{table}
-
 \section{General discussion}\label{sec:discussion}
@@
 
-\appendix
-\section{Illustration of the role of $\alpha$ in linguistic choice}\label{app:sig-example}
-To clarify the role of $\alpha$, notice that for two states and two messages the possible values for $H_{0}(s|m;L)$ are restricted to
-zero, $0.5$ and one. Consider the case of $H_0$ with $L_5$, which assigns the following
-probabilities to states after hearing messages:
-\begin{align*}
-  H_{0}(\messg \mid \state; L_5) & = \begin{blockarray}{lcc}
-    & \mygray{\ssome} & \mygray{\sall} \\
-    \begin{block}{l[cc]}
-      \mygray{\msome} & 0.5 & 0.5 \\
-      \mygray{\mall}  & 0 & 1 \\
-    \end{block}
-  \end{blockarray}
-\end{align*}
-A level-1 speaker's utilities are then:
-\begin{align*}
-  \begin{blockarray}{lcc}
-    & \mygray{\msome} & \mygray{\mall} \\
-    \begin{block}{l[cc]}
-      \mygray{\ssome} & 0.5^\alpha & 0^\alpha \\
-      \mygray{\sall}  & 0.5^\alpha & 1^\alpha \\
-    \end{block}
-  \end{blockarray}
-\end{align*}
-As the soft-max choice rule is sensitive to differences in utilities, no matter what $\lambda$
-is we will get $S_1(\msome \mid \ssome; L_5) = S_1(\mall \mid \sall; L_5)$ if $\alpha =
-1$. This means that we cannot distinguish pragmatically deviant choices (selecting a less
-informative message instead of a more informative one), from semantically deviant choices
-(selecting a false message instead of a true one). In contrast, if $\alpha > 1$ the former will be less likely than the latter -- and the other way around for $\alpha < 0$.
-
 \bibliographystyle{plainnat}

### 2017-03-09 paper/paper-v3.tex
small revision up to 3

@@
 \newcommand{\mnone}{\mymessg{none}}
+\newcommand{\Lall}{\mylang{all}}
 \newcommand{\Lbound}{\mylang{bound}}
@@
 
-\paragraph{Lexica and transmission bias.} We consider a state space with three states $\States = \set{\snone, \ssome, \sall}$. This space can be
+\paragraph{Lexica, types, and transmission bias.} We consider a state space with three states $\States = \set{\snone, \ssome, \sall}$. This space can be
 thought of as a partition of possible worlds into cells where none, some or all of the $A$s are
@@
 based on their truth or falsity in three world states, six of which are not contradictory
-(always false) or tautologous (always true). These are listed with mnemonic names in
+ or tautologous. These are listed with mnemonic names in
 Table~\ref{tab:concepts}.
@@
     & $\neg(A \cap B \neq \emptyset \wedge A \neq \emptyset)$
-    & \tb{$10$}
+    & $10$
     \\
@@
 A lexicon $L$ is a mapping $\Messgs \rightarrow C$ from messages to concepts. With three
-messages there are $6^3 = 216$ possible lexica. \tb{Many of these assign the same concept to
+messages there are $6^3 = 216$ possible lexica.
+Of these, three kinds of lexica are of particular relevance for our purpose.
+First, lexica that assign the same concept to more than message.
+Such lexica lack in expressivity under either signaling behavior but may be easier to acquire than others.
+Second, lexica that conventionalized upper-bounds to realize a (quasi-)partition of the relevant semantic space.
+As discussed in relation to $\Lbound$, users of these lexica need not resort to pragmatic reasoning to convey an upper-bound with weak scalar expressions.
+Instead, this information is already given by the semantics of their lexicon, giving them a functional advantage over others.
+Third and lastly, $\Llack$-style lexica that, paired with mutual reasoning, can be used to convey an upper-bound pragmatically.
+Overall, there are \hl{{\bf N}} lexica of the second kind and $12$ of the third.
+The following three lexica exemplify each kind, with $\Lall$ being a communicatively suboptimal lexicon that conventionalizes the meaning of ``all''  with each message (cf.
+Table \ref{tab:concepts}).
+\begin{align*}
+  \begin{blockarray}{lccc}
+    & & \underline{\Lall} & \\
+    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[ccc]}
+     \mygray{\snone}  & 0 & 0 & 0\\
+     \mygray{\ssome}  & 0 & 0 & 0\\
+    \mygray{\sall}   & 1 & 1 & 1\\
+    \end{block}
+  \end{blockarray} & &
+ \begin{blockarray}{lccc}
+    & & \underline{\Lbound} & \\
+    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[ccc]}
+       & 1 & 0 & 0\\
+       & 0 & 1 & 0\\
+       & 0 & 0 & 1\\
+    \end{block}
+  \end{blockarray} & &
+  \begin{blockarray}{lccc}
+    & & \underline{\Llack} & \\
+    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[ccc]}
+       & 1 & 0 & 0\\
+       & 0 & 1 & 0\\
+       & 0 & 1 & 1\\
+    \end{block}
+  \end{blockarray}
+\end{align*}
+
+Recall that types are a combination of a lexicon and a manner of language use: literal or pragmatic.
+Accordingly, there are $432$ types in this model.
+\hl{Here.
+Add footnote on redundancy and results holding for smaller spaces with $40$ types or the original $12$ from Br}
+
+In our setup, pragmatic $\Llack$-style speakers have a functional disadvantage over $\Lbound$-style ones as
+
+\tb{Many of these assign the same concept to
 more than one message. For simplicity, we restrict attention to only those lexica which avoid
@@
 restriction is fairly innocuous but practical.}
-
-As before, the type of a player is a combination of a lexicon and a manner of pragmatic
-language use: literal or pragmatic. With this, there are $432$ types in this model.
 
@@
 
-The space of possible lexica is given in Table \ref{tab:lexica}.
+The space of possible lexica is given in Table .
 These $(2,2)$-Boolean matrices are the simplest ones that allow us to make the contrast between the presence or absence of an upper-bound and the use of scalar implicatures precise.
@@
 The literal meaning of weak scalar expressions such as English {\em some} then corresponds to a message true of both rows in these fragments.
-While there are $16$ possible $(2,2)$-lexica, a number of them are identical both in terms of expressivity and on whether they codify upper-bounds lexically.
-The competition between such types is therefore determined by the initial configuration of a population.
-This fact can be obscured when averaging across outcomes.
-We therefore focus on this smaller representative subset.
-Simulations conducted with the full space confirm that the general results reported here do not hinge on this choice.
 
@@
 
-\begin{table}[t]
-\centering 
-\begin{tabular}{l c l}
-$L_1$ = $\begin{pmatrix} 0 & 0 \\ 1 & 1 \end{pmatrix}$ & 
-$L_2$ = $\begin{pmatrix} 1 & 1 \\ 0 & 0 \end{pmatrix}$ & 
-$L_3$ = $\begin{pmatrix} 1 & 1 \\ 1 & 1 \end{pmatrix}$\\[0.5cm]
-
-$L_4$ = $\begin{pmatrix} 1 & 0 \\ 0 & 1 \end{pmatrix}$ &
-$L_5$ = $\begin{pmatrix} 1 & 0 \\ 1 & 1 \end{pmatrix}$ &
-$L_6$ = $\begin{pmatrix} 1 & 1 \\ 0 & 1 \end{pmatrix}$
-\end{tabular}
-\caption{Space of possible lexica.}
-\label{tab:lexica}
-\end{table}
-
 \hl{Instead of this: Give an example of one of our target lexica, one of our competing (functional) lexica and one that is functionally deficient} Combining a linguistic behavior with each of these $6$ lexica yields a total of $12$ distinct types.
-Note in particular  that a type that has conventionalized upper-bounds to realize a (quasi-)partition of the relevant semantic space, such as $L_4$, will produce speaker behavior that is {\em almost} indistinguishable from that of a language that lacks upper-bounds, but with pragmatic users, such as $L_5$.
+, such as $L_4$, will produce speaker behavior that is {\em almost} indistinguishable from that of a language that lacks upper-bounds, but with pragmatic users, such as $L_5$.
 Almost, because there may be slight differences between the probability with which speakers would (erroneously) use a semantically false description and the probability with which speakers would (erroneously) use a pragmatically suboptimal description.

### 2017-03-10 paper/paper-v3.tex
revision up to analysis of R(M) effects

@@
 $l\ge1$ to modulate between posterior sampling and the MAP strategy. When $l = 1$ learners
-sample from the posterior. The learners propensity to maximize the posterior grows as $l$
+sample from the posterior. The learners' propensity to maximize the posterior grows as $l$
 increases.
@@
 \subsection{Analysis}
-We analyze the model's predictions in populations of types with one of the two signaling behaviors introduced earlier; literal or pragmatic.
-The former correspond to level $0$ reasoners and the latter to ones of level $1$.
-Higher level reasoning is not required to derive scalar implicatures from the lexica we consider here, nor do they leave room for substantial pragmatic refinement.
+The dynamics are initialized with an arbitrary distribution over types, constituting a population's first generation.
+We report the outcome of their development under different parameters and pressures after $50$ generations.
+This ensures that these outcomes correspond to developmental plateaus after which no noteworthy change is registered.
+As specified in \S\ref{sec:learnability}, the mutation matrix $Q$ can be obtained by considering all possible state-message sequences of length $k$.
+Given that this is intractable for large $k$, they were approximated by sampling $250$ sequences from each type's production probabilities.
 
-\paragraph{Lexica, types, and transmission bias.} We consider a state space with three states $\States = \set{\snone, \ssome, \sall}$. This space can be
+We begin setting the stage by defining the learning prior and the set of types under consideration.
+Drawing from our preceding discussion, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $\Llack$-like semantics.
+However, it is instructive to first inspect the effect of these pressures in isolation.
+Accordingly, we first showcase the effects that expressivity and learnability have on their own before turning to those resulting from their combination.
+
+\subsubsection{Inductive learning bias}
+There is a growing effort to develop empirically  testable  representational  languages that allow for the measure of semantic complexity.
+For instance, so-called {\em languages of thought} (LOT) have been put to test in various rational probabilistic models that show encouraging results (see, e.g., \citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012a} and \citealt{piantadosi+jacobs:2016} for recent discussion).
+At its core, a LOT defines a set of operations and composition rules from which lexical meaning can be derived.
+As a first approximation and for the sake of concreteness, we follow this approach to make the preference of learners for simpler semantic representations precise by defining the complexity of a concept as a function of its derivation cost in a weighed generative LOT.
+
+Our toy grammar of concepts is given in Table~\ref{tab:grammar}.
+This grammar uses basic set-theoretic operations to form expressions which can be evaluated as true or false in different world states (see below).
+Applications of generative  rules have a cost attached to them.
+(Alternatively, a probability.) Here we simply assume that Boolean combinations of concepts are more complex than ``atomic'' concepts and that otherwise each rule application adds the same cost unit.
+\begin{table}
+  \centering
+  \begin{align*}
+    C & \rightarrow_2 C \wedge C 
+    & 
+    X & \rightarrow_1 \set{A,B} \\
+    C & \rightarrow_2 \neg C 
+    & 
+    X & \rightarrow_1 X \cap X \\
+    C & \rightarrow_1 X \subseteq X
+    & 
+    X & \rightarrow_1 X \cup X \\
+    C & \rightarrow_1 X \neq \emptyset \\
+    C & \rightarrow_1 X = \emptyset     
+  \end{align*}
+  \caption{Toy grammar in a set-theoretic LOT with weighted rules.}
+  \label{tab:grammar}
+\end{table}
+
+The prior probability of a type is just the prior probability of its lexicon.
+The prior of a lexicon is a function of the complexity of the concepts in its image set.
+As motivated above, lexica that use simpler concepts are \emph{a priori} more likely.
+A simple way of defining such priors over lexica is:
+\begin{align*}
+  P(L)  & \propto \prod_{c \in Im(L)} P(c)   \ \text{, with} & 
+  P(c) & \propto \max_{c'}Compl(c') - Compl(c) + 1\,,
+\end{align*}
+
+where $Compl(c)$ is the complexity of each concept.
+There are other more sophisticated ways to define priors over lexica (see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a}) but the key assumption here, common to all of them, is that simple representational expressions should be favored and complex ones disfavored.
+Overall, we should stress that these details -- from generative grammar to complexity measure -- are stipulated and should therefore be regarded as an illustratory instantiation of our general assumptions instead of as strong empirical claims.
+%Additionally, this choice is intended to showcase how insights and representations from linguistics, cultural evolution, evolutionary game theory, and (probabilistic) language learning can fruitfully be combined.
+
+\subsubsection{Lexica, signaling behavior \& types}
+We consider a state space with three states $\States = \set{\snone, \ssome, \sall}$. This space can be
 thought of as a partition of possible worlds into cells where none, some or all of the $A$s are
@@
 based on their truth or falsity in three world states, six of which are not contradictory
- or tautologous. These are listed with mnemonic names in
-Table~\ref{tab:concepts}.
+ or tautological. These are listed with mnemonic names in
+Table~\ref{tab:concepts} together with their complexity according to the grammar given in \ref{tab:grammar} and $Compl(c)$.
 
@@
 messages there are $6^3 = 216$ possible lexica.
-Of these, three kinds of lexica are of particular relevance for our purpose.
+Of these, three kinds of lexica are of particular relevance to our purpose.
 First, lexica that assign the same concept to more than message.
-Such lexica lack in expressivity under either signaling behavior but may be easier to acquire than others.
-Second, lexica that conventionalized upper-bounds to realize a (quasi-)partition of the relevant semantic space.
-As discussed in relation to $\Lbound$, users of these lexica need not resort to pragmatic reasoning to convey an upper-bound with weak scalar expressions.
-Instead, this information is already given by the semantics of their lexicon, giving them a functional advantage over others.
+Such lexica lack in expressivity under either signaling behavior but may be easier to acquire and be favored by the learning bias.
+Second, lexica that conventionalize upper-bounds to realize a (quasi-)partition of the relevant semantic space.
+As discussed above in relation with $\Lbound$, users of these lexica need not resort to pragmatic reasoning to convey an upper-bound with weak scalar expressions.
+Instead, this information is already given by the semantics of their lexicon, giving them an advantage in terms of expressivity.
 Third and lastly, $\Llack$-style lexica that, paired with mutual reasoning, can be used to convey an upper-bound pragmatically.
-Overall, there are \hl{{\bf N}} lexica of the second kind and $12$ of the third.
-The following three lexica exemplify each kind, with $\Lall$ being a communicatively suboptimal lexicon that conventionalizes the meaning of ``all''  with each message (cf.
-Table \ref{tab:concepts}).
+Overall, there are \hl{{\bf N}} lexica of the second kind and $6$ of the third.
+The following three lexica exemplify each kind, with $\Lall$ being a lexicon that is not optimal for communication as it assigns all its messages to the same state $\sall$.
+This failure to be able to associate a state to single a form inevitably leads to a communicative disadvantage in its use.
 \begin{align*}
@@
     \end{block}
+% \\ & & P(\Lall)\rlap{ $\propto 9$} & \\
   \end{blockarray} & &
@@
     \end{block}
+%  \\& & P(\Lbound)\rlap{ $\propto 96$} & \\
   \end{blockarray} & &
@@
     \end{block}
+%   \\& & P(\Llack)\rlap{ $\propto 48$} & \\
   \end{blockarray}
@@
 
-Recall that types are a combination of a lexicon and a manner of language use: literal or pragmatic.
+As we consider the full space of lexica, some assign the same concept to more than one message, as $\Lall$, or lexicalize the same meanings but associate them with different messages, as the $6$ $\Llack$-style types.
+The main results reported in the following do not hinge on the inclusion of these lexica but such an unadulterated space does better illustrate the influence of the linguistic pressures at hand.
+
+Lastly, recall that types are a combination of a lexicon and a manner of language use.
+We analyze the model's predictions in populations of types with one of the two signaling behaviors introduced earlier; literal or pragmatic.
+The former correspond to level $0$ reasoners and the latter to ones of level $1$.
+Higher level reasoning is not required to derive scalar implicatures from the lexica we consider here, nor do they leave room for substantial pragmatic refinement.
 Accordingly, there are $432$ types in this model.
-\hl{Here.
-Add footnote on redundancy and results holding for smaller spaces with $40$ types or the original $12$ from Br}
 
-In our setup, pragmatic $\Llack$-style speakers have a functional disadvantage over $\Lbound$-style ones as
-
-\tb{Many of these assign the same concept to
-more than one message. For simplicity, we restrict attention to only those lexica which avoid
-expressive redundancy. Hardcoding such an \emph{mutual exclusivity bias}
-\citep[e.g.][]{Clark2009:Lexical-Meaning}, there are 20 non-redundant lexica. Functional
-pressure towards efficient communication will evidently favor non-redundant lexica, so this
-restriction is fairly innocuous but practical.}
-
-\begin{table}
-  \centering
-  \begin{align*}
-    C & \rightarrow_2 C \wedge C 
-    & 
-    X & \rightarrow_1 \set{A,B} \\
-    C & \rightarrow_2 \neg C 
-    & 
-    X & \rightarrow_1 X \cap X \\
-    C & \rightarrow_1 X \subseteq X
-    & 
-    X & \rightarrow_1 X \cup X \\
-    C & \rightarrow_1 X \neq \emptyset \\
-    C & \rightarrow_1 X = \emptyset     
-  \end{align*}
-  \caption{Toy grammar in a set-theoretic "language of thought"}
-  \label{tab:grammar}
-\end{table}
-
-\tb{ Note however that we do not represent the contrast between lexical representations explicitly.
-Instead, the bias is directly encoded in the learners' prior over types.\footnote{In principle this difference could be made precise with an adequate representational language, e.g., through measures over representational complexity such as minimal description length.
-There is a growing effort to develop such empirically  testable  representational  languages.
-For instance, the so-called {\em language of thought} has been put to test in various rational probabilistic models that show encouraging results (see e.g.
-\citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012a} and references therein).
-At present we decide against such an enrichment in favor of a stronger focus on the general predictions of the model and the interaction of its components.}}
-
-Following our assumption of an inductive bias that favors simpler lexical representations, the prior biases learners against lexica in which a message holds true only of the first row, i.e., against messages that lexicalize an upper-bound that rules out state \sall.
-All other semantics are assumed to be a priori equally probable.
-This is captured by $P(t_i) \propto n - c \cdot r$, where $n$ is the total number of states and $r$ is the number of messages only true of \ssome in $t_i$'s lexicon, $c \in [0,1]$.
-Increments in the value of $c$ thereby bring about a stronger bias against languages that lexicalize upper-bounds, i.e., $L_2, L_4$ and $L_6$.
-
-\tb{This entire paragraphs needs to add: mention to LOT, other possible priors in LOT and why we use ours} The prior probability of a type is just the prior probability of its lexicon. The prior of a
-lexicon is a function of the complexity of the concepts in its image set. Lexica that use
-simpler concepts are \emph{a priori} more likely. This can be motivated by assuming that
-learners beam search for suitable concepts to map onto overt signals by (probabilistically)
-considering simpler concepts first. Many ways for defining complexity of a concept are
-conceivable. If strong empirical claims were at stake here, empirically motivated measures of
-complexity should be used. For the sake of a non-trivial example, we follow
-\citet{piantadosi+etal:underreview} and related work to define complexity of a concept as a
-function of its derivation cost in a (weighed or probabilistic) generative ``language of
-thought''. For concreteness of example, consider the toy grammar of concepts in
-Table~\ref{tab:grammar}. This grammar uses basic set-theoretic operations to form expressions
-which can be evaluated as true or false in our three world states. Applications of generative
-rules have a cost attached to them. (Alternatively, a probability.) Here we simply assume that
-Boolean combinations of concepts are more complex than ``atomic'' concepts and that otherwise
-each rule application adds the same cost unit. Table~\ref{tab:concepts} lists, for each
-concept, the least complex formula derivable in this grammar that has the appropriate truth
-conditions. A simple way of defining priors over a lexicon is:
-\begin{align*}
-  P(L)  & = \prod_{c \in Im(L)} P(c)  \ \ \ \text{, with} & 
-  P(c) & \propto \max_{c'}Compl(c') - Compl(c') + 1\,,
-\end{align*}
-where $Compl(c)$ is the complexity of each concepts.
-
-The space of possible lexica is given in Table .
-These $(2,2)$-Boolean matrices are the simplest ones that allow us to make the contrast between the presence or absence of an upper-bound and the use of scalar implicatures precise.
-As illustrated in Section \ref{sec:languages+use}, one may think of the state corresponding to the first row of any such lexicon as a ``some but not all''-state, $\ssome$, and the second as an ``all''-state, $\sall$.
-The literal meaning of weak scalar expressions such as English {\em some} then corresponds to a message true of both rows in these fragments.
-
-Lexica $L_1$ to $L_3$ are not optimal for communication because they assign all their messages to the same state(s).
-This failure to be able to associate a state to single a form inevitably leads to a communicative disadvantage in their use.
-$L_4$ and $L_5$ are our target lexica.
-The former assings upper-bounded semantics to \msome (the first matrix's column) whereas the latter does not.
-Lastly, $L_6$ is similar to $L_5$ in that one message is true of the same state but differs from it in assigning upper-bounded semantics to \msome.
-
-\hl{Instead of this: Give an example of one of our target lexica, one of our competing (functional) lexica and one that is functionally deficient} Combining a linguistic behavior with each of these $6$ lexica yields a total of $12$ distinct types.
+\tb{Maybe note somewhere that pragmatic $\Lbound$ is functionally worse than literal/pragmatic $\Lbound$ but still very hard to distinguish when there's little data.
+But the latter has a learning advantage.
+Or do this while discussing each dynamic.
+See commented out below}
+%Combining a linguistic behavior with each of these $6$ lexica yields a total of $12$ distinct types.
 , such as $L_4$, will produce speaker behavior that is {\em almost} indistinguishable from that of a language that lacks upper-bounds, but with pragmatic users, such as $L_5$.
@@
 
-The dynamic is initialized with an arbitrary distribution over types, constituting the population's first generation.
-The results for each parameter setting were obtained from $1000$ independent runs, each consisting of $20$ generations.
-This corresponds to a developmental plateau after which no noteworthy change was registered.
-As specified in \S\ref{sec:learnability}, the mutation matrix $Q$ can be obtained by considering all possible state-message sequences of length $k$.
-Given that this is intractable for large $k$, matrices were approximated by sampling $10$ sequences from each type's production probabilities and a type's children being exposed only to this subset.
+\subsubsection{Expressivity only}
 
-Drawing from our preceding discussion, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $L_5$-like semantics.
-However, it is instructive to first inspect the effect of these pressures in isolation.
-For this purpose we focus our attention on three pragmatic types.\footnote{Pragmatic reasoning allows language users to refine their (possibly erroneous) choices.
-Therefore, it is advantageous even for types that codify more lexically.} Pragmatic $L_3$, a type that is lacking in expressivity but is a priori preferred for its lack of upper-bounds.
-Pragmatic $L_4$, a type that is functionally advantageous but biased against.
-And pragmatic $L_5$, combining virtues of the latter two.
-
-\paragraph{Expressivity only.} The replicator dynamic is sensitive to $\lambda$ and $\alpha$ as both have a bearing on a type's fitness.
+The replicator dynamic is sensitive to $\lambda$ and $\alpha$ as both have a bearing on a type's fitness.
 The influence of $\lambda$ is depicted in Figure \ref{fig:either-R-or-M}.A.

### 2017-03-13 paper/paper-v3.tex
plot-sketch

@@
  or tautological. These are listed with mnemonic names in
-Table~\ref{tab:concepts} together with their complexity according to the grammar given in \ref{tab:grammar} and $Compl(c)$.
+Table~\ref{tab:concepts} together with their complexity according to the grammar given in Table \ref{tab:grammar}.
 
@@
 Third and lastly, $\Llack$-style lexica that, paired with mutual reasoning, can be used to convey an upper-bound pragmatically.
-Overall, there are \hl{{\bf N}} lexica of the second kind and $6$ of the third.
+Overall, there are $6$ lexica of the second kind and $6$ of the third.
 The following three lexica exemplify each kind, with $\Lall$ being a lexicon that is not optimal for communication as it assigns all its messages to the same state $\sall$.
 This failure to be able to associate a state to single a form inevitably leads to a communicative disadvantage in its use.
+However, such a lexicon may be easy to acquire.
+
 \begin{align*}
@@
 
-As we consider the full space of lexica, some assign the same concept to more than one message, as $\Lall$, or lexicalize the same meanings but associate them with different messages, as the $6$ $\Llack$-style types.
+As we consider the full space of possible lexica, some assign the same concept to more than one message, as $\Lall$, or lexicalize the same meanings but associate them with different messages, as the $6$ $\Llack$-style types.
 The main results reported in the following do not hinge on the inclusion of these lexica but such an unadulterated space does better illustrate the influence of the linguistic pressures at hand.
@@
 
-\tb{Maybe note somewhere that pragmatic $\Lbound$ is functionally worse than literal/pragmatic $\Lbound$ but still very hard to distinguish when there's little data.
-But the latter has a learning advantage.
-Or do this while discussing each dynamic.
-See commented out below}
 %Combining a linguistic behavior with each of these $6$ lexica yields a total of $12$ distinct types.
@@
 
-The replicator dynamic is sensitive to $\lambda$ and $\alpha$ as both have a bearing on a type's fitness.
-The influence of $\lambda$ is depicted in Figure \ref{fig:either-R-or-M}.A.
-The less expressive $L_3$ speakers fare the worst and are influenced the least by change in $\lambda$.
-In contrast, low values of $\lambda$ result in a higher proportion of $L_4$ speakers relative to $L_5$.
-This is expected given the central role of rationality in producing more deterministic behavior in users of $L_5$-like languages.
-Consequently, as the rationality parameter increases the functional difference between $L_4$ and $L_5$ is leveled.
-Overall, the outcome from only a pressure towards expressivity approximates an even share of pragmatic $L_4, L_5$ and $L_6$ types.
-The latter follows the same trajectory as $L_5$ in Figure \ref{fig:either-R-or-M}.A.
-This illustrates an important issue in the evolution of meaning at the semantics-pragmatics interface: expressivity alone can not differentiate between (near-)functional equivalents to a degree that justifies the systematic prevalence of $L_5$-like semantics.
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig1-onlyr}
+\caption{Proportion of pragmatic $\Llack$-style types and of the incumbent type in $5$ independent populations after $50$ generations under only a pressure for expressivity.}
+\label{fig:only-R}
+\end{figure}
 
-%In particular, low $\alpha$ disadvantages types that rely on pragmatic reasoning to the gain of those that codify more semantically.
-The rationality parameter $\lambda$ has a similar effect for different reasons: Low rationality leads to a less pronounced preference for the choice(s) expected to succeed best in communication.
-That is, $\lambda$ regulates the strength by which users of $L_5$ associate non-upper-bounded $m_1$ exclusively with the ``some''-state $s_1$ over the ``all''-state $s_2$.
-Speakers of $L_4$ need not rely on $\lambda$ for this as the association of $m_1$ with $s_1$ is already part of their language's semantics.
+The replicator dynamic is sensitive to $\lambda$ as it influences signaling behavior and thereby has a bearing on a type's fitness as well.
+This influence is showcased in Figure \ref{fig:only-R}, which shows the proportion of pragmatic $\Llack$-style speakers and that of the type with the highest proportion, the incumbent, in $5$ independent populations.
+As is to be expected, less expressive types do not fare well under pressure for communicative success and those able to convey states unequivocally fare better.
+This difference increases with $\lambda$ as it promotes more deterministic linguistic behavior, widening the gap in expressivity between types.
+However, higher values of $\lambda$ alone do not lend a justification to the systematic prevalence of $\Llack$-like semantics paired with pragmatic reasoning.
+An inspection of $1000$ independent populations with $\lambda = 20$ shows a similar outcome to that hinted at by Figure \ref{fig:only-R}.
+Pragmatic $\Llack$ types only make up $11$ cases of incumbency, corresponding to a mean proportion of $0.003$ across populations.
+By contrast, $913$ incumbents are $\Lbound$-style users with close to an even share between literal ($454$) and pragmatic types ($459$), corresponding to a mean proportion of about $48\%$.
+Overall, incumbents of any type make up $50\%$ of the mean outcome, showcasing the strong dominance of $\Lbound$ semantics over other alternatives when only expressivity is at stake.
+
+\subsubsection{Learnability only}
 
@@
 \centering
-\includegraphics[scale=.5]{./fig1-only-R-or-M}
-\caption{Mean proportions of target types after $20$ generations in $1000$ populations with only a pressure for expressivity in A ($\alpha = 1$) and only for learnability in B ($\alpha =1, \lambda = 30, k = 5$, $l=1$).}
-\label{fig:either-R-or-M}
+\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig2-onlym-pr}
+\caption{Proportion of pragmatic $\Llack$-style types and of the incumbent type in $5$ independent populations after $50$ generations under only a pressure for expressivity.}
+\label{fig:only-R}
 \end{figure}
 
-\paragraph{Learnability only.} The effect of iterated learning with posterior sampling but without a pressure for expressivity is shown in Figure \ref{fig:either-R-or-M}.B.
+The effect of iterated learning with posterior sampling but without a pressure for expressivity is shown in Figure \ref{fig:either-R-or-M}.B.
 In line with our expectations, the share of $L_4$ speakers decreases as the bias against upper-bounds increases.

### 2017-03-13 code-LOT-extension/plots.py

@@ -8,54 +8,270 @@ import os
 import glob
 import sys
 
-def get_parameter_subfigs(type_list,list1,list2):
-    kind = 'rmd'
-    seq_length = 5
-    num_of_runs = 10
 
-    x_figs = len(list1)
-    y_figs = len(list2)
+def analysis(types,val1,val2,kind):
+    from lexica import get_lexica,get_lexica_bins
+    from rmd import get_type_bin
+    seq_length = 5
+    lex = get_lexica(3,3,False)
+    bins = get_lexica_bins(lex)
+
+    values_of_inc_in_bins = np.zeros(len(bins))
+    numb_of_inc = np.zeros(len(bins))
+
+    df = pd.read_csv('./results/%s-s3-m3-lam%d-a1-k%d-samples250-l%d-g50-meFalse.csv' % (kind,val1,seq_length,val2))
+    all_types = ['t_final'+str(z) for z in xrange(432)]
+    df = df[all_types]
+    for i in xrange(len(df)):
+        type_of_inc = df.iloc[i].idxmax(axis=1)
+        value = df.iloc[i][type_of_inc]
+        stripped_type = int(type_of_inc[7:]) #convert name into idx
+        bin_of_type = get_type_bin(stripped_type,bins)
+        values_of_inc_in_bins[bin_of_type] += value #add value to store in vector
+        numb_of_inc[bin_of_type] += 1 #add 1 to counter of times incumbent
+
+    return [values_of_inc_in_bins / len(df), numb_of_inc]
+
+
+def get_subfigs_incumbents(type_list,list1,list2,kind,prior=False):
+    seq_length = 5
+    num_of_runs = 5
+    
+    if not(kind == 'm'):
+        x_figs = len(list1)
+        y_figs = len(list2)
+    else:
+        x_figs = len(list2)
+        y_figs = len(list1)
     fig,axes = plt.subplots(nrows=y_figs,ncols=x_figs)
-    types_to_plot = ['t_final'+str(x) for x in type_list]
-#    factor = 1
+
+    if prior == True: #To add prior subfigure
+        fig,axes = plt.subplots(nrows=y_figs,ncols=x_figs+1)
+
+    if y_figs == 1:
+        axes = axes[np.newaxis] #add axis if 1d vector. Otherwise error
+
+    types_to_plot = ['t_final'+str(x) for x in type_list] + ['incumbent']
+
[…]

### 2017-03-14 paper/paper-v3.tex
slight revision + preliminary (ugly) plots

@@
 messages there are $6^3 = 216$ possible lexica.
-Of these, three kinds of lexica are of particular relevance to our purpose.
+Of these, three kinds of lexica are of particular relevance.
 First, lexica that assign the same concept to more than message.
-Such lexica lack in expressivity under either signaling behavior but may be easier to acquire and be favored by the learning bias.
+Such lexica lack in expressivity under either signaling behavior but may be favored by the learning bias.
 Second, lexica that conventionalize upper-bounds to realize a (quasi-)partition of the relevant semantic space.
@@
 Instead, this information is already given by the semantics of their lexicon, giving them an advantage in terms of expressivity.
-Third and lastly, $\Llack$-style lexica that, paired with mutual reasoning, can be used to convey an upper-bound pragmatically.
+Such lexica are expected to be the main contenders to the natural language like semantics that fail to lexicalize an upper-bound.
+Third and lastly, $\Llack$-style lexica that, paired with mutual reasoning, can convey an upper-bound pragmatically.
 Overall, there are $6$ lexica of the second kind and $6$ of the third.
@@
 This failure to be able to associate a state to single a form inevitably leads to a communicative disadvantage in its use.
-However, such a lexicon may be easy to acquire.
+This lexicon serves illustrative purposes in comparison to our target, $\Llack$, and its competitor $\Lbound$.
 
@@
 \includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig1-onlyr}
-\caption{Proportion of pragmatic $\Llack$-style types and of the incumbent type in $5$ independent populations after $50$ generations under only a pressure for expressivity.}
+\caption{Proportion of pragmatic $\Llack$-style types and incumbent types in $5$ independent populations after $50$ generations under only a pressure for expressivity.}
 \label{fig:only-R}
@@
 
-The replicator dynamic is sensitive to $\lambda$ as it influences signaling behavior and thereby has a bearing on a type's fitness as well.
-This influence is showcased in Figure \ref{fig:only-R}, which shows the proportion of pragmatic $\Llack$-style speakers and that of the type with the highest proportion, the incumbent, in $5$ independent populations.
-As is to be expected, less expressive types do not fare well under pressure for communicative success and those able to convey states unequivocally fare better.
-This difference increases with $\lambda$ as it promotes more deterministic linguistic behavior, widening the gap in expressivity between types.
-However, higher values of $\lambda$ alone do not lend a justification to the systematic prevalence of $\Llack$-like semantics paired with pragmatic reasoning.
-An inspection of $1000$ independent populations with $\lambda = 20$ shows a similar outcome to that hinted at by Figure \ref{fig:only-R}.
-Pragmatic $\Llack$ types only make up $11$ cases of incumbency, corresponding to a mean proportion of $0.003$ across populations.
-By contrast, $913$ incumbents are $\Lbound$-style users with close to an even share between literal ($454$) and pragmatic types ($459$), corresponding to a mean proportion of about $48\%$.
-Overall, incumbents of any type make up $50\%$ of the mean outcome, showcasing the strong dominance of $\Lbound$ semantics over other alternatives when only expressivity is at stake.
+The replicator dynamic is sensitive to $\lambda$ as it influences signaling behavior and thereby has a bearing on a type's fitness.
+This influence is showcased in Figure \ref{fig:only-R}, which depicts the proportion of pragmatic $\Llack$-style speakers in $5$ independent populations after $50$ generations, as well as that of the type with the highest proportion, the incumbent.
+As is to be expected, less expressive types do not fare well under pressure for communicative success and those able to convey states less equivocally fare better.
+This difference increases with $\lambda$ as this parameter promotes more deterministic linguistic behavior, widening the gap in expressivity between types.
+Importantly, high values of $\lambda$ alone do not lend a justification to the systematic prevalence of $\Llack$-like semantics paired with pragmatic reasoning that we set out to explain.
+An inspection of $1000$ independent populations with $\lambda = 20$ shows a similar outcome to that hinted at by Figure \ref{fig:only-R}: Pragmatic $\Llack$ types only make up $11$ cases of incumbency, corresponding to a mean proportion of $0.003$ across populations.
+By contrast, $913$ incumbents were $\Lbound$-style users with close to an even share between literal ($454$) and pragmatic types ($459$), corresponding to a mean proportion of about $48\%$ together.
+Overall, incumbents of any type make up $50\%$ of the mean outcome.
+This, again, showcases the strong dominance of $\Lbound$ semantics over other alternatives when only expressivity is at stake.
 
@@
 \includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig2-onlym-pr}
-\caption{Proportion of pragmatic $\Llack$-style types and of the incumbent type in $5$ independent populations after $50$ generations under only a pressure for expressivity.}
-\label{fig:only-R}
+\caption{Proportion of pragmatic $\Llack$-style types and incumbent types in $5$ independent populations after $50$ generations under only a pressure for learnability ($\lambda = 20, k = 5$).
+The types' prior probability is shown in the right-most plot.}
+\label{fig:only-M}
 \end{figure}
 
-The effect of iterated learning with posterior sampling but without a pressure for expressivity is shown in Figure \ref{fig:either-R-or-M}.B.
-In line with our expectations, the share of $L_4$ speakers decreases as the bias against upper-bounds increases.
-In turn, this benefits $L_3$ and, in particular, $L_5$.
-However, even a strong bias against lexical upper-bounds leads only to a moderate advantage of $L_5$ over $L_4$.
-More importantly, a pressure only towards learnability can promote functionally defective languages such as $L_3$.
+The effect of iterated learning without a pressure for expressivity using posterior sampling ($l = 1$) and a stronger tendency towards posterior maximization ($l = 15$) is shown in Figure \ref{fig:only-M} together with the prior over types.
+The latter illustrates that users of $\Llack$ are, while not the most favored types, advantaged by the inductive bias in virtue of the relatively simple semantics they conventionalize.
+Accordingly, a stronger propensity to maximize the posterior increases their proportion in the population in analogous fashion to increase in $\lambda$ under the replicator dynamic.
+However, while replication generally succeeds in driving out competing types of a kind, iterated learning leads to their coexistence as each of these types is passed on to next generations equal faithfulness.
+Although $\Llack$ is not the most favored type by the prior, it is the least semantically complex language in the type space that enables its users to convey each state with a single message when combined with pragmatic reasoning (provided sufficiently high $\lambda$).
+Put differently, it is easier to infer than functionally comparable lexica such as $\Lbound$ and it is less likely to be mistaken with others because its pragmatic use approximates a one-to-one form-meaning mapping.
+Consequently, as suggested by Figure \ref{fig:only-M}, in $1000$ analyzed independent populations all incumbents were pragmatic $\Llack$ users ($l = 15$ and $\lambda = 20$).
 
+\subsubsection{Expressivity and learnability.}
 Inspecting these pressures separately not only showcases the contribution of the model's components but also highlights some of their broader implications.
-First and foremost, neither dynamic on its own comes close to converging to a monomorphic population under most parameter configurations.
-For instance, while $L_4$ speakers can come to take over a substantial proportion of the population, this only happens in a restricted range of low degrees of rationality.
-Apart from polymorphy, both pressures make undesirable predictions in isolation.
-A pressure only towards expressivity leads to the expulsion of communicatively suboptimal $L_1$, $L_2$ and $L_3$ from the population but can not explain the regular selection of $L_5$-like semantics over either of its functionally similar alternatives.
-A pressure only towards learnability has a modest but clear effect in differentiating $L_5$ from these alternatives but fails to rule out functionally suboptimal types such as tautological $L_3$.
+First and foremost, neither dynamic on its own comes close to converging to a monomorphic population of a type of pragmatic $\Llack$ speakers.
+Under communicative pressure, the communicative advantage of $\Lbound$ leads to its selection when $\lambda$ is sufficiently high.
+Pressure only towards learnability has a modest but clear effect in differentiating $\Llack$ from such functionally similar but semantically more complex alternatives.
+However, this pressure on its own does not foment a propagation of these types across the population.
+Additionally, it leads to mixed populations that do not justify the selection of a particular type over others either.
 
-\begin{figure}
-\centering
-\includegraphics[scale=.5]{./fig2-rmd}
-\caption{Mean proportions of target types after $20$ generations in $1000$ populations across bias values $c \in [0,1]$ with $l =1$ in A and $l = 5$ in B ($\alpha =1, \lambda = 20, k = 5$).}
-\label{fig:cost}
-\end{figure}
-
-\paragraph{Expressivity and learnability.} Figure \ref{fig:cost} illustrates the effect of the learning bias for posterior sampling (\ref{fig:cost}.A) and slightly more MAP-like learning (\ref{fig:cost}.B) when pressured for both expressivity and learnability.
+ Figure \ref{fig:cost} illustrates the effect of the learning bias for posterior sampling (\ref{fig:cost}.A) and slightly more MAP-like learning (\ref{fig:cost}.B) when pressured for both expressivity and learnability.
 More detailed results for all types across a sample of $c$-values for $l = 1$ and $l = 5$ are presented in Table \ref{tab:numeric-results}.
@@
 
-\begin{table}
+\begin{figure}
 \centering
-\begin{tabular}{l | c c c c c| c c c c c c c c}
-\multicolumn{1}{c}{~} & \multicolumn{5}{c}{$l = 1$} & ~ & \multicolumn{5}{c}{$l = 5$}\\ \hline \hline
-  c        &  0  & .1  & .5  & .8  & .9 & ~ & 0 & .1 & .5 & .8 & .9\\ \hline \hline
-lit. $L_1$ & .03 & .03 & .04 & .04 & .04& ~ & 0 &0 & 0 & 0 & 0\\ 
-lit. $L_2$ & .03 & .03 & .02 & .01 & .04& ~ & 0& 0 & 0 & 0 & 0\\
-lit. $L_3$ & .03 & .03 & .04 & .04 & .04& ~ & 0& 0 & 0 & 0 & 0\\
-lit. $L_4$ & .07 & .07 & .06 & .06 & .05& ~ & 0& 0 & 0 & 0 & 0\\
-lit. $L_5$ & .04 & .05 & .05 & .06 & .06& ~ & 0& 0 & 0 & 0 & 0\\
-lit. $L_6$ & .04 & .04 & .04 & .04 & .03& ~ & 0& 0 & 0 & 0 & 0\\ \hline
-prg. $L_1$ & .03 & .03 & .04 & .04 & .04& ~ & 0& 0 & 0 & 0 & 0 \\
-prg. $L_2$ & .03 & .03 & .02 & .01 & .04& ~ & 0& 0 & 0 & 0 & 0 \\
-prg. $L_3$ & .03 & .03 & .04 & .04 & .04& ~ & 0& 0 & 0 & 0 & 0 \\ 
-prg. $L_4$ & .22 & .22 & .2 & .18 & .17& ~ & .33& .30 & .16 & .07 & .05 \\
-prg. $L_5$ & .22 & .23 & .27 & .3  & .32& ~ & .33& .39 & .68 & .86 & .90 \\
-prg. $L_6$ & .22 & .22 & .2 & .18 & .17& ~ & .33& .30 & .16 & .07 & .05
-\end{tabular}
-\caption{Mean proportions of types in $1000$ populations after $20$ generations across bias values $c \in [0,1]$ with $l =1$ and $l = 5$ ($\alpha = 1, \lambda = 30, k = 5$)}%, $\epsilon < 0.005$.}
-\label{tab:numeric-results}
-\end{table}
+\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig3-r+m}
+\caption{Proportion of pragmatic $\Llack$-style types and incumbent types in $5$ independent populations after $50$ generations under both pressures ($k = 5$).}
+\label{fig:only-M}
+\end{figure}
 
@@
 \centering
-\includegraphics[scale=.5]{../presentations/01heatmap}
-\caption{Mean proportion of pragmatic $L_5$ in $1000$ populations after $20$ generations ($\alpha = 1, \lambda = 30, k = 5$)}
+\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig4-incumbents-difference}
+%\caption{Mean proportion of pragmatic $L_5$ in $1000$ populations after $20$ generations ($\alpha = 1, \lambda = 30, k = 5$)}
 \label{fig:prior-posterior}

### 2017-03-14 code-LOT-extension/dynamics.py

@@ -2,16 +2,33 @@
 from rmd import run_dynamics
 
 ##### Parameters & setup #####
+#a = [1] # rate to control difference between semantic and pragmatic violations
+#lamb = [1,5,20,30] # soft-max parameter
+#seq_length = [1,5,10,15]  # length of observation sequences
+#samples = [250] #amount of k-length samples for each production type
+#l = [1,5,10,15] #prob-matching = 1, increments approach MAP
+#
+#g = [50] #number of generations per simulation run
+#r = [1000] #number of independent simulation runs
+#
+#dynamics = ['r','m','rmd'] #kind is the type of dynamics, either 'rmd', 'm' or 'r'
+#
+#states = 3 #number of states
+#messages = 3 #number of messages
+#me = [False] #mutual exclusivity
+
+
+# More focused run:
 a = [1] # rate to control difference between semantic and pragmatic violations
-lamb = [1,5,20,30] # soft-max parameter
-seq_length = [1,5,10,15]  # length of observation sequences
+lamb = [2,3,4,6,7,8,9,10,11,12,13,14,15,16,17,18,19] # soft-max parameter
+seq_length = [5]  # length of observation sequences
 samples = [250] #amount of k-length samples for each production type
-l = [1,5,10,15] #prob-matching = 1, increments approach MAP
+l = [2,3,4,6,7,8,9,11,12,13,14] #prob-matching = 1, increments approach MAP
 
 g = [50] #number of generations per simulation run
 r = [1000] #number of independent simulation runs
 
-dynamics = ['r','m','rmd'] #kind is the type of dynamics, either 'rmd', 'm' or 'r'
+dynamics = ['rmd'] #kind is the type of dynamics, either 'rmd', 'm' or 'r'
 
 states = 3 #number of states
 messages = 3 #number of messages

### 2017-03-14 code-LOT-extension/plots.py

@@ -113,23 +113,25 @@ def get_subfigs_incumbents(type_list,list1,list2,kind,prior=False):
         Y_rest.reverse()
         
         ax = fig.add_subplot(133)
+        ax.grid(False)
 
-        ax.barh(X,Y_target,color='grey')#,width=1, color='grey')#, orientation='horizontal')
+        ax.barh(X,Y_target,color='black')#, edgecolor='none')#,width=1, color='grey')#, orientation='horizontal')
         ax.barh(X,Y_rest,color='white')#,width=1, color='white' )#, orientation='horizontal')
 
-#        ylabels = [x for x in xrange(len(X))]
-#        ylabels[-5] = 'L-lack style'
-#        ylabels[45] = 'Other types'
-#        ax.set_yticklabels(ylabels)
+
+        ylabels = ["" for x in xrange(len(X))]
+        ylabels[9] = 'L-lack'
+        ylabels[5] = 'Other types'
+        ax.set_yticklabels(ylabels,rotation=90)
+
+#        for label in ax.get_yticklabels()[::3]:
+#            label.set_visible(False)
+
         for label in ax.get_xticklabels()[::2]:
             label.set_visible(False)
 
-        for label in ax.get_yticklabels()[::2]:
-            label.set_visible(False)
-
-
-
-
+        ax.tick_params(axis='both', which='major', labelsize=17)
+        
 #        xlabels = [str(int(x) + 1) for x in xlabels]
 #        ax.set_xticklabels(xlabels)
 
@@ -140,6 +142,7 @@ def get_subfigs_incumbents(type_list,list1,list2,kind,prior=False):
 
     for idx in xrange(len(plt.gcf().axes)):
         ax = plt.gcf().axes[idx]
+        ax.tick_params(axis='both', which='major', labelsize=17)
 
         #Only add top label to subplots that are on the first row
         if idx - x_figs < 0 and not(kind == 'm'):
@@ -148,10 +151,10 @@ def get_subfigs_incumbents(type_list,list1,list2,kind,prior=False):
             ax.xaxis.set_label_position('top')
 
         elif idx - x_figs < 0:
-            ax.set_xlabel('l = '+str(learning_labels[idx]),fontsize=20)
+            ax.set_xlabel('l = '+str(learning_labels[idx]),fontsize=25)
[…]

### 2017-03-15 paper/paper-v3.tex
minor additions new setup draft

@@
   to another.  We illustrate the model based on a case study showing that learning biases that favor simple semantic
-  representations can prevent the lexicalization of pragmatic inferences as long as mutual reasoning counteracts a potential loss in expressivity from such semantics. 
+  representations can prevent the lexicalization of pragmatic inferences as long as mutual reasoning counteracts a loss in expressivity otherwise incurred by the adoption of such semantics. 
 \end{abstract}
@@
 between the literal meaning of expressions and what they are intended and interpreted to convey
-is due to \citet{grice:1975}, characterizing pragmatic use and interpretation as a process of
+is due to \citet{grice:1975}, who characterizes pragmatic use and interpretation as a process of
 mutual reasoning about rational language use. For instance, under the assumption that the
@@
 \citep[e.g.,][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. While
-there is much dispute in this domain about many interesting details, a position endorsed by a
+there is much dispute in this domain about many details, a position endorsed by a
 clear majority is that a word like \emph{some} is underspecified to mean \emph{some and maybe
@@
 communicative efficiency and (ii) possible infidelity in the transmission of linguistic
-knowledge, caused by factors such as inductive learning biases and general aspects of learnability. Considering
+knowledge, caused by factors such as inductive learning biases and sparse data to learn from. Considering
 transmission of linguistic knowledge is important because neither semantic meanings nor
@@
 
-An elegant formal approach to capturing the interaction between expressivity and learnability
+An elegant formal approach to capture the interaction between expressivity and learnability
 is the \emph{replicator mutator dynamic}
@@
 meaning and pragmatic use, we need to fix what the relevant types are, how fitness is measured
-and how the mutation matrix is computed. These issues will be addressed, one by one, in the
+and how the mutation matrix is computed. These issues are addressed, one by one, in the
 following.
@@
 same sentence with \emph{some} replaced by \emph{all}.  
-\tb{I left $\snone$/$\mnone$ out because they are not relevant to the strengthening of $\msome$, but that means that it needs to be introduced a bit later.
-Feel free to add it if you think that's better, exposition-wise.}
-Lexica for this case would assign a
-truth value for each state-message pair. The following two lexica exemplify the distinction
+Lexica for this case would assign a Boolean truth value, either $0$ or $1$, to each state-message pair. The following two lexica exemplify the distinction
 between a lexicalized upper-bound for \emph{some} in $\Lbound$ and the widely assumed logical
@@
 states?} \tb{I think that's fine.
-$S_n(\cdot \mid \cdot)$ is not used beyond this section and the set $S$ only by mention of its members.
+$S_n(\cdot \mid \cdot)$ is not used beyond this section.
 Otherwise we can, e.g., relabel $M$ as $F$(orms) and use $M$(eanings) instead of $S$(tates).
-I'm fine with either alternative.}
+I'm fine with either option.}
 \begin{flalign}
@@
 in (\ref{s:level0}) also allows false messages to be sent with a small positive probability in
-analogy to the definition of pragmatic speakers in probabilistic pragmatic models. This is
-necessary to guarantee a mutation matrix with only positive entries (see below). If $\lambda = 1$ a literal
+analogy to the definition of pragmatic speakers in probabilistic pragmatic models.
+This is also necessary to guarantee a mutation matrix with only positive entries (see below).
+If $\lambda = 1$ a literal
 speaker with either lexicon $\Lbound$ or $\Llack$ produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
@@
 \sall;\Lbound) \approx .27$ for $\Lbound$ and $S_{0}(\msome \mid
-\sall;\Llack) \approx .5$ for $\Llack$. By contrast, if $\lambda = 20$, a literal speaker with either lexicon produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
+\sall;\Llack) = .5$ for $\Llack$. By contrast, if $\lambda = 20$, a literal speaker with either lexicon produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
 \ssome;L_{\text{bound},\text{lack}}) \approx 1$. But the probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
-\sall;\Lbound) \approx 1$ for $\Lbound$ whereas it is still $S_{0}(\msome \mid \sall;\Llack) \approx .5$ for $\Llack$.
+\sall;\Lbound) \approx 1$ for $\Lbound$ whereas it is still $S_{0}(\msome \mid \sall;\Llack) = .5$ because $\msome$ is semantically associated with both states by $\Llack$.
 
-Pragmatic behavior of level-$n+1$ is similar to its literal counterparts but uses the
+Pragmatic behavior of level $n+1$ is similar to its literal counterparts but uses the
 interpretation or production behavior of a level-$n$ player instead of the lexical meaning:
@@
 
-Particularly important to our purpose is the fact that players of a level greater than $0$ using $\Llack$ can {\em pragmatically} associate $\msome$ with $\ssome$ over $\sall$, in contrast to their literal counterparts.
+Particularly important to our purpose is the fact that players of a level greater than $0$ using $\Llack$ can come to {\em pragmatically} associate $\msome$ with $\ssome$ over $\sall$ in contrast to their literal counterparts.
 Intuitively, this is so because hearers reason a speaker to convey $\sall$ with $\mall$, as this has a higher chance of leading to mutual understanding than using underspecified $\msome$.
@@
 The converse reasoning pattern holds for the speaker.
+In this way, two distinct lexical conventions such as $\Lbound$ and $\Llack$ can give rise to similar observable linguistic behavior.
 
@@
 multiple languages may be compatible with the data learners are exposed to. These sources of
-uncertainty introduce variation in their transmission from one generation to the next. In particular, learning biases in the iterated transmission process can influence language evolution
+uncertainty introduce variation in the transmission of linguistic knowledge from one generation to the next.
+In particular, learning biases in the iterated transmission process can influence language evolution
 substantially.
@@
 the language of the teacher \citep[cf.][450]{griffiths+kalish:2007}. The extent of the prior's
-influence has been shown to heavily depend on the learning strategy assumed to underly the
+influence has been shown to heavily depend on the learning strategy assumed to underlie the
 inference process. On the one hand, early simulation results suggested that weak biases could
 be magnified by exposing learners to only small data samples \citep[e.g., in][]{brighton:2002}. On the other, \citeposs{griffiths+kalish:2007} mathematical
-characterization showed that iterated learning converged to the prior, i.e., the resulting
+characterization showed that iterated learning converged to the prior, i.e., that the resulting
 distribution over languages corresponds to the learners' prior distribution and is not
@@
 $i$ when observing the behavior of a type-$j$ agent. Fitness-based selection can be thought of
-as biological (fitness as expected relative number of offspring) or cultural (fitness of
+as biological (fitness as expected relative number of offspring) or cultural (fitness as 
 likelihood of being imitated or repeated). The types that the dynamic operates on are pairs
@@
 For instance, {\em some} is entailed by {\em all}.
-If it were true that `Chris owns all of Johnny Cash's albums', it would also be true that `Chris owns some of Johnny Cash's albums'.
+If it were true that ``Chris owns all of Johnny Cash's albums'', it would also be true that ``Chris owns some of Johnny Cash's albums''.
 However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what their use is normally taken to convey.
@@
 That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that stronger alternatives do not hold because the speaker used a weaker alternative instead.
-In this way, `Chris owns some of Johnny Cash's albums' is strengthened to convey that he owns some but not all albums.
+In this way, ``Chris owns some of Johnny Cash's albums'' is strengthened to convey that he owns some but not all albums.
 A bound that rules out stronger alternatives is thusly not codified in the lexical meaning of weak alternatives but instead pragmatically supplied.
@@
 The dynamics are initialized with an arbitrary distribution over types, constituting a population's first generation.
-We report the outcome of their development under different parameters and pressures after $50$ generations.
+In the following we report the outcome of their development under different parameters and pressures after $50$ generations.
 This ensures that these outcomes correspond to developmental plateaus after which no noteworthy change is registered.
 As specified in \S\ref{sec:learnability}, the mutation matrix $Q$ can be obtained by considering all possible state-message sequences of length $k$.
-Given that this is intractable for large $k$, they were approximated by sampling $250$ sequences from each type's production probabilities.
+Given that this is intractable for large $k$, the sets of data learners are exposed to are approximated by sampling $250$ $k$-length sequences from each type's production probabilities.
 
@@
 However, it is instructive to first inspect the effect of these pressures in isolation.
-Accordingly, we first showcase the effects that expressivity and learnability have on their own before turning to those resulting from their combination.
+We therefore first showcase the effects that expressivity and learnability have on their own before turning to those resulting from their combination.
 
-\subsubsection{Inductive learning bias}
+\subsubsection{Inductive learning bias against semantic complexity}
 There is a growing effort to develop empirically  testable  representational  languages that allow for the measure of semantic complexity.
@@
 At its core, a LOT defines a set of operations and composition rules from which lexical meaning can be derived.
-As a first approximation and for the sake of concreteness, we follow this approach to make the preference of learners for simpler semantic representations precise by defining the complexity of a concept as a function of its derivation cost in a weighed generative LOT.
+As a first approximation and for the sake of concreteness, we follow this approach to make the preference of learners for simpler semantic representations precise and define the complexity of a representation as a function of its derivation cost in a weighed generative LOT.
 
 Our toy grammar of concepts is given in Table~\ref{tab:grammar}.
-This grammar uses basic set-theoretic operations to form expressions which can be evaluated as true or false in different world states (see below).
+This grammar uses basic set-theoretic operations to form expressions which can be evaluated as true or false in different world states such as $\ssome$ or $\sall$ (see below).
 Applications of generative  rules have a cost attached to them.
-(Alternatively, a probability.) Here we simply assume that Boolean combinations of concepts are more complex than ``atomic'' concepts and that otherwise each rule application adds the same cost unit.
+(Alternatively, a probability.) Here we simply assume that Boolean combinations of concepts are more complex than atomic concepts and that otherwise each rule application adds the same cost unit.
 \begin{table}
@@
 The prior probability of a type is just the prior probability of its lexicon.
-The prior of a lexicon is a function of the complexity of the concepts in its image set.
+The prior of a lexicon is a function of the complexity of the lexical representations in its image set.
 As motivated above, lexica that use simpler concepts are \emph{a priori} more likely.
@@
 
-where $Compl(c)$ is the complexity of each concept.
+where $Compl(c)$ is the complexity of the minimal derivation length of each concept.
 There are other more sophisticated ways to define priors over lexica (see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a}) but the key assumption here, common to all of them, is that simple representational expressions should be favored and complex ones disfavored.
-Overall, we should stress that these details -- from generative grammar to complexity measure -- are stipulated and should therefore be regarded as an illustratory instantiation of our general assumptions instead of as strong empirical claims.
-%Additionally, this choice is intended to showcase how insights and representations from linguistics, cultural evolution, evolutionary game theory, and (probabilistic) language learning can fruitfully be combined.
+More generally, we should stress that these details -- from generative grammar to complexity measure -- are stipulated and should therefore be regarded as an illustratory instantiation of our general assumptions instead of as strong empirical claims.
+%Instead, this choice showcases how insights and representations from linguistics, cultural evolution, evolutionary game theory, and (probabilistic) language learning may fruitfully be combined.
 
@@
 Of these, three kinds of lexica are of particular relevance.
-First, lexica that assign the same concept to more than message.
-Such lexica lack in expressivity under either signaling behavior but may be favored by the learning bias.
+First, lexica that assign the same concept to more than one message.
+Such lexica lack in expressivity but may be favored by the learning bias.
 Second, lexica that conventionalize upper-bounds to realize a (quasi-)partition of the relevant semantic space.
-As discussed above in relation with $\Lbound$, users of these lexica need not resort to pragmatic reasoning to convey an upper-bound with weak scalar expressions.
+As discussed before in relation to $\Lbound$, users of these lexica need not resort to pragmatic reasoning to convey an upper-bound with weak scalar expressions.
 Instead, this information is already given by the semantics of their lexicon, giving them an advantage in terms of expressivity.
@@
 Overall, there are $6$ lexica of the second kind and $6$ of the third.
-The following three lexica exemplify each kind, with $\Lall$ being a lexicon that is not optimal for communication as it assigns all its messages to the same state $\sall$.
-This failure to be able to associate a state to single a form inevitably leads to a communicative disadvantage in its use.
-This lexicon serves illustrative purposes in comparison to our target, $\Llack$, and its competitor $\Lbound$.
+The following three lexica exemplify each kind, with $\Lall$ being a lexicon of the first kind.
+It is not optimal for communication as it assigns all its messages to the same state, $\sall$.
+This failure to associate a state to single a form inevitably leads to a communicative disadvantage in its use but it an serves illustrative purpose in comparison to our target, $\Llack$, and its competitor $\Lbound$, in virtue of being the simplest lexicon complexity-wise.
 
@@
 
-As we consider the full space of possible lexica, some assign the same concept to more than one message, as $\Lall$, or lexicalize the same meanings but associate them with different messages, as the $6$ $\Llack$-style types.
-The main results reported in the following do not hinge on the inclusion of these lexica but such an unadulterated space does better illustrate the influence of the linguistic pressures at hand.
+As we consider the full space of possible lexica, some assign the same concept to more than one message, as $\Lall$, or lexicalize the same meanings but associate them with different messages, as the $6$ $\Llack$-style lexica.
+The main results reported in the following do not hinge on the inclusion of these lexica but this an unaltered space does better illustrate the influence of the linguistic pressures at hand.
 
@@
 The replicator dynamic is sensitive to $\lambda$ as it influences signaling behavior and thereby has a bearing on a type's fitness.
-This influence is showcased in Figure \ref{fig:only-R}, which depicts the proportion of pragmatic $\Llack$-style speakers in $5$ independent populations after $50$ generations, as well as that of the type with the highest proportion, the incumbent.
+This influence is showcased in Figure \ref{fig:only-R}, which depicts the proportion of pragmatic $\Llack$-style speakers in $5$ independent populations after $50$ generations, as well as that of the type with the highest proportion in a population, the incumbent.
 As is to be expected, less expressive types do not fare well under pressure for communicative success and those able to convey states less equivocally fare better.
-This difference increases with $\lambda$ as this parameter promotes more deterministic linguistic behavior, widening the gap in expressivity between types.
-Importantly, high values of $\lambda$ alone do not lend a justification to the systematic prevalence of $\Llack$-like semantics paired with pragmatic reasoning that we set out to explain.
+This difference increases with $\lambda$ as it promotes more deterministic linguistic behavior, widening the gap in expressivity between types.
+Importantly, high values of $\lambda$ alone do not lend a justification to the systematic prevalence of a division of labor between semantics and pragmatics that we set out to explain.
 An inspection of $1000$ independent populations with $\lambda = 20$ shows a similar outcome to that hinted at by Figure \ref{fig:only-R}: Pragmatic $\Llack$ types only make up $11$ cases of incumbency, corresponding to a mean proportion of $0.003$ across populations.
@@
 Overall, incumbents of any type make up $50\%$ of the mean outcome.
-This, again, showcases the strong dominance of $\Lbound$ semantics over other alternatives when only expressivity is at stake.
+This shows the strong dominance of $\Lbound$-like semantics over other alternatives when only expressivity is at stake.
 
@@
 Put differently, it is easier to infer than functionally comparable lexica such as $\Lbound$ and it is less likely to be mistaken with others because its pragmatic use approximates a one-to-one form-meaning mapping.
-Consequently, as suggested by Figure \ref{fig:only-M}, in $1000$ analyzed independent populations all incumbents were pragmatic $\Llack$ users ($l = 15$ and $\lambda = 20$).
+Consequently, as suggested by Figure \ref{fig:only-M}, in $1000$ analyzed independent populations all incumbents were pragmatic $\Llack$ users provided sufficient rationality and a tendency towards posterior maximization ($l = 15$, $\lambda = 20$).
+By contrast, low values of $\lambda$ and $l$ reflect the inductive bias more directly, favoring functionally deficient but a priori preferred types such as literal $\Lall$.
+In the extreme, all out of $1000$ independent populations had this type as incumbent ($\lambda = 1$,$l = 1$).
 
@@
 First and foremost, neither dynamic on its own comes close to converging to a monomorphic population of a type of pragmatic $\Llack$ speakers.
-Under communicative pressure, the communicative advantage of $\Lbound$ leads to its selection when $\lambda$ is sufficiently high.
-Pressure only towards learnability has a modest but clear effect in differentiating $\Llack$ from such functionally similar but semantically more complex alternatives.
-However, this pressure on its own does not foment a propagation of these types across the population.
-Additionally, it leads to mixed populations that do not justify the selection of a particular type over others either.
+On the one hand, the slight communicative advantage of $\Lbound$'s lexicalization of an upper-bound over pragmatic use of $\Llack$ leads to the former's selection under a pressure for expressivity.
+On the other hand, pressure only towards learnability has a modest but clear effect of promoting pragmatic $\Llack$ over functionally similar but semantically more complex alternatives such as $\Lbound$.
+However, on its own learnability does not strongly foment the propagation of these types across the population.
+Furthermore, this pressure alone leads to heterogeneous populations that do not justify the selection of a particular type over others.
 
- Figure \ref{fig:cost} illustrates the effect of the learning bias for posterior sampling (\ref{fig:cost}.A) and slightly more MAP-like learning (\ref{fig:cost}.B) when pressured for both expressivity and learnability.
-More detailed results for all types across a sample of $c$-values for $l = 1$ and $l = 5$ are presented in Table \ref{tab:numeric-results}.
-These results show that a weak bias is sufficient to lead to a selection of $L_5$ over $L_4$.
-As when only learnability is considered, this effect increases with the bias' strength, provided $L_5$ users are pragmatic.
-Importantly, the addition of a pressure towards expressivity magnifies this effect and dampens the proliferation of functionally suboptimal types advantaged by the learning bias.
-As stressed above, this indicates that neither a learning bias nor functional pressure alone but their combination may lead to the lack of upper-bounds in the lexical meaning of scalar expressions.
+Figure \ref{fig:rmd} illustrates the combined effects of both pressures for a sample of $\lambda$ and $l$ values.
+Overall, these results show that an inductive learning bias for simpler semantics in tandem with functional pressure can lead to the selection of $\Llack$-like semantics over $\Lbound$.
+As when each pressure was considered in isolation, this effect increases with $\lambda$ and $l$.
+Importantly and by contrast to the effects of learnability alone, the addition of a pressure towards expressivity magnifies this effect and dampens the proliferation of similar or identical types.
+As stressed above, this suggests that neither a learning bias nor functional pressure alone but their combination may lead to the lack of upper-bounds in the lexical meaning of scalar expressions in a mostly monomorphic population.
+\tb{However, not even its combination with a pressure for expressivity completely drives out competing types under this inference strategy.
+This is not so for stronger posterior maximizing learning.
+The proportion to which a single pragmatic $\Llack$-style type increases grows with MAP-like learning.
+}
 
@@
 \caption{Proportion of pragmatic $\Llack$-style types and incumbent types in $5$ independent populations after $50$ generations under both pressures ($k = 5$).}
-\label{fig:only-M}
+\label{fig:rmd}
 \end{figure}
 
-Other than the involvement of both pressures, the resulting proportion of pragmatic $L_5$ speakers primarily hinges on three factors.
-First, the degree to which linguistic behavior is deterministic.
-This plays a role both for expressivity as well as in producing data that allows learners to discriminate this type from others.
-Second, the inductive bias, which controls the learners preference for simpler lexical representations.
-Lastly, the posterior parameter, as it magnifies the effects of the learning bias in tandem with replication.
-
-As showcased by Figure \ref{fig:cost}.A, posterior sampling can lead to the incumbency of pragmatic $L_5$.
-However, not even a strong favorable learning bias combined with a pressure for expressivity completely drives out competing types under this inference strategy.
-This is not so for stronger posterior maximizing learning.
-As shown in Figure \ref{fig:prior-posterior}, the range of bias values within which $L_5$ takes over the population increases with MAP-like learning.
-Put differently, the strength of the learning bias required for a given final proportion of $L_5$ speakers strongly depends on learners' inferential strategy.
-As for the effect of the other parameters not mentioned so far, changes in sequence length influence the population in a predictable way: smaller values lead to more heterogeneous populations that reflect the learner's prior more faithfully.
-Larger ones lead to more pronounced differences among equally preferred types.
-This is due to the fact that the likelihood that a small sequence was produced by any type is relatively uniform (modulo prior) compared to that of types with lexica $L_1$ - $L_3$ to produce larger sequences with the same state-message combination in contrast to pragmatic speakers of $L_4$ - $L_6$, or literal $L_4$.
-\tb{We need to discuss $\alpha$ here as well if noisy perception is taken out}
+As before, low $\lambda$ and $l$ lead to the incumbency of communicatively suboptimal types that learners are biased for, such as $\Lall$.
+An increase in $\lambda$ leads to the selection and incumbency of pragmatic $\Llack$ types but does not lead to monomorphic populations if learners sample from the posterior.
+Finally, a combination of sufficiently rational signaling behavior with some tendency for posterior maximization leads to increasingly monomorphic populations of types that convey upper-bounds pragmatically.
+A more succinct representation of the difference between the mean of the highest pragmatic $\Llack$-like types and the highest proportion of other types across $\lambda$ and $l$ values is given in Figure \ref{fig:diff}.
 
@@
 \includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig4-incumbents-difference}
-%\caption{Mean proportion of pragmatic $L_5$ in $1000$ populations after $20$ generations ($\alpha = 1, \lambda = 30, k = 5$)}
-\label{fig:prior-posterior}
+\caption{Difference between mean proportion of highest pragmatic $\Llack$ type and highest other type in $1000$ independent populations after $50$ generations under both pressures ($k = 5$).}
+\label{fig:diff}
 \end{figure}
+
+As for the effect of the sequence length $k$ not mentioned so far, it influences populations in a predictable way: small values lead to more heterogeneous populations that reflect the learner's prior more faithfully.
+This is due to the fact that the likelihood that a small sequence was produced by any type is relatively uniform (modulo prior).
+By contrast, larger values increasingly allow learners to differentiate types with different signaling behaviors.
+
+To recapitulate, other than the involvement of both pressures, the resulting proportion of pragmatic $\Llack$ speakers primarily hinges on three factors.
+First, the degree to which linguistic behavior is deterministic.
+This plays a role both in expressivity as well as in producing data that allows learners to discriminate this type from others.
+Second, the inductive bias, which leads learners to prefer simpler over more complex semantic representations in acquisition.
+Lastly, the posterior parameter, as it magnifies the effects of the learning bias in tandem with replication.
 

### 2017-03-17 paper/paper-v3.tex
tentative revision draft with correct plots

@@
 \centering
-\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig1-onlyr}
-\caption{Proportion of pragmatic $\Llack$-style types and incumbent types in $5$ independent populations after $50$ generations under only a pressure for expressivity.}
+\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig1-onlyr}
+\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a pressure for expressivity.}
 \label{fig:only-R}
@@
 \includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig2-onlym-pr}
-\caption{Proportion of pragmatic $\Llack$-style types and incumbent types in $5$ independent populations after $50$ generations under only a pressure for learnability ($\lambda = 20, k = 5$).
+\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a pressure for learnability ($\lambda = 20, k = 5$).
 The types' prior probability is shown in the right-most plot.}
@@
 
-The effect of iterated learning without a pressure for expressivity using posterior sampling ($l = 1$) and a stronger tendency towards posterior maximization ($l = 15$) is shown in Figure \ref{fig:only-M} together with the prior over types.
-The latter illustrates that users of $\Llack$ are, while not the most favored types, advantaged by the inductive bias in virtue of the relatively simple semantics they conventionalize.
-Accordingly, a stronger propensity to maximize the posterior increases their proportion in the population in analogous fashion to increase in $\lambda$ under the replicator dynamic.
-However, while replication generally succeeds in driving out competing types of a kind, iterated learning leads to their coexistence as each of these types is passed on to next generations equal faithfulness.
-Although $\Llack$ is not the most favored type by the prior, it is the least semantically complex language in the type space that enables its users to convey each state with a single message when combined with pragmatic reasoning (provided sufficiently high $\lambda$).
-Put differently, it is easier to infer than functionally comparable lexica such as $\Lbound$ and it is less likely to be mistaken with others because its pragmatic use approximates a one-to-one form-meaning mapping.
-Consequently, as suggested by Figure \ref{fig:only-M}, in $1000$ analyzed independent populations all incumbents were pragmatic $\Llack$ users provided sufficient rationality and a tendency towards posterior maximization ($l = 15$, $\lambda = 20$).
-By contrast, low values of $\lambda$ and $l$ reflect the inductive bias more directly, favoring functionally deficient but a priori preferred types such as literal $\Lall$.
-In the extreme, all out of $1000$ independent populations had this type as incumbent ($\lambda = 1$,$l = 1$).
+The effect of iterated learning without a pressure for expressivity with posterior sampling ($l = 1$) and a stronger tendency towards posterior maximization ($l = 15$) is shown in Figure \ref{fig:only-M} together with the prior over types.
+The latter shows that while users of $\Llack$ are not the most favored by the inductive bias (compared, e.g., to $\Lall$) they are nevertheless advantaged in virtue of the relatively simple semantics they conventionalize (compared, e.g., to $\Lbound$).
+Additionally, $\Llack$ enables its users to convey each state with a single message when combined with pragmatic reasoning provided sufficiently high $\lambda$.
+This makes it less likely to be confused with other types if the data is not too sparse ($k \geq 5$).
+Put differently, learners have a higher propensity to infer pragmatic $\Llack$ over types that might be hard or impossible to distinguish it from based on overt linguistic behavior, such as $\Lbound$, and $\Llack$ is also less likely to be mistaken with types with different linguistic behaviors because its pragmatic use approximates a one-to-one form-meaning mapping.
+As a consequence, a stronger propensity to maximize the posterior increases their proportion in the population.
+However, while we showed replication to generally succeed in driving out competing types of a kind, e.g., different pragmatic $\Llack$-style types, iterated learning leads to their coexistence as each of these types is passed on to next generations equal faithfulness and they do not stand in competition with each other.
+As already suggested by Figure \ref{fig:only-M}, in $1000$ analyzed independent populations all incumbents were pragmatic $\Llack$ users provided sufficient rationality and a tendency towards posterior maximization ($l = 15$, $\lambda = 20$).
+By contrast, low values of $\lambda$ and $l$ reflect the learners' inductive bias more directly, favoring functionally deficient but a priori preferred types such as literal $\Lall$.
+In the extreme, when $l = 1$ and $\lambda = 1$ all of $1000$ independent populations had $\Lall$ as incumbent.
 
@@
 Inspecting these pressures separately not only showcases the contribution of the model's components but also highlights some of their broader implications.
-First and foremost, neither dynamic on its own comes close to converging to a monomorphic population of a type of pragmatic $\Llack$ speakers.
+First and foremost, neither dynamic on its own comes close to converging to a monomorphic population of pragmatic $\Llack$ speakers.
 On the one hand, the slight communicative advantage of $\Lbound$'s lexicalization of an upper-bound over pragmatic use of $\Llack$ leads to the former's selection under a pressure for expressivity.
 On the other hand, pressure only towards learnability has a modest but clear effect of promoting pragmatic $\Llack$ over functionally similar but semantically more complex alternatives such as $\Lbound$.
-However, on its own learnability does not strongly foment the propagation of these types across the population.
+However, on its own learnability does not foment the propagation of these types across the population.
 Furthermore, this pressure alone leads to heterogeneous populations that do not justify the selection of a particular type over others.
-
-Figure \ref{fig:rmd} illustrates the combined effects of both pressures for a sample of $\lambda$ and $l$ values.
-Overall, these results show that an inductive learning bias for simpler semantics in tandem with functional pressure can lead to the selection of $\Llack$-like semantics over $\Lbound$.
-As when each pressure was considered in isolation, this effect increases with $\lambda$ and $l$.
-Importantly and by contrast to the effects of learnability alone, the addition of a pressure towards expressivity magnifies this effect and dampens the proliferation of similar or identical types.
-As stressed above, this suggests that neither a learning bias nor functional pressure alone but their combination may lead to the lack of upper-bounds in the lexical meaning of scalar expressions in a mostly monomorphic population.
-\tb{However, not even its combination with a pressure for expressivity completely drives out competing types under this inference strategy.
-This is not so for stronger posterior maximizing learning.
-The proportion to which a single pragmatic $\Llack$-style type increases grows with MAP-like learning.
-}
 
@@
 \includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig3-r+m}
-\caption{Proportion of pragmatic $\Llack$-style types and incumbent types in $5$ independent populations after $50$ generations under both pressures ($k = 5$).}
+\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under both pressures ($k = 5$).}
 \label{fig:rmd}
@@
 
-As before, low $\lambda$ and $l$ lead to the incumbency of communicatively suboptimal types that learners are biased for, such as $\Lall$.
+Figure \ref{fig:rmd} illustrates the combined effects of both pressures for a sample of $\lambda$ and $l$ values.
+These results show that an inductive learning bias for simpler semantics in tandem with functional pressure can lead to the selection of $\Llack$-like semantics over $\Lbound$.
+As when each pressure was considered in isolation, the proportion of these types increases with $\lambda$ and $l$.
+Crucially, the addition of a pressure towards expressivity magnifies the learning outcome and dampens the proliferation of similar or identical types.
+As stressed above, this suggests that neither a learning bias nor functional pressure alone but their combination may lead to the lack of upper-bounds in the lexical meaning of scalar expressions in a mostly monomorphic population.
+
+As before, low $\lambda$ and $l$ lead to the incumbency of communicatively suboptimal types that are a priori favored by learning, such as $\Lall$.
 An increase in $\lambda$ leads to the selection and incumbency of pragmatic $\Llack$ types but does not lead to monomorphic populations if learners sample from the posterior.
@@
 A more succinct representation of the difference between the mean of the highest pragmatic $\Llack$-like types and the highest proportion of other types across $\lambda$ and $l$ values is given in Figure \ref{fig:diff}.
+\tb{Say more}
 
@@
 \centering
-\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig4-incumbents-difference}
+\includegraphics[width=1\textwidth,height=6cm,keepaspectratio]{./plots/fig4-incumbents-difference}
 \caption{Difference between mean proportion of highest pragmatic $\Llack$ type and highest other type in $1000$ independent populations after $50$ generations under both pressures ($k = 5$).}
@@
 \subsection{Discussion}
-Under the assumption of a learning bias for simpler lexical representations, our results suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome communicative pressures and stabilize in a population.
+Under the assumption of a learning bias for simpler lexical representations, our results suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome communicative pressures and be majoritarily selected for in a population.
 This prediction hinges on three assumptions.
 First, that language is pressured toward both expressivity and learnability.
-Second, that language use is relatively deterministic.
-Lastly, that learners prefer simpler over more complex lexical representations.
-An important addendum to this third condition being that a combination of rationality in choice and maximization in learning requires a weaker bias towards simplicity.
-Under these conditions the selection of lexical meanings lacking upper-bounds in populations of pragmatic speakers is robust against parameter perturbations.
-This outcome is particularly encouraging in light of other advantages a lack of semantic upper-bounds may confer, as discussed at the beginning of Section \ref{sec:si-case-study}.
+Second, that language users favor expressions/interpretations expected to have a higher likelihood of being understood/meant over ones with a lower likelihood.
+Lastly, that learners prefer simpler over more complex lexical representations and exhibit a tendency towards the acquisition of the type that best explains the learning data.
+This outcome is particularly encouraging in light of other advantages a lack of semantic upper-bounds may confer, as discussed at the beginning of this section.
 
@@
 However, it not clear to what extent other types should be present in the final population.
-It seems reasonable to expect functionally suboptimal types $L_1$, $L_2$ and $L_3$ to be ruled out because they fail to enable their users to communicate effectively.
-However, this is not true of $L_4$.\footnote{$L_6$ presents a special case.
-In our current setup, it mirrors $L_5$ in enabling for the pragmatic strengthening of a message that does not codify an upper-bound lexically.
-However, this is achieved by ruling out $\ssome$ and not, as with scalar implicatures, $\sall$.
-$L_6$ speakers therefore strengthen a ``some''-message to convey something paraphrasable as `some but not [some but not all]'.
-The current representation of lexica as Boolean matrices is blind to this anomaly.} The prediction that natural language communities are homogeneous or that a single speaker may entertain $L_4$-like semantics for one scalar expression and $L_5$-like semantics for another is not implausible (cf.
+It seems reasonable to expect functionally suboptimal lexica such as $\Lall$ to be ruled out because they fail to enable their users to communicate effectively and are not evinced in natural language.
+However, this is not true of $\Lbound$.
+The prediction that natural language communities are homogeneous or that a single speaker may entertain $\Llack$-like semantics for one scalar expression and $\Lbound$-like semantics for another is not implausible (cf.
 \citealt{franke+degen:2016}).
-Alternatively, a stronger tendency for posterior maximization has to be assumed (see Figure \ref{fig:prior-posterior}).
+Alternatively, a stronger tendency for posterior maximization in learning has to be assumed.
 This empirical issue relates to other two aspects left undiscussed: disadvantages of pragmatic reasoning and the effect of state frequencies on the fossilization of pragmatic inferences.
@@
 Should cost play a role, then its effect is bound to depend on the frequency with which a given scalar expression is used.
-It is therefore plausible that frequently drawn scalar implicatures might fossilize to avoid cost, while infrequent ones are still derived on-line.
-This also opens a possible venue to address the preceding question about the expected presence of $L_4$-like semantics, but further empirical evidence is needed to assess these matters beyond speculation.
+It may be that frequently drawn scalar implicatures fossilize to avoid cost whereas infrequent ones are still derived on-line.
+This also opens a possible venue to address the preceding question about the expected presence of $\Lbound$-like semantics in a population, but further empirical evidence is needed to assess these matters.
 
 \section{General discussion}\label{sec:discussion}
-We laid out a model that combines game theoretical models of functional pressure towards efficient communication \citep{nowak+krakauer:1999}, effects of transmission perturbations on (iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012, franke+jaeger:2014} as well as different lexica \citep{bergen+etal:2012,bergen+etal:2016}.
+We laid out a model that combines game theoretical models of functional pressure towards efficient communication \citep{nowak+krakauer:1999}, effects of transmission perturbations on (iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012, franke+jaeger:2014} as well as different lexica \citep{bergen+etal:2012,bergen+etal:2016}, and \tb{mention LOT here}.
 This model generates predictions about lexicalization patterns and, more generally, effects of communicative pressures on the cultural evolution of language.
@@
 In the realm of inductive biases, we adopted the assumption that simpler semantic representations are more likely to be learned (cf.
-\citealt{chater+vitanyi:2003}).
+\citealt{chater+vitanyi:2003}), \tb{mention LOT here}.
 Under this view, semantics and pragmatics play a synergic role in that representational simplicity is supplemented by pragmatic reasoning to counteract functional disadvantages otherwise incurred.
 As a consequence, iterated transmission and use of language lead to a regularization that may explain the lack of lexicalization of systematic pragmatic enrichments.
-This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics.
-It offers an account of why (certain) pragmatic inferences fail to lexicalize.
-More generally, we showed that systematic noise in perception can produce outcomes that are similar from those generated by inductive biases.
+This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics by offering an account of why certain pragmatic inferences fail to lexicalize.
 
-The main innovations of the model are its modular separation of expressivity and learnability, allowing for their isolated and combined analysis, the learning process involving a joint inference over types of pragmatic behavior and lexical meaning, as well as in its accommodation of different transmission perturbations that go beyond learning biases.
+The main innovations of the model are its modular separation of expressivity and learnability, allowing for their isolated and combined analysis, the learning process involving a joint inference over types of pragmatic behavior and lexical meaning.
 The goal to decouple but model both expressivity and learnability has also recently been addressed by \citet{kirby+etal:2015}.
@@
 model expressivity as exerting its force only in the production of learning data.
-This model's expressivity parameter thereby fulfills a similar role  to high values of $\lambda$ in making speaker behavior more deterministic.
-In this way, it ``favors'' unambiguous languages.
-However, the degree of mutual understanding of interlocutors central to replication and to our notion of expressivity is not taken into consideration.
+On a general level, this model's expressivity parameter has a similar effect to $\lambda$ in that it can give a communicative advantage to unambiguous languages over ambiguous ones.
+\tb{Should we go into more detail here?
+I think it's fine if we leave it out but $\lambda$ has a broad motivation whereas Kirby et al.'s parameter is quite explicitly a penalty parameter against ambiguity.
+I mainly wanted to give an idea of what the pressure for expressivity in their model does rather than argue for why ours is more principled.} However, the degree of mutual understanding of interlocutors central to replication and to our notion of expressivity is not taken into consideration.
 That is, while our proposal combines bidirectional horizontal transmission with its vertical and unidirectional counterpart, Kirby et al.'s model only considers the latter's influence.
-Our reasoning behind the inclusion of the former lies in the empirical and theoretical observation that learnability alone can lead the selection of functionally defective languages, as illustrated by the tautological language $L_3$ in our analysis.
-This outcome has been reported in a number of laboratory experiments where the participants' task was to learn and subsequently reproduce the language produced by a previous participant, leading to a proliferation of languages that associated a large number of meanings with a single form (see e.g.
-\citealt{silvey+etal:2014} and experiment 1 in \citealt{kirby+etal:2008}).
-In contrast, experiments involving an interactive component have been found to foster languages that enable interlocutors to distinguish meanings more accurately  (e.g.
-\citealt{fay+etal:2013}; for a review of laboratory results under the iterated learning paradigm and further discussion see \citealt{kirby+etal:2015, tamariz+kirby:2016}).
+Our reasoning behind the inclusion of the former lies in the empirical and theoretical observation that learnability alone can lead the selection of functionally defective languages, as discussed in relation to $\Lall$ when $\lambda$ or $k$ are low.
+This outcome has been reported in a number of laboratory experiments where subjects learned and subsequently reproduced the language produced by a previous participant over multiple iterations, leading to a proliferation of ambiguous languages that associate a large number of meanings with a single form (see, e.g., \citealt{silvey+etal:2014} and experiment 1 in \citealt{kirby+etal:2008}).
+In contrast, experiments involving an interactive component have been found to foster languages that enable interlocutors to communicatively distinguish meanings more accurately  (e.g., \citealt{fay+etal:2013}; for a review of laboratory results under the iterated learning paradigm and further discussion see \citealt{kirby+etal:2015, tamariz+kirby:2016}).
 It is not evident how to compare these empirical findings given that they consider distinct meaning spaces, modes of transmission, iterations and feedback given to participants.
 However, we take these results to suggest that there is an important difference between a language generating learnable linguistic data and its actual performance as a means of information transfer.
-The former solely depends on the mechanism by which speakers associate form and meaning.
+\tb{Explain the next two sentences} The former solely depends on the mechanism by which speakers associate form and meaning.
 The latter additionally hinges on the addressee's linguistic experience and her ability to interpret linguistic input based on this experience.
 In sum, we contend that successful information transfer in a linguistic community is central to the adoption of a communication system and that this measure is not adequately reflected by production alone.
-
-The demonstration that noise can lead to regularized evolutionary outcomes that are similar to those generated by prior learning biases is relevant not for the case study at hand, but more so for the broader project of investigating the cultural evolution of language.
-On the one hand, the plurality of sources of transmission perturbations admitted by these models paints a cautionary tale for the design of studies that purport to provide explanatory accounts of linguistic phenomena.
-In particular when the outcome is interpreted as being informative about the perturbation assumed to generate it (cf.
-\citealt{tamariz+kirby:2016}).
-On the other, and most importantly, it showcases how regularities can arise as a byproduct of systematic noise rather than from standardly assumed inductive biases.
 
@@
 We set out to investigate this process by putting forward a model that combines a pressure toward efficient and successful information transfer with perturbations that may arise in the transmission of linguistic knowledge in acquisition.
-Additionally, we argued for the necessity of considering the role of pragmatics in investigations on the cultural evolution of meaning.
+\tb{Clarify and put this in better words.
+This is quite central:} Additionally, we argued for the necessity of considering the role of pragmatics in investigations on the cultural evolution of meaning.
 These components and their mutual influence were highlighted in a case study on the lack of lexical upper-bounds in weak scalar expressions that showed that, when pressured for learnability and expressivity, the former drives for simpler semantic representations inasmuch as pragmatics can compensate for lack of expressivity in use.
-That is, the relative learning advantage of simpler semantics in tandem with a functional pressure in use may offer an answer to why natural languages fail to lexicalize systematic pragmatic inferences.
-
-We also considered an alternate instantiation of the model, which shows that systematic noise in state perception can give rise to evolutionary outcomes that are similar to those predicted by inductive biases.
-This stresses the fact that that learning and typology are not necessarily close reflections of each other \citep{bowerman:2010}.
-In particular, language use and environmental factors can play an important role in language change, making them central variables in explanatory accounts of natural language properties.
+That is, the relative learning advantage of simpler semantics in combination with  functional pressure in use may offer an answer to why natural languages fail to lexicalize systematic pragmatic inferences.
 

### 2017-03-17 code-LOT-extension/plots.py

@@ -3,6 +3,8 @@ import pandas as pd
 import numpy as np 
 import matplotlib.pyplot as plt
 import seaborn as sns
+import matplotlib.patches as patches #to draw rectangles
+from mpl_toolkits.axes_grid.inset_locator import inset_axes, zoomed_inset_axes
 #sns.set_context(rc={'lines.markeredgewidth': 0.5})
 import os
 import glob
@@ -33,110 +35,240 @@ def analysis(types,val1,val2,kind):
     return [values_of_inc_in_bins / len(df), numb_of_inc]
 
 
-def get_subfigs_incumbents(type_list,list1,list2,kind,prior=False):
+def get_subfigs_replication(type_list,list1,list2):
     seq_length = 5
     num_of_runs = 5
     
-    if not(kind == 'm'):
-        x_figs = len(list1)
-        y_figs = len(list2)
-    else:
-        x_figs = len(list2)
-        y_figs = len(list1)
+    types_to_plot = ['t_final'+str(x) for x in type_list]  
+    restrict_to_final = ['t_final'+str(z) for z in xrange(432)] #as to avoid the incumbent to come from other columns
+
+    df1 = pd.read_csv('./results/%s-s3-m3-lam%d-a1-k%d-samples250-l%d-g50-meFalse.csv' % ('r',list1[0],seq_length,list2[0]))
+    df2 = pd.read_csv('./results/%s-s3-m3-lam%d-a1-k%d-samples250-l%d-g50-meFalse.csv' % ('r',list1[1],seq_length,list2[0]))
+    
+    df1 = df1[restrict_to_final]
+    df1['incumbent'] = df1.max(axis=1)
+
+    df2 = df2[restrict_to_final]
+    df2['incumbent'] = df2.max(axis=1)
+    
+    df1_large = df1[types_to_plot + ['incumbent']]
+    df1_large = df1_large.iloc[:num_of_runs]
+    Y1_large = df1_large.values
+                 
+    df2_large = df2[types_to_plot + ['incumbent']]
+    df2_large = df2_large.iloc[:num_of_runs]
+    Y2_large = df2_large.values
+
+    df1_targets = df1[types_to_plot]
+    df1_targets = df1_targets.iloc[:num_of_runs]
+    Y1_small = df1_targets.values
+
+    df2_targets = df2[types_to_plot]
+    df2_targets = df2_targets.iloc[:num_of_runs]
+    Y2_small = df2_targets.values
+
+
+
[…]

### 2017-03-18 paper/paper-v3.tex
minor revisions up to 3.1.5

@@
   conventional semantic meaning and general pragmatic reasoning applied
-  to the context of utterance. This implies that models of cultural evolution of meaning
+  to the context of utterance. This implies that models of the cultural evolution of meaning
   should likewise take into consideration that observable language use is a complex interaction
@@
 speaker is cooperative and relevant, ``I have to work'' may be interpreted as providing a
-reason why the speaker will not be able to accept the invitation, going beyond its literal
+reason why the speaker will not be able to accept an invitation, going beyond its literal
 meaning. Some of these enrichments are rather \emph{ad hoc}. Others show striking regularities,
@@
 
-A particularly productive and well studied class of pragmatic enrichments are scalar
+A particularly productive and well studied class of systematic pragmatic enrichments are scalar
 implicatures
@@
 If this majority view is correct, the question arises how such a division of labor between
-semantics and pragmatics could have evolved. Models of language evolution abound. There are
+semantics and pragmatics could have evolved and why it would be so pervasive across natural languages. Models of language evolution abound. There are
 simulation-based models studying the evolution of language in populations of communicating
@@
 communicative efficiency and (ii) possible infidelity in the transmission of linguistic
-knowledge, caused by factors such as inductive learning biases and sparse data to learn from. Considering
+knowledge, caused by factors such as inductive learning biases and sparse learning data. Considering
 transmission of linguistic knowledge is important because neither semantic meanings nor
@@
 faithful transmission across generations. Clearly, an unlearnable code will not make it past
-the one happy fellow who invented it. Importantly, even small biases implicit in acquisition
+the one happy fellow who invented it. Importantly, even small learning biases implicit in acquisition
 can build up and have quite striking effects on an evolving language in a process
@@
 While natural languages are pressured for both expressivity and learnability these forces may pull in opposite directions. Their opposition becomes particularly clear when
-considering the extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). A language with a
+considering the extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). A language consisting of a 
 single form-meaning association is easy to learn but lacking in expressivity. Conversely, a
-language that associates a distinct form with all possible meanings a speaker may want to
+language that lexicalizes a distinct form for each possible fine-grained meaning a speaker may want to
 convey is maximally expressive but challenging to acquire.
@@
 have an example from a lecture ready at hand; it's a simple coordination game in a
-one-population setting.} \tb{Yes, I think most readers would appreciate it. We can always cut it out again if the reviewers prefer more compression.}
+one-population setting.} \tb{Yes, I think that's a good idea.
+We can always cut it out again if the reviewers prefer more compression but let's explain more and cut later.
+The paper is relatively short at the moment.}
 
@@
 a prisoner's dilemma. \mf{maybe good to refer back to the example from before if there was
-  one?} \tb{Yes!} For our present purpose, types are identified by their cognitive make-up. Since we are
+  one?} \tb{Yes!} For our present purpose types are identified by their cognitive make-up. Since we are
 interested in the question under which conditions processes of cultural evolution will favor
@@
 same sentence with \emph{some} replaced by \emph{all}.  
-Lexica for this case would assign a Boolean truth value, either $0$ or $1$, to each state-message pair. The following two lexica exemplify the distinction
+Lexica for this case would assign a Boolean truth value, either $0$ for false or $1$ for true, to each state-message pair. The following two lexica exemplify the distinction
 between a lexicalized upper-bound for \emph{some} in $\Lbound$ and the widely assumed logical
@@
 \ssome;L_{\text{bound},\text{lack}}) \approx 1$. But the probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
-\sall;\Lbound) \approx 1$ for $\Lbound$ whereas it is still $S_{0}(\msome \mid \sall;\Llack) = .5$ because $\msome$ is semantically associated with both states by $\Llack$.
+\sall;\Lbound) \approx 1$ for $\Lbound$ whereas it is still $S_{0}(\msome \mid \sall;\Llack) = .5$ for $\Llack$ because it semantically associates $\msome$ with both states.
 
-Pragmatic behavior of level $n+1$ is similar to its literal counterparts but uses the
-interpretation or production behavior of a level-$n$ player instead of the lexical meaning:
+Pragmatic behavior of level $n+1$ is similar to its literal counterparts but bases linguistic choice in 
+interpretation or production on the expected behavior of a level-$n$ interlocutor instead of on lexical meaning:
 \begin{flalign}
@@
 
-Particularly important to our purpose is the fact that players of a level greater than $0$ using $\Llack$ can come to {\em pragmatically} associate $\msome$ with $\ssome$ over $\sall$ in contrast to their literal counterparts.
+Particularly important to our purpose is the fact that players of a level greater than $0$ using $\Llack$ can {\em pragmatically} associate $\msome$ with $\ssome$ over $\sall$,  in contrast to their literal counterparts.
 Intuitively, this is so because hearers reason a speaker to convey $\sall$ with $\mall$, as this has a higher chance of leading to mutual understanding than using underspecified $\msome$.
@@
 \citep[e.g.,][]{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Under a biological
-interpretation, the assumption is that organisms have a higher chance of survival and
+interpretation the assumption is that organisms have a higher chance of survival and
 reproduction if they are able to share and receive useful information via communication with
-peers. Under a cultural interpretation, the picture is that agents themselves strive towards
+peers. Under a cultural interpretation the picture is that agents themselves strive towards
 communicative success and therefore occasionally adapt or revise their behavior to achieve a
@@
 average success of $i$ when talking or listening to $j$. Assuming that agents are speakers half
-of the time this yields:
+of the time, this yields:
 \begin{align*}
@@
 speaker and as a hearer when communicating with $j$, defined as follows, where $n_i$ and $n_j$
-are type $i$'s and $j$'s pragmatic reasoning types and $L_i$ and $L_j$ are their lexica:
+are type $i$'s and type $j$'s pragmatic reasoning types and $L_i$ and $L_j$ are their lexica:
 \begin{flalign*}
@@
 serves as input for a new learner. Due to the pressure towards learnability it exerts, iterated
-learning generally leads to simpler and more regular languages (see \citealt{kirby+etal:2014}
+learning alone generally leads to simpler and more regular languages (see \citealt{kirby+etal:2014}
 and \citealt{tamariz+kirby:2016} for recent surveys).
@@
 inference process. On the one hand, early simulation results suggested that weak biases could
-be magnified by exposing learners to only small data samples \citep[e.g., in][]{brighton:2002}. On the other, \citeposs{griffiths+kalish:2007} mathematical
+be magnified by exposing learners to only small data samples \citep[e.g. in][]{brighton:2002}. On the other, \citeposs{griffiths+kalish:2007} mathematical
 characterization showed that iterated learning converged to the prior, i.e., that the resulting
@@
 That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that stronger alternatives do not hold because the speaker used a weaker alternative instead.
-In this way, ``Chris owns some of Johnny Cash's albums'' is strengthened to convey that he owns some but not all albums.
+In this way, ``Chris owns some of Johnny Cash's albums'' is strengthened to convey that he owns {\em some but not all} albums.
 A bound that rules out stronger alternatives is thusly not codified in the lexical meaning of weak alternatives but instead pragmatically supplied.
@@
 This is so because a rational speaker would use a more informative message when in such a state.
-Conversely, a pragmatic speaker will reason about her interlocutor's expected interpretation and use the messages at her disposition accordingly.
+Conversely, a pragmatic speaker will reason about her interlocutor's expected interpretation and use the messages at her disposition accordingly as to maximize communicative success.
 
@@
 That is, we ask why lexical meanings that lack upper-bounds and convey it pragmatically are regularly selected for over alternatives such as that of codifying the bound semantically.
-More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} and {\em big} were truth-conditionally incompatible with stronger alternatives such as {\em hot}, {\em and}, {\em all} and {\em huge}?
+More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} and {\em big} were truth-conditionally incompatible with stronger alternatives such as {\em hot}, {\em and}, {\em all} and {\em huge} to avoid misunderstanding?
 This question is particularly striking considering the number of expressions that license such inferences across natural languages.
@@
 This can happen when embedded in downward-entailing contexts, when the speaker is likely uncertain about whether the upper bounded reading is true, or when the distinction between an upper-bounded reading and the simple, only lower-bounded reading, is not relevant.
-For instance, if for all the speaker knows `Chris owns some of Johnny Cash's albums' but she does not know whether he owns all, then the use of {\em some} lacking an upper-bound succinctly conveys her uncertainty.
+For instance, if for all the speaker knows Chris owns some of Johnny Cash's albums but she does not know whether he owns all, then the use of {\em some} lacking an upper-bound succinctly conveys her uncertainty.
 This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: Should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then this is preferred over expressing it overtly through longer expressions, e.g., by saying {\em some but not all} explicitly.
@@
 Fourth, larger lexica are costly.
-Overall, neither condition seems convincing as a pivotal explanatory device for such a widespread phenomenon.
+Overall, neither condition seems convincing as a central explanatory device for such a widespread phenomenon.
 The first two conditions put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of unambiguous communication.
@@
 We begin setting the stage by defining the learning prior and the set of types under consideration.
-Drawing from our preceding discussion, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $\Llack$-like semantics.
-However, it is instructive to first inspect the effect of these pressures in isolation.
+Drawing from our preceding discussion, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $\Llack$-like semantics and the majority view of a division of labor between semantics and pragmatics.
+Before coming to this, it is instructive to first inspect the effect of these pressures in isolation.
 We therefore first showcase the effects that expressivity and learnability have on their own before turning to those resulting from their combination.
 
-\subsubsection{Inductive learning bias against semantic complexity}
+\subsubsection{An inductive learning bias in favor of semantic simplicity}
 There is a growing effort to develop empirically  testable representational  languages that allow for the measure of semantic complexity.
@@
 Our toy grammar of concepts is given in Table~\ref{tab:grammar}.
-This grammar uses basic set-theoretic operations to form expressions which can be evaluated as true or false in different world states such as $\ssome$ or $\sall$ (see below).
+This grammar uses basic set-theoretic operations to form expressions which can be evaluated as true or false in different world states such as $\ssome$ or $\sall$.
 Applications of generative  rules have a cost attached to them.
-(Alternatively, a probability.) Here we simply assume that Boolean combinations of concepts are more complex than atomic concepts and that otherwise each rule application adds the same cost unit.
+Here we simply assume that Boolean combinations of concepts are more complex than atomic concepts and that otherwise each rule application adds the same cost unit.
 \begin{table}
@@
 There are other more sophisticated ways to define priors over lexica (see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a}) but the key assumption here, common to all of them, is that simple representational expressions should be favored and complex ones disfavored.
-More generally, we should stress that these details -- from generative grammar to complexity measure -- are stipulated and should therefore be regarded as an illustratory instantiation of our general assumptions instead of as strong empirical claims.
-%Instead, this choice showcases how insights and representations from linguistics, cultural evolution, evolutionary game theory, and (probabilistic) language learning may fruitfully be combined.
+More generally, we should stress that these details -- from the generative grammar to its complexity measure -- are stipulated and are to be regarded as an illustratory instantiation of our general assumptions instead of as strong empirical claims.
 
 \subsubsection{Lexica, signaling behavior \& types}
-We consider a state space with three states $\States = \set{\snone, \ssome, \sall}$. This space can be
+We consider a state space with three states $\States = \set{\snone, \ssome, \sall}$.
+\tb{Should we say something about the inclusion of $\snone$?
+It may be more natural to introduce this state earlier but may unnecessarily complicate the lexica showcased in the previous section.} This space can be
 thought of as a partition of possible worlds into cells where none, some or all of the $A$s are
@@
 messages there are $6^3 = 216$ possible lexica.
-Of these, three kinds of lexica are of particular relevance.
+Some assign the same concept to more than one message and others lexicalize the same meanings but associate them with different messages.
+The main results reported in the following do not hinge on the inclusion of these lexica but such an unaltered space does better illustrate the influence and differences of the linguistic pressures we consider.
+Additionally, the competition of different languages that lexicalize the same concepts through different forms allows us to take serious the problem of only one being selected for even though multiple alternatives of equal expressivity {\em and} learnability exist.
+
+Out of these possible lexica, three kinds are of particular relevance.
 First, lexica that assign the same concept to more than one message.
-Such lexica lack in expressivity but may be favored by the learning bias.
-Second, lexica that conventionalize upper-bounds to realize a (quasi-)partition of the relevant semantic space.
+Such lexica lack in expressivity but are favored by the learning bias if they lexicalize simple concepts.
+Second, lexica that conventionalize upper-bounds to realize a (quasi-)partition of the state space.
 As discussed before in relation to $\Lbound$, users of these lexica need not resort to pragmatic reasoning to convey an upper-bound with weak scalar expressions.
-Instead, this information is already given by the semantics of their lexicon, giving them an advantage in terms of expressivity.
-Such lexica are expected to be the main contenders to the natural language like semantics that fail to lexicalize an upper-bound.
+Instead, this information is already provided by the semantics of their lexicon, giving them an advantage in terms of expressivity.
+Types with such lexica are expected to be the main contenders to the natural language like semantics that do not lexicalize an upper-bound for weak scalar expressions.
 Third and lastly, $\Llack$-style lexica that, paired with mutual reasoning, can convey an upper-bound pragmatically.
-Overall, there are $6$ lexica of the second kind and $6$ of the third.
-The following three lexica exemplify each kind, with $\Lall$ being a lexicon of the first kind.
-It is not optimal for communication as it assigns all its messages to the same state, $\sall$.
-This failure to associate a state to single a form inevitably leads to a communicative disadvantage in its use but it an serves illustrative purpose in comparison to our target, $\Llack$, and its competitor $\Lbound$, in virtue of being the simplest lexicon complexity-wise.
+There are six lexica of the second kind and six of the third.
+The following three lexica exemplify each kind, with $\Lall$ being a lexicon of the first.
+This lexicon is not optimal for communication as it assigns all its messages to the same state, $\sall$.
+This inevitably leads to a communicative disadvantage in its use but it may nevertheless have a learnability advantage over our target $\Llack$ and its competitor $\Lbound$ in virtue of being the simplest lexicon complexity-wise, making it a priori favored by learners.
 
@@
 
-As we consider the full space of possible lexica, some assign the same concept to more than one message, as $\Lall$, or lexicalize the same meanings but associate them with different messages, as the $6$ $\Llack$-style lexica.
-The main results reported in the following do not hinge on the inclusion of these lexica but this an unaltered space does better illustrate the influence of the linguistic pressures at hand.
-
-Lastly, recall that types are a combination of a lexicon and a manner of language use.
+Recall that types are a combination of a lexicon and a manner of language use.
 We analyze the model's predictions in populations of types with one of the two signaling behaviors introduced earlier; literal or pragmatic.
@@
 Accordingly, there are $432$ types in this model.
+Six are variants of our pragmatic $\Llack$ target and twelve are either literal or pragmatic speakers of variants of its main contender $\Lbound$.
 
@@
 \includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig1-onlyr}
-\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a pressure for expressivity.}
+\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a pressure for expressivity.
+The right-most plots zoom in on only the proportion of pragmatic $\Llack$ types.}
 \label{fig:only-R}
@@
 
-The replicator dynamic is sensitive to $\lambda$ as it influences signaling behavior and thereby has a bearing on a type's fitness.
-This influence is showcased in Figure \ref{fig:only-R}, which depicts the proportion of pragmatic $\Llack$-style speakers in $5$ independent populations after $50$ generations, as well as that of the type with the highest proportion in a population, the incumbent.
+Communicative efficiency is sensitive to $\lambda$ as it influences signaling behavior.
+It therefore has a central bearing on types' fitness.
+This is showcased in Figure \ref{fig:only-R}, which depicts the proportion of pragmatic $\Llack$-style speakers in $5$ independent populations after $50$ generations, as well as that of the type with the highest proportion in a given population, the incumbent.
 As is to be expected, less expressive types do not fare well under pressure for communicative success and those able to convey states less equivocally fare better.
-This difference increases with $\lambda$ as it promotes more deterministic linguistic behavior, widening the gap in expressivity between types.
-Importantly, high values of $\lambda$ alone do not lend a justification to the systematic prevalence of a division of labor between semantics and pragmatics that we set out to explain.
-An inspection of $1000$ independent populations with $\lambda = 20$ shows a similar outcome to that hinted at by Figure \ref{fig:only-R}: Pragmatic $\Llack$ types only make up $11$ cases of incumbency, corresponding to a mean proportion of $0.003$ across populations.
-By contrast, $913$ incumbents were $\Lbound$-style users with close to an even share between literal ($454$) and pragmatic types ($459$), corresponding to a mean proportion of about $48\%$ together.
-Overall, incumbents of any type make up $50\%$ of the mean outcome.
-This shows the strong dominance of $\Lbound$-like semantics over other alternatives when only expressivity is at stake.
+This difference increases with $\lambda$ as it promotes more rational linguistic behavior, widening the gap in expressivity between some types.
+As suggested by Figure \ref{fig:only-R} the incumbent in most populations is not one of the six pragmatic $\Llack$-style types.
+That is, a pressure only for expressivity does not lend a justification to the systematic prevalence of a division of labor between semantics and pragmatics that we set out to explain, under any $\lambda$ value.
+For instance, in $1000$ independent populations with $\lambda = 20$ pragmatic $\Llack$ types only make up $11$ cases of incumbency, corresponding to a mean proportion of $0.003$ across populations.
+By contrast, $913$ incumbents were $\Lbound$-style users with close to an even share between literal ($454$) and pragmatic types ($459$), corresponding to a mean proportion of about $.48$ taken together.
+Across types and populations incumbents make up $.50$ of the mean outcome.
 
@@
 The effect of iterated learning without a pressure for expressivity with posterior sampling ($l = 1$) and a stronger tendency towards posterior maximization ($l = 15$) is shown in Figure \ref{fig:only-M} together with the prior over types.
-The latter shows that while users of $\Llack$ are not the most favored by the inductive bias (compared, e.g., to $\Lall$) they are nevertheless advantaged in virtue of the relatively simple semantics they conventionalize (compared, e.g., to $\Lbound$).
-Additionally, $\Llack$ enables its users to convey each state with a single message when combined with pragmatic reasoning provided sufficiently high $\lambda$.
-This makes it less likely to be confused with other types if the data is not too sparse ($k \geq 5$).
-Put differently, learners have a higher propensity to infer pragmatic $\Llack$ over types that might be hard or impossible to distinguish it from based on overt linguistic behavior, such as $\Lbound$, and $\Llack$ is also less likely to be mistaken with types with different linguistic behaviors because its pragmatic use approximates a one-to-one form-meaning mapping.
+The prior shows that while users of $\Llack$ are not the most favored by the inductive bias (compared, e.g., to $\Lall$) they are nevertheless more advantaged than others such as $\Lbound$ in virtue of the relatively simple semantics they conventionalize.
+Crucially, $\Llack$ enables its users to convey each state with a single message when combined with pragmatic reasoning provided sufficiently high $\lambda$.
+This makes it less likely to be confused with other types if the learning data is not too sparse ($k \geq 5$).
+In sum, learners have a higher propensity to infer pragmatic $\Llack$ over types that might be hard or impossible to distinguish it from based on overt linguistic behavior, such as $\Lbound$, and $\Llack$ is also less likely to be mistaken with types with different linguistic behavior because its pragmatic use approximates a one-to-one form-meaning mapping.
 As a consequence, a stronger propensity to maximize the posterior increases their proportion in the population.
-However, while we showed replication to generally succeed in driving out competing types of a kind, e.g., different pragmatic $\Llack$-style types, iterated learning leads to their coexistence as each of these types is passed on to next generations equal faithfulness and they do not stand in competition with each other.
-As already suggested by Figure \ref{fig:only-M}, in $1000$ analyzed independent populations all incumbents were pragmatic $\Llack$ users provided sufficient rationality and a tendency towards posterior maximization ($l = 15$, $\lambda = 20$).
-By contrast, low values of $\lambda$ and $l$ reflect the learners' inductive bias more directly, favoring functionally deficient but a priori preferred types such as literal $\Lall$.
+
+In contrast to a pressure only for expressivity with higher $\lambda$, learnability does not succeed in selecting for a single type of a kind.
+For example, only one of the six pragmatic $\Lllack$ types.
+Instead, it leads to their coexistence.
+Each is passed on to next generations with the same faithfulness and, differently from a pressure for communicative success within a population, they do not stand in competition with each other.
+Nevertheless, as implicitly suggested by Figure \ref{fig:only-M}, in $1000$ independent populations all incumbents were pragmatic $\Llack$ users provided sufficient rationality in linguistic behavior.
+Low values of $\lambda$ instead level the differences in linguistic behavior across types and therefore reflect the learners' inductive bias more faithfully, favoring functionally deficient but a priori preferred types such as literal $\Lall$.
+A pressure for learnability alone may consequently lead to the selection of communicatively suboptimal types that are easier to learn.
 In the extreme, when $l = 1$ and $\lambda = 1$ all of $1000$ independent populations had $\Lall$ as incumbent.

### 2017-03-18 code-LOT-extension/plots.py
rev

@@ -170,37 +170,66 @@ def get_subfigs_mutation(type_list,list1,list2):
     from lexica import get_prior,get_lexica
     priors = get_prior(get_lexica(3,3,False))
     priors = list(priors)
+
     targets = [231,236,291,306,326,336]
-    #reorder prior
-    for t in targets:
-        priors.insert(0,priors.pop(t))
+    targets = targets + [x-216 for x in targets] #adding literal Llack
+
+    lbound = [225, 235, 255, 270, 325, 330]
+    lbound = lbound + [x-216 for x in lbound]
+
+    lall = [0,216]
+
+
 
     white_space = 10 #to separate from other priors
+    Y_target = [0 for _ in xrange(len(priors)+white_space*3)]
+    Y_lbound = [0 for _ in xrange(len(priors)+white_space*3)]
+    Y_all = [0 for _ in xrange(len(priors)+white_space*3)]
 
-    X = np.arange(len(priors)+white_space) 
-    Y_target = [0 for _ in xrange(len(priors)+white_space)]
-    for target in xrange(len(targets)):
-        Y_target[target] = priors[target]
-        priors[target] = 0
+    for t in xrange(len(targets)):
+        Y_target[t] = priors[targets[t]]
+
+    for t in xrange(len(lbound)):
+        Y_lbound[len(targets)+white_space+t] = priors[lbound[t]]
+    for t in xrange(len(lall)):
+        Y_all[len(targets)+white_space+len(lbound)+white_space+t] = priors[lall[t]]
+
+    all_types = targets+lbound+lall
+    all_types.sort()
+
+    #reorder prior
+    for t in all_types:
+        priors.insert(0,priors.pop(t))
+    for t in xrange(len(all_types)):
+        priors[t] = 0
+
    
-    Y_rest = priors[:len(targets)] + [0 for _ in xrange(white_space)] + priors[len(targets):]
+    Y_rest = priors[:len(targets)] + [0 for _ in xrange(white_space)] + priors[len(targets):len(targets)+len(lbound)] +\
+     [0 for _ in xrange(white_space)] + priors[len(targets)+len(lbound):len(all_types)] + [0 for _ in xrange(white_space)] + priors[len(all_types):]
+
+            
     Y_target.reverse()
[…]

### 2017-03-19 code-LOT-extension/plots.py
plots rev

@@ -99,32 +99,52 @@ def get_subfigs_replication(type_list,list1,list2):
     ax1_large.set_xlabel(r'$\lambda$ = '+str(list1[0]),fontsize=25)
     ax1_large.xaxis.set_label_position('top')
 
-    handles, labels = ax1_large.get_legend_handles_labels()
-    labels = ['prag. L-lack kind' for _ in xrange(len(types))] + ['Incumbent']
-    display = (0,6)
+    #Start counting simulations with 1 and not 0 on the x-axis and hide other ticks
+    xlabels = [item.get_text() for item in ax1_large.get_xticklabels()]
+    xlabels = [str(int(x) + 1) for x in xlabels]
+    ax1_large.set_xticklabels(xlabels)
 
-    ax1_large.legend([handle for i,handle in enumerate(handles) if i in display], \
-                  [label for i,label in enumerate(labels) if i in display], loc='best')
+
+#    handles, labels = ax1_large.get_legend_handles_labels()
+#    labels = ['prag. L-lack kind' for _ in xrange(len(types))] + ['Incumbent']
+#    display = (0,6)
+#
+#    ax1_large.legend([handle for i,handle in enumerate(handles) if i in display], \
+#                  [label for i,label in enumerate(labels) if i in display], loc='best')
 
     p1 = patches.Rectangle((0, 0), 1, 1, fc="green", alpha=al)
     p2 = patches.Rectangle((0, 0), 1, 1, fc="darkred",alpha=al)
-    ax1_large.legend([p1, p2], ['prag. L-lack','Other incumbent'],loc='best',prop={'size':20})
+    ax1_large.legend([p1, p2], ['prag. L-lack','Other incumbent'],loc='best',prop={'size':14})
 
     ax2_large.yaxis.set_major_formatter(FormatStrFormatter('%.2f'))
     ax2_large.tick_params(axis='both',which='major',labelsize=17)
     ax2_large.set_xlabel(r'$\lambda$ = '+str(list1[1]),fontsize=25)
     ax2_large.xaxis.set_label_position('top')
 
+    xlabels = [item.get_text() for item in ax2_large.get_xticklabels()]
+    xlabels = [str(int(x) + 1) for x in xlabels]
+    ax2_large.set_xticklabels(xlabels)
+
+
[…]

### 2017-03-20 paper/paper-v3.tex
rev
plots rev
revision draft

@@
 We begin setting the stage by defining the learning prior and the set of types under consideration.
-Drawing from our preceding discussion, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $\Llack$-like semantics and the majority view of a division of labor between semantics and pragmatics.
-Before coming to this, it is instructive to first inspect the effect of these pressures in isolation.
-We therefore first showcase the effects that expressivity and learnability have on their own before turning to those resulting from their combination.
+Drawing from our preceding discussion, functional pressure on successful communication combined with learning pressures in the form of a bias that favors simpler semantics may lead to the selection of $\Llack$-like semantics and to the majority view of a division of labor between semantics and pragmatics.
+We first showcase the effects that expressivity and learnability have in isolation before turning to those resulting from their combination.
 
-\subsubsection{An inductive learning bias in favor of semantic simplicity}
+\subsubsection{An inductive learning bias for semantic simplicity}
 There is a growing effort to develop empirically  testable representational  languages that allow for the measure of semantic complexity.
@@
 
-where $Compl(c)$ is the complexity of the minimal derivation length of each concept.
-There are other more sophisticated ways to define priors over lexica (see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a}) but the key assumption here, common to all of them, is that simple representational expressions should be favored and complex ones disfavored.
-More generally, we should stress that these details -- from the generative grammar to its complexity measure -- are stipulated and are to be regarded as an illustratory instantiation of our general assumptions instead of as strong empirical claims.
+where $Compl(c)$ is the complexity of the minimal derivation length of $c$ according to the grammar.
+There are many ways to define priors over lexica (see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a}) but the key assumption here, common to all of them, is that simple representational expressions should be favored over more complex ones.
+We should stress that these details -- from the generative grammar to its complexity measure -- are to be regarded as an instantiation of our general assumptions instead of as empirical claims in their own right.
 
@@
 First, lexica that assign the same concept to more than one message.
-Such lexica lack in expressivity but are favored by the learning bias if they lexicalize simple concepts.
+Such lexica lack in expressivity but are favored by the learning bias should they lexicalize simple concepts.
 Second, lexica that conventionalize upper-bounds to realize a (quasi-)partition of the state space.
@@
 Instead, this information is already provided by the semantics of their lexicon, giving them an advantage in terms of expressivity.
-Types with such lexica are expected to be the main contenders to the natural language like semantics that do not lexicalize an upper-bound for weak scalar expressions.
+Types with such lexica are expected to be the main contenders of the natural language like semantics that do not lexicalize an upper-bound for weak scalar expressions.
 Third and lastly, $\Llack$-style lexica that, paired with mutual reasoning, can convey an upper-bound pragmatically.
 There are six lexica of the second kind and six of the third.
-The following three lexica exemplify each kind, with $\Lall$ being a lexicon of the first.
+The following three lexica exemplify each kind, with $\Lall$ being a lexicon of the first kind.
 This lexicon is not optimal for communication as it assigns all its messages to the same state, $\sall$.
-This inevitably leads to a communicative disadvantage in its use but it may nevertheless have a learnability advantage over our target $\Llack$ and its competitor $\Lbound$ in virtue of being the simplest lexicon complexity-wise, making it a priori favored by learners.
+This inevitably leads to a communicative disadvantage in its use.
+However, it may have a learnability advantage over our target $\Llack$ and its competitor $\Lbound$ in virtue of being the simplest lexicon, making it a priori favored by learners.
 
@@
 Recall that types are a combination of a lexicon and a manner of language use.
-We analyze the model's predictions in populations of types with one of the two signaling behaviors introduced earlier; literal or pragmatic.
+We analyze the model's predictions in populations of types with one of the two behaviors introduced earlier; literal or pragmatic.
 The former correspond to level $0$ reasoners and the latter to ones of level $1$.
-Higher level reasoning is not required to derive scalar implicatures from the lexica we consider here, nor do they leave room for substantial pragmatic refinement.
-Accordingly, there are $432$ types in this model.
-Six are variants of our pragmatic $\Llack$ target and twelve are either literal or pragmatic speakers of variants of its main contender $\Lbound$.
+Higher level reasoning is not required to derive scalar implicatures from these lexica, nor do they leave room for substantial pragmatic refinement.
+Accordingly, we consider a total of $432$ types.
+Six are variants of our pragmatic $\Llack$ target and twelve are either literal or pragmatic speakers of variants of its main contender, $\Lbound$.
 
@@
 \subsubsection{Expressivity only} 
-
 \begin{figure}
@@
 Communicative efficiency is sensitive to $\lambda$ as it influences signaling behavior.
-It therefore has a central bearing on types' fitness.
 This is showcased in Figure \ref{fig:only-R}, which depicts the proportion of pragmatic $\Llack$-style speakers in $5$ independent populations after $50$ generations, as well as that of the type with the highest proportion in a given population, the incumbent.
@@
 This difference increases with $\lambda$ as it promotes more rational linguistic behavior, widening the gap in expressivity between some types.
-As suggested by Figure \ref{fig:only-R} the incumbent in most populations is not one of the six pragmatic $\Llack$-style types.
-That is, a pressure only for expressivity does not lend a justification to the systematic prevalence of a division of labor between semantics and pragmatics that we set out to explain, under any $\lambda$ value.
+As suggested by Figure \ref{fig:only-R}, the incumbent in most populations is not one of the six pragmatic $\Llack$-style types.
+That is, a pressure only for expressivity does not lend a justification to the systematic prevalence of a division of labor between semantics and pragmatics under any $\lambda$-value.
 For instance, in $1000$ independent populations with $\lambda = 20$ pragmatic $\Llack$ types only make up $11$ cases of incumbency, corresponding to a mean proportion of $0.003$ across populations.
 By contrast, $913$ incumbents were $\Lbound$-style users with close to an even share between literal ($454$) and pragmatic types ($459$), corresponding to a mean proportion of about $.48$ taken together.
-Across types and populations incumbents make up $.50$ of the mean outcome.
+%Across types and populations incumbents make up $.50$ of the mean outcome.
 
@@
 
-The effect of iterated learning without a pressure for expressivity with posterior sampling ($l = 1$) and a stronger tendency towards posterior maximization ($l = 15$) is shown in Figure \ref{fig:only-M} together with the prior over types.
+The effect of iterated learning without a pressure for expressivity using either posterior sampling ($l = 1$) or a stronger tendency towards posterior maximization ($l = 15$) is shown in Figure \ref{fig:only-M} together with the prior over types.
 The prior shows that while users of $\Llack$ are not the most favored by the inductive bias (compared, e.g., to $\Lall$) they are nevertheless more advantaged than others such as $\Lbound$ in virtue of the relatively simple semantics they conventionalize.
@@
 This makes it less likely to be confused with other types if the learning data is not too sparse ($k \geq 5$).
-In sum, learners have a higher propensity to infer pragmatic $\Llack$ over types that might be hard or impossible to distinguish it from based on overt linguistic behavior, such as $\Lbound$, and $\Llack$ is also less likely to be mistaken with types with different linguistic behavior because its pragmatic use approximates a one-to-one form-meaning mapping.
+Put differently, learners have a higher propensity to infer pragmatic $\Llack$ over types that might be hard or impossible to distinguish it from based on overt linguistic behavior, such as $\Lbound$, and $\Llack$ is also less likely to be mistaken with types with different observable behavior because its pragmatic use approximates a one-to-one form-meaning mapping.
 As a consequence, a stronger propensity to maximize the posterior increases their proportion in the population.
 
-In contrast to a pressure only for expressivity with higher $\lambda$, learnability does not succeed in selecting for a single type of a kind.
-For example, only one of the six pragmatic $\Lllack$ types.
+However, in contrast to a pressure only for expressivity with high $\lambda$, learnability does not succeed in selecting for a single type of a kind.
+For instance, for only one of the six pragmatic $\Llack$ types.
 Instead, it leads to their coexistence.
-Each is passed on to next generations with the same faithfulness and, differently from a pressure for communicative success within a population, they do not stand in competition with each other.
-Nevertheless, as implicitly suggested by Figure \ref{fig:only-M}, in $1000$ independent populations all incumbents were pragmatic $\Llack$ users provided sufficient rationality in linguistic behavior.
-Low values of $\lambda$ instead level the differences in linguistic behavior across types and therefore reflect the learners' inductive bias more faithfully, favoring functionally deficient but a priori preferred types such as literal $\Lall$.
+Each is passed on to next generations with the same faithfulness and, differently from a pressure for communicative success, they do not stand in competition with each other.
+
+As implicitly suggested by Figure \ref{fig:only-M}, in $1000$ independent populations all incumbents were pragmatic $\Llack$ users provided sufficient rationality in linguistic behavior, with each reaching approximately the same proportion of users in the population.
+Low values of $\lambda$ level the differences in observable behavior across types and therefore reflect the learners' inductive bias more faithfully, favoring functionally deficient but a priori preferred types such as literal $\Lall$.
 A pressure for learnability alone may consequently lead to the selection of communicatively suboptimal types that are easier to learn.
-In the extreme, when $l = 1$ and $\lambda = 1$ all of $1000$ independent populations had $\Lall$ as incumbent.
+In the extreme, when $l = 1$ and $\lambda = 1$ all of $1000$ independent populations had users of $\Lall$ as incumbents.
 
-\subsubsection{Expressivity and learnability.}
-Inspecting these pressures separately not only showcases the contribution of the model's components but also highlights some of their broader implications.
-First and foremost, neither dynamic on its own comes close to converging to a monomorphic population of pragmatic $\Llack$ speakers.
-On the one hand, the slight communicative advantage of $\Lbound$'s lexicalization of an upper-bound over pragmatic use of $\Llack$ leads to the former's selection under a pressure for expressivity.
-On the other hand, pressure only towards learnability has a modest but clear effect of promoting pragmatic $\Llack$ over functionally similar but semantically more complex alternatives such as $\Lbound$.
+\subsubsection{Expressivity and learnability}
+The separate inspection of each pressure not only showcases the contribution of the model's components but also highlights some of their broader implications for a division of labor between semantics and pragmatics.
+Neither dynamic on its own comes close to converging to a monomorphic population of pragmatic $\Llack$ speakers.
+When pressured toward expressivity, the slight communicative advantage of $\Lbound$'s lexicalization of an upper-bound over pragmatic use of $\Llack$ leads to the former's selection.
+When pressured toward learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more complex alternatives such as $\Lbound$.
 However, on its own learnability does not foment the propagation of these types across the population.
-Furthermore, this pressure alone leads to heterogeneous populations that do not justify the selection of a particular type over others.
+What is more, it leads to heterogeneous populations that foster an equal share of types of a kind, such as different pragmatic $\Llack$ types.
 
@@
 Figure \ref{fig:rmd} illustrates the combined effects of both pressures for a sample of $\lambda$ and $l$ values.
-These results show that an inductive learning bias for simpler semantics in tandem with functional pressure can lead to the selection of $\Llack$-like semantics over $\Lbound$.
-As when each pressure was considered in isolation, the proportion of these types increases with $\lambda$ and $l$.
-Crucially, the addition of a pressure towards expressivity magnifies the learning outcome and dampens the proliferation of similar or identical types.
-As stressed above, this suggests that neither a learning bias nor functional pressure alone but their combination may lead to the lack of upper-bounds in the lexical meaning of scalar expressions in a mostly monomorphic population.
+These results show that an inductive learning bias for simpler semantics in tandem with functional pressure can lead to the selection of $\Llack$-like semantics paired with pragmatic language use.
+The proportion of these types increases with $\lambda$ and $l$.
+Pressure towards expressivity magnifies the effects of iterated learning and dampens the proliferation of types of a kind that are equal in expressivity {\em and} learnability.
+A pressure towards learnability favors the transmission of simpler semantics and thereby makes types that exploit pragmatic reasoning to strengthen meaning more likely to be adopted over alternatives that do not.
+In short, neither learnability nor functional pressure alone but their combination may lead to the lack of upper-bounds in the lexical meaning of scalar expressions in predominantly monomorphic populations.
 
-As before, low $\lambda$ and $l$ lead to the incumbency of communicatively suboptimal types that are a priori favored by learning, such as $\Lall$.
-An increase in $\lambda$ leads to the selection and incumbency of pragmatic $\Llack$ types but does not lead to monomorphic populations if learners sample from the posterior.
+As before, low $\lambda$ and $l$ lead to the incumbency of communicatively suboptimal types that are a priori favored, such as $\Lall$.
+An increase in $\lambda$ leads to the selection and incumbency of pragmatic $\Llack$ but does not lead to monomorphic populations if learners sample from the posterior.
 Finally, a combination of sufficiently rational signaling behavior with some tendency for posterior maximization leads to increasingly monomorphic populations of types that convey upper-bounds pragmatically.
-A more succinct representation of the difference between the mean of the highest pragmatic $\Llack$-like types and the highest proportion of other types across $\lambda$ and $l$ values is given in Figure \ref{fig:diff}.
-\tb{Say more}
+The difference between the mean of the highest pragmatic $\Llack$-like type in $1000$ independent populations and the highest proportion of other types across $\lambda$ and $l$ values is shown in Figure \ref{fig:diff}.
 
@@
 \includegraphics[width=1\textwidth,height=6cm,keepaspectratio]{./plots/fig4-incumbents-difference}
-\caption{Difference between mean proportion of highest pragmatic $\Llack$ type and highest other type in $1000$ independent populations after $50$ generations under both pressures ($k = 5$).}
+\caption{Difference between mean proportion of highest pragmatic $\Llack$ type and highest other type in $1000$ independent populations after $50$ generations under both pressures ($k = 5$).
+\tb{I will run more simulations with higher $\lambda$ and $l$.
+I'm pretty sure the difference still increases.}}
 \label{fig:diff}
@@
 To recapitulate, other than the involvement of both pressures, the resulting proportion of pragmatic $\Llack$ speakers primarily hinges on three factors.
-First, the degree to which linguistic behavior is deterministic.
+First, the degree to which linguistic behavior is rational in that agents try to maximize communicative success.
 This plays a role both in expressivity as well as in producing data that allows learners to discriminate this type from others.
@@
 
-A lack of upper-bounded in the lexical meaning of weak scalar expressions constitutes the majority view in the literature.
+A lack of upper-bounds in the lexical meaning of weak scalar expressions constitutes the majority view in the literature.
 However, it not clear to what extent other types should be present in the final population.
-It seems reasonable to expect functionally suboptimal lexica such as $\Lall$ to be ruled out because they fail to enable their users to communicate effectively and are not evinced in natural language.
-However, this is not true of $\Lbound$.
-The prediction that natural language communities are homogeneous or that a single speaker may entertain $\Llack$-like semantics for one scalar expression and $\Lbound$-like semantics for another is not implausible (cf.
+It seems reasonable to expect functionally suboptimal lexica such as $\Lall$ to be ruled out but matters are different for $\Lbound$.
+The prediction that natural language communities are homogeneous or that speakers may entertain $\Llack$-like semantics for one scalar expression and $\Lbound$-like semantics for another is not implausible (cf.
 \citealt{franke+degen:2016}).
@@
 Should cost play a role, then its effect is bound to depend on the frequency with which a given scalar expression is used.
-It may be that frequently drawn scalar implicatures fossilize to avoid cost whereas infrequent ones are still derived on-line.
+It may be that frequently drawn scalar implicatures fossilize to avoid cost whereas infrequent ones are derived on-line to avoid complexity in acquisition.
 This also opens a possible venue to address the preceding question about the expected presence of $\Lbound$-like semantics in a population, but further empirical evidence is needed to assess these matters.
@@
 \section{General discussion}\label{sec:discussion}
-We laid out a model that combines game theoretical models of functional pressure towards efficient communication \citep{nowak+krakauer:1999}, effects of transmission perturbations on (iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012, franke+jaeger:2014} as well as different lexica \citep{bergen+etal:2012,bergen+etal:2016}, and \tb{mention LOT here}.
-This model generates predictions about lexicalization patterns and, more generally, effects of communicative pressures on the cultural evolution of language.
+We laid out a model that combines game theoretical models of functional pressure towards efficient communication \citep{nowak+krakauer:1999}, effects of transmission perturbations on (iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012, franke+jaeger:2014} as well as different lexica \citep{bergen+etal:2012,bergen+etal:2016}.
+This model generates predictions about the co-evolution of conventional meaning and pragmatic reasoning types and, more generally, effects of communicative pressures on the cultural evolution of language.
 
-We argued that the puzzle raised by semantics in light of pragmatics is hard to explain on purely functional grounds and that part of the answer may instead lie in the way transmission shapes the outcome of cultural evolution in tandem with a pressure for successful information transfer.
+We argued that the puzzle raised by semantics in light of pragmatics is hard to explain on purely functional grounds and that part of the answer may instead lie in the way successful transmission of linguistic knowledge across and within populations shapes the outcome of their co-evolution.
 In the realm of inductive biases, we adopted the assumption that simpler semantic representations are more likely to be learned (cf.
-\citealt{chater+vitanyi:2003}), \tb{mention LOT here}.
+\citealt{chater+vitanyi:2003}).
 Under this view, semantics and pragmatics play a synergic role in that representational simplicity is supplemented by pragmatic reasoning to counteract functional disadvantages otherwise incurred.
 As a consequence, iterated transmission and use of language lead to a regularization that may explain the lack of lexicalization of systematic pragmatic enrichments.
-This result is of particular relevance for the longstanding assumption of a divide and interaction between semantics and pragmatics by offering an account of why certain pragmatic inferences fail to lexicalize.
+This result is of particular relevance to the longstanding assumption of a divide and interaction between semantics and pragmatics by offering an account of why certain pragmatic inferences fail to lexicalize.
 
-The main innovations of the model are its modular separation of expressivity and learnability, allowing for their isolated and combined analysis, the learning process involving a joint inference over types of pragmatic behavior and lexical meaning.
+The main innovations of the model are its modular separation of expressivity and learnability, allowing for their isolated and combined analysis, as well as the learning process involving a joint inference over types of pragmatic behavior and lexical meaning.
 The goal to decouple but model both expressivity and learnability has also recently been addressed by \citet{kirby+etal:2015}.
@@
 model expressivity as exerting its force only in the production of learning data.
-On a general level, this model's expressivity parameter has a similar effect to $\lambda$ in that it can give a communicative advantage to unambiguous languages over ambiguous ones.
-\tb{Should we go into more detail here?
-I think it's fine if we leave it out but $\lambda$ has a broad motivation whereas Kirby et al.'s parameter is quite explicitly a penalty parameter against ambiguity.
-I mainly wanted to give an idea of what the pressure for expressivity in their model does rather than argue for why ours is more principled.} However, the degree of mutual understanding of interlocutors central to replication and to our notion of expressivity is not taken into consideration.
+The degree of mutual understanding of interlocutors central to replication and to our notion of expressivity is absent.
 That is, while our proposal combines bidirectional horizontal transmission with its vertical and unidirectional counterpart, Kirby et al.'s model only considers the latter's influence.
-Our reasoning behind the inclusion of the former lies in the empirical and theoretical observation that learnability alone can lead the selection of functionally defective languages, as discussed in relation to $\Lall$ when $\lambda$ or $k$ are low.
-This outcome has been reported in a number of laboratory experiments where subjects learned and subsequently reproduced the language produced by a previous participant over multiple iterations, leading to a proliferation of ambiguous languages that associate a large number of meanings with a single form (see, e.g., \citealt{silvey+etal:2014} and experiment 1 in \citealt{kirby+etal:2008}).
+Our reasoning behind the inclusion of the former lies in the empirical and theoretical observation that learnability alone can lead the selection of functionally defective languages, as showcased with $\Lall$.
+This outcome has been reported in a number of laboratory experiments where subjects learned and subsequently reproduced the language produced by a previous participant over multiple iterations, leading to a proliferation of ambiguous languages that associate many meanings with a single form (see, e.g., \citealt{silvey+etal:2014} and experiment 1 in \citealt{kirby+etal:2008}).
 In contrast, experiments involving an interactive component have been found to foster languages that enable interlocutors to communicatively distinguish meanings more accurately  (e.g., \citealt{fay+etal:2013}; for a review of laboratory results under the iterated learning paradigm and further discussion see \citealt{kirby+etal:2015, tamariz+kirby:2016}).
 It is not evident how to compare these empirical findings given that they consider distinct meaning spaces, modes of transmission, iterations and feedback given to participants.
-However, we take these results to suggest that there is an important difference between a language generating learnable linguistic data and its actual performance as a means of information transfer.
-\tb{Explain the next two sentences} The former solely depends on the mechanism by which speakers associate form and meaning.
-The latter additionally hinges on the addressee's linguistic experience and her ability to interpret linguistic input based on this experience.
-In sum, we contend that successful information transfer in a linguistic community is central to the adoption of a communication system and that this measure is not adequately reflected by production alone.
+However, on a general level they may suggest that there is an important difference between a language generating learnable linguistic data and its actual performance as a means of information transfer.
+That is, we contend that successful information transfer in a linguistic community is central to the adoption of a communication system and that this measure is not adequately reflected by production alone.
 
 \section{Conclusion}
-The cultural evolution of language is influenced by intertwined pressures.
-We set out to investigate this process by putting forward a model that combines a pressure toward efficient and successful information transfer with perturbations that may arise in the transmission of linguistic knowledge in acquisition.
-\tb{Clarify and put this in better words.
-This is quite central:} Additionally, we argued for the necessity of considering the role of pragmatics in investigations on the cultural evolution of meaning.
-These components and their mutual influence were highlighted in a case study on the lack of lexical upper-bounds in weak scalar expressions that showed that, when pressured for learnability and expressivity, the former drives for simpler semantic representations inasmuch as pragmatics can compensate for lack of expressivity in use.
+The cultural evolution of meaning is influenced by intertwined pressures.
+We set out to investigate this process by putting forward a model that combines a pressure toward successful information transfer with perturbations that may arise in the transmission of linguistic knowledge in acquisition.
+Its objects of selection and replication are pairs of lexical meanings and patterns of language use.
+This allows the model to trace the interaction between conventional meaning and pragmatic use.
+Additionally, it takes the challenge serious of neither semantics nor pragmatics being directly observable.
+Instead, learners need to infer these unobservables from overt data that results from their combination.
+These components and their mutual influence were highlighted in a case study on the lack of lexical upper-bounds in weak scalar expressions that showed that, when pressured for learnability and expressivity, the former force drives for simpler semantic representations inasmuch as pragmatics can compensate for lack of expressivity in use.
 That is, the relative learning advantage of simpler semantics in combination with functional pressure in use may offer an answer to why natural languages fail to lexicalize systematic pragmatic inferences.

### 2017-03-20 paper/paper-v3.tex
revision

@@
 that the speaker does not own all of them. This is because, if the speaker had them all, he
-could have used the stronger word \emph{all} instead of \emph{some} in his utterance, 
+could have used the word \emph{all} instead of \emph{some} in his utterance, 
 thereby making a more informative statement. Scalar implicatures, especially the
@@
 and its subsequent multiplication with a mutation matrix $Q$:
-
 \begin{align}
@@
 
-We distinguish between two kinds of pragmatic behavior. {\em Literal interlocutors} produce and
+We distinguish between two kinds of linguistic behavior. {\em Literal interlocutors} produce and
 interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
@@
 parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}. As $\lambda$
-increases, choices made in production are more rational in that higher values lead to behavior
-that is increasingly in line with expected utility maximization. Expected utility of a message
+increases, choices made in production are more rational in that they are increasingly in line with expected utility maximization. Expected utility of a message
 $\messg$ in state $\state$ for a level $n+1$ speaker is here defined as $H_{n}(s|m;L)$, the probability that the hearer will assign to or choose the correct meaning.
-Put differently, how likely the hearer is thought to infer $\state$ when sent $\messg$.
+Put differently, expected utility indicates how likely the hearer is thought to infer $\state$ when sent $\messg$.
 The formulation
@@
 
-Particularly important to our purpose is the fact that players of a level greater than $0$ using $\Llack$ can {\em pragmatically} associate $\msome$ with $\ssome$ over $\sall$,  in contrast to their literal counterparts.
-Intuitively, this is so because hearers reason a speaker to convey $\sall$ with $\mall$, as this has a higher chance of leading to mutual understanding than using underspecified $\msome$.
-Therefore they will expect $\msome$ to be more strongly associated with $\ssome$ than with $\sall$ as it is already possible to clearly convey the latter state with a different message.
+Particularly important to our purpose is the fact non-literal interlocutors using $\Llack$ can {\em pragmatically} associate $\msome$ with $\ssome$ over $\sall$,  in contrast to their literal counterparts.
+Intuitively, this is so because hearers reason a speaker to convey $\sall$ with $\mall$, as this message has a higher chance of leading to mutual understanding than using underspecified $\msome$.
+Therefore they will expect $\msome$ to be more strongly associated with $\ssome$ than with $\sall$ as it is already possible to clearly convey the latter state $\mall$.
 The converse reasoning pattern holds for the speaker.
 In this way, two distinct lexical conventions such as $\Lbound$ and $\Llack$ can give rise to similar observable linguistic behavior.
+In this case, associating $\msome$ with $\ssome$ and $\mall$ with $\sall$.
 
@@
 observation. Concretely, when agents of type $j$ want to adopt or imitate the linguistic
-behavior of type $i$, they observe the linguistic behavior of type $i$ and need to infer what
+behavior of type $i$, they observe the overt linguistic behavior of type $i$ and need to infer what
 their type is from that. Iterated learning is a process in which languages are learned repeatedly from the observation
@@
 This equation defines the frequency $x'_i$ of type $i$ at the next time step, based on its
-frequency $x_i$ before the step, its fitness $f_i$ and the probability that a learner infers
+frequency $x_i$ before the step, its fitness $f_i$, and the probability that a learner infers
 $i$ when observing the behavior of a type-$j$ agent. Fitness-based selection can be thought of
@@
 We begin setting the stage by defining the learning prior and the set of types under consideration.
-Drawing from our preceding discussion, functional pressure on successful communication combined with learning pressures in the form of a bias that favors simpler semantics may lead to the selection of $\Llack$-like semantics and to the majority view of a division of labor between semantics and pragmatics.
+Drawing from our preceding discussion, functional pressure on successful communication combined with learning pressures in the form of a bias that favors simpler semantics may lead to the selection of $\Llack$-like semantics and to a division of labor between semantics and pragmatics.
 We first showcase the effects that expressivity and learnability have in isolation before turning to those resulting from their combination.
@@
 
-where $Compl(c)$ is the complexity of the minimal derivation length of $c$ according to the grammar.
+where $Compl(c)$ is the complexity of the minimal derivation length of $c$ according to the LOT-grammar.
 There are many ways to define priors over lexica (see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a}) but the key assumption here, common to all of them, is that simple representational expressions should be favored over more complex ones.
@@
 \tb{Should we say something about the inclusion of $\snone$?
-It may be more natural to introduce this state earlier but may unnecessarily complicate the lexica showcased in the previous section.} This space can be
+It may be more natural to introduce this state earlier but may unnecessarily complicate the exposition of signaling behavior in the previous section.} This space can be
 thought of as a partition of possible worlds into cells where none, some or all of the $A$s are
@@
  or tautological. These are listed with mnemonic names in
-Table~\ref{tab:concepts} together with their complexity according to the grammar given in Table \ref{tab:grammar}.
+Table~\ref{tab:concepts} together with their complexity according to the grammar in Table \ref{tab:grammar}.
 
@@
 messages there are $6^3 = 216$ possible lexica.
-Some assign the same concept to more than one message and others lexicalize the same meanings but associate them with different messages.
-The main results reported in the following do not hinge on the inclusion of these lexica but such an unaltered space does better illustrate the influence and differences of the linguistic pressures we consider.
+Some assign the same concept to more than one message and others lexicalize the same concepts but associate them with different messages.
+The main results reported in the following do not hinge on the inclusion of these lexica but this unaltered space does better illustrate the influence and differences of the linguistic pressures we consider.
 Additionally, the competition of different languages that lexicalize the same concepts through different forms allows us to take serious the problem of only one being selected for even though multiple alternatives of equal expressivity {\em and} learnability exist.
@@
 First, lexica that assign the same concept to more than one message.
-Such lexica lack in expressivity but are favored by the learning bias should they lexicalize simple concepts.
+Such lexica lack in expressivity but may be favored by the learning prior should they lexicalize simple concepts.
 Second, lexica that conventionalize upper-bounds to realize a (quasi-)partition of the state space.

### 2017-03-21 paper/paper-v3.tex
revision new sim setup

@@
 
-Communicative efficiency is sensitive to $\lambda$ as it influences signaling behavior.
+Expressivity is sensitive to $\lambda$ as it influences signaling behavior.
 This is showcased in Figure \ref{fig:only-R}, which depicts the proportion of pragmatic $\Llack$-style speakers in $5$ independent populations after $50$ generations, as well as that of the type with the highest proportion in a given population, the incumbent.
-As is to be expected, less expressive types do not fare well under pressure for communicative success and those able to convey states less equivocally fare better.
-This difference increases with $\lambda$ as it promotes more rational linguistic behavior, widening the gap in expressivity between some types.
+Low $\lambda$ levels communicative differences between types, as even those that can differentiate between all states semantically or pragmatically do not always maximize expected communicative success.
+The outcome is that of a mixed population that allows for the survival of competing types.
+Conversely, higher $\lambda$ promotes more rational linguistic behavior, widening the gap in expressivity between types and promoting more homogeneous populations.
 As suggested by Figure \ref{fig:only-R}, the incumbent in most populations is not one of the six pragmatic $\Llack$-style types.
-That is, a pressure only for expressivity does not lend a justification to the systematic prevalence of a division of labor between semantics and pragmatics under any $\lambda$-value.
-For instance, in $1000$ independent populations with $\lambda = 20$ pragmatic $\Llack$ types only make up $11$ cases of incumbency, corresponding to a mean proportion of $0.003$ across populations.
+That is, a pressure only for expressivity does not lend a justification to the systematic prevalence of a lack of upper-bounds under any $\lambda$-value.
+For instance, in $1000$ independent populations with $\lambda = 20$ pragmatic $\Llack$ types only made up $11$ cases of incumbency, corresponding to a mean proportion of $0.003$ across populations.
 By contrast, $913$ incumbents were $\Lbound$-style users with close to an even share between literal ($454$) and pragmatic types ($459$), corresponding to a mean proportion of about $.48$ taken together.
@@
 The effect of iterated learning without a pressure for expressivity using either posterior sampling ($l = 1$) or a stronger tendency towards posterior maximization ($l = 15$) is shown in Figure \ref{fig:only-M} together with the prior over types.
-The prior shows that while users of $\Llack$ are not the most favored by the inductive bias (compared, e.g., to $\Lall$) they are nevertheless more advantaged than others such as $\Lbound$ in virtue of the relatively simple semantics they conventionalize.
+The prior shows that while users of $\Llack$ are not the most favored by the inductive bias (compared, e.g., to $\Lall$) they are nevertheless more advantaged than others, such as $\Lbound$, in virtue of the relatively simple semantics they conventionalize.
 Crucially, $\Llack$ enables its users to convey each state with a single message when combined with pragmatic reasoning provided sufficiently high $\lambda$.
@@
 
-However, in contrast to a pressure only for expressivity with high $\lambda$, learnability does not succeed in selecting for a single type of a kind.
+However, in contrast to a pressure only for expressivity with high $\lambda$, learnability alone does not succeed in selecting for a single type of a kind.
 For instance, for only one of the six pragmatic $\Llack$ types.
 Instead, it leads to their coexistence.
-Each is passed on to next generations with the same faithfulness and, differently from a pressure for communicative success, they do not stand in competition with each other.
-
-As implicitly suggested by Figure \ref{fig:only-M}, in $1000$ independent populations all incumbents were pragmatic $\Llack$ users provided sufficient rationality in linguistic behavior, with each reaching approximately the same proportion of users in the population.
-Low values of $\lambda$ level the differences in observable behavior across types and therefore reflect the learners' inductive bias more faithfully, favoring functionally deficient but a priori preferred types such as literal $\Lall$.
-A pressure for learnability alone may consequently lead to the selection of communicatively suboptimal types that are easier to learn.
+Each is passed on to next generation with the same faithfulness and, differently from a pressure for communicative success, they do not stand in competition with each other.
+As implicit in Figure \ref{fig:only-M}, in $1000$ independent populations all incumbents were pragmatic $\Llack$ users provided sufficiently high $\lambda$, with each reaching approximately the same proportion of users in the population.
+As with expressivity only, low values of $\lambda$ make the differences in observable behavior across types less pronounced and therefore reflect the learners' inductive bias more faithfully, favoring functionally deficient but a priori preferred types such as those that use $\Lall$.
+A pressure for learnability alone may consequently lead to a spread of communicatively suboptimal types that are easier to learn.
 In the extreme, when $l = 1$ and $\lambda = 1$ all of $1000$ independent populations had users of $\Lall$ as incumbents.
@@
 \subsubsection{Expressivity and learnability}
-The separate inspection of each pressure not only showcases the contribution of the model's components but also highlights some of their broader implications for a division of labor between semantics and pragmatics.
+The separate inspection of each pressure not only showcases the contribution of the model's components but also highlights some of their broader implications for the emergence of a division of labor between semantics and pragmatics.
 Neither dynamic on its own comes close to converging to a monomorphic population of pragmatic $\Llack$ speakers.
@@
 Pressure towards expressivity magnifies the effects of iterated learning and dampens the proliferation of types of a kind that are equal in expressivity {\em and} learnability.
-A pressure towards learnability favors the transmission of simpler semantics and thereby makes types that exploit pragmatic reasoning to strengthen meaning more likely to be adopted over alternatives that do not.
-In short, neither learnability nor functional pressure alone but their combination may lead to the lack of upper-bounds in the lexical meaning of scalar expressions in predominantly monomorphic populations.
+A pressure towards learnability favors the transmission of simpler semantics and thereby makes types that exploit pragmatic reasoning to strengthen lexical meaning more likely to be adopted over alternatives that do not.
+%In short, neither learnability nor functional pressure alone but their combination may lead to the lack of upper-bounds in the lexical meaning of scalar expressions in predominantly monomorphic populations.
 
@@
 To recapitulate, other than the involvement of both pressures, the resulting proportion of pragmatic $\Llack$ speakers primarily hinges on three factors.
-First, the degree to which linguistic behavior is rational in that agents try to maximize communicative success.
+First, the degree to which agents try to maximize communicative success.
 This plays a role both in expressivity as well as in producing data that allows learners to discriminate this type from others.
@@
 
-The main innovations of the model are its modular separation of expressivity and learnability, allowing for their isolated and combined analysis, as well as the learning process involving a joint inference over types of pragmatic behavior and lexical meaning.
+The main innovations of the model are its modular separation of expressivity and learnability, allowing for their isolated and combined analysis, learning involving a joint inference over types of pragmatic behavior and lexical meaning, and, accordingly, the possibility to trace the co-evolution of conventional semantics and pragmatic use.
 The goal to decouple but model both expressivity and learnability has also recently been addressed by \citet{kirby+etal:2015}.
@@
 The degree of mutual understanding of interlocutors central to replication and to our notion of expressivity is absent.
-That is, while our proposal combines bidirectional horizontal transmission with its vertical and unidirectional counterpart, Kirby et al.'s model only considers the latter's influence.
-Our reasoning behind the inclusion of the former lies in the empirical and theoretical observation that learnability alone can lead the selection of functionally defective languages, as showcased with $\Lall$.
+That is, while our proposal combines bidirectional horizontal transmission with its vertical and unidirectional counterpart, Kirby et al.
+only consider the latter's influence.
+Our reasoning behind the inclusion of the former lies in the empirical and theoretical observation that learnability alone can lead the selection of functionally defective languages, as showcased by $\Lall$.
 This outcome has been reported in a number of laboratory experiments where subjects learned and subsequently reproduced the language produced by a previous participant over multiple iterations, leading to a proliferation of ambiguous languages that associate many meanings with a single form (see, e.g., \citealt{silvey+etal:2014} and experiment 1 in \citealt{kirby+etal:2008}).
@@
 That is, the relative learning advantage of simpler semantics in combination with functional pressure in use may offer an answer to why natural languages fail to lexicalize systematic pragmatic inferences.
+And, more broadly, to a division of labors between semantics and pragmatics.
 

### 2017-03-22 paper/paper-v3.tex
intro polished and RMD example added (coordination game)

@@
-\documentclass[a4paper]{article}
+\documentclass[a4paper, 11pt]{article}
 
@@
 \usepackage{booktabs}
+\usepackage{pgfplots}
+\usepackage{subcaption}
+\pgfplotsset{compat=1.3}
+\usetikzlibrary{pgfplots.groupplots,decorations.markings}
+
 %\newcommand{\mvalueof}[1]{\llbracket#1\rrbracket}
@@
 \usepackage{xspace}
+
+\usepackage{tgtermes}
+\renewcommand{\baselinestretch}{1.2}
 
@@
 \begin{document}
+
+  %% for arrow head placement
+  \tikzset{->-/.style={decoration={
+  markings,
+  mark=at position #1 with {\arrow{>}}},postaction={decorate}}}
+  %%%
+
 \maketitle
@@
   According to standard linguistic theory, the meaning of an utterance is the product of
-  conventional semantic meaning and general pragmatic reasoning applied
-  to the context of utterance. This implies that models of the cultural evolution of meaning
-  should likewise take into consideration that observable language use is a complex interaction
-  of semantic representations and pragmatic use. To this end, we present a game theoretic model
-  of cultural language evolution where communicative pressures work on abstract semantic
-  representations and pragmatic patterns of use. Our model traces two evolutionary forces and
-  their interaction: (i) fitness-based pressure towards communicative efficiency and (ii)
-  systematic transmission perturbations when linguistic knowledge is transferred from one agent
-  to another.  We illustrate the model based on a case study showing that learning biases that favor simple semantic
-  representations can prevent the lexicalization of pragmatic inferences as long as mutual reasoning counteracts a loss in expressivity otherwise incurred by the adoption of such semantics. 
+  conventional semantic meaning and general pragmatic rules on language use. To investigate how
+  cultural evolution of language plays out under this picture of the semantics-pragmatics
+  interface, we present a game theoretic model of the competition between types of language
+  users, each endowed with her selection of lexical concepts and her own pragmatic dispositions
+  to act on them. Our model traces two evolutionary forces and their interaction: (i)
+  fitness-based pressure towards communicative efficiency and (ii) potential learning biases
+  during the transfer of linguistic knowledge. We illustrate the model based on a case study on
+  scalar implicatures. In this case study learning biases that favor simple semantic
+  representations can boost the evolution of more sophisticated pragmatic reasoning types and
+  so prevent the lexicalization of scalar implicatures.
 \end{abstract}
@@
 What is conveyed usually goes beyond what is said. A request for a blanket can be politely
-veiled by uttering ``I'm cold.'' The temporal succession of events can be communicated by the
-order in which conjuncts appear as in ``I traveled to Paris and got married.'' An invitation
-can be declined by saying ``I have to work.'' An influential explanation of the relation
-between the literal meaning of expressions and what they are intended and interpreted to convey
-is due to \citet{grice:1975}, who characterizes pragmatic use and interpretation as a process of
-mutual reasoning about rational language use. For instance, under the assumption that the
-speaker is cooperative and relevant, ``I have to work'' may be interpreted as providing a
-reason why the speaker will not be able to accept an invitation, going beyond its literal
-meaning. Some of these enrichments are rather \emph{ad hoc}. Others show striking regularities,
-such as the use of ability questions for polite requests (``Could you please \dots?''), or
-certain enrichments of lexical meanings such as \emph{and} to mean \emph{and then}.
+veiled by uttering ``I'm cold;'' temporal succession of events can be communicated by the order
+in which conjuncts appear as in ``I traveled to Paris and got married;'' an invitation can be
+declined by saying ``I have to work.'' An influential explanation of the relation between the
+literal meaning of expressions and what they may convey in context is due to
+\citet{grice:1975}, who characterizes pragmatic use and interpretation as a process of mutual
+reasoning about rational language use. For instance, under the assumption that the speaker is
+cooperative and relevant, ``I have to work'' may be interpreted as providing a reason why the
+speaker will not be able to accept an invitation, going beyond its literal meaning. Some of
+these enrichments are rather \emph{ad hoc}. Others show striking regularities, such as the use
+of ability questions for polite requests (``Could you please \dots?''), or certain enrichments
+of lexical meanings such as \emph{and} to convey \emph{and then}.
 
@@
 the utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
-that the speaker does not own all of them. This is because, if the speaker had them all, he
-could have used the word \emph{all} instead of \emph{some} in his utterance, 
-thereby making a more informative statement. Scalar implicatures, especially the
-inference from \emph{some} to \emph{some but not all}, have been studied extensively, both
-theoretically
+that the speaker does not own all of them. This is because, if the speaker owned them all, she
+could have used the word \emph{all} instead of \emph{some} in his utterance, thereby making a
+more informative statement. Scalar implicatures, especially the inference from \emph{some} to
+\emph{some but not all}, have been studied extensively, both theoretically
 \citep[e.g.,][]{Sauerland2004:Scalar-Implicat,ChierchiaFox2008:The-Grammatical,Rooyvan-RooijJagerde-Jager2012:Explaining-Quan}
@@
 \citep[e.g.,][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. While
-there is much dispute in this domain about many details, a position endorsed by a
-clear majority is that a word like \emph{some} is underspecified to mean \emph{some and maybe
-  all} and that the enrichment to \emph{some but not all} is part of some regular 
-process with roots in pragmatics.
+there is much dispute in this domain about many details, a position endorsed by a clear
+majority is that a scalar item like \emph{some} is underspecified to mean \emph{some and maybe
+  all} and that the enrichment to \emph{some but not all} is part of some regular process with
+roots in pragmatics.
 
 If this majority view is correct, the question arises how such a division of labor between
-semantics and pragmatics could have evolved and why it would be so pervasive across natural languages. Models of language evolution abound. There are
-simulation-based models studying the evolution of language in populations of communicating
-agents
+semantics and pragmatics could have evolved and why it would be so pervasive across natural
+languages. Models of language evolution abound. There are simulation-based models studying the
+evolution of language in populations of communicating agents
 \citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
 and there are mathematical models of language evolution, mostly coming from game theory
-\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}.
-Much of this work has focused on explaining basic properties such as compositionality and combinatoriality
-\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax}, but
-little attention has been paid to the interaction between conventional meaning and pragmatic
-use. What is more, many mathematical models explain evolved meaning as a regularity in the
-behavior of agents which maps objective states of the world to observable signals. There is no
-room in such a purely extensional approach to address the semantics-pragmatics division
-directly. Instead, we would need to look at richer representations of cognizing agents and their
-communicative interaction.
+\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
+of this work has focused on explaining basic properties such as compositionality and
+combinatoriality
+\citep[e.g.,][]{Batali1998:Computational-S,nowak+krakauer:1999,nowak+etal:2000,KirbyHurford2002:The-Emergence-o,kirby:2002,SmithKirby2003:Iterated-Learni,Gong2007:Language-Evolut,kirby+etal:2015,verhoef+etal:2014,Franke2015:Proto-Syntax},
+but little attention has been paid to the interaction between conventional meaning and
+pragmatic use. What is more, many mathematical models explain evolved meaning as a regularity
+in the overt behavior of agents. In contrast, we will here look at language users with a richer
+cognitive make-up and their dispositions to act.
 
-To fill this gap, we spell out a model of the co-evolution of conventional meaning and
+Towards this end, we spell out a model of the co-evolution of conventional meaning and
 pragmatic reasoning types. The objects of replication and selection are pairs of lexical
-meanings and general types of pragmatic behavior, which we represent using state-of-the-art
-probabilistic cognitive models of pragmatic language use
+meanings and general types of pragmatic behavior, which we represent using probabilistic models
+of pragmatic language use
 \citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
@@
 communicative efficiency and (ii) possible infidelity in the transmission of linguistic
-knowledge, caused by factors such as inductive learning biases and sparse learning data. Considering
-transmission of linguistic knowledge is important because neither semantic meanings nor
-pragmatic usage patterns are directly observable. Instead, language learners have to infer these
-unobservables from the observable behavior in which they result. We formalize this 
-process as a form of Bayesian inference. Our approach thereby contains a well-understood model
-of iterated Bayesian learning \citep{griffiths+kalish:2007} as a special case, but combines it
-with functional selection, here formalized as the most versatile dynamic from evolutionary game
-theory; the replicator dynamic
+knowledge, caused by factors such as inductive learning biases and sparse learning
+data. Considering transmission of linguistic knowledge is important because neither semantic
+meanings nor pragmatic usage patterns are directly observable. Instead, language learners have
+to infer these unobservables from the observable behavior in which they result. We formalize
+this process as a form of Bayesian inference. Our approach thereby contains a well-understood
+model of iterated Bayesian learning \citep{griffiths+kalish:2007} as a special case, but
+combines it with functional selection, here formalized as the most versatile dynamic from
+evolutionary game theory; the replicator dynamic
 \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this model.
@@
 show that inductive learning biases of Bayesian learners that favor simpler lexical meanings
-can lead to this outcome.
+can prevent the lexicalization of scalar inferences and lead to the emergence of Gricean-like
+pragmatic reasoning types. Results of this case study are critically assessed in the light of
+the assumptions that feed our model in Section~\ref{sec:discussion}.
 
-We see the main contribution of this work as conceptual and technical, not as a definite answer
-to the question why scalar implicatures emerge. It rather demonstrates how current
-probabilistic cognitive modeling of language use and evolutionary modeling can be fruitfully
-combined to study the co-evolution of semantics and pragmatics side-by-side. Reversely, the
-approach taken here may be seen as a first step towards giving an evolutionary rationale for
-empirically successful probabilistic models of language use that embrace the majority view of
-the division of labor between semantics and pragmatics. Section~\ref{sec:discussion} elaborates
-on these points.
+% We see the main contribution of this work as conceptual and technical, not as a definite answer
+% to the question why scalar implicatures emerge. It rather demonstrates how current
+% probabilistic cognitive modeling of language use and evolutionary modeling can be fruitfully
+% combined to study the co-evolution of semantics and pragmatics side-by-side. Reversely, the
+% approach taken here may be seen as a first step towards giving an evolutionary rationale for
+% empirically successful probabilistic models of language use that embrace the majority view of
+% the division of labor between semantics and pragmatics. Section~\ref{sec:discussion} elaborates
+% on these points.
 
-\section{Model}
+\section{A model of co-evolving lexical concepts and pragmatic behavior}
 \label{sec:model}
 
-\subsection{Expressivity and learnability at the semantics-pragmatics interface}
+\subsection{Expressivity and learnability}
 
@@
 
-The idea that language is an adaptation to serve a communicative function has played a pivotal
-role in synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of
-word frequency rankings as a result of competing hearer and speaker preferences \citep[e.g., in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
-  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation
-or reinforcement, favor more communicatively efficient types of behavior, languages are driven
+The idea that language is an adaptation to serve a communicative function is fundamental to
+many synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of word
+frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
+in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
+  piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation or
+reinforcement, favor more communicatively efficient types of behavior, languages are driven
 towards semantic expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
 pressure towards communicative efficiency is not the only force that shapes
-language. Learnability is another, as natural languages need to be learnable to survive their
-faithful transmission across generations. Clearly, an unlearnable code will not make it past
-the one happy fellow who invented it. Importantly, even small learning biases implicit in acquisition
-can build up and have quite striking effects on an evolving language in a process
-of iterated learning
+language. Learnability is another. As natural languages need to be learnable to survive their
+faithful transmission across generations. Even small learning biases implicit in acquisition
+can build up and have quite striking effects on an evolving language in a process of iterated
+learning
 \citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.
 
-While natural languages are pressured for both expressivity and learnability these forces may pull in opposite directions. Their opposition becomes particularly clear when
-considering the extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). A language consisting of a 
-single form-meaning association is easy to learn but lacking in expressivity. Conversely, a
-language that lexicalizes a distinct form for each possible fine-grained meaning a speaker may want to
-convey is maximally expressive but challenging to acquire.
+While natural languages are pressured for both expressivity and learnability these forces may
+pull in opposite directions. Their opposition becomes particularly clear when considering the
+extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). A language consisting of a single
+form-meaning association is easy to learn but lacking in expressivity. Conversely, a language
+that lexicalizes a distinct form for a large number of different meanings is highly expressive
+but challenging to acquire.
 
-An elegant formal approach to capture the interaction between expressivity and learnability
-is the \emph{replicator mutator dynamic}
+\subsection{The replicator mutator dynamic}
+
+An elegant formal approach to capture the interaction between expressivity and learnability is
+the \emph{replicator mutator dynamic}
 \citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. In
@@
 in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
-$i$ before the step, (ii) the fitness $f_i$ of each type $i$ before the step, and (iii) the
-probability $Q_{ji}$ that an agent who wants to imitate, adopt, or learn the type of an agent
-with type $j$ ends up acquiring type $i$:
+$i$ before the update step, (ii) the fitness $f_i$ of each type $i$ before the update, and
+(iii) the probability $Q_{ji}$ that an agent who wants to imitate, adopt, or learn the type of
+an agent with type $j$ ends up acquiring type $i$:
 \begin{align}
@@
 (\ref{eq:RMD_discrete}), equivalently (\ref{eq:RMD_discrete_recast}), contains a model of
-iterated learning \citep{griffiths+kalish:2007}. \mf{should we include a simple example here? I 
-have an example from a lecture ready at hand; it's a simple coordination game in a
-one-population setting.} \tb{Yes, I think that's a good idea.
-We can always cut it out again if the reviewers prefer more compression but let's explain more and cut later.
-The paper is relatively short at the moment.}
+iterated learning \citep{griffiths+kalish:2007}.
 
-Where our goal is an application of this dynamic to the case of co-evolution of semantic
-meaning and pragmatic use, we need to fix what the relevant types are, how fitness is measured
-and how the mutation matrix is computed. These issues are addressed, one by one, in the
-following.
+Consider the example of a simple coordination game. Agents are of two types: positive and
+negative. If agents of different type play with each other, they obtain a payoff of 0. If
+negative meets negative, each receives a payoff of 1. If positive meets positive, they get a
+payoff of 2. A population state is completely characterized by the proportion $x$ of
+negatives. The fitness of negatives is $f_n(x) = x$, that of positives is $f_p(x) = 2-2x$. The
+average fitness is $\Phi(x) = x f_n(x) + (1-x) f_p(x) = 3x^2 - 4x + 2$. Without mutation, the
+replicator dynamic will then update $x$ to $RD(x) = \nicefrac{x^2}{\Phi(x)}$. The replicator
+step is plotted in Figure~\ref{fig:Updates_RMS} as the blue line. Rest points, for which
+$RD(x)=x$, are at $x=0$, $x=1$ and $x= \nicefrac{2}{3}$. The former are attractors as nearby
+points converge to them. Points near $x=\nicefrac{2}{3}$ move towards 0 or 1. This is
+schematically pictured in the topmost phase portrait in Figure~\ref{fig:Phase_RD}.
+
+Adding mutation changes the dynamics and its rest points. Let's assume that $Q_{ji} = .9$ when
+$j=i$. The update effect of mutation alone is $M(x) = .9 x + .1 (1-x) = .8x + .1$ and is
+plotted as the linear green line in Figure~\ref{fig:Updates_RMS}. It has only one stable rest
+point at $x = 0.5$. If we first take the replicator dynamic update and then the mutation update
+in sequence, we obtain the replicator mutator dynamic $RMD(x) = M(RD(x)) = \nicefrac{.9x^2 -
+  .2x +.2}{3x^2 - 4x + 2}$, which is plotted in red in Figure~\ref{fig:Updates_RMS}. The rest
+points are at $x=.121$, $x=.903$ and $x=.609$. The former two are attractors (see bottom of
+Figure~\ref{fig:Phase_RD}).
+
+\begin{figure}[t]
+  \centering
+
+    \begin{subfigure}[b]{0.45\textwidth}
+      \centering
+      \begin{tikzpicture}
+      \begin{axis}[xlabel=$x$, ylabel = $x'$,
+                   grid=major, width=6.85cm, height = 5cm,
+                   legend style = {legend pos = north west},
+                   ymin=0,ymax=1, xmin=0,xmax=1]
+
+      \addplot[smooth, color=blue] {x^2 / (3*x^2 - 4*x + 2) + 0.01};
+      \addplot[smooth, color=green] {.8*x + .1};
+
+      \addplot[smooth, color=red] {(.9*x^2 - .2*x +.2)/(3*x^2 - 4*x + 2)};
+
+      \addplot[smooth, color=gray] {x};
+
+      \legend{RD, M, RMD}
+
+      \end{axis}
+    \end{tikzpicture}
+
+        \caption{Update functions: the population state $x$ is mapped onto $x'$ in one update step}
+        \label{fig:Updates_RMS}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{0.5\textwidth}    
+      \centering
+    \begin{tikzpicture}[x=120]
+      \node[draw=black, circle, fill = black, minimum size = 0.25cm]
+      (0) at (0,0) {};
+
+      \node[draw=black, circle, fill = black, minimum size = 0.25cm]
+      (1) at (1,0) {};
+
+      \draw[-, thick] (0) -- (1);
+
+      \node[draw=black, circle, fill = white, minimum size = 0.25cm, thick]
+      (mid) at (2/3,0) {};
+
+      \node[]  (0.label) at (0,-0.5) {0};
+
+      \node[]  (1.label) at (1,-0.5) {1};
+
+      \node[]  (label) at (0.5,-0.5) {$x$};
+
+      \node[]  (label) at (0.5,0.55) {RD};
+
+      \draw[->-=0.8,very thick] (mid) -- (0);
+
+      \draw[->-=0.8,very thick] (mid) -- (1);
+
+    \end{tikzpicture}
+
+    \vspace*{1cm}
+
+    \begin{tikzpicture}[x=120]
+
+      \node (0) at (-0.05,0) {};
+
+      \node (1) at (1.05,0) {};
+
+      \node[draw=black, circle, fill = black, minimum size = 0.25cm]
+      (S0) at (0.121,0) {};
+
+      \node[draw=black, circle, fill = black, minimum size = 0.25cm]
+      (S1) at (0.903,0) {};
+
+      \draw[-, thick] (0) -- (1);
+
+      \node[draw=black, circle, fill = white, minimum size = 0.25cm, thick]
+      (mid) at (0.609,0) {};
+
+      \node[]  (0.label) at (0,-0.3) {0};
+
+      \node[]  (1.label) at (1,-0.3) {1};
+
+      \node[]  (label) at (0.5,-0.5) {$x$};
+
+      \node[]  (label) at (0.5,0.55) {RMD};
+
+      \draw[->-=0.5,very thick] (mid) -- (S0);
+
+      \draw[->-=0.5,very thick] (mid) -- (S1);
+
+      \draw[->-=0.5,very thick] (0) -- (S0);
+
+      \draw[->-=0.5,very thick] (1) -- (S1);
+
+    \end{tikzpicture}
+
+    \caption{Phase portraits for RD and RMD: unstable rest points are hollow, attractors are
+      solid.}
+        \label{fig:Phase_RD}
+    \end{subfigure}
+
+  \caption{Example}
+  \label{fig:Example_RMD}
+\end{figure}
+
+While this was only an abstract and simple example, our goal is to apply the RMD to a more
+complex case of co-evolving of lexical concepts and pragmatic behavior. To do so, we need to
+fix three things: (i) what the relevant types are, (ii) how fitness derives from communicative
+success and (iii) how the mutation matrix is computed. These issues are addressed, one by one,
+in the following.
 

### 2017-03-22 paper/paper-v3.tex
model section polish

@@
 
-Types are what cultural evolution operates on. In standard applications of evolutionary game
-theory, types correspond to ways of acting in a game, e.g., either cooperating or defecting in
-a prisoner's dilemma. \mf{maybe good to refer back to the example from before if there was
-  one?} \tb{Yes!} For our present purpose types are identified by their cognitive make-up. Since we are
-interested in the question under which conditions processes of cultural evolution will favor
-specific divisions of labor between lexical meaning and pragmatic use, a type is a pair
-consisting of a lexicon and a pragmatic strategy.
+Types are what evolution operates on. In standard applications of evolutionary game theory,
+types are ways of acting in a game, e.g., either cooperating or defecting in a prisoner's
+dilemma, or, in our previous example, they are what defines an agent's fitness in interaction
+with others, such as being positive or negative (whatever that means). For our present
+purposes, types are identified more concretely by their cognitive make-up. In general terms, we
+are interested in the evolutionary competition between different lexical concepts and ways of
+using them in communication. A type in our setting is therefore a pair consisting of a lexicon
+and a pragmatic strategy.
+
+Agents play signaling games, in which the speaker wants to communicate a world state $s$ with a
+message $m$ to a hearer who receives $m$ but does not know $s$
+\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a set of
+states. A pragmatic type maps a lexicon onto a probabilistic speaker rule (a probabilistic
+choice of message for each state) and a probabilistic listener rule (a probabilistic choice of
+state for each message). There are many ways of making these general notions more
+concrete. Here is what we will assume in the remainder of this paper.
+
+\bigskip
+\bigskip
+\bigskip
 
@@
 rational language use
-\citep{franke:2009,frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
+\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
 capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,

### 2017-03-23 paper/paper-v3.tex
paper update

@@
 \usetikzlibrary{pgfplots.groupplots,decorations.markings}
+
+\newtheoremstyle{Satz}
+   {}                      %Space above
+   {1em}                   %Space below
+   {\normalfont}           %Body font
+   {}                      %Indent amount (empty = no indent,
+                           %\parindent = para indent)
+   {\normalfont}           %Thm head font
+   {.}                     %Punctuation after thm head
+   {.8em}                  %Space after thm head: " " = normal interword
+                           %space; \newline = linebreak
+   {\bfseries\thmname{#1}\thmnote{ (#3)}}
+                           %Thm head spec (can be left empty, meaning
+                           %`normal')
+
+\theoremstyle{Satz}
+\newtheorem{example}{Example}
 
@@
 
-\title{Tracing the cultural evolution of meaning at the semantics-pragmatics interface}
+\title{Co-evolution of lexical meaning \& pragmatic use}
 
@@
   interface, we present a game theoretic model of the competition between types of language
-  users, each endowed with her selection of lexical concepts and her own pragmatic dispositions
-  to act on them. Our model traces two evolutionary forces and their interaction: (i)
-  fitness-based pressure towards communicative efficiency and (ii) potential learning biases
-  during the transfer of linguistic knowledge. We illustrate the model based on a case study on
-  scalar implicatures. In this case study learning biases that favor simple semantic
+  users, each endowed with a selection of lexical concepts and a particular pragmatic
+  disposition to act on them. Our model traces two evolutionary forces and their interaction:
+  (i) fitness-based pressure towards communicative efficiency and (ii) potential learning
+  biases during the transfer of linguistic knowledge. We illustrate the model based on a case
+  study on scalar implicatures. In this case study learning biases that favor simple semantic
   representations can boost the evolution of more sophisticated pragmatic reasoning types and
@@
 in the overt behavior of agents. In contrast, we will here look at language users with a richer
-cognitive make-up and their dispositions to act.
-
-Towards this end, we spell out a model of the co-evolution of conventional meaning and
-pragmatic reasoning types. The objects of replication and selection are pairs of lexical
-meanings and general types of pragmatic behavior, which we represent using probabilistic models
-of pragmatic language use
+cognitive make-up.
+
+We spell out a model of the co-evolution of conventional meaning and pragmatic reasoning
+types. The objects of replication and selection are pairs of lexical meanings and general types
+of pragmatic behavior, which we represent using probabilistic models of pragmatic language use
 \citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
@@
 pressure towards communicative efficiency is not the only force that shapes
-language. Learnability is another. As natural languages need to be learnable to survive their
+language. Learnability is another. Natural languages need to be learnable to survive their
 faithful transmission across generations. Even small learning biases implicit in acquisition
@@
 
-Consider the example of a simple coordination game. Agents are of two types: positive and
-negative. If agents of different type play with each other, they obtain a payoff of 0. If
-negative meets negative, each receives a payoff of 1. If positive meets positive, they get a
-payoff of 2. A population state is completely characterized by the proportion $x$ of
-negatives. The fitness of negatives is $f_n(x) = x$, that of positives is $f_p(x) = 2-2x$. The
-average fitness is $\Phi(x) = x f_n(x) + (1-x) f_p(x) = 3x^2 - 4x + 2$. Without mutation, the
-replicator dynamic will then update $x$ to $RD(x) = \nicefrac{x^2}{\Phi(x)}$. The replicator
-step is plotted in Figure~\ref{fig:Updates_RMS} as the blue line. Rest points, for which
-$RD(x)=x$, are at $x=0$, $x=1$ and $x= \nicefrac{2}{3}$. The former are attractors as nearby
-points converge to them. Points near $x=\nicefrac{2}{3}$ move towards 0 or 1. This is
-schematically pictured in the topmost phase portrait in Figure~\ref{fig:Phase_RD}.
-
-Adding mutation changes the dynamics and its rest points. Let's assume that $Q_{ji} = .9$ when
-$j=i$. The update effect of mutation alone is $M(x) = .9 x + .1 (1-x) = .8x + .1$ and is
+\begin{example}
+  Consider a simple coordination game. Agents are of two types: positive and negative. If
+  agents of different type play with each other, they obtain a payoff of 0. If negative meets
+  negative, each receives a payoff of 1. If positive meets positive, they get a payoff of 2. A
+  population state is completely characterized by the proportion $x$ of negatives. The fitness
+  of negatives in population state $x$ is $f_n(x) = x$, that of positives is $f_p(x) = 2-2x$.
+  The average fitness is $\Phi(x) = x f_n(x) + (1-x) f_p(x) = 3x^2 - 4x + 2$. Without mutation,
+  the replicator dynamic will then update $x$ to $RD(x) = \nicefrac{x^2}{\Phi(x)}$. The update
+  function $RD(x)$ of the replicator step is plotted in Figure~\ref{fig:Updates_RMS} as the
+  blue line. Rest points, for which $RD(x)=x$, are at $x=0$, $x=1$ and $x= \nicefrac{2}{3}$.
+  The former are attractors as nearby points converge to them. Points near $x=\nicefrac{2}{3}$
+  move towards 0 or 1. This is schematically pictured in the topmost phase portrait in
+  Figure~\ref{fig:Phase_RD}.
+
+  Adding mutation changes the dynamics and its rest points. Let's assume that $Q_{ji} = .9$
+  when $j=i$. The update effect of mutation alone is $M(x) = .9 x + .1 (1-x) = .8x + .1$ and is
   plotted as the linear green line in Figure~\ref{fig:Updates_RMS}. It has only one stable rest
-point at $x = 0.5$. If we first take the replicator dynamic update and then the mutation update
-in sequence, we obtain the replicator mutator dynamic $RMD(x) = M(RD(x)) = \nicefrac{.9x^2 -
-  .2x +.2}{3x^2 - 4x + 2}$, which is plotted in red in Figure~\ref{fig:Updates_RMS}. The rest
-points are at $x=.121$, $x=.903$ and $x=.609$. The former two are attractors (see bottom of
-Figure~\ref{fig:Phase_RD}).
+  point at $x = 0.5$ (see Figure~\ref{fig:Phase_RD}). If we first take the replicator step and
+  then the mutation step in sequence, we obtain the replicator mutator dynamic
+  $RMD(x) = M(RD(x)) = \nicefrac{.9x^2 - .2x +.2}{3x^2 - 4x + 2}$, which is plotted in red in
+  Figure~\ref{fig:Updates_RMS}. The rest points are at $x=.121$, $x=.903$ and $x=.609$. The
+  former two are attractors (see Figure~\ref{fig:Phase_RD}).
+\end{example}
 
@@
   \centering
+
+  %% notice that the figure is hacked, because somehow the curve does not come out right under
+  %% PGFplots; I don't know why!?!?
 
@@
       \begin{axis}[xlabel=$x$, ylabel = $x'$,
-                   grid=major, width=6.85cm, height = 5cm,
+                   grid=major, width=6.85cm, height = 5.5cm,
                    legend style = {legend pos = north west},
@@
 
-      \addplot[smooth, color=blue] {x^2 / (3*x^2 - 4*x + 2) + 0.01};
+      \addplot[smooth, color=blue] {x^2 / (3*x^2 - 4*x + 2) + 0.019 * x};
       \addplot[smooth, color=green] {.8*x + .1};
@@
 
-    \vspace*{1cm}
+    \vspace*{0.35cm}
+
+    \begin{tikzpicture}[x=120]
+
+      \node (0) at (-0.05,0) {};
+
+      \node (1) at (1.05,0) {};
+
+      \draw[-, thick] (0) -- (1);
+
+      \node[draw=black, circle, fill = black, minimum size = 0.25cm, thick]
+      (mid) at (0.5,0) {};
+
+      \node[]  (0.label) at (0,-0.3) {0};
+
+      \node[]  (1.label) at (1,-0.3) {1};
+
+      \node[]  (label) at (0.5,-0.5) {$x$};
+
+      \node[]  (label) at (0.5,0.55) {M};
+
+      \draw[->-=0.5,very thick] (0) -- (mid);
+
+      \draw[->-=0.5,very thick] (1) -- (mid);
+
+    \end{tikzpicture}
+
+    \vspace*{0.35cm}
 
@@
 
-    \caption{Phase portraits for RD and RMD: unstable rest points are hollow, attractors are
+    \caption{Phase portraits for RD, M and RMD: unstable rest points are hollow, attractors are
       solid.}

### 2017-03-23 paper/paper-v3.tex
paper polishing: added examples for probabilistic behavior types

@@
 
-  Adding mutation changes the dynamics and its rest points. Let's assume that $Q_{ji} = .9$
+  Adding mutation changes the dynamic and its rest points. Let's assume that $Q_{ji} = .9$
   when $j=i$. The update effect of mutation alone is $M(x) = .9 x + .1 (1-x) = .8x + .1$ and is
@@
 
-While this was only an abstract and simple example, our goal is to apply the RMD to a more
-complex case of co-evolving of lexical concepts and pragmatic behavior. To do so, we need to
-fix three things: (i) what the relevant types are, (ii) how fitness derives from communicative
-success and (iii) how the mutation matrix is computed. These issues are addressed, one by one,
-in the following.
-
-\subsection{Types: Lexica and linguistic behavior}
+\subsection{Fitness \& learnability of lexical meanings \& pragmatic strategies}
+\label{sec:fitn--learn}
+
+Moving beyond an abstract example, our goal is to apply the RMD to investigate the co-evolution
+of lexical concepts and pragmatic behavior. To do so, we need to fix three things: (i) what the
+relevant types are, (ii) how fitness derives from communicative success and (iii) how the
+mutation matrix is computed. These issues are addressed, one by one, in the following.
+
+\subsubsection{Types: Lexica and pragmatic strategies}
 \label{sec:languages+use}
 
-Types are what evolution operates on. In standard applications of evolutionary game theory,
-types are ways of acting in a game, e.g., either cooperating or defecting in a prisoner's
-dilemma, or, in our previous example, they are what defines an agent's fitness in interaction
-with others, such as being positive or negative (whatever that means). For our present
-purposes, types are identified more concretely by their cognitive make-up. In general terms, we
-are interested in the evolutionary competition between different lexical concepts and ways of
-using them in communication. A type in our setting is therefore a pair consisting of a lexicon
-and a pragmatic strategy.
+Types are what evolution operates on. They define an agent's fitness, usually through a payoff
+accrued in single interactions with other agents. Often types can be identified as the possible
+acts in a game, e.g., either cooperating or defecting in a prisoner's dilemma. In other cases,
+they may be thought of as general properties of an agent that influences her fitness, such as
+being positive or negative in our previous example (whatever that means). For our present
+purposes, types are identified more concretely by specific assumptions about their cognitive
+make-up. Since we are interested in the evolutionary competition between different lexical
+concepts and ways of using them in communication, a type is here defined as a pair consisting
+of a lexicon and a pragmatic strategy.
 
@@
 message $m$ to a hearer who receives $m$ but does not know $s$
-\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a set of
-states. A pragmatic type maps a lexicon onto a probabilistic speaker rule (a probabilistic
-choice of message for each state) and a probabilistic listener rule (a probabilistic choice of
-state for each message). There are many ways of making these general notions more
-concrete. Here is what we will assume in the remainder of this paper.
-
-\bigskip
-\bigskip
-\bigskip
+\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a
+(possibly fuzzy) set of states. A pragmatic type maps a lexicon onto a probabilistic speaker
+rule (a probabilistic choice of message for each state) and a probabilistic listener rule (a
+probabilistic choice of state for each message). There are many ways of making these general
+notions more concrete. Here is what we will assume in the remainder of this paper.
 
@@
 short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
-same sentence with \emph{some} replaced by \emph{all}.  
-Lexica for this case would assign a Boolean truth value, either $0$ for false or $1$ for true, to each state-message pair. The following two lexica exemplify the distinction
-between a lexicalized upper-bound for \emph{some} in $\Lbound$ and the widely assumed logical
-semantics with only a lower-bound in $\Llack$.
+same sentence with \emph{some} replaced by \emph{all}.  Lexica for this case would assign a
+Boolean truth value, either $0$ for false or $1$ for true, to each state-message pair. The
+following two lexica are minimal examples for the distinction between a lexicalized upper-bound
+for \emph{some} in $\Lbound$ and the widely assumed logical semantics with only a lower-bound
+in $\Llack$.
 
@@
 
-We distinguish between two kinds of linguistic behavior. {\em Literal interlocutors} produce and
+Pragmatic types define dispositions of produce and interpret messages given a lexicon. We
+distinguish between two kinds of pragmatic types. {\em Literal interlocutors} produce and
 interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
@@
 capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
-level $0$, corresponds to literal language use. Pragmatic language users of level $n + 1$
-act (approximately) rational with respect to level-$n$ behavior of their
-interlocutors. (\ref{h:level0}) and (\ref{s:level0}) define probabilistic behavior of literal
-hearers and speakers respectively: \mf{is it a problem that $S$ denotes speakers and the set of
-states?} \tb{I think that's fine.
-$S_n(\cdot \mid \cdot)$ is not used beyond this section.
-Otherwise we can, e.g., relabel $M$ as $F$(orms) and use $M$(eanings) instead of $S$(tates).
-I'm fine with either option.}
-\begin{flalign}
-&H_{0}(s \mid m;L) \propto pr(s) L_{sm} \label{h:level0}\\
-&S_{0}(m \mid s;L) \propto \exp(\lambda \; L_{sm}) \label{s:level0}
-\end{flalign}
-
-According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state
-$s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in
-the following this prior is assumed to be uniform. A literal interpreter with lexicon $\Lbound$
-from above would assign $\ssome$ a probability of $H_0(\ssome \mid \msome; \Lbound) = 1$ after
-hearing $\msome$, while a literal interpreter with lexicon $\Llack$ would assign $\ssome$
-probability $H_0(\ssome \mid \msome; \Llack) = 0.5$. As usual in probabilistic pragmatics models, speaker behavior is regulated by a soft-max
-parameter $\lambda$, $\lambda \geq 1$ \citep{luce:1959,sutton+barto:1998}. As $\lambda$
-increases, choices made in production are more rational in that they are increasingly in line with expected utility maximization. Expected utility of a message
-$\messg$ in state $\state$ for a level $n+1$ speaker is here defined as $H_{n}(s|m;L)$, the probability that the hearer will assign to or choose the correct meaning.
-Put differently, expected utility indicates how likely the hearer is thought to infer $\state$ when sent $\messg$.
-The formulation
-in (\ref{s:level0}) also allows false messages to be sent with a small positive probability in
-analogy to the definition of pragmatic speakers in probabilistic pragmatic models.
-This is also necessary to guarantee a mutation matrix with only positive entries (see below).
-If $\lambda = 1$ a literal
-speaker with either lexicon $\Lbound$ or $\Llack$ produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
-\ssome;L_{\text{bound},\text{lack}}) \approx .73$. The probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
-\sall;\Lbound) \approx .27$ for $\Lbound$ and $S_{0}(\msome \mid
-\sall;\Llack) = .5$ for $\Llack$. By contrast, if $\lambda = 20$, a literal speaker with either lexicon produces $\msome$ in $\ssome$ with probability $S_{0}(\msome \mid
-\ssome;L_{\text{bound},\text{lack}}) \approx 1$. But the probability of producing $\msome$ in $\sall$ is $S_{0}(\msome \mid
-\sall;\Lbound) \approx 1$ for $\Lbound$ whereas it is still $S_{0}(\msome \mid \sall;\Llack) = .5$ for $\Llack$ because it semantically associates $\msome$ with both states.
-
-Pragmatic behavior of level $n+1$ is similar to its literal counterparts but bases linguistic choice in 
-interpretation or production on the expected behavior of a level-$n$ interlocutor instead of on lexical meaning:
-\begin{flalign}
-&H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
-&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)) \label{s:leveln}
-\end{flalign}
-
-Particularly important to our purpose is the fact non-literal interlocutors using $\Llack$ can {\em pragmatically} associate $\msome$ with $\ssome$ over $\sall$,  in contrast to their literal counterparts.
-Intuitively, this is so because hearers reason a speaker to convey $\sall$ with $\mall$, as this message has a higher chance of leading to mutual understanding than using underspecified $\msome$.
-Therefore they will expect $\msome$ to be more strongly associated with $\ssome$ than with $\sall$ as it is already possible to clearly convey the latter state $\mall$.
-The converse reasoning pattern holds for the speaker.
-In this way, two distinct lexical conventions such as $\Lbound$ and $\Llack$ can give rise to similar observable linguistic behavior.
-In this case, associating $\msome$ with $\ssome$ and $\mall$ with $\sall$.
-
-\subsection{Fitness \& fitness-based selection based on expressivity}\label{sec:expressivity}
-
-Most evolutionary dynamics assume that the proportion of type $i$ in a population will increase
-or decrease as a function of its relative fitness $f_i$. In the context of language evolution,
-fitness is frequently associated with expressivity, i.e., the ability to successfully
-communicate with other language users from the same population
+level $0$, corresponds to literal language use, as in Equations~(\ref{h:level0}) and
+(\ref{s:level0}). Pragmatic language users of level $n + 1$ act (approximately) rational with
+respect to level-$n$ behavior of their interlocutors, as in Equations~(\ref{h:leveln}) and
+(\ref{s:leveln}).
+\begin{alignat}{3}
+&H_{0}(s \mid m;L) &&\propto pr(s) L_{sm} \label{h:level0}\\
+&S_{0}(m \mid s;L) &&\propto \exp(\lambda \; L_{sm}) \label{s:level0}\\
+&H_{n+1}(s|m;L) &&\propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
+&S_{n+1}(m|s;L) &&\propto  \exp(\lambda \; H_{n}(s|m;L)) \label{s:leveln}
+\end{alignat}
+
+According to (\ref{h:level0}), a literal hearer's interpretation of a message depends on her
+lexicon and her prior over states, $pr \in \Delta(S)$, which is here assumed flat, for
+simplicity. Literal interpreters thereby choose an arbitrary true interpretation for each
+message. Pragmatic hearers, defined in (\ref{h:leveln}), use Bayes rule to weigh
+interpretations based on a conjecture about speaker behavior. Speaker behavior is regulated by
+a soft-max parameter $\lambda$, $\lambda \geq 0$ \citep{luce:1959,sutton+barto:1998}. As
+$\lambda$ increases, choices approximate strict maximization of expected utilities. Expected
+utility of a message $\messg$ in state $\state$ for a level $n+1$ speaker is here defined as
+$H_{n}(s|m;L)$, the probability that the hearer will assign to or choose the correct
+meaning. For literal speakers, utility only tracks truthfulness. Literal speakers choose any
+true message with equal probability but may send false messages as well with a probability
+dependent on $\lambda$.
+
+Here are some concrete examples. \mf{please double-check the numbers!} A literal interpreter with lexicon $\Lbound$ assigns $\ssome$
+a probability of $H_0(\ssome \mid \msome; \Lbound) = 1$ after hearing $\msome$, while a literal
+interpreter with $\Llack$ has $H_0(\ssome \mid \msome; \Llack) = 0.5$:
+\begin{align*}
+  H_0(\cdot \mid \cdot, \Lbound) & = \begin{blockarray}{lcc}
+    & \mygray{\ssome} & \mygray{\sall} \\
+    \begin{block}{l[cc]}
+      \mygray{\msome} & 1 & 0 \\
+      \mygray{\mall}  & 0 & 1 \\
+    \end{block}
+  \end{blockarray} &
+  H_0(\cdot \mid \cdot, \Llack) & = \begin{blockarray}{lcc}
+    & \mygray{\ssome} & \mygray{\sall} \\
+    \begin{block}{l[cc]}
+      \mygray{\msome} & .5 & .5 \\
+      \mygray{\mall}  & 0 & 1 \\
+    \end{block}
+  \end{blockarray} 
+\end{align*}
+If $\lambda = 1$ literal speaker behavior looks like this:
+\begin{align*}
+  S_0(\cdot \mid \cdot, \Lbound) & \approx \begin{blockarray}{lcc}
+    & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[cc]}
+      \mygray{\ssome} & .73 & .27 \\
+      \mygray{\sall}  & .27 & .73 \\
+    \end{block}
+  \end{blockarray} &
+  S_0(\cdot \mid \cdot, \Llack) & \approx \begin{blockarray}{lcc}
+    & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[cc]}
+      \mygray{\ssome} & .73 & .27 \\
+      \mygray{\sall}  & .5 & .5 \\
+    \end{block}
+  \end{blockarray} 
+\end{align*}
+If $\lambda = 20$, literal speakers will choose with less slack:
+\begin{align*}
+  S_0(\cdot \mid \cdot, \Lbound) & \approx \begin{blockarray}{lcc}
+    & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[cc]}
+      \mygray{\ssome} & 1 & 0 \\
+      \mygray{\sall}  & 0 & 1 \\
+    \end{block}
+  \end{blockarray} &
+  S_0(\cdot \mid \cdot, \Llack) & \approx \begin{blockarray}{lcc}
+    & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[cc]}
+      \mygray{\ssome} &  1 & 0 \\
+      \mygray{\sall}  & .5 & .5 \\
+    \end{block}
+  \end{blockarray} 
+\end{align*}
+Pragmatic hearers of level 1 have the following probabilistic interpretation behavior for
+$\lambda = 1$:
+\begin{align*}
+  H_1(\cdot \mid \cdot, \Lbound) & \approx \begin{blockarray}{lcc}
+    & \mygray{\ssome} & \mygray{\sall} \\
+    \begin{block}{l[cc]}
+      \mygray{\msome} & .73 & .27 \\
+      \mygray{\mall}  & .27 & .73 \\
+    \end{block}
+  \end{blockarray} &
+  H_1(\cdot \mid \cdot, \Llack) & \approx \begin{blockarray}{lcc}
+    & \mygray{\ssome} & \mygray{\sall} \\
+    \begin{block}{l[cc]}
+      \mygray{\msome} & .59 & .41 \\
+      \mygray{\mall}  & 0.35 & 0.65 \\
+    \end{block}
+  \end{blockarray} 
+\end{align*}
+For $\lambda = 20$, there wil less slack again:
+Pragmatic listeners of level 1 have the following probabilistic interpretation behavior for
+$\lambda = 1$:
+\begin{align*}
+  H_1(\cdot \mid \cdot, \Lbound) & \approx \begin{blockarray}{lcc}
+    & \mygray{\ssome} & \mygray{\sall} \\
+    \begin{block}{l[cc]}
+      \mygray{\msome} & 1 & 0 \\
+      \mygray{\mall}  & 0 & 1 \\
+    \end{block}
+  \end{blockarray} &
+  H_1(\cdot \mid \cdot, \Llack) & \approx \begin{blockarray}{lcc}
+    & \mygray{\ssome} & \mygray{\sall} \\
+    \begin{block}{l[cc]}
+      \mygray{\msome} & 0.67 & 0.33 \\
+      \mygray{\mall}  & 0 & 1 \\
+    \end{block}
+  \end{blockarray} 
+\end{align*}
+With $\lambda = 1$, pragmatic speakers of level 1 will have:
+\begin{align*}
+  S_1(\cdot \mid \cdot, \Lbound) & \approx \begin{blockarray}{lcc}
+    & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[cc]}
+      \mygray{\ssome} & .73 & .37 \\
+      \mygray{\sall}  & .37 & .73 \\
+    \end{block}
+  \end{blockarray} &
+  S_1(\cdot \mid \cdot, \Llack) & \approx \begin{blockarray}{lcc}
+    & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[cc]}
+      \mygray{\ssome} & .62 & .38 \\
+      \mygray{\sall}  & .38 & .62 \\
+    \end{block}
+  \end{blockarray} 
+\end{align*}
+For high $\lambda = 20$, pragmatic speakers of level 1 give us:
+\begin{align*}
+  S_1(\cdot \mid \cdot, \Lbound) & \approx \begin{blockarray}{lcc}
+    & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[cc]}
+      \mygray{\ssome} & 1 & 0 \\
+      \mygray{\sall}  & 0 & 1 \\
+    \end{block}
+  \end{blockarray} &
+  S_1(\cdot \mid \cdot, \Llack) & \approx \begin{blockarray}{lcc}
+    & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[cc]}
+      \mygray{\ssome} & 1 & 0 \\
+      \mygray{\sall}  & 0 & 1 \\
+    \end{block}
+  \end{blockarray} 
+\end{align*}
+
+It is important to note that pragmatic agents of level 1 using $\Llack$ associate $\msome$
+preferentially with $\ssome$, in contrast to their literal counterparts of level 0. This
+association is not perfect, and usually less strong than what agents with a lexicalized upper
+bound in $\Lbound$ can achieve, even without pragmatic reasoning. Higher oder iteration of
+reasoning types leads to stronger associations of $\msome$ and $\ssome$ also for the
+hearer. Still, the case study presented in Section~\ref{sec:si-case-study} will consider
+speaker and hearer behavior at levels 0 and 1, as the latter are the simplest pragmatic
+reasoning types which show a tendency to communicatively attuned pragmatic enrichment. Using
+only level 1 reasoning types is therefore a conservative choice that works against the
+fitness-based selection of pragmatic language use for a notion of fitness based on
+communicative efficiency, which is introduced next.
+
+\subsubsection{Fitness \& fitness-based selection based on expressivity}\label{sec:expressivity}
+
+Under the replicator dynamic the proportion of type $i$ in a population will increase or
+decrease as a function of its relative fitness $f_i$. In the context of language evolution,
+fitness is usually associated with expressivity, i.e., the ability to successfully communicate
+with other language users from the same population
 \citep[e.g.,][]{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Under a biological
@@
 peers. Under a cultural interpretation the picture is that agents themselves strive towards
-communicative success and therefore occasionally adapt or revise their behavior to achieve a
+communicative success and therefore occasionally adapt or revise their behavior to achieve
 higher communicative success (see \citealt[\S3.3]{benz+etal:2005b} for discussion).
 
-The replicator equation gives us the means to make the ensuing dynamics precise, without
-necessarily committing to a biological or cultural interpretation. As above, the proportion of types in a
-given population is codified in a vector $\vec{x}$, where $x_i$ is type $i$'s proportion. The
-fitness of type $i$ is its average expected communicative success, or \emph{expected
-  utility} (EU), given the frequencies of types in the current population:
+The replicator equation gives us the means to make the ensuing dynamic precise, without
+necessarily committing to a biological or cultural interpretation. As above, the proportion of
+types in a given population is codified in a vector $\vec{x}$, where $x_i$ is the proportion of
+type $i$. The fitness of type $i$ is its average expected communicative success, or
+\emph{expected utility} (EU), given the frequencies of types in the current population:
 \begin{align*}
@@
 The expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
-average success of $i$ when talking or listening to $j$. Assuming that agents are speakers half
-of the time, this yields:
+average success of $i$ when talking or listening to $j$. If agents are speakers half of the
+time, this yields:
 \begin{align*}
@@
 \begin{flalign*}
-  & \text{EU}_S(t_i,t_j)  = \sum_s P(s)\sum_m S_{n_i}(m \mid s;L_i) \sum_{s'} R_{n_j}(s' \mid m;L_j)
-  \delta(s,s') \\
- & \text{EU}_H(t_i,t_j)  = \text{EU}_S(t_j,t_i)
+  & \text{EU}_S(t_i,t_j)  = \sum_s P(s)\sum_m S_{n_i}(m \mid s;L_i) \sum_{s'} H_{n_j}(s' \mid m;L_j)
+  \delta(s,s') \,, \\
+ & \text{EU}_H(t_i,t_j)  = \text{EU}_S(t_j,t_i)\,.
 \end{flalign*}
-As usual for cooperative communication, $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.
-
-\subsection{Learnability}
+As usual, $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.
+
+\subsubsection{Learnability}
 \label{sec:learnability}
@@
 Languages are shaped not only by functionalist forces towards greater expressivity. Another
-important factor is the fidelity by which language is transmitted. Among others, linguistic
-production can be prone to errors, states or messages may be perceived incorrectly, and
-multiple languages may be compatible with the data learners are exposed to. These sources of
-uncertainty introduce variation in the transmission of linguistic knowledge from one generation to the next.
-In particular, learning biases in the iterated transmission process can influence language evolution
-substantially.
+important factor is the fidelity with which linguistic knowledge is transmitted. Among others,
+linguistic production can be prone to errors, states or messages may be perceived incorrectly,
+and multiple languages may be compatible with the data learners are exposed to. These sources
+of uncertainty introduce variation in the transmission of linguistic knowledge from one
+generation to the next. In particular, learning biases in the iterated transmission process can
+influence language evolution substantially.
 
 In biological evolution, where types are expressed genetically, transmission infidelity comes
-into the picture through infrequent and mostly random genetic mutations. However, an agent's lexicon
-and pragmatic reasoning behavior is not inherited genetically. They need to be learned from
-observation. Concretely, when agents of type $j$ want to adopt or imitate the linguistic
-behavior of type $i$, they observe the overt linguistic behavior of type $i$ and need to infer what
-their type is from that. Iterated learning is a process in which languages are learned repeatedly from the observation
-of linguistic behavior of agents who have acquired the language from observation and inference
-before as well. In the simplest case there is a single teacher and a single
-learner. After sufficient training the learner becomes a teacher and produces behavior that
-serves as input for a new learner. Due to the pressure towards learnability it exerts, iterated
-learning alone generally leads to simpler and more regular languages (see \citealt{kirby+etal:2014}
-and \citealt{tamariz+kirby:2016} for recent surveys).
+into the picture through infrequent and mostly random genetic mutations. However, an agent's
+lexicon and pragmatic reasoning behavior is likely not inherited genetically. They need to be
+learned from observation. Concretely, when agents of type $j$ want to adopt or imitate the
+linguistic behavior of type $i$, they observe the overt linguistic behavior of type $i$ and
+need to infer the covert type that most likely produced the visible behavior. Iterated learning
+is a process in which languages are learned repeatedly from the observation of linguistic
+behavior of agents who have themselves acquired the language from observation and inference. In
+the simplest case there is a single teacher and a single learner. After sufficient training the
+learner becomes a teacher and produces behavior that serves as input for a new learner. Due to
+the pressure towards learnability it exerts, iterated learning alone generally leads to simpler
+and more regular languages (see \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016} for
+recent surveys).
 
@@
 inference process. On the one hand, early simulation results suggested that weak biases could
-be magnified by exposing learners to only small data samples \citep[e.g. in][]{brighton:2002}. On the other, \citeposs{griffiths+kalish:2007} mathematical
-characterization showed that iterated learning converged to the prior, i.e., that the resulting
-distribution over languages corresponds to the learners' prior distribution and is not
-influenced by the amount of input given to them. This difference in predictions can be traced
-back to differences in the selection of hypotheses from the posterior. Griffith \& Kalish's
-convergence to the prior holds for learners that sample from the posterior. More deterministic
-strategies such as the adoption of the type with the highest posterior probability, so-called
-{\it maximum a posterior estimation} (MAP), increase the influence of both the prior and the
-data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we use a parameter
-$l\ge1$ to modulate between posterior sampling and the MAP strategy. When $l = 1$ learners
-sample from the posterior. The learners' propensity to maximize the posterior grows as $l$
-increases.
+be magnified by exposing learners to only small data samples
+\citep[e.g. in][]{brighton:2002}. On the other, \citeposs{griffiths+kalish:2007} mathematical
+characterization showed that iterated learning converges to the prior in the limit, i.e., that
+the resulting distribution over languages corresponds to the learners' prior distribution and
+is not influenced by the amount of input given to them. This difference in predictions can be
+traced back to differences in the selection of hypotheses from the posterior. Griffith \&
+Kalish's convergence to the prior holds for learners that sample from the posterior. More
+deterministic strategies such as the adoption of the type with the highest posterior
+probability, so-called {\it maximum a posterior estimation} (MAP), increase the influence of
+both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we
+use a parameter $l\ge1$ to modulate between posterior sampling and the MAP strategy. When
+$l = 1$ learners sample from the posterior. The learners' propensity to maximize the posterior
+grows as $l$ increases.
 
@@
 recover the true types that generated a given sequence with higher probability. The mutation
-matrix $Q$ of the replicator mutator dynamics in (\ref{eq:RMD_discrete}) can then be defined as
+matrix $Q$ of the replicator mutator dynamic in (\ref{eq:RMD_discrete}) can then be defined as
 follows: $Q_{ji}$ is the probability that a learner acquires type $i$ when learning from an
@@
 of communicative efficiency and iterated Bayesian learning. Their interaction is described by
-the discrete time replicator mutator dynamics in (\ref{eq:RMD_discrete}), repeated here:
+the discrete time replicator mutator dynamic in (\ref{eq:RMD_discrete}), repeated here:
 \begin{align*}
@@
 
-\section{Scalar implicatures}\label{sec:si-case-study}
-%
-Scalar implicatures are a particularly well-studied type of conventional pragmatic inferences.
-They are licensed for groups of expressions ordered in terms of informativity, here understood as an entailment induced order.
-For instance, {\em some} is entailed by {\em all}.
-If it were true that ``Chris owns all of Johnny Cash's albums'', it would also be true that ``Chris owns some of Johnny Cash's albums''.
-However, while weaker expressions such as {\em some} are truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what their use is normally taken to convey.
-Instead, the use of a less informative expression when a more informative one could have been used can license a defeasible inference that stronger alternatives do not hold (cf.
-\citealt{horn:1972,gazdar:1979}).
-That is, a hearer who assumes the speaker to be able and willing to provide all relevant information can infer that stronger alternatives do not hold because the speaker used a weaker alternative instead.
-In this way, ``Chris owns some of Johnny Cash's albums'' is strengthened to convey that he owns {\em some but not all} albums.
-A bound that rules out stronger alternatives is thusly not codified in the lexical meaning of weak alternatives but instead pragmatically supplied.
+\section{Scalar implicatures}
+\label{sec:si-case-study}
+
+Scalar implicatures are a particularly well-studied type of conventional pragmatic
+inferences. They are licensed for groups of expressions ordered in terms of informativity, here
+understood as an entailment induced order. For instance, {\em some} is entailed by {\em
+  all}. If it were true that ``Chris owns all of Johnny Cash's albums'', it would also be true
+that ``Chris owns some of Johnny Cash's albums''. However, while weaker expressions such as
+{\em some} are truth-conditionally compatible with stronger alternatives such as {\em all},
+this is not what their use is normally taken to convey. Instead, the use of a less informative
+expression when a more informative one could have been used can license a defeasible inference
+that stronger alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). That is, a hearer
+who assumes the speaker to be able and willing to provide all relevant information can infer
+that stronger alternatives do not hold because the speaker used a weaker alternative
+instead. In this way, ``Chris owns some of Johnny Cash's albums'' is strengthened to convey
+that he owns {\em some but not all} albums. A bound that rules out stronger alternatives is
+thusly not codified in the lexical meaning of weak alternatives but instead pragmatically
+supplied.
 

### 2017-03-23 paper/paper-v3.tex
revisions up to general discussion

@@
 frequency $x_i$ before the step, its fitness $f_i$, and the probability that a learner infers
-$i$ when observing the behavior of a type-$j$ agent. Fitness-based selection can be thought of
-as biological (fitness as expected relative number of offspring) or cultural (fitness as 
-likelihood of being imitated or repeated). The types that the dynamic operates on are pairs
-consisting of a lexicon and a pragmatic use pattern. A type's expressivity depends on its
-communicative efficiency within a population while its learnability depends on the fidelity by
-which it is inferred by new generations of learners. The learners' task is consequently to
-perform a joint inference over types of linguistic behavior and lexical meaning.
-
-\section{Scalar implicatures}
+$i$ when observing the behavior of a type-$j$ agent. Fitness-based selection is here thought of
+not as biological (fitness as expected relative number of offspring) but cultural (fitness as
+likelihood of being imitated or repeated) evolution, since the types that the dynamic operates
+on are pairs consisting of a lexicon and a pragmatic use pattern. A type's expressivity depends
+on its communicative efficiency within a population while its learnability depends on the
+fidelity by which it is inferred by new generations of learners. The learners' task is
+consequently to perform a joint inference over types of linguistic behavior and lexical
+meaning.
+
+The model has three parameters: $\lambda$ regulates the degree to which pragmatic speakers
+choose messages that maximize the chance of communicative success; $k$ is the number of
+observations for each language learner; $l$ regulates where the learners' inference behavior
+lies on a spectrum from probability matching to choice of the most likely parent type.
+
+\section{Case study: scalar implicatures}
 \label{sec:si-case-study}
 
-Scalar implicatures are a particularly well-studied type of conventional pragmatic
-inferences. They are licensed for groups of expressions ordered in terms of informativity, here
-understood as an entailment induced order. For instance, {\em some} is entailed by {\em
-  all}. If it were true that ``Chris owns all of Johnny Cash's albums'', it would also be true
-that ``Chris owns some of Johnny Cash's albums''. However, while weaker expressions such as
-{\em some} are truth-conditionally compatible with stronger alternatives such as {\em all},
-this is not what their use is normally taken to convey. Instead, the use of a less informative
-expression when a more informative one could have been used can license a defeasible inference
-that stronger alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). That is, a hearer
-who assumes the speaker to be able and willing to provide all relevant information can infer
-that stronger alternatives do not hold because the speaker used a weaker alternative
-instead. In this way, ``Chris owns some of Johnny Cash's albums'' is strengthened to convey
-that he owns {\em some but not all} albums. A bound that rules out stronger alternatives is
-thusly not codified in the lexical meaning of weak alternatives but instead pragmatically
-supplied.
-
-As noted earlier, this kind of strengthening is captured by the linguistic behavior of pragmatic types introduced in \S\ref{sec:languages+use}: A pragmatic hearer who reasons about the use of a message involving a weak scalar alternative will associate it more with a state in which stronger alternatives do not hold.
-This is so because a rational speaker would use a more informative message when in such a state.
-Conversely, a pragmatic speaker will reason about her interlocutor's expected interpretation and use the messages at her disposition accordingly as to maximize communicative success.
-
-Our initial question about the division of labor between semantics and pragmatics can be narrowed to the case of scalar implicatures by asking for a justification for the lack of lexical upper-bounds in weak scalar alternatives.
+The model of the previous section formalizes the evolutionary competition between different
+pairs of lexical concepts and ways of using them. This section looks at a concrete case study
+on scalar implicatures. Concretely, we will engage in the following formal thought experiment:
+if a population of language users could freely combine different lexica with different
+pragmatic strategies, what there the conditions under which the majority view of scalar
+implicatures could have evolved?
+
+Recall that the majority view is that scalar implicatures are non-lexicalized pragmatic
+enrichments. Scalar implicature triggers like \emph{some}, \emph{warm} or \emph{may} are
+semantically weak expressions for which logically stronger expressions are salient, e.g.,
+\emph{all}, \emph{hot} or \emph{must}. For instance, {\em some} is entailed by {\em all}. If
+the sentence ``Chris owns all of Johnny Cash's albums'' is true, then ``Chris owns some of
+Johnny Cash's albums'' is also true. However, while weaker expressions such as {\em some} are
+truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what
+their use is normally taken to convey. Instead, the use of a less informative expression when a
+more informative one could have been used can license a defeasible inference that stronger
+alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). That is, a hearer who assumes
+the speaker to be able and willing to provide all relevant information can infer that stronger
+alternatives do not hold because the speaker used a weaker alternative instead. In this way,
+``Chris owns some of Johnny Cash's albums'' is strengthened to convey that she owns {\em some
+  but not all} albums. The upper bound, so the majority view, is a pragmatic inference but not
+part of the conventional meaning. \mf{I wonder if this paragraph is really still necessary
+  after the introduction?}
+
+In the following we consider a specific application of the model from Section~\ref{sec:model}
+which allows to address the question if or when scalar inferences might (not) lexicalize. We
+consider what is perhaps one of the simplest non-trivial setups that speak to this matter and
+reflect on its limitations in Section~\ref{sec:discussion}. The setup is introduced in
+Section~\ref{sec:setup:-stat-conc}. Section~\ref{sec:simulation-results} describes the simulations
+and their results.
+
+% Our initial question about the division of labor between semantics and pragmatics can be narrowed to the case of scalar implicatures by asking for a justification for the lack of lexical upper-bounds in weak scalar alternatives.
 That is, we ask why lexical meanings that lack upper-bounds and convey it pragmatically are regularly selected for over alternatives such as that of codifying the bound semantically.
@@
 
-We see two main explanations for the lack of upper-bounds in the lexical meaning of weak scalar expressions.
-The first is that their truth-conditional compatibility with stronger expressions endows them with a broader range of applicability by allowing them to occur in contexts in which their upper-bounded reading is absent.
-This can happen when embedded in downward-entailing contexts, when the speaker is likely uncertain about whether the upper bounded reading is true, or when the distinction between an upper-bounded reading and the simple, only lower-bounded reading, is not relevant.
-For instance, if for all the speaker knows Chris owns some of Johnny Cash's albums but she does not know whether he owns all, then the use of {\em some} lacking an upper-bound succinctly conveys her uncertainty.
-This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: Should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then this is preferred over expressing it overtly through longer expressions, e.g., by saying {\em some but not all} explicitly.
-Importantly, although morphosyntactic disambiguation may be dispreferred due to its relative length and complexity \citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and override contextual cues that might otherwise mislead the hearer.
-In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so.
-Compare this with a hypothetical language that lexicalizes two expressions for each weak scalar expression -- one with and one lacking an upper-bound.
-We see four conditions along this functionalist explanation that may pressure languages for English-like semantics over this alternative.
-First, contextual cues are very reliable.
-Second, morphosyntactic disambiguation is seldom necessary.
-Third, morphosyntactic disambiguation is only marginally dispreferred.
-Fourth, larger lexica are costly.
-Overall, neither condition seems convincing as a central explanatory device for such a widespread phenomenon.
-The first two conditions put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of unambiguous communication.
-It is likely that human language users are very good at retrieving cues from context, but to stipulate that they are so good as to undercut the benefit of safe communication provided by this hypothetical alternative strikes us as too strong of an assumption.
-As for the third and fourth condition, these seem mostly like technical solutions without a proper empirical basis.
-
-Instead, the systematicity and typological spread of scalar implicatures together with the observation that monomorphemic expressions that lexically rule out stronger alternatives are unattested across languages (\citealt[252-267]{horn:1984}, \citealt{horn:1972,traugott:2004,vdAuwera:2010}) suggests that other forces may be at play.
-In what follows we investigate the predictions of our model under the assumption that the lack of lexicalization of scalar inferences may be accounted for by the relative representational simplicity of lexical meanings lacking an upper-bound over those that explicitly codify it.
-This difference is reflected in a learning bias towards more compressed lexical representations.
-That is, in a preference of learners for simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
-While we do not want to argue that functional aspects as the ones discussed above do not play a role, we do see a clear benefit in exploring whether matters of transmission biases would not give us additional explanatory leverage.
-
-\subsection{Analysis}
-The dynamics are initialized with an arbitrary distribution over types, constituting a population's first generation.
-In the following we report the outcome of their development under different parameters and pressures after $50$ generations.
-This ensures that these outcomes correspond to developmental plateaus after which no noteworthy change is registered.
-As specified in \S\ref{sec:learnability}, the mutation matrix $Q$ can be obtained by considering all possible state-message sequences of length $k$.
-Given that this is intractable for large $k$, the sets of data learners are exposed to are approximated by sampling $250$ $k$-length sequences from each type's production probabilities.
-
-We begin setting the stage by defining the learning prior and the set of types under consideration.
-Drawing from our preceding discussion, functional pressure on successful communication combined with learning pressures in the form of a bias that favors simpler semantics may lead to the selection of $\Llack$-like semantics and to a division of labor between semantics and pragmatics.
-We first showcase the effects that expressivity and learnability have in isolation before turning to those resulting from their combination.
+\subsection{Setup}
+\label{sec:setup:-stat-conc}
+
+To fill the model from Section~\ref{sec:model} with life, we need to specify the sets of
+states, messages and lexica. Additionally, we want to explore the effects of a learning bias in
+favor of simple lexical concepts. One way of motivating and formalizing such a bias is
+introduced below.
+
+\subsubsection{States, messages, concepts and Lexica}
+
+Consider a state space with three states $\States = \set{\snone, \ssome, \sall}$ and think of
+it as a partition of possible worlds into cells where none, some or all of the $A$s are $B$s,
+for some arbitrary fixed predicates $A$ and $B$. Eight concepts can be distinguished based on
+their truth or falsity in three world states, six of which are not contradictory or
+tautological.
+
+A lexicon $L$ is a mapping $\Messgs \rightarrow C$ from messages to concepts. With three
+messages there are $6^3 = 216$ possible lexica. Some assign the same concept to more than one
+message and others lexicalize the same concepts but associate them with different messages. Out
+of these possible lexica, three kinds are of particular relevance. First, lexica that assign
+the same concept to more than one message. Such lexica lack in expressivity but may be favored
+by particular learning biases nonetheless. Second, lexica that conventionalize upper-bounds to
+realize a (quasi-)partition of the state space. Finally, lexically that do not lexicalize an
+upper bound but allow it to be conveyed pragmatically due to the presence of a stronger lexical
+item. There are six lexica of the second kind and six of the third. The following three lexica
+exemplify each kind:
+\begin{align*}
+  \begin{blockarray}{lccc}
+    & & \underline{\Lall} & \\
+    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[ccc]}
+     \mygray{\snone}  & 0 & 0 & 0\\
+     \mygray{\ssome}  & 0 & 0 & 0\\
+    \mygray{\sall}   & 1 & 1 & 1\\
+    \end{block}
+% \\ & & P(\Lall)\rlap{ $\propto 9$} & \\
+  \end{blockarray} & &
+ \begin{blockarray}{lccc}
+    & & \underline{\Lbound} & \\
+    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[ccc]}
+       & 1 & 0 & 0\\
+       & 0 & 1 & 0\\
+       & 0 & 0 & 1\\
+    \end{block}
+%  \\& & P(\Lbound)\rlap{ $\propto 96$} & \\
+  \end{blockarray} & &
+  \begin{blockarray}{lccc}
+    & & \underline{\Llack} & \\
+    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
+    \begin{block}{l[ccc]}
+       & 1 & 0 & 0\\
+       & 0 & 1 & 0\\
+       & 0 & 1 & 1\\
+    \end{block}
+%   \\& & P(\Llack)\rlap{ $\propto 48$} & \\
+  \end{blockarray}
+\end{align*}
+Lexicon $\Lall$ is clearly bad for communication: all message and interpretation choices will
+be equally likely for all types; no information will be conveyed with it. Users of $\Lbound$
+need not resort to pragmatic reasoning to convey an upper-bound with weak scalar
+expressions. Instead, this information is already provided by the semantics of their lexicon,
+giving them an advantage in terms of expressivity. Types with such lexica may be expected to
+generally achieve higher communicative success than pragmatic users with a lexicon of type
+$\Llack$ (see Section~\ref{sec:languages+use}).
+
+Recall that types are a combination of a lexicon and a manner of language use. We analyze the
+model's predictions in populations of types with one of the two behaviors introduced earlier;
+literal or pragmatic. The former correspond to level $0$ reasoners and the latter to ones of
+level $1$. Accordingly, we consider a total of $432$ types. Six are variants of pragmatic
+language users with $\Llack$-like lexica. These will be our target types, because they
+represent lexica and language use that conform to the majority view of scalar
+implicatures. Twelve types are either literal or pragmatic speakers with lexical of the
+$\Lbound$ kind. These will be the target types main competitors in evolutionary competition.
 
 \subsubsection{An inductive learning bias for semantic simplicity}
-There is a growing effort to develop empirically  testable representational  languages that allow for the measure of semantic complexity.
-For instance, so-called {\em languages of thought} (LOT) have been put to test in various rational probabilistic models that show encouraging results (see, e.g., \citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012a} and \citealt{piantadosi+jacobs:2016} for recent discussion).
-At its core, a LOT defines a set of operations and composition rules from which lexical meaning can be derived.
-As a first approximation and for the sake of concreteness, we follow this approach to make the preference of learners for simpler semantic representations precise and define the complexity of a representation as a function of its derivation cost in a weighed generative LOT.
-
-Our toy grammar of concepts is given in Table~\ref{tab:grammar}.
-This grammar uses basic set-theoretic operations to form expressions which can be evaluated as true or false in different world states such as $\ssome$ or $\sall$.
-Applications of generative  rules have a cost attached to them.
-Here we simply assume that Boolean combinations of concepts are more complex than atomic concepts and that otherwise each rule application adds the same cost unit.
+\label{sec:an-induct-learn}
+
+There is a growing effort to develop empirically testable representational languages that allow
+for the measure of semantic complexity. For instance, so-called {\em languages of thought}
+(LOT) have been put to test in various rational probabilistic models that show encouraging
+results (see, e.g., \citealt{katz+etal:2008, piantadosi+etal:underreview,
+  piantadosi+etal:2012a} and \citealt{piantadosi+jacobs:2016} for recent discussion). At its
+core, a LOT defines a set of operations and composition rules from which concepts can be
+derived. As a first approximation and for the sake of concreteness, we follow this approach to
+make the preference of learners for simpler semantic representations precise and define the
+complexity of a representation as a function of its derivation cost in a weighed generative
+LOT.
+
 \begin{table}
@@
 
-The prior probability of a type is just the prior probability of its lexicon.
-The prior of a lexicon is a function of the complexity of the lexical representations in its image set.
-As motivated above, lexica that use simpler concepts are \emph{a priori} more likely.
-A simple way of defining such priors over lexica is:
-\begin{align*}
-  P(L)  & \propto \prod_{c \in Im(L)} P(c)   \ \text{, with} & 
-  P(c) & \propto \max_{c'}Compl(c') - Compl(c) + 1\,,
-\end{align*}
-
-where $Compl(c)$ is the complexity of the minimal derivation length of $c$ according to the LOT-grammar.
-There are many ways to define priors over lexica (see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a}) but the key assumption here, common to all of them, is that simple representational expressions should be favored over more complex ones.
-We should stress that these details -- from the generative grammar to its complexity measure -- are to be regarded as an instantiation of our general assumptions instead of as empirical claims in their own right.
-
-\subsubsection{Lexica, signaling behavior \& types}
-We consider a state space with three states $\States = \set{\snone, \ssome, \sall}$.
-\tb{Should we say something about the inclusion of $\snone$?
-It may be more natural to introduce this state earlier but may unnecessarily complicate the exposition of signaling behavior in the previous section.} This space can be
-thought of as a partition of possible worlds into cells where none, some or all of the $A$s are
-$B$s, for some arbitrary fixed predicates $A$ and $B$. Eight concepts can be distinguished
-based on their truth or falsity in three world states, six of which are not contradictory
- or tautological. These are listed with mnemonic names in
-Table~\ref{tab:concepts} together with their complexity according to the grammar in Table \ref{tab:grammar}.
+Our toy grammar of concepts is given in Table~\ref{tab:grammar}. This grammar uses basic
+set-theoretic operations to form expressions which can be evaluated as true or false in states
+$\snone$, $\ssome$ or $\sall$ from above. Applications of generative rules have a cost attached
+to them. Here we simply assume that Boolean combinations of concepts are more complex than
+atomic concepts and that otherwise each rule application adds the same cost
+unit. Table~\ref{tab:concepts} lists all six concepts relevant here, their truth conditions and
+the simplest formula that expresses this concept from the grammar in Table~\ref{tab:grammar}.
 
@@
 
-A lexicon $L$ is a mapping $\Messgs \rightarrow C$ from messages to concepts. With three
-messages there are $6^3 = 216$ possible lexica.
-Some assign the same concept to more than one message and others lexicalize the same concepts but associate them with different messages.
-The main results reported in the following do not hinge on the inclusion of these lexica but this unaltered space does better illustrate the influence and differences of the linguistic pressures we consider.
-Additionally, the competition of different languages that lexicalize the same concepts through different forms allows us to take serious the problem of only one being selected for even though multiple alternatives of equal expressivity {\em and} learnability exist.
-
-Out of these possible lexica, three kinds are of particular relevance.
-First, lexica that assign the same concept to more than one message.
-Such lexica lack in expressivity but may be favored by the learning prior should they lexicalize simple concepts.
-Second, lexica that conventionalize upper-bounds to realize a (quasi-)partition of the state space.
-As discussed before in relation to $\Lbound$, users of these lexica need not resort to pragmatic reasoning to convey an upper-bound with weak scalar expressions.
-Instead, this information is already provided by the semantics of their lexicon, giving them an advantage in terms of expressivity.
-Types with such lexica are expected to be the main contenders of the natural language like semantics that do not lexicalize an upper-bound for weak scalar expressions.
-Third and lastly, $\Llack$-style lexica that, paired with mutual reasoning, can convey an upper-bound pragmatically.
-There are six lexica of the second kind and six of the third.
-The following three lexica exemplify each kind, with $\Lall$ being a lexicon of the first kind.
-This lexicon is not optimal for communication as it assigns all its messages to the same state, $\sall$.
-This inevitably leads to a communicative disadvantage in its use.
-However, it may have a learnability advantage over our target $\Llack$ and its competitor $\Lbound$ in virtue of being the simplest lexicon, making it a priori favored by learners.
-
-\begin{align*}
-  \begin{blockarray}{lccc}
-    & & \underline{\Lall} & \\
-    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
-    \begin{block}{l[ccc]}
-     \mygray{\snone}  & 0 & 0 & 0\\
-     \mygray{\ssome}  & 0 & 0 & 0\\
-    \mygray{\sall}   & 1 & 1 & 1\\
-    \end{block}
-% \\ & & P(\Lall)\rlap{ $\propto 9$} & \\
-  \end{blockarray} & &
- \begin{blockarray}{lccc}
-    & & \underline{\Lbound} & \\
-    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
-    \begin{block}{l[ccc]}
-       & 1 & 0 & 0\\
-       & 0 & 1 & 0\\
-       & 0 & 0 & 1\\
-    \end{block}
-%  \\& & P(\Lbound)\rlap{ $\propto 96$} & \\
-  \end{blockarray} & &
-  \begin{blockarray}{lccc}
-    & & \underline{\Llack} & \\
-    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
-    \begin{block}{l[ccc]}
-       & 1 & 0 & 0\\
-       & 0 & 1 & 0\\
-       & 0 & 1 & 1\\
-    \end{block}
-%   \\& & P(\Llack)\rlap{ $\propto 48$} & \\
-  \end{blockarray}
-\end{align*}
-
-Recall that types are a combination of a lexicon and a manner of language use.
-We analyze the model's predictions in populations of types with one of the two behaviors introduced earlier; literal or pragmatic.
-The former correspond to level $0$ reasoners and the latter to ones of level $1$.
-Higher level reasoning is not required to derive scalar implicatures from these lexica, nor do they leave room for substantial pragmatic refinement.
-Accordingly, we consider a total of $432$ types.
-Six are variants of our pragmatic $\Llack$ target and twelve are either literal or pragmatic speakers of variants of its main contender, $\Lbound$.
-
-%Combining a linguistic behavior with each of these $6$ lexica yields a total of $12$ distinct types.
-, such as $L_4$, will produce speaker behavior that is {\em almost} indistinguishable from that of a language that lacks upper-bounds, but with pragmatic users, such as $L_5$.
-Almost, because there may be slight differences between the probability with which speakers would (erroneously) use a semantically false description and the probability with which speakers would (erroneously) use a pragmatically suboptimal description.
-Due to this possibly marginal difference between pragmatic $L_4$ and $L_5$, the selection of one type over the other is expected to mainly depend on how well each can be transmitted to new learners.
-Things are less clear for literal $L_5$ contrasted with literal/pragmatic $L_4$.
-The former has a learning advantage when learners are biased against upper-bounds but is expected to fare worse in communicative terms.
-
-\subsubsection{Expressivity only} 
+The complexity measure for lexical concepts from Table~\ref{tab:concepts} is used to define a
+learning bias that favor simpler concepts over more complex ones. The prior probability of a
+type is just the prior probability of its lexicon. The prior of a lexicon is a function of the
+complexity of the lexical representations in its image set. Lexica with simpler concepts have a
+higher prior. One simple way of defining such priors over lexica (and thereby types) is:
+\begin{align*}
+  P(L)  & \propto \prod_{c \in Im(L)} P(c)   \ \text{, with} & 
+  P(c) & \propto \max_{c'}Compl(c') - Compl(c) + 1\,,
+\end{align*}
+where $Compl(c)$ is the complexity of the minimal derivation length of $c$ according to the
+LOT-grammar (see Table~\ref{tab:concepts}). There are many ways to define priors over lexica
+(see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a}) but the key assumption here,
+common to all of them, is that simple representational expressions should be favored over more
+complex ones. We should stress that these details -- from the generative grammar to its
+complexity measure -- are to be regarded as one convenient operationalization of one general
+approach to explicating learning biases; this is not a commitment that this general approach is
+necessarily superior or that, within it, this particular instrumentalization is the single most
+plausible.
+
+\subsection{Simulation results}
+\label{sec:simulation-results}
+
+Recall that there are three parameters: soft-max parameter $\lambda$ affects how strongly
+pragmatic speakers favor more informative messages; the bottleneck size $k$ influences how
+faithfully learners can identify their teacher type; $l$ defines the learners disposition
+towards choosing the most likely teacher type from the posterior distribution. We expect that
+competitor types (types with lexica of the kind $\Lbound$) have a fitness advantage over target
+types (pragmatic agents with lexica of the kind $\Llack$), especially for very low levels of
+$\lambda$. Selection based on fitness alone may therefore not lead to prevalence of target
+types in the population. On the other hand, lexica of type $\Llack$ are simpler than those of
+type $\Lbound$ by the postulated measure from above. This may make them more likely to be
+adopted by learners, especially when $k$ is low so that teacher types are relatively
+indistinguishable based on their behavior in general and when $l$ is high. Still, lexica of the
+kind $\Lall$ are in turn more likely \emph{a priori} than lexica of the kind
+$\Llack$. Simulation results will shed light on the question whether target types can emerge
+and for which parameter constellations.
+
+To better understand the workings a pressure towards communicative efficiency and a pressure
+towards learnability, we look at the behavior of the replicator and mutator step first in
+isolation, and then in combination. All simulation runs are initialized with an arbitrary
+distribution over types, constituting a population's first generation. All reported results are
+the outcome of 50 update steps for each different constellation of parameter values. This
+ensures that these outcomes correspond to developmental plateaus in which change is, if not
+absent, then at least very slow, In other words, even if the resulting states do not correspond to
+an eventual attracting state, they characterize long almost stationary states in which the
+system remains for a very long time. As specified in \S\ref{sec:learnability}, the mutation
+matrix $Q$ can be obtained by considering all possible state-message sequences of length
+$k$. Given that this is intractable for large $k$, the sets of data learners are exposed to are
+approximated by sampling $250$ $k$-length sequences from each type's production probabilities.
+
+\subsubsection{Replication only: selection based on expressivity and communicative success} 
 \begin{figure}
@@
 
-Expressivity is sensitive to $\lambda$ as it influences signaling behavior.
-This is showcased in Figure \ref{fig:only-R}, which depicts the proportion of pragmatic $\Llack$-style speakers in $5$ independent populations after $50$ generations, as well as that of the type with the highest proportion in a given population, the incumbent.
-Low $\lambda$ levels communicative differences between types, as even those that can differentiate between all states semantically or pragmatically do not always maximize expected communicative success.
-The outcome is that of a mixed population that allows for the survival of competing types.
-Conversely, higher $\lambda$ promotes more rational linguistic behavior, widening the gap in expressivity between types and promoting more homogeneous populations.
-As suggested by Figure \ref{fig:only-R}, the incumbent in most populations is not one of the six pragmatic $\Llack$-style types.
-That is, a pressure only for expressivity does not lend a justification to the systematic prevalence of a lack of upper-bounds under any $\lambda$-value.
-For instance, in $1000$ independent populations with $\lambda = 20$ pragmatic $\Llack$ types only made up $11$ cases of incumbency, corresponding to a mean proportion of $0.003$ across populations.
-By contrast, $913$ incumbents were $\Lbound$-style users with close to an even share between literal ($454$) and pragmatic types ($459$), corresponding to a mean proportion of about $.48$ taken together.
-%Across types and populations incumbents make up $.50$ of the mean outcome.
-
-\subsubsection{Learnability only}
+As expected, selection based on communicative success is sensitive to $\lambda$ as it
+influences signaling behavior. This is showcased in Figure \ref{fig:only-R}, which depicts the
+proportion of target types in $5$ independent populations after $50$ generations, as well as
+that of the type with the highest proportion in a given population, the incumbent. With low
+$\lambda$ many types have very similar behavior, so that evolutionary selection lacks grip and
+becomes very slow. The result is a very long transition with near stagnancy in a rather
+homogeneous population with many types. Conversely, higher $\lambda$ promotes more rational
+linguistic behavior, widening the gap in expressivity between types and promoting more
+homogeneous populations. As suggested by Figure \ref{fig:only-R}, the incumbent in most
+populations is not one of the six pragmatic $\Llack$-style types. That is, a pressure only for
+expressivity does not lend a justification to the systematic prevalence of a lack of
+upper-bounds under any $\lambda$-value. For instance, in $1000$ independent populations with
+$\lambda = 20$ pragmatic $\Llack$ types only made up $11$ cases of incumbency, corresponding to
+a mean proportion of $0.003$ across populations. By contrast, $913$ incumbents were
+$\Lbound$-style users with close to an even share between literal ($454$) and pragmatic types
+($459$), corresponding to a mean proportion of about $.48$ taken
+together. %Across types and populations incumbents make up $.50$ of the mean outcome.
+
+\subsubsection{Iterated learning only}
 
@@
 
-The effect of iterated learning without a pressure for expressivity using either posterior sampling ($l = 1$) or a stronger tendency towards posterior maximization ($l = 15$) is shown in Figure \ref{fig:only-M} together with the prior over types.
-The prior shows that while users of $\Llack$ are not the most favored by the inductive bias (compared, e.g., to $\Lall$) they are nevertheless more advantaged than others, such as $\Lbound$, in virtue of the relatively simple semantics they conventionalize.
-Crucially, $\Llack$ enables its users to convey each state with a single message when combined with pragmatic reasoning provided sufficiently high $\lambda$.
-This makes it less likely to be confused with other types if the learning data is not too sparse ($k \geq 5$).
-Put differently, learners have a higher propensity to infer pragmatic $\Llack$ over types that might be hard or impossible to distinguish it from based on overt linguistic behavior, such as $\Lbound$, and $\Llack$ is also less likely to be mistaken with types with different observable behavior because its pragmatic use approximates a one-to-one form-meaning mapping.
-As a consequence, a stronger propensity to maximize the posterior increases their proportion in the population.
-
-However, in contrast to a pressure only for expressivity with high $\lambda$, learnability alone does not succeed in selecting for a single type of a kind.
-For instance, for only one of the six pragmatic $\Llack$ types.
-Instead, it leads to their coexistence.
-Each is passed on to next generation with the same faithfulness and, differently from a pressure for communicative success, they do not stand in competition with each other.
-As implicit in Figure \ref{fig:only-M}, in $1000$ independent populations all incumbents were pragmatic $\Llack$ users provided sufficiently high $\lambda$, with each reaching approximately the same proportion of users in the population.
-As with expressivity only, low values of $\lambda$ make the differences in observable behavior across types less pronounced and therefore reflect the learners' inductive bias more faithfully, favoring functionally deficient but a priori preferred types such as those that use $\Lall$.
-A pressure for learnability alone may consequently lead to a spread of communicatively suboptimal types that are easier to learn.
-In the extreme, when $l = 1$ and $\lambda = 1$ all of $1000$ independent populations had users of $\Lall$ as incumbents.
-
-\subsubsection{Expressivity and learnability}
-The separate inspection of each pressure not only showcases the contribution of the model's components but also highlights some of their broader implications for the emergence of a division of labor between semantics and pragmatics.
-Neither dynamic on its own comes close to converging to a monomorphic population of pragmatic $\Llack$ speakers.
-When pressured toward expressivity, the slight communicative advantage of $\Lbound$'s lexicalization of an upper-bound over pragmatic use of $\Llack$ leads to the former's selection.
-When pressured toward learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more complex alternatives such as $\Lbound$.
-However, on its own learnability does not foment the propagation of these types across the population.
-What is more, it leads to heterogeneous populations that foster an equal share of types of a kind, such as different pragmatic $\Llack$ types.
+The effect of iterated learning without a pressure for expressivity using either posterior
+sampling ($l = 1$) or a stronger tendency towards posterior maximization ($l = 15$) is shown in
+Figure \ref{fig:only-M} together with the prior over types. The prior shows that while users of
+$\Llack$ are not the most favored by the inductive bias (compared, e.g., to $\Lall$) they are
+nevertheless more advantaged than others, such as $\Lbound$, in virtue of the relatively simple
+semantics they conventionalize. Crucially, $\Llack$ enables its users to convey each state with
+a single message when combined with pragmatic reasoning provided sufficiently high
+$\lambda$. This makes it less likely to be confused with other types if the learning data is
+not too sparse ($k \geq 5$). Put differently, learners have a higher propensity to infer
+pragmatic $\Llack$ over types that might be hard or impossible to distinguish it based on overt
+linguistic behavior, such as $\Lbound$. Moreover, $\Llack$ is also less likely to be confused
+with types with different observable behavior because its pragmatic use approximates a
+one-to-one form-meaning mapping. As a consequence, a stronger propensity to maximize the
+posterior increases their proportion in the population.
+
+However, in contrast to a pressure only for expressivity with high $\lambda$, learnability
+alone does not succeed in selecting for a single prevalent type. Indeed, all six target types
+tend to coexist at roughly equal proportion. Each is passed on to next generation with the same
+faithfulness and, differently from a pressure for communicative success, they do not stand in
+competition with each other. As implicit in Figure \ref{fig:only-M}, in $1000$ independent
+populations all incumbents were pragmatic $\Llack$ users provided sufficiently high $\lambda$,
+with each reaching approximately the same proportion of users in the population. As with
+expressivity only, low values of $\lambda$ make the differences in observable behavior across
+types less pronounced and therefore reflect the learners' inductive bias more faithfully,
+favoring functionally deficient but a priori preferred types such as those that use $\Lall$. A
+pressure for learnability alone may consequently lead to a spread of communicatively suboptimal
+types that are easier to learn. In the extreme, when $l = 1$ and $\lambda = 1$ all of $1000$
+independent populations had users of $\Lall$ as incumbents.
+
+\subsubsection{Combining pressures of expressivity and learnability}
+
+Pressures for communicative success or learnability are not sufficient on their own to have a
+single target type dominate the population. When pressured toward expressivity, the slight
+communicative advantage of $\Lbound$ users leads to the its prevalence. When pressured toward
+learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more
+complex alternatives such as $\Lbound$. However, on its own learnability does not foment the
+propagation of these types across the population. What is more, it leads to heterogeneous
+populations that foster an equal share of types of a kind, such as different pragmatic $\Llack$
+types.
 
@@
 
-To recapitulate, other than the involvement of both pressures, the resulting proportion of pragmatic $\Llack$ speakers primarily hinges on three factors.
-First, the degree to which agents try to maximize communicative success.
-This plays a role both in expressivity as well as in producing data that allows learners to discriminate this type from others.
-Second, the inductive bias, which leads learners to prefer simpler over more complex semantic representations in acquisition.
-Lastly, the posterior parameter, as it magnifies the effects of the learning bias in tandem with replication.
-
-\subsection{Discussion}
-Under the assumption of a learning bias for simpler lexical representations, our results suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome communicative pressures and be majoritarily selected for in a population.
-This prediction hinges on three assumptions.
-First, that language is pressured toward both expressivity and learnability.
-Second, that language users favor expressions/interpretations expected to have a higher likelihood of being understood/meant over ones with a lower likelihood.
-Lastly, that learners prefer simpler over more complex lexical representations and exhibit a tendency towards the acquisition of the type that best explains the learning data.
-This outcome is particularly encouraging in light of other advantages a lack of semantic upper-bounds may confer, as discussed at the beginning of this section.
-
-A lack of upper-bounds in the lexical meaning of weak scalar expressions constitutes the majority view in the literature.
-However, it not clear to what extent other types should be present in the final population.
-It seems reasonable to expect functionally suboptimal lexica such as $\Lall$ to be ruled out but matters are different for $\Lbound$.
-The prediction that natural language communities are homogeneous or that speakers may entertain $\Llack$-like semantics for one scalar expression and $\Lbound$-like semantics for another is not implausible (cf.
-\citealt{franke+degen:2016}).
-Alternatively, a stronger tendency for posterior maximization in learning has to be assumed.
-This empirical issue relates to other two aspects left undiscussed: disadvantages of pragmatic reasoning and the effect of state frequencies on the fossilization of pragmatic inferences.
-We tacitly assumed pragmatic reasoning to come at no cost.
-However, there is experimental evidence that suggests  that the pragmatic derivation of upper-bounds costs effort and takes additional processing time (cf.
-\citealt{deNeys+schaeken:2007, huang+snedeker:2009}).
-This raises the question at which point such usage-based cost undercuts the learnability advantage of simpler semantic representations.
-Should cost play a role, then its effect is bound to depend on the frequency with which a given scalar expression is used.
-It may be that frequently drawn scalar implicatures fossilize to avoid cost whereas infrequent ones are derived on-line to avoid complexity in acquisition.
-This also opens a possible venue to address the preceding question about the expected presence of $\Lbound$-like semantics in a population, but further empirical evidence is needed to assess these matters.
+To recapitulate, other than the involvement of pressure on both expressivity and learnability,
+the resulting proportion of pragmatic $\Llack$ speakers primarily hinges on three
+factors. First, the degree to which agents try to maximize communicative success. This plays a
+role both in expressivity as well as in producing data that allows learners to discriminate
+this type from others. Second, the inductive bias, which leads learners to prefer simpler over
+more complex semantic representations in acquisition. Lastly, the posterior parameter, as it
+magnifies the effects of the learning bias in tandem with replication.
+
+\subsection{Summary}
+
+Under the assumption of a learning bias for simpler lexical representations, our results
+suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome
+communicative pressures and come to dominate a population. This prediction hinges on three
+assumptions. First, that language is pressured toward both expressivity and
+learnability. Second, that language users favor expressions/interpretations expected to have a
+higher likelihood of being understood/meant over ones with a lower likelihood. Lastly, that
+learners prefer simpler over more complex lexical representations and exhibit a tendency
+towards the acquisition of the type that best explains the learning data.
 
 \section{General discussion}\label{sec:discussion}
-We laid out a model that combines game theoretical models of functional pressure towards efficient communication \citep{nowak+krakauer:1999}, effects of transmission perturbations on (iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012, franke+jaeger:2014} as well as different lexica \citep{bergen+etal:2012,bergen+etal:2016}.
-This model generates predictions about the co-evolution of conventional meaning and pragmatic reasoning types and, more generally, effects of communicative pressures on the cultural evolution of language.
-
-We argued that the puzzle raised by semantics in light of pragmatics is hard to explain on purely functional grounds and that part of the answer may instead lie in the way successful transmission of linguistic knowledge across and within populations shapes the outcome of their co-evolution.
-In the realm of inductive biases, we adopted the assumption that simpler semantic representations are more likely to be learned (cf.
-\citealt{chater+vitanyi:2003}).
-Under this view, semantics and pragmatics play a synergic role in that representational simplicity is supplemented by pragmatic reasoning to counteract functional disadvantages otherwise incurred.
-As a consequence, iterated transmission and use of language lead to a regularization that may explain the lack of lexicalization of systematic pragmatic enrichments.
-This result is of particular relevance to the longstanding assumption of a divide and interaction between semantics and pragmatics by offering an account of why certain pragmatic inferences fail to lexicalize.
-
-The main innovations of the model are its modular separation of expressivity and learnability, allowing for their isolated and combined analysis, learning involving a joint inference over types of pragmatic behavior and lexical meaning, and, accordingly, the possibility to trace the co-evolution of conventional semantics and pragmatic use.
-The goal to decouple but model both expressivity and learnability has also recently been addressed by \citet{kirby+etal:2015}.
-In contrast to our proposal, Kirby et al.
-model expressivity as exerting its force only in the production of learning data.
-The degree of mutual understanding of interlocutors central to replication and to our notion of expressivity is absent.
-That is, while our proposal combines bidirectional horizontal transmission with its vertical and unidirectional counterpart, Kirby et al.
-only consider the latter's influence.
-Our reasoning behind the inclusion of the former lies in the empirical and theoretical observation that learnability alone can lead the selection of functionally defective languages, as showcased by $\Lall$.
-This outcome has been reported in a number of laboratory experiments where subjects learned and subsequently reproduced the language produced by a previous participant over multiple iterations, leading to a proliferation of ambiguous languages that associate many meanings with a single form (see, e.g., \citealt{silvey+etal:2014} and experiment 1 in \citealt{kirby+etal:2008}).
-In contrast, experiments involving an interactive component have been found to foster languages that enable interlocutors to communicatively distinguish meanings more accurately  (e.g., \citealt{fay+etal:2013}; for a review of laboratory results under the iterated learning paradigm and further discussion see \citealt{kirby+etal:2015, tamariz+kirby:2016}).
-It is not evident how to compare these empirical findings given that they consider distinct meaning spaces, modes of transmission, iterations and feedback given to participants.
-However, on a general level they may suggest that there is an important difference between a language generating learnable linguistic data and its actual performance as a means of information transfer.
-That is, we contend that successful information transfer in a linguistic community is central to the adoption of a communication system and that this measure is not adequately reflected by production alone.
+
+The approach introduced here combines game theoretic models of functional pressure towards
+efficient communication \citep{nowak+krakauer:1999}, effects of transmission perturbations on
+(iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener
+types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012,
+  franke+jaeger:2014} as well as reasoning about unobservable lexical representations
+\citep{bergen+etal:2012,bergen+etal:2016}. This model allows a structured conceptual
+investigation of the co-evolution of conventional meaning and pragmatic language use. Main
+innovations of the model are (i) its modular separation of expressivity and learnability,
+allowing for their isolated and combined analysis, (ii) the characterization of language
+learning as a joint inference over types of pragmatic behavior and lexical meaning, and (iii)
+the possibility to trace the co-evolution of conventional semantics and pragmatic use.
+
+The goal to decouple but model both expressivity and learnability has also recently been
+addressed by \citet{kirby+etal:2015}. In contrast to our proposal, Kirby et al. model
+expressivity as exerting its force only in the production of learning data. The degree of
+mutual understanding of interlocutors central to replication and to our notion of expressivity
+is absent. That is, while our proposal combines bidirectional horizontal transmission with its
+vertical and unidirectional counterpart, Kirby et al. only consider the latter's influence. Our
+reasoning behind the inclusion of the former lies in the empirical and theoretical observation
+that learnability alone can lead the selection of functionally defective languages, as
+showcased by $\Lall$. This outcome has been reported in a number of laboratory experiments
+where subjects learned and subsequently reproduced the language produced by a previous
+participant over multiple iterations, leading to a proliferation of ambiguous languages that
+associate many meanings with a single form (see, e.g., \citealt{silvey+etal:2014} and
+experiment 1 in \citealt{kirby+etal:2008}). In contrast, experiments involving an interactive
+component have been found to foster languages that enable interlocutors to communicatively
+distinguish meanings more accurately (e.g., \citealt{fay+etal:2013}; for a review of laboratory
+results under the iterated learning paradigm and further discussion see
+\citealt{kirby+etal:2015, tamariz+kirby:2016}). It is not evident how to compare these
+empirical findings given that they consider distinct meaning spaces, modes of transmission,
+iterations and feedback given to participants. However, on a general level they may suggest
+that there is an important difference between a language generating learnable linguistic data
+and its actual performance as a means of information transfer. That is, we contend that
+successful information transfer in a linguistic community is central to the adoption of a
+communication system and that this measure may not be adequately reflected by production alone.
+
+The main result of our case study is that types that correspond to the majority view of scalar
+implicatures (scalar readings are non-lexicalized pragmatic enrichments) can come to dominate a
+population. This can happen under the assumption that simpler semantic representations are more
+likely to be learned (cf. \citealt{chater+vitanyi:2003}). Pragmatic language use can then be
+recruited indirectly by a preference for simpler lexical concepts. Under this view, semantics
+and pragmatics play a synergic role: pragmatic use allows maintenance of simpler concepts;
+pressure towards representational simplicity indirectly promotes pragmatic over literal
+language use. As a consequence, iterated transmission and use of language lead to a
+regularization that may explain the lack of lexicalization of systematic pragmatic
+enrichments.
+
+While the results of this case study are interesting, they also raise a number of critical
+issues. First of all, it is not clear to what extent other types should be present in the final
+population. It seems reasonable to expect functionally suboptimal lexica such as $\Lall$ to be
+ruled out but matters are different for $\Lbound$. The prediction that natural language
+communities are homogeneous or that speakers may entertain $\Llack$-like semantics for one
+scalar expression and $\Lbound$-like semantics for another is not implausible
+(cf. \citealt{NOVECK ETC}). Alternatively, a stronger tendency for posterior
+maximization in learning has to be assumed. This empirical issue relates to other two aspects
+left undiscussed: disadvantages of pragmatic reasoning and the effect of state frequencies on
+the fossilization of pragmatic inferences. We tacitly assumed pragmatic reasoning to come at no
+cost. However, there is experimental evidence that suggests that the pragmatic derivation of
+upper-bounds costs effort and takes additional processing time
+(cf. \citealt{deNeys+schaeken:2007, huang+snedeker:2009}). This raises the question at which
+point such usage-based cost undercuts the learnability advantage of simpler semantic
+representations. Should cost play a role, then its effect is bound to depend on the frequency
+with which a given scalar expression is used. It may be that frequently drawn scalar
+implicatures fossilize to avoid cost whereas infrequent ones are derived on-line to avoid
+complexity in acquisition. This also opens a possible venue to address the preceding question
+about the expected presence of $\Lbound$-like semantics in a population, but further empirical
+evidence is needed to assess these matters.
 
@@
 
-\bibliographystyle{plainnat}
+\bibliographystyle{chicago}
 \bibliography{./bounds-rmd}
+
+\newpage
+
+\appendix
+
+\section{Snippets}
+
+\subsection{On alternative explanations}
+\label{sec:altern-expl}
+
+We see two main explanations for the lack of upper-bounds in the lexical meaning of weak scalar expressions.
+The first is that their truth-conditional compatibility with stronger expressions endows them with a broader range of applicability by allowing them to occur in contexts in which their upper-bounded reading is absent.
+This can happen when embedded in downward-entailing contexts, when the speaker is likely uncertain about whether the upper bounded reading is true, or when the distinction between an upper-bounded reading and the simple, only lower-bounded reading, is not relevant.
+For instance, if for all the speaker knows Chris owns some of Johnny Cash's albums but she does not know whether he owns all, then the use of {\em some} lacking an upper-bound succinctly conveys her uncertainty.
+This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: Should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then this is preferred over expressing it overtly through longer expressions, e.g., by saying {\em some but not all} explicitly.
+Importantly, although morphosyntactic disambiguation may be dispreferred due to its relative length and complexity \citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and override contextual cues that might otherwise mislead the hearer.
+In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so.
+Compare this with a hypothetical language that lexicalizes two expressions for each weak scalar expression -- one with and one lacking an upper-bound.
+We see four conditions along this functionalist explanation that may pressure languages for English-like semantics over this alternative.
+First, contextual cues are very reliable.
+Second, morphosyntactic disambiguation is seldom necessary.
+Third, morphosyntactic disambiguation is only marginally dispreferred.
+Fourth, larger lexica are costly.
+Overall, neither condition seems convincing as a central explanatory device for such a widespread phenomenon.
+The first two conditions put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of unambiguous communication.
+It is likely that human language users are very good at retrieving cues from context, but to stipulate that they are so good as to undercut the benefit of safe communication provided by this hypothetical alternative strikes us as too strong of an assumption.
+As for the third and fourth condition, these seem mostly like technical solutions without a proper empirical basis.
+
+Instead, the systematicity and typological spread of scalar implicatures together with the observation that monomorphemic expressions that lexically rule out stronger alternatives are unattested across languages (\citealt[252-267]{horn:1984}, \citealt{horn:1972,traugott:2004,vdAuwera:2010}) suggests that other forces may be at play.
+In what follows we investigate the predictions of our model under the assumption that the lack of lexicalization of scalar inferences may be accounted for by the relative representational simplicity of lexical meanings lacking an upper-bound over those that explicitly codify it.
+This difference is reflected in a learning bias towards more compressed lexical representations.
+That is, in a preference of learners for simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
+While we do not want to argue that functional aspects as the ones discussed above do not play a role, we do see a clear benefit in exploring whether matters of transmission biases would not give us additional explanatory leverage.
+
 \end{document}

### 2017-03-24 paper/paper-v3.tex
overhaul section 3 and 4; this version passed back to TB and RvR

@@
 The model of the previous section formalizes the evolutionary competition between different
-pairs of lexical concepts and ways of using them. This section looks at a concrete case study
-on scalar implicatures. Concretely, we will engage in the following formal thought experiment:
-if a population of language users could freely combine different lexica with different
-pragmatic strategies, what there the conditions under which the majority view of scalar
-implicatures could have evolved?
+pairs of lexical concepts and ways of using them. This section looks at a case study on scalar
+implicatures. It engages in a following formal thought experiment to address the question: if a
+population of language users could freely combine different lexica with different pragmatic
+strategies, what are conditions under which the majority view of scalar implicatures could have
+evolved?
 
@@
 more informative one could have been used can license a defeasible inference that stronger
-alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). That is, a hearer who assumes
-the speaker to be able and willing to provide all relevant information can infer that stronger
-alternatives do not hold because the speaker used a weaker alternative instead. In this way,
-``Chris owns some of Johnny Cash's albums'' is strengthened to convey that she owns {\em some
-  but not all} albums. The upper bound, so the majority view, is a pragmatic inference but not
-part of the conventional meaning. \mf{I wonder if this paragraph is really still necessary
-  after the introduction?}
+alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). In this way, ``Chris owns some
+of Johnny Cash's albums'' is strengthened to convey that she owns {\em some but not all}
+albums. According to the majority view, this is a pragmatic inference, not part of the
+conventional meaning.
 
@@
 To fill the model from Section~\ref{sec:model} with life, we need to specify the sets of
-states, messages and lexica. Additionally, we want to explore the effects of a learning bias in
-favor of simple lexical concepts. One way of motivating and formalizing such a bias is
-introduced below.
+states, messages and lexica (Section~\ref{sec:stat-mess-conc}). Additionally, we want to
+explore the effects of a learning bias in favor of simple lexical concepts. One way of
+motivating and formalizing such a bias is introduced below in
+Section~\ref{sec:an-induct-learn}.
 
 \subsubsection{States, messages, concepts and Lexica} 
+\label{sec:stat-mess-conc}
 
@@
 their truth or falsity in three world states, six of which are not contradictory or
-tautological.
+tautological (see Table~\ref{tab:concepts} below).
 
@@
 the same concept to more than one message. Such lexica lack in expressivity but may be favored
-by particular learning biases nonetheless. Second, lexica that conventionalize upper-bounds to
-realize a (quasi-)partition of the state space. Finally, lexically that do not lexicalize an
-upper bound but allow it to be conveyed pragmatically due to the presence of a stronger lexical
-item. There are six lexica of the second kind and six of the third. The following three lexica
-exemplify each kind:
+by particular learning biases nonetheless (see below). Second, lexica that conventionalize
+upper-bounds to realize a one-to-one mapping of messages to states. Finally, lexically that do
+not lexicalize an upper bound but allow it to be conveyed pragmatically due to the presence of
+a stronger lexical item. There are six lexica of the second kind and six of the third. The
+following three lexica exemplify each kind:
 \begin{align*}
@@
 Lexicon $\Lall$ is clearly bad for communication: all message and interpretation choices will
-be equally likely for all types; no information will be conveyed with it. Users of $\Lbound$
-need not resort to pragmatic reasoning to convey an upper-bound with weak scalar
-expressions. Instead, this information is already provided by the semantics of their lexicon,
-giving them an advantage in terms of expressivity. Types with such lexica may be expected to
-generally achieve higher communicative success than pragmatic users with a lexicon of type
-$\Llack$ (see Section~\ref{sec:languages+use}).
+be equally likely for all types; no information about the observed world state will be conveyed
+by its users. In contrast, users of $\Lbound$ can communicate world states perfectly, no matter
+whether they are literal or pragmatic users. Users of $\Llack$ can also communicate information
+about the actual world state but need pragmatic language use to approximate a one-to-one
+mapping between message use and states (see Section~\ref{sec:languages+use}).
 
@@
 model's predictions in populations of types with one of the two behaviors introduced earlier;
-literal or pragmatic. The former correspond to level $0$ reasoners and the latter to ones of
-level $1$. Accordingly, we consider a total of $432$ types. Six are variants of pragmatic
-language users with $\Llack$-like lexica. These will be our target types, because they
+literal or pragmatic. The former correspond to level $0$ reasoners and the latter to level
+$1$. Accordingly, we consider a total of $432$ types. Six are variants of pragmatic language
+users with $\Llack$-like lexica. We refer to these as \emph{target types}, because they
 represent lexica and language use that conform to the majority view of scalar
 implicatures. Twelve types are either literal or pragmatic speakers with lexical of the
-$\Lbound$ kind. These will be the target types main competitors in evolutionary competition.
+$\Lbound$ kind. We refer to these as \emph{competitor types}, because they are expected to be
+the target types' main competitors in evolutionary competition.
 
@@
 for the measure of semantic complexity. For instance, so-called {\em languages of thought}
-(LOT) have been put to test in various rational probabilistic models that show encouraging
+(LOTs) have been put to test in various rational probabilistic models that show encouraging
 results (see, e.g., \citealt{katz+etal:2008, piantadosi+etal:underreview,
@@
 derived. As a first approximation and for the sake of concreteness, we follow this approach to
-make the preference of learners for simpler semantic representations precise and define the
-complexity of a representation as a function of its derivation cost in a weighed generative
-LOT.
+motivate and formalize a preference of learners for simpler semantic representations
+\citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a,
+  kirby+etal:2015,piantadosi+etal:underreview}. A concept's complexity is a function of its
+derivation cost in a weighed generative LOT.
 
@@
     C & \rightarrow_2 C \wedge C 
-    & 
-    X & \rightarrow_1 \set{A,B} \\
-    C & \rightarrow_2 \neg C 
-    & 
-    X & \rightarrow_1 X \cap X \\
-    C & \rightarrow_1 X \subseteq X
-    & 
-    X & \rightarrow_1 X \cup X \\
-    C & \rightarrow_1 X \neq \emptyset \\
-    C & \rightarrow_1 X = \emptyset     
+    & C & \rightarrow_2 \neg C  \\ 
+    C & \rightarrow_1 X \subseteq X &
+    C & \rightarrow_1 X \neq \emptyset &
+    C & \rightarrow_1 X = \emptyset  \\
+    X & \rightarrow_1 \set{A,B} & 
+    X & \rightarrow_1 X \cap X & 
+    X & \rightarrow_1 X \cup X    
   \end{align*}
@@
 $\snone$, $\ssome$ or $\sall$ from above. Applications of generative rules have a cost attached
-to them. Here we simply assume that Boolean combinations of concepts are more complex than
-atomic concepts and that otherwise each rule application adds the same cost
-unit. Table~\ref{tab:concepts} lists all six concepts relevant here, their truth conditions and
-the simplest formula that expresses this concept from the grammar in Table~\ref{tab:grammar}.
+to them. Here we simply assume that the formation of Boolean combinations of concepts incurs 2
+cost units, while all other rule applications incur only 1 cost unit. Table~\ref{tab:concepts}
+lists all six concepts relevant here, their truth conditions and the simplest formula that
+expresses this concept from the grammar in Table~\ref{tab:grammar}.
 
@@
 \end{center}
-\caption{Available concepts and their minimal derivation length}
+\caption{Available concepts and their minimal derivation cost}
 \label{tab:concepts}
@@
 The complexity measure for lexical concepts from Table~\ref{tab:concepts} is used to define a
-learning bias that favor simpler concepts over more complex ones. The prior probability of a
+learning bias that favors simpler concepts over more complex ones. The prior probability of a
 type is just the prior probability of its lexicon. The prior of a lexicon is a function of the
@@
 \end{align*}
-where $Compl(c)$ is the complexity of the minimal derivation length of $c$ according to the
-LOT-grammar (see Table~\ref{tab:concepts}). There are many ways to define priors over lexica
-(see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a}) but the key assumption here,
-common to all of them, is that simple representational expressions should be favored over more
-complex ones. We should stress that these details -- from the generative grammar to its
-complexity measure -- are to be regarded as one convenient operationalization of one general
-approach to explicating learning biases; this is not a commitment that this general approach is
-necessarily superior or that, within it, this particular instrumentalization is the single most
-plausible.
+where $Compl(c)$ is the complexity of the minimal derivation cost of $c$ according to the
+LOT-grammar (see Table~\ref{tab:concepts}). This construal assigns the highest probability the
+above example for a lexicon of type $\Lall$, which only uses the simplest concept ``all'' for
+all messages. Lexica of type $\Llack$ are less likely, but more likely than $\Lbound$.
+
+There are many ways to define priors over lexica (see, e.g., \citealt{goodman+etal:2008,
+  piantadosi+etal:2012a}) but the key assumption here, common to all of them, is that simple
+representational expressions should be favored over more complex ones. We should stress that
+these details -- from the generative grammar to its complexity measure -- are to be regarded as
+one convenient operationalization of one general approach to explicating learning biases; this
+is not a commitment that this general approach is necessarily superior or that, within it, this
+particular instrumentalization is the single most plausible.
 
@@
 Recall that there are three parameters: soft-max parameter $\lambda$ affects how strongly
-pragmatic speakers favor more informative messages; the bottleneck size $k$ influences how
-faithfully learners can identify their teacher type; $l$ defines the learners disposition
-towards choosing the most likely teacher type from the posterior distribution. We expect that
-competitor types (types with lexica of the kind $\Lbound$) have a fitness advantage over target
-types (pragmatic agents with lexica of the kind $\Llack$), especially for very low levels of
-$\lambda$. Selection based on fitness alone may therefore not lead to prevalence of target
-types in the population. On the other hand, lexica of type $\Llack$ are simpler than those of
-type $\Lbound$ by the postulated measure from above. This may make them more likely to be
-adopted by learners, especially when $k$ is low so that teacher types are relatively
-indistinguishable based on their behavior in general and when $l$ is high. Still, lexica of the
-kind $\Lall$ are in turn more likely \emph{a priori} than lexica of the kind
-$\Llack$. Simulation results will shed light on the question whether target types can emerge
-and for which parameter constellations.
-
-To better understand the workings a pressure towards communicative efficiency and a pressure
+pragmatic speakers favor messages that appear best from their subjective point of view; the
+bottleneck size $k$ influences how faithfully learners can identify their teacher type; $l$
+defines the learners' disposition towards choosing the most likely teacher type from the
+posterior distribution. We expect that competitor types (types with lexica of the kind
+$\Lbound$) have a fitness advantage over target types (pragmatic agents with lexica of the kind
+$\Llack$), especially for very low levels of $\lambda$. Selection based on fitness alone may
+therefore not lead to prevalence of target types in the population. On the other hand, lexica
+of type $\Llack$ are simpler than those of type $\Lbound$ by the postulated measure from
+above. This may make them more likely to be adopted by learners, especially when $k$ is low so
+that different teacher types are relatively indistinguishable based on their behavior and when
+$l$ is high. Still, lexica of the kind $\Lall$ are in turn more likely \emph{a priori} than
+lexica of the kind $\Llack$. Simulation results will shed light on the question whether target
+types can emerge and for which parameter constellations.
+
+To better understand the workings of a pressure towards communicative efficiency and a pressure
 towards learnability, we look at the behavior of the replicator and mutator step first in
@@
 distribution over types, constituting a population's first generation. All reported results are
-the outcome of 50 update steps for each different constellation of parameter values. This
-ensures that these outcomes correspond to developmental plateaus in which change is, if not
-absent, then at least very slow, In other words, even if the resulting states do not correspond to
-an eventual attracting state, they characterize long almost stationary states in which the
-system remains for a very long time. As specified in \S\ref{sec:learnability}, the mutation
-matrix $Q$ can be obtained by considering all possible state-message sequences of length
-$k$. Given that this is intractable for large $k$, the sets of data learners are exposed to are
-approximated by sampling $250$ $k$-length sequences from each type's production probabilities.
+the outcome of 50 update steps. These outcomes correspond to developmental plateaus in which
+change is, if not absent, then at least very slow. In other words, even if the resulting states
+do not correspond to an eventual attracting state, they characterize long almost stationary
+states in which the system remains for a very long time. As specified in
+\S\ref{sec:learnability}, the mutation matrix $Q$ can be obtained by considering all possible
+state-message sequences of length $k$. Given that this is intractable for large $k$, the sets
+of data learners are exposed to are approximated by sampling $250$ $k$-length sequences from
+each type's production probabilities.
 
 \subsubsection{Replication only: selection based on expressivity and communicative success}
-\begin{figure}
+
+\begin{figure}[t]
 \centering
 \includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig1-onlyr}
-\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a pressure for expressivity.
-The right-most plots zoom in on only the proportion of pragmatic $\Llack$ types.}
+\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other
+  than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a
+  pressure for expressivity. The right-most plots zoom in on only the proportion of pragmatic
+  $\Llack$ types. \mf{not sure if ``incumbent'' is the right term here; I believe it means
+    ``someone who is already there and established'' but we mean ``type with the highest
+    proportion''; I used ``majority type'' in the main text; maybe change the plot?}}
 \label{fig:only-R}
@@
 As expected, selection based on communicative success is sensitive to $\lambda$ as it
-influences signaling behavior. This is showcased in Figure \ref{fig:only-R}, which depicts the
-proportion of target types in $5$ independent populations after $50$ generations, as well as
-that of the type with the highest proportion in a given population, the incumbent. With low
-$\lambda$ many types have very similar behavior, so that evolutionary selection lacks grip and
-becomes very slow. The result is a very long transition with near stagnancy in a rather
+influences signaling behavior, which in turn determines communicative success. This is
+showcased in Figure \ref{fig:only-R}, which depicts the proportion of target types in $5$
+independent populations after $50$ replicator steps. The plot also shows the proportion of the
+\emph{majority type}, i.e., the type with the highest proportion in the final population. With
+low $\lambda$ many types have very similar behavior, so that evolutionary selection lacks grip
+and becomes very slow. The result is a very long transition with near stagnancy in a rather
 homogeneous population with many types. Conversely, higher $\lambda$ promotes more rational
@@
 populations is not one of the six pragmatic $\Llack$-style types. That is, a pressure only for
-expressivity does not lend a justification to the systematic prevalence of a lack of
-upper-bounds under any $\lambda$-value. For instance, in $1000$ independent populations with
-$\lambda = 20$ pragmatic $\Llack$ types only made up $11$ cases of incumbency, corresponding to
-a mean proportion of $0.003$ across populations. By contrast, $913$ incumbents were
-$\Lbound$-style users with close to an even share between literal ($454$) and pragmatic types
-($459$), corresponding to a mean proportion of about $.48$ taken
-together. %Across types and populations incumbents make up $.50$ of the mean outcome.
+expressivity does not lead to a prevalence of target types under any $\lambda$-value. For
+instance, in $1000$ independent populations with $\lambda = 20$ only had $11$ cases in which
+the target type was the majority type, corresponding to a mean proportion of $0.003$ across
+populations. By contrast, in $913$ cases the majority types had $\Lbound$ with close to an even
+share between literal ($454$) and pragmatic types ($459$), corresponding to a mean proportion
+of about $.48$ taken together.
 
@@
 
-\begin{figure}
+\begin{figure}[t]
 \centering
 \includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig2-onlym-pr}
-\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a pressure for learnability ($\lambda = 20, k = 5$).
-The types' prior probability is shown in the right-most plot.}
+\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other
+  than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a
+  pressure for learnability ($\lambda = 20, k = 5$). The types' prior probability is shown in
+  the right-most plot. \mf{explain the topmost part of the right plot?!}}
 \label{fig:only-M}
@@
 nevertheless more advantaged than others, such as $\Lbound$, in virtue of the relatively simple
-semantics they conventionalize. Crucially, $\Llack$ enables its users to convey each state with
-a single message when combined with pragmatic reasoning provided sufficiently high
-$\lambda$. This makes it less likely to be confused with other types if the learning data is
-not too sparse ($k \geq 5$). Put differently, learners have a higher propensity to infer
-pragmatic $\Llack$ over types that might be hard or impossible to distinguish it based on overt
-linguistic behavior, such as $\Lbound$. Moreover, $\Llack$ is also less likely to be confused
-with types with different observable behavior because its pragmatic use approximates a
-one-to-one form-meaning mapping. As a consequence, a stronger propensity to maximize the
-posterior increases their proportion in the population.
-
-However, in contrast to a pressure only for expressivity with high $\lambda$, learnability
-alone does not succeed in selecting for a single prevalent type. Indeed, all six target types
-tend to coexist at roughly equal proportion. Each is passed on to next generation with the same
-faithfulness and, differently from a pressure for communicative success, they do not stand in
-competition with each other. As implicit in Figure \ref{fig:only-M}, in $1000$ independent
-populations all incumbents were pragmatic $\Llack$ users provided sufficiently high $\lambda$,
-with each reaching approximately the same proportion of users in the population. As with
-expressivity only, low values of $\lambda$ make the differences in observable behavior across
-types less pronounced and therefore reflect the learners' inductive bias more faithfully,
-favoring functionally deficient but a priori preferred types such as those that use $\Lall$. A
-pressure for learnability alone may consequently lead to a spread of communicatively suboptimal
-types that are easier to learn. In the extreme, when $l = 1$ and $\lambda = 1$ all of $1000$
-independent populations had users of $\Lall$ as incumbents.
+semantics they conventionalize (see Section~\ref{sec:an-induct-learn}). Crucially, $\Llack$
+enables its users to convey each state with a single message when combined with pragmatic
+reasoning provided sufficiently high $\lambda$. This makes it less likely to be confused with
+other types if the learning data is not too sparse ($k \geq 5$). Put differently, learners have
+a higher propensity to infer pragmatic $\Llack$ when the teacher's type produces very similar
+data, such as when using $\Lbound$. Moreover, $\Llack$ is also less likely to be confused with
+types with different observable behavior because its pragmatic use approximates a one-to-one
+form-meaning mapping. As a consequence, a stronger propensity to maximize the posterior
+increases their proportion in the population.
+
+However, in contrast to a pressure only for expressivity with high $\lambda$ (see
+Figure~\ref{fig:only-R}), learnability alone does not succeed in selecting for a single
+prevalent type. Indeed, all six target types tend to coexist at roughly equal proportion. Each
+is passed on to the next generation with the same faithfulness and, differently from a pressure
+for communicative success, they do not stand in competition with each other. In $1000$
+independent populations all majority types were target types provided, sufficiently high
+$\lambda$ \mf{don't understand this proviso: what $\lambda$ values were used in the 1000
+  runs? what's sufficiently high?}, with each reaching approximately the same proportion of users in the population. As
+with expressivity only, low values of $\lambda$ make the differences in observable behavior
+across types less pronounced and therefore reflect the learners' inductive bias more
+faithfully, favoring functionally deficient but a priori preferred types such as those that use
+$\Lall$. A pressure for learnability alone may consequently lead to a spread of communicatively
+suboptimal types that are easier to learn. In the extreme, when $l = 1$ and $\lambda = 1$ all
+of $1000$ independent populations had users of $\Lall$ as incumbents.
 
@@
 Pressures for communicative success or learnability are not sufficient on their own to have a
-single target type dominate the population. When pressured toward expressivity, the slight
-communicative advantage of $\Lbound$ users leads to the its prevalence. When pressured toward
+single target type dominate the population. When pressured for expressivity, the slight
+communicative advantage of $\Lbound$ users leads to the its prevalence. When pressured for
 learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more
 complex alternatives such as $\Lbound$. However, on its own learnability does not foment the
-propagation of these types across the population. What is more, it leads to heterogeneous
-populations that foster an equal share of types of a kind, such as different pragmatic $\Llack$
-types.
-
-\begin{figure}
+propagation of a single target type across the population.
+
+\begin{figure}[t]
 \centering
@@
 
-Figure \ref{fig:rmd} illustrates the combined effects of both pressures for a sample of $\lambda$ and $l$ values.
-These results show that an inductive learning bias for simpler semantics in tandem with functional pressure can lead to the selection of $\Llack$-like semantics paired with pragmatic language use.
-The proportion of these types increases with $\lambda$ and $l$.
-Pressure towards expressivity magnifies the effects of iterated learning and dampens the proliferation of types of a kind that are equal in expressivity {\em and} learnability.
-A pressure towards learnability favors the transmission of simpler semantics and thereby makes types that exploit pragmatic reasoning to strengthen lexical meaning more likely to be adopted over alternatives that do not.
-%In short, neither learnability nor functional pressure alone but their combination may lead to the lack of upper-bounds in the lexical meaning of scalar expressions in predominantly monomorphic populations.
-
-As before, low $\lambda$ and $l$ lead to the incumbency of communicatively suboptimal types that are a priori favored, such as $\Lall$.
-An increase in $\lambda$ leads to the selection and incumbency of pragmatic $\Llack$ but does not lead to monomorphic populations if learners sample from the posterior.
-Finally, a combination of sufficiently rational signaling behavior with some tendency for posterior maximization leads to increasingly monomorphic populations of types that convey upper-bounds pragmatically.
-The difference between the mean of the highest pragmatic $\Llack$-like type in $1000$ independent populations and the highest proportion of other types across $\lambda$ and $l$ values is shown in Figure \ref{fig:diff}.
-
-\begin{figure}
+Figure \ref{fig:rmd} illustrates the combined effects of both pressures for a sample of
+$\lambda$ and $l$ values. These results show that an inductive learning bias for simpler
+semantics in tandem with functional pressure can lead to the selection of a single target
+type. The proportion of a single majority target type increases with $\lambda$ and
+$l$. Pressure for communicative success magnifies the effects of iterated learning and dampens
+the proliferation of types of a kind that are equal in expressivity {\em and} learnability. A
+pressure towards learnability favors the transmission of simpler semantics and thereby selects
+for pragmatic language use.
+
+As before, low $\lambda$ and $l$ lead to the prevalence of communicatively suboptimal types
+that are a priori favored, such as $\Lall$. An increase in $\lambda$ leads to the selection of
+target types but does not lead to monomorphic populations if learners sample from the
+posterior. Finally, a combination of high $\lambda$ and $l$, leads to increasing proportions of
+a single majority target type. The difference between the mean of the highest pragmatic
+$\Llack$-like type in $1000$ independent populations and the highest proportion of other types
+across $\lambda$ and $l$ values is shown in Figure \ref{fig:diff}.
+
+\begin{figure}[t]
 \centering
-\includegraphics[width=1\textwidth,height=6cm,keepaspectratio]{./plots/fig4-incumbents-difference}
-\caption{Difference between mean proportion of highest pragmatic $\Llack$ type and highest other type in $1000$ independent populations after $50$ generations under both pressures ($k = 5$).
-\tb{I will run more simulations with higher $\lambda$ and $l$.
-I'm pretty sure the difference still increases.}}
+\includegraphics[width=1\textwidth,height=10cm,keepaspectratio]{./plots/fig4-incumbents-difference}
+\caption{Difference between mean proportion of highest pragmatic $\Llack$ type and highest
+  other type in $1000$ independent populations after $50$ generations under both pressures ($k
+  = 5$). \tb{I will run more simulations with higher $\lambda$ and $l$. I'm pretty sure the
+    difference still increases.} \mf{I think it's okay like that} \mf{the coloring is not that
+    easy to decipher here; would a two-dimensional surface be possible and look better?}}
 \label{fig:diff}
@@
 
-As for the effect of the sequence length $k$ not mentioned so far, it influences populations in a predictable way: small values lead to more heterogeneous populations that reflect the learner's prior more faithfully.
-This is due to the fact that the likelihood that a small sequence was produced by any type is relatively uniform (modulo prior).
-By contrast, larger values increasingly allow learners to differentiate types with different signaling behaviors.
+As for the effect of the sequence length $k$ not mentioned so far, it influences populations in
+a predictable way: small values lead to more heterogeneous populations that reflect the
+learner's prior more faithfully. This is due to the fact that the likelihood that a small
+sequence was produced by any type is relatively uniform (modulo prior). By contrast, larger
+values increasingly allow learners to differentiate types with different signaling behaviors.
 
@@
 the resulting proportion of pragmatic $\Llack$ speakers primarily hinges on three
-factors. First, the degree to which agents try to maximize communicative success. This plays a
-role both in expressivity as well as in producing data that allows learners to discriminate
-this type from others. Second, the inductive bias, which leads learners to prefer simpler over
-more complex semantic representations in acquisition. Lastly, the posterior parameter, as it
-magnifies the effects of the learning bias in tandem with replication.
-
-\subsection{Summary}
-
-Under the assumption of a learning bias for simpler lexical representations, our results
-suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome
-communicative pressures and come to dominate a population. This prediction hinges on three
-assumptions. First, that language is pressured toward both expressivity and
-learnability. Second, that language users favor expressions/interpretations expected to have a
-higher likelihood of being understood/meant over ones with a lower likelihood. Lastly, that
-learners prefer simpler over more complex lexical representations and exhibit a tendency
-towards the acquisition of the type that best explains the learning data.
+factors. First, the degree, captured by $\lambda$, to which agents try to maximize
+communicative success from their own subjective point of view. Second, the inductive bias,
+which leads learners to prefer simpler over more complex semantic representations in
+acquisition. Lastly, the learning behavior, captured by parameter $l$, where approximating a
+MAP strategy magnifies the effects of the learning bias in tandem with replication.
+
+In summary, target types, which represent the majority view of scalar implicatures, can come to
+dominate the population if three assumptions are met: (i) language is pressured toward both
+expressivity and learnability; (ii) pragmatic language use is an option; (iii) learners prefer
+simpler over more complex lexical representations and exhibit a tendency towards the
+acquisition of the type that best explains the learning data.
 
@@
   franke+jaeger:2014} as well as reasoning about unobservable lexical representations
-\citep{bergen+etal:2012,bergen+etal:2016}. This model allows a structured conceptual
-investigation of the co-evolution of conventional meaning and pragmatic language use. Main
-innovations of the model are (i) its modular separation of expressivity and learnability,
-allowing for their isolated and combined analysis, (ii) the characterization of language
-learning as a joint inference over types of pragmatic behavior and lexical meaning, and (iii)
-the possibility to trace the co-evolution of conventional semantics and pragmatic use.
-
-The goal to decouple but model both expressivity and learnability has also recently been
-addressed by \citet{kirby+etal:2015}. In contrast to our proposal, Kirby et al. model
-expressivity as exerting its force only in the production of learning data. The degree of
-mutual understanding of interlocutors central to replication and to our notion of expressivity
-is absent. That is, while our proposal combines bidirectional horizontal transmission with its
-vertical and unidirectional counterpart, Kirby et al. only consider the latter's influence. Our
-reasoning behind the inclusion of the former lies in the empirical and theoretical observation
-that learnability alone can lead the selection of functionally defective languages, as
-showcased by $\Lall$. This outcome has been reported in a number of laboratory experiments
-where subjects learned and subsequently reproduced the language produced by a previous
-participant over multiple iterations, leading to a proliferation of ambiguous languages that
-associate many meanings with a single form (see, e.g., \citealt{silvey+etal:2014} and
-experiment 1 in \citealt{kirby+etal:2008}). In contrast, experiments involving an interactive
-component have been found to foster languages that enable interlocutors to communicatively
-distinguish meanings more accurately (e.g., \citealt{fay+etal:2013}; for a review of laboratory
-results under the iterated learning paradigm and further discussion see
-\citealt{kirby+etal:2015, tamariz+kirby:2016}). It is not evident how to compare these
-empirical findings given that they consider distinct meaning spaces, modes of transmission,
-iterations and feedback given to participants. However, on a general level they may suggest
-that there is an important difference between a language generating learnable linguistic data
-and its actual performance as a means of information transfer. That is, we contend that
-successful information transfer in a linguistic community is central to the adoption of a
-communication system and that this measure may not be adequately reflected by production alone.
+\citep{bergen+etal:2012,bergen+etal:2016}. This allows a conceptual investigation of the
+co-evolution of conventional meaning and pragmatic language use. Main contributions of the
+model are (i) its modular separation of expressivity and learnability on evolutionary
+trajectories, (ii) the characterization of language learning as a joint inference over
+pragmatic behavior and lexical meaning, and (iii) the possibility to trace the co-evolution of
+conventional semantics and pragmatic use.
+
+\mf{can this abstract be shortened (by a factor of $.4$)?} The interaction of expressivity and
+learnability in language evolution has also recently been addressed by
+\citet{kirby+etal:2015}. In contrast to our proposal, Kirby et al. model expressivity as
+exerting its force only in the production of learning data. The degree of mutual understanding
+of interlocutors central to replication and to our notion of expressivity is absent. That is,
+while our proposal combines bidirectional horizontal transmission with its vertical and
+unidirectional counterpart, Kirby et al. only consider the latter's influence. Our reasoning
+behind the inclusion of the former lies in the empirical and theoretical observation that
+learnability alone can lead to the selection of functionally defective languages, as showcased
+by $\Lall$. This outcome has been reported in a number of laboratory experiments where subjects
+learned and subsequently reproduced the language produced by a previous participant over
+multiple iterations, leading to a proliferation of ambiguous languages that associate many
+meanings with a single form (see, e.g., \citealt{silvey+etal:2014} and experiment 1 in
+\citealt{kirby+etal:2008}). In contrast, experiments involving an interactive component have
+been found to foster languages that enable interlocutors to communicatively distinguish
+meanings more accurately (e.g., \citealt{fay+etal:2013}; for a review of laboratory results
+under the iterated learning paradigm and further discussion see \citealt{kirby+etal:2015,
+  tamariz+kirby:2016}). It is not evident how to compare these empirical findings given that
+they consider distinct meaning spaces, modes of transmission, iterations and feedback given to
+participants. However, on a general level they may suggest that there is an important
+difference between a language generating learnable linguistic data and its actual performance
+as a means of information transfer. That is, we contend that successful information transfer in
+a linguistic community is central to the adoption of a communication system and that this
+measure may not be adequately reflected by production alone.
 
@@
 population. This can happen under the assumption that simpler semantic representations are more
-likely to be learned (cf. \citealt{chater+vitanyi:2003}). Pragmatic language use can then be
+likely to be learned (cf. \citealt{chater+vitanyi:2003}). Pragmatic language use can be
 recruited indirectly by a preference for simpler lexical concepts. Under this view, semantics
@@
 language use. As a consequence, iterated transmission and use of language lead to a
-regularization that may explain the lack of lexicalization of systematic pragmatic
-enrichments.
+regularization that may explain the lack of lexicalization of systematic pragmatic enrichments.
 
 While the results of this case study are interesting, they also raise a number of critical
-issues. First of all, it is not clear to what extent other types should be present in the final
-population. It seems reasonable to expect functionally suboptimal lexica such as $\Lall$ to be
-ruled out but matters are different for $\Lbound$. The prediction that natural language
-communities are homogeneous or that speakers may entertain $\Llack$-like semantics for one
-scalar expression and $\Lbound$-like semantics for another is not implausible
-(cf. \citealt{NOVECK ETC}). Alternatively, a stronger tendency for posterior
-maximization in learning has to be assumed. This empirical issue relates to other two aspects
-left undiscussed: disadvantages of pragmatic reasoning and the effect of state frequencies on
-the fossilization of pragmatic inferences. We tacitly assumed pragmatic reasoning to come at no
-cost. However, there is experimental evidence that suggests that the pragmatic derivation of
-upper-bounds costs effort and takes additional processing time
-(cf. \citealt{deNeys+schaeken:2007, huang+snedeker:2009}). This raises the question at which
-point such usage-based cost undercuts the learnability advantage of simpler semantic
-representations. Should cost play a role, then its effect is bound to depend on the frequency
-with which a given scalar expression is used. It may be that frequently drawn scalar
-implicatures fossilize to avoid cost whereas infrequent ones are derived on-line to avoid
-complexity in acquisition. This also opens a possible venue to address the preceding question
-about the expected presence of $\Lbound$-like semantics in a population, but further empirical
-evidence is needed to assess these matters.
+issues. First of all, while many favorable parameter settings exist which lead to a prevalence
+of target types, other types are usually represented in non-negligible proportions. This may
+just be a technical quirk of the mutator step. But there is a related issue of empirical
+importance. Several experimental studies on scalar implicatures suggest that participants can
+be classified as either semantic or pragmatic users of, in particular, \emph{some}
+\citep[e.g.][]{BottNoveck2004:Some-Utterances,NieuwlandDitman2010:On-the-incremen,DegenTanenhaus2012:Processing-Scal}. The
+former consistently accept \emph{some} where \emph{all} would be true as well, the latter do
+not. Interestingly, in our simulations when a target type is the majority type an inflated
+proportion of the population uses compatible lexica with a lexicalized upper bound. \mf{Thomas,
+  please check this; I'm making it up; total conjecture; it might be true but I don't know for
+  sure} In other words, we do find a similar co-existence of semantic and pragmatic
+types. Whether this analogy has any further explanatory value is an interesting path for future
+exploration.
+
+Another important issue that is not addressed in the model are potential costs associated with
+pragmatic reasoning. Here, we simply assumed that literal and pragmatic reasoning strategies
+exist from the start and are equally costly to apply. In contrast, empirical results suggest
+that the computation of a scalar implicature may involve additional cognitive effort
+\citep[e.g.][]{BrehenyKatsos2006:Are-Generalised,deNeys+schaeken:2007,huang+snedeker:2009,Jr.Bailey2013:Possibly-all-of}. Extensions
+of the model presented here to include processing costs for pragmatic language use would be
+interesting future work. It seems plausible that effects of reasoning cost may trade off with
+the frequency with which a given scalar expression is used. It may be that frequently drawn
+scalar implicatures lexicalize to avoid cost, whereas infrequent ones are derived on-line to
+avoid more complex lexical concepts during acquisition. Such a prediction would lend itself to
+empirical testing in line with a recent interest in differences between various scalar
+implicature triggers \citep{Tielvan-TielMiltenburgvan-Miltenburg2014:Scalar-Diversit}.
+
+Our case study could be criticized as follows: all it shows is that scalar implicatures do not
+lexicalize because upper bounds are dispreferred concepts. This criticism would be too
+superficial and highly unjust. Dispreferred lexical concepts can thrive under evolutionary
+selection. Lexicalized upper-bounds can dominate a population because they may boost
+communicative efficiency. But they do not have to. Moreover, even without selective pressure
+for communicative efficiency, it is not the case that necessarily the types that are most
+likely \emph{a priori} will dominate. The dynamics of iterated learning are not that
+trivial. Iterated learning does not necessarily promote the \emph{a priori} more likely type,
+but tends to promote a type $t$ based on a gradient of how many other types might likely mutate
+into $t$, so to speak. Taken together, without an explicit model of the interaction between
+pressure for efficiency and learnability, it is far from trivial to judge whether or when
+preferred or dispreferred concepts can be adaptive. This is why a major contribution of this
+paper is the arrangement of many different ingredients into a joined model of the co-evolution
+of lexical meaning and pragmatic use.
+
+What is more, it is not that we just assumed a prior disadvantage of lexicalized upper
+bounds. We tried to motivate and formalize a general assumption about concepts' complexity
+witha a concrete, albeit provisional proposal. The specification of a learning bias in terms of
+a ``grammar of concepts'' can and should be seen critically, however. Much depends on the
+primitives of such a grammar. For instance, the concept ``none or all'' is the most complex in
+Table~\ref{tab:concepts}. But consider adding a primitive operation on sets $A \smile B$
+which is true if and only if $\neg(A \cap B \neq \emptyset \wedge A \neq \emptyset)$. The
+concept ``none or all'' would then be one of the simplest. Clearly, further research, empirical
+and conceptual, into the role of representational complexity, processing costs and learning
+biases is needed. The model here makes a clear and important contribution nonetheless: it shows
+how simplicity of concepts can interact with use and evolutionary selection to show that
+without pragmatic language users we would not see simpler concepts emerge in what may be a
+natural way. Future work should also include the possibility that conceptual simplicity may
+itself be a notion that is subject to evolutionary pressure
+\citep[c.f.][]{ThompsonKirby2016:Culture-Shapes-}.
+
+Finally, our case study should not be interpreted as a proposal for a definite explanation of
+how scalar implicatures evolved. Other factors should be considered eventually even if they
+will lead to much more complex modeling. One such factor is the observation that
+non-lexicalized upper bounds allow a broader range of applicability, e.g., when the speaker is
+not certain as to whether \emph{all} is true. This may suggest an alternative argument for why
+upper-bounded meanings do not conventionalize based on functional criteria only: should
+contextual cues provide enough information to the hearer to identify whether a bound is
+intended to be conveyed pragmatically, then this is preferred over expressing it overtly
+through longer expressions, e.g., by saying {\em some but not all} explicitly. Importantly,
+although morphosyntactic disambiguation may be dispreferred due to its relative length and
+complexity \citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and
+override contextual cues that might otherwise mislead the hearer. In a nutshell, this
+explanation posits that scalar implicatures fail to lexicalize because, all else being equal,
+speakers prefer to communicate as economically as possible and pragmatic reasoning enables them
+to do so. What this alternative explanation does not explain is why functional pressure does
+not lead to the emergence of different, equally costly lexical items to express different
+knowledge states of the speaker. Looking at pressure from learnability might come in again. The
+present work made a first start and gave a framework for exploring exactly these issues
+systematically.
 
 \section{Conclusion}
-The cultural evolution of meaning is influenced by intertwined pressures.
-We set out to investigate this process by putting forward a model that combines a pressure toward successful information transfer with perturbations that may arise in the transmission of linguistic knowledge in acquisition.
-Its objects of selection and replication are pairs of lexical meanings and patterns of language use.
-This allows the model to trace the interaction between conventional meaning and pragmatic use.
-Additionally, it takes the challenge serious of neither semantics nor pragmatics being directly observable.
-Instead, learners need to infer these unobservables from overt data that results from their combination.
-These components and their mutual influence were highlighted in a case study on the lack of lexical upper-bounds in weak scalar expressions that showed that, when pressured for learnability and expressivity, the former force drives for simpler semantic representations inasmuch as pragmatics can compensate for lack of expressivity in use.
-That is, the relative learning advantage of simpler semantics in combination with functional pressure in use may offer an answer to why natural languages fail to lexicalize systematic pragmatic inferences.
-And, more broadly, to a division of labors between semantics and pragmatics.
+The cultural evolution of meaning is influenced by intertwined pressures. We set out to
+investigate this process by putting forward a model that combines a pressure toward successful
+information transfer with perturbations that may arise in the transmission of linguistic
+knowledge in acquisition. Its objects of selection and replication are pairs of lexical
+meanings and patterns of language use. This allows the model to trace the interaction between
+conventional meaning and pragmatic use. Additionally, it takes the challenge serious of neither
+semantics nor pragmatics being directly observable. Instead, learners need to infer these
+unobservables from overt data that results from their combination.  These components and their
+mutual influence were highlighted in a case study on the lack of lexical upper-bounds in weak
+scalar expressions that showed that, when pressured for learnability and expressivity, the
+former force drives for simpler semantic representations inasmuch as pragmatics can compensate
+for lack of expressivity in use. That is, the relative learning advantage of simpler semantics
+in combination with functional pressure in use may offer an answer to why natural languages
+fail to lexicalize systematic pragmatic inferences. And, more broadly, to a division of labors
+between semantics and pragmatics.
 
@@
 
-\section{Snippets}
-
-\subsection{On alternative explanations}
-\label{sec:altern-expl}
-
-We see two main explanations for the lack of upper-bounds in the lexical meaning of weak scalar expressions.
-The first is that their truth-conditional compatibility with stronger expressions endows them with a broader range of applicability by allowing them to occur in contexts in which their upper-bounded reading is absent.
-This can happen when embedded in downward-entailing contexts, when the speaker is likely uncertain about whether the upper bounded reading is true, or when the distinction between an upper-bounded reading and the simple, only lower-bounded reading, is not relevant.
-For instance, if for all the speaker knows Chris owns some of Johnny Cash's albums but she does not know whether he owns all, then the use of {\em some} lacking an upper-bound succinctly conveys her uncertainty.
-This may suggest a functionalist argument for why upper-bounded meanings do not conventionalize: Should contextual cues provide enough information to the hearer to identify whether a bound is intended to be conveyed pragmatically, then this is preferred over expressing it overtly through longer expressions, e.g., by saying {\em some but not all} explicitly.
-Importantly, although morphosyntactic disambiguation may be dispreferred due to its relative length and complexity \citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and override contextual cues that might otherwise mislead the hearer.
-In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so.
-Compare this with a hypothetical language that lexicalizes two expressions for each weak scalar expression -- one with and one lacking an upper-bound.
-We see four conditions along this functionalist explanation that may pressure languages for English-like semantics over this alternative.
-First, contextual cues are very reliable.
-Second, morphosyntactic disambiguation is seldom necessary.
-Third, morphosyntactic disambiguation is only marginally dispreferred.
-Fourth, larger lexica are costly.
-Overall, neither condition seems convincing as a central explanatory device for such a widespread phenomenon.
-The first two conditions put a heavy burden on the ability to retrieve contextual cues to a degree that seems unlikely to undercut the benefit of unambiguous communication.
-It is likely that human language users are very good at retrieving cues from context, but to stipulate that they are so good as to undercut the benefit of safe communication provided by this hypothetical alternative strikes us as too strong of an assumption.
-As for the third and fourth condition, these seem mostly like technical solutions without a proper empirical basis.
-
-Instead, the systematicity and typological spread of scalar implicatures together with the observation that monomorphemic expressions that lexically rule out stronger alternatives are unattested across languages (\citealt[252-267]{horn:1984}, \citealt{horn:1972,traugott:2004,vdAuwera:2010}) suggests that other forces may be at play.
-In what follows we investigate the predictions of our model under the assumption that the lack of lexicalization of scalar inferences may be accounted for by the relative representational simplicity of lexical meanings lacking an upper-bound over those that explicitly codify it.
-This difference is reflected in a learning bias towards more compressed lexical representations.
-That is, in a preference of learners for simpler over more complex explanations of the data they witness \citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a, kirby+etal:2015,piantadosi+etal:underreview}.
-While we do not want to argue that functional aspects as the ones discussed above do not play a role, we do see a clear benefit in exploring whether matters of transmission biases would not give us additional explanatory leverage.
+\section{notes}
+
+\begin{itemize}
+\item \mf{The text frequently uses the term ``expressivity'' and ``pressure for/towards
+    expressivity''. I have mixed feelings about this. First of all, I don't know what
+    ``expressivity'' really is. It seems to relate to a lexicon: how much information can the
+    lexicon convey about the states. But we do not model selection of expressivity, but
+    selection of communicative success relative to the population state. The difference is very
+    pronounced: there can be several maximally expressive lexica with huge differences in their
+    communicatively success or expected utility for a given population state. We only
+    indirectly select for expressivity, and we don't necessarily select for expressivity of a
+    lexicon. \dots It may be that ``expressivity'' is very much Edinburgh terminology where
+    exactly the picture of communication success in the current population is missing. I might
+    be wrong though. Possibly, we should introduce this distinction (with this terminology) and
+    use it also in the general discussion to relate our work to those articles from the
+    iterated-learning community that bring in some pressure for ``expressivity'' but not
+    communicative efficiency (if that is what they do, I don't know?).}
+\end{itemize}
 

### 2017-03-24 paper/paper-v3.tex
main revision - some comments and changes in plots missing still

@@
   study on scalar implicatures. In this case study learning biases that favor simple semantic
-  representations can boost the evolution of more sophisticated pragmatic reasoning types and
+  representations can foster the evolution of more sophisticated pragmatic reasoning types and
   so prevent the lexicalization of scalar implicatures.
@@
 the utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
-that the speaker does not own all of them. This is because, if the speaker owned them all, she
+that the speaker does not own all of them. This is because, if the speaker instead owned them all, she
 could have used the word \emph{all} instead of \emph{some} in his utterance, thereby making a
@@
 model of evolutionary change in large and homogeneous populations
-\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. The
+\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. This
 approach allows us to study the interaction between (i) evolutionary pressure towards
@@
 language. Learnability is another. Natural languages need to be learnable to survive their
-faithful transmission across generations. Even small learning biases implicit in acquisition
+faithful transmission across generations. Furthermore, even small learning biases implicit in acquisition
 can build up and have quite striking effects on an evolving language in a process of iterated
@@
 
-While natural languages are pressured for both expressivity and learnability these forces may
+While natural languages are pressured for both expressivity and learnability, these forces may
 pull in opposite directions. Their opposition becomes particularly clear when considering the
@@
 \begin{example}
-  Consider a simple coordination game. Agents are of two types: positive and negative. If
+  Consider a simple coordination game. Agents are of two types: positive or negative. If
   agents of different type play with each other, they obtain a payoff of 0. If negative meets
@@
 relevant types are, (ii) how fitness derives from communicative success and (iii) how the
-mutation matrix is computed. These issues are addressed, one by one, in the following.
+mutation matrix $Q$ is computed. These issues are addressed, one by one, in the following.
 
@@
 (possibly fuzzy) set of states. A pragmatic type maps a lexicon onto a probabilistic speaker
-rule (a probabilistic choice of message for each state) and a probabilistic listener rule (a
+rule (a probabilistic choice of message for each state) and a probabilistic hearer rule (a
 probabilistic choice of state for each message). There are many ways of making these general
@@
 
-Pragmatic types define dispositions of produce and interpret messages given a lexicon. We
+Pragmatic types define dispositions to produce and interpret messages given a lexicon. We
 distinguish between two kinds of pragmatic types. {\em Literal interlocutors} produce and
@@
 
-Here are some concrete examples. \mf{please double-check the numbers!} A literal interpreter with lexicon $\Lbound$ assigns $\ssome$
+The following examples illustrate these behaviors. A literal interpreter with lexicon $\Lbound$ assigns $\ssome$
 a probability of $H_0(\ssome \mid \msome; \Lbound) = 1$ after hearing $\msome$, while a literal
@@
 \end{align*}
-If $\lambda = 20$, literal speakers will choose with less slack:
+If $\lambda = 20$, literal speakers will choose true utterances with less slack:
 \begin{align*}
@@
 \end{align*}
-For $\lambda = 20$, there wil less slack again:
-Pragmatic listeners of level 1 have the following probabilistic interpretation behavior for
-$\lambda = 1$:
+For $\lambda = 20$, there will less slack again:
 \begin{align*}
@@
     \begin{block}{l[cc]}
-      \mygray{\ssome} & .73 & .37 \\
-      \mygray{\sall}  & .37 & .73 \\
+      \mygray{\ssome} & .73 & .27 \\
+      \mygray{\sall}  & .27 & .73 \\
     \end{block}
@@
 association is not perfect, and usually less strong than what agents with a lexicalized upper
-bound in $\Lbound$ can achieve, even without pragmatic reasoning. Higher oder iteration of
+bound in $\Lbound$ can achieve, even without pragmatic reasoning. Higher order 
 reasoning types leads to stronger associations of $\msome$ and $\ssome$ also for the
@@
 \end{align*}
-The expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
+In signaling games, the expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
 average success of $i$ when talking or listening to $j$. If agents are speakers half of the
@@
 
-\subsubsection{States, messages, concepts and Lexica} 
+\subsubsection{States, messages, concepts and lexica} 
 \label{sec:stat-mess-conc}
@@
 where $Compl(c)$ is the complexity of the minimal derivation cost of $c$ according to the
-LOT-grammar (see Table~\ref{tab:concepts}). This construal assigns the highest probability the
-above example for a lexicon of type $\Lall$, which only uses the simplest concept ``all'' for
+LOT-grammar (see Table~\ref{tab:concepts}). Applied to our space of lexica, this construal assigns the highest probability 
+to a lexicon of type $\Lall$, which only uses the simplest concept ``all'' for
 all messages. Lexica of type $\Llack$ are less likely, but more likely than $\Lbound$.
@@
 Recall that there are three parameters: soft-max parameter $\lambda$ affects how strongly
-pragmatic speakers favor messages that appear best from their subjective point of view; the
+speakers favor messages that appear best from their subjective point of view; the
 bottleneck size $k$ influences how faithfully learners can identify their teacher type; $l$
@@
 \includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig1-onlyr}
-\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other
+\caption{Stacked proportion of pragmatic $\Llack$-style types and majority types, when other
   than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a
-  pressure for expressivity. The right-most plots zoom in on only the proportion of pragmatic
-  $\Llack$ types. \mf{not sure if ``incumbent'' is the right term here; I believe it means
-    ``someone who is already there and established'' but we mean ``type with the highest
-    proportion''; I used ``majority type'' in the main text; maybe change the plot?}}
+  pressure for expressivity. The right-most plots zoom in on only the proportion of target types. \tb{change plot to maj. type}}
 \label{fig:only-R}
@@
 linguistic behavior, widening the gap in expressivity between types and promoting more
-homogeneous populations. As suggested by Figure \ref{fig:only-R}, the incumbent in most
+homogeneous populations. As suggested by Figure \ref{fig:only-R}, the majority in most
 populations is not one of the six pragmatic $\Llack$-style types. That is, a pressure only for
 expressivity does not lead to a prevalence of target types under any $\lambda$-value. For
-instance, in $1000$ independent populations with $\lambda = 20$ only had $11$ cases in which
+instance, with $\lambda = 20$ $1000$ independent populations only had $11$ cases in which
 the target type was the majority type, corresponding to a mean proportion of $0.003$ across
@@
 \includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig2-onlym-pr}
-\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other
+\caption{Stacked proportion of pragmatic $\Llack$-style types and majority types, when other
   than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a
   pressure for learnability ($\lambda = 20, k = 5$). The types' prior probability is shown in
-  the right-most plot. \mf{explain the topmost part of the right plot?!}}
+  the right-most plot. \mf{explain the topmost part of the right plot?!} \tb{and change to maj. type in plot}}
 \label{fig:only-M}
@@
 for communicative success, they do not stand in competition with each other. In $1000$
-independent populations all majority types were target types provided, sufficiently high
-$\lambda$ \mf{don't understand this proviso: what $\lambda$ values were used in the 1000
-  runs? what's sufficiently high?}, with each reaching approximately the same proportion of users in the population. As
+independent populations with $\lambda = 20$ all majority types were target types,
+with each reaching approximately the same proportion of users in the population. As
 with expressivity only, low values of $\lambda$ make the differences in observable behavior
@@
 suboptimal types that are easier to learn. In the extreme, when $l = 1$ and $\lambda = 1$ all
-of $1000$ independent populations had users of $\Lall$ as incumbents.
+of $1000$ independent populations had users of $\Lall$ as majority types.
 
@@
 learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more
-complex alternatives such as $\Lbound$. However, on its own learnability does not foment the
+complex alternatives such as $\Lbound$. However, learnability alone does not foment the
 propagation of a single target type across the population.
@@
 \includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig3-r+m}
-\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under both pressures ($k = 5$).}
+\caption{Stacked proportion of pragmatic $\Llack$-style types and majority types, when other than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under both pressures ($k = 5$).
+\tb{change plot to maj.
+types}}
 \label{fig:rmd}
@@
 target types but does not lead to monomorphic populations if learners sample from the
-posterior. Finally, a combination of high $\lambda$ and $l$, leads to increasing proportions of
+posterior. Finally, a combination of high $\lambda$ and $l$ leads to increasing proportions of
 a single majority target type. The difference between the mean of the highest pragmatic
@@
   other type in $1000$ independent populations after $50$ generations under both pressures ($k
-  = 5$). \tb{I will run more simulations with higher $\lambda$ and $l$. I'm pretty sure the
-    difference still increases.} \mf{I think it's okay like that} \mf{the coloring is not that
+  = 5$).  \mf{the coloring is not that
     easy to decipher here; would a two-dimensional surface be possible and look better?}}
@@
 
-\mf{can this abstract be shortened (by a factor of $.4$)?} The interaction of expressivity and
+\mf{can this abstract be shortened (by a factor of $.4$)?} \tb{See notes at end} The interaction of expressivity and
 learnability in language evolution has also recently been addressed by
@@
 bounds. We tried to motivate and formalize a general assumption about concepts' complexity
-witha a concrete, albeit provisional proposal. The specification of a learning bias in terms of
+with a concrete, albeit provisional proposal. The specification of a learning bias in terms of
 a ``grammar of concepts'' can and should be seen critically, however. Much depends on the
@@
 itself be a notion that is subject to evolutionary pressure
-\citep[c.f.][]{ThompsonKirby2016:Culture-Shapes-}.
+\citep[cf.][]{ThompsonKirby2016:Culture-Shapes-}.
 
@@
 in combination with functional pressure in use may offer an answer to why natural languages
-fail to lexicalize systematic pragmatic inferences. And, more broadly, to a division of labors
+fail to lexicalize systematic pragmatic inferences. And, more broadly, to a division of labor
 between semantics and pragmatics.

### 2017-03-27 paper/paper-v3.tex
revision text

@@
 
-\subsection{Expressivity and learnability}
+\subsection{Communicative success and learnability}
 
@@
 
-While natural languages are pressured for both expressivity and learnability, these forces may
+While natural languages are pressured for both communicative success and learnability, these forces may
 pull in opposite directions. Their opposition becomes particularly clear when considering the
@@
 
-An elegant formal approach to capture the interaction between expressivity and learnability is
+An elegant formal approach to capture the interaction between communicative success and learnability is
 the \emph{replicator mutator dynamic}
@@
 
-\subsubsection{Fitness \& fitness-based selection based on expressivity}\label{sec:expressivity}
+\subsubsection{Fitness \& fitness-based selection based on communicative success}\label{sec:expressivity}
 
@@
 decrease as a function of its relative fitness $f_i$. In the context of language evolution,
-fitness is usually associated with expressivity, i.e., the ability to successfully communicate
+fitness is usually associated with the ability to successfully communicate
 with other language users from the same population
@@
 
-Languages are shaped not only by functionalist forces towards greater expressivity. Another
+Languages are shaped not only by functionalist forces towards greater communicative success. Another
 important factor is the fidelity with which linguistic knowledge is transmitted. Among others,
@@
 
-Expressivity and learnability are central to the cultural evolution of language. These
+Communicative success and learnability are central to the cultural evolution of language. These
 components can be modelled, respectively, as replication based on a measure of fitness in terms
@@
 likelihood of being imitated or repeated) evolution, since the types that the dynamic operates
-on are pairs consisting of a lexicon and a pragmatic use pattern. A type's expressivity depends
-on its communicative efficiency within a population while its learnability depends on the
+on are pairs consisting of a lexicon and a pragmatic use pattern. A type's communicative success depends
+on how well it communicates within a population while its learnability depends on the
 fidelity by which it is inferred by new generations of learners. The learners' task is
@@
 motivate and formalize a preference of learners for simpler semantic representations
-\citep{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a,
-  kirby+etal:2015,piantadosi+etal:underreview}. A concept's complexity is a function of its
+(\citealt{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a,kirby+etal:2015}, \citeauthor{piantadosi+etal:underreview} under review). A concept's complexity is a function of its
 derivation cost in a weighed generative LOT.
@@
   than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a
-  pressure for expressivity. The right-most plots zoom in on only the proportion of target types. \tb{change plot to maj. type}}
+  pressure for communicative success. The right-most plots zoom in on only the proportion of target types. \tb{change plot to maj. type}}
 \label{fig:only-R}
@@
 populations is not one of the six pragmatic $\Llack$-style types. That is, a pressure only for
-expressivity does not lead to a prevalence of target types under any $\lambda$-value. For
+communicative success does not lead to a prevalence of target types under any $\lambda$-value. For
 instance, with $\lambda = 20$ $1000$ independent populations only had $11$ cases in which
@@
 
-The effect of iterated learning without a pressure for expressivity using either posterior
+The effect of iterated learning without a pressure for communicative success using either posterior
 sampling ($l = 1$) or a stronger tendency towards posterior maximization ($l = 15$) is shown in
@@
 
-However, in contrast to a pressure only for expressivity with high $\lambda$ (see
+However, in contrast to a pressure only for communicative success with high $\lambda$ (see
 Figure~\ref{fig:only-R}), learnability alone does not succeed in selecting for a single
@@
 with each reaching approximately the same proportion of users in the population. As
-with expressivity only, low values of $\lambda$ make the differences in observable behavior
+with a pressure only for communicative success, low values of $\lambda$ make the differences in observable behavior
 across types less pronounced and therefore reflect the learners' inductive bias more
@@
 
-\subsubsection{Combining pressures of expressivity and learnability}
+\subsubsection{Combining pressures of communicative success and learnability}
 
 Pressures for communicative success or learnability are not sufficient on their own to have a
-single target type dominate the population. When pressured for expressivity, the slight
+single target type dominate the population. When pressured for communicative success, the slight
 communicative advantage of $\Lbound$ users leads to the its prevalence. When pressured for
@@
 
-To recapitulate, other than the involvement of pressure on both expressivity and learnability,
+To recapitulate, other than the involvement of pressure on both communicative success and learnability,
 the resulting proportion of pragmatic $\Llack$ speakers primarily hinges on three
@@
 dominate the population if three assumptions are met: (i) language is pressured toward both
-expressivity and learnability; (ii) pragmatic language use is an option; (iii) learners prefer
+communicative success and learnability; (ii) pragmatic language use is an option; (iii) learners prefer
 simpler over more complex lexical representations and exhibit a tendency towards the
@@
 co-evolution of conventional meaning and pragmatic language use. Main contributions of the
-model are (i) its modular separation of expressivity and learnability on evolutionary
+model are (i) its modular separation of communicative success and learnability on evolutionary
 trajectories, (ii) the characterization of language learning as a joint inference over
@@
 
-\mf{can this abstract be shortened (by a factor of $.4$)?} \tb{See notes at end} The interaction of expressivity and
-learnability in language evolution has also recently been addressed by
-\citet{kirby+etal:2015}. In contrast to our proposal, Kirby et al. model expressivity as
-exerting its force only in the production of learning data. The degree of mutual understanding
-of interlocutors central to replication and to our notion of expressivity is absent. That is,
-while our proposal combines bidirectional horizontal transmission with its vertical and
-unidirectional counterpart, Kirby et al. only consider the latter's influence. Our reasoning
-behind the inclusion of the former lies in the empirical and theoretical observation that
-learnability alone can lead to the selection of functionally defective languages, as showcased
-by $\Lall$. This outcome has been reported in a number of laboratory experiments where subjects
-learned and subsequently reproduced the language produced by a previous participant over
-multiple iterations, leading to a proliferation of ambiguous languages that associate many
-meanings with a single form (see, e.g., \citealt{silvey+etal:2014} and experiment 1 in
-\citealt{kirby+etal:2008}). In contrast, experiments involving an interactive component have
-been found to foster languages that enable interlocutors to communicatively distinguish
-meanings more accurately (e.g., \citealt{fay+etal:2013}; for a review of laboratory results
-under the iterated learning paradigm and further discussion see \citealt{kirby+etal:2015,
-  tamariz+kirby:2016}). It is not evident how to compare these empirical findings given that
-they consider distinct meaning spaces, modes of transmission, iterations and feedback given to
-participants. However, on a general level they may suggest that there is an important
-difference between a language generating learnable linguistic data and its actual performance
-as a means of information transfer. That is, we contend that successful information transfer in
-a linguistic community is central to the adoption of a communication system and that this
-measure may not be adequately reflected by production alone.
+With respect to (i), \citet{kirby+etal:2015} propose a comparable model of the interaction
+between a lexicon's expressivity and its learnability. One of its main differences to our proposal lies in 
+the former force exerting pressure only in the production of learning data. The pressure towards communicative success 
+within a population central to replication is absent. Put differently, we consider a pressure for mutual understanding 
+that may indirectly select for expressive types, i.e., those that can convey states unequivocally, whereas Kirby et al. only
+consider the bearing that the latter ability has on the production of learnable data. We see two main reasons
+for considering communicative success rather than expressivity, and to only considering it in learning. First, learnability alone can 
+lead to the acquisition of functionally defective languages, as showcased by $\Lall$ (cf.
+\citealt{kirby+etal:2008,silvey+etal:2014}, and \citealt{fay+etal:2013} for review of laboratory results).
+Second and more importantly, types may be equally expressive but their performance as means of information transfer crucially depends not only on themselves but on the population they find themselves in (cf.
+the competition of types of a kind in Figure \ref{fig:rmd} and their lack of competition in Figure \ref{fig:only-M}).
+That is, we contend that having an expressive lexicon that generates learnable data does not  in itself capture a type's arguably central communicative function of transferring information to peers.
+Under this view, a type's expressiveness is secondary to its ability to communicative successfully.
+Acquiring a maximally expressive type is of little use when it is not understood in its population.
 
@@
 not. Interestingly, in our simulations when a target type is the majority type an inflated
-proportion of the population uses compatible lexica with a lexicalized upper bound. \mf{Thomas,
-  please check this; I'm making it up; total conjecture; it might be true but I don't know for
-  sure} In other words, we do find a similar co-existence of semantic and pragmatic
-types. Whether this analogy has any further explanatory value is an interesting path for future
+proportion of the population uses compatible lexica with a lexicalized upper bound. Particularly in those parameter
+settings where the prior influences the outcome less. In other words, we do find a tendency toward a similar co-existence
+of semantic and pragmatic types. Whether this analogy has any further explanatory value is an interesting path for future
 exploration.
+\tb{I some qualifications.
+While it is indeed true that $\Lbound$'s proportion is inflated relative to other types in the hypothesis space, it is never the second most prolific type (which is what the above hopefully is not taken to suggest, otherwise it should be changed).
+As is to be expected, $\Lbound$ fares best when $l$ is low and $\lambda$ is high.
+Even then, it usually ranks around $5$-th with a mean proportion of its kind of about $0.01 - 0.05$.
+As $l$ increases, it gets pushed back.
+Second or third place are usually instead literal $\Llack$ or types that lexicalize ``all'', ``not all'' and ``none''.
+In short; yes, there is a inflated proportion but only compared to the whole population.
+Other particular types are usually more prolific.
+This is very likely due to the strong prior disadvantage of $\Llack$.
+In light of this I'm uncertain whether we want to keep this paragraph as is.
+I tend towards `yes' but let me know if you think otherwise.}
 
@@
 mutual influence were highlighted in a case study on the lack of lexical upper-bounds in weak
-scalar expressions that showed that, when pressured for learnability and expressivity, the
+scalar expressions that showed that, when pressured for learnability and communicative success, the
 former force drives for simpler semantic representations inasmuch as pragmatics can compensate

### 2017-03-27 code-LOT-extension/plots.py

@@ -432,6 +432,7 @@ def get_heatmap_diff_incumbents(type_list,list1,list2,kind):
 
 types = [231,236,291,306,326,336]
 
+
 ##Plot 1
 #list1 = [1,20] #lambda
 #list2 = [5] #posterior parameter
@@ -443,9 +444,9 @@ types = [231,236,291,306,326,336]
 #get_subfigs_mutation(types,list1,list2)
 
 ##Plot 3
-list1 = [1,5,20]
-list2 = [1,5,15]
-get_subfigs_rmd(types,list1,list2)
+#list1 = [1,5,20]
+#list2 = [1,5,15]
+#get_subfigs_rmd(types,list1,list2)
 
 ##Plot 4
 #kind = 'rmd'
@@ -453,4 +454,64 @@ get_subfigs_rmd(types,list1,list2)
 #list1 = [x for x in xrange(1,21)]
 #list2 = [x for x in xrange(1,16)]
 #get_heatmap_diff_incumbents(type_list,list1,list2,kind)
+
+
+##Who is the second largest beyond types?
+##This gives the mean of lbound-types across parameter values but doesn't say much about how well they compare to other types
+lbound = [225, 235, 255, 270, 325, 330]
+lbound = lbound + [x-216 for x in lbound]
+#id_bound = ['t_final'+str(x) for x in lbound]
+#
+#for lam in [1,5,10,15,20]:
+#    for l in [1,5,15]:
+#        df = pd.read_csv('./results/%s-s3-m3-lam%d-a1-k%d-samples250-l%d-g50-meFalse.csv' % ('rmd',lam,5,l))
+#        restrict_to_final = ['t_final'+str(z) for z in xrange(432)] #as to avoid the incumbent to come from other columns
+#        df = df[restrict_to_final]
+#        
+#        prop_bounds = 0 
+#        for row in xrange(len(df)):
+#            prop_bounds += sum([df.iloc[row][idb] for idb in id_bound])
+#        prop_bounds = prop_bounds / len(df)
+#        print '###'
+#        print 'lambda: ', lam, 'l: ', l
+#        print 'Summed proportion of lbound: ', prop_bounds
+#        print '###'
+        
+##Going by kinds (bins) instead:
+from lexica import get_lexica,get_lexica_bins
+from rmd import get_type_bin
+
+lex = get_lexica(3,3,False)
+bins = get_lexica_bins(lex)
+
[…]

### 2017-03-27 code-LOT-extension/plots.py
fixed label and spacing of plots

@@ -114,7 +114,7 @@ def get_subfigs_replication(type_list,list1,list2):
 
     p1 = patches.Rectangle((0, 0), 1, 1, fc="green", alpha=al)
     p2 = patches.Rectangle((0, 0), 1, 1, fc="darkred",alpha=al)
-    ax1_large.legend([p1, p2], ['prag. L-lack','Other incumbent'],loc='best',prop={'size':14})
+    ax1_large.legend([p1, p2], ['prag. L-lack','Other majority'],loc='best',prop={'size':14})
 
     ax2_large.yaxis.set_major_formatter(FormatStrFormatter('%.2f'))
     ax2_large.tick_params(axis='both',which='major',labelsize=17)
@@ -206,13 +206,32 @@ def get_subfigs_mutation(type_list,list1,list2):
     Y_lbound = [0 for _ in xrange(len(priors)+white_space*3)]
     Y_all = [0 for _ in xrange(len(priors)+white_space*3)]
 
-    for t in xrange(len(targets)):
-        Y_target[t] = priors[targets[t]]
+#    for t in xrange(len(targets)):
+#        Y_target[t] = priors[targets[t]]
+    #To interspread white space (done manually for convenience)
 
-    for t in xrange(len(lbound)):
-        Y_lbound[len(targets)+white_space+t] = priors[lbound[t]]
-    for t in xrange(len(lall)):
-        Y_all[len(targets)+white_space+len(lbound)+white_space+t] = priors[lall[t]]
+    Y_target[0] = priors[targets[0]]
+    Y_target[2] = priors[targets[1]]
+    Y_target[4] = priors[targets[2]]
+    Y_target[6] = priors[targets[3]]
+    Y_target[8] = priors[targets[4]]
+    Y_target[10] = priors[targets[5]]
+
+
+#    for t in xrange(len(lbound)):
+#        Y_lbound[len(targets)+white_space+t] = priors[lbound[t]]
+    Y_lbound[6+17+0] = priors[lbound[0]]
+    Y_lbound[6+17+2] = priors[lbound[1]]
+    Y_lbound[6+17+4] = priors[lbound[2]]
+    Y_lbound[6+17+6] = priors[lbound[3]]
+    Y_lbound[6+17+8] = priors[lbound[4]]
+    Y_lbound[6+17+10] = priors[lbound[5]]
+
+#    for t in xrange(len(lall)):
[…]

### 2017-03-27 code-LOT-extension/plots.py
plots draft

@@ -479,8 +479,27 @@ for i in xrange(len(list1)):
         data[i+1,j+1] = best_of_targets - best_of_others
 dPrime = data[1:,1:]
 
+from mpl_toolkits.mplot3d import Axes3D
+fig = plt.figure()
+ax = fig.add_subplot(111, projection='3d')
+
+xx,yy= np.meshgrid(range(len(data[0,1:])),range(len(data[1:,0])))
+ax.plot_surface(xx,yy,dPrime,cmap='summer',alpha=0.8) #'PrGn', 'winter', 'autumn'
+sns.set(font_scale=2)
+
+ax.set_xticklabels([x for x in xrange(1,16)])
+ax.set_yticklabels([x for x in xrange(1,21)])
+ax.set_ylabel('rationality parameter ('+r'$\lambda$'+')',fontsize=15)
+ax.set_xlabel('posterior parameter (l)',fontsize=15)
+ax.set_zlabel('Difference',fontsize=15)
+
+plt.show()
+
+#for row in xrange(np.shape(dPrime)[0]):
+#    ax.bar(data[0,1:], dPrime[row], zs=int(data[1:,0][row]),zdir='y')
+
+
 #    sns.set(font_scale=2)
-#    ax = sns.heatmap(dPrime, cmap='YlGnBu', xticklabels=list2, yticklabels=list1, annot_kws={"size": 20})# xticklabels=axlabels)#, annot=True) 
         
 #    ax.set_ylabel('rationality parameter ('+r'$\lambda$'+')',fontsize=30)#,fontsize=30)
 #    ax.set_xlabel('posterior parameter (l)',fontsize=30)#,fontsize=30)

### 2017-03-28 paper/paper-v3.tex
TB: full revision + new differences-plot. Version sent to RvR

@@
 pragmatic use. What is more, many mathematical models explain evolved meaning as a regularity
-in the overt behavior of agents. In contrast, we will here look at language users with a richer
+in the overt behavior of agents, abstracting from complex interactions between semantic representations and pragmatic use. In contrast, we will here look at language users with a richer
 cognitive make-up.
@@
 
-\subsection{Communicative success and learnability}
+\subsection{Communicative efficiency and learnability}
 
@@
 pressure towards communicative efficiency is not the only force that shapes
-language. Learnability is another. Natural languages need to be learnable to survive their
+language. Learnability is another as natural languages need to be learnable to survive their
 faithful transmission across generations. Furthermore, even small learning biases implicit in acquisition
@@
 
-While natural languages are pressured for both communicative success and learnability, these forces may
+While natural languages are pressured for both communicative efficiency and learnability, these forces may
 pull in opposite directions. Their opposition becomes particularly clear when considering the
@@
 
-An elegant formal approach to capture the interaction between communicative success and learnability is
+An elegant formal approach to capture the interaction between communicative efficiency and learnability is
 the \emph{replicator mutator dynamic}
@@
 
-  Adding mutation changes the dynamic and its rest points. Let's assume that $Q_{ji} = .9$
+  Adding mutation changes the dynamic and its rest points. Let us assume that $Q_{ji} = .9$
   when $j=i$. The update effect of mutation alone is $M(x) = .9 x + .1 (1-x) = .8x + .1$ and is
@@
   \centering
-
   %% notice that the figure is hacked, because somehow the curve does not come out right under
@@
 
-        \caption{Update functions: the population state $x$ is mapped onto $x'$ in one update step}
+        \caption{Update functions: the population state $x$ is mapped onto $x'$ in one update step.}
         \label{fig:Updates_RMS}
@@
 concepts and ways of using them in communication, a type is here defined as a pair consisting
-of a lexicon and a pragmatic strategy.
+of a lexicon and a pragmatic strategy of language use.
 
@@
 simplicity. Literal interpreters thereby choose an arbitrary true interpretation for each
-message. Pragmatic hearers, defined in (\ref{h:leveln}), use Bayes rule to weigh
+message. Pragmatic hearers, defined in (\ref{h:leveln}), instead use Bayes' rule to weigh
 interpretations based on a conjecture about speaker behavior. Speaker behavior is regulated by
@@
 \end{align*}
-If $\lambda = 1$ literal speaker behavior looks like this:
+If $\lambda = 1$ literal speaker behavior is as follows:
 \begin{align*}
@@
 
-It is important to note that pragmatic agents of level 1 using $\Llack$ associate $\msome$
+It is particularly important to note that pragmatic agents of level 1 using $\Llack$ associate $\msome$
 preferentially with $\ssome$, in contrast to their literal counterparts of level 0. This
@@
 fitness-based selection of pragmatic language use for a notion of fitness based on
-communicative efficiency, which is introduced next.
+communicative success, which is introduced next.
 
@@
 The model has three parameters: $\lambda$ regulates the degree to which pragmatic speakers
-choose messages that maximize the chance of communicative success; $k$ is the number of
+choose messages that maximize the chance of communicative success; $k$ is length of
 observations for each language learner; $l$ regulates where the learners' inference behavior
@@
 
-Recall that the majority view is that scalar implicatures are non-lexicalized pragmatic
+Recall that the majority view is that scalar implicatures are systematic non-lexicalized pragmatic
 enrichments. Scalar implicature triggers like \emph{some}, \emph{warm} or \emph{may} are
@@
 reflect on its limitations in Section~\ref{sec:discussion}. The setup is introduced in
-Section~\ref{sec:setup:-stat-conc}. Section~\ref{sec:simulation-results} describes the simulations
+Section~\ref{sec:setup:-stat-conc}. Section~\ref{sec:simulation-results} describes simulations
 and their results.
@@
 represent lexica and language use that conform to the majority view of scalar
-implicatures. Twelve types are either literal or pragmatic speakers with lexical of the
+implicatures. Twelve types are either literal or pragmatic speakers with lexica of the
 $\Lbound$ kind. We refer to these as \emph{competitor types}, because they are expected to be
-the target types' main competitors in evolutionary competition.
+the target types' main contenders in evolutionary competition.
 
@@
 (LOTs) have been put to test in various rational probabilistic models that show encouraging
-results (see, e.g., \citealt{katz+etal:2008, piantadosi+etal:underreview,
-  piantadosi+etal:2012a} and \citealt{piantadosi+jacobs:2016} for recent discussion). At its
+results (see, e.g., \citealt{katz+etal:2008}, \citeauthor{piantadosi+etal:underreview} under review; 2012, and \citealt{piantadosi+jacobs:2016} for recent discussion). At its
 core, a LOT defines a set of operations and composition rules from which concepts can be
@@
 motivate and formalize a preference of learners for simpler semantic representations
-(\citealt{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a,kirby+etal:2015}, \citeauthor{piantadosi+etal:underreview} under review). A concept's complexity is a function of its
-derivation cost in a weighed generative LOT.
+(\citealt{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a,kirby+etal:2015}, \citeauthor{piantadosi+etal:underreview} under review). 
+In a weighed generative LOT a concept's complexity is a function of its derivation cost.
 
@@
 set-theoretic operations to form expressions which can be evaluated as true or false in states
-$\snone$, $\ssome$ or $\sall$ from above. Applications of generative rules have a cost attached
+$\snone$, $\ssome$, or $\sall$ from above. Applications of generative rules have a cost attached
 to them. Here we simply assume that the formation of Boolean combinations of concepts incurs 2
@@
 
-The complexity measure for lexical concepts from Table~\ref{tab:concepts} is used to define a
+A complexity measure for lexical concepts from Table~\ref{tab:concepts} is used to define a
 learning bias that favors simpler concepts over more complex ones. The prior probability of a
 type is just the prior probability of its lexicon. The prior of a lexicon is a function of the
-complexity of the lexical representations in its image set. Lexica with simpler concepts have a
+complexity of the lexical representations in its image set. Lexica with simpler concepts accordingly have a
 higher prior. One simple way of defining such priors over lexica (and thereby types) is:
@@
 speakers favor messages that appear best from their subjective point of view; the
-bottleneck size $k$ influences how faithfully learners can identify their teacher type; $l$
+bottleneck size $k$ influences how faithfully learners tend to identify their teacher type; $l$
 defines the learners' disposition towards choosing the most likely teacher type from the
@@
 lexica of the kind $\Llack$. Simulation results will shed light on the question whether target
-types can emerge and for which parameter constellations.
+types can emerge, and for which parameter constellations.
 
@@
 
-\subsubsection{Replication only: selection based on expressivity and communicative success}
+\subsubsection{Replication only: selection based on communicative success}
 
@@
 \includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig1-onlyr}
-\caption{Stacked proportion of pragmatic $\Llack$-style types and majority types, when other
-  than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a
-  pressure for communicative success. The right-most plots zoom in on only the proportion of target types. \tb{change plot to maj. type}}
+\caption{Stacked proportion of target types and majority types, when other
+  than target type, in $5$ independent populations after $50$ generations under only a
+  pressure for communicative success. The right-most plots zoom in on only the proportion of target types.}
 \label{fig:only-R}
@@
 instance, with $\lambda = 20$ $1000$ independent populations only had $11$ cases in which
-the target type was the majority type, corresponding to a mean proportion of $0.003$ across
+the target type was the majority type, corresponding to a mean proportion of $.003$ across
 populations. By contrast, in $913$ cases the majority types had $\Lbound$ with close to an even
@@
 \includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig2-onlym-pr}
-\caption{Stacked proportion of pragmatic $\Llack$-style types and majority types, when other
-  than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a
-  pressure for learnability ($\lambda = 20, k = 5$). The types' prior probability is shown in
-  the right-most plot. \mf{explain the topmost part of the right plot?!} \tb{and change to maj. type in plot}}
+\caption{Stacked proportion of target types and majority types, when other than target type, in $5$ independent populations after $50$ generations under only a pressure for learnability ($\lambda = 20, k = 5$).
+The types' prior is shown in
+  the right-most plot. The top-most groupings correspond to types with lexica of the kind of $\Llack$ (green), $\Lbound$ (magenta), and $\Lall$ (gray).}
 \label{fig:only-M}
@@
 enables its users to convey each state with a single message when combined with pragmatic
-reasoning provided sufficiently high $\lambda$. This makes it less likely to be confused with
+reasoning and sufficiently high $\lambda$. This makes it less likely to be confused with
 other types if the learning data is not too sparse ($k \geq 5$). Put differently, learners have
 a higher propensity to infer pragmatic $\Llack$ when the teacher's type produces very similar
-data, such as when using $\Lbound$. Moreover, $\Llack$ is also less likely to be confused with
+data, such as when using $\Lbound$. Moreover, $\Llack$ is less likely to be confused with
 types with different observable behavior because its pragmatic use approximates a one-to-one
@@
 
-Pressures for communicative success or learnability are not sufficient on their own to have a
+Pressure for communicative success or learnability are not sufficient on their own to have a
 single target type dominate the population. When pressured for communicative success, the slight
@@
 \includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig3-r+m}
-\caption{Stacked proportion of pragmatic $\Llack$-style types and majority types, when other than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under both pressures ($k = 5$).
-\tb{change plot to maj.
-types}}
+\caption{Stacked proportion of target types and majority types, when other than target type, in $5$ independent populations after $50$ generations under both pressures ($k = 5$).}
 \label{fig:rmd}
@@
 semantics in tandem with functional pressure can lead to the selection of a single target
-type. The proportion of a single majority target type increases with $\lambda$ and
+type, and so to a division of labor between semantics and pragmatics.
+The proportion of a single majority target type increases with $\lambda$ and
 $l$. Pressure for communicative success magnifies the effects of iterated learning and dampens
@@
 posterior. Finally, a combination of high $\lambda$ and $l$ leads to increasing proportions of
-a single majority target type. The difference between the mean of the highest pragmatic
-$\Llack$-like type in $1000$ independent populations and the highest proportion of other types
-across $\lambda$ and $l$ values is shown in Figure \ref{fig:diff}.
+a single majority target type.
+This joint influence is summarized in Figure \ref{fig:diff}, which shows the mean difference between the highest target type and the highest proportion of a type of a different kind in $1000$ independent populations across $\lambda$ and $l$ values.
+Higher values of $\lambda$ and $l$ increase the prevalence of a single target type, whereas lower values lead to less pronounced differences with a valley resulting from low $\lambda$ and high $l$ (cf.
+Figure \ref{fig:rmd} with $\lambda = 1$ and $l=15$).
 
@@
 \centering
-\includegraphics[width=1\textwidth,height=10cm,keepaspectratio]{./plots/fig4-incumbents-difference}
-\caption{Difference between mean proportion of highest pragmatic $\Llack$ type and highest
-  other type in $1000$ independent populations after $50$ generations under both pressures ($k
-  = 5$).  \mf{the coloring is not that
-    easy to decipher here; would a two-dimensional surface be possible and look better?}}
+\includegraphics[width=1\textwidth,height=10cm,keepaspectratio]{./plots/fig4-3d-incumbents-difference} %alternatively without '-3d'
+\caption{Mean difference between proportion of highest target type and highest other type in $1000$ independent populations after $50$ generations under both pressures ($k = 5$).
+\tb{Note that this plot interpolates data, but I strongly doubt there's a hidden peak/valley that is not accounted for.
+Another option would be to make a more grid-like surface without interpolation, but the main point is to have a plot from which the trends that value combinations give rise to are easy to read off, which I hope this one does.
+Compare also with the alternative 1D heatmap in the appendix.}}
 \label{fig:diff}
@@
 acquisition. Lastly, the learning behavior, captured by parameter $l$, where approximating a
-MAP strategy magnifies the effects of the learning bias in tandem with replication.
-
-In summary, target types, which represent the majority view of scalar implicatures, can come to
+MAP strategy magnifies the effects of the learning bias in tandem with replication. More broadly,
+target types, which represent the majority view of scalar implicatures, can come to
 dominate the population if three assumptions are met: (i) language is pressured toward both
@@
 The approach introduced here combines game theoretic models of functional pressure towards
-efficient communication \citep{nowak+krakauer:1999}, effects of transmission perturbations on
+successful communication \citep{nowak+krakauer:1999}, effects of transmission perturbations on
 (iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener
@@
   franke+jaeger:2014} as well as reasoning about unobservable lexical representations
-\citep{bergen+etal:2012,bergen+etal:2016}. This allows a conceptual investigation of the
+\citep{bergen+etal:2012,bergen+etal:2016}. This allows for a conceptual investigation of the
 co-evolution of conventional meaning and pragmatic language use. Main contributions of the
@@
 the former force exerting pressure only in the production of learning data. The pressure towards communicative success 
-within a population central to replication is absent. Put differently, we consider a pressure for mutual understanding 
+within a population central to replication is absent. Put differently, here we considered a pressure for mutual understanding 
 that may indirectly select for expressive types, i.e., those that can convey states unequivocally, whereas Kirby et al. only
 consider the bearing that the latter ability has on the production of learnable data. We see two main reasons
-for considering communicative success rather than expressivity, and to only considering it in learning. First, learnability alone can 
+for considering communicative success rather than just expressivity, and to only considering it in learning. First, learnability alone can 
 lead to the acquisition of functionally defective languages, as showcased by $\Lall$ (cf.
@@
 Second and more importantly, types may be equally expressive but their performance as means of information transfer crucially depends not only on themselves but on the population they find themselves in (cf.
-the competition of types of a kind in Figure \ref{fig:rmd} and their lack of competition in Figure \ref{fig:only-M}).
+the competition of target types in Figure \ref{fig:rmd} and their lack of competition in Figure \ref{fig:only-M}).
 That is, we contend that having an expressive lexicon that generates learnable data does not  in itself capture a type's arguably central communicative function of transferring information to peers.
-Under this view, a type's expressiveness is secondary to its ability to communicative successfully.
-Acquiring a maximally expressive type is of little use when it is not understood in its population.
+Acquiring a maximally expressive type is of little use when it fails to be understood in its population.
 
@@
 exploration.
-\tb{I some qualifications.
-While it is indeed true that $\Lbound$'s proportion is inflated relative to other types in the hypothesis space, it is never the second most prolific type (which is what the above hopefully is not taken to suggest, otherwise it should be changed).
+\tb{While it is indeed true that $\Lbound$'s proportion is inflated relative to other types in the hypothesis space, it is never the second most prolific type (which is what the above hopefully is not taken to suggest, otherwise it should be changed).
 As is to be expected, $\Lbound$ fares best when $l$ is low and $\lambda$ is high.
@@
 Second or third place are usually instead literal $\Llack$ or types that lexicalize ``all'', ``not all'' and ``none''.
-In short; yes, there is a inflated proportion but only compared to the whole population.
-Other particular types are usually more prolific.
+In short; yes, there is a inflated proportion relative to the whole population, but other types are usually more prolific.
 This is very likely due to the strong prior disadvantage of $\Llack$.
 In light of this I'm uncertain whether we want to keep this paragraph as is.
-I tend towards `yes' but let me know if you think otherwise.}
+I currently tend towards `yes' but let me know if you think otherwise.}
 
@@
 selection. Lexicalized upper-bounds can dominate a population because they may boost
-communicative efficiency. But they do not have to. Moreover, even without selective pressure
-for communicative efficiency, it is not the case that necessarily the types that are most
+communicative success. But they do not have to. Moreover, even without selective pressure
+for communicative success, it is not the case that necessarily the types that are most
 likely \emph{a priori} will dominate. The dynamics of iterated learning are not that
@@
 into $t$, so to speak. Taken together, without an explicit model of the interaction between
-pressure for efficiency and learnability, it is far from trivial to judge whether or when
+pressure for success and learnability, it is far from trivial to judge whether or when
 preferred or dispreferred concepts can be adaptive. This is why a major contribution of this
@@
     iterated-learning community that bring in some pressure for ``expressivity'' but not
-    communicative efficiency (if that is what they do, I don't know?).}
+    communicative efficiency (if that is what they do, I don't know?).}\\
+    \tb{That's a good point.
+I adjusted the terminology but I don't think it warrants an explicit (terminological) discussion.
+Otherwise I feel we should define expressivity explicitly (e.g., as conditional entropy) and then exemplify what we mean by comparing $\Llack$ and $\Lbound$.
+That's probably uninteresting to readers beyond the IL audience, who, I hope, can infer the difference from our stress on communicative success and the brief discussion on Kirby et al.
+2015 in the discussion section.
+}
 \end{itemize}
 
+\tb{Alternative to Figure \ref{fig:diff}}:
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=1\textwidth,height=7.5cm,keepaspectratio]{./plots/fig4-incumbents-difference}
+\caption{Difference between mean proportion of highest pragmatic $\Llack$ type and highest
+  other type in $1000$ independent populations after $50$ generations under both pressures ($k
+  = 5$).}
+\end{figure}
+
 \end{document}

### 2017-03-28 code-LOT-extension/dynamics.py

@@ -17,13 +17,25 @@ from rmd import run_dynamics
 #messages = 3 #number of messages
 #me = [False] #mutual exclusivity
 
+#for alpha in a:
+#    for lam in lamb:
+#        for k in seq_length:
+#            for sample_amount in samples:
+#                for learning_parameter in l:
+#                    for gens in g:
+#                        for runs in r:
+#                            for kind in dynamics:
+#                                for m_excl in me:
+#                                    run_dynamics(alpha,lam,k,sample_amount,gens,runs,states,messages,learning_parameter,kind,m_excl) 
+#
+#### This is all that is needed to run either replication only, mutation only or both together #####
 
-# More focused run:
+#### Parameters & setup #####
 a = [1] # rate to control difference between semantic and pragmatic violations
-lamb = [2,3,4,6,7,8,9,10,11,12,13,14,15,16,17,18,19] # soft-max parameter
+lamb = [16,17,18,19,20,21,22,23,24,25,26,27,28,29,30] # soft-max parameter
 seq_length = [5]  # length of observation sequences
 samples = [250] #amount of k-length samples for each production type
-l = [2,3,4,6,7,8,9,11,12,13,14] #prob-matching = 1, increments approach MAP
+l = [x for x in xrange(1,16)] #prob-matching = 1, increments approach MAP
 
 g = [50] #number of generations per simulation run
 r = [1000] #number of independent simulation runs
@@ -44,8 +56,6 @@ for alpha in a:
                             for kind in dynamics:
                                 for m_excl in me:
                                     run_dynamics(alpha,lam,k,sample_amount,gens,runs,states,messages,learning_parameter,kind,m_excl) 
-
-#### This is all that is needed to run either replication only, mutation only or both together #####
 
 
 

### 2017-03-28 code-LOT-extension/plots.py

@@ -413,8 +413,6 @@ def get_subfigs_rmd(type_list,list1,list2):
     plt.show()
 
 
-        
-
 def get_heatmap_diff_incumbents(type_list,list1,list2,kind):
     all_types = ['t_final'+str(z) for z in xrange(432)]
     target_types = ['t_final'+str(x) for x in type_list]
@@ -449,68 +447,61 @@ def get_heatmap_diff_incumbents(type_list,list1,list2,kind):
     plt.show()
 
 
-kind = 'rmd'
-type_list = [231,236,291,306,326,336]
-list1 = [x for x in xrange(1,21)]
-list2 = [x for x in xrange(1,16)]
+def get_3d_diff_incumbents(type_list,list1,list2,kind):
+    all_types = ['t_final'+str(z) for z in xrange(432)]
+    target_types = ['t_final'+str(x) for x in type_list]
+    other_types = [x for x in all_types if x not in target_types]
+    seq_length = 5
+    
+    #data = np.zeros((len(list1)+1, len(list2)+1))
+    #for i in xrange(len(list1)):
+    #    for j in xrange(len(list2)):
+    #        #To avoid errors when adding cols to slices of copies:
+    #        dt = pd.read_csv('./results/%s-s3-m3-lam%d-a1-k%d-samples250-l%d-g50-meFalse.csv' % (kind,list1[i],seq_length,list2[j]))
+    #        do = pd.read_csv('./results/%s-s3-m3-lam%d-a1-k%d-samples250-l%d-g50-meFalse.csv' % (kind,list1[i],seq_length,list2[j]))
+    #
+    #        data[0,j+1] = list2[j]
+    #        data[i+1,0] = list1[i]
+    #        
+    #        dt = dt[target_types] 
+    #        do = do[other_types]
+    #        dt['incumbent_t'] = dt.max(axis=1) #incumbent amongst types in d_t
+    #        do['incumbent_o'] = do.max(axis=1) #incumbent amongst types in d_o
+    #        best_of_targets = dt['incumbent_t'].mean()
+    #        best_of_others = do['incumbent_o'].mean()
+    #        data[i+1,j+1] = best_of_targets - best_of_others
 
-#def get_3d_diff_incumbents(type_list,list1,list2,kind):
[…]

### 2017-03-29 paper/paper-v3.tex
revision TB + RvR

@@
   interface, we present a game theoretic model of the competition between types of language
-  users, each endowed with a selection of lexical concepts and a particular pragmatic
+  users, each endowed with a selection of lexical representations and a particular pragmatic
   disposition to act on them. Our model traces two evolutionary forces and their interaction:
@@
 
-\section{A model of co-evolving lexical concepts and pragmatic behavior}
+\section{A model of co-evolving lexical representations and pragmatic behavior}
 \label{sec:model}
@@
 Moving beyond an abstract example, our goal is to apply the RMD to investigate the co-evolution
-of lexical concepts and pragmatic behavior. To do so, we need to fix three things: (i) what the
+of lexical representations and pragmatic behavior. To do so, we need to fix three things: (i) what the
 relevant types are, (ii) how fitness derives from communicative success and (iii) how the
@@
 make-up. Since we are interested in the evolutionary competition between different lexical
-concepts and ways of using them in communication, a type is here defined as a pair consisting
+representations and ways of using them in communication, a type is here defined as a pair consisting
 of a lexicon and a pragmatic strategy of language use.
@@
 bound in $\Lbound$ can achieve, even without pragmatic reasoning. Higher order 
-reasoning types leads to stronger associations of $\msome$ and $\ssome$ also for the
+reasoning beyond level $1$ leads to stronger associations of $\msome$ and $\ssome$ also for the
 hearer. Still, the case study presented in Section~\ref{sec:si-case-study} will consider
@@
 reasoning types which show a tendency to communicatively attuned pragmatic enrichment. Using
-only level 1 reasoning types is therefore a conservative choice that works against the
+only level 1 reasoning types and possibly small $\lambda$ is therefore a conservative choice that works against the
 fitness-based selection of pragmatic language use for a notion of fitness based on
@@
 inference in which learners combine the likelihood of a type producing the witnessed learning
-input with prior inductive biases. Experimental and mathematical results on iterated learning
-suggest that the outcome of this process reflects learners' inductive biases
-\citep[e.g.,][]{kirby+etal:2014}. In a Bayesian setting these biases can be codified in a prior
+input with prior inductive biases. In a Bayesian setting these biases can be codified in a prior
 $P \in \Delta(T)$, which reflects the amount of data a learner requires to faithfully acquire
@@
 The model of the previous section formalizes the evolutionary competition between different
-pairs of lexical concepts and ways of using them. This section looks at a case study on scalar
+pairs of lexical representations and ways of using them. This section looks at a case study on scalar
 implicatures. It engages in a following formal thought experiment to address the question: if a
@@
 states, messages and lexica (Section~\ref{sec:stat-mess-conc}). Additionally, we want to
-explore the effects of a learning bias in favor of simple lexical concepts. One way of
+explore the effects of a learning bias in favor of simple lexical representations. One way of
 motivating and formalizing such a bias is introduced below in
@@
 
-\subsubsection{States, messages, concepts and lexica} 
+\subsubsection{States, messages, lexical representations, and lexica} 
 \label{sec:stat-mess-conc}
@@
 it as a partition of possible worlds into cells where none, some or all of the $A$s are $B$s,
-for some arbitrary fixed predicates $A$ and $B$. Eight concepts can be distinguished based on
+for some arbitrary fixed predicates $A$ and $B$. Eight lexical representations can be distinguished based on
 their truth or falsity in three world states, six of which are not contradictory or
@@
 
-A lexicon $L$ is a mapping $\Messgs \rightarrow C$ from messages to concepts. With three
-messages there are $6^3 = 216$ possible lexica. Some assign the same concept to more than one
-message and others lexicalize the same concepts but associate them with different messages. Out
+A lexicon $L$ is a mapping $\Messgs \rightarrow R$ from messages to representations. With three
+messages there are $6^3 = 216$ possible lexica. Some assign the same representations to more than one
+message and others lexicalize the same representations but associate them with different messages. Out
 of these possible lexica, three kinds are of particular relevance. First, lexica that assign
-the same concept to more than one message. Such lexica lack in expressivity but may be favored
+the same lexical representations to more than one message. Such lexica lack in expressivity but may be favored
 by particular learning biases nonetheless (see below). Second, lexica that conventionalize
@@
 not lexicalize an upper bound but allow it to be conveyed pragmatically due to the presence of
-a stronger lexical item. There are six lexica of the second kind and six of the third. The
-following three lexica exemplify each kind:
+a stronger lexical item. There are six lexica of the second kind and six of the third. 
+The following three lexica exemplify each kind:
 \begin{align*}
@@
 $\Lbound$ kind. We refer to these as \emph{competitor types}, because they are expected to be
-the target types' main contenders in evolutionary competition.
+the target types' main contenders in evolutionary competition. Finally, note that while variants of a kind
+lexicalize the same representations and use them in the same way when paired with identical linguistic behavior,
+they map different states to different messages. More informally, they speak different languages
+that lexicalize the same concepts. Consequently, more often than not types of a kind fail to communicate with each other.
 
@@
 results (see, e.g., \citealt{katz+etal:2008}, \citeauthor{piantadosi+etal:underreview} under review; 2012, and \citealt{piantadosi+jacobs:2016} for recent discussion). At its
-core, a LOT defines a set of operations and composition rules from which concepts can be
+core, a LOT defines a set of operations and composition rules from which lexical representations can be
 derived. As a first approximation and for the sake of concreteness, we follow this approach to
@@
 (\citealt{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a,kirby+etal:2015}, \citeauthor{piantadosi+etal:underreview} under review). 
-In a weighed generative LOT a concept's complexity is a function of its derivation cost.
+In a weighed generative LOT a representation's complexity is a function of its derivation cost.
 
@@
 
-Our toy grammar of concepts is given in Table~\ref{tab:grammar}. This grammar uses basic
+Our toy grammar of lexical representations is given in Table~\ref{tab:grammar}. This grammar uses basic
 set-theoretic operations to form expressions which can be evaluated as true or false in states
 $\snone$, $\ssome$, or $\sall$ from above. Applications of generative rules have a cost attached
-to them. Here we simply assume that the formation of Boolean combinations of concepts incurs 2
+to them. Here we simply assume that the formation of Boolean combinations of representations incurs 2
 cost units, while all other rule applications incur only 1 cost unit. Table~\ref{tab:concepts}
-lists all six concepts relevant here, their truth conditions and the simplest formula that
-expresses this concept from the grammar in Table~\ref{tab:grammar}.
+lists all six lexical representations relevant here, their truth conditions and the simplest formula that
+expresses this representation from the grammar in Table~\ref{tab:grammar}.
 
@@
 \end{center}
-\caption{Available concepts and their minimal derivation cost}
+\caption{Available lexical representations and their minimal derivation cost}
 \label{tab:concepts}
@@
 
-A complexity measure for lexical concepts from Table~\ref{tab:concepts} is used to define a
-learning bias that favors simpler concepts over more complex ones. The prior probability of a
+A complexity measure for lexical representations from Table~\ref{tab:concepts} is used to define a
+learning bias that favors simpler representations over more complex ones. The prior probability of a
 type is just the prior probability of its lexicon. The prior of a lexicon is a function of the
-complexity of the lexical representations in its image set. Lexica with simpler concepts accordingly have a
+complexity of the lexical representations in its image set. Lexica with simpler representations accordingly have a
 higher prior. One simple way of defining such priors over lexica (and thereby types) is:
@@
 LOT-grammar (see Table~\ref{tab:concepts}). Applied to our space of lexica, this construal assigns the highest probability 
-to a lexicon of type $\Lall$, which only uses the simplest concept ``all'' for
+to a lexicon of type $\Lall$, which only uses the simplest lexical representation ``all'' for
 all messages. Lexica of type $\Llack$ are less likely, but more likely than $\Lbound$.
@@
 \includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig1-onlyr}
-\caption{Stacked proportion of target types and majority types, when other
+\caption{Stacked proportion of target types, in shades of green, and majority types, when other
   than target type, in $5$ independent populations after $50$ generations under only a
@@
 \includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig2-onlym-pr}
-\caption{Stacked proportion of target types and majority types, when other than target type, in $5$ independent populations after $50$ generations under only a pressure for learnability ($\lambda = 20, k = 5$).
+\caption{Stacked proportion of target types, in shades of green, and majority types, when other than target type, in $5$ independent populations after $50$ generations under only a pressure for learnability ($\lambda = 20, k = 5$).
 The types' prior is shown in
-  the right-most plot. The top-most groupings correspond to types with lexica of the kind of $\Llack$ (green), $\Lbound$ (magenta), and $\Lall$ (gray).}
+  the right-most plot. The top-most groupings correspond to types, literal and pragmatic, with lexica of the kind of $\Llack$, $\Lbound$, and $\Lall$.}
 \label{fig:only-M}
@@
 single target type dominate the population. When pressured for communicative success, the slight
-communicative advantage of $\Lbound$ users leads to the its prevalence. When pressured for
+communicative advantage of $\Lbound$ users leads to the their prevalence. When pressured for
 learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more
@@
 \includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig3-r+m}
-\caption{Stacked proportion of target types and majority types, when other than target type, in $5$ independent populations after $50$ generations under both pressures ($k = 5$).}
+\caption{Stacked proportion of target types, in shades of green, and majority types, when other than target type, in $5$ independent populations after $50$ generations under both pressures ($k = 5$).}
 \label{fig:rmd}
@@
 likely to be learned (cf. \citealt{chater+vitanyi:2003}). Pragmatic language use can be
-recruited indirectly by a preference for simpler lexical concepts. Under this view, semantics
-and pragmatics play a synergic role: pragmatic use allows maintenance of simpler concepts;
+recruited indirectly by a preference for simpler lexical representations. Under this view, semantics
+and pragmatics play a synergic role: pragmatic use allows maintenance of simpler representations;
 pressure towards representational simplicity indirectly promotes pragmatic over literal
@@
 In light of this I'm uncertain whether we want to keep this paragraph as is.
-I currently tend towards `yes' but let me know if you think otherwise.}
+I currently tend towards `yes' but let me know if you think otherwise.
+Robert would leave it in as well.}
 
@@
 scalar implicatures lexicalize to avoid cost, whereas infrequent ones are derived on-line to
-avoid more complex lexical concepts during acquisition. Such a prediction would lend itself to
+avoid more complex lexical representations during acquisition. Such a prediction would lend itself to
 empirical testing in line with a recent interest in differences between various scalar
@@
 Our case study could be criticized as follows: all it shows is that scalar implicatures do not
-lexicalize because upper bounds are dispreferred concepts. This criticism would be too
-superficial and highly unjust. Dispreferred lexical concepts can thrive under evolutionary
+lexicalize because upper bounds are dispreferred lexical representations. This criticism would be too
+superficial and highly unjust. Dispreferred lexical representations can thrive under evolutionary
 selection. Lexicalized upper-bounds can dominate a population because they may boost
@@
 pressure for success and learnability, it is far from trivial to judge whether or when
-preferred or dispreferred concepts can be adaptive. This is why a major contribution of this
+preferred or dispreferred representations can be adaptive. This is why a major contribution of this
 paper is the arrangement of many different ingredients into a joined model of the co-evolution
@@
 What is more, it is not that we just assumed a prior disadvantage of lexicalized upper
-bounds. We tried to motivate and formalize a general assumption about concepts' complexity
+bounds. We tried to motivate and formalize a general assumption about lexical representations' complexity
 with a concrete, albeit provisional proposal. The specification of a learning bias in terms of
-a ``grammar of concepts'' can and should be seen critically, however. Much depends on the
-primitives of such a grammar. For instance, the concept ``none or all'' is the most complex in
+a ``grammar of representations'' can and should be seen critically, however. Much depends on the
+primitives of such a grammar. For instance, the lexical representation ``none or all'' is the most complex in
 Table~\ref{tab:concepts}. But consider adding a primitive operation on sets $A \smile B$
 which is true if and only if $\neg(A \cap B \neq \emptyset \wedge A \neq \emptyset)$. The
-concept ``none or all'' would then be one of the simplest. Clearly, further research, empirical
+lexical representation ``none or all'' would then be one of the simplest. Clearly, further research, empirical
 and conceptual, into the role of representational complexity, processing costs and learning
 biases is needed. The model here makes a clear and important contribution nonetheless: it shows
-how simplicity of concepts can interact with use and evolutionary selection to show that
-without pragmatic language users we would not see simpler concepts emerge in what may be a
-natural way. Future work should also include the possibility that conceptual simplicity may
+how simplicity of representations can interact with use and evolutionary selection to show that
+without pragmatic language users we would not see simpler representations emerge in what may be a
+natural way. Future work should also include the possibility that representational simplicity may
 itself be a notion that is subject to evolutionary pressure
@@
 not lead to the emergence of different, equally costly lexical items to express different
-knowledge states of the speaker. Looking at pressure from learnability might come in again. The
-present work made a first start and gave a framework for exploring exactly these issues
+knowledge states of the speaker (cf.
+\citealt[252-267]{horn:1984}, \citealt{horn:1972,traugott:2004,vdAuwera:2010}).
+For instance, two expressions for each weak scalar expression; one with and one lacking an upper-bound.
+Looking at pressure from learnability might come in again. The present work made a first start and gave a framework for exploring exactly these issues
 systematically.

### 2017-03-29 code-LOT-extension/plots.py

@@ -201,7 +201,7 @@ def get_subfigs_mutation(type_list,list1,list2):
 
 
 
-    white_space = 10 #to separate from other priors
+    white_space = 17 #to separate from other priors
     Y_target = [0 for _ in xrange(len(priors)+white_space*3)]
     Y_lbound = [0 for _ in xrange(len(priors)+white_space*3)]
     Y_all = [0 for _ in xrange(len(priors)+white_space*3)]
@@ -217,20 +217,36 @@ def get_subfigs_mutation(type_list,list1,list2):
     Y_target[8] = priors[targets[4]]
     Y_target[10] = priors[targets[5]]
 
+    Y_target[12] = priors[targets[6]]
+    Y_target[14] = priors[targets[7]]
+    Y_target[16] = priors[targets[8]]
+    Y_target[18] = priors[targets[9]]
+    Y_target[20] = priors[targets[10]]
+    Y_target[22] = priors[targets[11]]
+   
+
 
 #    for t in xrange(len(lbound)):
 #        Y_lbound[len(targets)+white_space+t] = priors[lbound[t]]
-    Y_lbound[6+17+0] = priors[lbound[0]]
-    Y_lbound[6+17+2] = priors[lbound[1]]
-    Y_lbound[6+17+4] = priors[lbound[2]]
-    Y_lbound[6+17+6] = priors[lbound[3]]
-    Y_lbound[6+17+8] = priors[lbound[4]]
-    Y_lbound[6+17+10] = priors[lbound[5]]
+    Y_lbound[len(targets)+17+0] = priors[lbound[0]]
+    Y_lbound[len(targets)+17+2] = priors[lbound[1]]
+    Y_lbound[len(targets)+17+4] = priors[lbound[2]]
+    Y_lbound[len(targets)+17+6] = priors[lbound[3]]
+    Y_lbound[len(targets)+17+8] = priors[lbound[4]]
+    Y_lbound[len(targets)+17+10] = priors[lbound[5]]
+
+    Y_lbound[len(targets)+17+12] = priors[lbound[6]]
+    Y_lbound[len(targets)+17+14] = priors[lbound[7]]
+    Y_lbound[len(targets)+17+16] = priors[lbound[8]]
+    Y_lbound[len(targets)+17+18] = priors[lbound[9]]
+    Y_lbound[len(targets)+17+20] = priors[lbound[10]]
+    Y_lbound[len(targets)+17+22] = priors[lbound[11]]
+
 
[…]

### 2017-04-02 paper/paper-v3.tex
finishing touches first paper draft

@@
 
-\author{%\bf NAME1 and NAME2\\
-    ( -- draft \today --- )
-}
+% \author{%\bf NAME1 and NAME2\\
+%     ( -- draft \today --- )
+% }
 
@@
   disposition to act on them. Our model traces two evolutionary forces and their interaction:
-  (i) fitness-based pressure towards communicative efficiency and (ii) potential learning
-  biases during the transfer of linguistic knowledge. We illustrate the model based on a case
-  study on scalar implicatures. In this case study learning biases that favor simple semantic
-  representations can foster the evolution of more sophisticated pragmatic reasoning types and
-  so prevent the lexicalization of scalar implicatures.
+  (i) pressure towards communicative efficiency and (ii) learning biases during the transfer of
+  linguistic knowledge. We illustrate the model based on a case study on scalar
+  implicatures. In this case study learning biases that favor simple semantic representations
+  can foster the evolution of more sophisticated pragmatic reasoning types and so prevent the
+  lexicalization of scalar implicatures.
 \end{abstract}
@@
 semantics and pragmatics could have evolved and why it would be so pervasive across natural
-languages. Models of language evolution abound. There are simulation-based models studying the
-evolution of language in populations of communicating agents
+languages. Models of language evolution abound. There are simulation-based models studying
+populations of communicating agents
 \citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
@@
 pragmatic use. What is more, many mathematical models explain evolved meaning as a regularity
-in the overt behavior of agents, abstracting from complex interactions between semantic representations and pragmatic use. In contrast, we will here look at language users with a richer
-cognitive make-up.
+in the overt behavior of agents, abstracting from complex interactions between semantic
+representations and pragmatic use. In contrast, we will here look at language users with a
+richer cognitive make-up.
 
 We spell out a model of the co-evolution of conventional meaning and pragmatic reasoning
-types. The objects of replication and selection are pairs of lexical meanings and general types
-of pragmatic behavior, which we represent using probabilistic models of pragmatic language use
+types. The objects of replication and selection are pairs consisting of a set of lexical
+meanings and a manner of pragmatic behavior. The latter are represented using probabilistic
+models of pragmatic language use
 \citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
@@
 \citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. This
-approach allows us to study the interaction between (i) evolutionary pressure towards
-communicative efficiency and (ii) possible infidelity in the transmission of linguistic
-knowledge, caused by factors such as inductive learning biases and sparse learning
-data. Considering transmission of linguistic knowledge is important because neither semantic
-meanings nor pragmatic usage patterns are directly observable. Instead, language learners have
-to infer these unobservables from the observable behavior in which they result. We formalize
-this process as a form of Bayesian inference. Our approach thereby contains a well-understood
-model of iterated Bayesian learning \citep{griffiths+kalish:2007} as a special case, but
-combines it with functional selection, here formalized as the most versatile dynamic from
-evolutionary game theory; the replicator dynamic
-\citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this model.
+approach allows us to study the interaction between (i) pressure towards communicative
+efficiency and (ii) infidelity in the transmission of linguistic knowledge, caused by factors
+such as inductive learning biases and sparse learning data. Considering transmission of
+linguistic knowledge is important because neither semantic meanings nor pragmatic usage
+patterns are directly observable. Instead, language learners have to infer these unobservables
+from the observable behavior in which they result. We formalize this process as a form of
+Bayesian inference. Our approach thereby contains as a special case a well-understood model of
+iterated Bayesian learning \citep{griffiths+kalish:2007}, but combines it with functional
+selection, here formalized as the most versatile dynamic from evolutionary game theory, the
+replicator dynamic \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces
+this model.
 
@@
 
-Pragmatic types define dispositions to produce and interpret messages given a lexicon. We
-distinguish between two kinds of pragmatic types. {\em Literal interlocutors} produce and
+Pragmatic strategies define dispositions to produce and interpret messages given a lexicon. We
+distinguish between two kinds of pragmatic strategies. {\em Literal interlocutors} produce and
 interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
@@
 According to (\ref{h:level0}), a literal hearer's interpretation of a message depends on her
-lexicon and her prior over states, $pr \in \Delta(S)$, which is here assumed flat, for
+lexicon and her prior over states, $pr \in \Delta(S)$, which is here assumed flat for
 simplicity. Literal interpreters thereby choose an arbitrary true interpretation for each
@@
 interpretations based on a conjecture about speaker behavior. Speaker behavior is regulated by
-a soft-max parameter $\lambda$, $\lambda \geq 0$ \citep{luce:1959,sutton+barto:1998}. As
+a soft-max parameter $\lambda \geq 0$ \citep{luce:1959,sutton+barto:1998}. As
 $\lambda$ increases, choices approximate strict maximization of expected utilities. Expected
@@
 
-It is particularly important to note that pragmatic agents of level 1 using $\Llack$ associate $\msome$
-preferentially with $\ssome$, in contrast to their literal counterparts of level 0. This
-association is not perfect, and usually less strong than what agents with a lexicalized upper
-bound in $\Lbound$ can achieve, even without pragmatic reasoning. Higher order 
-reasoning beyond level $1$ leads to stronger associations of $\msome$ and $\ssome$ also for the
+It is particularly important to note that pragmatic agents of level 1 using $\Llack$ associate
+$\msome$ preferentially with $\ssome$, in contrast to their literal counterparts of level
+0. This association is not perfect, and usually less strong than what agents with a lexicalized
+upper bound in $\Lbound$ can achieve, even without pragmatic reasoning. Higher order reasoning
+beyond level $1$ leads to stronger associations of $\msome$ and $\ssome$ also for the
 hearer. Still, the case study presented in Section~\ref{sec:si-case-study} will consider
@@
 reasoning types which show a tendency to communicatively attuned pragmatic enrichment. Using
-only level 1 reasoning types and possibly small $\lambda$ is therefore a conservative choice that works against the
-fitness-based selection of pragmatic language use for a notion of fitness based on
-communicative success, which is introduced next.
+only level 1 reasoning types and possibly small $\lambda$ is therefore a conservative choice
+that works against the fitness-based selection of pragmatic language use for a notion of
+fitness defined as communicative success, which is introduced next.
 
@@
 linguistic behavior of type $i$, they observe the overt linguistic behavior of type $i$ and
-need to infer the covert type that most likely produced the visible behavior. Iterated learning
+need to infer the covert type that most likely produced the observed behavior. Iterated learning
 is a process in which languages are learned repeatedly from the observation of linguistic
@@
 The model has three parameters: $\lambda$ regulates the degree to which pragmatic speakers
-choose messages that maximize the chance of communicative success; $k$ is length of
-observations for each language learner; $l$ regulates where the learners' inference behavior
-lies on a spectrum from probability matching to choice of the most likely parent type.
+choose messages that appear optimal from the point of view of the agent's own utility measure
+(which may be unrelated to the expected utility when communicating with a given population);
+$k$ is length of observations for each language learner; $l$ regulates where the learners'
+inference behavior lies on a spectrum from probability matching to choice of the most likely
+parent type.
 
@@
 The model of the previous section formalizes the evolutionary competition between different
-pairs of lexical representations and ways of using them. This section looks at a case study on scalar
-implicatures. It engages in a following formal thought experiment to address the question: if a
-population of language users could freely combine different lexica with different pragmatic
-strategies, what are conditions under which the majority view of scalar implicatures could have
-evolved?
-
-Recall that the majority view is that scalar implicatures are systematic non-lexicalized pragmatic
+sets of lexical representations and ways of using them. This section looks at a case study on
+scalar implicatures. It engages in a formal thought experiment to address the
+question: if a population of language users could freely combine different lexica with
+different pragmatic strategies, what are conditions under which the majority view of scalar
+implicatures could have evolved?
+
+Recall that the majority view is that scalar implicatures are non-lexicalized pragmatic
 enrichments. Scalar implicature triggers like \emph{some}, \emph{warm} or \emph{may} are
@@
 explore the effects of a learning bias in favor of simple lexical representations. One way of
-motivating and formalizing such a bias is introduced below in
-Section~\ref{sec:an-induct-learn}.
+motivating and formalizing such a bias is introduced in Section~\ref{sec:an-induct-learn}.
 
@@
 A lexicon $L$ is a mapping $\Messgs \rightarrow R$ from messages to representations. With three
-messages there are $6^3 = 216$ possible lexica. Some assign the same representations to more than one
-message and others lexicalize the same representations but associate them with different messages. Out
-of these possible lexica, three kinds are of particular relevance. First, lexica that assign
-the same lexical representations to more than one message. Such lexica lack in expressivity but may be favored
-by particular learning biases nonetheless (see below). Second, lexica that conventionalize
-upper-bounds to realize a one-to-one mapping of messages to states. Finally, lexically that do
-not lexicalize an upper bound but allow it to be conveyed pragmatically due to the presence of
-a stronger lexical item. There are six lexica of the second kind and six of the third. 
-The following three lexica exemplify each kind:
+messages there are $6^3 = 216$ possible lexica. Some assign the same representations to more
+than one message and others lexicalize the same representations but associate them with
+different messages. Out of these possible lexica, three kinds are of particular
+relevance. First, lexica that assign the same lexical representations to more than one
+message. Such lexica lack in expressivity but may be favored by particular learning biases
+nonetheless (see below). Second, lexica that conventionalize upper-bounds to realize a
+one-to-one mapping of messages to states. Finally, lexically that do not lexicalize an upper
+bound but allow it to be conveyed pragmatically due to the presence of a stronger lexical
+item. There are six lexica of the second kind and six of the third.  The following three lexica
+exemplify each kind:
 \begin{align*}
@@
 $\Lbound$ kind. We refer to these as \emph{competitor types}, because they are expected to be
-the target types' main contenders in evolutionary competition. Finally, note that while variants of a kind
-lexicalize the same representations and use them in the same way when paired with identical linguistic behavior,
-they map different states to different messages. More informally, they speak different languages
-that lexicalize the same concepts. Consequently, more often than not types of a kind fail to communicate with each other.
+the target types' main contenders in evolutionary competition. Finally, note that while
+different types may lexicalize the same representations, they may nevertheless map different
+states to different overt messages. More informally, they speak different languages that
+lexicalize the same concepts. Consequently, more often than not different variants of the same
+kind may fail to communicate with each other.
 
@@
 speakers favor messages that appear best from their subjective point of view; the
-bottleneck size $k$ influences how faithfully learners tend to identify their teacher type; $l$
+bottleneck size $k$ influences how faithfully learners can identify their teacher type; $l$
 defines the learners' disposition towards choosing the most likely teacher type from the
@@
 that different teacher types are relatively indistinguishable based on their behavior and when
-$l$ is high. Still, lexica of the kind $\Lall$ are in turn more likely \emph{a priori} than
+$l$ is high. Still, lexica of the kind $\Lall$ are in turn even more likely \emph{a priori} than
 lexica of the kind $\Llack$. Simulation results will shed light on the question whether target
@@
 change is, if not absent, then at least very slow. In other words, even if the resulting states
-do not correspond to an eventual attracting state, they characterize long almost stationary
-states in which the system remains for a very long time. As specified in
-\S\ref{sec:learnability}, the mutation matrix $Q$ can be obtained by considering all possible
-state-message sequences of length $k$. Given that this is intractable for large $k$, the sets
-of data learners are exposed to are approximated by sampling $250$ $k$-length sequences from
-each type's production probabilities.
+do not correspond to an eventual attracting state, they characterize almost stationary states
+in which the system remains for a very long time. As specified in \S\ref{sec:learnability}, the
+mutation matrix $Q$ can be obtained by considering all possible state-message sequences of
+length $k$. Given that this is intractable for large $k$, the sets of data which learners are
+exposed to are approximated by sampling $250$ $k$-length sequences from each type's production
+probabilities.
 
@@
 \caption{Stacked proportion of target types, in shades of green, and majority types, when other
-  than target type, in $5$ independent populations after $50$ generations under only a
-  pressure for communicative success. The right-most plots zoom in on only the proportion of target types.}
+  than a target type, in $5$ independent populations after $50$ generations under only a
+  pressure for communicative success. The right-most plots zoom in on only the proportion of
+  target types.}
 \label{fig:only-R}
@@
 
-As expected, selection based on communicative success is sensitive to $\lambda$ as it
+As expected, selection based on communicative success is sensitive to $\lambda$ since $\lambda$
 influences signaling behavior, which in turn determines communicative success. This is
@@
 and becomes very slow. The result is a very long transition with near stagnancy in a rather
-homogeneous population with many types. Conversely, higher $\lambda$ promotes more rational
-linguistic behavior, widening the gap in expressivity between types and promoting more
-homogeneous populations. As suggested by Figure \ref{fig:only-R}, the majority in most
-populations is not one of the six pragmatic $\Llack$-style types. That is, a pressure only for
-communicative success does not lead to a prevalence of target types under any $\lambda$-value. For
-instance, with $\lambda = 20$ $1000$ independent populations only had $11$ cases in which
-the target type was the majority type, corresponding to a mean proportion of $.003$ across
-populations. By contrast, in $913$ cases the majority types had $\Lbound$ with close to an even
-share between literal ($454$) and pragmatic types ($459$), corresponding to a mean proportion
-of about $.48$ taken together.
+homogeneous population with many types. Conversely, higher $\lambda$ promotes more
+less-stochastic pragmatic behavior, widening the gap in expressivity between types and
+promoting more homogeneous populations. As suggested by Figure \ref{fig:only-R}, the majority
+in most populations is not one of the six pragmatic $\Llack$-style types. That is, a pressure
+only for communicative success does not lead to a prevalence of target types under any
+$\lambda$-value. For instance, with $\lambda = 20$ $1000$ independent populations only had $11$
+cases in which the target type was the majority type, corresponding to a mean proportion of
+$.003$ across populations. By contrast, in $913$ cases the majority types had $\Lbound$ with
+close to an even share between literal ($454$) and pragmatic types ($459$), corresponding to a
+mean proportion of about $.48$ taken together.
 
@@
 \includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig2-onlym-pr}
-\caption{Stacked proportion of target types, in shades of green, and majority types, when other than target type, in $5$ independent populations after $50$ generations under only a pressure for learnability ($\lambda = 20, k = 5$).
-The types' prior is shown in
-  the right-most plot. The top-most groupings correspond to types, literal and pragmatic, with lexica of the kind of $\Llack$, $\Lbound$, and $\Lall$.}
+\caption{Stacked proportion of target types, in shades of green, and majority types, when other
+  than a target type, in $5$ independent populations after $50$ generations under only a
+  pressure for learnability ($\lambda = 20, k = 5$). The types' prior is shown in the
+  right-most plot. The top-most groupings correspond to types, literal and pragmatic, with
+  lexica of the kind of $\Llack$, $\Lbound$, and $\Lall$.}
 \label{fig:only-M}
@@
 
-Pressure for communicative success or learnability are not sufficient on their own to have a
+Pressure for communicative success and learnability are not sufficient on their own to have a
 single target type dominate the population. When pressured for communicative success, the slight
-communicative advantage of $\Lbound$ users leads to the their prevalence. When pressured for
+communicative advantage of $\Lbound$ users leads to their prevalence. When pressured for
 learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more
@@
 $l$. Pressure for communicative success magnifies the effects of iterated learning and dampens
-the proliferation of types of a kind that are equal in expressivity {\em and} learnability. A
+the proliferation of multiple types of a kind that are equal in expressivity {\em and} learnability. A
 pressure towards learnability favors the transmission of simpler semantics and thereby selects
@@
 \includegraphics[width=1\textwidth,height=10cm,keepaspectratio]{./plots/fig4-3d-incumbents-difference} %alternatively without '-3d'
-\caption{Mean difference between proportion of highest target type and highest other type in $1000$ independent populations after $50$ generations under both pressures ($k = 5$).
-\tb{Note that this plot interpolates data, but I strongly doubt there's a hidden peak/valley that is not accounted for.
-Another option would be to make a more grid-like surface without interpolation, but the main point is to have a plot from which the trends that value combinations give rise to are easy to read off, which I hope this one does.
-Compare also with the alternative 1D heatmap in the appendix.}}
+\caption{Mean difference between proportion of highest target type and highest other type in
+  $1000$ independent populations after $50$ generations under both pressures ($k = 5$).}
 \label{fig:diff}
@@
 
-As for the effect of the sequence length $k$ not mentioned so far, it influences populations in
-a predictable way: small values lead to more heterogeneous populations that reflect the
-learner's prior more faithfully. This is due to the fact that the likelihood that a small
-sequence was produced by any type is relatively uniform (modulo prior). By contrast, larger
-values increasingly allow learners to differentiate types with different signaling behaviors.
+Effects of manipulating the sequence length $k$ have not been addressed so far, but are rather
+predictable: small values lead to more heterogeneous populations that reflect the learner's
+prior more faithfully. This is due to the fact that the likelihood that a small sequence was
+produced by any type is relatively uniform (modulo prior). By contrast, larger values
+increasingly allow learners to differentiate types with different signaling behaviors.
 
@@
 With respect to (i), \citet{kirby+etal:2015} propose a comparable model of the interaction
-between a lexicon's expressivity and its learnability. One of its main differences to our proposal lies in 
-the former force exerting pressure only in the production of learning data. The pressure towards communicative success 
-within a population central to replication is absent. Put differently, here we considered a pressure for mutual understanding 
-that may indirectly select for expressive types, i.e., those that can convey states unequivocally, whereas Kirby et al. only
-consider the bearing that the latter ability has on the production of learnable data. We see two main reasons
-for considering communicative success rather than just expressivity, and to only considering it in learning. First, learnability alone can 
-lead to the acquisition of functionally defective languages, as showcased by $\Lall$ (cf.
-\citealt{kirby+etal:2008,silvey+etal:2014}, and \citealt{fay+etal:2013} for review of laboratory results).
-Second and more importantly, types may be equally expressive but their performance as means of information transfer crucially depends not only on themselves but on the population they find themselves in (cf.
-the competition of target types in Figure \ref{fig:rmd} and their lack of competition in Figure \ref{fig:only-M}).
-That is, we contend that having an expressive lexicon that generates learnable data does not  in itself capture a type's arguably central communicative function of transferring information to peers.
-Acquiring a maximally expressive type is of little use when it fails to be understood in its population.
+between a lexicon's expressivity and its learnability. A main difference is that here we
+considered a pressure for mutual understanding within a generation that may indirectly select
+for expressive types, i.e., those that can convey states unequivocally, whereas Kirby et
+al. only consider the bearing that the latter ability has on the production of learnable
+data. We see two main reasons for considering communicative success rather than just
+expressivity, and to only considering it in learning. First, learnability alone can lead to the
+acquisition of functionally defective languages, as showcased by $\Lall$
+(cf. \citealt{kirby+etal:2008,silvey+etal:2014}, and \citealt{fay+etal:2013} for review of
+laboratory results). Second and more importantly, types may be equally expressive but their
+performance as means of information transfer crucially depends not only on themselves but on
+the population they find themselves in (cf. the competition of target types in Figure
+\ref{fig:rmd} and their lack of competition in Figure \ref{fig:only-M}). That is, we contend
+that having an expressive lexicon that generates learnable data does not in itself capture a
+type's arguably central communicative function of transferring information to peers. Acquiring
+a maximally expressive type is of little use when it fails to be understood in its population.
 
@@
 not. Interestingly, in our simulations when a target type is the majority type an inflated
-proportion of the population uses compatible lexica with a lexicalized upper bound. Particularly in those parameter
-settings where the prior influences the outcome less. In other words, we do find a tendency toward a similar co-existence
-of semantic and pragmatic types. Whether this analogy has any further explanatory value is an interesting path for future
+proportion of the population uses compatible lexica with a lexicalized upper
+bound. Particularly in those parameter settings where the prior influences the outcome less. In
+other words, we do find a tendency toward a similar co-existence of semantic and pragmatic
+types. Whether this analogy has any further explanatory value is an interesting path for future
 exploration.
-\tb{While it is indeed true that $\Lbound$'s proportion is inflated relative to other types in the hypothesis space, it is never the second most prolific type (which is what the above hopefully is not taken to suggest, otherwise it should be changed).
-As is to be expected, $\Lbound$ fares best when $l$ is low and $\lambda$ is high.
-Even then, it usually ranks around $5$-th with a mean proportion of its kind of about $0.01 - 0.05$.
-As $l$ increases, it gets pushed back.
-Second or third place are usually instead literal $\Llack$ or types that lexicalize ``all'', ``not all'' and ``none''.
-In short; yes, there is a inflated proportion relative to the whole population, but other types are usually more prolific.
-This is very likely due to the strong prior disadvantage of $\Llack$.
-In light of this I'm uncertain whether we want to keep this paragraph as is.
-I currently tend towards `yes' but let me know if you think otherwise.
-Robert would leave it in as well.}
 
@@
 communicative success. But they do not have to. Moreover, even without selective pressure
-for communicative success, it is not the case that necessarily the types that are most
+for communicative success, it is not necessarily the case that the types that are most
 likely \emph{a priori} will dominate. The dynamics of iterated learning are not that
@@
 What is more, it is not that we just assumed a prior disadvantage of lexicalized upper
-bounds. We tried to motivate and formalize a general assumption about lexical representations' complexity
-with a concrete, albeit provisional proposal. The specification of a learning bias in terms of
-a ``grammar of representations'' can and should be seen critically, however. Much depends on the
-primitives of such a grammar. For instance, the lexical representation ``none or all'' is the most complex in
-Table~\ref{tab:concepts}. But consider adding a primitive operation on sets $A \smile B$
-which is true if and only if $\neg(A \cap B \neq \emptyset \wedge A \neq \emptyset)$. The
-lexical representation ``none or all'' would then be one of the simplest. Clearly, further research, empirical
-and conceptual, into the role of representational complexity, processing costs and learning
-biases is needed. The model here makes a clear and important contribution nonetheless: it shows
-how simplicity of representations can interact with use and evolutionary selection to show that
-without pragmatic language users we would not see simpler representations emerge in what may be a
-natural way. Future work should also include the possibility that representational simplicity may
+bounds. We tried to motivate and formalize a general assumption about lexical representations'
+complexity with a concrete, albeit provisional proposal. The specification of a learning bias
+in terms of a ``grammar of representations'' can and should be seen critically, however. Much
+depends on the primitives of such a grammar. For instance, the lexical representation ``none or
+all'' is the most complex in Table~\ref{tab:concepts}. But consider adding a new primitive
+relation between sets $A \smile B$ which is true if and only if
+$\neg(A \cap B \neq \emptyset \wedge A \neq \emptyset)$. The lexical representation ``none or
+all'' would then be one of the simplest. Clearly, further research, empirical and conceptual,
+into the role of representational complexity, processing costs and learning biases is
+needed. The model here makes a clear and important contribution nonetheless: it demonstrates
+how simplicity of representations can interact with use and evolutionary selection and shows
+that for simple representations to emerge it may require pragmatic strategies to compensate
+their potential expressive deficiencies. Hence model of co-evolving semantics and pragmatics is
+needed. Future work should also include the possibility that representational simplicity may
 itself be a notion that is subject to evolutionary pressure
@@
 non-lexicalized upper bounds allow a broader range of applicability, e.g., when the speaker is
-not certain as to whether \emph{all} is true. This may suggest an alternative argument for why
-upper-bounded meanings do not conventionalize based on functional criteria only: should
-contextual cues provide enough information to the hearer to identify whether a bound is
-intended to be conveyed pragmatically, then this is preferred over expressing it overtly
-through longer expressions, e.g., by saying {\em some but not all} explicitly. Importantly,
-although morphosyntactic disambiguation may be dispreferred due to its relative length and
-complexity \citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and
-override contextual cues that might otherwise mislead the hearer. In a nutshell, this
-explanation posits that scalar implicatures fail to lexicalize because, all else being equal,
-speakers prefer to communicate as economically as possible and pragmatic reasoning enables them
-to do so. What this alternative explanation does not explain is why functional pressure does
-not lead to the emergence of different, equally costly lexical items to express different
-knowledge states of the speaker (cf.
-\citealt[252-267]{horn:1984}, \citealt{horn:1972,traugott:2004,vdAuwera:2010}).
-For instance, two expressions for each weak scalar expression; one with and one lacking an upper-bound.
-Looking at pressure from learnability might come in again. The present work made a first start and gave a framework for exploring exactly these issues
-systematically.
+not certain as to whether \emph{all} is true. This may suggest an alternative and purely
+functionalist argument for why upper-bounded meanings do not conventionalize: should contextual
+cues provide enough information to the hearer to identify whether a bound is intended to be
+conveyed pragmatically, then this is preferred over expressing it overtly through longer
+expressions, e.g., by saying {\em some but not all} explicitly. Importantly, although
+morphosyntactic disambiguation may be dispreferred due to its relative length and complexity
+\citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and override
+contextual cues that might otherwise mislead the hearer. In a nutshell, this explanation posits
+that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to
+communicate as economically as possible and pragmatic reasoning enables them to do so. What
+this alternative argument does not explain is why functional pressure does not lead to the
+emergence of different, equally costly lexical items to express different knowledge states of
+the speaker (cf. \citealt[252-267]{horn:1984},
+\citealt{horn:1972,traugott:2004,vdAuwera:2010}). For instance, two expressions for each weak
+scalar expression; one with and one lacking an upper-bound.  Looking at pressure from
+learnability might come in again. The present work made a first start and gave a framework for
+exploring exactly these issues systematically.
 
@@
 meanings and patterns of language use. This allows the model to trace the interaction between
-conventional meaning and pragmatic use. Additionally, it takes the challenge serious of neither
-semantics nor pragmatics being directly observable. Instead, learners need to infer these
-unobservables from overt data that results from their combination.  These components and their
-mutual influence were highlighted in a case study on the lack of lexical upper-bounds in weak
-scalar expressions that showed that, when pressured for learnability and communicative success, the
-former force drives for simpler semantic representations inasmuch as pragmatics can compensate
-for lack of expressivity in use. That is, the relative learning advantage of simpler semantics
-in combination with functional pressure in use may offer an answer to why natural languages
-fail to lexicalize systematic pragmatic inferences. And, more broadly, to a division of labor
-between semantics and pragmatics.
+conventional meaning and pragmatic use. Additionally, it takes the challenge seriously of
+neither semantics nor pragmatics being directly observable. Instead, learners need to infer
+these unobservables from overt data that results from their combination.  These components and
+their mutual influence were highlighted in a case study on the lack of lexical upper-bounds in
+weak scalar expressions that showed that, when pressured for learnability and communicative
+success, the former force drives for simpler semantic representations inasmuch as pragmatics
+can compensate for lack of expressivity in use. That is, the relative learning advantage of
+simpler semantics in combination with functional pressure in use may offer an answer to why
+natural languages fail to lexicalize systematic pragmatic inferences. And, more broadly, to a
+division of labor between semantics and pragmatics.
 
@@
 
-\newpage
-
-\appendix
-
-\section{notes}
-
-\begin{itemize}
-\item \mf{The text frequently uses the term ``expressivity'' and ``pressure for/towards
-    expressivity''. I have mixed feelings about this. First of all, I don't know what
-    ``expressivity'' really is. It seems to relate to a lexicon: how much information can the
-    lexicon convey about the states. But we do not model selection of expressivity, but
-    selection of communicative success relative to the population state. The difference is very
-    pronounced: there can be several maximally expressive lexica with huge differences in their
-    communicatively success or expected utility for a given population state. We only
-    indirectly select for expressivity, and we don't necessarily select for expressivity of a
-    lexicon. \dots It may be that ``expressivity'' is very much Edinburgh terminology where
-    exactly the picture of communication success in the current population is missing. I might
-    be wrong though. Possibly, we should introduce this distinction (with this terminology) and
-    use it also in the general discussion to relate our work to those articles from the
-    iterated-learning community that bring in some pressure for ``expressivity'' but not
-    communicative efficiency (if that is what they do, I don't know?).}\\
-    \tb{That's a good point.
-I adjusted the terminology but I don't think it warrants an explicit (terminological) discussion.
-Otherwise I feel we should define expressivity explicitly (e.g., as conditional entropy) and then exemplify what we mean by comparing $\Llack$ and $\Lbound$.
-That's probably uninteresting to readers beyond the IL audience, who, I hope, can infer the difference from our stress on communicative success and the brief discussion on Kirby et al.
-2015 in the discussion section.
-}
-\end{itemize}
-
-\tb{Alternative to Figure \ref{fig:diff}}:
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=1\textwidth,height=7.5cm,keepaspectratio]{./plots/fig4-incumbents-difference}
-\caption{Difference between mean proportion of highest pragmatic $\Llack$ type and highest
-  other type in $1000$ independent populations after $50$ generations under both pressures ($k
-  = 5$).}
-\end{figure}
+% \newpage
+
+% \appendix
+
+% \begin{figure}[h!]
+% \centering
+% \includegraphics[width=1\textwidth,height=7.5cm,keepaspectratio]{./plots/fig4-incumbents-difference}
+% \caption{Difference between mean proportion of highest pragmatic $\Llack$ type and highest
+%   other type in $1000$ independent populations after $50$ generations under both pressures ($k
+%   = 5$).}
+% \end{figure}
 

### 2017-04-04 paper/paper-v3.tex
revisions based on RHs comments, take 1 + new plots take 1

@@
 If this majority view is correct, the question arises how such a division of labor between
-semantics and pragmatics could have evolved and why it would be so pervasive across natural
-languages. Models of language evolution abound. There are simulation-based models studying
+semantics and pragmatics could have evolved, why it would be so pervasive across natural
+languages and why it is that some expressions systematically draw from it while others semantically conventionalize. Models of language evolution abound. There are simulation-based models studying
 populations of communicating agents
@@
   The average fitness is $\Phi(x) = x f_n(x) + (1-x) f_p(x) = 3x^2 - 4x + 2$. Without mutation,
-  the replicator dynamic will then update $x$ to $RD(x) = \nicefrac{x^2}{\Phi(x)}$. The update
+  the replicator dynamic will then update $x$ to $RD(x) = \nicefrac{f(x)x}{\Phi(x)} = \nicefrac{x^2}{\Phi(x)}$. The update
   function $RD(x)$ of the replicator step is plotted in Figure~\ref{fig:Updates_RMS} as the
@@
 \citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a
-(possibly fuzzy) set of states. A pragmatic type maps a lexicon onto a probabilistic speaker
+(possibly fuzzy) set of states. A pragmatic type specifies a probabilistic speaker
 rule (a probabilistic choice of message for each state) and a probabilistic hearer rule (a
-probabilistic choice of state for each message). There are many ways of making these general
+probabilistic choice of state for each message) given a lexicon. There are many ways of making these general
 notions more concrete. Here is what we will assume in the remainder of this paper.
@@
 \begin{alignat}{3}
-&H_{0}(s \mid m;L) &&\propto pr(s) L_{sm} \label{h:level0}\\
-&S_{0}(m \mid s;L) &&\propto \exp(\lambda \; L_{sm}) \label{s:level0}\\
+&H_{0}(s \mid m;L) &&\propto pr(s) L_{[s,m]} \label{h:level0}\\
+&S_{0}(m \mid s;L) &&\propto \exp(\lambda \; L_{[s,m]}) \label{s:level0}\\
 &H_{n+1}(s|m;L) &&\propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
@@
 simplicity. Literal interpreters thereby choose an arbitrary true interpretation for each
-message. Pragmatic hearers, defined in (\ref{h:leveln}), instead use Bayes' rule to weigh
+message according to their lexicon. Pragmatic hearers, defined in (\ref{h:leveln}), instead use Bayes' rule to weigh
 interpretations based on a conjecture about speaker behavior. Speaker behavior is regulated by
@@
 \end{align*}
-If $\lambda = 1$ literal speaker behavior is as follows:
+By contrast, pragmatic hearers of level 1 have the following interpretative behavior for $\lambda = 1$:
+\begin{align*}
+  H_1(\cdot \mid \cdot, \Lbound) & \approx \begin{blockarray}{lcc}
+    & \mygray{\ssome} & \mygray{\sall} \\
+    \begin{block}{l[cc]}
+      \mygray{\msome} & .73 & .27 \\
+      \mygray{\mall}  & .27 & .73 \\
+    \end{block}
+  \end{blockarray} &
+  H_1(\cdot \mid \cdot, \Llack) & \approx \begin{blockarray}{lcc}
+    & \mygray{\ssome} & \mygray{\sall} \\
+    \begin{block}{l[cc]}
+      \mygray{\msome} & .59 & .41 \\
+      \mygray{\mall}  & 0.35 & 0.65 \\
+    \end{block}
+  \end{blockarray} 
+\end{align*}
+This is the outcome of reasoning about their level-$0$ speaker counterparts with $\lambda = 1$:
 \begin{align*}
@@
 \end{align*}
-If $\lambda = 20$, literal speakers will choose true utterances with less slack:
+With low $\lambda$ speakers choose true messages with more slack.
+Reasoning over this behavior therefore also results in a weaker association of messages with only true states in receivers, but also in a slightly stronger association of $\msome$ with $\ssome$ over $\sall$ for $\Llack$ users because $S_0(\msome | \ssome;\Llack) > S_0(\msome | \sall;\Llack)$.
+For $\lambda = 20$, there will be less slack in literal speaker behavior:
 \begin{align*}
@@
 \end{align*}
-Pragmatic hearers of level 1 have the following probabilistic interpretation behavior for
-$\lambda = 1$:
+And accordingly less slack in pragmatic interpretation:
 \begin{align*}
@@
     \begin{block}{l[cc]}
-      \mygray{\msome} & .73 & .27 \\
-      \mygray{\mall}  & .27 & .73 \\
+      \mygray{\msome} & 1 & 0 \\
+      \mygray{\mall}  & 0 & 1 \\
     \end{block}
@@
     \begin{block}{l[cc]}
-      \mygray{\msome} & .59 & .41 \\
-      \mygray{\mall}  & 0.35 & 0.65 \\
-    \end{block}
-  \end{blockarray} 
-\end{align*}
-For $\lambda = 20$, there will less slack again:
-\begin{align*}
-  H_1(\cdot \mid \cdot, \Lbound) & \approx \begin{blockarray}{lcc}
-    & \mygray{\ssome} & \mygray{\sall} \\
-    \begin{block}{l[cc]}
-      \mygray{\msome} & 1 & 0 \\
-      \mygray{\mall}  & 0 & 1 \\
-    \end{block}
-  \end{blockarray} &
-  H_1(\cdot \mid \cdot, \Llack) & \approx \begin{blockarray}{lcc}
-    & \mygray{\ssome} & \mygray{\sall} \\
-    \begin{block}{l[cc]}
       \mygray{\msome} & 0.67 & 0.33 \\
@@
 \end{align*}
-With $\lambda = 1$, pragmatic speakers of level 1 will have:
+Lastly, turning to types that have no bearing on hearer choices, with $\lambda = 1$ pragmatic speakers of level 1 have:
 \begin{align*}
@@
 \end{align*}
-For high $\lambda = 20$, pragmatic speakers of level 1 give us:
+For $\lambda = 20$, pragmatic speaker behavior of level 1 is as follows:
 \begin{align*}
@@
 In the following we consider a specific application of the model from Section~\ref{sec:model}
-which allows to address the question if or when scalar inferences might (not) lexicalize. We
+which allows us to address the question if or when scalar inferences might (not) lexicalize. We
 consider what is perhaps one of the simplest non-trivial setups that speak to this matter and
@@
 nonetheless (see below). Second, lexica that conventionalize upper-bounds to realize a
-one-to-one mapping of messages to states. Finally, lexically that do not lexicalize an upper
+one-to-one mapping of messages to states. Finally, lexica that do not lexicalize an upper
 bound but allow it to be conveyed pragmatically due to the presence of a stronger lexical
@@
 that having an expressive lexicon that generates learnable data does not in itself capture a
-type's arguably central communicative function of transferring information to peers. Acquiring
-a maximally expressive type is of little use when it fails to be understood in its population.
+type's arguably central communicative function of transferring information to peers. Taking communication
+into consideration in turn allows the model to be responsive to the task for which language is learned.
 
@@
 that for simple representations to emerge it may require pragmatic strategies to compensate
-their potential expressive deficiencies. Hence model of co-evolving semantics and pragmatics is
+their potential expressive deficiencies. Hence a model of co-evolving semantics and pragmatics is
 needed. Future work should also include the possibility that representational simplicity may
 itself be a notion that is subject to evolutionary pressure
-\citep[cf.][]{ThompsonKirby2016:Culture-Shapes-}.
+\citep[cf.][]{ThompsonKirby2016:Culture-Shapes-}, as well for the evolution of elements that define
+the agents' cognitive make-up; $\lambda$ and $l$.
 
@@
 scalar expression; one with and one lacking an upper-bound.  Looking at pressure from
-learnability might come in again. The present work made a first start and gave a framework for
+learnability might come in again. Beyond scalar implicatures, the model can generate predictions about
+likely lexicalization trajectories of pragmatic inferences, or a lack thereof. In this realm
+an interesting issue is whether proposed principles, such as the semantic conventionalization of once highly context-dependent inferences if they become regular enough \citep{levinson:2000,traugott:2004}, can be given
+a formal rationale and inform the postulated directionalities of change. The present work made a first start and gave a framework for
 exploring exactly these issues systematically.
@@
 
-% \appendix
+ \appendix
+\newpage
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/alt-fig1-onlyr}
+\caption{Representative population after $50$ generations under only a
+  pressure for communicative success. The right-most plots zoom in on only the proportion of
+  target types.}
+\label{fig:only-R}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/alt-fig2-onlym-pr}
+\caption{Representative population after $50$ generations under only a
+  pressure for learnability ($\lambda = 20, k = 5$). The types' prior is shown in the
+  right-most plot. The top-most groupings correspond to types, literal and pragmatic, with
+  lexica of the kind of $\Llack$, $\Lbound$, and $\Lall$.}
+\label{fig:only-M}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/alt-fig3-r+m}
+\caption{Representative population after $50$ generations under both pressures ($k = 5$).}
+\label{fig:rmd}
+\end{figure}
 

### 2017-04-04 code-LOT-extension/plots.py

@@ -193,7 +193,7 @@ def get_subfigs_mutation(type_list,list1,list2):
 
     targets = [231,236,291,306,326,336]
     targets = targets + [x-216 for x in targets] #adding literal Llack
-
+    
     lbound = [225, 235, 255, 270, 325, 330]
     lbound = lbound + [x-216 for x in lbound]
 
@@ -610,3 +610,370 @@ types = [231,236,291,306,326,336]
 #        print np.where(binned_props.argsort()[::-1] == 64)[0]
 #        print '###'
 #
+
+def get_subfigs_hist_replication(type_list,list1,list2):
+    seq_length = 5
+    num_of_runs = 5
+     
+    types_to_plot = ['t_final'+str(x) for x in type_list]  
+    targets_to_plot = ['t_final'+str(x) for x in type_list[:6]]  
+
+    restrict_to_final = ['t_final'+str(z) for z in xrange(432)] #as to avoid the incumbent to come from other columns
+
+    df1 = pd.read_csv('./results/%s-s3-m3-lam%d-a1-k%d-samples250-l%d-g50-meFalse.csv' % ('r',list1[0],seq_length,list2[0]))
+    df2 = pd.read_csv('./results/%s-s3-m3-lam%d-a1-k%d-samples250-l%d-g50-meFalse.csv' % ('r',list1[1],seq_length,list2[0]))
+    df3 = pd.read_csv('./results/%s-s3-m3-lam%d-a1-k%d-samples250-l%d-g50-meFalse.csv' % ('r',list1[2],seq_length,list2[0]))
+   
+    df1 = df1[restrict_to_final]
+    df1_majority = df1.max(axis=1).iloc[num_of_runs]
+
+    df2 = df2[restrict_to_final]
+    df2_majority = df2.max(axis=1).iloc[num_of_runs]
+ 
+    df3 = df3[restrict_to_final]
+    df3_majority = df3.max(axis=1).iloc[num_of_runs]
+   
+    df1_large = df1[types_to_plot]# + ['incumbent']]
+    df1_large = df1_large.iloc[num_of_runs]
+    Y1_large = df1_large.values
+                 
+    df2_large = df2[types_to_plot]# + ['incumbent']]
+    df2_large = df2_large.iloc[num_of_runs]
+    Y2_large = df2_large.values
+
+    df3_large = df3[types_to_plot]# + ['incumbent']]
+    df3_large = df3_large.iloc[num_of_runs]
+    Y3_large = df3_large.values
+
[…]

### 2017-04-05 code-LOT-extension/data-extraction.py
data for alt-plots

@@ -1,0 +1,28 @@
+import pandas as pd
+import numpy as np 
+import csv
+
+target_types = [231,236,291,306,326,336]
+list1 = [1,5,20] #lambda
+list2 = [1,5,15] #l
+def get_data_to_csv(target_types,list1,list2):
+    restrict_to_final = ['t_final'+str(z) for z in xrange(432)] #as to avoid the majority type to come from other columns
+    targets = ['t_final'+str(z) for z in target_types]
+    non_targets = [x for x in restrict_to_final if x not in targets]
+    seq_length = 5
+ 
+    f = csv.writer(open('./data-plot4.csv','wb'))
+    f.writerow(['runID','lambda','l','proportion largest target','proportion largest nontarget'])
+  
+    for x in xrange(len(list1)):
+        for y in xrange(len(list2)):
+            df = pd.read_csv('./results/%s-s3-m3-lam%d-a1-k%d-samples250-l%d-g50-meFalse.csv' % ('rmd',list1[y],seq_length,list2[x]))
+            
+            df['largest target'] = df[targets].max(axis=1)
+            df['largest non target'] = df[non_targets].max(axis=1)
+            for idx,row in df.iterrows():
+                f.writerow([str(row['runID']),str(row['lam']),str(row['l']),str(row['largest target']),str(row['largest non target'])])
+
+
+get_data_to_csv(target_types,list1,list2)
+

### 2017-04-12 code-LOT-extension/plot_result_summary.R
plot results

@@ -1,0 +1,71 @@
+library(tidyverse)
+library(reshape2)
+library(bootstrap)
+# quantile(bootstrap::bootstrap(x = mean, nboot = 1000, theta = function(x){mean(x)})$thetastar, 0.025)
+
+## only replication
+
+d = read.csv('ordered-data-plot2.csv') 
+e = melt(d, id.vars = c(1,2), variable.name = "strategy", value.name = "proportion") %>%
+  mutate(strategy = as.character(strategy))
+
+
+e$type = sapply(1:nrow(e), function(i) ifelse(strsplit(e$strategy[i], split = ".", fixed = T)[[1]][1] == "other", "other",
+                                              strsplit(e$strategy[i], split = ".", fixed = T)[[1]][2]))
+                       
+
+f = e %>% filter(type != "other") %>% arrange(runID, lambda, type, - proportion) %>% group_by(runID, lambda, type) %>% mutate(rank = 1:n())
+
+plot.data = f %>% group_by(lambda, type, rank) %>%
+  summarize(mean = mean(proportion),
+            bslo = quantile(bootstrap::bootstrap(x = proportion, nboot = 2000, theta = function(x){mean(x)})$thetastar, 0.025),
+            bshi = quantile(bootstrap::bootstrap(x = proportion, nboot = 2000, theta = function(x){mean(x)})$thetastar, 0.975))
+plot.data = plot.data %>% mutate(strategy = paste0(type, rank))
+
+results.plot = ggplot(plot.data, aes(x = strategy, y = mean, fill = type)) + geom_bar(stat = "identity") +
+  geom_errorbar(aes(ymin=bslo, ymax=bshi), width=0.5, position = "dodge") + facet_grid(. ~ lambda) + coord_flip()
+show(results.plot)
+
+## only mutation
+
+d = read.csv('ordered-data-plot3.csv') 
+e = melt(d, id.vars = c(1,2,3), variable.name = "strategy", value.name = "proportion") %>%
+  mutate(strategy = as.character(strategy))
+
+e$type = sapply(1:nrow(e), function(i) ifelse(strsplit(e$strategy[i], split = ".", fixed = T)[[1]][1] == "other", "other",
+                                              strsplit(e$strategy[i], split = ".", fixed = T)[[1]][2]))
+
+f = e %>% filter(type != "other") %>% arrange(runID, lambda, l, type, - proportion) %>% 
[…]

### 2017-04-12 code-LOT-extension/plot_result_summary.R
plot results

@@ -50,7 +50,7 @@ show(results.plot)
 
 ## mutation & replication
 
-d = read.csv('ordered-data-plot3.csv') 
+d = read.csv('ordered-data-plot4.csv') 
 e = melt(d, id.vars = c(1,2,3), variable.name = "strategy", value.name = "proportion") %>%
   mutate(strategy = as.character(strategy))
 

### 2017-04-12 code-LOT-extension/plot_result_summary.R
results plotting

@@ -14,7 +14,7 @@ e$type = sapply(1:nrow(e), function(i) ifelse(strsplit(e$strategy[i], split = ".
                                               strsplit(e$strategy[i], split = ".", fixed = T)[[1]][2]))
                        
 
-f = e %>% filter(type != "other") %>% arrange(runID, lambda, type, - proportion) %>% group_by(runID, lambda, type) %>% mutate(rank = 1:n())
+f = e %>% filter(rank >= 6) %>% arrange(runID, lambda, type, - proportion) %>% group_by(runID, lambda, type) %>% mutate(rank = 1:n())
 
 plot.data = f %>% group_by(lambda, type, rank) %>%
   summarize(mean = mean(proportion),
@@ -25,6 +25,8 @@ plot.data = plot.data %>% mutate(strategy = paste0(type, rank))
 results.plot = ggplot(plot.data, aes(x = strategy, y = mean, fill = type)) + geom_bar(stat = "identity") +
   geom_errorbar(aes(ymin=bslo, ymax=bshi), width=0.5, position = "dodge") + facet_grid(. ~ lambda) + coord_flip()
 show(results.plot)
+ggsave("plots/results_summary_replication.pdf", results.plot, width = 10, height = 10)
+
 
 ## only mutation
 
@@ -35,7 +37,7 @@ e = melt(d, id.vars = c(1,2,3), variable.name = "strategy", value.name = "propor
 e$type = sapply(1:nrow(e), function(i) ifelse(strsplit(e$strategy[i], split = ".", fixed = T)[[1]][1] == "other", "other",
                                               strsplit(e$strategy[i], split = ".", fixed = T)[[1]][2]))
 
-f = e %>% filter(type != "other") %>% arrange(runID, lambda, l, type, - proportion) %>% 
+f = e %>% filter(rank >= 6) %>% arrange(runID, lambda, l, type, - proportion) %>% 
   group_by(runID, lambda, l, type) %>% mutate(rank = 1:n())
 
 plot.data = f %>% group_by(l, lambda, type, rank) %>%
@@ -47,6 +49,7 @@ plot.data = plot.data %>% mutate(strategy = paste0(type, rank))
 results.plot = ggplot(plot.data, aes(x = strategy, y = mean, fill = type)) + geom_bar(stat = "identity") +
[…]

### 2017-04-12 paper/paper-v3.tex
slight changes
representative population plots integrated in text

@@
 \end{align*}
-Lastly, turning to types that have no bearing on hearer choices, with $\lambda = 1$ pragmatic speakers of level 1 have:
+Lastly, turning to types that have no bearing on the choices of hearers of level 1, with $\lambda = 1$ pragmatic speakers of level 1 have:
 \begin{align*}
@@
 \centering
-\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig1-onlyr}
-\caption{Stacked proportion of target types, in shades of green, and majority types, when other
-  than a target type, in $5$ independent populations after $50$ generations under only a
-  pressure for communicative success. The right-most plots zoom in on only the proportion of
-  target types.}
+\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/alt-fig1-onlyr}
+\caption{Proportion of target types, pragmatic competitor types, pragmatic $\Lall$, and the population's majority type, in representative populations after $50$ generations under only a
+  pressure for communicative success.}
 \label{fig:only-R}
@@
 influences signaling behavior, which in turn determines communicative success. This is
-showcased in Figure \ref{fig:only-R}, which depicts the proportion of target types in $5$
-independent populations after $50$ replicator steps. The plot also shows the proportion of the
+showcased in Figure \ref{fig:only-R}, which shows the proportion of target types, pragmatic competitor types and pragmatic $\Lall$ in a representative population after $50$ replicator steps.
+The plot also indicates the proportion of the
 \emph{majority type}, i.e., the type with the highest proportion in the final population. With
@@
 \centering
-\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig2-onlym-pr}
-\caption{Stacked proportion of target types, in shades of green, and majority types, when other
-  than a target type, in $5$ independent populations after $50$ generations under only a
+\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/alt-fig2-onlym-pr}
+\caption{Proportion of target types, pragmatic competitor types, pragmatic $\Lall$, and population's majority type in representative populations after $50$ generations under only a
   pressure for learnability ($\lambda = 20, k = 5$). The types' prior is shown in the
@@
 \centering
-\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig3-r+m}
-\caption{Stacked proportion of target types, in shades of green, and majority types, when other than target type, in $5$ independent populations after $50$ generations under both pressures ($k = 5$).}
+\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/alt-fig3-r+m}
+\caption{Proportion of target types, pragmatic competitor types, pragmatic $\Lall$, and the population's majority type in representative populations after $50$ generations under both pressures ($k = 5$).}
 \label{fig:rmd}
@@
 This joint influence is summarized in Figure \ref{fig:diff}, which shows the mean difference between the highest target type and the highest proportion of a type of a different kind in $1000$ independent populations across $\lambda$ and $l$ values.
-Higher values of $\lambda$ and $l$ increase the prevalence of a single target type, whereas lower values lead to less pronounced differences with a valley resulting from low $\lambda$ and high $l$ (cf.
+Higher values of $\lambda$ and $l$ increase the prevalence of a single target type, whereas lower values lead to less pronounced differences, with a valley resulting from low $\lambda$ and high $l$ (cf.
 Figure \ref{fig:rmd} with $\lambda = 1$ and $l=15$).
@@
 
-% \newpage
-
- \appendix
-\newpage
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/alt-fig1-onlyr}
-\caption{Representative population after $50$ generations under only a
-  pressure for communicative success. The right-most plots zoom in on only the proportion of
-  target types.}
-\label{fig:only-R}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/alt-fig2-onlym-pr}
-\caption{Representative population after $50$ generations under only a
-  pressure for learnability ($\lambda = 20, k = 5$). The types' prior is shown in the
-  right-most plot. The top-most groupings correspond to types, literal and pragmatic, with
-  lexica of the kind of $\Llack$, $\Lbound$, and $\Lall$.}
-\label{fig:only-M}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/alt-fig3-r+m}
-\caption{Representative population after $50$ generations under both pressures ($k = 5$).}
-\label{fig:rmd}
-\end{figure}
-
-% \begin{figure}[h!]
-% \centering
-% \includegraphics[width=1\textwidth,height=7.5cm,keepaspectratio]{./plots/fig4-incumbents-difference}
-% \caption{Difference between mean proportion of highest pragmatic $\Llack$ type and highest
-%   other type in $1000$ independent populations after $50$ generations under both pressures ($k
-%   = 5$).}
-% \end{figure}
-
 \end{document}

### 2017-04-12 code-LOT-extension/plots.py

@@ -656,14 +656,19 @@ def get_subfigs_hist_replication(type_list,list1,list2):
     plt.bar(X,Y1_large,color=colors,alpha=al)
     plt.axhline(y=df1_majority,linestyle='dashed')
     plt.text(0.05,df1_majority-0.0015,'majority type',color='royalblue',fontweight='extra bold')
+    ax1_large.tick_params(axis='both', which='major', labelsize=17)
+
 
     ax2_large = plt.subplot(1,3,2);
     plt.bar(X,Y2_large,color=colors,alpha=al)
     plt.axhline(y=df2_majority,linestyle='dashed')
+    ax2_large.tick_params(axis='both', which='major', labelsize=17)
+
 
     ax3_large = plt.subplot(1,3,3);
     plt.bar(X,Y3_large,color=colors,alpha=al)
     plt.axhline(y=df3_majority,linestyle='dashed')#, color=sns.color_palette()[1],linestyle ='dashed', markevery=10)
+    ax3_large.tick_params(axis='both', which='major', labelsize=17)
 
 #
     #Layout
@@ -681,13 +686,17 @@ def get_subfigs_hist_replication(type_list,list1,list2):
     p2 = patches.Rectangle((0, 0), 1, 1, fc="darkorange",alpha=al)
     p3 = patches.Rectangle((0, 0), 1, 1, fc="red",alpha=al)
 
-    ax1_large.legend([p1, p2, p3], ['Target types','Competitor types','prag. L-all'],loc='best',prop={'size':14})
+    ax1_large.legend([p1, p2, p3], ['Target types','Competitor types','prag. L-all'],loc=(0.0340644,0.69441),prop={'size':14})
 
     ax2_large.set_xlabel(r'$\lambda$ = '+str(list1[1]),fontsize=25)
     ax2_large.xaxis.set_label_position('top')
+#    ax2_large.text(5,-0.015,'Types',fontsize=20)
 
     ax3_large.set_xlabel(r'$\lambda$ = '+str(list1[2]),fontsize=25)
     ax3_large.xaxis.set_label_position('top')
+
+
+
 
     plt.tight_layout()
     plt.show()
@@ -842,9 +851,13 @@ def get_subfigs_hist_mutation(type_list,list1,list2):
     p3 = patches.Rectangle((0, 0), 1, 1, fc="red",alpha=al)
 
[…]

### 2017-04-12 paper/paper-v3.tex
sligth draft changes

@@
 semantics and pragmatics could have evolved, why it would be so pervasive across natural
-languages and why it is that some expressions systematically draw from it while others semantically conventionalize. Models of language evolution abound. There are simulation-based models studying
+languages, and why it is that some expressions systematically draw from it while others semantically conventionalize. Models of language evolution abound. There are simulation-based models studying
 populations of communicating agents
@@
 
-We spell out a model of the co-evolution of conventional meaning and pragmatic reasoning
-types. The objects of replication and selection are pairs consisting of a set of lexical
+We spell out a model of the co-evolution of conventional meaning and pragmatic reasoning.
+The objects of replication and selection are pairs consisting of a set of lexical
 meanings and a manner of pragmatic behavior. The latter are represented using probabilistic
@@
 Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. We
-discuss a setting in which the majority view of underspecified lexical meanings and pragmatic
+discuss a setting in which the majority view of underspecified lexical meanings and systematic pragmatic
 enrichments emerges if selection and transmission infidelity are combined. In particular, we
@@
 pressure towards communicative efficiency is not the only force that shapes
-language. Learnability is another as natural languages need to be learnable to survive their
+language. Learnability is another, as natural languages need to be learnable to survive their
 faithful transmission across generations. Furthermore, even small learning biases implicit in acquisition
@@
 \begin{example}
-  Consider a simple coordination game. Agents are of two types: positive or negative. If
+  Consider a simple and abstract coordination game. Agents are of two types: positive or negative. If
   agents of different type play with each other, they obtain a payoff of 0. If negative meets
@@
 \citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a
-(possibly fuzzy) set of states. A pragmatic type specifies a probabilistic speaker
+(possibly fuzzy) set of states. A pragmatic behavior specifies a probabilistic speaker
 rule (a probabilistic choice of message for each state) and a probabilistic hearer rule (a
@@
       \mygray{\msome} & .59 & .41 \\
-      \mygray{\mall}  & 0.35 & 0.65 \\
+      \mygray{\mall}  & .35 & .65 \\
     \end{block}
@@
 With low $\lambda$ speakers choose true messages with more slack.
-Reasoning over this behavior therefore also results in a weaker association of messages with only true states in receivers, but also in a slightly stronger association of $\msome$ with $\ssome$ over $\sall$ for $\Llack$ users because $S_0(\msome | \ssome;\Llack) > S_0(\msome | \sall;\Llack)$.
+Reasoning over this behavior therefore also results in a weaker association of messages with only true states in receivers, but also in a slightly stronger association of $\msome$ with $\ssome$ over $\sall$ for $\Llack$ users because they reason that $S_0(\msome | \ssome;\Llack) > S_0(\msome | \sall;\Llack)$.
 For $\lambda = 20$, there will be less slack in literal speaker behavior:
@@
 \end{align*}
-And accordingly less slack in pragmatic interpretation:
+And accordingly less slack in level $1$ pragmatic interpretation:
 \begin{align*}
@@
 \end{align*}
-For $\lambda = 20$, pragmatic speaker behavior of level 1 is as follows:
+For $\lambda = 20$, pragmatic speaker behavior of level 1 is instead as follows:
 \begin{align*}
@@
 the probability $P(d \mid t_j)$ with which sequence $d = \tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$ is observed depends on type $j$'s
-behavior:
+linguistic behavior:
 \begin{align*}
@@
 
-The model has three parameters: $\lambda$ regulates the degree to which pragmatic speakers
+The model has three parameters: $\lambda$ regulates the degree to which speakers
 choose messages that appear optimal from the point of view of the agent's own utility measure
@@
 $k$ is length of observations for each language learner; $l$ regulates where the learners'
-inference behavior lies on a spectrum from probability matching to choice of the most likely
+inference behavior lies on a spectrum from probability matching to acquisition of the most likely
 parent type.

### 2017-04-14 paper/cogsci-submission1.tex
submitted version of paper

--- 2017-04-12 paper/paper-v3.tex
@@
 
-\usepackage{geometry}
+\usepackage[margin=1in]{geometry}
 \usepackage{natbib}
@@
 %
-%\usepackage{setspace}
-%\onehalfspacing
+\usepackage{setspace}
+\doublespacing
 %
@@
 
-\begin{example}
-  Consider a simple and abstract coordination game. Agents are of two types: positive or negative. If
-  agents of different type play with each other, they obtain a payoff of 0. If negative meets
-  negative, each receives a payoff of 1. If positive meets positive, they get a payoff of 2. A
-  population state is completely characterized by the proportion $x$ of negatives. The fitness
-  of negatives in population state $x$ is $f_n(x) = x$, that of positives is $f_p(x) = 2-2x$.
-  The average fitness is $\Phi(x) = x f_n(x) + (1-x) f_p(x) = 3x^2 - 4x + 2$. Without mutation,
-  the replicator dynamic will then update $x$ to $RD(x) = \nicefrac{f(x)x}{\Phi(x)} = \nicefrac{x^2}{\Phi(x)}$. The update
-  function $RD(x)$ of the replicator step is plotted in Figure~\ref{fig:Updates_RMS} as the
-  blue line. Rest points, for which $RD(x)=x$, are at $x=0$, $x=1$ and $x= \nicefrac{2}{3}$.
-  The former are attractors as nearby points converge to them. Points near $x=\nicefrac{2}{3}$
-  move towards 0 or 1. This is schematically pictured in the topmost phase portrait in
-  Figure~\ref{fig:Phase_RD}.
-
-  Adding mutation changes the dynamic and its rest points. Let us assume that $Q_{ji} = .9$
-  when $j=i$. The update effect of mutation alone is $M(x) = .9 x + .1 (1-x) = .8x + .1$ and is
-  plotted as the linear green line in Figure~\ref{fig:Updates_RMS}. It has only one stable rest
-  point at $x = 0.5$ (see Figure~\ref{fig:Phase_RD}). If we first take the replicator step and
-  then the mutation step in sequence, we obtain the replicator mutator dynamic
-  $RMD(x) = M(RD(x)) = \nicefrac{.9x^2 - .2x +.2}{3x^2 - 4x + 2}$, which is plotted in red in
-  Figure~\ref{fig:Updates_RMS}. The rest points are at $x=.121$, $x=.903$ and $x=.609$. The
-  former two are attractors (see Figure~\ref{fig:Phase_RD}).
-\end{example}
-
 \begin{figure}[t]
@@
 
-  \caption{Example}
+  \caption{Example.}
   \label{fig:Example_RMD}
 \end{figure}
+
+\begin{example}
+  Consider a simple and abstract coordination game. Agents are of two types: positive or negative. If
+  agents of different type play with each other, they obtain a payoff of 0. If negative meets
+  negative, each receives a payoff of 1. If positive meets positive, they get a payoff of 2. A
+  population state is completely characterized by the proportion $x$ of negatives. The fitness
+  of negatives in population state $x$ is $f_n(x) = x$, that of positives is $f_p(x) = 2-2x$.
+  The average fitness is $\Phi(x) = x f_n(x) + (1-x) f_p(x) = 3x^2 - 4x + 2$. Without mutation,
+  the replicator dynamic will then update $x$ to $RD(x) = \nicefrac{f(x)x}{\Phi(x)} = \nicefrac{x^2}{\Phi(x)}$. The update
+  function $RD(x)$ of the replicator step is plotted in Fig.~\ref{fig:Updates_RMS} as the
+  blue line. Rest points, for which $RD(x)=x$, are at $x=0$, $x=1$ and $x= \nicefrac{2}{3}$.
+  The former are attractors as nearby points converge to them. Points near $x=\nicefrac{2}{3}$
+  move towards 0 or 1. This is schematically pictured in the topmost phase portrait in
+  Fig.~\ref{fig:Phase_RD}.
+
+  Adding mutation changes the dynamic and its rest points. Let us assume that $Q_{ji} = .9$
+  when $j=i$. The update effect of mutation alone is $M(x) = .9 x + .1 (1-x) = .8x + .1$ and is
+  plotted as the linear green line in Fig.~\ref{fig:Updates_RMS}. It has only one stable rest
+  point at $x = 0.5$ (see Fig.~\ref{fig:Phase_RD}). If we first take the replicator step and
+  then the mutation step in sequence, we obtain the replicator mutator dynamic
+  $RMD(x) = M(RD(x)) = \nicefrac{.9x^2 - .2x +.2}{3x^2 - 4x + 2}$, which is plotted in red in
+  Fig.~\ref{fig:Updates_RMS}. The rest points are at $x=.121$, $x=.903$ and $x=.609$. The
+  former two are attractors (see Fig.~\ref{fig:Phase_RD}).
+\end{example}
 
@@
 To fill the model from Section~\ref{sec:model} with life, we need to specify the sets of
-states, messages and lexica (Section~\ref{sec:stat-mess-conc}). Additionally, we want to
+states, messages and lexica we consider (Section~\ref{sec:stat-mess-conc}). Additionally, we want to
 explore the effects of a learning bias in favor of simple lexical representations. One way of
@@
 lexicalize the same concepts. Consequently, more often than not different variants of the same
-kind may fail to communicate with each other.
+kind fail to understand each other.
 
@@
   \begin{align*}
-    C & \rightarrow_2 C \wedge C 
-    & C & \rightarrow_2 \neg C  \\ 
-    C & \rightarrow_1 X \subseteq X &
-    C & \rightarrow_1 X \neq \emptyset &
-    C & \rightarrow_1 X = \emptyset  \\
+    R & \rightarrow_2 R \wedge R 
+    & R & \rightarrow_2 \neg R  \\ 
+    R & \rightarrow_1 X \subseteq X &
+    R & \rightarrow_1 X \neq \emptyset &
+    R & \rightarrow_1 X = \emptyset  \\
     X & \rightarrow_1 \set{A,B} & 
@@
 lists all six lexical representations relevant here, their truth conditions and the simplest formula that
-expresses this representation from the grammar in Table~\ref{tab:grammar}.
+expresses this representation in the grammar in Table~\ref{tab:grammar}.
 
@@
 \end{center}
-\caption{Available lexical representations and their minimal derivation cost}
+\caption{Available lexical representations and their minimal derivation cost.}
 \label{tab:concepts}
@@
 \begin{align*}
-  P(L)  & \propto \prod_{c \in Im(L)} P(c)   \ \text{, with} & 
-  P(c) & \propto \max_{c'}Compl(c') - Compl(c) + 1\,,
-\end{align*}
-where $Compl(c)$ is the complexity of the minimal derivation cost of $c$ according to the
+  P(L)  & \propto \prod_{r \in Im(L)} P(r)   \ \text{, with} & 
+  P(r) & \propto \max_{r'}Compl(r') - Compl(r) + 1\,,
+\end{align*}
+where $Compl(r)$ is the complexity of the minimal derivation cost of representation $r$ according to the
 LOT-grammar (see Table~\ref{tab:concepts}). Applied to our space of lexica, this construal assigns the highest probability 
@@
 
-As expected, selection based on communicative success is sensitive to $\lambda$ since $\lambda$
+Selection based on communicative success is sensitive to $\lambda$ since $\lambda$
 influences signaling behavior, which in turn determines communicative success. This is
-showcased in Figure \ref{fig:only-R}, which shows the proportion of target types, pragmatic competitor types and pragmatic $\Lall$ in a representative population after $50$ replicator steps.
+showcased in Fig.
+\ref{fig:only-R}, which shows the proportion of target types, pragmatic competitor types and pragmatic $\Lall$ in a representative population after $50$ replicator steps.
 The plot also indicates the proportion of the
@@
 and becomes very slow. The result is a very long transition with near stagnancy in a rather
-homogeneous population with many types. Conversely, higher $\lambda$ promotes more
+homogeneous population with many types. Conversely, higher $\lambda$ promotes 
 less-stochastic pragmatic behavior, widening the gap in expressivity between types and
-promoting more homogeneous populations. As suggested by Figure \ref{fig:only-R}, the majority
+promoting more homogeneous populations. As suggested by Fig. \ref{fig:only-R}, the majority
 in most populations is not one of the six pragmatic $\Llack$-style types. That is, a pressure
@@
 \includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/alt-fig2-onlym-pr}
-\caption{Proportion of target types, pragmatic competitor types, pragmatic $\Lall$, and population's majority type in representative populations after $50$ generations under only a
-  pressure for learnability ($\lambda = 20, k = 5$). The types' prior is shown in the
-  right-most plot. The top-most groupings correspond to types, literal and pragmatic, with
-  lexica of the kind of $\Llack$, $\Lbound$, and $\Lall$.}
+\caption{Proportion of target types, pragmatic competitor types, pragmatic $\Lall$, and the population's majority type in representative populations after $50$ generations under only a
+  pressure for learnability ($\lambda = 20, k = 5$). The learning prior is shown in the
+  right-most plot with top-most groupings corresponding to types, literal and pragmatic, with
+  lexica of kinds $\Llack$, $\Lbound$, and $\Lall$.}
 \label{fig:only-M}
@@
 sampling ($l = 1$) or a stronger tendency towards posterior maximization ($l = 15$) is shown in
-Figure \ref{fig:only-M} together with the prior over types. The prior shows that while users of
+Fig.\ref{fig:only-M} together with the prior over types. The prior shows that while users of
 $\Llack$ are not the most favored by the inductive bias (compared, e.g., to $\Lall$) they are
@@
 However, in contrast to a pressure only for communicative success with high $\lambda$ (see
-Figure~\ref{fig:only-R}), learnability alone does not succeed in selecting for a single
+Fig.~\ref{fig:only-R}), learnability alone does not succeed in selecting for a single
 prevalent type. Indeed, all six target types tend to coexist at roughly equal proportion. Each
@@
 
-Figure \ref{fig:rmd} illustrates the combined effects of both pressures for a sample of
+Fig. \ref{fig:rmd} illustrates the combined effects of both pressures for a sample of
 $\lambda$ and $l$ values. These results show that an inductive learning bias for simpler
@@
 a single majority target type.
-This joint influence is summarized in Figure \ref{fig:diff}, which shows the mean difference between the highest target type and the highest proportion of a type of a different kind in $1000$ independent populations across $\lambda$ and $l$ values.
+This joint influence is summarized in Fig.
+\ref{fig:diff}, which shows the mean difference between the highest target type and the highest proportion of a type of a different kind in $1000$ independent populations across $\lambda$ and $l$ values.
 Higher values of $\lambda$ and $l$ increase the prevalence of a single target type, whereas lower values lead to less pronounced differences, with a valley resulting from low $\lambda$ and high $l$ (cf.
-Figure \ref{fig:rmd} with $\lambda = 1$ and $l=15$).
+Fig.
+\ref{fig:rmd} with $\lambda = 1$ and $l=15$).
 
@@
 data. We see two main reasons for considering communicative success rather than just
-expressivity, and to only considering it in learning. First, learnability alone can lead to the
+expressivity, and over only considering it in learning. First, learnability alone can lead to the
 acquisition of functionally defective languages, as showcased by $\Lall$
-(cf. \citealt{kirby+etal:2008,silvey+etal:2014}, and \citealt{fay+etal:2013} for review of
+(cf. \citealt{kirby+etal:2008,silvey+etal:2014}, and see \citealt{fay+etal:2013} for review of
 laboratory results). Second and more importantly, types may be equally expressive but their
 performance as means of information transfer crucially depends not only on themselves but on
-the population they find themselves in (cf. the competition of target types in Figure
-\ref{fig:rmd} and their lack of competition in Figure \ref{fig:only-M}). That is, we contend
+the population they find themselves in (cf. the competition of target types in Fig. 
+\ref{fig:rmd} and their lack of competition in Fig. \ref{fig:only-M}). That is, we contend
 that having an expressive lexicon that generates learnable data does not in itself capture a
@@
 be classified as either semantic or pragmatic users of, in particular, \emph{some}
-\citep[e.g.][]{BottNoveck2004:Some-Utterances,NieuwlandDitman2010:On-the-incremen,DegenTanenhaus2012:Processing-Scal}. The
+\citep[e.g.,][]{BottNoveck2004:Some-Utterances,NieuwlandDitman2010:On-the-incremen,DegenTanenhaus2012:Processing-Scal}. The
 former consistently accept \emph{some} where \emph{all} would be true as well, the latter do
@@
 that the computation of a scalar implicature may involve additional cognitive effort
-\citep[e.g.][]{BrehenyKatsos2006:Are-Generalised,deNeys+schaeken:2007,huang+snedeker:2009,Jr.Bailey2013:Possibly-all-of}. Extensions
+\citep[e.g.,][]{BrehenyKatsos2006:Are-Generalised,deNeys+schaeken:2007,huang+snedeker:2009,Jr.Bailey2013:Possibly-all-of}. Extensions
 of the model presented here to include processing costs for pragmatic language use would be
@@
 into $t$, so to speak. Taken together, without an explicit model of the interaction between
-pressure for success and learnability, it is far from trivial to judge whether or when
+pressure for communicative success and learnability, it is far from trivial to judge whether or when
 preferred or dispreferred representations can be adaptive. This is why a major contribution of this
@@
 the speaker (cf. \citealt[252-267]{horn:1984},
-\citealt{horn:1972,traugott:2004,vdAuwera:2010}). For instance, two expressions for each weak
+\citealt{horn:1972,traugott:2004,vdAuwera:2010}). For instance, to the emergence of two expressions for each weak
 scalar expression; one with and one lacking an upper-bound.  Looking at pressure from
@@
 an interesting issue is whether proposed principles, such as the semantic conventionalization of once highly context-dependent inferences if they become regular enough \citep{levinson:2000,traugott:2004}, can be given
-a formal rationale and inform the postulated directionalities of change. The present work made a first start and gave a framework for
+a formal rationale and inform postulated directionalities of change. The present work made a first start and gave a framework for
 exploring exactly these issues systematically.
@@
 simpler semantics in combination with functional pressure in use may offer an answer to why
-natural languages fail to lexicalize systematic pragmatic inferences. And, more broadly, to a
+natural languages fail to lexicalize systematic pragmatic inferences. And, more broadly, lead to a
 division of labor between semantics and pragmatics.

### 2017-10-24 code-LOT-extension/co-evo/functions.py
co-evo phase+quiver draft

@@ -1,292 +1,0 @@
-import sys,os
-lib_path = os.path.abspath(os.path.join('..')) #specifying path for player module
-sys.path.append(lib_path) #specifying path for player module
-from player import LiteralPlayer, GriceanPlayer
-import numpy as np
-import matplotlib.pyplot as plt
-from itertools import product, combinations
-
-lam = 20
-type_indices = [0,2] #lb,ll,pb,pl
-bias_para = 2
-k = 5
-post_para = 15
-
-
-lexica = [np.array([[1.,0.],[0.,1.]]),np.array([[1.,0.],[1.,1.]])]
-typeList = [LiteralPlayer(lam,lex) for lex in lexica] + [GriceanPlayer(1,lam,lex) for lex in lexica]
-types = [typeList[x] for x in type_indices]
-
-def normalize(m):
-    return m / m.sum(axis=1)[:, np.newaxis]
-
-
-def get_u(types):
-    out = np.zeros([len(types), len(types)])
-    for i in xrange(len(types)):
-        for j in xrange(len(types)):
-            out[i,j] = (np.sum(types[i].sender_matrix * np.transpose(types[j].receiver_matrix)) / 2.) * 0.5 +\
-                       (np.sum(types[j].sender_matrix * np.transpose(types[j].receiver_matrix)) / 2.) * 0.5
-    return out
-
-U = get_u(types)
-
-def fitness(x_prop,U):
-    return np.sum((np.array([x_prop, 1-x_prop]) * U)[0])
-
-def overall_fitness(x_prop,U):
-    return (x_prop * np.sum((np.array([x_prop,1-x_prop]) * U)[0])) +\
-           ((1.-x_prop) * np.sum((np.array([x_prop,1-x_prop])* U)[1]))
-
-def replicator_step(x_prop):
-    return (x_prop * fitness(x_prop,U)) / overall_fitness(x_prop,U)
-
-
-def obs_counts(obs):
-    out = []
-    for i in xrange(len(obs)):
-        out.append([obs[i].count(j) for j in xrange(4)])
-    return out
-
-
-def get_likelihood(obs,likelihoods):
-    out = np.zeros([len(likelihoods), len(obs)]) # matrix to store results in
-    for lhi in range(len(likelihoods)):
-        for o in range(len(obs)):
-            flat_lhi = likelihoods[lhi].flatten()
[…]

### 2017-10-24 code-LOT-extension/co-evo/phase.py

@@ -1,0 +1,593 @@
+import sys,os
+lib_path = os.path.abspath(os.path.join('..')) #specifying path for player module
+sys.path.append(lib_path) #specifying path for player module
+from player import LiteralPlayer, GriceanPlayer
+import numpy as np
+import matplotlib.pyplot as plt
+from itertools import product, combinations
+import seaborn as sns
+
+lexica = [np.array([[1.,0.],[0.,1.]]),np.array([[1.,0.],[1.,1.]])]
+
+#lam = 20
+#type_indices = [0,1] #lb,ll,pb,pl
+#bias_para = 2
+#k = 8
+#post_para = 10
+
+#types = [typeList[x] for x in type_indices]
+def normalize(m):
+    return m / m.sum(axis=1)[:, np.newaxis]
+
+
+def get_u(types):
+    out = np.zeros([len(types), len(types)])
+    for i in xrange(len(types)):
+        for j in xrange(len(types)):
+            out[i,j] = (np.sum(types[i].sender_matrix * np.transpose(types[j].receiver_matrix)) / 2.) * 0.5 +\
+                       (np.sum(types[j].sender_matrix * np.transpose(types[j].receiver_matrix)) / 2.) * 0.5
+    return out
+
+def fitness(x_prop,U):
+    return np.sum((np.array([x_prop, 1-x_prop]) * U)[0])
+
+def overall_fitness(x_prop,U):
+    return (x_prop * np.sum((np.array([x_prop,1-x_prop]) * U)[0])) +\
+           ((1.-x_prop) * np.sum((np.array([x_prop,1-x_prop])* U)[1]))
+
+def replicator_step(x_prop,U):
+    return (x_prop * fitness(x_prop,U)) / overall_fitness(x_prop,U)
+
+def obs_counts(obs):
+    out = []
+    for i in xrange(len(obs)):
+        out.append([obs[i].count(j) for j in xrange(4)])
+    return out
+
+def get_likelihood(obs,likelihoods):
+    out = np.zeros([len(likelihoods), len(obs)]) # matrix to store results in
+    for lhi in range(len(likelihoods)):
+        for o in range(len(obs)):
+            flat_lhi = likelihoods[lhi].flatten()
+            out[lhi,o] = np.prod([flat_lhi[x]**obs[o][x] for x in xrange(len(obs[o]))]) 
+    return out
+
+
[…]

### 2017-10-24 code-LOT-extension/co-evo/quiver.py

@@ -1,0 +1,286 @@
+import sys,os
+lib_path = os.path.abspath(os.path.join('..')) #specifying path for player module
+sys.path.append(lib_path) #specifying path for player module
+from player import LiteralPlayer, GriceanPlayer
+import numpy as np
+import matplotlib.pyplot as plt
+from itertools import product, combinations
+
+lam = 20
+bias_para = 2
+k = 5
+post_para = 10
+
+
+lexica = [np.array([[1.,0.],[0.,1.]]),np.array([[1.,0.],[1.,1.]])]
+types = [LiteralPlayer(lam,lex) for lex in lexica] + [GriceanPlayer(1,lam,lex) for lex in lexica]
+
+def normalize(m):
+    return m / m.sum(axis=1)[:, np.newaxis]
+
+
+def get_u(types):
+    out = np.zeros([len(types), len(types)])
+    for i in xrange(len(types)):
+        for j in xrange(len(types)):
+            out[i,j] = (np.sum(types[i].sender_matrix * np.transpose(types[j].receiver_matrix)) / 2.) * 0.5 +\
+                       (np.sum(types[j].sender_matrix * np.transpose(types[j].receiver_matrix)) / 2.) * 0.5
+    return out
+
+def obs_counts(obs):
+    out = []
+    for i in xrange(len(obs)):
+        out.append([obs[i].count(j) for j in xrange(4)])
+    return out
+
+
+def get_likelihood(obs,likelihoods):
+    out = np.zeros([len(likelihoods), len(obs)]) # matrix to store results in
+    for lhi in range(len(likelihoods)):
+        for o in range(len(obs)):
+            flat_lhi = likelihoods[lhi].flatten()
+            out[lhi,o] = np.prod([flat_lhi[x]**obs[o][x] for x in xrange(len(obs[o]))]) 
+    return out
+
+
+def get_q(lexica_prior,learning_parameter,k):
+    likelihoods = [i.sender_matrix for i in types]
[…]

### 2017-10-25 code-LOT-extension/co-evo/co-evolution-notes.tex
sketch on co-evo presentation for min. types spaces

\documentclass[fleqn,reqno,10pt]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{mathrsfs}
\usepackage{nicefrac}
%\usepackage{stmaryrd}
%\usepackage{multicol}
\usepackage{graphicx}
\usepackage{color}
\usepackage{booktabs}
\usepackage{pgfplots}
\usepackage{subcaption}

\usepackage{blkarray}
\usepackage{xspace}

\usepackage{tgtermes}
\renewcommand{\baselinestretch}{1.2}

\newcommand{\mylang}[1]{\ensuremath{L_{\text{#1}}}\xspace} %meaningful states
\newcommand{\mystate}[1]{\ensuremath{\state_{\text{#1}}}\xspace} %meaningful states
\newcommand{\mymessg}[1]{\ensuremath{\messg_{\text{#1}}}\xspace} %meaningful messages
\newcommand{\messg}{\ensuremath{m}\xspace}		% single messages
\newcommand{\state}{\ensuremath{s}\xspace}		% single states

\newcommand{\Lbound}{\mylang{bound}}
\newcommand{\Llack}{\mylang{lack}}
\newcommand{\ssome}{\mystate{\ensuremath{\exists\neg\forall}}}
\newcommand{\sall}{\mystate{\ensuremath{\forall}}}
\newcommand{\snone}{\mystate{\ensuremath{\emptyset}}}
\newcommand{\msome}{\mymessg{some}}
\newcommand{\mall}{\mymessg{all}}

\definecolor{mygray}{cmyk}{0.35,0.35,0.35,0.35}
\newcommand{\mygray}[1]{{\textcolor{mygray}{#1}}}

\title{Notes on presentation of model for co-evolution}
\author{Thomas Brochhagen}
\date{}

\begin{document}
\maketitle

\subsection*{Motivation}

We would like to present our model in a cleaner and more intuitive way. As suggested by Michael, one way to do so is to focus first on a reduced type space to illustrate how the model works. As a reminder:

\begin{align*}
  \Lbound & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 0 & 1 \\
    \end{block}
  \end{blockarray} &
  \Llack & = \begin{blockarray}{lcc}
    & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[cc]}
      \mygray{\ssome} & 1 & 0 \\
      \mygray{\sall}  & 1 & 1 \\
    \end{block}
  \end{blockarray}
\end{align*}

These two lexica are paired with either level-$0$ {\em literal} behavior, or level-$1$ {\em pragmatic} behavior. Learnability-wise, if there is a bias for simpler semantics (no-upper bound), then the two types with $\Llack$ are favored {\em a priori}. Furthermore, both $\Lbound$ users are indiscernible when it comes to learning; they only differ in their receiver behavior. Utility-wise, literal $\Lbound$ fares slightly better than pragmatic $\Lbound$. If $\lambda$ is sufficiently high, then the difference between pragmatic $\Llack$ and both $\Lbound$ types is almost negligible. We want to visualize how (co-)evolution plays out by comparing four populations of two types each, where each pair of types differs only in (i) signaling behavior, or (ii) lexica:

\begin{enumerate}
  \item literal $\Lbound$ vs. literal $\Llack$;
  \item pragmatic $\Lbound$ vs. pragmatic $\Llack$;
  \item literal $\Lbound$ vs. pragmatic $\Lbound$;
  \item literal $\Llack$ vs. pragmatic $\Llack$.
\end{enumerate}

The other relevant parameters of the model are: soft-max parameter $\lambda$, posterior parameter $l$, sequence length $k$, and bias parameter $c$ (detailed below).

\subsection*{Learning bias}
In the paper, the learners' inductive bias is a function of the complexity of the formulae that types lexicalize. Introducing our measure and applying it only to this small type space is probably not the way to go. For illustratory purposes, we may instead assume that there's an abstract parameter (which we later make concrete using our measure) $c \geq 1$ that regulates how strongly $\Llack$ is preferred over $\Lbound$. In what follows, I assumed this to be $P(L_i) \propto c$ if $L_i$ is $\Llack$ and otherwise $1$. For instance, if we have  a type space with $\tau_i$ as literal $\Lbound$ and $\tau_j$ as pragmatic $\Llack$, then $P(\tau_i) \approx .33$ for $c = 2$. This is just for illustration, of course. We can change this at will. As far as I can see, the reduced type spaced behave in a very predictable manner. I set $c = 2$ for all the plots below.

\subsection*{General remarks}
First, I did not succeed in putting the model in an elegant closed form. The good news is that the type space is small enough so that we don't need to sample anything. In particular, we don't need to sample from speaker probabilities to approximate $Q$ for large $k$; we can look at all possible state-message sequences instead, meaning that we get the same result each time we apply replicator or mutator steps.

Second, there are some issues with phase-portrait-like representations. The main problem is that I have not (yet?) found an algorithm that succeeds in reliably finding stable points for mutator (or replicator-mutator) steps.

Third, one of the major disadvantages of the smaller type spaces is that (I fear that) they give the impression that the mutator-step is doing most the work. As shown below, $M(x)$ is very close to $M(RD(x))$. This is not true of the bigger type space, but makes sense in these reduced ones. After all, learning is what differentiates $\Llack$ from $\Lbound$. I'm concerned that these results may give a wrong impression.

Lastly, I think that all the representations I generated have some negative sides (details below). Before improving on them, I thought we should discuss these matters, making use of the fact that their flaws are evident.

\subsection*{Phase portraits}
\begin{figure}
\centering
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.4]{phases-lam01-bias2-l01-k5}
  \caption{}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.4]{phases-lam20-bias2-l10-k5}
  \caption{}
  \label{fig:sub2}
\end{subfigure}
\caption{Approximate phase-portraits for (a) $\lambda = 1, c = 2, l = 1, k = 5$ and (b) $\lambda = 20, c = 2, l =10, k =5$}
\label{fig:phase}
\end{figure}

Figure \ref{fig:phase} shows $8$ (approximate) portraits for applications of replicator steps only, mutator steps only, and replicator-mutator steps on our four type-pairings, for two parameter constellations. For instance, the upper-left-most portrait shows the dynamics for a type space consisting of literal $\Lbound$ ($x$) and literal $\Llack$ ($1-x$). We can see that, as long as there are some $\Lbound$ types in the population, the population progresses toward the extreme where only this type survives.

I used a couple of heuristics to generate these plots. First, I took $1000$ points in $[0,1]$ and applied the dynamic in question. Then, I checked if there were any points for which $D(x) = x$. As expected, the only stable points found in this way are the extremes for the replicator step: $RD(1) = 1$ and $RD(0) = 0$. To check whether these points are attractors, I looked at adjacent points to figure out whether they moved toward or away from that point. I did the same for the arrows, evenly spread across the axis.
\begin{itemize}
\item[{\bf The good}] It's easy to see how the dynamics play out. 
\item[{\bf The bad}] It's not easy to find stable points when mutation is involved. As mentioned above, I tried a couple of optimization algorithms but didn't get anywhere so far. Additionally, $M(x)$ and $M(RD(x))$ are (almost) indistinguishable. This is obviously an issue for all the depictions I played around with. It's not as obvious in these ones though.
\item[{\bf The ugly}] There are {\em a lot} of points where $RD(x) = x$ in the type space consisting of only literal and pragmatic $\Lbound$. This explains why these two phase portraits look so ugly; there are many overlapping circles.
\end{itemize}
{\bf My take:} I think these plots are nice but that they require the most cleaning. The major advantage I see here is that, even without finding stable points, it's easy to see where $M(\cdot)$ and $M(RD(\cdot))$ are headed (having less arrows would make this a lot easier to see). This representation slightly masks my big concern of mutator- and RM-steps being almost the same. As for the problem of too many stable points: We could be a little more selective about which points to depict (e.g., relative to the basin of attraction) or use other pictorial means (see the last set of plots).

\subsection*{Simple plots}
\begin{figure}
\centering
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.45]{plot-lam01-bias2-l01-k5}
  \caption{}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.45]{plot-lam20-bias2-l10-k5}
  \caption{}
  \label{fig:sub2}
\end{subfigure}
\caption{Evolutionary competition between pairs of types with (a) $\lambda = 1, c = 2, l = 1, k = 5$ and (b) $\lambda = 20, c = 2, l =10, k =5$. The $x$-axis shows proportion of $x$ before the update, the $y$-axis shows the result of a step of the dynamic on this $x$-value.}
\label{fig:plots}
\end{figure}

Figure \ref{fig:plots} shows $8$ plots using the same data as above. The main difference is that there's no information about stable points and that the directionality of change, particularly in the case of replication only, is harder to see.

\begin{itemize}
\item[{\bf The good}] It's very easy to see the effects of mutation only and the RMD. In particular, that (i) all of $Q$'s cells are equal to $.5$ in populations of only literal and pragmatic $\Lbound$, (ii) but that there is a clear difference in the learnability of literal and pragmatic $\Llack$ due to differences in their likelihoods when $\lambda > 1$.
\item[{\bf The bad}] It's not evident that there are differences across applications of the replicator step. Additionally, these plots may give the impression that replication {\em in general} is doing very little. Particularly when it comes to a comparison between $M(x)$ and $RMD(x)$. I must concede that it is true that replication is doing very little in these reduced type spaces.
\end{itemize}

{\bf My take:} I don't think that we should use plots like these as they don't serve the illustrative purpose we're after. I included them because they clarified a lot of questions I had when I first looked at the phase portraits.

\subsection*{Dynamics}
\begin{figure}
\centering
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.45]{quiver-lam1-c2-k5-l1}
  \caption{}
  \label{fig:sub1}
\end{subfigure}%
\begin{subfigure}{.6\textwidth}
  \centering
  \includegraphics[scale=0.45]{quiver-lam20-c2-k5-l10}
  \caption{}
  \label{fig:sub2}
\end{subfigure}
\caption{Evolutionary competition between pairs of types with (a) $\lambda = 1, c = 2, l = 1, k = 5$ and (b) $\lambda = 20, c = 2, l =10, k =5$. Coordinates $(x,y)$ show (proportion of pragmatic types, proportion of $\Llack$ types).}
\label{fig:quiver}
\end{figure}

Lastly, I wanted to try my hand at compressing the information more while trying to highlight the main take away of the dynamics applied to these reduced type spaces. Figure \ref{fig:quiver} is an illustration. I used the sketch Michael sent us with his comments as inspiration. We think of the type space as being $2$ dimensional. Along the four edges we have same types competing as we had in the previous plots. This $2D$ space is ambiguous once you move from the edges, so I only applied the dynamics on edges. More precisely, in each plot, the $x$-coordinate indicates the proportion of literal types and the $y$-coordinate indicates the proportion of $\Llack$ users. Points that are not on edges are ambiguous because, e.g., $(.5,.5)$ could mean $50\%$ are literal $\Lbound$ and $50\%$ are pragmatic $\Llack$ but also that there are $25\%$ of each of the four types in the population. The edges are unambiguous because each population consists of maximally two types, opposed with respect to a single dimension. This explains why replication in Figure \ref{fig:quiver} only moves along edges.

\begin{itemize}
\item[{\bf The good}] These plots look very similar to what Michael sketched out by hand. The RD plots are also a lot cleaner than the phase portraits I made above.
\item[{\bf The bad}] The ambiguity of $(x,y)$ away from edges.
\item[{\bf The ugly}] Its painfully evident that mutation steps and RMD steps are almost identical. Additionally, vectors are not scaled. Right now, they are automatically resized relative to the other vectors on their edge. This makes it easier to show small effects. If I scale the vectors then the RD plots look as if they consisted only of points because the mutator-steps are much larger. Note also that the movement from pragmatic $\Lbound$ to literal $\Lbound$ under replication only is only apparent when we look at enough points along the edge. However, this makes the plots look cluttered everywhere else. This is fixable and mostly cosmetic though.
\end{itemize}
{\bf My take:} I like these the most. Of course, we could also generate each edge separately and present them in a format similar to that of phase portraits above.

I couldn't come up with anything, but: Is there a space in which we could disentangle the ambiguity of the $2D$ space? That is, a shape that is as the regular $3$-simplex is to the regular $2$-simplex?

\end{document}

### 2017-10-25 code-LOT-extension/co-evo/phase.py

@@ -5,7 +5,7 @@ from player import LiteralPlayer, GriceanPlayer
 import numpy as np
 import matplotlib.pyplot as plt
 from itertools import product, combinations
-import seaborn as sns
+#import seaborn as sns
 
 lexica = [np.array([[1.,0.],[0.,1.]]),np.array([[1.,0.],[1.,1.]])]
 
@@ -394,7 +394,7 @@ def plot_phase_portraits(lam,bias_para,post_para,k):
     plt.show()
 
 
-#plot_phase_portraits(1,2,1,5)
+#plot_phase_portraits(20,2,10,5)
 
 def simple_plot(lam,bias_para,post_para,k):
     typeList = [LiteralPlayer(lam,lex) for lex in lexica] + [GriceanPlayer(1,lam,lex) for lex in lexica]
@@ -480,7 +480,7 @@ def simple_plot(lam,bias_para,post_para,k):
     plt.tight_layout()
     plt.show()
 
-#simple_plot(20,2,10,5)
+simple_plot(1,2,1,5)
 
 
         

### 2017-10-25 code-LOT-extension/co-evo/quiver.py

@@ -6,14 +6,12 @@ import numpy as np
 import matplotlib.pyplot as plt
 from itertools import product, combinations
 
-lam = 20
-bias_para = 2
-k = 5
-post_para = 10
+#lam = 20
+#bias_para = 2
+#k = 5
+#post_para = 10
 
 
-lexica = [np.array([[1.,0.],[0.,1.]]),np.array([[1.,0.],[1.,1.]])]
-types = [LiteralPlayer(lam,lex) for lex in lexica] + [GriceanPlayer(1,lam,lex) for lex in lexica]
 
 def normalize(m):
     return m / m.sum(axis=1)[:, np.newaxis]
@@ -43,7 +41,7 @@ def get_likelihood(obs,likelihoods):
     return out
 
 
-def get_q(lexica_prior,learning_parameter,k):
+def get_q(lexica_prior,learning_parameter,k,types):
     likelihoods = [i.sender_matrix for i in types]
     atomic_obs = [0,1,2,3] #0,1 upper-row; 2,3 are lower row
     D = list(product(atomic_obs,repeat=k))  
@@ -66,12 +64,12 @@ def learning_prior(types,c):
             out[i] = 1*c
     return out / np.sum(out)
 
-def rmd(p,u,q):
+def rmd(p,u,q,types):
     pPrime = p * [np.sum(u[t,] * p)  for t in xrange(len(types))]
     pPrime = pPrime / np.sum(pPrime)
     return np.dot(pPrime, q)
 #
-def rep(p,u):
+def rep(p,u,types):
     pPrime = p * [np.sum(u[t,] * p) for t in xrange(len(types))]
     return pPrime / np.sum(pPrime)
 
@@ -80,9 +78,7 @@ def mut(p,q):
 
 
 
-lexica_prior = learning_prior(types,bias_para)
-q = get_q(lexica_prior,post_para,k)
-u = get_u(types)
+
 
 
 def coords_x1(edge,pops):
@@ -118,91 +114,126 @@ def coords_y2(edge,pops):
     return [u,v]
 
 
-#def quiver_plot(lam,bias_para,post_para,k):
[…]

### 2017-11-26 code-LOT-extension/co-evo/co-evolution-notes2.tex
update, filled quiver plots for co-evo illustration

--- 2017-10-25 code-LOT-extension/co-evo/co-evolution-notes.tex
@@
 
-\title{Notes on presentation of model for co-evolution}
+\title{Notes 2.0 on presentation of model for co-evolution}
 \author{Thomas Brochhagen}
@@
 We can change this at will.
-As far as I can see, the reduced type spaced behave in a very predictable manner.
-I set $c = 2$ for all the plots below.
+As far as I can see, the reduced type spaced behaves in a very predictable manner.% I set $c = 2$ for all the plots below.
 
-\subsection*{General remarks}
-First, I did not succeed in putting the model in an elegant closed form.
-The good news is that the type space is small enough so that we don't need to sample anything.
-In particular, we don't need to sample from speaker probabilities to approximate $Q$ for large $k$; we can look at all possible state-message sequences instead, meaning that we get the same result each time we apply replicator or mutator steps.
+\subsection*{General remarks 2.0}
+There are three remarks to be made about this update.
 
-Second, there are some issues with phase-portrait-like representations.
-The main problem is that I have not (yet?) found an algorithm that succeeds in reliably finding stable points for mutator (or replicator-mutator) steps.
+First, the phase plots are a lot cleaner and all show stationary points.
+These were either calculated, or estimated as being in the middle of two points that changed in opposite directions.
+As before, the RD competition between literal and pragmatic $\Lbound$ is messy because there are many points for which $RD(x) = x$.
+On this front, I just changed white circles with black edges to red circles without edges for better visibility when many circles overlap.
 
-Third, one of the major disadvantages of the smaller type spaces is that (I fear that) they give the impression that the mutator-step is doing most the work.
-As shown below, $M(x)$ is very close to $M(RD(x))$.
-This is not true of the bigger type space, but makes sense in these reduced ones.
-After all, learning is what differentiates $\Llack$ from $\Lbound$.
-I'm concerned that these results may give a wrong impression.
+Second, when it comes to the direct competition of only two types on a line (phase plots), M and RMD still produce pretty similar outcomes.
+I tested a range of learning bias values and larger $k$ but there's not much change.
+This is not really troublesome (it being a consequence of there only being two types in a population), but I think this may disqualify the phase plots as candidates for a good illustration of the model's dynamics.
 
-Lastly, I think that all the representations I generated have some negative sides (details below).
-Before improving on them, I thought we should discuss these matters, making use of the fact that their flaws are evident.
+Third, I tried my hand at the 2D quiver plots with all types, not only at the edges but also at the center of the unit square.
+As Michael suggested, $(x,y)$-pairs in the square give the following proportions of types:
+\begin{itemize} 
+  \item $lit. \; \Lbound$: $(1-x) * (1-y)$ 
+  \item $lit. \; \Llack$: $(1-x) * y$
+  \item $prag.\; \Lbound$: $x * (1-y)$
+  \item $prag.\; \Llack$: $x * y$
+\end{itemize}
+
+Overall, I think these plots look good.
+Important, to me, is that (i) M and RMD are clearly distinguishable and that (ii) it's clear that RD with lowish $\lambda$ leads to $\Lbound$.
+Another plus is that it's easy to explain how this plot came about.
+On the negative side, I'm still a bit unhappy about having to assume that prag.
+and $\Llack$ are independent, but this is clearly better than just plotting the edges.
+The one important I am missing from these plots is clear movement on the lower edge, from $(1,0)$ to $(0,0)$, which would indicate movement from prag.
+$\Lbound$ to lit.
+$\Lbound$,  at least for RD-only.
+This would help us in driving the message home that pragmatic recruitment is not functionally advantageous for all types.
+I think the problem right now is that movement on this edge progresses in very small steps (if at all; cf.
+phase plots).
+These small steps disappear once we scale them with the bigger change that happens elsewhere in the square.
+Reinterpreting $(x,y)$ would not help on the edge itself, but maybe taking average trajectories would at least make points close to the edge move toward lit.
+$\Llack$ rather than toward its pragmatic counterpart.
 
@@
   \centering
-  \includegraphics[scale=0.4]{phases-lam01-bias2-l01-k5}
+  \includegraphics[scale=0.3]{phases-lam01-bias105-l01-k8}
   \caption{}
@@
   \centering
-  \includegraphics[scale=0.4]{phases-lam20-bias2-l10-k5}
+  \includegraphics[scale=0.3]{phases-lam20-bias105-l10-k8}
   \caption{}
@@
 \end{subfigure}
-\caption{Approximate phase-portraits for (a) $\lambda = 1, c = 2, l = 1, k = 5$ and (b) $\lambda = 20, c = 2, l =10, k =5$}
+\caption{Approximate phase-portraits for (a) $\lambda = 1, c = 1.05, l = 1, k = 8$ and (b) $\lambda = 20, c = 1.05, l =10, k =8$.
+Attractors are in black.
+Non-attractor points are in red for better visibility for lit.
+$\Lbound$ vs prag.
+$\Lbound$, which has many.}
 \label{fig:phase}
@@
 
-Figure \ref{fig:phase} shows $8$ (approximate) portraits for applications of replicator steps only, mutator steps only, and replicator-mutator steps on our four type-pairings, for two parameter constellations.
-For instance, the upper-left-most portrait shows the dynamics for a type space consisting of literal $\Lbound$ ($x$) and literal $\Llack$ ($1-x$).
+As before, Figure \ref{fig:phase} shows $8$ (approximate) portraits for applications of replicator steps only, mutator steps only, and replicator-mutator steps on our four type-pairings, for two parameter constellations ($c$ is now $1.05$ instead of $2$).
+For instance, the upper-left-most portrait shows the dynamic for a type space consisting of literal $\Lbound$ ($x$) and literal $\Llack$ ($1-x$).
 We can see that, as long as there are some $\Lbound$ types in the population, the population progresses toward the extreme where only this type survives.
 
-I used a couple of heuristics to generate these plots.
-First, I took $1000$ points in $[0,1]$ and applied the dynamic in question.
-Then, I checked if there were any points for which $D(x) = x$.
-As expected, the only stable points found in this way are the extremes for the replicator step: $RD(1) = 1$ and $RD(0) = 0$.
-To check whether these points are attractors, I looked at adjacent points to figure out whether they moved toward or away from that point.
-I did the same for the arrows, evenly spread across the axis.
-\begin{itemize}
-\item[{\bf The good}] It's easy to see how the dynamics play out. 
-\item[{\bf The bad}] It's not easy to find stable points when mutation is involved.
-As mentioned above, I tried a couple of optimization algorithms but didn't get anywhere so far.
-Additionally, $M(x)$ and $M(RD(x))$ are (almost) indistinguishable.
-This is obviously an issue for all the depictions I played around with.
-It's not as obvious in these ones though.
-\item[{\bf The ugly}] There are {\em a lot} of points where $RD(x) = x$ in the type space consisting of only literal and pragmatic $\Lbound$.
-This explains why these two phase portraits look so ugly; there are many overlapping circles.
-\end{itemize}
-{\bf My take:} I think these plots are nice but that they require the most cleaning.
-The major advantage I see here is that, even without finding stable points, it's easy to see where $M(\cdot)$ and $M(RD(\cdot))$ are headed (having less arrows would make this a lot easier to see).
-This representation slightly masks my big concern of mutator- and RM-steps being almost the same.
-As for the problem of too many stable points: We could be a little more selective about which points to depict (e.g., relative to the basin of attraction) or use other pictorial means (see the last set of plots).
-
-\subsection*{Simple plots}
-\begin{figure}
-\centering
-\begin{subfigure}{.6\textwidth}
-  \centering
-  \includegraphics[scale=0.45]{plot-lam01-bias2-l01-k5}
-  \caption{}
-  \label{fig:sub1}
-\end{subfigure}%
-\begin{subfigure}{.6\textwidth}
-  \centering
-  \includegraphics[scale=0.45]{plot-lam20-bias2-l10-k5}
-  \caption{}
-  \label{fig:sub2}
-\end{subfigure}
-\caption{Evolutionary competition between pairs of types with (a) $\lambda = 1, c = 2, l = 1, k = 5$ and (b) $\lambda = 20, c = 2, l =10, k =5$.
-The $x$-axis shows proportion of $x$ before the update, the $y$-axis shows the result of a step of the dynamic on this $x$-value.}
-\label{fig:plots}
-\end{figure}
-
-Figure \ref{fig:plots} shows $8$ plots using the same data as above.
-The main difference is that there's no information about stable points and that the directionality of change, particularly in the case of replication only, is harder to see.
-
-\begin{itemize}
-\item[{\bf The good}] It's very easy to see the effects of mutation only and the RMD.
-In particular, that (i) all of $Q$'s cells are equal to $.5$ in populations of only literal and pragmatic $\Lbound$, (ii) but that there is a clear difference in the learnability of literal and pragmatic $\Llack$ due to differences in their likelihoods when $\lambda > 1$.
-\item[{\bf The bad}] It's not evident that there are differences across applications of the replicator step.
-Additionally, these plots may give the impression that replication {\em in general} is doing very little.
-Particularly when it comes to a comparison between $M(x)$ and $RMD(x)$.
-I must concede that it is true that replication is doing very little in these reduced type spaces.
-\end{itemize}
-
-{\bf My take:} I don't think that we should use plots like these as they don't serve the illustrative purpose we're after.
-I included them because they clarified a lot of questions I had when I first looked at the phase portraits.
+Beyond having cleaner portraits, my take on these is as before: They are nice because it's easy to see how the dynamics play out but they do not highlight difference between $M(x)$ and $M(RD(x))$ very well.
+The most pronounced differences I have found are when $\lambda$ is high, $c$ low, and learners tend to sample from the posterior posterior sampling (e.g.; $\lambda = 20$, $c = 1.05$, $l = 1$).
+However, that's a rather esoteric parameter constellation to illustrate the dynamics with.
 
@@
   \centering
-  \includegraphics[scale=0.45]{quiver-lam1-c2-k5-l1}
+  \includegraphics[scale=0.4]{quiver-lam1-c105-k8-l1}
   \caption{}
@@
   \centering
-  \includegraphics[scale=0.45]{quiver-lam20-c2-k5-l10}
+  \includegraphics[scale=0.4]{quiver-lam20-c105-k8-l10}
   \caption{}
@@
 \end{subfigure}
-\caption{Evolutionary competition between pairs of types with (a) $\lambda = 1, c = 2, l = 1, k = 5$ and (b) $\lambda = 20, c = 2, l =10, k =5$.
+\caption{Evolutionary competition between pairs of types with (a) $\lambda = 1, c = 1.05, l = 1, k = 8$ and (b) $\lambda = 20, c = 1.05, l =10, k =8$.
 Coordinates $(x,y)$ show (proportion of pragmatic types, proportion of $\Llack$ types).}
-\label{fig:quiver}
+\label{fig:quivera}
 \end{figure}
 
-Lastly, I wanted to try my hand at compressing the information more while trying to highlight the main take away of the dynamics applied to these reduced type spaces.
-Figure \ref{fig:quiver} is an illustration.
-I used the sketch Michael sent us with his comments as inspiration.
-We think of the type space as being $2$ dimensional.
-Along the four edges we have same types competing as we had in the previous plots.
-This $2D$ space is ambiguous once you move from the edges, so I only applied the dynamics on edges.
-More precisely, in each plot, the $x$-coordinate indicates the proportion of literal types and the $y$-coordinate indicates the proportion of $\Llack$ users.
-Points that are not on edges are ambiguous because, e.g., $(.5,.5)$ could mean $50\%$ are literal $\Lbound$ and $50\%$ are pragmatic $\Llack$ but also that there are $25\%$ of each of the four types in the population.
-The edges are unambiguous because each population consists of maximally two types, opposed with respect to a single dimension.
-This explains why replication in Figure \ref{fig:quiver} only moves along edges.
+\begin{figure}
+\centering
+\begin{subfigure}{.6\textwidth}
+  \centering
+  \includegraphics[scale=0.4]{quiver-lam1-c200-k8-l1}
+  \caption{}
+  \label{fig:sub1b}
+\end{subfigure}%
+\begin{subfigure}{.6\textwidth}
+  \centering
+  \includegraphics[scale=0.4]{quiver-lam20-c200-k8-l10}
+  \caption{}
+  \label{fig:sub2b}
+\end{subfigure}
+\caption{Evolutionary competition between pairs of types with (a) $\lambda = 1, c = 2, l = 1, k = 8$ and (b) $\lambda = 20, c = 2, l =10, k =8$.
+Coordinates $(x,y)$ show (proportion of pragmatic types, proportion of $\Llack$ types).}
+\label{fig:quiverb}
+\end{figure}
 
-\begin{itemize}
-\item[{\bf The good}] These plots look very similar to what Michael sketched out by hand. The RD plots are also a lot cleaner than the phase portraits I made above.
-\item[{\bf The bad}] The ambiguity of $(x,y)$ away from edges.
-\item[{\bf The ugly}] Its painfully evident that mutation steps and RMD steps are almost identical.
-Additionally, vectors are not scaled.
-Right now, they are automatically resized relative to the other vectors on their edge.
-This makes it easier to show small effects.
-If I scale the vectors then the RD plots look as if they consisted only of points because the mutator-steps are much larger.
-Note also that the movement from pragmatic $\Lbound$ to literal $\Lbound$ under replication only is only apparent when we look at enough points along the edge.
-However, this makes the plots look cluttered everywhere else.
-This is fixable and mostly cosmetic though.
-\end{itemize}
-{\bf My take:} I like these the most. Of course, we could also generate each edge separately and present them in a format similar to that of phase portraits above.
-
-I couldn't come up with anything, but: Is there a space in which we could disentangle the ambiguity of the $2D$ space?
-That is, a shape that is as the regular $3$-simplex is to the regular $2$-simplex?
+Figure \ref{fig:quivera} ($c = 1.05$) and Figure \ref{fig:quiverb} ($c = 2$) illustrate the new filled unit squares.
+The main news are that we got rid of the ambiguity for $(x,y)$ when we move away from the edges (see above).
+Taking stock: the good is that (i) the contrast between RD, M, and RMD is clear, (ii) we can see that lit.
+$\Llack$ is functionally disadvantageous and that its pragmatic counterpart is so as well if $\lambda$ is low, (iii) that a functional advantage {\em and} a learnability advantage will select for pragmatic $\Llack$ but not one of these on its own.
+The bad is that we don't see the slight functional disadvantage that prag.
+$\Lbound$ suffers from when compared to its literal counterpart.
 

### 2017-11-26 code-LOT-extension/co-evo/quiver.py

@@ -114,6 +114,92 @@ def coords_y2(edge,pops):
     return [u,v]
 
 
+#def quiver_filled(lam,bias_para,post_para,k):
+lam = 20
+bias_para = 2
+post_para = 10
+k = 8
+
+lexica = [np.array([[1.,0.],[0.,1.]]),np.array([[1.,0.],[1.,1.]])]
+types = [LiteralPlayer(lam,lex) for lex in lexica] + [GriceanPlayer(1,lam,lex) for lex in lexica]
+lexica_prior = learning_prior(types,bias_para)
+Q = get_q(lexica_prior,post_para,k,types)
+U = get_u(types)
+
+fig,axs = plt.subplots(ncols=2,nrows=2,sharex=True,sharey=True)#,sharey=True,sharex=True)
+
+ax_n = [[0,0],[0,1],[1,0],[1,1]]
+
+x = np.linspace(0,1,20)
+y = np.linspace(0,1,20)
+#(X,Y) = np.meshgrid(x,y)
+#u,v = [], []
+
+for i in xrange(len(ax_n)-1):
+    u,v = np.zeros((len(x),len(y))), np.zeros((len(x),len(y)))
+    for tx in x:
+        for ty in y:
+            p = np.zeros(len(types))
+            p[0] = (1-tx) * (1-ty)
+            p[1] = (1-tx) * ty
+            p[2] = tx * (1-ty)
+            p[3] = tx * ty
+            if i == 0:
+                r = rep(p,U,types)
+                lbl = 'RD'
+            elif i == 1:
+                r = mut(p,Q)
+                lbl = 'MD'
+            else:
+                rmd(p,U,Q,types)
+                lbl = 'RMD'
+    
+            change_along_x =  r[2] + r[3] - tx #prag types
+            change_along_y =  r[1] + r[3] - ty #lack types
+            #u.append(change_along_x)
+            #v.append(change_along_y)
+            txidx = np.argwhere(x == tx)[0][0]
+            tyidx = np.argwhere(y == ty)[0][0]
+            u[tyidx,txidx] = change_along_x
+            v[tyidx,txidx] = change_along_y
+    
+    if i == 0:
[…]

### 2018-01-11 paper/cogsci-submission2.tex
revisions semi-draft

--- 2017-04-14 paper/cogsci-submission1.tex
@@
 \definecolor{Red}{RGB}{178,34,34}
+\definecolor{Steelblue}{RGB}{70,130,180}
 \newcommand{\mf}[1]{\textcolor{Red}{[MF: #1]}} 
 \newcommand{\tb}[1]{\textcolor[rgb]{.8,.33,.0}{[TB: #1]}}% prints in orange
+\newcommand{\rvv}[1]{\textcolor{Steelblue}{#1}}% prints in steel blue
+\definecolor{indianred}{RGB}{205,92,92}
+\definecolor{mygray}{cmyk}{0.35,0.35,0.35,0.35}
+\definecolor{goldish}{RGB}{118, 104, 154}
 
@@
   cultural evolution of language plays out under this picture of the semantics-pragmatics
-  interface, we present a game theoretic model of the competition between types of language
+  interface, we present a game-theoretic model of the competition between types of language
   users, each endowed with a selection of lexical representations and a particular pragmatic
@@
   linguistic knowledge. We illustrate the model based on a case study on scalar
-  implicatures. In this case study learning biases that favor simple semantic representations
-  can foster the evolution of more sophisticated pragmatic reasoning types and so prevent the
-  lexicalization of scalar implicatures.
+  implicatures.
+\rvv{The relationship between underspecified semantics and pragmatic inference that this study suggests is one of co-evolution.
+The maintenance of such semantics depends on pragmatic reasoning for their enrichment.
+In the opposite direction, complex semantics that can do away with mutual reasoning do not necessarily foster pragmatic reasoning and may even be encumbered by it.}
 \end{abstract}
@@
 \section{Introduction}\label{sec:introduction}
+\hl{2.
+This is true.
+We are grateful that this was pointed out, as our previous rendering could lead to confusion, e.g., as to why $L_{\text{all}}$, being the {\em a priori} more likely type to be inferred, did not come out strongly even when considering learnability alone.
+Throughout the article, we now stress more that what matters is the posterior -- and discuss the role of the prior mainly when contrasting targets and competitor, who crucially differ mostly in this respect.}
+
+\hl{3.
+These questions tie in with Reviewer $1$'s request for a clearer exposition and the explanation of why we speak of co-evolution in this case.
+In a nutshell, pressure for learnability alone does not put the types found in a population in competition.
+Instead, the population comes to be inhabited by types that are faithfully transmitted -- of which there might be many (e.g., all variants of a kind are equally learnable).
+What fitness-relative selection adds to this process is the competition that learnability alone lacks, independent of the type space we look at.
+Now, in the case of our type space we have a slight (modulo $\lambda$) communicative disadvantage for targets  {\em when in direct competition with competitors}, for instance.
+However, the fact that they are inferred more faithfully by na\"ive learners leads (i) to the existence of less competitors, counteracting this functional disadvantage, as well as (ii) to the existence of less variants of a kind (contra learning-alone).}
+
 What is conveyed usually goes beyond what is said. A request for a blanket can be politely
-veiled by uttering ``I'm cold;'' temporal succession of events can be communicated by the order
+veiled by uttering ``I'm cold;'' a temporal succession of events can be communicated by the order
 in which conjuncts appear as in ``I traveled to Paris and got married;'' an invitation can be
@@
 as well as experimentally
-\citep[e.g.,][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. While
-there is much dispute in this domain about many details, a position endorsed by a clear
-majority is that a scalar item like \emph{some} is underspecified to mean \emph{some and maybe
-  all} and that the enrichment to \emph{some but not all} is part of some regular process with
+\citep[e.g.,][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}.
+This makes them particularly suitable candidates for the study of the evolution of regular pragmatic inferences.
+While there is much dispute in this domain about many details, a position endorsed by a clear majority is that a scalar item like \emph{some} is underspecified to mean \emph{some and maybe all} and that the enrichment to \emph{some but not all} is part of some regular process with
 roots in pragmatics.
@@
 If this majority view is correct, the question arises how such a division of labor between
-semantics and pragmatics could have evolved, why it would be so pervasive across natural
-languages, and why it is that some expressions systematically draw from it while others semantically conventionalize. Models of language evolution abound. There are simulation-based models studying
+semantics and pragmatics could have evolved; why it would be so pervasive across natural
+languages; and why it is that some expressions systematically draw from it while others semantically conventionalize. Models of language evolution abound. There are simulation-based models studying
 populations of communicating agents
 \citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
-and there are mathematical models of language evolution, mostly coming from game theory
-\citep{lewis:1969,Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
+and there are mathematical models of language evolution, many coming from game theory
+\citep{Warneryd1993:Cheap-Talk-Coor,BlumeKim1993:Evolutionary-St,nowak+krakauer:1999,nowak:2006,Huttegger2007:Evolution-and-t,Skyrms2010:Signals}. Much
 of this work has focused on explaining basic properties such as compositionality and
@@
 representations and pragmatic use. In contrast, we will here look at language users with a
-richer cognitive make-up.
-
-We spell out a model of the co-evolution of conventional meaning and pragmatic reasoning.
+richer linguistic make-up. More precisely, We spell out a model of the co-evolution of conventional meaning and pragmatic reasoning.
 The objects of replication and selection are pairs consisting of a set of lexical
-meanings and a manner of pragmatic behavior. The latter are represented using probabilistic
-models of pragmatic language use
-\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
+meanings and a manner of pragmatic behavior, which together define agents' linguistic choices.
+Pragmatic behaviors are represented using probabilistic models of pragmatic language use \citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}.
+Replication
 and selection are described by the \emph{replicator mutator dynamic}, a general and established
 model of evolutionary change in large and homogeneous populations
-\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. This
+\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. This
 approach allows us to study the interaction between (i) pressure towards communicative
 efficiency and (ii) infidelity in the transmission of linguistic knowledge, caused by factors
-such as inductive learning biases and sparse learning data. Considering transmission of
-linguistic knowledge is important because neither semantic meanings nor pragmatic usage
-patterns are directly observable. Instead, language learners have to infer these unobservables
-from the observable behavior in which they result. We formalize this process as a form of
-Bayesian inference. Our approach thereby contains as a special case a well-understood model of
-iterated Bayesian learning \citep{griffiths+kalish:2007}, but combines it with functional
+such as inductive learning biases and sparse learning data \citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,KirbyGriffith2014:Iterated-Learni}.
+Considering the effects that the transmission of
+linguistic knowledge through (iterated) learning have is important because neither semantic meanings nor pragmatic usage
+patterns are directly observable. Instead, language learners have to infer these unobservables -- lexical meanings and manners of pragmatic behavior --
+from the observable behavior in which they result -- overt linguistic choice in a context of interaction. We formalize this process as a form of
+Bayesian inference. Our approach thereby contains a well-understood model of
+iterated Bayesian learning \citep{griffiths+kalish:2005,griffiths+kalish:2007}, but combines it with functional
 selection, here formalized as the most versatile dynamic from evolutionary game theory, the
@@
 show that inductive learning biases of Bayesian learners that favor simpler lexical meanings
-can prevent the lexicalization of scalar inferences and lead to the emergence of Gricean-like
-pragmatic reasoning types. Results of this case study are critically assessed in the light of
+may play a role in preventing the lexicalization of scalar inferences and thereby lead to the emergence of Gricean-like
+pragmatic reasoning types. The results of this case study are critically assessed in the light of
 the assumptions that feed our model in Section~\ref{sec:discussion}.
@@
 
-\subsection{Communicative efficiency and learnability}
+\subsection{Communicative success and learnability}
 
@@
 language. Learnability is another, as natural languages need to be learnable to survive their
-faithful transmission across generations. Furthermore, even small learning biases implicit in acquisition
-can build up and have quite striking effects on an evolving language in a process of iterated
+faithful transmission across generations.
+Furthermore, the effects of even small learning biases implicit in acquisition can build up and have quite striking impact on an evolving language in a process of iterated
 learning
@@
 While natural languages are pressured for both communicative efficiency and learnability, these forces may
-pull in opposite directions. Their opposition becomes particularly clear when considering the
-extreme (cf. \citealt{kemp+regier:2012,kirby+etal:2015}). A language consisting of a single
+pull in opposite directions \citep[\S7]{christiansen+chater:2008}. Their opposition becomes particularly clear when considering the
+extreme (\citealt{kemp+regier:2012,kirby+etal:2015}). A language consisting of a single
 form-meaning association is easy to learn but lacking in expressivity. Conversely, a language
@@
 \subsection{The replicator mutator dynamic}
-
 An elegant formal approach to capture the interaction between communicative efficiency and learnability is
 the \emph{replicator mutator dynamic}
-\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,Nowak2006:Evolutionary-Dy}. In
-its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$
-in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
+\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}.
+\rvv{The RMD is a population dynamic, which is to say that it describes change in the composition of a community made up of different types of agents.
+The population vector $\vec{x}$ codifies proportions or frequencies of types in a population, with $x_i$ being the frequency of type $i$.
+This population is infinite; its carrying capacity is held constant, $\sum\nolimits_i x_i = 1$.
+This allows us to track change that does not depend on varying population sizes nor their growth rates (but see, e.g., \citealt{nowak:2006} and \citealt[\S 5]{Skyrms2010:Signals} for alternatives that consider, among others, population size and its growth).} In its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$ in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
 $i$ before the update step, (ii) the fitness $f_i$ of each type $i$ before the update, and
-(iii) the probability $Q_{ji}$ that an agent who wants to imitate, adopt, or learn the type of
+(iii) the probability $Q_{ji}$ that an agent who observes the behavior of 
 an agent with type $j$ ends up acquiring type $i$:
@@
   \label{eq:RMD_discrete}
-  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}\,.
+  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_h x_h f_h}\,.
 \end{align}
-The RMD consists of two components: fitness-based selection and transmission biases. This
+The RMD consists of two components: fitness-based selection and transmission biases encoded in the so-called mutation matrix $Q$. This
 becomes most transparent when we consider an equivalent formulation in terms of a step-wise
@@
 \end{align*}
+%
+
 If the transmission matrix $Q$ is trivial in the sense that $Q_{ji}=1$ whenever $j=i$, the
-dynamic reduces to the replicator dynamic. The replicator dynamic is a model of fitness-based
-selection in which the relative frequency of type $i$ will increase with a gradient
-proportional to its average fitness in the population. This dynamic is popular and
-versatile because it can be derived from many abstract processes of biological and cultural
+dynamic reduces to the replicator dynamic.
+This situation is one where types are always replicated faithfully.
+The replicator dynamic is a model of fitness-based selection in which the relative frequency of type $i$ will increase with a gradient proportional to its average fitness in the population.
+In other words, selection comes into play whenever two or more types replicate at different rates due to differences in their fitness.
+This dynamic is popular and versatile because it can be derived from many abstract processes of biological and cultural
 transmission and selection \citep[for overview and several derivations
@@
 $f_i$ is the same for all types $i$, the replicator step is the identity map
-$ \left ( \text{RD}(\vec{x}) \right )_i = x_i$ and the dynamic reduces to a process of
-iteration of the transmission bias encoded in $Q$. In this way, the process in
-(\ref{eq:RMD_discrete}), equivalently (\ref{eq:RMD_discrete_recast}), contains a model of
-iterated learning \citep{griffiths+kalish:2007}.
+$ \left ( \text{RD}(\vec{x}) \right )_i = x_i$.
+In words, no difference in fitness translates into a lack of functional pressure: the population's type composition stays the same after the replicator step.
+In such cases the dynamic reduces to a process of iteration of the transmission bias encoded in $Q$: the rate by which one type changes into another after a generational turnover.
+As detailed below, in this way, the process in (\ref{eq:RMD_discrete}), and its equivalent in (\ref{eq:RMD_discrete_recast}), can contains a model of iterated learning \citep{griffiths+kalish:2005,griffiths+kalish:2007}.
 
@@
   Consider a simple and abstract coordination game. Agents are of two types: positive or negative. If
-  agents of different type play with each other, they obtain a payoff of 0. If negative meets
-  negative, each receives a payoff of 1. If positive meets positive, they get a payoff of 2. A
-  population state is completely characterized by the proportion $x$ of negatives. The fitness
+  agents of different type interact with each other, they obtain a payoff of $0$. If negative meets
+  negative, each receives a payoff of $1$. If positive meets positive, they get a payoff of $2$.  A population state is completely characterized by the proportion $x$ of negatives. The fitness
   of negatives in population state $x$ is $f_n(x) = x$, that of positives is $f_p(x) = 2-2x$.
   The average fitness is $\Phi(x) = x f_n(x) + (1-x) f_p(x) = 3x^2 - 4x + 2$. Without mutation,
-  the replicator dynamic will then update $x$ to $RD(x) = \nicefrac{f(x)x}{\Phi(x)} = \nicefrac{x^2}{\Phi(x)}$. The update
-  function $RD(x)$ of the replicator step is plotted in Fig.~\ref{fig:Updates_RMS} as the
-  blue line. Rest points, for which $RD(x)=x$, are at $x=0$, $x=1$ and $x= \nicefrac{2}{3}$.
+  the replicator dynamic will update $x$ to $\text{RD}(x) = \nicefrac{f(x)x}{\Phi(x)} = \nicefrac{x^2}{\Phi(x)}$. The update
+  function $\text{RD}(x)$ of the replicator step is plotted in Fig.~\ref{fig:Updates_RMS} as the
+  blue line. Rest points, for which $\text{RD}(x)=x$, are at $x=0$, $x=1$ and $x= \nicefrac{2}{3}$.
   The former are attractors as nearby points converge to them. Points near $x=\nicefrac{2}{3}$
-  move towards 0 or 1. This is schematically pictured in the topmost phase portrait in
-  Fig.~\ref{fig:Phase_RD}.
-
-  Adding mutation changes the dynamic and its rest points. Let us assume that $Q_{ji} = .9$
-  when $j=i$. The update effect of mutation alone is $M(x) = .9 x + .1 (1-x) = .8x + .1$ and is
-  plotted as the linear green line in Fig.~\ref{fig:Updates_RMS}. It has only one stable rest
-  point at $x = 0.5$ (see Fig.~\ref{fig:Phase_RD}). If we first take the replicator step and
+  also move towards 0 or 1. This is schematically pictured in the topmost phase portrait in
+  Fig.~\ref{fig:Phase_RD}. Adding mutation changes the dynamic and its rest points. Let us assume that $Q_{ji} = .9$
+  when $j=i$.
+This is the proportion of types that are replicated faithfully.
+Conversely, a proportion of $.1$ will change their type from positive to negative, and vice-versa.
+The update effect of mutation on its own are described by $\text{M}(x) = .9 x + .1 (1-x) = .8x + .1$, plotted as the linear green line in Fig.~\ref{fig:Updates_RMS}.
+As shown in Figure \ref{fig:Phase_RD}, in this example mutation alone has only one stable rest point.
+It is located at $x = .5$.
+If we first take the replicator step and
   then the mutation step in sequence, we obtain the replicator mutator dynamic
-  $RMD(x) = M(RD(x)) = \nicefrac{.9x^2 - .2x +.2}{3x^2 - 4x + 2}$, which is plotted in red in
+  $\text{RMD}(x) = \text{M}(\text{RD}(x)) = \nicefrac{.9x^2 - .2x +.2}{3x^2 - 4x + 2}$, which is plotted in red in
   Fig.~\ref{fig:Updates_RMS}. The rest points are at $x=.121$, $x=.903$ and $x=.609$. The
@@
 of lexical representations and pragmatic behavior. To do so, we need to fix three things: (i) what the
-relevant types are, (ii) how fitness derives from communicative success and (iii) how the
+relevant types are that populate $x$, (ii) how fitness derives from communicative success and (iii) how the
 mutation matrix $Q$ is computed. These issues are addressed, one by one, in the following.
@@
 \label{sec:languages+use}
-
 Types are what evolution operates on. They define an agent's fitness, usually through a payoff
@@
 acts in a game, e.g., either cooperating or defecting in a prisoner's dilemma. In other cases,
-they may be thought of as general properties of an agent that influences her fitness, such as
+they may be thought of as general properties of an agent that influence her fitness, such as
 being positive or negative in our previous example (whatever that means). For our present
-purposes, types are identified more concretely by specific assumptions about their cognitive
-make-up. Since we are interested in the evolutionary competition between different lexical
-representations and ways of using them in communication, a type is here defined as a pair consisting
-of a lexicon and a pragmatic strategy of language use.
+purposes, types are identified by specific assumptions about their cognitive
+make-up.
+More concretely, since we are interested in the evolutionary competition between different lexical representations and ways of using them in communication, a type is here defined as a pair consisting of a lexicon and a pragmatic strategy of language use.
 
@@
 rule (a probabilistic choice of message for each state) and a probabilistic hearer rule (a
-probabilistic choice of state for each message) given a lexicon. There are many ways of making these general
-notions more concrete. Here is what we will assume in the remainder of this paper.
+probabilistic choice of state for each message) given a lexicon.
+There are many ways of making these general notions more concrete.
+Broadly speaking, there are models of pragmatic language use coming from a game-theoretic tradition \citep[e.g.,][]{benz:2006,benz+vRooij:2007,jaeger:2007b,franke+jaeger:2014} and ones from a Bayesian tradition \citep[e.g.,][]{frank+goodman:2012,goodman+stuhlmueller:2013,GoodmanFrank2016:Pragmatic-Langu}.
+We will not focus on the particular design choices that make up each approach \citep{qing+franke:2015}.
+Instead, we will stress how the approach, common to all of these models, of pragmatic language use as {\em rational language use} can be embedded within the replicator mutator dynamic to answer evolutionary questions at the semantics-pragmatics interface.
+Here is what we will assume in the remainder of this paper.\footnote{Readers more familiar with the Bayesian tradition should note that the definitions below follow the game-theoretic tradition in assuming rational sender choice to consist in sending messages that maximize expected utility, rather than message informativity (see \citealt{qing+franke:2015} for detailed discussion).}
 
@@
 
-The following examples illustrate these behaviors. A literal interpreter with lexicon $\Lbound$ assigns $\ssome$
+In words, an agent's linguistic behavior -- what message to send when / which message to interpret how -- is defined by (i) her reasoning level and (ii) her lexicon.
+In what follows, this is what we identify an agent's type with in order to analyze under which conditions combinations of (i) and (ii) emerge through cultural evolution.
+Intuitively, literal behavior (level-$0$ reasoning) results from unreflected language use.
+Such agents produce and comprehend solely based on the truth-conditions of their lexicon.
+In a Gricean spirit, pragmatic behavior (level-$n+1$ reasoning) results from mutual reasoning about rational language use.
+Such agents produce and comprehend by reasoning about their interlocutor's likely behavior interpretation/production patterns.
+This reasoning is based on a conjecture resulting from their own behavior if they were in the role of their interlocutor.
+Higher order behavior is therefore, even if in tendency rational, quite simple.
+Crucially, we do not assume that agents know each other's types when interacting (e.g., what lexicon the interlocutor is using).
+
+The following examples illustrate these behaviors numerically. A literal interpreter with lexicon $\Lbound$ assigns $\ssome$
 a probability of $H_0(\ssome \mid \msome; \Lbound) = 1$ after hearing $\msome$, while a literal
@@
 
-It is particularly important to note that pragmatic agents of level 1 using $\Llack$ associate
-$\msome$ preferentially with $\ssome$, in contrast to their literal counterparts of level
-0. This association is not perfect, and usually less strong than what agents with a lexicalized
-upper bound in $\Lbound$ can achieve, even without pragmatic reasoning. Higher order reasoning
-beyond level $1$ leads to stronger associations of $\msome$ and $\ssome$ also for the
-hearer. Still, the case study presented in Section~\ref{sec:si-case-study} will consider
-speaker and hearer behavior at levels 0 and 1, as the latter are the simplest pragmatic
-reasoning types which show a tendency to communicatively attuned pragmatic enrichment. Using
-only level 1 reasoning types and possibly small $\lambda$ is therefore a conservative choice
-that works against the fitness-based selection of pragmatic language use for a notion of
-fitness defined as communicative success, which is introduced next.
+\rvv{There are two particularly important things to note.
+First, in contrast to their literal counterparts of level $0$, pragmatic agents of level $1$ using $\Llack$ associate $\msome$ preferentially with $\ssome$.
+This association is not perfect, and usually less strong than what agents with a lexicalized upper bound in $\Lbound$ can achieve -- with or without pragmatic reasoning.
+Higher order reasoning beyond level $1$ leads to stronger associations of $\msome$ and $\ssome$ also for the receiver.
+Still, the case study presented in Section~\ref{sec:si-case-study} will consider
+sender and receiver behavior at levels 0 and 1, as the latter are the simplest pragmatic reasoning types which show a tendency to communicatively attuned pragmatic enrichment.
+Using only level 1 reasoning and possibly small $\lambda$ is therefore a conservative choice that works against the fitness-based selection of pragmatic language use for a notion of fitness defined as communicative success, which is introduced next.
+Second, when it comes to competition between types of use of lexicon $\Lbound$, pragmatic reasoning at level $1$ is not advantageous.
+The reason for this is that literal use of $\Lbound$ already endows agents with a behavioral strategy that associates a single state with a single message (in tendency; depending on $\lambda$ for senders).
+For $\Lbound$-receivers of level $1$, reasoning over stochasticity introduced at $S_0$ will generally decrease the association of one state with one message.
+This decrease is only slight if $\lambda$ is high, but nevertheless present.
+That is to say, level-$1$ reasoning does not necessarily confer a functional advantage.
+For some types, such as users of $\Lbound$, literal signaling is preferable.}
 
@@
 In signaling games, the expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
-average success of $i$ when talking or listening to $j$. If agents are speakers half of the
+average success of $i$ when talking or listening to $j$. If, as standardly assumed, agents are speakers half of the
 time, this yields:
@@
 \end{flalign*}
-As usual, $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.
+As usual, we assume that agents are cooperative: $\delta(s,s') = 1$ iff $s = s'$ and $0$ otherwise.
+
+In words, expected utility $\text{EU}(t_i,t_j)$ quantifies how successful communication between agents of types $i$ and $j$ is, with each type being a combination of a lexicon and a reasoning level (Section~\ref{sec:language+use}).
+Fitness $f_i$ indicates how well type $i$ fares in a population given what other types are found in it: it is type $i$'s weighted expected utility.
 
@@
 \label{sec:learnability}
-
 Languages are shaped not only by functionalist forces towards greater communicative success. Another
@@
 In biological evolution, where types are expressed genetically, transmission infidelity comes
-into the picture through infrequent and mostly random genetic mutations. However, an agent's
+into the picture through infrequent and mostly random mutation and genetic drift \citep{kimura:1983}. However, an agent's
 lexicon and pragmatic reasoning behavior is likely not inherited genetically. They need to be
-learned from observation. Concretely, when agents of type $j$ want to adopt or imitate the
-linguistic behavior of type $i$, they observe the overt linguistic behavior of type $i$ and
-need to infer the covert type that most likely produced the observed behavior. Iterated learning
-is a process in which languages are learned repeatedly from the observation of linguistic
-behavior of agents who have themselves acquired the language from observation and inference. In
-the simplest case there is a single teacher and a single learner. After sufficient training the
-learner becomes a teacher and produces behavior that serves as input for a new learner. Due to
-the pressure towards learnability it exerts, iterated learning alone generally leads to simpler
+learned from observation.
+Concretely, when agents learn from type $j$, they observe the overt linguistic behavior of type $j$ and need to infer the covert type that most likely produced the observed behavior.
+As in biology, this transmission process is not perfectly accurate.
+
+Iterated learning is a process in which languages are learned repeatedly from the observation of linguistic behavior of agents who have themselves acquired their behavior from observation and inference.
+In the simplest case there is a single teacher and a single learner in each generation \citep[e.g.,][]{kirby:2001,brighton:2002}.
+After sufficient training the learner becomes a teacher and produces behavior that serves as input for a new learner.
+Due to the pressure towards learnability it exerts, iterated learning alone generally leads to simpler
 and more regular languages (see \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016} for
@@
 
-Following \citet{griffiths+kalish:2007} we model language acquisition as a process of Bayesian
+Following \citet{griffiths+kalish:2005,griffiths+kalish:2007} we model language acquisition as a process of Bayesian
 inference in which learners combine the likelihood of a type producing the witnessed learning
@@
 $P \in \Delta(T)$, which reflects the amount of data a learner requires to faithfully acquire
-the language of the teacher \citep[cf.][450]{griffiths+kalish:2007}. The extent of the prior's
+the language of the teacher \citep[cf.][450]{griffiths+kalish:2007}.
+Put differently, types favored by the prior are easier to learn because they require less data than less favored types to be inferred, even if they are both compatible with the learning data at hand.
+
+The extent of the prior's
 influence has been shown to heavily depend on the learning strategy assumed to underlie the
@@
 characterization showed that iterated learning converges to the prior in the limit, i.e., that
-the resulting distribution over languages corresponds to the learners' prior distribution and
+the resulting distribution over types corresponds to the learners' prior distribution and
 is not influenced by the amount of input given to them. This difference in predictions can be
@@
 grows as $l$ increases.
+
+The way the posterior over types is formed and how learners employ it thereafter are important design choices.
+There are other alternatives as well.
+For instance, instead of adopting a single type, learners could keep her uncertainty over types and base her production behavior on the formed posterior \citep{burkett+griffiths:2010}.
+A straightforward operationalization of this idea would then see production defined as a two-step process wherein a speaker first samples from her posterior over types and then produces an utterance based on the type she sampled, as done in \citep{kirby+etal:2015}.
+Beyond added complexity to the model, a disadvantage of this approach is that it could a speaker that first used a Hungarian utterance to speak Quechua in the next, so to speak.
+We stick to the simpler assumption of a single type being acquired, which is defined as follows.
 
@@
 \end{flalign*}
-The acquisition probability $F(t_i \mid
-d)$ given datum $d$ is obtained by probability matching $l = 1$ or a tendency towards choosing
-the most likely type $l > 1$ from the posterior distribution $P(\cdot \mid d)$ over types given
-the data, which is calculated by Bayes' rule:
+
+$Q$ is a stochastic $n \times n$ matrix, with $n$ equal to the amount of types.
+The acquisition probability $F(t_i \mid  d)$ given datum $d$ is obtained by probability matching $l = 1$ or a tendency towards choosing the most likely type $l > 1$ from the posterior distribution $P(\cdot \mid d)$ over types given the data, which is calculated by Bayes' rule:
 \begin{flalign*}
@@
 \end{flalign*}
+
+\subsection{Example}
+Section~\ref{sec:si-case-study} shows how these components and their interaction can shed light on the evolution of scalar implicatures.
+Before moving on, we will illustrate the model's workings in a constrained type space in order to showcase the effect that pressure for communicative success and learnability have on types.
+To this end, we focus only on the four types from Section~\ref{sec:languages+use}.
+These types are made up of a combination of either lexicon $\Llack$ or $\Lbound$ and either literal (level-$0$) or pragmatic (level-$1$) behavior.
+
+\begin{figure}
+\centering
+    \begin{tikzpicture}
+
+    \draw[-,dashed,gray] (0,0) -- (5,0) -- (5,5) -- (0,5) -- (0,0);
+
+    \draw (0,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Lbound$}}] {};
+    \draw (5,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Lbound$}}] {};
+    \draw (5,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Llack$}}] {};
+    \draw (0,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Llack$}}] {};
+
+    \draw[<-, thick, indianred] (0.75,0.25) -- (4.25,0.25) node[midway, above] {\footnotesize functional};
+    \draw[->, thick, indianred] (0.75,4.75) -- (4.25,4.75) node[midway, below] {\footnotesize functional};
+    \draw[->, thick, indianred] (0.25,4.25) -- (0.25,0.75) node[midway, below, rotate=90] {\footnotesize functional};
+    \draw[->, thick, indianred] (4.75,4.25) -- (4.75,0.75) node[midway, above, rotate=90] {\footnotesize functional};
+
+  \path[->, thick, densely dashdotted, Steelblue] (0.75,0.75) edge [bend left] node[sloped,below] {\footnotesize learning\textsubscript{flat}} (2.25,2.45);
+  \path[->, thick, densely dashdotted, Steelblue] (4.25,0.75) edge [bend right] node[sloped,below] {\footnotesize leaning\textsubscript{flat}} (2.75,2.45);
+  \path[->, thick, densely dashdotted, Steelblue] (0.75,4.25) edge [bend right] node[sloped,above] {\footnotesize leaning\textsubscript{flat}} (2.25,2.75);
+  \path[->, thick, densely dashdotted, Steelblue] (4.25,4.25) edge [bend left] node[sloped,above] {\footnotesize leaning\textsubscript{flat}} (2.75,2.75);
+
+    \draw[->, thick, goldish] (0.75,-0.25) -- (2.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
+    \draw[<-, thick, goldish] (2.75,-0.25) -- (4.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
+    \draw[->, thick, goldish] (-0.25,0.75) -- (-0.25,4.25) node[midway, above,rotate=90] {\footnotesize learning\textsubscript{biased}};
+    \draw[->, thick, goldish] (5.25,0.75) -- (5.25,4.25) node[midway, below,rotate=90] {\footnotesize learning\textsubscript{bias}};
+    \draw[->, thick, goldish] (0.75,5.25) -- (4.25,5.25) node[midway, above] {\footnotesize learning\textsubscript{biased}};
+
+    \end{tikzpicture}
+\caption{Sketch of dynamics on a two-dimensional type space with four types.
+The $x$-axis represents the proportion of types with $\Llack$; the $y$-axis that of pragmatic ones.
+Arrows sketch the directionality of population trajectories through the space under functional pressure (replicator steps) and learnability pressure (mutator steps), both with a flat prior and one that favors $\Llack$.}
+\label{fig:sketch}
+\end{figure}
+
+A population will be made up of frequencies of these four types.
+We may think about such a population as a point in a two-dimensional space.
+In Figure \ref{fig:sketch} this space is outlined by the grey dotted square.
+The first dimension determines the proportion of types that lexicalize ambiguous $\msome$ (the $x$-axis in Figure \ref{fig:sketch}).
+That is, the proportion of types with $\Llack$.
+The second dimension determines the proportion of pragmatic language users (the $y$-axis in Figure \ref{fig:sketch}).
+Populations at the either corner of this space are monomorphic, consisting purely of a single type.
+If we move from away from edges toward the center of the space, then this two-dimensional representation underspecifies the proportion of types found in a population.
+The space's center, for instance, could be composed 50\% of pragmatic $\Llack$ and of an equal proportion of literal $\Lbound$; but it could equally well be a population that has each type with the same frequency.
+This will be sufficient for our purposes, as we are only interested in showcasing the rough trajectories of populations through this space after replicator or mutator steps.
+%The space's corners (and edges) are unequivocal: the lower-left corner corresponds to an population of only literal $\Lbound$ users; the lower-right corner to all pragmatic $\Lbound$ users; the upper-left to literal $\Llack$ users only; and the upper-right corner to the majority all pragmatic $\Llack$ users.
+
+From Section~\ref{sec:languages+use}, we already know some things about this type space.
+When it comes to communicative success, types with $\Lbound$ have a functional advantage over $\Llack$; literal $\Lbound$ has a slight advantage over pragmatic $\Lbound$; all have a sizable advantage over literal $\Llack$.
+The red arrows in Figure \ref{fig:sketch} sketch out these relationships.
+They depict the rough trajectories we may expect if there is only pressure for communicative success.
+How much each replicator step moves a population ultimately depends on the rationality parameter $\lambda$.
+In particular, functional differences decrease as $\lambda$ increases: if $\lambda$ is sufficiently high, all types except for literal $\Llack$ speakers will show a strong tendency to associate $\ssome$ with $\msome$ and $\sall$ with $\mall$.
+As a consequence, when it comes to learnability and transmission fidelity, the likelihood that any of the three remaining types produces a given datum $d$ will be approximately equal.
+In other words, the likelihood offers little differentiation for learners if $\lambda$ is high.
+We can therefore expect that, if the prior is flat, no type is passed on very faithfully.
+This would lead populations to gravitate toward the space's center after mutator steps.
+This is sketched by the blue dash-dotted arrows in Figure~\ref{fig:sketch}.
+Neither prior nor posterior alone, but their combination, are what determines the effects of (iterated) Bayesian learning.
+If, instead of a flat prior, there is a prior that favors, say, $\Llack$ over $\Lbound$ we can expect mutator steps to be skewed toward the upper-half of the space.
+This is suggested by the purple arrows in Figure~\ref{fig:sketch}.
+Taking stock, the effects of iterated learning will depend not only on $\lambda$, affecting the likelihood, but also on the prior, how strong the bias is, and the degree to which learners tend to maximize the posterior, controlled by parameter $l$.
+
+Figure \ref{fig:quiver} shows actual trajectories, showcasing how they are modulated by parameter values.
+The replicator steps make the functional disadvantage of $\Llack$ against $\Lbound$ apparent, particularly for literal types and/or low $\lambda$.
+They also highlight that the functional difference between literal and pragmatic $\Lbound$ is small.
+Under functional pressure only, populations move toward the lower part of the square but the steps become small as the edge, representing populations of only $\Llack$, are reached.
+As for the mutator diagrams, there are two main things to note.
+As expected, a small bias for a type will lead to larger proportions to adopt it provided that likelihood offers little to no differentiation between types.
+This is the situation the right-most space in Figure \ref{fig:quiver} depicts.
+However, the bias is far from absolute.
+If, instead, certain types are more likely to be recovered based on the linguistic input they produce, they will be passed on more faithfully.
+This is why, for instance, populations move toward a majority of $\Llack$-types for $\lambda = 5$ and $l = 10$.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig-contourb105k5.png}
+\caption{Dynamics on a two-dimensional type space with four types.
+The $x$-axis represents the proportion of types with $\Llack$; the $y$-axis that of pragmatic ones.
+Arrows indicate directionality of trajectories after (a) replicator steps and (b) mutator steps with a prior that favors $\Llack$ by a factor of $1.05$.
+Colored contours show the proportion of pragmatic $\Llack$ users after a step \tb{I don't think the contours are adding too much, so they could be removed.
+Then again, I think they are rather non-intrusive and look nice.}}
+
+\label{fig:quiver}
+\end{figure}
+
+Figures \ref{fig:sketch} and \ref{fig:quiver} depict replicator and mutator steps in isolation to better showcase their effects.
+Their combination will expectedly yield different outcomes, which we now turn to in a case study on scalar implicatures.
 
@@
 \begin{align*}
-  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}\,.
+  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_h x_h f_h}\,.
 \end{align*}
@@
 inference behavior lies on a spectrum from probability matching to acquisition of the most likely
-parent type.
+teacher type.
 
@@
 states to different overt messages. More informally, they speak different languages that
-lexicalize the same concepts. Consequently, more often than not different variants of the same
+lexicalize the same concepts. Consequently, more often than not different types of the same
 kind fail to understand each other.
@@
 \label{sec:an-induct-learn}
+
+\hl{The discussion on derivation costs is not very satisfying on my opinion.
+For instance, the condition of "simple representations being favored over more complex ones" is compatible with many more cost specifications.}
+
+\hl{We do not believe that, ultimately, we should assume the non-parsimonious view that there are different complexity measures for different structures.
+However, with previous work in iterated learning tradition, we do assume that (i) inductive learning biases can play a role in shaping culturally transmitted knowledge and that, as the reviewer notes, (ii) biases can be of manifold nature; another well-studied example being the mutual exclusivity bias, which could be expected to work against targets and in favor of competitors.}
+
+\hl{At present, it is impossible to disentangle the complex interaction of multiple biases.
+We therefore focus on the effects of a single plausible contributing factor to a preference of target lexica over competitors: a well-motivated bias favoring simplicity \citep{feldman:2000,chater+vitanyi:2003, piantadosi+etal:2012a,kirby+etal:2015,piantadosi+etal:underreview}.
+By assumption, this translates to a preference not to lexicalize linguistic material over lexicalizing it; simpler formulae over complex ones.
+As mentioned in Section 3.1.2, we acknowledge that the complexity measure that the prior is based on  is just a convenient operationalization of this particular hypothesis.}
+
+\hl{Lastly, we should stress that, just as models of iterated learning make different predictions under the assumption of different inductive biases, so does our model.
+The assumption and operationalization of a particular bias should, of course, always be seen critically (Section 4).}
 
@@
 \label{sec:simulation-results}
+\hl{Change plots and captions from prag. Llackers to lit. Llackers}
+
+\hl{The figures (especially Figure 4, but the others as well) are difficult to read, because of the changing scales.
+I would recommend finding a way to share the same scale between subfigures, or make it visually clearer that the subfigures cannot be directly compared.}
+
+\hl{Mention that they have the same $k$, $\lambda$ and $l$: They do not have to but they do in our case study.
+Variation across parameter values would mean that we would have many more types: one for each combination of possible lexicon, pragmatic strategy, $\lambda$-, $l$- and $k$-value.
+This would allow us to trace the evolutionary trajectory (and competition) of not only lexica and pragmatic strategies, which is our goal here, but of all these factors combined.
+As our focus is not to show that, say, a particular $\lambda$ value leads to a particular evolutionary outcome we restrict our attention to a homogeneous population where all agents have the same parameter values.
+We discuss this issue in Section~3.2.
+Maybe take this from thesis.}
+
+\hl{We have thoroughly revised the article to make the contrast between these three predictions clearer, as well as the predictions themselves.
+Concretely, This should be the new figure as well as a clearer contrast when discussing the simulation outcomes.
+THe questions are: Why do learnability pressures not have the same effect in the joint learnability/fitness model, as they do in the lesioned model?
+Why is high speaker optimality necessary for scalar implicatures to evolve, when it does not have this effect in the fitness-only model?}
 
@@
 The plot also indicates the proportion of the
-\emph{majority type}, i.e., the type with the highest proportion in the final population. With
+\emph{majority type}: the type with the highest proportion in the final population. With
 low $\lambda$ many types have very similar behavior, so that evolutionary selection lacks grip
@@
 
-\subsubsection{Iterated learning only}
+\subsubsection{Iterated learning only: transmission fidelity}
 
@@
 independent populations with $\lambda = 20$ all majority types were target types,
-with each reaching approximately the same proportion of users in the population. As
+with each reaching approximately the same proportion of users in the population.
+
+As
 with a pressure only for communicative success, low values of $\lambda$ make the differences in observable behavior
 across types less pronounced and therefore reflect the learners' inductive bias more
-faithfully, favoring functionally deficient but a priori preferred types such as those that use
+faithfully. This favors functionally deficient but a priori preferred types such as those that use
 $\Lall$. A pressure for learnability alone may consequently lead to a spread of communicatively
@@
 of $1000$ independent populations had users of $\Lall$ as majority types.
+For higher $\lambda$, it becomes clear that a high prior (e.g., $\Lall$) is not the only thing that counts for learnability.
+As soon as there is information for learners to discern whether one type is more likely to have generated the data (depending on $\lambda$ and $k$), it becomes paramount for types to produce data that makes them easily identifiable if they are to be transmitted faithfully.
+This is to say that the learning prior is important but far from being the only element that regulates how iterated learning plays out.
 
@@
 Pressure for communicative success and learnability are not sufficient on their own to have a
-single target type dominate the population. When pressured for communicative success, the slight
-communicative advantage of $\Lbound$ users leads to their prevalence. When pressured for
+single target type dominate the population. When pressured for communicative success, the  communicative advantage of $\Lbound$ users leads to their prevalence. When pressured for
 learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more
@@
 target types but does not lead to monomorphic populations if learners sample from the
-posterior. Finally, a combination of high $\lambda$ and $l$ leads to increasing proportions of
+posterior (see the uppermost row in Figure \ref{fig:rmd}). Finally, a combination of high $\lambda$ and $l$ leads to increasing proportions of
 a single majority target type.
@@
 acquisition. Lastly, the learning behavior, captured by parameter $l$, where approximating a
-MAP strategy magnifies the effects of the learning bias in tandem with replication. More broadly,
+MAP strategy magnifies the effects of the learning bias in tandem with replication.
+
+More broadly,
 target types, which represent the majority view of scalar implicatures, can come to
@@
 
-The approach introduced here combines game theoretic models of functional pressure towards
+\hl{As now made explicit in \hl{Section~xx.xx}, we chose three messages for illustrative purposes mainly.
+In particular, we wanted to inspect a type space where targets are not the most likely {\em a priori} as this may mislead readers into thinking that learnability alone, in particular the prior alone, drives the outcome.
+This is related to our answer to reviewer 2, comment 7.
+The reason why we did not go beyond $n>2$ is for computational tractability, as calculating $Q$, even if approximated, is expensive.
+The question whether increasing the type space will decrease the frequency to which we expect to see pragmatic {\em some} users is interesting.
+It now figures in Section~X.X.
+At present, we have no answer.
+}
+
+The approach introduced here combines game-theoretic models of functional pressure towards
 successful communication \citep{nowak+krakauer:1999}, effects of transmission perturbations on
@@
 conventional semantics and pragmatic use.
+
+\hl{Redo around here. See commented out}
+%With respect to (i), in Section \ref{4sec:success-vs-expressivity} we discussed a comparable model proposed by \citet{kirby+etal:2015} and highlighted the difference between a type's expressivity and functional pressure.
+As showcased in Section \ref{4sec:si-case-study}, the latter pressure can indirectly select for expressive types, i.e., those that can convey states unequivocally.
+By contrast, Kirby et al.'s model only considers the bearing that expressivity has on the production of learnable data.
+We see three main reasons for considering utility rather than just expressivity; and, relatedly, to consider the effects that communication {\em and} learning have on an evolving linguistic system, rather than only learning.
+First, learning alone can  promote populations with non-negligible proportions of functionally defective types.
+This is true both of simulations, e.g.
+our $\Lall$-types in Figure \ref{fig:only-M}, as well as of laboratory experiments with human subjects (see \S\ref{4sec:success-vs-expressivity} and references within).
+Second and more importantly, types may be equally expressive but their performance as means of information transfer crucially depend not only on themselves but on the population they find themselves in.
+That is, we contend that adopting an expressive type that generates learnable data does not in itself capture a type's arguably central communicative function of transferring information to peers.
+Taking communication into consideration allows the model to be responsive to the task for which language is learned.
+Lastly, chains of iterated learning alone do not put types in direct competition.
+Accordingly, learning can lead to polymorphous populations in which multiple variants of a type coexist (compare the competition of target types in Figure \ref{fig:rmd} and their lack of competition in Figure \ref{fig:only-M}).
+This latter point of course depends on the particulars of a model's learning component.
+Here, we explored the effects that simple transmission chains have, but it is possible to add further complexity to this process.
+For example, by letting learners learn from multiple teachers \citep{burkett+griffiths:2010,brochhagen+etal:2016}.
 
@@
 
-Finally, our case study should not be interpreted as a proposal for a definite explanation of
+Finally, this case study should not be interpreted as a proposal for a definite explanation of
 how scalar implicatures evolved. Other factors should be considered eventually even if they
 will lead to much more complex modeling. One such factor is the observation that
-non-lexicalized upper bounds allow a broader range of applicability, e.g., when the speaker is
+non-lexicalized upper bounds allow a broader range of applicability. For example, when the speaker is
 not certain as to whether \emph{all} is true. This may suggest an alternative and purely
@@
 conveyed pragmatically, then this is preferred over expressing it overtly through longer
-expressions, e.g., by saying {\em some but not all} explicitly. Importantly, although
+expressions. For example, by saying {\em some but not all} explicitly. Importantly, although
 morphosyntactic disambiguation may be dispreferred due to its relative length and complexity
 \citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and override
-contextual cues that might otherwise mislead the hearer. In a nutshell, this explanation posits
-that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to
-communicate as economically as possible and pragmatic reasoning enables them to do so. What
-this alternative argument does not explain is why functional pressure does not lead to the
-emergence of different, equally costly lexical items to express different knowledge states of
-the speaker (cf. \citealt[252-267]{horn:1984},
-\citealt{horn:1972,traugott:2004,vdAuwera:2010}). For instance, to the emergence of two expressions for each weak
-scalar expression; one with and one lacking an upper-bound.  Looking at pressure from
-learnability might come in again. Beyond scalar implicatures, the model can generate predictions about
-likely lexicalization trajectories of pragmatic inferences, or a lack thereof. In this realm
-an interesting issue is whether proposed principles, such as the semantic conventionalization of once highly context-dependent inferences if they become regular enough \citep{levinson:2000,traugott:2004}, can be given
-a formal rationale and inform postulated directionalities of change. The present work made a first start and gave a framework for
-exploring exactly these issues systematically.
+contextual cues that might otherwise mislead the hearer.
+In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so (cf.
+Chapter \ref{chapter:ambiguity}).
+What this alternative argument does not explain is why functional pressure does not lead to the emergence of different, equally costly lexical items to express different knowledge states of the speaker (\citealt[252-267]{horn:1984},
+\citealt{horn:1972,traugott:2004,vdAuwera:2010}).
+For instance, to the emergence of two expressions for each weak scalar expression; one with and one lacking an upper-bound.
+That is, it does not explain why English and other languages do not have a monomorphemic dual expression for, e.g., {\em some} that lexicalizes an upper-bound.
+If this hypothetical expression existed, it could be deployed to signal that the speaker knows that {\em some but not all} holds, and unbounded {\em some} could exclusively signal epistemic uncertainty.
+Looking at pressure from learnability might come in again.
+
+Beyond scalar implicatures, the model can generate predictions about likely lexicalization trajectories of pragmatic inferences, or a lack thereof.
+In this realm an interesting issue is whether proposed principles, such as the semantic conventionalization of once highly context-dependent inferences if they become regular enough \citep{levinson:2000,traugott:2004}, can be given
+a formal rationale and inform postulated directionalities of change. The present chapter made a first start and gave a framework for exploring these issues systematically.
 

### 2018-01-11 paper/rv-letter1.tex
revisions semi-draft

\documentclass[12pt,a4paper]{article}
\usepackage{tikz}
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    linkcolor={red!50!black},
    citecolor={blue!50!black},
    urlcolor={blue!80!black}
}
\usepackage{subcaption}
\usepackage{multicol}
\usepackage{nicefrac}
\usepackage{multirow}
\usepackage{fixltx2e}
\usepackage[round]{natbib}
\bibpunct[:]{(}{)}{,}{a}{}{;}

	\newcommand{\tuple}[1]{\ensuremath{\left\langle #1 \right\rangle}} 
	\newcommand{\citeposs}[2][]{\citeauthor{#2}'s (\citeyear[#1]{#2})}
	\newcommand{\citeposss}[2][]{\citeauthor{#2}' (\citeyear[#1]{#2})}
	\newcommand{\hl}[1]{\textcolor[rgb]{.8,.33,.0}{#1}}% prints in orange

\usepackage{mathcomp}
\usepackage{amssymb}
\usepackage{mathrsfs}
\usepackage[fleqn]{amsmath}

\usepackage{mdframed}

\begin{document}

\newcounter{reviewerCounter}
\newcounter{reviewerCommentCounter}[reviewerCounter]
	\date{}

\title{\Large Co-evolution of lexical meaning \& pragmatic use\\ {\bf Revision Cover letter}}

	\maketitle
\normalsize

We would like to thank the reviewers for their helpful comments and suggestions. They are very welcome and have certainly helped improve the article.

In general terms, we have taken care to make the exposition of the model clearer, to clarify the  motivations underlying the modeling choices we make in the case study, and to clarify how the components of the model lead to the results they do. In what follows we address the reviewers' content-related comments and questions one by one, with pointers to respective changes effected on the manuscript.

\vspace{1cm}

%

\noindent\rule{\textwidth}{1pt}
\stepcounter{reviewerCounter} %increase counter
\stepcounter{reviewerCommentCounter} %increase counter

\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25,frametitle= Reviewer \thereviewerCounter~comment \thereviewerCommentCounter \hfill ~~({\it co-evolution})]
the co-evolutionary dynamic mentioned in the title could be better justified [...] In co-evolutionary dynamics, the evolution of one factor (here, the lexicon) should have an impact on the evolution of another (here, pragmatic strategy), and vice versa; the paper does not show evolutionary dynamics (change over time) for either of these factors, not does it show how evolution of a single factor favours the evolution of the other. The model looks more like adaptation of a kind of linguistic type (a lexicon + a pragmatic strategy) to selection pressures for communication and for learnability than co-evolution.  Please justify your characterization of the process modeled as co-evolution or change to adaptation.

\end{mdframed}

\hl{DL. TO DO AFTER FIGURE RMD} This is an important point that we now stress and highlight more throughout the paper. Particularly in \hl{here, here, and here}. The discussion surrounding Figure \hl{xxx} further illustrates how one factor has an impact on the evolution of another.

In a nutshell: a pragmatic strategy (level-$1$ reasoning) favors the evolution of underspecified semantics of the target kind ($L_{\text{lack}}$) as it allows for the maintenance of simpler lexical representations that are easier to learn without functional disadvantages otherwise incurred in communication (e.g., by level-$0$ reasoners). In the other direction, underspecified semantics of the target kind favor the evolution of a pragmatic disposition to act on them. When paired with other semantics, pragmatic refinement can instead be (close to a) neutral trait (e.g., $L_{\text{bound}}$ paired with high $\lambda$) or it can even lead to  functional disadvantages when compared with level-$0$ reasoning (e.g., $L_{\text{bound}}$ with low $\lambda$). Both cases are depicted in Figure \hl{xx.xx}.

%

\vspace{.75cm}
\noindent\rule{\textwidth}{1pt}
\stepcounter{reviewerCommentCounter} %increase counter

\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25,frametitle= Reviewer \thereviewerCounter~comment \thereviewerCommentCounter \hfill ~~({\it modeling choices})]
the mapping between the model and the example of choice (scalar implicatures) could be better motivated
\end{mdframed}

As we discuss in Section~1, our case study focuses on scalar implicatures because they are well-studied systematic pragmatic inferences that have received much attention, both theoretically and experimentally. This makes them particularly suitable candidates for the study of the evolution of regular pragmatic inferences because there are many findings that can guide the application and evaluation of our model to the study of these inferences. This ties in with our particular choices to model pragmatic language use of scalar expressions. As we now make more explicit \hl{in \S 2.3.1}, these draw from a growing game-theoretic and Bayesian tradition \citep[e.g.,][]{franke:2009,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. More generally, we have added further background and motivation to our modelling choices throughout the article.

%
\vspace{.75cm}
\noindent\rule{\textwidth}{1pt}
\vspace{.1cm}

\stepcounter{reviewerCommentCounter} %increase counter
Reviewer \thereviewerCounter~asks for a clearer description of the model. As noted earlier, we have taken care to make the exposition more accessible and to guide the reader more through the technical details of Section 2. \hl{In particular, Section 2.xx is new and should clarify much}. We took the reviewer's questions, briefly answered below, as guidance for the changes made in Section~2.

\vspace{.5cm}
\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25]
How is generation transmission implemented?
\end{mdframed}
As it is unlikely that lexical representations and/or pragmatic reasoning are inherited genetically, we model the transmission of linguistic knowledge across generations as  iterated Bayesian learning. Technical details and motivations are given in \S 2.3.3.

\vspace{.5cm}
\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25]
Do agents in a population die little by little and are replaced by new agents? Or all at once?
\end{mdframed}
\hl{As now made more explicit in \S2.2}, steps of the discrete-time replicator mutator dynamic correspond to generational turn-overs of the entire population (definitions (1) and (2)). That is to say, all agents die at once, being replaced by the following generation in a single swoop.

\vspace{.5cm}
\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25]
Do the models consider just a single teacher and a single learner at a time (I don't think this is the case, since the results refer speak of ``majority type, i.e., the type with the highest proportion in the final population'')?
\end{mdframed}
The parenthetical remark is correct. The amount of teachers of a given type is proportional to the frequency of this type in the population after the replicator step. More generally, the population is infinite so there are infinite learners/teachers. This was implicit in the definitions of the replicator mutator dynamic in (1) and (2) before, and is now explicitly stated in Section~2.2 in the prose that motivates and explicates these definitions.

\vspace{.5cm}
\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25]
 How many agents were there per population? And per generation?
\end{mdframed}
The prose surrounding definitions (1) and (2) in Section 2.2 now makes explicit that the population is infinite. This allows us to track change that does not depend on varying population sizes nor their growth rate. This is why we speak of proportions of agents of a type, rather than of there being a particular number of agents of a type.

\vspace{.5cm}
  \begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25]
Within one run of the model, do all agents in a population have the same lambda, l and k? If so, please this homogeneous population, and discuss the possible consequences of variation in the population in these parameters.
\end{mdframed}
While our model is well-compatible with the assumption that every agent comes with her own parameter values, the reviewer is correct in pointing out that we assume they all to have the same values. Variation across agents would mean many more types: one for each combination of a possible lexicon, pragmatic strategy, $\lambda$-, $l$- and $k$-value. This would allow us to trace the evolutionary trajectory of not only lexica and pragmatic strategies, which is our goal here, but of all these factors combined. As our focus is not to show that, say, a particular $\lambda$ value leads to a particular evolutionary outcome we restrict our attention to a homogeneous population where all agents have the same parameter values. \hl{We discuss this issue in Section~3.2}.

\vspace{.5cm}
\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25]
Does the model assume that learners observe not only the behaviour (the message, e.g. agent A hearing agent B say: "I own some of JC's albums") but also the true state of the world (i.e. agent A also knows whether agent B owns all or some-but-not-all albums)? Or do they get feedback on their communicative success? In other words, how do they learn the correct state-message mappings that form the lexicon? 
\end{mdframed}
Learners witness sequences of data $d$. Such a sequence is composed of message-state pairings. In other words, learners do witness the true state of the world (Section~2.3.3, {\em learnability}).

The probability of the learner receiving learning input $d$ depends only on the likelihood of the teacher producing it. This is determined by the teacher's production probabilities; the speaker behavior of her type. Put differently, communicative success does not come into play, only speaker behavior. It would certainly be possible to let learners witness proficient users communicate and to have them infer types from such interactions instead. However, we wanted to keep our assumptions close to the well-studied assumptions standardly made in the iterated (Bayesian) learning tradition. Iterated learning only considers production, rather than production and comprehension (\hl{see discussion that should make this explicit in the discussion section - maybe draw from thesis}). Details on how learners infer types are given in Section~2.3.3.

\vspace{.5cm}
\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25]
How exactly are pragmatic strategies inherited/inferred by learners?    I think all of this may be in the formulas, but it would be useful if it was expressed in prose too.
\end{mdframed}
As with the other questions, we agree and have made these matters more explicit. This matter is particularly important. We now stress it throughout the article.

In short, a type is a combination of a lexicon and a pragmatic disposition to act on it. Together, both ingredients define an agent's linguistic behavior. A learner faced with input $d$ (see above) calculates $P(\tau \mid d)$, the probability that a type $\tau$ produced input $d$. Crucially, neither pragmatic strategies nor lexical meanings are directly observable, so they can only be faithfully recovered insofar as the behavior evidenced by a type is (in tendency) attributable to only one pairing of a lexicon and a pragmatic strategy.

\vspace{.5cm}
\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25]
Could you express in prose how literal and pragmatic use work in practice? I am not sure I understand why you call the strategies `pragmatic', given that pragmatics depends on the context, and there is no context here. I would like to see the use of the term `pragmatic' justified further in the paper.
\end{mdframed}
We follow the Gricean tradition of viewing pragmatic inference as effected by mutual reasoning about rational language use (Section~1). This may involve the recruitment of contextual information, as the reviewer notes, but can also involve pragmatic enrichments that result purely from reasoning about linguistic alternatives (``why did the speaker say {\em some} instead of {\em all}''). It is in the latter sense that reasoning beyond level $0$ is pragmatic. Classically, theoretical analyses of scalar implicatures put their emphasis on reasoning about linguistic choice rather than on contextual information \citep[e.g.,][]{horn:1972,gazdar:1979, franke:2009, GoodmanStuhlmuller2013:Knowledge-and-I} and we chose to do the same.

\vspace{.5cm}
\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25]
I thought $\lambda$ was the parameter that modeled communication, so it should not be present in the iterated-learning-without-communication models; why is $\lambda$ set to 20 here, and why discuss lower levels of lambda? 
\end{mdframed}
This parameter regulates the linguistic production behavior of types, i.e., speaker behavior. Speaker behavior is relevant both for communication with hearers as well as for the learning input that a teacher produces. When considering only iterated Bayesian learning $\lambda$ accordingly also plays a role because different $\lambda$ values yield different likelihoods of producing data (see above and Section~2.3.3).

\vspace{.5cm}
\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25]
Also [in the Subsection {\em learnability only}], why isn't the proportion of Lall highest, given that it has the lowest complexity (highest prior)?
\end{mdframed}
%
Due to their stochastic speaker behavior, the data produced by $L_{\text{all}}$-teachers tends to be compatible with the behavior of many other types. Intuitively, the data such teachers produce are all over the place and do not do a good job in setting them apart from other types. Consequently, even if the prior favors $L_{\text{all}}$, this type is not transmitted very faithfully; learners reason that it could also come from other types and may adopt those instead. If there are types that are transmitted more faithfully, the population will, over time, transition to these types rather than to $L_{\text{all}}$. As we put it in Section~4, iterated learning does not necessarily promote the {\em a priori} more likely type, but tends to promote a type $t$ based on a gradient of how many other types might likely mutate into $t$, so to speak. We stress this in Section~3.2.2 and Section~4.

\vspace{1cm}
\noindent\rule{\textwidth}{1pt}
\stepcounter{reviewerCommentCounter} %increase counter

\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25,frametitle= Reviewer \thereviewerCounter~comment \thereviewerCommentCounter \hfill ~~({\it LOT \& model relation})]
%
This model rests heavily on the assumption that the representation or meaning of an expression is a logical formula like those in Table 2, and that these formulas yield a complexity hierarchy.  This may be valid for scalar implicature, as you argue, but how realistic is it cognitively for other linguistic structures or for other socially learned items? Should we assume the non-parsimonious view that there are different complexity measures for different linguistic structures? It would be nice to see another linguistic or cultural example that could be fitted with this model.
%
\end{mdframed}

\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25]
Learnability [...] is operationalized as the complexity of the minimal logical formula for each state. Another possible measure of the difficulty/ease of learning would be comparing the regularity/systematicity of the lexicon types given in 3.1.1 (measured e.g. as mutual information (Cornish, Tamariz, Kirby 2009; using the mantel test, Kirby, Tamariz, Cornish, Smith 2015). Yet another approach to complexity would be to assume that the meanings are mutually exclusive categories without internal structure, or to assume a preference or mutual exclusivity. In these cases, I expect L-bound (competitor), to be easier to learn than L-lack (target). Please justify your choice of learnability measure in the face of these alternatives.
\end{mdframed}

We do not believe that, ultimately, we should assume the non-parsimonious view that there are different complexity measures for different structures. However, with previous work in iterated learning tradition, we do assume that (i) inductive learning biases can play a role in shaping culturally transmitted knowledge and that, as the reviewer notes, (ii) biases can be of manifold nature; another well-studied example being the mutual exclusivity bias, which could be expected to work against targets and in favor of competitors.

At present, it is impossible to disentangle the complex interaction of multiple biases. We therefore focus on the effects of a single plausible contributing factor to a preference of target lexica over competitors: a well-motivated bias favoring simplicity \citep{feldman:2000,chater+vitanyi:2003, piantadosi+etal:2012a,kirby+etal:2015,piantadosi+etal:underreview}. By assumption, this translates to a preference not to lexicalize linguistic material over lexicalizing it; simpler formulae over complex ones. As mentioned in Section 3.1.2, we acknowledge that the complexity measure that the prior is based on  is just a convenient operationalization of this particular hypothesis.

Lastly, we should stress that, just as models of iterated learning make different predictions under the assumption of different inductive biases, so does our model. The assumption and operationalization of a particular bias should, of course, always be seen critically (Section 4).

%

\vspace{0.5cm}
\noindent\rule{\textwidth}{1pt}
\stepcounter{reviewerCommentCounter} %increase counter

\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25,frametitle= Reviewer \thereviewerCounter~comment \thereviewerCommentCounter \hfill ~~({\it transmission fidelity})]
%
What is your measure of fidelity/infidelity? Is it just 0 if the teacher and learner are of a different type and 1 if they are of the same? Could you have a more graded measure of fidelity, e.g. taking into account if two types are of the same 'kind' or not? 
\end{mdframed}
%
We understand transmission fidelity as the transition probability of acquiring type $i$ when learning from type $j$ (Section~2.3.3). This is the value of a cell in the learning matrix $Q$. In this case, $Q_{ji}$. As discussed in Section~2.2, if $Q_{ji} = 1$ then $i$ is always acquired when learning from type $j$. On the other extreme, $Q_{ji} = 0$ means that $i$ is never acquired from $j$. $Q$ is a stochastic matrix and we we usually see neither extreme but rather, as suggested by the reviewer, a graded notion of transmission fidelity (Section~2.2). The details on how $Q_{ji}$ is computed are given in Section 2.3.3. We have expanded these sections to make these details clearer in the prose.
%

\vspace{0.5cm}

\noindent\rule{\textwidth}{1pt}
\stepcounter{reviewerCounter} %increase counter
\stepcounter{reviewerCommentCounter} %increase counter

\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25,frametitle= Reviewer \thereviewerCounter~comment \thereviewerCommentCounter \hfill ~~({\it Interpretation of results})]
%
The results are interesting, though somewhat difficult to interpret. First, pragmatic language use does not evolve in the absence of learnability pressures, due to its slightly lower fitness. Second, pragmatic language use can arise from learnability pressures alone, though only for sufficiently ``rational'' learners. Third, pragmatic language use can evolve given both fitness and learnability pressures, though only if both learners and language speakers are both sufficiently optimal. The first two results are fairly straightforward, but the third (which is the most important in the paper) is more puzzling. Why do learnability pressures not have the same effect in the joint learnability/fitness model, as they do in the lesioned model? Why is high speaker optimality necessary for scalar implicatures to evolve, when it does not have this effect in the fitness-only model?
\end{mdframed}

We have thoroughly revised the article to make the contrast between these three predictions clearer, as well as the predictions themselves. Concretely, \hl{say where this is more explicit now. This should be the new figure as well as a clearer contrast when discussing the simulation outcomes}.

These questions tie in with Reviewer $1$'s request for a clearer exposition and the explanation of why we speak of co-evolution in this case (reviewer 1, comment 1). In a nutshell, pressure for learnability alone does not put the types found in a population in competition. Instead, the population comes to be inhabited by types that are faithfully transmitted -- of which there might be many (e.g., all variants of a kind are equally learnable; if there are many variants of a type that is, in tendency, faithfully transmitted the population will stay polymorphic). What fitness-relative selection adds to this process is the competition that learnability alone lacks, independent of the type space we look at. Now, in the case of our type space we have a slight (modulo $\lambda$) communicative disadvantage for targets  when in direct competition with competitors, for instance. However, the fact that targets are inferred more faithfully by na\"ive learners leads (i) to the gradual existence of less competitors, counteracting this, if slight, functional disadvantage, as well as (ii) to the existence of less variants of a kind (contra the {\em learning alone} dynamic).
%

\vspace{0.5cm}

\noindent\rule{\textwidth}{1pt}
\stepcounter{reviewerCommentCounter} %increase counter

\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25,frametitle= Reviewer \thereviewerCounter~comment \thereviewerCommentCounter \hfill ~~({\it Model predictions})]
%
I agree with its claim that the results are non-trivial; it is not clear a priori that there should exist any regimes where learnability and fitness pressures balance in order to produce scalar implicatures. The paper does not, however, address why the pressures do balance in this case. Are there general theorems (from previous work) about the replicator mutator dynamic that could be brought to bear here? How do learnability and fitness pressures trade-off in general? Is there anything more general that can be said about conditions under which pragmatic language use will evolve? Does the current example generalize to other cases of pragmatic language use? The paper does not need to answer these questions in an exhaustive manner, but greater scientific understanding of the model is desired.
\end{mdframed}

As noted above, we have expanded our analysis of the model and put much more emphasis on how the components that feed it play a role in driving particular evolutionary outcomes. See, in particular, \hl{sections this and that}. 
%

\vspace{0.5cm}

\noindent\rule{\textwidth}{1pt}
\stepcounter{reviewerCommentCounter} %increase counter

\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25,frametitle= Reviewer \thereviewerCounter~comment \thereviewerCommentCounter \hfill ~~({\it co-evolution and pragmatic maintenance})]
%
It makes sense for agents to acquire their grammar/lexicon by learning, as these are social conventions. I believe, however, that the paper also assumes that an agent's pragmatic type (i.e. whether they are literal or pragmatic speakers) is also determined by learning. This is more difficult to interpret. It is reasonable for an agent to learn whether the other agents in a population are behaving pragmatically. Even if the other agents are not behaving pragmatically, however, it will be rational for the learner to behave pragmatically. This is a property of the recursive definition of the pragmatics model. The pragmatic speaker is defined so as to select utterances that will be correctly interpreted by the literal listener. If an agent learns that their interlocutors are all literal agents, then it would be rational for them to pragmatically reason about these literal agents. It is therefore not clear why the agents should, as currently proposed, copy the inferred pragmatic type of their interlocutor.
\end{mdframed}

It is not necessarily true that it will always be rational for agents to behave pragmatically.  \hl{We now make explicit in Section~2}, but the main thing to note is that a pragmatic level-$1$ hearer using lexicon $L_{\text{bound}}$ will reason over the behavior of a {\em soft-maximizing} level-$0$ speaker of $L_{\text{bound}}$. The lower $\lambda$, the more the stochasticity percolates to level-$1$ pragmatic interpretation. By contrast, literal level-$0$ hearers of $L_{\text{bound}}$ have a one-to-one form-meaning mapping from the get go. Put differently, pragmatic reasoning can actually encumber some hearers, depending on their lexicon. Particularly if $\lambda$ is low.

%
\vspace{0.5cm}

\noindent\rule{\textwidth}{1pt}
\stepcounter{reviewerCommentCounter} %increase counter

\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25,frametitle= Reviewer \thereviewerCounter~comment \thereviewerCommentCounter \hfill ~~({\it agent simplicity})]
%
The paper assumes that the speaker (and listener) know the type of the player that they're playing against. This does not seem like a natural assumption. Unless the speaker has repeated interactions with the listener (which is not assumed in the paper), the pragmatic type of the other agent is a latent property. One could imagine the speaker calculating expected utility with respect to the distribution on players that they expect (and speakers of different types choosing different strategies against this distribution).

%
\end{mdframed}

As we now clarify in \hl{Section~n} and stress a couple of times throughout the article, we assume very little sophistication from our agents. In particular, they do {\em not} know the type of the player that the are playing against. They simply behave according to their subjective point of view in a boundedly rational fashion. For example, the linguistic choices of a level-$1$ speaker of lexicon $L_{\text{some}}$ do not change depending on whom she interacts with; they are always defined as (boundedly) rational choice relative to the interpretative behavior of a level-$0$ user of $L_{\text{some}}$ (\hl{point to definitions of behavior in section 2}).
%

\vspace{0.5cm}

\noindent\rule{\textwidth}{1pt}
\stepcounter{reviewerCommentCounter} %increase counter

\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25,frametitle= Reviewer \thereviewerCounter~comment \thereviewerCommentCounter \hfill ~~({\it posterior type sampling})]
%
The learner in this model always selects a particular grammar/pragmatic type. An alternative would be for them to maintain uncertainty about the type (which is optimal behavior from a Bayesian perspective). I suspect that this model is intractable (or at least much less manageable) than the proposed one, but it would be worth noting that this is a substantive choice point in the model.
\end{mdframed}
We now highlight that this is a design choice and that there are alternatives in Section~2.3.3.

\vspace{0.5cm}

\noindent\rule{\textwidth}{1pt}
\stepcounter{reviewerCommentCounter} %increase counter

\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25,frametitle= Reviewer \thereviewerCounter~comment \thereviewerCommentCounter \hfill ~~({\it definitions in rational language use})]
%
The current pragmatic model is somewhat non-standard. In particular, neither the literal nor pragmatic speakers in Equations 3-6 use an information-theoretic utility function, as is standard in the literature. One consequence of this is that there is now an asymmetry between the literal speaker and listener: the literal listener never interprets utterances in a manner inconsistent with their literal meanings, while the literal speaker will sometimes use utterances in a non-literal way. I do not think this needs to be changed -- the current utility function is perfectly sensible -- but this is worth noting in the paper. It would also be desirable to know (possibly in future work) whether the modeling results are robust to switching to information-theoretic utility functions.

%
\end{mdframed}
\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25]
%
A related issue is in the fitness definition in Section 2.3.2. If one adopts the alternate utility function above for the agents' behavior, then it would be natural to have fitness scale in a logarithmic manner as well. In this case, the information-theoretic utility function would have the following interpretation: the fitness of a language is determined by how well it allows speakers to communicate their beliefs, rather than how often it leads the listener to make the correct guess about the world. Similar robustness questions apply here as above.

%
\end{mdframed}
There are many alternatives when it comes to definitions of rational language use, many of which do not use an information-theoretic utility function (see \citealt{qing+franke:2015} for an overview and discussion up to $2015$). As noted in our answer to Reviewer 1's comment 2, we have taken care to highlight that our choices draw from this literature and to mention that there are other alternatives in the same spirit.

%

\vspace{0.5cm}

\noindent\rule{\textwidth}{1pt}
\stepcounter{reviewerCommentCounter} %increase counter

\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25,frametitle= Reviewer \thereviewerCounter~comment \thereviewerCommentCounter \hfill ~~({\it Learnability})]
%
Throughout, the paper mostly equates the learnability of a grammar with its complexity/prior probability. In general, however, the relationship may be more complex. Certain grammars may require less data than others to be statistically identified. This is probably not an issue in the current case study, given the simplicity of the grammars and the large amount of data relative to grammar size, but this is likely to be an issue in other cases.

%
\end{mdframed}

This is true. We are grateful that this was pointed out, as our previous rendering could lead to confusion, e.g., as to why $L_{\text{all}}$, being the {\em a priori} more likely type to be inferred, did not come out strongly even when considering learnability alone. \hl{Throughout the article, we now stress more that what matters is the posterior -- and discuss the role of the prior mainly when contrasting targets and competitor, who crucially differ mostly in this respect.}

%

\noindent\rule{\textwidth}{1pt}
\stepcounter{reviewerCounter} %increase counter
\stepcounter{reviewerCommentCounter} %increase counter

\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25,frametitle= Reviewer \thereviewerCounter~comment \thereviewerCommentCounter \hfill ~~({\it expected utility})]
%
what would be the impact of assuming that fitness is not contributed equally by being successful as speaker and being successful as a perceiver?

%
\end{mdframed}

This is a standard assumption in the literature lacking reasons to believe that there is an asymmetry in how often, on average, agents use and hear a type of expression. In the case of scalar implicatures, we see no reason to assume such an asymmetric contribution to fitness.

As for concrete impact: The main contrast that we are interested in, i.e., competition between targets and competitors, does not hinge on agents being speakers half of the time. However, the contrast between pragmatic and literal competitor types does, because their receiver behaviors differ. Assuming that successful receiver behavior contributes differently than sender behavior would therefore modulate differences between these types.

%
\noindent\rule{\textwidth}{1pt}
\stepcounter{reviewerCommentCounter} %increase counter

\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25,frametitle= Reviewer \thereviewerCounter~comment \thereviewerCommentCounter \hfill ~~({\it message amount})]
%
what motivates modelling just three messages and not just two (``some'' and ``all'') or any $n>3$? One obvious prediction: $n=2$ will lead to larger prevalence of lack types, whereas increasing $n$ beyond $3$ will do the opposite. More concretely: is there any non-trivial relative frequency of some/all that could falsify the results obtained in the paper?

%
\end{mdframed}

As now made explicit in \hl{Section~xx.xx}, we chose three messages for illustrative purposes mainly. In particular, we wanted to inspect a type space where targets are not the most likely {\em a priori} as this may mislead readers into thinking that learnability alone, in particular the prior alone, drives the outcome. This is related to our answer to reviewer 2, comment 7. The reason why we did not go beyond $n>2$ is for computational tractability, as calculating $Q$, even if approximated, is expensive. The question whether increasing the type space will decrease the frequency to which we expect to see pragmatic {\em some} users is interesting. It now figures in \hl{Section~X.X}. At present, we have no answer.

\noindent\rule{\textwidth}{1pt}
\stepcounter{reviewerCommentCounter} %increase counter

\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25,frametitle= Reviewer \thereviewerCounter~comment \thereviewerCommentCounter \hfill ~~({\it LOT \& complexity})]
%
The discussion on derivation costs is not very satisfying on my opinion. For instance, the condition of "simple representations being favored over more complex ones" is compatible with many more cost specifications.

%
\end{mdframed}

We have expanded this section to further motivate our choices. Relatedly, see our answer to Reviewer 1's comment 4 above.

%\noindent\rule{\textwidth}{1pt}
%\stepcounter{reviewerCommentCounter} %increase counter
%\stepcounter{reviewerCounter} %increase counter
%
%\begin{mdframed}[backgroundcolor=gray!25,linecolor=gray!25,frametitle= Reviewer \thereviewerCounter~comment \thereviewerCommentCounter \hfill ~~({\it topic of the issue})]
%%
%reviewer comment
%%
%\end{mdframed}
%
%
%and here we answer
%%
%
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{apalike}
\bibliography{./bounds-rmd}

\end{document}

### 2018-01-11 code-LOT-extension/co-evo/paper-toy-demo.py

@@ -1,0 +1,235 @@
+import sys,os
+lib_path = os.path.abspath(os.path.join('..')) #specifying path for player module
+sys.path.append(lib_path) #specifying path for player module
+from player import LiteralPlayer, GriceanPlayer
+import numpy as np
+import matplotlib.pyplot as plt
+from itertools import product, combinations
+
+def normalize(m):
+    return m / m.sum(axis=1)[:, np.newaxis]
+
+def get_u(types):
+    out = np.zeros([len(types), len(types)])
+    for i in xrange(len(types)):
+        for j in xrange(len(types)):
+            out[i,j] = (np.sum(types[i].sender_matrix * np.transpose(types[j].receiver_matrix)) / 2.) * 0.5 +\
+                       (np.sum(types[j].sender_matrix * np.transpose(types[j].receiver_matrix)) / 2.) * 0.5
+    return out
+
+def obs_counts(obs):
+    out = []
+    for i in xrange(len(obs)):
+        out.append([obs[i].count(j) for j in xrange(4)])
+    return out
+
+def get_likelihood(obs,likelihoods):
+    out = np.zeros([len(likelihoods), len(obs)]) # matrix to store results in
+    for lhi in range(len(likelihoods)):
+        for o in range(len(obs)):
+            flat_lhi = likelihoods[lhi].flatten()
+            out[lhi,o] = np.prod([flat_lhi[x]**obs[o][x] for x in xrange(len(obs[o]))]) 
+    return out
+
+def get_q(lexica_prior,learning_parameter,k,types):
+    likelihoods = [i.sender_matrix for i in types]
+    atomic_obs = [0,1,2,3] #0,1 upper-row; 2,3 are lower row
+    D = list(product(atomic_obs,repeat=k))  
+    D = obs_counts(D)
+    
+    lhs = get_likelihood(D,likelihoods)
+    post = normalize(lexica_prior * np.transpose(lhs))
+    parametrized_post = normalize(post**learning_parameter)
+    return normalize(np.dot(lhs,parametrized_post))
+
+
+def learning_prior(types,c):
+    out = np.zeros(len(types))
+    for i in xrange(len(types)):
+        lx = types[i].lexicon
[…]

### 2018-01-12 paper/cogsci-submission2.tex
revision draft 2/3

@@
 The update effect of mutation on its own are described by $\text{M}(x) = .9 x + .1 (1-x) = .8x + .1$, plotted as the linear green line in Fig.~\ref{fig:Updates_RMS}.
-As shown in Figure \ref{fig:Phase_RD}, in this example mutation alone has only one stable rest point.
+As shown in Fig.~\ref{fig:Phase_RD}, in this example mutation alone has only one stable rest point.
 It is located at $x = .5$.
@@
 
-\subsection{Example}
+\subsection{Example}\label{subsec:example}
 Section~\ref{sec:si-case-study} shows how these components and their interaction can shed light on the evolution of scalar implicatures.
@@
 We may think about such a population as a point in a two-dimensional space.
-In Figure \ref{fig:sketch} this space is outlined by the grey dotted square.
-The first dimension determines the proportion of types that lexicalize ambiguous $\msome$ (the $x$-axis in Figure \ref{fig:sketch}).
+In Fig.~\ref{fig:sketch} this space is outlined by the grey dotted square.
+The first dimension determines the proportion of types that lexicalize ambiguous $\msome$ (the $x$-axis in Fig.~\ref{fig:sketch}).
 That is, the proportion of types with $\Llack$.
-The second dimension determines the proportion of pragmatic language users (the $y$-axis in Figure \ref{fig:sketch}).
+The second dimension determines the proportion of pragmatic language users (the $y$-axis in Fig.~\ref{fig:sketch}).
 Populations at the either corner of this space are monomorphic, consisting purely of a single type.
@@
 When it comes to communicative success, types with $\Lbound$ have a functional advantage over $\Llack$; literal $\Lbound$ has a slight advantage over pragmatic $\Lbound$; all have a sizable advantage over literal $\Llack$.
-The red arrows in Figure \ref{fig:sketch} sketch out these relationships.
+The red arrows in Fig.~\ref{fig:sketch} sketch out these relationships.
 They depict the rough trajectories we may expect if there is only pressure for communicative success.
@@
 This would lead populations to gravitate toward the space's center after mutator steps.
-This is sketched by the blue dash-dotted arrows in Figure~\ref{fig:sketch}.
+This is sketched by the blue dash-dotted arrows in Fig.~\ref{fig:sketch}.
 Neither prior nor posterior alone, but their combination, are what determines the effects of (iterated) Bayesian learning.
 If, instead of a flat prior, there is a prior that favors, say, $\Llack$ over $\Lbound$ we can expect mutator steps to be skewed toward the upper-half of the space.
-This is suggested by the purple arrows in Figure~\ref{fig:sketch}.
+This is suggested by the purple arrows in Fig.~\ref{fig:sketch}.
 Taking stock, the effects of iterated learning will depend not only on $\lambda$, affecting the likelihood, but also on the prior, how strong the bias is, and the degree to which learners tend to maximize the posterior, controlled by parameter $l$.
 
-Figure \ref{fig:quiver} shows actual trajectories, showcasing how they are modulated by parameter values.
+Fig.~\ref{fig:quiver} shows actual trajectories, showcasing how they are modulated by parameter values.
 The replicator steps make the functional disadvantage of $\Llack$ against $\Lbound$ apparent, particularly for literal types and/or low $\lambda$.
@@
 As expected, a small bias for a type will lead to larger proportions to adopt it provided that likelihood offers little to no differentiation between types.
-This is the situation the right-most space in Figure \ref{fig:quiver} depicts.
+This is the situation the right-most space in Fig.~\ref{fig:quiver} depicts.
 However, the bias is far from absolute.
@@
 
-Figures \ref{fig:sketch} and \ref{fig:quiver} depict replicator and mutator steps in isolation to better showcase their effects.
+Fig.~\ref{fig:sketch} and \ref{fig:quiver} depict replicator and mutator steps in isolation to better showcase their effects.
 Their combination will expectedly yield different outcomes, which we now turn to in a case study on scalar implicatures.
@@
 question: if a population of language users could freely combine different lexica with
-different pragmatic strategies, what are conditions under which the majority view of scalar
+different dispositions to act on them, what are conditions under which the majority view of scalar
 implicatures could have evolved?
@@
 it as a partition of possible worlds into cells where none, some or all of the $A$s are $B$s,
-for some arbitrary fixed predicates $A$ and $B$. Eight lexical representations can be distinguished based on
-their truth or falsity in three world states, six of which are not contradictory or
-tautological (see Table~\ref{tab:concepts} below).
+for some arbitrary fixed predicates $A$ and $B$.
+Enlarging the setup from Section~\ref{sec:languages+use} by another state will enable us to discern the difference between three kinds of lexica we will use to unravel the model's dynamics (see below).
+Eight lexical representations can be distinguished based on their truth or falsity in three world states, six of which are not contradictory or tautological (see Table~\ref{tab:concepts} below).
 
@@
 Recall that types are a combination of a lexicon and a manner of language use. We analyze the
-model's predictions in populations of types with one of the two behaviors introduced earlier;
-literal or pragmatic. The former correspond to level $0$ reasoners and the latter to level
-$1$. Accordingly, we consider a total of $432$ types. Six are variants of pragmatic language
-users with $\Llack$-like lexica. We refer to these as \emph{target types}, because they
+model's predictions in populations of types with one of the two behaviors introduced earlier:
+literal or pragmatic. The former correspond to level-$0$ reasoners and the latter to level $1$. Accordingly, there is a total of $432$ types. Six types are variants of pragmatic language
+users with $\Llack$-like lexica. We refer to these as \emph{target types} because they
 represent lexica and language use that conform to the majority view of scalar
-implicatures. Twelve types are either literal or pragmatic speakers with lexica of the
+implicatures. Twelve types are either literal or pragmatic users of lexica of the
 $\Lbound$ kind. We refer to these as \emph{competitor types}, because they are expected to be
@@
 \label{sec:an-induct-learn}
-
-\hl{The discussion on derivation costs is not very satisfying on my opinion.
-For instance, the condition of "simple representations being favored over more complex ones" is compatible with many more cost specifications.}
-
-\hl{We do not believe that, ultimately, we should assume the non-parsimonious view that there are different complexity measures for different structures.
-However, with previous work in iterated learning tradition, we do assume that (i) inductive learning biases can play a role in shaping culturally transmitted knowledge and that, as the reviewer notes, (ii) biases can be of manifold nature; another well-studied example being the mutual exclusivity bias, which could be expected to work against targets and in favor of competitors.}
-
-\hl{At present, it is impossible to disentangle the complex interaction of multiple biases.
-We therefore focus on the effects of a single plausible contributing factor to a preference of target lexica over competitors: a well-motivated bias favoring simplicity \citep{feldman:2000,chater+vitanyi:2003, piantadosi+etal:2012a,kirby+etal:2015,piantadosi+etal:underreview}.
-By assumption, this translates to a preference not to lexicalize linguistic material over lexicalizing it; simpler formulae over complex ones.
-As mentioned in Section 3.1.2, we acknowledge that the complexity measure that the prior is based on  is just a convenient operationalization of this particular hypothesis.}
-
-\hl{Lastly, we should stress that, just as models of iterated learning make different predictions under the assumption of different inductive biases, so does our model.
-The assumption and operationalization of a particular bias should, of course, always be seen critically (Section 4).}
 
@@
 (\citealt{feldman:2000, chater+vitanyi:2003, piantadosi+etal:2012a,kirby+etal:2015}, \citeauthor{piantadosi+etal:underreview} under review). 
-In a weighed generative LOT a representation's complexity is a function of its derivation cost.
+In a weighted generative LOT a representation's complexity is a function of its derivation cost.
 
@@
 
-There are many ways to define priors over lexica (see, e.g., \citealt{goodman+etal:2008,
-  piantadosi+etal:2012a}) but the key assumption here, common to all of them, is that simple
-representational expressions should be favored over more complex ones. We should stress that
-these details -- from the generative grammar to its complexity measure -- are to be regarded as
-one convenient operationalization of one general approach to explicating learning biases; this
-is not a commitment that this general approach is necessarily superior or that, within it, this
-particular instrumentalization is the single most plausible.
+The menu of inductive biases argued to shape language acquisition is steadily being refined.
+Apart from simplicity, prominent examples include mutual exclusivity \citep{merriman+bowman:1989,clark:2009}, regularization \citep{hudson+etal:2005}, and generalization \citep{smith:2011}.
+Even viewed in isolation, there are still many ways to define priors over lexica (see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a, kirby+etal:2015}).
+The key assumption here, common to all ways to define simplicity-biased priors, is that simple representational expressions should be favored over more complex ones.
+However, in view of the menu of possible operationalizations, other biases and their potential interplay, we should stress that these details -- from the generative grammar to its complexity measure -- are to be regarded as one convenient operationalization of one general approach to explicating learning biases; this is not a commitment that this general approach is necessarily superior or that, within it, this particular instrumentalization is the single most plausible.
+We return to this issue in relation to our results in Section~\ref{sec:an-induct-learn}.
 
@@
 I would recommend finding a way to share the same scale between subfigures, or make it visually clearer that the subfigures cannot be directly compared.}
-
-\hl{Mention that they have the same $k$, $\lambda$ and $l$: They do not have to but they do in our case study.
-Variation across parameter values would mean that we would have many more types: one for each combination of possible lexicon, pragmatic strategy, $\lambda$-, $l$- and $k$-value.
-This would allow us to trace the evolutionary trajectory (and competition) of not only lexica and pragmatic strategies, which is our goal here, but of all these factors combined.
-As our focus is not to show that, say, a particular $\lambda$ value leads to a particular evolutionary outcome we restrict our attention to a homogeneous population where all agents have the same parameter values.
-We discuss this issue in Section~3.2.
-Maybe take this from thesis.}
 
@@
 
-Recall that there are three parameters: soft-max parameter $\lambda$ affects how strongly
+Recall that there are three parameters, which we hold constant across types: soft-max parameter $\lambda$ affects how strongly
 speakers favor messages that appear best from their subjective point of view; the
@@
 posterior distribution. We expect that competitor types (types with lexica of the kind
-$\Lbound$) have a fitness advantage over target types (pragmatic agents with lexica of the kind
-$\Llack$), especially for very low levels of $\lambda$. Selection based on fitness alone may
+$\Lbound$) have a fitness advantage over target types (pragmatic types with lexica of the kind
+$\Llack$), especially for very low levels of $\lambda$ (see Fig~\ref{fig:sketch}). Selection based on fitness alone may
 therefore not lead to prevalence of target types in the population. On the other hand, lexica
 of type $\Llack$ are simpler than those of type $\Lbound$ by the postulated measure from
-above. This may make them more likely to be adopted by learners, especially when $k$ is low so
-that different teacher types are relatively indistinguishable based on their behavior and when
-$l$ is high. Still, lexica of the kind $\Lall$ are in turn even more likely \emph{a priori} than
-lexica of the kind $\Llack$. Simulation results will shed light on the question whether target
+above.
+This may make them more likely to be adopted by learners, especially when $l$ is high.
+Still, lexica of the kind $\Lall$ are in turn even more likely \emph{a priori} than lexica of the kind $\Llack$.
+As showcased in Section~\ref{subsec:example}, the prior is not all that counts for learnability.
+Simulation results will shed light on the question whether target
 types can emerge, and for which parameter constellations.
@@
 change is, if not absent, then at least very slow. In other words, even if the resulting states
-do not correspond to an eventual attracting state, they characterize almost stationary states
-in which the system remains for a very long time. As specified in \S\ref{sec:learnability}, the
+do not correspond to an eventual attracting state, they characterize population states in which the system remains for a long time. As specified in Section~\ref{sec:learnability}, the
 mutation matrix $Q$ can be obtained by considering all possible state-message sequences of
@@
 \includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/alt-fig1-onlyr}
-\caption{Proportion of target types, pragmatic competitor types, pragmatic $\Lall$, and the population's majority type, in representative populations after $50$ generations under only a
-  pressure for communicative success.}
+\caption{Proportion of target types, \hl{pragmatic} competitor types, \hl{pragmatic} $\Lall$, and the population's majority type, in representative populations after $50$ generations under only a
+  pressure for communicative success.
+\hl{Flip plots, change plot generation to show the highest 6 Lbound users and the highest Lall, irrespective of ling.
+behavior, change text surrounding these things}}
 \label{fig:only-R}
@@
 showcased in Fig.
-\ref{fig:only-R}, which shows the proportion of target types, pragmatic competitor types and pragmatic $\Lall$ in a representative population after $50$ replicator steps.
+\ref{fig:only-R}, which shows the proportion of target types, \hl{pragmatic} competitor types and \hl{pragmatic} $\Lall$ in a representative population after $50$ replicator steps.
 The plot also indicates the proportion of the
@@
 homogeneous population with many types. Conversely, higher $\lambda$ promotes 
-less-stochastic pragmatic behavior, widening the gap in expressivity between types and
+less-stochastic linguistic behavior, widening the gap in expressivity between some types and
 promoting more homogeneous populations. As suggested by Fig. \ref{fig:only-R}, the majority
@@
 sampling ($l = 1$) or a stronger tendency towards posterior maximization ($l = 15$) is shown in
-Fig.\ref{fig:only-M} together with the prior over types. The prior shows that while users of
+Fig.~\ref{fig:only-M} together with the prior over types. The prior shows that while users of
 $\Llack$ are not the most favored by the inductive bias (compared, e.g., to $\Lall$) they are
@@
 target types but does not lead to monomorphic populations if learners sample from the
-posterior (see the uppermost row in Figure \ref{fig:rmd}). Finally, a combination of high $\lambda$ and $l$ leads to increasing proportions of
+posterior (see the uppermost row in Fig.~\ref{fig:rmd}). Finally, a combination of high $\lambda$ and $l$ leads to increasing proportions of
 a single majority target type.
@@
 \section{General discussion}\label{sec:discussion}
+
+\hl{We do not believe that, ultimately, we should assume the non-parsimonious view that there are different complexity measures for different structures.
+However, with previous work in iterated learning tradition, we do assume that (i) inductive learning biases can play a role in shaping culturally transmitted knowledge and that, as the reviewer notes, (ii) biases can be of manifold nature; another well-studied example being the mutual exclusivity bias, which could be expected to work against targets and in favor of competitors.}
+
+\hl{At present, it is impossible to disentangle the complex interaction of multiple biases.
+We therefore focus on the effects of a single plausible contributing factor to a preference of target lexica over competitors: a well-motivated bias favoring simplicity \citep{feldman:2000,chater+vitanyi:2003, piantadosi+etal:2012a,kirby+etal:2015,piantadosi+etal:underreview}.
+By assumption, this translates to a preference not to lexicalize linguistic material over lexicalizing it; simpler formulae over complex ones.
+As mentioned in Section 3.1.2, we acknowledge that the complexity measure that the prior is based on  is just a convenient operationalization of this particular hypothesis.}
+
+\hl{Lastly, we should stress that, just as models of iterated learning make different predictions under the assumption of different inductive biases, so does our model.
+The assumption and operationalization of a particular bias should, of course, always be seen critically (Section 4).}
 
@@
 }
+
+\hl{Mention that they have the same $k$, $\lambda$ and $l$: They do not have to but they do in our case study.
+Variation across parameter values would mean that we would have many more types: one for each combination of possible lexicon, pragmatic strategy, $\lambda$-, $l$- and $k$-value.
+This would allow us to trace the evolutionary trajectory (and competition) of not only lexica and pragmatic strategies, which is our goal here, but of all these factors combined.
+As our focus is not to show that, say, a particular $\lambda$ value leads to a particular evolutionary outcome we restrict our attention to a homogeneous population where all agents have the same parameter values.
+We discuss this issue in Section~3.2.
+Maybe take this from thesis.}
 

### 2018-01-14 paper/cogsci-submission2.tex
revision draft + letter

@@
 \section{Introduction}\label{sec:introduction}
-\hl{2.
-This is true.
-We are grateful that this was pointed out, as our previous rendering could lead to confusion, e.g., as to why $L_{\text{all}}$, being the {\em a priori} more likely type to be inferred, did not come out strongly even when considering learnability alone.
-Throughout the article, we now stress more that what matters is the posterior -- and discuss the role of the prior mainly when contrasting targets and competitor, who crucially differ mostly in this respect.}
-
-\hl{3.
-These questions tie in with Reviewer $1$'s request for a clearer exposition and the explanation of why we speak of co-evolution in this case.
-In a nutshell, pressure for learnability alone does not put the types found in a population in competition.
-Instead, the population comes to be inhabited by types that are faithfully transmitted -- of which there might be many (e.g., all variants of a kind are equally learnable).
-What fitness-relative selection adds to this process is the competition that learnability alone lacks, independent of the type space we look at.
-Now, in the case of our type space we have a slight (modulo $\lambda$) communicative disadvantage for targets  {\em when in direct competition with competitors}, for instance.
-However, the fact that they are inferred more faithfully by na\"ive learners leads (i) to the existence of less competitors, counteracting this functional disadvantage, as well as (ii) to the existence of less variants of a kind (contra learning-alone).}
 
@@
 
-\hl{We have thoroughly revised the article to make the contrast between these three predictions clearer, as well as the predictions themselves.
-Concretely, This should be the new figure as well as a clearer contrast when discussing the simulation outcomes.
-THe questions are: Why do learnability pressures not have the same effect in the joint learnability/fitness model, as they do in the lesioned model?
-Why is high speaker optimality necessary for scalar implicatures to evolve, when it does not have this effect in the fitness-only model?}
-
 Recall that there are three parameters, which we hold constant across types: soft-max parameter $\lambda$ affects how strongly
@@
 \includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/alt-fig2-onlym-pr}
-\caption{Proportion of target types, pragmatic competitor types, pragmatic $\Lall$, and the population's majority type in representative populations after $50$ generations under only a
+\caption{Proportion of target types, \hl{pragmatic competitor} types, \hl{pragmatic} $\Lall$, and the population's majority type in representative populations after $50$ generations under only a
   pressure for learnability ($\lambda = 20, k = 5$). The learning prior is shown in the
@@
 a higher propensity to infer pragmatic $\Llack$ when the teacher's type produces very similar
-data, such as when using $\Lbound$. Moreover, $\Llack$ is less likely to be confused with
+data, such as when using $\Lbound$ (see Section~\ref{subsec:example} and Fig.~\ref{fig:sketch}). Moreover, $\Llack$ is less likely to be confused with
 types with different observable behavior because its pragmatic use approximates a one-to-one
@@
 Fig.~\ref{fig:only-R}), learnability alone does not succeed in selecting for a single
-prevalent type. Indeed, all six target types tend to coexist at roughly equal proportion. Each
+prevalent type; all six target types tend to coexist at roughly equal proportion. Each
 is passed on to the next generation with the same faithfulness and, differently from a pressure
@@
 
-As
-with a pressure only for communicative success, low values of $\lambda$ make the differences in observable behavior
-across types less pronounced and therefore reflect the learners' inductive bias more
-faithfully. This favors functionally deficient but a priori preferred types such as those that use
-$\Lall$. A pressure for learnability alone may consequently lead to a spread of communicatively
-suboptimal types that are easier to learn. In the extreme, when $l = 1$ and $\lambda = 1$ all
+As with a pressure only for communicative success, low values of $\lambda$ make the differences in observable behavior across types less pronounced.
+This makes differences in the likelihood of particular types having generated a learning input less pronounced.
+Therefore, low $\lambda$ leads populations to reflect the learners' inductive bias more faithfully.
+This favors functionally deficient but a priori preferred types such as those that use  $\Lall$.
+A pressure for learnability alone may consequently lead to a spread of communicatively suboptimal types that lexicalize simpler semantics.
+In the extreme, when $l = 1$ and $\lambda = 1$ all
 of $1000$ independent populations had users of $\Lall$ as majority types.
-For higher $\lambda$, it becomes clear that a high prior (e.g., $\Lall$) is not the only thing that counts for learnability.
+For higher $\lambda$, it becomes clear that a high prior (e.g., $\Lall$) is not the only thing that counts when it comes to learnability (see Section~\ref{subsec:example}).
 As soon as there is information for learners to discern whether one type is more likely to have generated the data (depending on $\lambda$ and $k$), it becomes paramount for types to produce data that makes them easily identifiable if they are to be transmitted faithfully.
@@
 \includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/alt-fig3-r+m}
-\caption{Proportion of target types, pragmatic competitor types, pragmatic $\Lall$, and the population's majority type in representative populations after $50$ generations under both pressures ($k = 5$).}
+\caption{Proportion of target types, \hl{pragmatic} competitor types, \hl{pragmatic} $\Lall$, and the population's majority type in representative populations after $50$ generations under both pressures ($k = 5$).}
 \label{fig:rmd}
@@
 $l$. Pressure for communicative success magnifies the effects of iterated learning and dampens
-the proliferation of multiple types of a kind that are equal in expressivity {\em and} learnability. A
-pressure towards learnability favors the transmission of simpler semantics and thereby selects
-for pragmatic language use.
+the proliferation of multiple variants of a kind that are equal in expressivity {\em and} learnability.
+A pressure towards learnability favors the transmission of simpler semantics and thereby selects for pragmatic language use in these types.
+Pragmatic behavior, however, is not selected in types that can do away without pragmatic reasoning such as $\Lbound$ users (see Section~\ref{subsec:example}).
 
@@
 that are a priori favored, such as $\Lall$. An increase in $\lambda$ leads to the selection of
-target types but does not lead to monomorphic populations if learners sample from the
-posterior (see the uppermost row in Fig.~\ref{fig:rmd}). Finally, a combination of high $\lambda$ and $l$ leads to increasing proportions of
-a single majority target type.
+target types but does not lead to monomorphic populations if learners sample from the posterior (see the uppermost row in Fig.~\ref{fig:rmd}).
+Finally, a combination of high $\lambda$ and $l$ leads to increasing proportions of a single majority target type.
 This joint influence is summarized in Fig.
@@
 
+When pressured for communicative success only, target types are outcompeted by competitor types.
+When pressured for learnability, populations are polymorphic due to lack of competition; minimally among variants of a kind.
+However, due to the prior, learners that are exposed to the input of either targets or competitors tend to infer acquire the former type.
+The consequence of combining both pressures is then that the (slight) functional disadvantage of targets is counterbalanced by an advantage in learnability.
+This leads to large proportions of targets in a population and, due to the competition among types applied by each replicator steps, to more monomorphic populations.
+
 To recapitulate, other than the involvement of pressure on both communicative success and learnability,
@@
 
-More broadly,
-target types, which represent the majority view of scalar implicatures, can come to
+More broadly, target types, which represent the majority view of scalar implicatures, can come to
 dominate the population if three assumptions are met: (i) language is pressured toward both
-communicative success and learnability; (ii) pragmatic language use is an option; (iii) learners prefer
+communicative success {\em and} learnability; (ii) pragmatic language use is an option; (iii) learners prefer
 simpler over more complex lexical representations and exhibit a tendency towards the
@@
 \section{General discussion}\label{sec:discussion}
-
-\hl{We do not believe that, ultimately, we should assume the non-parsimonious view that there are different complexity measures for different structures.
-However, with previous work in iterated learning tradition, we do assume that (i) inductive learning biases can play a role in shaping culturally transmitted knowledge and that, as the reviewer notes, (ii) biases can be of manifold nature; another well-studied example being the mutual exclusivity bias, which could be expected to work against targets and in favor of competitors.}
-
-\hl{At present, it is impossible to disentangle the complex interaction of multiple biases.
-We therefore focus on the effects of a single plausible contributing factor to a preference of target lexica over competitors: a well-motivated bias favoring simplicity \citep{feldman:2000,chater+vitanyi:2003, piantadosi+etal:2012a,kirby+etal:2015,piantadosi+etal:underreview}.
-By assumption, this translates to a preference not to lexicalize linguistic material over lexicalizing it; simpler formulae over complex ones.
-As mentioned in Section 3.1.2, we acknowledge that the complexity measure that the prior is based on  is just a convenient operationalization of this particular hypothesis.}
-
-\hl{Lastly, we should stress that, just as models of iterated learning make different predictions under the assumption of different inductive biases, so does our model.
-The assumption and operationalization of a particular bias should, of course, always be seen critically (Section 4).}
-
-\hl{As now made explicit in \hl{Section~xx.xx}, we chose three messages for illustrative purposes mainly.
-In particular, we wanted to inspect a type space where targets are not the most likely {\em a priori} as this may mislead readers into thinking that learnability alone, in particular the prior alone, drives the outcome.
-This is related to our answer to reviewer 2, comment 7.
-The reason why we did not go beyond $n>2$ is for computational tractability, as calculating $Q$, even if approximated, is expensive.
-The question whether increasing the type space will decrease the frequency to which we expect to see pragmatic {\em some} users is interesting.
-It now figures in Section~X.X.
-At present, we have no answer.
-}
-
-\hl{Mention that they have the same $k$, $\lambda$ and $l$: They do not have to but they do in our case study.
-Variation across parameter values would mean that we would have many more types: one for each combination of possible lexicon, pragmatic strategy, $\lambda$-, $l$- and $k$-value.
-This would allow us to trace the evolutionary trajectory (and competition) of not only lexica and pragmatic strategies, which is our goal here, but of all these factors combined.
-As our focus is not to show that, say, a particular $\lambda$ value leads to a particular evolutionary outcome we restrict our attention to a homogeneous population where all agents have the same parameter values.
-We discuss this issue in Section~3.2.
-Maybe take this from thesis.}
-
 The approach introduced here combines game-theoretic models of functional pressure towards
@@
 trajectories, (ii) the characterization of language learning as a joint inference over
-pragmatic behavior and lexical meaning, and (iii) the possibility to trace the co-evolution of
+linguistic behavior and lexical meaning, and (iii) the possibility to trace the co-evolution of
 conventional semantics and pragmatic use.
 
-\hl{Redo around here. See commented out}
-%With respect to (i), in Section \ref{4sec:success-vs-expressivity} we discussed a comparable model proposed by \citet{kirby+etal:2015} and highlighted the difference between a type's expressivity and functional pressure.
-As showcased in Section \ref{4sec:si-case-study}, the latter pressure can indirectly select for expressive types, i.e., those that can convey states unequivocally.
-By contrast, Kirby et al.'s model only considers the bearing that expressivity has on the production of learnable data.
-We see three main reasons for considering utility rather than just expressivity; and, relatedly, to consider the effects that communication {\em and} learning have on an evolving linguistic system, rather than only learning.
+With respect to (i), \citet{kirby+etal:2015} propose a comparable model of the interaction
+between a lexicon's expressivity and its learnability. A main difference is that here we
+considered a pressure for mutual understanding within a generation.
+More precisely, expected utility.
+This pressure can indirectly select for expressive types, those that can convey states unequivocally, whereas Kirby et al.
+only consider the bearing that the latter ability has on the production of learnable data.
+We see three main reasons for considering communicative success rather than just
+expressivity, and for looking at communication and learning rather than just learning.
 First, learning alone can  promote populations with non-negligible proportions of functionally defective types.
-This is true both of simulations, e.g.
-our $\Lall$-types in Figure \ref{fig:only-M}, as well as of laboratory experiments with human subjects (see \S\ref{4sec:success-vs-expressivity} and references within).
-Second and more importantly, types may be equally expressive but their performance as means of information transfer crucially depend not only on themselves but on the population they find themselves in.
-That is, we contend that adopting an expressive type that generates learnable data does not in itself capture a type's arguably central communicative function of transferring information to peers.
+This is true both of simulations, e.g., our $\Lall$-types in Fig.~\ref{fig:only-M}, as well as of laboratory experiments with human subjects (see \citealt{kirby+etal:2008,silvey+etal:2014}, and see \citealt{fay+etal:2013} for review of laboratory results).
+Second and more importantly, types may be equally expressive but their performance as means of information transfer crucially depend not only on themselves but on the population they find themselves in (compare the competition of target types in Fig.
+\ref{fig:rmd} and their lack of competition in Fig.
+\ref{fig:only-M}).
+That is, we contend that adopting an expressive type that generates learnable data does not, in itself, capture a type's arguably central communicative function of transferring information to peers.
 Taking communication into consideration allows the model to be responsive to the task for which language is learned.
 Lastly, chains of iterated learning alone do not put types in direct competition.
-Accordingly, learning can lead to polymorphous populations in which multiple variants of a type coexist (compare the competition of target types in Figure \ref{fig:rmd} and their lack of competition in Figure \ref{fig:only-M}).
-This latter point of course depends on the particulars of a model's learning component.
-Here, we explored the effects that simple transmission chains have, but it is possible to add further complexity to this process.
-For example, by letting learners learn from multiple teachers \citep{burkett+griffiths:2010,brochhagen+etal:2016}.
-
-With respect to (i), \citet{kirby+etal:2015} propose a comparable model of the interaction
-between a lexicon's expressivity and its learnability. A main difference is that here we
-considered a pressure for mutual understanding within a generation that may indirectly select
-for expressive types, i.e., those that can convey states unequivocally, whereas Kirby et
-al. only consider the bearing that the latter ability has on the production of learnable
-data. We see two main reasons for considering communicative success rather than just
-expressivity, and over only considering it in learning. First, learnability alone can lead to the
-acquisition of functionally defective languages, as showcased by $\Lall$
-(cf. \citealt{kirby+etal:2008,silvey+etal:2014}, and see \citealt{fay+etal:2013} for review of
-laboratory results). Second and more importantly, types may be equally expressive but their
-performance as means of information transfer crucially depends not only on themselves but on
-the population they find themselves in (cf. the competition of target types in Fig. 
-\ref{fig:rmd} and their lack of competition in Fig. \ref{fig:only-M}). That is, we contend
-that having an expressive lexicon that generates learnable data does not in itself capture a
-type's arguably central communicative function of transferring information to peers. Taking communication
-into consideration in turn allows the model to be responsive to the task for which language is learned.
+Accordingly, learning can lead to polymorphous populations in which multiple variants of a type coexist \citep{nowak:2006}.
 
 The main result of our case study is that types that correspond to the majority view of scalar
-implicatures (scalar readings are non-lexicalized pragmatic enrichments) can come to dominate a
-population. This can happen under the assumption that simpler semantic representations are more
-likely to be learned (cf. \citealt{chater+vitanyi:2003}). Pragmatic language use can be
-recruited indirectly by a preference for simpler lexical representations. Under this view, semantics
+implicatures -- scalar readings are non-lexicalized pragmatic enrichments -- can come to dominate a
+population.
+This can happen if pragmatic language use is recruited indirectly by a preference for simpler lexical representations (relative to more complex ones that lead to comparable overt linguistic behavior without pragmatic language use).
+Under this view, semantics
 and pragmatics play a synergic role: pragmatic use allows maintenance of simpler representations;
@@
 issues. First of all, while many favorable parameter settings exist which lead to a prevalence
-of target types, other types are usually represented in non-negligible proportions. This may
-just be a technical quirk of the mutator step. But there is a related issue of empirical
+of target types, other types are usually represented in non-negligible proportions (see Fig.~\ref{fig:rmd} and \ref{fig:diff}). This may
+just be a technical quirk of the mutator step; but there is a related issue of empirical
 importance. Several experimental studies on scalar implicatures suggest that participants can
@@
 selection. Lexicalized upper-bounds can dominate a population because they may boost
-communicative success. But they do not have to. Moreover, even without selective pressure
+communicative success. But they do not have to. Moreover, as stressed throughout this investigation, even without selective pressure
 for communicative success, it is not necessarily the case that the types that are most
@@
 contextual cues that might otherwise mislead the hearer.
-In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so (cf.
-Chapter \ref{chapter:ambiguity}).
+In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so.
 What this alternative argument does not explain is why functional pressure does not lead to the emergence of different, equally costly lexical items to express different knowledge states of the speaker (\citealt[252-267]{horn:1984},
@@
 For instance, to the emergence of two expressions for each weak scalar expression; one with and one lacking an upper-bound.
-That is, it does not explain why English and other languages do not have a monomorphemic dual expression for, e.g., {\em some} that lexicalizes an upper-bound.
+That is, this argument does not explain why English and other languages do not have a monomorphemic dual for, e.g., {\em some} that lexicalizes an upper-bound.
 If this hypothetical expression existed, it could be deployed to signal that the speaker knows that {\em some but not all} holds, and unbounded {\em some} could exclusively signal epistemic uncertainty.
@@
 In this realm an interesting issue is whether proposed principles, such as the semantic conventionalization of once highly context-dependent inferences if they become regular enough \citep{levinson:2000,traugott:2004}, can be given
-a formal rationale and inform postulated directionalities of change. The present chapter made a first start and gave a framework for exploring these issues systematically.
+a formal rationale and inform postulated directionalities of change. The present investigation made a first start and gave a framework for exploring these issues systematically.
 
@@
 The cultural evolution of meaning is influenced by intertwined pressures. We set out to
-investigate this process by putting forward a model that combines a pressure toward successful
+investigate this process by putting forward a model that combines pressure toward successful
 information transfer with perturbations that may arise in the transmission of linguistic
 knowledge in acquisition. Its objects of selection and replication are pairs of lexical
-meanings and patterns of language use. This allows the model to trace the interaction between
+meanings and patterns of language use. This allows the model to trace the evolutionary interaction between
 conventional meaning and pragmatic use. Additionally, it takes the challenge seriously of
@@
 their mutual influence were highlighted in a case study on the lack of lexical upper-bounds in
-weak scalar expressions that showed that, when pressured for learnability and communicative
-success, the former force drives for simpler semantic representations inasmuch as pragmatics
+weak scalar expression. This study showed that, when pressured for learnability and communicative
+success, the former force can drive for simpler semantic representations inasmuch as pragmatics
 can compensate for lack of expressivity in use. That is, the relative learning advantage of

### 2018-01-14 paper/rv-letter1.tex
revision draft + letter

@@
 
-We would like to thank the reviewers for their helpful comments and suggestions. They are very welcome and have certainly helped improve the article.
+We would like to thank the reviewers for their helpful comments and suggestions. They are very welcome and have certainly helped improve the manuscript.
 
-In general terms, we have taken care to make the exposition of the model clearer, to clarify the  motivations underlying the modeling choices we make in the case study, and to clarify how the components of the model lead to the results they do.
-In what follows we address the reviewers' content-related comments and questions one by one, with pointers to respective changes effected on the manuscript.
+We have taken care to improve the exposition of the model; to clarify the  motivations underlying the modeling choices we make in the case study, and to make more precise and explicit how the components of the model lead to the results they do.
+In what follows we address the reviewers' main comments and questions one by one, with pointers to respective changes effected on the manuscript.
 
@@
 the co-evolutionary dynamic mentioned in the title could be better justified [...] In co-evolutionary dynamics, the evolution of one factor (here, the lexicon) should have an impact on the evolution of another (here, pragmatic strategy), and vice versa; the paper does not show evolutionary dynamics (change over time) for either of these factors, not does it show how evolution of a single factor favours the evolution of the other.
-The model looks more like adaptation of a kind of linguistic type (a lexicon + a pragmatic strategy) to selection pressures for communication and for learnability than co-evolution.
-Please justify your characterization of the process modeled as co-evolution or change to adaptation.
+[...] Please justify your characterization of the process modeled as co-evolution or change to adaptation.
 
@@
 
-\hl{DL.
-TO DO AFTER FIGURE RMD} This is an important point that we now stress and highlight more throughout the paper.
-Particularly in \hl{here, here, and here}.
-The discussion surrounding Figure \hl{xxx} further illustrates how one factor has an impact on the evolution of another.
-
+This is an important point that we now stress and highlight more throughout the paper.
+In particular the discussion surrounding Figure $2$ and $3$ in the new Section~2.4 makes precise how one factor has an impact on the evolution of the other.
 In a nutshell: a pragmatic strategy (level-$1$ reasoning) favors the evolution of underspecified semantics of the target kind ($L_{\text{lack}}$) as it allows for the maintenance of simpler lexical representations that are easier to learn without functional disadvantages otherwise incurred in communication (e.g., by level-$0$ reasoners).
@@
 When paired with other semantics, pragmatic refinement can instead be (close to a) neutral trait (e.g., $L_{\text{bound}}$ paired with high $\lambda$) or it can even lead to  functional disadvantages when compared with level-$0$ reasoning (e.g., $L_{\text{bound}}$ with low $\lambda$).
-Both cases are depicted in Figure \hl{xx.xx}.
 
@@
 As we discuss in Section~1, our case study focuses on scalar implicatures because they are well-studied systematic pragmatic inferences that have received much attention, both theoretically and experimentally.
-This makes them particularly suitable candidates for the study of the evolution of regular pragmatic inferences because there are many findings that can guide the application and evaluation of our model to the study of these inferences.
+This makes them suitable candidates for the study of the evolution of regular pragmatic inferences because there are many findings that can guide the application and evaluation of our model to the study of these inferences.
 This ties in with our particular choices to model pragmatic language use of scalar expressions.
-As we now make more explicit \hl{in \S 2.3.1}, these draw from a growing game-theoretic and Bayesian tradition \citep[e.g.,][]{franke:2009,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}.
-More generally, we have added further background and motivation to our modelling choices throughout the article.
+As we now make more explicit in \S 2.3.1, these draw from a growing game-theoretic and Bayesian tradition \citep[e.g.,][]{franke:2009,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}.
 
@@
 \stepcounter{reviewerCommentCounter} %increase counter
-Reviewer \thereviewerCounter~asks for a clearer description of the model.
+\noindent Reviewer \thereviewerCounter~asks for a clearer description of the model.
 As noted earlier, we have taken care to make the exposition more accessible and to guide the reader more through the technical details of Section 2.
-\hl{In particular, Section 2.xx is new and should clarify much}.
-We took the reviewer's questions, briefly answered below, as guidance for the changes made in Section~2.
+In particular, the newly added Section 2.4 should help unravel the model.
+We took the following questions, briefly answered below, as guidance for changes made in Section~2.
 
@@
 \end{mdframed}
-\hl{As now made more explicit in \S2.2}, steps of the discrete-time replicator mutator dynamic correspond to generational turn-overs of the entire population (definitions (1) and (2)).
+As now made more explicit in \S2.2, steps of the discrete-time replicator mutator dynamic correspond to generational turn-overs of the entire population (definitions (1) and (2)).
 That is to say, all agents die at once, being replaced by the following generation in a single swoop.
@@
 The amount of teachers of a given type is proportional to the frequency of this type in the population after the replicator step.
-More generally, the population is infinite so there are infinite learners/teachers.
-This was implicit in the definitions of the replicator mutator dynamic in (1) and (2) before, and is now explicitly stated in Section~2.2 in the prose that motivates and explicates these definitions.
+The population is infinite so there are infinite learners/teachers.
+This was implicit in the definitions of the replicator mutator dynamic in (1) and (2), and is now explicitly stated in Section~2.2 in the prose that motivates and explicates these definitions.
 
@@
 This allows us to track change that does not depend on varying population sizes nor their growth rate.
-This is why we speak of proportions of agents of a type, rather than of there being a particular number of agents of a type.
+This is why we speak of proportions of a type, rather than of there being a particular number of agents of a type.
 
@@
 \end{mdframed}
-While our model is well-compatible with the assumption that every agent comes with her own parameter values, the reviewer is correct in pointing out that we assume they all to have the same values.
+While our model is well-compatible with the assumption that every agent comes with her own parameter values, the reviewer is correct in pointing out that we assume they all to have the same values (explicit in \S3.2).
 Variation across agents would mean many more types: one for each combination of a possible lexicon, pragmatic strategy, $\lambda$-, $l$- and $k$-value.
 This would allow us to trace the evolutionary trajectory of not only lexica and pragmatic strategies, which is our goal here, but of all these factors combined.
-As our focus is not to show that, say, a particular $\lambda$ value leads to a particular evolutionary outcome we restrict our attention to a homogeneous population where all agents have the same parameter values.
-\hl{We discuss this issue in Section~3.2}.
+As mentioned in Section~$4$, this would be interesting future research.
 
@@
 This is determined by the teacher's production probabilities; the speaker behavior of her type.
-Put differently, communicative success does not come into play, only speaker behavior.
+Communicative success accordingly does not come into play, only speaker behavior.
 It would certainly be possible to let learners witness proficient users communicate and to have them infer types from such interactions instead.
 However, we wanted to keep our assumptions close to the well-studied assumptions standardly made in the iterated (Bayesian) learning tradition.
-Iterated learning only considers production, rather than production and comprehension (\hl{see discussion that should make this explicit in the discussion section - maybe draw from thesis}).
+Iterated learning only considers production, rather than production and comprehension.
 Details on how learners infer types are given in Section~2.3.3.
@@
 \end{mdframed}
-As with the other questions, we agree and have made these matters more explicit. This matter is particularly important. We now stress it throughout the article.
-
-In short, a type is a combination of a lexicon and a pragmatic disposition to act on it.
+In short, a type is a combination of a lexicon and a disposition to act on it.
 Together, both ingredients define an agent's linguistic behavior.
 A learner faced with input $d$ (see above) calculates $P(\tau \mid d)$, the probability that a type $\tau$ produced input $d$.
-Crucially, neither pragmatic strategies nor lexical meanings are directly observable, so they can only be faithfully recovered insofar as the behavior evidenced by a type is (in tendency) attributable to only one pairing of a lexicon and a pragmatic strategy.
+Crucially, neither pragmatic strategies nor lexical meanings are directly observable, so they can only be faithfully recovered inasmuch as the {\em overt} behavior evidenced by a type is (in tendency) attributable to only one, {\em covert}, pairing of a lexicon and a pragmatic strategy.
 
@@
 This may involve the recruitment of contextual information, as the reviewer notes, but can also involve pragmatic enrichments that result purely from reasoning about linguistic alternatives (``why did the speaker say {\em some} instead of {\em all}'').
-It is in the latter sense that reasoning beyond level $0$ is pragmatic.
-Classically, theoretical analyses of scalar implicatures put their emphasis on reasoning about linguistic choice rather than on contextual information \citep[e.g.,][]{horn:1972,gazdar:1979, franke:2009, GoodmanStuhlmuller2013:Knowledge-and-I} and we chose to do the same.
+It is in the latter sense that reasoning beyond level $0$ is pragmatic (see \S2.31.).
+Classically, theoretical analyses of scalar implicatures put their emphasis on reasoning about linguistic choice rather than on contextual information \citep[e.g.,][]{horn:1972,gazdar:1979, franke:2009, GoodmanStuhlmuller2013:Knowledge-and-I}.
+We chose to do the same.
 
@@
 Intuitively, the data such teachers produce are all over the place and do not do a good job in setting them apart from other types.
-Consequently, even if the prior favors $L_{\text{all}}$, this type is not transmitted very faithfully; learners reason that it could also come from other types and may adopt those instead.
+Consequently, even if the prior favors $L_{\text{all}}$, this type is not transmitted very faithfully; learners reason that the input they get from $L_{\text{all}}$-teachers could also come from other types and may adopt those instead.
 If there are types that are transmitted more faithfully, the population will, over time, transition to these types rather than to $L_{\text{all}}$.
@@
 
-We have thoroughly revised the article to make the contrast between these three predictions clearer, as well as the predictions themselves.
-Concretely, \hl{say where this is more explicit now.
-This should be the new figure as well as a clearer contrast when discussing the simulation outcomes}.
-
+We have thoroughly revised the article to make the contrast between these three predictions clearer (see Section~3.2, particularly 3.2.3; but also Section~2.4, which illustrates how the dynamics behave in a smaller type space).
 These questions tie in with Reviewer $1$'s request for a clearer exposition and the explanation of why we speak of co-evolution in this case (reviewer 1, comment 1).
@@
 Now, in the case of our type space we have a slight (modulo $\lambda$) communicative disadvantage for targets  when in direct competition with competitors, for instance.
-However, the fact that targets are inferred more faithfully by na\"ive learners leads (i) to the gradual existence of less competitors, counteracting this, if slight, functional disadvantage, as well as (ii) to the existence of less variants of a kind (contra the {\em learning alone} dynamic).
+However, the fact that targets are inferred more faithfully by na\"ive learners leads (i) to the gradual existence of less competitors, counteracting this functional disadvantage, as well as (ii) to the existence of less variants of a kind (contra the {\em learning alone} dynamic).
 %
@@
 
-It is not necessarily true that it will always be rational for agents to behave pragmatically.
-\hl{We now make explicit in Section~2}, but the main thing to note is that a pragmatic level-$1$ hearer using lexicon $L_{\text{bound}}$ will reason over the behavior of a {\em soft-maximizing} level-$0$ speaker of $L_{\text{bound}}$.
+It is not necessarily true that it will always be rational for agents to behave pragmatically (see the end of Section~2.3.1 but also the discussion surrounding Figure~3a).
+The main thing to note is that a pragmatic level-$1$ hearer using lexicon $L_{\text{bound}}$ will reason about the behavior of a {\em soft-maximizing} level-$0$ speaker of $L_{\text{bound}}$.
 The lower $\lambda$, the more the stochasticity percolates to level-$1$ pragmatic interpretation.
@@
 
-As we now clarify in \hl{Section~n} and stress a couple of times throughout the article, we assume very little sophistication from our agents.
-In particular, they do {\em not} know the type of the player that the are playing against.
+As we now clarify in Section~2.3.1 and stress a couple of times throughout the article, we assume very little sophistication from our agents.
+In particular, they do not know the type of the player that the are playing against.
 They simply behave according to their subjective point of view in a boundedly rational fashion.
-For example, the linguistic choices of a level-$1$ speaker of lexicon $L_{\text{some}}$ do not change depending on whom she interacts with; they are always defined as (boundedly) rational choice relative to the interpretative behavior of a level-$0$ user of $L_{\text{some}}$ (\hl{point to definitions of behavior in section 2}).
+For example, the linguistic choices of a level-$1$ speaker of lexicon $L_{\text{some}}$ do not change depending on whom she interacts with.
+They are always defined as (boundedly) rational choice relative to the interpretative behavior of a level-$0$ user of $L_{\text{some}}$ (see (3) --(6) in \S2.3.1).
 %
@@
 
-This is true.
-We are grateful that this was pointed out, as our previous rendering could lead to confusion, e.g., as to why $L_{\text{all}}$, being the {\em a priori} more likely type to be inferred, did not come out strongly even when considering learnability alone.
-\hl{Throughout the article, we now stress more that what matters is the posterior -- and discuss the role of the prior mainly when contrasting targets and competitor, who crucially differ mostly in this respect.}
+This is an important point.
+We agree that some fragments of our previous manuscript could lead to confusion, e.g., as to why $L_{\text{all}}$, being the {\em a priori} more likely type to be inferred, did not come out strongly even when considering learnability alone.
+Throughout the article, we now stress more that what matters is the posterior and explain the role of the likelihood more.
 
@@
 
-This is a standard assumption in the literature lacking reasons to believe that there is an asymmetry in how often, on average, agents use and hear a type of expression.
-In the case of scalar implicatures, we see no reason to assume such an asymmetric contribution to fitness.
-
-As for concrete impact: The main contrast that we are interested in, i.e., competition between targets and competitors, does not hinge on agents being speakers half of the time.
-However, the contrast between pragmatic and literal competitor types does, because their receiver behaviors differ.
-Assuming that successful receiver behavior contributes differently than sender behavior would therefore modulate differences between these types.
+This is a standard assumption in the literature (lacking reasons to believe that there is an asymmetry in how often, on average, agents use and hear a type of expression.) In the case of scalar implicatures, we see no reason to assume such an asymmetric contribution to fitness.
+As for concrete impact: The main contrast that we are interested in (competition between targets and competitors) does not hinge on agents being speakers half of the time.
+However, e.g., the contrast between pragmatic and literal competitor types does as they differ only in their receiver behaviors.
+Assuming that receiver behavior contributes differently than sender behavior would therefore modulate differences between these types.
 
@@
 
-As now made explicit in \hl{Section~xx.xx}, we chose three messages for illustrative purposes mainly.
+As now made explicit in Section~3.1, we chose three messages for illustrative purposes mainly.
 In particular, we wanted to inspect a type space where targets are not the most likely {\em a priori} as this may mislead readers into thinking that learnability alone, in particular the prior alone, drives the outcome.

### 2018-01-14 code-LOT-extension/plots.py
barh for fig-onlyr and fig-onlym, with max. competitor and lall

@@ -469,24 +469,6 @@ def get_3d_diff_incumbents(type_list,list1,list2,kind):
     other_types = [x for x in all_types if x not in target_types]
     seq_length = 5
     
-    #data = np.zeros((len(list1)+1, len(list2)+1))
-    #for i in xrange(len(list1)):
-    #    for j in xrange(len(list2)):
-    #        #To avoid errors when adding cols to slices of copies:
-    #        dt = pd.read_csv('./results/%s-s3-m3-lam%d-a1-k%d-samples250-l%d-g50-meFalse.csv' % (kind,list1[i],seq_length,list2[j]))
-    #        do = pd.read_csv('./results/%s-s3-m3-lam%d-a1-k%d-samples250-l%d-g50-meFalse.csv' % (kind,list1[i],seq_length,list2[j]))
-    #
-    #        data[0,j+1] = list2[j]
-    #        data[i+1,0] = list1[i]
-    #        
-    #        dt = dt[target_types] 
-    #        do = do[other_types]
-    #        dt['incumbent_t'] = dt.max(axis=1) #incumbent amongst types in d_t
-    #        do['incumbent_o'] = do.max(axis=1) #incumbent amongst types in d_o
-    #        best_of_targets = dt['incumbent_t'].mean()
-    #        best_of_others = do['incumbent_o'].mean()
-    #        data[i+1,j+1] = best_of_targets - best_of_others
-
     data = np.genfromtxt('differences-data.csv',delimiter=',')
     dPrime = data[1:,1:]
     sns.set_style("whitegrid")
@@ -614,10 +596,7 @@ types = [231,236,291,306,326,336]
 def get_subfigs_hist_replication(type_list,list1,list2):
     seq_length = 5
     num_of_runs = 5
-     
-    types_to_plot = ['t_final'+str(x) for x in type_list]  
-    targets_to_plot = ['t_final'+str(x) for x in type_list[:6]]  
-
+    
     restrict_to_final = ['t_final'+str(z) for z in xrange(432)] #as to avoid the incumbent to come from other columns
 
     df1 = pd.read_csv('./results/%s-s3-m3-lam%d-a1-k%d-samples250-l%d-g50-meFalse.csv' % ('r',list1[0],seq_length,list2[0]))
@@ -632,70 +611,132 @@ def get_subfigs_hist_replication(type_list,list1,list2):
  
     df3 = df3[restrict_to_final]
[…]

### 2018-01-14 paper/cogsci-submission2.tex
small modifications + new barh plots

@@
 \subsection{Example}\label{subsec:example}
+\hl{Mention that x,y independent}
 Section~\ref{sec:si-case-study} shows how these components and their interaction can shed light on the evolution of scalar implicatures.
@@
 \centering
-\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/alt-fig1-onlyr}
+%\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/alt-fig1-onlyr}
+\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/barh-onlyr}
 \caption{Proportion of target types, \hl{pragmatic} competitor types, \hl{pragmatic} $\Lall$, and the population's majority type, in representative populations after $50$ generations under only a
@@
 \centering
-\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/alt-fig2-onlym-pr}
+%\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/alt-fig2-onlym-pr}
+\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/barh-onlym}
+
 \caption{Proportion of target types, \hl{pragmatic competitor} types, \hl{pragmatic} $\Lall$, and the population's majority type in representative populations after $50$ generations under only a

### 2018-01-15 paper/cogsci-submission2.tex
first full draft for revision and revision letter

@@
   disposition to act on them. Our model traces two evolutionary forces and their interaction:
-  (i) pressure towards communicative efficiency and (ii) learning biases during the transfer of
+  (i) pressure towards communicative efficiency and (ii) transmission perturbations during the acquisition of
   linguistic knowledge. We illustrate the model based on a case study on scalar
-  implicatures.
-\rvv{The relationship between underspecified semantics and pragmatic inference that this study suggests is one of co-evolution.
-The maintenance of such semantics depends on pragmatic reasoning for their enrichment.
-In the opposite direction, complex semantics that can do away with mutual reasoning do not necessarily foster pragmatic reasoning and may even be encumbered by it.}
+  implicatures, which suggests the relationship between underspecified semantics and pragmatic inference to be one of co-evolution.
+The maintenance of underspecified semantics depends on pragmatic reasoning for their enrichment.
+In the opposite direction, complex semantics that can do away with mutual reasoning may not foster pragmatic reasoning, and may even be encumbered by it.
 \end{abstract}
@@
 semantics and pragmatics could have evolved; why it would be so pervasive across natural
-languages; and why it is that some expressions systematically draw from it while others semantically conventionalize. Models of language evolution abound. There are simulation-based models studying
+languages; and why it is that some expressions systematically draw from it while others semantically conventionalize.
+
+Models of language evolution abound. There are simulation-based models studying
 populations of communicating agents
@@
 representations and pragmatic use. In contrast, we will here look at language users with a
-richer linguistic make-up. More precisely, We spell out a model of the co-evolution of conventional meaning and pragmatic reasoning.
+richer linguistic make-up. More precisely, we spell out a model of the evolution of conventional meaning and pragmatic reasoning.
 The objects of replication and selection are pairs consisting of a set of lexical
@@
 Considering the effects that the transmission of
-linguistic knowledge through (iterated) learning have is important because neither semantic meanings nor pragmatic usage
-patterns are directly observable. Instead, language learners have to infer these unobservables -- lexical meanings and manners of pragmatic behavior --
-from the observable behavior in which they result -- overt linguistic choice in a context of interaction. We formalize this process as a form of
+linguistic knowledge through (iterated) learning have is important because neither semantic meaning nor pragmatic usage
+patterns are directly observable. Instead, language learners have to infer these unobservables from the observable behavior in which they result. We formalize this process as a form of
 Bayesian inference. Our approach thereby contains a well-understood model of
@@
 may play a role in preventing the lexicalization of scalar inferences and thereby lead to the emergence of Gricean-like
-pragmatic reasoning types. The results of this case study are critically assessed in the light of
+pragmatic reasoning types. The results of this study are critically assessed in the light of
 the assumptions that feed our model in Section~\ref{sec:discussion}.
@@
 
-\section{A model of co-evolving lexical representations and pragmatic behavior}
+\section{A model of evolving lexical representations and pragmatic behavior}
 \label{sec:model}
@@
 The idea that language is an adaptation to serve a communicative function is fundamental to
-many synchronic and diachronic analyses at least since \citeposs{zipf:1949} explanation of word
+many synchronic and diachronic analyses, at least since \citeposs{zipf:1949} explanation of word
 frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
@@
 extreme (\citealt{kemp+regier:2012,kirby+etal:2015}). A language consisting of a single
-form-meaning association is easy to learn but lacking in expressivity. Conversely, a language
-that lexicalizes a distinct form for a large number of different meanings is highly expressive
+form-meaning association is easy to learn but likely bad for communication with peers. Conversely, a language
+that lexicalizes distinct forms for a large number of different meanings may be better for communication 
 but challenging to acquire.
@@
 \citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}.
-\rvv{The RMD is a population dynamic, which is to say that it describes change in the composition of a community made up of different types of agents.
+The RMD is a population dynamic, which is to say that it describes change in the composition of a community made up of different types of agents.
 The population vector $\vec{x}$ codifies proportions or frequencies of types in a population, with $x_i$ being the frequency of type $i$.
 This population is infinite; its carrying capacity is held constant, $\sum\nolimits_i x_i = 1$.
-This allows us to track change that does not depend on varying population sizes nor their growth rates (but see, e.g., \citealt{nowak:2006} and \citealt[\S 5]{Skyrms2010:Signals} for alternatives that consider, among others, population size and its growth).} In its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$ in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
+This allows us to track change that does not depend on varying population sizes nor their growth rates (but see, e.g., \citealt{nowak:2006} and \citealt[\S 5]{Skyrms2010:Signals} for alternatives that consider, among others, population size and growth).
+In its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$ in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
 $i$ before the update step, (ii) the fitness $f_i$ of each type $i$ before the update, and
@@
 becomes most transparent when we consider an equivalent formulation in terms of a step-wise
-application of the discrete-time replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$
-and its subsequent multiplication with a mutation matrix $Q$:
+application of the discrete-time replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$ and its subsequent multiplication with the mutation matrix $Q$:
 \begin{align}
@@
 In such cases the dynamic reduces to a process of iteration of the transmission bias encoded in $Q$: the rate by which one type changes into another after a generational turnover.
-As detailed below, in this way, the process in (\ref{eq:RMD_discrete}), and its equivalent in (\ref{eq:RMD_discrete_recast}), can contains a model of iterated learning \citep{griffiths+kalish:2005,griffiths+kalish:2007}.
+As detailed below, in this way, the process in (\ref{eq:RMD_discrete}), and its equivalent in (\ref{eq:RMD_discrete_recast}), can contain a model of iterated learning \citep{griffiths+kalish:2005,griffiths+kalish:2007}.
 
@@
   blue line. Rest points, for which $\text{RD}(x)=x$, are at $x=0$, $x=1$ and $x= \nicefrac{2}{3}$.
-  The former are attractors as nearby points converge to them. Points near $x=\nicefrac{2}{3}$
+  The former are attractors; nearby points converge to them. Points near $x=\nicefrac{2}{3}$
   also move towards 0 or 1. This is schematically pictured in the topmost phase portrait in
@@
 
-Moving beyond an abstract example, our goal is to apply the RMD to investigate the co-evolution
+Moving beyond an abstract example, our goal is to apply the RMD to investigate the evolution
 of lexical representations and pragmatic behavior. To do so, we need to fix three things: (i) what the
@@
 make-up.
-More concretely, since we are interested in the evolutionary competition between different lexical representations and ways of using them in communication, a type is here defined as a pair consisting of a lexicon and a pragmatic strategy of language use.
+More concretely, since we are interested in the evolutionary competition between different lexical representations and ways of deploying them in communication, a type is here defined as a pair consisting of a lexicon and a linguistic strategy of language use.
 
@@
 \citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a
-(possibly fuzzy) set of states. A pragmatic behavior specifies a probabilistic speaker
+(possibly fuzzy) set of states. A linguistic behavior specifies a probabilistic speaker
 rule (a probabilistic choice of message for each state) and a probabilistic hearer rule (a
 probabilistic choice of state for each message) given a lexicon.
-There are many ways of making these general notions more concrete.
-Broadly speaking, there are models of pragmatic language use coming from a game-theoretic tradition \citep[e.g.,][]{benz:2006,benz+vRooij:2007,jaeger:2007b,franke+jaeger:2014} and ones from a Bayesian tradition \citep[e.g.,][]{frank+goodman:2012,goodman+stuhlmueller:2013,GoodmanFrank2016:Pragmatic-Langu}.
-We will not focus on the particular design choices that make up each approach \citep{qing+franke:2015}.
+There are many ways of making these general notions concrete.
+There are models of rational language use coming from a game-theoretic tradition \citep[e.g.,][]{benz:2006,benz+vRooij:2007,jaeger:2007b,franke+jaeger:2014} and ones from a Bayesian tradition \citep[e.g.,][]{frank+goodman:2012,goodman+stuhlmueller:2013,GoodmanFrank2016:Pragmatic-Langu}.
+We will not focus on the particular design choices that make up each approach (on this see \citealt{qing+franke:2015}).
 Instead, we will stress how the approach, common to all of these models, of pragmatic language use as {\em rational language use} can be embedded within the replicator mutator dynamic to answer evolutionary questions at the semantics-pragmatics interface.
-Here is what we will assume in the remainder of this paper.\footnote{Readers more familiar with the Bayesian tradition should note that the definitions below follow the game-theoretic tradition in assuming rational sender choice to consist in sending messages that maximize expected utility, rather than message informativity (see \citealt{qing+franke:2015} for detailed discussion).}
+Here is what we will assume in the remainder of this paper.\footnote{Readers more familiar with the Bayesian tradition than the game-theoretic one should note that the definitions below follow the latter in assuming rational sender choice to consist in sending messages that maximize expected utility, rather than message informativity (see \citealt{qing+franke:2015} for discussion).}
 
@@
 following two lexica are minimal examples for the distinction between a lexicalized upper-bound
-for \emph{some} in $\Lbound$ and the widely assumed logical semantics with only a lower-bound
+for \emph{some}, in $\Lbound$, and the widely assumed logical semantics with only a lower-bound,
 in $\Llack$.
@@
 
-Pragmatic strategies define dispositions to produce and interpret messages given a lexicon. We
-distinguish between two kinds of pragmatic strategies. {\em Literal interlocutors} produce and
+Linguistic strategies define dispositions to produce and interpret messages given a lexicon. We
+distinguish between two kinds strategies. {\em Literal interlocutors} produce and
 interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
@@
 
-In words, an agent's linguistic behavior -- what message to send when / which message to interpret how -- is defined by (i) her reasoning level and (ii) her lexicon.
-In what follows, this is what we identify an agent's type with in order to analyze under which conditions combinations of (i) and (ii) emerge through cultural evolution.
+In words, an agent's linguistic behavior -- what message to send when or which message to interpret how -- is defined by (i) her reasoning level and (ii) her lexicon.
+In what follows, this is what we identify an agent's type with, in order to analyze under which conditions combinations of (i) and (ii) emerge through cultural evolution.
 Intuitively, literal behavior (level-$0$ reasoning) results from unreflected language use.
@@
 In a Gricean spirit, pragmatic behavior (level-$n+1$ reasoning) results from mutual reasoning about rational language use.
-Such agents produce and comprehend by reasoning about their interlocutor's likely behavior interpretation/production patterns.
-This reasoning is based on a conjecture resulting from their own behavior if they were in the role of their interlocutor.
+Such agents produce and comprehend by reasoning about their interlocutor's likely interpretation/production patterns.
 Higher order behavior is therefore, even if in tendency rational, quite simple.
-Crucially, we do not assume that agents know each other's types when interacting (e.g., what lexicon the interlocutor is using).
+It does not assume agents to know each other's types when interacting (e.g., what lexicon the interlocutor is using); they simply act based on a conjecture of their own behavior if they were in the role of their interlocutor.
 
@@
 
-\rvv{There are two particularly important things to note.
+There are two particularly important things to note.
 First, in contrast to their literal counterparts of level $0$, pragmatic agents of level $1$ using $\Llack$ associate $\msome$ preferentially with $\ssome$.
@@
 The reason for this is that literal use of $\Lbound$ already endows agents with a behavioral strategy that associates a single state with a single message (in tendency; depending on $\lambda$ for senders).
-For $\Lbound$-receivers of level $1$, reasoning over stochasticity introduced at $S_0$ will generally decrease the association of one state with one message.
+For $\Lbound$-receivers of level $1$, reasoning over stochasticity introduced at $S_0( \cdot \mid \cdot, \Lbound)$ will generally decrease the association of one state with one message.
 This decrease is only slight if $\lambda$ is high, but nevertheless present.
 That is to say, level-$1$ reasoning does not necessarily confer a functional advantage.
-For some types, such as users of $\Lbound$, literal signaling is preferable.}
+For some types, such as users of $\Lbound$, literal signaling is preferable.
 
@@
 
-In words, expected utility $\text{EU}(t_i,t_j)$ quantifies how successful communication between agents of types $i$ and $j$ is, with each type being a combination of a lexicon and a reasoning level (Section~\ref{sec:language+use}).
+In words, expected utility $\text{EU}(t_i,t_j)$ quantifies how successful communication between agents of types $i$ and $j$ is, with each type's behavior resulting from a combination of a lexicon and a reasoning level (Section~\ref{sec:languages+use}).
 Fitness $f_i$ indicates how well type $i$ fares in a population given what other types are found in it: it is type $i$'s weighted expected utility.
@@
 Put differently, types favored by the prior are easier to learn because they require less data than less favored types to be inferred, even if they are both compatible with the learning data at hand.
-
 The extent of the prior's
@@
 
-The way the posterior over types is formed and how learners employ it thereafter are important design choices.
+The way the posterior over types is formed and how learners employ it are important design choices.
 There are other alternatives as well.
-For instance, instead of adopting a single type, learners could keep her uncertainty over types and base her production behavior on the formed posterior \citep{burkett+griffiths:2010}.
-A straightforward operationalization of this idea would then see production defined as a two-step process wherein a speaker first samples from her posterior over types and then produces an utterance based on the type she sampled, as done in \citep{kirby+etal:2015}.
-Beyond added complexity to the model, a disadvantage of this approach is that it could a speaker that first used a Hungarian utterance to speak Quechua in the next, so to speak.
-We stick to the simpler assumption of a single type being acquired, which is defined as follows.
-
-Let $D$ be the set of possible data that learners may be exposed to. This set $D$ contains all
+For instance, instead of adopting a single type, learners could keep their uncertainty over types and base their production behavior on the posterior over types they form \citep{burkett+griffiths:2010}.
+A straightforward operationalization of this idea would then see production defined as a two-step process wherein a speaker first samples from her posterior and then produces utterances based on the type she sampled, as done in \citealt{kirby+etal:2015}.
+Beyond adding complexity to the model, a disadvantage of this approach is that it could happen that a speaker that first uttered a Hungarian utterance speaks Quechua in the next, so to speak.
+We stick to the simpler assumption of a single type being acquired, which we now define.
+
+Let $D$ be the set of possible data that learners may be exposed to. This set contains all
 sequences of state-message pairs of length $k$, e.g.,
@@
 follows: $Q_{ji}$ is the probability that a learner acquires type $i$ when learning from an
-agent of type $j$. The learner observes a length-$k$ sequence $d$ of state-message pairs, but
+agent of type $j$. The learner observes length-$k$ sequences $d$ of state-message pairs, but
 the probability $P(d \mid t_j)$ with which sequence $d = \tuple{\tuple{s_1,m_1},\dots , \tuple{s_k,m_k}}$ is observed depends on type $j$'s
@@
 
-$Q$ is a stochastic $n \times n$ matrix, with $n$ equal to the amount of types.
-The acquisition probability $F(t_i \mid  d)$ given datum $d$ is obtained by probability matching $l = 1$ or a tendency towards choosing the most likely type $l > 1$ from the posterior distribution $P(\cdot \mid d)$ over types given the data, which is calculated by Bayes' rule:
+The acquisition probability $F(t_i \mid  d)$ given datum $d$ is obtained by probability matching $l = 1$ or a tendency towards choosing the most likely type $l > 1$ from the posterior distribution $P(\cdot \mid d)$ over types given the data.
+This is calculated by Bayes' rule:
 \begin{flalign*}
@@
 \subsection{Example}\label{subsec:example}
-\hl{Mention that x,y independent}
-Section~\ref{sec:si-case-study} shows how these components and their interaction can shed light on the evolution of scalar implicatures.
-Before moving on, we will illustrate the model's workings in a constrained type space in order to showcase the effect that pressure for communicative success and learnability have on types.
+The case study in Section~\ref{sec:si-case-study} shows how these components and their interaction can shed light on the evolution of scalar implicatures.
+Before moving on, this section illustrates the model's workings in a constrained type space to better trace the effects that pressure for communicative success and learnability can have on an evolving population.
 To this end, we focus only on the four types from Section~\ref{sec:languages+use}.
-These types are made up of a combination of either lexicon $\Llack$ or $\Lbound$ and either literal (level-$0$) or pragmatic (level-$1$) behavior.
+Recall that these types are made up of a combination of either lexicon $\Llack$ or $\Lbound$ and either literal (level-$0$) or pragmatic (level-$1$) behavior.
 
@@
 The $x$-axis represents the proportion of types with $\Llack$; the $y$-axis that of pragmatic ones.
-Arrows sketch the directionality of population trajectories through the space under functional pressure (replicator steps) and learnability pressure (mutator steps), both with a flat prior and one that favors $\Llack$.}
+Arrows sketch population trajectories under functional pressure (replicator steps) and learnability pressure (mutator steps) with either a flat prior or one that favors $\Llack$.}
 \label{fig:sketch}
@@
 A population will be made up of frequencies of these four types.
-We may think about such a population as a point in a two-dimensional space.
-In Fig.~\ref{fig:sketch} this space is outlined by the grey dotted square.
-The first dimension determines the proportion of types that lexicalize ambiguous $\msome$ (the $x$-axis in Fig.~\ref{fig:sketch}).
+We may think of such a population as a point in a two-dimensional space.
+The grey dotted square in Fig.~\ref{fig:sketch} outlines this space.
+The space's first dimension determines the proportion of types that lexicalize ambiguous $\msome$ (the $x$-axis in Fig.~\ref{fig:sketch}).
 That is, the proportion of types with $\Llack$.
 The second dimension determines the proportion of pragmatic language users (the $y$-axis in Fig.~\ref{fig:sketch}).
-Populations at the either corner of this space are monomorphic, consisting purely of a single type.
-If we move from away from edges toward the center of the space, then this two-dimensional representation underspecifies the proportion of types found in a population.
-The space's center, for instance, could be composed 50\% of pragmatic $\Llack$ and of an equal proportion of literal $\Lbound$; but it could equally well be a population that has each type with the same frequency.
-This will be sufficient for our purposes, as we are only interested in showcasing the rough trajectories of populations through this space after replicator or mutator steps.
-%The space's corners (and edges) are unequivocal: the lower-left corner corresponds to an population of only literal $\Lbound$ users; the lower-right corner to all pragmatic $\Lbound$ users; the upper-left to literal $\Llack$ users only; and the upper-right corner to the majority all pragmatic $\Llack$ users.
+Populations at the corners of this space are monomorphic; all agents in them are of a single type.
+If we move from away from edges toward the center of the space, then this two-dimensional representation underspecifies the exact proportions of types found in a population.
+The space's center, for instance, could be composed 50\% of pragmatic $\Llack$ and of an equal proportion of literal $\Lbound$.
+But it could equally well be a population that has each type with the same frequency.
+This level of granularity will nevertheless be sufficient for our purposes, as we are only interested in sketching rough trajectories after replicator or mutator steps.
 
 From Section~\ref{sec:languages+use}, we already know some things about this type space.
-When it comes to communicative success, types with $\Lbound$ have a functional advantage over $\Llack$; literal $\Lbound$ has a slight advantage over pragmatic $\Lbound$; all have a sizable advantage over literal $\Llack$.
+When it comes to communicative success: types with $\Lbound$ have a functional advantage over $\Llack$; literal $\Lbound$ has a slight advantage over pragmatic $\Lbound$; all have a sizable advantage over literal $\Llack$.
 The red arrows in Fig.~\ref{fig:sketch} sketch out these relationships.
-They depict the rough trajectories we may expect if there is only pressure for communicative success.
-How much each replicator step moves a population ultimately depends on the rationality parameter $\lambda$.
-In particular, functional differences decrease as $\lambda$ increases: if $\lambda$ is sufficiently high, all types except for literal $\Llack$ speakers will show a strong tendency to associate $\ssome$ with $\msome$ and $\sall$ with $\mall$.
-As a consequence, when it comes to learnability and transmission fidelity, the likelihood that any of the three remaining types produces a given datum $d$ will be approximately equal.
-In other words, the likelihood offers little differentiation for learners if $\lambda$ is high.
-We can therefore expect that, if the prior is flat, no type is passed on very faithfully.
+They depict the rough trajectories we can expect if there is only pressure for communicative success.
+How much each replicator step moves a population through this space ultimately depends on the rationality parameter $\lambda$.
+In particular, note that functional differences decrease as $\lambda$ increases: if $\lambda$ is sufficiently high, all types except for literal $\Llack$ speakers will show a strong tendency to associate $\ssome$ with $\msome$ and $\sall$ with $\mall$.
+As a consequence, when it comes to learnability and transmission fidelity, the likelihood that any of these three types produces a given datum $d$ will be approximately equal.
+In other words, if $\lambda$ is high the likelihood offers little differentiation between pragmatic $\Llack$ and any disposition to use $\Lbound$ for learners.
+We can therefore expect that, if the prior is flat, no type is passed on very faithfully in this space.
 This would lead populations to gravitate toward the space's center after mutator steps.
 This is sketched by the blue dash-dotted arrows in Fig.~\ref{fig:sketch}.
-Neither prior nor posterior alone, but their combination, are what determines the effects of (iterated) Bayesian learning.
+Neither prior nor likelihood alone, but their combination, are what determines the effects of (iterated) Bayesian learning.
 If, instead of a flat prior, there is a prior that favors, say, $\Llack$ over $\Lbound$ we can expect mutator steps to be skewed toward the upper-half of the space.
 This is suggested by the purple arrows in Fig.~\ref{fig:sketch}.
-Taking stock, the effects of iterated learning will depend not only on $\lambda$, affecting the likelihood, but also on the prior, how strong the bias is, and the degree to which learners tend to maximize the posterior, controlled by parameter $l$.
-
-Fig.~\ref{fig:quiver} shows actual trajectories, showcasing how they are modulated by parameter values.
-The replicator steps make the functional disadvantage of $\Llack$ against $\Lbound$ apparent, particularly for literal types and/or low $\lambda$.
-They also highlight that the functional difference between literal and pragmatic $\Lbound$ is small.
-Under functional pressure only, populations move toward the lower part of the square but the steps become small as the edge, representing populations of only $\Llack$, are reached.
-As for the mutator diagrams, there are two main things to note.
-As expected, a small bias for a type will lead to larger proportions to adopt it provided that likelihood offers little to no differentiation between types.
-This is the situation the right-most space in Fig.~\ref{fig:quiver} depicts.
-However, the bias is far from absolute.
-If, instead, certain types are more likely to be recovered based on the linguistic input they produce, they will be passed on more faithfully.
-This is why, for instance, populations move toward a majority of $\Llack$-types for $\lambda = 5$ and $l = 10$.
+Taking stock, the effects of iterated learning will depend not only on $\lambda$, affecting the likelihood, but also on the prior (how strong of a bias there is) and the degree to which learners tend to maximize the posterior (controlled by parameter $l$).
 
@@
 The $x$-axis represents the proportion of types with $\Llack$; the $y$-axis that of pragmatic ones.
-Arrows indicate directionality of trajectories after (a) replicator steps and (b) mutator steps with a prior that favors $\Llack$ by a factor of $1.05$.
-Colored contours show the proportion of pragmatic $\Llack$ users after a step \tb{I don't think the contours are adding too much, so they could be removed.
-Then again, I think they are rather non-intrusive and look nice.}}
+Arrows indicate directionality of trajectories after (a) replicator steps and (b) mutator steps with a prior that favors types with $\Llack$ by a factor of $1.05$ over those with $\Lbound$.
+Colored contours show the proportion of pragmatic $\Llack$ users after a step \tb{I don't think the contours are adding much, so they could be removed.
+Then again, I think they are rather non-intrusive, look nice, and illustrate how the step-size, depending on where we are in the space, varies.}}
 
@@
 
+Fig.~\ref{fig:quiver} shows actual evolutionary trajectories in this type space.
+The replicator steps make the functional disadvantage of $\Llack$ against $\Lbound$ apparent, particularly for its literal users and/or low $\lambda$.
+They also highlight that the functional difference between literal and pragmatic $\Lbound$ is small.
+Under functional pressure only, populations move toward the lower part of the square but the steps become small as the lower edge, representing populations composed only  of $\Llack$ users, are reached.
+As for the mutator diagrams, there are two main things to note.
+As expected, a small bias for a type will lead to larger proportions to adopt it, provided that likelihood offers little to no differentiation between types.
+This is the situation the right-most space in Fig.~\ref{fig:quiver} depicts.
+However, the bias is far from absolute.
+If, instead, certain types are more likely to be recovered based on the linguistic input they produce, they will be passed on more faithfully.
+This is why, for instance, populations move toward a majority of $\Llack$-types for $\lambda = 5$ and $l = 10$.
+
 Fig.~\ref{fig:sketch} and \ref{fig:quiver} depict replicator and mutator steps in isolation to better showcase their effects.
-Their combination will expectedly yield different outcomes, which we now turn to in a case study on scalar implicatures.
+Their combination will yield different outcomes, which we now turn to in a case study on scalar implicatures.
+\tb{Also: In this type space RMD way too similar to MD to make it worth the ink.
+It raises more questions than it answers so I left it out as the plots look identical to those in Fig.~3b}
 
@@
 (which may be unrelated to the expected utility when communicating with a given population);
-$k$ is length of observations for each language learner; $l$ regulates where the learners'
+$k$ is the length of observations received by each language learner; $l$ regulates where the learners'
 inference behavior lies on a spectrum from probability matching to acquisition of the most likely
@@
 for some arbitrary fixed predicates $A$ and $B$.
-Enlarging the setup from Section~\ref{sec:languages+use} by another state will enable us to discern the difference between three kinds of lexica we will use to unravel the model's dynamics (see below).
+Enlarging the setup from Section~\ref{sec:languages+use} by another state in this way enables us to discern the difference between three kinds of lexica which we will use to unravel the model's dynamics (see below).
 Eight lexical representations can be distinguished based on their truth or falsity in three world states, six of which are not contradictory or tautological (see Table~\ref{tab:concepts} below).
@@
 Lexicon $\Lall$ is clearly bad for communication: all message and interpretation choices will
-be equally likely for all types; no information about the observed world state will be conveyed
+be equally likely; no information about the observed world state will be conveyed
 by its users. In contrast, users of $\Lbound$ can communicate world states perfectly, no matter
@@
 model's predictions in populations of types with one of the two behaviors introduced earlier:
-literal or pragmatic. The former correspond to level-$0$ reasoners and the latter to level $1$. Accordingly, there is a total of $432$ types. Six types are variants of pragmatic language
+literal or pragmatic. The former correspond to level-$0$ reasoners and the latter to level $1$. Accordingly, there is a total of $432$ types. Six types are pragmatic language
 users with $\Llack$-like lexica. We refer to these as \emph{target types} because they
@@
 $\Lbound$ kind. We refer to these as \emph{competitor types}, because they are expected to be
-the target types' main contenders in evolutionary competition. Finally, note that while
+the target types' main contenders in evolutionary competition.
+
+Note that while
 different types may lexicalize the same representations, they may nevertheless map different
@@
 Apart from simplicity, prominent examples include mutual exclusivity \citep{merriman+bowman:1989,clark:2009}, regularization \citep{hudson+etal:2005}, and generalization \citep{smith:2011}.
-Even viewed in isolation, there are still many ways to define priors over lexica (see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a, kirby+etal:2015}).
-The key assumption here, common to all ways to define simplicity-biased priors, is that simple representational expressions should be favored over more complex ones.
-However, in view of the menu of possible operationalizations, other biases and their potential interplay, we should stress that these details -- from the generative grammar to its complexity measure -- are to be regarded as one convenient operationalization of one general approach to explicating learning biases; this is not a commitment that this general approach is necessarily superior or that, within it, this particular instrumentalization is the single most plausible.
-We return to this issue in relation to our results in Section~\ref{sec:an-induct-learn}.
+Even when these biases are considered in isolation, there are many ways in which they can be translated into priors over types.
+The key assumption here, common to all simplicity-biased priors, is that simple representational expressions should be favored over more complex ones (see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a, kirby+etal:2015}).
+In view of the menu of possible operationalizations, other biases and their potential interplay, we should stress that these details -- from the generative grammar to its complexity measure -- are to be regarded as one convenient operationalization of one general approach to explicating learning biases; this is not a commitment that this general approach is necessarily superior or that, within it, this particular instrumentalization is the single most plausible.
+We return to this issue in Section~\ref{sec:discussion}.
 
@@
 \label{sec:simulation-results}
-\hl{Change plots and captions from prag. Llackers to lit. Llackers}
-
-\hl{The figures (especially Figure 4, but the others as well) are difficult to read, because of the changing scales.
-I would recommend finding a way to share the same scale between subfigures, or make it visually clearer that the subfigures cannot be directly compared.}
-
-Recall that there are three parameters, which we hold constant across types: soft-max parameter $\lambda$ affects how strongly
+Recall that there are three parameters held constant across types: soft-max parameter $\lambda$ affects how strongly
 speakers favor messages that appear best from their subjective point of view; the
@@
 Still, lexica of the kind $\Lall$ are in turn even more likely \emph{a priori} than lexica of the kind $\Llack$.
-As showcased in Section~\ref{subsec:example}, the prior is not all that counts for learnability.
+In any case, as showcased in Section~\ref{subsec:example}, the prior is not all that counts for learnability.
 Simulation results will shed light on the question whether target
@@
 \includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/barh-onlyr}
-\caption{Proportion of target types, \hl{pragmatic} competitor types, \hl{pragmatic} $\Lall$, and the population's majority type, in representative populations after $50$ generations under only a
-  pressure for communicative success.
-\hl{Flip plots, change plot generation to show the highest 6 Lbound users and the highest Lall, irrespective of ling.
-behavior, change text surrounding these things}}
+\caption{Proportion of target types, the six competitor types with the highest proportion, the highest type of the $\Lall$-kind, and the population's majority type, in representative populations after $50$ generations under only a
+  pressure for communicative success.}
 \label{fig:only-R}
@@
 showcased in Fig.
-\ref{fig:only-R}, which shows the proportion of target types, \hl{pragmatic} competitor types and \hl{pragmatic} $\Lall$ in a representative population after $50$ replicator steps.
+\ref{fig:only-R}, which shows the proportion of target types, the highest competitor types and the highest type with $\Lall$ in a representative population after $50$ replicator steps.
 The plot also indicates the proportion of the
-\emph{majority type}: the type with the highest proportion in the final population. With
+\emph{majority type}: the type with the highest proportion in the population. With
 low $\lambda$ many types have very similar behavior, so that evolutionary selection lacks grip
@@
 cases in which the target type was the majority type, corresponding to a mean proportion of
-$.003$ across populations. By contrast, in $913$ cases the majority types had $\Lbound$ with
+$.003$ across populations. By contrast, in $913$ cases the majority types had $\Lbound$ with,
 close to an even share between literal ($454$) and pragmatic types ($459$), corresponding to a
@@
 
-\subsubsection{Iterated learning only: transmission fidelity}
+\subsubsection{Iterated learning only: transmission fidelity and learnability}
 
@@
 
-\caption{Proportion of target types, \hl{pragmatic competitor} types, \hl{pragmatic} $\Lall$, and the population's majority type in representative populations after $50$ generations under only a
+\caption{Proportion of target types, the six competitor types with the highest proportion, the highest type of the $\Lall$-kind, and the population's majority type in representative populations after $50$ generations under only a
   pressure for learnability ($\lambda = 20, k = 5$). The learning prior is shown in the
@@
 reasoning and sufficiently high $\lambda$. This makes it less likely to be confused with
-other types if the learning data is not too sparse ($k \geq 5$). Put differently, learners have
-a higher propensity to infer pragmatic $\Llack$ when the teacher's type produces very similar
-data, such as when using $\Lbound$ (see Section~\ref{subsec:example} and Fig.~\ref{fig:sketch}). Moreover, $\Llack$ is less likely to be confused with
+other types if the learning data is not too sparse ($k \geq 5$). What is more, learners have
+a propensity to infer pragmatic $\Llack$ when the teacher's type produces very similar
+data, such as when using $\Lbound$ (see Section~\ref{subsec:example}). Moreover, $\Llack$ is less likely to be confused with
 types with different observable behavior because its pragmatic use approximates a one-to-one
@@
 of $1000$ independent populations had users of $\Lall$ as majority types.
-For higher $\lambda$, it becomes clear that a high prior (e.g., $\Lall$) is not the only thing that counts when it comes to learnability (see Section~\ref{subsec:example}).
-As soon as there is information for learners to discern whether one type is more likely to have generated the data (depending on $\lambda$ and $k$), it becomes paramount for types to produce data that makes them easily identifiable if they are to be transmitted faithfully.
-This is to say that the learning prior is important but far from being the only element that regulates how iterated learning plays out.
+For higher $\lambda$, it becomes clear that a high prior is not the only thing that counts when it comes to learnability (see Section~\ref{subsec:example}).
+As soon as there is information for learners to discern whether one type is more likely to have generated the data (depending on $\lambda$ and $k$), it becomes paramount for types to produce data that makes them easily identifiable if they are to be inferred more often.
 
@@
 \includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/alt-fig3-r+m}
-\caption{Proportion of target types, \hl{pragmatic} competitor types, \hl{pragmatic} $\Lall$, and the population's majority type in representative populations after $50$ generations under both pressures ($k = 5$).}
+\caption{Proportion of target types, the six competitor types with the highest proportion, the highest type of the $\Lall$-kind, and the population's majority type in representative populations after $50$ generations under both pressures ($k = 5$).}
 \label{fig:rmd}
@@
 $l$. Pressure for communicative success magnifies the effects of iterated learning and dampens
-the proliferation of multiple variants of a kind that are equal in expressivity {\em and} learnability.
-A pressure towards learnability favors the transmission of simpler semantics and thereby selects for pragmatic language use in these types.
-Pragmatic behavior, however, is not selected in types that can do away without pragmatic reasoning such as $\Lbound$ users (see Section~\ref{subsec:example}).
+the proliferation of multiple types of a kind, equal in expressivity and learnability.
+A pressure towards learnability favors the transmission of simpler semantics and thereby indirectly selects for pragmatic language use in these types.
+Pragmatic behavior, however, is not necessarily selected for in types that can do away without pragmatic reasoning such as $\Lbound$ users (see Section~\ref{subsec:example}).
 
@@
 When pressured for communicative success only, target types are outcompeted by competitor types.
-When pressured for learnability, populations are polymorphic due to lack of competition; minimally among variants of a kind.
-However, due to the prior, learners that are exposed to the input of either targets or competitors tend to infer acquire the former type.
+When pressured for learnability, populations are polymorphic due to lack of competition; minimally among types of a kind.
+However, due to the prior, learners that are exposed to the input of either targets or competitors tend to acquire the former.
 The consequence of combining both pressures is then that the (slight) functional disadvantage of targets is counterbalanced by an advantage in learnability.
@@
 \citep{bergen+etal:2012,bergen+etal:2016}. This allows for a conceptual investigation of the
-co-evolution of conventional meaning and pragmatic language use. Main contributions of the
+(co-)evolution of conventional meaning and pragmatic language use. Main contributions of the
 model are (i) its modular separation of communicative success and learnability on evolutionary
@@
 between a lexicon's expressivity and its learnability. A main difference is that here we
-considered a pressure for mutual understanding within a generation.
-More precisely, expected utility.
+considered communicative pressure for mutual understanding.
 This pressure can indirectly select for expressive types, those that can convey states unequivocally, whereas Kirby et al.
@@
 Lastly, chains of iterated learning alone do not put types in direct competition.
-Accordingly, learning can lead to polymorphous populations in which multiple variants of a type coexist \citep{nowak:2006}.
+Accordingly, learning leads to polymorphous populations in which multiple types of a kind coexist \citep{nowak:2006}.
 
@@
 preferred or dispreferred representations can be adaptive. This is why a major contribution of this
-paper is the arrangement of many different ingredients into a joined model of the co-evolution
+paper is the arrangement of many different ingredients into a joined model of the (co-)evolution
 of lexical meaning and pragmatic use.

### 2018-01-15 paper/rv-letter1.tex
first full draft for revision and revision letter

@@
 This is an important point that we now stress and highlight more throughout the paper.
-In particular the discussion surrounding Figure $2$ and $3$ in the new Section~2.4 makes precise how one factor has an impact on the evolution of the other.
+In particular the material surrounding Figure $2$ and $3$ in the new Section~2.4 illustrates how one factor has an impact on the evolution of the other.
 In a nutshell: a pragmatic strategy (level-$1$ reasoning) favors the evolution of underspecified semantics of the target kind ($L_{\text{lack}}$) as it allows for the maintenance of simpler lexical representations that are easier to learn without functional disadvantages otherwise incurred in communication (e.g., by level-$0$ reasoners).
-In the other direction, underspecified semantics of the target kind favor the evolution of a pragmatic disposition to act on them.
+In the other direction, underspecified semantics of the target kind also favors the evolution of a pragmatic disposition to act on them.
 When paired with other semantics, pragmatic refinement can instead be (close to a) neutral trait (e.g., $L_{\text{bound}}$ paired with high $\lambda$) or it can even lead to  functional disadvantages when compared with level-$0$ reasoning (e.g., $L_{\text{bound}}$ with low $\lambda$).
+In this way both features benefit from each other.
 
@@
 As we discuss in Section~1, our case study focuses on scalar implicatures because they are well-studied systematic pragmatic inferences that have received much attention, both theoretically and experimentally.
-This makes them suitable candidates for the study of the evolution of regular pragmatic inferences because there are many findings that can guide the application and evaluation of our model to the study of these inferences.
-This ties in with our particular choices to model pragmatic language use of scalar expressions.
+This makes them suitable candidates for the study of the evolution of regular pragmatic inferences; past research can guide our modeling choices and aid in evaluating our application to the study of these inferences.
+This ties in with our particular choices to model the pragmatic use of scalar expressions.
 As we now make more explicit in \S 2.3.1, these draw from a growing game-theoretic and Bayesian tradition \citep[e.g.,][]{franke:2009,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}.
@@
 As noted earlier, we have taken care to make the exposition more accessible and to guide the reader more through the technical details of Section 2.
-In particular, the newly added Section 2.4 should help unravel the model.
-We took the following questions, briefly answered below, as guidance for changes made in Section~2.
+In particular, the newly added Section 2.4 should help readers unravel the model.
+We used the following questions from Reviewer 1, briefly answered below, to guide the changes made in Section~2.
 
@@
 The amount of teachers of a given type is proportional to the frequency of this type in the population after the replicator step.
-The population is infinite so there are infinite learners/teachers.
+The population is infinite so there are technically infinite learners/teachers.
 This was implicit in the definitions of the replicator mutator dynamic in (1) and (2), and is now explicitly stated in Section~2.2 in the prose that motivates and explicates these definitions.
@@
 This allows us to track change that does not depend on varying population sizes nor their growth rate.
-This is why we speak of proportions of a type, rather than of there being a particular number of agents of a type.
+This is why we speak of proportions of a type in a population, rather than of there being a particular number of agents of a type.
 
@@
 \end{mdframed}
-While our model is well-compatible with the assumption that every agent comes with her own parameter values, the reviewer is correct in pointing out that we assume they all to have the same values (explicit in \S3.2).
-Variation across agents would mean many more types: one for each combination of a possible lexicon, pragmatic strategy, $\lambda$-, $l$- and $k$-value.
+While our model is compatible with the assumption that every agent comes with her own parameter values, the reviewer is correct in pointing out that we assume they all to have the same values (explicit in \S3.2).
+Variation across agents would mean having many more types: one for each combination of a possible lexicon, pragmatic strategy, $\lambda$-, $l$- and $k$-value.
 This would allow us to trace the evolutionary trajectory of not only lexica and pragmatic strategies, which is our goal here, but of all these factors combined.
@@
 A learner faced with input $d$ (see above) calculates $P(\tau \mid d)$, the probability that a type $\tau$ produced input $d$.
-Crucially, neither pragmatic strategies nor lexical meanings are directly observable, so they can only be faithfully recovered inasmuch as the {\em overt} behavior evidenced by a type is (in tendency) attributable to only one, {\em covert}, pairing of a lexicon and a pragmatic strategy.
+Crucially, neither linguistic strategies nor lexical meanings are directly observable, so they can only be faithfully recovered inasmuch as the {\em overt} behavior evidenced by a type -- the data it produces -- is (in tendency) attributable to only one, {\em covert}, pairing of a lexicon and a linguistic strategy.
 
@@
 This may involve the recruitment of contextual information, as the reviewer notes, but can also involve pragmatic enrichments that result purely from reasoning about linguistic alternatives (``why did the speaker say {\em some} instead of {\em all}'').
-It is in the latter sense that reasoning beyond level $0$ is pragmatic (see \S2.31.).
-Classically, theoretical analyses of scalar implicatures put their emphasis on reasoning about linguistic choice rather than on contextual information \citep[e.g.,][]{horn:1972,gazdar:1979, franke:2009, GoodmanStuhlmuller2013:Knowledge-and-I}.
-We chose to do the same.
+It is in the latter sense that reasoning beyond level $0$ is pragmatic (see \S2.3.1).
+Classically, theoretical analyses of scalar implicatures have put their emphasis on reasoning about linguistic choice rather than on contextual information \citep[e.g.,][]{horn:1972,gazdar:1979, franke:2009, GoodmanStuhlmuller2013:Knowledge-and-I}.
+We chose to do the same here.
 
@@
 Speaker behavior is relevant both for communication with hearers as well as for the learning input that a teacher produces.
-When considering only iterated Bayesian learning $\lambda$ accordingly also plays a role because different $\lambda$ values yield different likelihoods of producing data (see above and Section~2.3.3).
+When considering only iterated Bayesian learning $\lambda$ accordingly also plays a role because different $\lambda$ values yield different likelihoods of producing particular data (see above and Section~2.3.3).
 
@@
 
-We do not believe that, ultimately, we should assume the non-parsimonious view that there are different complexity measures for different structures.
-However, with previous work in iterated learning tradition, we do assume that (i) inductive learning biases can play a role in shaping culturally transmitted knowledge and that, as the reviewer notes, (ii) biases can be of manifold nature; another well-studied example being the mutual exclusivity bias, which could be expected to work against targets and in favor of competitors.
+We do not believe that, ultimately, the non-parsimonious view that there are different complexity measures for different structures should be assumed.
+However, with previous work in the iterated learning tradition, we do assume that (i) inductive learning biases can play a role in shaping culturally transmitted knowledge and that, as the reviewer notes, (ii) biases can be of manifold nature; another well-studied example being the mutual exclusivity bias, which could be expected to work against targets and in favor of competitors.
 
@@
 By assumption, this translates to a preference not to lexicalize linguistic material over lexicalizing it; simpler formulae over complex ones.
-As mentioned in Section 3.1.2, we acknowledge that the complexity measure that the prior is based on  is just a convenient operationalization of this particular hypothesis.
+As mentioned in Section 3.1.2, the complexity measure that the prior is based on is just a convenient operationalization of this particular hypothesis.
 
-Lastly, we should stress that, just as models of iterated learning make different predictions under the assumption of different inductive biases, so does our model.
+We should stress that, just as models of iterated learning make different predictions under the assumption of different inductive biases, so does our model.
 The assumption and operationalization of a particular bias should, of course, always be seen critically (Section 4).
@@
 %
-We understand transmission fidelity as the transition probability of acquiring type $i$ when learning from type $j$ (Section~2.3.3).
+We understand transmission fidelity as the probability of acquiring type $i$ when learning from type $i$ (Section~2.3.3).
 This is the value of a cell in the learning matrix $Q$.
-In this case, $Q_{ji}$.
-As discussed in Section~2.2, if $Q_{ji} = 1$ then $i$ is always acquired when learning from type $j$.
-On the other extreme, $Q_{ji} = 0$ means that $i$ is never acquired from $j$.
+In this case, $Q_{ii}$.
+As discussed in Section~2.2, if $Q_{ii} = 1$ then $i$ is always acquired when learning from type $i$.
+On the other extreme, $Q_{ii} = 0$ means that $i$ is never acquired from $i$.
 $Q$ is a stochastic matrix and we we usually see neither extreme but rather, as suggested by the reviewer, a graded notion of transmission fidelity (Section~2.2).
-The details on how $Q_{ji}$ is computed are given in Section 2.3.3.
-We have expanded these sections to make these details clearer in the prose.
+The details on how $Q$ is computed are given in Section 2.3.3.
+We have expanded this section to make these details clearer in the prose.
 %
@@
 
-We have thoroughly revised the article to make the contrast between these three predictions clearer (see Section~3.2, particularly 3.2.3; but also Section~2.4, which illustrates how the dynamics behave in a smaller type space).
+We have thoroughly revised the manuscript to make the contrast between these three predictions clearer (see Section~3.2, particularly 3.2.3; but also Section~2.4, which illustrates how the dynamics behave in a smaller type space).
 These questions tie in with Reviewer $1$'s request for a clearer exposition and the explanation of why we speak of co-evolution in this case (reviewer 1, comment 1).
-In a nutshell, pressure for learnability alone does not put the types found in a population in competition.
-Instead, the population comes to be inhabited by types that are faithfully transmitted -- of which there might be many (e.g., all variants of a kind are equally learnable; if there are many variants of a type that is, in tendency, faithfully transmitted the population will stay polymorphic).
+In a nutshell, pressure for learnability alone does not put the types in a population in competition.
+Instead, the population comes to be inhabited by types that are recovered from learning input of replicated types more often -- of which there might be many (e.g., all variants of a kind are equally learnable; if there are many variants of a type that is, in tendency, acquired more often, then the population will stay polymorphic).
 What fitness-relative selection adds to this process is the competition that learnability alone lacks, independent of the type space we look at.
-Now, in the case of our type space we have a slight (modulo $\lambda$) communicative disadvantage for targets  when in direct competition with competitors, for instance.
-However, the fact that targets are inferred more faithfully by na\"ive learners leads (i) to the gradual existence of less competitors, counteracting this functional disadvantage, as well as (ii) to the existence of less variants of a kind (contra the {\em learning alone} dynamic).
+Now, in the case of our type space we have a slight (modulo $\lambda$) communicative disadvantage for targets relative to competitors, for instance.
+However, the fact that targets are inferred more more often by na\"ive learners, also from those that learn from competitors, leads (i) to the gradual existence of less competitors, counteracting this functional disadvantage, as well as (ii) to the existence of only one variant of a kind (contra what we get in the {\em learning alone} condition).
 %
@@
 As noted above, we have expanded our analysis of the model and put much more emphasis on how the components that feed it play a role in driving particular evolutionary outcomes.
-See, in particular, \hl{sections this and that}.
+See, in particular, Section 2.4, Section 3.2, and Section 4.
 %
@@
 
-It is not necessarily true that it will always be rational for agents to behave pragmatically (see the end of Section~2.3.1 but also the discussion surrounding Figure~3a).
+It is not necessarily true that it will always be advantageous for agents to behave pragmatically (see the end of Section~2.3.1 but also the discussion surrounding Figure~3a).
 The main thing to note is that a pragmatic level-$1$ hearer using lexicon $L_{\text{bound}}$ will reason about the behavior of a {\em soft-maximizing} level-$0$ speaker of $L_{\text{bound}}$.
-The lower $\lambda$, the more the stochasticity percolates to level-$1$ pragmatic interpretation.
+The lower $\lambda$, the more the stochasticity percolates from the level-$0$ speaker to level-$1$ pragmatic interpretation.
 By contrast, literal level-$0$ hearers of $L_{\text{bound}}$ have a one-to-one form-meaning mapping from the get go.
@@
 %
-The paper assumes that the speaker (and listener) know the type of the player that they're playing against.
+The paper assumes that the speaker (and listener) know the type of the player that they're interacting with.
 This does not seem like a natural assumption.
@@
 For example, the linguistic choices of a level-$1$ speaker of lexicon $L_{\text{some}}$ do not change depending on whom she interacts with.
-They are always defined as (boundedly) rational choice relative to the interpretative behavior of a level-$0$ user of $L_{\text{some}}$ (see (3) --(6) in \S2.3.1).
+They are always defined as (boundedly) rational choice relative to the interpretative behavior of a level-$0$ user of $L_{\text{some}}$ (see definitions (3) --(6) in \S2.3.1).
 %
@@
 \end{mdframed}
-There are many alternatives when it comes to definitions of rational language use, many of which do not use an information-theoretic utility function (see \citealt{qing+franke:2015} for an overview and discussion up to $2015$).
+There are many alternatives when it comes to definitions of rational language use.
+The definitions that draw more strongly from a game-theoretic tradition standardly do not make use of an information-theoretic utility function (see \citealt{qing+franke:2015} for an overview and discussion up to $2015$).
 As noted in our answer to Reviewer 1's comment 2, we have taken care to highlight that our choices draw from this literature and to mention that there are other alternatives in the same spirit.
@@
 This is an important point.
-We agree that some fragments of our previous manuscript could lead to confusion, e.g., as to why $L_{\text{all}}$, being the {\em a priori} more likely type to be inferred, did not come out strongly even when considering learnability alone.
-Throughout the article, we now stress more that what matters is the posterior and explain the role of the likelihood more.
+We agree that some fragments of our previous manuscript could lead to confusion, e.g., as to why $L_{\text{all}}$, being the {\em a priori} more likely type to be inferred, did not come out strongly even in the {\em learning only} condition.
+Throughout the article, we now stress more that what matters is the posterior, and put emphasis on the role of the likelihood.
 
@@
 
-This is a standard assumption in the literature (lacking reasons to believe that there is an asymmetry in how often, on average, agents use and hear a type of expression.) In the case of scalar implicatures, we see no reason to assume such an asymmetric contribution to fitness.
+This is a standard assumption in the literature (lacking reasons to believe that there is an asymmetry in how often, on average, agents use and hear a type of expression.) In the case of scalar implicatures, we see no reason to assume such an asymmetric contribution.
 As for concrete impact: The main contrast that we are interested in (competition between targets and competitors) does not hinge on agents being speakers half of the time.
-However, e.g., the contrast between pragmatic and literal competitor types does as they differ only in their receiver behaviors.
+However, e.g., the contrast between pragmatic and literal competitor types does, as they differ only in their receiver behaviors.
 Assuming that receiver behavior contributes differently than sender behavior would therefore modulate differences between these types.
@@
 In particular, we wanted to inspect a type space where targets are not the most likely {\em a priori} as this may mislead readers into thinking that learnability alone, in particular the prior alone, drives the outcome.
-This is related to our answer to reviewer 2, comment 7.
-The reason why we did not go beyond $n>2$ is for computational tractability, as calculating $Q$, even if approximated, is expensive.
+A larger type space also better allows us to showcase that multiple kinds of a type may exist and how this affects evolutionary outcomes.
+The reason why we did not go beyond $3$ messages is for computational tractability.
+Calculating $Q$, even if approximated, is expensive.
 The question whether increasing the type space will decrease the frequency to which we expect to see pragmatic {\em some} users is interesting.
-It now figures in \hl{Section~X.X}.
 At present, we have no answer.
@@
 
-We have expanded this section to further motivate our choices. Relatedly, see our answer to Reviewer 1's comment 4 above.
+We have expanded this section to further motivate our choices (see also the discussion in Section~4). Relatedly, see our answer to Reviewer 1's comment 4 above.
 

### 2018-03-08 paper/cogsci-submission2.tex
minimal edits resubmission

@@
 that the speaker does not own all of them. This is because, if the speaker instead owned them all, she
-could have used the word \emph{all} instead of \emph{some} in his utterance, thereby making a
+could have used the word \emph{all} instead of \emph{some} in her utterance, thereby making a
 more informative statement. Scalar implicatures, especially the inference from \emph{some} to
@@
 \subsection{The replicator mutator dynamic}
-An elegant formal approach to capture the interaction between communicative efficiency and learnability is
-the \emph{replicator mutator dynamic}
-\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}.
-The RMD is a population dynamic, which is to say that it describes change in the composition of a community made up of different types of agents.
-The population vector $\vec{x}$ codifies proportions or frequencies of types in a population, with $x_i$ being the frequency of type $i$.
-This population is infinite; its carrying capacity is held constant, $\sum\nolimits_i x_i = 1$.
-This allows us to track change that does not depend on varying population sizes nor their growth rates (but see, e.g., \citealt{nowak:2006} and \citealt[\S 5]{Skyrms2010:Signals} for alternatives that consider, among others, population size and growth).
-In its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$ in a population at the next time step as a function of: (i) the frequency $x_i$ of each type
-$i$ before the update step, (ii) the fitness $f_i$ of each type $i$ before the update, and
-(iii) the probability $Q_{ji}$ that an agent who observes the behavior of 
-an agent with type $j$ ends up acquiring type $i$:
+An elegant formal approach to capture the interaction between communicative efficiency and
+learnability is the \emph{replicator mutator dynamic}
+\citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. The
+RMD is a population dynamic, which is to say that it describes change in the composition of a
+community made up of different types of agents. The population vector $\vec{x}$ codifies
+relative frequencies of types in a population, where $x_i$ is the relative frequency of type
+$i$ and $\sum\nolimits_i x_i = 1$. The dynamic assumes a virtually infinite population, so that
+it is able to abstract away from what individual agents do and how they change their behavior
+over time. The result is an abstract description of the most probable trajectory of change in
+$\vec{x}$. In its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of
+each type $i$ in a population at the next time step as a function of: (i) the frequency $x_i$
+of each type $i$ before the update step, (ii) the fitness $f_i$ of each type $i$ before the
+update, and (iii) the probability $Q_{ji}$ that an agent who observes the behavior of an agent
+with type $j$ ends up acquiring type $i$:
 \begin{align}
@@
 \centering
-    \begin{tikzpicture}
-
-    \draw[-,dashed,gray] (0,0) -- (5,0) -- (5,5) -- (0,5) -- (0,0);
-
-    \draw (0,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Lbound$}}] {};
-    \draw (5,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Lbound$}}] {};
-    \draw (5,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Llack$}}] {};
-    \draw (0,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Llack$}}] {};
-
-    \draw[<-, thick, indianred] (0.75,0.25) -- (4.25,0.25) node[midway, above] {\footnotesize functional};
-    \draw[->, thick, indianred] (0.75,4.75) -- (4.25,4.75) node[midway, below] {\footnotesize functional};
-    \draw[->, thick, indianred] (0.25,4.25) -- (0.25,0.75) node[midway, below, rotate=90] {\footnotesize functional};
-    \draw[->, thick, indianred] (4.75,4.25) -- (4.75,0.75) node[midway, above, rotate=90] {\footnotesize functional};
-
-  \path[->, thick, densely dashdotted, Steelblue] (0.75,0.75) edge [bend left] node[sloped,below] {\footnotesize learning\textsubscript{flat}} (2.25,2.45);
-  \path[->, thick, densely dashdotted, Steelblue] (4.25,0.75) edge [bend right] node[sloped,below] {\footnotesize leaning\textsubscript{flat}} (2.75,2.45);
-  \path[->, thick, densely dashdotted, Steelblue] (0.75,4.25) edge [bend right] node[sloped,above] {\footnotesize leaning\textsubscript{flat}} (2.25,2.75);
-  \path[->, thick, densely dashdotted, Steelblue] (4.25,4.25) edge [bend left] node[sloped,above] {\footnotesize leaning\textsubscript{flat}} (2.75,2.75);
-
-    \draw[->, thick, goldish] (0.75,-0.25) -- (2.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
-    \draw[<-, thick, goldish] (2.75,-0.25) -- (4.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
-    \draw[->, thick, goldish] (-0.25,0.75) -- (-0.25,4.25) node[midway, above,rotate=90] {\footnotesize learning\textsubscript{biased}};
-    \draw[->, thick, goldish] (5.25,0.75) -- (5.25,4.25) node[midway, below,rotate=90] {\footnotesize learning\textsubscript{bias}};
-    \draw[->, thick, goldish] (0.75,5.25) -- (4.25,5.25) node[midway, above] {\footnotesize learning\textsubscript{biased}};
-
-    \end{tikzpicture}
+    % \begin{tikzpicture}
+
+  %   \draw[-,dashed,gray] (0,0) -- (5,0) -- (5,5) -- (0,5) -- (0,0);
+
+  %   \draw (0,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Lbound$}}] {};
+  %   \draw (5,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Lbound$}}] {};
+  %   \draw (5,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Llack$}}] {};
+  %   \draw (0,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Llack$}}] {};
+
+  %   \draw[<-, thick, indianred] (0.75,0.25) -- (4.25,0.25) node[midway, above] {\footnotesize functional};
+  %   \draw[->, thick, indianred] (0.75,4.75) -- (4.25,4.75) node[midway, below] {\footnotesize functional};
+  %   \draw[->, thick, indianred] (0.25,4.25) -- (0.25,0.75) node[midway, below, rotate=90] {\footnotesize functional};
+  %   \draw[->, thick, indianred] (4.75,4.25) -- (4.75,0.75) node[midway, above, rotate=90] {\footnotesize functional};
+
+  % \path[->, thick, densely dashdotted, Steelblue] (0.75,0.75) edge [bend left] node[sloped,below] {\footnotesize learning\textsubscript{flat}} (2.25,2.45);
+  % \path[->, thick, densely dashdotted, Steelblue] (4.25,0.75) edge [bend right] node[sloped,below] {\footnotesize leaning\textsubscript{flat}} (2.75,2.45);
+  % \path[->, thick, densely dashdotted, Steelblue] (0.75,4.25) edge [bend right] node[sloped,above] {\footnotesize leaning\textsubscript{flat}} (2.25,2.75);
+  % \path[->, thick, densely dashdotted, Steelblue] (4.25,4.25) edge [bend left] node[sloped,above] {\footnotesize leaning\textsubscript{flat}} (2.75,2.75);
+
+  %   \draw[->, thick, goldish] (0.75,-0.25) -- (2.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
+  %   \draw[<-, thick, goldish] (2.75,-0.25) -- (4.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
+  %   \draw[->, thick, goldish] (-0.25,0.75) -- (-0.25,4.25) node[midway, above,rotate=90] {\footnotesize learning\textsubscript{biased}};
+  %   \draw[->, thick, goldish] (5.25,0.75) -- (5.25,4.25) node[midway, below,rotate=90] {\footnotesize learning\textsubscript{bias}};
+  %   \draw[->, thick, goldish] (0.75,5.25) -- (4.25,5.25) node[midway, above] {\footnotesize learning\textsubscript{biased}};
+
+  %   \end{tikzpicture}
 \caption{Sketch of dynamics on a two-dimensional type space with four types.

### 2018-03-09 paper/cogsci-submission2.tex
elaboration on the RMD

@@
 RMD is a population dynamic, which is to say that it describes change in the composition of a
-community made up of different types of agents. The population vector $\vec{x}$ codifies
-relative frequencies of types in a population, where $x_i$ is the relative frequency of type
-$i$ and $\sum\nolimits_i x_i = 1$. The dynamic assumes a virtually infinite population, so that
-it is able to abstract away from what individual agents do and how they change their behavior
-over time. The result is an abstract description of the most probable trajectory of change in
-$\vec{x}$. In its simplest, discrete-time formulation, the RMD defines the frequency $x'_i$ of
-each type $i$ in a population at the next time step as a function of: (i) the frequency $x_i$
-of each type $i$ before the update step, (ii) the fitness $f_i$ of each type $i$ before the
-update, and (iii) the probability $Q_{ji}$ that an agent who observes the behavior of an agent
-with type $j$ ends up acquiring type $i$:
+community made up of different types of agents. The dynamic assumes a virtually infinite
+population, so that it is able to abstract away from what individual agents do and how they
+change their individual behavior over time. The result is an abstract description of the most
+probable trajectory of change in the population vector $\vec{x}$, which codifies relative
+frequencies of types, so that $x_i$ is the relative frequency of type $i$ and
+$\sum\nolimits_i x_i = 1$. In its simplest, discrete-time formulation, the RMD defines the
+frequency $x'_i$ of each type $i$ in the population at the next time step as a function of: (i)
+the frequency $x_i$ of each type $i$ before the update step, (ii) the fitness $f_i$ of each
+type $i$ before the update, and (iii) the probability $Q_{ji}$ that an agent who observes the
+behavior of an agent with type $j$ ends up acquiring type $i$:
 \begin{align}
@@
 
-If the transmission matrix $Q$ is trivial in the sense that $Q_{ji}=1$ whenever $j=i$, the
-dynamic reduces to the replicator dynamic.
-This situation is one where types are always replicated faithfully.
-The replicator dynamic is a model of fitness-based selection in which the relative frequency of type $i$ will increase with a gradient proportional to its average fitness in the population.
-In other words, selection comes into play whenever two or more types replicate at different rates due to differences in their fitness.
-This dynamic is popular and versatile because it can be derived from many abstract processes of biological and cultural
-transmission and selection \citep[for overview and several derivations
-see][]{Sandholm2010:Population-Game}, including conditional imitation
+If there is no mutation, so that types are always replicated faithfully with $Q_{ji}=1$
+whenever $j=i$, the dynamic reduces to the replicator dynamic. If fitness
+$f_i$ is the same for all types $i$, the replicator step is the identity map
+$ \left ( \text{RD}(\vec{x}) \right )_i = x_i$ and the replicator mutator dynamics reduces to a
+process of iterated mutation. Let us have a look at each component in isolation: first
+replication, then mutation.
+
+According to the replicator dynamic the relative frequency $x_i$ of type $i$ increases
+proportional to its average fitness in the population. This dynamic is popular and versatile
+because it can be derived from many abstract processes of biological and cultural transmission
+and selection \citep[for overview and several derivations
+see][]{Sandholm2010:Population-Game}. Under a biological interpretation, the model describes
+population-level changes when the number of offspring in single-parent replication is
+proportional to the parent's fitness. Under a cultural interpretation, examples of concrete
+agent-level processes that lead to population-level changes as described by the replicator
+dynamic include conditional imitation
 \citep[e.g.,][]{Helbing1996:A-Stochastic-Be,Schlag1998:Why-Imitate-and} or reinforcement
-learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. If fitness
-$f_i$ is the same for all types $i$, the replicator step is the identity map
-$ \left ( \text{RD}(\vec{x}) \right )_i = x_i$.
-In words, no difference in fitness translates into a lack of functional pressure: the population's type composition stays the same after the replicator step.
-In such cases the dynamic reduces to a process of iteration of the transmission bias encoded in $Q$: the rate by which one type changes into another after a generational turnover.
-As detailed below, in this way, the process in (\ref{eq:RMD_discrete}), and its equivalent in (\ref{eq:RMD_discrete_recast}), can contain a model of iterated learning \citep{griffiths+kalish:2005,griffiths+kalish:2007}.
+learning \citep[e.g.,][]{BorgersSarin997:Learning-Throug,Beggs2005:On-the-Converge}. In
+conditional imitation, for instance, each agent carries a fixed type $i$ which it may change to
+a type $j$ when it happens to observe another agent using type $j$. Imitation occurs with a
+probability that is a monotonic function of how well $j$ fares against the population: the
+better the behavior $j$, the more likely it is to be imitated.
+
+A crucial assumption in all of the derivations of the replicator dynamic as a biological or
+cultural process of evolution is that agents themselves are rather limited in their strategic
+capacities. They are, for example, not able to innovate a strategy, i.e., to just choose a
+type, that is optimal given the current population state. They are also not able to
+strategically adapt their behavior to any particular agent they are interacting with on some
+given occasion. Such minimalism about the strategic capacities of agents is a conservative
+methodological choice. It allows us to study ---as the formal thought experiment that any such
+model is--- whether whatever we would like to explain could have evolved also in the absence of
+specific strategic or optimal agent-level reasoning. In other words, given its versatile
+interpretability, the replicator dynamic is an abstract, austere and maximally neutral
+description of how many specific kinds of global fitness-based selection play out in large
+populations of agents who are not themselves optimizing their behavior strategically.
+
+If all types have equal fitness no matter what there is no functional pressure and the
+replicator mutator dynamic reduces to a process of iterated mutation, the bias of which is
+encoded in mutation matrix $Q$. Mutation can be interpreted as biological or cultural as
+well. Under a biological interpretation where types are genetically encoded behavioral traits,
+$Q_{ij}$ is the probability by which offspring of type $i$ genetically mutates into type
+$j$. Under a cultural interpretation where types are transmitted by observation or learning,
+such as in conditional imitation, $Q_{ij}$ is the probability that an agent trying to learn,
+adopt or imitate the behavior of an agent with type $i$ acquires type $j$. As detailed below,
+in this way, the process in (\ref{eq:RMD_discrete}), and its equivalent in
+(\ref{eq:RMD_discrete_recast}), can contain a model of iterated learning
+\citep{griffiths+kalish:2005,griffiths+kalish:2007}.
+
+A biological or cultural interpretation with either of the replicator or mutator component need
+not necessarily imply the same kind of interpretation for the other. In the applications to
+language evolution here, we clearly prefer a cultural interpretation of the mutator component
+as the perturbation induces by learnability of linguistic traits. Nothing much hinges on
+whether the replicator component is interpreted as a biological process or a cultural
+process. Fitness-based selection may either be thought of as the outcome of actual reproductive
+success (increased number of offspring which learn a language from their single parent), or, as
+we slightly prefer here, as a process of cultural evolution, such as conditional imitation for
+instance.
+
+we prefer the
 

### 2018-03-09 paper/cogsci-submission2.tex
elaboration on the RMD
moved example into appendix and added derivation of RMD

@@
 \citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. The
-RMD is a population dynamic, which is to say that it describes change in the composition of a
-community made up of different types of agents. The dynamic assumes a virtually infinite
-population, so that it is able to abstract away from what individual agents do and how they
-change their individual behavior over time. The result is an abstract description of the most
-probable trajectory of change in the population vector $\vec{x}$, which codifies relative
-frequencies of types, so that $x_i$ is the relative frequency of type $i$ and
-$\sum\nolimits_i x_i = 1$. In its simplest, discrete-time formulation, the RMD defines the
-frequency $x'_i$ of each type $i$ in the population at the next time step as a function of: (i)
-the frequency $x_i$ of each type $i$ before the update step, (ii) the fitness $f_i$ of each
-type $i$ before the update, and (iii) the probability $Q_{ji}$ that an agent who observes the
-behavior of an agent with type $j$ ends up acquiring type $i$:
+RMD is a population dynamic. It describes the average change in the composition of a community
+of agents of different types. The dynamic assumes a virtually infinite population, so that it
+is able to abstract away from what individual agents do and how they change their individual
+behavior over time. The result is an abstract description of the most probable trajectory of
+change in the population vector $\vec{x}$, which codifies relative frequencies of types, so
+that $x_i$ is the relative frequency of type $i$ and $\sum\nolimits_i x_i = 1$. In its
+simplest discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$ in
+the population at the next time step as a function of: (i) the frequency $x_i$ of each type $i$
+before the update step, (ii) the fitness $f_i$ of each type $i$ before the update, and (iii)
+the probability $Q_{ji}$ that an agent who observes the behavior of an agent with type $j$ ends
+up acquiring type $i$:
 \begin{align}
   \label{eq:RMD_discrete}
-  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_h x_h f_h}\,.
+  x'_i = \sum_j \frac{x_jf_j Q_{ji}}{\Phi}\,,
 \end{align}
-The RMD consists of two components: fitness-based selection and transmission biases encoded in the so-called mutation matrix $Q$. This
-becomes most transparent when we consider an equivalent formulation in terms of a step-wise
-application of the discrete-time replicator dynamic \citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$ and its subsequent multiplication with the mutation matrix $Q$:
+where $\Phi = \sum_k x_k f_k$ is the average fitness in the population.
+
+The RMD consists of two components: fitness-based selection and transmission biases encoded in
+the so-called mutation matrix $Q$. This becomes most transparent when we consider an equivalent
+formulation in terms of a step-wise application of the discrete-time replicator dynamic
+\citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$ and its
+subsequent multiplication with the mutation matrix $Q$:
 \begin{align}
@@
       \left ( \text{RD}(\vec{x}) \right )_i 
-         = \frac{x_i f_i}{\sum_k x_k f_k}
+         = \frac{x_i f_i}{\Phi}
  \ \ \ \ \text{and} \ \ \ \ 
@@
 %
-
 If there is no mutation, so that types are always replicated faithfully with $Q_{ji}=1$
-whenever $j=i$, the dynamic reduces to the replicator dynamic. If fitness
-$f_i$ is the same for all types $i$, the replicator step is the identity map
-$ \left ( \text{RD}(\vec{x}) \right )_i = x_i$ and the replicator mutator dynamics reduces to a
-process of iterated mutation. Let us have a look at each component in isolation: first
-replication, then mutation.
+whenever $j=i$, the dynamic reduces to the replicator dynamic. If fitness is the same for all
+types, $f_i = f_j$ for all $i$ and $j$, the dynamics reduces to a process of iterated
+mutation. Appendix~\ref{sec:minim-example-repl} provides an abstract minimal example of how the
+replicator mutator dynamic works.
+
+\subsection{Interpreting the replicator mutator dynamic}
+
+The replicator mutator dynamic is an abstract description of population-level changes. It
+abstracts away from specific assumptions about how agents interact. To understand this better,
+the following looks at each component in isolation: first replication, then mutation.
 
@@
 see][]{Sandholm2010:Population-Game}. Under a biological interpretation, the model describes
-population-level changes when the number of offspring in single-parent replication is
+population-level changes when the number of offspring under single-parent preproduction is
 proportional to the parent's fitness. Under a cultural interpretation, examples of concrete
@@
 probability that is a monotonic function of how well $j$ fares against the population: the
-better the behavior $j$, the more likely it is to be imitated.
-
-A crucial assumption in all of the derivations of the replicator dynamic as a biological or
-cultural process of evolution is that agents themselves are rather limited in their strategic
-capacities. They are, for example, not able to innovate a strategy, i.e., to just choose a
-type, that is optimal given the current population state. They are also not able to
-strategically adapt their behavior to any particular agent they are interacting with on some
-given occasion. Such minimalism about the strategic capacities of agents is a conservative
-methodological choice. It allows us to study ---as the formal thought experiment that any such
-model is--- whether whatever we would like to explain could have evolved also in the absence of
-specific strategic or optimal agent-level reasoning. In other words, given its versatile
-interpretability, the replicator dynamic is an abstract, austere and maximally neutral
-description of how many specific kinds of global fitness-based selection play out in large
-populations of agents who are not themselves optimizing their behavior strategically.
-
-If all types have equal fitness no matter what there is no functional pressure and the
-replicator mutator dynamic reduces to a process of iterated mutation, the bias of which is
-encoded in mutation matrix $Q$. Mutation can be interpreted as biological or cultural as
+better the behavior $j$, the more likely it is to be imitated.\footnote{A crucial assumption in
+  all of the derivations of the replicator dynamic as a biological or cultural process of
+  evolution is that agents themselves are rather limited in their strategic capacities. They
+  are, for example, not able to innovate a strategy, i.e., to just choose a type, that is
+  optimal given the current population state. They are also not able to strategically adapt
+  their behavior to any particular agent they are interacting with on some given occasion. Such
+  minimalism about the strategic capacities of agents is a conservative methodological
+  choice. It allows us to study ---as the formal thought experiment that any such model is---
+  how fitness-based selection changes a population in the absence of optimal agent-level
+  reasoning and it abstracts away from specific assumptions about how fitness-based selection
+  might be realized concretely.}
+
+If all types have equal fitness at all times, the replicator mutator dynamic reduces to a
+process of iterated mutation. Mutation can be interpreted as biological or cultural as
 well. Under a biological interpretation where types are genetically encoded behavioral traits,
-$Q_{ij}$ is the probability by which offspring of type $i$ genetically mutates into type
-$j$. Under a cultural interpretation where types are transmitted by observation or learning,
-such as in conditional imitation, $Q_{ij}$ is the probability that an agent trying to learn,
-adopt or imitate the behavior of an agent with type $i$ acquires type $j$. As detailed below,
+$Q_{ji}$ is the probability by which offspring of type $j$ genetically mutates into type
+$i$. Under a cultural interpretation where types are transmitted by observation or learning,
+such as in conditional imitation, $Q_{ji}$ is the probability that an agent trying to learn,
+adopt or imitate the behavior of an agent with type $j$ acquires type $i$. As detailed below,
 in this way, the process in (\ref{eq:RMD_discrete}), and its equivalent in
@@
 
-A biological or cultural interpretation with either of the replicator or mutator component need
-not necessarily imply the same kind of interpretation for the other. In the applications to
-language evolution here, we clearly prefer a cultural interpretation of the mutator component
-as the perturbation induces by learnability of linguistic traits. Nothing much hinges on
-whether the replicator component is interpreted as a biological process or a cultural
-process. Fitness-based selection may either be thought of as the outcome of actual reproductive
-success (increased number of offspring which learn a language from their single parent), or, as
-we slightly prefer here, as a process of cultural evolution, such as conditional imitation for
-instance.
-
-we prefer the
-
-\begin{figure}[t]
-  \centering
-  %% notice that the figure is hacked, because somehow the curve does not come out right under
-  %% PGFplots; I don't know why!?!?
-
-    \begin{subfigure}[b]{0.45\textwidth}
-      \centering
-      \begin{tikzpicture}
-      \begin{axis}[xlabel=$x$, ylabel = $x'$,
-                   grid=major, width=6.85cm, height = 5.5cm,
-                   legend style = {legend pos = north west},
-                   ymin=0,ymax=1, xmin=0,xmax=1]
-
-      \addplot[smooth, color=blue] {x^2 / (3*x^2 - 4*x + 2) + 0.019 * x};
-      \addplot[smooth, color=green] {.8*x + .1};
-
-      \addplot[smooth, color=red] {(.9*x^2 - .2*x +.2)/(3*x^2 - 4*x + 2)};
-
-      \addplot[smooth, color=gray] {x};
-
-      \legend{RD, M, RMD}
-
-      \end{axis}
-    \end{tikzpicture}
-
-        \caption{Update functions: the population state $x$ is mapped onto $x'$ in one update step.}
-        \label{fig:Updates_RMS}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.5\textwidth}    
-      \centering
-    \begin{tikzpicture}[x=120]
-      \node[draw=black, circle, fill = black, minimum size = 0.25cm]
-      (0) at (0,0) {};
-
-      \node[draw=black, circle, fill = black, minimum size = 0.25cm]
-      (1) at (1,0) {};
-
-      \draw[-, thick] (0) -- (1);
-
-      \node[draw=black, circle, fill = white, minimum size = 0.25cm, thick]
-      (mid) at (2/3,0) {};
-
-      \node[]  (0.label) at (0,-0.5) {0};
-
-      \node[]  (1.label) at (1,-0.5) {1};
-
-      \node[]  (label) at (0.5,-0.5) {$x$};
-
-      \node[]  (label) at (0.5,0.55) {RD};
-
-      \draw[->-=0.8,very thick] (mid) -- (0);
-
-      \draw[->-=0.8,very thick] (mid) -- (1);
-
-    \end{tikzpicture}
-
-    \vspace*{0.35cm}
-
-    \begin{tikzpicture}[x=120]
-
-      \node (0) at (-0.05,0) {};
-
-      \node (1) at (1.05,0) {};
-
-      \draw[-, thick] (0) -- (1);
-
-      \node[draw=black, circle, fill = black, minimum size = 0.25cm, thick]
-      (mid) at (0.5,0) {};
-
-      \node[]  (0.label) at (0,-0.3) {0};
-
-      \node[]  (1.label) at (1,-0.3) {1};
-
-      \node[]  (label) at (0.5,-0.5) {$x$};
-
-      \node[]  (label) at (0.5,0.55) {M};
-
-      \draw[->-=0.5,very thick] (0) -- (mid);
-
-      \draw[->-=0.5,very thick] (1) -- (mid);
-
-    \end{tikzpicture}
-
-    \vspace*{0.35cm}
-
-    \begin{tikzpicture}[x=120]
-
-      \node (0) at (-0.05,0) {};
-
-      \node (1) at (1.05,0) {};
-
-      \node[draw=black, circle, fill = black, minimum size = 0.25cm]
-      (S0) at (0.121,0) {};
-
-      \node[draw=black, circle, fill = black, minimum size = 0.25cm]
-      (S1) at (0.903,0) {};
-
-      \draw[-, thick] (0) -- (1);
-
-      \node[draw=black, circle, fill = white, minimum size = 0.25cm, thick]
-      (mid) at (0.609,0) {};
-
-      \node[]  (0.label) at (0,-0.3) {0};
-
-      \node[]  (1.label) at (1,-0.3) {1};
-
-      \node[]  (label) at (0.5,-0.5) {$x$};
-
-      \node[]  (label) at (0.5,0.55) {RMD};
-
-      \draw[->-=0.5,very thick] (mid) -- (S0);
-
-      \draw[->-=0.5,very thick] (mid) -- (S1);
-
-      \draw[->-=0.5,very thick] (0) -- (S0);
-
-      \draw[->-=0.5,very thick] (1) -- (S1);
-
-    \end{tikzpicture}
-
-    \caption{Phase portraits for RD, M and RMD: unstable rest points are hollow, attractors are
-      solid.}
-        \label{fig:Phase_RD}
-    \end{subfigure}
-
-  \caption{Example.}
-  \label{fig:Example_RMD}
-\end{figure}
-
-\begin{example}
-  Consider a simple and abstract coordination game. Agents are of two types: positive or negative. If
-  agents of different type interact with each other, they obtain a payoff of $0$. If negative meets
-  negative, each receives a payoff of $1$. If positive meets positive, they get a payoff of $2$.  A population state is completely characterized by the proportion $x$ of negatives. The fitness
-  of negatives in population state $x$ is $f_n(x) = x$, that of positives is $f_p(x) = 2-2x$.
-  The average fitness is $\Phi(x) = x f_n(x) + (1-x) f_p(x) = 3x^2 - 4x + 2$. Without mutation,
-  the replicator dynamic will update $x$ to $\text{RD}(x) = \nicefrac{f(x)x}{\Phi(x)} = \nicefrac{x^2}{\Phi(x)}$. The update
-  function $\text{RD}(x)$ of the replicator step is plotted in Fig.~\ref{fig:Updates_RMS} as the
-  blue line. Rest points, for which $\text{RD}(x)=x$, are at $x=0$, $x=1$ and $x= \nicefrac{2}{3}$.
-  The former are attractors; nearby points converge to them. Points near $x=\nicefrac{2}{3}$
-  also move towards 0 or 1. This is schematically pictured in the topmost phase portrait in
-  Fig.~\ref{fig:Phase_RD}. Adding mutation changes the dynamic and its rest points. Let us assume that $Q_{ji} = .9$
-  when $j=i$.
-This is the proportion of types that are replicated faithfully.
-Conversely, a proportion of $.1$ will change their type from positive to negative, and vice-versa.
-The update effect of mutation on its own are described by $\text{M}(x) = .9 x + .1 (1-x) = .8x + .1$, plotted as the linear green line in Fig.~\ref{fig:Updates_RMS}.
-As shown in Fig.~\ref{fig:Phase_RD}, in this example mutation alone has only one stable rest point.
-It is located at $x = .5$.
-If we first take the replicator step and
-  then the mutation step in sequence, we obtain the replicator mutator dynamic
-  $\text{RMD}(x) = \text{M}(\text{RD}(x)) = \nicefrac{.9x^2 - .2x +.2}{3x^2 - 4x + 2}$, which is plotted in red in
-  Fig.~\ref{fig:Updates_RMS}. The rest points are at $x=.121$, $x=.903$ and $x=.609$. The
-  former two are attractors (see Fig.~\ref{fig:Phase_RD}).
-\end{example}
+In the following application to language evolution, we prefer a cultural interpretation of the
+mutator component as the perturbation induced by how easy different linguistic traits are
+learnable from observation of language use. Nothing much hinges on whether the replicator
+component is interpreted as a biological process or a cultural process. Fitness-based selection
+may either be thought of as the outcome of actual reproductive success (increased number of
+offspring which learn a language from their single parent), or as a process of cultural
+evolution, such as conditional imitation for instance. We believe that this flexibility of
+interpretation is an asset. It abstracts away from specific assumptions and focuses on a
+general, albeit abstract description of fitness-based selection in general. Still, to have
+something concrete in mind when interpreting the following application, it is also important to
+be concrete. Appendix~\ref{sec:deriv-repl-mutat} therefore derives the replicator mutator
+dynamic from a specific scheme of agent-level conditional imitation.
 
@@
 
+\appendix
+
+\section{A minimal example of the replicator mutator dynamic}
+\label{sec:minim-example-repl}
+
+\begin{figure}[t]
+  \centering
+  %% notice that the figure is hacked, because somehow the curve does not come out right under
+  %% PGFplots; I don't know why!?!?
+
+    \begin{subfigure}[b]{0.45\textwidth}
+      \centering
+      \begin{tikzpicture}
+      \begin{axis}[xlabel=$x$, ylabel = $x'$,
+                   grid=major, width=6.85cm, height = 5.5cm,
+                   legend style = {legend pos = north west},
+                   ymin=0,ymax=1, xmin=0,xmax=1]
+
+      \addplot[smooth, color=blue] {x^2 / (3*x^2 - 4*x + 2) + 0.019 * x};
+      \addplot[smooth, color=green] {.8*x + .1};
+
+      \addplot[smooth, color=red] {(.9*x^2 - .2*x +.2)/(3*x^2 - 4*x + 2)};
+
+      \addplot[smooth, color=gray] {x};
+
+      \legend{RD, M, RMD}
+
+      \end{axis}
+    \end{tikzpicture}
+
+        \caption{Update functions: the population state $x$ is mapped onto $x'$ in one update step.}
+        \label{fig:Updates_RMS}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{0.5\textwidth}    
+      \centering
+    \begin{tikzpicture}[x=120]
+      \node[draw=black, circle, fill = black, minimum size = 0.25cm]
+      (0) at (0,0) {};
+
+      \node[draw=black, circle, fill = black, minimum size = 0.25cm]
+      (1) at (1,0) {};
+
+      \draw[-, thick] (0) -- (1);
+
+      \node[draw=black, circle, fill = white, minimum size = 0.25cm, thick]
+      (mid) at (2/3,0) {};
+
+      \node[]  (0.label) at (0,-0.5) {0};
+
+      \node[]  (1.label) at (1,-0.5) {1};
+
+      \node[]  (label) at (0.5,-0.5) {$x$};
+
+      \node[]  (label) at (0.5,0.55) {RD};
+
+      \draw[->-=0.8,very thick] (mid) -- (0);
+
+      \draw[->-=0.8,very thick] (mid) -- (1);
+
+    \end{tikzpicture}
+
+    \vspace*{0.35cm}
+
+    \begin{tikzpicture}[x=120]
+
+      \node (0) at (-0.05,0) {};
+
+      \node (1) at (1.05,0) {};
+
+      \draw[-, thick] (0) -- (1);
+
+      \node[draw=black, circle, fill = black, minimum size = 0.25cm, thick]
+      (mid) at (0.5,0) {};
+
+      \node[]  (0.label) at (0,-0.3) {0};
+
+      \node[]  (1.label) at (1,-0.3) {1};
+
+      \node[]  (label) at (0.5,-0.5) {$x$};
+
+      \node[]  (label) at (0.5,0.55) {M};
+
+      \draw[->-=0.5,very thick] (0) -- (mid);
+
+      \draw[->-=0.5,very thick] (1) -- (mid);
+
+    \end{tikzpicture}
+
+    \vspace*{0.35cm}
+
+    \begin{tikzpicture}[x=120]
+
+      \node (0) at (-0.05,0) {};
+
+      \node (1) at (1.05,0) {};
+
+      \node[draw=black, circle, fill = black, minimum size = 0.25cm]
+      (S0) at (0.121,0) {};
+
+      \node[draw=black, circle, fill = black, minimum size = 0.25cm]
+      (S1) at (0.903,0) {};
+
+      \draw[-, thick] (0) -- (1);
+
+      \node[draw=black, circle, fill = white, minimum size = 0.25cm, thick]
+      (mid) at (0.609,0) {};
+
+      \node[]  (0.label) at (0,-0.3) {0};
+
+      \node[]  (1.label) at (1,-0.3) {1};
+
+      \node[]  (label) at (0.5,-0.5) {$x$};
+
+      \node[]  (label) at (0.5,0.55) {RMD};
+
+      \draw[->-=0.5,very thick] (mid) -- (S0);
+
+      \draw[->-=0.5,very thick] (mid) -- (S1);
+
+      \draw[->-=0.5,very thick] (0) -- (S0);
+
+      \draw[->-=0.5,very thick] (1) -- (S1);
+
+    \end{tikzpicture}
+
+    \caption{Phase portraits for RD, M and RMD: unstable rest points are hollow, attractors are
+      solid.}
+        \label{fig:Phase_RD}
+    \end{subfigure}
+
+  \caption{Example.}
+  \label{fig:Example_RMD}
+\end{figure}
+
+Consider a simple and abstract coordination game. Agents are of two types: positive or
+negative. If agents of different type interact with each other, they obtain a payoff of $0$. If
+negative meets negative, each receives a payoff of $1$. If positive meets positive, they get a
+payoff of $2$.  A population state is completely characterized by the proportion $x$ of
+negatives. The fitness of negatives in population state $x$ is $f_n(x) = x$, that of positives
+is $f_p(x) = 2-2x$.  The average fitness is
+$\Phi(x) = x f_n(x) + (1-x) f_p(x) = 3x^2 - 4x + 2$. Without mutation, the replicator dynamic
+will update $x$ to $\text{RD}(x) = \nicefrac{f(x)x}{\Phi(x)} = \nicefrac{x^2}{\Phi(x)}$. The
+update function $\text{RD}(x)$ of the replicator step is plotted in Fig.~\ref{fig:Updates_RMS}
+as the blue line. Rest points, for which $\text{RD}(x)=x$, are at $x=0$, $x=1$ and
+$x= \nicefrac{2}{3}$.  The former are attractors; nearby points converge to them. Points near
+$x=\nicefrac{2}{3}$ also move towards 0 or 1. This is schematically pictured in the topmost
+phase portrait in Fig.~\ref{fig:Phase_RD}. Adding mutation changes the dynamic and its rest
+points. Let us assume that $Q_{ji} = .9$ when $j=i$. This is the proportion of types that are
+replicated faithfully. Conversely, a proportion of $.1$ will change their type from positive to
+negative, and vice-versa. The update effect of mutation on its own are described by
+$\text{M}(x) = .9 x + .1 (1-x) = .8x + .1$, plotted as the linear green line in
+Fig.~\ref{fig:Updates_RMS}. As shown in Fig.~\ref{fig:Phase_RD}, in this example mutation alone
+has only one stable rest point. It is located at $x = .5$. If we first take the replicator step
+and then the mutation step in sequence, we obtain the replicator mutator dynamic
+$\text{RMD}(x) = \text{M}(\text{RD}(x)) = \nicefrac{.9x^2 - .2x +.2}{3x^2 - 4x + 2}$, which is
+plotted in red in Fig.~\ref{fig:Updates_RMS}. The rest points are at $x=.121$, $x=.903$ and
+$x=.609$. The former two are attractors (see Fig.~\ref{fig:Phase_RD}).
+
+\section{Deriving the replicator mutator dynamic from conditional imitation}
+\label{sec:deriv-repl-mutat}
+
+Consider a huge, virtually infinite population of agents. At any point in time each agent has a
+fixed type $i$ which determines its fitness $f_i$ at that moment in time when playing with an
+arbitrary member of the population. The relative frequency of type $i$ is $x_i$ with
+$\sum_i x_i = 1$. The average population fitness is $\Phi = \sum_k x_k f_k$. The mutation
+probability from $j$ to $i$ is $Q_{ji}$, with $\sum_k Q_{jk} = 1$. The proportion of type $i$
+at the next time step $x'_i$ is given by the discrete time replicator dynamic:
+\begin{align}
+  \label{eq:RMD_discrete_repeated}
+  x'_i = \sum_j \frac{x_jf_j Q_{ji}}{\Phi}\,,
+\end{align}
+We will derive this dynamic from an agent-level update scheme of conditional imitation. There
+are many variants on this derivation. Our choice is to be as simple as
+possible. \citet{Sandholm2010:Population-Game} provides much more detail and discusses many
+interesting variants.
+
+A simple conditional imitation dynamic runs as follows. Every now and then an random agent is
+selected at random from the population and given a chance to possibly change its behavior. Call
+this agent (perhaps slightly misleadingly) \emph{learner} $L$. The learner chooses a random
+agent from the population to possibly imitate. Call this agent (somewhat more misleadingly)
+\emph{teacher} $T$. Whether $L$ actually chooses to imitate the behavior of $T$ depends on how
+good $T$'s behavior is at the moment. For simplicity, we simply assume that $L$ adopts $T$'s
+type with a probability given by $f_j$, where $j$ is the type of $T$.\footnote{We assume here
+  that $f_i \in [0;1]$ for all $i$. This is actually the case for the definition of fitness as
+  communicative success from Section~\ref{sec:fitn--learn}. See
+  \citet{Sandholm2010:Population-Game} for more complex alternative conditional update rules
+  that still derive the same population dynamic.} 
+If $L$ chooses to imitate $T$'s behavior, it may not be able to directly observe $T$'s actual
+type. Rather it may have to infer $T$'s latent type from its overt behavior. The probability
+that $L$ then actually acquires type $k$ when it chose to imitate a teacher of type $j$ is
+$Q_{jk}$. In sum, we assume that learners have (at least roughly) an estimate of how good
+another agent's behavior is, decide to imitate that behavior conditionally on how good it
+actually is and the adopt whatever mechanism (type) they infer may cause the observable
+behavior.
+
+This model of occasional conditional imitation gives rise to expected changes in relative
+frequencies in the population. Given a huge population, the expected probability
+$P(k \rightarrow i)$ that some agent of type $k$ switches by conditional imitation to type $i$
+is the product of probabilities for the following four events:
+\begin{enumerate}
+\item a learner $L$ of type $k$ is sampled, which happens with probability $x_k$
+\item a teacher $T$ of type $j$ is sampled, which happens with probability $x_j$
+\item $L$ chooses to imitate $T$, which happens with probability $f_j$
+\item $L$ infers $T$'s type to be $i$, which happens with probability $Q_{ji}$
+\end{enumerate}
+The average expected change $\dot{x}_i$ in proportion of type $i$ is then:
+\begin{align*}
+  \dot{x}_i & = \sum_k P(k \rightarrow i) - \sum_k P(i \rightarrow k) \\ 
+  & = \sum_k x_k \sum_j x_j f_j Q_{ji} -   \sum_k x_i \sum_j x_j f_j Q_{jk} \\
+  & = \sum_j x_j f_j Q_{ji} -   x_i \sum_j x_j f_j \\
+  & =   (x'_i - x_i) \Phi
+\end{align*}
+In this way, the discrete-time replicator mutator equation in (\ref{eq:RMD_discrete_repeated})
+captures the expected difference in relative frequency of type $i$ from one discrete time step
+to the next (relative to current baseline fitness $\Phi$, which is constant for all types, and
+so only governs the absolute speed of change).
+
 \bibliographystyle{chicago}

### 2018-03-09 paper/cogsci-submission2.tex
restructured sections

@@
 
-The replicator mutator dynamic is an abstract description of population-level changes. It
-abstracts away from specific assumptions about how agents interact. To understand this better,
-the following looks at each component in isolation: first replication, then mutation.
+The RMD is an abstract description of population-level changes. It abstracts away from specific
+assumptions about how agents interact. To understand this better, the following looks at each
+component in isolation: first replication, then mutation.
 
@@
 mutator component as the perturbation induced by how easy different linguistic traits are
-learnable from observation of language use. Nothing much hinges on whether the replicator
-component is interpreted as a biological process or a cultural process. Fitness-based selection
-may either be thought of as the outcome of actual reproductive success (increased number of
-offspring which learn a language from their single parent), or as a process of cultural
-evolution, such as conditional imitation for instance. We believe that this flexibility of
-interpretation is an asset. It abstracts away from specific assumptions and focuses on a
-general, albeit abstract description of fitness-based selection in general. Still, to have
-something concrete in mind when interpreting the following application, it is also important to
-be concrete. Appendix~\ref{sec:deriv-repl-mutat} therefore derives the replicator mutator
-dynamic from a specific scheme of agent-level conditional imitation.
+learnable from observation of language use. Nothing of current relevance hinges on whether the
+replicator component is interpreted as a biological process or a cultural process. We believe
+that this flexibility of interpretation is an asset. It abstracts away from specific
+assumptions and focuses on a general, albeit abstract description of fitness-based selection in
+general. Still, to have something concrete in mind when interpreting the following application,
+it is also important to be concrete. Appendix~\ref{sec:deriv-repl-mutat} therefore derives the
+RMD from a specific scheme of agent-level conditional imitation.
 
@@
 
-Moving beyond an abstract example, our goal is to apply the RMD to investigate the evolution
-of lexical representations and pragmatic behavior. To do so, we need to fix three things: (i) what the
-relevant types are that populate $x$, (ii) how fitness derives from communicative success and (iii) how the
-mutation matrix $Q$ is computed. These issues are addressed, one by one, in the following.
+Our goal is to apply the RMD to the evolution of lexical representations and pragmatic
+behavior. This requires determining three things: (i) what the relevant types are, (ii) how
+fitness of a type derives from its communicative success and (iii) how the mutation matrix $Q$
+is computed. These issues are addressed, one by one, in the following.
 
@@
 they may be thought of as general properties of an agent that influence her fitness, such as
-being positive or negative in our previous example (whatever that means). For our present
-purposes, types are identified by specific assumptions about their cognitive
-make-up.
-More concretely, since we are interested in the evolutionary competition between different lexical representations and ways of deploying them in communication, a type is here defined as a pair consisting of a lexicon and a linguistic strategy of language use.
+being positive or negative in our previous example (whatever that means). Since we are
+interested in the evolutionary competition between different lexical representations and ways
+of deploying them in communication, a type is here defined as a pair consisting of a lexicon
+and a linguistic strategy of language use.
 
@@
 \citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a
-(possibly fuzzy) set of states. A linguistic behavior specifies a probabilistic speaker
-rule (a probabilistic choice of message for each state) and a probabilistic hearer rule (a
-probabilistic choice of state for each message) given a lexicon.
-There are many ways of making these general notions concrete.
-There are models of rational language use coming from a game-theoretic tradition \citep[e.g.,][]{benz:2006,benz+vRooij:2007,jaeger:2007b,franke+jaeger:2014} and ones from a Bayesian tradition \citep[e.g.,][]{frank+goodman:2012,goodman+stuhlmueller:2013,GoodmanFrank2016:Pragmatic-Langu}.
-We will not focus on the particular design choices that make up each approach (on this see \citealt{qing+franke:2015}).
-Instead, we will stress how the approach, common to all of these models, of pragmatic language use as {\em rational language use} can be embedded within the replicator mutator dynamic to answer evolutionary questions at the semantics-pragmatics interface.
-Here is what we will assume in the remainder of this paper.\footnote{Readers more familiar with the Bayesian tradition than the game-theoretic one should note that the definitions below follow the latter in assuming rational sender choice to consist in sending messages that maximize expected utility, rather than message informativity (see \citealt{qing+franke:2015} for discussion).}
+(possibly fuzzy) set of states. A linguistic behavior specifies a probabilistic speaker rule (a
+probabilistic choice of message for each state) and a probabilistic hearer rule (a
+probabilistic choice of state for each message) given a lexicon. There are many ways of making
+these general notions concrete. Since our main emphasis is on how different kinds of language
+can be embedded within the replicator mutator dynamic, we focus here on a general formulation
+of speaker and listener behavior which is as simple as possible but still in line with both a
+game-theoretic tradition
+\citep[e.g.,][]{benz:2006,benz+vRooij:2007,jaeger:2007b,franke+jaeger:2014} and a Bayesian
+tradition
+\citep[e.g.,][]{frank+goodman:2012,goodman+stuhlmueller:2013,GoodmanFrank2016:Pragmatic-Langu}. Here
+is what we will assume in the remainder of this paper.
 
@@
 
-In words, an agent's linguistic behavior -- what message to send when or which message to interpret how -- is defined by (i) her reasoning level and (ii) her lexicon.
-In what follows, this is what we identify an agent's type with, in order to analyze under which conditions combinations of (i) and (ii) emerge through cultural evolution.
-Intuitively, literal behavior (level-$0$ reasoning) results from unreflected language use.
-Such agents produce and comprehend solely based on the truth-conditions of their lexicon.
-In a Gricean spirit, pragmatic behavior (level-$n+1$ reasoning) results from mutual reasoning about rational language use.
-Such agents produce and comprehend by reasoning about their interlocutor's likely interpretation/production patterns.
-Higher order behavior is therefore, even if in tendency rational, quite simple.
-It does not assume agents to know each other's types when interacting (e.g., what lexicon the interlocutor is using); they simply act based on a conjecture of their own behavior if they were in the role of their interlocutor.
+In words, an agent's linguistic behavior -- what message to send when or which message to
+interpret how -- is defined by (i) her reasoning level and (ii) her lexicon. In what follows,
+this is what we identify an agent's type with, in order to analyze under which conditions
+combinations of (i) and (ii) emerge through cultural evolution. Intuitively, literal behavior
+(level-$0$ reasoning) results from unreflected language use. Such agents produce and comprehend
+solely based on the truth-conditions of their lexicon. In a Gricean spirit, pragmatic behavior
+(level-$n+1$ reasoning) results from reasoning about rational language use. Such agents produce
+and comprehend by reasoning about how they themselves would interpret or use
+expressions. Higher order behavior is therefore, even if in tendency rational, quite simple. It
+does not assume agents to know their interlocutor's actual type when interacting (e.g., what
+lexicon the interlocutor is using); they simply act based on a conjecture about language use
+derived from their own (solipsistic) perspective.
 
@@
 decrease as a function of its relative fitness $f_i$. In the context of language evolution,
-fitness is usually associated with the ability to successfully communicate
-with other language users from the same population
-\citep[e.g.,][]{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Under a biological
-interpretation the assumption is that organisms have a higher chance of survival and
-reproduction if they are able to share and receive useful information via communication with
-peers. Under a cultural interpretation the picture is that agents themselves strive towards
-communicative success and therefore occasionally adapt or revise their behavior to achieve
-higher communicative success (see \citealt[\S3.3]{benz+etal:2005b} for discussion).
-
-The replicator equation gives us the means to make the ensuing dynamic precise, without
-necessarily committing to a biological or cultural interpretation. As above, the proportion of
-types in a given population is codified in a vector $\vec{x}$, where $x_i$ is the proportion of
-type $i$. The fitness of type $i$ is its average expected communicative success, or
-\emph{expected utility} (EU), given the frequencies of types in the current population:
+fitness is usually associated with the ability to successfully communicate with other language
+users from the same population \citep[e.g.,][]{nowak+krakauer:1999,nowak+etal:2000,
+  nowak+etal:2002}. Concretely, the fitness of type $i$ is its average expected communicative
+success, or \emph{expected utility} (EU), given the relative frequencies $x_j$ of types $j$ in
+the current population:
 \begin{align*}
@@
 linguistic production can be prone to errors, states or messages may be perceived incorrectly,
-and multiple languages may be compatible with the data learners are exposed to. These sources
+and multiple linguistic hypotheses may be compatible with the data learners are exposed to. These sources
 of uncertainty introduce variation in the transmission of linguistic knowledge from one
@@
 In biological evolution, where types are expressed genetically, transmission infidelity comes
-into the picture through infrequent and mostly random mutation and genetic drift \citep{kimura:1983}. However, an agent's
-lexicon and pragmatic reasoning behavior is likely not inherited genetically. They need to be
-learned from observation.
-Concretely, when agents learn from type $j$, they observe the overt linguistic behavior of type $j$ and need to infer the covert type that most likely produced the observed behavior.
-As in biology, this transmission process is not perfectly accurate.
+into the picture through infrequent and mostly random mutation and genetic drift
+\citep{kimura:1983}. However, an agent's lexicon and pragmatic reasoning behavior is likely not
+inherited genetically. They need to be learned from observation, such as in conditional
+imitation (see Appendix~\ref{sec:deriv-repl-mutat}).  Concretely, when agents learn from or
+imitate type $j$, they observe the overt linguistic behavior of type $j$ and need to infer the
+covert type that most likely produced the observed behavior.
 
@@
 
-Following \citet{griffiths+kalish:2005,griffiths+kalish:2007} we model language acquisition as a process of Bayesian
-inference in which learners combine the likelihood of a type producing the witnessed learning
-input with prior inductive biases. In a Bayesian setting these biases can be codified in a prior
-$P \in \Delta(T)$, which reflects the amount of data a learner requires to faithfully acquire
-the language of the teacher \citep[cf.][450]{griffiths+kalish:2007}.
-Put differently, types favored by the prior are easier to learn because they require less data than less favored types to be inferred, even if they are both compatible with the learning data at hand.
-The extent of the prior's
-influence has been shown to heavily depend on the learning strategy assumed to underlie the
-inference process. On the one hand, early simulation results suggested that weak biases could
-be magnified by exposing learners to only small data samples
-\citep[e.g. in][]{brighton:2002}. On the other, \citeposs{griffiths+kalish:2007} mathematical
-characterization showed that iterated learning converges to the prior in the limit, i.e., that
-the resulting distribution over types corresponds to the learners' prior distribution and
-is not influenced by the amount of input given to them. This difference in predictions can be
-traced back to differences in the selection of hypotheses from the posterior. Griffith \&
-Kalish's convergence to the prior holds for learners that sample from the posterior. More
-deterministic strategies such as the adoption of the type with the highest posterior
-probability, so-called {\it maximum a posterior estimation} (MAP), increase the influence of
-both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we
-use a parameter $l\ge1$ to modulate between posterior sampling and the MAP strategy. When
-$l = 1$ learners sample from the posterior. The learners' propensity to maximize the posterior
-grows as $l$ increases.
-
-The way the posterior over types is formed and how learners employ it are important design choices.
-There are other alternatives as well.
-For instance, instead of adopting a single type, learners could keep their uncertainty over types and base their production behavior on the posterior over types they form \citep{burkett+griffiths:2010}.
-A straightforward operationalization of this idea would then see production defined as a two-step process wherein a speaker first samples from her posterior and then produces utterances based on the type she sampled, as done in \citealt{kirby+etal:2015}.
-Beyond adding complexity to the model, a disadvantage of this approach is that it could happen that a speaker that first uttered a Hungarian utterance speaks Quechua in the next, so to speak.
-We stick to the simpler assumption of a single type being acquired, which we now define.
+Following \citet{griffiths+kalish:2005,griffiths+kalish:2007} we model language acquisition as
+a process of Bayesian inference in which learners combine the likelihood of a type producing
+the witnessed learning input with prior inductive biases. In a Bayesian setting these biases
+can be codified in a prior $P \in \Delta(T)$, which reflects the amount of data a learner
+requires to faithfully acquire the language of the teacher
+\citep[cf.][450]{griffiths+kalish:2007}. Put differently, types favored by the prior are easier
+to learn because they require less data than less favored types to be inferred, even if they
+are both compatible with the learning data at hand. The extent of the prior's influence has
+been shown to heavily depend on the learning strategy assumed to underlie the inference
+process. On the one hand, early simulation results suggested that weak biases could be
+magnified by exposing learners to only small data samples \citep[e.g. in][]{brighton:2002}. On
+the other, \citeposs{griffiths+kalish:2007} mathematical characterization showed that iterated
+learning converges to the prior in the limit, i.e., that the resulting distribution over types
+corresponds to the learners' prior distribution and is not influenced by the amount of input
+given to them. This difference in predictions can be traced back to differences in the
+selection of hypotheses from the posterior. Griffith \& Kalish's convergence to the prior holds
+for learners that sample from the posterior. More deterministic strategies such as the adoption
+of the type with the highest posterior probability, so-called {\it maximum a posterior
+  estimation} (MAP), increase the influence of both the prior and the data
+\citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we use a parameter $l\ge1$ to
+modulate between posterior sampling and the MAP strategy. When $l = 1$ learners sample from the
+posterior. The learners' propensity to maximize the posterior grows as $l$
+increases.\footnote{The way the posterior over types is formed and how learners employ it are
+  important design choices. There are other alternatives as well. For instance, instead of
+  adopting a single type, learners could keep their uncertainty over types and base their
+  production behavior on the posterior over types they form \citep{burkett+griffiths:2010}. A
+  straightforward operationalization of this idea would then see production defined as a
+  two-step process wherein a speaker first samples from her posterior and then produces
+  utterances based on the type she sampled, as done in \citet{kirby+etal:2015}. Beyond adding
+  complexity to the model, a disadvantage of this approach is that it could happen that a
+  speaker that first uttered a Hungarian utterance speaks Quechua in the next, so to speak. We
+  stick to the simpler assumption of a single type being acquired, which we now define.}
 
@@
 
-\subsection{Example}\label{subsec:example}
-The case study in Section~\ref{sec:si-case-study} shows how these components and their interaction can shed light on the evolution of scalar implicatures.
-Before moving on, this section illustrates the model's workings in a constrained type space to better trace the effects that pressure for communicative success and learnability can have on an evolving population.
-To this end, we focus only on the four types from Section~\ref{sec:languages+use}.
-Recall that these types are made up of a combination of either lexicon $\Llack$ or $\Lbound$ and either literal (level-$0$) or pragmatic (level-$1$) behavior.
-
-\begin{figure}
-\centering
-    % \begin{tikzpicture}
-
-  %   \draw[-,dashed,gray] (0,0) -- (5,0) -- (5,5) -- (0,5) -- (0,0);
-
-  %   \draw (0,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Lbound$}}] {};
-  %   \draw (5,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Lbound$}}] {};
-  %   \draw (5,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Llack$}}] {};
-  %   \draw (0,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Llack$}}] {};
-
-  %   \draw[<-, thick, indianred] (0.75,0.25) -- (4.25,0.25) node[midway, above] {\footnotesize functional};
-  %   \draw[->, thick, indianred] (0.75,4.75) -- (4.25,4.75) node[midway, below] {\footnotesize functional};
-  %   \draw[->, thick, indianred] (0.25,4.25) -- (0.25,0.75) node[midway, below, rotate=90] {\footnotesize functional};
-  %   \draw[->, thick, indianred] (4.75,4.25) -- (4.75,0.75) node[midway, above, rotate=90] {\footnotesize functional};
-
-  % \path[->, thick, densely dashdotted, Steelblue] (0.75,0.75) edge [bend left] node[sloped,below] {\footnotesize learning\textsubscript{flat}} (2.25,2.45);
-  % \path[->, thick, densely dashdotted, Steelblue] (4.25,0.75) edge [bend right] node[sloped,below] {\footnotesize leaning\textsubscript{flat}} (2.75,2.45);
-  % \path[->, thick, densely dashdotted, Steelblue] (0.75,4.25) edge [bend right] node[sloped,above] {\footnotesize leaning\textsubscript{flat}} (2.25,2.75);
-  % \path[->, thick, densely dashdotted, Steelblue] (4.25,4.25) edge [bend left] node[sloped,above] {\footnotesize leaning\textsubscript{flat}} (2.75,2.75);
-
-  %   \draw[->, thick, goldish] (0.75,-0.25) -- (2.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
-  %   \draw[<-, thick, goldish] (2.75,-0.25) -- (4.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
-  %   \draw[->, thick, goldish] (-0.25,0.75) -- (-0.25,4.25) node[midway, above,rotate=90] {\footnotesize learning\textsubscript{biased}};
-  %   \draw[->, thick, goldish] (5.25,0.75) -- (5.25,4.25) node[midway, below,rotate=90] {\footnotesize learning\textsubscript{bias}};
-  %   \draw[->, thick, goldish] (0.75,5.25) -- (4.25,5.25) node[midway, above] {\footnotesize learning\textsubscript{biased}};
-
-  %   \end{tikzpicture}
-\caption{Sketch of dynamics on a two-dimensional type space with four types.
-The $x$-axis represents the proportion of types with $\Llack$; the $y$-axis that of pragmatic ones.
-Arrows sketch population trajectories under functional pressure (replicator steps) and learnability pressure (mutator steps) with either a flat prior or one that favors $\Llack$.}
-\label{fig:sketch}
-\end{figure}
-
-A population will be made up of frequencies of these four types.
-We may think of such a population as a point in a two-dimensional space.
-The grey dotted square in Fig.~\ref{fig:sketch} outlines this space.
-The space's first dimension determines the proportion of types that lexicalize ambiguous $\msome$ (the $x$-axis in Fig.~\ref{fig:sketch}).
-That is, the proportion of types with $\Llack$.
-The second dimension determines the proportion of pragmatic language users (the $y$-axis in Fig.~\ref{fig:sketch}).
-Populations at the corners of this space are monomorphic; all agents in them are of a single type.
-If we move from away from edges toward the center of the space, then this two-dimensional representation underspecifies the exact proportions of types found in a population.
-The space's center, for instance, could be composed 50\% of pragmatic $\Llack$ and of an equal proportion of literal $\Lbound$.
-But it could equally well be a population that has each type with the same frequency.
-This level of granularity will nevertheless be sufficient for our purposes, as we are only interested in sketching rough trajectories after replicator or mutator steps.
-
-From Section~\ref{sec:languages+use}, we already know some things about this type space.
-When it comes to communicative success: types with $\Lbound$ have a functional advantage over $\Llack$; literal $\Lbound$ has a slight advantage over pragmatic $\Lbound$; all have a sizable advantage over literal $\Llack$.
-The red arrows in Fig.~\ref{fig:sketch} sketch out these relationships.
-They depict the rough trajectories we can expect if there is only pressure for communicative success.
-How much each replicator step moves a population through this space ultimately depends on the rationality parameter $\lambda$.
-In particular, note that functional differences decrease as $\lambda$ increases: if $\lambda$ is sufficiently high, all types except for literal $\Llack$ speakers will show a strong tendency to associate $\ssome$ with $\msome$ and $\sall$ with $\mall$.
-As a consequence, when it comes to learnability and transmission fidelity, the likelihood that any of these three types produces a given datum $d$ will be approximately equal.
-In other words, if $\lambda$ is high the likelihood offers little differentiation between pragmatic $\Llack$ and any disposition to use $\Lbound$ for learners.
-We can therefore expect that, if the prior is flat, no type is passed on very faithfully in this space.
-This would lead populations to gravitate toward the space's center after mutator steps.
-This is sketched by the blue dash-dotted arrows in Fig.~\ref{fig:sketch}.
-Neither prior nor likelihood alone, but their combination, are what determines the effects of (iterated) Bayesian learning.
-If, instead of a flat prior, there is a prior that favors, say, $\Llack$ over $\Lbound$ we can expect mutator steps to be skewed toward the upper-half of the space.
-This is suggested by the purple arrows in Fig.~\ref{fig:sketch}.
-Taking stock, the effects of iterated learning will depend not only on $\lambda$, affecting the likelihood, but also on the prior (how strong of a bias there is) and the degree to which learners tend to maximize the posterior (controlled by parameter $l$).
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig-contourb105k5.png}
-\caption{Dynamics on a two-dimensional type space with four types.
-The $x$-axis represents the proportion of types with $\Llack$; the $y$-axis that of pragmatic ones.
-Arrows indicate directionality of trajectories after (a) replicator steps and (b) mutator steps with a prior that favors types with $\Llack$ by a factor of $1.05$ over those with $\Lbound$.
-Colored contours show the proportion of pragmatic $\Llack$ users after a step \tb{I don't think the contours are adding much, so they could be removed.
-Then again, I think they are rather non-intrusive, look nice, and illustrate how the step-size, depending on where we are in the space, varies.}}
-
-\label{fig:quiver}
-\end{figure}
-
-Fig.~\ref{fig:quiver} shows actual evolutionary trajectories in this type space.
-The replicator steps make the functional disadvantage of $\Llack$ against $\Lbound$ apparent, particularly for its literal users and/or low $\lambda$.
-They also highlight that the functional difference between literal and pragmatic $\Lbound$ is small.
-Under functional pressure only, populations move toward the lower part of the square but the steps become small as the lower edge, representing populations composed only  of $\Llack$ users, are reached.
-As for the mutator diagrams, there are two main things to note.
-As expected, a small bias for a type will lead to larger proportions to adopt it, provided that likelihood offers little to no differentiation between types.
-This is the situation the right-most space in Fig.~\ref{fig:quiver} depicts.
-However, the bias is far from absolute.
-If, instead, certain types are more likely to be recovered based on the linguistic input they produce, they will be passed on more faithfully.
-This is why, for instance, populations move toward a majority of $\Llack$-types for $\lambda = 5$ and $l = 10$.
-
-Fig.~\ref{fig:sketch} and \ref{fig:quiver} depict replicator and mutator steps in isolation to better showcase their effects.
-Their combination will yield different outcomes, which we now turn to in a case study on scalar implicatures.
-\tb{Also: In this type space RMD way too similar to MD to make it worth the ink.
-It raises more questions than it answers so I left it out as the plots look identical to those in Fig.~3b}
-
 \subsection{Model summary}
@@
 consider what is perhaps one of the simplest non-trivial setups that speak to this matter and
-reflect on its limitations in Section~\ref{sec:discussion}. The setup is introduced in
+reflect on their limitations in Section~\ref{sec:discussion}. Section
+\ref{sec:setup:-stat-conc} first considers a minimal example intended to isolate effects of
+replication and mutation and to gain insights into the sense in which semantic meaning and
+pragmatic use can be said to co-evolve. We then turn to a larger and more realistic setup in
 Section~\ref{sec:setup:-stat-conc}. Section~\ref{sec:simulation-results} describes simulations
-and their results.
-
-% Our initial question about the division of labor between semantics and pragmatics can be narrowed to the case of scalar implicatures by asking for a justification for the lack of lexical upper-bounds in weak scalar alternatives.
-That is, we ask why lexical meanings that lack upper-bounds and convey it pragmatically are regularly selected for over alternatives such as that of codifying the bound semantically.
-More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} and {\em big} were truth-conditionally incompatible with stronger alternatives such as {\em hot}, {\em and}, {\em all} and {\em huge} to avoid misunderstanding?
-This question is particularly striking considering the number of expressions that license such inferences across natural languages.
+for this case and their results.
+
+\section{Co-evolution of semantics \& pragmatics in a restricted type space}
+\label{subsec:example}
+
+This section illustrates how replication and mutation act on the evolution of semantic meaning
+and pragmatic use in the simplest non-trivial type space. Our goal is to sharpen intuitions
+about the sense in which semantics and pragmatics might be said to co-evolve.
+
+\begin{figure}
+\centering
+    % \begin{tikzpicture}
+
+  %   \draw[-,dashed,gray] (0,0) -- (5,0) -- (5,5) -- (0,5) -- (0,0);
+
+  %   \draw (0,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Lbound$}}] {};
+  %   \draw (5,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Lbound$}}] {};
+  %   \draw (5,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Llack$}}] {};
+  %   \draw (0,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Llack$}}] {};
+
+  %   \draw[<-, thick, indianred] (0.75,0.25) -- (4.25,0.25) node[midway, above] {\footnotesize functional};
+  %   \draw[->, thick, indianred] (0.75,4.75) -- (4.25,4.75) node[midway, below] {\footnotesize functional};
+  %   \draw[->, thick, indianred] (0.25,4.25) -- (0.25,0.75) node[midway, below, rotate=90] {\footnotesize functional};
+  %   \draw[->, thick, indianred] (4.75,4.25) -- (4.75,0.75) node[midway, above, rotate=90] {\footnotesize functional};
+
+  % \path[->, thick, densely dashdotted, Steelblue] (0.75,0.75) edge [bend left] node[sloped,below] {\footnotesize learning\textsubscript{flat}} (2.25,2.45);
+  % \path[->, thick, densely dashdotted, Steelblue] (4.25,0.75) edge [bend right] node[sloped,below] {\footnotesize learning\textsubscript{flat}} (2.75,2.45);
+  % \path[->, thick, densely dashdotted, Steelblue] (0.75,4.25) edge [bend right] node[sloped,above] {\footnotesize learning\textsubscript{flat}} (2.25,2.75);
+  % \path[->, thick, densely dashdotted, Steelblue] (4.25,4.25) edge [bend left] node[sloped,above] {\footnotesize learning\textsubscript{flat}} (2.75,2.75);
+
+  %   \draw[->, thick, goldish] (0.75,-0.25) -- (2.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
+  %   \draw[<-, thick, goldish] (2.75,-0.25) -- (4.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
+  %   \draw[->, thick, goldish] (-0.25,0.75) -- (-0.25,4.25) node[midway, above,rotate=90] {\footnotesize learning\textsubscript{biased}};
+  %   \draw[->, thick, goldish] (5.25,0.75) -- (5.25,4.25) node[midway, below,rotate=90] {\footnotesize learning\textsubscript{biased}};
+  %   \draw[->, thick, goldish] (0.75,5.25) -- (4.25,5.25) node[midway, above] {\footnotesize learning\textsubscript{biased}};
+
+  %   \end{tikzpicture}
+\caption{Sketch of dynamics on a two-dimensional type space with four types. The $y$-axis
+  represents the proportion of types with $\Llack$; the $x$-axis that of pragmatic ones. Arrows
+  sketch population trajectories under functional pressure (replicator steps, red arrows) and
+  learnability pressure (mutator steps) with either a flat prior (blue dash-dotted arrows) or a
+  prior that favors $\Llack$ (solid purple arrows).}
+\label{fig:sketch}
+\end{figure}
+
+We focus only on the four types from Section~\ref{sec:languages+use}. Recall that these types
+are made up of a combination of either lexicon $\Llack$ or $\Lbound$ and either literal
+(level-$0$) or pragmatic (level-$1$) behavior.
+
+The population under consideration consists of only the four types from
+Section~\ref{sec:languages+use}. Recall that these types are made up of a combination of either
+lexicon $\Llack$ or $\Lbound$ and either literal (level-$0$) or pragmatic (level-$1$)
+behavior. We may think of such a population as a point in a two-dimensional space. The grey
+dotted square in Fig.~\ref{fig:sketch} outlines this space. The space's first dimension
+determines the proportion of types that lexicalize ambiguous $\msome$ (the $y$-axis in
+Fig.~\ref{fig:sketch}). That is, the proportion of types with $\Llack$. The second dimension
+determines the proportion of pragmatic language users (the $x$-axis in
+Fig.~\ref{fig:sketch}). Populations at the corners of this space are monomorphic; all agents in
+them are of a single type. Points away from the corners represent mixed populations of various
+types.\footnote{This two-dimensional representation underspecifies the exact proportions of
+  types found in a mixed population. The space's center, for instance, could be composed 50\%
+  of pragmatic $\Llack$ and of an equal proportion of literal $\Lbound$. But it could equally
+  well be a population that has each type with the same frequency. This level of granularity
+  will nevertheless be sufficient for our purposes, as we are only interested in sketching
+  rough trajectories after replicator or mutator steps.}
+
+From Section~\ref{sec:languages+use}, we already know some things about this type space. When
+it comes to communicative success: types with $\Lbound$ have a functional advantage over
+$\Llack$; literal $\Lbound$ has a slight advantage over pragmatic $\Lbound$; all have a sizable
+advantage over literal $\Llack$. The red arrows in Fig.~\ref{fig:sketch} sketch out these
+relationships. They depict the rough trajectories we can expect if there is only pressure for
+communicative success. How much each replicator step moves a population through this space
+ultimately depends on the rationality parameter $\lambda$. In particular, note that functional
+differences decrease as $\lambda$ increases: if $\lambda$ is sufficiently high, all types
+except for literal $\Llack$ speakers will show a strong tendency to associate $\ssome$ with
+$\msome$ and $\sall$ with $\mall$.
+
+As a consequence, when it comes to learnability and transmission fidelity, the likelihood that
+any of these three types produces a given datum $d$ will be approximately equal. We can
+therefore expect that, if the prior is flat, no type is passed on very faithfully in this
+space. This would lead populations to gravitate toward the space's center after mutator
+steps. This is sketched by the blue dash-dotted arrows in Fig.~\ref{fig:sketch}. Neither prior
+nor likelihood alone, but their combination, are what determines the effects of (iterated)
+Bayesian learning. If, instead of a flat prior, there is a prior that favors, say, $\Llack$
+over $\Lbound$ we can expect mutator steps to be skewed toward the upper-half of the
+space. This is suggested by the solid purple arrows in Fig.~\ref{fig:sketch}. Taking stock, the
+effects of iterated learning will depend not only on $\lambda$, affecting the likelihood, but
+also on the prior (how strong of a bias there is) and the degree to which learners tend to
+maximize the posterior (controlled by parameter $l$).
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=1.08\textwidth,height=10cm, keepaspectratio]{./plots/fig-contourb105k5.png}
+\caption{Dynamics on a two-dimensional type space with four types. The $y$-axis represents the
+  proportion of types with $\Llack$; the $x$-axis that of pragmatic ones. Arrows indicate
+  directionality of trajectories after (a) replicator steps and (b) mutator steps with a prior
+  that favors types with $\Llack$ by a factor of $1.05$ over those with $\Lbound$. Colored
+  contours show the proportion of pragmatic $\Llack$ users after a step. \tb{I don't think the
+    contours are adding much, so they could be removed. Then again, I think they are rather
+    non-intrusive, look nice, and illustrate how the step-size, depending on where we are in
+    the space, varies.} \mf{I think that the contours are sweet. But I worried about the size
+  of the arrows. I made the picture a bit bigger}}
+
+\label{fig:quiver}
+\end{figure}
+
+Fig.~\ref{fig:quiver} shows actual evolutionary trajectories in this type space. The replicator
+steps make the functional disadvantage of $\Llack$ against $\Lbound$ apparent, particularly for
+its literal users and/or low $\lambda$. They also highlight that the functional difference
+between literal and pragmatic $\Lbound$ is small. Under functional pressure only, populations
+move toward the lower part of the square but the steps become small as the lower edge,
+representing populations composed only of $\Llack$ users, are reached. As for the mutator
+diagrams, there are two main things to note. As expected, a small bias for a type will lead to
+larger proportions to adopt it, provided that likelihood offers little to no differentiation
+between types (see lower left space in Figure~\ref{fig:quiver}). However, the bias is far from
+absolute. If, instead, certain types are more likely to be recovered based on the linguistic
+input they produce, they will be passed on more faithfully. This is why, for instance,
+populations move toward a majority of $\Llack$-types for $\lambda = 5$ and $l = 10$.\mf{last
+  sentence: I don't get it; looks to me as if they are going the other direction; only in the
+  lower right case do I see a shift in that direction}
+
+Fig.~\ref{fig:sketch} and \ref{fig:quiver} depict replicator and mutator steps in isolation to
+better showcase their effects. Their combination will yield different outcomes, which we now
+turn to in a case study with a larger type space and more motivation for particular assumptions
+about learning biases.
 

### 2018-03-10 paper/cogsci-submission2.tex
minor polishing of first parts; added paragraph on co-evolution

@@
 The idea that language is an adaptation to serve a communicative function is fundamental to
-many synchronic and diachronic analyses, at least since \citeposs{zipf:1949} explanation of word
-frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
+many synchronic and diachronic analyses, at least since \citeposs{zipf:1949} explanation of
+word frequency rankings as a result of competing hearer and speaker preferences \citep[e.g.,
 in][]{martinet:1962, horn:1984,jaeger+vRooij:2007,jaeger:2007,
@@
 language. Learnability is another, as natural languages need to be learnable to survive their
-faithful transmission across generations.
-Furthermore, the effects of even small learning biases implicit in acquisition can build up and have quite striking impact on an evolving language in a process of iterated
-learning
+faithful transmission across generations. Furthermore, the effects of even small learning
+biases can have striking effects on an evolving language in a process of iterated learning
 \citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.
@@
 \subsection{The replicator mutator dynamic}
-An elegant formal approach to capture the interaction between communicative efficiency and
-learnability is the \emph{replicator mutator dynamic}
+An elegant formal approach to capture the interaction between selection gor communicative
+efficiency and learnability is the \emph{replicator mutator dynamic}
 \citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. The
@@
 of agents of different types. The dynamic assumes a virtually infinite population, so that it
-is able to abstract away from what individual agents do and how they change their individual
-behavior over time. The result is an abstract description of the most probable trajectory of
-change in the population vector $\vec{x}$, which codifies relative frequencies of types, so
-that $x_i$ is the relative frequency of type $i$ and $\sum\nolimits_i x_i = 1$. In its
-simplest discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$ in
-the population at the next time step as a function of: (i) the frequency $x_i$ of each type $i$
-before the update step, (ii) the fitness $f_i$ of each type $i$ before the update, and (iii)
-the probability $Q_{ji}$ that an agent who observes the behavior of an agent with type $j$ ends
-up acquiring type $i$:
+is able to abstract away from what individual agents do. The result is an abstract description
+of the most probable trajectory of change in the population vector $\vec{x}$, which codifies
+relative frequencies of types, so that $x_i$ is the relative frequency of type $i$ and
+$\sum\nolimits_i x_i = 1$. In its simplest discrete-time formulation, the RMD defines the
+frequency $x'_i$ of each type $i$ in the population at the next time step as a function of: (i)
+the frequency $x_i$ of each type $i$ before the update step, (ii) the fitness $f_i$ of each
+type $i$ before the update, and (iii) the probability $Q_{ji}$ that an agent who observes the
+behavior of an agent with type $j$ ends up acquiring type $i$:
 \begin{align}
@@
 formulation in terms of a step-wise application of the discrete-time replicator dynamic
-\citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$ and its
+\citep{TaylorJonker1978:Evolutionary-St} on the initial population vector $\vec{x}$ and a
 subsequent multiplication with the mutation matrix $Q$:
@@
 If there is no mutation, so that types are always replicated faithfully with $Q_{ji}=1$
-whenever $j=i$, the dynamic reduces to the replicator dynamic. If fitness is the same for all
-types, $f_i = f_j$ for all $i$ and $j$, the dynamics reduces to a process of iterated
+whenever $j=i$, the RMD reduces to the replicator dynamic. If fitness is the same for all
+types, $f_i = f_j$ for all $i$ and $j$, the RMD reduces to a process of iterated
 mutation. Appendix~\ref{sec:minim-example-repl} provides an abstract minimal example of how the
@@
 proportional to its average fitness in the population. This dynamic is popular and versatile
-because it can be derived from many abstract processes of biological and cultural transmission
-and selection \citep[for overview and several derivations
+because it can be derived from many processes of biological and cultural transmission and
+selection \citep[for overview and several derivations
 see][]{Sandholm2010:Population-Game}. Under a biological interpretation, the model describes
@@
 that this flexibility of interpretation is an asset. It abstracts away from specific
-assumptions and focuses on a general, albeit abstract description of fitness-based selection in
-general. Still, to have something concrete in mind when interpreting the following application,
-it is also important to be concrete. Appendix~\ref{sec:deriv-repl-mutat} therefore derives the
-RMD from a specific scheme of agent-level conditional imitation.
+assumptions and focuses on a general and versatile description of fitness-based
+selection. Still, to have something concrete in mind when interpreting the following
+application, Appendix~\ref{sec:deriv-repl-mutat} derives the RMD from a specific scheme of
+agent-level conditional imitation.
 
@@
 
-Under the replicator dynamic the proportion of type $i$ in a population will increase or
-decrease as a function of its relative fitness $f_i$. In the context of language evolution,
-fitness is usually associated with the ability to successfully communicate with other language
-users from the same population \citep[e.g.,][]{nowak+krakauer:1999,nowak+etal:2000,
-  nowak+etal:2002}. Concretely, the fitness of type $i$ is its average expected communicative
-success, or \emph{expected utility} (EU), given the relative frequencies $x_j$ of types $j$ in
-the current population:
+n the context of language evolution, fitness is usually associated with the ability to
+successfully communicate with other language users from the same population
+\citep[e.g.,][]{nowak+krakauer:1999,nowak+etal:2000, nowak+etal:2002}. Concretely, the fitness
+of type $i$ is its average expected communicative success, or \emph{expected utility} (EU),
+given the relative frequencies $x_j$ of types $j$ in the current population:
 \begin{align*}
@@
 frequency $x_i$ before the step, its fitness $f_i$, and the probability that a learner infers
-$i$ when observing the behavior of a type-$j$ agent. Fitness-based selection is here thought of
-not as biological (fitness as expected relative number of offspring) but cultural (fitness as
-likelihood of being imitated or repeated) evolution, since the types that the dynamic operates
-on are pairs consisting of a lexicon and a pragmatic use pattern. A type's communicative success depends
-on how well it communicates within a population while its learnability depends on the
-fidelity by which it is inferred by new generations of learners. The learners' task is
-consequently to perform a joint inference over types of linguistic behavior and lexical
-meaning.
+$i$ when observing the behavior of a type-$j$ agent. Fitness-based can be thought of as
+biological (fitness as expected relative number of offspring) or cultural evolution (fitness as
+likelihood of being imitated or repeated). A type's communicative success depends on how well
+it communicates within a population while its learnability depends on the fidelity by which it
+is inferred by new generations of learners. The learners' task is consequently to perform a
+joint inference over types of linguistic behavior and lexical meaning.
 
@@
 
-The model of the previous section formalizes the evolutionary competition between different
-sets of lexical representations and ways of using them. This section looks at a case study on
-scalar implicatures. It engages in a formal thought experiment to address the
-question: if a population of language users could freely combine different lexica with
-different dispositions to act on them, what are conditions under which the majority view of scalar
-implicatures could have evolved?
+This section engages in a formal thought experiment to address the question: if a population of
+language users could freely combine different lexica with different dispositions to act on
+them, what are conditions under which the majority view of scalar implicatures could have
+evolved?
 
@@
 semantically weak expressions for which logically stronger expressions are salient, e.g.,
-\emph{all}, \emph{hot} or \emph{must}. For instance, {\em some} is entailed by {\em all}. If
-the sentence ``Chris owns all of Johnny Cash's albums'' is true, then ``Chris owns some of
-Johnny Cash's albums'' is also true. However, while weaker expressions such as {\em some} are
-truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what
-their use is normally taken to convey. Instead, the use of a less informative expression when a
-more informative one could have been used can license a defeasible inference that stronger
-alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). In this way, ``Chris owns some
-of Johnny Cash's albums'' is strengthened to convey that she owns {\em some but not all}
-albums. According to the majority view, this is a pragmatic inference, not part of the
-conventional meaning.
+\emph{all}, \emph{hot} or \emph{must}. For instance, {\em some} is entailed by {\em
+  all}. Still, the use of a less informative expression (\emph{some}) when a more informative
+one (\emph{all}) could have been used can license a defeasible inference (\emph{some but not
+  all}) that stronger alternative is false (cf. \citealt{horn:1972,gazdar:1979}). According to
+the majority view, this is a pragmatic inference, not part of the conventional meaning.
 
@@
 which allows us to address the question if or when scalar inferences might (not) lexicalize. We
-consider what is perhaps one of the simplest non-trivial setups that speak to this matter and
-reflect on their limitations in Section~\ref{sec:discussion}. Section
-\ref{sec:setup:-stat-conc} first considers a minimal example intended to isolate effects of
-replication and mutation and to gain insights into the sense in which semantic meaning and
-pragmatic use can be said to co-evolve. We then turn to a larger and more realistic setup in
+consider what are perhaps two of the simplest non-trivial setups that speak to this matter and
+reflect on their limitations in Section~\ref{sec:discussion}. Section \ref{subsec:example}
+first considers a minimal example intended to isolate effects of replication and mutation and
+to gain insights into the sense in which semantic meaning and pragmatic use can be said to
+co-evolve. We then turn to a larger and more realistic setup in
 Section~\ref{sec:setup:-stat-conc}. Section~\ref{sec:simulation-results} describes simulations
@@
 
-\section{Co-evolution of semantics \& pragmatics in a restricted type space}
+\subsection{Co-evolution of semantics \& pragmatics in a restricted type space}
 \label{subsec:example}
@@
 \end{figure}
-
-We focus only on the four types from Section~\ref{sec:languages+use}. Recall that these types
-are made up of a combination of either lexicon $\Llack$ or $\Lbound$ and either literal
-(level-$0$) or pragmatic (level-$1$) behavior.
 
@@
 From Section~\ref{sec:languages+use}, we already know some things about this type space. When
-it comes to communicative success: types with $\Lbound$ have a functional advantage over
+it comes to communicative success, types with $\Lbound$ have a functional advantage over
 $\Llack$; literal $\Lbound$ has a slight advantage over pragmatic $\Lbound$; all have a sizable
@@
 communicative success. How much each replicator step moves a population through this space
-ultimately depends on the rationality parameter $\lambda$. In particular, note that functional
-differences decrease as $\lambda$ increases: if $\lambda$ is sufficiently high, all types
-except for literal $\Llack$ speakers will show a strong tendency to associate $\ssome$ with
-$\msome$ and $\sall$ with $\mall$.
+ultimately depends on the rationality parameter $\lambda$. Functional differences decrease as
+$\lambda$ increases: if $\lambda$ is sufficiently high, all types except for literal $\Llack$
+speakers will show a strong tendency to associate $\ssome$ with $\msome$ and $\sall$ with
+$\mall$.
 
@@
 over $\Lbound$ we can expect mutator steps to be skewed toward the upper-half of the
-space. This is suggested by the solid purple arrows in Fig.~\ref{fig:sketch}. Taking stock, the
-effects of iterated learning will depend not only on $\lambda$, affecting the likelihood, but
-also on the prior (how strong of a bias there is) and the degree to which learners tend to
-maximize the posterior (controlled by parameter $l$).
+space. This is suggested by the solid purple arrows in Fig.~\ref{fig:sketch}. \mf{Should we
+  include here a sentence on why the top-most arrow for biased learning points to pragmatic
+  L-lack?} Taking stock, the effects of iterated learning will depend not only on $\lambda$,
+affecting the likelihood, but also on the prior (how strong of a bias there is) and the degree
+to which learners tend to maximize the posterior (controlled by parameter $l$).
 
@@
 Fig.~\ref{fig:sketch} and \ref{fig:quiver} depict replicator and mutator steps in isolation to
-better showcase their effects. Their combination will yield different outcomes, which we now
-turn to in a case study with a larger type space and more motivation for particular assumptions
-about learning biases.
+better showcase their effects. How exactly these forces shape a population over time also
+hinges on which types are present. Especially the effects of learning biases affected by
+this. If, hypothetically, we add many functionally dispreferred types which all happen to
+mutate with a high probability to the same type, this can strongly bias the dynamics in favor
+of this type. \mf{maybe a sentence here about the top-most arrow in Figure 1?} For this reason,
+we turn to a larger setup next with a larger and more realistic set of types and a suggestion
+for how to formalize potential learning biases in lexical inferences.
+
+Despite its simplicity, the diagram in Figure~\ref{fig:sketch} makes clear the sense in which
+semantics and pragmatics might co-evolve. Pragmatic language use will not evolve under a strong
+and fully expressive semantics (bottom part of Figure~\ref{fig:sketch}). A weak semantic
+convention will not evolve without disposition towards pragmatic enrichments (left part of
+Figure~\ref{fig:sketch}). But together a weak semantics and a pragmatic disposition for
+enrichment can co-evolve.
 
@@
 probability from $j$ to $i$ is $Q_{ji}$, with $\sum_k Q_{jk} = 1$. The proportion of type $i$
-at the next time step $x'_i$ is given by the discrete time replicator dynamic:
+at the next time step $x'_i$ is given by the discrete-time replicator mutator dynamic:
 \begin{align}
@@
 possible. \citet{Sandholm2010:Population-Game} provides much more detail and discusses many
-interesting variants.
-
-A simple conditional imitation dynamic runs as follows. Every now and then an random agent is
-selected at random from the population and given a chance to possibly change its behavior. Call
-this agent (perhaps slightly misleadingly) \emph{learner} $L$. The learner chooses a random
-agent from the population to possibly imitate. Call this agent (somewhat more misleadingly)
+further interesting derivations of similar evolutionary dynamics.
+
+A simple conditional imitation dynamic runs as follows. Every now and then an agent is selected
+at random from the population and given a chance to possibly change its behavior. Call this
+agent (perhaps slightly misleadingly) \emph{learner} $L$. The learner chooses a random agent
+from the population to possibly imitate. Call this agent (somewhat more misleadingly)
 \emph{teacher} $T$. Whether $L$ actually chooses to imitate the behavior of $T$ depends on how
-good $T$'s behavior is at the moment. For simplicity, we simply assume that $L$ adopts $T$'s
-type with a probability given by $f_j$, where $j$ is the type of $T$.\footnote{We assume here
-  that $f_i \in [0;1]$ for all $i$. This is actually the case for the definition of fitness as
-  communicative success from Section~\ref{sec:fitn--learn}. See
-  \citet{Sandholm2010:Population-Game} for more complex alternative conditional update rules
-  that still derive the same population dynamic.} 
-If $L$ chooses to imitate $T$'s behavior, it may not be able to directly observe $T$'s actual
+good $T$'s behavior is at the moment. For simplicity, we assume that $L$ adopts $T$'s type with
+probability $f_j$, where $j$ is the type of $T$.\footnote{We assume here that $f_i \in [0;1]$
+  for all $i$. This is actually the case for the definition of fitness as communicative success
+  from Section~\ref{sec:fitn--learn}. See \citet{Sandholm2010:Population-Game} for more complex
+  alternative conditional update rules that still derive the same population dynamic.}  If $L$
+chooses to imitate $T$'s behavior, it may not be able to directly observe $T$'s actual
 type. Rather it may have to infer $T$'s latent type from its overt behavior. The probability
-that $L$ then actually acquires type $k$ when it chose to imitate a teacher of type $j$ is
+that $L$ actually acquires type $k$ when it chose to imitate a teacher of type $j$ is
 $Q_{jk}$. In sum, we assume that learners have (at least roughly) an estimate of how good
-another agent's behavior is, decide to imitate that behavior conditionally on how good it
-actually is and the adopt whatever mechanism (type) they infer may cause the observable
-behavior.
+another agent's behavior is, decide to imitate that behavior conditionally on how good it is
+and the adopt whatever mechanism (type) they infer may cause the observed behavior.
 

### 2018-03-12 paper/cogsci-submission2.tex
minor edits last sections (shortening)

@@
   learnability pressure (mutator steps) with either a flat prior (blue dash-dotted arrows) or a
-  prior that favors $\Llack$ (solid purple arrows).}
+  prior that favors $\Llack$ (solid purple arrows).\mf{I did not change the upper-most arrow
+    for biased learnability, because I cannot compile this anyway; I did change some typos in
+    arrow labelling though.}}
 \label{fig:sketch}
@@
 
-To fill the model from Section~\ref{sec:model} with life, we need to specify the sets of
-states, messages and lexica we consider (Section~\ref{sec:stat-mess-conc}). Additionally, we want to
-explore the effects of a learning bias in favor of simple lexical representations. One way of
+The effects of learning biases can depend on the set of types. We therefore consider next
+larger and more realistic sets of states, messages and lexica. Additionally, we want to explore
+the effects of a learning bias in favor of simple lexical representations. One way of
 motivating and formalizing such a bias is introduced in Section~\ref{sec:an-induct-learn}.
@@
 it as a partition of possible worlds into cells where none, some or all of the $A$s are $B$s,
-for some arbitrary fixed predicates $A$ and $B$.
-Enlarging the setup from Section~\ref{sec:languages+use} by another state in this way enables us to discern the difference between three kinds of lexica which we will use to unravel the model's dynamics (see below).
-Eight lexical representations can be distinguished based on their truth or falsity in three world states, six of which are not contradictory or tautological (see Table~\ref{tab:concepts} below).
+for some arbitrary fixed predicates $A$ and $B$. Eight lexical representations can be
+distinguished based on their truth or falsity in three world states, six of which are not
+contradictory or tautological (see Table~\ref{tab:concepts} below).
 
@@
 
-The menu of inductive biases argued to shape language acquisition is steadily being refined.
-Apart from simplicity, prominent examples include mutual exclusivity \citep{merriman+bowman:1989,clark:2009}, regularization \citep{hudson+etal:2005}, and generalization \citep{smith:2011}.
-Even when these biases are considered in isolation, there are many ways in which they can be translated into priors over types.
-The key assumption here, common to all simplicity-biased priors, is that simple representational expressions should be favored over more complex ones (see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a, kirby+etal:2015}).
-In view of the menu of possible operationalizations, other biases and their potential interplay, we should stress that these details -- from the generative grammar to its complexity measure -- are to be regarded as one convenient operationalization of one general approach to explicating learning biases; this is not a commitment that this general approach is necessarily superior or that, within it, this particular instrumentalization is the single most plausible.
-We return to this issue in Section~\ref{sec:discussion}.
+The menu of inductive biases argued to shape language acquisition is steadily being
+refined. Apart from simplicity, prominent examples include mutual exclusivity
+\citep{merriman+bowman:1989,clark:2009}, regularization \citep{hudson+etal:2005}, and
+generalization \citep{smith:2011}. Even when these biases are considered in isolation, there
+are many ways in which they can be translated into priors over types. The key assumption here,
+common to all simplicity-biased priors, is that simple representational expressions should be
+favored over more complex ones (see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a,
+  kirby+etal:2015}). In view of these numerous possibilities, we should stress that these
+details -- from the generative grammar to its complexity measure -- are to be regarded as one
+convenient operationalization of one general approach to explicating learning biases; this is
+not a commitment that this general approach is necessarily superior or that, within it, this
+particular instrumentalization is the single most plausible. We return to this issue in
+Section~\ref{sec:discussion}.
 
@@
 \label{sec:simulation-results}
-Recall that there are three parameters held constant across types: soft-max parameter $\lambda$ affects how strongly
-speakers favor messages that appear best from their subjective point of view; the
-bottleneck size $k$ influences how faithfully learners can identify their teacher type; $l$
-defines the learners' disposition towards choosing the most likely teacher type from the
-posterior distribution. We expect that competitor types (types with lexica of the kind
+Recall that there are three parameters held constant across types: soft-max parameter $\lambda$
+affects how strongly speakers favor messages that appear best from their subjective point of
+view; the bottleneck size $k$ influences how faithfully learners can identify their teacher
+type; $l$ defines the learners' disposition towards choosing the most likely teacher type from
+the posterior distribution. Based on considerations and results from
+Section~\ref{subsec:example}, we expect that competitor types (types with lexica of the kind
 $\Lbound$) have a fitness advantage over target types (pragmatic types with lexica of the kind
-$\Llack$), especially for very low levels of $\lambda$ (see Fig~\ref{fig:sketch}). Selection based on fitness alone may
-therefore not lead to prevalence of target types in the population. On the other hand, lexica
-of type $\Llack$ are simpler than those of type $\Lbound$ by the postulated measure from
-above.
-This may make them more likely to be adopted by learners, especially when $l$ is high.
-Still, lexica of the kind $\Lall$ are in turn even more likely \emph{a priori} than lexica of the kind $\Llack$.
-In any case, as showcased in Section~\ref{subsec:example}, the prior is not all that counts for learnability.
-Simulation results will shed light on the question whether target
-types can emerge, and for which parameter constellations.
+$\Llack$), especially for very low levels of $\lambda$ (see Fig~\ref{fig:sketch}). Selection
+based on fitness alone may therefore not lead to prevalence of target types in the
+population. On the other hand, lexica of type $\Llack$ are simpler than those of type $\Lbound$
+by the postulated measure from above. This may make them more likely to be adopted by learners,
+especially when $l$ is high. Still, lexica of the kind $\Lall$ are in turn even more likely
+\emph{a priori} than lexica of the kind $\Llack$. Simulation results will shed light on the
+question whether target types can emerge, and for which parameter constellations.
 
@@
 \includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/barh-onlyr}
-\caption{Proportion of target types, the six competitor types with the highest proportion, the highest type of the $\Lall$-kind, and the population's majority type, in representative populations after $50$ generations under only a
-  pressure for communicative success.}
+\caption{Proportion of target types, the six competitor types with the highest proportion, the
+  most frequent type of the $\Lall$-kind, and the population's majority type, in representative
+  populations after $50$ generations under only a pressure for communicative success
+  (replicator steps).}
 \label{fig:only-R}
@@
 
-Selection based on communicative success is sensitive to $\lambda$ since $\lambda$
-influences signaling behavior, which in turn determines communicative success. This is
-showcased in Fig.
-\ref{fig:only-R}, which shows the proportion of target types, the highest competitor types and the highest type with $\Lall$ in a representative population after $50$ replicator steps.
-The plot also indicates the proportion of the
-\emph{majority type}: the type with the highest proportion in the population. With
-low $\lambda$ many types have very similar behavior, so that evolutionary selection lacks grip
-and becomes very slow. The result is a very long transition with near stagnancy in a rather
-homogeneous population with many types. Conversely, higher $\lambda$ promotes 
-less-stochastic linguistic behavior, widening the gap in expressivity between some types and
-promoting more homogeneous populations. As suggested by Fig. \ref{fig:only-R}, the majority
-in most populations is not one of the six pragmatic $\Llack$-style types. That is, a pressure
-only for communicative success does not lead to a prevalence of target types under any
-$\lambda$-value. For instance, with $\lambda = 20$ $1000$ independent populations only had $11$
+Selection based on communicative success is sensitive to $\lambda$ since $\lambda$ influences
+signaling behavior, which in turn determines communicative success. Fig. \ref{fig:only-R} shows
+the proportion of target types, the highest competitor types and the highest type with $\Lall$
+in a representative population after $50$ replicator steps. The plot also indicates the
+proportion of the \emph{majority type}: the type with the highest proportion in the
+population. With low $\lambda$ many types have very similar behavior, so that evolutionary
+selection lacks grip and becomes very slow. The result is a very long transition with near
+stagnancy in a rather homogeneous population with many types. Conversely, higher $\lambda$
+promotes less-stochastic linguistic behavior, widening the gap in expressivity between some
+types and promoting more homogeneous populations. As suggested by Fig. \ref{fig:only-R}, the
+majority in most populations is not one of the six pragmatic $\Llack$-style types. That is, a
+pressure only for communicative success does not lead to a prevalence of target types under any
+$\lambda$-value. For instance, with $\lambda = 20$, $1000$ independent populations only had $11$
 cases in which the target type was the majority type, corresponding to a mean proportion of
-$.003$ across populations. By contrast, in $913$ cases the majority types had $\Lbound$ with,
+$.003$ across populations. By contrast, in $913$ cases the majority types had $\Lbound$ with
 close to an even share between literal ($454$) and pragmatic types ($459$), corresponding to a
-mean proportion of about $.48$ taken together.
+mean proportion of about $.48$ taken together. In sum, fitness-based selection of single types
+requires sufficiently high $\lambda$ but does not often single out types with an underspecified
+semantic representation. \mf{rewrite this paragraph or section}
 
@@
 
-The effect of iterated learning without a pressure for communicative success using either posterior
-sampling ($l = 1$) or a stronger tendency towards posterior maximization ($l = 15$) is shown in
-Fig.~\ref{fig:only-M} together with the prior over types. The prior shows that while users of
-$\Llack$ are not the most favored by the inductive bias (compared, e.g., to $\Lall$) they are
-nevertheless more advantaged than others, such as $\Lbound$, in virtue of the relatively simple
-semantics they conventionalize (see Section~\ref{sec:an-induct-learn}). Crucially, $\Llack$
-enables its users to convey each state with a single message when combined with pragmatic
-reasoning and sufficiently high $\lambda$. This makes it less likely to be confused with
-other types if the learning data is not too sparse ($k \geq 5$). What is more, learners have
-a propensity to infer pragmatic $\Llack$ when the teacher's type produces very similar
-data, such as when using $\Lbound$ (see Section~\ref{subsec:example}). Moreover, $\Llack$ is less likely to be confused with
-types with different observable behavior because its pragmatic use approximates a one-to-one
-form-meaning mapping. As a consequence, a stronger propensity to maximize the posterior
-increases their proportion in the population.
+The effect of iterated learning without a pressure for communicative success using either
+posterior sampling ($l = 1$) or a stronger tendency towards posterior maximization ($l = 15$)
+is shown in Fig.~\ref{fig:only-M} together with the prior over types. The prior shows that
+while users of $\Llack$ are not the most favored by the inductive bias (compared, e.g., to
+$\Lall$) they are nevertheless more advantaged than others, such as $\Lbound$, in virtue of the
+relatively simple semantics they conventionalize. Crucially, $\Llack$ enables its users to
+convey each state with a single message when combined with pragmatic reasoning and sufficiently
+high $\lambda$. This makes it less likely to be confused with other types if the learning data
+is not too sparse ($k \geq 5$). What is more, learners have a propensity to infer pragmatic
+$\Llack$ when the teacher's type produces very similar data, such as when using $\Lbound$ (see
+Section~\ref{subsec:example}). Moreover, $\Llack$ is less likely to be confused with types with
+different observable behavior because its pragmatic use approximates a one-to-one form-meaning
+mapping. As a consequence, a stronger propensity to maximize the posterior increases their
+proportion in the population.
 
@@
 
-Fig. \ref{fig:rmd} illustrates the combined effects of both pressures for a sample of
-$\lambda$ and $l$ values. These results show that an inductive learning bias for simpler
-semantics in tandem with functional pressure can lead to the selection of a single target
-type, and so to a division of labor between semantics and pragmatics.
-The proportion of a single majority target type increases with $\lambda$ and
-$l$. Pressure for communicative success magnifies the effects of iterated learning and dampens
-the proliferation of multiple types of a kind, equal in expressivity and learnability.
-A pressure towards learnability favors the transmission of simpler semantics and thereby indirectly selects for pragmatic language use in these types.
-Pragmatic behavior, however, is not necessarily selected for in types that can do away without pragmatic reasoning such as $\Lbound$ users (see Section~\ref{subsec:example}).
+Combining pressures for communicative efficiency and learnability can lead to the selection of
+a single target type (see Fig.~\ref{fig:rmd}). The proportion of a single majority target type
+increases with $\lambda$ and $l$.
 
@@
 that are a priori favored, such as $\Lall$. An increase in $\lambda$ leads to the selection of
-target types but does not lead to monomorphic populations if learners sample from the posterior (see the uppermost row in Fig.~\ref{fig:rmd}).
-Finally, a combination of high $\lambda$ and $l$ leads to increasing proportions of a single majority target type.
-This joint influence is summarized in Fig.
-\ref{fig:diff}, which shows the mean difference between the highest target type and the highest proportion of a type of a different kind in $1000$ independent populations across $\lambda$ and $l$ values.
-Higher values of $\lambda$ and $l$ increase the prevalence of a single target type, whereas lower values lead to less pronounced differences, with a valley resulting from low $\lambda$ and high $l$ (cf.
-Fig.
-\ref{fig:rmd} with $\lambda = 1$ and $l=15$).
+target types but does not necessarily lead to monomorphic populations if learners sample from
+the posterior (see the uppermost row in Fig.~\ref{fig:rmd}). Finally, a combination of high
+$\lambda$ and $l$ leads to increasing proportions of a single majority target type. This joint
+influence is summarized in Fig. \ref{fig:diff}, which shows the mean difference between the
+most frequent target type and the proportion of the most frequent non-target type of a
+different kind in $1000$ independent populations across $\lambda$ and $l$ values. Higher values
+of $\lambda$ and $l$ increase the prevalence of a single target type, whereas lower values lead
+to less pronounced differences, with a valley resulting from low $\lambda$ and high $l$
+(cf. Fig. \ref{fig:rmd} with $\lambda = 1$ and $l=15$).
 
@@
 
-When pressured for communicative success only, target types are outcompeted by competitor types.
-When pressured for learnability, populations are polymorphic due to lack of competition; minimally among types of a kind.
-However, due to the prior, learners that are exposed to the input of either targets or competitors tend to acquire the former.
-The consequence of combining both pressures is then that the (slight) functional disadvantage of targets is counterbalanced by an advantage in learnability.
-This leads to large proportions of targets in a population and, due to the competition among types applied by each replicator steps, to more monomorphic populations.
-
-To recapitulate, other than the involvement of pressure on both communicative success and learnability,
-the resulting proportion of pragmatic $\Llack$ speakers primarily hinges on three
-factors. First, the degree, captured by $\lambda$, to which agents try to maximize
-communicative success from their own subjective point of view. Second, the inductive bias,
-which leads learners to prefer simpler over more complex semantic representations in
-acquisition. Lastly, the learning behavior, captured by parameter $l$, where approximating a
-MAP strategy magnifies the effects of the learning bias in tandem with replication.
-
-More broadly, target types, which represent the majority view of scalar implicatures, can come to
-dominate the population if three assumptions are met: (i) language is pressured toward both
-communicative success {\em and} learnability; (ii) pragmatic language use is an option; (iii) learners prefer
-simpler over more complex lexical representations and exhibit a tendency towards the
-acquisition of the type that best explains the learning data.
+In sum, when pressured for communicative success only, target types are outperformed by
+competitor types. When pressured for learnability only, populations are polymorphic due to lack
+of competition. When we combine both pressures, the (slight) functional disadvantage of targets
+is counterbalanced by an advantage in learnability. This leads to large proportions of targets
+and, due to the competition among types applied by each replicator steps, to more monomorphic
+populations. To conclude, target types can come to dominate the population if three assumptions
+are met: (i) language is pressured toward both communicative success {\em and} learnability;
+(ii) pragmatic language use is an option; (iii) learners prefer simpler over more complex
+lexical representations and exhibit a tendency towards the acquisition of the type that best
+explains the learning data.
 
@@
 types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012,
-  franke+jaeger:2014} as well as reasoning about unobservable lexical representations
+  franke+jaeger:2014}, as well as reasoning about unobservable lexical representations
 \citep{bergen+etal:2012,bergen+etal:2016}. This allows for a conceptual investigation of the
@@
 between a lexicon's expressivity and its learnability. A main difference is that here we
-considered communicative pressure for mutual understanding.
-This pressure can indirectly select for expressive types, those that can convey states unequivocally, whereas Kirby et al.
-only consider the bearing that the latter ability has on the production of learnable data.
-We see three main reasons for considering communicative success rather than just
-expressivity, and for looking at communication and learning rather than just learning.
-First, learning alone can  promote populations with non-negligible proportions of functionally defective types.
-This is true both of simulations, e.g., our $\Lall$-types in Fig.~\ref{fig:only-M}, as well as of laboratory experiments with human subjects (see \citealt{kirby+etal:2008,silvey+etal:2014}, and see \citealt{fay+etal:2013} for review of laboratory results).
-Second and more importantly, types may be equally expressive but their performance as means of information transfer crucially depend not only on themselves but on the population they find themselves in (compare the competition of target types in Fig.
-\ref{fig:rmd} and their lack of competition in Fig.
-\ref{fig:only-M}).
-That is, we contend that adopting an expressive type that generates learnable data does not, in itself, capture a type's arguably central communicative function of transferring information to peers.
-Taking communication into consideration allows the model to be responsive to the task for which language is learned.
-Lastly, chains of iterated learning alone do not put types in direct competition.
-Accordingly, learning leads to polymorphous populations in which multiple types of a kind coexist \citep{nowak:2006}.
+considered communicative pressure for mutual understanding. This pressure can indirectly select
+for expressive types, those that can convey states unequivocally, whereas Kirby et al. only
+consider the bearing that the latter ability has on the production of learnable data. We see
+three main reasons for considering communicative success rather than just expressivity, and for
+looking at communication and learning rather than just learning. First, learning alone can
+promote populations with non-negligible proportions of functionally defective types. This is
+true both of simulations and of laboratory experiments with human subjects (see
+\citealt{kirby+etal:2008,silvey+etal:2014}, and see \citealt{fay+etal:2013} for review of
+laboratory results). Second and more importantly, types may be equally expressive but their
+performance as a means of information transfer depends not only on themselves but on the
+population they find themselves in (compare the competition of target types in Fig.
+\ref{fig:rmd} and their lack of competition in Fig. \ref{fig:only-M}). That is, we contend that
+adopting an expressive type that generates learnable data does not, in itself, capture a type's
+arguably central communicative function of transferring information to peers. Taking
+communication into consideration allows the model to be responsive to the task for which
+language is learned. Lastly, chains of iterated learning alone do not put types in direct
+competition. Accordingly, learning leads to polymorphous populations in which multiple types of
+a kind coexist \citep{nowak:2006}.
 
 The main result of our case study is that types that correspond to the majority view of scalar
-implicatures -- scalar readings are non-lexicalized pragmatic enrichments -- can come to dominate a
-population.
-This can happen if pragmatic language use is recruited indirectly by a preference for simpler lexical representations (relative to more complex ones that lead to comparable overt linguistic behavior without pragmatic language use).
-Under this view, semantics
-and pragmatics play a synergic role: pragmatic use allows maintenance of simpler representations;
-pressure towards representational simplicity indirectly promotes pragmatic over literal
-language use. As a consequence, iterated transmission and use of language lead to a
-regularization that may explain the lack of lexicalization of systematic pragmatic enrichments.
+implicatures -- scalar readings are non-lexicalized pragmatic enrichments -- can come to
+dominate a population. This can happen if pragmatic language use is recruited indirectly by a
+preference for simpler lexical representations (relative to more complex ones that lead to
+comparable overt linguistic behavior without pragmatic language use). Under this view,
+semantics and pragmatics play a synergic role and can co-evolve: pragmatic use allows
+maintenance of simpler representations; pressure towards representational simplicity indirectly
+promotes pragmatic over literal language use.
 
@@
 issues. First of all, while many favorable parameter settings exist which lead to a prevalence
-of target types, other types are usually represented in non-negligible proportions (see Fig.~\ref{fig:rmd} and \ref{fig:diff}). This may
-just be a technical quirk of the mutator step; but there is a related issue of empirical
-importance. Several experimental studies on scalar implicatures suggest that participants can
-be classified as either semantic or pragmatic users of, in particular, \emph{some}
+of target types, other types are usually represented in non-negligible proportions as well (see
+Fig.~\ref{fig:rmd} and \ref{fig:diff}). This may just be a technical quirk of the mutator step;
+but there is a related issue of empirical importance. Several experimental studies on scalar
+implicatures suggest that participants can be classified as either semantic or pragmatic users
+of, in particular, \emph{some}
 \citep[e.g.,][]{BottNoveck2004:Some-Utterances,NieuwlandDitman2010:On-the-incremen,DegenTanenhaus2012:Processing-Scal}. The
@@
 not. Interestingly, in our simulations when a target type is the majority type an inflated
-proportion of the population uses compatible lexica with a lexicalized upper
-bound. Particularly in those parameter settings where the prior influences the outcome less. In
-other words, we do find a tendency toward a similar co-existence of semantic and pragmatic
+proportion of the population uses compatible lexica with a lexicalized upper bound. In other
+words, we find a tendency toward a similar co-existence of semantic and pragmatic
 types. Whether this analogy has any further explanatory value is an interesting path for future
@@
 
-Our case study could be criticized as follows: all it shows is that scalar implicatures do not
-lexicalize because upper bounds are dispreferred lexical representations. This criticism would be too
-superficial and highly unjust. Dispreferred lexical representations can thrive under evolutionary
-selection. Lexicalized upper-bounds can dominate a population because they may boost
-communicative success. But they do not have to. Moreover, as stressed throughout this investigation, even without selective pressure
-for communicative success, it is not necessarily the case that the types that are most
-likely \emph{a priori} will dominate. The dynamics of iterated learning are not that
-trivial. Iterated learning does not necessarily promote the \emph{a priori} more likely type,
-but tends to promote a type $t$ based on a gradient of how many other types might likely mutate
-into $t$, so to speak. Taken together, without an explicit model of the interaction between
-pressure for communicative success and learnability, it is far from trivial to judge whether or when
-preferred or dispreferred representations can be adaptive. This is why a major contribution of this
-paper is the arrangement of many different ingredients into a joined model of the (co-)evolution
-of lexical meaning and pragmatic use.
-
-What is more, it is not that we just assumed a prior disadvantage of lexicalized upper
-bounds. We tried to motivate and formalize a general assumption about lexical representations'
+% Our case study could be criticized as follows: all it shows is that scalar implicatures do not
+% lexicalize because upper bounds are dispreferred lexical representations. This criticism would be too
+% superficial and highly unjust. Dispreferred lexical representations can thrive under evolutionary
+% selection. Lexicalized upper-bounds can dominate a population because they may boost
+% communicative success. But they do not have to. Moreover, as stressed throughout this investigation, even without selective pressure
+% for communicative success, it is not necessarily the case that the types that are most
+% likely \emph{a priori} will dominate. The dynamics of iterated learning are not that
+% trivial. Iterated learning does not necessarily promote the \emph{a priori} more likely type,
+% but tends to promote a type $t$ based on a gradient of how many other types might likely mutate
+% into $t$, so to speak. Taken together, without an explicit model of the interaction between
+% pressure for communicative success and learnability, it is far from trivial to judge whether or when
+% preferred or dispreferred representations can be adaptive. This is why a major contribution of this
+% paper is the arrangement of many different ingredients into a joined model of the (co-)evolution
+% of lexical meaning and pragmatic use.
+
+We tried to motivate and formalize a general assumption about lexical representations'
 complexity with a concrete, albeit provisional proposal. The specification of a learning bias
@@
 that for simple representations to emerge it may require pragmatic strategies to compensate
-their potential expressive deficiencies. Hence a model of co-evolving semantics and pragmatics is
-needed. Future work should also include the possibility that representational simplicity may
+their potential expressive deficiencies. Hence a model of co-evolving semantics and pragmatics
+is needed. Future work should also include the possibility that representational simplicity may
 itself be a notion that is subject to evolutionary pressure
-\citep[cf.][]{ThompsonKirby2016:Culture-Shapes-}, as well for the evolution of elements that define
-the agents' cognitive make-up; $\lambda$ and $l$.
-
-Finally, this case study should not be interpreted as a proposal for a definite explanation of
-how scalar implicatures evolved. Other factors should be considered eventually even if they
-will lead to much more complex modeling. One such factor is the observation that
-non-lexicalized upper bounds allow a broader range of applicability. For example, when the speaker is
-not certain as to whether \emph{all} is true. This may suggest an alternative and purely
+\citep[cf.][]{ThompsonKirby2016:Culture-Shapes-}, as well for the evolution of elements that
+define the agents' cognitive make-up; $\lambda$ and $l$.
+
+This case study should not be interpreted as a proposal for a definite explanation of how
+scalar implicatures evolved. Other factors should be considered eventually even if they will
+lead to much more complex modeling. One such factor is the observation that non-lexicalized
+upper bounds allow a broader range of applicability. For example, when the speaker is not
+certain as to whether \emph{all} is true. This may suggest an alternative and purely
 functionalist argument for why upper-bounded meanings do not conventionalize: should contextual
@@
 \citep{piantadosi+etal:2012b}, it allows speakers to enforce an upper-bound and override
-contextual cues that might otherwise mislead the hearer.
-In a nutshell, this explanation posits that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to communicate as economically as possible and pragmatic reasoning enables them to do so.
-What this alternative argument does not explain is why functional pressure does not lead to the emergence of different, equally costly lexical items to express different knowledge states of the speaker (\citealt[252-267]{horn:1984},
-\citealt{horn:1972,traugott:2004,vdAuwera:2010}).
-For instance, to the emergence of two expressions for each weak scalar expression; one with and one lacking an upper-bound.
-That is, this argument does not explain why English and other languages do not have a monomorphemic dual for, e.g., {\em some} that lexicalizes an upper-bound.
-If this hypothetical expression existed, it could be deployed to signal that the speaker knows that {\em some but not all} holds, and unbounded {\em some} could exclusively signal epistemic uncertainty.
-Looking at pressure from learnability might come in again.
-
-Beyond scalar implicatures, the model can generate predictions about likely lexicalization trajectories of pragmatic inferences, or a lack thereof.
-In this realm an interesting issue is whether proposed principles, such as the semantic conventionalization of once highly context-dependent inferences if they become regular enough \citep{levinson:2000,traugott:2004}, can be given
-a formal rationale and inform postulated directionalities of change. The present investigation made a first start and gave a framework for exploring these issues systematically.
+contextual cues that might otherwise mislead the hearer. In a nutshell, this explanation posits
+that scalar implicatures fail to lexicalize because, all else being equal, speakers prefer to
+communicate as economically as possible and pragmatic reasoning enables them to do so. What
+this alternative argument does not explain is why functional pressure does not lead to the
+emergence of different, equally costly lexical items to express different knowledge states of
+the speaker (\citealt[252-267]{horn:1984},
+\citealt{horn:1972,traugott:2004,vdAuwera:2010}). That is, this argument does not explain why
+English and other languages do not have a monomorphemic dual for, e.g., {\em some} that
+lexicalizes an upper-bound. If this hypothetical expression existed, it could be deployed to
+signal that the speaker knows that {\em some but not all} holds, and unbounded {\em some} could
+exclusively signal epistemic uncertainty. Looking at pressure from learnability might come in
+again.
+
+Beyond scalar implicatures, the model can generate predictions about likely lexicalization
+trajectories of pragmatic inferences, or a lack thereof. In this realm an interesting issue is
+whether proposed principles, such as the semantic conventionalization of once highly
+context-dependent inferences, if they become regular enough
+\citep{levinson:2000,traugott:2004}, can be given a formal rationale and inform postulated
+directionalities of change. The present investigation made a first start and gave a framework
+for exploring these issues systematically.
 

### 2018-03-13 paper/cogsci-submission2.tex
thorough revision first half; appendix added; clarified effect of l>1

@@
   interface, we present a game-theoretic model of the competition between types of language
-  users, each endowed with a selection of lexical representations and a particular pragmatic
+  users, each endowed with certain lexical representations and a particular pragmatic
   disposition to act on them. Our model traces two evolutionary forces and their interaction:
-  (i) pressure towards communicative efficiency and (ii) transmission perturbations during the acquisition of
-  linguistic knowledge. We illustrate the model based on a case study on scalar
-  implicatures, which suggests the relationship between underspecified semantics and pragmatic inference to be one of co-evolution.
-The maintenance of underspecified semantics depends on pragmatic reasoning for their enrichment.
-In the opposite direction, complex semantics that can do away with mutual reasoning may not foster pragmatic reasoning, and may even be encumbered by it.
+  (i) pressure towards communicative efficiency and (ii) transmission perturbations during the
+  acquisition of linguistic knowledge. We illustrate the model based on a case study on scalar
+  implicatures, which suggests that the relationship between underspecified semantics and
+  pragmatic inference is one of co-evolution. The evolution of underspecified semantics
+  requires pragmatic reasoning, while pragmatic reasoning may not evolve if lexical meanings do
+  not require pragmatic enrichment.
 \end{abstract}
@@
 the utterance of a sentence like ``I own some of Johnny Cash's albums'' will be taken to mean
-that the speaker does not own all of them. This is because, if the speaker instead owned them all, she
-could have used the word \emph{all} instead of \emph{some} in her utterance, thereby making a
-more informative statement. Scalar implicatures, especially the inference from \emph{some} to
-\emph{some but not all}, have been studied extensively, both theoretically
+that the speaker does not own all of them. This is because, if the speaker instead owned them
+all, she could have used the word \emph{all} instead of \emph{some} in her utterance, thereby
+making a more informative statement. Scalar implicatures, especially the inference from
+\emph{some} to \emph{some but not all}, have been studied extensively, both theoretically
 \citep[e.g.,][]{Sauerland2004:Scalar-Implicat,ChierchiaFox2008:The-Grammatical,Rooyvan-RooijJagerde-Jager2012:Explaining-Quan}
 as well as experimentally
-\citep[e.g.,][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}.
-This makes them particularly suitable candidates for the study of the evolution of regular pragmatic inferences.
-While there is much dispute in this domain about many details, a position endorsed by a clear majority is that a scalar item like \emph{some} is underspecified to mean \emph{some and maybe all} and that the enrichment to \emph{some but not all} is part of some regular process with
-roots in pragmatics.
-
-If this majority view is correct, the question arises how such a division of labor between
-semantics and pragmatics could have evolved; why it would be so pervasive across natural
-languages; and why it is that some expressions systematically draw from it while others semantically conventionalize.
-
-Models of language evolution abound. There are simulation-based models studying
-populations of communicating agents
+\citep[e.g.,][]{BottNoveck2004:Some-Utterances,huang+snedeker:2009,GrodnerKlein2010:Some-and-Possib,GoodmanStuhlmuller2013:Knowledge-and-I,DegenTanenhaus2012:Processing-Scal}. This
+makes them particularly suitable candidates for the study of the evolution of regular pragmatic
+inferences.  While there is much dispute in this domain about many details, a position endorsed
+by a clear majority is that a scalar item like \emph{some} is underspecified to mean \emph{some
+  and maybe all} and that the enrichment to \emph{some but not all} is part of some regular
+process with roots in pragmatics. If this majority view is correct, the question arises how
+such a division of labor between semantics and pragmatics could have evolved.
+
+Models of language evolution abound. There are simulation-based models studying populations of
+communicating agents
 \citep{Hurford1989:Biological-Evol,Steels1995:A-Self-Organizi,LenaertsJansen2005:The-Evolutionar,SteelsBelpaeme2005:Coordinating-Pe,BaronchelliPuglisi2008:Cultural-route-,steels:2011,SpikeStadler2016:Minimal-Require}
@@
 in the overt behavior of agents, abstracting from complex interactions between semantic
-representations and pragmatic use. In contrast, we will here look at language users with a
-richer linguistic make-up. More precisely, we spell out a model of the evolution of conventional meaning and pragmatic reasoning.
-The objects of replication and selection are pairs consisting of a set of lexical
-meanings and a manner of pragmatic behavior, which together define agents' linguistic choices.
-Pragmatic behaviors are represented using probabilistic models of pragmatic language use \citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}.
-Replication
+representations and pragmatic use.
+
+In contrast, we will here look at language users with a richer linguistic make-up. The objects
+of replication and selection in our model are pairs consisting of a set of lexical meanings and
+a manner of pragmatic behavior, which together define an agent's linguistic choices. Types of
+pragmatic behavior are represented using probabilistic models of pragmatic language use
+\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
 and selection are described by the \emph{replicator mutator dynamic}, a general and established
@@
 efficiency and (ii) infidelity in the transmission of linguistic knowledge, caused by factors
-such as inductive learning biases and sparse learning data \citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,KirbyGriffith2014:Iterated-Learni}.
-Considering the effects that the transmission of
-linguistic knowledge through (iterated) learning have is important because neither semantic meaning nor pragmatic usage
-patterns are directly observable. Instead, language learners have to infer these unobservables from the observable behavior in which they result. We formalize this process as a form of
-Bayesian inference. Our approach thereby contains a well-understood model of
-iterated Bayesian learning \citep{griffiths+kalish:2005,griffiths+kalish:2007}, but combines it with functional
-selection, here formalized as the most versatile dynamic from evolutionary game theory, the
-replicator dynamic \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces
-this model.
-
-Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. We
-discuss a setting in which the majority view of underspecified lexical meanings and systematic pragmatic
-enrichments emerges if selection and transmission infidelity are combined. In particular, we
-show that inductive learning biases of Bayesian learners that favor simpler lexical meanings
-may play a role in preventing the lexicalization of scalar inferences and thereby lead to the emergence of Gricean-like
-pragmatic reasoning types. The results of this study are critically assessed in the light of
-the assumptions that feed our model in Section~\ref{sec:discussion}.
+such as inductive learning biases and sparse learning data
+\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,KirbyGriffith2014:Iterated-Learni}. Considering
+the effects that the transmission of linguistic knowledge through (iterated) learning have is
+important because neither semantic meaning nor pragmatic usage patterns are directly
+observable. Instead, language learners have to infer these unobservables from the observable
+behavior in which they result. We formalize this process as a form of Bayesian inference. Our
+approach thereby contains a well-understood model of iterated Bayesian learning
+\citep{griffiths+kalish:2005,griffiths+kalish:2007}, but combines it with functional selection,
+here formalized as the most versatile dynamic from evolutionary game theory, the replicator
+dynamic \citep{TaylorJonker1978:Evolutionary-St}. Section~\ref{sec:model} introduces this
+model.
+
+Section~\ref{sec:si-case-study} applies this model to a case study on scalar implicatures. It
+shows that inductive learning biases of Bayesian learners that favor simpler lexical meanings
+may play a role in preventing the lexicalization of scalar inferences and thereby lead to the
+emergence of Gricean-like pragmatic reasoning types, but only if there is selective pressure
+for communicative efficiency and learnability. The results of this study are critically
+assessed in the light of the assumptions that feed our model in Section~\ref{sec:discussion}.
 
@@
 \subsection{The replicator mutator dynamic}
-An elegant formal approach to capture the interaction between selection gor communicative
+An elegant formal approach to capture the interaction between selection for communicative
 efficiency and learnability is the \emph{replicator mutator dynamic}
 \citep{Hofbauer1985:The-Selection-M,nowak+etal:2000,NowakKomarova2001:Evolution-of-Un,hofbauer+sigmund:2003,nowak:2006}. The
-RMD is a population dynamic. It describes the average change in the composition of a community
-of agents of different types. The dynamic assumes a virtually infinite population, so that it
-is able to abstract away from what individual agents do. The result is an abstract description
-of the most probable trajectory of change in the population vector $\vec{x}$, which codifies
-relative frequencies of types, so that $x_i$ is the relative frequency of type $i$ and
-$\sum\nolimits_i x_i = 1$. In its simplest discrete-time formulation, the RMD defines the
-frequency $x'_i$ of each type $i$ in the population at the next time step as a function of: (i)
-the frequency $x_i$ of each type $i$ before the update step, (ii) the fitness $f_i$ of each
-type $i$ before the update, and (iii) the probability $Q_{ji}$ that an agent who observes the
-behavior of an agent with type $j$ ends up acquiring type $i$:
+RMD is a population dynamic which describes the average change in the composition of a
+community of agents of different types. The dynamic assumes a virtually infinite population, so
+that it is able to abstract away from what individual agents do. It describes the most probable
+trajectory of change in the population vector $\vec{x}$, which codifies relative frequencies of
+types, so that $x_i$ is the relative frequency of type $i$ and $\sum\nolimits_i x_i = 1$. In
+its simplest discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$
+in the population at the next time step as a function of: (i) the frequency $x_i$ of each type
+$i$ before the update step, (ii) the fitness $f_i$ of each type $i$ before the update, and
+(iii) the probability $Q_{ji}$ that an agent who observes the behavior of an agent with type
+$j$ ends up acquiring type $i$ (with $\sum_k Q_{jk} = 1$ for all $j$):
 \begin{align}
@@
 types, $f_i = f_j$ for all $i$ and $j$, the RMD reduces to a process of iterated
-mutation. Appendix~\ref{sec:minim-example-repl} provides an abstract minimal example of how the
-replicator mutator dynamic works.
+mutation. Appendix~\ref{sec:minim-example-repl} provides an abstract minimal example.
 
@@
 The RMD is an abstract description of population-level changes. It abstracts away from specific
-assumptions about how agents interact. To understand this better, the following looks at each
-component in isolation: first replication, then mutation.
+assumptions about how agents interact and what causes selection and mutation. To understand
+this better, the following looks at each component in isolation: first replication, then
+mutation.
 
@@
 see][]{Sandholm2010:Population-Game}. Under a biological interpretation, the model describes
-population-level changes when the number of offspring under single-parent preproduction is
+population-level changes when the number of offspring under single-parent reproduction is
 proportional to the parent's fitness. Under a cultural interpretation, examples of concrete
@@
 a type $j$ when it happens to observe another agent using type $j$. Imitation occurs with a
-probability that is a monotonic function of how well $j$ fares against the population: the
-better the behavior $j$, the more likely it is to be imitated.\footnote{A crucial assumption in
-  all of the derivations of the replicator dynamic as a biological or cultural process of
-  evolution is that agents themselves are rather limited in their strategic capacities. They
-  are, for example, not able to innovate a strategy, i.e., to just choose a type, that is
-  optimal given the current population state. They are also not able to strategically adapt
-  their behavior to any particular agent they are interacting with on some given occasion. Such
-  minimalism about the strategic capacities of agents is a conservative methodological
-  choice. It allows us to study ---as the formal thought experiment that any such model is---
-  how fitness-based selection changes a population in the absence of optimal agent-level
-  reasoning and it abstracts away from specific assumptions about how fitness-based selection
-  might be realized concretely.}
+probability that is a monotonic function of $f_j$, i.e., of how well $j$ fares against the
+population: the better the behavior $j$, the more likely it is to be imitated.\footnote{A
+  crucial assumption in all of the derivations of the replicator dynamic as a biological or
+  cultural process of evolution is that agents themselves are rather limited in their strategic
+  capacities. They are, for example, not able to innovate a strategy, i.e., to just choose a
+  type, that is optimal given the current population state. They are also not able to
+  strategically adapt their behavior to any particular agent they are interacting with on some
+  given occasion. Such minimalism about the strategic capacities of agents is a conservative
+  methodological choice. It allows us to study ---as the formal thought experiment that any
+  such model is--- how fitness-based selection changes a population in the absence of optimal
+  agent-level reasoning and it abstracts away from specific assumptions about how fitness-based
+  selection might be realized concretely.}
 
@@
 \label{sec:languages+use}
-Types are what evolution operates on. They define an agent's fitness, usually through a payoff
-accrued in single interactions with other agents. Often types can be identified as the possible
-acts in a game, e.g., either cooperating or defecting in a prisoner's dilemma. In other cases,
-they may be thought of as general properties of an agent that influence her fitness, such as
-being positive or negative in our previous example (whatever that means). Since we are
-interested in the evolutionary competition between different lexical representations and ways
-of deploying them in communication, a type is here defined as a pair consisting of a lexicon
-and a linguistic strategy of language use.
+Types are what evolution operates on. A type is what determines an agent's behavior and thereby
+an agent's fitness, so that it makes sense to speak of the fitness of a type itself. To study
+the joint evolution of semantic meaning and pragmatic use, types are pairs consisting of a
+lexicon and a linguistic strategy of language use.
 
@@
 message $m$ to a hearer who receives $m$ but does not know $s$
-\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a
-(possibly fuzzy) set of states. A linguistic behavior specifies a probabilistic speaker rule (a
-probabilistic choice of message for each state) and a probabilistic hearer rule (a
-probabilistic choice of state for each message) given a lexicon. There are many ways of making
-these general notions concrete. Since our main emphasis is on how different kinds of language
-can be embedded within the replicator mutator dynamic, we focus here on a general formulation
-of speaker and listener behavior which is as simple as possible but still in line with both a
-game-theoretic tradition
+\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a set of
+states. A linguistic behavior specifies a probabilistic speaker rule (a probabilistic choice of
+message for each state) and a probabilistic hearer rule (a probabilistic choice of state for
+each message) given a lexicon. There are many ways of making these general notions
+concrete. In the following we aim at a general formulation of speaker and listener
+behavior which is as simple as possible but still in line with both a game-theoretic tradition
 \citep[e.g.,][]{benz:2006,benz+vRooij:2007,jaeger:2007b,franke+jaeger:2014} and a Bayesian
 tradition
-\citep[e.g.,][]{frank+goodman:2012,goodman+stuhlmueller:2013,GoodmanFrank2016:Pragmatic-Langu}. Here
-is what we will assume in the remainder of this paper.
+\citep[e.g.,][]{frank+goodman:2012,goodman+stuhlmueller:2013,GoodmanFrank2016:Pragmatic-Langu}.
 
@@
 short for a sentence like \emph{Chris owns some of Johnny Cash's albums} and $\mall$ for the
-same sentence with \emph{some} replaced by \emph{all}.  Lexica for this case would assign a
-Boolean truth value, either $0$ for false or $1$ for true, to each state-message pair. The
-following two lexica are minimal examples for the distinction between a lexicalized upper-bound
-for \emph{some}, in $\Lbound$, and the widely assumed logical semantics with only a lower-bound,
-in $\Llack$.
-
+same sentence with \emph{some} replaced by \emph{all}.  Lexica then assign a Boolean truth
+value, either $0$ for false or $1$ for true, to each state-message pair. The following two
+lexica are minimal examples for the distinction between a lexicalized upper-bound for
+\emph{some}, in $\Lbound$, and the widely assumed logical semantics with only a lower-bound, in
+$\Llack$.
 \begin{align*}
@@
 simplicity. Literal interpreters thereby choose an arbitrary true interpretation for each
-message according to their lexicon. Pragmatic hearers, defined in (\ref{h:leveln}), instead use Bayes' rule to weigh
-interpretations based on a conjecture about speaker behavior. Speaker behavior is regulated by
-a soft-max parameter $\lambda \geq 0$ \citep{luce:1959,sutton+barto:1998}. As
-$\lambda$ increases, choices approximate strict maximization of expected utilities. Expected
-utility of a message $\messg$ in state $\state$ for a level $n+1$ speaker is here defined as
-$H_{n}(s|m;L)$, the probability that the hearer will assign to or choose the correct
-meaning. For literal speakers, utility only tracks truthfulness. Literal speakers choose any
-true message with equal probability but may send false messages as well with a probability
-dependent on $\lambda$.
+message according to their lexicon. Pragmatic hearers, defined in (\ref{h:leveln}), instead use
+Bayes' rule to weigh interpretations based on a conjecture about speaker behavior. Speaker
+behavior is regulated by a soft-max parameter $\lambda \geq 0$
+\citep{luce:1959,sutton+barto:1998}. As $\lambda$ increases, choices approximate strict
+maximization of expected utilities. Expected utility of a message $\messg$ in state $\state$
+for a level $n+1$ speaker is here defined as $H_{n}(s|m;L)$, the probability that the hearer
+will assign to or choose the correct meaning.\footnote{An alternative choice for the speaker's
+  utility function is $\log H_{n}(s|m;L)$. Conceptually, the difference is that without the
+  logarithm, utilities are derived from action choices, while with the logarithm utilities are
+  derived from how closely the listener's beliefs (after hearing an utterance) align with the
+  speaker's, as measure by Kullback-Leibler divergence
+  \cite{frank+goodman:2012,qing+franke:2015}. In an evolutionary context, where fitness comes
+  from actual interactions with the environment, not just beliefs held about it, the
+  formulation without the logarithm may be slightly more appealing. Moreover, log-based speaker
+  utilities cause the speaker never to use false utterances. Since we aim here for a situation
+  in which any type has a reasonable chance to be mistaken for any other based on their
+  production behavior (see Section~\ref{sec:learnability}), we want a definition of speaker
+  types which also send any message in any state with positive probability.} For literal
+speakers, utility only tracks truthfulness. Literal speakers choose any true message with equal
+probability but may send false messages as well with a probability dependent on $\lambda$.
 
@@
 
-There are two particularly important things to note.
-First, in contrast to their literal counterparts of level $0$, pragmatic agents of level $1$ using $\Llack$ associate $\msome$ preferentially with $\ssome$.
-This association is not perfect, and usually less strong than what agents with a lexicalized upper bound in $\Lbound$ can achieve -- with or without pragmatic reasoning.
-Higher order reasoning beyond level $1$ leads to stronger associations of $\msome$ and $\ssome$ also for the receiver.
-Still, the case study presented in Section~\ref{sec:si-case-study} will consider
-sender and receiver behavior at levels 0 and 1, as the latter are the simplest pragmatic reasoning types which show a tendency to communicatively attuned pragmatic enrichment.
-Using only level 1 reasoning and possibly small $\lambda$ is therefore a conservative choice that works against the fitness-based selection of pragmatic language use for a notion of fitness defined as communicative success, which is introduced next.
-Second, when it comes to competition between types of use of lexicon $\Lbound$, pragmatic reasoning at level $1$ is not advantageous.
-The reason for this is that literal use of $\Lbound$ already endows agents with a behavioral strategy that associates a single state with a single message (in tendency; depending on $\lambda$ for senders).
-For $\Lbound$-receivers of level $1$, reasoning over stochasticity introduced at $S_0( \cdot \mid \cdot, \Lbound)$ will generally decrease the association of one state with one message.
-This decrease is only slight if $\lambda$ is high, but nevertheless present.
-That is to say, level-$1$ reasoning does not necessarily confer a functional advantage.
-For some types, such as users of $\Lbound$, literal signaling is preferable.
+In contrast to their literal counterparts of level $0$, pragmatic agents of level $1$ who use
+$\Llack$ associate $\msome$ preferentially with $\ssome$. This association is not perfect, and
+usually less strong than what agents with a lexicalized upper bound in $\Lbound$ can achieve --
+with or without pragmatic reasoning. Higher order reasoning beyond level $1$ leads to stronger
+associations of $\msome$ and $\ssome$ also for the receiver. Still, the case study presented in
+Section~\ref{sec:si-case-study} will consider sender and receiver behavior at levels 0 and 1,
+as the latter are the simplest pragmatic reasoning types which show a tendency to
+communicatively attuned pragmatic enrichment. Using only level 1 reasoning and possibly small
+$\lambda$ is therefore a conservative choice that works \emph{against} the fitness-based
+selection of pragmatic language use for a notion of fitness defined as communicative success,
+which is introduced next.
+
+When it comes to competition between types of use of lexicon $\Lbound$, pragmatic reasoning at
+level $1$ is not advantageous. The reason for this is that literal use of $\Lbound$ already
+endows agents with a behavioral strategy that associates a single state with a single message
+(in tendency; depending on $\lambda$ for senders). For $\Lbound$-receivers of level $1$,
+reasoning over stochasticity introduced at $S_0( \cdot \mid \cdot, \Lbound)$ will generally
+decrease the association of one state with one message. This decrease is only slight if
+$\lambda$ is high, but nevertheless present. That is to say, level-$1$ reasoning does not
+necessarily confer a functional advantage. For some types, such as users of $\Lbound$, literal
+signaling is preferable. In sum, the definition of types introduced here constitute conditions
+that are purposefully averse to what we would like to show; the evolution of is not immediate
+just based on fitness-based selection, to which we turn next.
 
@@
 
-n the context of language evolution, fitness is usually associated with the ability to
+In the context of language evolution, fitness is usually associated with the ability to
 successfully communicate with other language users from the same population
@@
 \end{align*}
-In signaling games, the expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
+The expected utility $\text{EU}(t_i,t_j)$ for type $i$ when communicating with type $j$ is the
 average success of $i$ when talking or listening to $j$. If, as standardly assumed, agents are speakers half of the
@@
 
-In words, expected utility $\text{EU}(t_i,t_j)$ quantifies how successful communication between agents of types $i$ and $j$ is, with each type's behavior resulting from a combination of a lexicon and a reasoning level (Section~\ref{sec:languages+use}).
-Fitness $f_i$ indicates how well type $i$ fares in a population given what other types are found in it: it is type $i$'s weighted expected utility.
+In words, expected utility $\text{EU}(t_i,t_j)$ quantifies how successful communication between
+agents of types $i$ and $j$ is, with each type's behavior resulting from a combination of a
+lexicon and a reasoning level (Section~\ref{sec:languages+use}). Fitness $f_i$ indicates how
+well type $i$ fares in a population $\vec{x}$ where the probability of meeting a type $j$ is
+$x_j$.
 
@@
 \label{sec:learnability}
-Languages are shaped not only by functionalist forces towards greater communicative success. Another
-important factor is the fidelity with which linguistic knowledge is transmitted. Among others,
-linguistic production can be prone to errors, states or messages may be perceived incorrectly,
-and multiple linguistic hypotheses may be compatible with the data learners are exposed to. These sources
-of uncertainty introduce variation in the transmission of linguistic knowledge from one
-generation to the next. In particular, learning biases in the iterated transmission process can
-influence language evolution substantially.
-
-In biological evolution, where types are expressed genetically, transmission infidelity comes
-into the picture through infrequent and mostly random mutation and genetic drift
-\citep{kimura:1983}. However, an agent's lexicon and pragmatic reasoning behavior is likely not
-inherited genetically. They need to be learned from observation, such as in conditional
-imitation (see Appendix~\ref{sec:deriv-repl-mutat}).  Concretely, when agents learn from or
-imitate type $j$, they observe the overt linguistic behavior of type $j$ and need to infer the
-covert type that most likely produced the observed behavior.
-
-Iterated learning is a process in which languages are learned repeatedly from the observation of linguistic behavior of agents who have themselves acquired their behavior from observation and inference.
-In the simplest case there is a single teacher and a single learner in each generation \citep[e.g.,][]{kirby:2001,brighton:2002}.
-After sufficient training the learner becomes a teacher and produces behavior that serves as input for a new learner.
-Due to the pressure towards learnability it exerts, iterated learning alone generally leads to simpler
-and more regular languages (see \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016} for
+
+Languages are shaped not only by functionalist forces towards greater communicative
+success. Another important factor is the fidelity with which linguistic knowledge is
+transmitted. In biological evolution, where types are expressed genetically, transmission
+infidelity comes into the picture through infrequent and mostly random mutation and genetic
+drift \citep{kimura:1983}. However, an agent's lexicon and a disposition for pragmatic
+reasoning are likely not inherited genetically. They need to be learned from observation, such
+as in conditional imitation (see Appendix~\ref{sec:deriv-repl-mutat}).  Concretely, when agents
+learn from or imitate type $j$, they observe the overt linguistic behavior of type $j$ and need
+to infer the covert type that most likely produced the observed behavior.
+
+Iterated learning is a process in which languages are learned repeatedly from the observation
+of linguistic behavior of agents who have themselves acquired their behavior from observation
+and inference. In the simplest case there is a single teacher and a single learner in each
+generation \citep[e.g.,][]{kirby:2001,brighton:2002}. After sufficient training the learner
+becomes a teacher and produces behavior that serves as input for a new learner. Due to the
+pressure towards learnability it exerts, iterated learning alone generally leads to simpler and
+more regular languages (see \citealt{kirby+etal:2014} and \citealt{tamariz+kirby:2016} for
 recent surveys).
@@
 to learn because they require less data than less favored types to be inferred, even if they
-are both compatible with the learning data at hand. The extent of the prior's influence has
-been shown to heavily depend on the learning strategy assumed to underlie the inference
-process. On the one hand, early simulation results suggested that weak biases could be
-magnified by exposing learners to only small data samples \citep[e.g. in][]{brighton:2002}. On
-the other, \citeposs{griffiths+kalish:2007} mathematical characterization showed that iterated
-learning converges to the prior in the limit, i.e., that the resulting distribution over types
+are both compatible with the learning data at hand.
+
+The extent of the prior's influence has been shown to heavily depend on the learning strategy
+assumed to underlie the inference process. Early simulation results suggested that weak biases
+could be magnified by exposing learners to only small data samples
+\citep[e.g. in][]{brighton:2002}. \citet{griffiths+kalish:2007} showed that iterated learning
+converges to the prior in the limit, i.e., that the resulting distribution over types
 corresponds to the learners' prior distribution and is not influenced by the amount of input
-given to them. This difference in predictions can be traced back to differences in the
-selection of hypotheses from the posterior. Griffith \& Kalish's convergence to the prior holds
-for learners that sample from the posterior. More deterministic strategies such as the adoption
-of the type with the highest posterior probability, so-called {\it maximum a posterior
-  estimation} (MAP), increase the influence of both the prior and the data
+given to them, if learners select random hypothesis from their posterior beliefs about which
+types may have generated the learning data that they saw. More deterministic strategies such as
+the adoption of the type with the highest posterior probability, so-called {\it maximum a
+  posterior estimation} (MAP), increase the influence of both the prior and the data
 \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we use a parameter $l\ge1$ to
@@
   speaker that first uttered a Hungarian utterance speaks Quechua in the next, so to speak. We
-  stick to the simpler assumption of a single type being acquired, which we now define.}
+  therefore maintain the simpler assumption of a single type being acquired in the present
+  context, and postpone further investigation of more complex type-adoption schemes to future
+  research.}
 
@@
 \end{align*}
-where, as before, $n_j$ is $j$'s pragmatic reasoning type and $L_j$ is $j$'s lexicon. For a
+where, as before, $n_j$ is $j$'s pragmatic reasoning level and $L_j$ is $j$'s lexicon. For a
 given observation $d$, the probability of acquiring type $i$ is $F(t_i \mid d)$, so that:
 \begin{flalign*}
-  Q_{ji} \propto \sum_{d \in D} P(d \mid t_j) F(t_i \mid d)\,.
+  Q_{ji} = \sum_{d \in D} P(d \mid t_j) F(t_i \mid d)\,.
 \end{flalign*}
@@
 
+When $l=1$ learners sample a type to adopt from their posterior distribution which they obtain
+from observing the teacher produce utterances in particular states. From
+\citet{griffiths+kalish:2007}, we know that under $l=1$ iterated Bayesian learning will
+converge to a population proportion that reflects that prior over types. If $l>1$ learners tend
+to pick types which are \emph{a posteriori} more likely with a higher probability than less
+likely types. Even in the absence of a prior bias, iterated Bayesian learning with $l>1$ can
+lead to populations in which a particular type $t_i$ has a higher proportion than other
+types. For $l>1$, much hinges on particular asymmetries in production likelihoods
+$P(d \mid t_j)$, and therefore on the composition of the set of types in general. Intuitively
+put, if there are many types $t_j$ which all happen to produce data that is most likely to be
+produced by $t_i$, this can lead to a larger proportion of $t_i$ in the long run even when
+priors over types are uniform. Appendix~\ref{sec:example-iter-bayes} provides an abstract
+example.
+
 \subsection{Model summary}
@@
 This equation defines the frequency $x'_i$ of type $i$ at the next time step, based on its
-frequency $x_i$ before the step, its fitness $f_i$, and the probability that a learner infers
-$i$ when observing the behavior of a type-$j$ agent. Fitness-based can be thought of as
+frequency $x_i$ before the update step, its fitness $f_i$, and the probability that a learner
+infers $i$ when observing the behavior of a type-$j$ agent. Fitness-based can be thought of as
 biological (fitness as expected relative number of offspring) or cultural evolution (fitness as
-likelihood of being imitated or repeated). A type's communicative success depends on how well
-it communicates within a population while its learnability depends on the fidelity by which it
-is inferred by new generations of learners. The learners' task is consequently to perform a
-joint inference over types of linguistic behavior and lexical meaning.
+likelihood of being imitated or repeated, see Appendix~\ref{sec:deriv-repl-mutat} for a
+concrete example). A type's communicative success depends on how well it communicates within a
+population while its learnability depends on the fidelity by which it is inferred by new
+generations of learners. The learners' task is consequently to perform a joint inference over
+types of linguistic behavior and lexical meaning.
 
@@
 them, what are conditions under which the majority view of scalar implicatures could have
-evolved?
-
-Recall that the majority view is that scalar implicatures are non-lexicalized pragmatic
-enrichments. Scalar implicature triggers like \emph{some}, \emph{warm} or \emph{may} are
-semantically weak expressions for which logically stronger expressions are salient, e.g.,
+evolved? --- Recall that the majority view is that scalar implicatures are non-lexicalized
+pragmatic enrichments. Scalar implicature triggers like \emph{some}, \emph{warm} or \emph{may}
+are semantically weak expressions for which logically stronger expressions are salient, e.g.,
 \emph{all}, \emph{hot} or \emph{must}. For instance, {\em some} is entailed by {\em
@@
 
-In the following we consider a specific application of the model from Section~\ref{sec:model}
-which allows us to address the question if or when scalar inferences might (not) lexicalize. We
-consider what are perhaps two of the simplest non-trivial setups that speak to this matter and
-reflect on their limitations in Section~\ref{sec:discussion}. Section \ref{subsec:example}
-first considers a minimal example intended to isolate effects of replication and mutation and
-to gain insights into the sense in which semantic meaning and pragmatic use can be said to
-co-evolve. We then turn to a larger and more realistic setup in
+In the following we consider what are perhaps two of the simplest non-trivial setups that speak
+to this matter and reflect on their limitations in Section~\ref{sec:discussion}. Section
+\ref{subsec:example} first investigates a minimal example intended to isolate effects of
+replication and mutation and to gain insights into the sense in which semantic meaning and
+pragmatic use can be said to co-evolve. We then turn to a larger and more realistic setup in
 Section~\ref{sec:setup:-stat-conc}. Section~\ref{sec:simulation-results} describes simulations
@@
 \label{subsec:example}
-
-This section illustrates how replication and mutation act on the evolution of semantic meaning
-and pragmatic use in the simplest non-trivial type space. Our goal is to sharpen intuitions
-about the sense in which semantics and pragmatics might be said to co-evolve.
 
@@
 
-The population under consideration consists of only the four types from
-Section~\ref{sec:languages+use}. Recall that these types are made up of a combination of either
-lexicon $\Llack$ or $\Lbound$ and either literal (level-$0$) or pragmatic (level-$1$)
-behavior. We may think of such a population as a point in a two-dimensional space. The grey
-dotted square in Fig.~\ref{fig:sketch} outlines this space. The space's first dimension
-determines the proportion of types that lexicalize ambiguous $\msome$ (the $y$-axis in
-Fig.~\ref{fig:sketch}). That is, the proportion of types with $\Llack$. The second dimension
-determines the proportion of pragmatic language users (the $x$-axis in
-Fig.~\ref{fig:sketch}). Populations at the corners of this space are monomorphic; all agents in
-them are of a single type. Points away from the corners represent mixed populations of various
-types.\footnote{This two-dimensional representation underspecifies the exact proportions of
-  types found in a mixed population. The space's center, for instance, could be composed 50\%
-  of pragmatic $\Llack$ and of an equal proportion of literal $\Lbound$. But it could equally
-  well be a population that has each type with the same frequency. This level of granularity
-  will nevertheless be sufficient for our purposes, as we are only interested in sketching
-  rough trajectories after replicator or mutator steps.}
-
-From Section~\ref{sec:languages+use}, we already know some things about this type space. When
-it comes to communicative success, types with $\Lbound$ have a functional advantage over
+This section illustrates how replication and mutation act on the evolution of semantic meaning
+and pragmatic use in the simplest non-trivial type space, which consists of only the four types
+from Section~\ref{sec:languages+use}: a type has either lexicon $\Llack$ or $\Lbound$ and is
+either a literal (level-$0$) or a pragmatic (level-$1$) language user. We may think of such a
+population as a point in a two-dimensional space. The grey dotted square in
+Fig.~\ref{fig:sketch} outlines this space. The space's first dimension determines the
+proportion of types that lexicalize ambiguous $\msome$ (the $y$-axis in
+Fig.~\ref{fig:sketch}). The second dimension determines the proportion of pragmatic language
+users (the $x$-axis in Fig.~\ref{fig:sketch}). Populations at the corners of this space are
+monomorphic; all agents in them are of a single type. Points away from the corners represent
+mixed populations of various types.\footnote{This two-dimensional representation underspecifies
+  the exact proportions of types found in a mixed population. The space's center, for instance,
+  could be composed in equal parts of pragmatic $\Llack$ and of literal $\Lbound$. But it could
+  equally well be a population that has all four types in equal proportions. This level of
+  granularity will nevertheless be sufficient for our purposes, as we are only interested in
+  sketching rough trajectories after replicator or mutator steps.} We can predict general
+trends in evolutionary trajectories in this space based on what we know from the literature and
+the setup from Section~\ref{sec:fitn--learn}, even before any further simulation results.
+
+When it comes to communicative success, types with $\Lbound$ have a functional advantage over
 $\Llack$; literal $\Lbound$ has a slight advantage over pragmatic $\Lbound$; all have a sizable
-advantage over literal $\Llack$. The red arrows in Fig.~\ref{fig:sketch} sketch out these
-relationships. They depict the rough trajectories we can expect if there is only pressure for
-communicative success. How much each replicator step moves a population through this space
-ultimately depends on the rationality parameter $\lambda$. Functional differences decrease as
-$\lambda$ increases: if $\lambda$ is sufficiently high, all types except for literal $\Llack$
-speakers will show a strong tendency to associate $\ssome$ with $\msome$ and $\sall$ with
-$\mall$.
-
-As a consequence, when it comes to learnability and transmission fidelity, the likelihood that
-any of these three types produces a given datum $d$ will be approximately equal. We can
-therefore expect that, if the prior is flat, no type is passed on very faithfully in this
-space. This would lead populations to gravitate toward the space's center after mutator
-steps. This is sketched by the blue dash-dotted arrows in Fig.~\ref{fig:sketch}. Neither prior
-nor likelihood alone, but their combination, are what determines the effects of (iterated)
-Bayesian learning. If, instead of a flat prior, there is a prior that favors, say, $\Llack$
-over $\Lbound$ we can expect mutator steps to be skewed toward the upper-half of the
-space. This is suggested by the solid purple arrows in Fig.~\ref{fig:sketch}. \mf{Should we
-  include here a sentence on why the top-most arrow for biased learning points to pragmatic
-  L-lack?} Taking stock, the effects of iterated learning will depend not only on $\lambda$,
-affecting the likelihood, but also on the prior (how strong of a bias there is) and the degree
-to which learners tend to maximize the posterior (controlled by parameter $l$).
+advantage over literal $\Llack$. The red arrows in Fig.~\ref{fig:sketch} sketch trajectories we
+can expect if there is only pressure for communicative success. How much each replicator step
+moves a population through this space ultimately depends on the rationality parameter
+$\lambda$. Functional differences decrease as $\lambda$ increases: if $\lambda$ is sufficiently
+high, all types except for literal $\Llack$ speakers will show a strong tendency to associate
+$\ssome$ with $\msome$ and $\sall$ with $\mall$.
+
+Under probability matching $l=1$, iterated Bayesian learning converges to the prior over types
+\citep{griffiths+kalish:2007}. If the prior on types is uniform, populations gravitate toward
+the space's center under repeated mutator steps. This is sketched by the blue dash-dotted
+arrows in Fig.~\ref{fig:sketch}. If instead the prior favors $\Llack$ over $\Lbound$, we expect
+repeated mutator steps to drive the population toward the upper-half of the space. This is
+suggested by the solid purple arrows in Fig.~\ref{fig:sketch}. The system's predictions are
+more difficult to predict when $l>1$, because now individual mutator steps may also be
+influenced not only by the priors, but also by the differential likelihood with which a
+particular type $i$ can be confused for any other type $j$ (see Section~\ref{sec:learnability}
+and Appendix~\ref{sec:example-iter-bayes}).
+
+Despite its simplicity, the diagram in Figure~\ref{fig:sketch} makes clear the sense in which
+semantics and pragmatics might co-evolve. Pragmatic language use will not evolve under a strong
+and fully expressive semantics (bottom part of Figure~\ref{fig:sketch}). A weak semantic
+convention will not evolve without disposition towards pragmatic enrichments (left part of
+Figure~\ref{fig:sketch}). But together a weak semantics and a pragmatic disposition for
+enrichment can co-evolve. These abstract theoretical considerations also suggest that we need
+both fitness based selection of communicative efficiency and pressure towards learnability to
+see underspecified semantics and pragmatic language use emerge together.
 
@@
 Fig.~\ref{fig:quiver} shows actual evolutionary trajectories in this type space. The replicator
-steps make the functional disadvantage of $\Llack$ against $\Lbound$ apparent, particularly for
-its literal users and/or low $\lambda$. They also highlight that the functional difference
-between literal and pragmatic $\Lbound$ is small. Under functional pressure only, populations
-move toward the lower part of the square but the steps become small as the lower edge,
-representing populations composed only of $\Llack$ users, are reached. As for the mutator
-diagrams, there are two main things to note. As expected, a small bias for a type will lead to
-larger proportions to adopt it, provided that likelihood offers little to no differentiation
-between types (see lower left space in Figure~\ref{fig:quiver}). However, the bias is far from
-absolute. If, instead, certain types are more likely to be recovered based on the linguistic
-input they produce, they will be passed on more faithfully. This is why, for instance,
-populations move toward a majority of $\Llack$-types for $\lambda = 5$ and $l = 10$.\mf{last
-  sentence: I don't get it; looks to me as if they are going the other direction; only in the
-  lower right case do I see a shift in that direction}
-
-Fig.~\ref{fig:sketch} and \ref{fig:quiver} depict replicator and mutator steps in isolation to
-better showcase their effects. How exactly these forces shape a population over time also
-hinges on which types are present. Especially the effects of learning biases affected by
-this. If, hypothetically, we add many functionally dispreferred types which all happen to
-mutate with a high probability to the same type, this can strongly bias the dynamics in favor
-of this type. \mf{maybe a sentence here about the top-most arrow in Figure 1?} For this reason,
-we turn to a larger setup next with a larger and more realistic set of types and a suggestion
-for how to formalize potential learning biases in lexical inferences.
-
-Despite its simplicity, the diagram in Figure~\ref{fig:sketch} makes clear the sense in which
-semantics and pragmatics might co-evolve. Pragmatic language use will not evolve under a strong
-and fully expressive semantics (bottom part of Figure~\ref{fig:sketch}). A weak semantic
-convention will not evolve without disposition towards pragmatic enrichments (left part of
-Figure~\ref{fig:sketch}). But together a weak semantics and a pragmatic disposition for
-enrichment can co-evolve.
+steps shown in the top row of Fig.~\ref{fig:quiver} make the functional disadvantage of
+$\Llack$ against $\Lbound$ apparent, particularly for its literal users and/or low
+$\lambda$. They also highlight that the functional difference between literal and pragmatic
+$\Lbound$ is small. Under functional pressure only, populations move toward the lower part of
+the square but the steps become small as the lower edge, representing populations composed only
+of $\Llack$ users, are reached. As for the mutator diagrams (bottom row of
+Fig.~\ref{fig:quiver}), there are three main things to note. First, as expected from the
+literature, a small bias for a type will lead to larger proportions to adopt it under $l=1$
+(see lower left space in Figure~\ref{fig:quiver}). Second, if $l>0$, the dynamic is influenced
+heavily also by asymmetries in production likelihoods. Third, these asymmetries in the
+production likelihoods are in turn influenced by $\lambda$, so that $\lambda$ also has an
+effect on iterated Bayesian learning (compare the middle and right-most diagram in the bottom
+row of Fig.~\ref{fig:quiver}).
+
+Fig.~\ref{fig:sketch} and \ref{fig:quiver} investigate replicator and mutator steps in
+isolation. It remains to investigate how exactly these forces shape a population over time when
+they apply together. This will also hinge on which types are present because the space of all
+types determines whether production likelihoods asymmetrically favor certain types over others
+(when $l>1$; see Appendix~\ref{sec:example-iter-bayes}). For this reason, we turn to a larger
+setup next with a larger and more realistic set of types. Additionally, we want to explore
+the effects of a learning bias in favor of simple lexical representations. One way of
+motivating and formalizing such a bias is introduced in Section~\ref{sec:an-induct-learn}.
 
@@
 \label{sec:setup:-stat-conc}
-
-The effects of learning biases can depend on the set of types. We therefore consider next
-larger and more realistic sets of states, messages and lexica. Additionally, we want to explore
-the effects of a learning bias in favor of simple lexical representations. One way of
-motivating and formalizing such a bias is introduced in Section~\ref{sec:an-induct-learn}.
 
@@
 
+\section{Example of iterated Bayesian learning}
+\label{sec:example-iter-bayes}
+
+We consider three types and three possible data observations they can produce. Types are
+equally likely \emph{a priori}, but show different and crucially asymmetric production
+behavior. The likelihood with which type $t_i$ produces datum $d_j$ is $LH_{ij}$:
+\begin{align*}
+  LH & = \begin{blockarray}{lccc}
+    & d_1 & d_2 & d_3 \\
+    \begin{block}{l[ccc]}
+      t_1 & 1-2\epsilon & \epsilon & \epsilon \\
+      t_2 & \delta & 1-(\epsilon + \delta) & \epsilon \\
+      t_2 & \delta & \epsilon & 1-(\epsilon + \delta)  \\
+    \end{block}
+  \end{blockarray}
+\end{align*}
+If $\delta > \epsilon$, production likelihoods are asymmetrically skewed to possibly favor
+$t_1$, in the sense that both types $t_2$ and $t_3$  produce behavior that is strong
+evidence for $t_1$ more often than behavior that is strong evidence for any other
+type other then themselves. Type $t_1$ makes no such ``asymmetric mistakes''.
+
+Using the definitions from Section~\ref{sec:learnability}, we calculate mutation matrices $Q$ for
+$l=1$ and $l=10$ when $\epsilon = 0.1$ and $\delta = 0.2$:
+\begin{align*}
+  Q_{l = 1} & \approx \begin{blockarray}{lccc}
+    & t_1 & t_2 & t_3 \\
+    \begin{block}{l[ccc]}
+      t_1 & 0.56 & 0.22 & 0.22 \\
+      t_2 & 0.22 & 0.59 & 0.19 \\
+      t_2 & 0.22 & 0.19 & 0.59  \\
+    \end{block}
+  \end{blockarray} & 
+  Q_{l = 10} & \approx \begin{blockarray}{lccc}
+    & t_1 & t_2 & t_3 \\
+    \begin{block}{l[ccc]}
+      t_1 & 0.8 & 0.1 & 0.1 \\
+      t_2 & 0.2 & 0.7 & 0.1 \\
+      t_2 & 0.2 & 0.1 & 0.7  \\
+    \end{block}
+  \end{blockarray}
+\end{align*}
+Despite flat priors over types, the mutation matrix for $l=10$ is skewed towards $t_1$ in the
+sense that the probability of any type mutating to $t_1$, $\sum_i Q_{i1} = 1.2$, is higher than
+the probability of $\sum_i Q_{ij} = 0.8$ for $j \in \set{2,3}$. For $l=1$, these probabilities
+are all equal for all three types. Consequently, under iterated Bayesian learning with $l=1$
+the population will converge to a uniform distribution over all three types in the limit, but
+for $l=10$ it converges to a skewed population vector $\vec{x} \approx \tuple{0.5, 0.25, 0.25}$
+where $t_1$ is a clear majority type. This demonstrates how iterated Bayesian learning with
+$l>1$ can be sensitive to the set of types represented in the model, in particular when
+production likelihoods are asymmetrically in favor of some types. With probability matching
+learners ($l = 1$), these considerations do not play a role, as the system will gravitate to
+the prior distribution over types nonetheless.
+
 \bibliographystyle{chicago}

### 2018-03-13 code-LOT-extension/plot_result_summary.R

@@ -14,7 +14,7 @@ e$type = sapply(1:nrow(e), function(i) ifelse(strsplit(e$strategy[i], split = ".
                                               strsplit(e$strategy[i], split = ".", fixed = T)[[1]][2]))
                        
 
-f = e %>% filter(rank >= 6) %>% arrange(runID, lambda, type, - proportion) %>% group_by(runID, lambda, type) %>% mutate(rank = 1:n())
+f = e %>% filter(type != "other") %>% arrange(runID, lambda, type, - proportion) %>% group_by(runID, lambda, type) %>% mutate(rank = 1:n())
 
 plot.data = f %>% group_by(lambda, type, rank) %>%
   summarize(mean = mean(proportion),
@@ -37,7 +37,7 @@ e = melt(d, id.vars = c(1,2,3), variable.name = "strategy", value.name = "propor
 e$type = sapply(1:nrow(e), function(i) ifelse(strsplit(e$strategy[i], split = ".", fixed = T)[[1]][1] == "other", "other",
                                               strsplit(e$strategy[i], split = ".", fixed = T)[[1]][2]))
 
-f = e %>% filter(rank >= 6) %>% arrange(runID, lambda, l, type, - proportion) %>% 
+f = e %>% filter(type != "other") %>% arrange(runID, lambda, l, type, - proportion) %>% 
   group_by(runID, lambda, l, type) %>% mutate(rank = 1:n())
 
 plot.data = f %>% group_by(l, lambda, type, rank) %>%
@@ -60,7 +60,7 @@ e = melt(d, id.vars = c(1,2,3), variable.name = "strategy", value.name = "propor
 e$type = sapply(1:nrow(e), function(i) ifelse(strsplit(e$strategy[i], split = ".", fixed = T)[[1]][1] == "other", "other",
                                               strsplit(e$strategy[i], split = ".", fixed = T)[[1]][2]))
 
-f = e %>% filter(rank >= 6) %>% arrange(runID, lambda, l, type, - proportion) %>% 
+f = e %>% filter(type != "other") %>% arrange(runID, lambda, l, type, - proportion) %>% 
   group_by(runID, lambda, l, type) %>% mutate(rank = 1:n())
 
 plot.data = f %>% group_by(l, lambda, type, rank) %>%

### 2018-03-17 paper/cogsci-submission2.tex
minor edits intro and RMD section

@@
   According to standard linguistic theory, the meaning of an utterance is the product of
-  conventional semantic meaning and general pragmatic rules on language use. To investigate how
-  cultural evolution of language plays out under this picture of the semantics-pragmatics
-  interface, we present a game-theoretic model of the competition between types of language
-  users, each endowed with certain lexical representations and a particular pragmatic
-  disposition to act on them. Our model traces two evolutionary forces and their interaction:
-  (i) pressure towards communicative efficiency and (ii) transmission perturbations during the
-  acquisition of linguistic knowledge. We illustrate the model based on a case study on scalar
-  implicatures, which suggests that the relationship between underspecified semantics and
-  pragmatic inference is one of co-evolution. The evolution of underspecified semantics
-  requires pragmatic reasoning, while pragmatic reasoning may not evolve if lexical meanings do
-  not require pragmatic enrichment.
+  conventional semantic meaning and general pragmatic rules on language use. We investigate how
+  such a division of labor between semantics and pragmatics could evolve under general
+  processes of selection and learning. We present a game-theoretic model of the competition
+  between types of language users, each endowed with certain lexical representations and a
+  particular pragmatic disposition to act on them. Our model traces two evolutionary forces and
+  their interaction: (i) pressure towards communicative efficiency and (ii) transmission
+  perturbations during the acquisition of linguistic knowledge. We illustrate the model based
+  on a case study on scalar implicatures, which suggests that the relationship between
+  underspecified semantics and pragmatic inference is one of co-evolution.
 \end{abstract}
@@
 
-In contrast, we will here look at language users with a richer linguistic make-up. The objects
-of replication and selection in our model are pairs consisting of a set of lexical meanings and
-a manner of pragmatic behavior, which together define an agent's linguistic choices. Types of
-pragmatic behavior are represented using probabilistic models of pragmatic language use
+In contrast, we will here represent explicitly language users' representations of lexical
+meanings and their own particular manner of manner of production and interpretation based on
+these lexical representation. Different types of pragmatic behavior are represented using
+probabilistic models of pragmatic language use
 \citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}. Replication
@@
 such as inductive learning biases and sparse learning data
-\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,KirbyGriffith2014:Iterated-Learni}. Considering
-the effects that the transmission of linguistic knowledge through (iterated) learning have is
-important because neither semantic meaning nor pragmatic usage patterns are directly
+\citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,KirbyGriffith2014:Iterated-Learni}.
+
+It is important to consider the effects of transmission of linguistic knowledge through
+(iterated) learning because neither semantic meaning nor pragmatic usage patterns are directly
 observable. Instead, language learners have to infer these unobservables from the observable
@@
 for communicative efficiency and learnability. The results of this study are critically
-assessed in the light of the assumptions that feed our model in Section~\ref{sec:discussion}.
+assessed in the light of the simplifying assumptions inherent in our abstract model in Section~\ref{sec:discussion}.
 
@@
 
-The emergence and change of linguistic structure is influenced by many intertwined
-factors. These range from biological and socio-ecological to cultural ones \citep{benz+etal:2005b,steels:2011,tamariz+kirby:2016}. Social and ecological pressures determine communicative needs, while
-biology determines the architecture that enables and constrains the means by which they can be
-fulfilled. In the following, our focus lies on cultural aspects, wherein processes of
-linguistic change are viewed as shaped by language use and its transmission, i.e., as a result
-of a process of cultural evolution
+The emergence and change of linguistic structure is influenced by many factors in complex ways
+\citep{benz+etal:2005b,steels:2011,tamariz+kirby:2016}. Social and ecological pressures
+determine communicative needs, while biology determines the architecture that enables and
+constrains the means by which they can be fulfilled. In the following, our focus lies on
+cultural aspects, wherein processes of linguistic change are viewed as shaped by language use
+and its transmission, i.e., as a result of a process of cultural evolution
 \citep{Pagel2009:Human-Language-,ThompsonKirby2016:Culture-Shapes-}.
@@
   piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation or
-reinforcement, favor more communicatively efficient types of behavior, languages are driven
-towards semantic expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But
-pressure towards communicative efficiency is not the only force that shapes
-language. Learnability is another, as natural languages need to be learnable to survive their
-faithful transmission across generations. Furthermore, the effects of even small learning
-biases can have striking effects on an evolving language in a process of iterated learning
+reinforcement, favors communicative efficiency, languages are driven towards semantic
+expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But pressure towards
+communicative efficiency is not the only force that shapes language. Learnability is another.
+Natural languages need to be learnable to survive their faithful transmission across
+generations. Furthermore, the effects of even small learning biases can have striking effects
+on an evolving language in a process of iterated learning
 \citep{KirbyHurford2002:The-Emergence-o,SmithKirby2003:Iterated-Learni,kirby+etal:2014}.
@@
 that it is able to abstract away from what individual agents do. It describes the most probable
-trajectory of change in the population vector $\vec{x}$, which codifies relative frequencies of
-types, so that $x_i$ is the relative frequency of type $i$ and $\sum\nolimits_i x_i = 1$. In
-its simplest discrete-time formulation, the RMD defines the frequency $x'_i$ of each type $i$
-in the population at the next time step as a function of: (i) the frequency $x_i$ of each type
-$i$ before the update step, (ii) the fitness $f_i$ of each type $i$ before the update, and
-(iii) the probability $Q_{ji}$ that an agent who observes the behavior of an agent with type
-$j$ ends up acquiring type $i$ (with $\sum_k Q_{jk} = 1$ for all $j$):
+trajectory of change in the population vector $\vec{x}$, in which each component $x_i$ gives
+the relative frequency of type $i$ and $\sum\nolimits_i x_i = 1$. In its simplest discrete-time
+formulation, the RMD defines the frequency $x'_i$ of each type $i$ in the population at the
+next time step as a function of: (i) the frequency $x_i$ of each type $i$ before the update
+step, (ii) the fitness $f_i$ of each type $i$ before the update, and (iii) the probability
+$Q_{ji}$ that an agent who observes the behavior of an agent with type $j$ ends up acquiring
+type $i$ (with $\sum_k Q_{jk} = 1$ for all $j$):
 \begin{align}
@@
 
-The RMD is an abstract description of population-level changes. It abstracts away from specific
-assumptions about how agents interact and what causes selection and mutation. To understand
-this better, the following looks at each component in isolation: first replication, then
-mutation.
+The RMD is an abstract description of population-level changes. It abstracts away, to the
+extent possible, from specific assumptions about how agents interact and what exactly causes
+selection and mutation. To understand this better, the following looks at each component in
+isolation: first replication, then mutation.
 
@@
   given occasion. Such minimalism about the strategic capacities of agents is a conservative
-  methodological choice. It allows us to study ---as the formal thought experiment that any
-  such model is--- how fitness-based selection changes a population in the absence of optimal
-  agent-level reasoning and it abstracts away from specific assumptions about how fitness-based
-  selection might be realized concretely.}
+  methodological choice. It is one thing to be able to demonstrate that phenomenon $X$ is what
+  a rational agent would do. This is useful but may not always be the most satisfactory answer
+  to the question how $X$ happened to evolve. It is another thing to show that even if no agent
+  was individually capable of seeing $X$ as the rational thing to do, $X$ would still evolve if
+  every agent at least sometimes, locally and myopically tried to optimize its behavior, e.g.,
+  by imitating what seems to be working well for others \citep{Skyrms2010:Signals}.}
 
@@
 such as in conditional imitation, $Q_{ji}$ is the probability that an agent trying to learn,
-adopt or imitate the behavior of an agent with type $j$ acquires type $i$. As detailed below,
-in this way, the process in (\ref{eq:RMD_discrete}), and its equivalent in
-(\ref{eq:RMD_discrete_recast}), can contain a model of iterated learning
-\citep{griffiths+kalish:2005,griffiths+kalish:2007}.
+adopt or imitate the behavior of an agent with type $j$ acquires type $i$.
 
@@
 mutator component as the perturbation induced by how easy different linguistic traits are
-learnable from observation of language use. Nothing of current relevance hinges on whether the
-replicator component is interpreted as a biological process or a cultural process. We believe
-that this flexibility of interpretation is an asset. It abstracts away from specific
-assumptions and focuses on a general and versatile description of fitness-based
+learnable from observation of language use. In this way, the RMD contains a chain of iterated
+(Bayesian) learning as a special case
+\citep{griffiths+kalish:2005,griffiths+kalish:2007}. Nothing of current relevance hinges on
+whether the replicator component is interpreted as a biological process or a cultural
+process. We believe that this flexibility of interpretation is an asset. It abstracts away from
+specific assumptions and focuses on a general and versatile description of fitness-based
 selection. Still, to have something concrete in mind when interpreting the following

### 2018-03-17 paper/cogsci-submission2.tex
fixes up to sec 3

@@
 \label{sec:languages+use}
-Types are what evolution operates on. A type is what determines an agent's behavior and thereby
+Types are what evolution promotes or demotes. A type determines an agent's behavior and thereby
 an agent's fitness, so that it makes sense to speak of the fitness of a type itself. To study
@@
 message for each state) and a probabilistic hearer rule (a probabilistic choice of state for
-each message) given a lexicon. There are many ways of making these general notions
-concrete. In the following we aim at a general formulation of speaker and listener
-behavior which is as simple as possible but still in line with both a game-theoretic tradition
-\citep[e.g.,][]{benz:2006,benz+vRooij:2007,jaeger:2007b,franke+jaeger:2014} and a Bayesian
-tradition
-\citep[e.g.,][]{frank+goodman:2012,goodman+stuhlmueller:2013,GoodmanFrank2016:Pragmatic-Langu}.
+each message) given a lexicon.
 
@@
 Linguistic strategies define dispositions to produce and interpret messages given a lexicon. We
-distinguish between two kinds strategies. {\em Literal interlocutors} produce and
-interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
-instead engage in mutual reasoning to inform their choices. Recent probabilistic models of
-rational language use
-\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
-capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
-level $0$, corresponds to literal language use, as in Equations~(\ref{h:level0}) and
-(\ref{s:level0}). Pragmatic language users of level $n + 1$ act (approximately) rational with
-respect to level-$n$ behavior of their interlocutors, as in Equations~(\ref{h:leveln}) and
-(\ref{s:leveln}).
+distinguish between two kinds strategies. {\em Literal interlocutors} produce and interpret
+messages literally, being guided only by their lexica. {\em Pragmatic interlocutors} instead
+engage in mutual reasoning to inform their choices. Recent game-theoretic
+\citep[e.g.,][]{benz:2006,benz+vRooij:2007,jaeger:2007b,franke+jaeger:2014} and probabilistic
+models of rational language use
+\citep[e.g.,][]{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
+capture different types of pragmatic behavior in a reasoning hierarchy. In the following we aim
+at a general formulation of speaker and listener behavior which is as simple and as practical
+as possible for our current purposes, but
+still in line with both the game-theoretic and the Bayesian traditions.
+
+The hierarchy's bottom, level $0$, corresponds to literal language use, as in
+Equations~(\ref{h:level0}) and (\ref{s:level0}). Pragmatic language users of level $n + 1$ act
+(approximately) rational with respect to level-$n$ behavior of their interlocutors, as in
+Equations~(\ref{h:leveln}) and (\ref{s:leveln}).
 \begin{alignat}{3}
@@
 will assign to or choose the correct meaning.\footnote{An alternative choice for the speaker's
-  utility function is $\log H_{n}(s|m;L)$. Conceptually, the difference is that without the
-  logarithm, utilities are derived from action choices, while with the logarithm utilities are
-  derived from how closely the listener's beliefs (after hearing an utterance) align with the
-  speaker's, as measure by Kullback-Leibler divergence
-  \cite{frank+goodman:2012,qing+franke:2015}. In an evolutionary context, where fitness comes
-  from actual interactions with the environment, not just beliefs held about it, the
+  utility function, frequently made in Bayesian approaches after \citet{frank+goodman:2012}, is
+  $\log H_{n}(s|m;L)$. Conceptually, the difference is that without the logarithm, utilities
+  are derived from action choices, while with the logarithmic utilities are derived from how
+  closely the listener's beliefs (after hearing an utterance) align with the speaker's, as
+  measure by Kullback-Leibler divergence. In an evolutionary context, where fitness comes from
+  actual interactions with the environment, not just solemn beliefs held about it, the
   formulation without the logarithm may be slightly more appealing. Moreover, log-based speaker
@@
   in which any type has a reasonable chance to be mistaken for any other based on their
-  production behavior (see Section~\ref{sec:learnability}), we want a definition of speaker
-  types which also send any message in any state with positive probability.} For literal
-speakers, utility only tracks truthfulness. Literal speakers choose any true message with equal
-probability but may send false messages as well with a probability dependent on $\lambda$.
+  production behavior (see Section~\ref{sec:learnability}), we want speaker types which send
+  any message in any state with positive probability.} For literal speakers, utility only
+tracks truthfulness. Literal speakers choose any true message with equal probability but may
+send false messages as well with a probability dependent on $\lambda$.
 
@@
 as the latter are the simplest pragmatic reasoning types which show a tendency to
-communicatively attuned pragmatic enrichment. Using only level 1 reasoning and possibly small
-$\lambda$ is therefore a conservative choice that works \emph{against} the fitness-based
-selection of pragmatic language use for a notion of fitness defined as communicative success,
-which is introduced next.
+communicatively attuned pragmatic enrichment.
 
@@
 
-Languages are shaped not only by functionalist forces towards greater communicative
-success. Another important factor is the fidelity with which linguistic knowledge is
-transmitted. In biological evolution, where types are expressed genetically, transmission
-infidelity comes into the picture through infrequent and mostly random mutation and genetic
-drift \citep{kimura:1983}. However, an agent's lexicon and a disposition for pragmatic
-reasoning are likely not inherited genetically. They need to be learned from observation, such
-as in conditional imitation (see Appendix~\ref{sec:deriv-repl-mutat}).  Concretely, when agents
-learn from or imitate type $j$, they observe the overt linguistic behavior of type $j$ and need
-to infer the covert type that most likely produced the observed behavior.
+In biological evolution, where types are expressed genetically, transmission infidelity comes
+into the picture through infrequent and mostly random mutation and genetic drift
+\citep{kimura:1983}. However, an agent's lexicon and a disposition for pragmatic reasoning are
+likely not inherited genetically. They need to be learned from observation, such as in
+conditional imitation (see Appendix~\ref{sec:deriv-repl-mutat}).  Concretely, when agents learn
+from or imitate type $j$, they observe the overt linguistic behavior of type $j$ and need to
+infer the covert type that most likely produced the observed behavior.
 
@@
 requires to faithfully acquire the language of the teacher
-\citep[cf.][450]{griffiths+kalish:2007}. Put differently, types favored by the prior are easier
-to learn because they require less data than less favored types to be inferred, even if they
-are both compatible with the learning data at hand.
-
-The extent of the prior's influence has been shown to heavily depend on the learning strategy
-assumed to underlie the inference process. Early simulation results suggested that weak biases
-could be magnified by exposing learners to only small data samples
-\citep[e.g. in][]{brighton:2002}. \citet{griffiths+kalish:2007} showed that iterated learning
-converges to the prior in the limit, i.e., that the resulting distribution over types
-corresponds to the learners' prior distribution and is not influenced by the amount of input
-given to them, if learners select random hypothesis from their posterior beliefs about which
-types may have generated the learning data that they saw. More deterministic strategies such as
-the adoption of the type with the highest posterior probability, so-called {\it maximum a
-  posterior estimation} (MAP), increase the influence of both the prior and the data
-\citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we use a parameter $l\ge1$ to
-modulate between posterior sampling and the MAP strategy. When $l = 1$ learners sample from the
-posterior. The learners' propensity to maximize the posterior grows as $l$
-increases.\footnote{The way the posterior over types is formed and how learners employ it are
-  important design choices. There are other alternatives as well. For instance, instead of
-  adopting a single type, learners could keep their uncertainty over types and base their
-  production behavior on the posterior over types they form \citep{burkett+griffiths:2010}. A
-  straightforward operationalization of this idea would then see production defined as a
-  two-step process wherein a speaker first samples from her posterior and then produces
-  utterances based on the type she sampled, as done in \citet{kirby+etal:2015}. Beyond adding
-  complexity to the model, a disadvantage of this approach is that it could happen that a
-  speaker that first uttered a Hungarian utterance speaks Quechua in the next, so to speak. We
-  therefore maintain the simpler assumption of a single type being acquired in the present
-  context, and postpone further investigation of more complex type-adoption schemes to future
-  research.}
+\citep[cf.][450]{griffiths+kalish:2007}.
+
+The prior's influence depends on particulars of the learning or inference process. Early
+simulation results suggested that weak biases could be magnified by exposing learners to only
+small data samples \citep[e.g. in][]{brighton:2002}. \citet{griffiths+kalish:2007} proved that
+if learners adopt a random type as a sample from their posterior beliefs about which types may
+have generated the learning data that they saw, the population distribution of types will
+eventually come to match the learners' prior distribution over types exactly. More
+deterministic strategies such as the adoption of the type with the highest posterior
+probability, so-called {\it maximum a posterior estimation} (MAP), increase the influence of
+both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we
+use a parameter $l\ge1$ to modulate between posterior sampling and the MAP strategy. When
+$l = 1$ learners sample from the posterior. The learners' propensity to maximize the posterior
+grows as $l$ increases.\footnote{The way the posterior over types is formed and how learners
+  employ it are important design choices. There are other alternatives as well. For instance,
+  instead of adopting a single type, learners could keep their uncertainty over types and base
+  their production behavior on the posterior over types they form
+  \citep{burkett+griffiths:2010}. A straightforward operationalization of this idea would then
+  see production defined as a two-step process wherein a speaker first samples from her
+  posterior and then produces utterances based on the type she sampled, as done in
+  \citet{kirby+etal:2015}. Beyond adding complexity to the model, a disadvantage of this
+  approach is that it could happen that a speaker that first uttered a Hungarian utterance
+  speaks Quechua in the next, so to speak. We therefore maintain the simpler assumption of a
+  single type being acquired in the present context, and postpone further investigation of more
+  complex type-adoption schemes to future research.}
 
@@
 \citet{griffiths+kalish:2007}, we know that under $l=1$ iterated Bayesian learning will
-converge to a population proportion that reflects that prior over types. If $l>1$ learners tend
-to pick types which are \emph{a posteriori} more likely with a higher probability than less
-likely types. Even in the absence of a prior bias, iterated Bayesian learning with $l>1$ can
-lead to populations in which a particular type $t_i$ has a higher proportion than other
-types. For $l>1$, much hinges on particular asymmetries in production likelihoods
+converge to a population proportion that exactly matches the agents' prior over types. If $l>1$
+learners tend to pick types which are \emph{a posteriori} more likely with a higher probability
+than less likely types. Even in the absence of a prior bias, iterated Bayesian learning with
+$l>1$ can lead to populations in which a particular type $t_i$ has a higher proportion than
+other types. For $l>1$, much hinges on particular asymmetries in production likelihoods
 $P(d \mid t_j)$, and therefore on the composition of the set of types in general. Intuitively
-put, if there are many types $t_j$ which all happen to produce data that is most likely to be
-produced by $t_i$, this can lead to a larger proportion of $t_i$ in the long run even when
-priors over types are uniform. Appendix~\ref{sec:example-iter-bayes} provides an abstract
-example.
+put, if there are many types $t_j$ which all happen to (``erroneously'') produce data that is
+most likely produced (``correctly'') by $t_i$, this can lead to a larger proportion of $t_i$ in
+the long run even when priors over types are uniform. Appendix~\ref{sec:example-iter-bayes}
+provides an abstract example for these likelihood driven effects.
 

### 2018-03-17 paper/cogsci-submission2.tex
minor fixes first sections

@@
 \end{align*}
-If $\delta > \epsilon$, production likelihoods are asymmetrically skewed to possibly favor
-$t_1$, in the sense that both types $t_2$ and $t_3$  produce behavior that is strong
-evidence for $t_1$ more often than behavior that is strong evidence for any other
-type other then themselves. Type $t_1$ makes no such ``asymmetric mistakes''.
+If $\delta > \epsilon$, production likelihoods are asymmetrically skewed to favor $t_1$, in the
+sense that both types $t_2$ and $t_3$ produce behavior that is strong evidence for $t_1$ more
+often than behavior that is strong evidence for any other type different from themselves. Type
+$t_1$ makes no such ``asymmetric mistakes'' and is more faithfully recognized based on its
+production behavior than the other types.
 

### 2018-03-18 paper/cogsci-submission2.tex
paper edits whole paper

@@
 
-This section engages in a formal thought experiment to address the question: if a population of
-language users could freely combine different lexica with different dispositions to act on
-them, what are conditions under which the majority view of scalar implicatures could have
-evolved? --- Recall that the majority view is that scalar implicatures are non-lexicalized
-pragmatic enrichments. Scalar implicature triggers like \emph{some}, \emph{warm} or \emph{may}
-are semantically weak expressions for which logically stronger expressions are salient, e.g.,
-\emph{all}, \emph{hot} or \emph{must}. For instance, {\em some} is entailed by {\em
-  all}. Still, the use of a less informative expression (\emph{some}) when a more informative
-one (\emph{all}) could have been used can license a defeasible inference (\emph{some but not
-  all}) that stronger alternative is false (cf. \citealt{horn:1972,gazdar:1979}). According to
-the majority view, this is a pragmatic inference, not part of the conventional meaning.
-
-In the following we consider what are perhaps two of the simplest non-trivial setups that speak
-to this matter and reflect on their limitations in Section~\ref{sec:discussion}. Section
-\ref{subsec:example} first investigates a minimal example intended to isolate effects of
-replication and mutation and to gain insights into the sense in which semantic meaning and
-pragmatic use can be said to co-evolve. We then turn to a larger and more realistic setup in
+This section addresses the question under which conditions the division of labor, which was
+described in Section~\ref{sec:introduction}, between underspecified semantics and pragmatic
+enrichment could have evolved. We consider what are perhaps two of the simplest non-trivial
+setups that speak to this matter and reflect on their limitations in
+Section~\ref{sec:discussion}. Section \ref{subsec:example} first investigates a minimal example
+intended to isolate effects of replication and mutation and to gain insights into the sense in
+which semantic meaning and pragmatic use could be said to co-evolve. We then turn to a larger and
+more realistic setup in
 Section~\ref{sec:setup:-stat-conc}. Section~\ref{sec:simulation-results} describes simulations
@@
 and pragmatic use in the simplest non-trivial type space, which consists of only the four types
-from Section~\ref{sec:languages+use}: a type has either lexicon $\Llack$ or $\Lbound$ and is
+from Section~\ref{sec:languages+use}: a type has either lexicon $\Llack$ or $\Lbound$ and it is
 either a literal (level-$0$) or a pragmatic (level-$1$) language user. We may think of such a
@@
 enrichment can co-evolve. These abstract theoretical considerations also suggest that we need
-both fitness based selection of communicative efficiency and pressure towards learnability to
+both fitness-based selection of communicative efficiency and pressure towards learnability to
 see underspecified semantics and pragmatic language use emerge together.
@@
 steps shown in the top row of Fig.~\ref{fig:quiver} make the functional disadvantage of
-$\Llack$ against $\Lbound$ apparent, particularly for its literal users and/or low
-$\lambda$. They also highlight that the functional difference between literal and pragmatic
-$\Lbound$ is small. Under functional pressure only, populations move toward the lower part of
-the square but the steps become small as the lower edge, representing populations composed only
-of $\Llack$ users, are reached. As for the mutator diagrams (bottom row of
-Fig.~\ref{fig:quiver}), there are three main things to note. First, as expected from the
-literature, a small bias for a type will lead to larger proportions to adopt it under $l=1$
-(see lower left space in Figure~\ref{fig:quiver}). Second, if $l>0$, the dynamic is influenced
-heavily also by asymmetries in production likelihoods. Third, these asymmetries in the
-production likelihoods are in turn influenced by $\lambda$, so that $\lambda$ also has an
-effect on iterated Bayesian learning (compare the middle and right-most diagram in the bottom
-row of Fig.~\ref{fig:quiver}).
-
-Fig.~\ref{fig:sketch} and \ref{fig:quiver} investigate replicator and mutator steps in
+$\Llack$ against $\Lbound$ apparent. With increasing $\lambda$, this difference pragmatic users
+of $\Lbound$ are no longer selected against by fitness-based selection.
+
+As for the iterated learning only, the diagrams in the bottom row of Fig.~\ref{fig:quiver} show
+the case of a prior which favors types with underspecified semantics ($\Llack$) over those with
+lexicalized upper bounds ($\Lbound$) by a factor of $1.05$. There are three main things to
+note. First, as expected from the literature, a small bias for a type will lead to larger
+proportions of types who adopt it under $l=1$ (see lower left space in Figure~\ref{fig:quiver},
+where the system gravitates indeed to a population where the relative frequency of $\Llack$
+over $\Lbound$ is exactly $1.05$). Second, if $l>0$, the dynamic is influenced heavily also by
+asymmetries in production likelihoods. Third, these asymmetries in the production likelihoods
+are in turn influenced by $\lambda$, so that $\lambda$ also has an effect on iterated Bayesian
+learning (compare the middle and right-most diagram in the bottom row of
+Fig.~\ref{fig:quiver}).
+
+Fig.s~\ref{fig:sketch} and \ref{fig:quiver} investigate replicator and mutator steps in
 isolation. It remains to investigate how exactly these forces shape a population over time when
 they apply together. This will also hinge on which types are present because the space of all
-types determines whether production likelihoods asymmetrically favor certain types over others
-(when $l>1$; see Appendix~\ref{sec:example-iter-bayes}). For this reason, we turn to a larger
-setup next with a larger and more realistic set of types. Additionally, we want to explore
-the effects of a learning bias in favor of simple lexical representations. One way of
-motivating and formalizing such a bias is introduced in Section~\ref{sec:an-induct-learn}.
+types determines how much production likelihoods asymmetrically favor certain types over others
+(when $l>1$; see Appendix~\ref{sec:example-iter-bayes}). For this reason, we turn to a somewhat
+more realistic but still manageable setup next with a larger set of types.
 
@@
 nonetheless (see below). Second, lexica that conventionalize upper-bounds to realize a
-one-to-one mapping of messages to states. Finally, lexica that do not lexicalize an upper
-bound but allow it to be conveyed pragmatically due to the presence of a stronger lexical
-item. There are six lexica of the second kind and six of the third.  The following three lexica
-exemplify each kind:
+one-to-one mapping of messages to states. Finally, lexica that do not lexicalize upper bounds
+but would allow for perfect communication under additional pragmatic strengthening. There are
+six lexica of the second kind and six of the third.  The following three lexica exemplify each
+kind:
 \begin{align*}
@@
 \end{align*}
-Lexicon $\Lall$ is clearly bad for communication: all message and interpretation choices will
-be equally likely; no information about the observed world state will be conveyed
-by its users. In contrast, users of $\Lbound$ can communicate world states perfectly, no matter
-whether they are literal or pragmatic users. Users of $\Llack$ can also communicate information
-about the actual world state but need pragmatic language use to approximate a one-to-one
-mapping between message use and states (see Section~\ref{sec:languages+use}).
+% Lexicon $\Lall$ is clearly bad for communication: all message and interpretation choices will
+% be equally likely; no information about the observed world state will be conveyed
+% by its users. In contrast, users of $\Lbound$ can communicate world states perfectly, no matter
+% whether they are literal or pragmatic users. Users of $\Llack$ can also communicate information
+% about the actual world state but need pragmatic language use to approximate a one-to-one
+% mapping between message use and states (see Section~\ref{sec:languages+use}).
 
@@
 
-Note that while
-different types may lexicalize the same representations, they may nevertheless map different
-states to different overt messages. More informally, they speak different languages that
-lexicalize the same concepts. Consequently, more often than not different types of the same
-kind fail to understand each other.
+Note that while different types may lexicalize the same representations, they may nevertheless
+map different states to different overt messages. Consequently, different types of the same
+kind can fail to understand each other completely.
 
@@
 
-A complexity measure for lexical representations from Table~\ref{tab:concepts} is used to define a
-learning bias that favors simpler representations over more complex ones. The prior probability of a
-type is just the prior probability of its lexicon. The prior of a lexicon is a function of the
-complexity of the lexical representations in its image set. Lexica with simpler representations accordingly have a
-higher prior. One simple way of defining such priors over lexica (and thereby types) is:
+The complexity measures for lexical representations from Table~\ref{tab:concepts} are used to
+define a learning bias that favors simpler representations over more complex ones. The prior
+probability of a type is just the prior probability of its lexicon. The prior of a lexicon is a
+function of the complexity of the lexical representations in its image set. Lexica with simpler
+representations accordingly have a higher prior. One simple way of defining such priors over
+lexica (and thereby types) is:
 \begin{align*}
@@
 convenient operationalization of one general approach to explicating learning biases; this is
-not a commitment that this general approach is necessarily superior or that, within it, this
-particular instrumentalization is the single most plausible. We return to this issue in
-Section~\ref{sec:discussion}.
+not a implicit commitment to the claim that this particular instrumentalization is the single
+most plausible. We merely want to be concrete to have a maximally concrete working example in
+which the priors over types are systematically related to the complexity of lexical
+representations. We return to this issue in Section~\ref{sec:discussion}.
 
@@
 \emph{a priori} than lexica of the kind $\Llack$. Simulation results will shed light on the
-question whether target types can emerge, and for which parameter constellations.
-
-To better understand the workings of a pressure towards communicative efficiency and a pressure
-towards learnability, we look at the behavior of the replicator and mutator step first in
-isolation, and then in combination. All simulation runs are initialized with an arbitrary
-distribution over types, constituting a population's first generation. All reported results are
-the outcome of 50 update steps. These outcomes correspond to developmental plateaus in which
-change is, if not absent, then at least very slow. In other words, even if the resulting states
-do not correspond to an eventual attracting state, they characterize population states in which the system remains for a long time. As specified in Section~\ref{sec:learnability}, the
-mutation matrix $Q$ can be obtained by considering all possible state-message sequences of
-length $k$. Given that this is intractable for large $k$, the sets of data which learners are
-exposed to are approximated by sampling $250$ $k$-length sequences from each type's production
-probabilities.
+question whether target types can emerge, and for which parameter constellations this is
+likely.
+
+As before, we first look at the behavior of the replicator and mutator step in isolation, and
+then in combination. All simulation runs are initialized with an arbitrary distribution over
+types, constituting a population's first generation. All reported results are the outcome of 50
+update steps. These outcomes correspond to developmental plateaus in which change is, if not
+absent, then at least very slow. In other words, even if the resulting states do not correspond
+to an eventual attracting state, they characterize population states in which the system
+remains for a long time. As specified in Section~\ref{sec:learnability}, the mutation matrix
+$Q$ can be obtained by considering all possible state-message sequences of length $k$. Given
+that this is intractable for large $k$, the sets of data which learners are exposed to are
+approximated by sampling $250$ $k$-length sequences from each type's production probabilities.
 
@@
 requires sufficiently high $\lambda$ but does not often single out types with an underspecified
-semantic representation. \mf{rewrite this paragraph or section}
+semantic representation.
 
@@
 \end{figure}
+
+\mf{Possibly change plot on the left in Figure~\ref{fig:only-M} and adapt text
+  accordingly. Write that we know that happens under $l = 5$ in the long run (even if we do not
+  see it after 50 steps yet, if that is indeed so). If it is so, we could leave the plot and
+  say: long term development is nice, but when fitness-based selection interferes at every step
+  we do not care so much for infinite time, but rather for short time (actually one-step). If
+  it is not so, insert a plot for $l>1$ that shows something interesting.}
 
@@
 
-As with a pressure only for communicative success, low values of $\lambda$ make the differences in observable behavior across types less pronounced.
-This makes differences in the likelihood of particular types having generated a learning input less pronounced.
-Therefore, low $\lambda$ leads populations to reflect the learners' inductive bias more faithfully.
-This favors functionally deficient but a priori preferred types such as those that use  $\Lall$.
-A pressure for learnability alone may consequently lead to a spread of communicatively suboptimal types that lexicalize simpler semantics.
-In the extreme, when $l = 1$ and $\lambda = 1$ all
-of $1000$ independent populations had users of $\Lall$ as majority types.
-For higher $\lambda$, it becomes clear that a high prior is not the only thing that counts when it comes to learnability (see Section~\ref{subsec:example}).
-As soon as there is information for learners to discern whether one type is more likely to have generated the data (depending on $\lambda$ and $k$), it becomes paramount for types to produce data that makes them easily identifiable if they are to be inferred more often.
+As with a pressure only for communicative success, low values of $\lambda$ make the differences
+in observable behavior across types less pronounced. This makes differences in the likelihood
+of particular types having generated a learning input less pronounced. Therefore, low $\lambda$
+leads populations to reflect the learners' inductive bias more faithfully. This favors
+functionally deficient but a priori preferred types such as those that use $\Lall$. A pressure
+for learnability alone may consequently lead to a spread of communicatively suboptimal types
+that lexicalize simpler semantics. In the extreme, when $l = 1$ and $\lambda = 1$ all of $1000$
+independent populations had users of $\Lall$ as majority types. For higher $\lambda$, it
+becomes clear that a high prior is not the only thing that counts when it comes to learnability
+(see Section~\ref{subsec:example}). As soon as there is information for learners to discern
+whether one type is more likely to have generated the data (depending on $\lambda$ and $k$), it
+becomes paramount for types to produce data that makes them easily identifiable if they are to
+be inferred more often.
+
+In sum, when pressured for learnability, pragmatic $\Llack$ is promoted over functionally
+similar but semantically more complex alternatives such as $\Lbound$. However, learnability
+alone does not foment the propagation of a single target type across the population, because it
+does not differentiate between different ways of mapping the same semantic representations onto
+different overt signals.
 
 \subsubsection{Combining pressures of communicative success and learnability}
-
-Pressure for communicative success and learnability are not sufficient on their own to have a
-single target type dominate the population. When pressured for communicative success, the  communicative advantage of $\Lbound$ users leads to their prevalence. When pressured for
-learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more
-complex alternatives such as $\Lbound$. However, learnability alone does not foment the
-propagation of a single target type across the population.
 
@@
 
-Combining pressures for communicative efficiency and learnability can lead to the selection of
-a single target type (see Fig.~\ref{fig:rmd}). The proportion of a single majority target type
-increases with $\lambda$ and $l$.
+Pressure for communicative success and learnability are not sufficient on their own to have a
+single target type dominate the population. But the combination of both pressures can lead to
+the selection of a single target type (see Fig.~\ref{fig:rmd}). The proportion of a single
+majority target type increases with $\lambda$ and $l$.
 
@@
 prior more faithfully. This is due to the fact that the likelihood that a small sequence was
-produced by any type is relatively uniform (modulo prior). By contrast, larger values
-increasingly allow learners to differentiate types with different signaling behaviors.
+produced by any type is relatively uniform. By contrast, larger values increasingly allow
+learners to differentiate types with different signaling behaviors.
 
@@
 
-This case study should not be interpreted as a proposal for a definite explanation of how
-scalar implicatures evolved. Other factors should be considered eventually even if they will
-lead to much more complex modeling. One such factor is the observation that non-lexicalized
-upper bounds allow a broader range of applicability. For example, when the speaker is not
-certain as to whether \emph{all} is true. This may suggest an alternative and purely
+This case study is a first attempt at an explanation of how scalar implicatures evolved. But
+other factors, which are presently not taken into account, should be considered eventually even
+if they will lead to much more complex modeling. One such factor is the observation that
+non-lexicalized upper bounds allow a broader range of applicability, e.g., when the speaker is
+not certain as to whether \emph{all} is true. This may suggest an alternative and purely
 functionalist argument for why upper-bounded meanings do not conventionalize: should contextual
@@
 conveyed pragmatically, then this is preferred over expressing it overtly through longer
-expressions. For example, by saying {\em some but not all} explicitly. Importantly, although
+expressions, e.g., by saying {\em some but not all} explicitly. Importantly, although
 morphosyntactic disambiguation may be dispreferred due to its relative length and complexity
@@
 knowledge in acquisition. Its objects of selection and replication are pairs of lexical
-meanings and patterns of language use. This allows the model to trace the evolutionary interaction between
-conventional meaning and pragmatic use. Additionally, it takes the challenge seriously of
-neither semantics nor pragmatics being directly observable. Instead, learners need to infer
-these unobservables from overt data that results from their combination.  These components and
-their mutual influence were highlighted in a case study on the lack of lexical upper-bounds in
-weak scalar expression. This study showed that, when pressured for learnability and communicative
-success, the former force can drive for simpler semantic representations inasmuch as pragmatics
-can compensate for lack of expressivity in use. That is, the relative learning advantage of
-simpler semantics in combination with functional pressure in use may offer an answer to why
-natural languages fail to lexicalize systematic pragmatic inferences. And, more broadly, lead to a
-division of labor between semantics and pragmatics.
+meanings and patterns of language use. This allows the model to trace the evolutionary
+interaction between conventional meaning and pragmatic use. Additionally, it takes the
+challenge seriously of neither semantics nor pragmatics being directly observable. Instead,
+learners need to infer these unobservables from overt data that results from their combination.
+These components and their mutual influence were highlighted in a case study on the lack of
+lexical upper-bounds in weak scalar expression. This study showed that, when pressured for
+learnability and communicative success, the former force can drive for simpler semantic
+representations inasmuch as pragmatics can compensate for lack of expressivity in use. That is,
+the relative learning advantage of simpler semantics in combination with functional pressure in
+use may offer an answer to why natural languages fail to lexicalize systematic pragmatic
+inferences. Conversely, by appealing to economy of lexical representations, this model also
+suggests a rationale for why a system of pragmatic enrichment could have evolved in the first
+place. The resulting picture is one of co-evolution: the division of labor between semantics
+and pragmatics could have evolved because underspecified lexical meanings are easier, yet can
+only be maintained with a mechanism of pragmatic enrichment; pragmatic reasoning, however,
+would not evolve in the absence of lexically ambiguous representations.
 

### 2018-03-18 paper/rv-letter1.tex
repsonse letter edits

@@
 
+\definecolor{Red}{RGB}{178,34,34}
+\definecolor{Steelblue}{RGB}{70,130,180}
+\newcommand{\mf}[1]{\textcolor{Red}{[MF: #1]}} 
+\newcommand{\tb}[1]{\textcolor[rgb]{.8,.33,.0}{[TB: #1]}}% prints in orange
+\newcommand{\rvv}[1]{\textcolor{Steelblue}{#1}}% prints in steel blue
+
 \title{\Large Co-evolution of lexical meaning \& pragmatic use\\ {\bf Revision Cover letter}}
@@
 
-We would like to thank the reviewers for their helpful comments and suggestions. They are very welcome and have certainly helped improve the manuscript.
+We would like to thank the reviewers for their helpful comments and suggestions which have
+certainly helped improve the manuscript. The main changes in response to the reviewers comments
+are the following.
 
-We have taken care to improve the exposition of the model; to clarify the  motivations underlying the modeling choices we make in the case study, and to make more precise and explicit how the components of the model lead to the results they do.
-In what follows we address the reviewers' main comments and questions one by one, with pointers to respective changes effected on the manuscript.
+\begin{itemize}
+\item We have tried to improve the exposition of the model, both on the technical side and on
+  the way the model can be interpreted (re reviewer 1's concerns about how to interpret
+  particular formulas and the model as a whole). There are now three very short appendix
+  chapters, each with formal detail about how to better understand the abstract model. There is
+  also the new section~2.3 dedicated to an explicit discussion of how to interpret the replicator
+  mutator dynamic and a formal derivation of one particular agent-level update process which
+  gives rise to this dynamic at the population level (appendix B).
+\item Where previously we only had one simulation-based case study, we now complement these
+  results with a more in-depth case study, presented in the new section 3.1, which focuses on a
+  subset of types present in the larger case study presented in sections 3.2 and 3.3. This
+  helps tie in our results closer to known results from the literature (re reviewer 2's comment
+  2) and also helps understand the inner workings of the model better for the case at hand. It
+  also makes clearer in which sense we may speak of ``co-evolution of semantics and
+  pragmatics'' (in response to reviewer 1's comment).
+\item To compensate for the added material, we also tried to shorten the paper. We tried to
+  avoid repetitive formulations, delegated more detailed explanations to footnotes and omitted
+  certain passages that we feel are interesting, but are possibly also too distracting in a
+  paper that is already quite long and also partly rather technical.
+\end{itemize}
+
+In what follows we address the reviewers' main comments and questions one by one, with pointers
+to respective changes effected on the manuscript.
 
@@
 
-This is an important point that we now stress and highlight more throughout the paper.
-In particular the material surrounding Figure $2$ and $3$ in the new Section~2.4 illustrates how one factor has an impact on the evolution of the other.
-In a nutshell: a pragmatic strategy (level-$1$ reasoning) favors the evolution of underspecified semantics of the target kind ($L_{\text{lack}}$) as it allows for the maintenance of simpler lexical representations that are easier to learn without functional disadvantages otherwise incurred in communication (e.g., by level-$0$ reasoners).
-In the other direction, underspecified semantics of the target kind also favors the evolution of a pragmatic disposition to act on them.
-When paired with other semantics, pragmatic refinement can instead be (close to a) neutral trait (e.g., $L_{\text{bound}}$ paired with high $\lambda$) or it can even lead to  functional disadvantages when compared with level-$0$ reasoning (e.g., $L_{\text{bound}}$ with low $\lambda$).
-In this way both features benefit from each other.
+This is an important point that we now stress and highlight more throughout the paper. In
+particular, section 3.1 illustrates how one factor has an impact on the evolution of the
+other. In a nutshell: a pragmatic strategy (level-$1$ reasoning) favors the evolution of
+underspecified semantics of the target kind ($L_{\text{lack}}$) as it allows for the
+maintenance of simpler lexical representations that are easier to learn without functional
+disadvantages otherwise incurred in communication (e.g., by level-$0$ reasoners). In the other
+direction, underspecified semantics of the target kind also favors the evolution of a pragmatic
+disposition to act on them. When paired with other semantics, pragmatic refinement can instead
+be (close to a) neutral trait (e.g., $L_{\text{bound}}$ paired with high $\lambda$) or it can
+even lead to functional disadvantages when compared with level-$0$ reasoning (e.g.,
+$L_{\text{bound}}$ with low $\lambda$). In this way both features benefit from each other.
 
@@
 
-As we discuss in Section~1, our case study focuses on scalar implicatures because they are well-studied systematic pragmatic inferences that have received much attention, both theoretically and experimentally.
-This makes them suitable candidates for the study of the evolution of regular pragmatic inferences; past research can guide our modeling choices and aid in evaluating our application to the study of these inferences.
-This ties in with our particular choices to model the pragmatic use of scalar expressions.
-As we now make more explicit in \S 2.3.1, these draw from a growing game-theoretic and Bayesian tradition \citep[e.g.,][]{franke:2009,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}.
+As we discuss in Section~1, our case study focuses on scalar implicatures because they are
+well-studied systematic pragmatic inferences that have received much attention, both
+theoretically and experimentally. This makes them suitable candidates for the study of the
+evolution of regular pragmatic inferences; past research can guide our modeling choices and aid
+in evaluating our application to the study of these inferences. This ties in with our
+particular choices to model the pragmatic use of scalar expressions. As we now make more
+explicit in \S 2.3.1, these draw from a growing game-theoretic and Bayesian tradition
+\citep[e.g.,][]{franke:2009,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}.
 
@@
 \stepcounter{reviewerCommentCounter} %increase counter
-\noindent Reviewer \thereviewerCounter~asks for a clearer description of the model.
-As noted earlier, we have taken care to make the exposition more accessible and to guide the reader more through the technical details of Section 2.
-In particular, the newly added Section 2.4 should help readers unravel the model.
-We used the following questions from Reviewer 1, briefly answered below, to guide the changes made in Section~2.
+\noindent Reviewer \thereviewerCounter~asks for a clearer description of the model. As noted
+earlier, we have taken care to make the exposition more accessible and to guide the reader more
+through the technical details of Section 2. In particular, the newly added sections 2.3, 3.1,
+and the appendix sections should help readers unravel the model and understand its conceptual
+interpretation and inner mechanisms much better than before. We used the following questions
+from Reviewer 1, briefly answered below, to guide the changes made in this new version.
 
@@
 \end{mdframed}
-As it is unlikely that lexical representations and/or pragmatic reasoning are inherited genetically, we model the transmission of linguistic knowledge across generations as  iterated Bayesian learning.
-Technical details and motivations are given in \S 2.3.3.
+As it is unlikely that lexical representations and/or pragmatic reasoning are inherited
+genetically, we model the transmission of linguistic knowledge across generations as  iterated
+Bayesian learning. Technical details and motivations are given in \S 2.3 and appendix B.
 
@@
 \end{mdframed}
-As now made more explicit in \S2.2, steps of the discrete-time replicator mutator dynamic correspond to generational turn-overs of the entire population (definitions (1) and (2)).
-That is to say, all agents die at once, being replaced by the following generation in a single swoop.
+This question clearly demonstrates a failure on our part to make clear enough how the
+replicator mutator dynamic can be interpreted. To answer the question posed here directly: all
+of the above possibilities \emph{could} give rise to the replicator dynamic. It is a deliberate
+choice of abstract evolutionary modeling to \emph{not} to have to commit to any particular
+model, when it is also possible to answer what would happen under \emph{many} different regimes
+of fitness-based selection. To give a concrete interpretation of the population-level dynamic
+nonetheless, appendix B sketches the mean-field derivation of one form of an agent-level
+conditional imitation scheme.
 
@@
 \end{mdframed}
-The parenthetical remark is correct.
-The amount of teachers of a given type is proportional to the frequency of this type in the population after the replicator step.
-The population is infinite so there are technically infinite learners/teachers.
-This was implicit in the definitions of the replicator mutator dynamic in (1) and (2), and is now explicitly stated in Section~2.2 in the prose that motivates and explicates these definitions.
+The parenthetical remark is correct. The amount of teachers of a given type is proportional to
+the frequency of this type in the population after the replicator step. The population is
+infinite so there are technically infinitely many learners/teachers. This was implicit in the
+definitions of the replicator mutator dynamic in (1) and (2), and is now more explicitly stated
+in sections 2.2 and 2.3.
 
@@
 \end{mdframed}
-The prose surrounding definitions (1) and (2) in Section 2.2 now makes explicit that the population is infinite.
-This allows us to track change that does not depend on varying population sizes nor their growth rate.
-This is why we speak of proportions of a type in a population, rather than of there being a particular number of agents of a type.
+The prose surrounding definitions (1) and (2) in Section 2.2 and 2.3 now makes explicit that
+the population is infinite. This allows us to track change that does not depend on varying
+population sizes nor their growth rate. This is why we speak of proportions of a type in a
+population, rather than of there being a particular number of agents of a type.
 
@@
 \end{mdframed}
-While our model is compatible with the assumption that every agent comes with her own parameter values, the reviewer is correct in pointing out that we assume they all to have the same values (explicit in \S3.2).
-Variation across agents would mean having many more types: one for each combination of a possible lexicon, pragmatic strategy, $\lambda$-, $l$- and $k$-value.
-This would allow us to trace the evolutionary trajectory of not only lexica and pragmatic strategies, which is our goal here, but of all these factors combined.
-As mentioned in Section~$4$, this would be interesting future research.
+While our model is compatible with the assumption that every agent comes with her own parameter
+values, the reviewer is correct in pointing out that we assume they all to have the same values
+(explicit in \S3.3). Variation across agents would mean having many more types: one for each
+combination of a possible lexicon, pragmatic strategy, $\lambda$-, $l$- and $k$-value. This
+would allow us to trace the evolutionary trajectory of not only lexica and pragmatic
+strategies, which is our goal here, but of all these factors combined. As mentioned in
+Section~$4$, this would be interesting future research, but we cannot go into this here for
+reasons of space and a huge increase of complexity in the resulting model.
 
@@
 \end{mdframed}
-Learners witness sequences of data $d$. Such a sequence is composed of message-state pairings. In other words, learners do witness the true state of the world (Section~2.3.3, {\em learnability}).
+Learners witness sequences of data $d$. Such a sequence is composed of message-state
+pairings. In other words, learners do witness the true state of the world (Section~2.3.3, {\em
+  learnability}). This, however, is a widely made assumption in the literature on iterated
+learning.
 
-The probability of the learner receiving learning input $d$ depends only on the likelihood of the teacher producing it.
-This is determined by the teacher's production probabilities; the speaker behavior of her type.
-Communicative success accordingly does not come into play, only speaker behavior.
-It would certainly be possible to let learners witness proficient users communicate and to have them infer types from such interactions instead.
-However, we wanted to keep our assumptions close to the well-studied assumptions standardly made in the iterated (Bayesian) learning tradition.
-Iterated learning only considers production, rather than production and comprehension.
-Details on how learners infer types are given in Section~2.3.3.
+Additionally, some interpretations of the replicator mutator dynamic would require agents to
+have some rudimentary access to how good their own and/or other agents' behavior is. For
+example, the conditional imitation scheme sketched in appendix B assumes that agents imitate
+with a probability that depends on how good the possibly to be imitated behavior is. A
+biological interpretation or an interpretation in terms of reinforcement learning, for example,
+would not require this. In other words, even if this assumption is felt to be too strong, it is
+not necessary to discredit the model proposed here, because it is compatible with many
+agent-level update schemes, some of which are very innocuous when it comes to what each agent
+knows or sees or learns.
 
@@
 \end{mdframed}
-This parameter regulates the linguistic production behavior of types, i.e., speaker behavior.
-Speaker behavior is relevant both for communication with hearers as well as for the learning input that a teacher produces.
-When considering only iterated Bayesian learning $\lambda$ accordingly also plays a role because different $\lambda$ values yield different likelihoods of producing particular data (see above and Section~2.3.3).
+This parameter regulates the linguistic production behavior of types, i.e., speaker
+behavior. Speaker behavior is relevant both for communication with hearers as well as for the
+learning input that a teacher produces. When considering only iterated Bayesian learning
+$\lambda$ accordingly also plays a role because different $\lambda$ values yield different
+likelihoods of producing particular data. Section 3.2 and appendix B zoom into this topic and
+make this point more explicit. We regret that this was not ideally transparent in the previous
+version.
 
@@
 %
+\mf{We need to possibly rewrite this.}
 Due to their stochastic speaker behavior, the data produced by $L_{\text{all}}$-teachers tends to be compatible with the behavior of many other types.

### 2018-03-19 paper/rv-letter1.tex
repsonse letter

@@
 Learners witness sequences of data $d$. Such a sequence is composed of message-state
-pairings. In other words, learners do witness the true state of the world (Section~2.3.3, {\em
+pairings. In other words, learners do witness the true state of the world (Section~2.4.3, {\em
   learnability}). This, however, is a widely made assumption in the literature on iterated
@@
 
-We do not believe that, ultimately, the non-parsimonious view that there are different complexity measures for different structures should be assumed.
-However, with previous work in the iterated learning tradition, we do assume that (i) inductive learning biases can play a role in shaping culturally transmitted knowledge and that, as the reviewer notes, (ii) biases can be of manifold nature; another well-studied example being the mutual exclusivity bias, which could be expected to work against targets and in favor of competitors.
+First of all, the model does not rest on this assumption at all; only the particular application does.
+Following previous work in the iterated learning tradition, we do assume that (i) inductive learning biases can play a role in shaping culturally transmitted knowledge and that, as the reviewer notes, (ii) biases can be of manifold nature.
 
-At present, it is impossible to disentangle the complex interaction of multiple biases.
-We therefore focus on the effects of a single plausible contributing factor to a preference of target lexica over competitors: a well-motivated bias favoring simplicity \citep{feldman:2000,chater+vitanyi:2003, piantadosi+etal:2012a,kirby+etal:2015,piantadosi+etal:underreview}.
-By assumption, this translates to a preference not to lexicalize linguistic material over lexicalizing it; simpler formulae over complex ones.
-As mentioned in Section 3.1.2, the complexity measure that the prior is based on is just a convenient operationalization of this particular hypothesis.
+On the other hand, it is true that the model itself poses constraints on what might count as a plausible inductive bias.
+One well-studied example of an inductive bias from the iterated learning tradition is the mutual exclusivity bias.
+This bias should be expected to work against targets and in favor of competitors, as the reviewer says.
+Yet in the context of our model a bias for mutual exclusivity might be hard to motivate, because the model itself has an explicit component of fitness-based selection for communicative efficiency.
+This component should take the part of selecting for mutual exclusivity.
+In other words, the division of labor is shifted here: not every constraint is as easily justifiable; the model tries to give an explanation in terms or fitness-based selection for some of the effects that certain biases are meant to explain in systems which only look at iterated learning.
+
+What counts for this model is chiefly a bias that is \emph{not} motivated by the effects it has on communicative efficiency.
+Instead we only focus on the effects of a single plausible contributing factor to a preference of target lexica over competitors: a well-motivated bias favoring representational simplicity of single concepts, not lexica as a whole \citep{feldman:2000,chater+vitanyi:2003, piantadosi+etal:2012a,kirby+etal:2015,piantadosi+etal:underreview}.
+How to spell out such a lexical simplicity bias is, we readily admit, an open question.
+As mentioned in Section 3.1.2, the complexity measure in terms of a Language of Thought is just a convenient operationalization of this particular hypothesis for the sake of a concrete working example.
 
 We should stress that, just as models of iterated learning make different predictions under the assumption of different inductive biases, so does our model.
-The assumption and operationalization of a particular bias should, of course, always be seen critically (Section 4).
-
-%
+The assumption and operationalization of a particular bias should, of course, always be seen critically.
+We criticize our own setup and assumptions in Section 4.
 
@@
 %
-We understand transmission fidelity as the probability of acquiring type $i$ when learning from type $i$ (Section~2.3.3).
+We understand transmission fidelity as the probability of acquiring type $i$ when learning from type $i$ (Section~2.4.3).
 This is the value of a cell in the learning matrix $Q$.
@@
 $Q$ is a stochastic matrix and we we usually see neither extreme but rather, as suggested by the reviewer, a graded notion of transmission fidelity (Section~2.2).
-The details on how $Q$ is computed are given in Section 2.3.3.
+The details on how $Q$ is computed are given in Section 2.4.3.
 We have expanded this section to make these details clearer in the prose.
@@
 
-We have thoroughly revised the manuscript to make the contrast between these three predictions clearer (see Section~3.2, particularly 3.2.3; but also Section~2.4, which illustrates how the dynamics behave in a smaller type space).
+We have thoroughly revised the manuscript to make the contrast between these three predictions clearer (see Section~3.3, particularly 3.3.3; but also Section~3.1, which illustrates how the dynamics behave in a smaller type space; as well as the discussion of iterated learning alone and its dependence on $\lambda$ and $l$ in Sections 2.4.3 and appendix C).
 These questions tie in with Reviewer $1$'s request for a clearer exposition and the explanation of why we speak of co-evolution in this case (reviewer 1, comment 1).
@@
 Now, in the case of our type space we have a slight (modulo $\lambda$) communicative disadvantage for targets relative to competitors, for instance.
-However, the fact that targets are inferred more more often by na\"ive learners, also from those that learn from competitors, leads (i) to the gradual existence of less competitors, counteracting this functional disadvantage, as well as (ii) to the existence of only one variant of a kind (contra what we get in the {\em learning alone} condition).
+However, the fact that targets are inferred more often by na\"ive learners (by asymmetries in production likelihoods, see appendix C; a tendency which is also mediated by $\lambda$ and $l$), leads (i) to the gradual existence of less competitors, counteracting this functional disadvantage, as well as (ii) to the existence of only one variant of a kind (contra what we get in the {\em learning alone} condition).
 %
@@
 As noted above, we have expanded our analysis of the model and put much more emphasis on how the components that feed it play a role in driving particular evolutionary outcomes.
-See, in particular, Section 2.4, Section 3.2, and Section 4.
+Section 3.1 can be seen a direct response to this comment, where we try to deduce from what is known from the literature as much insight as possible about the workings of the dynamic in this case.
+Unfortunately, owing to the complexity of the case study we are not able to offer a mathematical characterization of, say, the stable rest points of the dynamic.
+(Notice that for this purpose we should switch to the continuous time formulation, which would also extend the paper further.)
 %
@@
 
-It is not necessarily true that it will always be advantageous for agents to behave pragmatically (see the end of Section~2.3.1 but also the discussion surrounding Figure~3a).
+It is not necessarily true that it will always be advantageous for agents to behave pragmatically (see the end of Section~2.4.1 but also the discussion surrounding Figure~3a).
 The main thing to note is that a pragmatic level-$1$ hearer using lexicon $L_{\text{bound}}$ will reason about the behavior of a {\em soft-maximizing} level-$0$ speaker of $L_{\text{bound}}$.
 The lower $\lambda$, the more the stochasticity percolates from the level-$0$ speaker to level-$1$ pragmatic interpretation.
-By contrast, literal level-$0$ hearers of $L_{\text{bound}}$ have a one-to-one form-meaning mapping from the get go.
+By contrast, literal level-$0$ hearers of $L_{\text{bound}}$ have a one-to-one form-meaning mapping from the start.
 Put differently, pragmatic reasoning can actually encumber some hearers, depending on their lexicon.
 Particularly if $\lambda$ is low.
+
+More to the point of the question, even if higher levels of pragmatic reasoning were always at least as good as the next lower level, the current population-level dynamic rests on \emph{minimal} assumptions about agent-level rationality.
+We have expanded on this point, which was definitely unclear in the first version, in section 2.3.
+In particular footnote 1 is almost a direct reply to this excellent comment.
+So, our agents are deliberately \emph{not} assumed to know the type of their interlocutor, but merely iterate pragmatic reasoning based on their own behavior.
+They would also not play a best response to the population average.
+The dynamic assumes that agents do not innovate strategies (like higher-order pragmatic reasoning).
+The dynamic is built on the idea that fitness-based selection is a non-innovative, gradient and proportional amplifier of whatever works well.
+Agents might never find out what works well, e.g., when fitness is just a measure of the proportion of offspring (a biological interpretation).
+Or they find out implicitly based on what works for them when they explore different ways of behavior (reinforcement learning).
+Or they occasionally realize that other agents are communicatively successful to a certain degree (e.g., conditional imitation based on success).
+Yet other interpretations are possible as well.
 
@@
 
-As we now clarify in Section~2.3.1 and stress a couple of times throughout the article, we assume very little sophistication from our agents.
+Indeed, this was regrettably not clear enough in the previous version.
+As we now emphasize in Sections~2.3 and 2.4.1, we assume very little sophistication from our agents.
 In particular, they do not know the type of the player that the are playing against.
@@
 \end{mdframed}
-We now highlight that this is a design choice and that there are alternatives in Section~2.3.3.
+We now highlight that this is a design choice and that there are alternatives in Section~2.4.3, footnote 3.
 
@@
 
+More concretely, there is a technical reason why we would like all speaker types to produce any message with positive probability: we do not want to make types too easily identifiable by even small sets of observed data, so that iterated learning has some work to do.
+Moreover, we belief that while a logarithmic definition of speaker utilities is just fine, when it comes to evolutionary fitness, a definition in terms of beliefs is not.
+Fitness should track genuine observable interaction with the world.
+Just holding a ``good belief'' gives no countable or observable advantage unless it is acted upon in some manner.
+
 %
@@
 
-This is an important point.
-We agree that some fragments of our previous manuscript could lead to confusion, e.g., as to why $L_{\text{all}}$, being the {\em a priori} more likely type to be inferred, did not come out strongly even in the {\em learning only} condition.
-Throughout the article, we now stress more that what matters is the posterior, and put emphasis on the role of the likelihood.
+The agents in this model do not actually learn the ``grammar'' of concepts which is used only as a tool for fixing one concrete measure of complexity of lexical concepts.
+It is these lexical concepts that are learned.
+We have tried to make this clearer by more careful reformulation.
 
@@
 
-As now made explicit in Section~3.1, we chose three messages for illustrative purposes mainly.
+As now made explicit in Section~3.2, we chose three messages for illustrative purposes mainly.
 In particular, we wanted to inspect a type space where targets are not the most likely {\em a priori} as this may mislead readers into thinking that learnability alone, in particular the prior alone, drives the outcome.
@@
 At present, we have no answer.
+
+The new section 3.1 now also looks at the case with just two messages that the reviewer brings into play.
+It is useful to contrast exactly why the effects of iterated learning hinge on the set of all types (the point made in connection with appendix C).
 

### 2018-03-19 mf_stuff/mutation_appendix_asymmetric_LH.r
R script for appendix on asymmetric LH

@@ -1,0 +1,24 @@
+library(tidyverse)
+
+# 3 types, 3 data
+# types t2 and t3 are more likely to produce data that is evidence for t1
+e = 0.1
+d = 0.2
+lh = matrix(c(1 - 2*e, e, e, d, 1-(d+e), e, d, e, 1-(d+e)), nrow = 3, byrow = T) %>% prop.table(1)
+
+# posterior over types for each datum:
+l = 10
+
+post = prop.table(t(lh),1)
+post = prop.table(post^l,1)
+
+q = matrix(0, nrow = 3, ncol = 3)
+
+for (i in 1:3) {
+  for (j in 1:3) {
+    q[i,j] = sum(post[,j] * lh[i,])
+  }
+}
+q %>% round(.,2) %>%  show()
+eigen(t(q)) %>% show()
+eigen(t(q))$vectors[,1] / sum(eigen(t(q))$vectors[,1])

### 2018-03-28 paper/rv-letter1.tex
cover lett answer

@@
 \begin{itemize}
-\item We have tried to improve the exposition of the model, both on the technical side and on
+\item We improved the exposition of the model, both on the technical side and on
   the way the model can be interpreted (re reviewer 1's concerns about how to interpret
@@
   pragmatics'' (in response to reviewer 1's comment).
-\item To compensate for the added material, we also tried to shorten the paper. We tried to
+\item To compensate for the added material, we also shortened the paper. We tried to
   avoid repetitive formulations, delegated more detailed explanations to footnotes and omitted
@@
 theoretically and experimentally. This makes them suitable candidates for the study of the
-evolution of regular pragmatic inferences; past research can guide our modeling choices and aid
+evolution of regular pragmatic inferences because past research can guide our modeling choices and aid
 in evaluating our application to the study of these inferences. This ties in with our
@@
 \stepcounter{reviewerCommentCounter} %increase counter
-\noindent Reviewer \thereviewerCounter~asks for a clearer description of the model. As noted
-earlier, we have taken care to make the exposition more accessible and to guide the reader more
+\noindent Reviewer \thereviewerCounter~asks for a clearer description of the model. As mentioned above, we have taken care to make the exposition more accessible and to guide the reader more
 through the technical details of Section 2. In particular, the newly added sections 2.3, 3.1,
@@
 Together, both ingredients define an agent's linguistic behavior.
-A learner faced with input $d$ (see above) calculates $P(\tau \mid d)$, the probability that a type $\tau$ produced input $d$.
+A learner faced with input $d$ (see above) calculates $P(\tau \mid d)$, the probability that type $\tau$ produced input $d$.
 Crucially, neither linguistic strategies nor lexical meanings are directly observable, so they can only be faithfully recovered inasmuch as the {\em overt} behavior evidenced by a type -- the data it produces -- is (in tendency) attributable to only one, {\em covert}, pairing of a lexicon and a linguistic strategy.
@@
 What fitness-relative selection adds to this process is the competition that learnability alone lacks, independent of the type space we look at.
-Now, in the case of our type space we have a slight (modulo $\lambda$) communicative disadvantage for targets relative to competitors, for instance.
+In the case of our type space we have a slight (modulo $\lambda$) communicative disadvantage for targets relative to competitors, for instance.
 However, the fact that targets are inferred more often by na\"ive learners (by asymmetries in production likelihoods, see appendix C; a tendency which is also mediated by $\lambda$ and $l$), leads (i) to the gradual existence of less competitors, counteracting this functional disadvantage, as well as (ii) to the existence of only one variant of a kind (contra what we get in the {\em learning alone} condition).
@@
 More concretely, there is a technical reason why we would like all speaker types to produce any message with positive probability: we do not want to make types too easily identifiable by even small sets of observed data, so that iterated learning has some work to do.
-Moreover, we belief that while a logarithmic definition of speaker utilities is just fine, when it comes to evolutionary fitness, a definition in terms of beliefs is not.
+Moreover, we believe that while a logarithmic definition of speaker utilities is just fine, when it comes to evolutionary fitness, a definition in terms of beliefs is not.
 Fitness should track genuine observable interaction with the world.

### 2018-03-29 paper/deprecated_data/generating_code/1rmd-singlescalar.py
plots with rerun experimentation

@@ -1,163 +1,0 @@
-#####
-#RMD with parametrized iterated parental learning 
-#1 pairs of scalar items, six lexica per scalar pair. 
-#2 possible signaling behaviors: literal or gricean
-#12 types (6 lexica * 2 signaling behaviors) per signaling pair
-#6 literal types, 6 gricean types, 12 types in total
-#####
-
-import numpy as np
-np.set_printoptions(threshold=np.nan)
-from random import sample
-from itertools import product
-from player import LiteralPlayer,GriceanPlayer
-import sys 
-import datetime
-import csv
-
-#####
-l1,l2,l3,l4,l5,l6 = np.array( [[0.,0.],[1.,1.]] ), np.array( [[1.,1.],[0.,0.]] ), np.array( [[1.,1.],[1.,1.]] ), np.array( [[0.,1.],[1.,0.]] ), np.array( [[0.,1.],[1.,1.]] ), np.array( [[1.,1.],[1.,0.]] )
-
-alpha = 1 # rate to control difference between semantic and pragmatic violations
-cost = 0.1 # cost for LOT-concept with upper bound
-lam = 30 # soft-max parameter
-k = 3  # length of observation sequences
-sample_amount = 15 #amount of k-length samples for each production type
-
-gens = 20 #number of generations per simulation run
-runs = 50 #number of independent simulation runs
-
-states = 2 #number of states
-messages = 2 #number of messages
-
-learning_parameter = 1 #prob-matching = 1, increments approach MAP
-state_freq = np.ones(states) / float(states) #frequency of states s_1,...,s_n 
-
-
-f = csv.writer(open('./results/singlescalar-a%d-c%.2f-l%d-k%d-samples%d-learn%.2f-g%d-r%d.csv' %(alpha,cost,lam,k,sample_amount, learning_parameter, gens,runs),'wb')) #file to store mean results
-
[…]

### 2018-03-29 paper/deprecated_data/generating_code/3rmd-singlescalar_withNoise.py

@@ -1,187 +1,0 @@
-#####
-#RMD with parametrized iterated parental learning 
-#1 pairs of scalar items, six lexica per scalar pair. 
-#2 possible signaling behaviors: literal or gricean
-#12 types (6 lexica * 2 signaling behaviors) per signaling pair
-#6 literal types, 6 gricean types, 12 types in total
-#####
-
-import numpy as np
-np.set_printoptions(threshold=np.nan)
-from random import sample
-from itertools import product
-from player import LiteralPlayer,GriceanPlayer
-import sys 
-import datetime
-import csv
-
-#####
-l1,l2,l3,l4,l5,l6 = np.array( [[0.,0.],[1.,1.]] ), np.array( [[1.,1.],[0.,0.]] ), np.array( [[1.,1.],[1.,1.]] ), np.array( [[0.,1.],[1.,0.]] ), np.array( [[0.,1.],[1.,1.]] ), np.array( [[1.,1.],[1.,0.]] )
-
-alpha = 20 # rate to control difference between semantic and pragmatic violations
-cost = 0 # cost for LOT-concept with upper bound
-lam = 10 # soft-max parameter
-k = 5  # length of observation sequences
-sample_amount = 10 #amount of k-length samples for each production type
-epsilon  = .5 # probability of perceiving S-all, when true state is S-sbna
-delta = .1 # probability of perceiving S-sbna, when true state is S-all
-learning_parameter = 10 #prob-matching = 1, increments approach MAP
-
-gens = 1000 #number of generations per simulation run
-runs = 50 #number of independent simulation runs
-
-states = 2 #number of states
-messages = 2 #number of messages
-state_freq = np.ones(states) / float(states) #frequency of states s_1,...,s_n 
-
-
-print '#Starting, ', datetime.datetime.now()
-
-t1,t2,t3,t4,t5,t6 = LiteralPlayer(alpha,lam,l1), LiteralPlayer(alpha,lam,l2), LiteralPlayer(alpha,lam,l3), LiteralPlayer(alpha,lam,l4), LiteralPlayer(alpha,lam,l5), LiteralPlayer(alpha,lam,l6)
[…]

### 2018-03-29 paper/deprecated_data/generating_code/all_sample_1rmd-singlescalar.py

@@ -1,186 +1,0 @@
-#####
-#RMD with parametrized iterated parental learning 
-#1 pairs of scalar items, six lexica per scalar pair. 
-#2 possible signaling behaviors: literal or gricean
-#12 types (6 lexica * 2 signaling behaviors) per signaling pair
-#6 literal types, 6 gricean types, 12 types in total
-#####
-
-import numpy as np
-np.set_printoptions(threshold=np.nan)
-from random import sample
-from itertools import product, combinations_with_replacement
-from player import LiteralPlayer,GriceanPlayer
-import sys 
-import datetime
-import csv
-
-#####
-l1,l2,l3,l4,l5,l6 = np.array( [[0.,0.],[1.,1.]] ), np.array( [[1.,1.],[0.,0.]] ), np.array( [[1.,1.],[1.,1.]] ), np.array( [[0.,1.],[1.,0.]] ), np.array( [[0.,1.],[1.,1.]] ), np.array( [[1.,1.],[1.,0.]] )
-
-alpha = 1 # rate to control difference between semantic and pragmatic violations
-cost = 0.1 # cost for LOT-concept with upper bound
-lam = 30 # soft-max parameter
-k = 3  # length of observation sequences
-sample_amount = 15 #amount of k-length samples for each production type
-
-gens = 20 #number of generations per simulation run
-runs = 50 #number of independent simulation runs
-
-states = 2 #number of states
-messages = 2 #number of messages
-
-learning_parameter = 1 #prob-matching = 1, increments approach MAP
-state_freq = np.ones(states) / float(states) #frequency of states s_1,...,s_n 
-
-
-f = csv.writer(open('./results/singlescalar-a%d-c%.2f-l%d-k%d-samples%d-learn%.2f-g%d-r%d.csv' %(alpha,cost,lam,k,sample_amount, learning_parameter, gens,runs),'wb')) #file to store mean results
-
[…]

### 2018-03-29 paper/deprecated_data/generating_code/modsinglescalar.py

@@ -1,156 +1,0 @@
-#####
-#RMD with parametrized iterated parental learning 
-#1 pairs of scalar items, six lexica per scalar pair. 
-#2 possible signaling behaviors: literal or gricean
-#12 types (6 lexica * 2 signaling behaviors) per signaling pair
-#6 literal types, 6 gricean types, 12 types in total
-#####
-
-import numpy as np
-np.set_printoptions(threshold=np.nan)
-from random import sample
-from itertools import product
-from player import LiteralPlayer,GriceanPlayer
-import sys 
-import datetime
-import csv
-
-def normalize(m):
-    return m / m.sum(axis=1)[:, np.newaxis]
-
-def summarize_counts(lst,states,messages):
-    """summarize counts for tuples of k-states and k-messages""" 
-    counter = [0 for _ in xrange(states**messages)]
-    for i in xrange(len(lst)):
-        s,m = lst[i][0] *2, lst[i][1]
-        counter[s+m] += 1
-    return counter
-
-def get_obs(k,states,messages,likelihoods,state_freq,sample_amount):
-    """Returns summarized counts of k-length <s_i,m_j> production observations as [#(<s_0,m_0>), #(<s_0,m_1), #(<s_1,m_0>, #(s_1,m_1)], ...]] = k"""
-    s = list(xrange(states))
-    m = list(xrange(messages))
-    atomic_observations = list(product(s,m))
-   
-    obs = [] #store all produced k-length (s,m) sequences 
-    for t in xrange(12):
-        produced_obs = [] #store k-length (s,m) sequences of a type
-        production_vector = likelihoods[t].flatten()
-        doubled_state_freq = np.column_stack((state_freq,state_freq)).flatten() # P(s)
-        sample_vector = production_vector * doubled_state_freq #P(s) * P(m|s,t_i)
-        for i in xrange(sample_amount):
-            sample_t = [np.random.choice(len(atomic_observations),p=sample_vector) for _ in xrange(k)]
-            sampled_obs = [atomic_observations[i] for i in sample_t]
-            produced_obs.append(summarize_counts(sampled_obs,states,messages))
-        obs.append(produced_obs)
-    return obs
-
-
[…]

### 2018-03-29 paper/deprecated_data/generating_code/modsinglescalaronlyfitness.py

@@ -1,136 +1,0 @@
-#####
-#RMD with parametrized iterated parental learning 
-#1 pairs of scalar items, six lexica per scalar pair. 
-#2 possible signaling behaviors: literal or gricean
-#12 types (6 lexica * 2 signaling behaviors) per signaling pair
-#6 literal types, 6 gricean types, 12 types in total
-#####
-
-import numpy as np
-np.set_printoptions(threshold=np.nan)
-from random import sample
-from itertools import product
-from player import LiteralPlayer,GriceanPlayer
-import sys 
-import datetime
-import csv
-
-def normalize(m):
-    return m / m.sum(axis=1)[:, np.newaxis]
-
-def summarize_counts(lst,states,messages):
-    """summarize counts for tuples of k-states and k-messages""" 
-    counter = [0 for _ in xrange(states**messages)]
-    for i in xrange(len(lst)):
-        s,m = lst[i][0] *2, lst[i][1]
-        counter[s+m] += 1
-    return counter
-
-def get_obs(k,states,messages,likelihoods,state_freq,sample_amount):
-    """Returns summarized counts of k-length <s_i,m_j> production observations as [#(<s_0,m_0>), #(<s_0,m_1), #(<s_1,m_0>, #(s_1,m_1)], ...]] = k"""
-    s = list(xrange(states))
-    m = list(xrange(messages))
-    atomic_observations = list(product(s,m))
-   
-    obs = [] #store all produced k-length (s,m) sequences 
-    for t in xrange(12):
-        produced_obs = [] #store k-length (s,m) sequences of a type
-        production_vector = likelihoods[t].flatten()
-        doubled_state_freq = np.column_stack((state_freq,state_freq)).flatten() # P(s)
-        sample_vector = production_vector * doubled_state_freq #P(s) * P(m|s,t_i)
-        for i in xrange(sample_amount):
-            sample_t = [np.random.choice(len(atomic_observations),p=sample_vector) for _ in xrange(k)]
-            sampled_obs = [atomic_observations[i] for i in sample_t]
-            produced_obs.append(summarize_counts(sampled_obs,states,messages))
-        obs.append(produced_obs)
-    return obs
-
-
-def get_likelihood(obs,likelihoods):
[…]

### 2018-03-29 paper/deprecated_data/generating_code/modsinglescalaronlylearning.py

@@ -1,132 +1,0 @@
-#####
-#RMD with parametrized iterated parental learning 
-#1 pairs of scalar items, six lexica per scalar pair. 
-#2 possible signaling behaviors: literal or gricean
-#12 types (6 lexica * 2 signaling behaviors) per signaling pair
-#6 literal types, 6 gricean types, 12 types in total
-#####
-
-import numpy as np
-np.set_printoptions(threshold=np.nan)
-from random import sample
-from itertools import product
-from player import LiteralPlayer,GriceanPlayer
-import sys 
-import datetime
-import csv
-
-def normalize(m):
-    return m / m.sum(axis=1)[:, np.newaxis]
-
-def summarize_counts(lst,states,messages):
-    """summarize counts for tuples of k-states and k-messages""" 
-    counter = [0 for _ in xrange(states**messages)]
-    for i in xrange(len(lst)):
-        s,m = lst[i][0] *2, lst[i][1]
-        counter[s+m] += 1
-    return counter
-
-def get_obs(k,states,messages,likelihoods,state_freq,sample_amount):
-    """Returns summarized counts of k-length <s_i,m_j> production observations as [#(<s_0,m_0>), #(<s_0,m_1), #(<s_1,m_0>, #(s_1,m_1)], ...]] = k"""
-    s = list(xrange(states))
-    m = list(xrange(messages))
-    atomic_observations = list(product(s,m))
-   
-    obs = [] #store all produced k-length (s,m) sequences 
-    for t in xrange(12):
-        produced_obs = [] #store k-length (s,m) sequences of a type
-        production_vector = likelihoods[t].flatten()
-        doubled_state_freq = np.column_stack((state_freq,state_freq)).flatten() # P(s)
-        sample_vector = production_vector * doubled_state_freq #P(s) * P(m|s,t_i)
-        for i in xrange(sample_amount):
-            sample_t = [np.random.choice(len(atomic_observations),p=sample_vector) for _ in xrange(k)]
-            sampled_obs = [atomic_observations[i] for i in sample_t]
-            produced_obs.append(summarize_counts(sampled_obs,states,messages))
-        obs.append(produced_obs)
-    return obs
-
-
-def get_likelihood(obs,likelihoods):
[…]

### 2018-03-29 paper/deprecated_data/generating_code/new_sims.py

@@ -1,69 +1,0 @@
-import modsinglescalar as rmd
-import modsinglescalaronlyfitness as fit
-import modsinglescalaronlylearning as learner
-#parameter order for run(*) is run(alpha,cost,lam,k,sample_amount, learning_parameter,gens,runs)
-
-
-def cost_to_learning(): #results 1
-    a = 1
-    lam = 30
-    cost = [x/100. for x in xrange(100)]
-    seq_length = 5
-    sample_amount = 10
-    learning_parameter = [x for x in xrange(1,11)]
-    gens = 20
-    runs = 10000
-    
-    for c in cost:
-        for learn in learning_parameter:
-            rmd.run(a,c,lam,seq_length,learn,gens,runs)
-
-cost_to_learning()
-
-def ration_to_seq():
-    a = 1
-    lam = [x for x in xrange(1,50)]
-    c = .4
-    seq_length = [x for x in xrange(1,20)]
-    sample_amount = 10
-    learn = 5
-    gens = 20
-    runs = 10000
-    
-    for l in lam:
-        for k in seq_length:
-            rmd.run(a,c,l,k,sample_amount,learn,gens,runs)
-
-def alpha_to_seq():
-    a = [x/100. for x in xrange(1,101)]
-    lam = 20
-    c = .4
-    seq_length = [x for x in xrange(1,20)]
-    sample_amount = 20
-    learn = 3
-    gens = 10
-    runs = 10000
-    
-    for al in a:
-        for k in seq_length:
-            rmd.run(al,c,l,k,sample_amount,learn,gens,runs)
-
-def only_r_and_m():
-    a = 1
-    lam = 30
-    cost = [x/100. for x in xrange(100)]
-    learn = 1
-    seq_length = 5
-    sample_amount = 10
-    gens = 20
-    runs = 10000
-    
-    for c in cost:
-        print 'learn-only'
-        learner.run(a,c,lam,seq_length,sample_amount,learn,gens,runs)
-        print 'fit-only'
-        fit.run(a,c,lam,seq_length,sample_amount,learn,gens,runs)
-
-
-
-#only_r_and_m()

### 2018-03-29 paper/deprecated_data/generating_code/player.py

@@ -1,51 +1,0 @@
-import numpy as np
-
-def normalize(m):
-    m = m / m.sum(axis=1)[:, np.newaxis]
-    m[np.isnan(m)] = 0.
-    return m
-
-class LiteralPlayer:
-    def __init__(self,alpha,lam,lexicon):
-        self.alpha = alpha
-        self.lam = lam
-        self.lexicon = lexicon
-        self.sender_matrix = self.sender_selection_matrix(lexicon)
-        self.receiver_matrix =  self.receiver_selection_matrix()
-   
-    def sender_selection_matrix(self,l):
-        m = np.zeros(np.shape(l))
-        for i in range(np.shape(l)[0]):
-            for j in range(np.shape(l)[1]):
-                m[i,j] = np.exp(self.lam * l[i,j])
-        return normalize(m)
-
-    def receiver_selection_matrix(self):
-        """Take transposed lexicon and normalize row-wise (prior over states plays no role as it's currently uniform)"""
-        m = normalize(np.transpose(self.lexicon))
-        for r in range(np.shape(m)[0]):
-            if sum(m[r]) == 0:
-                for c in range(np.shape(m)[1]):
-                    m[r,c] = 1. / np.shape(m)[0]
-
-        return m
-
-
-
-#######################################
-class GriceanPlayer:
-    def __init__(self,alpha, lam, lexicon):
-        self.alpha = alpha
-        self.lam = lam
-        self.lexicon = lexicon
-        self.sender_matrix = self.sender_selection_matrix(lexicon)
-        self.receiver_matrix = self.receiver_selection_matrix()
-
-    def sender_selection_matrix(self,l):
-        literalListener = normalize(np.transpose(l))
-        utils = np.transpose(literalListener)
-        return normalize(np.exp(self.lam*np.power(utils, self.alpha)))
-
-    def receiver_selection_matrix(self):
-        """Take transposed sender matrix and normalize row-wise (prior over states plays no role as it's currently uniform)"""
-        return normalize(np.transpose(self.sender_matrix))

### 2018-03-29 paper/deprecated_data/generating_code/plot-fitness.py

@@ -1,274 +1,0 @@
-###
-import pandas as pd
-import numpy as np 
-import matplotlib.pyplot as plt
-import seaborn as sns
-#sns.set_context(rc={'lines.markeredgewidth': 0.5})
-import os
-import glob
-import sys
-###
-###Column headers:###
-#run_ID,t1_initial,t2_initial,t3_initial,t4_initial,t5_initial,t6_initial,t7_initial,t8_initial,t9_initial,t10_initial,t11_initial,t12_initial,
-#alpha,prior_cost_c,lambda,k,sample_amount,learning_parameter,generations,
-#t1_final,t2_final,t3_final,t4_final,t5_final,t6_final,t7_final,t8_final,t9_final,t10_final,t11_final,t12_final
-#####################
-
-
-def fitness_only_plot():
-    print 'Loading data fitness'
-    df = pd.read_csv('./mean-1000games-replication-only.csv')
-
-    
-    a = 1
-    lam = 30
-    
-    final_groupA = df.loc[:,'alpha':]
-    
-    groupA = final_groupA.loc[final_groupA['alpha'] == a]
-    
-    t_finalA = groupA.groupby(['lambda'])
-    t_finalA = t_finalA[['t1_final','t2_final','t3_final','t4_final','t5_final','t6_final','t7_final','t8_final','t9_final','t10_final','t11_final','t12_final']].agg(np.average) 
-    
-    d1 = t_finalA.loc[:,('t1_final','t2_final','t3_final', 't4_final', 't5_final', 't6_final')]
-    d2 = t_finalA.loc[:,('t9_final','t10_final', 't11_final')]
-    
-    ax2 = plt.plot(d2)
-    #ax1 = plt.plot(d1,marker='d',markevery=5)
-    plt.title(r'($\alpha$ = %d)' %(a))
-    plt.ylabel("Proportion in population")
-    plt.xlabel(r'Rationality parameter $\lambda$')
-    
-    plt.legend(["pragmatic L-taut", "pragmatic L-bound","pragmatic L-lack"], loc='best')
-    plt.xlim(0,48)
-    plt.show() 
-
-
-def learning_only_plot():
-    print 'Loading data learn'
-    dl = pd.read_csv('./mean-1000games-mutation-only.csv')
-    
-    a = 1
-    lam = 30
-    k = 5
-    sample = 10
-    learn = 1
-    
-    final_groupB = dl.loc[:,'alpha':]
[…]

### 2018-03-29 paper/deprecated_data/plots.py

@@ -1,275 +1,0 @@
-###
-import pandas as pd
-import numpy as np 
-import matplotlib.pyplot as plt
-import seaborn as sns
-#sns.set_context(rc={'lines.markeredgewidth': 0.5})
-import os
-import glob
-import sys
-###
-###Column headers:###
-#run_ID,t1_initial,t2_initial,t3_initial,t4_initial,t5_initial,t6_initial,t7_initial,t8_initial,t9_initial,t10_initial,t11_initial,t12_initial,
-#alpha,prior_cost_c,lambda,k,sample_amount,learning_parameter,generations,
-#t1_final,t2_final,t3_final,t4_final,t5_final,t6_final,t7_final,t8_final,t9_final,t10_final,t11_final,t12_final
-#####################
-
-
-def fitness_only_plot():
-    print 'Loading data fitness'
-    df = pd.read_csv('./mean-1000games-replication-only.csv')
-
-    
-    a = 1
-    lam = 30
-    
-    final_groupA = df.loc[:,'alpha':]
-    
-    groupA = final_groupA.loc[final_groupA['alpha'] == a]
-    
-    t_finalA = groupA.groupby(['lambda'])
-    t_finalA = t_finalA[['t1_final','t2_final','t3_final','t4_final','t5_final','t6_final','t7_final','t8_final','t9_final','t10_final','t11_final','t12_final']].agg(np.average) 
-    
-    d1 = t_finalA.loc[:,('t1_final','t2_final','t3_final', 't4_final', 't5_final', 't6_final')]
-    d2 = t_finalA.loc[:,('t9_final','t10_final', 't11_final')]
-    
-    ax2 = plt.plot(d2)
-    #ax1 = plt.plot(d1,marker='d',markevery=5)
-    plt.title(r'($\alpha$ = %d)' %(a))
-    plt.ylabel("Proportion in population")
-    plt.xlabel(r'Rationality parameter $\lambda$')
-    
-    plt.legend(["prag. L-taut", "prag. L-bound","prag. L-lack"], loc='best')
-    plt.xlim(0,48)
-    plt.show() 
-
-
-def learning_only_plot():
-    print 'Loading data learn'
-    dl = pd.read_csv('./mean-1000games-mutation-only.csv')
-    
-    a = 1
-    lam = 30
-    k = 5
-    sample = 10
-    learn = 1
-    
-    final_groupB = dl.loc[:,'alpha':]
[…]

### 2018-03-30 paper/cogsci-submission2.tex
final manuscript+letter changes TB

@@
 
-In contrast, we will here represent explicitly language users' representations of lexical
-meanings and their own particular manner of manner of production and interpretation based on
-these lexical representation. Different types of pragmatic behavior are represented using
+In contrast, we will here explicitly model language users' representations of lexical
+meanings and their own particular manner of production and interpretation based on
+these lexical representations. Different types of pragmatic behavior -- ways of deploying meaning in interaction -- are represented using
 probabilistic models of pragmatic language use
@@
   piantadosi:2014,kirby+etal:2015}. If processes of selection, such as conditional imitation or
-reinforcement, favors communicative efficiency, languages are driven towards semantic
+reinforcement, favor communicative efficiency, languages are driven towards semantic
 expressivity \citep[e.g.,][]{nowak+krakauer:1999,Skyrms2010:Signals}. But pressure towards
@@
 that lexicalizes distinct forms for a large number of different meanings may be better for communication 
-but challenging to acquire.
+but more challenging to acquire.
 
@@
   was individually capable of seeing $X$ as the rational thing to do, $X$ would still evolve if
-  every agent at least sometimes, locally and myopically tried to optimize its behavior, e.g.,
+  every agent at least sometimes, locally and myopically, tried to optimize its behavior, e.g.,
   by imitating what seems to be working well for others \citep{Skyrms2010:Signals}.}
@@
 \label{sec:languages+use}
-Types are what evolution promotes or demotes. A type determines an agent's behavior and thereby
+Types are what evolution promotes or demotes. Its type determines an agent's behavior and thereby
 an agent's fitness, so that it makes sense to speak of the fitness of a type itself. To study
@@
 According to (\ref{h:level0}), a literal hearer's interpretation of a message depends on her
-lexicon and her prior over states, $pr \in \Delta(S)$, which is here assumed flat for
+lexicon and her prior over states, $pr \in \Delta(S)$, which is in the following assumed flat for
 simplicity. Literal interpreters thereby choose an arbitrary true interpretation for each
@@
 expressions. Higher order behavior is therefore, even if in tendency rational, quite simple. It
-does not assume agents to know their interlocutor's actual type when interacting (e.g., what
+does not assume agents to know their interlocutor's type when interacting (e.g., what
 lexicon the interlocutor is using); they simply act based on a conjecture about language use
@@
 With low $\lambda$ speakers choose true messages with more slack.
-Reasoning over this behavior therefore also results in a weaker association of messages with only true states in receivers, but also in a slightly stronger association of $\msome$ with $\ssome$ over $\sall$ for $\Llack$ users because they reason that $S_0(\msome | \ssome;\Llack) > S_0(\msome | \sall;\Llack)$.
+Reasoning over this behavior therefore also results in a weaker association of messages with only true states in receivers, but also in a slightly stronger association of $\msome$ with $\ssome$ over $\sall$ for $\Llack$ users, because they reason that $S_0(\msome | \ssome;\Llack) > S_0(\msome | \sall;\Llack)$.
 For $\lambda = 20$, there will be less slack in literal speaker behavior:
@@
 signaling is preferable. In sum, the definition of types introduced here constitute conditions
-that are purposefully averse to what we would like to show; the evolution of is not immediate
+that are purposefully averse to what we would like to show; the evolution of a division of labor between semantics and pragmatics is not immediate
 just based on fitness-based selection, to which we turn next.
@@
 frequency $x_i$ before the update step, its fitness $f_i$, and the probability that a learner
-infers $i$ when observing the behavior of a type-$j$ agent. Fitness-based can be thought of as
+infers $i$ when observing the behavior of a type-$j$ agent. Fitness-based replication can be thought of as
 biological (fitness as expected relative number of offspring) or cultural evolution (fitness as
 likelihood of being imitated or repeated, see Appendix~\ref{sec:deriv-repl-mutat} for a
-concrete example). A type's communicative success depends on how well it communicates within a
+concrete example). A type's communicative success depends on how well it communicates within its 
 population while its learnability depends on the fidelity by which it is inferred by new
@@
 \centering
-    % \begin{tikzpicture}
-
-  %   \draw[-,dashed,gray] (0,0) -- (5,0) -- (5,5) -- (0,5) -- (0,0);
-
-  %   \draw (0,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Lbound$}}] {};
-  %   \draw (5,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Lbound$}}] {};
-  %   \draw (5,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Llack$}}] {};
-  %   \draw (0,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Llack$}}] {};
-
-  %   \draw[<-, thick, indianred] (0.75,0.25) -- (4.25,0.25) node[midway, above] {\footnotesize functional};
-  %   \draw[->, thick, indianred] (0.75,4.75) -- (4.25,4.75) node[midway, below] {\footnotesize functional};
-  %   \draw[->, thick, indianred] (0.25,4.25) -- (0.25,0.75) node[midway, below, rotate=90] {\footnotesize functional};
-  %   \draw[->, thick, indianred] (4.75,4.25) -- (4.75,0.75) node[midway, above, rotate=90] {\footnotesize functional};
-
-  % \path[->, thick, densely dashdotted, Steelblue] (0.75,0.75) edge [bend left] node[sloped,below] {\footnotesize learning\textsubscript{flat}} (2.25,2.45);
-  % \path[->, thick, densely dashdotted, Steelblue] (4.25,0.75) edge [bend right] node[sloped,below] {\footnotesize learning\textsubscript{flat}} (2.75,2.45);
-  % \path[->, thick, densely dashdotted, Steelblue] (0.75,4.25) edge [bend right] node[sloped,above] {\footnotesize learning\textsubscript{flat}} (2.25,2.75);
-  % \path[->, thick, densely dashdotted, Steelblue] (4.25,4.25) edge [bend left] node[sloped,above] {\footnotesize learning\textsubscript{flat}} (2.75,2.75);
-
-  %   \draw[->, thick, goldish] (0.75,-0.25) -- (2.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
-  %   \draw[<-, thick, goldish] (2.75,-0.25) -- (4.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
-  %   \draw[->, thick, goldish] (-0.25,0.75) -- (-0.25,4.25) node[midway, above,rotate=90] {\footnotesize learning\textsubscript{biased}};
-  %   \draw[->, thick, goldish] (5.25,0.75) -- (5.25,4.25) node[midway, below,rotate=90] {\footnotesize learning\textsubscript{biased}};
-  %   \draw[->, thick, goldish] (0.75,5.25) -- (4.25,5.25) node[midway, above] {\footnotesize learning\textsubscript{biased}};
-
-  %   \end{tikzpicture}
+     \begin{tikzpicture}
+
+     \draw[-,dashed,gray] (0,0) -- (5,0) -- (5,5) -- (0,5) -- (0,0);
+
+     \draw (0,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Lbound$}}] {};
+     \draw (5,0) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Lbound$}}] {};
+     \draw (5,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]east:{\tiny 100\%} \\[-0.25cm] {\tiny prag. $\Llack$}}] {};
+     \draw (0,5) node[fill,circle,gray,inner sep=1mm,scale=0.6,label={[gray,align=left,label distance=0.1mm]west:{\tiny 100\%} \\[-0.25cm] {\tiny lit. $\Llack$}}] {};
+
+     \draw[<-, thick, indianred] (0.75,0.25) -- (4.25,0.25) node[midway, above] {\footnotesize functional};
+     \draw[->, thick, indianred] (0.75,4.75) -- (4.25,4.75) node[midway, below] {\footnotesize functional};
+     \draw[->, thick, indianred] (0.25,4.25) -- (0.25,0.75) node[midway, below, rotate=90] {\footnotesize functional};
+     \draw[->, thick, indianred] (4.75,4.25) -- (4.75,0.75) node[midway, above, rotate=90] {\footnotesize functional};
+
+   \path[->, thick, densely dashdotted, Steelblue] (0.75,0.75) edge [bend left] node[sloped,below] {\footnotesize learning\textsubscript{flat}} (2.25,2.45);
+   \path[->, thick, densely dashdotted, Steelblue] (4.25,0.75) edge [bend right] node[sloped,below] {\footnotesize learning\textsubscript{flat}} (2.75,2.45);
+   \path[->, thick, densely dashdotted, Steelblue] (0.75,4.25) edge [bend right] node[sloped,above] {\footnotesize learning\textsubscript{flat}} (2.25,2.75);
+   \path[->, thick, densely dashdotted, Steelblue] (4.25,4.25) edge [bend left] node[sloped,above] {\footnotesize learning\textsubscript{flat}} (2.75,2.75);
+
+     \draw[->, thick, goldish] (0.75,-0.25) -- (2.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
+     \draw[<-, thick, goldish] (2.75,-0.25) -- (4.25,-0.25) node[midway, below] {\footnotesize learning\textsubscript{biased}};
+     \draw[->, thick, goldish] (-0.25,0.75) -- (-0.25,4.25) node[midway, above,rotate=90] {\footnotesize learning\textsubscript{biased}};
+     \draw[->, thick, goldish] (5.25,0.75) -- (5.25,4.25) node[midway, below,rotate=90] {\footnotesize learning\textsubscript{biased}};
+     \draw[->, thick, goldish] (0.75,5.25) -- (2.25,5.25) node[midway, above] {\footnotesize learning\textsubscript{biased}};
+     \draw[<-, thick, goldish] (2.75,5.25) -- (4.25,5.25) node[midway, above] {\footnotesize learning\textsubscript{biased}};
+
+     \end{tikzpicture}
 \caption{Sketch of dynamics on a two-dimensional type space with four types. The $y$-axis
@@
   learnability pressure (mutator steps) with either a flat prior (blue dash-dotted arrows) or a
-  prior that favors $\Llack$ (solid purple arrows).\mf{I did not change the upper-most arrow
-    for biased learnability, because I cannot compile this anyway; I did change some typos in
-    arrow labelling though.}}
+  prior that favors $\Llack$ (solid purple arrows).}
 \label{fig:sketch}
@@
 trends in evolutionary trajectories in this space based on what we know from the literature and
-the setup from Section~\ref{sec:fitn--learn}, even before any further simulation results.
+the setup from Section~\ref{sec:fitn--learn}, even before any simulation results.
 
@@
 advantage over literal $\Llack$. The red arrows in Fig.~\ref{fig:sketch} sketch trajectories we
-can expect if there is only pressure for communicative success. How much each replicator step
+can accordingly expect if there is only pressure for communicative success. How much each replicator step
 moves a population through this space ultimately depends on the rationality parameter
@@
 
-Under probability matching $l=1$, iterated Bayesian learning converges to the prior over types
+Under probability matching, $l=1$, iterated Bayesian learning converges to the prior over types
 \citep{griffiths+kalish:2007}. If the prior on types is uniform, populations gravitate toward
@@
 more difficult to predict when $l>1$, because now individual mutator steps may also be
-influenced not only by the priors, but also by the differential likelihood with which a
+influenced not only by the prior, but also by the differential likelihood with which a
 particular type $i$ can be confused for any other type $j$ (see Section~\ref{sec:learnability}
@@
 Despite its simplicity, the diagram in Figure~\ref{fig:sketch} makes clear the sense in which
-semantics and pragmatics might co-evolve. Pragmatic language use will not evolve under a strong
+semantics and pragmatics might co-evolve. Pragmatic language use will not evolve under strong
 and fully expressive semantics (bottom part of Figure~\ref{fig:sketch}). A weak semantic
@@
   that favors types with $\Llack$ by a factor of $1.05$ over those with $\Lbound$. Colored
-  contours show the proportion of pragmatic $\Llack$ users after a step. \tb{I don't think the
-    contours are adding much, so they could be removed. Then again, I think they are rather
-    non-intrusive, look nice, and illustrate how the step-size, depending on where we are in
-    the space, varies.} \mf{I think that the contours are sweet. But I worried about the size
-  of the arrows. I made the picture a bit bigger}}
+  contours show the proportion of pragmatic $\Llack$ users after a step.}
 
@@
 steps shown in the top row of Fig.~\ref{fig:quiver} make the functional disadvantage of
-$\Llack$ against $\Lbound$ apparent. With increasing $\lambda$, this difference pragmatic users
-of $\Lbound$ are no longer selected against by fitness-based selection.
+$\Llack$ against $\Lbound$ apparent. With increasing $\lambda$, this difference is mitigated for pragmatic users of $\Lbound$, who are no longer selected against by fitness-based selection.
 
@@
 where the system gravitates indeed to a population where the relative frequency of $\Llack$
-over $\Lbound$ is exactly $1.05$). Second, if $l>0$, the dynamic is influenced heavily also by
+over $\Lbound$ is exactly $1.05$). Second, if $l>1$, the dynamic is influenced heavily also by
 asymmetries in production likelihoods. Third, these asymmetries in the production likelihoods
@@
 (when $l>1$; see Appendix~\ref{sec:example-iter-bayes}). For this reason, we turn to a somewhat
-more realistic but still manageable setup next with a larger set of types.
+more realistic but still manageable setup with a larger set of types next.
 
@@
 model's predictions in populations of types with one of the two behaviors introduced earlier:
-literal or pragmatic. The former correspond to level-$0$ reasoners and the latter to level $1$. Accordingly, there is a total of $432$ types. Six types are pragmatic language
+literal or pragmatic. The former correspond to level-$0$ reasoners and the latter to ones of level $1$. Accordingly, there is a total of $432$ types. Six types are pragmatic language
 users with $\Llack$-like lexica. We refer to these as \emph{target types} because they
@@
 map different states to different overt messages. Consequently, different types of the same
-kind can fail to understand each other completely.
+kind will fail to understand each other completely.
 
@@
 convenient operationalization of one general approach to explicating learning biases; this is
-not a implicit commitment to the claim that this particular instrumentalization is the single
-most plausible. We merely want to be concrete to have a maximally concrete working example in
+not an implicit commitment to the claim that this particular instrumentalization is the single
+most plausible. We merely want to have a maximally concrete working example in
 which the priors over types are systematically related to the complexity of lexical
@@
 \centering
-%\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/alt-fig1-onlyr}
-\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/barh-onlyr}
+\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig3}
 \caption{Proportion of target types, the six competitor types with the highest proportion, the
@@
 signaling behavior, which in turn determines communicative success. Fig. \ref{fig:only-R} shows
-the proportion of target types, the highest competitor types and the highest type with $\Lall$
-in a representative population after $50$ replicator steps. The plot also indicates the
+the proportion of target types, the highest competitor types and the highest type with an $\Lall$-style lexicon in a representative population after $50$ replicator steps. The plot also indicates the
 proportion of the \emph{majority type}: the type with the highest proportion in the
@@
 promotes less-stochastic linguistic behavior, widening the gap in expressivity between some
-types and promoting more homogeneous populations. As suggested by Fig. \ref{fig:only-R}, the
-majority in most populations is not one of the six pragmatic $\Llack$-style types. That is, a
+types and promoting more homogeneous populations.
+As suggested by Fig.
+\ref{fig:only-R}, under replication only the majority in most populations is not one of the six pragmatic $\Llack$-style types.
+That is, a
 pressure only for communicative success does not lead to a prevalence of target types under any
 $\lambda$-value. For instance, with $\lambda = 20$, $1000$ independent populations only had $11$
-cases in which the target type was the majority type, corresponding to a mean proportion of
+cases in which a target type was the majority type, corresponding to a mean proportion of
 $.003$ across populations. By contrast, in $913$ cases the majority types had $\Lbound$ with
@@
 \centering
-%\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/alt-fig2-onlym-pr}
-\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/barh-onlym}
+\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig4}
 
@@
   pressure for learnability ($\lambda = 20, k = 5$). The learning prior is shown in the
-  right-most plot with top-most groupings corresponding to types, literal and pragmatic, with
+  right-most plot, with top-most groupings corresponding to types, literal and pragmatic, with
   lexica of kinds $\Llack$, $\Lbound$, and $\Lall$.}
@@
 
-\mf{Possibly change plot on the left in Figure~\ref{fig:only-M} and adapt text
-  accordingly. Write that we know that happens under $l = 5$ in the long run (even if we do not
-  see it after 50 steps yet, if that is indeed so). If it is so, we could leave the plot and
-  say: long term development is nice, but when fitness-based selection interferes at every step
-  we do not care so much for infinite time, but rather for short time (actually one-step). If
-  it is not so, insert a plot for $l>1$ that shows something interesting.}
-
-The effect of iterated learning without a pressure for communicative success using either
-posterior sampling ($l = 1$) or a stronger tendency towards posterior maximization ($l = 15$)
-is shown in Fig.~\ref{fig:only-M} together with the prior over types. The prior shows that
-while users of $\Llack$ are not the most favored by the inductive bias (compared, e.g., to
+Recall that iterated learning without pressure for communicative success and posterior sampling ($l = 1$) converges to a population that mirrors the prior \citep{griffiths+kalish:2007}, shown in the left-most pane of Fig.~\ref{fig:only-M}.
+Its effects when learners exhibit a stronger tendency towards posterior maximization are illustrated in the other two plots of Fig.~\ref{fig:only-M}.
+The prior shows that while users of $\Llack$ are not the most favored by the inductive bias (compared, e.g., to
 $\Lall$) they are nevertheless more advantaged than others, such as $\Lbound$, in virtue of the
-relatively simple semantics they conventionalize. Crucially, $\Llack$ enables its users to
+relatively simple semantics they conventionalize. In contrast to, e.g., $\Lall$, however, $\Llack$ enables its users to
 convey each state with a single message when combined with pragmatic reasoning and sufficiently
 high $\lambda$. This makes it less likely to be confused with other types if the learning data
-is not too sparse ($k \geq 5$). What is more, learners have a propensity to infer pragmatic
-$\Llack$ when the teacher's type produces very similar data, such as when using $\Lbound$ (see
-Section~\ref{subsec:example}). Moreover, $\Llack$ is less likely to be confused with types with
-different observable behavior because its pragmatic use approximates a one-to-one form-meaning
-mapping. As a consequence, a stronger propensity to maximize the posterior increases their
-proportion in the population.
-
-However, in contrast to a pressure only for communicative success with high $\lambda$ (see
+is not too sparse ($k \geq 5$) and $\lambda$ not too low ($\lambda \geq 5$).
+What is more, learners have a propensity to infer pragmatic $\Llack$ when the teacher's type produces very similar data, such as when using $\Lbound$ (see Section~\ref{subsec:example}).
+As a consequence, a stronger propensity to maximize the posterior increases their
+proportion in the population. In $1000$
+independent populations with $\lambda = 20$ and $l = 5$ all majority types were target types,
+with each reaching approximately the same proportion of users in the population.
+
+Crucially, in contrast to a pressure only for communicative success with high $\lambda$ (see
 Fig.~\ref{fig:only-R}), learnability alone does not succeed in selecting for a single
@@
 is passed on to the next generation with the same faithfulness and, differently from a pressure
-for communicative success, they do not stand in competition with each other. In $1000$
-independent populations with $\lambda = 20$ all majority types were target types,
-with each reaching approximately the same proportion of users in the population.
+for communicative success, they do not stand in competition with each other.
 
@@
 for learnability alone may consequently lead to a spread of communicatively suboptimal types
-that lexicalize simpler semantics. In the extreme, when $l = 1$ and $\lambda = 1$ all of $1000$
-independent populations had users of $\Lall$ as majority types. For higher $\lambda$, it
+that lexicalize simpler semantics. For higher $\lambda$ and at least a slight tendency to maximize the posterior, it
 becomes clear that a high prior is not the only thing that counts when it comes to learnability
-(see Section~\ref{subsec:example}). As soon as there is information for learners to discern
+(see Section~\ref{subsec:example} and Appendix \ref{sec:example-iter-bayes}). As soon as there is information for learners to discern
 whether one type is more likely to have generated the data (depending on $\lambda$ and $k$), it
@@
 \centering
-\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/alt-fig3-r+m}
+\includegraphics[width=1\textwidth,height=8cm,keepaspectratio]{./plots/fig5}
 \caption{Proportion of target types, the six competitor types with the highest proportion, the highest type of the $\Lall$-kind, and the population's majority type in representative populations after $50$ generations under both pressures ($k = 5$).}
@@
 that are a priori favored, such as $\Lall$. An increase in $\lambda$ leads to the selection of
-target types but does not necessarily lead to monomorphic populations if learners sample from
-the posterior (see the uppermost row in Fig.~\ref{fig:rmd}). Finally, a combination of high
-$\lambda$ and $l$ leads to increasing proportions of a single majority target type. This joint
+target types but does not necessarily lead to monomorphic populations if learners' tendency to maximize the posterior is not very strong (see the uppermost row in Fig.~\ref{fig:rmd}) or absent.
+Finally, a combination of high $\lambda$ and $l$ leads to increasing proportions of a single majority target type.
+This joint
 influence is summarized in Fig. \ref{fig:diff}, which shows the mean difference between the
-most frequent target type and the proportion of the most frequent non-target type of a
-different kind in $1000$ independent populations across $\lambda$ and $l$ values. Higher values
+most frequent target type and the proportion of the most frequent non-target type in $1000$ independent populations across $\lambda$ and $l$ values. Higher values
 of $\lambda$ and $l$ increase the prevalence of a single target type, whereas lower values lead
 to less pronounced differences, with a valley resulting from low $\lambda$ and high $l$
-(cf. Fig. \ref{fig:rmd} with $\lambda = 1$ and $l=15$).
+(cf. distance to the majority type in Fig. \ref{fig:rmd} for $\lambda = 1$).
 
@@
 \centering
-\includegraphics[width=1\textwidth,height=10cm,keepaspectratio]{./plots/fig4-3d-incumbents-difference} %alternatively without '-3d'
+\includegraphics[width=1\textwidth,height=10cm,keepaspectratio]{./plots/fig6} %alternatively without '-3d'
 \caption{Mean difference between proportion of highest target type and highest other type in
@@
 \ref{fig:rmd} and their lack of competition in Fig. \ref{fig:only-M}). That is, we contend that
-adopting an expressive type that generates learnable data does not, in itself, capture a type's
+the adoption and retention rate of an expressive type that generates learnable data does not, in itself, capture a type's
 arguably central communicative function of transferring information to peers. Taking
@@
 language is learned. Lastly, chains of iterated learning alone do not put types in direct
-competition. Accordingly, learning leads to polymorphous populations in which multiple types of
+competition. Accordingly, learning alone leads to polymorphous populations in which multiple types of
 a kind coexist \citep{nowak:2006}.
@@
 
-Another important issue that is not addressed in the model are potential costs associated with
+Another important issue that is not addressed in the present model are potential costs associated with
 pragmatic reasoning. Here, we simply assumed that literal and pragmatic reasoning strategies
@@
 \citep[cf.][]{ThompsonKirby2016:Culture-Shapes-}, as well for the evolution of elements that
-define the agents' cognitive make-up; $\lambda$ and $l$.
+define the agents' cognitive make-up: $\lambda$ and $l$.
 
@@
 are all equal for all three types. Consequently, under iterated Bayesian learning with $l=1$
-the population will converge to a uniform distribution over all three types in the limit, but
+the population will eventually converge to a uniform distribution over all three types in the limit, but
 for $l=10$ it converges to a skewed population vector $\vec{x} \approx \tuple{0.5, 0.25, 0.25}$

### 2018-03-30 paper/rv-letter1.tex
final manuscript+letter changes TB

@@
 %
-\mf{We need to possibly rewrite this.}
+As now made much more explicit in the discussion surrounding iterated learning and the {\em learnability only} condition, a population that approximates the prior (with, e.g., $L_{\text{all}}$ as the highest) is only guaranteed for $l = 1$ (in line with the analytic results from \citealt{griffiths+kalish:2007}).
+For $l > 1$ matters are different.
+The reason why we do not see $L_{\text{all}}$ come out as the majority type are asymmetries in types' production behavior (see appendix C and \S3.3.1).
 Due to their stochastic speaker behavior, the data produced by $L_{\text{all}}$-teachers tends to be compatible with the behavior of many other types.
 Intuitively, the data such teachers produce are all over the place and do not do a good job in setting them apart from other types.
+In fact, the data may be ``misleading'' and suggest a different type.
 Consequently, even if the prior favors $L_{\text{all}}$, this type is not transmitted very faithfully; learners reason that the input they get from $L_{\text{all}}$-teachers could also come from other types and may adopt those instead.
-If there are types that are transmitted more faithfully, the population will, over time, transition to these types rather than to $L_{\text{all}}$.
+If there are types that are transmitted more faithfully under the same parameter constellation, the populationmay, over time, transition to these types rather than to $L_{\text{all}}$.
 As we put it in Section~4, iterated learning does not necessarily promote the {\em a priori} more likely type, but tends to promote a type $t$ based on a gradient of how many other types might likely mutate into $t$, so to speak.
-We stress this in Section~3.2.2 and Section~4.
+We now stress this more in Section~3.2.2, Section~4, and appendix C.
 

### 2018-03-30 code-LOT-extension/plots.py

@@ -961,9 +961,9 @@ list2 = [5,15]
 #get_subfigs_hist_mutation(types,list1,list2)
 
 #Plot 3:
-list1 = [1,2,20]
+list1 = [1,10,20]
 list2 = [5,10,15]
-#get_subfigs_hist_rmd(types,list1,list2)
+get_subfigs_hist_rmd(types,list1,list2)
 
 
 ##Plot 4 in 3D:
